### 2022-10-17 comp_or.tex
Infeasible solutions

% !TeX spellcheck = en_GB
\documentclass[preprint,authoryear]{elsarticle}

\usepackage[left]{lineno}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{multirow}

\usepackage{lineno,hyperref}
\modulolinenumbers[5]

%\usepackage{hyphenat}
\usepackage[none]{hyphenat}

\usepackage{url}
\usepackage{booktabs}

\usepackage{tikz}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsthm}

\usepackage{caption}
\captionsetup{labelfont=bf}
\captionsetup{skip=0pt} % vertical distance from the table or figure

\usepackage{pgfplots}
\pgfplotsset{compat=newest}
\usepackage{algorithm}

\usepackage[noend]{algpseudocode}

\usepackage[margin=0.9in]{geometry}

\usepackage[normalem]{ulem}
\usepackage{xcolor}

\newcommand{\Comentar}[1]{\State {\cmmt{#1}}}
\newcommand{\Break}{\State \textbf{break}}
\renewcommand{\Return}{\State \textbf{return}~}
\renewcommand{\algorithmicensure}{\textbf{Parameters:}}

\renewcommand\algorithmicthen{}
\renewcommand\algorithmicdo{}

\algblock{ForEach}{EndFor}
\algblock{ForDownTo}{EndFor}
\algblock{ForTo}{EndFor}

\newcommand{\Block}[1]{\State #1 \{}
\newcommand{\EndBlock}{\State \}}

%----- begin celio
\usepackage{scrextend}
\newcommand{\boldm}[1] {\mathversion{bold}#1\mathversion{normal}}
\newcommand{\round}[1]{\ensuremath{\lfloor#1\rceil}}
\usepackage{setspace}
\usepackage{array}
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}

\newcommand{\specialcell}[2][c]{%
	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

%----- end celio

\journal{European Journal of Operational Research}

\begin{document}

\begin{frontmatter}

\title{Air cargo load and route planning in pickup and delivery operations}

\author{A.C.P.~Mesquita}
\ead{celio@ita.br}

\author{C.A.A.~Sanches\corref{cor1}}
\ead{alonso@ita.br}
\cortext[cor1]{Corresponding author.}

\address {Instituto Tecnol\'{o}gico de Aeron\'{a}utica - DCTA/ITA/IEC\\
Pra\c{c}a Mal. Eduardo Gomes, 50\\
S\~{a}o Jos\'{e} dos Campos - SP - 12.228-900 - Brazil}

\begin{abstract}

In the scenario of aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading for rapid take-off and mission accomplishment, especially in times of crisis, public calamity, client contract short lead times, or any external pressure for immediate take-off. Also, there is no technological assistance to help the load and trip planners with the huge number of demands for transport in each hub. This enables other risks such as improper delivery, excessive fuel burn, and longer than necessary turn-around time. We contributed by modelling and solving a new problem of planning the loading and routing of an aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs. This new hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modelled using standardised pallets in fixed positions, obeying the centre of gravity constraints, delivering each item to its destination, and minimising fuel consumption costs. We also contributed by carrying out multiple experiments with a commercial solver and four well-known meta-heuristics on synthetic data based on real data from the {\it Brazilian Air Force}\/ transportation history. These challenging benchmark instances are made publicly available. We also created a heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it was the unique method that managed to solve all testing scenarios.

% Reescrever

\end{abstract}

\begin{keyword}
Air Cargo \sep Load Planning \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Vehicle Routing
\end{keyword}

\end{frontmatter}

\label{sec1}
\section{Introduction}

Air cargo transport involves several sub-problems that are difficult to solve. Recently, \cite[p. 401]{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP) and {\it Weight and Balance Problem} (WBP). Several aspects were considered in this modelling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc. They also presented a comprehensive bibliographic survey of solving methods that had been developed in different situations.

However, there are still other important challenges in air cargo transport that go beyond the definition of the ACLPP, especially with regard to the flight itinerary and the loading and unloading at each destination (or node) of this travel plan. In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Pickup and Delivery Problem} (PDP), and the search for the lowest cost route, which is the well-known {\it Travelling Salesman Problem} (TSP).

Considering air cargo transport, Table \ref{tab:sa} lists the main works in the literature and the corresponding sub-problems addressed. We also indicate whether the dimensions of the items were taken into account ({\bf 3D} or {\bf 2D}) and which solution method was used: heuristics ({\bf Heu}), integers ({\bf Int}), or linear programming ({\bf Lin}).

\begin{table}[H]
	\centering
	\caption{Air cargo transport: literature, problems and features}  \label{tab:sa}
	\scriptsize
	\renewcommand{\arraystretch}{1.1} % alturas das linhas
	\begin{tabular}{r|cc|cc|cc|ccc}
		\toprule
		                            & {\bf APP}  & {\bf WBP}  &  {\bf PDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} \\
		\midrule
		\cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$ \\
		\cite{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$ \\
		\cite{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ \\
		\cite{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$ \\
		\cite{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		\cite{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
		\cite{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$ \\
		\cite{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$ \\
		\cite{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		\cite{MesquitaCunha2011}   & $.$ & $.$    & $\bigstar$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\		
		\cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		\cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
		\cite{Vancroonemburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		\cite{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		\cite{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
		\cite{PaquaySchynsLimbourg2016,PaquayLimbourgSchynsOliveira2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$ \\
		\cite{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$ \\
		\cite{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$  \\
		\cite{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$  \\
		\cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
		{\bf This work}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$  \\
		\bottomrule 
	\end{tabular}
	\normalsize 
\end{table}

As can be seen, so far \cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem. Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a pre-defined flight plan restricted to two legs. It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modelled through a container sequencing problem. Referring directly to this work, \cite[p. 409]{BrandtStefan2019} comment: {\it However, not even these sub-problems are acceptably solved for real-world problem sizes or the models omit some practically relevant constraints}.

There are real situations that are much more complex. In this work, we consider a practical case in Brazil, which is the largest economy in Latin America. Due to its dimensions, this country has the largest air market on the continent with $2,499$\/ registered airports, of which $1,911$\/ are private and $588$\/ are public. Although it is an immense distribution network, airlift missions consider 3 to 5 nodes per flight plan. Throughout this work, we address routes with up to 7 nodes, as can be seen in Table \ref{tab:costs} and Figure \ref{fig:nodes}.

\begin{table}[H]

	\begin{minipage}{0.05\linewidth}

	\end{minipage}\hfill % these two lines must be close to each other
	\begin{minipage}{0.45\linewidth}

		\caption{Brazilian airports distances ($km$)}  \label{tab:costs}
		\centering

		\footnotesize

		\newcolumntype{Y}{>{\centering\arraybackslash}p{0.09\textwidth}}
		\newcolumntype{X}{>{\centering}p{0.09\textwidth}}

		\begin{tabular}{X X X X X X X Y}
			\toprule
			Node & $l_0$ & $l_1$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ & $l_6$ \\
			IATA*   & GRU   & GIG   & SSA   & CNF   & CWB   & BSB   & REC \\	
			\midrule	
			GRU     & 0	    &343	&1,439   &504    &358    &866    &2,114\\
			GIG	    & 343	&0	    &1,218   &371    &677    &935    &1,876\\
			SSA	    & 1,439	&1,218	&0	    &938    &1,788   &1,062   &676\\
			CNF	    & 504	&371	&938	&0	    &851    &606    &1,613\\
			CWB	    & 358	&677	&1,788	&851	&0	    &1,084   &2,462\\
			BSB	    & 866	&935	&1,062	&606	&1,084	&0	    &1,658\\
			REC	    & 2,114	&1,876	&676	&1,613	&2,462	&1,658	&0\\
			\bottomrule
			\multicolumn{8}{c}{*International Air Transport Association}\\
			\multicolumn{8}{c}{\small\textsuperscript{Source: www.airportdistancecalculator.com}}\\
		\end{tabular}
		\normalsize

	\end{minipage}\hfill % these two lines must be close to each other
	\begin{minipage}{0.50\linewidth}
		\centering
		\includegraphics[scale=0.25]{Images/nodes.png}
		\captionof{figure}{A route between Brazilian airports}
		\label{fig:nodes}		
	\end{minipage}
\end{table}

In the {\it Brazilian Air Force}\/ missions, hundreds of items can be carried at each node, where the objectives are to prioritise the transport of the most important items and minimise the cost of fuel along the route. As standardised pallets are used with predefined positions on the aircraft, it is possible to carry out loading and unloading at each node in around two hours. However, there is no technological assistance that guarantees the achievement of these objectives.

This work proposes a method that attends these objectives: a pallet building and arrangement plan with a routing option that maximises the benefit-cost ratio for the smooth execution of pickup and delivery transport missions. We develop a heuristic that can be executed on a simple handheld computer (like a laptop or a tablet) and that provides a solution quickly enough to keep this cargo handling time under an hour. This heuristic will reduce the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets, and picking up and delivering at each node. To the best of our knowledge, this is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP and TSP has been addressed. This new problem is named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem} (ACLP+RPDP).

This article is organised into six more sections. In Section \ref{sec2}, we give a brief review of the literature. In Section \ref{sec3}, we present the problem context and assumptions and, in Section \ref{sec4}, the mathematical model and how we dealt with its issues. In Section \ref{sec5}, we describe the elaborate algorithms, whose results are presented in Section \ref{sec6}. Finally, our conclusions are in Section \ref{sec7}.

\section{Related literature}
\label{sec2}

In this section, we briefly describe the characteristics of the main works related to air cargo transport, following the chronological order of Table \ref{tab:sa}.

\cite{LarsenMikkelsen1979} developed an interactive procedure for loading 14 types of Boeing 747 into a two-leg flight plan. Seven types of items were considered to be allocated in 17 to 42 positions. With non-linear programming and heuristics, they present a solution that minimises positioning changes in the intermediate node, optimising the load balancing in the aircraft.

\cite{Brosh1981} addressed the problem of planning the allocation of cargo on an aircraft. Considering volume, weight and structural constraints, the author finds the optimal load layout through a fractional programming problem.

\cite{Kevin1992} developed a multi-criteria optimization approach to load the {\it C-130} aircraft of the {\it Canadian Air Force}. Based on integer programming, this model provides timely planning and improves airlift support for combat operations, solving WBP with pallets in fixed positions, and considering 20 different items.

\cite{Heidelberg1998} developed a heuristic for 2D packing in air loading, comparing it with methods for solving the {\it Bin Packing Problem}. Authors conclude that the classical algorithms are inadequate in this context, because they ignore the aircraft balancing constraints.

%\cite{Gueret2003} proposed a method for loading military aircraft (specifically, the french military) for airlift operations. They used a two-phased solution method consisting of heuristics that quickly compute suboptimal initial solutions for a 2-dimensional bin packing and a local search algorithm to improve upon the initial solution, while ignoring the aircraft center of gravity.

\cite{MongeauBes2003} presented a method based on linear integer programming to solve the problem of choosing and positioning containers on the {\it Airbus 340-300}. Safety and stability restrictions were considered, with the objective of minimising fuel consumption.

\cite{fok2004optimizing} developed a web-based application to make efficient use of space and load balancing for an air cargo company. Based on an analysis of historical data, an operational load planning with mathematical optimization is obtained. This container load planning is usually done roughly 2 hours before departure, when all cargo details are in place.

\cite{Chan2006} carried out a case study with heterogeneous pallets. In order to minimise the total cost of shipping, they developed a 3D packing heuristic, with a loading plan for each pallet. Although the authors do not consider load balancing or positioning of pallets in the cargo hold, this method is relevant in commercial and industrial applications, where cargo items tend to be less dense.

\cite{KaluznyBohdanL2009Oalb} developed a mixed integer linear programming model to arrange a set of items in a military context that optimises the load balance.

\cite{Verstichel2011} solved WBP by selecting the most profitable subset of containers to be loaded onto an aircraft using mixed-integer programming. Experimental results on real-life data showed significant improvements compared to those obtained manually by an experienced planner.

\cite{MesquitaCunha2011} presented a heuristic for a real problem of the {\it Brazilian Air Force}, which consists of defining transport routes with simultaneous collection and delivery from a central distribution terminal.

\cite{Limbourg2012} developed a mixed-integer program for optimally rearranging a set of pallets into a compartmentalised cargo aircraft, specifically the {\it Boeing 747}.

\cite{RoesenerHall2014} solved APP and WBP as an integer programming problem, which also allows items to be loaded into pallets according to a specific orientation (e.g., this side up).

\cite{Vancroonemburg2014} presented a mixed integer linear programming model that selects the most profitable pallets, satisfying safety and load balancing constraints on the {\it Boeing 747-400}. Using a solver, authors solved real problems in less than an hour.

As already mentioned, \cite{LurkinSchyns2015} was the first work that simultaneously modelled WBP and PDP in air cargo transport. The authors demonstrated that this problem is NP-hard and performed some experiments with real data, noting that their model offers better results than those obtained manually.

\cite{RoesenerBarnes2016} proposed a heuristic to solve the {\it Dynamic Airlift Loading Problem} (DALP). Given a set of palletized cargo items that require transport between two nodes in a time frame, the objective of this problem is to select an efficient subset of aircraft, partition the pallets into aircraft loads and assign them to allowable positions on those aircraft.

\cite{PaquaySchynsLimbourg2016} presented a mathematical modelling to optimise the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format. Its objective is to maximise the volume used in containers, considering load balancing restrictions, the presence of fragile items and the possibility of rotating these boxes. \cite{PaquayLimbourgSchynsOliveira2018} developed some heuristics to solve this problem.

\cite{YangLiuGao2018} modelled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing in order to minimise fuel consumption.

\cite{wong2020}A developed a mathematical model and a tool based on mixed integer programming for optimising cargo in aircraft with different pallet configurations. Balance restrictions and the presence of dangerous items were considered. \cite{eugene2021} integrated this tool to a digital simulation model, with a visualisation and validation system, based on sensors that alert about load deviations.

\cite{zhao2021} proposed a new modelling for WBP based on mixed integer programming. Instead of focusing on the centre of gravity (CG) deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.

As can be seen, none of these works address air cargo palletization and load balancing with route optimization in a multi-leg flight plan. This is the objective of our work: to model and elaborate heuristics for a real problem that simultaneously involves 4 intractable sub-problems: APP, WBP, PDP and TSP.

\section{Problem context and assumptions}
\label{sec3}

In this section, we describe the context of the problem addressed in this work, as well as the assumptions considered.

\subsection{Operational premises}

As we are dealing with an extremely complex and diverse problem, we decided to establish some simplifying characteristics:

\begin{itemize}

	\item At each node of the flight plan, the items to be allocated are characterised by weight, volume, scores, and previously known destinations, but do not have dimensions. We leave the consideration of 2D or 3D items for future work.

	\item We also disregarded {\it hazardous} items, which eventually could be treated as high score items and other specific constraints.

	\item We considered a unique pallet type: the {\it 463L Master Pallet}, a common size platform for bundling and moving air cargo. It is the primary air cargo pallet for more than 70 Air Forces and many air transport companies. This pallet has a capacity of $4500 kg$ and $13.7 m^3$, is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$. For more information, see {\tt www.463LPallet.com}.

	\item All items allocated on a pallet must have the same destination. A pallet which has not yet reached its destination may receive more items, although it is known that these operations of removing restraining nets increase handling time and the risk of improper delivery. We do not consider oversized cargo in this work, but only cargoes that fit on these pallets.

	\item Finally, as we are interested in minimising fuel costs, we disregarded some lower ones as {\it handling operation costs}.

\end{itemize}

Throughout this text, we call a {\it consolidated item} a set of items of the same destination stacked on a pallet and covered with a restraining net. It is considered unique, having the same attributes of its components, whose values are the sum of individual scores, weights and volumes. See Figure \ref{fig:larger2}. To ensure accuracy in pickup and delivery operations, consolidated items must remain on board until their destination.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.4]{Images/large.png}
	\caption{Consolidated items on pallets inside a {\it Boeing C-17}}
	\small\textsuperscript{Source: https://www.sacprogram.org/en/Pages/Boeing-C-17-Globemaster-III.aspx}
	\label{fig:larger2}
\end{figure}

\subsection{Aircraft and load balancing}

We consider real scenarios with a smaller or a larger aircraft with payloads of $26,000 kg$ or $75,000 kg$ respectively. Both layouts are represented in Figures \ref{fig:smaller} and \ref{fig:larger}, where the pallets are identified by $p_i$.

\begin{figure}[!h]
	\centering

	\begin{tikzpicture}[scale=1.4, samples=100]

		\filldraw[fill=blue!5!white, draw=black] (0, 0) rectangle (5.15, 1.4);
		\filldraw[fill=gray!20!white, draw=black] (0.2, 0.2) rectangle node{$p_7$} (1.0, 1.2); 
		\filldraw[fill=gray!20!white, draw=black] (1.2, 0.2) rectangle node{$p_6$} (2.0, 1.2); 
		\filldraw[fill=gray!20!white, draw=black] (2.2, 0.2) rectangle node{$p_5$} (3.0, 1.2); 
		\filldraw[fill=gray!20!white, draw=black] (3.2, 0.2) rectangle node{$p_4$} (4.0, 1.2); 
		\filldraw[fill=gray!20!white, draw=black] (4.2, 0.2) rectangle node{$p_3$} (5.0, 1.2); 
		\filldraw[fill=blue!5!white,  draw=black] (5.15,  0) rectangle             (7.25, 1.4);
		\filldraw[fill=gray!20!white, draw=black] (5.3, 0.2) rectangle node{$p_2$} (6.1, 1.2); 
		\filldraw[fill=gray!20!white, draw=black] (6.3, 0.2) rectangle node{$p_1$} (7.1, 1.2);

	\end{tikzpicture}

	\begin{tikzpicture}[scale=1.4, samples=100]
		\draw[black, thick][<-]  (1, 2.17) node[anchor=east, font=\fontsize{7}{3.5}\selectfont]{Forward} -- (5.5, 2.17) node[anchor=west, font=\fontsize{7}{3.5}\selectfont]{Ramp} ;
	\end{tikzpicture}

	\caption{Smaller aircraft layout}
	%\small\textsuperscript{Source: the author}	
	\label{fig:smaller}
\end{figure}

\begin{figure}[!h]
	\centering

	\begin{tikzpicture}[scale=1.2, samples=100]

		\filldraw[fill=blue!5!white, draw=black] (0, 0) rectangle (10.23, 2.6);

		\filldraw[fill=gray!20!white, draw=black] (0.20, 1.40) rectangle node{$p_{17}$} (1.43, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (0.20, 0.20) rectangle node{$p_{18}$} (1.43, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (1.63, 1.40) rectangle node{$p_{15}$} (2.86, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (1.63, 0.20) rectangle node{$p_{16}$} (2.86, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (3.06, 1.40) rectangle node{$p_{13}$} (4.29, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (3.06, 0.20) rectangle node{$p_{14}$} (4.29, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (4.49, 1.40) rectangle node{$p_{11}$} (5.72, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (4.49, 0.20) rectangle node{$p_{12}$} (5.72, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (5.92, 1.40) rectangle node{$p_{9}$} (7.15, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (5.92, 0.20) rectangle node{$p_{10}$} (7.15, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (7.35, 1.40) rectangle node{$p_{7}$} (8.58, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (7.35, 0.20) rectangle node{$p_{8}$} (8.58, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (8.78, 1.40) rectangle node{$p_{5}$} (10.1, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (8.78, 0.20) rectangle node{$p_{6}$} (10.1, 1.20);

		\filldraw[fill=blue!5!white, draw=black] (10.23, 0) rectangle (13.27, 2.6);
		\filldraw[fill=gray!20!white, draw=black] (10.41, 1.40) rectangle node{$p_{3}$} (11.64, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (10.41, 0.20) rectangle node{$p_{4}$} (11.64, 1.20);
		\filldraw[fill=gray!20!white, draw=black] (11.84, 1.40) rectangle node{$p_{1}$} (13.07, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (11.84, 0.20) rectangle node{$p_{2}$} (13.07, 1.20);

	\end{tikzpicture}

	\begin{tikzpicture}[scale=1.2, samples=100]
		\draw[black, thick][<-]  (1, 2.77) node[anchor=east, font=\fontsize{7}{3.5}\selectfont]{Forward} -- (9, 2.77) node[anchor=west, font=\fontsize{7}{3.5}\selectfont]{Ramp} ;
	\end{tikzpicture}

	\caption{Larger aircraft layout} \label{fig:larger}
\end{figure}

In both cases, the torque applied to the aircraft must keep its CG in the operational range, which corresponds to a percentage of the {\it Mean Aerodynamic Chord} \footnote{Chord is the distance between the leading and trailing edges of the wing, measured parallel to the normal airflow over the wing. The average length of the chord is known as the {\it Mean Aerodynamic Chord} (MAC).}: $0.556m$ in the smaller aircraft and $1.17m$ in the larger one. See Figure \ref{fig:lateral}.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.22]{Images/lateral.png}
	\caption{Aircraft longitudinal cut showing red lines as pallets}
	%\small\textsuperscript{Source: the author}
	\label{fig:lateral}
\end{figure}

Tables \ref{tab:smaller} and \ref{tab:larger} show the parameters in both cases. {\it CGx} and {\it CGy} refer to the relative distances of pallet centroids (in metres) in relation to the CG of aircraft along both axes. In both aircraft, as the ramps have an inclination of 25 degrees, we made the necessary corrections in {\it CGx}, {\it Weight} and {\it Volume limits} of the corresponding pallets. The monetary costs of both aircraft are also indicated: per unit of distance in flights between legs ($c_d$) and per deviation in the CG ($c_g$). It is important to consider that $c_g$ tends to zero as the aircraft attitude tends to level.

\begin{table}[H]
	\centering
	\caption{Smaller aircraft parameters}  \label{tab:smaller}
	\footnotesize
	\begin{tabular}{c | c c c c c c c}
		\toprule
		\textbf{Limits}  & \multicolumn{3}{c}{$Payload$: $26,000kg$} & \multicolumn{4}{c}{$limit^{CG}_{long}$: $0.556m$} \\
		\midrule
		\bf{Pallets} & $p_7$ & $p_6$ & $p_5$ & $p_4$ & $p_3$ & $p_2$ & $p_1$ \\
		\midrule
		{\bf CGx ($m$)}     & -5.10 & -2.70 & -0.30   & 2.10 & 4.50 & 6.25 & 8.39  \\
		\midrule
		{\bf Weight limits ($kg$)}  & 4,500  &  4,500 &  4,500   & 4,500 & 4,500 & 4,000 & 3,500 \\ 
		{\bf Volume limits ($m^3$)} & 13.7  &  13.7 &  13.7   & 13.7 & 13.7 & 8.9 & 6.9 \\ 
		\midrule

		\textbf{Costs}  & \multicolumn{4}{c}{$c_d$: US\$ $1.100/km$ } &	\multicolumn{3}{c}{$c_g = 0.05$} \\

		\bottomrule
	\end{tabular}
	\normalsize 
\end{table}

\begin{table}[H]
	\centering
	\caption{Larger aircraft parameters}  \label{tab:larger}
	\footnotesize
	\begin{tabular}{c | c c c c c c c c c}
		\toprule
		\textbf{Limits}& \multicolumn{3}{c}{$Payload$: $75,000kg$} & \multicolumn{3}{c}{$limit^{CG}_{long}$: $1.170m$} &
		\multicolumn{3}{c}{$limit^{CG}_{lat}$: $0.19m$} \\
		\midrule
		\multirow{2}{*}{\textbf{Pallets}}  & $p_{17}$ & $p_{15}$ & $p_{13}$ & $p_{11}$ & $p_{9}$ & $p_{7}$ & $p_{5}$ & $p_{3}$ & $p_{1}$ \\
		& $p_{18}$ & $p_{16}$ & $p_{14}$ & $p_{12}$ & $p_{10}$ & $p_{8}$ & $p_{6}$ & $p_{4}$ & $p_{2}$ \\
		\midrule 
		\multirow{2}{*}{\textbf{CGx ($m$)}} & -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\
		& -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\			
		\midrule 
		\multirow{2}{*}{\textbf{CGy ($m$)}}  & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 \\
		& -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 \\	
		\midrule
		{\bf Weight limits ($kg$)}      &   4,500   &    4,500  &   4,500   &  4,500    & 4,500     & 4,500     & 4,500     & 4,000    & 3,000   \\
		{\bf Volume limits ($m^3$)}   &   14.8   &   14.8   &  14.8    &  14.8    & 14.8     & 14.8     & 14.8     & 10.0    & 7.0 \\	
		\midrule

		\textbf{Costs}  & \multicolumn{5}{c}{ $c_d$: US\$ $4.900/km$ } &	\multicolumn{4}{c}{$c_g = 0.05$} \\

		\bottomrule
	\end{tabular}
	\normalsize 
\end{table}

We also make the following assumptions:
\begin{itemize}
	\item on each pallet, the items are distributed in such a way that their CG coincides with the centroid of the pallet;
	\item the CG of the payload must be at a maximum longitudinal distance of $limit^{CG}_{long}$ from the CG of the aircraft;
	\item in the larger aircraft, the CG of the payload must be at a maximum lateral distance of $limit^{CG}_{lat}$ from the CG of the aircraft;
	\item in the larger aircraft, pallets are distributed in two identical rows (with odd and even indices, respectively), and their centroids are at a distance $d^{CG}_{pallet}$ from the centre-line of the aircraft.
\end{itemize}

\subsection{Problem Summary}

\newcolumntype{M}{>{\raggedright}p{0.03\textwidth}}
\newcolumntype{T}{>{\raggedright\arraybackslash}p{0.9\textwidth}}

\bgroup
\def\arraystretch{1.2}
\begin{table}[H]
	\centering
	\small
	\begin{tabular}{MT}
		&Informally, ACLP+RPDP can be summarised as follows:\\
		\midrule
		max &  (items score sum) / (tour cost) of picked up and delivered items at each node on a tour  \\
		\midrule
s.t.    & Along a tour, the set of unvisited nodes is updated. \\
		& In each node, an item may be included in at most one pallet.\\		
		& In each node, consolidated items are composed of items with the same destination. \\
		& Weight, volume and score of a consolidated item are the corresponding sum of their components.\\
		& Consolidated items remain on board until their destinations.\\	
		& Consolidated items can only be included in the same pallet if their destinations are the same.\\
		& Only items destined for the remaining nodes can be loaded.  \\
		& The lateral and longitudinal torques must be within the operational range of the aircraft.\\
		& Weight and volume limitations of pallets must be respected.\\
		& The total weight must be less than the aircraft payload or the total pallet capacity, whichever is the lowest.\\	
		\midrule
	\end{tabular}
	\normalsize
\end{table}
\egroup

\section{The mathematical modelling}
\label{sec4}

Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modelling of ACLP+RPDP, which is one of the contributions of our work.

Let $L = \{ l_0, l_1, \ldots, l_K \}$ be the set of $K+1$ nodes (or destinations), where $l_0$ is the origin and end of a flight plan. Let $d(l_i,l_j)$ be the distance from $l_i$ to $l_j$, where $0 \leq i,j \leq K$. By definition, $d(l_i,l_i)=0$. Let $L_k$ be the set of remaining nodes when the aircraft is in $l_k$, $0 \leq k \leq K$. Therefore, $L_0=L$ and $L_K = \{ l_0 \}$.

Let $C=\{c_{ij}\}$ be the cost matrix of flights, where $c_{ij} = c_d*d(l_i,l_j), 0 \leq i,j \leq K$.

Let $S_K = \{s: \{1, \dots, K\} \rightarrow \{1, \dots, K\} \}$ be the set of $K!$ permutations, which correspond to all possible tours (or itineraries) that have $l_0$ as origin and end, passing through the others $K$ nodes.

Let $M = \{p_1, p_2, \ldots, p_m \}$ the set of $m$ pallets. Each pallet $p_i$, $1 \leq i \leq m$, has weight capacity $p_i.w$, volume capacity $p_i.v$, pallet destinations $p_i.to[k]$, $0 \leq k \leq K$, and distance to the CG of aircraft $p_i.d$. $p_i.to[k]$ denotes that pallet $p_i$ may assume a different destination in each node $k$.

Let $N_k = \{t^k_1, t^k_2, \ldots, t^k_{n_k} \}$ be the set of $n_k$ items to be loaded in node $l_k$, $0 \leq k \leq K$. Each item $t^k_j$, $1 \leq j \leq n_k$, has score $t^k_j.s$, weight $t^k_j.w$, volume $t^k_j.v$, and destination $t^k_j.to \in L_k$. Let $N = \bigcup_{0 \leq k \leq K} N_k$ be the set of items of all nodes along a tour.

Let $Q_k = \{a^k_1, a^k_2, \ldots, a^k_{m_k} \}$ be the set of consolidated items loaded in $m_k \leq m$ pallets when the aircraft arrives at node $l_k$, with $0 \leq k \leq K$. $a^k_i$, $1 \leq i \leq m_k$, is the group of picked-up items that were allocated on pallet $p_i$ in some of the previous nodes. $a^k_i$ has total weight $a^k_i.w$, total volume $a^k_i.v$, and destination $a^k_i.to \in L_k \cup \{l_k\}$. If $a^k_i.to = l_k$, then $a^k_i$ is unloaded, and $p_i$ will be available for reloading; otherwise, $a^k_i$ remains on the aircraft, eventually in another pallet, and with items of $N_k$ having the same destination.

Let $X_{ij}^k$ and $Y_{iq}^k$ be binary variables, where $0 \leq k \leq K$, $1 \leq j \leq n_k$, $1 \leq i \leq m$ and $1 \leq q \leq m_k$. $X_{ij}^k = 1$ if $t_j^k$ is assigned to $p_i$ in node $l_k$, and 0 otherwise. $Y_{iq}^k = 1$ if $a_q^k$ is assigned to $p_i$ in node $l_k$, and 0 otherwise. By definition, $Y_{iq}^0=0$. Allocations of items or consolidated items to pallets in node $l_k$ can be seen as a bipartite graph $G_k(V_k, E_k)$, where $V_k = M \cup N_k \cup Q_k$, $E_k = E^N_k \cup E^Q_k$, $(p_i, t_j^k) \in E^N_k$ if $X_{ij}^k = 1$, and $(p_i, a_q^k) \in E^Q_k$ if $Y_{iq}^k = 1$.

The mathematical modelling of this problem is described in the equations below.

\begin{equation} \label{eq:maxf}
	\max_{\pi \in S_K} f_\pi(\tilde{s},\tilde{c})
\end{equation}

\begin{equation} \label{eq:scores}
	\tilde{s} = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.s
\end{equation}

\begin{equation} \label{eq:costs}
	\tilde{c} = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} [ c_{\pi(k), \pi(k+1)}\times(1+c_g\times|\epsilon_k|) ] + c_{\pi(K),0}\times(1+c_g\times|\epsilon_K|)
\end{equation}

\begin{equation} \label{eq:maxW}
	maxW = \min(Payload, \sum_{i=1}^{m}p_i.w)
\end{equation}

\begin{equation} \label{eq:tau}
\tau_k = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w +  \sum_{q=1}^{m} Y_{iq}^k \times a_q^k.w)];\ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:eps}
\epsilon_k = \frac{\tau_k}{maxW \times limit^{CG}_{long}};\ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:pdp11}
	L_0 = L; L_{k} = L_{k-1} - \{l_{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:pdp13}
	Y^0_{iq} = 0; a^0_i.w = 0; a^0_i.v = 0; a^0_i.to = -1;\ i,q \in \{1, 2, \ldots, m\}
\end{equation}

\begin{equation} \label{eq:pdp12}
	X_{ij}^k = 0 \ \mbox{if} \ t_j^k.to \notin L_k; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}; \ k \in \{1,2, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:pdp9}
	Y_{iq}^k = 0 \ \mbox{if} \ a_i^k.to \notin L_k; \ i \in \{1, 2, \ldots, m\}; \ q \in \{1, 2, \ldots, m_k\}; \ k \in \{ 1, 2, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:cons2}
	a_i^{\pi(k+1)}.w = \sum_{j=1}^{n_k} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w;  \ i \in \{1, 2, \ldots, m_k\}; k \in \{0, 1, \ldots, K-1\}
\end{equation}

\begin{equation} \label{eq:cons3}
	a_i^{\pi(k+1)}.v = \sum_{j=1}^{n_k} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.v + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.v;  \ i \in \{1, 2, \ldots, m_k\}; k \in \{0, 1, \ldots, K-1\}
\end{equation}

\begin{equation} \label{eq:cons5}
	a_i^{\pi(k+1)}.to = t^{\pi(k)}_j.to  \ \mbox{if} \ X^{\pi(k)}_{ij} = 1  \ \mbox{and} \ t^{\pi(k)}_j.to \in L_{k+1};  \ i \in \{1, 2, \ldots, m_k\}; \ j \in \{1, 2, \ldots, n_k\}; \ k \in \{0,1, \ldots, K-1\}
\end{equation}

\begin{equation} \label{eq:LatIt}
	LatIt_k = \sum_{i=1}^{m} \sum_{j=1}^{n_k} ( X_{ij}^k \times t_j^k.w \times (i\%2) - X_{ij}^k \times t_j^k.w \times (i+1)\%2 ); \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:LatCons}
	LatCons_k =  \sum_{i=1}^{m} \sum_{q=1}^{m_k}  ( Y_{iq}^k \times a_q^k.w \times (i\%2) - Y_{iq}^k \times a_q^k.w \times (i+1)\%2); \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:torqlat}
	s.t.: d^{CG}_{pallet} \times | LatIt_k + LatCons_k | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:torqlong}
	s.t.: |\tau_k| \leq maxW \times limit^{CG}_{long};\ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:payload}
	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w ) \leq maxW; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:app2}
	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m_k\}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:app3}
	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.v + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m_k\}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:app4}
	s.t.: \sum_{i=1}^{m} X_{ij}^k \leq 1; \ j \in \{1, 2, \ldots, n_k\}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:app5}
	s.t.:  Y_{iq}^k = 1 \ \mbox{if} \ a^k_q.to \in L_k; \ q \in \{1, 2, \ldots, m_k\}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:pdp8}
	s.t.: p_i.to[k] = t^k_j.to\ \mbox{if} \ X_{ij}^k = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}; \ k \in \{1, 2, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:pdp2}
	s.t.:  p_i.to[k] = a^k_q.to\ \mbox{if} \ Y_{iq}^k = 1; \ i \in \{1, 2, \ldots, m\};\ q \in \{1, 2, \ldots, m_k\}; \ k \in \{1, 2, \ldots, K\}
\end{equation}

The objective of this problem is to find a permutation $\pi \in S_K$\/ that maximises the function $f_\pi(\tilde{s},\tilde{c})$ \ref{eq:maxf}. In this way, the flight plan will be $l_0, l_{\pi(1)}, \ldots, l_{\pi(K)}, l_0$. $\tilde{s}$\/ is the total score of transported items \ref{eq:scores} and  $\tilde{c}$\/ is the total cost of fuel consumed \ref{eq:costs}. As can be seen, $\tilde{c}$\/ corresponds to the fuel consumption due to the flights carried out and the CG deviation of the transported cargo. Throughout this work, for simplicity, we use $f=\tilde{s}/\tilde{c}$.

The maximum load will be the minimum between the payload and the capacity supported by the pallets \ref{eq:maxW}. Considering the maximum longitudinal distance allowed for the CG, all torques \ref{eq:tau} and deviations \ref{eq:eps} are calculated.

For each step of the flight plan, the set of unvisited nodes is updated \ref{eq:pdp11}. Although there are no items consolidated at the beginning of the flight plan, we defined these variables for ease of notation \ref{eq:pdp13}. Items destined outside the rest of the flight plan will not be loaded (\ref{eq:pdp12} and \ref{eq:pdp9}).

Consolidated items appear when there are items on the pallets that will not be unloaded on the next node. Its weights \ref{eq:cons2} and volumes \ref{eq:cons3} correspond to all the items that were on the pallet, since all these items have the same destination. On subsequent nodes, consolidated items can be allocated with other items of same destination \ref{eq:cons5}.

Equations \ref{eq:LatIt} and \ref{eq:LatCons} respectively, are applied only to the larger aircraft, and calculate the lateral torques of items and consolidated items loaded in both rows of pallets, whose constraint is described in \ref{eq:torqlat}. Similarly, \ref{eq:torqlong} is the longitudinal torque constraint, which is applied to both aircraft sizes.

The weight limitation of the aircraft must be respected \ref{eq:payload}. The sum of weights \ref{eq:app2} and volumes \ref{eq:app3} in each pallet must not exceed its capacity. Each item is associated with a pallet at most \ref{eq:app4}.

Consolidated items remain on board \ref{eq:app5} until their destinations. At each node, an item (\ref{eq:pdp8}) and a consolidated item (\ref{eq:pdp2}) must only be allocated to a pallet if the destinations are the same.

\section{Resolution strategy}
\label{sec5}

Once the assumptions of this work and the mathematical modelling of the problem are presented, it is easy to see that ACLP+RPDP is NP-hard. In a similar way to \cite[p. 6]{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in $l_0$, and no items in $l_1$. Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is NP-complete.

Real cases are more complex as they have hundreds of different items in each node and involve three intractable sub-problems: APP, WBP and PDP. Through the mathematical modelling presented in the previous section, we verify that {\it Mixed-Integer Programming}\/ (MIP) is not able to solve these cases in feasible time. Thus, it is necessary to adopt some strategy to find a viable solution, not necessarily optimal, that seeks to maximise the objective function $f$.

Our strategy is based on the fact that, in real cases, $K$\/ is usually small. Specifically, we will consider $K \leq 6$\/ throughout this work, which is a higher value than usual in {\it Brazilian Air Force} missions. As a result, if we have fast node-by-node solutions that allow us to construct a complete tour, we will be able to test all possible $K!$\/ tours and thus select the one that provides the best value for the $f$\/ function.

On the other hand, our tactic will be, at each shipping node, to predefine the destinations of the pallets at that node. In this way, we will reserve a number of pallets proportional to the volume demanded by each destination at the shipping node. We could have used another criterion, but it was observed in the experiments that volume is more constrictive in airlift. Once the destinations of the pallets are defined, we will use heuristics and a MIP solver to find the best possible node-by-node solutions. This strategy is summarized in Algorithm \ref{alg:main}.

\begin{algorithm}[H]
	\caption{$ACLP+RPDP (scenario,volume)$}  \label{alg:main}
	\begin{algorithmic}[1]
		\State Let $L, M, C$ be according to $scenario$ \label{main:LMC}
		\State $N \gets ItemsGeneration(scenario,volume)$ \label{main:items}
		\For {each $method$}
			\For {each $\pi \in S_K$} \label{main:loop1}
				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method )$ \label{main:method}
			\EndFor \label{main:loop2}
			\State $answer[scenario,volume,method] \gets \max f$ \label{main:f}
		\EndFor
		\State $ {\bf return}\ answer$
	\end{algorithmic}
\end{algorithm}

In this algorithm, there are six values for the $scenario$\/ parameter, according to Table \ref{tab:scenarios}, which defines $K$, the sets of nodes, the aircraft, the pallets and the costs from Tables \ref{tab:smaller} or \ref{tab:larger} that will be used (line \ref{main:LMC}).

The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $l_k$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n_k} t^k_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.v$). This parameter is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).

$method$\/ corresponds to a MIP solver or one of the heuristics that we will present in subsection \ref{methods}. The loop of lines \ref{main:loop1}-\ref{main:loop2} goes through all permutations $\pi$, where the node-by-node resolutions are performed by $SolveTour$, whose result is stored in $f_{\pi}$. The best result among all $K!$\/ tours will be the answer for $scenario$, $volume$\/ and $method$\/ (line \ref{main:f}).

\vspace{2.0mm}
\begin{table}[H]
	\centering
	\caption{Testing scenarios}  \label{tab:scenarios}
	\begin{tabular}{c c c c }
		\toprule
		{\bf Scenario} & {$K$} & {$L$} & {\bf Aircraft} \\		
		\midrule
		1 & 2    & \{$l_0$, $l_1$, $l_2$\}                                 & smaller \\
		2 & 2    & \{$l_0$, $l_1$, $l_2$\}                                 & larger  \\
		3 & 3    & \{$l_0$, $l_1$, $l_2$, $l_3$\}                          & larger  \\
		4 & 4    & \{$l_0$, $l_1$, $l_2$, $l_3$, $l_4$\}                   & larger  \\
		5 & 5    & \{$l_0$, $l_1$, $l_2$, $l_3$, $l_4$, $l_5$\}            & larger  \\
		6 & 6    & \{$l_0$, $l_1$, $l_2$, $l_3$, $l_4$, $l_5$, $l_6$\}     & larger  \\
		\bottomrule
	\end{tabular}
\end{table}

Next, we will present two subsections: in the first we explain how $SolveTour$ is executed, while in the second we will present the heuristics developed for node-by-node resolutions.

\subsection{SolveTour algorithm}
\label{tour}

In addition to the set of nodes, pallets, costs and items, $SolveTour$, described in Algorithm \ref{alg:tour}, receives the parameter $method$, which corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and the parameter $\pi$, which is a permutation that defines the order of visits in this tour.

As we mentioned in the previous section, all tours start and end at $l_0$\/ (lines \ref{tour:pi1}-\ref{tour:pi2}).
After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ flights (lines \ref{tour:loop1}-\ref{tour:loop2}). Initially we set pallets destination as $-1$\/ (line \ref{tour:-1}). When the aircraft is at node $l_0$, the initial graph $G_1$\/ is empty because it has no consolidated items \ref{tour:g11}. Otherwise, the set $L_k$\/ of remaining nodes is updated (line \ref{tour:lk1}), and $UpdateConsolidated$\/ (line \ref{tour:dest}) returns the set of consolidated items that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation. This allocation is stored in graph $G_1$\/ (line \ref{tour:g12}).

In the context of this work, we know that $m>K$, once the aircraft has $7$\/ or $18$\/ pallets and $K\leq 6$, allowing there to be at least one pallet for each node to be visited. $SetPalletsDestination$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node, without changing the pallets destination with consolidated items.

Finally, $SolveNode$\/ includes the edges corresponding to the items shipped at the current node, returning the graph $G_2$\/ (line \ref{tour:node}). The score and the CG deviation of this graph are calculated (line \ref{tour:analyse}) and accumulated (lines \ref{tour:score2}-\ref{tour:cost2}), allowing the final result of this tour (line \ref{tour:f}).

\begin{algorithm}[H]
	\caption{$SolveTour(\pi, L, M, C, N, method)$}  \label{alg:tour}

	\begin{algorithmic}[1]

		\State $\pi(0) \gets 0$ \label{tour:pi1}
		\State $\pi(K+1) \gets 0$ \label{tour:pi2}
		\State $score \gets 0$ \label{tour:score}
		\State $cost \gets 0$ \label{tour:cost}
		\For {$k\gets0$ to $K$} \label{tour:loop1}		
			\For{$i \gets 1$ to $m$}
				\State $p_i.to[k] \gets -1$ \label{tour:-1}
			\EndFor	
			\If {$k = 0$}
				\State $L_0 \gets L$
				\State Let $G_1(M \cup N_0, \varnothing)$ \label{tour:g11}
			\Else
				\State $L_k \gets L_k - \pi(k)$  \label{tour:lk1}			
				\State $Q_{\pi(k)}, E^Q_{\pi(k)}, M \gets UpdateConsolidated(\pi(k))$ \label{tour:dest}			
				\State Let $G_1(M \cup N_{\pi(k)} \cup Q_{\pi(k)}, E^Q_{\pi(k)})$ \label{tour:g12}
			\EndIf  \label{tour:lk2}	
			\State $M \gets SetPalletsDestination( \pi(k) )$ \label{tour:dest2}		
			\State $G_2 \gets SolveNode(method, \pi(k), G_1)$ \label{tour:node}
			\State $s, \epsilon \gets ScoreAndDeviation(\pi(k), G_2)$ \label{tour:analyse}
			\State $score \gets score + s$ \label{tour:score2}
			\State $cost \gets cost + c_{\pi(k),\pi(k+1)} * (1 + c_g * |\epsilon|)$ \label{tour:cost2} 
		\EndFor  \label{tour:loop2}
		\State $ {\bf return}\ score / cost$ \label{tour:f}

	\end{algorithmic}
\end{algorithm}

$UpdateConsolidated$, described in Algorithm \ref{alg:cons}, finds the best allocation for the consolidated items that remain on board. Initially, the set $Q$\/ is created, with the consolidated items that did not reach their destination (lines \ref{cons:remain1}-\ref{cons:remain2}). Then $MinCGDeviation$\/ (line \ref{cons:minCG}) is run through a MIP solver with the equations \ref{eq:torque}, \ref{eq:embarked} and \ref {eq:one}, to relocate the consolidated items on the pallets minimizing torque and ensuring that they all remain on board, one on each pallet. As there are few variables, the MIP solver returns an allocation $E^Q_k$\/ very quickly. Finally, the destination of each pallet with consolidated items is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).

\begin{algorithm}[H]
	\caption{$UpdateConsolidated(k)$}  \label{alg:cons}
	\begin{algorithmic}[1]

		\State Let $Q_k  = \{ a^k_1, a^k_2, \ldots, a^k_{m_k} \}$ 
		\State $Q \gets \varnothing$ \label{cons:remain1}
		\For{$q\gets1$ to $m_k$} 
			\If {$a_q^k.to \in L_k$} 
				\State $Q \gets Q \cup \{a_q^k\}$
			\EndIf
		\EndFor	 
		\State $m_k \gets |Q|$ \label{cons:remain2}
		\State $E^Q_k \gets MinCGDeviation(k, Q)$ \label{cons:minCG}
		\For{$i \gets 1$ to $m$} \label{cons:Ybegin}
			\For{$q \gets 1$ to $m_k$}
				\If{$(p_i, a_q^k) \in E^Q_k$} 
					\State $p_i.to[k] \gets a_q^k.to$
				\EndIf
			\EndFor		
		\EndFor \label{cons:Yend}
		\State ${\bf return}\ Q, E^Q_k, M$ 
	\end{algorithmic}
\end{algorithm}

\begin{equation} \label{eq:torque}
	\mbox{minimize}\ | \sum_{i=1}^{m} \sum_{q=1}^{m_k} Y^k_{iq} \times p_i.d \times a_q^k.w |,\ \mbox{where}\ Q  = \{ a^k_1, a^k_2, \ldots, a^k_{m_k} \} 
\end{equation}

\begin{equation} \label{eq:embarked}
	s.t.: \sum_{i=1}^{m} Y^k_{iq} = 1;\ q \in \{1,2,\ldots,m_k\}
\end{equation}

\begin{equation} \label{eq:one}
	s.t.: \sum_{q=1}^{m_k} Y^k_{iq} \leq 1;\ i \in \{1,2,\ldots,m\}
\end{equation}

$SetPalletsDestination$, which set pallets destination not yet defined, is described in Algorithm \ref{alg:dest}. The vectors $vol[0..K]$\/ and $PalCons[0..K]$\/ store the total volume of items and the number of pallets with consolidated items destined for each node, respectively (lines \ref{dest:vector1}-\ref{dest:vector2}). The destinations of each pallet are defined proportionally to the volume of items, regarding the pallets with consolidated items (lines \ref{dest:propor1}-\ref{dest:propor2}). The node with the maximum volume demand defines the destination of any remaining pallets (lines \ref{dest:max1}-\ref{dest:max2}).

\begin{algorithm}[H]
	\caption{$SetPalletsDestination(k)$}  \label{alg:dest}
	\begin{algorithmic}[1]

		\For{$x \gets 0$ to $K$} \label{dest:vector1}
			\State $vol[x] \gets 0$ 
			\State $PalCons[x]  \gets 0$
		\EndFor
		\State $max \gets 0$ \label{dest:max}
		\State $total \gets 0$

		\For{$j \gets 1$ to $n_k$}
			\State $d \gets t_j^k.to$
			\If {$d \in L_k$} 
				\State $vol[d] \gets vol[d] + t_j^k.v$ 
				\State $total \gets total + t_j^k.v$ 
				\If {$vol[d] > vol[max]$}
					\State $max \gets d$ 
				\EndIf
			\EndIf
		\EndFor
		\For{$q \gets 1$ to $m_k$}
			\State $d \gets a_q^k.to$
			\State $PalCons[d] \gets PalCons[d]+1$
			\If {$d \in L_k$} 
				\State $vol[d] \gets vol[d] + a_q^k.v$
				\State $total \gets total + a_q^k.v$ 
				\If {$vol[d] > vol[max]$}
					\State $max \gets d$ 
				\EndIf
			\EndIf
		\EndFor \label{dest:vector2}
		\For{$x \gets 0$ to $K$} \label{dest:propor1}
			\If {$vol[x] \neq 0$}
				\State $quant \gets \max \{1, \lfloor{ m \times vol[x]/total}\rfloor \}-PalCons[x]$ 
				\State $np \gets 0$
				\For{$i \gets 1$ to $m$}

					\If{$np = quant$}
						\State ${\bf break}$
					\EndIf
					\If{$p_i.to[k] = -1$}
						\State $p_i.to[x] \gets x$
						\State $np \gets np + 1$
					\EndIf
				\EndFor

			\EndIf 
		\EndFor \label{dest:propor2}
		\For{$i \gets 1$ to $m$} \label{dest:max1}
			\If{$p_i.to[k] = -1$}
				\State $p_i.to[x] \gets max$
			\EndIf
		\EndFor \label{dest:max2}
		\State ${\bf return}\ M$

	\end{algorithmic}
\end{algorithm}

$ScoreAndDeviation$\/ is described in Algorithm \ref{alg:eval}, which evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation. It consists of a loop that goes through all the pallets (lines \ref{eval:loop1}-\ref{eval:loop2}), accumulating the scores (lines \ref{eval:score1}-\ref{eval:score2} ) and the torques (lines \ref{eval:eps1}-\ref{eval:eps2}) of the shipped items, allowing the final calculation of the CG deviation (lines \ref{eval:torque1}-\ref{eval:torque2}).

\begin{algorithm}[H]
	\caption{$ScoreAndDeviation(k, G)$}  \label{alg:eval}

	\begin{algorithmic}[1]
		\State Let $G(V_k, E^N_k \cup E^Q_k)$
		\State Let $X_{ij}^k = 1$ if $(p_i, t_j^k) \in E^N_k$, $1 \leq i \leq m$, $1 \leq j \leq n_k$
		\State Let $Y_{iq}^k = 1$ if $(p_i, a_q^k) \in E^Q_k$, $1 \leq i \leq m$, $1 \leq q \leq m_k$
		\State $s \gets 0$
		\State $\epsilon \gets 0$
		\State $weight \gets 0$
		\For{$i \gets 1$ to $m$}	\label{eval:loop1}
			\State $weight \gets weight + p_i.w$		
			\For{$j \gets 1$ to $n_k$}
				\If{$X_{ij}^k = 1$}
					\State $s \gets s + t_{ij}^k.s$ \label{eval:score1}
					\State $\epsilon \gets \epsilon + t_{ij}^k.w * p_i.d$ \label{eval:eps1}
				\EndIf
			\EndFor				
			\For{$q \gets 1$ to $m_k$}
				\If{$Y_{iq}^k = 1$}
					\State $s \gets s + a_{iq}^k.s$   \label{eval:score2}
					\State $\epsilon \gets \epsilon + a_{iq}^k.w * p_i.d$  \label{eval:eps2}
				\EndIf
			\EndFor	   \label{eval:loop2}
		\EndFor
		\State $maxWeight \gets \min(Payload, weight)$ \label{eval:torque1}
		\State $maxTorque \gets maxWeight \times limit^{CG}_{long}$
		\State $\epsilon \gets \epsilon/maxTorque$ \label{eval:torque2}
		\State ${\bf return}\ s, \epsilon$	
	\end{algorithmic}
\end{algorithm}

\subsection{Node-by-node solutions}
\label{methods}

In this subsection we present implementations of $SolveNode(method,k, G)$, where $method$\/ corresponds to a MIP solver or a heuristic, $k$\/ is the index of the current node and $G$\/ is the allocation graph of the consolidated items that remain on board at node $l_k$.

An important issue on meta-heuristics is that they are known for the hard work in tuning parameters. For this, when we use the word {\it empirically}\/ throughout the text, we want to say that we accomplished many tests with the hardest instances to fit these values. This does not guarantee best results for all instances, but, on average, overall results are acceptable. Another important observation is that our heuristics were designed based on traditional ideas from the literature \cite{NiarFreville1997,Fidanova2006,Laabadi2018,Alonso2019,Zhan2020} and on the experiences learned during tests.

To guarantee that performance comparisons are made correctly, our implementations use the same elements described below.

\newcolumntype{C}{>{\raggedright}p{0.18\textwidth}}
\newcolumntype{D}{>{\raggedright\arraybackslash}p{0.82\textwidth}}

\bgroup
\def\arraystretch{1.2}
\begin{table}[H]
	\centering
	\small
	\begin{tabular}{CD}

		{\it Data structure:}       & Items, pallets and edges are arrays of \emph{python} class objects, which are initialized with the same data loaded from text files. \\

		{\it Common procedures:}    & All heuristics use the same procedures for selecting and inserting edges, and printing solutions.  \\

		{\it Stopping criterion:} & All heuristics have the same time limit of $0.7$ seconds per node.\\

	\end{tabular}
	\normalsize
\end{table}
\egroup

The selection of edges for $E^N_k$\/ uses the {\it edge attractiveness}\/ $\theta^k_{ij}$\/ \ref{eq:edge}, which can be understood as the tendency to allocate the item $t^k_j$\/ to the pallet $p_i$. It is directly proportional to the score, and inversely proportional to the volume and the torque of each item.

\begin{equation} \label{eq:edge}
	\theta^k_{ij}= \frac{t^k_j.s}{t^k_j.v}\times(1-\frac{t^k_j.w\times|p_i.d|}{\max_s\{t^k_s.w\}\times\max_q\{|p_q.d|\}});\ i \in \{1,2,\ldots,m\},\ j \in \{1,2,\ldots,n_k\}
\end{equation}

We use four types of random selections:
\begin{itemize}
	\item $RandomReal(r_1,r_2)$: randomly selects a real number in $[r_1,r_2]$, where $r_1$ and $r_2$ are real numbers;
	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$, where $i_1$ and $i_2$ are integer numbers;
	\item $RandomChoice(set)$: randomly selects an element from $set$;
	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
\end{itemize}

We also developed Algorithm \ref{alg:greedy}, which generates a greedy allocation of the items available in node $l_k$, according to the non-ascending order of $\theta^k_{ij}$, and considering the consolidated items already shipped. Edges are included in this allocation as long as they respect feasibility constraints. Furthermore, the volume of each pallet $p_i$\/ cannot exceed $p_i.v \times limit$, where $ 0 < limit \leq 1$\/ is a parameter.

\begin{algorithm}[H]
	\caption{ $Greedy(k, G, limit)$}  \label{alg:greedy}

	\begin{algorithmic}[1]
		\State Let $G(V_k, E^Q_k)$
		\State $E^N_k \gets \varnothing$ 			
		\For{$i \gets 1$ to $m$}
			\State $pv[i] \gets 0$ 		
		\EndFor
		\For{$q \gets 1$ to $m_k$}
			\If{$(p_i, a_q^k) \in E^Q_k$} 
				\State $pv[i] \gets pv[i] + a_q^k.v$ 
			\EndIf		
		\EndFor		
		\For {each $e_{ij}^k$ in non-ascending order of $\theta^k_{ij}$}
			\If{($E^N_k \cup \{e_{ij}^k\}$ is feasible) {\bf and} ($pv[i] \leq p_i.v \times limit$)} 
				\State $E^N_k \gets E^N_k \cup \{e_{ij}^k\}$ 
				\State $pv[i] \gets pv[i] + t_j^k.v$ 	
			\EndIf
		\EndFor
		\State ${\bf return}\ G(V_k, E^N_k \cup E^Q_k)$

	\end{algorithmic}
\end{algorithm}

To compare allocation graphs generated by the heuristics, we choice the one that offer a higher associated score, since the CG deviation is limited by the restrictions. In the pseudocodes presented below, given an allocation graph $G$\/ for the node $l_k$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation(k,G) $, described in Algorithm \ref{alg:eval}.

Below, we present six implementations for $SolveNode$: the first using a MIP solver, and five with developed heuristics: {\it Ant Colony Optimization}\/ (ACO), {\it Noising Method Optimization}\/ (NMO), {\it Tabu Search}\/ (TS), {\it Greedy Randomized Adaptive Search Procedure}\/ (GRASP) and {\it Shims}.

\subsubsection{MIP solver}
\label{solver}

In order to find a viable solution for ACLP+RPDP, the strategy adopted in $SolveTour$\/ previously defines the values of some variables. Concretely, all permutations between the nodes are tested, the set of nodes to be visited is updated, the consolidated items that remain on board are reallocated to minimize the CG deviation, and the pallets destination is determined according to the volume of items available for shipment. In this way, the mathematical modelling for $SolveNode(MIP,k,G)$\/ becomes simpler, which finds an allocation of available items in node $l_k$\/ using previously defined values of $L_k$, $p_i.to[k]$, $Y^k_{iq}$\/ and $a^k_q$.

This simplified modelling is described in the equations \ref{eq:newf} to \ref{eq:lastnew}, which maximizes the score of the $t^k_j$\/ items to be shipped, maintaining load balancing. The variables $X^k_{ij}$\/ define the set of edges $E^N_k$, which will be added to the graph $G$.

\begin{equation} \label{eq:newf}
	\max f_s = \sum_{i=1}^{m} \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.s
\end{equation}

\begin{equation} 
	maxW = \min(Payload, \sum_{i=1}^{m}p_i.w)
\end{equation}

\begin{equation} 
	\epsilon_k = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w +  \sum_{q=1}^{m} Y_{iq}^k \times a_q^k.w)]/(maxW \times limit^{CG}_{long})
\end{equation}

\begin{equation}
	X_{ij}^k = 0 \ \mbox{if} \ t_j^k.to \notin L_k; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}
\end{equation}

\begin{equation} 
	LatIt_k = \sum_{i=1}^{m} \sum_{j=1}^{n_k} ( X_{ij}^k \times t_j^k.w \times (i\%2) - X_{ij}^k \times t_j^k.w \times (i+1)\%2 )
\end{equation}

\begin{equation} 
	LatCons_k =  \sum_{i=1}^{m} \sum_{q=1}^{m_k}  ( Y_{iq}^k \times a_q^k.w \times (i\%2) - Y_{iq}^k \times a_q^k.w \times (i+1)\%2)
\end{equation}

\begin{equation} 
	s.t.: d^{CG}_{pallet} \times | LatIt_k + LatCons_k | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}
\end{equation}

\begin{equation}
	s.t.: |\tau_k| \leq maxW \times limit^{CG}_{long}
\end{equation}

\begin{equation} 
	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w ) \leq maxW
\end{equation}

\begin{equation} 
	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m_k\}
\end{equation}

\begin{equation} 
	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.v + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m_k\}
\end{equation}

\begin{equation} 
	s.t.: \sum_{i=1}^{m} X_{ij}^k \leq 1; \ j \in \{1, 2, \ldots, n_k\}
\end{equation}

\begin{equation} \label{eq:lastnew}
	s.t.: p_i.to[k] = t^k_j.to\ \mbox{if} \ X_{ij}^k = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}
\end{equation}

\subsubsection{Ant Colony Optimization (ACO)}

In ACO \cite{Dorigo1992}, a population of $A$\/ ants performs independent search, where each ant $a$\/ finds its own allocation graph $G^a$. Each edge $e^k_{ij}=(p_i,t^k_j)$\/ has a general attractiveness $ga^k_{ij} = (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$, where $\tau^k_{ij}$\/ is the pheromone of this edge, and $\alpha$\/ and $\beta$\/ are constants. If $e^k_{ij}$\/ is included in $G^a$, ant $a$\/ increases the pheromone $\tau^k_{ij}$. Algorithm \ref{alg:aco} presents this heuristic.

\begin{algorithm}[H]

	\caption{ $SolveNode(ACO, k, G)$ } \label{alg:aco}
	\begin{algorithmic}[1]
		\State Let $G(V_k, E^Q_k)$
		\State $G^*(V_k, E^{N*}_k \cup E^Q_k) \gets Greedy(k, G, 1)$  \label{aco:greedy}
		\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^*(V_k,E^{N*}_k \cup E^Q_k)$
		\State $A \gets 4$ \label{aco:ants}
		\State $ga^k_{ij}   \gets 0$, $1 \leq i \leq m$, $1 \leq j \leq n_k$  \label{aco:init1}
		\State $\tau^k_{ij} \gets 1$, $1 \leq i \leq m$, $1 \leq j \leq n_k$   \label{aco:init2}
		\State $\alpha    \gets 1.0$ \label{aco:alpha}
		\State $\beta     \gets 4.0$ \label{aco:beta}
		\State $\rho      \gets 0.2$ \label{aco:rho}
		\State $stagnant \gets 0$
		\While{($stagnant < 3$) {\bf and} ($runtime < 0.7s$)} \label{aco:while}
			\For{$a \gets 1$ to $A$} \label{aco:g1}
				\State $G^a(V_k, E^{Na}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$
				\State $E \gets \{ (p_i,t^k_j) \mid (p_i,t^k_j) \not\in E^{Na}_k \}$
				\While{$|E| > 0$}
					\State $e \gets Roulette(E)$ biased through $ga^k_{ij}$  \label{aco:roulette}
					\State $E \gets E - \{e\}$		
					\If{$E^{Na}_k \cup \{e\}$ is feasible}
						\State $ E^{Na}_k \gets E^{Na}_k \cup \{e\} $ \label{aco:ant2}
					\EndIf	
				\EndWhile	
				\If{$f_s(G^{a}) > f_s(G^{+})$} \label{aco:gen1}
					\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^a(V_k, E^{Na}_k \cup E^Q_k)$  \label{aco:gen2}
						\For{each $(p_i,t^k_j) \in E^{N+}_k$}
							\State $\tau^k_{ij} \gets \tau^k_{ij} - \Delta\tau^k_{ij}$ \label{aco:phero1}
							\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{aco:ga1}
						\EndFor
				\EndIf 
			\EndFor \label{aco:g2}
			\If{$f_s(G^+) > f_s(G^*)$} \label{aco:best1}
				\State $stagnant \gets 0$
				\State $G^*(V_k, E^{N*}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$
				\For{each $(p_i,t^k_j) \in E^{N*}_k$}
					\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{aco:rho1}
					\State $\tau^k_{ij} \gets \tau^k_{ij} + \Delta\tau^k_{ij}$ \label{aco:phero2}
					\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{aco:ga2}
				\EndFor			
			\Else
				\State $stagnant \gets stagnant + 1$ \label{aco:best2}
			\EndIf
			\For{all $\tau^k_{ij}$} 
				\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{aco:rho2}
			\EndFor	
		\EndWhile	
		\State ${\bf return}$ $G^*(V_k, E^{N*}_k \cup E^Q_k)$		
	\end{algorithmic}
\end{algorithm}

A greedy solution is obtained in line \ref{aco:greedy}, being initially stored in $G^*$\/ and $G^+$. The number $A$\/ of ants was empirically set to 4 (line \ref{aco:ants}), yielding solutions with a good relation quality/time. The $m \times n_k$\/ matrices $\tau^k_{ij}$ and $ga^k_{ij}$\/ (lines \ref{aco:init1}-\ref{aco:init2}) are used and updated by all ants.

In each generation, $A$\/ ants are created (lines \ref{aco:g1}-\ref{aco:g2}). The number of generations depends on the time limit or stagnation in improving results (line \ref{aco:while}). Each ant $a$\/ elaborates its allocation graph $G^a$, considering the current values of $ga^k_{ij}$. To enforce solution diversification, if $(p_i,t^k_j)$ is included in $G^a$\/ then pheromone $\tau^k_{ij}$\/ is decreased (line \ref{aco:phero1}). Similarly, the values of $ga^k_{ij}$\/ are also updated (line \ref{aco:ga1}), where the exponents $\alpha$\/ and $\beta$\/ were chosen empirically: $\alpha$ was set to 1.0 and $\beta$\/ to $4.0$\/ (lines \ref{aco:alpha}-\ref{aco:beta}). For further explanation, see \cite{DorigoManiezzoColorni1996}.

The best solution of each generation is stored in $G^+$\/ (lines \ref{aco:gen1}-\ref{aco:gen2}), and the update of $G^*$\/ and the stopping criterion are performed on lines \ref{aco:best1}-\ref{aco:best2}. Each edge $(p_i,t^k_j)$\/ in some solution has its corresponding pheromone updated in a cumulative way (line \ref{aco:phero2}), that is, more pheromone will be given to the most relevant edges, where the increment is defined in \ref{eq:delta_tau}. The same with $ga^k_{ij}$\/ (line \ref{aco:ga2}).

\begin{equation} \label{eq:delta_tau}
	\Delta\tau^k_{ij} = \frac{ ga^k_{ij}\times t^k_j.s }{ A \times t^k_j.v}
\end{equation}

Pheromone $\tau^k_{ij}$\/ evaporates (lines \ref{aco:rho1} and \ref{aco:rho2}), where $\rho$\/ is a evaporation rate, which can be seen as a mechanism that delays premature convergence towards a sub-optimal solution. $\rho$\/ was empirically set to $0.2$\/ (line \ref{aco:rho}).

\subsubsection{Shims}

Finally, we present a new heuristic designed specifically for ACLP+RPDP, which we named \emph{Shims}. Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}). This strategy is based on a practical observation: usually, subsets of smaller and lighter items are saved for later adjustments to the remaining clearances.

\begin{table}[H]

	\begin{minipage}{0.08\linewidth}

	\end{minipage}\hfill % these two lines must be close to each other
	\begin{minipage}{0.34\linewidth}

		\includegraphics[scale=0.25]{Images/shims.png}
		\captionof{figure}{Shims of various thicknesses}
		\small\textsuperscript{Source: www.mscdirect.com/product/details/70475967}
		\label{fig:shims}

	\end{minipage}\hfill % these two lines must be close to each other
	\begin{minipage}{0.58\linewidth}

		\begin{tikzpicture}[scale=0.75, samples=100]
			\filldraw[fill=blue!3!white, draw=black] (0, 0) rectangle (12.5, 5.5);
			\draw[->] (.2, .6) -- coordinate (x axis mid) (12, .6);
			\node at (5, 0.3) {$\eta_1[i]$};
			\node at (5, 2.5) {|};
			\node at (6, 2.5) {shims};
			\node at (7, 2.5) {|};
			\node at (7, 0.3) {$\eta_2[i]$};
			\node at (0.3, 5) {$\theta_{ij}^k$};
			\draw[->] (.6, .2) -- coordinate (y axis mid) (0.6, 5.3);
			\node at (12, 0.3) {$e_{ij}^k$};
			\draw[smooth, domain = 0.09:2, color=black, thick] plot (.3+1/\x,{4.2+log2(\x)});
			\draw[->, dashed] (5, 0.6)--(5, 2.0);
			\draw[->, dashed] (7, 0.6)--(7, 1.5);
		\end{tikzpicture}

		\captionof{figure}{$n_k$\/ edges $e_{ij}^k$\/ of $p_i$\/ sorted by $\theta_{ij}^k$\/ in non-ascending order}
		\label{fig:whip}		
	\end{minipage}
\end{table}

In this heuristic, described in Algorithm \ref{alg:shims}, pallets $p_i$\/ are considered in non-descending order of $|p_i.d|$ (line \ref{shims:pallets}). For each $p_i$, the $n_k$\/ possible edges $e_{ ij}^k$\/ are considered in non-increasing order of $\theta_{ij}^k$\/ (line \ref{shims:edges}) in three phases: the greedy phase (lines \ref{shims:phase1a}-\ref{shims:phase1b}), the composition phase of the shims (lines \ref{shims:phase2a}-\ref{shims:phase2b}) and the selection phase of the best shim (lines \ref{shims:phase3a}-\ref{shims:phase3b}).

Figure \ref{fig:whip} represents the $n_k$\/ possible edges $e_{ij}^k$\/ of $p_i$\/ sorted by $\theta_{ij}^k$. \emph{Shims}\/ starts with a greedy solution, stopping at the edge with index $\eta_1[i]$ close to the local optimum (first phase). Then, considering even the edge with the index $\eta_2[i]$, it elaborates different possible complements for this pallet (second phase) and selects the best of these complements (third phase).

%Examples:
%$\to$ for a small problem if limit is 0.85 for the first phase of the solution process, remains 0.15 for second and third phases.
%$\to$ for a big problem if it is 0.98 for the first phase, it is 0.02 for the remaining phases.

\begin{algorithm}[H]
	\caption{ $SolveNode(Shims, k, G)$}  \label{alg:shims}
	\begin{algorithmic}[1]
		\State Let $G(V_k, E^Q_k)$
		\State Sort $M$ by $|p_i.d|$ in non-descending order \label{shims:pallets}
		\State $E^N_k \gets \varnothing$ 
		\State Let $E[1..m][1..n_k]$, where $E[i]$ is an array with $n_k$ edges $(p_i,t^k_j)$ sorted by $\theta_{ij}^k$ in non-ascending order, $1 \leq i \leq m$ \label{shims:edges}
		\State $gap \gets 2000/n_k*m$ \label{shims:gap} 
		\For{$i \gets 1$ to $m$} 
			\State $pv[i] \gets 0$		\label{shims:phase1a}
			\State $\eta_1[i] \gets 0$
			\While{($p_i.v - pv[i] > gap*p_i.v$) {\bf and} ($\eta_1[i]<n_k$)} \label{shims1:slack}
				\State $\eta_1[i] \gets \eta_1[i] + 1$ \label{shims:eta1}
				\State $e^k_{ij} \gets E[i][\eta_1[i]]$
				\If{($e^k_{ij}$ is feasible) {\bf and} ($t^k_j.v \leq p_i.v - pv[i]$)} 
					\State $E^N_k \gets E^N_k \cup \{e^k_{ij}\}$
					\State $pv[i] \gets pv[i] + t^k_j.v$ 
				\EndIf
			\EndWhile 
			\State $slack[i] \gets p_i.v - pv[i]$
			\State $\eta_2[i] \gets \eta_1[i]$
			\Repeat
				\State $\eta_2[i] \gets \eta_2[i] + 1$  \label{shims:eta2}
				\State $e^k_{ij} \gets E[i][\eta_2[i]]$
				\State $pv[i] \gets pv[i] + t^k_j.v$
			\Until{($\eta_2[i] = n_k$) {\bf or} ($ pv[i] \geq (1 + 3*gap)*p_i.v)$)} \label{shims:phase1b}

\State $Set \gets \varnothing$  \label{shims:phase2a}
\State $q \gets 1$
\State $shim[q] \gets \varnothing$
\State $volume \gets 0$
\For{$x \gets \eta_1[i]$ to $\eta_2[i]$}
\State $e_{ij}^k \gets E[i][x]$
\If {($e_{ij}^k \not\in E^N_k \cup shim[q]$) {\bf and} ($e_{ij}^k$ is feasible)}
\If{$t_j^k.v + volume \leq slack[i]$} \label{bincapacity}
\State $shim[q] \gets shim[q] \cup \{e_{ij}^k\}$	
\State $volume \gets volume + t_j^k.v$	
\Else \label{newbins}
\State $Set \gets Set \cup shim[q] $
\State $q \gets q+1$
\State $shim[q] \gets \varnothing$
\State $volume \gets 0$
\If{$t_j^k.v \leq slack[i]$} 
\State $shim[q] \gets shim[q] \cup \{e_{ij}^k\}$	
\State $volume \gets volume + t_j^k.v$
\EndIf
\EndIf
\EndIf
\EndFor   \label{shims:phase2b}

			\State $sh_w \gets S$, where $S \in Set$ and $\sum_{e_{ab}^k \in S} t_b^k.w$ is maximum \label{shims:phase3a}
			\State $sh_v \gets S$, where $S \in Set$ and $\sum_{e_{ab}^k \in S} t_b^k.v$ is maximum
			\State $sh_{best} \gets S$, where $S \in \{sh_w, sh_v\}$ and $\sum_{e_{ab}^k \in x} t_b^k.s$ is maximum
			\State $E^N_k \gets E^N_k \cup sh_{best}$ \label{shims:phase3b}
		\EndFor
		\State ${\bf return}$ $G(V_k, E^N_k \cup E^Q_k)$
	\end{algorithmic}
\end{algorithm}

In the first phase (lines \ref{shims:phase1a}-\ref{shims:phase1b}), a greedy and partial solution for $p_i$\/ is constructed by edges inclusion, until a certain slack or clearance is reached. This slack was empirically defined as $gap * p_i.v$\/ (line \ref{shims1:slack}), where $gap = 2000/n_k*m$\/ (line \ref{shims:gap}).

In the second phase (lines \ref{shims:phase2a}-\ref{shims:phase2b}), a set of shims named $Set$\/ is created, where each shim is formed by a group of edges in the range $[\eta_1[i],\eta_2[i]]$, whose total volume is limited by $slack[i]$. In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it Next-Fit}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \cite{JohnsonGarey1985}. Basically, shims are created by accumulating the following edges, taking $slack[i]$\/ as a limit.

In the third phase (lines \ref{shims:phase3a}-\ref{shims:phase3b}), the best shim in $Set$\/ is chosen. Initially, two shims are found: $sh_w$\/ with larger weight and $sh_v$\/ with larger volume. Among the two, the shim with the highest score will be chosen and its edges are inserted into $E^N_k$.

\section{Implementation and results}
\label{sec6}

This section is composed of two parts: the generation of test instances and the results obtained in our implementations.

\subsection{Instances generation}
\label{items}

As we are dealing with a new problem, which until now had not been modelled in the literature, we have to create our own benchmarks. For this, we based on the characteristics of real airlifts carried out by the {\em Brazilian Air Force}, as described below.

In the military airlift carried out in Brazil from 2008 to 2010, 23\% of the items weighed between $10kg$ and $20kg$, 22\% from $21kg$ to $40kg$, 24\% from $41kg$ to $80kg$, 23\% from $81kg$ to $200kg$, and 8\% between $201kg$ and $340kg$. These five groups of items are described in Table \ref{tab:weights}. On the other hand, the average density of these items is approximately $246 kg/m^3$.

\begin{table}[H]
		\centering
		\caption{Items weight distribution ($kg$) in Brazil}  \label{tab:weights}
		\begin{tabular}{c c c c }
			\toprule
			$item$ & $P$ & $low$ & $high$  \\		
			\midrule
			1              & 0.23           & 10  & 20\\
			2              & 0.22           & 21  & 40\\
			3              & 0.24           & 41  & 80 \\		
			4              & 0.23           & 81  & 200\\
			5              & 0.08           & 201 & 340 \\
			\bottomrule
		\end{tabular}
\end{table}

$ItemsGeneration$, which generates $N$, is described in Algorithm \ref{alg:itemsgen}. The parameter $scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and the parameter $volume$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}). To avoid simply loading all items, we use $volume > 1$.

\begin{algorithm}[H]
	\caption{$ItemsGeneration(scenario,volume)$}  \label{alg:itemsgen}
	\begin{algorithmic}[1]
		\State Let $L$ and $M$ be according to $scenario$ \label{ig:LM}
		\State $limit \gets volume \times \sum_{i=1}^{m} p_i.v$ \label{ig:extended}
		\For{$k \gets 0$ to $K$}
			\State $N_k \gets \varnothing$
			\State $j \gets 0$
			\State $vol \gets 0$	\label{ig:totals}	
			\While{$vol < limit$}
				\State $j \gets j+1$
				\Repeat
					\State $t_j^k.to \gets RandomInt(0, K)$ \label{ig:dest}
				\Until{$t_j^k.to \neq k$}
				\State $x = Roulette(item)$ biased through $P$ \label{ig:weight1}	
				\State $t_j^k.w \gets RandomReal(low(x), high(x))$        \label{ig:weight2}		
				\State $t_j^k.s \gets \round{100 \times (1 - \log_{10}(RandomInt(1, 9)))} $ \label{ig:score}
				\State $t_j^k.v \gets t_j^k.w / RandomReal(148, 344)$ \label{ig:volume}
				\State $vol \gets vol + t_j^k.v$ 
				\State $N_k \gets N_k \cup \{t_j^k\}$ 
			\EndWhile
			\State $n_k \gets j$
		\EndFor
		\State $N \gets \bigcup_{0 \leq k \leq K} N_k$ 
		\State ${\bf return}$ $N$
	\end{algorithmic}
\end{algorithm}

For each generated $t^k_j$\/ item, its destination is uniformly random selected (line \ref{ig:dest}), its weight has a distribution according to Table \ref{tab:weights} (lines \ref{ig:weight1}-\ref{ig:weight2}), its score varies $100$\/ (highest) and $5$\/ (lowest) according to a logarithmic scale (line \ref{ig:score}, and its volume is randomly defined from the density, where we allow a variation of 40\% more or less than the average density of $246 kg/m^3$\/ (line \ref{ig:volume}).

\subsection{Results obtained}

In the tests performed, we used a 64-bit, 16GB, 3.6GHz, eight-core processor with {\it Linux Ubuntu 22.04.1 LTS 64-bit} as the operational system and {\it Python 3.10.4}\/ as the programming language. We also used the well-known solver {\it Gurobi}\/ (www.gurobi.com), version 9.5.2.

\begin{table}[H]
	\centering
	\caption{xxx} 
	\footnotesize

\begin{tabular}{c c c c} 
	\toprule 
	\rowcolor{Gray}\multicolumn{4}{l}{Node 0: GRU solution} \\ 
	$Pallet$ & $From:To$ & $Kg$ & $m^3$ \\ 
	\midrule 
	1 & GRU:GIG & 2504 & 10.7 \\ 
	2 & GRU:GIG & 2405 & 10.4 \\ 
	3 & GRU:GIG & 3710 & 15.2 \\ 
	4 & GRU:SSA & 3386 & 14.3 \\ 
	5 & GRU:SSA & 3835 & 15.6 \\ 
	6 & GRU:SSA & 3124 & 12.6 \\ 
	7 & GRU:SSA & 3518 & 13.8 \\ 
	\midrule 
	Score: 8911.0 & Weight: 0.86 & Volume: 1.10 & Torque: 2.24 \\ 
	\toprule 
	\rowcolor{Gray}\multicolumn{4}{l}{Node 1: GIG solution} \\ 
	$Pallet$ & $From:To$ & $Kg$ & $m^3$ \\ 
	\midrule 
	1 & GIG:SSA & 2585 & 12.2 \\ 
	2 & GIG:SSA & 3603 & 17.0 \\ 
	3 & GIG:SSA & 2881 & 12.8 \\ 
	4 & GIG:SSA & 3039 & 13.7 \\ 
	5 & GIG:GRU & 2075 &  8.9 \\ 
	6 & GIG:GRU & 3478 & 14.5 \\ 
	7 & GIG:GRU & 1632 &  7.4 \\ 
	\midrule 
	Score: 3546.0 & Weight: 0.74 & Volume: 1.03 & Torque: 3.13 \\ 
	\toprule 
	\rowcolor{Gray}\multicolumn{4}{l}{Node 2: SSA solution} \\ 
	$Pallet$ & $From:To$ & $Kg$ & $m^3$ \\ 
	\midrule 
	1 & SSA:GRU & 3027 & 13.2 \\ 
	2 & SSA:GRU & 1287 &  6.2 \\ 
	3 & SSA:GRU & 3044 & 15.0 \\ 
	4 & SSA:GRU & 1939 &  9.1 \\ 
	5 & SSA:GRU & 2010 &  8.9 \\ 
	6 & SSA:GRU & 2686 & 11.8 \\ 
	7 & SSA:GRU & 1410 &  5.8 \\ 
	\midrule 
	Score: 3632.0 & Weight: 0.59 & Volume: 0.83 & Torque: 2.50 \\ 
	\bottomrule 
	\multicolumn{4}{l}{Tour 0\ \ \ \ \ S/C: 4.291\ \ \ \ \ \ \ \ \ \ elapsed: 0.05s\ \ \ \ \ method:\ Greedy} \\ 
	\bottomrule 
\end{tabular}

	\normalsize

\end{table}

\section{Conclusions}
\label{sec7}

xxxxx

\section*{Acknowledgments}

This research was partially supported by \textit{S\~{a}o Paulo Research Foundation} (FAPESP, grant 2016/01860-1).

%\bibliographystyle{elsarticle-num} % descomentar antes de submeter
\bibliographystyle{elsarticle-harv} % comentar antes de submeter

%\section*{References}

\bibliography{references}

\end{document}

### 2022-10-17 main.py

@@ -10,6 +10,7 @@ import optcgcons
 import shims
 import shims_mp
 import aco
+import aco_mp
 import greedy
 
 
@@ -183,11 +184,11 @@ if __name__ == "__main__":
     # methods.DATA = "data50"
     # methods.DATA = "data100"
 
-    scenario = 2
+    scenario = 1
 
     if scenario == 1:
-        instances = [1,2,3,4,5,6,7]
-        # instances = [1]
+        # instances = [1,2,3,4,5,6,7]
+        instances = [1]
     if scenario == 2:
         # instances = [1,2,3,4,5,6,7]
         instances = [1]

### 2022-11-02 latex/comp_or.tex
ACO working in no_append folder

--- 2022-10-17 comp_or.tex
@@
 
+\vspace{2.0mm}
+
 \begin{table}[H]
 	\centering
-	\caption{xxx} 
+	\caption{Methods and scenarios with $volume =1.5$}  \label{tab:50}
 	\footnotesize
-
-\begin{tabular}{c c c c} 
+	\begin{tabular}{c|c|cccccc|c}
 		\toprule
-	\rowcolor{Gray}\multicolumn{4}{l}{Node 0: GRU solution} \\ 
-	$Pallet$ & $From:To$ & $Kg$ & $m^3$ \\ 
-	\midrule 
-	1 & GRU:GIG & 2504 & 10.7 \\ 
-	2 & GRU:GIG & 2405 & 10.4 \\ 
-	3 & GRU:GIG & 3710 & 15.2 \\ 
-	4 & GRU:SSA & 3386 & 14.3 \\ 
-	5 & GRU:SSA & 3835 & 15.6 \\ 
-	6 & GRU:SSA & 3124 & 12.6 \\ 
-	7 & GRU:SSA & 3518 & 13.8 \\ 
-	\midrule 
-	Score: 8911.0 & Weight: 0.86 & Volume: 1.10 & Torque: 2.24 \\ 
+		{$method$}                        & {\bf Results}    & {\bf 1}     &  {\bf 2}    & {\bf 3}      & {\bf 4}      & {\bf 5}      & {\bf 6}     & \specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
 		\toprule
-	\rowcolor{Gray}\multicolumn{4}{l}{Node 1: GIG solution} \\ 
-	$Pallet$ & $From:To$ & $Kg$ & $m^3$ \\ 
-	\midrule 
-	1 & GIG:SSA & 2585 & 12.2 \\ 
-	2 & GIG:SSA & 3603 & 17.0 \\ 
-	3 & GIG:SSA & 2881 & 12.8 \\ 
-	4 & GIG:SSA & 3039 & 13.7 \\ 
-	5 & GIG:GRU & 2075 &  8.9 \\ 
-	6 & GIG:GRU & 3478 & 14.5 \\ 
-	7 & GIG:GRU & 1632 &  7.4 \\ 
-	\midrule 
-	Score: 3546.0 & Weight: 0.74 & Volume: 1.03 & Torque: 3.13 \\ 
-	\toprule 
-	\rowcolor{Gray}\multicolumn{4}{l}{Node 2: SSA solution} \\ 
-	$Pallet$ & $From:To$ & $Kg$ & $m^3$ \\ 
-	\midrule 
-	1 & SSA:GRU & 3027 & 13.2 \\ 
-	2 & SSA:GRU & 1287 &  6.2 \\ 
-	3 & SSA:GRU & 3044 & 15.0 \\ 
-	4 & SSA:GRU & 1939 &  9.1 \\ 
-	5 & SSA:GRU & 2010 &  8.9 \\ 
-	6 & SSA:GRU & 2686 & 11.8 \\ 
-	7 & SSA:GRU & 1410 &  5.8 \\ 
-	\midrule 
-	Score: 3632.0 & Weight: 0.59 & Volume: 0.83 & Torque: 2.50 \\ 
-	\bottomrule 
-	\multicolumn{4}{l}{Tour 0\ \ \ \ \ S/C: 4.291\ \ \ \ \ \ \ \ \ \ elapsed: 0.05s\ \ \ \ \ method:\ Greedy} \\ 
+
+		\multirow{2}{*}{\textbf{ACO}}     & $f$             & 21.93       & {\bf 6.77}& 8.85      & 9.65       & 11.75      & 11.54      &  0.934 \\ %21.927   &   19.65s
+		                                  &{\bf runtime (s)}& 20          & 4        & 17       & 86        & 513       & 3,596     &  2.714 \\ 
+		\midrule[.1pt]
+
+		\multirow{2}{*}{\textbf{Shims}}   & $f$             & 22.28       & {\bf 6.77} & 8.92     & 9.60       & 11.78      & 11.61      &  0.948       \\ 
+		                                  &{\bf runtime (s)}& <1         & 2         & 8       & 31        & 166       & 1,087     &  {\bf 8.856} \\
+
+		\midrule[.1pt]
+
+		\multirow{2}{*}{\textbf{ACO-mp}}  & $f$             & 9.43       & {\bf 6.77}& 8.85      & 9.65       & 11.75      & 11.54      &  0.934 \\ 
+                                          &{\bf runtime (s)}& 3         & 4        & 17       & 86        & 513       & 3,596     &  2.714 \\ 
+        \midrule[.1pt]
+
+        \multirow{2}{*}{\textbf{Shims-mp}} & $f$             & 10.18       & {\bf 6.77} & 8.92     & 9.60       & 11.78      & 11.61      &  0.948       \\ 
+                                           &{\bf runtime (s)}& $< 1$         & 2         & 8       & 31        & 166       & 1,087     &  {\bf 8.856} \\
+
         \bottomrule		

### 2022-11-02 no_append/aco.py

@@ -3,6 +3,7 @@ import time
 import math
 import numpy as np
 import methods as mno
+import copy
 
 ALPHA = 1   # pheromone exponent (linearly affects attractivity)
 BETA  = 4   # heuristic exponent (exponentialy affects attractivity)
@@ -66,49 +67,40 @@ def Solve( pallets, items, startTime, cfg, k):  # items include kept on board
     print("\nAnt Colony Optimization for ACLP+RPDP")
 
     antsField = mno.mountEdges(pallets, items, cfg)
-    Nbhood    = mno.edgesCopy(antsField)
+    edges     = mno.mountEdges(pallets, items, cfg)
 
-    # initialize the best solution so far        1.0 totally greedy
-    Gbest = mno.Solution(Nbhood, pallets, items, 1.0, cfg, k)
+    # initialize the best solution so far           1.0 totally greedy
+    Gbest = mno.Solution(antsField, pallets, items, 1.00, cfg, k)
+    Ginit = mno.Solution(    edges, pallets, items, 0.85, cfg, k)
 
     numPallets = len(pallets)
     numItems   = len(items)
- 
-    stagnant = 0
     numAnts = 0
-    while stagnant < NANTS:# and (time() - startTime) < mno.SEC_BREAK:
+    stagnant = 0
+
+    while stagnant < NANTS and (time.perf_counter() - startTime) < mno.SEC_BREAK:
 
         Glocal = Gbest
-   
-        for _ in np.arange(NANTS): # sequential ants
 
-            # if (time() - startTime) > mno.SEC_BREAK:
-            #     break
-        
+        for _ in np.arange(NANTS):
+
+            if (time.perf_counter() - startTime) > mno.SEC_BREAK:
+                break
+
             numAnts += 1
             
-            Nbhood = mno.edgesCopy(antsField)
+            Gant = Ginit
 
-            # initilize a solution for each ant
-            Gant = mno.Solution(Nbhood, pallets, items, 0.9, cfg, k)
+            Nbhood   = [ce for ce in Gant.Edges if not ce.InSol]
+            attracts = [ce.Attract for ce in Nbhood]
 
-            # antsField edges are a heritage from previous ants
-            # prepare for the random proportional selection in pickFromNbhood method
-            attracts   = [ce.Attract for ce in antsField]
-
-            # builds or completes a solution by exploring the entire items neighborhood
             while Nbhood:
 
-                # if ((time() - startTime) > mno.SEC_BREAK):
[…]

### 2022-11-08 aco.py
Preparing to experimentation through ssh

@@ -26,6 +26,8 @@ def rouletteSelection2(values):
 
 # getDeltaTau calculates the pheromone to be dropped by each on ants tracks
 def getDeltaTau(score, bestSoFar, numAnts):
+    if bestSoFar == 0:
+        bestSoFar = 1000
     DeltaTau = (score - bestSoFar)/bestSoFar # at this point DeltaTau may be positive ou negative learning
     DeltaTau /= numAnts*numAnts
     return DeltaTau
@@ -127,6 +129,9 @@ def Solve( pallets, items, startTime, cfg, k, limit):  # items include kept on b
 
         updatePheroAttract(Glocal.S, Gbest.S, antsField, NANTS)
 
+
+    if initialS == 0:
+        initialS = 1000
 
     print(f"Used {numAnts} ants | ratio {Glocal.S/initialS:.3f} | {improvements} improvements")
 

### 2022-11-08 methods.py

@@ -5,8 +5,6 @@ import math
 import numpy as np
 from time import time
 
-NCPU = os.cpu_count()
-
 SEC_BREAK = 3.
 
 CITIES = ["GRU", "GIG", "SSA", "CNF", "CWB", "BSB", "REC"]

### 2022-11-08 testing.py

@@ -102,13 +102,13 @@ E = []
 startNodeTime = time.perf_counter()
 
 if method == "Greedy":
-    E = greedy.Solve(pallets, items, cfg, 0)
+    E = greedy.Solve(pallets, items, cfg, k)
 
 if method == "Shims_p":
-    E = shims_p.Solve(pallets, items, cfg, 0, limit)
+    E = shims_p.Solve(pallets, items, cfg, k, limit)
 
 if method == "Shims":
-    E = shims.Solve(pallets, items, cfg, 0, limit)
+    E = shims.Solve(pallets, items, cfg, k, limit)
 
 if method == "ACO":
     E =   aco.Solve( pallets, items, startNodeTime, cfg, k, limit)
@@ -165,11 +165,14 @@ if len(E) > 0:
 
     epsilom = tau/cfg.maxTorque
 
+    # greedyScore = 50842.0 # data50
+
     sol += f"Score: {sNodeAccum}\t"
     sol += f"Weight: {wNodeAccum/cfg.weiCap:.2f}\t"
     sol += f"Volume: {vNodeAccum/cfg.volCap:.2f}\t"
     sol += f"Torque: {epsilom:.2f}\n"
     sol += f"Elapsed: {elapsed:.2f}\n"
+    # sol += f"Ratio: {sNodeAccum/greedyScore:.3f}\n"
 
     state = "Feasible"
     for n in itemsCount:

### 2022-11-09 latex/comp_or.tex
Preparing to experimentation through ssh
Update commands

@@
 \documentclass[preprint,authoryear]{elsarticle}
+% sudo apt-get install -y texlive-publishers
 
@@
 \pgfplotsset{compat=newest}
-\usepackage{algorithm}
+\usepackage{algorithm} % sudo apt-get install -y texlive-science
 
@@
 
-\journal{European Journal of Operational Research}
+\journal{Computers \& Operations Research}
 
@@
 
-\title{Air cargo load and route planning in pickup and delivery operations}
+\title{Parallel air palletization and tour planning for simultaneous pickup and delivery}
 
@@
 Also, there is no technological assistance to help the load and trip planners with the huge number of demands for transport in each hub.
+
 This enables other risks such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
-We contributed by modelling and solving a new problem of planning the loading and routing of an aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
-This new hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modelled using standardised pallets in fixed positions, obeying the centre of gravity constraints, delivering each item to its destination, and minimising fuel consumption costs.
-We also contributed by carrying out multiple experiments with a commercial solver and four well-known meta-heuristics on synthetic data based on real data from the {\it Brazilian Air Force}\/ transportation history.
+We contributed by modeling and solving a new problem of planning the loading and routing of an aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
+
+This new hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modeled using standardized pallets in fixed positions, obeying the center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
+
+We also contributed by carrying out multiple experiments with a commercial solver and the well-known Ant Colony Optimization on synthetic data based on real data from the {\it Brazilian Air Force}\/ transportation history.
+
 These challenging benchmark instances are made publicly available.
-We also created a heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it was the unique method that managed to solve all testing scenarios.
+We also created a parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it was the unique method that managed to solve all testing scenarios.
 
@@
 \begin{keyword}
-Air Cargo \sep Load Planning \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Vehicle Routing
+Air Cargo \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Vehicle Routing \sep Parallel computing
 \end{keyword}
@@
 401]{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP) and {\it Weight and Balance Problem} (WBP).
-Several aspects were considered in this modelling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc.
+Several aspects were considered in this modeling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc.
 They also presented a comprehensive bibliographic survey of solving methods that had been developed in different situations.
@@
 However, there are still other important challenges in air cargo transport that go beyond the definition of the ACLPP, especially with regard to the flight itinerary and the loading and unloading at each destination (or node) of this travel plan.
-In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Pickup and Delivery Problem} (PDP), and the search for the lowest cost route, which is the well-known {\it Travelling Salesman Problem} (TSP).
+In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Pickup and Delivery Problem} (PDP), and the search for the lowest cost route, which is the well-known {\it Traveling Salesman Problem} (TSP).
 
@@
 Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a pre-defined flight plan restricted to two legs.
-It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modelled through a container sequencing problem.
+It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem.
 Referring directly to this work, \cite[p.
@@
 
-In the {\it Brazilian Air Force}\/ missions, hundreds of items can be carried at each node, where the objectives are to prioritise the transport of the most important items and minimise the cost of fuel along the route.
-As standardised pallets are used with predefined positions on the aircraft, it is possible to carry out loading and unloading at each node in around two hours.
+In the {\it Brazilian Air Force}\/ missions, hundreds of items can be carried at each node, where the objectives are to prioritize the transport of the most important items and minimize the cost of fuel along the route.
+As standardized pallets are used with predefined positions on the aircraft, it is possible to carry out loading and unloading at each node in around two hours.
 However, there is no technological assistance that guarantees the achievement of these objectives.
 
-This work proposes a method that attends these objectives: a pallet building and arrangement plan with a routing option that maximises the benefit-cost ratio for the smooth execution of pickup and delivery transport missions.
+This work proposes a method that attends these objectives: a pallet building and arrangement plan with a routing option that maximizes the benefit-cost ratio for the smooth execution of pickup and delivery transport missions.
 We develop a heuristic that can be executed on a simple handheld computer (like a laptop or a tablet) and that provides a solution quickly enough to keep this cargo handling time under an hour.
@@
 
-This article is organised into six more sections.
+This article is organized into six more sections.
 In Section \ref{sec2}, we give a brief review of the literature.
@@
 Seven types of items were considered to be allocated in 17 to 42 positions.
-With non-linear programming and heuristics, they present a solution that minimises positioning changes in the intermediate node, optimising the load balancing in the aircraft.
+With non-linear programming and heuristics, they present a solution that minimizes positioning changes in the intermediate node, optimizing the load balancing in the aircraft.
 
@@
 
-%\cite{Gueret2003} proposed a method for loading military aircraft (specifically, the french military) for airlift operations.
-They used a two-phased solution method consisting of heuristics that quickly compute suboptimal initial solutions for a 2-dimensional bin packing and a local search algorithm to improve upon the initial solution, while ignoring the aircraft center of gravity.
-
 \cite{MongeauBes2003} presented a method based on linear integer programming to solve the problem of choosing and positioning containers on the {\it Airbus 340-300}.
-Safety and stability restrictions were considered, with the objective of minimising fuel consumption.
+Safety and stability restrictions were considered, with the objective of minimizing fuel consumption.
 
@@
 
-\cite{KaluznyBohdanL2009Oalb} developed a mixed integer linear programming model to arrange a set of items in a military context that optimises the load balance.
+\cite{KaluznyBohdanL2009Oalb} developed a mixed integer linear programming model to arrange a set of items in a military context that optimizes the load balance.
 
@@
 
-\cite{Limbourg2012} developed a mixed-integer program for optimally rearranging a set of pallets into a compartmentalised cargo aircraft, specifically the {\it Boeing 747}.
+\cite{Limbourg2012} developed a mixed-integer program for optimally rearranging a set of pallets into a compartmentalized cargo aircraft, specifically the {\it Boeing 747}.
 
@@
 
-As already mentioned, \cite{LurkinSchyns2015} was the first work that simultaneously modelled WBP and PDP in air cargo transport.
+As already mentioned, \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and PDP in air cargo transport.
 The authors demonstrated that this problem is NP-hard and performed some experiments with real data, noting that their model offers better results than those obtained manually.
@@
 
-\cite{PaquaySchynsLimbourg2016} presented a mathematical modelling to optimise the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
+\cite{PaquaySchynsLimbourg2016} presented a mathematical modeling to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
 Its objective is to maximise the volume used in containers, considering load balancing restrictions, the presence of fragile items and the possibility of rotating these boxes.
@@
 
-\cite{YangLiuGao2018} modelled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing in order to minimise fuel consumption.
-
-\cite{wong2020}A developed a mathematical model and a tool based on mixed integer programming for optimising cargo in aircraft with different pallet configurations.
+\cite{YangLiuGao2018} modeled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing in order to minimise fuel consumption.
+
+\cite{wong2020}A developed a mathematical model and a tool based on mixed integer programming for optimizing cargo in aircraft with different pallet configurations.
 Balance restrictions and the presence of dangerous items were considered.
-\cite{eugene2021} integrated this tool to a digital simulation model, with a visualisation and validation system, based on sensors that alert about load deviations.
-
-\cite{zhao2021} proposed a new modelling for WBP based on mixed integer programming.
-Instead of focusing on the centre of gravity (CG) deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
+\cite{eugene2021} integrated this tool to a digital simulation model, with a visualization and validation system, based on sensors that alert about load deviations.
+
+\cite{zhao2021} proposed a new modeling for WBP based on mixed integer programming.
+Instead of focusing on the center of gravity (CG) deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
 
@@
 
-	\item At each node of the flight plan, the items to be allocated are characterised by weight, volume, scores, and previously known destinations, but do not have dimensions.
+	\item At each node of the flight plan, the items to be allocated are characterized by weight, volume, scores, and previously known destinations, but do not have dimensions.
 We leave the consideration of 2D or 3D items for future work.
@@
 
-	\item Finally, as we are interested in minimising fuel costs, we disregarded some lower ones as {\it handling operation costs}.
+	\item Finally, as we are interested in minimizing fuel costs while keeping the CG in its operational range, we disregarded some lower ones as handling costs.
 
@@
 Tables \ref{tab:smaller} and \ref{tab:larger} show the parameters in both cases.
-{\it CGx} and {\it CGy} refer to the relative distances of pallet centroids (in metres) in relation to the CG of aircraft along both axes.
+{\it CGx} and {\it CGy} refer to the relative distances of pallet centroids (in meters) in relation to the CG of aircraft along both axes.
 In both aircraft, as the ramps have an inclination of 25 degrees, we made the necessary corrections in {\it CGx}, {\it Weight} and {\it Volume limits} of the corresponding pallets.
@@
 	\item in the larger aircraft, the CG of the payload must be at a maximum lateral distance of $limit^{CG}_{lat}$ from the CG of the aircraft;
-	\item in the larger aircraft, pallets are distributed in two identical rows (with odd and even indices, respectively), and their centroids are at a distance $d^{CG}_{pallet}$ from the centre-line of the aircraft.
+	\item in the larger aircraft, pallets are distributed in two identical rows (with odd and even indices, respectively), and their centroids are at a distance $d^{CG}_{pallet}$ from the center-line of the aircraft.
 \end{itemize}
@@
 	\begin{tabular}{MT}
-		&Informally, ACLP+RPDP can be summarised as follows:\\
-		\midrule
-		max &  (items score sum) / (tour cost) of picked up and delivered items at each node on a tour  \\
+		&Informally, ACLP+RPDP can be summarized as follows:\\
+		\midrule
+		max &  (items score sum) / (tour cost) of picked up and delivered items on a tour  \\
 		\midrule
@@
 
-Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modelling of ACLP+RPDP, which is one of the contributions of our work.
+Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modeling of ACLP+RPDP, which is one of the contributions of our work.
 
@@
 
-The mathematical modelling of this problem is described in the equations below.
+The mathematical modeling of this problem is described in the equations below.
 
@@
 
-The objective of this problem is to find a permutation $\pi \in S_K$\/ that maximises the function $f_\pi(\tilde{s},\tilde{c})$ \ref{eq:maxf}.
+The objective of this problem is to find a permutation $\pi \in S_K$\/ that maximizes the function $f_\pi(\tilde{s},\tilde{c})$ \ref{eq:maxf}.
 In this way, the flight plan will be $l_0, l_{\pi(1)}, \ldots, l_{\pi(K)}, l_0$.
@@
 
-Once the assumptions of this work and the mathematical modelling of the problem are presented, it is easy to see that ACLP+RPDP is NP-hard.
+Once the assumptions of this work and the mathematical modeling of the problem are presented, it is easy to see that ACLP+RPDP is NP-hard.
 In a similar way to \cite[p.
@@
 Real cases are more complex as they have hundreds of different items in each node and involve three intractable sub-problems: APP, WBP and PDP.
-Through the mathematical modelling presented in the previous section, we verify that {\it Mixed-Integer Programming}\/ (MIP) is not able to solve these cases in feasible time.
-Thus, it is necessary to adopt some strategy to find a viable solution, not necessarily optimal, that seeks to maximise the objective function $f$.
+Through the mathematical modeling presented in the previous section, we verify that {\it Mixed-Integer Programming}\/ (MIP) is not able to solve these cases in feasible time.
+Thus, it is necessary to adopt some strategy to find a viable solution, not necessarily optimal, that seeks to maximize the objective function $f$.
 
@@
 
-\newcolumntype{C}{>{\raggedright}p{0.18\textwidth}}
-\newcolumntype{D}{>{\raggedright\arraybackslash}p{0.82\textwidth}}
+\newcolumntype{C}{>{\raggedright}p{0.19\textwidth}}
+\newcolumntype{D}{>{\raggedright\arraybackslash}p{0.81\textwidth}}
 
@@
 
-		{\it Data structure:}       & Items, pallets and edges are arrays of \emph{python} class objects, which are initialized with the same data loaded from text files. \\
-
-		{\it Common procedures:}    & All heuristics use the same procedures for selecting and inserting edges, and printing solutions.  \\
-
-		{\it Stopping criterion:} & All heuristics have the same time limit of $0.7$ seconds per node.\\
+		{\it Data structure:}     & Items, pallets and edges are arrays of \emph{Python} class objects, which are initialized with the same data loaded from text files. \\
+
+		{\it Heuristic's procedures:}  & All heuristics use the same procedures for selecting and inserting edges.  \\
+
+		{\it Stopping criterion:} & All methods have the same time limit of $0.7$ seconds per node.\\
+
+		{\it Common procedures:}  & All methods have the same procedures for: solution printing, pallets destinations and initial parameters setting, and solution integrity checking.\\
 
@@
 \begin{itemize}
-	\item $RandomReal(r_1,r_2)$: randomly selects a real number in $[r_1,r_2]$, where $r_1$ and $r_2$ are real numbers;
-	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$, where $i_1$ and $i_2$ are integer numbers;
+	\item $RandomReal(r_1,r_2)$: randomly selects a real number in $[r_1,r_2]$;
+	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$;
 	\item $RandomChoice(set)$: randomly selects an element from $set$;
@@
 
-To compare allocation graphs generated by the heuristics, we choice the one that offer a higher associated score, since the CG deviation is limited by the restrictions.
-In the pseudocodes presented below, given an allocation graph $G$\/ for the node $l_k$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation(k,G) $, described in Algorithm \ref{alg:eval}.
-
-Below, we present six implementations for $SolveNode$: the first using a MIP solver, and five with developed heuristics: {\it Ant Colony Optimization}\/ (ACO), {\it Noising Method Optimization}\/ (NMO), {\it Tabu Search}\/ (TS), {\it Greedy Randomized Adaptive Search Procedure}\/ (GRASP) and {\it Shims}.
+To compare allocation graphs generated by the heuristics, we choose the one that offer a higher associated score, since the CG deviation is limited by the restrictions.
+
+In the pseudo-codes presented below, given an allocation graph $G$\/ for the node $l_k$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation(k,G) $, described in Algorithm \ref{alg:eval}.
+
+Below, we present five implementations for $SolveNode$: the first using a MIP solver, and four with developed heuristics: {\it Ant Colony Optimization}\/ (ACO), {\it Parallel Ant Colony Optimization}\/ (ACO-p),  and {\it Shims},  and {\it Parallel Shims (Shims-p)}.
 
@@
 
-In order to find a viable solution for ACLP+RPDP, the strategy adopted in $SolveTour$\/ previously defines the values of some variables.
+To find a viable solution for ACLP+RPDP, the strategy adopted in $SolveTour$\/ previously defines the values of some variables.
 Concretely, all permutations between the nodes are tested, the set of nodes to be visited is updated, the consolidated items that remain on board are reallocated to minimize the CG deviation, and the pallets destination is determined according to the volume of items available for shipment.
-In this way, the mathematical modelling for $SolveNode(MIP,k,G)$\/ becomes simpler, which finds an allocation of available items in node $l_k$\/ using previously defined values of $L_k$, $p_i.to[k]$, $Y^k_{iq}$\/ and $a^k_q$.
+In this way, the mathematical modeling for $SolveNode(MIP,k,G)$\/ becomes simpler, which finds an allocation of available items in node $l_k$\/ using previously defined values of $L_k$, $p_i.to[k]$, $Y^k_{iq}$\/ and $a^k_q$.
 
@@
 
-As we are dealing with a new problem, which until now had not been modelled in the literature, we have to create our own benchmarks.
+As we are dealing with a new problem, which until now had not been modeled in the literature, we have to create our own benchmarks.
 For this, we based on the characteristics of real airlifts carried out by the {\em Brazilian Air Force}, as described below.
@@
 
-In the tests performed, we used a 64-bit, 16GB, 3.6GHz, eight-core processor with {\it Linux Ubuntu 22.04.1 LTS 64-bit} as the operational system and {\it Python 3.10.4}\/ as the programming language.
+In the tests performed, we used a 64-bit, 128GiB, 3.6GHz, 48-core processor, 2 threads per core, Intel(R) Xeon(R), with {\it Linux CentOS 8} as the operational system and {\it Python 3.10.4}\/ as the programming language.
 We also used the well-known solver {\it Gurobi}\/ (www.gurobi.com), version 9.5.2.
+
+For parallelization, we used the {\it Python multiprocessing} package, which implements {\it process-based parallelism}.
+According to https://docs.python.org/3/library/multiprocessing.html, Multiprocessing is a package that supports spawning processes using an API similar to the threading module.
+The multiprocessing package offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
+Because of this, the multiprocessing module lets the programmer take full advantage of the fact that a machine has more than one processor. This decision for multiprocessing relies on the need to exchange information among processes, which would not be possible with thread-based parallelism.
+
+Process: The operating system’s spawned and controlled entity that encapsulates an executing application.
+A process has two main functions.
+The first is to act as the resource holder for the application, and the second is to execute the instructions of the application.
+
+— PAGE 271, THE ART OF CONCURRENCY, 2009. The Art of Concurrency: A Thread Monkey's Guide to Writing Parallel Applications 1st Edition, Kindle Edition by Clay Breshears  (Author)
+
+parallel shared memory algorithms
 
@@
 	\centering
-	\caption{Methods and scenarios with $volume =1.5$}  \label{tab:50}
+	\caption{Methods and scenarios with $volume=1.2$}  \label{tab:20}
 	\footnotesize
-	\begin{tabular}{c|c|cccccc|c}
+	\begin{tabular}{l|c|cccccc|c}
 		\toprule
@@
 
-		\multirow{2}{*}{\textbf{ACO}}     & $f$             & 21.93       & {\bf 6.77}& 8.85      & 9.65       & 11.75      & 11.54      &  0.934 \\ %21.927   &   19.65s
-		                                  &{\bf runtime (s)}& 20          & 4        & 17       & 86        & 513       & 3,596     &  2.714 \\ 
+		\multirow{2}{*}{\textbf{Shims}}   & $f$              & 6.47        & 3.90        & 3.30         & 3.25         & 3.13         & 6.03        &   \\ 
+		                                  &{\bf run-time (s)}& <1          & 1           & 2            & 18           & 46           & 138         &   \\ %206s
 		\midrule[.1pt]
 
-		\multirow{2}{*}{\textbf{Shims}}   & $f$             & 22.28       & {\bf 6.77} & 8.92     & 9.60       & 11.78      & 11.61      &  0.948       \\ 
-		                                  &{\bf runtime (s)}& <1         & 2         & 8       & 31        & 166       & 1,087     &  {\bf 8.856} \\
-
+		\multirow{2}{*}{\textbf{Shims-p}} & $f$              & 6.47        & 3.90        & 3.30         & 3.25         & 3.13         & 6.03        &   \\ 
+		                                  &{\bf run-time (s)}& 3           & 4           & 3            & 12           & 64           & 213         &   \\ %299s
 		\midrule[.1pt]
 
-		\multirow{2}{*}{\textbf{ACO-mp}}  & $f$             & 9.43       & {\bf 6.77}& 8.85      & 9.65       & 11.75      & 11.54      &  0.934 \\ 
-                                          &{\bf runtime (s)}& 3         & 4        & 17       & 86        & 513       & 3,596     &  2.714 \\ 
+		\multirow{2}{*}{\textbf{ACO}}     & $f$              & 6.19        & 3.63        & 3.21         &        &       &      &   \\ 
+                                          &{\bf run-time (s)}& 5           & 239         & 572          &        &       &      &   \\ 
         \midrule[.1pt]
 
-        \multirow{2}{*}{\textbf{Shims-mp}} & $f$             & 10.18       & {\bf 6.77} & 8.92     & 9.60       & 11.78      & 11.61      &  0.948       \\ 
-                                           &{\bf runtime (s)}& $< 1$         & 2         & 8       & 31        & 166       & 1,087     &  {\bf 8.856} \\
+        \multirow{2}{*}{\textbf{ACO-p}}   & $f$              & 6.04        & 4.23        & 3.61         &         &       &       &   \\ 
+                                          &{\bf run-time (s)}& 3           & 72          & 196          &         &       &       &   \\
 

### 2022-11-09 aco_p.py

@@ -62,7 +62,8 @@ def Solve( pallets, items, startTime, cfg, k, limit):  # items include kept on b
 
         aco.updatePheroAttract(Glocal.S, Gbest.S, antsField, aco.NANTS)
 
-    print(f"Used {numAnts} ants | ratio {Glocal.S/initialS:.3f} | {improvements} improvements")
+    if initialS > 0:
+        print(f"Used {numAnts} ants | ratio {Glocal.S/initialS:.3f} | {improvements} improvements")
 
     return mno.getSolMatrix(Gbest.Edges, numPallets, numItems)
 

### 2022-11-09 testing.py

@@ -7,22 +7,22 @@ import aco
 import aco_p
 import greedy
 
-# mno.DATA = "data20"
+mno.DATA = "data20"
 # mno.DATA = "data50"
 # mno.DATA = "data100"
 
-import sys
-scenario  =   int(sys.argv[1])
-method    =    f"{sys.argv[2]}"
-mno.NCPU  =   int(sys.argv[3])
-limit     = float(sys.argv[4])
-mno.DATA  =    f"{sys.argv[5]}"
+# import sys
+# scenario  =   int(sys.argv[1])
+# method    =    f"{sys.argv[2]}"
+# mno.NCPU  =   int(sys.argv[3])
+# limit     = float(sys.argv[4])
+# mno.DATA  =    f"{sys.argv[5]}"
 
-# scenario = 1
+scenario = 1
 
 # method = "Shims"
 # method = "Shims_p"
-# method = "ACO"
+method = "ACO"
 # method = "ACO_p"
 # method = "Greedy"
 
@@ -104,17 +104,21 @@ startNodeTime = time.perf_counter()
 if method == "Greedy":
     E = greedy.Solve(pallets, items, cfg, k)
 
+lim = 0.99 # for Shims only
+if scenario > 1:
+    lim = 0.85    
+
 if method == "Shims_p":
-    E = shims_p.Solve(pallets, items, cfg, k, limit)
+    E = shims_p.Solve(pallets, items, cfg, k, lim)
 
 if method == "Shims":
-    E = shims.Solve(pallets, items, cfg, k, limit)
+    E = shims.Solve(pallets, items, cfg, k, lim)
 
 if method == "ACO":
-    E =   aco.Solve( pallets, items, startNodeTime, cfg, k, limit)
+    E =   aco.Solve( pallets, items, startNodeTime, cfg, k, 0.95)
 
 if method == "ACO_p":
-    E = aco_p.Solve( pallets, items, startNodeTime, cfg, k, limit)
+    E = aco_p.Solve( pallets, items, startNodeTime, cfg, k, 0.95)
 
 elapsed = time.perf_counter() - startNodeTime
 

### 2022-11-10 latex/comp_or.tex
Experimentation initiated

@@
 
-\title{Parallel air palletization and tour planning for simultaneous pickup and delivery}
+\title{Multiprocessing air palletization and tour planning for simultaneous pickup and delivery}
 
@@
 These challenging benchmark instances are made publicly available.
-We also created a parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it was the unique method that managed to solve all testing scenarios.
+We also created a multiprocessing computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it was the unique method that managed to solve all testing scenarios.
 
@@
 
-Below, we present five implementations for $SolveNode$: the first using a MIP solver, and four with developed heuristics: {\it Ant Colony Optimization}\/ (ACO), {\it Parallel Ant Colony Optimization}\/ (ACO-p),  and {\it Shims},  and {\it Parallel Shims (Shims-p)}.
+Below, we present five implementations for $SolveNode$: the first using a MIP solver, and four with developed heuristics: {\it Ant Colony Optimization}\/ (ACO), {\it Multiprocessed Ant Colony Optimization}\/ (ACO-p),  and {\it Shims},  and {\it Multiprocessed Shims (Shims-p)}.
 
@@
 
-		\multirow{2}{*}{\textbf{Shims}}   & $f$              & 6.47        & 3.90        & 3.30         & 3.25         & 3.13         & 6.03        &   \\ 
-		                                  &{\bf run-time (s)}& <1          & 1           & 2            & 18           & 46           & 138         &   \\ %206s
+		\multirow{2}{*}{\textbf{Shims}}   & $f$              & 6.24        & 4.07        &          &          &          &         &   \\ 
+		                                  &{\bf run-time (s)}& <1          & 1           &             &            &            &          &   \\ %206s
 		\midrule[.1pt]
 
-		\multirow{2}{*}{\textbf{Shims-p}} & $f$              & 6.47        & 3.90        & 3.30         & 3.25         & 3.13         & 6.03        &   \\ 
-		                                  &{\bf run-time (s)}& 3           & 4           & 3            & 12           & 64           & 213         &   \\ %299s
+		\multirow{2}{*}{\textbf{Shims-p}} & $f$              & 6.59        & 3.95        &          &          &          &         &   \\ 
+		                                  &{\bf run-time (s)}& 2           & 12          &             &            &            &          &   \\ %299s
 		\midrule[.1pt]
 
-		\multirow{2}{*}{\textbf{ACO}}     & $f$              & 6.19        & 3.63        & 3.21         &        &       &      &   \\ 
-                                          &{\bf run-time (s)}& 5           & 239         & 572          &        &       &      &   \\ 
+		\multirow{2}{*}{\textbf{ACO}}     & $f$              & 6.05        &         &          &        &       &      &   \\ 
+                                          &{\bf run-time (s)}& 26          &          &           &        &       &      &   \\ 
         \midrule[.1pt]
 
-        \multirow{2}{*}{\textbf{ACO-p}}   & $f$              & 6.04        & 4.23        & 3.61         &         &       &       &   \\ 
-                                          &{\bf run-time (s)}& 3           & 72          & 196          &         &       &       &   \\
+        \multirow{2}{*}{\textbf{ACO-p}}   & $f$              & 5.90        &         &          &         &       &       &   \\ 
+                                          &{\bf run-time (s)}& 12          &           &           &         &       &       &   \\
 

### 2022-11-10 shims.py

@@ -86,6 +86,8 @@ def getBestShims(p, notInSol, sol, limit, numItems, maxTorque, k):
                 break
         else:
             sh = Shim(numItems) # create a new Shim
+            if sh.shimIsFeasible(oe, sol, maxTorque, k): # try to insert the edge "oe" in the new Shim
+                sh.AppendEdge(oe)
             Set.append(sh)
 
     # all Set of edges are feasible, but one has a better score
@@ -128,14 +130,10 @@ def Compute(edges, pallets, items, limit, cfg, k) :
     #edges not in sol groupped by pallets
     notInSol = [ [] for _ in range(len(pallets)) ]
 
-    for p in (pallets):
-        notInSol[p.ID] = [e for e in sol.Edges if e.Pallet.ID == p.ID and not e.InSol  ]
-
-	# pallets closer to the CG are completed first
-    pallets.sort(key=lambda x: abs(x.D), reverse=False)
-
 	# get the best shim that fills the slack
     for p in (pallets):
+
+        notInSol[p.ID] = [e for e in sol.Edges if e.Pallet.ID == p.ID and not e.InSol  ]
 
 		# get the best shim of edges             greedy limit
         BestShims = getBestShims(p, notInSol[p.ID], sol, limit, len(items), cfg.maxTorque, k)
@@ -152,6 +150,7 @@ def Compute(edges, pallets, items, limit, cfg, k) :
             if sol.isFeasible(ce, 1.0, cfg, k):
                 sol.putInSol(ce)
                 counter += 1
+
     print(f"{counter} edges included by the local search\n")
 
     return sol
@@ -165,6 +164,9 @@ def Solve(pallets, items, cfg, k, limit): # items include kept on board
     numPallets = len(pallets)
 
     edges = mno.mountEdges(pallets, items, cfg)
+
+	# pallets closer to the CG are completed first
+    pallets.sort(key=lambda x: abs(x.D), reverse=False)      
 
     sol = Compute(edges, pallets, items, limit, cfg, k)
 

### 2022-11-10 latex/comp_or.tex
Edges heuristic calculation changed

@@
 It is directly proportional to the score, and inversely proportional to the volume and the torque of each item.
+The 3000 value was empirically set to keep the attractiveness in the range [0, 1], to be compatible with the Ant Colony Optimization pheromone range.
 
 \begin{equation} \label{eq:edge}
-	\theta^k_{ij}= \frac{t^k_j.s}{t^k_j.v}\times(1-\frac{t^k_j.w\times|p_i.d|}{\max_s\{t^k_s.w\}\times\max_q\{|p_q.d|\}});\ i \in \{1,2,\ldots,m\},\ j \in \{1,2,\ldots,n_k\}
+	\theta^k_{ij}= \frac{t^k_j.s}{t^k_j.v \times 3000}\times(1-\frac{t^k_j.w\times|p_i.d|}{\max_s\{t^k_s.w\}\times\max_q\{|p_q.d|\}});\ i \in \{1,2,\ldots,m\},\ j \in \{1,2,\ldots,n_k\}
 \end{equation}
@@
 
-Below, we present five implementations for $SolveNode$: the first using a MIP solver, and four with developed heuristics: {\it Ant Colony Optimization}\/ (ACO), {\it Multiprocessed Ant Colony Optimization}\/ (ACO-p),  and {\it Shims},  and {\it Multiprocessed Shims (Shims-p)}.
+Below, we present five implementations for $SolveNode$: the first using a MIP solver, and four with developed heuristics: {\it Ant Colony Optimization}\/ (ACO), {\it Multiprocessed Ant Colony Optimization}\/ ({\it mpACO}),  and {\it Shims},  and {\it Multiprocessed Shims (mpShims)}.
+
+It is important to say that the parallel ACO ({\it pACO}) is a different implementation where employs threads instead of processes to solve each ant solution in parallel.
+We chose {\it mpACO} for our approach because of the need to share the pheromone array between ants/processes and the facility to handle the possibility of race condition.
 
@@
 
-In the tests performed, we used a 64-bit, 128GiB, 3.6GHz, 48-core processor, 2 threads per core, Intel(R) Xeon(R), with {\it Linux CentOS 8} as the operational system and {\it Python 3.10.4}\/ as the programming language.
-We also used the well-known solver {\it Gurobi}\/ (www.gurobi.com), version 9.5.2.
+All the experiments are performed on a 64-bit, 128GiB, 3.6GHz, 48-core processor, 2 threads per core, Intel Xeon, with {\it Linux CentOS 8} as the operational system and {\it Python 3.10.4}\/ as the programming language.
 
@@
 		\toprule
-		{$method$}                        & {\bf Results}    & {\bf 1}     &  {\bf 2}    & {\bf 3}      & {\bf 4}      & {\bf 5}      & {\bf 6}     & \specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
+		{$method$}                        &{\bf Results}    &{\bf 1}     &{\bf 2}    &{\bf 3}      &{\bf 4}      &{\bf 5}      &{\bf 6}     &\specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
 		\toprule
 
-		\multirow{2}{*}{\textbf{Shims}}   & $f$              & 6.24        & 4.07        &          &          &          &         &   \\ 
+		\multirow{2}{*}{\textbf{Shims}}   & $f$              & 5.40        & 2.94        &          &          &          &         &   \\ 
 		                                  &{\bf run-time (s)}& <1          & 1           &             &            &            &          &   \\ %206s
@@
 
-		\multirow{2}{*}{\textbf{Shims-p}} & $f$              & 6.59        & 3.95        &          &          &          &         &   \\ 
+		\multirow{2}{*}{\textbf{Shims-p}} & $f$              & 5.72        & 3.17        &          &          &          &         &   \\ 
 		                                  &{\bf run-time (s)}& 2           & 12          &             &            &            &          &   \\ %299s
@@
 
-		\multirow{2}{*}{\textbf{ACO}}     & $f$              & 6.05        &         &          &        &       &      &   \\ 
-                                          &{\bf run-time (s)}& 26          &          &           &        &       &      &   \\ 
+		\multirow{2}{*}{\textbf{ACO}}     & $f$              & 5.64        &         &          &        &       &      &   \\ 
+                                          &{\bf run-time (s)}& 23          &          &           &        &       &      &   \\ 
         \midrule[.1pt]
 
-        \multirow{2}{*}{\textbf{ACO-p}}   & $f$              & 5.90        &         &          &         &       &       &   \\ 
-                                          &{\bf run-time (s)}& 12          &           &           &         &       &       &   \\
+        \multirow{2}{*}{\textbf{ACO-p}}   & $f$              & {\bf 5.81}        &         &          &         &       &       &   \\ 
+                                          &{\bf run-time (s)}& 16          &           &           &         &       &       &   \\
 
@@
 
+\begin{figure}[H]
+	\centering
+	\includegraphics[scale=0.4]{Images/multithreads.png}
+	\caption{sdf sdf sdf sdf}
+	\small\textsuperscript{Source: sdfsdf}
+	\label{fig:multithreads}
+\end{figure}
+
 \section{Conclusions}

### 2022-11-10 aco.py

@@ -39,25 +39,26 @@ def getDeltaTau(score, bestSoFar, numAnts):
 # and edge attractiveness according to the its solution value
 def updatePheroAttract(score, bestSoFar, edges, numAnts):
 
-    # evaporate some pheromone from all edges
-    for id, e in enumerate(edges):
-        edges[id].Pheromone = math.sqrt(e.Pheromone) / 1.25 # RHO = 0.2
-        edges[id].updateAttract(ALPHA, BETA)
-
     deltaTau = getDeltaTau(score, bestSoFar, numAnts)
 
     maxAttract = 0.
     if deltaTau < 1:
         # update pheromone level in all edges
-        for id, _ in enumerate(edges):
+        for id, e in enumerate(edges):
+
+            #evaporate some pheromone 
+            edges[id].Pheromone = math.sqrt(e.Pheromone) / 1.25 # RHO = 0.2
+
             edges[id].Pheromone += deltaTau
             if edges[id].Pheromone < 0.:
                 edges[id].Pheromone = 0.
+
             edges[id].updateAttract(ALPHA, BETA)
 
             if edges[id].Attract > maxAttract:
                 maxAttract = edges[id].Attract
 
+        # normalize attractivity
         for id, _ in enumerate(edges):
             edges[id].Attract /= maxAttract
 

### 2022-11-10 testing.py

@@ -9,8 +9,8 @@ import greedy
 
 mno.NCPU = 8
 
-# mno.DATA = "data20"
-mno.DATA = "data50"
+mno.DATA = "data20"
+# mno.DATA = "data50"
 # mno.DATA = "data100"
 
 # import sys
@@ -26,7 +26,7 @@ scenario = 1
 # method = "Shims"
 # method = "Shims_p"
 # method = "ACO"
-method = "ACO_p"
+# method = "ACO_p"
 
 # data20 Shims best limit = 0.10 | 0.02s-0.04s
 # data20 ACO   best limit = 0.50

### 2022-11-10 latex/comp_or.tex
Paper updated

@@
 
-\title{Multiprocessing air palletization and tour planning for simultaneous pickup and delivery}
+\title{Parallel air palletization and tour planning for simultaneous pickup and delivery}
 
@@
 Also, there is no technological assistance to help the load and trip planners with the huge number of demands for transport in each hub.
-
 This enables other risks such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
 We contributed by modeling and solving a new problem of planning the loading and routing of an aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
-
 This new hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modeled using standardized pallets in fixed positions, obeying the center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
-
-We also contributed by carrying out multiple experiments with a commercial solver and the well-known Ant Colony Optimization on synthetic data based on real data from the {\it Brazilian Air Force}\/ transportation history.
-
+We also contributed by carrying out multiple experiments with the well-known Ant Colony Optimization on synthetic data based on real data from the {\it Brazilian Air Force}\/ transportation history.
 These challenging benchmark instances are made publicly available.
@@
 We could have used another criterion, but it was observed in the experiments that volume is more constrictive in airlift.
-Once the destinations of the pallets are defined, we will use heuristics and a MIP solver to find the best possible node-by-node solutions.
+Once the destinations of the pallets are defined, we will use serial and multiprocessing heuristics to find the best possible node-by-node solutions.
 This strategy is summarized in Algorithm \ref{alg:main}.
@@
 
-$method$\/ corresponds to a MIP solver or one of the heuristics that we will present in subsection \ref{methods}.
+$method$\/ corresponds to one of the heuristics that we will present in subsection \ref{methods}.
 The loop of lines \ref{main:loop1}-\ref{main:loop2} goes through all permutations $\pi$, where the node-by-node resolutions are performed by $SolveTour$, whose result is stored in $f_{\pi}$.
@@
 
-In addition to the set of nodes, pallets, costs and items, $SolveTour$, described in Algorithm \ref{alg:tour}, receives the parameter $method$, which corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and the parameter $\pi$, which is a permutation that defines the order of visits in this tour.
+In addition to the set of nodes, pallets, costs and items, $SolveTour$, described in Algorithm \ref{alg:tour}, receives the parameter $method$, which corresponds to a heuristic for solving the node-by-node problems, and the parameter $\pi$, which is a permutation that defines the order of visits in this tour.
 
@@
 
-In this subsection we present implementations of $SolveNode(method,k, G)$, where $method$\/ corresponds to a MIP solver or a heuristic, $k$\/ is the index of the current node and $G$\/ is the allocation graph of the consolidated items that remain on board at node $l_k$.
+In this subsection we present implementations of $SolveNode(method,k, G)$, where $method$\/ corresponds to a heuristic, $k$\/ is the index of the current node and $G$\/ is the allocation graph of the consolidated items that remain on board at node $l_k$.
 
@@
 
-		{\it Heuristic's procedures:}  & All heuristics use the same procedures for selecting and inserting edges.  \\
-
 		{\it Stopping criterion:} & All methods have the same time limit of $0.7$ seconds per node.\\
 
-		{\it Common procedures:}  & All methods have the same procedures for: solution printing, pallets destinations and initial parameters setting, and solution integrity checking.\\
+		{\it Common procedures:}  & All methods have the same procedures for: solution printing, pallets destinations and initial parameters setting, solution integrity checking, and the same procedures for selecting and inserting edges.\\
 
@@
 	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$;
-	\item $RandomChoice(set)$: randomly selects an element from $set$;
 	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
@@
 
-Below, we present five implementations for $SolveNode$: the first using a MIP solver, and four with developed heuristics: {\it Ant Colony Optimization}\/ (ACO), {\it Multiprocessed Ant Colony Optimization}\/ ({\it mpACO}),  and {\it Shims},  and {\it Multiprocessed Shims (mpShims)}.
+Below, we present four implementations for $SolveNode$: {\it Ant Colony Optimization}\/ (ACO), {\it Multiprocessed Ant Colony Optimization}\/ ({\it mpACO}),  and {\it Shims},  and {\it Multiprocessed Shims (mpShims)}.
 
@@
 We chose {\it mpACO} for our approach because of the need to share the pheromone array between ants/processes and the facility to handle the possibility of race condition.
-
-\subsubsection{MIP solver}
-\label{solver}
-
-To find a viable solution for ACLP+RPDP, the strategy adopted in $SolveTour$\/ previously defines the values of some variables.
-Concretely, all permutations between the nodes are tested, the set of nodes to be visited is updated, the consolidated items that remain on board are reallocated to minimize the CG deviation, and the pallets destination is determined according to the volume of items available for shipment.
-In this way, the mathematical modeling for $SolveNode(MIP,k,G)$\/ becomes simpler, which finds an allocation of available items in node $l_k$\/ using previously defined values of $L_k$, $p_i.to[k]$, $Y^k_{iq}$\/ and $a^k_q$.
-
-This simplified modelling is described in the equations \ref{eq:newf} to \ref{eq:lastnew}, which maximizes the score of the $t^k_j$\/ items to be shipped, maintaining load balancing.
-The variables $X^k_{ij}$\/ define the set of edges $E^N_k$, which will be added to the graph $G$.
-
-\begin{equation} \label{eq:newf}
-	\max f_s = \sum_{i=1}^{m} \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.s
-\end{equation}
-
-\begin{equation} 
-	maxW = \min(Payload, \sum_{i=1}^{m}p_i.w)
-\end{equation}
-
-\begin{equation} 
-	\epsilon_k = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w +  \sum_{q=1}^{m} Y_{iq}^k \times a_q^k.w)]/(maxW \times limit^{CG}_{long})
-\end{equation}
-
-\begin{equation}
-	X_{ij}^k = 0 \ \mbox{if} \ t_j^k.to \notin L_k; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}
-\end{equation}
-
-\begin{equation} 
-	LatIt_k = \sum_{i=1}^{m} \sum_{j=1}^{n_k} ( X_{ij}^k \times t_j^k.w \times (i\%2) - X_{ij}^k \times t_j^k.w \times (i+1)\%2 )
-\end{equation}
-
-\begin{equation} 
-	LatCons_k =  \sum_{i=1}^{m} \sum_{q=1}^{m_k}  ( Y_{iq}^k \times a_q^k.w \times (i\%2) - Y_{iq}^k \times a_q^k.w \times (i+1)\%2)
-\end{equation}
-
-\begin{equation} 
-	s.t.: d^{CG}_{pallet} \times | LatIt_k + LatCons_k | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}
-\end{equation}
-
-\begin{equation}
-	s.t.: |\tau_k| \leq maxW \times limit^{CG}_{long}
-\end{equation}
-
-\begin{equation} 
-	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w ) \leq maxW
-\end{equation}
-
-\begin{equation} 
-	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m_k\}
-\end{equation}
-
-\begin{equation} 
-	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.v + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m_k\}
-\end{equation}
-
-\begin{equation} 
-	s.t.: \sum_{i=1}^{m} X_{ij}^k \leq 1; \ j \in \{1, 2, \ldots, n_k\}
-\end{equation}
-
-\begin{equation} \label{eq:lastnew}
-	s.t.: p_i.to[k] = t^k_j.to\ \mbox{if} \ X_{ij}^k = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}
-\end{equation}
 
@@
 
-For parallelization, we used the {\it Python multiprocessing} package, which implements {\it process-based parallelism}.
-According to https://docs.python.org/3/library/multiprocessing.html, Multiprocessing is a package that supports spawning processes using an API similar to the threading module.
+For parallelization, we used the {\it Python multiprocessing} package, which implements {\it process-based parallelism or parallel shared memory algorithm}.
+According to \cite{multiprocessing}, Multiprocessing is a package that supports spawning processes using an API similar to the threading module.
+
 The multiprocessing package offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
-Because of this, the multiprocessing module lets the programmer take full advantage of the fact that a machine has more than one processor. This decision for multiprocessing relies on the need to exchange information among processes, which would not be possible with thread-based parallelism.
-
-Process: The operating system’s spawned and controlled entity that encapsulates an executing application.
-A process has two main functions.
-The first is to act as the resource holder for the application, and the second is to execute the instructions of the application.
-
-— PAGE 271, THE ART OF CONCURRENCY, 2009. The Art of Concurrency: A Thread Monkey's Guide to Writing Parallel Applications 1st Edition, Kindle Edition by Clay Breshears  (Author)
-
-parallel shared memory algorithms
+Because of this, the multiprocessing module lets the programmer take full advantage of the fact that a machine has more than one processor. This decision for multiprocessing relies on the need to exchange information among processes, which would be harder to handle with thread-based parallelism.
+
+According to \cite[p.271]{Breshears2009}, a Process is the operating system’s spawned and controlled entity that encapsulates an executing application.
+A process has two main functions: The first is to act as the resource holder for the application, and the second is to execute the instructions of the application.
 

### 2022-11-11 latex/comp_or.tex
SEC_BREAK to secBreak and NCPU to numCpu

@@
 This does not guarantee best results for all instances, but, on average, overall results are acceptable.
-Another important observation is that our heuristics were designed based on traditional ideas from the literature \cite{NiarFreville1997,Fidanova2006,Laabadi2018,Alonso2019,Zhan2020} and on the experiences learned during tests.
+Another important observation is that our heuristics were designed based on traditional ideas from the literature \cite{Fidanova2006} and on the experiences learned during tests.
 
@@
 
-		\multirow{2}{*}{\textbf{Shims-p}} & $f$              & 5.72        & 3.17        &          &          &          &         &   \\ 
-		                                  &{\bf run-time (s)}& 2           & 12          &             &            &            &          &   \\ %299s
+		\multirow{2}{*}{\textbf{Shims-p}} & $f$              & 5.88        & 3.16        &          &          &          &         &   \\ 
+		                                  &{\bf run-time (s)}& 1           & 10          &             &            &            &          &   \\ %299s
 		\midrule[.1pt]
 
-		\multirow{2}{*}{\textbf{ACO}}     & $f$              & 5.64        &         &          &        &       &      &   \\ 
-                                          &{\bf run-time (s)}& 23          &          &           &        &       &      &   \\ 
+		\multirow{2}{*}{\textbf{ACO}}     & $f$              & 5.77        &         &          &        &       &      &   \\ 
+                                          &{\bf run-time (s)}& 14          &          &           &        &       &      &   \\ 
         \midrule[.1pt]
 
-        \multirow{2}{*}{\textbf{ACO-p}}   & $f$              & {\bf 5.81}        &         &          &         &       &       &   \\ 
-                                          &{\bf run-time (s)}& 16          &           &           &         &       &       &   \\
+        \multirow{2}{*}{\textbf{ACO-p}}   & $f$              & 5.82        &         &          &         &       &       &   \\ 
+                                          &{\bf run-time (s)}& 22          &           &           &         &       &       &   \\
 

### 2022-11-11 aco.py

@@ -7,7 +7,7 @@ import random
 
 ALPHA = 1   # pheromone exponent (linearly affects attractivity)
 BETA  = 4   # heuristic exponent (exponentialy affects attractivity)
-NANTS = mno.NCPU
+NANTS = 6
 
 def rouletteSelection(values): # at least 15 times faster than randomChoice
     max = sum(values)
@@ -42,7 +42,7 @@ def updatePheroAttract(score, bestSoFar, edges, numAnts):
     deltaTau = getDeltaTau(score, bestSoFar, numAnts)
 
     maxAttract = 0.
-    if deltaTau < 1:
+    if abs(deltaTau) < 1:
         # update pheromone level in all edges
         for id, e in enumerate(edges):
 
@@ -71,9 +71,7 @@ def pickFromNbhood(nbhood, values):
     values.pop(i)
     return e
 
-def Solve( pallets, items, startTime, cfg, k, limit):  # items include kept on board
-
-    SEC_BREAK = NANTS
+def Solve( pallets, items, startTime, cfg, k, limit, secBreak):  # items include kept on board
 
     print("\nAnt Colony Optimization for ACLP+RPDP")
 
@@ -91,7 +89,7 @@ def Solve( pallets, items, startTime, cfg, k, limit):  # items include kept on b
     stagnant = 0
     improvements = 0
 
-    while stagnant <= 3:# and (time.perf_counter() - startTime) < SEC_BREAK:
+    while stagnant <= 3 and (time.perf_counter() - startTime) < secBreak:
 
         Glocal = mno.Solution(antsField, pallets, items, limit, cfg, k)
 
@@ -99,7 +97,7 @@ def Solve( pallets, items, startTime, cfg, k, limit):  # items include kept on b
 
         for _ in np.arange(NANTS):
 
-            if (time.perf_counter() - startTime) > SEC_BREAK:
+            if (time.perf_counter() - startTime) > secBreak:
                 break
 
             numAnts += 1

### 2022-11-11 aco_p.py

@@ -36,7 +36,7 @@ def antSolve(Gant, cfg, k, antsField, outLock, bestScore):
 def antQueue( Gant, cfg, k, antsField, outQueue, outLock, bestScore ):
     outQueue.put( antSolve(Gant, cfg, k, antsField, outLock, bestScore) )
 
-def Solve( pallets, items, startTime, cfg, k, limit):  # items include kept on board
+def Solve( pallets, items, startTime, cfg, k, limit, numCPU, secBreak):  # items include kept on board
 
     print("\nMultiprocess Ant Colony Optimization for ACLP+RPDP")
 
@@ -55,11 +55,11 @@ def Solve( pallets, items, startTime, cfg, k, limit):  # items include kept on b
     stagnant = 0
     improvements = 0
    
-    while stagnant < aco.NANTS and (time.perf_counter() - startTime) < mno.SEC_BREAK:    
+    while stagnant <= 3 and (time.perf_counter() - startTime) < secBreak:    
 
         Glocal = mno.Solution(antsField, pallets, items, limit, cfg, k)
 
-        procs = [None for _ in range(aco.NANTS)]
+        procs = [None for _ in range(numCPU)]
         outQueue = mp.Queue()
         outLock  = mp.Lock()
         for i, p in enumerate(procs):
@@ -69,10 +69,12 @@ def Solve( pallets, items, startTime, cfg, k, limit):  # items include kept on b
         for p in procs:
             numAnts+=1
             p.start()
-        sols = [outQueue.get() for _ in procs if (time.perf_counter() - startTime) < mno.SEC_BREAK]
+        sols = [outQueue.get() for _ in procs if (time.perf_counter() - startTime) < secBreak]
+
+        outQueue = None
 
         for Gant in sols:
-            if Gant.S > Glocal.S:
+            if Gant != None and Gant.S > Glocal.S:
                 Glocal = Gant
 
         if Glocal.S > Gbest.S:

### 2022-11-11 main.py

@@ -14,7 +14,7 @@ import aco_p
 import greedy
 
 
-def solveTour(scenario, instance, pi, tour, method, pallets, cfg):
+def solveTour(scenario, instance, pi, tour, method, pallets, cfg, numCPU, secBreak):
     """
     Solves one tour
     """
@@ -113,16 +113,16 @@ def solveTour(scenario, instance, pi, tour, method, pallets, cfg):
             E = greedy.Solve(pallets, items, cfg, k)
 
         if method == "Shims_p":
-            E = shims_p.Solve(pallets, items, cfg, k, minLim)
+            E = shims_p.Solve(pallets, items, cfg, k, minLim, numCPU)
 
         if method == "Shims":
             E = shims.Solve(pallets, items, cfg, k, limit)
 
         if method == "ACO":
-            E =   aco.Solve( pallets, items, startNodeTime, cfg, k, limit)
+            E =   aco.Solve( pallets, items, startNodeTime, cfg, k, limit, secBreak)
 
         if method == "ACO_p":
-            E = aco_p.Solve( pallets, items, startNodeTime, cfg, k, limit) 
+            E = aco_p.Solve( pallets, items, startNodeTime, cfg, k, limit, numCPU, secBreak) 
 
         nodeElapsed = time() - startNodeTime
 
@@ -198,14 +198,13 @@ if __name__ == "__main__":
     import sys
     scenario     =   int(sys.argv[1])
     method       =    f"{sys.argv[2]}"
-    methods.NCPU =   int(sys.argv[3])
-    # limit        = float(sys.argv[4])
+    numCPU       =   int(sys.argv[3])
     methods.DATA =    f"{sys.argv[4]}"
 
     # clear cache
     # find . | grep -E "(__pycache__|\.pyc|\.pyo$)" | xargs rm -rf
 
-    methods.SEC_BREAK = 1.5
+    secBreak = 0.7
 
     # methods.DATA = "data20"
     # methods.DATA = "data50"
@@ -271,7 +270,7 @@ if __name__ == "__main__":
         searchTime = 0
         for pi, tour in enumerate(tours):
 
-            broke = solveTour(scenario, instance, pi, tour, method, pallets, cfg) # writeTourSol is True or False
+            broke = solveTour(scenario, instance, pi, tour, method, pallets, cfg, numCPU, secBreak) # writeTourSol is True or False
 
             searchTime += tour.elapsed
 

### 2022-11-11 shims_p.py

@@ -35,7 +35,7 @@ def Compute(edges, pallets, items, limit, cfg, k) :
     return sol
 
 
-def Solve(pallets, items, cfg, k, minLim): # items include kept on board
+def Solve(pallets, items, cfg, k, minLim, numCPU): # items include kept on board
 
     print(f"\nParallel Shims for ACLP+RPDP")
 
@@ -47,12 +47,12 @@ def Solve(pallets, items, cfg, k, minLim): # items include kept on board
 	# pallets closer to the CG are completed first
     pallets.sort(key=lambda x: abs(x.D), reverse=False)      
 
-    procs = [None for _ in range(mno.NCPU)]
+    procs = [None for _ in range(numCPU)]
     outQueue = mp.Queue()
 
     for i, p in enumerate(procs):
 
-        limit = minLim + (1.0 - minLim)*(i+1)/mno.NCPU
+        limit = minLim + (1.0 - minLim)*(i+1)/numCPU
 
         # create a child process for each limit
         procs[i] = mp.Process( target=shimsQueue,args=( edges, pallets, items, limit, cfg, k, outQueue  ) )

### 2022-11-11 latex/comp_or.tex
New table for processes performance

@@
 \newcommand{\Comentar}[1]{\State {\cmmt{#1}}}
-\newcommand{\Break}{\State \textbf{break}}
-\renewcommand{\Return}{\State \textbf{return}~}
-\renewcommand{\algorithmicensure}{\textbf{Parameters:}}
+\newcommand{\Break}{\State \bf {break}}
+\renewcommand{\Return}{\State \bf {return}~}
+\renewcommand{\algorithmicensure}{\bf {Parameters:}}
 
@@
 		\toprule
-		\textbf{Limits}  & \multicolumn{3}{c}{$Payload$: $26,000kg$} & \multicolumn{4}{c}{$limit^{CG}_{long}$: $0.556m$} \\
-		\midrule
-		\bf{Pallets} & $p_7$ & $p_6$ & $p_5$ & $p_4$ & $p_3$ & $p_2$ & $p_1$ \\
+		\bf {Limits}  & \multicolumn{3}{c}{$Payload$: $26,000kg$} & \multicolumn{4}{c}{$limit^{CG}_{long}$: $0.556m$} \\
+		\midrule
+		\bf {Pallets} & $p_7$ & $p_6$ & $p_5$ & $p_4$ & $p_3$ & $p_2$ & $p_1$ \\
 		\midrule
@@
 
-		\textbf{Costs}  & \multicolumn{4}{c}{$c_d$: US\$ $1.100/km$ } &	\multicolumn{3}{c}{$c_g = 0.05$} \\
+		\bf {Costs}  & \multicolumn{4}{c}{$c_d$: US\$ $1.100/km$ } &	\multicolumn{3}{c}{$c_g = 0.05$} \\
 
@@
 		\toprule
-		\textbf{Limits}& \multicolumn{3}{c}{$Payload$: $75,000kg$} & \multicolumn{3}{c}{$limit^{CG}_{long}$: $1.170m$} &
+		\bf {Limits}& \multicolumn{3}{c}{$Payload$: $75,000kg$} & \multicolumn{3}{c}{$limit^{CG}_{long}$: $1.170m$} &
 		\multicolumn{3}{c}{$limit^{CG}_{lat}$: $0.19m$} \\
 		\midrule
-		\multirow{2}{*}{\textbf{Pallets}}  & $p_{17}$ & $p_{15}$ & $p_{13}$ & $p_{11}$ & $p_{9}$ & $p_{7}$ & $p_{5}$ & $p_{3}$ & $p_{1}$ \\
+		\multirow{2}{*}{\bf {Pallets}}  & $p_{17}$ & $p_{15}$ & $p_{13}$ & $p_{11}$ & $p_{9}$ & $p_{7}$ & $p_{5}$ & $p_{3}$ & $p_{1}$ \\
 		& $p_{18}$ & $p_{16}$ & $p_{14}$ & $p_{12}$ & $p_{10}$ & $p_{8}$ & $p_{6}$ & $p_{4}$ & $p_{2}$ \\
 		\midrule 
-		\multirow{2}{*}{\textbf{CGx ($m$)}} & -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\
+		\multirow{2}{*}{\bf {CGx ($m$)}} & -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\
 		& -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\			
 		\midrule 
-		\multirow{2}{*}{\textbf{CGy ($m$)}}  & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 \\
+		\multirow{2}{*}{\bf {CGy ($m$)}}  & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 \\
 		& -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 \\	
@@
 
-		\textbf{Costs}  & \multicolumn{5}{c}{ $c_d$: US\$ $4.900/km$ } &	\multicolumn{4}{c}{$c_g = 0.05$} \\
+		\bf {Costs}  & \multicolumn{5}{c}{ $c_d$: US\$ $4.900/km$ } &	\multicolumn{4}{c}{$c_g = 0.05$} \\
 
@@
 
-For parallelization, we used the {\it Python multiprocessing} package, which implements {\it process-based parallelism or parallel shared memory algorithm}.
-According to \cite{multiprocessing}, Multiprocessing is a package that supports spawning processes using an API similar to the threading module.
-
-The multiprocessing package offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
-Because of this, the multiprocessing module lets the programmer take full advantage of the fact that a machine has more than one processor. This decision for multiprocessing relies on the need to exchange information among processes, which would be harder to handle with thread-based parallelism.
+For parallelization, we used the {\it Python multiprocessing} package, which implements {\it process-based parallelism} or {\it parallel shared memory algorithm}.
+According to \cite{multiprocessing}, {\it Multiprocessing} is a package that supports spawning processes using an API similar to the {\it Python Threading} package.
+
+The {\it Multiprocessing} package (mp) offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
+Because of this, the {\it Multiprocessing} module lets the programmer take full advantage of the fact that a machine has more than one processor, 48 in our computer. This decision for {\it Multiprocessing} relies on the need to exchange information among processes (the pheromone trails or some information regarding load balancing, in our case), which would be harder to handle with thread-based parallelism.
 
@@
 	\centering
-	\caption{Methods and scenarios with $volume=1.2$}  \label{tab:20}
-	\footnotesize
-	\begin{tabular}{l|c|cccccc|c}
+	\caption{Performance for different number of concurrent processes with $1.2 \times volume$}  \label{tab:procs}
+%	\footnotesize
+	\scriptsize
+	\begin{tabular}{c| c | c | c c c c c c }
 	\toprule
-		{$method$}                        &{\bf Results}    &{\bf 1}     &{\bf 2}    &{\bf 3}      &{\bf 4}      &{\bf 5}      &{\bf 6}     &\specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
+	\textbf{Scenario} & $method$ &{\bf Results}     & \textbf{1} & \textbf{4} & \textbf{8} & \textbf{16} & \textbf{24} & \textbf{32}  \\
 	\toprule
-
-		\multirow{2}{*}{\textbf{Shims}}   & $f$              & 5.40        & 2.94        &          &          &          &         &   \\ 
-		                                  &{\bf run-time (s)}& <1          & 1           &             &            &            &          &   \\ %206s
+	\multirow{4}{*}{\bf {1}}  &\multirow{2}{*}{\it {mpShims}}  &$f$               & 1 & 4 & 8 & 16 & 24 & 32 \\
+	                          &                                &{\bf run-time (s)}& 1 & 4 & 8 & 16 & 24 & 32 \\
+                              \cmidrule{2-9}
+	                          &\multirow{2}{*}{\it {mpACO}}    &$f$               & 1 & 4 & 8 & 16 & 24 & 32 \\
+	                          &                                &{\bf run-time (s)}& 1 & 4 & 8 & 16 & 24 & 32 \\
+	\cmidrule{1-9}
+	\multirow{4}{*}{\bf {2}}  &\multirow{2}{*}{\it {mpShims}} &$f$               & 1 & 4 & 8 & 16 & 24 & 32 \\
+	                          &   &{\bf run-time (s)}& 1 & 4 & 8 & 16 & 24 & 32 \\
+                              \cmidrule{2-9}
+	                          &\multirow{2}{*}{\it {mpACO}}   &$f$               & 1 & 4 & 8 & 16 & 24 & 32 \\
+	                          &   &{\bf run-time (s)}& 1 & 4 & 8 & 16 & 24 & 32 \\				
+	\bottomrule		
+\end{tabular}
+
+\normalsize
+
+\end{table}
+
+\vspace{2.0mm}
+
+\begin{table}[H]
+	\centering
+	\caption{Methods and scenarios with $1.2 \times volume$}  \label{tab:20}
+%	\footnotesize
+\scriptsize
+	\begin{tabular}{r|c|c|cccccc|c}
+		\toprule
+		{$method$}                     &{\bf Procs}              &{\bf Results}     &{\bf 1}     &{\bf 2}    &{\bf 3}      &{\bf 4}      &{\bf 5}      &{\bf 6}     &\specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
+		\toprule
+
+		\multirow{2}{*}{\it {Shims}}   &\multirow{2}{*}{\bf {1}} & $f$              & 5.40        & 2.94    & 2.93         &          &          &         &   \\ 
+		                               &                         &{\bf run-time (s)}& <1          & 1       & 4            &            &            &          &   \\ %206s
 		\midrule[.1pt]
 
-		\multirow{2}{*}{\textbf{Shims-p}} & $f$              & 5.88        & 3.16        &          &          &          &         &   \\ 
-		                                  &{\bf run-time (s)}& 1           & 10          &             &            &            &          &   \\ %299s
+		\multirow{2}{*}{\it {mpShims}} &\multirow{2}{*}{\bf {24}}& $f$              & 5.88        & 3.16    & 3.16         &          &          &         &   \\ 
+		                               &                         &{\bf run-time (s)}& 1           & 10      & 41           &            &            &          &   \\ %299s
 		\midrule[.1pt]
 
-		\multirow{2}{*}{\textbf{ACO}}     & $f$              & 5.77        &         &          &        &       &      &   \\ 
-                                          &{\bf run-time (s)}& 14          &          &           &        &       &      &   \\ 
+		\multirow{2}{*}{\it {ACO}}     &\multirow{2}{*}{\bf {1}} & $f$              & 5.77        & 3.49    &          &        &       &      &   \\ 
+                                       &                         &{\bf run-time (s)}& 14          & 808     &           &        &       &      &   \\ 
         \midrule[.1pt]
 
-        \multirow{2}{*}{\textbf{ACO-p}}   & $f$              & 5.82        &         &          &         &       &       &   \\ 
-                                          &{\bf run-time (s)}& 22          &           &           &         &       &       &   \\
+        \multirow{2}{*}{\it {mpACO}}   &\multirow{2}{*}{\bf {32}}& $f$              & 5.82        & 3.12    &          &         &       &       &   \\ 
+                                       &                         &{\bf run-time (s)}& 22          & 473     &           &         &       &       &   \\
 

### 2022-11-12 latex/comp_or.tex
Performance plots added

@@
 \pgfplotsset{compat=newest}
+\usepgfplotslibrary{units}
+\usepackage{tikz}
+
 \usepackage{algorithm} % sudo apt-get install -y texlive-science
@@
 	\scriptsize
-	\begin{tabular}{c| c | c | c c c c c c }
+	\begin{tabular}{c| c | c | c c c c c c c c }
 	\toprule
-	\textbf{Scenario} & $method$ &{\bf Results}     & \textbf{1} & \textbf{4} & \textbf{8} & \textbf{16} & \textbf{24} & \textbf{32}  \\
+	{\it Scenario} & $method$ &{\bf numProcs}    & \textbf{1} & \textbf{4} & \textbf{8} & \textbf{16} & \textbf{24} & \textbf{32} & \textbf{40} & \textbf{48}  \\
 	\toprule
-	\multirow{4}{*}{\bf {1}}  &\multirow{2}{*}{\it {mpShims}}  &$f$               & 1 & 4 & 8 & 16 & 24 & 32 \\
-	                          &                                &{\bf run-time (s)}& 1 & 4 & 8 & 16 & 24 & 32 \\
-                              \cmidrule{2-9}
-	                          &\multirow{2}{*}{\it {mpACO}}    &$f$               & 1 & 4 & 8 & 16 & 24 & 32 \\
-	                          &                                &{\bf run-time (s)}& 1 & 4 & 8 & 16 & 24 & 32 \\
-	\cmidrule{1-9}
-	\multirow{4}{*}{\bf {2}}  &\multirow{2}{*}{\it {mpShims}} &$f$               & 1 & 4 & 8 & 16 & 24 & 32 \\
-	                          &   &{\bf run-time (s)}& 1 & 4 & 8 & 16 & 24 & 32 \\
-                              \cmidrule{2-9}
-	                          &\multirow{2}{*}{\it {mpACO}}   &$f$               & 1 & 4 & 8 & 16 & 24 & 32 \\
-	                          &   &{\bf run-time (s)}& 1 & 4 & 8 & 16 & 24 & 32 \\				
+	\multirow{4}{*}{\it {1}}  &\multirow{2}{*}{\it {mpShims}}  &{\bf scores}      & 10909 & 10936 & 11022 & 11003 & 11033 & 11021 & 11013 & 11033 \\
+	                          &                                &{\bf run-time (s)}& 0.12 & 0.23 & 0.44 & 0.65 & 1.13 & 1.74 & 2.32 & 3.68  \\
+                              \cmidrule{2-11}
+	                          &\multirow{2}{*}{\it {mpACO}}    &{\bf scores}               & 4 & 4 & 8 & 16 & 24 & 32&& \\
+	                          &                                &{\bf run-time (s)}& 4 & 4 & 8 & 16 & 24 & 32&& \\
+	\cmidrule{1-11}
+	\multirow{4}{*}{\it {2}}  &\multirow{2}{*}{\it {mpShims}} &{\bf scores}       & 40016 & 40071 & 40071 & 40078 & 40078 & 40073 & 40083 & 40083 \\
+	                          &                               &{\bf run-time (s)} & 2.72 & 2.85 & 2.15 & 1.95 & 3.10 & 20.32 & 13.95 & 17.24  \\
+                              \cmidrule{2-11}
+	                          &\multirow{2}{*}{\it {mpACO}}   &{\bf scores}       & 4 & 4 & 8 & 16 & 24 & 32&& \\
+	                          &   &{\bf run-time (s)}         & 4 & 4 & 8 & 16 & 24 & 32&& \\				
 	\bottomrule		
@@
 \end{table}
+
+%np  time    score
+
+\begin{figure}[h!]
+	\begin{center}
+		\begin{tikzpicture}[scale=0.8, samples=100]
+			\begin{axis}[
+				%width=\linewidth, % Scale the plot to \linewidth
+				grid=major, 
+				grid style={dashed,gray!30},
+				xlabel={\bf Scenario 1 computation times (s)},
+				ylabel={\bf Scores},
+				nodes near coords,
+				enlargelimits=0.2,
+				legend pos=south east,
+				y tick label style={
+						/pgf/number format/.cd,
+						fixed,
+						fixed zerofill,
+						precision=3,
+						/tikz/.cd,
+					},				
+				]
+
+				\addplot +[
+					only marks,
+					point meta=explicit symbolic,
+					nodes near coords = {$\pgfmathprintnumber[frac]\np$},
+					visualization depends on={\thisrow{np} \as \np},
+					font=\small,
+				]
+				table[x=time, y=score, col sep=comma] {mpShims1.csv};
+
+				\legend{\it parallel processes}
+			\end{axis}
+		\end{tikzpicture}%
+~%
+%
+		\begin{tikzpicture}[scale=0.8, samples=100]
+			\begin{axis}[
+				%width=\linewidth, % Scale the plot to \linewidth
+				grid=major, 
+				grid style={dashed,gray!30},
+				xlabel={\bf Scenario 2 computation times (s)},
+				ylabel={\bf Scores},
+				nodes near coords,
+				enlargelimits=0.2,
+				legend pos=south east,
+				y tick label style={
+						/pgf/number format/.cd,
+						fixed,
+						fixed zerofill,
+						precision=3,
+						/tikz/.cd,
+					}
+				]
+
+				\addplot +[
+					only marks,
+					point meta=explicit symbolic,
+					nodes near coords = {$\pgfmathprintnumber[frac]\np$},
+					visualization depends on={\thisrow{np} \as \np},
+					font=\small,
+				]
+				table[x=time, y=score, col sep=comma] {mpShims2.csv};
+
+				\legend{\it parallel processes}
+			\end{axis}
+		\end{tikzpicture}
+
+		\caption{{\it mpShims} performance on {\it Scenarios 1 and 2}.}
+	\end{center}
+
+\end{figure}\label{fig:mpShims}
+
+Observing Figure \ref{fig:mpShims}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is 8. For {\it Scenario 2}, the best number is 16.
 
@@
 
-		\multirow{2}{*}{\it {ACO}}     &\multirow{2}{*}{\bf {1}} & $f$              & 5.77        & 3.49    &          &        &       &      &   \\ 
-                                       &                         &{\bf run-time (s)}& 14          & 808     &           &        &       &      &   \\ 
+		\multirow{2}{*}{\it {ACO}}     &\multirow{2}{*}{\bf {1}} & $f$              & 5.77        & 3.49    & 3.02         &        &       &      &   \\ 
+                                       &                         &{\bf run-time (s)}& 14          & 808     & 2398          &        &       &      &   \\ 
         \midrule[.1pt]
@@
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=0.4]{Images/multithreads.png}
-	\caption{sdf sdf sdf sdf}
-	\small\textsuperscript{Source: sdfsdf}
-	\label{fig:multithreads}
-\end{figure}
-
 \section{Conclusions}

### 2022-11-12 latex/comp_or.tex
All plots working

@@
 
-\usepackage{tikz}
-\usetikzlibrary{calc}
-\usetikzlibrary{positioning}
-
 \usepackage{mathtools}
@@
 \captionsetup{skip=0pt} % vertical distance from the table or figure
-
-\usepackage{pgfplots}
-\pgfplotsset{compat=newest}
-\usepgfplotslibrary{units}
-\usepackage{tikz}
 
@@
 	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+
+\usepackage{tikz}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+
+\usepackage{pgfplots}
+
+\pgfplotsset{
+	compat=1.9,
+	myplotstyle/.style={
+		legend cell align=left,
+		every axis plot/.append style={thick},
+		legend style={font=\small,draw=none},
+		enlargelimits=0.2,
+		legend pos=south east,
+		y tick label style={ /pgf/number format/.cd, fixed, fixed zerofill, precision=3, /tikz/.cd, },
+		grid=major,		
+	},
+}
 
@@
 
-%np  time    score
-
 \begin{figure}[h!]
@@
 			\begin{axis}[
-				%width=\linewidth, % Scale the plot to \linewidth
-				grid=major, 
-				grid style={dashed,gray!30},
+				myplotstyle,
 				xlabel={\bf Scenario 1 computation times (s)},
 				ylabel={\bf Scores},
-				nodes near coords,
-				enlargelimits=0.2,
-				legend pos=south east,
-				y tick label style={
-						/pgf/number format/.cd,
-						fixed,
-						fixed zerofill,
-						precision=3,
-						/tikz/.cd,
-					},				
 				]
-
-				\addplot +[
+				\addplot+[
 				only marks,	
-					point meta=explicit symbolic,
-					nodes near coords = {$\pgfmathprintnumber[frac]\np$},
-					visualization depends on={\thisrow{np} \as \np},
-					font=\small,
+				nodes near coords = {\numProcs{}},
+				visualization depends on = {value \thisrow{np} \as \numProcs},						
 				]
-				table[x=time, y=score, col sep=comma] {mpShims1.csv};
+				table[x=time, y=score, col sep=comma] {mpShims1data20.csv};
 
@@
 			\begin{axis}[
-				%width=\linewidth, % Scale the plot to \linewidth
-				grid=major, 
-				grid style={dashed,gray!30},
+				myplotstyle,
 				xlabel={\bf Scenario 2 computation times (s)},
 				ylabel={\bf Scores},
-				nodes near coords,
-				enlargelimits=0.2,
-				legend pos=south east,
-				y tick label style={
-						/pgf/number format/.cd,
-						fixed,
-						fixed zerofill,
-						precision=3,
-						/tikz/.cd,
-					}
 				]
-
-				\addplot +[
+				\addplot+[
 				only marks,	
-					point meta=explicit symbolic,
-					nodes near coords = {$\pgfmathprintnumber[frac]\np$},
-					visualization depends on={\thisrow{np} \as \np},
-					font=\small,
+				nodes near coords = {\numProcs{}},
+				visualization depends on = {value \thisrow{np} \as \numProcs},						
 				]
-				table[x=time, y=score, col sep=comma] {mpShims2.csv};
+				table[x=time, y=score, col sep=comma] {mpShims2data20.csv};
 
@@
 
-		\caption{{\it mpShims} performance on {\it Scenarios 1 and 2}.}
+		\caption{ {\it mpShims} performance with $1.2 \times volume$. }
 	\end{center}
-
-\end{figure}\label{fig:mpShims}
-
-Observing Figure \ref{fig:mpShims}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is 8. For {\it Scenario 2}, the best number is 16.
+	\label{fig:mpShimsData20}
+\end{figure}
+
+Observing Figure \ref{fig:mpShimsData20}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is 8. For {\it Scenario 2}, the best number is 16.
+
+\begin{figure}[h!]
+	\begin{center}
+		\begin{tikzpicture}[scale=0.8, samples=100]
+			\begin{axis}[
+				myplotstyle,
+				xlabel={\bf Scenario 1 computation times (s)},
+				ylabel={\bf Scores},
+				]
+				\addplot+[
+				only marks,	
+				nodes near coords = {\numProcs{}},
+				visualization depends on = {value \thisrow{np} \as \numProcs},						
+				]
+				table[x=time, y=score, col sep=comma] {mpShims1data50.csv};
+
+				\legend{\it parallel processes}
+			\end{axis}
+		\end{tikzpicture}%
+		~%
+		%
+		\begin{tikzpicture}[scale=0.8, samples=100]
+			\begin{axis}[
+				myplotstyle,
+				xlabel={\bf Scenario 2 computation times (s)},
+				ylabel={\bf Scores},
+				]
+				\addplot+[
+				only marks,	
+				nodes near coords = {\numProcs{}},
+				visualization depends on = {value \thisrow{np} \as \numProcs},						
+				]
+				table[x=time, y=score, col sep=comma] {mpShims2data50.csv};
+
+				\legend{\it parallel processes}
+			\end{axis}
+		\end{tikzpicture}
+
+		\caption{{\it mpShims} performance with $1.5 \times volume$.}
+	\end{center}
+	\label{fig:mpShimsData50}
+\end{figure}
+
+Observing Figure \ref{fig:mpShimsData50}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is 8. For {\it Scenario 2}, the best number is 16.
+
+\begin{figure}[h!]
+
+	\begin{center}
+		\begin{tikzpicture}[scale=0.8, samples=100]
+			\begin{axis}[
+				myplotstyle,
+				xlabel={\bf Scenario 1 computation times (s)},
+				ylabel={\bf Scores},
+				]
+				\addplot+[
+				only marks,	
+				nodes near coords = {\numProcs{}},
+				visualization depends on = {value \thisrow{np} \as \numProcs},						
+				]
+				table[x=time, y=score, col sep=comma] {Shims_p1data100.csv};
+
+				\legend{\it parallel processes}
+			\end{axis}
+		\end{tikzpicture}%
+		~%
+		%
+		\begin{tikzpicture}[scale=0.8, samples=100]
+			\begin{axis}[
+				myplotstyle,
+				xlabel={\bf Scenario 2 computation times (s)},
+				ylabel={\bf Scores},
+				]
+				\addplot+[
+				only marks,	
+				nodes near coords = {\numProcs{}},
+				visualization depends on = {value \thisrow{np} \as \numProcs},						
+				]
+				table[x=time, y=score, col sep=comma] {Shims_p2data100.csv};
+
+				\legend{\it parallel processes}
+			\end{axis}
+		\end{tikzpicture}
+
+		\caption{{\it mpShims} performance with $2.0 \times volume$.}
+	\end{center}
+	\label{fig:mpShimsData100}
+\end{figure}
+
+Observing Figure \ref{fig:mpShimsData100}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is 8.
+For {\it Scenario 2}, the best number is 16.
 

### 2022-11-12 testing.py

@@ -218,12 +218,37 @@ while tries:
 print(f"average elapsed = {totElapsed/(numTries):.2f} | average score = {sumScores/numTries:.0f}")
 
 if method == "Shims_p" or method == "ACO_p":
-    timesLatex = ""
-    for t in times:
-        timesLatex += f"& {t:.2f} " 
-    print(timesLatex)
+    # timesLatex = ""
+    # for t in times:
+    #     timesLatex += f"& {t:.2f} " 
+    # print(timesLatex)
 
-    scoresLatex = ""
-    for s in scores:
-        scoresLatex += f"& {s:.0f} " 
-    print(scoresLatex)
+    # scoresLatex = ""
+    # for s in scores:
+    #     scoresLatex += f"& {s:.0f} " 
+    # print(scoresLatex)
+
+    text = "np,time,score\n"
+    for i, _ in enumerate(numCpus):
+        text += f"{numCpus[i]},{times[i]:.2f},{scores[i]:.0f}\n"
+
+    fname = f"./latex/{method}{scenario}{mno.DATA}.csv"
+
+    writer = open(fname,'w+') # + creates the file, if not exists
+
+    try:
+        writer.write(text)
+    finally:
+        writer.close() 
+
+    print(text)
+
+# np,  time,    score
+# 1,	0.41,	50914
+# 4,	0.77,	50937
+# 8,	1.21,	50916
+# 16,	2.16,	50929
+# 24,	3.05,	50951
+# 32,	4.41,	50941
+# 40,	4.52,	50944
+# 48,	6.15,	50934

### 2022-11-13 latex/comp_or.tex
2 - abs(epsilon) removed

@@
 
-\vspace{2.0mm}
-
-\begin{table}[H]
-	\centering
-	\caption{Performance for different number of concurrent processes with $1.2 \times volume$}  \label{tab:procs}
-%	\footnotesize
-	\scriptsize
-	\begin{tabular}{c| c | c | c c c c c c c c }
-	\toprule
-	{\it Scenario} & $method$ &{\bf numProcs}    & \textbf{1} & \textbf{4} & \textbf{8} & \textbf{16} & \textbf{24} & \textbf{32} & \textbf{40} & \textbf{48}  \\
-	\toprule
-	\multirow{4}{*}{\it {1}}  &\multirow{2}{*}{\it {mpShims}}  &{\bf scores}      & 10909 & 10936 & 11022 & 11003 & 11033 & 11021 & 11013 & 11033 \\
-	                          &                                &{\bf run-time (s)}& 0.12 & 0.23 & 0.44 & 0.65 & 1.13 & 1.74 & 2.32 & 3.68  \\
-                              \cmidrule{2-11}
-	                          &\multirow{2}{*}{\it {mpACO}}    &{\bf scores}               & 4 & 4 & 8 & 16 & 24 & 32&& \\
-	                          &                                &{\bf run-time (s)}& 4 & 4 & 8 & 16 & 24 & 32&& \\
-	\cmidrule{1-11}
-	\multirow{4}{*}{\it {2}}  &\multirow{2}{*}{\it {mpShims}} &{\bf scores}       & 40016 & 40071 & 40071 & 40078 & 40078 & 40073 & 40083 & 40083 \\
-	                          &                               &{\bf run-time (s)} & 2.72 & 2.85 & 2.15 & 1.95 & 3.10 & 20.32 & 13.95 & 17.24  \\
-                              \cmidrule{2-11}
-	                          &\multirow{2}{*}{\it {mpACO}}   &{\bf scores}       & 4 & 4 & 8 & 16 & 24 & 32&& \\
-	                          &   &{\bf run-time (s)}         & 4 & 4 & 8 & 16 & 24 & 32&& \\				
-	\bottomrule		
-\end{tabular}
-
-\normalsize
-
-\end{table}
-
-\begin{figure}[h!]
+\begin{figure}[H]
 	\begin{center}
@@
 				xlabel={\bf Scenario 1 computation times (s)},
-				ylabel={\bf Scores},
+				ylabel={\bf Score},
 				]
@@
 				xlabel={\bf Scenario 2 computation times (s)},
-				ylabel={\bf Scores},
 				]
@@
 
+	\end{center}
 	\caption{ {\it mpShims} performance with $1.2 \times volume$. }
-	\end{center}
-	\label{fig:mpShimsData20}
+	\label{fig:mpShims2data20}
 \end{figure}
 
-Observing Figure \ref{fig:mpShimsData20}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is 8. For {\it Scenario 2}, the best number is 16.
-
-\begin{figure}[h!]
+Observing Figure \ref{fig:mpShims2data20}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is \textbf{1}.
+For {\it Scenario 2}, the best number is \textbf{4}.
+
+\begin{figure}[H]
 	\begin{center}
@@
 
+	\end{center}
 	\caption{{\it mpShims} performance with $1.5 \times volume$.}
-	\end{center}
-	\label{fig:mpShimsData50}
+	\label{fig:mpShims2data50}
 \end{figure}
 
-Observing Figure \ref{fig:mpShimsData50}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is 8. For {\it Scenario 2}, the best number is 16.
-
-\begin{figure}[h!]
+Observing Figure \ref{fig:mpShims2data50}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{1}. For {\it Scenario 2}, the best number is \textbf{8}.
+
+\begin{figure}[H]
 
@@
 				]
-				table[x=time, y=score, col sep=comma] {Shims_p1data100.csv};
+				table[x=time, y=score, col sep=comma] {mpACO1data20.csv};
 
@@
 				]
-				table[x=time, y=score, col sep=comma] {Shims_p2data100.csv};
+				table[x=time, y=score, col sep=comma] {mpACO2data20.csv};
 
@@
 
-		\caption{{\it mpShims} performance with $2.0 \times volume$.}
 	\end{center}
-	\label{fig:mpShimsData100}
+	\caption{{\it mpACO} performance with $1.2 \times volume$.}
+	\label{fig:mpACO2data20}
 \end{figure}
 
-Observing Figure \ref{fig:mpShimsData100}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is 8.
-For {\it Scenario 2}, the best number is 16.
+Observing Figure \ref{fig:mpACO2data20}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{40}. For {\it Scenario 2}, the best number is \textbf{1}.
+
+\begin{figure}[H]
+
+	\begin{center}
+		\begin{tikzpicture}[scale=0.8, samples=100]
+			\begin{axis}[
+				myplotstyle,
+				xlabel={\bf Scenario 1 computation times (s)},
+				ylabel={\bf Score},
+				]
+				\addplot+[
+				only marks,	
+				nodes near coords = {\numProcs{}},
+				visualization depends on = {value \thisrow{np} \as \numProcs},						
+				]
+				table[x=time, y=score, col sep=comma] {mpACO1data50.csv};
+
+				\legend{\it parallel processes}
+			\end{axis}
+		\end{tikzpicture}%
+		~%
+		%
+		\begin{tikzpicture}[scale=0.8, samples=100]
+			\begin{axis}[
+				myplotstyle,
+				xlabel={\bf Scenario 2 computation times (s)},
+				]
+				\addplot+[
+				only marks,	
+				nodes near coords = {\numProcs{}},
+				visualization depends on = {value \thisrow{np} \as \numProcs},						
+				]
+				table[x=time, y=score, col sep=comma] {mpACO2data50.csv};
+
+				\legend{\it parallel processes}
+			\end{axis}
+		\end{tikzpicture}
+
+	\end{center}
+	\caption{{\it mpACO} performance with $1.5 \times volume$.}
+	\label{fig:mpACO2data50}
+\end{figure}
+
+Observing Figure \ref{fig:mpACO2data50}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{1}. For {\it Scenario 2}, the best number is \textbf{16}.
 

### 2022-11-13 aco_p.py

@@ -69,9 +69,10 @@ def Solve( pallets, items, startTime, cfg, k, limit, numCPU, secBreak):  # items
         for p in procs:
             numAnts+=1
             p.start()
-        sols = [outQueue.get() for _ in procs if (time.perf_counter() - startTime) < secBreak]
+        sols = [outQueue.get() for _ in procs]
 
         outQueue = None
+        outLock  = None
 
         for Gant in sols:
             if Gant != None and Gant.S > Glocal.S:
@@ -86,8 +87,8 @@ def Solve( pallets, items, startTime, cfg, k, limit, numCPU, secBreak):  # items
 
         aco.updatePheroAttract(Glocal.S, Gbest.S, antsField, aco.NANTS)
 
-    if initialS > 0:
-        print(f"Used {numAnts} ants | ratio {Glocal.S/initialS:.3f} | {improvements} improvements")
+    # if initialS > 0:
+        # print(f"Used {numAnts} ants | ratio {Glocal.S/initialS:.3f} | {improvements} improvements")
 
     return mno.getSolMatrix(Gbest.Edges, numPallets, numItems)
 

### 2022-11-14 latex/comp_or.tex
Queue give up worked fine

@@
 				]
-				table[x=time, y=score, col sep=comma] {mpShims1data20.csv};
+				table[x=time, y=score, col sep=comma] {./csv/Shims_p1data20.csv};
 
@@
 				]
-				table[x=time, y=score, col sep=comma] {mpShims2data20.csv};
+				table[x=time, y=score, col sep=comma] {./csv/Shims_p2data20.csv};
 
@@
 	\caption{ {\it mpShims} performance with $1.2 \times volume$. }
-	\label{fig:mpShims2data20}
+	\label{fig:mpShimsdata20}
 \end{figure}
 
-Observing Figure \ref{fig:mpShims2data20}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is \textbf{1}.
+Observing Figure \ref{fig:mpShimsdata20}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is \textbf{1}.
 For {\it Scenario 2}, the best number is \textbf{4}.
@@
 				]
-				table[x=time, y=score, col sep=comma] {mpShims1data50.csv};
+				table[x=time, y=score, col sep=comma] {./csv/Shims_p1data50.csv};
 
@@
 				]
-				table[x=time, y=score, col sep=comma] {mpShims2data50.csv};
+				table[x=time, y=score, col sep=comma] {./csv/Shims_p2data50.csv};
 
@@
 	\caption{{\it mpShims} performance with $1.5 \times volume$.}
-	\label{fig:mpShims2data50}
+	\label{fig:mpShimsdata50}
 \end{figure}
 
-Observing Figure \ref{fig:mpShims2data50}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{1}. For {\it Scenario 2}, the best number is \textbf{8}.
+Observing Figure \ref{fig:mpShimsdata50}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{1}. For {\it Scenario 2}, the best number is \textbf{8}.
 
@@
 				]
-				table[x=time, y=score, col sep=comma] {mpACO1data20.csv};
+				table[x=time, y=score, col sep=comma] {./csv/ACO_p1data20.csv};
 
@@
 				]
-				table[x=time, y=score, col sep=comma] {mpACO2data20.csv};
+				table[x=time, y=score, col sep=comma] {./csv/ACO_p2data20.csv};
 
@@
 	\caption{{\it mpACO} performance with $1.2 \times volume$.}
-	\label{fig:mpACO2data20}
+	\label{fig:mpACOdata20}
 \end{figure}
 
-Observing Figure \ref{fig:mpACO2data20}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{40}. For {\it Scenario 2}, the best number is \textbf{1}.
+Observing Figure \ref{fig:mpACOdata20}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{40}. For {\it Scenario 2}, the best number is \textbf{1}.
 
@@
 				]
-				table[x=time, y=score, col sep=comma] {mpACO1data50.csv};
+				table[x=time, y=score, col sep=comma] {./csv/ACO_p1data50.csv};
 
@@
 				]
-				table[x=time, y=score, col sep=comma] {mpACO2data50.csv};
+				table[x=time, y=score, col sep=comma] {./csv/ACO_p2data50.csv};
 
@@
 	\caption{{\it mpACO} performance with $1.5 \times volume$.}
-	\label{fig:mpACO2data50}
+	\label{fig:mpACOdata50}
 \end{figure}
 
-Observing Figure \ref{fig:mpACO2data50}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{1}. For {\it Scenario 2}, the best number is \textbf{16}.
+Observing Figure \ref{fig:mpACOdata50}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{1}. For {\it Scenario 2}, the best number is \textbf{16}.
 

### 2022-11-14 aco_p.py

@@ -1,6 +1,4 @@
 import time
-import math
-import numpy as np
 import methods as mno
 import multiprocessing as mp
 import aco
@@ -47,7 +45,7 @@ def Solve( pallets, items, startTime, cfg, k, limit, numCPU, secBreak):  # items
 
     bestScore = Gbest.S
 
-    initialS = Gbest.S
+    # initialS = Gbest.S
 
     numPallets = len(pallets)
     numItems   = len(items)
@@ -60,22 +58,28 @@ def Solve( pallets, items, startTime, cfg, k, limit, numCPU, secBreak):  # items
         Glocal = mno.Solution(antsField, pallets, items, limit, cfg, k)
 
         procs = [None for _ in range(numCPU)]
+
+
         outQueue = mp.Queue()
+
         outLock  = mp.Lock()
         for i, p in enumerate(procs):
-
             # create a child process for each ant
             procs[i] = mp.Process( target=antQueue,args=( Glocal, cfg, k, antsField, outQueue, outLock, bestScore  ) )
+        
         for p in procs:
             numAnts+=1
             p.start()
-        sols = [outQueue.get() for _ in procs]
 
-        outQueue = None
-        outLock  = None
+        sols = [outQueue.get() for _ in procs if time.perf_counter() - startTime < secBreak ]
+       
+        for p in procs:
+            p.terminate()
+               
+        outQueue.close()
 
         for Gant in sols:
-            if Gant != None and Gant.S > Glocal.S:
+            if Gant.S > Glocal.S:
                 Glocal = Gant
 
         if Glocal.S > Gbest.S:

### 2022-11-15 latex/comp_or.tex
ACO_p experiments completed

@@
 These challenging benchmark instances are made publicly available.
-We also created a multiprocessing computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it was the unique method that managed to solve all testing scenarios.
+We also created a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as { \color{red} it was the unique method that managed to solve all testing scenarios.}
 
@@
 However, there are still other important challenges in air cargo transport that go beyond the definition of the ACLPP, especially with regard to the flight itinerary and the loading and unloading at each destination (or node) of this travel plan.
-In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Pickup and Delivery Problem} (PDP), and the search for the lowest cost route, which is the well-known {\it Traveling Salesman Problem} (TSP).
+In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Simultaneous Pickup and Delivery Problem} (SPDP), and the search for the most profitable route, which is the well-known {\it Traveling Salesman Problem} (TSP), similar to the proposed by \cite{kaspi2019}, where the profit per unit time is defined as the profit gained after completing the tour divided by the total time required to complete the tour.
 
@@
 		\toprule
-		                            & {\bf APP}  & {\bf WBP}  &  {\bf PDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} \\
+		                            & {\bf APP}  & {\bf WBP}  &  {\bf SPDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} \\
 		\midrule
@@
 		\cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
+
+		\cite{kaspi2019} & $.$ & $.$  & $.$ & $\bigstar$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
+
 		\cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
@@
 
+\cite{kaspi2019} define and solve a new extension of the TSP to maximize the financial contribution per invested time and present an optimal iterative solution procedure for the problem which converges after a limited number of iterations.
+
 \cite{RoesenerHall2014} solved APP and WBP as an integer programming problem, which also allows items to be loaded into pallets according to a specific orientation (e.g., this side up).
@@
 
-As already mentioned, \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and PDP in air cargo transport.
+As already mentioned, \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and SPDP in air cargo transport.
 The authors demonstrated that this problem is NP-hard and performed some experiments with real data, noting that their model offers better results than those obtained manually.
@@
 As can be seen, none of these works address air cargo palletization and load balancing with route optimization in a multi-leg flight plan.
-This is the objective of our work: to model and elaborate heuristics for a real problem that simultaneously involves 4 intractable sub-problems: APP, WBP, PDP and TSP.
+This is the objective of our work: to model and elaborate heuristics for a real problem that simultaneously involves 4 intractable sub-problems: APP, WBP, SPDP and TSP.
 
@@
 It is the primary air cargo pallet for more than 70 Air Forces and many air transport companies.
-This pallet has a capacity of $4500 kg$ and $13.7 m^3$, is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$.
+This pallet has a capacity of $4500 kg$ and $14.8 m^3$, is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$.
 For more information, see {\tt www.463LPallet.com}.
@@
 
-Real cases are more complex as they have hundreds of different items in each node and involve three intractable sub-problems: APP, WBP and PDP.
+Real cases are more complex as they have hundreds of different items in each node and involve three intractable sub-problems: APP, WBP and SPDP.
 Through the mathematical modeling presented in the previous section, we verify that {\it Mixed-Integer Programming}\/ (MIP) is not able to solve these cases in feasible time.
@@
 		\State $s \gets 0$
-		\State $\epsilon \gets 0$
+		\State $\tau \gets 0$
 		\State $weight \gets 0$
@@
 					\State $s \gets s + t_{ij}^k.s$ \label{eval:score1}
-					\State $\epsilon \gets \epsilon + t_{ij}^k.w * p_i.d$ \label{eval:eps1}
+					\State $\tau \gets \tau + t_{ij}^k.w * p_i.d$ \label{eval:eps1}
 				\EndIf
@@
 					\State $s \gets s + a_{iq}^k.s$   \label{eval:score2}
-					\State $\epsilon \gets \epsilon + a_{iq}^k.w * p_i.d$  \label{eval:eps2}
+					\State $\tau \gets \tau + a_{iq}^k.w * p_i.d$  \label{eval:eps2}
 				\EndIf
@@
 		\State $maxTorque \gets maxWeight \times limit^{CG}_{long}$
-		\State $\epsilon \gets \epsilon/maxTorque$ \label{eval:torque2}
+		\State $\epsilon \gets \tau/maxTorque$ \label{eval:torque2}
 		\State ${\bf return}\ s, \epsilon$	
@@
 This does not guarantee best results for all instances, but, on average, overall results are acceptable.
-Another important observation is that our heuristics were designed based on traditional ideas from the literature \cite{Fidanova2006} and on the experiences learned during tests.
+Another important observation is that our {\it ACO} implementations were designed based on traditional ideas from the literature \cite{Fidanova2006} and on the experiences learned during tests, plus the special features to process-based parallelism.
 
@@
 
-All the experiments are performed on a 64-bit, 128GiB, 3.6GHz, 48-core processor, 2 threads per core, Intel Xeon, with {\it Linux CentOS 8} as the operational system and {\it Python 3.10.4}\/ as the programming language.
+%All the experiments are performed on a 64-bit, 128GiB, 3.6GHz, 48-core processor, 2 threads per core, Intel Xeon, with {\it Linux CentOS 8} as the operational system and {\it Python 3.10.4}\/ as the programming language.
+
+All the experiments are performed on a 64-bit, 16GiB, 3.2GHz, 8-core processor, 2 threads per core, AMD FX-8320E, with {\it Linux Ubuntu 22.04} as the operational system and {\it Python 3.10.4}\/ as the programming language.
 
@@
 
-The {\it Multiprocessing} package (mp) offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
-Because of this, the {\it Multiprocessing} module lets the programmer take full advantage of the fact that a machine has more than one processor, 48 in our computer. This decision for {\it Multiprocessing} relies on the need to exchange information among processes (the pheromone trails or some information regarding load balancing, in our case), which would be harder to handle with thread-based parallelism.
+There is an optimal number of threads for each problem in each environment. This is the reason for the experiments ahead.
+
+It is known that working concurrently opens up synchronization issues.
+But the {\it Multiprocessing} package (mp) offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
+Because of this, the {\it Multiprocessing} module lets the programmer take full advantage of the fact that a machine has more than one processor, normally capable of 2 threads each.
+
+The decision for {\it Multiprocessing} relies on the need to exchange information among processes (the pheromone trails or some information regarding load balancing, in our case), which would be harder to handle with thread-based parallelism.
+For this, we used the multiprocessing queue feature, that is thread safe.
+
+Communication between processes can be achieved via {\it multiprocessing.Queue} and {\it multiprocessing.Pipe} which permits regular inter-process communication -IPC.
+(https://uwpce-pythoncert.github.io/Py300/threading-multiprocessing.html)
 
@@
 A process has two main functions: The first is to act as the resource holder for the application, and the second is to execute the instructions of the application.
+
+For each red dot closest to the upper left corner, we call it {\it Utopia}, an impossible result with zero time duration and the best possible score.
 
@@
 				table[x=time, y=score, col sep=comma] {./csv/Shims_p1data20.csv};
-
-				\legend{\it parallel processes}
 			\end{axis}
+			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
 		\end{tikzpicture}%
@@
 				table[x=time, y=score, col sep=comma] {./csv/Shims_p2data20.csv};
-
-				\legend{\it parallel processes}
 			\end{axis}
+			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);		
 		\end{tikzpicture}
@@
 
-Observing Figure \ref{fig:mpShimsdata20}, it may be noticed that the best number of processes for {\it Scenario 1} (closest to the upper left corner) is \textbf{1}.
+Observing Figure \ref{fig:mpShimsdata20}, it may be noticed that the best number of processes for {\it Scenario 1} (the closest to the {\it Utopia}) is \textbf{6}.
 For {\it Scenario 2}, the best number is \textbf{4}.
@@
 				table[x=time, y=score, col sep=comma] {./csv/Shims_p1data50.csv};
-
-				\legend{\it parallel processes}
 			\end{axis}
+			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
 		\end{tikzpicture}%
@@
 				table[x=time, y=score, col sep=comma] {./csv/Shims_p2data50.csv};
-
-				\legend{\it parallel processes}
 			\end{axis}
+			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
 		\end{tikzpicture}
-
 	\end{center}
@@
 
-Observing Figure \ref{fig:mpShimsdata50}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{1}. For {\it Scenario 2}, the best number is \textbf{8}.
-
 \begin{figure}[H]
-
 	\begin{center}
@@
 				table[x=time, y=score, col sep=comma] {./csv/ACO_p1data20.csv};
-
-				\legend{\it parallel processes}
 			\end{axis}
+			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
 		\end{tikzpicture}%
@@
 				table[x=time, y=score, col sep=comma] {./csv/ACO_p2data20.csv};
-
-				\legend{\it parallel processes}
 			\end{axis}
+			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
 		\end{tikzpicture}
-
 	\end{center}
@@
 \end{figure}
-
-Observing Figure \ref{fig:mpACOdata20}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{40}. For {\it Scenario 2}, the best number is \textbf{1}.
 
@@
 				table[x=time, y=score, col sep=comma] {./csv/ACO_p1data50.csv};
-
-				\legend{\it parallel processes}
 			\end{axis}
+			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
 		\end{tikzpicture}%
@@
 				table[x=time, y=score, col sep=comma] {./csv/ACO_p2data50.csv};
-
-				\legend{\it parallel processes}
 			\end{axis}
+			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
 		\end{tikzpicture}
-
 	\end{center}
@@
 
-Observing Figure \ref{fig:mpACOdata50}, it may be noticed that the best number of processes for {\it Scenario 1} is \textbf{1}. For {\it Scenario 2}, the best number is \textbf{16}.
+Observing figures \ref{fig:mpShimsdata20}, \ref{fig:mpShimsdata50}, \ref{fig:mpACOdata20}, \ref{fig:mpACOdata50}, it may be noticed that the best overall number of processes for all scenarios is \textbf{?}, because in average it is the number that most of the times is closer to the {\it Utopia} red bullet.
 
@@
 
-		\multirow{2}{*}{\it {mpShims}} &\multirow{2}{*}{\bf {24}}& $f$              & 5.88        & 3.16    & 3.16         &          &          &         &   \\ 
+		\multirow{2}{*}{\it {mpShims}} &\multirow{2}{*}{\bf {?}}& $f$              & 5.88        & 3.16    & 3.16         &          &          &         &   \\ 
 		                               &                         &{\bf run-time (s)}& 1           & 10      & 41           &            &            &          &   \\ %299s
@@
 
-        \multirow{2}{*}{\it {mpACO}}   &\multirow{2}{*}{\bf {32}}& $f$              & 5.82        & 3.12    &          &         &       &       &   \\ 
+        \multirow{2}{*}{\it {mpACO}}   &\multirow{2}{*}{\bf {?}}& $f$              & 5.82        & 3.12    &          &         &       &       &   \\ 
                                        &                         &{\bf run-time (s)}& 22          & 473     &           &         &       &       &   \\

### 2022-11-15 aco.py

@@ -4,6 +4,7 @@ import math
 import numpy as np
 import methods as mno
 import random
+from scipy import stats
 
 ALPHA = 1   # pheromone exponent (linearly affects attractivity)
 BETA  = 4   # heuristic exponent (exponentialy affects attractivity)
@@ -25,43 +26,40 @@ def randomChoice(values):
     return np.random.choice(indexes, p=probabilities)
 
 # getDeltaTau calculates the pheromone to be dropped by each on ants tracks
-def getDeltaTau(score, bestSoFar, numAnts):
-    if bestSoFar == 0:
-        bestSoFar = 1000
+def getDeltaTau(score, bestSoFar):
+
     DeltaTau = (score - bestSoFar)/bestSoFar # at this point DeltaTau may be positive ou negative learning
-    DeltaTau /= numAnts*numAnts
-    return DeltaTau
+    return DeltaTau*30.
 
 
 # Ant System (AS), the classic method that uses a random proportional state transition rule, while the pheromone is deposited
 # by all ants proportionally to their solution quality and is evaporated in all the components.
 # each Ant makes use of "updatePheroAttract" to update pheromones
 # and edge attractiveness according to the its solution value
-def updatePheroAttract(score, bestSoFar, edges, numAnts):
+def updatePheroAttract(score, bestSoFar, edges):
 
-    deltaTau = getDeltaTau(score, bestSoFar, numAnts)
+    deltaTau = getDeltaTau(score, bestSoFar)
 
-    maxAttract = 0.
-    if abs(deltaTau) < 1:
-        # update pheromone level in all edges
-        for id, e in enumerate(edges):
+    Pheromone = [] 
 
-            #evaporate some pheromone 
-            edges[id].Pheromone = math.sqrt(e.Pheromone) / 1.25 # RHO = 0.2
+    # update pheromone level in all edges
+    for id, e in enumerate(edges):
 
+        #evaporate some pheromone 
+        edges[id].Pheromone = math.sqrt(e.Pheromone) / 1.4 # RHO = 0.2
+
+        # update pheromone level
+        if e.Pheromone + deltaTau > 0 and e.Pheromone + deltaTau <= 1:
             edges[id].Pheromone += deltaTau
-            if edges[id].Pheromone < 0.:
-                edges[id].Pheromone = 0.
+        
+        Pheromone.append(e.Pheromone)
+        
+        # update the general attractiveness
[…]

### 2022-11-15 tests/byvalue.py

@@ -1,27 +1,22 @@
 
-def set_list(list):
+def set_list(list): # by value
 	list = ["A", "B", "C"]
 	return list
 
-def add(list):
+my_list = ["X","Y"]
+
+new_list = set_list(my_list)
+
+print(new_list)
+
+
+def add(list): # by reference
 	list.append("D")
 	return list
 
 my_list = ["E"]
-add(my_list)
-set_list(my_list)
-print(my_list)
 
+new_list = add(my_list)
 
-def set_list(list):
-	list = 5
-	return list
+print(new_list)
 
-def add(list):
-	list += 1
-	return list
-
-my_list = 2
-add(my_list)
-set_list(my_list)
-print(my_list)

### 2022-11-15 latex/comp_or.tex
Starting to change paper to multiprocess

@@
 
-Below, we present four implementations for $SolveNode$: {\it Ant Colony Optimization}\/ (ACO), {\it Multiprocessed Ant Colony Optimization}\/ ({\it mpACO}),  and {\it Shims},  and {\it Multiprocessed Shims (mpShims)}.
-
-It is important to say that the parallel ACO ({\it pACO}) is a different implementation where employs threads instead of processes to solve each ant solution in parallel.
+Below, we present 2 implementations for $SolveNode$: {\it Multiprocess Ant Colony Optimization}\/ ({\it mpACO})  and a new parallel heuristic that we propose called {\it Multiprocess Shims (mpShims)}.
+
+It is important to say that the parallel ACO ({\it pACO}) is a different implementation where employs threads instead of processes to solve each ant solution in parallel (for a detailed information on parallel ACO see \cite{Pedemonte2011}).
 We chose {\it mpACO} for our approach because of the need to share the pheromone array between ants/processes and the facility to handle the possibility of race condition.
 
-\subsubsection{Ant Colony Optimization (ACO)}
+\subsubsection{Multiprocess Ant Colony Optimization (mpACO)}
 
@@
 
-	\caption{ $SolveNode(ACO, k, G)$ } \label{alg:aco}
+	\caption{ $SolveNode(mpACO, k, G)$ } \label{alg:aco}
 	\begin{algorithmic}[1]
@@
 		\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^*(V_k,E^{N*}_k \cup E^Q_k)$
-		\State $A \gets 4$ \label{aco:ants}
+		\State $A \gets$ a number of ants or parallel processes to be determined for each iteration \label{aco:ants}
 		\State $ga^k_{ij}   \gets 0$, $1 \leq i \leq m$, $1 \leq j \leq n_k$  \label{aco:init1}
@@
 		\While{($stagnant < 3$) {\bf and} ($runtime < 0.7s$)} \label{aco:while}
+
+		\State $publish$
+		\State $solutions \gets subscribe()$
+
 			\For{$a \gets 1$ to $A$} \label{aco:g1}
@@
 
-\subsubsection{Shims}
-
-Finally, we present a new heuristic designed specifically for ACLP+RPDP, which we named \emph{Shims}.
+\subsubsection{Multiprocess Shims - mpShims}
+
+Finally, we present a new heuristic designed specifically for ACLP+RPDP, which we named \emph{mpShims}.
 Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}).
@@
 
-Communication between processes can be achieved via {\it multiprocessing.Queue} and {\it multiprocessing.Pipe} which permits regular inter-process communication -IPC.
-(https://uwpce-pythoncert.github.io/Py300/threading-multiprocessing.html)
+Communication between processes can be achieved via {\it multiprocessing.Queue} and {\it multiprocessing.Pipe} which permits regular inter-process communication -IPC. (\cite{multiprocessing}).
 

### 2022-11-16 latex/comp_or.tex
The job computer was 10 times faster.

@@
 		y tick label style={ /pgf/number format/.cd, fixed, fixed zerofill, precision=3, /tikz/.cd, },
+		x tick label style={ /pgf/number format/.cd, fixed, fixed zerofill, precision=1, /tikz/.cd, },
 		grid=major,		
@@
 
-%All the experiments are performed on a 64-bit, 128GiB, 3.6GHz, 48-core processor, 2 threads per core, Intel Xeon, with {\it Linux CentOS 8} as the operational system and {\it Python 3.10.4}\/ as the programming language.
-
-All the experiments are performed on a 64-bit, 16GiB, 3.2GHz, 8-core processor, 2 threads per core, AMD FX-8320E, with {\it Linux Ubuntu 22.04} as the operational system and {\it Python 3.10.4}\/ as the programming language.
+The testing experiments are performed on a 64-bit, 16GiB, 3.2GHz, 8-core processor, 2 threads per core, AMD FX-8320E, with {\it Linux Ubuntu 22.04} as the operational system and {\it Python 3.10.4}\/ as the programming language.
+
+The final experiments are performed on a 64-bit, 128GiB, 3.6GHz, 48-core processor, 2 threads per core, Intel Xeon, with {\it Linux CentOS 8} as the operational system and {\it Python 3.8.2}\/ as the programming language.
 
@@
 
-There is an optimal number of threads for each problem in each environment. This is the reason for the experiments ahead.
+According to \cite[p.271]{Breshears2009}, a {\it Process} is the operating system’s spawned and controlled entity that encapsulates an executing application.
+A process has two main jobs: the first is to hold the application's resources, and the second is to carry out the application's instructions.
 
@@
 The decision for {\it Multiprocessing} relies on the need to exchange information among processes (the pheromone trails or some information regarding load balancing, in our case), which would be harder to handle with thread-based parallelism.
-For this, we used the multiprocessing queue feature, that is thread safe.
-
-Communication between processes can be achieved via {\it multiprocessing.Queue} and {\it multiprocessing.Pipe} which permits regular inter-process communication -IPC. (\cite{multiprocessing}).
-
-According to \cite[p.271]{Breshears2009}, a Process is the operating system’s spawned and controlled entity that encapsulates an executing application.
-A process has two main functions: The first is to act as the resource holder for the application, and the second is to execute the instructions of the application.
-
-For each red dot closest to the upper left corner, we call it {\it Utopia}, an impossible result with zero time duration and the best possible score.
+For this, we used the multiprocessing queue feature that, according to \cite{multiprocessing}, is thread safe.
+
+By Amdahl's law, there is an optimal number of parallel executions for each problem in each environment.
+While working at IBM in 1967, Gene Amdahl developed the foundation for what became known as Amdahl's Law or Amdahl's Argument.
+Essentially, the law states that while a process can be decomposed into steps that may then be run in parallel, the time taken for the whole process will be significantly limited by the steps that remain serialized.
+So, we decided to run some tests to find the best number of parallel processes for each of this work's operational scenarios.
+
+For each red bullet closest to the upper left corner, we call it {\it Utopia}, an impossible result with zero time duration and a possible best score.
+The blue bullets are experiments performed with the number of parallel processes indicated.
+The shortest distance from {\it Utopia} may reveal the most efficient parallel processes for the scenario (see Table \ref{tab:scenarios}).
 

### 2022-11-17 latex/comp_or.tex
Plots order change

@@
 				myplotstyle,
-				xlabel={\bf Scenario 1 computation times (s)},
+				xlabel={\bf mpShims computation times (s)},
 				ylabel={\bf Score},
@@
 			myplotstyle,
-				xlabel={\bf Scenario 2 computation times (s)},
+			xlabel={\bf mpACO computation times (s)},
+			ylabel={\bf Score},
 			]
@@
 			]
-				table[x=time, y=score, col sep=comma] {./csv/Shims_p2data20.csv};
+			table[x=time, y=score, col sep=comma] {./csv/ACO_p1data20.csv};
 		\end{axis}
@@
 	\end{center}
-	\caption{ {\it mpShims} performance with $1.2 \times volume$. }
-	\label{fig:mpShimsdata20}
+	\caption{ Scenario 1 performances with $1.2 \times volume$. }
+	\label{fig:scenario1data20}
 \end{figure}
 
-Observing Figure \ref{fig:mpShimsdata20}, it may be noticed that the best number of processes for {\it Scenario 1} (the closest to the {\it Utopia}) is \textbf{6}.
+Observing Figure \ref{fig:scenario1data20}, it may be noticed that the best number of processes for {\it Scenario 1} (the closest to the {\it Utopia}) is \textbf{6}.
 For {\it Scenario 2}, the best number is \textbf{4}.
@@
 				myplotstyle,
-				xlabel={\bf Scenario 2 computation times (s)},
-				]
-				\addplot+[
-				only marks,	
-				nodes near coords = {\numProcs{}},
-				visualization depends on = {value \thisrow{np} \as \numProcs},						
-				]
-				table[x=time, y=score, col sep=comma] {./csv/Shims_p2data50.csv};
-			\end{axis}
-			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-		\end{tikzpicture}
-	\end{center}
-	\caption{{\it mpShims} performance with $1.5 \times volume$.}
-	\label{fig:mpShimsdata50}
-\end{figure}
-
-\begin{figure}[H]
-	\begin{center}
-		\begin{tikzpicture}[scale=0.8, samples=100]
-			\begin{axis}[
-				myplotstyle,
-				xlabel={\bf Scenario 1 computation times (s)},
+				xlabel={\bf mpACO computation times (s)},
 				ylabel={\bf Score},
@@
 				]
-				table[x=time, y=score, col sep=comma] {./csv/ACO_p1data20.csv};
+				table[x=time, y=score, col sep=comma] {./csv/ACO_p1data50.csv};
+			\end{axis}
+			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
+		\end{tikzpicture}
+
+	\end{center}
+	\caption{ Scenario 1 performances with $1.5 \times volume$. }
+	\label{fig:scenario1data50}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+		\begin{tikzpicture}[scale=0.8, samples=100]
+			\begin{axis}[
+				myplotstyle,
+				xlabel={\bf mpShims computation times (s)},
+				ylabel={\bf Score},
+				]
+				\addplot+[
+				only marks,	
+				nodes near coords = {\numProcs{}},
+				visualization depends on = {value \thisrow{np} \as \numProcs},						
+				]
+				table[x=time, y=score, col sep=comma] {./csv/Shims_p2data20.csv};
 			\end{axis}
@@
 				myplotstyle,
-				xlabel={\bf Scenario 2 computation times (s)},
+				xlabel={\bf mpACO computation times (s)},
+				ylabel={\bf Score},
 				]
@@
 		\end{tikzpicture}
+
 	\end{center}
-	\caption{{\it mpACO} performance with $1.2 \times volume$.}
-	\label{fig:mpACOdata20}
+	\caption{ Scenario 2 performances with $1.2 \times volume$. }
+	\label{fig:scenario2data20}
 \end{figure}
@@
 \begin{figure}[H]
-
 	\begin{center}
@@
 				]
-				table[x=time, y=score, col sep=comma] {./csv/ACO_p1data50.csv};
+				table[x=time, y=score, col sep=comma] {./csv/Shims_p2data50.csv};
 			\end{axis}
@@
 		\end{tikzpicture}
+
 	\end{center}
-	\caption{{\it mpACO} performance with $1.5 \times volume$.}
-	\label{fig:mpACOdata50}
+	\caption{ Scenario 2 performances with $1.5 \times volume$. }
+	\label{fig:scenario2data50}
 \end{figure}
 
-Observing figures \ref{fig:mpShimsdata20}, \ref{fig:mpShimsdata50}, \ref{fig:mpACOdata20}, \ref{fig:mpACOdata50}, it may be noticed that the best overall number of processes for all scenarios is \textbf{?}, because in average it is the number that most of the times is closer to the {\it Utopia} red bullet.
+%Observing figures \ref{fig:mpShimsdata20}, \ref{fig:mpShimsdata50}, \ref{fig:mpACOdata20}, \ref{fig:mpACOdata50}, it may be noticed that the best overall number of processes for all scenarios is \textbf{?}, because in average it is the number that most of the times is closer to the {\it Utopia} red bullet.
 

### 2022-11-17 latex/comp_or.tex
Shared data in parallel processing

@@
 Below, we present 2 implementations for $SolveNode$: {\it Multiprocess Ant Colony Optimization}\/ ({\it mpACO})  and a new parallel heuristic that we propose called {\it Multiprocess Shims (mpShims)}.
+
+According to \cite[p.226]{manfrin2006}, there are five topologies for multi-process interchange of information: {\it fully-connected}, where the master process broadcasts to all remaining child processes; {\it replace-worst}, where the best-so-far solution process broadcasts only to the current worst solution process; {\it hypercube}, where processes are connected as a hypercube, and a vertex process broadcasts only to the connected vertices; {\it ring}, in which one process only sends a message to the next process connected to it; and {\it parallel independent runs (PIR)}, in which there are no communication costs and the best solution is chosen among all processes.
+They concluded that the PIR topology presented, in average, the smallest distance from the optimal solutions.
 
@@
 Because of this, the {\it Multiprocessing} module lets the programmer take full advantage of the fact that a machine has more than one processor, normally capable of 2 threads each.
-
-The decision for {\it Multiprocessing} relies on the need to exchange information among processes (the pheromone trails or some information regarding load balancing, in our case), which would be harder to handle with thread-based parallelism.
-For this, we used the multiprocessing queue feature that, according to \cite{multiprocessing}, is thread safe.
 
@@
 For {\it Scenario 2}, the best number is \textbf{4}.
-
-\begin{figure}[H]
-	\begin{center}
-		\begin{tikzpicture}[scale=0.8, samples=100]
-			\begin{axis}[
-				myplotstyle,
-				xlabel={\bf mpShims computation times (s)},
-				ylabel={\bf Score},
-				]
-				\addplot+[
-				only marks,	
-				nodes near coords = {\numProcs{}},
-				visualization depends on = {value \thisrow{np} \as \numProcs},						
-				]
-				table[x=time, y=score, col sep=comma] {./csv/Shims_p1data50.csv};
-			\end{axis}
-			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-		\end{tikzpicture}%
-		~%
-		%
-		\begin{tikzpicture}[scale=0.8, samples=100]
-			\begin{axis}[
-				myplotstyle,
-				xlabel={\bf mpACO computation times (s)},
-				ylabel={\bf Score},
-				]
-				\addplot+[
-				only marks,	
-				nodes near coords = {\numProcs{}},
-				visualization depends on = {value \thisrow{np} \as \numProcs},						
-				]
-				table[x=time, y=score, col sep=comma] {./csv/ACO_p1data50.csv};
-			\end{axis}
-			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-		\end{tikzpicture}
-
-	\end{center}
-	\caption{ Scenario 1 performances with $1.5 \times volume$. }
-	\label{fig:scenario1data50}
-\end{figure}
-
-\begin{figure}[H]
-	\begin{center}
-		\begin{tikzpicture}[scale=0.8, samples=100]
-			\begin{axis}[
-				myplotstyle,
-				xlabel={\bf mpShims computation times (s)},
-				ylabel={\bf Score},
-				]
-				\addplot+[
-				only marks,	
-				nodes near coords = {\numProcs{}},
-				visualization depends on = {value \thisrow{np} \as \numProcs},						
-				]
-				table[x=time, y=score, col sep=comma] {./csv/Shims_p2data20.csv};
-			\end{axis}
-			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-		\end{tikzpicture}%
-		~%
-		%
-		\begin{tikzpicture}[scale=0.8, samples=100]
-			\begin{axis}[
-				myplotstyle,
-				xlabel={\bf mpACO computation times (s)},
-				ylabel={\bf Score},
-				]
-				\addplot+[
-				only marks,	
-				nodes near coords = {\numProcs{}},
-				visualization depends on = {value \thisrow{np} \as \numProcs},						
-				]
-				table[x=time, y=score, col sep=comma] {./csv/ACO_p2data20.csv};
-			\end{axis}
-			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-		\end{tikzpicture}
-
-	\end{center}
-	\caption{ Scenario 2 performances with $1.2 \times volume$. }
-	\label{fig:scenario2data20}
-\end{figure}
-
-\begin{figure}[H]
-	\begin{center}
-		\begin{tikzpicture}[scale=0.8, samples=100]
-			\begin{axis}[
-				myplotstyle,
-				xlabel={\bf mpShims computation times (s)},
-				ylabel={\bf Score},
-				]
-				\addplot+[
-				only marks,	
-				nodes near coords = {\numProcs{}},
-				visualization depends on = {value \thisrow{np} \as \numProcs},						
-				]
-				table[x=time, y=score, col sep=comma] {./csv/Shims_p2data50.csv};
-			\end{axis}
-			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-		\end{tikzpicture}%
-		~%
-		%
-		\begin{tikzpicture}[scale=0.8, samples=100]
-			\begin{axis}[
-				myplotstyle,
-				xlabel={\bf mpACO computation times (s)},
-				ylabel={\bf Score},
-				]
-				\addplot+[
-				only marks,	
-				nodes near coords = {\numProcs{}},
-				visualization depends on = {value \thisrow{np} \as \numProcs},						
-				]
-				table[x=time, y=score, col sep=comma] {./csv/ACO_p2data50.csv};
-			\end{axis}
-			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-		\end{tikzpicture}
-
-	\end{center}
-	\caption{ Scenario 2 performances with $1.5 \times volume$. }
-	\label{fig:scenario2data50}
-\end{figure}
-
-%Observing figures \ref{fig:mpShimsdata20}, \ref{fig:mpShimsdata50}, \ref{fig:mpACOdata20}, \ref{fig:mpACOdata50}, it may be noticed that the best overall number of processes for all scenarios is \textbf{?}, because in average it is the number that most of the times is closer to the {\it Utopia} red bullet.
+Table \ref{tab:numProcs} presents all methods, volumes percentages and scenarios numbers of best parallel processes to be run to solve all tours.
+
+To build Table \ref{tab:numProcs}, it was necessary to determine the number of parallel processes that best solve each node. For this we created a procedure as in Algorithm \ref{alg:bestnum}.
+
+\begin{algorithm}[H]
+	\caption{$getBestNPP(procs, scores, runtimes)$}  \label{alg:bestnum}
+	\begin{algorithmic}[1]
+		\State $maxS   \gets max(scores)$
+		\State $minS   \gets min(scores)$
+		\State $maxR \gets max(runtimes)$
+		\State $minR \gets min(runtimes)$
+
+		\State $bestNPP \gets 0$, best number of parallel processes
+		\State $minD \gets 999999999.9$
+
+		\For{$i$ in $procs$}
+			\State $y = ( maxS - scores[i] )\ /\ (maxS   - minS  )$
+			\State $x = ( runtimes[i] - minR )\ /\ (maxR - minR)$
+			\State $h = \sqrt{x^2 + y^2}$, distance from {\it Utopia}
+			\If{$minD$ > $h$}
+				\State $minD \gets h$
+				\State $bestNPP \gets procs[i]$
+			\EndIf
+
+		\EndFor
+
+		\State ${\bf return}$ $bestNPP$
+	\end{algorithmic}
+\end{algorithm}
+
+\vspace{2.0mm}
+
+\begin{table}[H]
+\centering
+\caption{$bestNPP$ found for each scenario}  \label{tab:numProcs}
+	\footnotesize
+%\scriptsize
+\begin{tabular}{rccccccc}
+\toprule
+                               &               &\multicolumn{6}{c}{\bf Scenarios} \\
+                               \cmidrule{3-8}
+{$method$}                     &{\it volume}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 6} \\
+\toprule
+
+\multirow{3}{*}{\it {mpShims}} &$1.2$& 6       & 2     & 0     & 0     & 0     & 0    \\ 
+                               &$1.5$& 4       & 4     & 0     & 0     & 0     & 0    \\
+                               &$2.0$& 0       & 0     & 0     & 0     & 0     & 0    \\
+\midrule[.1pt]
+
+\multirow{3}{*}{\it {mpACO}}   &$1.2$& 4       & 2     & 0     & 0     & 0     & 0    \\ 
+                               &$1.5$& 6       & 1     & 0     & 0     & 0     & 0    \\ 
+                               &$2.0$& 0       & 0     & 0     & 0     & 0     & 0    \\                               
+\bottomrule		
+\end{tabular}
+
+\normalsize
+
+\end{table}
+
+Table \ref{tab:numProcs} presents the best numbers of parallel processes for each method, scenario, and volume surplus.
 
@@
 \caption{Methods and scenarios with $1.2 \times volume$}  \label{tab:20}
-%	\footnotesize
-\scriptsize
-	\begin{tabular}{r|c|c|cccccc|c}
+\footnotesize
+%\scriptsize
+\begin{tabular}{rccccccc|c}
 \toprule
-		{$method$}                     &{\bf Procs}              &{\bf Results}     &{\bf 1}     &{\bf 2}    &{\bf 3}      &{\bf 4}      &{\bf 5}      &{\bf 6}     &\specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
+                                &                &\multicolumn{6}{c}{\bf Scenarios} &\\
+\cmidrule{3-8}		
+{$method$}                      &{\bf Results}   &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 6}&\specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
 \toprule
-
-		\multirow{2}{*}{\it {Shims}}   &\multirow{2}{*}{\bf {1}} & $f$              & 5.40        & 2.94    & 2.93         &          &          &         &   \\ 
-		                               &                         &{\bf run-time (s)}& <1          & 1       & 4            &            &            &          &   \\ %206s
-		\midrule[.1pt]
-
-		\multirow{2}{*}{\it {mpShims}} &\multirow{2}{*}{\bf {?}}& $f$              & 5.88        & 3.16    & 3.16         &          &          &         &   \\ 
-		                               &                         &{\bf run-time (s)}& 1           & 10      & 41           &            &            &          &   \\ %299s
-		\midrule[.1pt]
-
-		\multirow{2}{*}{\it {ACO}}     &\multirow{2}{*}{\bf {1}} & $f$              & 5.77        & 3.49    & 3.02         &        &       &      &   \\ 
-                                       &                         &{\bf run-time (s)}& 14          & 808     & 2398          &        &       &      &   \\ 
-        \midrule[.1pt]
-
-        \multirow{2}{*}{\it {mpACO}}   &\multirow{2}{*}{\bf {?}}& $f$              & 5.82        & 3.12    &          &         &       &       &   \\ 
-                                       &                         &{\bf run-time (s)}& 22          & 473     &           &         &       &       &   \\
-
+\multirow{2}{*}{\it {mpShims}} & $f$              & 5.88 & 3.16    & 3.16 &      &      &      &   \\ 
+                               &{\bf run-time (s)}& 1    & 10      & 41   &      &      &      &   \\
+\midrule
+
+\multirow{2}{*}{\it {mpACO}}   & $f$              & 5.82 & 3.12    &      &      &      &      &   \\ 
+                               &{\bf run-time (s)}& 22   & 473     &      &      &      &      &   \\ 
 \bottomrule		
@@
 \section{Conclusions}
-\label{sec7}
 

### 2022-11-17 shims_p.py

@@ -34,7 +34,6 @@ def Compute(edges, pallets, items, limit, cfg, k) :
 
     return sol
 
-
 def Solve(pallets, items, cfg, k, minLim, numProcs): # items include kept on board
 
     print(f"\nParallel Shims for ACLP+RPDP")

### 2022-11-17 tests/comm_process.py

@@ -1,0 +1,43 @@
+import multiprocessing
+
+def square_list(mylist, result, square_sum):
+	"""
+	function to square a given list
+	"""
+	# append squares of mylist to result array
+	for idx, num in enumerate(mylist):
+		result[idx] = num * num
+
+	# square_sum value
+	square_sum.value = sum(result)
+
+	# print result Array
+	print("Result(in process p1): {}".format(result[:]))
+
+	# print square_sum Value
+	print("Sum of squares(in process p1): {}".format(square_sum.value))
+
+if __name__ == "__main__":
+	# input list
+	mylist = [1,2,3,4]
+
+	# creating Array of int data type with space for 4 integers
+	result = multiprocessing.Array('i', 4)
+
+	# creating Value of int data type
+	square_sum = multiprocessing.Value('i')
+
+	# creating new process
+	p1 = multiprocessing.Process(target=square_list, args=(mylist, result, square_sum))
+
+	# starting process
+	p1.start()
+
+	# wait until the process is finished
+	p1.join()
+
+	# print result array
+	print("Result(in main program): {}".format(result[:]))
+
+	# print square_sum Value
+	print("Sum of squares(in main program): {}".format(square_sum.value))

### 2022-11-18 latex/comp_or.tex
Continue tests in Calypso

@@
 
-\subsection{Instances generation}
+\subsection{Items generation}
 \label{items}
@@
 
+We test our methods with 126 different instances because we have six operational scenarios with seven randomly generated item sets for three volume surpluses (1.2, 1.5, and 2.0 times aircraft volume capacities).
+
 \subsection{Results obtained}
 
-The testing experiments are performed on a 64-bit, 16GiB, 3.2GHz, 8-core processor, 2 threads per core, AMD FX-8320E, with {\it Linux Ubuntu 22.04} as the operational system and {\it Python 3.10.4}\/ as the programming language.
-
-The final experiments are performed on a 64-bit, 128GiB, 3.6GHz, 48-core processor, 2 threads per core, Intel Xeon, with {\it Linux CentOS 8} as the operational system and {\it Python 3.8.2}\/ as the programming language.
+%The testing experiments are performed on a 64-bit, 16GiB, 3.2GHz, 8-core processor, 2 threads per core, AMD FX-8320E, with {\it Linux Ubuntu 22.04} as the operational system and {\it Python 3.10.4}\/ as the programming language.
+
+The final experiments are performed on a 64-bit, 128GiB, 3.6GHz, 48-core processor, 2 threads per core, Intel Xeon, with {\it Linux CentOS 8} as the operational system and {\it Python 3.8.2} as the programming language.
 
@@
 \begin{algorithm}[H]
-	\caption{$getBestNPP(procs, scores, runtimes)$}  \label{alg:bestnum}
+	\caption{$getBestNPP(scores, runtimes, procs)$}  \label{alg:bestnum}
 	\begin{algorithmic}[1]
@@
 
-		\For{$i$ in $procs$}
-			\State $y = ( maxS - scores[i] )\ /\ (maxS   - minS  )$
-			\State $x = ( runtimes[i] - minR )\ /\ (maxR - minR)$
-			\State $h = \sqrt{x^2 + y^2}$, distance from {\it Utopia}
+		\For{$i \in |scores|$}
+			\State $y = ( maxS - scores[i] )\ /\ (maxS   - minS  )$, the Y axis leg of Figure \ref{fig:scenario1data20}
+			\State $x = ( runtimes[i] - minR )\ /\ (maxR - minR)$. the X axis leg
+			\State $h = \sqrt{x^2 + y^2}$, the hypotenuse, distance from {\it Utopia}
 			\If{$minD$ > $h$}
@@
 
-\multirow{3}{*}{\it {mpShims}} &$1.2$& 6       & 2     & 0     & 0     & 0     & 0    \\ 
-                               &$1.5$& 4       & 4     & 0     & 0     & 0     & 0    \\
-                               &$2.0$& 0       & 0     & 0     & 0     & 0     & 0    \\
+\multirow{3}{*}{\it {mpShims}} &$1.2$& 6       & 4     & 4     & 4     & 2     & 2    \\ 
+                               &$1.5$& 4       & 4     & 4     & 4     & 4     & 4    \\
+                               &$2.0$& 4       & 4     & 4     & 2     & 2     & 4    \\
 \midrule[.1pt]
 
-\multirow{3}{*}{\it {mpACO}}   &$1.2$& 4       & 2     & 0     & 0     & 0     & 0    \\ 
+\multirow{3}{*}{\it {mpACO}}   &$1.2$& 10      & 8     & 10    & 6     & 10     & 10    \\ 
                                &$1.5$& 6       & 1     & 0     & 0     & 0     & 0    \\ 
@@
 Table \ref{tab:numProcs} presents the best numbers of parallel processes for each method, scenario, and volume surplus.
+As all numbers are less than or equal to 8, any hand-held computer with 8 cores can yield efficient solutions.
 

### 2022-11-18 aco.py

@@ -36,11 +36,11 @@ def getDeltaTau(score, bestSoFar):
 # by all ants proportionally to their solution quality and is evaporated in all the components.
 # each Ant makes use of "updatePheroAttract" to update pheromones
 # and edge attractiveness according to the its solution value
-def updatePheroAttract(score, bestSoFar, edges):
+def updatePheroAttract(score, bestSoFar, edges, mpAttract):
 
     deltaTau = getDeltaTau(score, bestSoFar)
 
-    Pheromone = [] 
+    # Pheromone = [] 
 
     # update pheromone level in all edges
     for id, e in enumerate(edges):
@@ -52,10 +52,12 @@ def updatePheroAttract(score, bestSoFar, edges):
         if e.Pheromone + deltaTau > 0 and e.Pheromone + deltaTau <= 1:
             edges[id].Pheromone += deltaTau
         
-        Pheromone.append(e.Pheromone)
+        # Pheromone.append(e.Pheromone)
         
         # update the general attractiveness
         edges[id].updateAttract(ALPHA, BETA)
+
+        mpAttract[id] = edges[id].Attract
 
     # mean = stats.tmean(Pheromone)
     # sdev = stats.tstd(Pheromone)

### 2022-11-18 aco_p.py

@@ -1,16 +1,19 @@
 import time
 import methods as mno
 import multiprocessing as mp
+# import threading as thrd
+
 import aco
 
 # Process-based parallelism
 
-
-
-def antSolve(Gant, cfg, k, antsField, bestScore):
+def antSolve(Gant, cfg, k, antsField, bestScore, mpAttract):
     
     Nbhood   = [ce for ce in Gant.Edges if not ce.InSol]
     attracts = [ce.Attract for ce in Nbhood]
+
+    # for i, e in enumerate(Nbhood): # https://superfastpython.com/multiprocessing-shared-ctypes-in-python/
+    #     attracts[i] = mpAttract[e.ID]
 
     while Nbhood:
 
@@ -19,14 +22,14 @@ def antSolve(Gant, cfg, k, antsField, bestScore):
         if Gant.isFeasible(ce, 1.0, cfg, k):
             Gant.putInSol(ce)
     
-    aco.updatePheroAttract(Gant.S, bestScore, antsField)
+    aco.updatePheroAttract(Gant.S, bestScore, antsField, mpAttract)
 
     return Gant
 
-def enqueue( antsQueue,     Gant, cfg, k, antsField, bestScore ):
-    antsQueue.put( antSolve(Gant, cfg, k, antsField, bestScore) )
+def enqueue( antsQueue,     Gant, cfg, k, antsField, bestScore, mpAttract ):
+    antsQueue.put( antSolve(Gant, cfg, k, antsField, bestScore, mpAttract) )
 
-def publish(antsQueue, procs, limits, antsField, pallets, items, cfg, k, bestScore ):
+def publish(antsQueue, procs, limits, antsField, pallets, items, cfg, k, bestScore, mpAttract ):
 
     for i, _ in enumerate(procs):
 
@@ -35,7 +38,8 @@ def publish(antsQueue, procs, limits, antsField, pallets, items, cfg, k, bestSco
         Glocal = mno.Solution(antsField, pallets, items, limit, cfg, k)
 
         # create a child process for each ant
-        procs[i] = mp.Process( target=enqueue, args=( antsQueue, Glocal, cfg, k, antsField, bestScore ) )
+        procs[i] = mp.Process( target=enqueue, args=( antsQueue, Glocal, cfg, k, antsField, bestScore, mpAttract ) ) # faster than thread
+        # procs[i] = thrd.Thread( target=enqueue, args=( antsQueue, Glocal, cfg, k, antsField, bestScore ) )
[…]

### 2022-11-19 latex/comp_or.tex
Citing Mesquita and Sanches 2023

@@
-% !TeX spellcheck = en_GB
+% !TeX spellcheck = en_US
 \documentclass[preprint,authoryear]{elsarticle}
@@
 This enables other risks such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
-We contributed by modeling and solving a new problem of planning the loading and routing of an aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
-This new hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modeled using standardized pallets in fixed positions, obeying the center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
+We contributed by modelling and solving a new problem of planning the loading and routing of an aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
+This new hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modelled using standardized pallets in fixed positions, obeying the center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
 We also contributed by carrying out multiple experiments with the well-known Ant Colony Optimization on synthetic data based on real data from the {\it Brazilian Air Force}\/ transportation history.
@@
 401]{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP) and {\it Weight and Balance Problem} (WBP).
-Several aspects were considered in this modeling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc.
+Several aspects were considered in this modelling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc.
 They also presented a comprehensive bibliographic survey of solving methods that had been developed in different situations.
@@
 In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Simultaneous Pickup and Delivery Problem} (SPDP), and the search for the most profitable route, which is the well-known {\it Traveling Salesman Problem} (TSP), similar to the proposed by \cite{kaspi2019}, where the profit per unit time is defined as the profit gained after completing the tour divided by the total time required to complete the tour.
+
+\cite{MesquitaSanches2023} explored these problems by solving them in an integrated form.
 
@@
 	\scriptsize
-	\renewcommand{\arraystretch}{1.1} % alturas das linhas
-	\begin{tabular}{r|cc|cc|cc|ccc}
+	\renewcommand{\arraystretch}{1.1} % alturas das linhas \specialcell{{\bf Parallel}\\{\bf heuristics}}
+	\begin{tabular}{r|cc|cc|cc|ccc|c}
 		\toprule
-		                            & {\bf APP}  & {\bf WBP}  &  {\bf SPDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} \\
-		\midrule
-		\cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$ \\
-		\cite{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$ \\
-		\cite{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$ \\
-		\cite{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
-		\cite{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$ \\
-		\cite{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$ \\
-		\cite{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{MesquitaCunha2011}   & $.$ & $.$    & $\bigstar$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\		
-		\cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-
-		\cite{kaspi2019} & $.$ & $.$  & $.$ & $\bigstar$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
-
-		\cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{Vancroonemburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
-		\cite{PaquaySchynsLimbourg2016,PaquayLimbourgSchynsOliveira2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$ \\
-		\cite{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$ \\
-		\cite{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$  \\
-		\cite{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$  \\
-		\cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
-		{\bf This work}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$  \\
+		                            & {\bf APP}  & {\bf WBP}  &  {\bf SPDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} & \specialcell{{\bf Parallel}\\{\bf heuristics}}\\
+		\midrule
+		\cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$&  $.$ \\
+		\cite{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$&  $.$ \\
+		\cite{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ &  $.$\\
+		\cite{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$&  $.$ \\
+		\cite{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ &  $.$\\
+		\cite{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$&  $.$ \\	
+		\cite{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$&  $.$ \\
+		\cite{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$&  $.$ \\
+		\cite{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$&  $.$ \\
+		\cite{MesquitaCunha2011}   & $.$ & $.$    & $\bigstar$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ &  $.$\\		
+		\cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$&  $.$ \\
+
+		\cite{kaspi2019} & $.$ & $.$  & $.$ & $\bigstar$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$&  $.$ \\
+
+		\cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$&  $.$ \\
+		\cite{Vancroonemburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$&  $.$ \\
+		\cite{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$&  $.$ \\
+		\cite{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$&  $.$ \\
+		\cite{PaquaySchynsLimbourg2016,PaquayLimbourgSchynsOliveira2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$&  $.$ \\
+		\cite{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$&  $.$ \\
+		\cite{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$&  $.$  \\
+		\cite{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$&  $.$  \\
+		\cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$&  $.$ \\
+		\cite{MesquitaSanches2023}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$&  $.$  \\
+		{\bf This work}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$&  $\bigstar$  \\
 		\bottomrule 
@@
 Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a pre-defined flight plan restricted to two legs.
-It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem.
+It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modelled through a container sequencing problem.
 Referring directly to this work, \cite[p.
@@
 
-As already mentioned, \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and SPDP in air cargo transport.
+As already mentioned, \cite{LurkinSchyns2015} was the first work that simultaneously modelled WBP and SPDP in air cargo transport.
 The authors demonstrated that this problem is NP-hard and performed some experiments with real data, noting that their model offers better results than those obtained manually.
@@
 
-\cite{PaquaySchynsLimbourg2016} presented a mathematical modeling to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
+\cite{PaquaySchynsLimbourg2016} presented a mathematical modelling to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
 Its objective is to maximise the volume used in containers, considering load balancing restrictions, the presence of fragile items and the possibility of rotating these boxes.
@@
 
-\cite{YangLiuGao2018} modeled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing in order to minimise fuel consumption.
+\cite{YangLiuGao2018} modelled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing in order to minimise fuel consumption.
 
@@
 
-\cite{zhao2021} proposed a new modeling for WBP based on mixed integer programming.
+\cite{zhao2021} proposed a new modelling for WBP based on mixed integer programming.
 Instead of focusing on the center of gravity (CG) deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
 
-As can be seen, none of these works address air cargo palletization and load balancing with route optimization in a multi-leg flight plan.
-This is the objective of our work: to model and elaborate heuristics for a real problem that simultaneously involves 4 intractable sub-problems: APP, WBP, SPDP and TSP.
+\cite{MesquitaSanches2023} contributed with a complex model and elaborate heuristics for the ACLP-RPDP that simultaneously solve 4 intractable sub-problems: APP, WBP, SPDP and TSP.
+They also compare the performances of four well known heuristics: Ant Colony Optimization, the Noising Method Optimization, the Greedy Randomized Adaptive Search Procedure, and Tabu Search.
+
+As can be seen, except for \cite{MesquitaSanches2023}, the remaining works do not address air cargo palletization and load balancing with route optimization in a multi-leg flight plan, and none of them employ parallelization into their algorithms.
+This is the objective of our work: to reshape \cite{MesquitaSanches2023} solution process and algorithms to accommodate parallel features which can improve solution quality and performance.
 
@@
 
-Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modeling of ACLP+RPDP, which is one of the contributions of our work.
+Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modelling of ACLP+RPDP, which is one of the contributions of our work.
 
@@
 
-The mathematical modeling of this problem is described in the equations below.
+The mathematical modelling of this problem is described in the equations below.
 
@@
 
-Once the assumptions of this work and the mathematical modeling of the problem are presented, it is easy to see that ACLP+RPDP is NP-hard.
+Once the assumptions of this work and the mathematical modelling of the problem are presented, it is easy to see that ACLP+RPDP is NP-hard.
 In a similar way to \cite[p.
@@
 Real cases are more complex as they have hundreds of different items in each node and involve three intractable sub-problems: APP, WBP and SPDP.
-Through the mathematical modeling presented in the previous section, we verify that {\it Mixed-Integer Programming}\/ (MIP) is not able to solve these cases in feasible time.
+Through the mathematical modelling presented in the previous section, we verify that {\it Mixed-Integer Programming}\/ (MIP) is not able to solve these cases in feasible time.
 Thus, it is necessary to adopt some strategy to find a viable solution, not necessarily optimal, that seeks to maximize the objective function $f$.
@@
 
-As we are dealing with a new problem, which until now had not been modeled in the literature, we have to create our own benchmarks.
+As we are dealing with a new problem, which until now had not been modelled in the literature, we have to create our own benchmarks.
 For this, we based on the characteristics of real airlifts carried out by the {\em Brazilian Air Force}, as described below.
@@
 \multirow{3}{*}{\it {mpACO}}   &$1.2$& 10      & 8     & 10    & 6     & 10    & 10   \\ 
-                               &$1.5$& 6       & 1     & 0     & 0     & 0     & 0    \\ 
-                               &$2.0$& 0       & 0     & 0     & 0     & 0     & 0    \\                               
+                               &$1.5$& 10      & 6     & 12    & 8     & 10    & 8    \\
+                               &$2.0$& 10      & 8     & 2     & 10    & 6     & 6    \\                               
 \bottomrule		
@@
 Table \ref{tab:numProcs} presents the best numbers of parallel processes for each method, scenario, and volume surplus.
-As all numbers are less than or equal to 8, any hand-held computer with 8 cores can yield efficient solutions.
+As all numbers are less than or equal to 12, any hand-held computer with 8 cores can yield efficient solutions, as normally each core may be capable of 2 threads.
+
+To find these values we submitted to test 1, 2, 4, 6, 8, 10, 12, 14, 16 parallel processes.
 
@@
 %\scriptsize
-\begin{tabular}{rccccccc|c}
+\begin{tabular}{rccccccccc}
 \toprule
-                                &                &\multicolumn{6}{c}{\bf Scenarios} &\\
-\cmidrule{3-8}		
-{$method$}                      &{\bf Results}   &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 6}&\specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
+                                &{\bf Volume}         &                  &\multicolumn{6}{c}{\bf Scenarios}              &{\bf Normalized}\\
+\cmidrule{4-9}		
+{$method$}                      &{\bf surplus}        &{\bf Results}     &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 6}&{\bf Speed-up} \\
 \toprule
-\multirow{2}{*}{\it {mpShims}} & $f$              & 5.88 & 3.16    & 3.16 &      &      &      &   \\ 
-                               &{\bf run-time (s)}& 1    & 10      & 41   &      &      &      &   \\
-\midrule
-
-\multirow{2}{*}{\it {mpACO}}   & $f$              & 5.82 & 3.12    &      &      &      &      &   \\ 
-                               &{\bf run-time (s)}& 22   & 473     &      &      &      &      &   \\ 
+\multirow{6}{*}{\it {mpShims}} &\multirow{2}{*}{$1.2$}& $f$              & 5.70  & 3.09  & 2.94  & 2.90  & 2.79  & 5.70  &   \\
+                               &                      &{\bf run-time (s)}& 1     & 3     & 7     & 26    & 97    & 267   &   \\
+\cmidrule{2-10}                               
+                               &\multirow{2}{*}{$1.5$}& $f$              & 8.15  & 4.22  & 4.40  & 4.01  & 3.51  & 7.08  &\\
+                               &                      &{\bf run-time (s)}& 1     & 5     & 12    & 45    & 244   & 737   &\\
+\cmidrule{2-10}                                 
+                               &\multirow{2}{*}{$2.0$}& $f$              &&&&&&&\\
+                               &                      &{\bf run-time (s)}&&&&&&&\\
+\midrule
+
+\multirow{6}{*}{\it {mpACO}}   &\multirow{2}{*}{$1.2$}& $f$              &  &     &      &      &      &      &   \\
+                               &                      &{\bf run-time (s)}&  &     &      &      &      &      &   \\
+\cmidrule{2-10}                               
+                               &\multirow{2}{*}{$1.5$}& $f$              &&&&&&&\\
+                               &                      &{\bf run-time (s)}&&&&&&&\\
+\cmidrule{2-10}                               
+                               &\multirow{2}{*}{$2.0$}& $f$              &&&&&&&\\
+                               &                      &{\bf run-time (s)}&&&&&&&\\
 \bottomrule		
 \end{tabular}
-
 \normalsize
-
 \end{table}
@@
 \section{Conclusions}
+\label{sec7}
 
@@
 
-This research was partially supported by \textit{S\~{a}o Paulo Research Foundation} (FAPESP, grant 2016/01860-1).
+This research was partially supported by \textit{São Paulo Research Foundation} (FAPESP, grant 2016/01860-1).
 

### 2022-11-19 aco.py

@@ -36,7 +36,7 @@ def getDeltaTau(score, bestSoFar):
 # by all ants proportionally to their solution quality and is evaporated in all the components.
 # each Ant makes use of "updatePheroAttract" to update pheromones
 # and edge attractiveness according to the its solution value
-def updatePheroAttract(score, bestSoFar, edges, mpAttract):
+def updatePheroAttract(score, bestSoFar, edges):
 
     deltaTau = getDeltaTau(score, bestSoFar)
 
@@ -57,7 +57,7 @@ def updatePheroAttract(score, bestSoFar, edges, mpAttract):
         # update the general attractiveness
         edges[id].updateAttract(ALPHA, BETA)
 
-        mpAttract[id] = edges[id].Attract
+        # mpAttract[id] = edges[id].Attract
 
     # mean = stats.tmean(Pheromone)
     # sdev = stats.tstd(Pheromone)

### 2022-11-19 methods.py

@@ -279,6 +279,15 @@ def getTours(num, costs, threshold):
 
     return tours2
 
+bestNPP = [
+[6, 4, 4, 4, 2, 2,],
+[4, 4, 4, 4, 4, 4],
+[4, 4, 4, 2, 2, 4],
+[10, 8, 10, 6, 10, 10],
+[10, 6, 12, 8, 10, 8],
+[10, 8, 2, 10, 6, 6]
+]
+
 def mountEdges(pallets, items, cfg, Alpha=1, Beta=4):
    
     # items include kept on board, are from this node (k), and destined to unattended nodes
@@ -287,7 +296,7 @@ def mountEdges(pallets, items, cfg, Alpha=1, Beta=4):
 
     edges = [None for _ in np.arange(m*n)]
 
-    mpAttract = mp.Array('d', [0.5 for _ in np.arange(m*n)])
+    # mpAttract = mp.Array('d', [0.5 for _ in np.arange(m*n)])
 
     i = 0
     for p in pallets:           
@@ -304,7 +313,7 @@ def mountEdges(pallets, items, cfg, Alpha=1, Beta=4):
     for i, e in enumerate(edges):
         edges[i].ID = i
 
-    return edges, mpAttract
+    return edges
     
 def loadPallets(cfg):
     """

### 2022-11-19 testing.py

@@ -14,6 +14,7 @@ method   = f"{sys.argv[1]}"
 mno.DATA = f"{sys.argv[2]}"
 
 scenarios = [1,2,3,4,5,6]
+# scenarios = [1]
 bests = []
 
 for scenario in scenarios:
@@ -67,7 +68,7 @@ for scenario in scenarios:
     totElapsed = 0
     sumScores = 0
 
-    numProcs = [1,2,4,6,8,10]
+    numProcs = [1,2,4,6,8,10,12,14,16]
 
     if method == "Shims" or method == "ACO" or method == "Greedy":
         numProcs = [1]

### 2022-11-20 latex/comp_or.tex
Paper 2 update

@@
 		\For {each $method$}
+
+			\State $numProcs \gets $, the number of parallel processes to be used
+
 			\For {each $\pi \in S_K$} \label{main:loop1}
-				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method )$ \label{main:method}
+				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, numProcs )$ \label{main:method}
 			\EndFor \label{main:loop2}
@@
 \begin{algorithm}[H]
-	\caption{$SolveTour(\pi, L, M, C, N, method)$}  \label{alg:tour}
+	\caption{$SolveTour(\pi, L, M, C, N, method, numProcs)$}  \label{alg:tour}
 
@@
 			\State $M \gets SetPalletsDestination( \pi(k) )$ \label{tour:dest2}		
-			\State $G_2 \gets SolveNode(method, \pi(k), G_1)$ \label{tour:node}
+			\State $G_2 \gets SolveNode(method, \pi(k), G_1, numProcs)$ \label{tour:node}
 			\State $s, \epsilon \gets ScoreAndDeviation(\pi(k), G_2)$ \label{tour:analyse}
@@
 
-	\caption{ $SolveNode(mpACO, k, G)$ } \label{alg:aco}
+	\caption{ $SolveNode(mpACO, k, G, numProcs)$ } \label{alg:aco}
 	\begin{algorithmic}[1]
@@
 Figure \ref{fig:whip} represents the $n_k$\/ possible edges $e_{ij}^k$\/ of $p_i$\/ sorted by $\theta_{ij}^k$.
-\emph{Shims}\/ starts with a greedy solution, stopping at the edge with index $\eta_1[i]$ close to the local optimum (first phase).
+\emph{mpShims}\/ starts with a greedy solution, stopping at the edge with index $\eta_1[i]$ close to the local optimum (first phase).
 Then, considering even the edge with the index $\eta_2[i]$, it elaborates different possible complements for this pallet (second phase) and selects the best of these complements (third phase).
 
-%Examples:
-%$\to$ for a small problem if limit is 0.85 for the first phase of the solution process, remains 0.15 for second and third phases.
-%$\to$ for a big problem if it is 0.98 for the first phase, it is 0.02 for the remaining phases.
-
 \begin{algorithm}[H]
-	\caption{ $SolveNode(Shims, k, G)$}  \label{alg:shims}
+	\caption{ $SolveNode(mpShims, k, G, numProcs)$}  \label{alg:shims}
 	\begin{algorithmic}[1]
@@
 		\State Let $E[1..m][1..n_k]$, where $E[i]$ is an array with $n_k$ edges $(p_i,t^k_j)$ sorted by $\theta_{ij}^k$ in non-ascending order, $1 \leq i \leq m$ \label{shims:edges}
-		\State $gap \gets 2000/n_k*m$ \label{shims:gap} 
+
+		\State $limits \gets getLimits(0.75,\ 0.98,\ numProcs)$, a list of unique limits, 1 for each process, between 0.75 and 0.98
+
 		\For{$i \gets 1$ to $m$} 
@@
 \cmidrule{2-10}                                 
-                               &\multirow{2}{*}{$2.0$}& $f$              &&&&&&&\\
-                               &                      &{\bf run-time (s)}&&&&&&&\\
+                               &\multirow{2}{*}{$2.0$}& $f$              & 11.31 & 6.29  & 5.54  & 5.83  & 5.05  & 10.63 &\\
+                               &                      &{\bf run-time (s)}& 2     & 14    & 34    & 106   & 477   & 2417  &\\
 \midrule

### 2022-11-20 methods.py

@@ -148,6 +148,7 @@ class Solution(object):
  
         return True
 
+# gets a list of unique limits, 1 for each process, greater than the minimum defined limit
 def getLimits(minLim, numProcs): # pid: 0 - numProcs-1
 
     limitSet = set()
@@ -278,15 +279,6 @@ def getTours(num, costs, threshold):
     toursInt = None
 
     return tours2
-
-bestNPP = [
-[6, 4, 4, 4, 2, 2,],
-[4, 4, 4, 4, 4, 4],
-[4, 4, 4, 2, 2, 4],
-[10, 8, 10, 6, 10, 10],
-[10, 6, 12, 8, 10, 8],
-[10, 8, 2, 10, 6, 6]
-]
 
 def mountEdges(pallets, items, cfg, Alpha=1, Beta=4):
    

### 2022-11-20 latex/comp_or.tex
mpACO algorithm update

@@
 		\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^*(V_k,E^{N*}_k \cup E^Q_k)$
-		\State $A \gets$ a number of ants or parallel processes to be determined for each iteration \label{aco:ants}
+		\State $A \gets numProcs$ \label{aco:numProcs}
 		\State $ga^k_{ij}   \gets 0$, $1 \leq i \leq m$, $1 \leq j \leq n_k$  \label{aco:init1}
@@
 		\State $stagnant \gets 0$
-		\While{($stagnant < 3$) {\bf and} ($runtime < 0.7s$)} \label{aco:while}
-
-		\State $publish$
-		\State $solutions \gets subscribe()$
-
-			\For{$a \gets 1$ to $A$} \label{aco:g1}
-				\State $G^a(V_k, E^{Na}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$
-				\State $E \gets \{ (p_i,t^k_j) \mid (p_i,t^k_j) \not\in E^{Na}_k \}$
-				\While{$|E| > 0$}
-					\State $e \gets Roulette(E)$ biased through $ga^k_{ij}$  \label{aco:roulette}
-					\State $E \gets E - \{e\}$		
-					\If{$E^{Na}_k \cup \{e\}$ is feasible}
-						\State $ E^{Na}_k \gets E^{Na}_k \cup \{e\} $ \label{aco:ant2}
-					\EndIf	
+		\While{ $stagnant < 3$ } \label{aco:while}
+
+			\State $AntsQueue \gets \{\ \}$ \label{aco:AntsQueue}
+
+			\For{$1$ {\bf to} $A$} \label{aco:AntsSolve}	
+				\State $AntSolve(G^+,\ AntsQueue,\ ga^k_{ij})$
+			\EndFor
+
+			\State $counter \gets A$
+
+			\While{$runtime < 0.7s$ {\bf and} $counter > 0$} \label{aco:while2}
+
+				\State $G^{a} \gets AntsQueue.Pop()$	
+				\If{$f_s(G^{a}) > f_s(G^{+})$}
+					\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^a(V_k, E^{Na}_k \cup E^Q_k)$ 
+				\EndIf
+
+				\State $counter \gets counter - 1$
+
 			\EndWhile
-				\If{$f_s(G^{a}) > f_s(G^{+})$} \label{aco:gen1}
-					\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^a(V_k, E^{Na}_k \cup E^Q_k)$  \label{aco:gen2}
-						\For{each $(p_i,t^k_j) \in E^{N+}_k$}
-							\State $\tau^k_{ij} \gets \tau^k_{ij} - \Delta\tau^k_{ij}$ \label{aco:phero1}
-							\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{aco:ga1}
-						\EndFor
-				\EndIf 
-			\EndFor \label{aco:g2}
+
 			\If{$f_s(G^+) > f_s(G^*)$} \label{aco:best1}
@@
 				\State $G^*(V_k, E^{N*}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$		
-				\For{each $(p_i,t^k_j) \in E^{N*}_k$}
-					\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{aco:rho1}
-					\State $\tau^k_{ij} \gets \tau^k_{ij} + \Delta\tau^k_{ij}$ \label{aco:phero2}
-					\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{aco:ga2}
-				\EndFor			
 			\Else
-				\State $stagnant \gets stagnant + 1$ \label{aco:best2}
+				\State $stagnant \gets stagnant + 1$ 
 			\EndIf
@@
 				\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{aco:rho2}
+				\State $\tau^k_{ij} \gets \tau^k_{ij} + \Delta\tau$ \label{aco:phero2}
+				\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{aco:ga2}				
 			\EndFor	
@@
 A greedy solution is obtained in line \ref{aco:greedy}, being initially stored in $G^*$\/ and $G^+$.
-The number $A$\/ of ants was empirically set to 4 (line \ref{aco:ants}), yielding solutions with a good relation quality/time.
-The $m \times n_k$\/ matrices $\tau^k_{ij}$ and $ga^k_{ij}$\/ (lines \ref{aco:init1}-\ref{aco:init2}) are used and updated by all ants.
-
-In each generation, $A$\/ ants are created (lines \ref{aco:g1}-\ref{aco:g2}).
-The number of generations depends on the time limit or stagnation in improving results (line \ref{aco:while}).
-Each ant $a$\/ elaborates its allocation graph $G^a$, considering the current values of $ga^k_{ij}$.
-To enforce solution diversification, if $(p_i,t^k_j)$ is included in $G^a$\/ then pheromone $\tau^k_{ij}$\/ is decreased (line \ref{aco:phero1}).
-Similarly, the values of $ga^k_{ij}$\/ are also updated (line \ref{aco:ga1}), where the exponents $\alpha$\/ and $\beta$\/ were chosen empirically: $\alpha$ was set to 1.0 and $\beta$\/ to $4.0$\/ (lines \ref{aco:alpha}-\ref{aco:beta}).
-For further explanation, see \cite{DorigoManiezzoColorni1996}.
-
-The best solution of each generation is stored in $G^+$\/ (lines \ref{aco:gen1}-\ref{aco:gen2}), and the update of $G^*$\/ and the stopping criterion are performed on lines \ref{aco:best1}-\ref{aco:best2}.
-Each edge $(p_i,t^k_j)$\/ in some solution has its corresponding pheromone updated in a cumulative way (line \ref{aco:phero2}), that is, more pheromone will be given to the most relevant edges, where the increment is defined in \ref{eq:delta_tau}.
-The same with $ga^k_{ij}$\/ (line \ref{aco:ga2}).
+The number $A$\/ of ants was set to the number of parallel processes (line \ref{aco:numProcs}).
+
+The $m \times n_k$\/ matrices $\tau^k_{ij}$ and $ga^k_{ij}$\/ (lines \ref{aco:init1}-\ref{aco:init2}) are used and updated by all ants. For further explanation, see \cite{DorigoManiezzoColorni1996}.
+
+The number of iterations depends on the time limit or stagnation in improving results (line \ref{aco:while}).
+At the beginning os each iteration (\ref{aco:AntsQueue}) a queue of ants ($AntsQueue$) is created to hold ants parallel solutions generated in the child processes.
+
+The local solution $G^+$ is sent to A child processes (\ref{aco:AntsSolve}) almost simultaneously.
+While there are any ant working on an solution or the run-time is still less then maximum (\ref{aco:while2}), it hangs in the $AntsQueue$ until there is a solution available.
+
+If there is a global improvement (\ref{aco:best1}) the best so far solution ($G^*(V_k, E^{N*}_k \cup E^Q_k)$) is updated.
+At the end of each iteration pheromone levels (\ref{aco:rho2}) are evaporated, than updated with the new calculated $\Delta\tau$ (line \ref{aco:phero2}), and according to Equation \ref{eq:delta_tau}, and the general attractiveness ( \ref{aco:ga2}) is also updated.
 
 \begin{equation} \label{eq:delta_tau}
-	\Delta\tau^k_{ij} = \frac{ ga^k_{ij}\times t^k_j.s }{ A \times t^k_j.v}
-\end{equation}
-
-Pheromone $\tau^k_{ij}$\/ evaporates (lines \ref{aco:rho1} and \ref{aco:rho2}), where $\rho$\/ is a evaporation rate, which can be seen as a mechanism that delays premature convergence towards a sub-optimal solution.
-$\rho$\/ was empirically set to $0.2$\/ (line \ref{aco:rho}).
+	\Delta\tau = \frac{f_s(G^{+}) - f_s(G^{*})}{A \times f_s(G^{*})}
+\end{equation}
+
+Many ACO applications update pheromone levels and the general attractiveness only when there is a solution improvement.
+Differently, this work updates in both cases, exploring also the negative learning approach proposed by \cite{math9040361}.
+
+\begin{algorithm}[H] 
+	\caption{ $AntSolve(G^+,\ AntsQueue,\ ga^k_{ij})$ } \label{alg:AntSolve}
+	\begin{algorithmic}[1]
+		\State $G^a(V_k, E^{Na}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$
+		\State $E \gets \{ (p_i,t^k_j) \mid (p_i,t^k_j) \not\in E^{Na}_k \}$
+		\While{$|E| > 0$}
+			\State $e \gets Roulette(E)$ biased through $ga^k_{ij}$  \label{AntSolve:roulette}
+			\State $E \gets E - \{e\}$		
+			\If{$E^{Na}_k \cup \{e\}$ is feasible}
+				\State $ E^{Na}_k \gets E^{Na}_k \cup \{e\} $ \label{AntSolve:ant}
+			\EndIf	
+		\EndWhile
+		\State $AntsQueue \gets AntsQueue \cup G^a(V_k, E^{Na}_k \cup E^Q_k)$
+	\end{algorithmic}
+\end{algorithm}
 

### 2022-11-20 latex/comp_or.tex
Parallel algorithms almost ready

@@
 
-We also developed Algorithm \ref{alg:greedy}, which generates a greedy allocation of the items available in node $l_k$, according to the non-ascending order of $\theta^k_{ij}$, and considering the consolidated items already shipped.
-Edges are included in this allocation as long as they respect feasibility constraints.
-Furthermore, the volume of each pallet $p_i$\/ cannot exceed $p_i.v \times limit$, where $ 0 < limit \leq 1$\/ is a parameter.
-
-\begin{algorithm}[H]
-	\caption{ $Greedy(k, G, limit)$}  \label{alg:greedy}
-
-	\begin{algorithmic}[1]
-		\State Let $G(V_k, E^Q_k)$
-		\State $E^N_k \gets \varnothing$ 			
-		\For{$i \gets 1$ to $m$}
-			\State $pv[i] \gets 0$ 		
-		\EndFor
-		\For{$q \gets 1$ to $m_k$}
-			\If{$(p_i, a_q^k) \in E^Q_k$} 
-				\State $pv[i] \gets pv[i] + a_q^k.v$ 
-			\EndIf		
-		\EndFor		
-		\For {each $e_{ij}^k$ in non-ascending order of $\theta^k_{ij}$}
-			\If{($E^N_k \cup \{e_{ij}^k\}$ is feasible) {\bf and} ($pv[i] \leq p_i.v \times limit$)} 
-				\State $E^N_k \gets E^N_k \cup \{e_{ij}^k\}$ 
-				\State $pv[i] \gets pv[i] + t_j^k.v$ 	
-			\EndIf
-		\EndFor
-		\State ${\bf return}\ G(V_k, E^N_k \cup E^Q_k)$
-
-	\end{algorithmic}
-\end{algorithm}
-
 To compare allocation graphs generated by the heuristics, we choose the one that offer a higher associated score, since the CG deviation is limited by the restrictions.
@@
 		\State Let $G(V_k, E^Q_k)$
-		\State $G^*(V_k, E^{N*}_k \cup E^Q_k) \gets Greedy(k, G, 1)$  \label{aco:greedy}
+		\State $G^*(V_k, E^{N*}_k \cup E^Q_k), \_ \gets Greedy(k, G, 0.9)$  \label{aco:greedy}
 		\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^*(V_k,E^{N*}_k \cup E^Q_k)$
@@
 A greedy solution is obtained in line \ref{aco:greedy}, being initially stored in $G^*$\/ and $G^+$.
+The argument 0.9 generates an initial and poor solution for ACO to improve.
+The second return value denoted by underscore is not used by ACO.
+
 The number $A$\/ of ants was set to the number of parallel processes (line \ref{aco:numProcs}).
@@
 Differently, this work updates in both cases, exploring also the negative learning approach proposed by \cite{math9040361}.
+
+We also developed Algorithm \ref{alg:greedy}, which generates a greedy allocation of the items available in node $l_k$, according to the non-ascending order of $\theta^k_{ij}$, and considering the consolidated items already shipped.
+Edges are included in this allocation as long as they respect feasibility constraints.
+Furthermore, the volume of each pallet $p_i$\/ cannot exceed $p_i.v \times limit$, where $ 0 < limit \leq 1$\/ is a given parameter.
+
+\begin{algorithm}[H]
+	\caption{ $Greedy(k, G, limit)$}  \label{alg:greedy}
+
+	\begin{algorithmic}[1]
+		\State Let $G(V_k, E^Q_k)$
+
+		\State $\eta_1 \gets \{0\} \times m$ \label{greedy:eta}
+		\State $volume \gets \{0\} \times m$
+
+		\State $E^N_k \gets \varnothing$
+
+		\For{$q \gets 1$ to $m$}
+			\If{$(p_i, a_q^k) \in E^Q_k$} 
+				\State $volume[i] \gets volume[i] + a_q^k.v$ 
+			\EndIf		
+		\EndFor
+
+		\For {each $e_{ij}^k$ in non-ascending order of $\theta^k_{ij}$}
+			\If{($E^N_k \cup \{e_{ij}^k\}$ is feasible) {\bf and} ($volume[i] \leq p_i.v \times limit$)} 
+				\State $E^N_k \gets E^N_k \cup \{e_{ij}^k\}$ 
+				\State $volume[i] \gets volume[i] + t_j^k.v$
+				\State $\eta_1[i] \gets \eta_1[i] + 1$ 	
+			\EndIf
+		\EndFor
+		\State ${\bf return}\ G(V_k, E^N_k \cup E^Q_k),\ \eta_1$
+
+	\end{algorithmic}
+\end{algorithm}
+
+In Algorithm \ref{alg:greedy}, line \ref{greedy:eta} defines an array with the same size as pallets array to save the set of last edges inserted in the greedy solution.
 
@@
 In this heuristic, described in Algorithm \ref{alg:shims}, pallets $p_i$\/ are considered in non-descending order of $|p_i.d|$ (line \ref{shims:pallets}).
-For each $p_i$, the $n_k$\/ possible edges $e_{ ij}^k$\/ are considered in non-increasing order of $\theta_{ij}^k$\/ (line \ref{shims:edges}) in three phases: the greedy phase (lines \ref{shims:phase1a}-\ref{shims:phase1b}), the composition phase of the shims (lines \ref{shims:phase2a}-\ref{shims:phase2b}) and the selection phase of the best shim (lines \ref{shims:phase3a}-\ref{shims:phase3b}).
 
 Figure \ref{fig:whip} represents the $n_k$\/ possible edges $e_{ij}^k$\/ of $p_i$\/ sorted by $\theta_{ij}^k$.
-\emph{mpShims}\/ starts with a greedy solution, stopping at the edge with index $\eta_1[i]$ close to the local optimum (first phase).
+\emph{mpShims}\/ starts with a greedy solution, stopping at the edge with index $\eta_1$ close to the local optimum (first phase).
 Then, considering even the edge with the index $\eta_2[i]$, it elaborates different possible complements for this pallet (second phase) and selects the best of these complements (third phase).
@@
 	\begin{algorithmic}[1]
-		\State Let $G(V_k, E^Q_k)$
+
+		\State $G^* \gets G(V_k, E^Q_k)$
+
 		\State Sort $M$ by $|p_i.d|$ in non-descending order \label{shims:pallets}
-		\State $E^N_k \gets \varnothing$ 
+
+		\State $E^N_k \gets \{\}$
+
 		\State Let $E[1..m][1..n_k]$, where $E[i]$ is an array with $n_k$ edges $(p_i,t^k_j)$ sorted by $\theta_{ij}^k$ in non-ascending order, $1 \leq i \leq m$ \label{shims:edges}
@@
 
-		\For{$i \gets 1$ to $m$} 
-			\State $pv[i] \gets 0$		\label{shims:phase1a}
-			\State $\eta_1[i] \gets 0$
-			\While{($p_i.v - pv[i] > gap*p_i.v$) {\bf and} ($\eta_1[i]<n_k$)} \label{shims1:slack}
-				\State $\eta_1[i] \gets \eta_1[i] + 1$ \label{shims:eta1}
-				\State $e^k_{ij} \gets E[i][\eta_1[i]]$
-				\If{($e^k_{ij}$ is feasible) {\bf and} ($t^k_j.v \leq p_i.v - pv[i]$)} 
-					\State $E^N_k \gets E^N_k \cup \{e^k_{ij}\}$
-					\State $pv[i] \gets pv[i] + t^k_j.v$ 
-				\EndIf
+		\State $shimsQueue \gets \{\}$, a queue of shims solutions to be used by the child processes
+
+		\For{$lim \in limits$}
+			\State $shimsSolve(G(V_k, E^Q_k),\ M,\ shimsQueue,\ lim,\ E)$
+		\EndFor
+
+		\State $counter \gets |limits|$
+
+		\While{$runtime < 0.7s$ {\bf and} $counter > 0$}
+
+			\State $G \gets shimsQueue.Pop()$	
+			\If{$f_s(G) > f_s(G^*)$}
+				\State $G^* \gets G$ 
+			\EndIf
+
+			\State $counter \gets counter - 1$
+
 		\EndWhile
-			\State $slack[i] \gets p_i.v - pv[i]$
-			\State $\eta_2[i] \gets \eta_1[i]$
-			\Repeat
-				\State $\eta_2[i] \gets \eta_2[i] + 1$  \label{shims:eta2}
-				\State $e^k_{ij} \gets E[i][\eta_2[i]]$
-				\State $pv[i] \gets pv[i] + t^k_j.v$
-			\Until{($\eta_2[i] = n_k$) {\bf or} ($ pv[i] \geq (1 + 3*gap)*p_i.v)$)} \label{shims:phase1b}
-
-\State $Set \gets \varnothing$  \label{shims:phase2a}
-\State $q \gets 1$
-\State $shim[q] \gets \varnothing$
-\State $volume \gets 0$
-\For{$x \gets \eta_1[i]$ to $\eta_2[i]$}
-\State $e_{ij}^k \gets E[i][x]$
-\If {($e_{ij}^k \not\in E^N_k \cup shim[q]$) {\bf and} ($e_{ij}^k$ is feasible)}
-\If{$t_j^k.v + volume \leq slack[i]$} \label{bincapacity}
-\State $shim[q] \gets shim[q] \cup \{e_{ij}^k\}$	
-\State $volume \gets volume + t_j^k.v$	
-\Else \label{newbins}
-\State $Set \gets Set \cup shim[q] $
-\State $q \gets q+1$
-\State $shim[q] \gets \varnothing$
-\State $volume \gets 0$
-\If{$t_j^k.v \leq slack[i]$} 
-\State $shim[q] \gets shim[q] \cup \{e_{ij}^k\}$	
-\State $volume \gets volume + t_j^k.v$
-\EndIf
-\EndIf
-\EndIf
-\EndFor   \label{shims:phase2b}
-
-			\State $sh_w \gets S$, where $S \in Set$ and $\sum_{e_{ab}^k \in S} t_b^k.w$ is maximum \label{shims:phase3a}
-			\State $sh_v \gets S$, where $S \in Set$ and $\sum_{e_{ab}^k \in S} t_b^k.v$ is maximum
-			\State $sh_{best} \gets S$, where $S \in \{sh_w, sh_v\}$ and $\sum_{e_{ab}^k \in x} t_b^k.s$ is maximum
-			\State $E^N_k \gets E^N_k \cup sh_{best}$ \label{shims:phase3b}
-		\EndFor
-		\State ${\bf return}$ $G(V_k, E^N_k \cup E^Q_k)$
+
+		\State ${\bf return}$ $G^*(V_k, E^N_k \cup E^Q_k)$
 	\end{algorithmic}
@@
 
-In the first phase (lines \ref{shims:phase1a}-\ref{shims:phase1b}), a greedy and partial solution for $p_i$\/ is constructed by edges inclusion, until a certain slack or clearance is reached.
-This slack was empirically defined as $gap * p_i.v$\/ (line \ref{shims1:slack}), where $gap = 2000/n_k*m$\/ (line \ref{shims:gap}).
-
-In the second phase (lines \ref{shims:phase2a}-\ref{shims:phase2b}), a set of shims named $Set$\/ is created, where each shim is formed by a group of edges in the range $[\eta_1[i],\eta_2[i]]$, whose total volume is limited by $slack[i]$.
 In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it Next-Fit}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \cite{JohnsonGarey1985}.
@@
 
-In the third phase (lines \ref{shims:phase3a}-\ref{shims:phase3b}), the best shim in $Set$\/ is chosen.
-Initially, two shims are found: $sh_w$\/ with larger weight and $sh_v$\/ with larger volume.
-Among the two, the shim with the highest score will be chosen and its edges are inserted into $E^N_k$.
+\begin{algorithm}[H]
+	\caption{ $shimsSolve(G,\ M,\ shimsQueue,\ limit,\ E)$}  \label{alg:shimsSolve}
+	\begin{algorithmic}[1]
+
+		\State $G(V_k,\ E^Q_k \cup E^N_k),\ \eta_1 \gets Greedy(k,\ G(V_k,\ E^Q_k),\ limit)$ \label{first_phase}
+
+		\For{$i \gets 1$ to $m$}
+
+			\State $\eta_2[i] \gets \eta_1[i]$
+
+			\State $volume \gets 0$
+
+			\Repeat \label{second_phase_begin}
+				\State $\eta_2[i] \gets \eta_2[i] + 1$ 
+				\State $e^k_{ij} \gets E[i][\eta_2[i]]$
+				\State $volume \gets volume + t^k_j.v$
+			\Until{($\eta_2[i] = n_k$) {\bf or} ($ volume \geq (2 - limit)*p_i.v)$)} \label{second_phase_end}
+
+			\State $slack[i] \gets (1-limit) \times p_i.v$
+
+			\State $E^N_k \gets E^N_k \cup getBestShims(i,\ \eta_1,\ \eta_2,\ E,\ k,\ slack[i])$ \label{third_phase}
+
+		\EndFor
+
+		\State $shimsQueue \gets shimsQueue \cup \{G(V_k, E^N_k \cup E^Q_k)\}$ \label{enqueue}
+
+	\end{algorithmic}
+\end{algorithm}
+
+Algorithm \ref{alg:shimsSolve}, line \ref{first_phase} corresponds to Shims' first phase. Lines  \ref{second_phase_begin}-\ref{second_phase_end} correspond to the Shims' second phase.
+Line \ref{third_phase} corresponds to Shims' third phase. In the last line, \ref{enqueue}, {\it mpShims} puts the solution in the queue to be retrieved by the main process.
+
+\begin{algorithm}[H]
+	\caption{ $getBestShims(i,\ \eta_1,\ \eta_2,\ E,\ k,\ slack)$ }  \label{alg:getBestShims}
+	\begin{algorithmic}[1]
+
+		\State $volume \gets 0$
+
+		\State $b \gets 1$
+		\State $shims[b] \gets \{\}$
+		\State $Set \gets Set \cup \{shims[b]\}$
+
+		\For{$x \gets \eta_1$ to $\eta_2[i]$}
+
+			\State $NewShims \gets$ {\bf True}
+			\State $e_{ij}^k \gets E[i][x]$
+
+			\For{$shims \in Set$}
+
+				\If { $e_{ij}^k \not\in (E^N_k \cup shims)$ {\bf and} $e_{ij}^k$ is feasible  {\bf and}  $(t_j^k.v + volume) \leq slack[i]$}
+
+					\State $shims \gets shims \cup \{e_{ij}^k\}$
+					\State $volume \gets volume + t_j^k.v$
+					\State $NewShims \gets$ {\bf False}
+
+					\State {\bf break}
+
+				\EndIf
+
+			\EndFor
+
+			\If{$NewShims$} \label{ffd:new1}
+				\State $volume \gets 0$
+				\State $b \gets b + 1$
+				\State $shims[b] \gets \{\}$
+				\State $shims[b] \gets shims[b] \cup \{e_{ij}^k\}$
+				\State $Set \gets Set \cup \{shims[b]\}$
+			\EndIf
+
+		\EndFor
+
+		\State $sh_w \gets shims$, where $shims \in Set$ and $\sum_{e_{ab}^k \in shims} t_b^k.w$ is maximum 		
+		\State $sh_v \gets shims$, where $shims \in Set$ and $\sum_{e_{ab}^k \in shims} t_b^k.v$ is maximum
+		\State $sh_{best} \gets shims$, where $shims \in \{sh_w, sh_v\}$ and $\sum_{e_{ab}^k \in x} t_b^k.s$ is maximum
+
+		\State ${\bf return}$ $sh_{best}$
+
+	\end{algorithmic}
+\end{algorithm}
 

### 2022-11-21 latex/comp_or.tex
Exchanged result table layout
Parallel shims described

@@
 
-In this heuristic, described in Algorithm \ref{alg:shims}, pallets $p_i$\/ are considered in non-descending order of $|p_i.d|$ (line \ref{shims:pallets}).
-
 Figure \ref{fig:whip} represents the $n_k$\/ possible edges $e_{ij}^k$\/ of $p_i$\/ sorted by $\theta_{ij}^k$.
-\emph{mpShims}\/ starts with a greedy solution, stopping at the edge with index $\eta_1$ close to the local optimum (first phase).
+\emph{mpShims}\/ starts with a greedy solution, stopping at the edge with index in $\eta_1[i]$ close to the local optimum (first phase).
 Then, considering even the edge with the index $\eta_2[i]$, it elaborates different possible complements for this pallet (second phase) and selects the best of these complements (third phase).
@@
 \begin{algorithm}[H]
-	\caption{ $SolveNode(mpShims, k, G, numProcs)$}  \label{alg:shims}
+	\caption{ $SolveNode(mpShims, k, G, numProcs)$}  \label{alg:mpshims}
 	\begin{algorithmic}[1]
 
-		\State $G^* \gets G(V_k, E^Q_k)$
-
-		\State Sort $M$ by $|p_i.d|$ in non-descending order \label{shims:pallets}
-
-		\State $E^N_k \gets \{\}$
-
-		\State Let $E[1..m][1..n_k]$, where $E[i]$ is an array with $n_k$ edges $(p_i,t^k_j)$ sorted by $\theta_{ij}^k$ in non-ascending order, $1 \leq i \leq m$ \label{shims:edges}
-
-		\State $limits \gets getLimits(0.75,\ 0.98,\ numProcs)$, a list of unique limits, 1 for each process, between 0.75 and 0.98
-
-		\State $shimsQueue \gets \{\}$, a queue of shims solutions to be used by the child processes
+		\State $G^* \gets G(V_k, E^Q_k)$ \label{best_so_far}
+
+		\State Sort $M$ by $|p_i.d|$ in non-descending order \label{mpshims:pallets}
+
+		\State $E^N_k \gets \{\}$ \label{mpshims:items}
+
+		\State Let $E[1..m][1..n_k]$, where $E[i]$ is an array with $n_k$ edges $(p_i,t^k_j)$ sorted by $\theta_{ij}^k$ in non-ascending order, $1 \leq i \leq m$ \label{mpshims:edges}
+
+		\State $limits \gets getLimits(0.75,\ 0.98,\ numProcs)$ \label{mpshims:limits}
+
+		\State $shimsQueue \gets \{\}$ \label{mpshims:queue}
 
 		\For{$lim \in limits$}
-			\State $shimsSolve(G(V_k, E^Q_k),\ M,\ shimsQueue,\ lim,\ E)$
+			\State $shimsSolve(G(V_k, E^Q_k),\ M,\ shimsQueue,\ lim,\ E)$ \label{mpshims:send}
 		\EndFor
@@
 
-			\State $G \gets shimsQueue.Pop()$	
+			\State $G \gets shimsQueue.Pop()$	\label{mpshims:wait}
 			\If{$f_s(G) > f_s(G^*)$}
@@
 
-		\State ${\bf return}$ $G^*(V_k, E^N_k \cup E^Q_k)$
+		\State ${\bf return}$ $G^*(V_k, E^N_k \cup E^Q_k)$ \label{mpshims:return}
 	\end{algorithmic}
@@
 
-In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it Next-Fit}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \cite{JohnsonGarey1985}.
-Basically, shims are created by accumulating the following edges, taking $slack[i]$\/ as a limit.
+Algorithm \ref{alg:mpshims} describes the {\it SolveNode} method when it receives {\it mpShims} as argument.
+
+Line \ref{best_so_far} defines the best solution so far which is formed by a semi-loaded aircraft where pallets have been unloaded and some remain on board ($E^Q_k$) with consolidated items.
+
+The Shims solution starts with the set of pallets closer to the center of gravity (\ref{mpshims:pallets}).
+
+A new empty set of edges is created in line \ref{mpshims:items}.
+
+Edges with items in node "k" to be loaded into not full pallets are defined in line \ref{mpshims:edges}.
+
+Line \ref{mpshims:limits} creates a list of unique limits with values between 0.75 and 0.98, that will be sent to different parallel processes to generate diverse solutions.
+
+A queue of shims solutions to be used by the child processes to send solution to the main process is defined in line \ref{mpshims:queue}.
+
+The iteration in \ref{mpshims:send} publishes or triggers parallel processes to solve $G(V_k, E^Q_k)$.
+
+In line \ref{mpshims:wait} the program hangs until each child process put a solution into the queue. Each solution is collected by the main process and compared with the best so far.
+
+Finally, in line \ref{mpshims:return}, the best solution found is returned.
 
@@
 Algorithm \ref{alg:shimsSolve}, line \ref{first_phase} corresponds to Shims' first phase. Lines  \ref{second_phase_begin}-\ref{second_phase_end} correspond to the Shims' second phase.
-Line \ref{third_phase} corresponds to Shims' third phase. In the last line, \ref{enqueue}, {\it mpShims} puts the solution in the queue to be retrieved by the main process.
+Line \ref{third_phase} corresponds to Shims' third phase. In the last line, \ref{enqueue}, {\it mpShims} puts the solution into the queue to be retrieved by the main process.
 
@@
 		\State $b \gets 1$
-		\State $shims[b] \gets \{\}$
-		\State $Set \gets Set \cup \{shims[b]\}$
-
-		\For{$x \gets \eta_1$ to $\eta_2[i]$}
-
-			\State $NewShims \gets$ {\bf True}
+		\State $shims[b] \gets \{\}$  \label{empty_shims}
+		\State $Set \gets Set \cup \{shims[b]\}$ \label{empty_set}
+
+		\For{$x \gets \eta_1[i]$ to $\eta_2[i]$} \label{edges_indexes}
+
+			\State $NewShims \gets$ {\bf True} \label{new_shims}
 			\State $e_{ij}^k \gets E[i][x]$
 
-			\For{$shims \in Set$}
+			\For{$shims \in Set$} \label{shims_set}
 
@@
 					\State $volume \gets volume + t_j^k.v$
-					\State $NewShims \gets$ {\bf False}
+					\State $NewShims \gets$ {\bf False} \label{new_shims_false}
 
@@
 
-			\If{$NewShims$} \label{ffd:new1}
+			\If{$NewShims$} \label{new_shims2}
 				\State $volume \gets 0$
@@
 
-		\State $sh_w \gets shims$, where $shims \in Set$ and $\sum_{e_{ab}^k \in shims} t_b^k.w$ is maximum 		
-		\State $sh_v \gets shims$, where $shims \in Set$ and $\sum_{e_{ab}^k \in shims} t_b^k.v$ is maximum
-		\State $sh_{best} \gets shims$, where $shims \in \{sh_w, sh_v\}$ and $\sum_{e_{ab}^k \in x} t_b^k.s$ is maximum
+		\State $sh_w \gets shims$, where $shims \in Set$ and $\sum_{e_{ab}^k \in shims} t_b^k.w$ is maximum  \label{best_weight}		
+		\State $sh_v \gets shims$, where $shims \in Set$ and $\sum_{e_{ab}^k \in shims} t_b^k.v$ is maximum \label{best_volume}
+		\State $sh_{best} \gets shims$, where $shims \in \{sh_w, sh_v\}$ and $\sum_{e_{ab}^k \in x} t_b^k.s$ is maximum \label{best_score}
 
@@
 
+Algorithm \ref{alg:getBestShims} describes the procedure do get the best Shims, the best subset of edges to fill each pallet gap.
+This algorithm follows the logic of the {\it First-Fit Decreasing} algorithm as described by \cite{JohnsonGarey1985}.
+
+Line \ref{empty_shims} creates the first empty shims.
+Line \ref{empty_set} creates the first set of shims from where the best shims will be chosen.
+
+Line \ref{edges_indexes} iterate in the edges indexes range to find subsets of shims that fit into each pallet gap.
+
+Initially, new shims creation is permitted (\ref{new_shims}), but it will only be created if the last edge could not be included in any shims from the set (\ref{new_shims2}).
+
+For each edge, iterate in all sets in a try to include it in any of the previous shims (\ref{shims_set}).
+As the last edge was included, forbid a new shims creation (\ref{new_shims_false}).
+
+Finally, select the best weight shims (\ref{best_weight}), the best volume shims (\ref{best_volume}), and the best score shims (\ref{best_score}) between these 2.
+The best score shims edges will be returned.
+
 \section{Implementation and results}
@@
 
-This section is composed of two parts: the generation of test instances and the results obtained in our implementations.
-
-\subsection{Items generation}
-\label{items}
-
-As we are dealing with a new problem, which until now had not been modelled in the literature, we have to create our own benchmarks.
-For this, we based on the characteristics of real airlifts carried out by the {\em Brazilian Air Force}, as described below.
-
-In the military airlift carried out in Brazil from 2008 to 2010, 23\% of the items weighed between $10kg$ and $20kg$, 22\% from $21kg$ to $40kg$, 24\% from $41kg$ to $80kg$, 23\% from $81kg$ to $200kg$, and 8\% between $201kg$ and $340kg$.
-These five groups of items are described in Table \ref{tab:weights}.
-On the other hand, the average density of these items is approximately $246 kg/m^3$.
-
-\begin{table}[H]
-		\centering
-		\caption{Items weight distribution ($kg$) in Brazil}  \label{tab:weights}
-		\begin{tabular}{c c c c }
-			\toprule
-			$item$ & $P$ & $low$ & $high$  \\		
-			\midrule
-			1              & 0.23           & 10  & 20\\
-			2              & 0.22           & 21  & 40\\
-			3              & 0.24           & 41  & 80 \\		
-			4              & 0.23           & 81  & 200\\
-			5              & 0.08           & 201 & 340 \\
-			\bottomrule
-		\end{tabular}
-\end{table}
-
-$ItemsGeneration$, which generates $N$, is described in Algorithm \ref{alg:itemsgen}.
-The parameter $scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and the parameter $volume$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}).
-To avoid simply loading all items, we use $volume > 1$.
-
-\begin{algorithm}[H]
-	\caption{$ItemsGeneration(scenario,volume)$}  \label{alg:itemsgen}
-	\begin{algorithmic}[1]
-		\State Let $L$ and $M$ be according to $scenario$ \label{ig:LM}
-		\State $limit \gets volume \times \sum_{i=1}^{m} p_i.v$ \label{ig:extended}
-		\For{$k \gets 0$ to $K$}
-			\State $N_k \gets \varnothing$
-			\State $j \gets 0$
-			\State $vol \gets 0$	\label{ig:totals}	
-			\While{$vol < limit$}
-				\State $j \gets j+1$
-				\Repeat
-					\State $t_j^k.to \gets RandomInt(0, K)$ \label{ig:dest}
-				\Until{$t_j^k.to \neq k$}
-				\State $x = Roulette(item)$ biased through $P$ \label{ig:weight1}	
-				\State $t_j^k.w \gets RandomReal(low(x), high(x))$        \label{ig:weight2}		
-				\State $t_j^k.s \gets \round{100 \times (1 - \log_{10}(RandomInt(1, 9)))} $ \label{ig:score}
-				\State $t_j^k.v \gets t_j^k.w / RandomReal(148, 344)$ \label{ig:volume}
-				\State $vol \gets vol + t_j^k.v$ 
-				\State $N_k \gets N_k \cup \{t_j^k\}$ 
-			\EndWhile
-			\State $n_k \gets j$
-		\EndFor
-		\State $N \gets \bigcup_{0 \leq k \leq K} N_k$ 
-		\State ${\bf return}$ $N$
-	\end{algorithmic}
-\end{algorithm}
-
-For each generated $t^k_j$\/ item, its destination is uniformly random selected (line \ref{ig:dest}), its weight has a distribution according to Table \ref{tab:weights} (lines \ref{ig:weight1}-\ref{ig:weight2}), its score varies $100$\/ (highest) and $5$\/ (lowest) according to a logarithmic scale (line \ref{ig:score}, and its volume is randomly defined from the density, where we allow a variation of 40\% more or less than the average density of $246 kg/m^3$\/ (line \ref{ig:volume}).
-
-We test our methods with 126 different instances because we have six operational scenarios with seven randomly generated item sets for three volume surpluses (1.2, 1.5, and 2.0 times aircraft volume capacities).
+As we are dealing with a new problem, which has been modeled by \cite{MesquitaSanches2023}, we use the same benchmarks of their work, which are based on the characteristics of real airlifts carried out by the {\em Brazilian Air Force}.
+
+We test our methods with 126 different instances: six operational scenarios with 2 aircraft sizes and 3 to 7 nodes; and seven randomly generated item sets for three volume surpluses (1.2, 1.5, and 2.0 times aircraft volume capacities).
 
@@
 	\centering
-\caption{Methods and scenarios with $1.2 \times volume$}  \label{tab:20}
+	\caption{Volume surpluses, methods and scenarios results}  \label{tab:results}
 	\footnotesize
 	%\scriptsize
-\begin{tabular}{rccccccccc}
+	\begin{tabular}{cccccccccc}
 		\toprule
-                                &{\bf Volume}         &                  &\multicolumn{6}{c}{\bf Scenarios}              &{\bf Normalized}\\
+		              &            &                  &\multicolumn{6}{c}{\bf Scenarios}              &{\bf Normalized}\\
 		\cmidrule{4-9}		
-{$method$}                      &{\bf surplus}        &{\bf Results}     &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 6}&{\bf Speed-up} \\
+		{\bf Surplus} & {$method$} &{\bf Results}     &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 6}&{\bf Speed-up} \\
 		\toprule
-\multirow{6}{*}{\it {mpShims}} &\multirow{2}{*}{$1.2$}& $f$              & 5.70  & 3.09  & 2.94  & 2.90  & 2.79  & 5.70  &   \\
-                               &                      &{\bf run-time (s)}& 1     & 3     & 7     & 26    & 97    & 267   &   \\
+
+		\multirow{4}{*}{$1.2$} &\multirow{2}{*}{\it {mpShims}} & $f$                & 5.70  & 3.09  & 2.94  & 2.90  & 2.79  & 5.70  & 0.91  \\%2312
+		                       &                               & {\bf run-time (s)} & 1     & 3     & 7     & 26    & 97    & 267   & 70  \\%390
 		                       \cmidrule{2-10}
-                               &\multirow{2}{*}{$1.5$}& $f$              & 8.15  & 4.22  & 4.40  & 4.01  & 3.51  & 7.08  &\\
-                               &                      &{\bf run-time (s)}& 1     & 5     & 12    & 45    & 244   & 737   &\\
+		                       &\multirow{2}{*}{\it {mpACO}}   & $f$                & 6.42  & 3.78  &  3.19 & 3.17  & 2.97  & 5.87  & 1.00  \\ %2540
+                               &                               & {\bf run-time (s)} & 16    & 282   &  314  & 1334  & 6945  & 18883 & 1  \\ %27162
+		\midrule
+		\multirow{4}{*}{$1.5$} &\multirow{2}{*}{\it {mpShims}} & $f$                & 8.15  & 4.22  & 4.40  & 4.01  & 3.51  & 7.08  &   \\
+		                       &                               & {\bf run-time (s)} & 1     & 5     & 12    & 45    & 244   & 737   &   \\
 		\cmidrule{2-10}
-                               &\multirow{2}{*}{$2.0$}& $f$              & 11.31 & 6.29  & 5.54  & 5.83  & 5.05  & 10.63 &\\
-                               &                      &{\bf run-time (s)}& 2     & 14    & 34    & 106   & 477   & 2417  &\\
-\midrule
-
-\multirow{6}{*}{\it {mpACO}}   &\multirow{2}{*}{$1.2$}& $f$              &  &     &      &      &      &      &   \\
-                               &                      &{\bf run-time (s)}&  &     &      &      &      &      &   \\
+		                       &\multirow{2}{*}{\it {mpACO}}   & $f$                &&&&&&&\\
+		                       &                               & {\bf run-time (s)} &&&&&&&\\
+		\midrule
+		\multirow{4}{*}{$2.0$} &\multirow{2}{*}{\it {mpShims}} & $f$                & 11.31 & 6.29  & 5.54  & 5.83  & 5.05  & 10.63 &   \\
+		                       &                               & {\bf run-time (s)} & 2     & 14    & 34    & 106   & 477   & 2417  &   \\
 		\cmidrule{2-10}
-                               &\multirow{2}{*}{$1.5$}& $f$              &&&&&&&\\
-                               &                      &{\bf run-time (s)}&&&&&&&\\
-\cmidrule{2-10}                               
-                               &\multirow{2}{*}{$2.0$}& $f$              &&&&&&&\\
-                               &                      &{\bf run-time (s)}&&&&&&&\\
+		                       &\multirow{2}{*}{\it {mpACO}}   & $f$                &&&&&&&\\
+		                       &                               & {\bf run-time (s)} &&&&&&&\\
+
 		\bottomrule		

### 2022-11-21 latex/comp_or.tex
Paper update

@@
 
+\usepackage{float}% http://ctan.org/pkg/float
+
 %----- end celio
@@
 
-In the scenario of aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading for rapid take-off and mission accomplishment, especially in times of crisis, public calamity, client contract short lead times, or any external pressure for immediate take-off.
-Also, there is no technological assistance to help the load and trip planners with the huge number of demands for transport in each hub.
-This enables other risks such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
-We contributed by modelling and solving a new problem of planning the loading and routing of an aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
-This new hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modelled using standardized pallets in fixed positions, obeying the center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
-We also contributed by carrying out multiple experiments with the well-known Ant Colony Optimization on synthetic data based on real data from the {\it Brazilian Air Force}\/ transportation history.
-These challenging benchmark instances are made publicly available.
+Transport aviation faces risks of cargo imbalance in the set of circumstances of aerial pickup and shipment of goods in a distribution system given the urgency required for loading for fast takeoff and mission accomplishment, notably in moments of crisis, public emergency, client contract short turnaround times, or any external forces for instant takeoff.
+Additionally, there is no computer support to aid load and trip programmers with the massive transportation demands in each location.
+
+Other dangers are made possible as a result, including poor delivery, excessive fuel consumption, and longer turnaround times than necessary.
+
+We developed an efficient solution by means of computer parallelism to the problem involving scheduling the loading and routing of an airplane in a tour of simultaneous pickup and delivery at intermediate hubs while taking into account a utility score, weight and balance rules, and fuel usage.
+
+This hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem} considers using standardized pallets in fixed positions, obeying the center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
+
+We also contributed by carrying out multiple experiments with the well-known Ant Colony Optimization on synthetic data based on real data from the {\it Brazilian Air Force}\/ transportation history and a new procedure to minimize the distance from the next node destined pallet to the ramp door.
+
 We also created a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as { \color{red} it was the unique method that managed to solve all testing scenarios.}
-
-% Reescrever
 
@@
 Air cargo transport involves several sub-problems that are difficult to solve.
-Recently, \cite[p.
-401]{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP) and {\it Weight and Balance Problem} (WBP).
-Several aspects were considered in this modelling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc.
-They also presented a comprehensive bibliographic survey of solving methods that had been developed in different situations.
-
-However, there are still other important challenges in air cargo transport that go beyond the definition of the ACLPP, especially with regard to the flight itinerary and the loading and unloading at each destination (or node) of this travel plan.
-In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Simultaneous Pickup and Delivery Problem} (SPDP), and the search for the most profitable route, which is the well-known {\it Traveling Salesman Problem} (TSP), similar to the proposed by \cite{kaspi2019}, where the profit per unit time is defined as the profit gained after completing the tour divided by the total time required to complete the tour.
-
-\cite{MesquitaSanches2023} explored these problems by solving them in an integrated form.
+Recently, \cite{MesquitaSanches2023} modeled and solved the {\it Air Cargo Load Planning with Routing, Pickup and Delivery} (ACLP+RPDP) composed by four sub-problems: {\it Build-up Scheduling Problem}, {\it Air Palletization Problem} (APP), {\it Weight and Balance Problem} (WBP), and a special case of the {\it Traveling Salesman Problem}, similar to the proposed by \cite{kaspi2019}, where the profit per unit time is defined as the profit gained after completing the tour divided by the total time (cost in our case) required to complete the tour.
+
+However, there are still other important challenges in air cargo transport that go beyond the definition of the ACLP+RPDP, especially with regard to algorithms performance and the easiness of loading operations at each destination.
 
@@
 		\cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$&  $.$ \\
+
 		\cite{MesquitaSanches2023}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$&  $.$  \\
+
 		{\bf This work}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$&  $\bigstar$  \\
@@
 
-This work proposes a method that attends these objectives: a pallet building and arrangement plan with a routing option that maximizes the benefit-cost ratio for the smooth execution of pickup and delivery transport missions.
-We develop a heuristic that can be executed on a simple handheld computer (like a laptop or a tablet) and that provides a solution quickly enough to keep this cargo handling time under an hour.
-This heuristic will reduce the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets, and picking up and delivering at each node.
-To the best of our knowledge, this is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP and TSP has been addressed.
-This new problem is named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem} (ACLP+RPDP).
+\cite{MesquitaSanches2023} proposed a method that attends these objectives: a pallet building and arrangement plan with a routing option that maximizes the benefit-cost ratio for the smooth execution of pickup and delivery transport missions.
+They develop a heuristic that can be executed on a simple handheld computer (like a laptop or a tablet) and that provides a solution quickly enough to keep this cargo handling time under an hour.
+This work seeks to improve their results in terms of quality and performance by the use of computer parallelism and by minimizing the distance between the next node destined pallets and the cargo door.
+
+These improvements in their heuristics are meant to reduce the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets, and picking up and delivering at each node.
+To the best of our knowledge, that is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP and TSP has been addressed, and the first time ACLP+RPDP is solved with computer parallelism.
 
@@
 \cite{MongeauBes2003} presented a method based on linear integer programming to solve the problem of choosing and positioning containers on the {\it Airbus 340-300}.
-Safety and stability restrictions were considered, with the objective of minimizing fuel consumption.
+Safety and stability constraints were considered, with the objective of minimizing fuel consumption.
 
@@
 
-As already mentioned, \cite{LurkinSchyns2015} was the first work that simultaneously modelled WBP and SPDP in air cargo transport.
+As already mentioned, \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and SPDP in air cargo transport.
 The authors demonstrated that this problem is NP-hard and performed some experiments with real data, noting that their model offers better results than those obtained manually.
@@
 
-\cite{PaquaySchynsLimbourg2016} presented a mathematical modelling to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
-Its objective is to maximise the volume used in containers, considering load balancing restrictions, the presence of fragile items and the possibility of rotating these boxes.
+\cite{PaquaySchynsLimbourg2016} presented a mathematical modeling to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
+Its objective is to maximise the volume used in containers, considering load balancing constraints, the presence of fragile items and the possibility of rotating these boxes.
 \cite{PaquayLimbourgSchynsOliveira2018} developed some heuristics to solve this problem.
@@
 \cite{wong2020}A developed a mathematical model and a tool based on mixed integer programming for optimizing cargo in aircraft with different pallet configurations.
-Balance restrictions and the presence of dangerous items were considered.
+Balance constraints and the presence of dangerous items were considered.
 \cite{eugene2021} integrated this tool to a digital simulation model, with a visualization and validation system, based on sensors that alert about load deviations.
@@
 \cite{MesquitaSanches2023} contributed with a complex model and elaborate heuristics for the ACLP-RPDP that simultaneously solve 4 intractable sub-problems: APP, WBP, SPDP and TSP.
-They also compare the performances of four well known heuristics: Ant Colony Optimization, the Noising Method Optimization, the Greedy Randomized Adaptive Search Procedure, and Tabu Search.
-
-As can be seen, except for \cite{MesquitaSanches2023}, the remaining works do not address air cargo palletization and load balancing with route optimization in a multi-leg flight plan, and none of them employ parallelization into their algorithms.
-This is the objective of our work: to reshape \cite{MesquitaSanches2023} solution process and algorithms to accommodate parallel features which can improve solution quality and performance.
+They also compare the performances of four well known heuristics: {\it Ant Colony Optimization}, the {\it Noising Method Optimization}, the {\it Greedy Randomized Adaptive Search Procedure}, and {\it Tabu Search}.
+They also create a new heuristic called {\it Shims} which is fast and may be run in a handheld computer.
+They did not use computer parallelism to improve performance, nor tryed to minimize the distances from the next nodes destined pallets to the cargo ramp door.
+
+As can be seen, except for \cite{MesquitaSanches2023}, the remaining works do not address air cargo palletization and load balancing with route optimization in a multi-leg flight plan, and none of them employ parallelization in their algorithms.
+This is the objective of our work: to reshape \cite{MesquitaSanches2023} solution process and algorithms to accommodate parallel features that can improve solution quality and performance, and also include a new feature that is to minimize the distances between the next node's destined pallets and the cargo ramp door.
 
@@
 
+	\item Pallets with destinations set to the next node should be put as near as possible to the cargo ramp door.
+
 	\item Finally, as we are interested in minimizing fuel costs while keeping the CG in its operational range, we disregarded some lower ones as handling costs.
@@
 See Figure \ref{fig:larger2}.
-To ensure accuracy in pickup and delivery operations, consolidated items must remain on board until their destination.
+Consolidated items must stay on board until they reach their destination to maintain accuracy in pickup and delivery procedures.
 
@@
 	\centering
-	\includegraphics[scale=0.4]{Images/large.png}
-	\caption{Consolidated items on pallets inside a {\it Boeing C-17}}
-	\small\textsuperscript{Source: https://www.sacprogram.org/en/Pages/Boeing-C-17-Globemaster-III.aspx}
+	\includegraphics[scale=0.4]{Images/consolidated.jpg}
+	\caption{Consolidated items on 463L pallets inside a {\it Boeing C-17}}
+	\small\textsuperscript{Source: From Wikimedia Commons, the free media repository}
 	\label{fig:larger2}
@@
 
-We consider real scenarios with a smaller or a larger aircraft with payloads of $26,000 kg$ or $75,000 kg$ respectively.
+We consider real scenarios with a smaller or a larger aircraft with payloads of 26,000 kg or 75,000 kg respectively.
 Both layouts are represented in Figures \ref{fig:smaller} and \ref{fig:larger}, where the pallets are identified by $p_i$.
@@
 
-In both cases, the torque applied to the aircraft must keep its CG in the operational range, which corresponds to a percentage of the {\it Mean Aerodynamic Chord} \footnote{Chord is the distance between the leading and trailing edges of the wing, measured parallel to the normal airflow over the wing.
+In both cases, the torque applied to the aircraft must keep its CG in the operational range, which corresponds to a percentage of the {\it Mean Aerodynamic Chord} \footnote{Chord is the distance between the leading and trailing edges of the wing, measured parallel to the normal airflow over the wing \cite[p.18]{HoughtonCarpenter2003}.
 The average length of the chord is known as the {\it Mean Aerodynamic Chord} (MAC).}: $0.556m$ in the smaller aircraft and $1.17m$ in the larger one.
@@
 	\centering
-	\includegraphics[scale=0.22]{Images/lateral.png}
+	\includegraphics[scale=0.24]{Images/lateral.png}
 	\caption{Aircraft longitudinal cut showing red lines as pallets}
@@
 
-We also make the following assumptions:
+We accept the same assumptions as stated by \cite{MesquitaSanches2023}:
 \begin{itemize}
@@
 		& The total weight must be less than the aircraft payload or the total pallet capacity, whichever is the lowest.\\	
+		& The pallets destined for the next node of the tour should be put as near as possible to the cargo ramp door (our inclusion).\\		
 		\midrule
@@
 
-Our strategy is based on the fact that, in real cases, $K$\/ is usually small.
-Specifically, we will consider $K \leq 6$\/ throughout this work, which is a higher value than usual in {\it Brazilian Air Force} missions.
+\cite{MesquitaSanches2023} strategy is based on the fact that, in real cases, $K$\/ is usually small.
+Specifically, they will consider $K \leq 6$\/ throughout their and our work, which is a higher value than usual in {\it Brazilian Air Force} missions.
 As a result, if we have fast node-by-node solutions that allow us to construct a complete tour, we will be able to test all possible $K!$\/ tours and thus select the one that provides the best value for the $f$\/ function.
 
-On the other hand, our tactic will be, at each shipping node, to predefine the destinations of the pallets at that node.
+The tactic will be, at each shipping node, to predefine the destinations of the pallets at that node.
 In this way, we will reserve a number of pallets proportional to the volume demanded by each destination at the shipping node.
 We could have used another criterion, but it was observed in the experiments that volume is more constrictive in airlift.
-Once the destinations of the pallets are defined, we will use serial and multiprocessing heuristics to find the best possible node-by-node solutions.
-This strategy is summarized in Algorithm \ref{alg:main}.
+Once the destinations of the pallets are defined, we will use serial and Multi-processing heuristics to find the best possible node-by-node solutions.
+This strategy is summarized in Algorithm \ref{alg:main}, retrieved from \cite{MesquitaSanches2023}.
 
@@
 The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $l_k$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n_k} t^k_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.v$).
-This parameter is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).
+This parameter is passed to $ItemsGeneration$ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section.
 
@@
 	\centering
-	\caption{Testing scenarios}  \label{tab:scenarios}
+	\caption{Testing scenarios retrieved from \cite{MesquitaSanches2023}.}  \label{tab:scenarios}
 	\begin{tabular}{c c c c }
@@
 
-$UpdateConsolidated$, described in Algorithm \ref{alg:cons}, finds the best allocation for the consolidated items that remain on board.
-Initially, the set $Q$\/ is created, with the consolidated items that did not reach their destination (lines \ref{cons:remain1}-\ref{cons:remain2}).
-Then $MinCGDeviation$\/ (line \ref{cons:minCG}) is run through a MIP solver with the equations \ref{eq:torque}, \ref{eq:embarked} and \ref {eq:one}, to relocate the consolidated items on the pallets minimizing torque and ensuring that they all remain on board, one on each pallet.
-As there are few variables, the MIP solver returns an allocation $E^Q_k$\/ very quickly.
-Finally, the destination of each pallet with consolidated items is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).
-
-\begin{algorithm}[H]
-	\caption{$UpdateConsolidated(k)$}  \label{alg:cons}
-	\begin{algorithmic}[1]
-
-		\State Let $Q_k  = \{ a^k_1, a^k_2, \ldots, a^k_{m_k} \}$ 
-		\State $Q \gets \varnothing$ \label{cons:remain1}
-		\For{$q\gets1$ to $m_k$} 
-			\If {$a_q^k.to \in L_k$} 
-				\State $Q \gets Q \cup \{a_q^k\}$
-			\EndIf
-		\EndFor	 
-		\State $m_k \gets |Q|$ \label{cons:remain2}
-		\State $E^Q_k \gets MinCGDeviation(k, Q)$ \label{cons:minCG}
-		\For{$i \gets 1$ to $m$} \label{cons:Ybegin}
-			\For{$q \gets 1$ to $m_k$}
-				\If{$(p_i, a_q^k) \in E^Q_k$} 
-					\State $p_i.to[k] \gets a_q^k.to$
-				\EndIf
-			\EndFor		
-		\EndFor \label{cons:Yend}
-		\State ${\bf return}\ Q, E^Q_k, M$ 
-	\end{algorithmic}
-\end{algorithm}
-
-\begin{equation} \label{eq:torque}
-	\mbox{minimize}\ | \sum_{i=1}^{m} \sum_{q=1}^{m_k} Y^k_{iq} \times p_i.d \times a_q^k.w |,\ \mbox{where}\ Q  = \{ a^k_1, a^k_2, \ldots, a^k_{m_k} \} 
-\end{equation}
-
-\begin{equation} \label{eq:embarked}
-	s.t.: \sum_{i=1}^{m} Y^k_{iq} = 1;\ q \in \{1,2,\ldots,m_k\}
-\end{equation}
-
-\begin{equation} \label{eq:one}
-	s.t.: \sum_{q=1}^{m_k} Y^k_{iq} \leq 1;\ i \in \{1,2,\ldots,m\}
-\end{equation}
-
-$SetPalletsDestination$, which set pallets destination not yet defined, is described in Algorithm \ref{alg:dest}.
-The vectors $vol[0..K]$\/ and $PalCons[0..K]$\/ store the total volume of items and the number of pallets with consolidated items destined for each node, respectively (lines \ref{dest:vector1}-\ref{dest:vector2}).
-The destinations of each pallet are defined proportionally to the volume of items, regarding the pallets with consolidated items (lines \ref{dest:propor1}-\ref{dest:propor2}).
-The node with the maximum volume demand defines the destination of any remaining pallets (lines \ref{dest:max1}-\ref{dest:max2}).
-
-\begin{algorithm}[H]
-	\caption{$SetPalletsDestination(k)$}  \label{alg:dest}
-	\begin{algorithmic}[1]
-
-		\For{$x \gets 0$ to $K$} \label{dest:vector1}
-			\State $vol[x] \gets 0$ 
-			\State $PalCons[x]  \gets 0$
-		\EndFor
-		\State $max \gets 0$ \label{dest:max}
-		\State $total \gets 0$
-
-		\For{$j \gets 1$ to $n_k$}
-			\State $d \gets t_j^k.to$
-			\If {$d \in L_k$} 
-				\State $vol[d] \gets vol[d] + t_j^k.v$ 
-				\State $total \gets total + t_j^k.v$ 
-				\If {$vol[d] > vol[max]$}
-					\State $max \gets d$ 
-				\EndIf
-			\EndIf
-		\EndFor
-		\For{$q \gets 1$ to $m_k$}
-			\State $d \gets a_q^k.to$
-			\State $PalCons[d] \gets PalCons[d]+1$
-			\If {$d \in L_k$} 
-				\State $vol[d] \gets vol[d] + a_q^k.v$
-				\State $total \gets total + a_q^k.v$ 
-				\If {$vol[d] > vol[max]$}
-					\State $max \gets d$ 
-				\EndIf
-			\EndIf
-		\EndFor \label{dest:vector2}
-		\For{$x \gets 0$ to $K$} \label{dest:propor1}
-			\If {$vol[x] \neq 0$}
-				\State $quant \gets \max \{1, \lfloor{ m \times vol[x]/total}\rfloor \}-PalCons[x]$ 
-				\State $np \gets 0$
-				\For{$i \gets 1$ to $m$}
-
-					\If{$np = quant$}
-						\State ${\bf break}$
-					\EndIf
-					\If{$p_i.to[k] = -1$}
-						\State $p_i.to[x] \gets x$
-						\State $np \gets np + 1$
-					\EndIf
-				\EndFor
-
-			\EndIf 
-		\EndFor \label{dest:propor2}
-		\For{$i \gets 1$ to $m$} \label{dest:max1}
-			\If{$p_i.to[k] = -1$}
-				\State $p_i.to[x] \gets max$
-			\EndIf
-		\EndFor \label{dest:max2}
-		\State ${\bf return}\ M$
-
-	\end{algorithmic}
-\end{algorithm}
-
-$ScoreAndDeviation$\/ is described in Algorithm \ref{alg:eval}, which evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.
-It consists of a loop that goes through all the pallets (lines \ref{eval:loop1}-\ref{eval:loop2}), accumulating the scores (lines \ref{eval:score1}-\ref{eval:score2} ) and the torques (lines \ref{eval:eps1}-\ref{eval:eps2}) of the shipped items, allowing the final calculation of the CG deviation (lines \ref{eval:torque1}-\ref{eval:torque2}).
-
-\begin{algorithm}[H]
-	\caption{$ScoreAndDeviation(k, G)$}  \label{alg:eval}
-
-	\begin{algorithmic}[1]
-		\State Let $G(V_k, E^N_k \cup E^Q_k)$
-		\State Let $X_{ij}^k = 1$ if $(p_i, t_j^k) \in E^N_k$, $1 \leq i \leq m$, $1 \leq j \leq n_k$
-		\State Let $Y_{iq}^k = 1$ if $(p_i, a_q^k) \in E^Q_k$, $1 \leq i \leq m$, $1 \leq q \leq m_k$
-		\State $s \gets 0$
-		\State $\tau \gets 0$
-		\State $weight \gets 0$
-		\For{$i \gets 1$ to $m$}	\label{eval:loop1}
-			\State $weight \gets weight + p_i.w$		
-			\For{$j \gets 1$ to $n_k$}
-				\If{$X_{ij}^k = 1$}
-					\State $s \gets s + t_{ij}^k.s$ \label{eval:score1}
-					\State $\tau \gets \tau + t_{ij}^k.w * p_i.d$ \label{eval:eps1}
-				\EndIf
-			\EndFor				
-			\For{$q \gets 1$ to $m_k$}
-				\If{$Y_{iq}^k = 1$}
-					\State $s \gets s + a_{iq}^k.s$   \label{eval:score2}
-					\State $\tau \gets \tau + a_{iq}^k.w * p_i.d$  \label{eval:eps2}
-				\EndIf
-			\EndFor	   \label{eval:loop2}
-		\EndFor
-		\State $maxWeight \gets \min(Payload, weight)$ \label{eval:torque1}
-		\State $maxTorque \gets maxWeight \times limit^{CG}_{long}$
-		\State $\epsilon \gets \tau/maxTorque$ \label{eval:torque2}
-		\State ${\bf return}\ s, \epsilon$	
-	\end{algorithmic}
-\end{algorithm}
+$UpdateConsolidated$ finds the best allocation for the consolidated items that remain on board. Initially, the set $Q$ is created, with the consolidated items that did not reach their destination.
+
+Then $MinCGDeviation$ is run through a MIP solver to relocate the consolidated items on the pallets minimizing torque and ensuring that they all remain on board, one on each pallet.
+As there are few variables, the MIP solver returns an allocation $E^Q_k$ very quickly.
+
+$SetPalletsDestination$ sets pallets destination not yet defined.
+The total volume of items and the number of pallets with consolidated items destined for each node.
+The destinations of each pallet are defined proportionally to the volume of items, regarding the pallets with consolidated items.
+
+$ScoreAndDeviation$ evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.
+It consists of a loop that goes through all the pallets, accumulating the scores and the torques of the shipped items, allowing the final calculation of the CG deviation.
+
+$UpdateConsolidated$, $MinCGDeviation$, $SetPalletsDestination$, and $ScoreAndDeviation$ are thoroughly described by \cite{MesquitaSanches2023}.
 
@@
 
-The selection of edges for $E^N_k$\/ uses the {\it edge attractiveness}\/ $\theta^k_{ij}$\/ \ref{eq:edge}, which can be understood as the tendency to allocate the item $t^k_j$\/ to the pallet $p_i$.
+The selection of edges for $E^N_k$\/ uses the {\it edge attractiveness}\/ $\theta^k_{ij}$, Equation \ref{eq:edge}, which can be understood as the tendency to allocate the item $t^k_j$\/ to the pallet $p_i$.
 It is directly proportional to the score, and inversely proportional to the volume and the torque of each item.
-The 3000 value was empirically set to keep the attractiveness in the range [0, 1], to be compatible with the Ant Colony Optimization pheromone range.
+The 3000 value was empirically set to keep the attractiveness in the range [0, 1], to be in the same order of magnitude of the Ant Colony Optimization pheromone range.
 
@@
 
-To compare allocation graphs generated by the heuristics, we choose the one that offer a higher associated score, since the CG deviation is limited by the restrictions.
-
-In the pseudo-codes presented below, given an allocation graph $G$\/ for the node $l_k$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation(k,G) $, described in Algorithm \ref{alg:eval}.
-
-Below, we present 2 implementations for $SolveNode$: {\it Multiprocess Ant Colony Optimization}\/ ({\it mpACO})  and a new parallel heuristic that we propose called {\it Multiprocess Shims (mpShims)}.
+To compare allocation graphs generated by the heuristics, we choose the one that offer a higher associated score, since the CG deviation is limited by the constraints.
+
+In the pseudo-codes presented below, given an allocation graph $G$\/ for the node $l_k$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation(k,G) $, described in \cite{MesquitaSanches2023}.
+
+Below, we present 2 implementations for $SolveNode$: {\it Multi-process Ant Colony Optimization}\/ ({\it mpACO})  and a new parallel heuristic that we propose called {\it Multi-process Shims (mpShims)}.
 
@@
 It is important to say that the parallel ACO ({\it pACO}) is a different implementation where employs threads instead of processes to solve each ant solution in parallel (for a detailed information on parallel ACO see \cite{Pedemonte2011}).
-We chose {\it mpACO} for our approach because of the need to share the pheromone array between ants/processes and the facility to handle the possibility of race condition.
-
-\subsubsection{Multiprocess Ant Colony Optimization (mpACO)}
-
-In ACO \cite{Dorigo1992}, a population of $A$\/ ants performs independent search, where each ant $a$\/ finds its own allocation graph $G^a$.
+We chose {\it mpACO} for our approach because we employed the {\it Fully-Connected} topology which requires to share the pheromone array between the multi-process ants and the facility to handle the possibility of race condition.
+
+\subsubsection{Multi-process Ant Colony Optimization (mpACO)}
+
+In ACO \cite{Dorigo1992}, a population of $A$\/ ants performs independent search (sequential or in parallel), where each ant $a$\/ finds its own allocation graph $G^a$.
 Each edge $e^k_{ij}=(p_i,t^k_j)$\/ has a general attractiveness $ga^k_{ij} = (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$, where $\tau^k_{ij}$\/ is the pheromone of this edge, and $\alpha$\/ and $\beta$\/ are constants.
 If $e^k_{ij}$\/ is included in $G^a$, ant $a$\/ increases the pheromone $\tau^k_{ij}$.
-Algorithm \ref{alg:aco} presents this heuristic.
+
+Algorithm \ref{alg:aco} presents this heuristic in its parallel multi-process version as we have implemented.
+
+$SolveNode(mpACO, k, G, numProcs)$ is the main process, while $AntSolve(G^+,\ AntsQueue,\ ga^k_{ij})$ is executed the child concurrent processes.
 
@@
 	\begin{algorithmic}[1]
-		\State Let $G(V_k, E^Q_k)$
-		\State $G^*(V_k, E^{N*}_k \cup E^Q_k), \_ \gets Greedy(k, G, 0.9)$  \label{aco:greedy}
+		\State Let $G(V_k, E^Q_k)$ \Comment{This is a partial solution with only consolidated on board}
+		\State $G^*(V_k, E^{N*}_k \cup E^Q_k), \_ \gets Greedy(k, G, 0.9)$  \label{aco:greedy} \Comment{This partial solution is completed by a greedy algorithm}
 		\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^*(V_k,E^{N*}_k \cup E^Q_k)$
@@
 Differently, this work updates in both cases, exploring also the negative learning approach proposed by \cite{math9040361}.
-
-We also developed Algorithm \ref{alg:greedy}, which generates a greedy allocation of the items available in node $l_k$, according to the non-ascending order of $\theta^k_{ij}$, and considering the consolidated items already shipped.
+We did not compare these performances because it is out of the scope of this work.
+
+We also developed Algorithm \ref{alg:greedy}, which generates a greedy allocation of the items available in node $l_k$, according to the non-ascending order of $\theta^k_{ij}$, and considering the consolidated items already shipped ($E^Q_k$).
 Edges are included in this allocation as long as they respect feasibility constraints.
@@
 In Algorithm \ref{alg:greedy}, line \ref{greedy:eta} defines an array with the same size as pallets array to save the set of last edges inserted in the greedy solution.
+The set $\eta_1$ is returned together with the greedy solution.
 
@@
 		\State $G^a(V_k, E^{Na}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$
-		\State $E \gets \{ (p_i,t^k_j) \mid (p_i,t^k_j) \not\in E^{Na}_k \}$
-		\While{$|E| > 0$}
+		\State $E \gets \{ (p_i,t^k_j) \mid (p_i,t^k_j) \not\in E^{Na}_k \}$ \label{AntSolve:neighborhood}
+		\While{$|E| > 0$} \label{AntSolve:not_empty}
 			\State $e \gets Roulette(E)$ biased through $ga^k_{ij}$  \label{AntSolve:roulette}
@@
 		\EndWhile
-		\State $AntsQueue \gets AntsQueue \cup G^a(V_k, E^{Na}_k \cup E^Q_k)$
+
+\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{AntSolve:rho}
+\State $\tau^k_{ij} \gets \tau^k_{ij} + \Delta\tau$ \label{AntSolve:phero}
+\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{AntSolve:ga}
+
+		\State $AntsQueue \gets AntsQueue \cup G^a(V_k, E^{Na}_k \cup E^Q_k)$ \label{AntSolve:put_in_queue}
 	\end{algorithmic}
@@
 
-\subsubsection{Multiprocess Shims - mpShims}
+Algorithm \ref{alg:AntSolve}, which will be executed by a child parallel process, receives an initial and poor solution ($G^+(V_k, E^{N+}_k \cup E^Q_k)$) generated greedily to be completed and improved by {\it mpACO}.
+
+Line \ref{AntSolve:neighborhood} expresses the creation of the neighborhood, the set of candidate edges.
+
+While the neighborhood is not empty (\ref{AntSolve:not_empty}), choose an edge (\ref{AntSolve:roulette}), delete it from the neighborhood and include it in the solution, if feasible (\ref{AntSolve:ant}).
+
+Finally, put the solution in the queue (\ref{AntSolve:put_in_queue}) to be retrieved by the main process.
+
+\subsubsection{Multi-process Shims - mpShims}
 
@@
 
-For parallelization, we used the {\it Python multiprocessing} package, which implements {\it process-based parallelism} or {\it parallel shared memory algorithm}.
-According to \cite{multiprocessing}, {\it Multiprocessing} is a package that supports spawning processes using an API similar to the {\it Python Threading} package.
+For parallelization, we used the {\it Python Multi-processing} package, which implements {\it process-based parallelism} or {\it parallel shared memory algorithm}.
+According to \cite{Multi-processing}, {\it Multi-processing} is a package that supports spawning processes using an API similar to the {\it Python Threading} package.
 
@@
 It is known that working concurrently opens up synchronization issues.
-But the {\it Multiprocessing} package (mp) offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
-Because of this, the {\it Multiprocessing} module lets the programmer take full advantage of the fact that a machine has more than one processor, normally capable of 2 threads each.
+But the {\it Multi-processing} package (mp) offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
+Because of this, the {\it Multi-processing} module lets the programmer take full advantage of the fact that a machine has more than one processor, normally capable of 2 threads each.
 

### 2022-11-22 latex2/comp_or.tex
Paper 2 update

--- 2022-11-21 latex/comp_or.tex
@@
 }
+
+\usepackage{float}% http://ctan.org/pkg/float
 
@@
 
-$UpdateConsolidated$, described in Algorithm \ref{alg:cons}, finds the best allocation for the consolidated items that remain on board.
-Initially, the set $Q$\/ is created, with the consolidated items that did not reach their destination (lines \ref{cons:remain1}-\ref{cons:remain2}).
-Then $MinCGDeviation$\/ (line \ref{cons:minCG}) is run through a MIP solver with the equations \ref{eq:torque}, \ref{eq:embarked} and \ref {eq:one}, to relocate the consolidated items on the pallets minimizing torque and ensuring that they all remain on board, one on each pallet.
-As there are few variables, the MIP solver returns an allocation $E^Q_k$\/ very quickly.
-Finally, the destination of each pallet with consolidated items is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).
-
-\begin{algorithm}[H]
-	\caption{$UpdateConsolidated(k)$}  \label{alg:cons}
-	\begin{algorithmic}[1]
-
-		\State Let $Q_k  = \{ a^k_1, a^k_2, \ldots, a^k_{m_k} \}$ 
-		\State $Q \gets \varnothing$ \label{cons:remain1}
-		\For{$q\gets1$ to $m_k$} 
-			\If {$a_q^k.to \in L_k$} 
-				\State $Q \gets Q \cup \{a_q^k\}$
-			\EndIf
-		\EndFor	 
-		\State $m_k \gets |Q|$ \label{cons:remain2}
-		\State $E^Q_k \gets MinCGDeviation(k, Q)$ \label{cons:minCG}
-		\For{$i \gets 1$ to $m$} \label{cons:Ybegin}
-			\For{$q \gets 1$ to $m_k$}
-				\If{$(p_i, a_q^k) \in E^Q_k$} 
-					\State $p_i.to[k] \gets a_q^k.to$
-				\EndIf
-			\EndFor		
-		\EndFor \label{cons:Yend}
-		\State ${\bf return}\ Q, E^Q_k, M$ 
-	\end{algorithmic}
-\end{algorithm}
-
-\begin{equation} \label{eq:torque}
-	\mbox{minimize}\ | \sum_{i=1}^{m} \sum_{q=1}^{m_k} Y^k_{iq} \times p_i.d \times a_q^k.w |,\ \mbox{where}\ Q  = \{ a^k_1, a^k_2, \ldots, a^k_{m_k} \} 
-\end{equation}
-
-\begin{equation} \label{eq:embarked}
-	s.t.: \sum_{i=1}^{m} Y^k_{iq} = 1;\ q \in \{1,2,\ldots,m_k\}
-\end{equation}
-
-\begin{equation} \label{eq:one}
-	s.t.: \sum_{q=1}^{m_k} Y^k_{iq} \leq 1;\ i \in \{1,2,\ldots,m\}
-\end{equation}
-
-$SetPalletsDestination$, which set pallets destination not yet defined, is described in Algorithm \ref{alg:dest}.
-The vectors $vol[0..K]$\/ and $PalCons[0..K]$\/ store the total volume of items and the number of pallets with consolidated items destined for each node, respectively (lines \ref{dest:vector1}-\ref{dest:vector2}).
-The destinations of each pallet are defined proportionally to the volume of items, regarding the pallets with consolidated items (lines \ref{dest:propor1}-\ref{dest:propor2}).
-The node with the maximum volume demand defines the destination of any remaining pallets (lines \ref{dest:max1}-\ref{dest:max2}).
-
-\begin{algorithm}[H]
-	\caption{$SetPalletsDestination(k)$}  \label{alg:dest}
-	\begin{algorithmic}[1]
-
-		\For{$x \gets 0$ to $K$} \label{dest:vector1}
-			\State $vol[x] \gets 0$ 
-			\State $PalCons[x]  \gets 0$
-		\EndFor
-		\State $max \gets 0$ \label{dest:max}
-		\State $total \gets 0$
-
-		\For{$j \gets 1$ to $n_k$}
-			\State $d \gets t_j^k.to$
-			\If {$d \in L_k$} 
-				\State $vol[d] \gets vol[d] + t_j^k.v$ 
-				\State $total \gets total + t_j^k.v$ 
-				\If {$vol[d] > vol[max]$}
-					\State $max \gets d$ 
-				\EndIf
-			\EndIf
-		\EndFor
-		\For{$q \gets 1$ to $m_k$}
-			\State $d \gets a_q^k.to$
-			\State $PalCons[d] \gets PalCons[d]+1$
-			\If {$d \in L_k$} 
-				\State $vol[d] \gets vol[d] + a_q^k.v$
-				\State $total \gets total + a_q^k.v$ 
-				\If {$vol[d] > vol[max]$}
-					\State $max \gets d$ 
-				\EndIf
-			\EndIf
-		\EndFor \label{dest:vector2}
-		\For{$x \gets 0$ to $K$} \label{dest:propor1}
-			\If {$vol[x] \neq 0$}
-				\State $quant \gets \max \{1, \lfloor{ m \times vol[x]/total}\rfloor \}-PalCons[x]$ 
-				\State $np \gets 0$
-				\For{$i \gets 1$ to $m$}
-
-					\If{$np = quant$}
-						\State ${\bf break}$
-					\EndIf
-					\If{$p_i.to[k] = -1$}
-						\State $p_i.to[x] \gets x$
-						\State $np \gets np + 1$
-					\EndIf
-				\EndFor
-
-			\EndIf 
-		\EndFor \label{dest:propor2}
-		\For{$i \gets 1$ to $m$} \label{dest:max1}
-			\If{$p_i.to[k] = -1$}
-				\State $p_i.to[x] \gets max$
-			\EndIf
-		\EndFor \label{dest:max2}
-		\State ${\bf return}\ M$
-
-	\end{algorithmic}
-\end{algorithm}
-
-$ScoreAndDeviation$\/ is described in Algorithm \ref{alg:eval}, which evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.
-It consists of a loop that goes through all the pallets (lines \ref{eval:loop1}-\ref{eval:loop2}), accumulating the scores (lines \ref{eval:score1}-\ref{eval:score2} ) and the torques (lines \ref{eval:eps1}-\ref{eval:eps2}) of the shipped items, allowing the final calculation of the CG deviation (lines \ref{eval:torque1}-\ref{eval:torque2}).
-
-\begin{algorithm}[H]
-	\caption{$ScoreAndDeviation(k, G)$}  \label{alg:eval}
-
-	\begin{algorithmic}[1]
-		\State Let $G(V_k, E^N_k \cup E^Q_k)$
-		\State Let $X_{ij}^k = 1$ if $(p_i, t_j^k) \in E^N_k$, $1 \leq i \leq m$, $1 \leq j \leq n_k$
-		\State Let $Y_{iq}^k = 1$ if $(p_i, a_q^k) \in E^Q_k$, $1 \leq i \leq m$, $1 \leq q \leq m_k$
-		\State $s \gets 0$
-		\State $\tau \gets 0$
-		\State $weight \gets 0$
-		\For{$i \gets 1$ to $m$}	\label{eval:loop1}
-			\State $weight \gets weight + p_i.w$		
-			\For{$j \gets 1$ to $n_k$}
-				\If{$X_{ij}^k = 1$}
-					\State $s \gets s + t_{ij}^k.s$ \label{eval:score1}
-					\State $\tau \gets \tau + t_{ij}^k.w * p_i.d$ \label{eval:eps1}
-				\EndIf
-			\EndFor				
-			\For{$q \gets 1$ to $m_k$}
-				\If{$Y_{iq}^k = 1$}
-					\State $s \gets s + a_{iq}^k.s$   \label{eval:score2}
-					\State $\tau \gets \tau + a_{iq}^k.w * p_i.d$  \label{eval:eps2}
-				\EndIf
-			\EndFor	   \label{eval:loop2}
-		\EndFor
-		\State $maxWeight \gets \min(Payload, weight)$ \label{eval:torque1}
-		\State $maxTorque \gets maxWeight \times limit^{CG}_{long}$
-		\State $\epsilon \gets \tau/maxTorque$ \label{eval:torque2}
-		\State ${\bf return}\ s, \epsilon$	
-	\end{algorithmic}
-\end{algorithm}
+$UpdateConsolidated$ finds the best allocation for the consolidated items that remain on board. Initially, the set $Q$ is created, with the consolidated items that did not reach their destination.
+
+Then $MinCGDeviation$ is run through a MIP solver to relocate the consolidated items on the pallets minimizing torque and ensuring that they all remain on board, one on each pallet.
+As there are few variables, the MIP solver returns an allocation $E^Q_k$ very quickly.
+
+$SetPalletsDestination$, which set pallets destination not yet defined.
+The total volume of items and the number of pallets with consolidated items destined for each node.
+The destinations of each pallet are defined proportionally to the volume of items, regarding the pallets with consolidated items.
+
+$ScoreAndDeviation$ evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.
+It consists of a loop that goes through all the pallets, accumulating the scores and the torques of the shipped items, allowing the final calculation of the CG deviation.
+
+$UpdateConsolidated$, $MinCGDeviation$, $SetPalletsDestination$, and $ScoreAndDeviation$ are well described by \cite{MesquitaSanches2023}.
 
@@
 
-In ACO \cite{Dorigo1992}, a population of $A$\/ ants performs independent search, where each ant $a$\/ finds its own allocation graph $G^a$.
+In ACO \cite{Dorigo1992}, a population of $A$\/ ants performs independent search (sequential or in parallel), where each ant $a$\/ finds its own allocation graph $G^a$.
 Each edge $e^k_{ij}=(p_i,t^k_j)$\/ has a general attractiveness $ga^k_{ij} = (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$, where $\tau^k_{ij}$\/ is the pheromone of this edge, and $\alpha$\/ and $\beta$\/ are constants.
 If $e^k_{ij}$\/ is included in $G^a$, ant $a$\/ increases the pheromone $\tau^k_{ij}$.
-Algorithm \ref{alg:aco} presents this heuristic.
+
+Algorithm \ref{alg:aco} presents this heuristic in its parallel multi-process version as we have implemented.
+
+$SolveNode(mpACO, k, G, numProcs)$ is the main process, while $AntSolve(G^+,\ AntsQueue,\ ga^k_{ij})$ is executed the child concurrent processes.
 
@@
 		\EndWhile
-		\State $AntsQueue \gets AntsQueue \cup G^a(V_k, E^{Na}_k \cup E^Q_k)$
+		\State $AntsQueue \gets AntsQueue \cup G^a(V_k, E^{Na}_k \cup E^Q_k)$ \label{AntSolve:put_in_queue}
 	\end{algorithmic}
 \end{algorithm}
+
+Algorithm \ref{alg:AntSolve}, which will be executed by a child parallel process, receives an initial and poor solution ($G^+(V_k, E^{N+}_k \cup E^Q_k)$) generated greedily to be completed and improved by {\it mpACO}.
+
+Line \ref{AntSolve:neighborhood} expresses the creation of the neighborhood, the set of candidate edges.
+
+While the neighborhood is not empty (\ref{AntSolve:not_empty}), choose an edge (\ref{AntSolve:roulette}), delete it from the neighborhood and include it in the solution, if feasible (\ref{AntSolve:ant}).
+
+Finally, put the solution in the queue (\ref{AntSolve:put_in_queue}) to be retrieved by the main process.
 

### 2022-11-22 latex2/grossmanSPAC_homework.tex
Paper 2 update

\documentclass{article}
\usepackage{graphicx}
\usepackage{fullpage}

\parskip 1ex
\parindent 0pt

% an environment for problems that numbers them
\newcounter{exnum}[section]
\newenvironment{problem}{{\vskip 0.1in
   \noindent \bf Problem\addtocounter{exnum}{1}~\arabic{exnum}.~}}{\vskip 0.1in}

% a blunt instrument for making subproblems be (a), (b), ...
\renewcommand{\labelenumi}{(\alph{enumi})}

\begin{document}

\ 
\vfil
\begin{center}
{\Large Some Homework Problems For the Material in ``A Sophomoric
  Introduction to Shared-Memory Parallelism and Concurrency''}
\vfil
\ 
\vfil
For more information, see
http://www.cs.washington.edu/homes/djg/teachingMaterials.
\end{center}
\vfil

\newpage

\begin{problem}{\bf Fork-Join Parallelism: Longest Series}

Consider the problem of finding the longest sequence of some number in
an array of numbers:\\ {\tt longest\_sequence(i,arr)} returns the
longest number of consecutive {\tt i} in {\tt arr}.  For example, if\\
{\tt arr} is {\tt \{2,17,17,8,17,17,17,0,17,1\}} then
{\tt longest\_sequence(17,arr)} is {\tt 3} and 
{\tt longest\_sequence(9,arr)} is {\tt 0}.

\begin{enumerate}
\item In pseudocode, give a parallel fork-join algorithm for
  implementing {\tt longest\_sequence}.  Your algorithm should have
  work $O(n)$ and span $O(\log n)$ where $n$ is the length of the
  array.  Do \emph{not} employ a sequential cut-off: your base case
  should process an array range containing one element.  Hint: Use
  this definition:
\begin{verbatim}
  class Result {
    int numLeftEdge;
    int numRightEdge;
    int numLongest;
    boolean entireRange;
    Result(int l, int r, int m, boolean a) {
      numLeftEdge=l; numRightEdge=r; numLongest=m; entireRange=a;
    }
  }
\end{verbatim}
For example, {\tt numLeftEdge} should represent the length of the
sequence at the \emph{beginning} of the range processed by a
subproblem.  Think carefully about how to combine results.
\item In English, describe how you would make your answer to part (a)
  more efficient by using a sequential cut-off.  In pseudocode, show
  the code you would use below this cut-off.
\end{enumerate}

\end{problem}

\newpage

\begin{problem}{\bf Fork-Join Parallelism: Leftmost Occurrence of Substring}

Consider the problem of finding the leftmost occurrence of the sequence
of characters {\tt cseRox} in an array of characters, returning the index
of the leftmost occurrence or {\tt -1} if there is none.  For example,
the answer for the sequence {\tt cseRhellocseRoxmomcseRox} is {\tt 9}.

\begin{enumerate}
\item In English (though some high-level pseudocode will probably help),
  describe a fork-join algorithm similar in design to your solution in problem
  1.  Use a sequential cut-off of at least 6 (the
  length of {\tt cseRox}) and explain why this
  significantly simplifies your solution.  Notice you still must deal
  with the leftmost occurrence being ``split'' across two recursive
  subproblems.
\item Give a much simpler fork-join solution to the problem that avoids the
  possibility of a ``split'' by using slightly overlapping
  subproblems.  Assume a larger sequential cut-off, for example 100.
  Give your solution precisely in pseudocode. Avoid off-by-one errors.
\end{enumerate}

\end{problem}

\newpage

\begin{problem}{\bf Amdahl's Law: Graphing the Pain}

  Use a graphing program such as a spreadsheet to plot the following
  implications of Amdahl's Law.  Turn in the graphs and tables with
  the data.

\begin{enumerate}
\item Consider the speed-up ($T_1/T_P$) where $P=256$ of a program
  with sequential portion $S$ where the portion $1-S$ enjoys perfect
  linear speed-up.  Plot the speed-up as $S$ ranges from .01 (1\%
  sequential) to .25 (25\% sequential).
\item Consider again the speed-up of a program with sequential portion $S$
  where the portion $1-S$ enjoys perfect linear speed-up.  This time,
  hold $S$ constant and vary the number of processors $P$ from 2 to
  32.  On the same graph, show three curves, one each for $S=.01$,
  $S=.1$, and $S=.25$.
\end{enumerate}

\end{problem}

\newpage

\begin{problem}{\bf Parallel Prefix and Pack}

  In this problem, the input is an array of strings and the output is
  an array of integers.  The output has the length of each string in
  the input, but empty strings are filtered out.  For example:
\begin{verbatim}
   [ "", "", "cse", "rox", "", "homework", "", "7", "" ]
\end{verbatim}
produces output:
\begin{verbatim}
   [ 3, 3, 8, 1]
\end{verbatim}
  A parallel algorithm can solve this problem in $O(\log n)$ span and
  $O(n)$ work by doing a parallel map to produce a bit vector, followed
  by a parallel prefix over the bit vector, followed by a parallel map
  to produce the output.

  Show the intermediate steps for the algorithm described above on the
  example above.  For each step, show the tree of recursive task
  objects that would be created (where a node's child is for two
  problems of half the size) and the fields each node needs.  Do not
  use a sequential cut-off.  Show three separate trees (for the three
  steps).  Explain briefly what each field represents.

  Note that because the input length is not a power of two, the tree
  will not have all its leaves at exactly the same height.
\end{problem}

\newpage

\begin{problem}{\bf Parallel Quicksort}

  Lecture presented a parallel version of quicksort with
  \emph{best-case} $O(\log^2 n)$ span and $O(n\log n)$ work.  This
  algorithm used parallelism for the two recursive sorting calls and
  the partition.

  \begin{enumerate}
  \item For the algorithm from lecture, what is the asymptotic \emph{worst-case}
  span and work.  Justify your answer.

  \item Suppose we use the parallel partition part of the algorithm,
  but perform the two recursive calls in sequence rather than in
  parallel.
  \begin{enumerate}
  \item[i.] What is the asymptotic best-case span and work?  Justify your answer.
  \item[ii.] What is the asymptotic worst-case span and work? Justify your answer.
  \end{enumerate}
  \end{enumerate}

\end{problem}

\newpage

\begin{problem}{\bf Another Wrong Bank Account}

Note: The purpose of this problem is to show you something you should
not do because it does not work.

Consider this pseudocode for a bank account supporting concurrent
access:
\begin{verbatim}
class BankAccount {
  private int balance = 0;
  private Lock lk = new Lock();
  int getBalance() {
    lk.acquire();
    int ans = balance;
    lk.release();
    return ans;
  }
  void setBalance(int x) {
    lk.acquire();
    balance = x;
    lk.release(); 
  }
  void withdraw(int amount) {
    lk.acquire(); 
    int b = getBalance();
    if(amount > b) {
      lk.release();
      throw new WithdrawTooLargeException();
    }
    setBalance(b - amount);
    lk.release(); 
  }
}
\end{verbatim}
The code above is wrong if locks are \emph{not} re-entrant.
Consider the \emph{absolutely horrible idea} of ``fixing'' this
problem by rewriting the {\tt withdraw} method to be:
\begin{verbatim}
  void withdraw(int amount) {
    lk.acquire(); 
    lk.release();
    int b = getBalance();
    lk.acquire();
    if(amount > b) {
      lk.release();
      throw new WithdrawTooLargeException();
    }
    lk.release();
    setBalance(b - amount);
    lk.acquire();
    lk.release(); 
  }
\end{verbatim}

\begin{enumerate}
\item Explain how this approach prevents blocking forever unlike the
  original code.
\item Show this approach is incorrect by giving an interleaving of two
  threads in which a withdrawal is forgotten.
\end{enumerate}

\end{problem}

\newpage

\begin{problem}{\bf Concurrent Queue with Two Stacks}

Consider this Java implementation of a queue with two stacks.  We do
not show the entire stack implementation, but assume it is correct.
Notice the stack has synchronized methods but the queue does not.
The queue is incorrect in a concurrent setting.
\begin{verbatim}
class Stack<E> {                              class Queue<E> {
  ...                                           Stack<E> in  = new Stack<E>();
  synchronized boolean isEmpty() { ... }        Stack<E> out = new Stack<E>();
  synchronized E pop() { ... }                  void enqueue(E x){ in.push(x); }
  synchronized void push(E x) { ... }           E dequeue() {
}                                                 if(out.isEmpty()) {
                                                    while(!in.isEmpty()) {
                                                      out.push(in.pop());
                                                    }
                                                  }
                                                  return out.pop();
                                                }
                                              }
\end{verbatim}
\begin{enumerate}
\item Show the queue is incorrect by showing an interleaving that
  meets the following criteria:
  \begin{enumerate}
  \item[i.] Only one thread ever performs {\tt enqueue} operations and
  that thread enqueues numbers in increasing order (1, 2, 3, ...).
  \item[ii.] There is a thread that performs two {\tt dequeue} operations
  such that its first {\tt dequeue} returns a number larger than its
  second {\tt dequeue}, which should never happen.
  \item[iii.] Every {\tt dequeue} succeeds (the queue is never empty).
  \end{enumerate}
  Your solution can use 1 or more additional threads that perform {\tt
    dequeue} operations.
\item A simple fix would make {\tt enqueue} and {\tt dequeue}
  synchronized methods.  Explain why this would never allow an {\tt
  enqueue} and {\tt dequeue} to happen at the same time.
\item To allow an {\tt enqueue} and a {\tt dequeue} to
  operate on a queue at the same time (at least when {\tt out} is not empty),
  we could try either of the approaches below for {\tt
    dequeue}.  For each, show an interleaving that demonstrates the
  approach is broken.  Your interleaving should satisfy the three
  properties listed in part (a).
\begin{verbatim}
  E dequeue() {                                E dequeue() {
    synchronized(out) {                          synchronized(in) {
      if(out.isEmpty()) {                          if(out.isEmpty()) {
        while(!in.isEmpty()) {                       while(!in.isEmpty()) {
          out.push(in.pop());                          out.push(in.pop());
        }                                            }
      }                                            }
      return out.pop();                           }
    }                                             return out.pop();
  }                                            } 
\end{verbatim}
\item Provide a solution, based on two stacks as above, that correctly allows
an {\tt enqueue} and a {\tt dequeue} to happen at the same time, at
least when {\tt out} is not empty.  Your solution should define {\tt dequeue} and
involve multiple locks.
\end{enumerate}

\end{problem}

\newpage

\begin{problem}{\bf Simple Concurrency with B-Trees}

  Note: Real databases and file systems use very fancy fine-grained
  synchronization for B-Trees such as ``hand-over-hand locking''
  (which we did not discuss), but this problem considers some
  relatively simpler approaches.

  Suppose we have a B Tree supporting operations {\tt insert} and {\tt
    lookup}.  A simple way to synchronize threads accessing the tree
  would be to have one lock for the entire tree that both operations
  acquire/release.
\begin{enumerate}
\item Suppose instead we have one lock per node in the tree. Each
  operation acquires the locks along the path to the leaf it needs
  and then at the end releases all these locks.  Explain why this
  allegedly more fine-grained approach provides absolutely no benefit.
\item Now suppose we have one readers/writer lock per node and {\tt
    lookup} acquires a read lock for each node it encounters whereas
  {\tt insert} acquires a write lock for each node it encounters.  How
  does this provide more concurrent access than the approach in part
  (a)?  Is it any better than having one readers/writer lock for the
  whole tree (explain)?
\item Now suppose we modify the approach in part (b) so that {\tt
    insert} acquires a write lock \emph{only for the leaf node} (and
    read locks for other nodes it encounters).  How would
  this approach increase concurrent access?  When would this be
  \emph{incorrect}?  Explain how to fix this approach without changing the
  asymptotic complexity of {\tt insert} by detecting when it is
  incorrect and in (only) those cases, starting the {\tt insert} over
  using the approach in part (b) for that {\tt insert}.  Why would
  reverting to the approach in part (b) be fairly rare?
\end{enumerate}

\end{problem}

\newpage

\begin{problem}{\bf Concurrent/Parallel Graph Traversal}

\begin{enumerate}
\item In Java or pseudocode, define an \emph{unbounded stack} with
  operations {\tt push} and {\tt pop} that can be used by threads
  concurrently.  Make the element type generic and use a linked list
  for the underlying implementation.  A {\tt pop} should not raise an
  error.  Instead it should wait until the stack is not empty.  
  Use a condition variable.  (Java detail: The {\tt wait}
  method throws an exception that Java requires you catch, but it
  doesn't matter if you include this detail in your solution.)
\item The method {\tt traverseFrom} in the code below uses your answer
  to part (a) to apply the {\tt doIt} method to every node reachable
  from some node in a graph.  The order {\tt doIt} is applied to nodes
  does not matter.  The code uses 4 threads to try to improve
  performance.  All you need to do is write the {\tt run} method
  (probably around 10 lines) for {\tt GraphWalker} using the other
  code.  Requirements:
  \begin{itemize}
  \item No node should have {\tt doIt} applied to it more than once.
  You can assume the {\tt visited} field of each node is initially
  {\tt false} for this purpose.
  \item Use the stack to hold nodes that still need to be processed.
  That way threads not busy completing a {\tt doIt} call can find
  useful work.
  \item Avoid all synchronization errors.  (Hint: Your code does not
  need any \emph{explicit} synchronization since it can call other
  code that already performs synchronization.)  
  \item Because {\tt doIt} may be expensive, do not hold any locks while
  calling it.
  \end{itemize}
  Note that after all reachable nodes have been processed, all threads
  will be blocked forever waiting for the stack to become non-empty.
  That is fine for a homework problem.  Also note that you
  may or may not find {\tt isVisited} helpful.
\end{enumerate}

\begin{tabular}{ll}
\begin{minipage}[t]{.48\linewidth}
\begin{verbatim}
class GraphNode {
  boolean visited = false;
  GraphNode[] neighbors; 
  // ... other fields, constructor, 
  // etc. omitted ...
  synchronized boolean isVisited() { 
    return visited; 
  }
  synchronized boolean wasAndSetVisited(){ 
    boolean ans = visited;
    visited = true;
    return ans;
  }
}
\end{verbatim}
\end{minipage}
&
\begin{minipage}[t]{.48\linewidth}
\begin{verbatim}
class GraphWalker extends java.lang.Thread {
  Stack<GraphNode> stk;
  GraphWalker(Stack<GraphNode> s) {
    stk = s;
  }
  void doIt(GraphNode n) {/*...*/}

  // ... FOR YOU ...

}
class Main {
  public static final int NUM_THREADS = 4;
  void traverseFrom(GraphNode root) {
    Stack<GraphNode> stk = 
       new Stack<GraphNode>();
    stk.push(root);
    for(int i = 0; i < NUM_THREADS; i++)
      new GraphWalker(stk).start();
  }
}
\end{verbatim}
\end{minipage}
\end{tabular}

\end{problem}

\end{document}

### 2022-11-24 latex/comp_or.tex
Algorithms shape updated

@@
 \begin{algorithm}[H]
-	\caption{$ACLP+RPDP (scenario,volume)$}  \label{alg:main}
+	\caption{Solves ACLP+RPDP for a scenario with certain volume surplus (1.2, 1.5, or 2.0)}  \label{alg:main}
 	\begin{algorithmic}[1]
+
+		\Procedure{$ACLP+RPDP$}{$scenario,\ surplus,\ numProcs$}
+
 		\State Let $L, M, C$ be according to $scenario$ \label{main:LMC}
-		\State $N \gets ItemsGeneration(scenario,volume)$ \label{main:items}
+		\State $N \gets ItemsGeneration(scenario,\ surplus)$ \label{main:items}
 		\For {each $method$}
-
-			\State $numProcs \gets $, the number of parallel processes to be used
 
@@
 			\EndFor \label{main:loop2}
-			\State $answer[scenario,volume,method] \gets \max f$ \label{main:f}
+			\State $answer[scenario,surplus,method] \gets \max f$ \label{main:f}
 		\EndFor
-		\State $ {\bf return}\ answer$
+
+		\Return $answer$
+
+		\EndProcedure
 	\end{algorithmic}
@@
 \begin{algorithm}[H]
-	\caption{$SolveTour(\pi, L, M, C, N, method, numProcs)$}  \label{alg:tour}
+	\caption{ Solves the sequence of nodes in the tour indexed by $\pi$ }  \label{alg:tour}
 
 	\begin{algorithmic}[1]
+
+		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method, numProcs$}
 
@@
 		\EndFor  \label{tour:loop2}
-		\State $ {\bf return}\ score / cost$ \label{tour:f}
+		\Return $score / cost$ \label{tour:f}
+
+		\EndProcedure
 
@@
 
-	\caption{ $SolveNode(mpACO, k, G, numProcs)$ } \label{alg:aco}
+	\caption{ Solves a node with the {\it mpACO} method } \label{alg:aco}
 	\begin{algorithmic}[1]
+
+		\Procedure{$SolveNode$}{$mpACO, k, G, numProcs$}
+
 		\State Let $G(V_k, E^Q_k)$ \Comment{This is a partial solution with only consolidated on board}
@@
 
-			\While{$runtime < 0.7s$ {\bf and} $counter > 0$} \label{aco:while2}
+			\While{$counter > 0$} \label{aco:while2}
+
+				\State $G^{a} \gets \varnothing$
+
+				\While{$runtime < 0.7s$} \Comment{Wait until there is a solution in the queue}
 
 					\State $G^{a} \gets AntsQueue.Pop()$
+
+					\If{$G^{a} \ne \varnothing$}				
+						\Break
+					\EndIf
+
+				\EndWhile
+
 				\If{$f_s(G^{a}) > f_s(G^{+})$}
@@
 		\EndWhile	
-		\State ${\bf return}$ $G^*(V_k, E^{N*}_k \cup E^Q_k)$		
+		\Return $G^*(V_k,\ E^{N*}_k \cup E^Q_k)$
+
+		\EndProcedure		
 	\end{algorithmic}
@@
 \begin{algorithm}[H]
-	\caption{ $Greedy(k, G, limit)$}  \label{alg:greedy}
+	\caption{Mount a greedy solution until the volume limit for each pallet}  \label{alg:greedy}
 
 	\begin{algorithmic}[1]
+
+		\Procedure{$Greedy$}{$k, G, limit$}
+
 		\State Let $G(V_k, E^Q_k)$
@@
 			\If{$(p_i, a_q^k) \in E^Q_k$} 
-				\State $volume[i] \gets volume[i] + a_q^k.v$ 
+				\State $volume[q] \gets volume[q] + a_q^k.v$ 
 			\EndIf		
@@
 		\EndFor
-		\State ${\bf return}\ G(V_k, E^N_k \cup E^Q_k),\ \eta_1$
+
+		\Return $G(V_k, E^N_k \cup E^Q_k),\ \eta_1$ \Comment{return the solution and the last item indexes array }
+
+		\EndProcedure
 
@@
 \begin{algorithm}[H] 
-	\caption{ $AntSolve(G^+,\ AntsQueue,\ ga^k_{ij})$ } \label{alg:AntSolve}
+	\caption{Each ant solves a node in parallel} \label{alg:AntSolve}
 	\begin{algorithmic}[1]
+
+		\Procedure{$AntSolve$}{$G^+,\ AntsQueue,\ ga^k_{ij})$}
+
 		\State $G^a(V_k, E^{Na}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$
@@
 		\While{$|E| > 0$} \label{AntSolve:not_empty}
+
 			\State $e \gets Roulette(E)$ biased through $ga^k_{ij}$  \label{AntSolve:roulette}
@@
 			\EndIf
+
 		\EndWhile
 
-\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{AntSolve:rho}
-\State $\tau^k_{ij} \gets \tau^k_{ij} + \Delta\tau$ \label{AntSolve:phero}
-\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{AntSolve:ga}
-
-		\State $AntsQueue \gets AntsQueue \cup G^a(V_k, E^{Na}_k \cup E^Q_k)$ \label{AntSolve:put_in_queue}
+%\For{$\tau^k_{ij}, \theta^k_{ij} \in E^{Na}_k$}
+%\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{AntSolve:rho}
+%\State $\tau^k_{ij} \gets \tau^k_{ij} + \Delta\tau$ \label{AntSolve:phero}  \Comment{$\Delta\tau \gets$ Equation \ref{eq:delta_tau}}
+%\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{AntSolve:ga}		
+%\EndFor
+
+		\State $AntsQueue \gets AntsQueue \cup G^a(V_k,\ E^{Na}_k \cup E^Q_k)$ \label{AntSolve:put_in_queue}
+
+		\EndProcedure
 	\end{algorithmic}
@@
 
-Algorithm \ref{alg:AntSolve}, which will be executed by a child parallel process, receives an initial and poor solution ($G^+(V_k, E^{N+}_k \cup E^Q_k)$) generated greedily to be completed and improved by {\it mpACO}.
+Algorithm \ref{alg:AntSolve}, which will be executed by a child parallel process, receives an initial and poor solution $G^+(V_k, E^{N+}_k \cup E^Q_k)$ generated greedily to be completed and improved by {\it mpACO}.
 
@@
 
-While the neighborhood is not empty (\ref{AntSolve:not_empty}), choose an edge (\ref{AntSolve:roulette}), delete it from the neighborhood and include it in the solution, if feasible (\ref{AntSolve:ant}).
-
-Finally, put the solution in the queue (\ref{AntSolve:put_in_queue}) to be retrieved by the main process.
+While the neighborhood is not empty (line \ref{AntSolve:not_empty}), choose an edge (line \ref{AntSolve:roulette}), delete it from the neighborhood and include it in the solution, if feasible (line \ref{AntSolve:ant}).
+
+Finally, put the solution in the queue (line \ref{AntSolve:put_in_queue}) to be retrieved by the main process.
+
+%Note that $ga^k_{ij}$ is updated by each parallel ant, which may influence the solution of the others during run-time.
+This fully connected topology may increase run-time due to the overhead in communications handled by {\it Python API Multiprocessing} but it is expected that this will yield better solutions.
 
@@
 \begin{algorithm}[H]
-	\caption{ $SolveNode(mpShims, k, G, numProcs)$}  \label{alg:mpshims}
+	\caption{Solve a node with {\it mpShims}}  \label{alg:mpshims}
 	\begin{algorithmic}[1]
+
+		\Procedure{$SolveNode$}{$mpShims, k, G, numProcs$}
 
@@
 
-		\While{$runtime < 0.7s$ {\bf and} $counter > 0$}
-
-			\State $G \gets shimsQueue.Pop()$	\label{mpshims:wait}
+		\While{$counter > 0$}
+
+			\State $G \gets \varnothing$
+
+			\While{$runtime < 0.7s$} \Comment{Wait until there is a solution in the queue}
+
+			\State $G \gets shimsQueue.Pop()$
+				\If{$G \ne \varnothing$}
+					\Break
+				\EndIf
+			\EndWhile
+
 			\If{$f_s(G) > f_s(G^*)$}
@@
 			\EndIf
-
 			\State $counter \gets counter - 1$
@@
 
-		\State ${\bf return}$ $G^*(V_k, E^N_k \cup E^Q_k)$ \label{mpshims:return}
+		\Return $G^*(V_k, E^N_k \cup E^Q_k)$ \label{mpshims:return}
+
+		\EndProcedure
 	\end{algorithmic}
@@
 \begin{algorithm}[H]
-	\caption{ $shimsSolve(G,\ M,\ shimsQueue,\ limit,\ E)$}  \label{alg:shimsSolve}
+	\caption{Shims solving method to be run in parallel}  \label{alg:shimsSolve}
 	\begin{algorithmic}[1]
+
+		\Procedure{$shimsSolve$}{$G,\ M,\ shimsQueue,\ limit,\ E$}
 
@@
 
+		\EndProcedure
+
 	\end{algorithmic}
@@
 \begin{algorithm}[H]
-	\caption{ $getBestShims(i,\ \eta_1,\ \eta_2,\ E,\ k,\ slack)$ }  \label{alg:getBestShims}
+	\caption{Get the best shims of edges that fills each pallet gap}  \label{alg:getBestShims}
 	\begin{algorithmic}[1]
+
+		\Procedure{$getBestShims$}{$i,\ \eta_1,\ \eta_2,\ E,\ k,\ slack$}
 
@@
 
-		\State ${\bf return}$ $sh_{best}$
+		\Return $sh_{best}$
+
+		\EndProcedure
 
@@
 
-Observing Figure \ref{fig:scenario1data20}, it may be noticed that the best number of processes for {\it Scenario 1} (the closest to the {\it Utopia}) is \textbf{6}.
-For {\it Scenario 2}, the best number is \textbf{4}.
+Observing Figure \ref{fig:scenario1data20}, it may be noticed that the best number of processes for {\it Scenario 1} solved with {\it mpShims} (the closest to the {\it Utopia}) is \textbf{6}.
+Solved with {\it mpACO}, the best number is \textbf{8}.
 Table \ref{tab:numProcs} presents all methods, volumes percentages and scenarios numbers of best parallel processes to be run to solve all tours.
@@
 \begin{algorithm}[H]
-	\caption{$getBestNPP(scores, runtimes, procs)$}  \label{alg:bestnum}
+	\caption{Get the best number of parallel processes}  \label{alg:bestnum}
 	\begin{algorithmic}[1]
+
+		\Procedure{$getBestNPP$}{$scores,\ runtimes,\ procs$}
+
 		\State $maxS   \gets max(scores)$
@@
 		\For{$i \in |scores|$}
-			\State $y = ( maxS - scores[i] )\ /\ (maxS   - minS  )$, the Y axis leg of Figure \ref{fig:scenario1data20}
-			\State $x = ( runtimes[i] - minR )\ /\ (maxR - minR)$. the X axis leg
-			\State $h = \sqrt{x^2 + y^2}$, the hypotenuse, distance from {\it Utopia}
+			\State $y = ( maxS - scores[i] )\ /\ (maxS   - minS  )$ \Comment{the Y axis leg}
+			\State $x = ( runtimes[i] - minR )\ /\ (maxR - minR)$ \Comment{the X axis leg}
+			\State $h = \sqrt{x^2 + y^2}$  \Comment{distance from {\it Utopia}}
 			\If{$minD$ > $h$}
@@
 
-		\State ${\bf return}$ $bestNPP$
+		\Return $bestNPP$
+
+		\EndProcedure
 	\end{algorithmic}
@@
 
-		\multirow{4}{*}{$1.2$} &\multirow{2}{*}{\it {mpShims}} & $f$                & 5.70  & 3.09  & 2.94  & 2.90  & 2.79  & 5.70  & 0.91  \\%2312
+		\multirow{6}{*}{$1.2$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$                & 5.70  & 3.09  & 2.94  & 2.90  & 2.79  & 5.70  & 0.91  \\%2312
 		                       &                                 & {\bf run-time (s)} & 1     & 3     & 7     & 26    & 97    & 267   & 70  \\%390
 		                       \cmidrule{2-10}
-		                       &\multirow{2}{*}{\it {mpACO}}   & $f$                & 6.42  & 3.78  &  3.19 & 3.17  & 2.97  & 5.87  & 1.00  \\ %2540
+		                       &\multirow{2}{*}{ $mpACO^{60s}$}   & $f$                & 6.42  & 3.78  &  3.19 & 3.17  & 2.97  & 5.87  & 1.00  \\ %2540
                                &                                 & {\bf run-time (s)} & 16    & 282   &  314  & 1334  & 6945  & 18883 & 1  \\ %27162
-		\midrule
-		\multirow{4}{*}{$1.5$} &\multirow{2}{*}{\it {mpShims}} & $f$                & 8.15  & 4.22  & 4.40  & 4.01  & 3.51  & 7.08  &   \\
+		\cmidrule{2-10}		                       
+							   &\multirow{2}{*}{ $mpACO^{0.7s}$}  & $f$                &&&&&&&\\
+							   &                                 & {\bf run-time (s)} &&&&&&&\\                               
+		\midrule
+		\multirow{6}{*}{$1.5$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$                & 8.15  & 4.22  & 4.40  & 4.01  & 3.51  & 7.08  &   \\
 		                       &                                 & {\bf run-time (s)} & 1     & 5     & 12    & 45    & 244   & 737   &   \\
 		\cmidrule{2-10}
-		                       &\multirow{2}{*}{\it {mpACO}}   & $f$                &&&&&&&\\
+		                       &\multirow{2}{*}{ $mpACO^{60s}$}   & $f$                & 9.50  & 5.01  & 4.60  & 4.28  &&&\\
+		                       &                                 & {\bf run-time (s)} & 9     & 344   & 787   & 1525  &&&\\
+		\cmidrule{2-10}		                       
+							   &\multirow{2}{*}{ $mpACO^{0.7s}$}  & $f$                &&&&&&&\\
 							   &                                 & {\bf run-time (s)} &&&&&&&\\		                       
 		\midrule
-		\multirow{4}{*}{$2.0$} &\multirow{2}{*}{\it {mpShims}} & $f$                & 11.31 & 6.29  & 5.54  & 5.83  & 5.05  & 10.63 &   \\
+		\multirow{6}{*}{$2.0$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$                & 11.31 & 6.29  & 5.54  & 5.83  & 5.05  & 10.63 &   \\
 		                       &                                 & {\bf run-time (s)} & 2     & 14    & 34    & 106   & 477   & 2417  &   \\
 		\cmidrule{2-10}
-		                       &\multirow{2}{*}{\it {mpACO}}   & $f$                &&&&&&&\\
+		                       &\multirow{2}{*}{ $mpACO^{60s}$}   & $f$                &&&&&&&\\
 		                       &                                 & {\bf run-time (s)} &&&&&&&\\
-
+		\cmidrule{2-10}		                       
+		                       &\multirow{2}{*}{ $mpACO^{0.7s}$}  & $f$                &&&&&&&\\
+							   &                                 & {\bf run-time (s)} &&&&&&&\\
 		\bottomrule
+
 	\end{tabular}
+\textsuperscript{The numbers 0.7s and 60s as superscript on the methods names represent the time limit per node.}	
 \normalsize

### 2022-11-25 latex/comp_or.tex
Paper 2 update
Pheromone update changed

@@
 
-		\Procedure{$ACLP+RPDP$}{$scenario,\ surplus,\ numProcs$}
+		\Procedure{$ACLP+RPDP$}{$scenario,\ surplus,\ numProcs,\ timeLim$}
 
@@
 			\For {each $\pi \in S_K$} \label{main:loop1}
-				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, numProcs )$ \label{main:method}
+				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, numProcs,\ timeLim )$ \label{main:method}
 			\EndFor \label{main:loop2}
@@
 
-		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method, numProcs$}
+		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method, numProcs,\ timeLim$}
 
@@
 			\State $M \gets SetPalletsDestination( \pi(k) )$ \label{tour:dest2}		
-			\State $G_2 \gets SolveNode(method, \pi(k), G_1, numProcs)$ \label{tour:node}
+			\State $G_2 \gets SolveNode(method, \pi(k), G_1, numProcs,\ timeLim)$ \label{tour:node}
 			\State $s, \epsilon \gets ScoreAndDeviation(\pi(k), G_2)$ \label{tour:analyse}
@@
 
-In this subsection we present implementations of $SolveNode(method,k, G)$, where $method$\/ corresponds to a heuristic, $k$\/ is the index of the current node and $G$\/ is the allocation graph of the consolidated items that remain on board at node $l_k$.
+In this subsection we present implementations of $SolveNode(...)$, where $method$\/ corresponds to a heuristic, $k$\/ is the index of the current node and $G$\/ is the allocation graph of the consolidated items that remain on board at node $l_k$.
 
@@
 
-$SolveNode(mpACO, k, G, numProcs)$ is the main process, while $AntSolve(G^+,\ AntsQueue,\ ga^k_{ij})$ is executed the child concurrent processes.
+$SolveNode(mpACO, ...)$ is the main process, while $AntSolve(...)$ is executed the child concurrent processes.
 
@@
 
-		\Procedure{$SolveNode$}{$mpACO, k, G, numProcs$}
+		\Procedure{$SolveNode$}{$mpACO, k, G, numProcs,\ timeLim$}
 
@@
 
-				\While{$runtime < 0.7s$} \Comment{Wait until there is a solution in the queue}
+				\While{$runtime < timeLim$} \Comment{Wait until there is a solution in the queue}
 
@@
 
-		\Procedure{$SolveNode$}{$mpShims, k, G, numProcs$}
+		\Procedure{$SolveNode$}{$mpShims, k, G, numProcs,\ timeLim$}
 
@@
 
-			\While{$runtime < 0.7s$} \Comment{Wait until there is a solution in the queue}
+			\While{$runtime < timeLim$} \Comment{Wait until there is a solution in the queue}
 

### 2022-11-25 shims.py
Tests with parallel pallets were not good

@@ -24,7 +24,7 @@ class Shim(object):
         self.Edges.append(e)
         self.Included[e.Item.ID] = 1        
 
-    def shimIsFeasible(self, ce, sol, maxTorque, k ):
+    def shimIsFeasible(self, ce, sol, maxTorque, k, torque=0.0 ):
 
         if ce.Item.ID > len(self.Included)-1:
             return False
@@ -43,7 +43,12 @@ class Shim(object):
         if sol.PAV[ce.Pallet.ID] + self.V + ce.Item.V > ce.Pallet.V:
             return False #this item volume would exceed pallet volumetric limit. Equation 18
         
-        newTorque = self.T + sol.T + ce.Torque
+        newTorque = self.T + ce.Torque
+        if torque == 0.0:
+            newTorque += sol.T
+        else:
+            newTorque += torque
+
         if abs(newTorque) > maxTorque:
             return False
         
@@ -53,7 +58,7 @@ class Shim(object):
 
 # solve a Subset Selection Problem for this pallet, by selecting
 # the best shim and including its edges in solution.
-def getBestShims(p, notInSol, sol, limit, numItems, maxTorque, k):
+def getBestShims(p, notInSol, sol, limit, numItems, maxTorque, k, torque=0.0):
 
     # create the first shim
     sh = Shim(numItems)
@@ -83,12 +88,12 @@ def getBestShims(p, notInSol, sol, limit, numItems, maxTorque, k):
     # First Fit Decrease - equivalente ao KP, mas mais rápido
     for i, oe in enumerate(outter):
         for sh in Set:
-            if sh.shimIsFeasible(oe, sol, maxTorque, k):
+            if sh.shimIsFeasible(oe, sol, maxTorque, k, torque):
                 sh.AppendEdge(oe)
                 break
         else:
             sh = Shim(numItems) # create a new Shim
-            if sh.shimIsFeasible(oe, sol, maxTorque, k): # try to insert the edge "oe" in the new Shim
+            if sh.shimIsFeasible(oe, sol, maxTorque, k, torque): # try to insert the edge "oe" in the new Shim
                 sh.AppendEdge(oe)
             Set.append(sh)
 

### 2022-11-25 testing.py

@@ -20,8 +20,8 @@ bests = []
 for scenario in scenarios:
 
     if scenario == 1:
-        # instances = [1,2,3,4,5,6,7]
-        instances = [1]
+        instances = [1,2,3,4,5,6,7]
+        # instances = [1]
     if scenario == 2:
         instances = [1,2,3,4,5,6,7]
         # instances = [1]
@@ -68,8 +68,7 @@ for scenario in scenarios:
     totElapsed = 0
     sumScores = 0
 
-    numProcs = [1,2,4,6,8,10,12,14,16]
-    # numProcs = [8]
+    numProcs = [1,4,8,12,16]
 
     if method == "Shims" or method == "ACO" or method == "Greedy":
         numProcs = [1]
@@ -116,8 +115,8 @@ for scenario in scenarios:
             if method == "Greedy":
                 E = greedy.Solve(pallets, items, cfg, k)
 
-            if method == "Shims_p": # 6 CPU
-                E = shims_p.Solve(pallets, items, cfg, k, minLim, procs)
+            if method == "Shims_p":
+                E = shims_p.Solve(pallets, items, cfg, k, minLim, procs)           
 
             if method == "Shims":
                 E = shims.Solve(pallets, items, cfg, k, limit)
@@ -211,7 +210,7 @@ for scenario in scenarios:
                     runtimes[ix]  += elapsed
                     scores[ix] += sNodeAccum # * (2-abs(epsilom))
 
-    if method == "Shims_p" or method == "ACO_p":
+    if len(numProcs) > 1:
 
         text = "np,time,score\n"
         for i, _ in enumerate(numProcs):
@@ -241,14 +240,16 @@ for scenario in scenarios:
 
         for i, _ in enumerate(numProcs):
 
-            yLeg = (maxScore - scores[i])/(maxScore - minScore)
-            xLeg = (runtimes[i] - minTime)  /(maxTime  - minTime)
+            if (maxScore - minScore) and (maxTime  - minTime) > 0:
 
-            distance = math.sqrt(xLeg**2 + yLeg**2)
+                yLeg = (maxScore - scores[i])/(maxScore - minScore)
+                xLeg = (runtimes[i] - minTime)  /(maxTime  - minTime)
 
-            if minDist > distance:
-                minDist = distance
-                bestNumProcs = numProcs[i]
+                distance = math.sqrt(xLeg**2 + yLeg**2)
+
+                if minDist > distance:
[…]

### 2022-11-27 latex/comp_or.tex
Tests with parallel pallets were not good
Shims_p is better with 1 process

@@
 
+\begin{figure}[H]
+	\begin{center}
+		\begin{tikzpicture}[scale=0.8, samples=100]
+			\begin{axis}[
+				myplotstyle,
+				xlabel={\bf mpShims computation times (s)},
+				ylabel={\bf Score},
+				]
+				\addplot+[
+				only marks,	
+				nodes near coords = {\numProcs{}},
+				visualization depends on = {value \thisrow{np} \as \numProcs},						
+				]
+				table[x=time, y=score, col sep=comma] {./csv/Shims_p1data50.csv};
+			\end{axis}
+			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
+		\end{tikzpicture}%
+		~%
+		%
+		\begin{tikzpicture}[scale=0.8, samples=100]
+			\begin{axis}[
+				myplotstyle,
+				xlabel={\bf mpACO computation times (s)},
+				ylabel={\bf Score},
+				]
+				\addplot+[
+				only marks,	
+				nodes near coords = {\numProcs{}},
+				visualization depends on = {value \thisrow{np} \as \numProcs},						
+				]
+				table[x=time, y=score, col sep=comma] {./csv/ACO_p1data50.csv};
+			\end{axis}
+			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
+		\end{tikzpicture}
+
+	\end{center}
+	\caption{ Scenario 1 performances with $1.5 \times volume$. }
+	\label{fig:scenario1data50}
+\end{figure}
+
 To build Table \ref{tab:numProcs}, it was necessary to determine the number of parallel processes that best solve each node. For this we created a procedure as in Algorithm \ref{alg:bestnum}.
@@
 
-		\multirow{6}{*}{$1.2$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$                & 5.70  & 3.09  & 2.94  & 2.90  & 2.79  & 5.70  & 0.91  \\%2312
+		\multirow{8}{*}{$1.2$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$               &   &   &   &   &   &   &   \\%
+                               &                                 & {\bf run-time (s)} &      &      &      &     &     &    &   \\%\cmidrule{2-10}
+
+		                       &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$               & 5.70  & 3.09  & 2.94  & 2.90  & 2.79  & 5.70  & 0.91  \\%2312
 		                       &                                 & {\bf run-time (s)} & 1     & 3     & 7     & 26    & 97    & 267   & 70  \\%390
@@
 		\midrule
-		\multirow{6}{*}{$1.5$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$                & 8.15  & 4.22  & 4.40  & 4.01  & 3.51  & 7.08  &   \\
-		                       &                                 & {\bf run-time (s)} & 1     & 5     & 12    & 45    & 244   & 737   &   \\
+		\multirow{6}{*}{$1.5$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$               & 8.15  & 4.22  & 4.40  & 4.01  & 3.51  & 7.08  & 0.91  \\%3137
+		                       &                                 & {\bf run-time (s)} & 1     & 5     & 12    & 45    & 244   & 737   & 30  \\%1026
 		\cmidrule{2-10}
-		                       &\multirow{2}{*}{ $mpACO^{60s}$}   & $f$                & 9.50  & 5.01  & 4.60  & 4.28  &&&\\
-		                       &                                 & {\bf run-time (s)} & 9     & 344   & 787   & 1525  &&&\\
+		                       &\multirow{2}{*}{ $mpACO^{60s}$}   & $f$               & 9.50  & 5.01  & 4.60  & 4.28  & 3.78  & 7.15  & 1.00 \\ %3432
+
+		                       &                                 & {\bf run-time (s)} & 9     & 344   & 787   & 1525  & 7525  & 21443 & 1 \\%30493
 		\cmidrule{2-10}		                       
@@
 		\midrule
-		\multirow{6}{*}{$2.0$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$                & 11.31 & 6.29  & 5.54  & 5.83  & 5.05  & 10.63 &   \\
-		                       &                                 & {\bf run-time (s)} & 2     & 14    & 34    & 106   & 477   & 2417  &   \\
+		\multirow{6}{*}{$2.0$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$               & 11.31 & 6.29  & 5.54  & 5.83  & 5.05  & 10.63 & 0.94  \\%4465
+		                       &                                 & {\bf run-time (s)} & 2     & 14    & 34    & 106   & 477   & 2417  & 10  \\%3000
 		\cmidrule{2-10}
-		                       &\multirow{2}{*}{ $mpACO^{60s}$}   & $f$                &&&&&&&\\
-		                       &                                 & {\bf run-time (s)} &&&&&&&\\
+		                       &\multirow{2}{*}{ $mpACO^{60s}$}   & $f$               & 12.70 & 7.09  & 5.97  & 5.91  & 5.31  & 10.69 & 1.00  \\%4767
+		                       &                                 & {\bf run-time (s)} & 5     & 189   & 784   & 1456  & 7758  & 20330 & 1  \\%29544
 		\cmidrule{2-10}		                       

### 2022-11-27 methods.py

@@ -156,7 +156,7 @@ def getLimits(minLim, numProcs): # pid: 0 - numProcs-1
 
     limitSet = set()
 
-    maxLim = 0.98
+    maxLim = 0.99
 
     delta = (maxLim - minLim)/(numProcs+1)
 

### 2022-11-27 shims_p.py

@@ -1,5 +1,4 @@
 import methods as mno
-import numpy as np
 import multiprocessing as mp
 import shims
 
@@ -20,22 +19,14 @@ def Compute(edges, pallets, items, limit, cfg, k) :
 
 		# get the best shim of edges             greedy limit
         BestShims = shims.getBestShims(p, notInSol[p.ID], sol, limit, len(items), cfg.maxTorque, k)
-
-        # move the best shim of edges to solution
+ 
         for e in BestShims:
             sol.putInSol(e)
-            notInSol[e.Pallet.ID].remove(e)
-
-    # local search: Maybe still exists any edge that fit in solution
-    for nis in notInSol: # for each row
-        for ce in nis:
-            if sol.isFeasible(ce, 1.0, cfg, k):
-                sol.putInSol(ce)
 
     return sol
 
 # parallel Shims based on different limits
-def Solve(pallets, items, cfg, k, minLim, numProcs): # items include kept on board
+def Solve(pallets, items, cfg, k, limit, numProcs): # items include kept on board
 
     print(f"\nParallel Shims for ACLP+RPDP")
 
@@ -47,15 +38,10 @@ def Solve(pallets, items, cfg, k, minLim, numProcs): # items include kept on boa
 	# pallets closer to the CG are completed first
     pallets.sort(key=lambda x: abs(x.D), reverse=False)
 
-    # set of unique limit values
-    limits = mno.getLimits(minLim, numProcs)    
-
-    procs = [None for _ in limits]
+    procs = [None for _ in range(numProcs)]
     outQueue = mp.Queue()
 
     for i, p in enumerate(procs):
-
-        limit = limits[i]
 
         # create a child process for each limit
         procs[i] = mp.Process( target=shimsEnqueue,args=( outQueue, edges, pallets, items, limit, cfg, k  ) )
@@ -71,6 +57,15 @@ def Solve(pallets, items, cfg, k, minLim, numProcs): # items include kept on boa
         if sol.S > bestScore:
             bestScore = sol.S
             bestID = i
+
+    # local search
+    counter = 0
+    for e in sols[bestID].Edges:
+        if not e.InSol and sol.isFeasible(e, 1.03, cfg, k):
+            sol.putInSol(e)
+            counter += 1
+
+    print(f"{counter} extra edges put in the best solution")
 
     return mno.getSolMatrix(sols[bestID].Edges, numPallets, numItems)
         

### 2022-11-28 NewShims/methods.py
New parallel pallets shims working fine.

@@ -557,7 +557,7 @@ def writeTourSol(method, scenario, instance, pi, tour, cfg, pallets, cons, write
         sol += '\\bottomrule \n'
         sol += '\end{tabular} \n'
 
-    print(sol)
+    # print(sol)
 
     if write:
         dirname = f"./results/{surplus}/{method}_{scenario}"

### 2022-11-28 NewShims/shims_p.py

@@ -1,12 +1,135 @@
 import methods as mno
 import multiprocessing as mp
-import shims
+import numpy as np
+
+# A Shim is a thin and often tapered or wedged piece of material, used to fill small gaps or spaces between objects.
+# Set are typically used in order to support, adjust for better fit, or provide a level surface.
+# Set may also be used as spacers to fill gaps between parts subject to wear.
+
+# Shims fit in a pallet slack
+class Shim(object):
+    def __init__(self, numItems):
+        self.W        = 0   # total weight of the shim set (must fit the slack)
+        self.V        = 0.0 # total volume of the shim set (must fit the slack)
+        self.S        = 0   # score of the shim set
+        self.T        = 0.0 # torque of the shim set
+        self.Edges    = []
+        self.Included = [ 0  for _ in np.arange(numItems) ]   # items included in shim set edges
+        self.numItems = numItems
+
+    def AppendEdge(self, e):
+        self.W += e.Item.W
+        self.V += e.Item.V
+        self.S += e.Item.S
+        self.T += float(e.Item.W) * e.Pallet.D
+        self.Edges.append(e)
+        self.Included[e.Item.ID] = 1        
+
+    def shimIsFeasible(self, ce, sol, maxTorque, k):
+
+        # if ce.Item.ID > len(self.Included)-1:
+        #     return False
+
+        if self.Included[ce.Item.ID] > 0 or sol.Included[ce.Item.ID] > 0:
+            return False # this item was already inserted.
+          
+        # if ce.Item.To != ce.Pallet.Dests[k]:
+        #     return False #item and pallet destinations are different. Equation 21
+        
+        # Pallet Acumulated Weight
+        # if sol.PAW[ce.Pallet.ID] + self.W + ce.Item.W > ce.Pallet.W:
+        #     return False #this item weight would exceed pallet weight limit. Equation 17
+        
+        # Pallet Acumulated Volume
+        if sol.PAV[ce.Pallet.ID] + self.V + ce.Item.V > ce.Pallet.V*1.03:
+            return False #this item volume would exceed pallet volumetric limit. Equation 18
+        
+        # newTorque = self.T + sol.T + ce.Torque
+        # if abs(newTorque) > maxTorque:
+        #     return False
+        
+        return True
+        
+
+
+# solve a Subset Selection Problem for this pallet, by selecting
+# the best shim and including its edges in solution.
[…]

### 2022-11-29 NewShims/greedy_p.py
New shims still not working

@@ -64,6 +64,8 @@ def getBestShims(pallet, limit, items, k, solTorque, cfg):
             sh.putItem(item, k, solTorque, cfg) # insert the item in the new Shim
             Set.append(sh)
 
+    # print(len(whip), len(Set))
+
     bestScore = 0
     bestIndex = 0
     for i, shims in enumerate(Set):
@@ -100,10 +102,11 @@ class Pallet(object):
             self.PCV += item.V
             self.InSol[item.ID] = True
 
-            with solTorque.get_lock():
-                solTorque.value += deltaTau
+            # with solTorque.get_lock():
+            solTorque.value += deltaTau
 
             self.S += item.S
+
 
 def loadPallets(cfg, items):
 
@@ -222,14 +225,15 @@ if __name__ == "__main__":
     for i, _ in enumerate(procs):
         pallets[i] = palletsQueue.get( timeout = 0.7 )
 
-    printPallets(pallets, cfg, solTorque, "\n---Greedy solution---")
+    printPallets(pallets, cfg, solTorque, f"\n---Greedy solution---{limit}---{len(items)} items")
 
 
-
+    counter = 0
     for i, pallet in enumerate(pallets):
         shims = getBestShims(pallet, limit, items, k, solTorque, cfg)
 
         for item in shims.Items:
             pallets[i].putItem(item, limit, k, solTorque, cfg)
+            counter += 1
 
-    printPallets(pallets, cfg, solTorque, "\n---Shims solution---")
+    printPallets(pallets, cfg, solTorque, f"\n---Shims solution---{counter} items")

### 2022-11-29 main.py

@@ -202,8 +202,8 @@ if __name__ == "__main__":
         [0, 10, 8,  2, 10,  6,  6]  # ACO_p   2.0
     ]
 
-    scenarios = [1,2,3,4,5,6]
-    # scenarios = [5,6]
+    # scenarios = [1,2,3,4,5,6]
+    scenarios = [5,6]
     secBreak  = 0.7
 
     dists = methods.loadDistances()

### 2022-12-08 NewShims/robot.c
Trying without queue, with lock

@@ -1,0 +1,213 @@
+#include <Ultrasonic.h>
+ 
+/* Definições dos GPIOs para leitura do sensor ultrasonico */
+#define GPIO_TRIGGER    12
+#define GPIO_ECHO   11
+ 
+/* Definições de operação do sensor ultrasônico */
+#define MINIMUM_DISTANCE 10.0 //cm
+#define TIME_BETWEEN_READINGS   250  //ms
+ 
+/* Definições para controle dos dois motores */
+#define IN_1    3
+#define IN_2    4
+#define IN_3    5
+#define IN_4    6
+ 
+/* Definições dos motores a serem controlados */
+#define MOTOR_A 0x00
+#define MOTOR_B 0x01
+ 
+/* Definições das ações dos motores */
+#define BREAK    0x00
+#define ANTI_CLOCKWISE_MOVE  0x01
+#define CLOCKWISE_MOVE   0x02
+#define STOP   0x03
+ 
+/* Definições de sentido de giro (em caso de obstáculo) */
+#define ANTI_CLOCKWISE  0x00
+#define CLOCKWISE_TURN  0x01
+ 
+/* Definições do desvio de objetos */
+#define WAITING_OBSTACLE    0x00
+#define TURNING 0x01
+ 
+/* Variáveis e objetos globais */
+Ultrasonic ultrasonic(GPIO_TRIGGER, GPIO_ECHO);
+char last_side_turned = ANTI_CLOCKWISE;
+char state_obstacle_freed = WAITING_OBSTACLE;
+ 
+/* Protótipos */
+void motor_gpio_config(void);
+void motor_control(char motor, char action);
+float read_distance(void);
+void state_machine_obstacle(float obstacle_distance);
+ 
+/* Função: configura GPIOs de controle do L298N como output
+ * Parâmetros: nenhum
+ * Retorno: nenhum
+ */
+void motor_gpio_config(void)
+{
+    pinMode(IN_1, OUTPUT);
+    pinMode(IN_2, OUTPUT);
+    pinMode(IN_3, OUTPUT);
+    pinMode(IN_4, OUTPUT);
+}
+ 
+/* Função: controle um motor (freia, movimento anti-horário, movimento horário
+ *         ou ponto morto)
+ * Parâmetros: motor a ser controlado e ação desejada
+ * Retorno: nenhum
+ */
+void motor_control(char motor, char action)
+{
+    int gpio_1_motor = 0;
+    int gpio_2_motor = 0;
+ 
[…]

### 2022-12-13 NewShims/greedy_p.py
New tests accomplished but Shims_mp still does not work within the main file

@@ -1,317 +1,0 @@
-
-import methods as mno
-import numpy as np
-import multiprocessing as mp
-import time
-import ctypes as ct
-
-# A Shims fits in a pallet slack
-class Shims(object):
-    def __init__(self, pallet, numItems):
-        self.Pallet = pallet # this shims is one of the possibles for this pallet
-        self.SCW    = 0   # shims current weight (must fit the slack)
-        self.SCV    = 0.0 # shims current volume (must fit the slack)
-        self.SCS    = 0   # shims current score
-        self.InSol  = [0 for _ in np.arange(numItems)] # 1 if an item is in the shims
-        self.SCT    = 0. # shims current torque
-
-    def putItem(self, item):
-                 
-            self.SCW += item.W
-            self.SCV += item.V
-            self.SCS += item.S
-            self.InSol[item.ID] = 1
-            self.SCT += float(item.W) * float(self.Pallet.D)
-
-    def isFeasible(self, item, k, solTorque, cfg): # check constraints
-
-        if item.To != self.Pallet.Dests[k]:
-            return False
-
-        if self.Pallet.PCV + self.SCV + item.V > self.Pallet.V:
-            return False
-
-        deltaTau = float(item.W) * float(self.Pallet.D)
-        if abs(solTorque.value + self.SCT + deltaTau) > cfg.maxTorque:
-            return False
-
-        return True        
-
-# create a set of shims for this pallet and selects the best shims
-def getBestShims(pallet, items, limit, k, solTorque, solItems, cfg):
-
-    # create the first shim
-    sh = Shims(pallet, len(items))
-    Set = [sh]
-
-    whipLen = pallet.V * (3. - 2*limit) # whip length
-
-    vol = 0.
-    whip = []
-    
-    for item in items:
-        if solItems[item.ID] == -1: # not allocated in any pallet
-            vol += item.V
-            whip.append(item)
-            if vol > whipLen:
-                break
-
-    # First Fit Decrease - equivalente ao KP, mas mais rápido
-    # included1 = 0
-    # included2 = 0
-    for item in whip:
-        newShims = True
-        for sh in Set:
-            if sh.isFeasible(item, k, solTorque, cfg):
-                sh.putItem(item)
-                newShims = False
[…]

### 2022-12-13 shims_p.py

@@ -1,74 +1,0 @@
-import methods as mno
-import multiprocessing as mp
-import shims
-
-def shimsEnqueue(outQueue, edges, pallets, items, limit, cfg, k ):
-    outQueue.put( Compute( edges, pallets, items, limit, cfg, k) )
-
-def Compute(edges, pallets, items, limit, cfg, k) :
-
-    sol = mno.Solution(edges, pallets, items, limit, cfg, k)
-
-    #edges not in sol groupped by pallets
-    notInSol = [ [] for _ in range(len(pallets)) ]
-
-	# get the best shim that fills the slack
-    for p in (pallets):
-
-        notInSol[p.ID] = [e for e in sol.Edges if e.Pallet.ID == p.ID and not e.InSol  ]
-
-		# get the best shim of edges             greedy limit
-        BestShims = shims.getBestShims(p, notInSol[p.ID], sol, limit, len(items), cfg.maxTorque, k)
- 
-        for e in BestShims:
-            sol.putInSol(e)
-
-    return sol
-
-# parallel Shims based on different limits
-def Solve(pallets, items, cfg, k, limit, numProcs): # items include kept on board
-
-    print(f"\nParallel Shims for ACLP+RPDP")
-
-    numItems   = len(items)
-    numPallets = len(pallets)
-
-    edges = mno.mountEdges(pallets, items, cfg)
-
-	# pallets closer to the CG are completed first
-    pallets.sort(key=lambda x: abs(x.D), reverse=False)
-
-    procs = [None for _ in range(numProcs)]
-    outQueue = mp.Queue()
-
-    for i, p in enumerate(procs):
-
-        # create a child process for each limit
-        procs[i] = mp.Process( target=shimsEnqueue,args=( outQueue, edges, pallets, items, limit, cfg, k  ) )
-        
-    for p in procs:
-        p.start()
-        
-    sols = [outQueue.get() for _ in procs]
-
-    bestScore = 0
-    bestID    = 0
-    for i, sol in enumerate(sols):
-        if sol.S > bestScore:
-            bestScore = sol.S
-            bestID = i
-
-    # local search
-    counter = 0
-    for e in sols[bestID].Edges:
-        if not e.InSol and sol.isFeasible(e, 1.03, cfg, k):
-            sol.putInSol(e)
-            counter += 1
-
-    print(f"{counter} extra edges put in the best solution")
-
-    return mno.getSolMatrix(sols[bestID].Edges, numPallets, numItems)
[…]

### 2023-01-13 common.py
Testing Gurobi

@@ -206,7 +206,7 @@ def getTours(num, costs, threshold):
         nodes = []
         cost = 0.
       
-        for j, nid in enumerate(tour[:-1]):
+        for j, nid in enumerate(tour):
             n = Node(nid, 0.)
             nodes.append( n )
 

### 2023-01-14 latex/comp_or.tex
New parallel pallets shims working fine.
Started parallel greedy
New shims still not working
Trying without queue, with lock
New tests accomplished but Shims_mp still does not work within the main file
Testing Gurobi
Check Gurobi results on testing

@@
 
-%\usepackage{hyphenat}
 \usepackage[none]{hyphenat}
@@
 Transport aviation faces risks of cargo imbalance in the set of circumstances of aerial pickup and shipment of goods in a distribution system given the urgency required for loading for fast takeoff and mission accomplishment, notably in moments of crisis, public emergency, client contract short turnaround times, or any external forces for instant takeoff.
-Additionally, there is no computer support to aid load and trip programmers with the massive transportation demands in each location.
+Furthermore, there is no computer support available to aid load and trip programmers with the massive transportation demands in each location.
 
@@
 %\scriptsize
-	\begin{tabular}{cccccccccc}
+\begin{tabular}{crcccccccc}
 \toprule
@@
 
-		\multirow{8}{*}{$1.2$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$               &   &   &   &   &   &   &   \\%
-                               &                                 & {\bf run-time (s)} &      &      &      &     &     &    &   \\%\cmidrule{2-10}
-
-		                       &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$               & 5.70  & 3.09  & 2.94  & 2.90  & 2.79  & 5.70  & 0.91  \\%2312
-		                       &                                 & {\bf run-time (s)} & 1     & 3     & 7     & 26    & 97    & 267   & 70  \\%390
+\multirow{10}{*}{$1.2$} &\multirow{2}{*}{ $Shims$}  & $f$                &       &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
 \cmidrule{2-10}
-		                       &\multirow{2}{*}{ $mpACO^{60s}$}   & $f$               & 6.42  & 3.78  &  3.19 & 3.17  & 2.97  & 5.87  & 1.00  \\ %2540
-                               &                                 & {\bf run-time (s)} & 16    & 282   &  314  & 1334  & 6945  & 18883 & 1  \\ %27162
+&\multirow{2}{*}{ $mpShims$}                        & $f$                &       &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
+
 \cmidrule{2-10}
-							   &\multirow{2}{*}{ $mpACO^{0.7s}$}  & $f$               &&&&&&&\\
-							   &                                 & {\bf run-time (s)} &&&&&&&\\                               
-		\midrule
-		\multirow{6}{*}{$1.5$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$               & 8.15  & 4.22  & 4.40  & 4.01  & 3.51  & 7.08  & 0.91  \\%3137
-		                       &                                 & {\bf run-time (s)} & 1     & 5     & 12    & 45    & 244   & 737   & 30  \\%1026
+&\multirow{2}{*}{ $ACO$}                            & $f$                & 2.85  &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & 15    &       &        &       &       &       &   \\%
 \cmidrule{2-10}		                       
-		                       &\multirow{2}{*}{ $mpACO^{60s}$}   & $f$               & 9.50  & 5.01  & 4.60  & 4.28  & 3.78  & 7.15  & 1.00 \\ %3432
-
-		                       &                                 & {\bf run-time (s)} & 9     & 344   & 787   & 1525  & 7525  & 21443 & 1 \\%30493
+&\multirow{2}{*}{ $mpACO$}                          & $f$                & 3.63  &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & 11    &       &        &       &       &       &   \\%
+
 \cmidrule{2-10}		                       
-							   &\multirow{2}{*}{ $mpACO^{0.7s}$}  & $f$               &&&&&&&\\
-							   &                                 & {\bf run-time (s)} &&&&&&&\\		                       
-		\midrule
-		\multirow{6}{*}{$2.0$} &\multirow{2}{*}{ $mpShims^{0.7s}$}& $f$               & 11.31 & 6.29  & 5.54  & 5.83  & 5.05  & 10.63 & 0.94  \\%4465
-		                       &                                 & {\bf run-time (s)} & 2     & 14    & 34    & 106   & 477   & 2417  & 10  \\%3000
+&\multirow{2}{*}{ $Gurobi$}                         & $f$                & 6.35  &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & 2     &       &        &       &       &       &   \\%
+
+\midrule
+\multirow{10}{*}{$1.5$} &\multirow{2}{*}{ $Shims$}  & $f$                &       &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
 \cmidrule{2-10}
-		                       &\multirow{2}{*}{ $mpACO^{60s}$}   & $f$               & 12.70 & 7.09  & 5.97  & 5.91  & 5.31  & 10.69 & 1.00  \\%4767
-		                       &                                 & {\bf run-time (s)} & 5     & 189   & 784   & 1456  & 7758  & 20330 & 1  \\%29544
+&\multirow{2}{*}{ $mpShims$}                        & $f$                &       &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
+
 \cmidrule{2-10}
-		                       &\multirow{2}{*}{ $mpACO^{0.7s}$}  & $f$                &&&&&&&\\
-							   &                                 & {\bf run-time (s)} &&&&&&&\\
+&\multirow{2}{*}{ $ACO$}                            & $f$                & 2.85  &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & 15    &       &        &       &       &       &   \\%
+\cmidrule{2-10}		                       
+&\multirow{2}{*}{ $mpACO$}                          & $f$                & 3.63  &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & 11    &       &        &       &       &       &   \\%
+
+\cmidrule{2-10}		                       
+&\multirow{2}{*}{ $Gurobi$}                         & $f$                & 6.35  &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & 2     &       &        &       &       &       &   \\%
+
+\midrule
+\multirow{10}{*}{$2.0$} &\multirow{2}{*}{ $Shims$}  & $f$                &       &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
+\cmidrule{2-10}
+&\multirow{2}{*}{ $mpShims$}                        & $f$                &       &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
+
+\cmidrule{2-10}
+&\multirow{2}{*}{ $ACO$}                            & $f$                & 2.85  &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & 15    &       &        &       &       &       &   \\%
+\cmidrule{2-10}		                       
+&\multirow{2}{*}{ $mpACO$}                          & $f$                & 3.63  &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & 11    &       &        &       &       &       &   \\%
+
+\cmidrule{2-10}		                       
+&\multirow{2}{*}{ $Gurobi$}                         & $f$                & 6.35  &       &        &       &       &       &   \\%
+&                                                   & {\bf run-time (s)} & 2     &       &        &       &       &       &   \\%
+
 \bottomrule
@@
 \end{tabular}
-\textsuperscript{The numbers 0.7s and 60s as superscript on the methods names represent the time limit per node.}	
+%\textsuperscript{The numbers 0.7s and 60s as superscript on the methods names represent the time limit per node.}	
 \normalsize

### 2023-02-04 latex/comp_or.tex
mpShims still with values "better" than optimal

@@
 
-The final experiments are performed on a 64-bit, 128GiB, 3.6GHz, 48-core processor, 2 threads per core, Intel Xeon, with {\it Linux CentOS 8} as the operational system and {\it Python 3.8.2} as the programming language.
+The final experiments are performed on a 64-bit, 16GiB, 3.6GHz, 8-core AMD® Fx-8320e processor, 2 threads per core, with {\it Linux Ubuntu 20} as the operational system and {\it Python 3.8.2} as the programming language.
 
 For parallelization, we used the {\it Python Multi-processing} package, which implements {\it process-based parallelism} or {\it parallel shared memory algorithm}.
-According to \cite{Multi-processing}, {\it Multi-processing} is a package that supports spawning processes using an API similar to the {\it Python Threading} package.
+According to \cite{multiprocessing}, {\it Multi-processing} is a package that supports spawning processes using an API similar to the {\it Python Threading} package.
 
@@
 Essentially, the law states that while a process can be decomposed into steps that may then be run in parallel, the time taken for the whole process will be significantly limited by the steps that remain serialized.
-So, we decided to run some tests to find the best number of parallel processes for each of this work's operational scenarios.
-
-For each red bullet closest to the upper left corner, we call it {\it Utopia}, an impossible result with zero time duration and a possible best score.
-The blue bullets are experiments performed with the number of parallel processes indicated.
-The shortest distance from {\it Utopia} may reveal the most efficient parallel processes for the scenario (see Table \ref{tab:scenarios}).
-
-\begin{figure}[H]
-	\begin{center}
-		\begin{tikzpicture}[scale=0.8, samples=100]
-			\begin{axis}[
-				myplotstyle,
-				xlabel={\bf mpShims computation times (s)},
-				ylabel={\bf Score},
-				]
-				\addplot+[
-				only marks,	
-				nodes near coords = {\numProcs{}},
-				visualization depends on = {value \thisrow{np} \as \numProcs},						
-				]
-				table[x=time, y=score, col sep=comma] {./csv/Shims_p1data20.csv};
-			\end{axis}
-			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-		\end{tikzpicture}%
-		~%
+
 %
-		\begin{tikzpicture}[scale=0.8, samples=100]
-		\begin{axis}[
-			myplotstyle,
-			xlabel={\bf mpACO computation times (s)},
-			ylabel={\bf Score},
-			]
-			\addplot+[
-			only marks,	
-			nodes near coords = {\numProcs{}},
-			visualization depends on = {value \thisrow{np} \as \numProcs},						
-			]
-			table[x=time, y=score, col sep=comma] {./csv/ACO_p1data20.csv};
-		\end{axis}
-		\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-	\end{tikzpicture}
-
-	\end{center}
-	\caption{ Scenario 1 performances with $1.2 \times volume$. }
-	\label{fig:scenario1data20}
-\end{figure}
-
-Observing Figure \ref{fig:scenario1data20}, it may be noticed that the best number of processes for {\it Scenario 1} solved with {\it mpShims} (the closest to the {\it Utopia}) is \textbf{6}.
-Solved with {\it mpACO}, the best number is \textbf{8}.
-Table \ref{tab:numProcs} presents all methods, volumes percentages and scenarios numbers of best parallel processes to be run to solve all tours.
-
-\begin{figure}[H]
-	\begin{center}
-		\begin{tikzpicture}[scale=0.8, samples=100]
-			\begin{axis}[
-				myplotstyle,
-				xlabel={\bf mpShims computation times (s)},
-				ylabel={\bf Score},
-				]
-				\addplot+[
-				only marks,	
-				nodes near coords = {\numProcs{}},
-				visualization depends on = {value \thisrow{np} \as \numProcs},						
-				]
-				table[x=time, y=score, col sep=comma] {./csv/Shims_p1data50.csv};
-			\end{axis}
-			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-		\end{tikzpicture}%
-		~%
+%\begin{figure}[H]
+%	\begin{center}
+%		\begin{tikzpicture}[scale=0.8, samples=100]
+%			\begin{axis}[
+%				myplotstyle,
+%				xlabel={\bf mpShims computation times (s)},
+%				ylabel={\bf Score},
+%				]
+%				\addplot+[
+%				only marks,	
+%				nodes near coords = {\numProcs{}},
+%				visualization depends on = {value \thisrow{np} \as \numProcs},						
+%				]
+%				table[x=time, y=score, col sep=comma] {./csv/Shims_p1data20.csv};
+%			\end{axis}
+%			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
+%		\end{tikzpicture}%
+%		~%
+%		%
+%		\begin{tikzpicture}[scale=0.8, samples=100]
+%		\begin{axis}[
+%			myplotstyle,
+%			xlabel={\bf mpACO computation times (s)},
+%			ylabel={\bf Score},
+%			]
+%			\addplot+[
+%			only marks,	
+%			nodes near coords = {\numProcs{}},
+%			visualization depends on = {value \thisrow{np} \as \numProcs},						
+%			]
+%			table[x=time, y=score, col sep=comma] {./csv/ACO_p1data20.csv};
+%		\end{axis}
+%		\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
+%	\end{tikzpicture}
 %		
-		\begin{tikzpicture}[scale=0.8, samples=100]
-			\begin{axis}[
-				myplotstyle,
-				xlabel={\bf mpACO computation times (s)},
-				ylabel={\bf Score},
-				]
-				\addplot+[
-				only marks,	
-				nodes near coords = {\numProcs{}},
-				visualization depends on = {value \thisrow{np} \as \numProcs},						
-				]
-				table[x=time, y=score, col sep=comma] {./csv/ACO_p1data50.csv};
-			\end{axis}
-			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-		\end{tikzpicture}
-
-	\end{center}
-	\caption{ Scenario 1 performances with $1.5 \times volume$. }
-	\label{fig:scenario1data50}
-\end{figure}
-
-To build Table \ref{tab:numProcs}, it was necessary to determine the number of parallel processes that best solve each node. For this we created a procedure as in Algorithm \ref{alg:bestnum}.
-
-\begin{algorithm}[H]
-	\caption{Get the best number of parallel processes}  \label{alg:bestnum}
-	\begin{algorithmic}[1]
-
-		\Procedure{$getBestNPP$}{$scores,\ runtimes,\ procs$}
-
-		\State $maxS   \gets max(scores)$
-		\State $minS   \gets min(scores)$
-		\State $maxR \gets max(runtimes)$
-		\State $minR \gets min(runtimes)$
-
-		\State $bestNPP \gets 0$, best number of parallel processes
-		\State $minD \gets 999999999.9$
-
-		\For{$i \in |scores|$}
-			\State $y = ( maxS - scores[i] )\ /\ (maxS   - minS  )$ \Comment{the Y axis leg}
-			\State $x = ( runtimes[i] - minR )\ /\ (maxR - minR)$ \Comment{the X axis leg}
-			\State $h = \sqrt{x^2 + y^2}$  \Comment{distance from {\it Utopia}}
-			\If{$minD$ > $h$}
-				\State $minD \gets h$
-				\State $bestNPP \gets procs[i]$
-			\EndIf
-
-		\EndFor
-
-		\Return $bestNPP$
-
-		\EndProcedure
-	\end{algorithmic}
-\end{algorithm}
-
-\vspace{2.0mm}
-
-\begin{table}[H]
-\centering
-\caption{$bestNPP$ found for each scenario}  \label{tab:numProcs}
-	\footnotesize
-%\scriptsize
-\begin{tabular}{rccccccc}
-\toprule
-                               &               &\multicolumn{6}{c}{\bf Scenarios} \\
-                               \cmidrule{3-8}
-{$method$}                     &{\it volume}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 6} \\
-\toprule
-
-\multirow{3}{*}{\it {mpShims}} &$1.2$& 6       & 4     & 4     & 4     & 2     & 2    \\ 
-                               &$1.5$& 4       & 4     & 4     & 4     & 4     & 4    \\
-                               &$2.0$& 4       & 4     & 4     & 2     & 2     & 4    \\
-\midrule[.1pt]
-
-\multirow{3}{*}{\it {mpACO}}   &$1.2$& 10      & 8     & 10    & 6     & 10    & 10   \\ 
-                               &$1.5$& 10      & 6     & 12    & 8     & 10    & 8    \\
-                               &$2.0$& 10      & 8     & 2     & 10    & 6     & 6    \\                               
-\bottomrule		
-\end{tabular}
-
-\normalsize
-
-\end{table}
-
-Table \ref{tab:numProcs} presents the best numbers of parallel processes for each method, scenario, and volume surplus.
-As all numbers are less than or equal to 12, any hand-held computer with 8 cores can yield efficient solutions, as normally each core may be capable of 2 threads.
-
-To find these values we submitted to test 1, 2, 4, 6, 8, 10, 12, 14, 16 parallel processes.
+%	\end{center}
+%	\caption{ Scenario 1 performances with $1.2 \times volume$. }
+%	\label{fig:scenario1data20}
+%\end{figure}
 
@@
 
-\multirow{10}{*}{$1.2$} &\multirow{2}{*}{ $Shims$}  & $f$                &       &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
+\multirow{6}{*}{$1.2$} &\multirow{2}{*}{ Gurobi}& $f$                & 9.75       &  6.34    &  7.61  &   7.43    &       &       &   \\%
+&                                               & {\bf run-time (s)} &  8         &   17     &  30    &  114     &       &       &   \\%
+
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                        & $f$                &       &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
+&\multirow{2}{*}{ $Shims$}                      & $f$                & 7.00       &  4.20    &  4.12  &  3.54     &       &       &   \\%
+&                                               & {\bf run-time (s)} &  < 1       &   3      &    5   &   19    &       &       &   \\%
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $ACO$}                            & $f$                & 2.85  &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & 15    &       &        &       &       &       &   \\%
-\cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpACO$}                          & $f$                & 3.63  &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & 11    &       &        &       &       &       &   \\%
-
-\cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Gurobi$}                         & $f$                & 6.35  &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & 2     &       &        &       &       &       &   \\%
-
-\midrule
-\multirow{10}{*}{$1.5$} &\multirow{2}{*}{ $Shims$}  & $f$                &       &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
-\cmidrule{2-10}
-&\multirow{2}{*}{ $mpShims$}                        & $f$                &       &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
-
-\cmidrule{2-10}
-&\multirow{2}{*}{ $ACO$}                            & $f$                & 2.85  &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & 15    &       &        &       &       &       &   \\%
-\cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpACO$}                          & $f$                & 3.63  &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & 11    &       &        &       &       &       &   \\%
-
-\cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Gurobi$}                         & $f$                & 6.35  &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & 2     &       &        &       &       &       &   \\%
-
-\midrule
-\multirow{10}{*}{$2.0$} &\multirow{2}{*}{ $Shims$}  & $f$                &       &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
-\cmidrule{2-10}
-&\multirow{2}{*}{ $mpShims$}                        & $f$                &       &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & < 1   &       &        &       &       &       &   \\%
-
-\cmidrule{2-10}
-&\multirow{2}{*}{ $ACO$}                            & $f$                & 2.85  &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & 15    &       &        &       &       &       &   \\%
-\cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpACO$}                          & $f$                & 3.63  &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & 11    &       &        &       &       &       &   \\%
-
-\cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Gurobi$}                         & $f$                & 6.35  &       &        &       &       &       &   \\%
-&                                                   & {\bf run-time (s)} & 2     &       &        &       &       &       &   \\%
+&\multirow{2}{*}{ $mpShims$}                    & $f$                & {\bf 9.85} &  7.52    &  8.33  &  7.82     &       &       &   \\%
+&                                               & {\bf run-time (s)} &  < 1       &  < 1     &    2   &  8     &       &       &   \\%
+
+\midrule
 
@@
 \end{tabular}
-%\textsuperscript{The numbers 0.7s and 60s as superscript on the methods names represent the time limit per node.}	
+%\textsuperscript{$Gurobi^*$ was run with a time limit of 60s per node.}	
 \normalsize
@@
 
+$Gurobi^*$ was run with a time limit of 60s per node. The remaining methods were run with a time limit of 1.6s per node, the worst time {\it Shims} took to solve a node.
+
 \section{Conclusions}

### 2023-02-04 main.py

@@ -10,15 +10,16 @@ import mpACO
 import optcgcons
 import mipGRB
 
-shimsThreshold = 0.25 # best solume threshold
-acoThreshold   = 0.25
+volThreshold = 0.25 # best solume threshold
+volThreshold   = 0.25
 
 def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus):
     """
     Solves one tour
     """
-    print(f"----- Tour {pi},", end='')
+    writeConsFile = False
 
+    print(f"----- Tour {pi},", end='')
     for node in tour.nodes:
         print(f" {node.ICAO}", end='')
     print()
@@ -32,18 +33,22 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus)
 
     base = len(tour.nodes)-1
 
+    epsilom = 0.
+
     for k, node in enumerate(tour.nodes):  # solve each node sequentialy
+
 
         for i, p in enumerate(pallets):
             pallets[i].reset(cfg.numNodes)
 
-        # solution global torque to be shared and changed by all pallets concurrently
-        solTorque = mp.Value('d', 0.0) # a multiprocessing double type variable
+        # solution global torque to be shared and changed by all processes concurrently
+        nodeTorque = mp.Value('d', 0.0) # a multiprocessing double type variable
+        nodeVol    = 0.0
         
         # initialize- the accumulated values
-        sNodeAccum = 0.
-        wNodeAccum = 0.
-        vNodeAccum = 0.
+        if writeConsFile:
+            wNodeAccum = 0.
+            vNodeAccum = 0.
 
         # L_k destination nodes set
         unattended = [n.ID for n in tour.nodes[k+1:]]
@@ -74,6 +79,7 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus)
             # Pallets destinations are also set, according to kept on board in new positions
             # Kept P is not -2 anymore, but the pallet ID.
             if len(kept) > 0:
+                # This time, without OptCGCons, no infeasibilities were found.
                 # optcgcons.OptCGCons(kept, pallets, cfg.maxTorque, "GRB", k)
 
                 # N: number of items to embark
[…]

### 2023-02-04 mipGRB.py

@@ -7,7 +7,7 @@ import multiprocessing as mp
 import numpy as np
 
     
-def Solve( pallets, items, cfg, k, secBreak, solTorque, solDict, itemsDict ):
+def Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict ):
 
     # print("W-PCW\tV-PCV")
     # for p in pallets:
@@ -36,10 +36,7 @@ def Solve( pallets, items, cfg, k, secBreak, solTorque, solDict, itemsDict ):
             ]                                                  for i in set_M
         ]
          
-    # consolidated are included in items
-    itemsScore = xsum( X[i][j] * items[j].S for i in set_M for j in set_N )
-
-    mod.objective = maximize( itemsScore )
+    mod.objective = maximize( xsum( X[i][j] * items[j].S for i in set_M for j in set_N ) )
 
     # CONSTRAINTS ----------------------------------------------------------------------------
     
@@ -98,9 +95,9 @@ def Solve( pallets, items, cfg, k, secBreak, solTorque, solDict, itemsDict ):
     if mod.num_solutions:   
 
         # reset empty pallets torque
-        solTorque.value = 0.0
+        nodeTorque.value = 0.0
         for i in set_M:
-            solTorque.value += 140.0 * pallets[i].D         
+            nodeTorque.value += 140.0 * pallets[i].D         
 
             for j in set_N:
 
@@ -110,7 +107,7 @@ def Solve( pallets, items, cfg, k, secBreak, solTorque, solDict, itemsDict ):
 
                     solDict["solMatrix"][N*i+j] = 1
 
-                    solTorque.value += items[j].W * pallets[i].D
+                    nodeTorque.value += items[j].W * pallets[i].D
 
                     if itemsDict["mpItems"][j] == 0:
                         itemsDict["mpItems"][j] = 1

### 2023-02-04 mpACO.py

@@ -16,6 +16,30 @@ import copy
 
 import common
 
+def copyPallets(pallets):
+    array = [None for _ in pallets]
+    for i, p in enumerate(pallets):
+        array[i] = common.Pallet(p.ID, p.D, p.V, p.W, 1)
+        array[i].Dest = p.Dest
+        array[i].PCW  = p.PCW 
+        array[i].PCV  = p.PCV
+        array[i].PCS  = p.PCS
+    return array
+     
+def copySolDict(solDict):
+    N_M = len(solDict["solMatrix"])
+    solMatrix = mp.Array('i', [0 for _ in np.arange(N_M)] ) 
+    for pos, v in enumerate(solDict["solMatrix"]):
+        solMatrix[pos] = v
+    return dict(solMatrix=solMatrix)
+
+def copyItemsDict(itemsDict):
+    N = len(itemsDict["mpItems"])
+    mpItems = mp.Array('i', [0 for _ in np.arange(N)] ) 
+    for pos, v in enumerate(itemsDict["mpItems"]):
+        mpItems[pos] = v
+    return dict(mpItems=mpItems)
+
 # Process-based parallelism
 
 ALPHA = 1 # pheromone exponent
@@ -134,7 +158,7 @@ def antSolve(antPallets, items, cfg, k, secBreak, antTorque, antSolDict, Attract
                         values.pop(j)        
 
 
-def Solve( pallets, items, cfg, k, limit, secBreak, mode, solTorque, solDict, itemsDict ):
+def Solve( pallets, items, cfg, k, limit, secBreak, mode, nodeTorque, solDict, itemsDict ):
 
     # to control the general attractiveness for the tournament selection
     Attract = mp.Array('d', np.arange(len(items)))
@@ -166,15 +190,15 @@ def Solve( pallets, items, cfg, k, limit, secBreak, mode, solTorque, solDict, it
     print(f"{len(items)} items  {len(pallets)} pallets")
 
     for i, _ in enumerate(pallets):               
-        common.fillPallet(pallets[i], items, k, solTorque, solDict, cfg, limit, itemsDict, lock)
+        common.fillPallet(pallets[i], items, k, nodeTorque, solDict, cfg, limit, itemsDict, lock)
         bestScore  += pallets[i].PCS
         if abs(pallets[i].D) > maxD:
             maxD = abs(pallets[i].D)
     
-    initPallets   = common.copyPallets(pallets)              
-    initTorque    = solTorque
[…]

### 2023-02-05 mpShims.py
Test everything again. Still volume unfeasibilities.

@@ -61,7 +61,7 @@ def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict,
     sh = Shims(pallet, len(whip))
     Set = [sh]
 
-    # First Fit Decrease - equivalente ao KP, mas mais rápido
+    # First Fit Decrease - faster than KP
     whip.sort(key=lambda x: abs(x.V), reverse=True)
 
     for whipPos, item in enumerate(whip):
@@ -86,7 +86,7 @@ def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict,
             bestIndex = i
 
     for item in Set[bestIndex].Items:
-        if item != None:
+        if item != None and pallet.isFeasible(item, 1.0, k, nodeTorque,  cfg, itemsDict, lock):
             pallet.putItem(item, nodeTorque, solDict, N, itemsDict, lock)
 
 
@@ -110,7 +110,7 @@ def Solve(pallets, items, cfg, k, threshold, secBreak, mode, nodeTorque, solDict
         procs = [None for _ in pallets] # each pallets has its own process
 
         # parallel greedy phase
-        for i, p in enumerate(pallets):
+        for i, _ in enumerate(pallets):
             procs[i] = mp.Process( target=common.fillPallet, args=( pallets[i], items, k,\
                  nodeTorque, solDict, cfg, threshold, itemsDict, lock) )
             time.sleep(0.001)
@@ -137,19 +137,10 @@ def Solve(pallets, items, cfg, k, threshold, secBreak, mode, nodeTorque, solDict
             common.fillPallet( pallets[i], items, k, nodeTorque, solDict, cfg, threshold, itemsDict, lock) 
             getBestShims(      pallets[i], items, k, nodeTorque, solDict, cfg, surplus,   itemsDict, lock)
 
-    # local search itemsDict["mpItems"][j] == 0 and\
+    # local search
     counter = 0
     for i, _ in enumerate(pallets):
         counter += common.fillPallet( pallets[i], items, k, nodeTorque, solDict, cfg, 1.0, itemsDict, lock) 
-
-    # counter = 0
-    # N = len(items)
-    # for i, _ in enumerate(pallets):
-    #     for item in items:
-    #         j = item.ID
-    #         if pallets[i].isFeasible(items[j], threshold, k, nodeTorque, cfg, itemsDict, lock):
[…]

### 2023-02-07 latex/comp_or.tex
Test everything again. Still volume unfeasibilities.
Shim still weak

@@
 \cmidrule{4-9}		
-{\bf Surplus}           & {$method$}                &{\bf Results}       &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 6}&{\bf Speed-up} \\
+{\bf Surplus}           & {$method$}            &{\bf Results}       &{\bf 1}      &{\bf 2}    &{\bf 3}      &{\bf 4}     &{\bf 5}    &{\bf 6}&{\bf Speed-up} \\
 \toprule
 
-\multirow{6}{*}{$1.2$} &\multirow{2}{*}{ Gurobi}& $f$                & 9.75       &  6.34    &  7.61  &   7.43    &       &       &   \\%
-&                                               & {\bf run-time (s)} &  8         &   17     &  30    &  114     &       &       &   \\%
+\multirow{6}{*}{$1.2$} &\multirow{2}{*}{ Gurobi}& $f$                & {\bf 12.00} & {\bf 8.28} & {\bf 12.19} &  13.38     &       &       &   \\%
+&                                               & {\bf run-time (s)} &  8          &   19       &  35         &  133     &    315   &       &   \\%
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                      & $f$                & 7.00       &  4.20    &  4.12  &  3.54     &       &       &   \\%
-&                                               & {\bf run-time (s)} &  < 1       &   3      &    5   &   19    &       &       &   \\%
+&\multirow{2}{*}{ $Shims$}                      & $f$                & 8.02        &  6.00      &  6.00       &  5.57      &       &       &   \\%
+&                                               & {\bf run-time (s)} &  < 1        &   2        &    3        &   12       &   38    &       &   \\%
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                    & $f$                & {\bf 9.85} &  7.52    &  8.33  &  7.82     &       &       &   \\%
-&                                               & {\bf run-time (s)} &  < 1       &  < 1     &    2   &  8     &       &       &   \\%
+&\multirow{2}{*}{ $mpShims$}                    & $f$                & 8.58        &  6.87      &  10.90      & 12.99 &  {\bf }     &       &   \\%
+&                                               & {\bf run-time (s)} &  < 1        &   2        &    3        &  9        &    27   &       &   \\%
 
@@
 
-$Gurobi^*$ was run with a time limit of 60s per node. The remaining methods were run with a time limit of 1.6s per node, the worst time {\it Shims} took to solve a node.
+All methods were run with a time limit of 1.8s per node, the worst time {\it Shims} took to solve a node, considering all scenarios and instances.
 

### 2023-02-07 main.py

@@ -140,7 +140,7 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus)
 
         Y = np.reshape(solDict["solMatrix"], (-1, N)) # N number of items (columns)
 
-
+        nodeScore = 0
         for i, row in enumerate(Y):
             for j, X_ij in enumerate(row):
                 if X_ij:
@@ -156,12 +156,19 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus)
                         wNodeAccum += float(items[j].W)
                         vNodeAccum += float(items[j].V)
 
-                    tour.score += items[j].S
+                    # tour.score += items[j].S
+                    nodeScore  += items[j].S
                     nodeVol    += items[j].V
 
         nodeVol /= cfg.volCap
+        if nodeVol > 1.0: # if volume infeasible, no score
+            nodeScore /= nodeVol
 
         epsilon = nodeTorque.value/cfg.maxTorque
+        if abs(epsilon) > 1.0: # if torque infeasible, no score
+            nodeScore /= abs(epsilon)
+
+        tour.score += nodeScore
 
         tour.AvgVol    += nodeVol
         tour.AvgTorque += epsilon
@@ -214,14 +221,14 @@ if __name__ == "__main__":
     volThreshold = 0.95
 
     # scenarios = [1,2,3,4,5,6]
-    scenarios = [1]
+    scenarios = [5,6]
 
     surplus   = "data20"
     # surplus   = "data50"
     # surplus   = "data100"
 
-    # methods = ["Shims","mpShims","GRB"]
-    methods = ["Shims"]
+    methods = ["Shims","mpShims","GRB"]
+    # methods = ["Shims"]
     # methods = ["mpShims"]
     # methods = ["GRB"]
 
@@ -229,8 +236,8 @@ if __name__ == "__main__":
 
         for scenario in scenarios:
 
-            # instances = [1,2,3,4,5,6,7]
-            instances = [1]
+            instances = [1,2,3,4,5,6,7]
+            # instances = [1]
 
             cfg = common.Config(scenario)
             

### 2023-02-07 mpShims.py

@@ -133,13 +133,13 @@ def Solve(pallets, items, cfg, k, threshold, secBreak, mode, nodeTorque, solDict
 
         for i, _ in enumerate(pallets):
             common.fillPallet( pallets[i], items, k, nodeTorque, solDict, cfg, threshold, itemsDict, lock) 
-            # getBestShims(      pallets[i], items, k, nodeTorque, solDict, cfg, surplus,   itemsDict, lock)
+            getBestShims(      pallets[i], items, k, nodeTorque, solDict, cfg, surplus,   itemsDict, lock)
 
     # local search
-    # counter = 0
-    # for i, _ in enumerate(pallets):
-    #     counter += common.fillPallet( pallets[i], items, k, nodeTorque, solDict, cfg, 1.0, itemsDict, lock) 
-    # print(f"---> {counter} items inserted by the local search.")
+    counter = 0
+    for i, _ in enumerate(pallets):
+        counter += common.fillPallet( pallets[i], items, k, nodeTorque, solDict, cfg, 1.0, itemsDict, lock) 
+    print(f"---> {counter} items inserted by the local search.")
 
 
 if __name__ == "__main__":

### 2023-02-08 mpShims.py
Knapsack DP algorithm will be tested

@@ -1,7 +1,11 @@
 import numpy as np
 import multiprocessing as mp
 import time
+import math
+
+import knapsack as kp
 import common
+import optcgcons
 
 # A Shims fits in a pallet slack
 class Shims(object):
@@ -13,13 +17,13 @@ class Shims(object):
         self.SCT    = 0.     # shims current torque
         self.Items  = [None for _ in np.arange(whipLen)]
 
-    def putItem(self, item, whipPos):
+    def putItem(self, item, w):
                  
             self.SCW += item.W
             self.SCV += item.V
             self.SCS += item.S
             self.SCT += float(item.W) * float(self.Pallet.D)
-            self.Items[whipPos] = item
+            self.Items[w] = item
 
     def isFeasible(self, item, k, nodeTorque, cfg, lock): # check constraints
 
@@ -39,7 +43,7 @@ class Shims(object):
         return ret
 
 # create a set of shims for this pallet and selects the best shims
-def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict, lock):
+def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict, lock, tipo):
 
     maxVol = pallet.V * surplus
 
@@ -61,36 +65,52 @@ def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict,
     sh = Shims(pallet, len(whip))
     Set = [sh]
 
-    # First Fit Decrease - faster than KP
-    whip.sort(key=lambda x: abs(x.V), reverse=True)
+    if tipo == "KP":
+        capacity = math.floor((pallet.V - pallet.PCV)*100)
+        if capacity > 0:
+            volumes = [0]*len(whip)
+            scores  = [0]*len(whip)
+            for w, item in enumerate(whip):
+                volumes[w] = math.floor(item.V*100)
+                scores[w]  = item.S
+                    
+            indexes = kp.Solve(capacity, volumes, scores)
 
-    for whipPos, item in enumerate(whip):
-        newShims = True
-        for sh in Set:
-            if sh.isFeasible(item, k, nodeTorque, cfg, lock):
-                sh.putItem(item, whipPos)
-                newShims = False
-                break
+            for w, item in enumerate(whip):
+                if indexes[w] == 1 and pallet.isFeasible(item, 1.0, k, nodeTorque, cfg, itemsDict, lock):
[…]

### 2023-02-08 optcgcons.py

@@ -15,7 +15,6 @@ def OptCGCons(kept, pallets, maxTorque, k):
 
     torque1 = sum( X[i][j] * (kept[j].W *    pallets[i].D) for i in PalletsRange for j in KeptRange ) 
     torque2 = sum( X[i][j] * (kept[j].W * -1*pallets[i].D) for i in PalletsRange for j in KeptRange ) 
-
     mod.setObjective(torque1 + torque2)
 
     mod.ModelSense = GRB.MINIMIZE
@@ -31,18 +30,57 @@ def OptCGCons(kept, pallets, maxTorque, k):
 
     mod.optimize() 
 
-    tTotAccum = 0
+    torque = 0
     for i, _ in enumerate(pallets):
         pallets[i].Dest[k] = -1 # reset pallets destinations from this node
         for j in KeptRange:
             if X[i][j].x >= 0.99:
-                tTotAccum += float(kept[j].W) * pallets[i].D
+                torque += float(kept[j].W) * pallets[i].D
                 pallets[i].Dest[k] = kept[j].To
                 kept[j].P = i # put the consolidated in the best position to minimize torque                        
 
-    print(f"--- {common.CITIES[k]} CG deviation minimized {tTotAccum/maxTorque:.2f}")
+    print(f"--- {common.CITIES[k]} CG deviation minimized {torque/maxTorque:.2f}")
 
-    return tTotAccum
+    return torque
+
+# After solved, optimize consolidated positions to minimize CG deviation
+def OptCG(pallets, k, nodeTorque):
+
+    cons = [common.Item(p.ID, -2, p.PCW, p.PCS, p.PCV, k, p.Dest[k]) for p in pallets]
+
+    ConsRange    = range(len(cons))
+    PalletsRange = range(len(pallets))
+
+    mod = gp.Model()
+    mod.setParam('OutputFlag', 0)
+
+    X = [ [ mod.addVar(name=f"X[{i}],[{j}]", vtype=GRB.BINARY) for j in ConsRange ] for i in PalletsRange ]      
+
+    torque1 = sum( X[i][j] * (cons[j].W *    pallets[i].D) for i in PalletsRange for j in ConsRange ) 
+    torque2 = sum( X[i][j] * (cons[j].W * -1*pallets[i].D) for i in PalletsRange for j in ConsRange ) 
+    mod.setObjective(torque1 + torque2)
+
+    mod.ModelSense = GRB.MINIMIZE
+
+    for j in ConsRange:
+        mod.addConstr(
[…]

### 2023-02-09 latex/comp_or.tex
Knapsack DP algorithm will be tested
bETTER RESULTS

--- 2023-02-04 latex/comp_or.tex
@@
 
-We also created a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as { \color{red} it was the unique method that managed to solve all testing scenarios.}
+We also created a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it surpassed all Shims results and some of the Gurobi results, all methods under a time limit.
 
@@
                                                                       \cmidrule{4-9}		
-{\bf Surplus}           & {$method$}                &{\bf Results}       &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 6}&{\bf Speed-up} \\
+{\bf Surplus}         & {$method$}               & {\bf Results}      &{\bf 1}      &{\bf 2}     &{\bf 3}      &{\bf 4}      &{\bf 5}       &{\bf 6}      &{\bf Speed-up} \\
 \toprule
 
-\multirow{6}{*}{$1.2$} &\multirow{2}{*}{ Gurobi}& $f$                & 9.75       &  6.34    &  7.61  &   7.43    &       &       &   \\%
-&                                               & {\bf run-time (s)} &  8         &   17     &  30    &  114     &       &       &   \\%
+\multirow{7}{*}{$1.2$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 12.00       & {\bf 8.38} & {\bf 12.19} & 13.38       &  14.42      &           &   \\%
+                      &                          & {\it run-time (s)} &  8          &   14       &  25         &  95         &    217      &  1719       &   \\%
 
                       \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                      & $f$                & 7.00       &  4.20    &  4.12  &  3.54     &       &       &   \\%
-&                                               & {\bf run-time (s)} &  < 1       &   3      &    5   &   19    &       &       &   \\%
+                      &\multirow{2}{*}{ $Shims$} & $f$                & 8.37        &  6.40      &  7.68       &  8.05       &   7.72      & 6.74        & \\%
+                      &                          & {\it run-time (s)} &  < 1        &  < 1       &    3        &   11         &   27       &  342        &\\%
 
                       \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                    & $f$                & {\bf 9.85} &  7.52    &  8.33  &  7.82     &       &       &   \\%
-&                                               & {\bf run-time (s)} &  < 1       &  < 1     &    2   &  8     &       &       &   \\%
+                      &\multirow{2}{*}{ $mpShims$}& $f$               & {\bf 12.24} &  7.95      &  11.59      & {\bf 13.80} & {\bf 16.16} & {\bf 13.74} &   \\%
+                      &                           & {\it run-time (s)}&  < 1        &  < 1       &    2        &   6         &    15       &   165       &   \\%
 
@@
 \end{tabular}
-%\textsuperscript{$Gurobi^*$ was run with a time limit of 60s per node.}	
+%\textsuperscript{asdf sd .}	
 \normalsize
@@
 
-$Gurobi^*$ was run with a time limit of 60s per node. The remaining methods were run with a time limit of 1.6s per node, the worst time {\it Shims} took to solve a node.
+For {\it Gurobi} in {\bf Scenario 1}, 50\% of the results were optimal, but the remaining were suboptimal due to the time limit.
+This explains that {\it mpShims} surpassed {\it Gurobi} results.
+The same applies to scenarios 4, 5 and 6, where {\it Gurobi} obtained less than 30\% optimal results.
 

### 2023-02-09 testing.py

@@ -62,8 +62,9 @@ surplus = "data20"
 
 # method = "Shims"  
 # method = "mpShims"
-# method = "GRB"
-method = "GRB*"
+tipo = "FFD"
+
+method = "GRB"
 
 scenario = 1
 
@@ -232,10 +233,10 @@ for inst in instances:
     objValue = 0.0
  
     if method == "mpShims":
-        mpShims.Solve(pallets, items, cfg, k, limit, secBreak, "p", nodeTorque, solDict, itemsDict)
+        mpShims.Solve(pallets, items, cfg, k, limit, secBreak, "p", nodeTorque, solDict, itemsDict, tipo)
 
     if method == "Shims":            
-        mpShims.Solve(pallets, items, cfg, k, limit, secBreak, "s", nodeTorque, solDict, itemsDict)         
+        mpShims.Solve(pallets, items, cfg, k, limit, secBreak, "s", nodeTorque, solDict, itemsDict, tipo)         
 
     if method == "mpACO":       
         mpACO.Solve(pallets,   items, cfg, k, limit, secBreak, "p", nodeTorque, solDict, itemsDict) 
@@ -246,11 +247,7 @@ for inst in instances:
     if method == "GRB":       
         mipGRB.Solve(pallets,  items, cfg, k,        secBreak,      nodeTorque, solDict, itemsDict) 
     
-    if method == "GRB*":       
-        mipGRB.Solve( pallets, items, cfg, k,        secBreak,      nodeTorque, solDict, itemsDict, True) 
-
     elapsed = time.perf_counter() - startNodeTime
-
 
     print("ID\tDest\tPCW\tPCV\tPCS")
     for p in pallets:

### 2023-02-09 latex/comp_or.tex
Update the summary.

@@
 
-Transport aviation faces risks of cargo imbalance in the set of circumstances of aerial pickup and shipment of goods in a distribution system given the urgency required for loading for fast takeoff and mission accomplishment, notably in moments of crisis, public emergency, client contract short turnaround times, or any external forces for instant takeoff.
-Furthermore, there is no computer support available to aid load and trip programmers with the massive transportation demands in each location.
-
-Other dangers are made possible as a result, including poor delivery, excessive fuel consumption, and longer turnaround times than necessary.
-
-We developed an efficient solution by means of computer parallelism to the problem involving scheduling the loading and routing of an airplane in a tour of simultaneous pickup and delivery at intermediate hubs while taking into account a utility score, weight and balance rules, and fuel usage.
+Parallelism is crucial for mobile computers to be capable of doing complicated operations while conserving energy.
+By leveraging parallelism, handheld computers can divide tasks into numerous smaller sections and execute these parts simultaneously, so utilizing multiple cores of the processor and offering a substantial performance improvement.
+This enables for more complicated applications and speedier performance, making mobile computers more competent than ever before.
+
+In this work, we explore computer parallelism in solving the problem involving scheduling the loading and routing of an airplane in a tour of simultaneous pickup and delivery at intermediate hubs while taking into account a utility score, weight and balance rules, and fuel usage.
 
@@
 
-We also contributed by carrying out multiple experiments with the well-known Ant Colony Optimization on synthetic data based on real data from the {\it Brazilian Air Force}\/ transportation history and a new procedure to minimize the distance from the next node destined pallet to the ramp door.
-
-We also created a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it surpassed all Shims results and some of the Gurobi results, all methods under a time limit.
+We also contributed by carrying out multiple experiments with the Shims heuristic on synthetic data based on real data from the {\it Brazilian Air Force} transportation history and a new procedure to minimize the distance from the next node destined pallet to the ramp door.
+
+We converted Shims into a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it surpassed all Shims results and some of the integer programming results, all methods under a time limit to guarantee that a distribution plan is yielded in less than one hour.
 
@@
 
-\multirow{7}{*}{$1.2$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 12.00       & {\bf 8.38} & {\bf 12.19} & 13.38       &  14.42      &           &   \\%
+\multirow{7}{*}{$1.2$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 12.00       & {\bf 8.38} & {\bf 12.19} & 13.38       &  14.42      & 13.25          &   \\%
                       &                          & {\it run-time (s)} &  8          &   14       &  25         &  95         &    217      &  1719       &   \\%

### 2023-02-09 latex/comp_or.tex
Running data50 solutions

@@
 \cite{MesquitaCunha2011} presented a heuristic for a real problem of the {\it Brazilian Air Force}, which consists of defining transport routes with simultaneous collection and delivery from a central distribution terminal.
+They used the {\it Scatter Search} metaheuristic as solution method.
 
@@
 
-An important issue on meta-heuristics is that they are known for the hard work in tuning parameters.
-For this, when we use the word {\it empirically}\/ throughout the text, we want to say that we accomplished many tests with the hardest instances to fit these values.
-This does not guarantee best results for all instances, but, on average, overall results are acceptable.
-Another important observation is that our {\it ACO} implementations were designed based on traditional ideas from the literature \cite{Fidanova2006} and on the experiences learned during tests, plus the special features to process-based parallelism.
-
 To guarantee that performance comparisons are made correctly, our implementations use the same elements described below.
@@
 It is directly proportional to the score, and inversely proportional to the volume and the torque of each item.
-The 3000 value was empirically set to keep the attractiveness in the range [0, 1], to be in the same order of magnitude of the Ant Colony Optimization pheromone range.
 
@@
 
-Below, we present 2 implementations for $SolveNode$: {\it Multi-process Ant Colony Optimization}\/ ({\it mpACO})  and a new parallel heuristic that we propose called {\it Multi-process Shims (mpShims)}.
+Below, we present the implementation for $SolveNode$, the new parallel heuristic that we propose called {\it Multi-process Shims (mpShims)}.
 
@@
 They concluded that the PIR topology presented, in average, the smallest distance from the optimal solutions.
-
-It is important to say that the parallel ACO ({\it pACO}) is a different implementation where employs threads instead of processes to solve each ant solution in parallel (for a detailed information on parallel ACO see \cite{Pedemonte2011}).
-We chose {\it mpACO} for our approach because we employed the {\it Fully-Connected} topology which requires to share the pheromone array between the multi-process ants and the facility to handle the possibility of race condition.
-
-\subsubsection{Multi-process Ant Colony Optimization (mpACO)}
-
-In ACO \cite{Dorigo1992}, a population of $A$\/ ants performs independent search (sequential or in parallel), where each ant $a$\/ finds its own allocation graph $G^a$.
-Each edge $e^k_{ij}=(p_i,t^k_j)$\/ has a general attractiveness $ga^k_{ij} = (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$, where $\tau^k_{ij}$\/ is the pheromone of this edge, and $\alpha$\/ and $\beta$\/ are constants.
-If $e^k_{ij}$\/ is included in $G^a$, ant $a$\/ increases the pheromone $\tau^k_{ij}$.
-
-Algorithm \ref{alg:aco} presents this heuristic in its parallel multi-process version as we have implemented.
-
-$SolveNode(mpACO, ...)$ is the main process, while $AntSolve(...)$ is executed the child concurrent processes.
-
-\begin{algorithm}[H]
-
-	\caption{ Solves a node with the {\it mpACO} method } \label{alg:aco}
-	\begin{algorithmic}[1]
-
-		\Procedure{$SolveNode$}{$mpACO, k, G, numProcs,\ timeLim$}
-
-		\State Let $G(V_k, E^Q_k)$ \Comment{This is a partial solution with only consolidated on board}
-		\State $G^*(V_k, E^{N*}_k \cup E^Q_k), \_ \gets Greedy(k, G, 0.9)$  \label{aco:greedy} \Comment{This partial solution is completed by a greedy algorithm}
-		\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^*(V_k,E^{N*}_k \cup E^Q_k)$
-		\State $A \gets numProcs$ \label{aco:numProcs}
-		\State $ga^k_{ij}   \gets 0$, $1 \leq i \leq m$, $1 \leq j \leq n_k$  \label{aco:init1}
-		\State $\tau^k_{ij} \gets 1$, $1 \leq i \leq m$, $1 \leq j \leq n_k$   \label{aco:init2}
-		\State $\alpha    \gets 1.0$ \label{aco:alpha}
-		\State $\beta     \gets 4.0$ \label{aco:beta}
-		\State $\rho      \gets 0.2$ \label{aco:rho}
-		\State $stagnant \gets 0$
-		\While{ $stagnant < 3$ } \label{aco:while}
-
-			\State $AntsQueue \gets \{\ \}$ \label{aco:AntsQueue}
-
-			\For{$1$ {\bf to} $A$} \label{aco:AntsSolve}	
-				\State $AntSolve(G^+,\ AntsQueue,\ ga^k_{ij})$
-			\EndFor
-
-			\State $counter \gets A$
-
-			\While{$counter > 0$} \label{aco:while2}
-
-				\State $G^{a} \gets \varnothing$
-
-				\While{$runtime < timeLim$} \Comment{Wait until there is a solution in the queue}
-
-					\State $G^{a} \gets AntsQueue.Pop()$
-
-					\If{$G^{a} \ne \varnothing$}				
-						\Break
-					\EndIf
-
-				\EndWhile
-
-				\If{$f_s(G^{a}) > f_s(G^{+})$}
-					\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^a(V_k, E^{Na}_k \cup E^Q_k)$ 
-				\EndIf
-
-				\State $counter \gets counter - 1$
-
-			\EndWhile
-
-			\If{$f_s(G^+) > f_s(G^*)$} \label{aco:best1}
-				\State $stagnant \gets 0$
-				\State $G^*(V_k, E^{N*}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$		
-			\Else
-				\State $stagnant \gets stagnant + 1$ 
-			\EndIf
-			\For{all $\tau^k_{ij}$} 
-				\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{aco:rho2}
-				\State $\tau^k_{ij} \gets \tau^k_{ij} + \Delta\tau$ \label{aco:phero2}
-				\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{aco:ga2}				
-			\EndFor	
-		\EndWhile	
-		\Return $G^*(V_k,\ E^{N*}_k \cup E^Q_k)$
-
-		\EndProcedure		
-	\end{algorithmic}
-\end{algorithm}
-
-A greedy solution is obtained in line \ref{aco:greedy}, being initially stored in $G^*$\/ and $G^+$.
-The argument 0.9 generates an initial and poor solution for ACO to improve.
-The second return value denoted by underscore is not used by ACO.
-
-The number $A$\/ of ants was set to the number of parallel processes (line \ref{aco:numProcs}).
-
-The $m \times n_k$\/ matrices $\tau^k_{ij}$ and $ga^k_{ij}$\/ (lines \ref{aco:init1}-\ref{aco:init2}) are used and updated by all ants. For further explanation, see \cite{DorigoManiezzoColorni1996}.
-
-The number of iterations depends on the time limit or stagnation in improving results (line \ref{aco:while}).
-At the beginning os each iteration (\ref{aco:AntsQueue}) a queue of ants ($AntsQueue$) is created to hold ants parallel solutions generated in the child processes.
-
-The local solution $G^+$ is sent to A child processes (\ref{aco:AntsSolve}) almost simultaneously.
-While there are any ant working on an solution or the run-time is still less then maximum (\ref{aco:while2}), it hangs in the $AntsQueue$ until there is a solution available.
-
-If there is a global improvement (\ref{aco:best1}) the best so far solution ($G^*(V_k, E^{N*}_k \cup E^Q_k)$) is updated.
-At the end of each iteration pheromone levels (\ref{aco:rho2}) are evaporated, than updated with the new calculated $\Delta\tau$ (line \ref{aco:phero2}), and according to Equation \ref{eq:delta_tau}, and the general attractiveness ( \ref{aco:ga2}) is also updated.
-
-\begin{equation} \label{eq:delta_tau}
-	\Delta\tau = \frac{f_s(G^{+}) - f_s(G^{*})}{A \times f_s(G^{*})}
-\end{equation}
-
-Many ACO applications update pheromone levels and the general attractiveness only when there is a solution improvement.
-Differently, this work updates in both cases, exploring also the negative learning approach proposed by \cite{math9040361}.
-We did not compare these performances because it is out of the scope of this work.
 
@@
 The set $\eta_1$ is returned together with the greedy solution.
-
-\begin{algorithm}[H] 
-	\caption{Each ant solves a node in parallel} \label{alg:AntSolve}
-	\begin{algorithmic}[1]
-
-		\Procedure{$AntSolve$}{$G^+,\ AntsQueue,\ ga^k_{ij})$}
-
-		\State $G^a(V_k, E^{Na}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$
-		\State $E \gets \{ (p_i,t^k_j) \mid (p_i,t^k_j) \not\in E^{Na}_k \}$ \label{AntSolve:neighborhood}
-		\While{$|E| > 0$} \label{AntSolve:not_empty}
-
-			\State $e \gets Roulette(E)$ biased through $ga^k_{ij}$  \label{AntSolve:roulette}
-			\State $E \gets E - \{e\}$		
-			\If{$E^{Na}_k \cup \{e\}$ is feasible}
-				\State $ E^{Na}_k \gets E^{Na}_k \cup \{e\} $ \label{AntSolve:ant}
-			\EndIf
-
-		\EndWhile
-
-%\For{$\tau^k_{ij}, \theta^k_{ij} \in E^{Na}_k$}
-%\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{AntSolve:rho}
-%\State $\tau^k_{ij} \gets \tau^k_{ij} + \Delta\tau$ \label{AntSolve:phero}  \Comment{$\Delta\tau \gets$ Equation \ref{eq:delta_tau}}
-%\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{AntSolve:ga}		
-%\EndFor
-
-		\State $AntsQueue \gets AntsQueue \cup G^a(V_k,\ E^{Na}_k \cup E^Q_k)$ \label{AntSolve:put_in_queue}
-
-		\EndProcedure
-	\end{algorithmic}
-\end{algorithm}
-
-Algorithm \ref{alg:AntSolve}, which will be executed by a child parallel process, receives an initial and poor solution $G^+(V_k, E^{N+}_k \cup E^Q_k)$ generated greedily to be completed and improved by {\it mpACO}.
-
-Line \ref{AntSolve:neighborhood} expresses the creation of the neighborhood, the set of candidate edges.
-
-While the neighborhood is not empty (line \ref{AntSolve:not_empty}), choose an edge (line \ref{AntSolve:roulette}), delete it from the neighborhood and include it in the solution, if feasible (line \ref{AntSolve:ant}).
-
-Finally, put the solution in the queue (line \ref{AntSolve:put_in_queue}) to be retrieved by the main process.
-
-%Note that $ga^k_{ij}$ is updated by each parallel ant, which may influence the solution of the others during run-time.
-This fully connected topology may increase run-time due to the overhead in communications handled by {\it Python API Multiprocessing} but it is expected that this will yield better solutions.
 
@@
 
-%
-%\begin{figure}[H]
-%	\begin{center}
-%		\begin{tikzpicture}[scale=0.8, samples=100]
-%			\begin{axis}[
-%				myplotstyle,
-%				xlabel={\bf mpShims computation times (s)},
-%				ylabel={\bf Score},
-%				]
-%				\addplot+[
-%				only marks,	
-%				nodes near coords = {\numProcs{}},
-%				visualization depends on = {value \thisrow{np} \as \numProcs},						
-%				]
-%				table[x=time, y=score, col sep=comma] {./csv/Shims_p1data20.csv};
-%			\end{axis}
-%			\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-%		\end{tikzpicture}%
-%		~%
-%		%
-%		\begin{tikzpicture}[scale=0.8, samples=100]
-%		\begin{axis}[
-%			myplotstyle,
-%			xlabel={\bf mpACO computation times (s)},
-%			ylabel={\bf Score},
-%			]
-%			\addplot+[
-%			only marks,	
-%			nodes near coords = {\numProcs{}},
-%			visualization depends on = {value \thisrow{np} \as \numProcs},						
-%			]
-%			table[x=time, y=score, col sep=comma] {./csv/ACO_p1data20.csv};
-%		\end{axis}
-%		\fill [red] (canvas cs:x=0.1cm,y=5.6cm) circle (2pt);
-%	\end{tikzpicture}
-%		
-%	\end{center}
-%	\caption{ Scenario 1 performances with $1.2 \times volume$. }
-%	\label{fig:scenario1data20}
-%\end{figure}
-
 \vspace{2.0mm}
@@
 \cmidrule{2-10}		                       
-                      &\multirow{2}{*}{ $mpShims$}& $f$               & {\bf 12.24} &  7.95      &  11.59      & {\bf 13.80} & {\bf 16.16} & {\bf 13.74} &   \\%
-                      &                           & {\it run-time (s)}&  < 1        &  < 1       &    2        &   6         &    15       &   165       &   \\%
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 12.24} &  7.95       &  11.59      & {\bf 13.80} & {\bf 16.16} & {\bf 13.74}  &   \\%
+&                                                & {\it run-time (s)} &  < 1        &  < 1        &    2        &   6         &    15       &   165        &   \\%
+
+\midrule
+
+\multirow{7}{*}{$1.5$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 16.49       & 0        & 0 & 13.38       &  0      & 0        &   \\%
+&                                                & {\it run-time (s)} &  12         &   14        &  25         &  95         &    217      &  1719        &   \\%
+
+\cmidrule{2-10}		                       
+&\multirow{2}{*}{ $Shims$}                       & $f$                & 12.53       &  8.31       &  10.04      &  10.94      &   11.72     & 10.13        & \\%
+&                                                & {\it run-time (s)} &  < 1        &  3          &    6        &   24        &   70        &  419         &\\%
+
+\cmidrule{2-10}		                       
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 16.59} & {\bf 10.49} &  16.82      & 18.46        & 22.19      & {\bf 0}  &   \\%
+&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   10         &   27       &   165        &   \\%
 

### 2023-02-09 main.py

@@ -167,10 +167,12 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
         nodeVol /= cfg.volCap
         if nodeVol > 1.0: # if volume infeasible, no score
             nodeScore /= nodeVol
+            nodeVol = 1.0
 
         epsilon = nodeTorque.value/cfg.maxTorque
         if abs(epsilon) > 1.0: # if torque infeasible, no score
             nodeScore /= abs(epsilon)
+            epsilon = 1.0
 
         tour.score += nodeScore
 
@@ -225,10 +227,10 @@ if __name__ == "__main__":
     volThreshold = 0.92 # 0.92 best for scenario 1
 
     # scenarios = [1,2,3,4,5,6]
-    scenarios = [5]
+    scenarios = [2,3,4,5,6]
 
-    surplus   = "data20"
-    # surplus   = "data50"
+    # surplus   = "data20"
+    surplus   = "data50"
     # surplus   = "data100"
 
     methods = ["Shims","mpShims","GRB"]

### 2023-02-10 latex/comp_or.tex
data50 experiments completed

@@
 
+We ran Algorithm \ref{alg:main} in the 6 scenarios described in Table \ref{tab:scenarios}, considering 3 methods for node-by-node resolution: {\it Gurobi}, {\it Shims} (Algorithm \ref{alg:shims}), and {\it mpShims} (Algorithm \ref{alg:mpshims}).
+
+In generating the items in each node, we consider 3 values for the parameter ${\it surplus}$.
+The results obtained for the function $f$, with the corresponding runtime in seconds, are shown in Tables \ref{tab:results} (${\it surplus} = 1.2$, ${\it surplus} = 1.5$, and ${\it surplus} = 2.0$).
+
+For each $method$, $scenario$ and $surplus$, 7 different instances were generated. Therefore, 126 tests (6 scenarios $\times$ 3 volumes $\times$ 7 instances) were performed for each method.
+
+The average tour values were presented for $f$ and, for the runtime, the worst result obtained.
+To facilitate the comparison between the methods, we added a last column in these tables, where two values are indicated:
+\begin{itemize}
+	\item {\bf Normalized}: value between 0 and 1, which corresponds to the ratio between the sum of $f$\/ values obtained by the method in all scenarios and the sum of the best values obtained among all methods in all scenarios.
+The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
+	\item {\bf Speed-up}: ratio of the sums of the worst runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
+\end{itemize}
+
+In each $scenario$, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.
+
 \vspace{2.0mm}
@@
                                                  \cmidrule{4-9}		
-{\bf Surplus}         & {$method$}               & {\bf Results}      &{\bf 1}      &{\bf 2}      &{\bf 3}      &{\bf 4}      &{\bf 5}       &{\bf 6}      &{\bf Speed-up} \\
+{\it surplus}         & {$method$}               & {\bf Results}      &{\bf 1}      &{\bf 2}      &{\bf 3}      &{\bf 4}      &{\bf 5}       &{\bf 6}      &{\bf Speed-up} \\
 \toprule
 
-\multirow{7}{*}{$1.2$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 12.00       & {\bf 8.38}  & {\bf 12.19} & 13.38       &  14.42      & 13.25        &   \\%
-&                                                & {\it run-time (s)} &  8          &   14        &  25         &  95         &    217      &  1719        &   \\%
+\multirow{7}{*}{$1.2$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 12.00       & {\bf 8.38}  & {\bf 12.19} & 13.38       &  14.42      & 13.25        & 0.96 \\% 7651
+&                                                & {\it run-time (s)} &  8          &   14        &  25         &  95         &    217      &  1719        & 1.0  \\%2078
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 8.37        &  6.40       &  7.68       &  8.05       &   7.72      & 6.74         & \\%
-&                                                & {\it run-time (s)} &  < 1        &  < 1        &    3        &   11        &   27        &  342         &\\%
+&\multirow{2}{*}{ $Shims$}                       & $f$                & 8.37        &  6.40       &  7.68       &  8.05       &   7.72      & 6.74         & 0.56 \\%
+&                                                & {\it run-time (s)} &  < 1        &  < 1        &    3        &   11        &   27        &  342         & 5.4   \\%385
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 12.24} &  7.95       &  11.59      & {\bf 13.80} & {\bf 16.16} & {\bf 13.74}  &   \\%
-&                                                & {\it run-time (s)} &  < 1        &  < 1        &    2        &   6         &    15       &   165        &   \\%
-
-\midrule
-
-\multirow{7}{*}{$1.5$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 16.49       & 0        & 0 & 13.38       &  0      & 0        &   \\%
-&                                                & {\it run-time (s)} &  12         &   14        &  25         &  95         &    217      &  1719        &   \\%
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 12.24} &  7.95       &  11.59      & {\bf 13.80} & {\bf 16.16} & {\bf 13.74}  & {\bf 0.99} \\%
+&                                                & {\it run-time (s)} &  < 1        &  < 1        &    2        &   6         &    15       &   165        & {\bf 10.9}  \\% 190
+
+\midrule
+
+\multirow{7}{*}{$1.5$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 16.49       & {\bf 11.60} & {\bf 16.68} & 18.33       &  20.10      & 17.94        & 0.97  \\%10402
+&                                                & {\it run-time (s)} &  12         &   25        &  45         &  169        &   437       &  2377        & 1.0  \\%3065
+
+\cmidrule{2-10}		                       
+&\multirow{2}{*}{ $Shims$}                       & $f$                & 12.53       &  8.31       &  10.04      &  10.94      &   11.72     & 10.13        & 0.61 \\%
+&                                                & {\it run-time (s)} &  < 1        &  3          &    6        &   24        &   70        &  419         & 5.9 \\%523
+
+\cmidrule{2-10}		                       
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 16.71} & 10.86       & {\bf 16.68} & {\bf 18.84} & {\bf 21.30} & {\bf 19.39}  & {\bf 1.00}  \\%
+&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   9         &   26        &   158        & {\bf 15.4}  \\%199
+
+\midrule
+\multirow{7}{*}{$2.0$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 16.49       & {\bf 11.60} & {\bf 16.68} & 18.33       &  20.10      & 17.94        &   \\%
+&                                                & {\it run-time (s)} &  12         &   25        &  45         &  169        &   437       &  2377        &   \\%
 
@@
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 16.59} & {\bf 10.49} &  16.82      & 18.46        & 22.19      & {\bf 0}  &   \\%
-&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   10         &   27       &   165        &   \\%
-
-\midrule
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 16.71} & 10.86       & {\bf 16.68} & {\bf 18.84} & {\bf 21.30} & {\bf 19.39}  &   \\%
+&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   9         &   26        &   158        &   \\%
 
@@
 
-For {\it Gurobi} in {\bf Scenario 1}, 50\% of the results were optimal, but the remaining were suboptimal due to the time limit.
+For {\it Gurobi} in some scenarios, less than 50\% of the results were optimal, and the remaining were suboptimal due to the time limit.
 This explains that {\it mpShims} surpassed {\it Gurobi} results.
-The same applies to scenarios 4, 5 and 6, where {\it Gurobi} obtained less than 30\% optimal results.
 

### 2023-02-11 latex/comp_or.tex
data100 almost finished
Results updated on latex

@@
 
-We ran Algorithm \ref{alg:main} in the 6 scenarios described in Table \ref{tab:scenarios}, considering 3 methods for node-by-node resolution: {\it Gurobi}, {\it Shims} (Algorithm \ref{alg:shims}), and {\it mpShims} (Algorithm \ref{alg:mpshims}).
+We ran Algorithm \ref{alg:main} in the 6 scenarios described in Table \ref{tab:scenarios}, considering 3 methods for node-by-node resolution: the {\it Gurobi} MIP solver, {\it Shims}, and {\it mpShims} (Algorithm \ref{alg:mpshims}).
 
@@
 \midrule
-\multirow{7}{*}{$2.0$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 16.49       & {\bf 11.60} & {\bf 16.68} & 18.33       &  20.10      & 17.94        &   \\%
-&                                                & {\it run-time (s)} &  12         &   25        &  45         &  169        &   437       &  2377        &   \\%
+\multirow{7}{*}{$2.0$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 24.04       & {\bf 17.33} & 24.62       & {\bf 26.36} &  28.88      & 25.89        & 0.98  \\%14962
+&                                                & {\it run-time (s)} &  14         &   34        &  62         &  236        &   609       & 3431         & 1.0  \\%4386
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 12.53       &  8.31       &  10.04      &  10.94      &   11.72     & 10.13        & \\%
-&                                                & {\it run-time (s)} &  < 1        &  3          &    6        &   24        &   70        &  419         &\\%
+&\multirow{2}{*}{ $Shims$}                       & $f$                & 18.41       &  13.37       &  15.90      &  16.17      &   17.08     & 15.96        & 0.65 \\%
+&                                                & {\it run-time (s)} &  < 1        &  4          &    7        &   30         &   85        &  501         & 7.0 \\%628
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 16.71} & 10.86       & {\bf 16.68} & {\bf 18.84} & {\bf 21.30} & {\bf 19.39}  &   \\%
-&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   9         &   26        &   158        &   \\%
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 24.28} & 16.54       & {\bf 25.53} & 25.87        & {\bf 29.10} & {\bf 27.02}  & {\bf 0.99} \\%
+&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   12         &   32        &   188        & {\bf 18.4}  \\%238
 
@@
 
-For {\it Gurobi} in some scenarios, less than 50\% of the results were optimal, and the remaining were suboptimal due to the time limit.
-This explains that {\it mpShims} surpassed {\it Gurobi} results.
+For the MIP solver in some scenarios, less than 50\% of the results were optimal, and the remaining were suboptimal due to the time limit.
+This explains that {\it mpShims} surpassed the MIP solver results.
+
+In \cite{MesquitaSanches2023}, {\it Shims} surpassed the MIP solver, on average, because the time limit they used (0.7s) was sufficient to guarantee a worst-case solution in less than 30min.
+This decision caused most of the MIP solver results to be poor.
+In this work, we used as a time limit the worst {\it Shims} run-time, which was 1.8s.
+This made the MIP solver yield optimal or near-optimal solutions.
+This is good for the reader to have an idea of the real performance differences among the methods.
+
+{\it mpShims} presented the best performance, as it was slightly better in terms of quality than the MIP solver, and 15 times faster on average.
 

### 2023-02-12 latex/comp_or.tex
Dist from ramp door model infeasible

@@
 
+\subsection{A possible improvement}
+
+\cite{LurkinSchyns2015} considered an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem.
+But, as both aircrafts dealt with in this work have only the ramp door, we tried to minimize the distance of travel of pallets inside the cargo bay, favoring unloading in the next node, by a post optimization of pallets positions to favor this distance minimization.
+
 \section{Conclusions}

### 2023-02-12 main.py

@@ -10,7 +10,7 @@ import mpACO
 import optcgcons
 import mipGRB
 
-def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus, tipo, numOptDict, rampDistCG):
+def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus, tipo, numOptDict, rampDistCG, distRampDict):
     """
     Solves one tour
     """
@@ -138,19 +138,19 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
         if modStatus == 2: # 2: optimal
             numOptDict["numOpt"] += 1
 
+        optcgcons.OptRampDist(pallets, k, tour, rampDistCG, cfg)
+
         nodeElapsed = time.perf_counter() - startNodeTime
 
         tour.elapsed += nodeElapsed
 
         Y = np.reshape(solDict["solMatrix"], (-1, N)) # N number of items (columns)
-
-        nodeSumDistsFromRamp = 0
-        
+       
         nodeScore = 0
         for i, row in enumerate(Y):
             
             if pallets[i].Dest[k] == k+1:
-                nodeSumDistsFromRamp += rampDistCG - pallets[i].D # distance from the pallet to the ramp door
+                distRampDict['distRamp'] += rampDistCG - pallets[i].D # distance from the pallet to the ramp door
 
             for j, X_ij in enumerate(row):
                 if X_ij:
@@ -181,7 +181,8 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
 
         print(f"----- node {node.ICAO},", end='')
         print(f" score {tour.score:.0f}, cost {tour.cost:.0f}, vol {nodeVol:.2f}, torque {epsilon:.2f} -----")
-        print(f"nodeSumDistsFromRamp: {nodeSumDistsFromRamp}")
+        print(f"--- sum of dists from ramp: {distRampDict['distRamp']:.1f} in {node.ICAO}\n")
+
 
         if writeConsFile:
 
@@ -228,27 +229,27 @@ if __name__ == "__main__":
     volThreshold = 0.92 # 0.92 best for scenario 1
 
     # scenarios = [1,2,3,4,5,6]
-    scenarios = [6]
+    scenarios = [1]
 
-    # surplus   = "data20"
+    surplus   = "data20"
     # surplus   = "data50"
-    surplus   = "data100"
[…]

### 2023-02-12 testing.py

@@ -60,21 +60,19 @@ surplus = "data20"
 # method = "ACO"
 # method = "mpACO"
 
-# method = "Shims"  
+method = "Shims"  
 # method = "mpShims"
 tipo = "FFD"
 
-method = "GRB"
+# method = "GRB"
 
 scenario = 1
 
 # instances = [1,2,3,4,5,6,7]
 instances = [1]
 
-# limit    = 0.95 # best
-limit    = 0.95
-secBreak = 0.7 # seconds
-# secBreak = 60
+limit    = 0.92
+secBreak = 1.8 # seconds
 
 
 # --- distances and costs matrix ---
@@ -88,7 +86,7 @@ for i, cols in enumerate(dists):
     for j, dist in enumerate(cols):
         costs[i][j] = cfg.kmCost*dist
 
-pallets = common.loadPallets(cfg)
+pallets, rampDistCG = common.loadPallets(cfg)
 lock = mp.Lock
 
 # pallets capacities
@@ -142,8 +140,8 @@ for inst in instances:
 
     # N = first cons ID
     # cons = common.loadNodeCons(surplus, scenario, inst, pi, prevNode, N )
-    # cons = testingGetCons(N, True) # for testing only
-    cons = testingGetCons(N, False) # False: no randomness in consolidated generation
+    cons = testingGetCons(N, True) # for testing only
+    # cons = testingGetCons(N, False) # False: no randomness in consolidated generation
 
     # --- from main
     # cons = []
@@ -253,6 +251,8 @@ for inst in instances:
     for p in pallets:
         print(f"{p.ID}\t{p.Dest[k]}\t{p.PCW:.0f}\t{p.PCV:.2f}\t{p.PCS:.0f}")
     print("Pallets filled-up.\n")
+
+    optcgcons.OptRampDist(pallets, k, tour, rampDistCG, cfg)
 
     # Validate the solution for this node
 

### 2023-02-12 latex/comp_or.tex
Experimentation with dist from ramp door have started

@@
 
+Let $\pi$ be a tour in the set of permutations of nodes.
+
+Let $maxD$ be the distance from the last ramp pallet to the center of gravity.
+
+Let $Q  = \{ a^{\pi(k_1)}, a^{\pi(k_2)}, \ldots, a^{\pi(k_{m})} \}$ be the set of consolidated items assembled on the set of $m$ pallets in node $\pi(k)$.
+
+Let $Z^{\pi(k)}_{iq}$ be the set of decision variables that relates pallet $i$ to consolidated $q$ in node $\pi(k)$.
+
+\begin{equation} \label{eq:distobj}
+	\mbox{minimize} \sum_{i=1}^{m} \sum_{q=1}^{m} Z^{\pi(k)}_{iq} \times ( maxD - p_i.d) ,\ \mbox{{\bf if}}\ a^{\pi(k)}_q.to = \pi(k+1) \}
+\end{equation}
+
+\begin{equation} \label{eq:distcons}
+	s.t.: \sum_{i=1}^{m} Z^{\pi(k)}_{iq} = 1;\ q \in \{1,2,\ldots,m\}
+\end{equation}
+
+\begin{equation} \label{eq:distpallets}
+	s.t.: \sum_{q=1}^{m} Z^{\pi(k)}_{iq}=1;\ i \in \{1,2,\ldots,m\}
+\end{equation}
+
+\begin{equation} \label{eq:distpalletw}
+	p_i.w = 140 + a^{\pi(k)}_q.w ;\ i \in \{1,2,\ldots,m\};\ q \in \{1,2,\ldots,m\}
+\end{equation}
+
+\begin{equation} \label{eq:disttorque}
+	s.t.:\ | \sum\ p_i.d * p_i.w\ | \leq maxW \times limit^{CG}_{long} ;\ i \in \{1,2,\ldots,m\}
+\end{equation}
+
+Equation \ref{eq:distobj} minimizes the sum of distances of the consolidated in node $\pi(k)$ destined to the next node $\pi(k+1)$; Equation \ref{eq:distcons} states that each consolidated will be assigned to exactly one pallet; Equation \ref{eq:distpallets} states that each pallet will receive one consolidated; Equation \ref{eq:distpalletw} calculates the total weight of each pallet; and Equation \ref{eq:disttorque} guarantees that the torque limit will not be exceeded.
+
+As the number of pallets was small for an integer program solver, the increased time was negligible.
+
+Columns {\bf Before} and {\bf After} on Table \ref{tab:distRampDoor} refer to the application of the integer programming solver according to equations \ref{eq:distobj} to \ref{eq:disttorque}.
+
+\vspace{2.0mm}
+
+\begin{table}[H]
+\centering
+\caption{Distance from ramp door minimization}  \label{tab:distRampDoor}
+\footnotesize
+%\scriptsize
+\begin{tabular}{cccc}
+\toprule
+
+{\bf Surplus} & {\bf Scenario} & {\bf Before (m)} & {\bf After (m)} \\
+\midrule
+\multirow{6}{*}{$1.2$}&1 & 7.07 & 7.07 \\
+&                     2 & 41.23 & 37.52 \\
+&                     3 & 24.03 & 24.64 \\
+&                     4 & 26.84 & After \\
+&                     5 & Before & After \\
+&                     6 & Before & After \\
+\midrule
+\multirow{6}{*}{$1.5$}&1 & 7.07 & 7.07 \\
+&                     2 & 41.23 & 37.52 \\
+&                     3 & 24.03 & 24.64 \\
+&                     4 & 26.84 & After \\
+&                     5 & Before & After \\
+&                     6 & Before & After \\
+\midrule
+\multirow{6}{*}{$2.0$}&1 & 7.07 & 7.07 \\
+&                     2 & 41.23 & 37.52 \\
+&                     3 & 24.03 & 24.64 \\
+&                     4 & 26.84 & After \\
+&                     5 & Before & After \\
+&                     6 & Before & After \\
+\midrule
+&                     {\bf Average} & Before & After \\
+\bottomrule	
+\end{tabular}
+%\textsuperscript{asdf sd .}	
+\normalsize
+\end{table}
+
 \section{Conclusions}

### 2023-02-12 latex/comp_or.tex
Reviewing the mathematical formulation

@@
 
-Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modelling of ACLP+RPDP, which is one of the contributions of our work.
+Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modelling of ACLP+RPDP.
 
@@
 
-Let $S_K = \{s: \{1, \dots, K\} \rightarrow \{1, \dots, K\} \}$ be the set of $K!$ permutations, which correspond to all possible tours (or itineraries) that have $l_0$ as origin and end, passing through the others $K$ nodes.
+Let $S_K = \{s: \{1, \dots, K\} \rightarrow \{1, \dots, K\} \}$ be the set of $K!$ permutations, which correspond to all possible tours (or itineraries) that have $l_0$ as origin and end, passing through the other $K$ nodes.
 
@@
 
-		{\it Stopping criterion:} & All methods have the same time limit of $0.7$ seconds per node.\\
+		{\it Stopping criterion:} & All methods have the same time limit of $1.8$ seconds per node.\\
 
@@
 \end{equation}
-
-We use four types of random selections:
-\begin{itemize}
-	\item $RandomReal(r_1,r_2)$: randomly selects a real number in $[r_1,r_2]$;
-	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$;
-	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
-\end{itemize}
 
@@
 
-Let $Q  = \{ a^{\pi(k_1)}, a^{\pi(k_2)}, \ldots, a^{\pi(k_{m})} \}$ be the set of consolidated items assembled on the set of $m$ pallets in node $\pi(k)$.
+Let $Q^{\pi(k)}  = \{ a^{\pi(k)}_1, a^{\pi(k)}_2, \ldots, a^{\pi(k)}_m \}$ be the set of consolidated items assembled on the set of $m$ pallets in node $\pi(k)$.
 
@@
 
-As the number of pallets was small for an integer program solver, the increased time was negligible.
+As the number of pallets was small (7 and 18) for an integer program solver, the increased time was negligible (less than 1s).
 
@@
 %\scriptsize
-\begin{tabular}{cccc}
+\begin{tabular}{ccccc}
 \toprule
 
-{\bf Surplus} & {\bf Scenario} & {\bf Before (m)} & {\bf After (m)} \\
-\midrule
-\multirow{6}{*}{$1.2$}&1 & 7.07 & 7.07 \\
-&                     2 & 41.23 & 37.52 \\
-&                     3 & 24.03 & 24.64 \\
-&                     4 & 26.84 & After \\
-&                     5 & Before & After \\
-&                     6 & Before & After \\
-\midrule
-\multirow{6}{*}{$1.5$}&1 & 7.07 & 7.07 \\
-&                     2 & 41.23 & 37.52 \\
-&                     3 & 24.03 & 24.64 \\
-&                     4 & 26.84 & After \\
-&                     5 & Before & After \\
-&                     6 & Before & After \\
-\midrule
-\multirow{6}{*}{$2.0$}&1 & 7.07 & 7.07 \\
-&                     2 & 41.23 & 37.52 \\
-&                     3 & 24.03 & 24.64 \\
-&                     4 & 26.84 & After \\
-&                     5 & Before & After \\
-&                     6 & Before & After \\
-\midrule
-&                     {\bf Average} & Before & After \\
+{\it surplus} & {\bf Scenario} & {\bf Before (m)} & {\bf After (m)}  & {\bf Improvement (\%)}  \\
+\midrule
+\multirow{6}{*}{$1.2$}& 1 & 31.7  & 29.0  & 8.3\\
+&                       2 & 195.9 & 195.9 & 0.0\\
+&                       3 & 156.0 & 153.6 & 1.5 \\
+&                       4 & 143.3 & 143.3 & 0.0 \\
+&                       5 & 123.2 & 122.5 & 0.5 \\
+&                       6 & 106.5 & 105.7 & 0.8 \\
+\midrule
+\multirow{6}{*}{$1.5$}& 1 &  31.8 &  28.7 & 9.6 \\
+&                       2 & 197.0 & 197.0 & 0.0 \\
+&                       3 & 156.3 & 156.3 & 0.0 \\
+&                       4 & 143.8 & 143.8 & 0.0\\
+&                       5 & 125.0 & 124.2 & 0.7 \\
+&                       6 & 108.2 & 107.6 & 0.6\\
+\midrule
+\multirow{6}{*}{$2.0$}& 1 & 7.07 & 7.07 & \\
+&                       2 & 41.23 & 37.52 & \\
+&                       3 & 24.03 & 24.64 & \\
+&                       4 & 26.84 & After & \\
+&                       5 & Before & After & \\
+&                       6 & Before & After & \\
+\midrule
+&                         &        & {\bf Average} & 0.0\\
 \bottomrule	

### 2023-02-12 common.py

@@ -41,7 +41,7 @@ class Config(object):
 class Node(object):
     def __init__(self, id, tau):
         self.ID      = id
-        self.next    = 0
+        # self.next    = 0
         self.tau     = tau # node sum of torques
         self.ICAO = CITIES[0]
         if id < len(CITIES):
@@ -57,8 +57,8 @@ class Item(object):
         self.W  = w  # weight
         self.S  = s  # score
         self.V  = v  # volume
-        self.Frm = frm  # from
-        self.To = to # destination
+        self.Frm = frm  # origin node ID
+        self.To = to # destination node ID
         self.Attr = 0.0
 
 class Pallet(object):
@@ -67,7 +67,7 @@ class Pallet(object):
         self.D  = d  # centroid distance to CG
         self.V  = v  # volume threshold
         self.W  = w  # weight threshold
-        self.Dest = np.full(numNodes, -1)
+        self.Dest = np.full(numNodes, -1) # nodes IDs indexed by "k"
         self.PCW = 0 # pallet current weight
         self.PCV = 0.
         self.PCS = 0.
@@ -142,7 +142,7 @@ def loadPallets(cfg):
         pallets.append( Pallet(id, d, vol, wei, cfg.numNodes) )
         id += 1
    
-    return pallets, dists[0]
+    return pallets, dists[0] # ramp door distance from CG
 
 def fillPallet(pallet, items, k, nodeTorque, solDict, cfg, threshold, itemsDict, lock, torqueSurplus=1.0):
     N = len(items)
@@ -346,7 +346,7 @@ def loadNodeItems(scenario, instance, node, unatended, surplus): # unatended, fu
 def setPalletsDestinations(items, pallets, nodes, k, L_k):
 
     vols  = [0]*len(nodes)
-    max   = 0
+    max   = 0 # node with maximum volume demand
     total = 0
 
     # all items from all nodes

### 2023-02-12 optcgcons.py

@@ -33,6 +33,7 @@ def OptCGCons(kept, pallets, maxTorque, k):
     torque = 0
     for i, _ in enumerate(pallets):
         pallets[i].Dest[k] = -1 # reset pallets destinations from this node
+        torque += 140.0 * pallets[i].D
         for j in KeptRange:
             if X[i][j].x >= 0.99:
                 torque += float(kept[j].W) * pallets[i].D
@@ -56,8 +57,8 @@ def OptCG(pallets, k, nodeTorque):
 
     X = [ [ mod.addVar(name=f"X[{i}],[{j}]", vtype=GRB.BINARY) for j in ConsRange ] for i in PalletsRange ]      
 
-    torque1 = sum( X[i][j] * (cons[j].W *    pallets[i].D) for i in PalletsRange for j in ConsRange ) 
-    torque2 = sum( X[i][j] * (cons[j].W * -1*pallets[i].D) for i in PalletsRange for j in ConsRange ) 
+    torque1 = sum( X[i][j] * ((140+cons[j].W) *    pallets[i].D) for i in PalletsRange for j in ConsRange ) 
+    torque2 = sum( X[i][j] * ((140+cons[j].W) * -1*pallets[i].D) for i in PalletsRange for j in ConsRange ) 
     mod.setObjective(torque1 + torque2)
 
     mod.ModelSense = GRB.MINIMIZE
@@ -84,7 +85,7 @@ def OptCG(pallets, k, nodeTorque):
 
 
 # After solved, minimize the sum of ramp door distances for the next node pallets
-def OptRampDist(pallets, k, tour, rampDistCG, cfg):
+def OptRampDist(pallets, k, tour, rampDistCG, cfg, nodeTorque):
 
     palletWeights = [0 for _ in pallets]
 
@@ -95,6 +96,9 @@ def OptRampDist(pallets, k, tour, rampDistCG, cfg):
 
     ConsRange    = range(len(cons))
     PalletsRange = range(len(pallets))
+
+    for j in ConsRange:
+        palletWeights[j] = 140 + cons[j].W 
 
     mod = gp.Model()
     mod.setParam('OutputFlag', 0)
@@ -111,7 +115,6 @@ def OptRampDist(pallets, k, tour, rampDistCG, cfg):
         mod.addConstr(
             sum(X[i][j] for i in PalletsRange) == 1
         )
-        palletWeights[j] = 140 + cons[j].W   
     for i in PalletsRange:                
[…]

### 2023-02-12 latex/comp_or.tex
Paper update in the model

@@
 Let $S_K = \{s: \{1, \dots, K\} \rightarrow \{1, \dots, K\} \}$ be the set of $K!$ permutations, which correspond to all possible tours (or itineraries) that have $l_0$ as origin and end, passing through the other $K$ nodes.
+{\color{blue} $k$ is the node index and $\pi(k)$ is the node in position $k$ of tour $\pi$}.
 
 Let $M = \{p_1, p_2, \ldots, p_m \}$ the set of $m$ pallets.
-Each pallet $p_i$, $1 \leq i \leq m$, has weight capacity $p_i.w$, volume capacity $p_i.v$, pallet destinations $p_i.to[k]$, $0 \leq k \leq K$, and distance to the CG of aircraft $p_i.d$.
-$p_i.to[k]$ denotes that pallet $p_i$ may assume a different destination in each node $k$.
-
-Let $N_k = \{t^k_1, t^k_2, \ldots, t^k_{n_k} \}$ be the set of $n_k$ items to be loaded in node $l_k$, $0 \leq k \leq K$.
-Each item $t^k_j$, $1 \leq j \leq n_k$, has score $t^k_j.s$, weight $t^k_j.w$, volume $t^k_j.v$, and destination $t^k_j.to \in L_k$.
-Let $N = \bigcup_{0 \leq k \leq K} N_k$ be the set of items of all nodes along a tour.
-
-Let $Q_k = \{a^k_1, a^k_2, \ldots, a^k_{m_k} \}$ be the set of consolidated items loaded in $m_k \leq m$ pallets when the aircraft arrives at node $l_k$, with $0 \leq k \leq K$.
-$a^k_i$, $1 \leq i \leq m_k$, is the group of picked-up items that were allocated on pallet $p_i$ in some of the previous nodes.
-$a^k_i$ has total weight $a^k_i.w$, total volume $a^k_i.v$, and destination $a^k_i.to \in L_k \cup \{l_k\}$.
-If $a^k_i.to = l_k$, then $a^k_i$ is unloaded, and $p_i$ will be available for reloading; otherwise, $a^k_i$ remains on the aircraft, eventually in another pallet, and with items of $N_k$ having the same destination.
-
-Let $X_{ij}^k$ and $Y_{iq}^k$ be binary variables, where $0 \leq k \leq K$, $1 \leq j \leq n_k$, $1 \leq i \leq m$ and $1 \leq q \leq m_k$.
-$X_{ij}^k = 1$ if $t_j^k$ is assigned to $p_i$ in node $l_k$, and 0 otherwise.
-$Y_{iq}^k = 1$ if $a_q^k$ is assigned to $p_i$ in node $l_k$, and 0 otherwise.
+Each pallet $p_i$, $1 \leq i \leq m$, has weight capacity $p_i.w$, volume capacity $p_i.v$, pallet destinations $p_i.to[{\pi(k)}]$, $0 \leq k \leq K$, and distance to the CG of aircraft $p_i.d$.
+$p_i.to[{\pi(k)}]$ denotes that pallet $p_i$ may assume a different destination in each node ${\pi(k)}$.
+
+Let $N_{\pi(k)} = \{t^{\pi(k)}_1, t^{\pi(k)}_2, \ldots, t^{\pi(k)}_{n_{\pi(k)}} \}$ be the set of $n_{\pi(k)}$ items to be loaded in node $l_{\pi(k)}$, $0 \leq k \leq K$.
+Each item $t^{\pi(k)}_j$, $1 \leq j \leq n_{\pi(k)}$, has score $t^{\pi(k)}_j.s$, weight $t^{\pi(k)}_j.w$, volume $t^{\pi(k)}_j.v$, and destination $t^{\pi(k)}_j.to \in L_{\pi(k)}$.
+Let $N = \bigcup_{0 \leq k \leq K} N_{\pi(k)}$ be the set of items of all nodes along a tour.
+
+Let $Q_{\pi(k)} = \{a^{\pi(k)}_1, a^{\pi(k)}_2, \ldots, a^{\pi(k)}_{m_{\pi(k)}} \}$ be the set of consolidated items loaded in $m_k \leq m$ pallets when the aircraft arrives at node $l_{\pi(k)}$, with $0 \leq k \leq K$.
+$a^{\pi(k)}_i$, $1 \leq i \leq m_{\pi(k)}$, is the group of picked-up items that were allocated on pallet $p_i$ in some of the previous nodes.
+$a^{\pi(k)}_i$ has total weight $a^{\pi(k)}_i.w$, total volume $a^{\pi(k)}_i.v$, and destination $a^{\pi(k)}_i.to \in L_{\pi(k)} \cup \{l_{\pi(k)}\}$.
+If $a^{\pi(k)}_i.to = l_{\pi(k)}$, then $a^{\pi(k)}_i$ is unloaded, and $p_i$ will be available for reloading; otherwise, $a^{\pi(k)}_i$ remains on the aircraft, eventually in another pallet, and with items of $N_{\pi(k)}$ having the same destination.
+
+Let $X_{ij}^{\pi(k)}$ and $Y_{iq}^{\pi(k)}$ be binary variables, where $0 \leq k \leq K$, $1 \leq j \leq n_{\pi(k)}$, $1 \leq i \leq m$ and $1 \leq q \leq m_{\pi(k)}$.
+$X_{ij}^{\pi(k)} = 1$ if $t_j^{\pi(k)}$ is assigned to $p_i$ in node $l_{\pi(k)}$, and 0 otherwise.
+$Y_{iq}^{\pi(k)} = 1$ if $a_q^{\pi(k)}$ is assigned to $p_i$ in node $l_{\pi(k)}$, and 0 otherwise.
 By definition, $Y_{iq}^0=0$.
-Allocations of items or consolidated items to pallets in node $l_k$ can be seen as a bipartite graph $G_k(V_k, E_k)$, where $V_k = M \cup N_k \cup Q_k$, $E_k = E^N_k \cup E^Q_k$, $(p_i, t_j^k) \in E^N_k$ if $X_{ij}^k = 1$, and $(p_i, a_q^k) \in E^Q_k$ if $Y_{iq}^k = 1$.
+Allocations of items or consolidated items to pallets in node $l_{\pi(k)}$ can be seen as a bipartite graph $G_{\pi(k)}(V_{\pi(k)}, E_{\pi(k)})$, where $V_{\pi(k)} = M \cup N_{\pi(k)} \cup Q_{\pi(k)}$, $E_{\pi(k)} = E^N_{\pi(k)} \cup E^Q_{\pi(k)}$, $(p_i, t_j^{\pi(k)}) \in E^N_{\pi(k)}$ if $X_{ij}^{\pi(k)} = 1$, and $(p_i, a_q^{\pi(k)}) \in E^Q_{\pi(k)}$ if $Y_{iq}^{\pi(k)} = 1$.
 
@@
 \begin{equation} \label{eq:scores}
-	\tilde{s} = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.s
+	\tilde{s} = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.s
 \end{equation}
@@
 \begin{equation} \label{eq:costs}
-	\tilde{c} = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} [ c_{\pi(k), \pi(k+1)}\times(1+c_g\times|\epsilon_k|) ] + c_{\pi(K),0}\times(1+c_g\times|\epsilon_K|)
+	\tilde{c} = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} [ c_{\pi(k), \pi(k+1)}\times(1+c_g\times|\epsilon_{\pi(k)}|) ] + c_{\pi(K),0}\times(1+c_g\times|\epsilon_K|)
 \end{equation}
@@
 \begin{equation} \label{eq:tau}
-\tau_k = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w +  \sum_{q=1}^{m} Y_{iq}^k \times a_q^k.w)];\ k \in \{0, 1, \ldots, K\}
+\tau_{\pi(k)} = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w +  \sum_{q=1}^{m} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w)];\ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:eps}
-\epsilon_k = \frac{\tau_k}{maxW \times limit^{CG}_{long}};\ k \in \{0, 1, \ldots, K\}
+\epsilon_{\pi(k)} = \frac{\tau_{\pi(k)}}{maxW \times limit^{CG}_{long}};\ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp11}
-	L_0 = L; L_{k} = L_{k-1} - \{l_{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
+	L_0 = L; L_{\pi(k)} = L_{{\pi(k-1)}} - \{l_{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp12}
-	X_{ij}^k = 0 \ \mbox{if} \ t_j^k.to \notin L_k; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}; \ k \in \{1,2, \ldots, K\}
+	X_{ij}^{\pi(k)} = 0 \ \mbox{if} \ t_j^{\pi(k)}.to \notin L_{\pi(k)}; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_{\pi(k)}\}; \ k \in \{1,2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp9}
-	Y_{iq}^k = 0 \ \mbox{if} \ a_i^k.to \notin L_k; \ i \in \{1, 2, \ldots, m\}; \ q \in \{1, 2, \ldots, m_k\}; \ k \in \{ 1, 2, \ldots, K\}
+	Y_{iq}^{\pi(k)} = 0 \ \mbox{if} \ a_i^{\pi(k)}.to \notin L_{\pi(k)}; \ i \in \{1, 2, \ldots, m\}; \ q \in \{1, 2, \ldots, m_{\pi(k)}\}; \ k \in \{ 1, 2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:cons2}
-	a_i^{\pi(k+1)}.w = \sum_{j=1}^{n_k} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w;  \ i \in \{1, 2, \ldots, m_k\}; k \in \{0, 1, \ldots, K-1\}
+	a_i^{\pi(k+1)}.w = \sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_{\pi(k)}} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w;  \ i \in \{1, 2, \ldots, m_{\pi(k)}\}; k \in \{0, 1, \ldots, K-1\}
 \end{equation}
@@
 \begin{equation} \label{eq:cons3}
-	a_i^{\pi(k+1)}.v = \sum_{j=1}^{n_k} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.v + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.v;  \ i \in \{1, 2, \ldots, m_k\}; k \in \{0, 1, \ldots, K-1\}
+	a_i^{\pi(k+1)}.v = \sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.v + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.v;  \ i \in \{1, 2, \ldots, m_{\pi(k)}\}; k \in \{0, 1, \ldots, K-1\}
 \end{equation}
@@
 \begin{equation} \label{eq:cons5}
-	a_i^{\pi(k+1)}.to = t^{\pi(k)}_j.to  \ \mbox{if} \ X^{\pi(k)}_{ij} = 1  \ \mbox{and} \ t^{\pi(k)}_j.to \in L_{k+1};  \ i \in \{1, 2, \ldots, m_k\}; \ j \in \{1, 2, \ldots, n_k\}; \ k \in \{0,1, \ldots, K-1\}
+	a_i^{\pi(k+1)}.to = t^{\pi(k)}_j.to  \ \mbox{if} \ X^{\pi(k)}_{ij} = 1  \ \mbox{and} \ t^{\pi(k)}_j.to \in L_{\pi(k+1)};  \ i \in \{1, 2, \ldots, m_{\pi(k)}\}; \ j \in \{1, 2, \ldots, n_{\pi(k)}\}; \ k \in \{0,1, \ldots, K-1\}
 \end{equation}
@@
 \begin{equation} \label{eq:LatIt}
-	LatIt_k = \sum_{i=1}^{m} \sum_{j=1}^{n_k} ( X_{ij}^k \times t_j^k.w \times (i\%2) - X_{ij}^k \times t_j^k.w \times (i+1)\%2 ); \ k \in \{0, 1, \ldots, K\}
+	LatIt_{\pi(k)} = \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi(k)}} ( X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w \times (i\%2) - X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w \times (i+1)\%2 ); \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:LatCons}
-	LatCons_k =  \sum_{i=1}^{m} \sum_{q=1}^{m_k}  ( Y_{iq}^k \times a_q^k.w \times (i\%2) - Y_{iq}^k \times a_q^k.w \times (i+1)\%2); \ k \in \{0, 1, \ldots, K\}
+	LatCons_{\pi(k)} =  \sum_{i=1}^{m} \sum_{q=1}^{m_{\pi(k)}}  ( Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w \times (i\%2) - Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w \times (i+1)\%2); \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:torqlat}
-	s.t.: d^{CG}_{pallet} \times | LatIt_k + LatCons_k | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: d^{CG}_{pallet} \times | LatIt_{\pi(k)} + LatCons_{\pi(k)} | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:torqlong}
-	s.t.: |\tau_k| \leq maxW \times limit^{CG}_{long};\ k \in \{0, 1, \ldots, K\}
+	s.t.: |\tau_{\pi(k)}| \leq maxW \times limit^{CG}_{long};\ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:payload}
-	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w ) \leq maxW; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^k.w ) \leq maxW; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app2}
-	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m_k\}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m_{\pi(k)}\}; \ {\pi(k)} \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app3}
-	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.v + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m_k\}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.v + \sum_{q=1}^{m_{\pi(k)}} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m_{\pi(k)}\}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app4}
-	s.t.: \sum_{i=1}^{m} X_{ij}^k \leq 1; \ j \in \{1, 2, \ldots, n_k\}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{i=1}^{m} X_{ij}^{\pi(k)} \leq 1; \ j \in \{1, 2, \ldots, n_{\pi(k)}\}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app5}
-	s.t.:  Y_{iq}^k = 1 \ \mbox{if} \ a^k_q.to \in L_k; \ q \in \{1, 2, \ldots, m_k\}; \ k \in \{0, 1, \ldots, K\}
+	s.t.:  Y_{iq}^{\pi(k)} = 1 \ \mbox{if} \ a^{\pi(k)}_q.to \in L_{\pi(k)}; \ q \in \{1, 2, \ldots, m_{\pi(k)}\}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp8}
-	s.t.: p_i.to[k] = t^k_j.to\ \mbox{if} \ X_{ij}^k = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}; \ k \in \{1, 2, \ldots, K\}
+	s.t.: p_i.to[{\pi(k)}] = t^{\pi(k)}_j.to\ \mbox{if} \ X_{ij}^{\pi(k)} = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp2}
-	s.t.:  p_i.to[k] = a^k_q.to\ \mbox{if} \ Y_{iq}^k = 1; \ i \in \{1, 2, \ldots, m\};\ q \in \{1, 2, \ldots, m_k\}; \ k \in \{1, 2, \ldots, K\}
+	s.t.:  p_i.to[{\pi(k)}] = a^{\pi(k)}_q.to\ \mbox{if} \ Y_{iq}^{\pi(k)} = 1; \ i \in \{1, 2, \ldots, m\};\ q \in \{1, 2, \ldots, m_{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
 \end{equation}
@@
 \midrule
-\multirow{6}{*}{$2.0$}& 1 & 7.07 & 7.07 & \\
-&                       2 & 41.23 & 37.52 & \\
-&                       3 & 24.03 & 24.64 & \\
-&                       4 & 26.84 & After & \\
-&                       5 & Before & After & \\
-&                       6 & Before & After & \\
-\midrule
-&                         &        & {\bf Average} & 0.0\\
+\multirow{6}{*}{$2.0$}& 1 &  32.8 & 30.3 & 7.7 \\
+&                       2 & 198.6 & 198.6 & 0.0 \\
+&                       3 & 163.9 & 163.9 & 0.0 \\
+&                       4 & 143.5 & 143.5 & 0.0 \\
+&                       5 & 129.1 & 128.5 & 0.5 \\
+&                       6 & 112.6 & 111.2 & 1.3 \\
+\midrule
+&                         &        & {\bf Average} & {\bf 1.75}\\
 \bottomrule	

### 2023-02-13 latex/comp_or.tex
Experimentation finished. Writing paper.

@@
 \usepackage{amsthm}
+%\usepackage{amsmath}
 
@@
 6]{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in $l_0$, and no items in $l_1$.
-Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is NP-complete.
+Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is NP-hard.
 
 Real cases are more complex as they have hundreds of different items in each node and involve three intractable sub-problems: APP, WBP and SPDP.
-Through the mathematical modelling presented in the previous section, we verify that {\it Mixed-Integer Programming}\/ (MIP) is not able to solve these cases in feasible time.
+Through the mathematical modeling presented in the previous section, we verify that {\it Mixed-Integer Programming}\/ (MIP) is not able to solve these cases in feasible time.
 Thus, it is necessary to adopt some strategy to find a viable solution, not necessarily optimal, that seeks to maximize the objective function $f$.
@@
 As a result, if we have fast node-by-node solutions that allow us to construct a complete tour, we will be able to test all possible $K!$\/ tours and thus select the one that provides the best value for the $f$\/ function.
+
+{\color{blue} Each node, except the base, inherits information from the previous nodes (pallets that must remain on board), and must solved taking into account the remaining not visited nodes.
+One more reason for a node-by-node approach.}
 
@@
 
-		\Procedure{$ACLP+RPDP$}{$scenario,\ surplus,\ numProcs,\ timeLim$}
+		\Procedure{$ACLP+RPDP$}{$scenario,\ surplus,\ timeLim$}
 
@@
 			\For {each $\pi \in S_K$} \label{main:loop1}
-				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, numProcs,\ timeLim )$ \label{main:method}
+				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, \ timeLim )$ \label{main:method}
 			\EndFor \label{main:loop2}
@@
 
-The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $l_k$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n_k} t^k_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.v$).
+The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $l_{\pi(k)}$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n_{\pi(k)}} t^{\pi(k)}_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.v$).
 This parameter is passed to $ItemsGeneration$ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section.
@@
 When the aircraft is at node $l_0$, the initial graph $G_1$\/ is empty because it has no consolidated items \ref{tour:g11}.
-Otherwise, the set $L_k$\/ of remaining nodes is updated (line \ref{tour:lk1}), and $UpdateConsolidated$\/ (line \ref{tour:dest}) returns the set of consolidated items that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation.
+Otherwise, the set $L_k$\/ of remaining nodes when departing from $\pi(k)$ is updated (line \ref{tour:lk1}), and $UpdateConsolidated$\/ (line \ref{tour:dest}) returns the set of consolidated items that have not yet reached their destinations and remain on board, rearranging them on the pallets to minimize CG deviation.
 This allocation is stored in graph $G_1$\/ (line \ref{tour:g12}).
@@
 In the context of this work, we know that $m>K$, once the aircraft has $7$\/ or $18$\/ pallets and $K\leq 6$, allowing there to be at least one pallet for each node to be visited.
-$SetPalletsDestination$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node, without changing the pallets destination with consolidated items.
+$SetPalletsDestination$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands caused by the items to be embarked in the current node, without changing the pallets destination with consolidated items.
 
@@
 \begin{algorithm}[H]
-	\caption{ Solves the sequence of nodes in the tour indexed by $\pi$ }  \label{alg:tour}
+	\caption{ Solves the sequence of nodes of tour $\pi$ }  \label{alg:tour}
 
@@
 
-		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method, numProcs,\ timeLim$}
+		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method, timeLim$}
 
@@
 			\For{$i \gets 1$ to $m$}
-				\State $p_i.to[k] \gets -1$ \label{tour:-1}
+				\State $p_i.to[k] \gets -1$ \label{tour:-1} \Comment{reset this pallet destination}
 			\EndFor	
@@
 			\State $M \gets SetPalletsDestination( \pi(k) )$ \label{tour:dest2}		
-			\State $G_2 \gets SolveNode(method, \pi(k), G_1, numProcs,\ timeLim)$ \label{tour:node}
+			\State $G_2 \gets SolveNode(method, \pi(k), G_1, timeLim)$ \label{tour:node}
 			\State $s, \epsilon \gets ScoreAndDeviation(\pi(k), G_2)$ \label{tour:analyse}
@@
 
-\subsection{Node-by-node solutions}
+\subsection{Node-by-node resolution}
 \label{methods}
 
-In this subsection we present implementations of $SolveNode(...)$, where $method$\/ corresponds to a heuristic, $k$\/ is the index of the current node and $G$\/ is the allocation graph of the consolidated items that remain on board at node $l_k$.
+In this subsection we present implementations of $SolveNode(...)$, where $method$\/ corresponds to a heuristic, $k$\/ is the index of the current node $l_{\pi(k)}$ and $G$\/ is the allocation graph of the consolidated items that remain on board at $l_{\pi(k)}$.
 
@@
 
-The selection of edges for $E^N_k$\/ uses the {\it edge attractiveness}\/ $\theta^k_{ij}$, Equation \ref{eq:edge}, which can be understood as the tendency to allocate the item $t^k_j$\/ to the pallet $p_i$.
+The selection of edges for $E^N_{\pi(k)}$\/ uses the {\it edge attractiveness}\/ $\theta^{\pi(k)}_{ij}$, Equation \ref{eq:edge}, which can be understood as the tendency to allocate the item $t^{\pi(k)}_j$\/ to the pallet $p_i$.
 It is directly proportional to the score, and inversely proportional to the volume and the torque of each item.
@@
 \begin{equation} \label{eq:edge}
-	\theta^k_{ij}= \frac{t^k_j.s}{t^k_j.v \times 3000}\times(1-\frac{t^k_j.w\times|p_i.d|}{\max_s\{t^k_s.w\}\times\max_q\{|p_q.d|\}});\ i \in \{1,2,\ldots,m\},\ j \in \{1,2,\ldots,n_k\}
+	\theta^{\pi(k)}_{ij}= \frac{t^{\pi(k)}_j.s}{t^{\pi(k)}_j.v \times 3000}\times(1-\frac{t^{\pi(k)}_j.w\times|p_i.d|}{\max_s\{t^{\pi(k)}_s.w\}\times\max_q\{|p_q.d|\}});\ i \in \{1,2,\ldots,m\},\ j \in \{1,2,\ldots,n_k\}
 \end{equation}
@@
 
-In the pseudo-codes presented below, given an allocation graph $G$\/ for the node $l_k$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation(k,G) $, described in \cite{MesquitaSanches2023}.
+In the pseudo-codes presented below, given an allocation graph $G$\/ for the node $l_{\pi(k)}$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation({\pi(k)},G) $, described in \cite{MesquitaSanches2023}.
 
@@
 
-We also developed Algorithm \ref{alg:greedy}, which generates a greedy allocation of the items available in node $l_k$, according to the non-ascending order of $\theta^k_{ij}$, and considering the consolidated items already shipped ($E^Q_k$).
-Edges are included in this allocation as long as they respect feasibility constraints.
-Furthermore, the volume of each pallet $p_i$\/ cannot exceed $p_i.v \times limit$, where $ 0 < limit \leq 1$\/ is a given parameter.
-
-\begin{algorithm}[H]
-	\caption{Mount a greedy solution until the volume limit for each pallet}  \label{alg:greedy}
-
-	\begin{algorithmic}[1]
-
-		\Procedure{$Greedy$}{$k, G, limit$}
-
-		\State Let $G(V_k, E^Q_k)$
-
-		\State $\eta_1 \gets \{0\} \times m$ \label{greedy:eta}
-		\State $volume \gets \{0\} \times m$
-
-		\State $E^N_k \gets \varnothing$
-
-		\For{$q \gets 1$ to $m$}
-			\If{$(p_i, a_q^k) \in E^Q_k$} 
-				\State $volume[q] \gets volume[q] + a_q^k.v$ 
-			\EndIf		
-		\EndFor
-
-		\For {each $e_{ij}^k$ in non-ascending order of $\theta^k_{ij}$}
-			\If{($E^N_k \cup \{e_{ij}^k\}$ is feasible) {\bf and} ($volume[i] \leq p_i.v \times limit$)} 
-				\State $E^N_k \gets E^N_k \cup \{e_{ij}^k\}$ 
-				\State $volume[i] \gets volume[i] + t_j^k.v$
-				\State $\eta_1[i] \gets \eta_1[i] + 1$ 	
-			\EndIf
-		\EndFor
-
-		\Return $G(V_k, E^N_k \cup E^Q_k),\ \eta_1$ \Comment{return the solution and the last item indexes array }
-
-		\EndProcedure
-
-	\end{algorithmic}
-\end{algorithm}
-
-In Algorithm \ref{alg:greedy}, line \ref{greedy:eta} defines an array with the same size as pallets array to save the set of last edges inserted in the greedy solution.
-The set $\eta_1$ is returned together with the greedy solution.
-
-\subsubsection{Multi-process Shims - mpShims}
+\subsection{Multi-process Shims - mpShims}
 
@@
 			\node at (7, 0.3) {$\eta_2[i]$};
-			\node at (0.3, 5) {$\theta_{ij}^k$};
+			\node at (0.3, 5) {$\theta_{ij}^{\pi(k)}$};
 			\draw[->] (.6, .2) -- coordinate (y axis mid) (0.6, 5.3);
-			\node at (12, 0.3) {$e_{ij}^k$};
+			\node at (12, 0.3) {$e_{ij}^{\pi(k)}$};
 			\draw[smooth, domain = 0.09:2, color=black, thick] plot (.3+1/\x,{4.2+log2(\x)});
@@
 
-		\captionof{figure}{$n_k$\/ edges $e_{ij}^k$\/ of $p_i$\/ sorted by $\theta_{ij}^k$\/ in non-ascending order}
+		\captionof{figure}{$n_{\pi(k)}$\/ edges $e_{ij}^{\pi(k)}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi(k)}$\/ in non-ascending order}
 		\label{fig:whip}		
@@
 
-Figure \ref{fig:whip} represents the $n_k$\/ possible edges $e_{ij}^k$\/ of $p_i$\/ sorted by $\theta_{ij}^k$.
-\emph{mpShims}\/ starts with a greedy solution, stopping at the edge with index in $\eta_1[i]$ close to the local optimum (first phase).
+Figure \ref{fig:whip} represents the $n_{\pi(k)}$\/ possible edges $e_{ij}^{\pi(k)}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi(k)}$.
+\emph{Shims}\/ starts with a greedy solution, stopping at the edge with index in $\eta_1[i]$ close to the local optimum (first phase).
 Then, considering even the edge with the index $\eta_2[i]$, it elaborates different possible complements for this pallet (second phase) and selects the best of these complements (third phase).
 
+{\color{blue} In this work, we included a new phase between the first and the second: a minimization of the CG deviation.
+An integer programming procedure that helps improve the overall {\it mpShims} results.
+
+$1^{st}$ - Greedy phase
+
+$2^{st}$ - Minimization of CG deviation
+
+$3^{st}$ - Shims construction
+
+$4^{st}$ - best Shims selection
+
+All executed in a multi-process runtime, accessing concurrently, reading and writing, the list of items and the current value of the CG deviation.
+
+As to the multi-process runtime, we considered each pallet as a process.
+}
+
+\subsection{mpShims}
+
 \begin{algorithm}[H]
-	\caption{Solve a node with {\it mpShims}}  \label{alg:mpshims}
+	\caption{Mount a greedy solution until the volume limit for each pallet}  \label{alg:mpShims}
+
 	\begin{algorithmic}[1]
 
-		\Procedure{$SolveNode$}{$mpShims, k, G, numProcs,\ timeLim$}
-
-		\State $G^* \gets G(V_k, E^Q_k)$ \label{best_so_far}
-
-		\State Sort $M$ by $|p_i.d|$ in non-descending order \label{mpshims:pallets}
-
-		\State $E^N_k \gets \{\}$ \label{mpshims:items}
-
-		\State Let $E[1..m][1..n_k]$, where $E[i]$ is an array with $n_k$ edges $(p_i,t^k_j)$ sorted by $\theta_{ij}^k$ in non-ascending order, $1 \leq i \leq m$ \label{mpshims:edges}
-
-		\State $limits \gets getLimits(0.75,\ 0.98,\ numProcs)$ \label{mpshims:limits}
-
-		\State $shimsQueue \gets \{\}$ \label{mpshims:queue}
-
-		\For{$lim \in limits$}
-			\State $shimsSolve(G(V_k, E^Q_k),\ M,\ shimsQueue,\ lim,\ E)$ \label{mpshims:send}
+		\Procedure{$mpShims$}{${N_{\pi(k)}, \pi(k)}, limit, G$}
+
+		\State Let $G(E^{Q_{\pi(k)}})$ \Comment{the solution comes with some consolidated on board}
+
+		\State $surplus \gets 1 + 3 \times (1 - limit)$
+
+		\State $nodeTorque \gets 0$
+
+		\State $ts \gets 2.0$ \Comment{torque surplus} \label{mpShims:ts}
+
+		\State $lock \gets$ multi-processing lock \Comment{for shared data among processes to avoid race condition}
+
+		\State $proc \gets \{p_1,p_2,..., p_m\}$  \Comment{each pallet has its own process}
+
+		\For{$i \gets 1$ to $m$}
+		\State $proc_i \gets Greedy(p_i,\ N_{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ limit,\ lock,\ ts )$
+		\State $proc_i.start$
 		\EndFor
 
-		\State $counter \gets |limits|$
-
-		\While{$counter > 0$}
-
-			\State $G \gets \varnothing$
-
-			\While{$runtime < timeLim$} \Comment{Wait until there is a solution in the queue}
-
-			\State $G \gets shimsQueue.Pop()$
-				\If{$G \ne \varnothing$}
-					\Break
-				\EndIf
-			\EndWhile
-
-			\If{$f_s(G) > f_s(G^*)$}
-				\State $G^* \gets G$ 
-			\EndIf
-			\State $counter \gets counter - 1$
-
-		\EndWhile
-
-		\Return $G^*(V_k, E^N_k \cup E^Q_k)$ \label{mpshims:return}
+		\For{$i \gets 1$ to $m$}
+		\State $proc_i.join$
+		\EndFor
+
+		\State $minCGdev(M,\ \pi(k),\ nodeTorque)$  \label{mpShims:minCGdev} \Comment{minimize the CG deviation.}
+
+		\For{$i \gets 1$ to $m$}
+		\State $proc_i \gets getBestShims(p_i,\ N_{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ surplus,\ lock,\ ts )$
+		\State $proc_i.start$
+		\EndFor
+
+		\For{$i \gets 1$ to $m$}
+		\State $proc_i.join$
+
+		\EndFor
+
+		\Return $G(E^{N_{\pi(k)}} \cup E^{Q_{\pi(k)}})$
 
 		\EndProcedure
+
 	\end{algorithmic}
@@
 
-Algorithm \ref{alg:mpshims} describes the {\it SolveNode} method when it receives {\it mpShims} as argument.
-
-Line \ref{best_so_far} defines the best solution so far which is formed by a semi-loaded aircraft where pallets have been unloaded and some remain on board ($E^Q_k$) with consolidated items.
-
-The Shims solution starts with the set of pallets closer to the center of gravity (\ref{mpshims:pallets}).
-
-A new empty set of edges is created in line \ref{mpshims:items}.
-
-Edges with items in node "k" to be loaded into not full pallets are defined in line \ref{mpshims:edges}.
-
-Line \ref{mpshims:limits} creates a list of unique limits with values between 0.75 and 0.98, that will be sent to different parallel processes to generate diverse solutions.
-
-A queue of shims solutions to be used by the child processes to send solution to the main process is defined in line \ref{mpshims:queue}.
-
-The iteration in \ref{mpshims:send} publishes or triggers parallel processes to solve $G(V_k, E^Q_k)$.
-
-In line \ref{mpshims:wait} the program hangs until each child process put a solution into the queue. Each solution is collected by the main process and compared with the best so far.
-
-Finally, in line \ref{mpshims:return}, the best solution found is returned.
+Line \ref{mpShims:ts} defines the torque surplus ($ts$).
+For this phase, the solution may generate torque infeasible solutions with doubled torque, which will be repaired by the next phase, the minimization of the CG deviation.
+
+Line \ref{mpShims:minCGdev} presents the procedure to minimize the CG deviation and repair the infeasible greedy solution.
+
+\subsubsection{Shims first phase}
+
+Algorithm \ref{alg:greedy} generates a greedy allocation of the items available in node $l_{\pi(k)}$, according to the non-ascending order of $\theta^{\pi(k)}_{ij}$, and considering the consolidated items already shipped ($E^Q_{\pi(k)}$).
+Edges are included in this allocation as long as they respect feasibility constraints.
+Furthermore, the volume of each pallet $p_i$\/ cannot exceed $p_i.v \times limit$, where $ 0 < limit \leq 1$\/ is a given parameter.
 
 \begin{algorithm}[H]
-	\caption{Shims solving method to be run in parallel}  \label{alg:shimsSolve}
+	\caption{Mount a greedy solution until the volume limit for each pallet}  \label{alg:greedy}
+
 	\begin{algorithmic}[1]
 
-		\Procedure{$shimsSolve$}{$G,\ M,\ shimsQueue,\ limit,\ E$}
-
-		\State $G(V_k,\ E^Q_k \cup E^N_k),\ \eta_1 \gets Greedy(k,\ G(V_k,\ E^Q_k),\ limit)$ \label{first_phase}
-
-		\For{$i \gets 1$ to $m$}
-
-			\State $\eta_2[i] \gets \eta_1[i]$
-
-			\State $volume \gets 0$
-
-			\Repeat \label{second_phase_begin}
-				\State $\eta_2[i] \gets \eta_2[i] + 1$ 
-				\State $e^k_{ij} \gets E[i][\eta_2[i]]$
-				\State $volume \gets volume + t^k_j.v$
-			\Until{($\eta_2[i] = n_k$) {\bf or} ($ volume \geq (2 - limit)*p_i.v)$)} \label{second_phase_end}
-
-			\State $slack[i] \gets (1-limit) \times p_i.v$
-
-			\State $E^N_k \gets E^N_k \cup getBestShims(i,\ \eta_1,\ \eta_2,\ E,\ k,\ slack[i])$ \label{third_phase}
-
+		\Procedure{$Greedy$}{${\pi(k)}, G, limit$}
+
+		\State Let $G(V_{\pi(k)}, E^Q_{\pi(k)})$
+
+		\State $\eta_1 \gets \{0\} \times m$ \label{greedy:eta}
+		\State $volume \gets \{0\} \times m$
+
+		\State $E^N_{\pi(k)} \gets \varnothing$
+
+		\For{$q \gets 1$ to $m$}
+		\If{$(p_i, a_q^{\pi(k)}) \in E^Q_{\pi(k)}$} 
+		\State $volume[q] \gets volume[q] + a_q^{\pi(k)}.v$ 
+		\EndIf		
 		\EndFor
 
-		\State $shimsQueue \gets shimsQueue \cup \{G(V_k, E^N_k \cup E^Q_k)\}$ \label{enqueue}
+		\For {each $e_{ij}^{\pi(k)}$ in non-ascending order of $\theta^{\pi(k)}_{ij}$}
+		\If{($E^N_{\pi(k)} \cup \{e_{ij}^{\pi(k)}\}$ is feasible) {\bf and} ($volume[i] \leq p_i.v \times limit$)} 
+		\State $E^N_{\pi(k)} \gets E^N_{\pi(k)} \cup \{e_{ij}^{\pi(k)}\}$ 
+		\State $volume[i] \gets volume[i] + t_j^{\pi(k)}.v$
+		\State $\eta_1[i] \gets \eta_1[i] + 1$ 	
+		\EndIf
+		\EndFor
+
+		\Return $G(V_{\pi(k)}, E^N_{\pi(k)} \cup E^Q_{\pi(k)}),\ \eta_1$ \Comment{return the partial solution and the last item indexes array }
 
@@
 
-Algorithm \ref{alg:shimsSolve}, line \ref{first_phase} corresponds to Shims' first phase. Lines  \ref{second_phase_begin}-\ref{second_phase_end} correspond to the Shims' second phase.
-Line \ref{third_phase} corresponds to Shims' third phase. In the last line, \ref{enqueue}, {\it mpShims} puts the solution into the queue to be retrieved by the main process.
+In Algorithm \ref{alg:greedy}, line \ref{greedy:eta} defines an array with the same size as pallets array to save the set of last edges inserted in the greedy solution.
+The set $\eta_1$ is returned together with the greedy solution.
+
+\subsubsection{Shims second phase: improvement on minimizing CG deviation}
+
+Between the greedy and the {\it Shims} selection phases we inserted, in this work, a method to minimize the CG deviation, as stated by the model that follows.
+
+Let $\pi$ be a tour in the set of permutations of nodes.
+
+Let $Q^{\pi(k)}  = \{ a^{\pi(k)}_1, a^{\pi(k)}_2, \ldots, a^{\pi(k)}_m \}$ be the set of consolidated items assembled on the set of $m$ pallets in node $\pi(k)$.
+There is a bi-univocal relation between pallets and consolidated.
+
+Let $H^{\pi(k)}_{iq}$ be the set of decision variables that relates pallet $i$ to consolidated $q$ in node $\pi(k)$.
+
+\begin{equation} \label{eq:distobjshims}
+	\mbox{minimize}\ \Bigg |  \sum_{i=1}^{m} \sum_{q=1}^{m} H^{\pi(k)}_{iq} \times  (140 + a^{\pi(k)}_q.w) * p_i.d\ \Bigg | 
+\end{equation}
+
+\begin{equation} \label{eq:distconsshims}
+	s.t.: \sum_{i=1}^{m} H^{\pi(k)}_{iq} = 1;\ q \in \{1,2,\ldots,m\}
+\end{equation}
+
+\begin{equation} \label{eq:distpalletsshims}
+	s.t.: \sum_{q=1}^{m} H^{\pi(k)}_{iq}=1;\ i \in \{1,2,\ldots,m\}
+\end{equation}
+
+Equation \ref{eq:distobjshims} minimizes the aircraft absolute torque in node $\pi(k)$, where 140 is the weight of an empty pallet, $a^{\pi(k)}_q.w$ the weight of the consolidated, and $p_i.d$ the distance between the pallet centroid to the aircraft CG; Equation \ref{eq:distconsshims} states that each consolidated will be assigned to exactly one pallet; and Equation \ref{eq:distpalletsshims} states that each pallet will receive one consolidated.
+
+\subsubsection{Shims third and fourth phases}
 
 \begin{algorithm}[H]
-	\caption{Get the best shims of edges that fills each pallet gap}  \label{alg:getBestShims}
+	\caption{Mount shims of edges that fills each pallet gap and return the best shims}  \label{alg:getBestShims}
 	\begin{algorithmic}[1]
@@
 
-\multirow{7}{*}{$1.2$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 12.00       & {\bf 8.38}  & {\bf 12.19} & 13.38       &  14.42      & 13.25        & 0.96 \\% 7651
+\multirow{7}{*}{$1.2$}&\multirow{2}{*}{ $Gurobi$}& $f$                & {\bf 12.00} & {\bf 8.38}  & {\bf 12.19} & {\bf 13.38} & {\bf 14.42} & 13.25        & {\bf 0.99} \\% 7412
 &                                                & {\it run-time (s)} &  8          &   14        &  25         &  95         &    217      &  1719        &  1.0       \\%2078
@@
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 8.37        &  6.40       &  7.68       &  8.05       &   7.72      & 6.74         & 0.56 \\%
-&                                                & {\it run-time (s)} &  < 1        &  < 1        &    3        &   11        &   27        &  342         & 5.4   \\%385
+&\multirow{2}{*}{ $Shims$}                       & $f$                & 9.66        &  6.37       &  9.58       &  10.87      &   12.24     & 11.94        &  0.82\\%
+&                                                & {\it run-time (s)} &  < 1        &  3          &    5        &   11        &   27        & 342          &  5.3\\%389
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 12.24} &  7.95       &  11.59      & {\bf 13.80} & {\bf 16.16} & {\bf 13.74}  & {\bf 0.99} \\%
-&                                                & {\it run-time (s)} &  < 1        &  < 1        &    2        &   6         &    15       &   165        & {\bf 10.9}  \\% 190
-
-\midrule
-
-\multirow{7}{*}{$1.5$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 16.49       & {\bf 11.60} & {\bf 16.68} & 18.33       &  20.10      & 17.94        & 0.97  \\%10402
-&                                                & {\it run-time (s)} &  12         &   25        &  45         &  169        &   437       &  2377        & 1.0  \\%3065
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & 11.90       &  7.61       &  10.86      & 13.13       & 13.75       & {\bf 13.75}  &  0.96\\%
+&                                                & {\it run-time (s)} &  < 1        &  < 1        &    3        &  9          &  26         & 156          &  {\bf 10.6}\\%
+
+\midrule
+
+\multirow{7}{*}{$1.5$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 16.49       & {\bf 11.60} & {\bf 16.68} & {\bf 18.33} & {\bf 20.10} & 17.94        & {\bf 0.99}\\%10209
+&                                                & {\it run-time (s)} &  12         &   25        &  45         &  169        &   437       &  2377        & 1.0\\%3065
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 12.53       &  8.31       &  10.04      &  10.94      &   11.72     & 10.13        & 0.61 \\%
-&                                                & {\it run-time (s)} &  < 1        &  3          &    6        &   24        &   70        &  419         & 5.9 \\%523
+&\multirow{2}{*}{ $Shims$}                       & $f$                & 13.71       &  9.06       &  13.54      &  15.21      &   17.09     &  17.15       & 0.84\\%
+&                                                & {\it run-time (s)} &  < 1        &  3          &    6        &   24        &   70        &  419         & 5.9\\%
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 16.71} & 10.86       & {\bf 16.68} & {\bf 18.84} & {\bf 21.30} & {\bf 19.39}  & {\bf 1.00}  \\%
-&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   9         &   26        &   158        & {\bf 15.4}  \\%199
-
-\midrule
-\multirow{7}{*}{$2.0$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 24.04       & {\bf 17.33} & 24.62       & {\bf 26.36} &  28.88      & 25.89        & 0.98  \\%14962
-&                                                & {\it run-time (s)} &  14         &   34        &  62         &  236        &   609       & 3431         & 1.0  \\%4386
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 16.85} & 10.65       & {\bf 16.68} & 17.05       & 19.72       & {\bf 18.53}  & 0.98\\%
+&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   10        &   28        &   173        & {\bf 14.1}\\%
+
+\midrule
+\multirow{7}{*}{$2.0$}&\multirow{2}{*}{ $Gurobi$}& $f$                & {\bf 24.04} & {\bf 17.33} & {\bf 24.62} & {\bf 26.36} & {\bf 28.88} & {\bf 25.89}  & {\bf 1.00}\\%14712
+&                                                & {\it run-time (s)} &  14         &   34        &  62         &  236        &  609        & 3431         & 1.0\\%4836
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 18.41       &  13.37       &  15.90      &  16.17      &   17.08     & 15.96        & 0.65 \\%
-&                                                & {\it run-time (s)} &  < 1        &  4          &    7        &   30         &   85        &  501         & 7.0 \\%628
+&\multirow{2}{*}{ $Shims$}                       & $f$                & 20.42       &  13.89      &  20.36      &  22.71      &   25.13     & 24.69        & 0.86\\%
+&                                                & {\it run-time (s)} &  < 1        &  4          &    7        &   30        &   85        &  501         & 7.7\\%
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 24.28} & 16.54       & {\bf 25.53} & 25.87        & {\bf 29.10} & {\bf 27.02}  & {\bf 0.99} \\%
-&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   12         &   32        &   188        & {\bf 18.4}  \\%238
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & 23.12       & 14.89       & 22.08       & 23.96       & 26.98       & 25.29        &  0.93\\%
+&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   12        &   33        &  203         & {\bf 19.0}\\%
 
@@
 For the MIP solver in some scenarios, less than 50\% of the results were optimal, and the remaining were suboptimal due to the time limit.
-This explains that {\it mpShims} surpassed the MIP solver results.
+This explains that {\it mpShims} surpassed some of the MIP solver results.
 
 In \cite{MesquitaSanches2023}, {\it Shims} surpassed the MIP solver, on average, because the time limit they used (0.7s) was sufficient to guarantee a worst-case solution in less than 30min.
-This decision caused most of the MIP solver results to be poor.
+This decision caused most of the MIP solver results to be poor, making {\it Shims} outstanding as it was the unique method to perform well with all scenarios and instances.
+
 In this work, we used as a time limit the worst {\it Shims} run-time, which was 1.8s.
@@
 
-{\it mpShims} presented the best performance, as it was slightly better in terms of quality than the MIP solver, and 15 times faster on average.
-
-\subsection{A possible improvement}
+On average, {\it mpShims} is 14 times faster than the MIP solver, and it reached 96\% of {\it Gurobi} $f$ values, on average, which is good for a tool of operational use.
+The use of parallelism is another important contribution of this work.
+It has improved {\it Shims} results and also its speed.
+
+\subsection{Yet another improvement}
 
@@
 
-As the number of pallets was small (7 and 18) for an integer program solver, the increased time was negligible (less than 1s).
+As the number of pallets was small (7 or 18) for an integer program solver, the increased time was negligible (less than 1s).
 
 Columns {\bf Before} and {\bf After} on Table \ref{tab:distRampDoor} refer to the application of the integer programming solver according to equations \ref{eq:distobj} to \ref{eq:disttorque}.
+Column {\bf Improvement} is the percentage of improvements in distances minimization.
 
@@
 \centering
-\caption{Distance from ramp door minimization}  \label{tab:distRampDoor}
+\caption{Minimization of pallets distances from the ramp door}  \label{tab:distRampDoor}
 \footnotesize
@@
 \midrule
-\multirow{6}{*}{$1.2$}& 1 & 31.7  & 29.0  & 8.3\\
-&                       2 & 195.9 & 195.9 & 0.0\\
-&                       3 & 156.0 & 153.6 & 1.5 \\
-&                       4 & 143.3 & 143.3 & 0.0 \\
-&                       5 & 123.2 & 122.5 & 0.5 \\
-&                       6 & 106.5 & 105.7 & 0.8 \\
-\midrule
-\multirow{6}{*}{$1.5$}& 1 &  31.8 &  28.7 & 9.6 \\
-&                       2 & 197.0 & 197.0 & 0.0 \\
-&                       3 & 156.3 & 156.3 & 0.0 \\
-&                       4 & 143.8 & 143.8 & 0.0\\
-&                       5 & 125.0 & 124.2 & 0.7 \\
-&                       6 & 108.2 & 107.6 & 0.6\\
-\midrule
-\multirow{6}{*}{$2.0$}& 1 &  32.8 & 30.3 & 7.7 \\
-&                       2 & 198.6 & 198.6 & 0.0 \\
-&                       3 & 163.9 & 163.9 & 0.0 \\
-&                       4 & 143.5 & 143.5 & 0.0 \\
-&                       5 & 129.1 & 128.5 & 0.5 \\
-&                       6 & 112.6 & 111.2 & 1.3 \\
-\midrule
-&                         &        & {\bf Average} & {\bf 1.75}\\
+\multirow{6}{*}{$1.2$}& 1 &  33  &  16 & 52\\
+&                       2 & 194  & 154 & 21\\
+&                       3 & 149  & 119 & 20 \\
+&                       4 & 135  & 103 & 23 \\
+&                       5 & 121  &  84 & 31 \\
+&                       6 & 105  &  71 & 33 \\
+\midrule
+\multirow{6}{*}{$1.5$}& 1 &  32 &  18 & 44 \\
+&                       2 & 198 & 135 & 32 \\
+&                       3 & 156 & 128 & 18 \\
+&                       4 & 137 & 106 & 22\\
+&                       5 & 120 &  87 & 28 \\
+&                       6 & 106 &  77 & 27 \\
+\midrule
+\multirow{6}{*}{$2.0$}& 1 &  33 &  20 & 38 \\
+&                       2 & 199 & 142 & 28 \\
+&                       3 & 156 & 122 & 22 \\
+&                       4 & 138 & 106 & 23 \\
+&                       5 & 123 &  84 & 31 \\
+&                       6 & 109 &  75 & 31 \\
+\midrule
+&                         &        & {\bf Average} & {\bf 29}\\
 \bottomrule	
@@
 
+This last procedure favored ground operations as it reduced in 29\% the distances pallets will be moved on disembarking operations.
+This last step is not part of {\it mpShims} and may be applied as a post improvement for any method.
+This is another important contribution of this work.
+
 \section{Conclusions}

### 2023-02-13 main.py

@@ -146,7 +146,7 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
             if p.Dest[k] == next.ID:
                 beforeDict['Before'] += rampDistCG - p.D # distance from the pallet to the ramp door
 
-        optcgcons.OptRampDist(pallets, k, tour, rampDistCG, cfg, nodeTorque)
+        optcgcons.minRampDist(pallets, k, tour, rampDistCG, cfg, nodeTorque)
 
         for p in pallets:
             if p.Dest[k] == next.ID:
@@ -244,8 +244,8 @@ if __name__ == "__main__":
 
     # methods = ["Shims","mpShims","GRB"]
     
-    # methods = ["Shims"]
-    methods = ["mpShims"]
+    methods = ["Shims"]
+    # methods = ["mpShims"]
     # tipo = "KP"
     tipo = "FFD"
 

### 2023-02-13 optcgcons.py

@@ -40,17 +40,20 @@ def OptCGCons(kept, pallets, maxTorque, k):
                 pallets[i].Dest[k] = kept[j].To
                 kept[j].P = i # put the consolidated in the best position to minimize torque                        
 
-    print(f"--- {common.CITIES[k]} CG deviation minimized {torque/maxTorque:.2f}")
+    # print(f"--- {common.CITIES[k]} CG deviation minimized {torque/maxTorque:.2f}")
 
     return torque
 
 # After solved, optimize consolidated positions to minimize CG deviation
-def OptCG(pallets, k, nodeTorque):
+def minCGdev(pallets, k, nodeTorque, cfg):
 
     cons = [common.Item(p.ID, -2, p.PCW, p.PCS, p.PCV, k, p.Dest[k]) for p in pallets]
 
     ConsRange    = range(len(cons))
     PalletsRange = range(len(pallets))
+
+    for i in PalletsRange:
+        pallets[i].reset(cfg.numNodes)
 
     mod = gp.Model()
     mod.setParam('OutputFlag', 0)
@@ -77,15 +80,16 @@ def OptCG(pallets, k, nodeTorque):
     nodeTorque.value = mod.objVal
 
     for i, _ in enumerate(pallets):
-
         for j in ConsRange:
             if X[i][j].x >= 0.99:
-                pallets[i].Dest[k] = cons[j].To
                 cons[j].P = i # put the consolidated in the best position to minimize torque                        
-
+                pallets[i].Dest[k] = cons[j].To                
+                pallets[i].PCW     = cons[j].W
+                pallets[i].PCV     = cons[j].V
+                pallets[i].PCS     = cons[j].S
 
 # After solved, minimize the sum of ramp door distances for the next node pallets
-def OptRampDist(pallets, k, tour, rampDistCG, cfg, nodeTorque):
+def minRampDist(pallets, k, tour, rampDistCG, cfg, nodeTorque):
 
     palletWeights = [0 for _ in pallets]
 
@@ -98,7 +102,8 @@ def OptRampDist(pallets, k, tour, rampDistCG, cfg, nodeTorque):
     PalletsRange = range(len(pallets))
 
     for j in ConsRange:
-        palletWeights[j] = 140 + cons[j].W 
+        palletWeights[j] = 140 + cons[j].W
+        pallets[j].reset(cfg.numNodes)
 
     mod = gp.Model()
     mod.setParam('OutputFlag', 0)
[…]

### 2023-02-14 latex/comp_or.tex
mpShims main processes well described

@@
 
+{\color{blue}
 Figure \ref{fig:whip} represents the $n_{\pi(k)}$\/ possible edges $e_{ij}^{\pi(k)}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi(k)}$.
 \emph{Shims}\/ starts with a greedy solution, stopping at the edge with index in $\eta_1[i]$ close to the local optimum (first phase).
-Then, considering even the edge with the index $\eta_2[i]$, it elaborates different possible complements for this pallet (second phase) and selects the best of these complements (third phase).
-
-{\color{blue} In this work, we included a new phase between the first and the second: a minimization of the CG deviation.
-An integer programming procedure that helps improve the overall {\it mpShims} results.
-
-$1^{st}$ - Greedy phase
+Then, considering even the edge with the index $\eta_2[i]$, it elaborates different possible complements for this pallet and selects the best of these complements (second phase).
+
+Although the second phase solves a knapsack problem, an algorithm for solving a bin packing problem was used (the First Fit Decreasing - FFD), and the most profitable bin (shims) was selected.
+This method was proved to be faster than a dynamic program solution for the knapsack problem.
+
+	In this work, we included a new phase after the first: a minimization of the CG deviation. An integer programming procedure that helps improve the overall {\it mpShims} results.
+
+$1^{st}$ - Greedy phase with torque surplus
 
@@
 
-$3^{st}$ - Shims construction
-
-$4^{st}$ - best Shims selection
+$3^{st}$ - Shims construction and best Shims selection
 
@@
 \begin{algorithm}[H]
-	\caption{Mount a greedy solution until the volume limit for each pallet}  \label{alg:mpShims}
+	\caption{$mpShims$ main process}  \label{alg:mpShims}
 
@@
 
-		\State Let $G(E^{Q_{\pi(k)}})$ \Comment{the solution comes with some consolidated on board}
+		\State Let $G(M,\ Q_{\pi(k)},\ E_{\pi(k)})$ \label{mpShims:g1} \Comment{the initial solution solution}
 
@@
 
-		\State $lock \gets$ multi-processing lock \Comment{for shared data among processes to avoid race condition}
+		\State $lock \gets$ multi-processing lock  \label{mpShims:lock} \Comment{avoid race condition}
 
@@
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets Greedy(p_i,\ N_{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ limit,\ lock,\ ts )$
+		\State $proc_i \gets Greedy(p_i,\ N_{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ limit,\ lock,\ ts )$   \label{mpShims:greedy}
 		\State $proc_i.start$
@@
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i.join$
+		\State $proc_i.join$ \label{mpShims:join1}
 		\EndFor
@@
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets getBestShims(p_i,\ N_{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ surplus,\ lock,\ ts )$
+		\State $proc_i \gets getBestShims(p_i,\ N_{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ surplus,\ lock,\ ts )$ \label{mpShims:getBestShims}
 		\State $proc_i.start$
@@
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i.join$
+		\State $proc_i.join$ \label{mpShims:join2}
 
@@
 
-		\Return $G(E^{N_{\pi(k)}} \cup E^{Q_{\pi(k)}})$
+		\Return $G(M,\ N_{\pi(k)} \cup Q_{\pi(k)},\ E_{\pi(k)})$ \label{mpShims:return}
 
@@
 
+Line \ref{mpShims:g1} the initial solution comes with some consolidated on board, a bipartite graph between pallets and consolidated.
+
 Line \ref{mpShims:ts} defines the torque surplus ($ts$).
-For this phase, the solution may generate torque infeasible solutions with doubled torque, which will be repaired by the next phase, the minimization of the CG deviation.
-
-Line \ref{mpShims:minCGdev} presents the procedure to minimize the CG deviation and repair the infeasible greedy solution.
+For the greedy phase, the solution may generate torque infeasible solutions with doubled torque ($2.0$), which will be repaired by the next phase, the minimization of the CG deviation.
+
+Line \ref{mpShims:lock} creates a lock feature for shared data among processes to avoid race condition.
+
+Line \ref{mpShims:greedy} calls the greedy method which updates $p_i$, $N_{\pi(k)}$, $nodeTorque$, and $G$.
+
+Line \ref{mpShims:join1} wait for all $Greedy$ processes to finish.
+
+Line \ref{mpShims:minCGdev} calls the $minCGdev$ method which updates the $nodeTorque$ and $G$. $minCGdev$ minimizes the CG deviation and repairs the greedy solution if it is infeasible.
+
+Line \ref{mpShims:getBestShims} calls the $getBestShims$ method which updates $p_i$, $N_{\pi(k)}$, $nodeTorque$, and $G$.
+
+Line \ref{mpShims:join2} wait for all $getBestShims$ processes to finish.
+
+Line \ref{mpShims:return} returns a complete solution, a bipartite graph between pallets and consolidated or items.
 

### 2023-02-18 latex/irace.tex
TSP python update

% !Rnw weave = knitr
%%% DO NOT EDIT the .tex file directly since it is generated from the .Rnw
%%% sources.
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{irace package: User Guide}
%\VignetteDepends{knitr}
%\VignetteCompiler{knitr}
\synctex=1
\RequirePackage[dvipsnames]{xcolor}
\documentclass[a4paper,english]{article}
\usepackage[a4paper]{geometry} % It saves some pages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{ifthen}
\newboolean{Release}
\setboolean{Release}{true}
\usepackage{calc}
\usepackage{afterpage}
\usepackage{algorithm,algorithmic}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{xspace}
\usepackage{amsmath,amssymb}
\usepackage{relsize}
\usepackage{fancyvrb}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\usepackage[numbers]{natbib}
\usepackage[nottoc]{tocbibind}
%% For autoref
\hypersetup{
	colorlinks,
	linkcolor={red!50!black},
	citecolor={blue!50!black},
	urlcolor={blue!70!black}
}
\addto\extrasenglish{%
	\def\sectionautorefname{Section}
	\let\subsectionautorefname\sectionautorefname
	\let\subsubsectionautorefname\sectionautorefname
}
\usepackage[titletoc, title]{appendix}
% Fix use with \autoref
\newcommand*{\Appendixautorefname}{Appendix}
\usepackage{tocloft}
\setlength{\cftsubsecnumwidth}{3em}% Set length of number width in ToC for \subsection
\usepackage[inline]{enumitem}
\setlist[enumerate]{leftmargin=*,widest=00}
\setlist[itemize]{leftmargin=1.5em}

%% FIXME: listing is very limited, we should use 'minted'
\usepackage{listings}
\lstdefinestyle{BashInputStyle}{
	language=bash,%
	basicstyle=\ttfamily,%
	numbers=none,%
	frame=tb,%
	rulecolor=\color{lightgray},
	%  framesep=1ex,
	framexleftmargin=1ex,
	columns=fullflexible,%
	backgroundcolor=\color{yellow!05},%
	linewidth=\linewidth,%
	% xleftmargin=1\linewidth,%
	identifierstyle=\color{darkgray},%
	keywordstyle=\color{darkgray},%
	keywordstyle={[2]\color{Cyan}},%
	keywordstyle={[3]\color{olive}},%
	stringstyle=\color{MidnightBlue},%
	commentstyle=\color{RedOrange},%
	morestring=[b]',%
	showstringspaces=false
}

\DefineVerbatimEnvironment{Code}{Verbatim}{}
\DefineVerbatimEnvironment{CodeInput}{Verbatim}{fontshape=rm}
\DefineVerbatimEnvironment{CodeOutput}{Verbatim}{}
\newenvironment{CodeChunk}{}{}

\newcommand{\IRACEHOME}[1]{\hyperlink{irace_home}{\path{$IRACE_HOME}}\path{#1}}

\providecommand{\keywords}[1]{\textbf{\textit{Index terms---}} #1}

% Simple font selection is not good enough.  For example, |\texttt{--}|
% gives `\texttt{--}', i.e., an endash in typewriter font.  Hence, we
% need to turn off ligatures, which currently only happens for commands
% |\code| and |\samp| and the ones derived from them.  Hyphenation is
% another issue; it should really be turned off inside |\samp|.  And
% most importantly, \LaTeX{} special characters are a nightmare.  E.g.,
% one needs |\~{}| to produce a tilde in a file name marked by |\file|.
% Perhaps a few years ago, most users would have agreed that this may be
% unfortunate but should not be changed to ensure consistency.  But with
% the advent of the WWW and the need for getting `|~|' and `|#|' into
% URLs, commands which only treat the escape and grouping characters
% specially have gained acceptance
\makeatletter
\DeclareRobustCommand\code{\bgroup\@makeother\_\@makeother\~\@makeother\$\@noligs\@codex}
\def\@codex#1{\texorpdfstring%
	{{\normalfont\ttfamily\hyphenchar\font=-1 #1}}%
	{#1}\egroup}
\makeatother

\let\proglang=\textsf
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}
\newcommand{\aR}{\proglang{R}\xspace}
\newcommand{\MATLAB}{\proglang{MATLAB}\xspace}
\newcommand{\eg}{e.g.,\xspace}
\newcommand{\SoftwarePackage}{\pkg}
\newcommand{\ACOTSP}{\SoftwarePackage{ACOTSP}\xspace}

%% How to use this command:
% Parameter with one short switch: \defparameter[short]{paramName}{long}{default}
% Parameter without short switch:  \defparameter{paramName}{long}{default}
% Parameter without switch:  \defparameter{paramName}{}{default}
\newcommand{\defparameter}[4][]{%
	\item[\code{#2}]\hypertarget{opt:#2}{} ~~ %
	\ifthenelse{\equal{#3}{}}{}{%
		\emph{flag:} %
		\ifthenelse{\equal{#1}{}}{}{%
			\code{-#1}~~~\emph{or}~~~}%
		\code{-{}-#3} ~~ }%
	\emph{default:}~\texttt{#4} \\
}
\newcommand{\parameter}[1]{\hyperlink{opt:#1}{\code{#1}}}
%\usepackage{showlabels}
%\showlabels{hypertarget}

\newcommand{\irace}{\pkg{irace}\xspace}
\newcommand{\Irace}{\pkg{Irace}\xspace}
\newcommand{\race}{\pkg{race}\xspace}
\newcommand{\FRACE}{\text{F-Race}\xspace}
\newcommand{\IFRACE}{\text{I/F-Race}\xspace}
\newcommand{\PyImp}{\pkg{PyImp}\xspace}
\newcommand{\iraceversion}{\Sexpr{packageVersion("irace")}}

\newcommand{\Niter}{\ensuremath{N^\text{iter}}\xspace}
\newcommand{\Nparam}{\ensuremath{{N^\text{param}}}\xspace}
\newcommand{\iter}{\ensuremath{j}\xspace}
\newcommand{\Budget}{\ensuremath{B}\xspace}
\newcommand{\Budgetj}{\ensuremath{\Budget_{\iter}}\xspace}
\newcommand{\Bused}{\ensuremath{\Budget_\text{used}}\xspace}
\newcommand{\Ncand}[1][]{\ensuremath{N_{#1}}\xspace}
\newcommand{\Mui}{\ensuremath{\mu_{\iter}}\xspace}
\newcommand{\Nmin}{\ensuremath{N^\text{min}}\xspace}
\newcommand{\Nsurv}{\ensuremath{N^\text{surv}}\xspace}
\newcommand{\Nelite}{\ensuremath{N^\text{elite}}\xspace}
\newcommand{\Nnew}{\ensuremath{N^\text{new}}\xspace}

\newcommand{\bmax}{\ensuremath{b^\text{max}}\xspace}
\newcommand{\bmin}{\ensuremath{b^\text{min}}\xspace}
\newcommand{\Celite}{\ensuremath{\Theta^\text{elite}}\xspace}

\ifthenelse {\boolean{Release}}{%
	\newcommand{\MANUEL}[1]{}
	\newcommand{\LESLIE}[1]{}
	\newcommand{\THOMAS}[1]{}
}{%
	\newcommand{\MANUEL}[1]{{\footnotesize\noindent\textbf{\color{red}[~MANUEL: #1~]}}}
	\newcommand{\LESLIE}[1]{\footnote{\noindent\textbf{[ LESLIE: #1 ]}}}
	\newcommand{\THOMAS}[1]{\footnote{\noindent\textbf{[ THOMAS: #1 ]}}}
}
\newcommand{\hide}[1]{}

\usepackage{tcolorbox}
\newcommand{\infoicon}{%
	\parbox[c]{0.75cm}{\includegraphics[keepaspectratio=true,width=0.75cm]{light-bulb-icon}}%
	\hspace{1em}}
\newcommand{\warningicon}{%
	\parbox[c]{0.75cm}{\includegraphics[keepaspectratio=true,width=0.75cm]{Warning-icon}}%
	\hspace{1em}}

\definecolor{LightGray}{RGB}{193,193,193}
\definecolor{LightYellow}{RGB}{253,247,172}

\newlength\macroiconwidth
\newenvironment{xwarningbox}{%
	\setlength{\fboxrule}{3.0\fboxrule}%
	\setlength{\fboxsep}{0\fboxsep}%
	\begin{tcolorbox}[colback=LightYellow,colframe=LightGray,boxrule=\fboxrule,boxsep=\fboxsep]%
		\infoicon%
		\settowidth{\macroiconwidth}{\infoicon}%
		\begin{minipage}[c]{\columnwidth - \macroiconwidth - 2.0\fboxrule - 2.0\fboxsep}
			\raggedright\footnotesize
			%
		}{%
		\end{minipage}
	\end{tcolorbox}
	%
}

\begin{document}

	<<include=FALSE>>=
	library(knitr)
	@

	\author{Manuel L\'opez-Ib\'a\~nez, Leslie P\'erez C\'aceres, J\'er\'emie Dubois-Lacoste,\\
		Thomas St\"utzle and Mauro Birattari
		\\IRIDIA, CoDE, Universit\'e Libre de Bruxelles, Brussels, Belgium}

	\title{The \irace Package: User Guide}
	\date{Version \iraceversion, \today}
	%\keywords{automatic
		%  algorithm configuration, racing, parameter tuning, \aR}

	\maketitle

	\tableofcontents

	%Load files needed for examples
	<<exampleload,eval=TRUE,include=FALSE>>=
	library("irace")
	load("examples.Rdata")
	load("irace-acotsp.Rdata")
	load("log-ablation.Rdata")
	options(width = 70)
	@
	\newpage

	%%
	%%
	%%
	%% General info
	%%
	%%
	%%
	\section{General information}
	\MANUEL{Some things could be taken from the intro of the irace paper and reformulated.}
	\MANUEL{It would be good to mention that not only opt algorithms can be configured with irace, we say this in the paper.}

	\subsection{Background}
	\MANUEL{I would add a paragraph defining what is irace (a bit longer than the abstract above) and references to the literature so people can find more info. The first reference should be the irace TR.} \LESLIE{Here i guess we should say why tune an algorithm is a good idea, and why using irace is a better one.}
	The \irace package implements an \emph{iterated racing} procedure,
	which is an extension of Iterated F-race (\IFRACE)~\cite{BirYuaBal2010:emaoa}.  The main use of
	\irace is the automatic configuration of optimization and decision algorithms,
	that is, finding the most appropriate settings of an algorithm given a 
	set of instances of a problem. However, it may also be useful for 
	configuring other types of algorithms when performance depends on 
	the used parameter settings. It builds upon the \pkg{race} package 
	by Birattari and it is implemented in \aR. The \irace
	package is available from CRAN:
	%
	\begin{center}
		\url{https://cran.r-project.org/package=irace}
	\end{center}
	%
	More information about \irace is available at
	\url{http://iridia.ulb.ac.be/irace}.

	\subsection{Version}
	The current version of the \irace package is \iraceversion. Previous
	versions of the package can also be found in the \href{https://cran.r-project.org/package=irace}{CRAN website}.

	The algorithm underlying the current version of \irace and its motivation are
	described by \citet{LopDubPerStuBir2016irace}. The \textbf{adaptive capping
		mechanism} available from version $3.0$ is described by
	\citet{PerLopHooStu2017:lion}. Details of the implementation before version 2.0
	can be found in a previous technical report~\cite{LopDubStu2011irace}.
	%
	\begin{xwarningbox}
		Versions of \irace before 2.0 are not compatible with the file formats
		detailed in this document.
	\end{xwarningbox}

	\subsection{License}

	The \irace package is Copyright \copyright{} \the\year\ and distributed under the GNU General Public
	License version 3.0 (\url{http://www.gnu.org/licenses/gpl-3.0.en.html}).
	The \irace package is free software (software libre): You can redistribute it and/or modify it under the terms
	of the GNU General Public License as published by the Free Software Foundation,
	either version 3 of the License, or (at your option) any later version.

	The \irace package is distributed in the hope that it will be useful, but
	WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or
	FITNESS FOR A PARTICULAR PURPOSE.

	Please be aware that the fact that this program is released as Free Software
	does not excuse you from scientific propriety, which obligates you to give
	appropriate credit! If you write a scientific paper describing research that
	made substantive use of this program, it is your obligation as a scientist to
	(a) mention the fashion in which this software was used in the Methods section;
	(b) mention the algorithm in the References section. The appropriate citation
	is:

	\begin{itemize}[leftmargin=3em]
		\item[] Manuel López-Ibáñez, Jérémie Dubois-Lacoste, Leslie Pérez Cáceres,
		Thomas Stützle, and Mauro Birattari. The \irace package: Iterated Racing for
		Automatic Algorithm Configuration. \emph{Operations Research Perspectives}, 3:43--58, 2016.
		doi:~\href{http://dx.doi.org/10.1016/j.orp.2016.09.002}{10.1016/j.orp.2016.09.002}

	\end{itemize}

	\section{Before starting}
	\MANUEL{I think this could be a bit more detailed by defining what is a parameter, a configuration, an instance, etc. but ok for now.}

	The \irace package provides an automatic configuration tool for
	tuning optimization algorithms, that is, automatically finding good configurations for the parameters values
	of a (target) algorithm saving the effort that normally requires manual
	tuning.

	\begin{figure}[t]
		\centering
		\includegraphics[width=0.6\textwidth]{irace-scheme}
		\caption{Scheme of \irace flow of information.}
		\label{fig:irace-scheme}
	\end{figure}

	Figure~\ref{fig:irace-scheme} gives a general scheme of how \irace works.
	\Irace receives as input a \emph{parameter space definition} corresponding to the
	parameters of the target algorithm that will be tuned, a set of \emph{instances}
	for which the parameters must be tuned for and a set of options for \irace that define the \emph{configuration scenario}. Then, \irace searches in the parameter search space for good performing algorithm
	configurations by executing the target algorithm on different instances and
	with different parameter configurations. A \parameter{targetRunner} must be provided to execute the target algorithm with a
	specific parameter configuration ($\theta$) and instance ($i$). The \parameter{targetRunner} function (or program) acts
	as an interface between the execution of the target algorithm and \irace: It
	receives the instance and configuration as arguments and must return the
	evaluation of the execution of the target algorithm.

	The following user guide contains guidelines for installing \irace, defining
	configuration scenarios, and using \irace to automatically configure your
	algorithms.

	%%
	%%
	%%
	%% Installation
	%%
	%%
	%%

	\section{Installation}

	\subsection{System requirements}
	\begin{itemize}
		\item \aR ($\text{version} \geq 3.2.0$) is required for running irace, but you don't
		need to know the \aR language to use it.  \aR is freely available and you can
		download it from the \aR project website
		(\url{https://www.r-project.org}). See \autoref{sec:installation} for a
		quick installation guide of \aR.

		\item For GNU/Linux and OS X, the command-line executable
		\code{parallel-irace} requires GNU Bash.  Individual examples may require additional software.
	\end{itemize}

	\subsection{\irace installation} \label{sec:irace install}

	The \irace package can be installed automatically within \aR or
	by manual download and installation. We advise to use the automatic
	installation unless particular circumstances do not allow it. The instructions
	to install \irace with the two mentioned methods are the following:

	\subsubsection[Install automatically within R]{Install automatically within \aR{}}

	Execute the following line in the \aR console to install the package:
	<<R_irace_install, prompt=FALSE, eval=FALSE>>=
	install.packages("irace")
	@
	Select a mirror close to your location, and test the installation in the \aR console with:
	<<R_irace_launch,eval=FALSE, prompt=FALSE>>=
	library("irace")
	q() # To exit R
	@

	Alternatively, within the \aR graphical interface, you may use
	the \code{Packages and data->Package installer} menu on OS X or the \code{Packages} menu on Windows.

	\subsubsection{Manual download and installation}

	From the \irace package CRAN website
	(\url{https://cran.r-project.org/package=irace}), download one of the three
	versions available depending on your operating system:
	\begin{itemize}
		\item \code{irace_\iraceversion.tar.gz} (Unix/BSD/GNU/Linux)
		\item \code{irace_\iraceversion.tgz} (OS X)
		\item \code{irace_\iraceversion.zip} (Windows)
	\end{itemize}

	To install the package on GNU/Linux and OS X, you must execute the following
	command at the shell (replace \code{<package>} with the path to the downloaded file, either \code{irace_\iraceversion.tar.gz} or \code{irace_\iraceversion.zip}):
	%
	\begin{lstlisting}[style=BashInputStyle]
		R CMD INSTALL <package>
	\end{lstlisting}

	To install the package on Windows, open \aR and execute the following line on
	the \aR console (replace \code{<package>} with the path to the downloaded file \code{irace_\iraceversion.zip}):
	%\LESLIE{Check that this actually works on internet says that  this: \code{Rscript -e "install.packages('foo.zip', repos = NULL)"} also works}
	%
	<<install_win1,eval=FALSE, prompt=FALSE>>=
	install.packages("<package>", repos = NULL)
	@

	If the previous installation instructions fail because of insufficient
	permissions and you do not have sufficient admin rights to install \irace
	system-wide, then you need to force a local installation.

	\subsubsection{Local installation}

	Let's assume you wish to install \irace on a path denoted by
	\code{<R_LIBS_USER>}, which is a filesystem path for which you have sufficient
	rights. This directory \textbf{must} exist before attempting the
	installation. Moreover, you must provide to \aR the path to this
	library when loading the package. However, the latter can be avoided by adding
	the path to the system variable \code{R_LIBS} or to the \aR internal variable
	\code{.libPaths}, as we will see below.\footnote{%
		On Windows, see also
		\url{https://cran.r-project.org/bin/windows/base/rw-FAQ.html\#I-don_0027t-have-permission-to-write-to-the-R_002d3_002e3_002e1_005clibrary-directory}.}

	On GNU/Linux or OS X, execute the following commands to install the
	package on a local directory:

	\begin{lstlisting}[style=BashInputStyle]
		export R_LIBS_USER="<R_LIBS_USER>"
		# Create R_LIBS_USER if it doesn't exist
		mkdir $R_LIBS_USER
		# Replace <package> with the path to the downloaded file.
		R CMD INSTALL --library=$R_LIBS_USER <package>
		# Tell R where to find R_LIBS_USER
		export R_LIBS=${R_LIBS_USER}:${R_LIBS}
	\end{lstlisting}

	On Windows, you can install the package on a local directory by executing the
	following lines in the \aR console:

	<<install_win2,eval=FALSE, prompt=FALSE>>=
	# Replace <package> with the path to the downloaded file.
	# Replace <R_LIBS_USER> with the path used for installation.
	install.packages("<package>", repos = NULL, lib = "<R_LIBS_USER>")
	# Tell R where to find R_LIBS_USER.
	# This must be executed for every new session.
	.libPaths(c("<R_LIBS_USER>", .libPaths()))
	@

	\subsubsection{Testing the installation and invoking irace}

	Once \irace has been installed, load the package and test that the installation
	was successful by opening an \aR console and executing:

	<<R_irace_test1, prompt=FALSE, eval=FALSE>>=
	# Load the package
	library("irace")
	# Obtain the installation path
	system.file(package = "irace")
	@

	The last command must print out the filesystem path  where \irace is
	installed.  In the remainder of this guide, the variable
	\code{\$IRACE_HOME} is used to denote this path. When executing any provided
	command that includes the \code{\$IRACE_HOME} variable do not forget to replace
	this variable with the installation path of \irace.

	On GNU/Linux or OS X, you can let the operating system know where to find \irace
	by defining the \code{\$IRACE_HOME} variable and adding it to the system
	\code{PATH}. Append the following commands to \path{~/.bash_profile}, \path{~/.bashrc} or
	\path{~/.profile}:
	%
	%<<linux_irace_path1,engine='bash',eval=FALSE>>=
	\begin{lstlisting}[style=BashInputStyle]
		# Replace <IRACE_HOME> with the irace installation path
		export IRACE_HOME=<IRACE_HOME>
		export PATH=${IRACE_HOME}/bin/:$PATH
		# Tell R where to find R_LIBS_USER
		# Use the following line only if local installation was forced
		export R_LIBS=${R_LIBS_USER}:${R_LIBS}
	\end{lstlisting}
	%@

	Then, open a new terminal and launch \irace as follows:

	%<<linux_irace_help,engine='bash',eval=FALSE>>=
	\begin{lstlisting}[style=BashInputStyle]
		irace --help
	\end{lstlisting}
	%@

	On Windows, you need to add both \aR and the installation path of \irace to the
	environment variable \code{PATH}. To edit the \code{PATH}, search for
	``Environment variables'' in the control panel, edit \code{PATH} and add
	a string similar to \path{C:\R_PATH\bin;C:\IRACE_HOME\bin\x64\} where \code{R_PATH}
		is the installation path of \aR and \code{IRACE_HOME} is the installation path
		of \irace. If \irace was installed locally, you also need to edit the
		environment variable \code{R_LIBS} to add \code{R_LIBS_USER}. Then, open a new
		terminal (run program \code{cmd.exe}) and launch \irace as:
		%
		%<<win_irace_help,engine='bash',eval=FALSE>>=
		\begin{lstlisting}[style=BashInputStyle]
			irace.exe --help
		\end{lstlisting}
		%@

		Alternatively, you may directly invoke \irace from within the \aR
		console by executing:
		<<windows_irace_help ,eval=FALSE, prompt=FALSE>>=
		library("irace")
		irace.cmdline("--help")
		@

		\section{Running irace}\label{sec:execution}

		Before performing the tuning of your algorithm, it is necessary to define a
		tuning scenario that will give \irace all the necessary information to optimize the
		parameters of the algorithm. The tuning scenario is composed of the following elements:

		\begin{enumerate}
			\item Target algorithm parameter description (see \autoref{sec:target parameters}).
			\item Target algorithm runner (see \autoref{sec:runner}).
			\item Training instances list (see \autoref{sec:training})
			\item \irace options (see \autoref{sec:irace options}).
			\item \textit{Optional:} Initial configurations (see \autoref{sec:initial}).
			\item \textit{Optional:} Forbidden configurations (see \autoref{sec:forbidden}).
			\item \textit{Optional:} Target algorithm evaluator (see \autoref{sec:evaluator}).
		\end{enumerate}

		These scenario elements can be provided as plain text files
		or as \aR objects. This user guide provides examples of both types,
		but we advise the use of plain text files, which we consider the simpler
		option.

		For a step-by-step guide to create the scenario elements for your target algorithm
		continue to \autoref{sec:step}. For an example execution of \irace using the
		\ACOTSP scenario go to \autoref{sec:example}.

		\subsection{Step-by-step setup guide}\label{sec:step}
		This section provides a guide to setup a basic execution of \irace. The
		template files provided in the package (\IRACEHOME{/templates}) will be
		used as basis for creating your new scenario. Please follow carefully the
		indications provided in each step and in the template files used; if you have
		doubts check the the sections that describe each option in detail.

		\begin{enumerate}[leftmargin=*]
			\item Create a directory (\eg~\path{~/tuning/}) for the scenario setup. This directory will contain
			all the files that describe the scenario. On GNU/Linux or OS X, you can do this as follows:
			%<<dir1,engine='bash',eval=FALSE>>=
			\begin{lstlisting}[style=BashInputStyle]
				mkdir ~/tuning
				cd ~/tuning
			\end{lstlisting}
			%@

			\item Copy all the template files from the \IRACEHOME{/templates/}
			directory to the scenario directory.
			%<<copy1,engine='bash',eval=FALSE>>=
			\begin{lstlisting}[style=BashInputStyle]
				# $IRACE_HOME is the installation directory of irace.
				cp $IRACE_HOME/templates/*.tmpl  ~/tuning/
			\end{lstlisting}
			%@

			% Remember that \IRACEHOME{} is the path to the installation
			% directory of \irace. It can be obtained in the \aR console with:
			%
			% <<irace_path,eval=FALSE, prompt=FALSE>>=
			% library("irace")
			% system.file(package = "irace")
			% @

			\item For each template in your tuning directory, remove the \code{.tmpl}
			suffix, and modify them following the next steps.

			\item Define the target algorithm parameters to be tuned by following the
			instructions in \code{parameters.txt}.  Available parameter types and other
			guidelines can be found in \autoref{sec:target parameters}.

			\item \textit{Optional}: Define the initial parameter configuration(s) of your
			algorithm, which  allows you to provide good starting configurations (if you know
			some) for the tuning. Follow the instructions in \code{configurations.txt} and set \parameter{configurationsFile}\code{="configurations.txt"} in \code{scenario.txt}. More
			information in \autoref{sec:initial}. If you do not need to
			define initial configurations remove this file from the directory.

			\item \textit{Optional}: Define forbidden parameter value combinations,
			that is, configurations that \irace must not consider in the tuning.
			Follow the instructions in \code{forbidden.txt} and update \code{scenario.txt} with \parameter{forbiddenFile} \code{=} \code{"forbidden.txt"}. More information about
			forbidden configurations in \autoref{sec:forbidden}. 
			If you do not need to define forbidden configurations remove this file from
			the directory.

			\item Place the instances you would like to use for the tuning of your
			algorithm in the folder \path{~/tuning/Instances/}. In addition, you can
			create a file (\eg~\code{instances-list.txt}) that specifies which instances
			from that directory should be run and which instance-specific parameters to
			use. To use such an instance file, set the appropriate option in
			\code{scenario.txt}, e.g., \parameter{trainInstancesFile} \code{= "instances-list.txt"}.
			See \autoref{sec:training} for guidelines.

			\item Uncomment and assign in \code{scenario.txt} only the options for which you
			need a value different from the default.
			%% MANUEL: I'm not sure what this means.
			% The names of the template files match the default names of the scenario
			% options.
			Some common parameters that you might want to adjust are:
			\begin{description}
				\item[\parameter{execDir}] (\code{--exec-dir}): the directory in which \irace will execute the
				target algorithm; the default value is the current directory.
				%\item[\parameter{logFile}] (\code{--log-file}): a file the name of the results \aR data file that produces \irace.
				\item[\parameter{maxExperiments}] (\code{--max-experiments}): the maximum number of executions
				of the target algorithm that \irace will perform.
				\item[\parameter{maxTime}] (\code{--max-time}): maximum total execution time in seconds for the executions of \code{targetRunner}. In this case, \code{targetRunner} must return two values: cost and time. Note that you must provide either \parameter{maxTime} or \parameter{maxExperiments}.
			\end{description}
			For setting the tuning budget see \autoref{sec:budget}. For more information on \irace options and their default values, see \autoref{sec:irace options}.

			\item Modify the \code{target-runner} script to run your algorithm. This script
			must execute your algorithm with the parameters and instance specified by
			\irace and return the evaluation of the execution and \textit{optionally} the execution 
			time (\code{cost [time]}). When the \parameter{maxTime} option is used, returning \code{time} is mandatory.
			The \code{target-runner} template is written in \proglang{GNU Bash}
			scripting language, which can be executed easily in GNU/Linux and OS X
			systems. However, you may use any other programming language. As an example,
			we provide a \proglang{Python} example in the directory
			\IRACEHOME{/examples/python}. %
			Follow these instructions to adjust the given \code{target-runner} template
			to your algorithm:
			\begin{enumerate}
				\item Set the \code{EXE} variable with the path to the executable of the target algorithm.
				\item Set the \code{FIXED_PARAMS} if you need extra arguments in the
				execution line of your algorithm. An example could be the time that your
				algorithm is required to run
				(\code{FIXED_PARAMS}\hspace{0pt}\code{=}\hspace{0pt}\code{"--time 60"}) or
				the number of evaluations required (\code{FIXED_PARAMS}\hspace{0pt}\code{=}\hspace{0pt}\code{"--evaluations 10000"}).

				\item The line provided in the template executes the executable described in the \code{EXE} variable.
				%
				\begin{center}
					\code{\$EXE \$\{FIXED_PARAMS\} -i \$\{INSTANCE\} --seed \$\{SEED\} \$\{CONFIG_PARAMS\}}
				\end{center}
				%
				You must change this line according to the way your algorithm is executed. In this
				example, the algorithm receives the instance to solve with the flag
				\code{-i} and the seed of the random number generator with the flag \code{--seed}. The variable
				\code{CONFIG_PARAMS} adds to the command line the parameters that \irace has
				given for the execution.  You must set the command line execution as
				needed. For example, the instance might not need a flag and might need to be
				the first argument:
				\begin{center}
					\code{\$EXE \$\{INSTANCE\} \$\{FIXED_PARAMS\} --seed \$\{SEED\} \$\{CONFIG_PARAMS\}}
				\end{center}
				The output of your algorithm is saved to the file defined in the
				\code{\$STDOUT} variable, and error output is saved in the file given by
				\code{\$STDERR}. The line:
				\begin{center}
					\code{if [ -s "\${STDOUT}" ]; then}
				\end{center}
				checks if the file containing the output of your algorithm is not
				empty. The example provided in the template assumes that your algorithm
				prints in the last output line the best result found (only a number). The
				line:
				\begin{center}
					\code{COST=\$(cat \$\{STDOUT\} | grep -e '\^{}[[:space:]]*[+-]\textbackslash{}?[0-9]' | cut -f1)}
				\end{center}
				parses the output of your algorithm to obtain the result from the last line. The \code{target-runner}
				script must return \textbf{only} one number. In the template example, the result is returned with
				\code{echo "\$COST"} (assuming \parameter{maxExperiments} is used) and the used files are deleted.

				\begin{xwarningbox}
					The \code{target-runner} script must be executable.
				\end{xwarningbox}

				You can test the target runner from the \aR console by checking the scenario as 
				explained earlier in \autoref{sec:execution}.

				If you have problems related to the \code{target-runner} script when executing
				\irace, see \autoref{sec:check list} for a check list to help diagnose common
				problems. For more information about the \parameter{targetRunner}, please see
				\autoref{sec:runner},
			\end{enumerate}

			\item \textit{Optional}: Modify the \code{target-evaluator} file. This is rarely needed and the \code{target-runner} template does not use it. \autoref{sec:evaluator} explains when a \parameter{targetEvaluator} is needed and how to define it.

			\item The \irace executable provides an option (\parameter{-{}-check}) to
			check that the scenario is correctly defined. We recommend to perform a
			check every time you create a new scenario. When performing the check,
			\irace will verify that the scenario and parameter definitions are
			correct and will test the execution of the target algorithm. To check
			your scenario execute the following commands:

			\begin{itemize}
				\item From the command-line (on Windows, execute \code{irace.bat}):

				%<<irace_check,engine='bash', eval=FALSE>>=
				\begin{lstlisting}[style=BashInputStyle]
					# $IRACE_HOME is the installation directory of irace.
					$IRACE_HOME/bin/irace --scenario scenario.txt --check
				\end{lstlisting}
				%@

				\item Or from the \aR console:

				<<irace_R_check, eval=FALSE, prompt=FALSE>>=
				library("irace")
				scenario <- readScenario(filename = "scenario.txt",
				scenario = defaultScenario())
				checkIraceScenario(scenario = scenario)
				@

			\end{itemize}

			\item  Once all the scenario elements are prepared you can execute \irace, either using the command-line wrappers provided by the package
			or directly from the \aR console:

			\begin{itemize}
				\item{%
					\textbf{From the command-line console}, call the command (on Windows, you should execute
					\code{irace.exe}):
					\begin{lstlisting}[style=BashInputStyle]
						cd ~/tuning/
						# $IRACE_HOME is the installation directory of irace
						# By default, irace reads scenario.txt, you can specify a different file
						# with --scenario.
						$IRACE_HOME/bin/irace
					\end{lstlisting}
					For this example we assume that the needed scenario files have been set
					properly in the \code{scenario.txt} file using the options described in
					\autoref{sec:irace options}. Most \irace options can be specified
					in the command line or directly in the \code{scenario.txt} file.
				}
				\item{
					\textbf{From the \aR console}, evaluate:

					<<irace_R_exe, eval=FALSE, prompt=FALSE>>=
					library("irace")
					# Go to the directory containing the scenario files
					setwd("~/tuning")
					scenario <- readScenario(filename = "scenario.txt",
					scenario = defaultScenario())
					irace.main(scenario = scenario)
					@
				}
			\end{itemize}

			This will perform one run of \irace. See the output of \code{irace --help} in the command-line or \code{irace.usage()} in \aR for quick information on
			additional \irace parameters. For more information about \irace options, see \autoref{sec:irace options}.

		\end{enumerate}

		\begin{xwarningbox}
			Command-line options override the same options specified in the \code{scenario.txt} file.
		\end{xwarningbox}

		\subsection{Setup example for ACOTSP}\label{sec:example}

		The \ACOTSP tuning example can be found in the package installation at 
		\IRACEHOME{/examples/acotsp}.
		%
		Additionally, a number of example scenarios can be found in the \code{examples} folder. More
		examples of tuning scenarios can be found in the Algorithm Configuration Library (AClib, \url{http://www.aclib.net/}).

		In this section, we describe how to execute the \ACOTSP scenario. If you wish to start setting up
		your own scenario, continue to the next section. For this example, we assume
		a GNU/Linux system but making the necessary changes in the commands and \parameter{targetRunner},
		it can be executed in any system that has a \proglang{C} compiler.
		%\MANUEL{I don't think this is true, since the target-runner script needs Bash}
		To execute this scenario follow these steps:

		\begin{enumerate}
			\item Create a directory for the tuning (\eg~\path{~/tuning/}) and copy the example scenario files
			located in the \code{examples} folder to the created directory:
			%<<dir0,engine='bash',eval=FALSE>>=
			\begin{lstlisting}[style=BashInputStyle]
				mkdir ~/tuning
				cd ~/tuning
				# $IRACE_HOME is the installation directory of irace.
				cp $IRACE_HOME/examples/acotsp/* ~/tuning/
			\end{lstlisting}
			%@

			\item Download the training instances from \url{http://iridia.ulb.ac.be/irace/} to the \path{~/tuning/} directory.
			\item Create the instance directory (\eg~\path{~/tuning/Instances}) and decompress the instance files on it.

			%<<instance0,engine='bash',eval=FALSE>>=
			\begin{lstlisting}[style=BashInputStyle]
				mkdir ~/tuning/Instances/
				cd ~/tuning/
				tar -xvf tsp-instances-training.tar.bz2 Instances/
			\end{lstlisting}
			%@

			\item Download the \ACOTSP software from \url{http://www.aco-metaheuristic.org/aco-code/} to the \path{~/tuning/}
			directory and compile it.
			%<<acotsp0,engine='bash',eval=FALSE>>=
			\begin{lstlisting}[style=BashInputStyle]
				cd ~/tuning/
				tar -xvf ACOTSP-1.03.tgz
				cd ~/tuning/ACOTSP-1.03
				make
			\end{lstlisting}
			%@
			\item Create a directory for the executable and copy it:

			%<<acotsp1,engine='bash',eval=FALSE>>=
			\begin{lstlisting}[style=BashInputStyle]
				mkdir ~/bin/
				cp ~/tuning/ACOTSP-1.03/acotsp ~/bin/
			\end{lstlisting}
			%@

			\item Create a directory for executing the experiments and execute \irace:

			%<<runexample,engine='bash',eval=FALSE>>=
			\begin{lstlisting}[style=BashInputStyle]
				mkdir ~/tuning/acotsp-arena/
				cd ~/tuning/
				# $IRACE_HOME is the installation directory of irace.
				$IRACE_HOME/bin/irace
			\end{lstlisting}
			%@

			\item Or you can also execute \irace from the \aR console using:

			<<runexample2,prompt=FALSE,eval=FALSE>>=
			library("irace")
			setwd("~/tuning/")
			irace.cmdline()
			@

		\end{enumerate}

		%%
		%%
		%%
		%% Scenario settings
		%%
		%%
		%%

		\section{Defining a configuration scenario}\label{sec:scenario}
		\subsection{Target algorithm parameters} \label{sec:target parameters}

		The parameters of the target algorithm are defined by a parameter file as
		described in \autoref{sec:parameters file}. Optionally, when executing
		\irace from the \aR console, the parameters can be specified directly as an \aR
		object (see \autoref{sec:parameters object}). For defining your parameters
		follow the guidelines provided in the following sections.

		\subsubsection{Parameter types}

		Each target parameter has an associated type that defines its domain and the
		way \irace handles them internally. Understanding the nature of the domains of
		the target parameters is important to select appropriate types. The four basic
		types supported by \irace are the following:

		\begin{itemize}
			\item \textit{Real} parameters are numerical parameters that can take
			floating-point values within a given range. The range is
			specified as an interval `\code{(<lower bound>,<upper
				bound>)}'. This interval is closed, that is, the parameter value
			may eventually be one of the bounds. The possible values are rounded
			to a number of \emph{decimal places} specified by
			option \parameter{digits}. For example, given the default
			number of digits of $4$, the values $0.12345$ and
			$0.12341$ are both rounded to $0.1234$.
			% However, the values  $0.00001$ and $0.00005$ remain the same.
			Selected real-valued parameters can be optionally sampled on
			a logarithmic scale (base $e$).

			\item \textit{Integer} parameters are numerical parameters that can take only
			integer values within the given range. Their range is specified as the range
			of real parameters and they can also be optionally sampled on a
			logarithmic scale (base $e$).

			\item \textit{Categorical} parameters are defined by a set of possible values
			specified as `\code{(<value 1>,} \code{...,} \code{<value n>)}'. The values
			are quoted or unquoted character strings. Empty strings and strings
			containing commas or spaces must be quoted.

			\item \emph{Ordinal} parameters are defined by an \emph{ordered} set
			of possible values in the same format as for categorical
			parameters. They are handled internally as integer parameters, where
			the integers correspond to the indexes of the values.

		\end{itemize}

		\subsubsection{Parameter domains}

		For each target parameter, an interval or a set of values must be defined
		according to its type, as described above. There is no limit for the size of
		the set or the length of the interval, but keep in mind that larger domains
		could increase the difficulty of the tuning task. Choose always values that you
		consider relevant for the tuning. In case of doubt, we recommend to choose larger 
		intervals, as occasionally best parameter settings may be not intuitive a priori. 
		All intervals are considered as closed intervals.

		It is possible to define parameters that will have always the same value.
		Such ``\emph{fixed}'' parameters will not be tuned but their values are used
		when executing the target algorithm and they are affected by constraints
		defined on them. All fixed parameters must be defined as categorical
		parameters and have a domain of one element.

		\subsubsection{Conditional parameters}

		Conditional parameters are active only when others have certain values. These
		dependencies define a hierarchical relation between parameters. For example,
		the target algorithm may have a parameter \code{localsearch} that takes values
		\code{(sa, ts)} and another parameter \code{ts-length} that only needs to be
		set if the first parameter takes precisely the value \code{ts}. Thus, parameter
		\code{ts-length} is conditional on \code{localsearch == "ts"}.

		\subsubsection{Parameter file format}\label{sec:parameters file}

		For simplicity, the description of the parameters space is given as a
		table. Each line of the table defines a configurable parameter
		%
		\begin{center}
			\code{<name>\ <label>\ <type>\ <range>\ [ | <condition>\ ] }
		\end{center}
		where each field is defined as follows:
		%
		\begin{center}
			\renewcommand{\arraystretch}{1.2}
			\begin{tabularx}{0.98\linewidth}{@{}rX@{}}
				\code{<name>} & The name of the parameter as an unquoted
				alphanumeric string, e.g., `\code{ants}'.\\
				%
				\code{<label>}& A \emph{label} for this parameter. This is a
				string that will be passed together with the parameter
				to \parameter{targetRunner}. In the default \parameter{targetRunner}
				provided with the package (\autoref{sec:runner}), this is the
				command-line switch used to pass the value of this parameter, for
				instance `\code{"-{}-ants "}'.\newline
				The value of the parameter is concatenated \emph{without
					separator} to the label when
				invoking \parameter{targetRunner}, thus \emph{any whitespace in the label is significant}. Following the same example, when parameter \code{ants} takes value \code{5}, the default targetRunner will pass the parameter as \code{"-{}-ants 5"}.\\
				%
				\code{{<type>}} & The type of the parameter, either
				\textit{integer}, \textit{real}, \textit{ordinal} or
				\textit{categorical}, given as a single letter: `\code{i}',
				`\code{r}', `\code{o}' or `\code{c}'. Numerical parameters
				can be sampled using a natural logarithmic scale with '\code{i,log}' and
				'\code{r,log}' (without spaces) for integer and real
				parameters, respectively.\\
				%
				\code{{<range>}} & The range or set of values of the parameter delimited by
				parentheses. \eg~\code{(0,1)} or \code{(a,b,c,d)}.\\
				%

				\code{{<condition>}} & An optional \emph{condition} that determines whether the parameter is
				enabled or disabled, thus making the parameter conditional. If the
				condition evaluates to false, then no value is assigned to this
				parameter, and neither the parameter value nor the corresponding
				label are passed to \parameter{targetRunner}. The condition must follow the same syntax as those for specifying forbidden configurations (\autoref{sec:forbidden}), that is, it must be a
				valid \aR logical expression\footnote{For a list of \aR operators
					see: \url{https://stat.ethz.ch/R-manual/R-devel/library/base/html/Syntax.html}}. 
				The condition may contain the name
				of other parameters as long as the dependency graph does not
				contain any cycle. Otherwise, \irace will detect the cycle and
				stop with an error. \\
			\end{tabularx}
		\end{center}

		As an example, Figure~\ref{fig:acotsp_parameters} shows the parameters file
		of the \ACOTSP scenario.

		\begin{figure}[!hbt]
			\small\centering%
			\begin{CodeInput}
				# name      switch            type     values               [conditions (using R syntax)]
				algorithm   "--"              c        (as,mmas,eas,ras,acs)
				localsearch "--localsearch "  c        (0, 1, 2, 3)
				alpha       "--alpha "        r        (0.00, 5.00)
				beta        "--beta "         r        (0.00, 10.00)
				rho         "--rho  "         r        (0.01, 1.00)
				ants        "--ants "         i        (5, 100)
				nnls        "--nnls "         i        (5, 50)              | localsearch %in% c(1, 2, 3)
				q0          "--q0 "           r        (0.0, 1.0)           | algorithm == "acs"
				dlb         "--dlb "          c        (0, 1)               | localsearch %in% c(1,2,3)
				rasrank     "--rasranks "     i        (1, 100)             | algorithm == "ras"
				elitistants "--elitistants "  i        (1, 750)             | algorithm == "eas"
			\end{CodeInput}
			\caption{Parameter file (\code{parameters.txt}) for tuning \ACOTSP.}\label{fig:acotsp_parameters}
		\end{figure}

		\subsubsection{Parameters R format}\label{sec:parameters object}
		The target parameters are stored in an \aR~\code{list} that you can obtain from the \aR console using the following command:

		<<readParameters,prompt=FALSE, eval=FALSE>>=
		parameters <- readParameters(file = "parameters.txt")
		@

		See the help of the \code{readParameters} function (\code{?readParameters}) for more information.
		The structure of the parameter list that is created is as follows:

		\begin{center}
			\renewcommand{\arraystretch}{1.2}
			\begin{tabularx}{0.98\linewidth}{@{}rX}
				\code{names}        & Vector that contains the names of the parameters. \\
				\code{types}        & Vector that contains the type of each parameter 'i', 'c', 'r', 'o'. \\
				\code{switches}     & Vector that contains the labels of the parameters. e.g., switches to be used for the
				parameters on the command line. \\
				\code{domain}       & List of vectors, where each vector may contain two
				values (minimum, maximum) for real and integer parameters, or
				a set of values for categorical and ordinal parameters.\\
				\code{conditions}   & List of \aR logical expressions, with variables
				corresponding to parameter names.\\
				\code{isFixed}      & Logical vector that specifies which parameter is fixed
				and, thus, it does not need to be tuned.\\
				\code{transform}    & Vector that contains the transformation of each parameter. Currently, it
				can take values \code{``''} (no transformation, default) of \code{``log''}
				(natural logarithmic transformation).\\
				\code{nbParameters} & An integer, the total number of parameters.\\
				\code{nbFixed}      & An integer, the number of parameters with a fixed value.\\
				\code{nbVariable}   & Number of variable (i.e., to be tuned) parameters.\\
			\end{tabularx}
		\end{center}

		The following example shows the structure of the \code{parameters} \aR object
		for the \code{algorithm}, \code{ants} and \code{q0} parameters of the \ACOTSP scenario:

		<<parametersetup,eval=TRUE,include=FALSE>>=
		# Setup example
		parameters <- irace::readParameters(text='
		algorithm       "--"                 c		(as,mmas,eas,ras,acs)
		ants            "--ants "            i	  	(5, 100)
		q0              "--q0 "              r  	(0.0, 1.0) 		| algorithm %in% c("acs")
		')
		@

		<<parameterlist,eval=TRUE, prompt=TRUE, size='normalsize', comment="">>=
		str(parameters, vec.len = 10)
		@

		\subsection{Target algorithm runner}
		\label{sec:runner}

		The evaluation of a candidate configuration on a single instance is done by means of a user-given auxiliary
		program or, alternatively, a user-given \aR function. The function (or program name) is specified by the option
		\parameter{targetRunner}. The \parameter{targetRunner} must return the
		cost value (\eg cost of the best solution found) of the evaluation; unless computing the cost requires information from all the configurations evaluated on an instance, \eg when evaluating multi-objective algorithms with unknown normalisation bounds (see \autoref{sec:evaluator} for details).

		\begin{xwarningbox}
			The objective of \irace is to minimize the cost value returned by the target
			algorithm.  If you wish to maximize, you can multiply the cost by
			\code{-1} before returning it to \irace.
		\end{xwarningbox}

		\subsubsection{Target runner executable program}

		When \parameter{targetRunner} is an auxiliary executable program, it
		is invoked for each candidate configuration, passing as arguments:
		%
		\begin{center}
			\small\code{<id.configuration>\ <id.instance>\ <seed>\ <instance>\ [bound]\ <configuration>}
		\end{center}
		%
		\begin{center}
			\renewcommand{\arraystretch}{1.2}
			\begin{tabularx}{1\linewidth}{@{}rX}
				\code{id.configuration}  & an alphanumeric string that uniquely identifies a configuration;\\
				\code{id.instance} & an alphanumeric string that uniquely identifies an instance;\\
				\code{seed} & seed for the random number generator to be used for this evaluation, ignore the seed for deterministic algorithms;\\
				\code{instance} & string giving the instance to be used for this evaluation;\\
				\code{bound} & optional execution time bound. Only provided when the \parameter{maxBound} option is set in the scenario, see \autoref{sec:capping};\\
				\code{configuration} & the pairs parameter label-value that
				describe this candidate configuration. Typically given as command-line
				switches to be passed to the executable program.\\
			\end{tabularx}
		\end{center}

		The experiment list shown in \autoref{sec:runner fx}, would result in the following execution line:

		%<<target_runner_ex1, engine='bash', eval=FALSE>>=
		\begin{lstlisting}[style=BashInputStyle]
			target-runner 1 113 734718556 /home/user/instances/tsp/2000-533.tsp \
			--eas --localsearch 0 --alpha 2.92 --beta 3.06 --rho 0.6 --ants 80
		\end{lstlisting}
		%@

		The command line switches that describe the candidate configuration are
		constructed by appending to each parameter label (switch), \emph{without
			separator}, the value of the parameter, following the order given in the
		parameter table. The program \parameter{targetRunner} must print a real
		number, which corresponds to the cost measure of the candidate configuration
		for the given instance and optionally its execution time (mandatory when \parameter{maxTime} 
		is used and/or when the \parameter{capping} option is enabled). The working directory of 
		\parameter{targetRunner} is set to the execution directory specified by the option
		\parameter{execDir}. This allows the user to execute independent runs of \irace
		in parallel using different values for \parameter{execDir}, without the runs
		interfering with each other.

		\subsubsection{Target runner R function}
		\label{sec:runner fx}

		When \parameter{targetRunner} is an \aR function, it is invoked
		for each candidate configuration as:
		%
		<<targetRunner,prompt=FALSE, eval=FALSE>>=
		targetRunner(experiment, scenario)
		@
		%
		\noindent where \code{experiment} is a list that contains  information about
		configuration and instance to execute one experiment, and \code{scenario} is the scenario
		list. The structure of the \code{experiment} list is as follows:

		\begin{center}
			\renewcommand{\arraystretch}{1.2}
			\begin{tabularx}{1\linewidth}{@{}rX}
				\code{id.configuration}  & an alphanumeric string that uniquely identifies a configuration;\\
				\code{id.instance} & an alphanumeric string that uniquely identifies an instance;\\
				\code{seed} & seed to be used for this evaluation;\\
				\code{instance} & string giving the instance to be used for this evaluation;\\
				\code{bound} &  optional execution time bound;\\
				\code{configuration}    & 1-row data frame with a column per parameter name;\\
				\code{switches}         & vector of parameter switches (labels) in the order of parameters used in \code{configuration}.\\

			\end{tabularx}
		\end{center}

		The following is an example of an experiment list for the \ACOTSP scenario:

		<<experimentlist,eval=TRUE,size='normalsize', prompt=TRUE,  comment="">>=
		print(experiment)
		@

		If \parameter{targetEvaluator} is \code{NULL}, then
		the \parameter{targetRunner} function must return a list with at least one
		element \code{"cost"}, the numerical value corresponding to the evaluation of
		the given configuration on the given instance. A cost of \code{Inf} is accepted
		and results in the immediate rejection of the configuration (see
		\autoref{sec:reject}).

		If the scenario option \parameter{maxTime} is non-zero or if the \parameter{capping} 
		option is enabled, then the list must contain at least another element \code{"time"} that 
		reports the execution time for this call to \code{targetRunner}.

		The return list may also contain the following optional elements that are used
		by \irace for reporting errors in \code{targetRunner}:
		%
		\begin{center}
			\renewcommand{\arraystretch}{1.2}
			\begin{tabularx}{1\linewidth}{@{}rX}
				\code{error} & is a string used to report an error;\\
				\code{outputRaw} & is a string used to report the raw output of calls to
				an external program or function;\\
				\code{call} & is a string used to report how \code{targetRunner} called 
				an external program or function;\\
			\end{tabularx}
		\end{center}

		\subsection{Target evaluator} \label{sec:evaluator}

		Normally, \parameter{targetRunner} returns the cost of the execution of a
		candidate configuration (see \autoref{sec:runner}). However, there are cases
		when the cost evaluation must be delayed until all candidate configurations in
		a race have been executed on a instance.

		The \parameter{targetEvaluator} option defines an auxiliary program (or an \aR
		function) that allows postponing the evaluations of the candidate
		configurations. For each instance seen, the program \parameter{targetEvaluator} is only invoked after all
		the calls to \parameter{targetRunner} for all alive candidate configurations on
		the same instance have already finished.

		\begin{xwarningbox}
			When using \parameter{targetEvaluator}, \parameter{targetRunner} must not return the evaluation
			of the configuration. If \parameter{maxTime} is used, \parameter{targetRunner} must return only 
			execution time.
		\end{xwarningbox}

		As an example, \parameter{targetEvaluator} may be used to dynamically find
		normalization bounds for the output returned by an algorithm for each
		individual instance. In this case, \parameter{targetRunner} will save the
		output of the algorithm, then the first call to \parameter{targetEvaluator}
		will examine the output produced by all calls to \parameter{targetRunner} for
		the same instance, update the normalization bounds and return the normalized
		output. Subsequent calls to \parameter{targetEvaluator} for the same instance
		will simply return the normalized output.

		A similar need arises when using quality measures for multi-objective
		optimization algorithms, such as the hypervolume, which typically
		require specifying reference points or sets. By
		using \parameter{targetEvaluator}, it is possible to dynamically compute the
		reference points or sets while \irace is running. Examples are provided at
		\path{examples/hypervolume}. See also \autoref{sec:multi objective} for
		more information on how to tune multi-objective algorithms.

		\subsubsection{Target evaluator executable program}

		When \parameter{targetEvaluator} is an auxiliary executable program, it
		is invoked for each candidate with the following arguments:
		%
		\begin{center}
			\small\code{<id.configuration>\ <id.instance>\ <seed>\ <instance>\ <num.configs>\ <all.conf.id>}
		\end{center}
		%
		\begin{center}
			\renewcommand{\arraystretch}{1.2}
			\begin{tabularx}{1\linewidth}{@{}rX}
				\code{id.configuration}  & an alphanumeric string that uniquely identifies a configuration;\\
				\code{id.instance} & an alphanumeric string that uniquely identifies an instance;\\
				\code{seed} & seed to be used for this evaluation;\\
				\code{instance} & string giving the instance to be used for this evaluation;\\
				\code{num.configs} & number of alive candidate configurations;\\
				\code{all.conf.id}    & list of IDs of the alive configurations separated by whitespace.\\
			\end{tabularx}
		\end{center}

		The \parameter{targetEvaluator} executable must print a numerical value
		corresponding to the cost measure of the candidate configuration on
		the given instance.

		\subsubsection{Target evaluator R function}

		When \parameter{targetEvaluator} is an \aR function, it is invoked
		for each candidate configuration as:%
		%
		<<targetEvaluator, prompt=FALSE, eval=FALSE>>=
		targetEvaluator(experiment, num.configurations, all.conf.id,
		scenario, target.runner.call)
		@
		%
		\noindent where \code{experiment} is a list that contains information about one
		experiment (see \autoref{sec:runner fx}), \code{num.configurations}
		is the number of configurations alive in the race, \code{all.conf.id} is
		the vector of IDs of the alive configurations,  \code{scenario} is
		the scenario list and \code{target.tunner.call} is the string of the
		\parameter{targetRunner} execution line.

		The function \parameter{targetEvaluator} must return a list with one element
		\code{"cost"}, the numerical value corresponding to the cost measure of the
		given configuration on the given instance.

		The return list may also contain the following optional elements that are used
		by \irace for reporting errors in \code{targetEvaluator}:
		%
		\begin{center}
			\renewcommand{\arraystretch}{1.2}
			\begin{tabularx}{1\linewidth}{@{}rX}
				\code{error} & is a string used to report an error;\\
				\code{outputRaw} & is a string used to report the raw output of calls to
				an external program or function;\\
				\code{call} & is a string used to report how \code{targetEvaluator} called 
				an external program or function;\\
			\end{tabularx}
		\end{center}

		\subsection{Training instances}\label{sec:training}

		The \irace options \parameter{trainInstancesDir} and \parameter{trainInstancesFile}
		specify where to find the training instances.
		%
		By default, the value of \parameter{trainInstancesFile} is empty. This means that \irace will consider all files within the directory given by \parameter{trainInstancesDir} (by default \path{./Instances}) as training instances.

		Otherwise, the value of \parameter{trainInstancesFile} may specify a text
		file. The format of this file is one instance per line. Within each line, elements separated by white-space will be parsed as separate arguments to be supplied to \parameter{targetRunner}. This allows defining instance-specific parameter settings. Quoted strings will be parsed as a single argument.
		%
		The following example shows a training instance file for the \ACOTSP scenario:

		\begin{figure}[!ht]
			\centering%
			\begin{minipage}{0.4\textwidth}
				\begin{CodeInput}
					# Example training instances file
					100/100-1_100-2.tsp --time 1
					100/100-1_100-3.tsp --time 2
					100/100-1_100-4.tsp --time 3
				\end{CodeInput}
			\end{minipage}
			\caption{Training instances file for tuning \ACOTSP.}\label{fig:acotsp_training}
		\end{figure}

		The value of \parameter{trainInstancesDir} is always prefixed to the instance
		name, that is, the instances names are treated as relative to this directory.
		%
		For example, given the above file as \parameter{trainInstancesFile} and the default value of  \parameter{trainInstancesDir} (\code{./Instances}), then a possible invocation of \parameter{targetRunner} would be:
		%
		%<<target_runner_cmd, engine='bash', eval=FALSE>>=
		\begin{lstlisting}[style=BashInputStyle]
			target-runner 1 113 734718 ./Instances/100/100-1_100-2.tsp --time 1 \
			--alpha 2.92 ...
		\end{lstlisting}
		%@

		Training instances do not need to be files, \irace just passes the elements of each line as arguments to \parameter{targetRunner}, thus each line may denote the name of a benchmark function or a label, plus instance-specific settings, that the target algorithm understands. Each line may even be the command-line parameters required to call an instance generator within \parameter{targetRunner}. When the instances do not represent actual files, then \parameter{trainInstancesDir} is usually set to the empty string (\code{--train-instances-dir=""}). For example,

		\begin{center}
			\begin{minipage}{0.4\linewidth}
				\begin{CodeInput}
					# Example training instances file
					rosenbrock_20 --function=12 --nvar 20
					rosenbrock_30 --function=12 --nvar 30
					rastrigin_20 --function=15 --nvar 20
					rastrigin_30 --function=15 --nvar 30
				\end{CodeInput}
			\end{minipage}
		\end{center}
		%\caption{Example of training instances file where instances do not correspond to files.}\label{fig:nofile_training}
		%\end{figure}

		Optionally, when executing \irace from the \aR console, the list of instances might be provided explicitly by means of the variable \code{scenario$instances}. Thus, the previous example would be equivalent to:

		<<instance1, prompt=FALSE, eval=FALSE>>=
		scenario$instances <- c("rosenbrock_20 --function=12 --nvar 20",
		"rosenbrock_40 --function=12 --nvar 30",
		"rastrigin_20 --function=15 --nvar 20",
		"rastrigin_40 --function=15 --nvar 30")
		@

		By default, \irace assumes that the target algorithm is stochastic (the value
		of the option \parameter{deterministic} is $0$), thus, the same configuration
		can be executed more than once on the same instance and obtain different
		results.  In this case, \irace generates pairs \code{(instance, seed)} by
		generating a random seed for each instance. In other words, configurations
		evaluated on the same instance use the same random seed. This is a well-known
		variance reduction technique called \emph{common random
			numbers}~\citep{McG1992vrt}. If all available pairs are used within a run of
		\irace, new pairs are generated with different seeds, that is, a configuration
		evaluated more than once per instance will use different random seeds.

		If \parameter{deterministic} is set to $1$, then each instance will be used at
		most once per race. This setting should only be used for target algorithms that
		do not have a stochastic behavior and, therefore, executing the target
		algorithm on the same instance several times with different seeds does not make
		sense.
		\begin{xwarningbox}
			If \parameter{deterministic} is active and the number of training instances
			provided to \irace is less than \parameter{firstTest} (default: 5), no
			statistical test will be performed on the race.
		\end{xwarningbox}

		Finally, \irace randomly re-orders the sequence of instances provided. This
		random sampling may be disabled by using the option \parameter{sampleInstances}
		(\code{--sample-instances 0}) if keeping the order provided in the instance
		file is important.

		\begin{xwarningbox}
			We advise to always sample instances to prevent biasing the tuning due to the
			instance order. See also \autoref{sec:het}
		\end{xwarningbox}

		\subsection{Initial configurations} \label{sec:initial}

		The scenario option \parameter{configurationsFile} allows specifying a text
		file that contains an initial set of configurations to start the execution of
		\irace. If the number of initial configurations supplied in the file is less
		than the number of configurations required by \irace in the first iteration,
		additional configurations will be sampled uniformly at random.

		The format of the configurations file is one configuration per line, and one
		parameter value per column. The first line must give the parameter name
		corresponding to each column (names must match those given in the parameters
		file).  Each configuration must satisfy the parameter conditions (\code{NA}
		should be used for those parameters that are not enabled for a given
		configuration) and not be forbidden by the constraints that define forbidden
		configurations (\autoref{sec:forbidden}), if any.

		Figure~\ref{fig:acotsp_default} gives an example file that corresponds to the
		\ACOTSP scenario.

		\begin{figure}[!ht]
			\centering
			\begin{minipage}{0.7\linewidth}
				\footnotesize
				\begin{CodeInput}
					## Initial candidate configuration for irace
					algorithm localsearch alpha beta rho  ants nnls dlb q0 rasrank elitistants
					as        0           1.0   1.0  0.95 10   NA   NA  0  NA      NA
				\end{CodeInput}
			\end{minipage}
			\caption{Initial configuration file (\code{default.txt}) for tuning \ACOTSP.}\label{fig:acotsp_default}
		\end{figure}

		We advise to use this feature when a default configuration of the target algorithm exists or when
		different sets of good parameter values are known. This will allow \irace to start the search
		from those parameter values and attempt to improve their performance.

		\subsection{Forbidden configurations}\label{sec:forbidden}

		The scenario option \parameter{forbiddenFile} specifies a text file containing
		logical expressions of parameter values that valid configurations should not
		satisfy, that is, no configuration that satisfies any of these logical
		expressions will be evaluated by \irace. This is useful when some combination
		of parameter values could cause the target algorithm to crash, consume
		excessive CPU time or memory, or when it is known that they do no produce
		satisfactory results.

		The format of the forbidden configurations file is one constraint per line,
		where each constraint is a logical expression (in \aR syntax) containing
		parameter names as defined by the \parameter{parameterFile}
		(\autoref{sec:target parameters}), values and logical operators. For a list of \aR logical operators see:
		\begin{center}
			\url{https://stat.ethz.ch/R-manual/R-devel/library/base/html/Syntax.html}
		\end{center}

		If a parameter configuration is generated that makes any of the logical
		expressions evaluate to \code{TRUE}, then the configuration is considered
		forbidden and it is discarded. Figure~\ref{fig:acotsp_forbidden} shows an
		example file that corresponds to the \ACOTSP scenario.

		\begin{figure}[!ht]
			\centering
			\begin{minipage}{0.5\linewidth}
				\begin{CodeInput}
					## Examples of valid logical operators are:
					## ==  !=  >=  <=  >  <  &  |  !  %in%
					(alpha == 0.0) & (beta == 0.0)
				\end{CodeInput}
			\end{minipage}
			\caption{Forbidden configurations file (\code{forbidden.txt}) for tuning \ACOTSP.}\label{fig:acotsp_forbidden}
		\end{figure}

		\begin{xwarningbox}
			If initial configuration are provided (\autoref{sec:initial}), they must also comply with the
			constraints defined in \parameter{forbiddenFile}.
		\end{xwarningbox}

		\begin{xwarningbox}
			Categorical and ordinal parameters are always treated as strings. Given a parameter like:
			\begin{CodeInput}
				a "" c (0, 5, 10, 20)
			\end{CodeInput}
			then, a condition like \code{a > 10} will be true when \code{a} is \code{5},
			because comparisons between strings are lexicographic and \code{"10"} is sorted
			before \code{"5"}. As a work-around, you can convert the string to numeric in
			the condition with \code{as.numeric(a)}.
		\end{xwarningbox}

		\subsection{Repairing configurations}\label{sec:repairconf}

		In some problems, the parameter values require complex constraints that cannot
		be implemented by constraints defined in \parameter{forbiddenFile}
		(\autoref{sec:forbidden}).
		%
		The scenario option \parameter{repairConfiguration} can be set to a
		user-defined \aR function that takes a single configuration generated by irace
		and returns a ``\emph{repaired}'' configuration, thus allowing the
		implementation of any rules necessary to satisfy arbitrary constraints on
		parameter values.
		%
		The \parameter{repairConfiguration} function is called after generating a
		configuration and before checking for forbidden configurations. The first
		argument is a 1-row \code{data.frame} with parameter names as the column names,
		the second argument is the \code{parameters} list (\autoref{sec:parameters
			object}), and the third argument is the scenario
		variable \parameter{digits}. An example that makes all real-valued parameters
		sum up to one would be:
		%
		<<repairEx,prompt=FALSE, eval=FALSE>>=
		repairConfiguration = function (configuration, parameters, digits)
		{
			isreal <- parameters$type[colnames(configuration)] %in% "r"
			configuration[isreal] <- configuration[isreal] / sum(configuration[isreal])
			return(configuration)
		}
		@
		The following example forces  three specific parameters to be in increasing order:
		%
		<<repairEx2,prompt=FALSE, eval=FALSE>>=
		repairConfiguration = function (configuration, parameters, digits)
		{
			columns <- c("p1","p2","p3")
			# cat("Before"); print(configuration)
			configuration[columns] <- sort(configuration[columns])
			# cat("After"); print(configuration)
			return(configuration)
		}
		@

		The above code can be specified directly in the \parameter{scenarioFile}, by default \code{scenario.txt}.

		%%
		%%
		%%
		%% Parallel processing
		%%
		%%
		%%

		\section{Parallelization}\label{sec:parallel}
		A single run of \irace can be done much faster by executing the calls
		to \parameter{targetRunner} (the runs of the target algorithm) in
		parallel. There are four ways to parallelize a single run of \irace:

		\begin{enumerate}
			\item \textbf{Parallel processes}: The option \parameter{parallel} allows
			executing in parallel, within a single computer, the calls
			to \parameter{targetRunner}, by means of the \pkg{parallel} \aR package. For
			example, adding \code{--parallel N} to the command line of \irace will launch
			in parallel up to $N$ calls of the target algorithm.

			\item \textbf{MPI}: By enabling the option \parameter{mpi}, calls
			to \parameter{targetRunner} will be executed in parallel by using the message
			passing interface (MPI) protocol (requires the \pkg{Rmpi} \aR package). In
			this case, the option \parameter{parallel} controls the number of slave nodes
			used by \irace. For example, adding \code{--mpi 1 --parallel N} to the
			command-line will create $N$ slaves + 1 master, and execute up to $N$ calls of
			\parameter{targetRunner} in parallel.

			The user is responsible for setting up the required MPI environment.  MPI is
			commonly available in computing clusters and requires launching \irace in
			some particular way. An example script for using MPI mode in a SGE cluster
			is given at \IRACEHOME{/bin/parallel-irace-mpi}.

			By default, \irace dynamically balances the load among nodes, however, this
			may significantly increase communication overhead in some parallel
			environments, where disabling \parameter{loadBalancing} may be faster.

			\item \textbf{Batch jobs clusters}: Some computing clusters work by submitting
			jobs to a batch queue and waiting for the jobs to finish. With the
			option \parameter{batchmode} (\code{--batchmode} \code{[sge|pbs|} \code{torque|slurm]}),
			\irace will launch in parallel as many calls of \parameter{targetRunner} as
			possible (\parameter{parallel} can be used to set a limit) and use a
			cluster-specific method to wait for jobs to finish. If your cluster type is
			not supported or not working as expected, please contact us and we will
			gladly add support for it.

			\begin{xwarningbox}
				In this mode, \irace must run in the submission node of the cluster, and
				hence, \irace should not be submitted to the cluster as a job (that is,
				neither \code{qsub} nor \code{squeue} should be used to invoke \irace
				itself). The user must call the appropriate job submission command (e.g., \code{qsub}) from within
				\parameter{targetRunner} with the appropriate settings for their cluster,
				otherwise \parameter{targetRunner} will not submit jobs to the
				cluster. The script must return a single string: The job ID that allows
				\irace to determine the status of the running job. Moreover, the use of a
				separate
				\parameter{targetEvaluator} script is required to evaluate the results of
				\parameter{targetRunner} and return them to \irace.
			\end{xwarningbox}
			See the examples in
			\IRACEHOME{/examples/batchmode-cluster/}.

			\item \parameter{targetRunnerParallel}: This option allows users to fully
			control the parallelization of the execution of \parameter{targetRunner}. 
			Its value must be an \aR function that will be invoked by \irace as follows:

			<<targetRunnerParallel,prompt=FALSE, eval=FALSE>>=
			targetRunnerParallel(experiments, exec.target.runner, scenario, target.runner)
			@
			%
			where \code{scenario} is the list describing the configuration scenario
			(\autoref{sec:scenario}); \code{experiments} is a list that describes the
			configurations and instances to be executed (see \autoref{sec:runner} for a
			description); \code{target.runner} is the function that calls the target
			algorithm and it is the same as \parameter{targetRunner}, if the latter is a
			function, or it is a call to \code{target.runner.default}, if
			\parameter{targetRunner} is the path to an executable; and
			\code{exec.target.runner} is an internal function within \irace that takes
			care of executing \code{target.runner}, check its output and, possibly,
			retry in case of error (see \parameter{targetRunnerRetries}). The
			\parameter{targetRunnerParallel} function should call the given
			\code{target.runner} function for each element in the \code{experiments}
			list, possibly using \code{exec.target.runner} as a wrapper. A trivial
			example would be:

			<<targetRunnerParallel2,prompt=FALSE, eval=FALSE>>=
			targetRunnerParallel <- function(experiments, exec.target.runner, scenario)
			{
				return (lapply(experiments, exec.target.runner, scenario = scenario,
				target.runner = target.runner))
			}
			@
			%
			However, the user is free to set up the calls in any way, perhaps
			implementing its own replacement for \code{target.runner} and/or
			\code{exec.target.runner}.

			The only requirement is that the \parameter{targetRunnerParallel} function
			must return a list of the same length as \code{experiments}, where each
			element is the output expected from the corresponding call
			to \parameter{targetRunner} (see \autoref{sec:runner}). The following is an
			example of the output of a call to \parameter{targetRunnerParallel} with 2
			experiments, in which the execution time is not reported:

			<<targetRunnerParallel3,prompt=FALSE, eval=TRUE>>=
			print(output)
			@
			%

		\end{enumerate}

		%%
		%%
		%%
		%% Testing
		%%
		%%
		%%

		\section{Testing (Validation) of configurations}\label{sec:testing}

		Once the tuning process is finished, \irace returns a set of configurations
		corresponding to the elite configurations at the end of the run, ordered from
		best to worst. In order to evaluate the generality of these configurations
		without looking at their performance on the training set, \irace offers the
		possibility of evaluating these configurations on a test instance set,
		typically different from the training set used during the tuning phase. These
		evaluations will use the same settings for parallel
		execution, \parameter{targetRunner} and \parameter{targetEvaluator}.

		The test instance set can be specified by the
		options \parameter{testInstancesDir} and \parameter{testInstancesFile}, or by
		setting directly the variable \code{scenario\$testInstances}, which behave the
		same as their counterparts for the training instances
		(\autoref{sec:training}). In particular, each test instance is assigned a
		different seed in the same way as done for the training instances. In
		principle, \irace evaluates each configuration on each testing instance just
		once, because evaluating one run on $n$ instances is always better than
		evaluating $n'$ runs on $n/n'$ instances~\citep{IRIDIA-2004-001}. However, if
		the number of instances is limited, one can always duplicate instances as
		needed in the \parameter{testInstancesFile}, and \irace will assign a different
		random seed to each instance.

		The options \parameter{testNbElites} and \parameter{testIterationElites}
		control which configurations are evaluated during the testing phase. In
		particular, setting \code{testIterationElites = 1} will test not only the final
		set of elite configurations (those returned at the end of the training phase),
		but also the set of elites at the end of each race
		(iteration). The option \parameter{testNbElites} limits the maximum number of 
		configurations considered within each set. Some examples:
		\begin{itemize}
			\item \code{testIterationElites = 0; testNbElites = 1} means that only the best
			configuration found during the run of \irace, the final best, will be used in
			the testing phase.
			\item \code{testIterationElites = 1; testNbElites = 1} will test, in addition to the final best, the best configuration found at each iteration.
			\item \code{testIterationElites = 1; testNbElites = 2} will test the two best
			configurations found at each iteration, in addition to the final best and
			second-best configurations.
		\end{itemize}

		The testing can be also (re-)executed at a later time by using the following
		\aR command:
		%
		<<testing_r, prompt=FALSE, eval=FALSE>>=
		testing.main(logFile = "./irace.Rdata")
		@
		%
		The above line will load the scenario setup from \parameter{logFile} to
		perform the testing.  The testing results will be stored in the \aR object
		\code{iraceResults\$testing}, which is saved in the file specified by
		\code{scenario\$logFile}. The structure of the object is described in
		\autoref{sec:output r}.  For examples on how to analyse the results see
		\autoref{sec:analysis}.

		Another alternative is to test a specific set of configurations using the command-line option \parameter{-{}-only-test} as follows:
		\begin{lstlisting}[style=BashInputStyle]
			irace --only-test configurations.txt
		\end{lstlisting}
		where \path{configurations.txt} has the same format as the 
		set of initial configurations (\autoref{sec:initial}).

		%%
		%%
		%%
		%% Recovery
		%%
		%%
		%%
		\section{Recovering \irace runs}\label{sec:recovery}

		Problems like power cuts, hardware malfunction or the need to use computational
		power for other tasks may occur during the execution of \irace, terminating a
		run before completion. At the end of each iteration, \irace saves an \aR data
		file (\parameter{logFile}, by default \code{"./irace.Rdata"}) that not only
		contains information about the tuning progress (\autoref{sec:output r}), but
		also internal information that allows recovering an incomplete execution.

		To recover an incomplete \irace run, set the option \parameter{recoveryFile} to
		the log file previously produced, and \irace will continue the execution from
		the last saved iteration. The state of the random generator is saved and
		loaded, therefore, as long as the execution is continued in the same machine,
		the obtained results will be exactly the same as executing \irace in one step
		(external factors, such as CPU load and disk caches, may affect the
		target algorithm and that may affect the results).  You can specify
		the \parameter{recoveryFile} from the command-line or from the scenario file,
		and execute \irace as described in \autoref{sec:execution}. For example, from
		the command-line use:

		\begin{lstlisting}[style=BashInputStyle]
			irace --recovery-file "./irace-backup.Rdata"
		\end{lstlisting}

		\begin{xwarningbox}
			When recovering a previous run, \irace will try to save data on the file
			specified by the \parameter{logFile} option. Thus, you must specify different
			files for \parameter{logFile} and \parameter{recoveryFile}.  Before
			recovering, we strongly advise to rename the saved \aR data file as in the
			example above, which uses \code{"irace-backup.Rdata"}.
		\end{xwarningbox}

		\hide{%

			%\LESLIE{Actually, there other variables we allow to change...what we shoudl say about this?}
			% MANUEL: I would say nothing for now until we fix this properly.
			When recovery is done you can modify some \irace options from the command-line
			or from the scenario file.

			\begin{itemize}
				\item \parameter{execDir}
				\item \parameter{logFile}
				\item \parameter{debugLevel}
				\item \parameter{parallel}
				\item \parameter{loadBalancing}
				\item \parameter{mpi}
				\item \parameter{batchmode}
			\end{itemize}

			It is not possible (and in most cases not correct) to change the scenario options, given that the
			previous results can become invalid. However there are some cases in which this options must be changed,
			for example, if you have instances that are files, it might be the case that the instances are not
			available at the same location. To change the location of them you should modify directly the \aR data file,
			please be careful not to change the order of the instances, because this would make the results obtained by
			\irace invalid. To change the instances path, open the \aR console and use:

			<<change_recover, prompt=TRUE, eval=FALSE>>=
			load ("~/tuning/irace.Rdata")
			new.path <- "~/experiments/tuning/instances/"
			iraceResults$scenario$instances <-
			paste (new.path,
			basename(iraceResults$scenario$instances),
			sep="")
			save (iraceResults, file="~/tuning/irace.Rdata")
			@

			This example can also can be applied to the target execution files \parameter{targetRunner} and \parameter{targetEvaluator}.
		}

		\begin{xwarningbox}
			Do not change anything in the log file or the scenario file before
			recovering, as it may have unexpected effects on the recovered run of
			\irace. In case of doubt, please contact us first (\autoref{sec:contact}). In particular, it is not possible to continue a run of \irace by recovering with a larger budget. Results will \textbf{not} be the same as running \irace from the start with the largest budget.  An alternative is to use the final configurations from one run as the initial configurations of a new run.
		\end{xwarningbox}

		\begin{xwarningbox}
			If your scenario uses \parameter{targetEvaluator} (\autoref{sec:evaluator})
			and \parameter{targetEvaluator} requires files created
			by \parameter{targetRunner}, then recovery will fail if those files are not
			present in the \parameter{execDir} directory. This can happen, for example,
			if you recover from a different directory than the one from which irace was
			initially executed, or when \parameter{execDir} is set to a temporary
			directory for every irace run. Thus, you need to copy the contents of the
			previous \parameter{execDir} into the new one.
		\end{xwarningbox}

		%%
		%%
		%%
		%% Output and analysis of results
		%%
		%%
		%%
		\section{Output and results}

		During its execution, \irace prints information about the progress of the
		tuning in the standard output.  Additionally, after each iteration, an \aR data
		file is saved (\parameter{logFile} option) containing the state of \irace.

		\subsection{Text output}\label{sec:output text}

		Figure~\ref{fig:output} shows the output, up to the end of the first iteration,
		of a run of elitist \irace applied to the \ACOTSP scenario with 1000
		evaluations as budget.

		\afterpage{\clearpage
			\begin{figure}[p]
				\RecustomVerbatimCommand{\VerbatimInput}{VerbatimInput}%
				{fontsize=\scriptsize,
					%
					frame=none,  % top and bottom rule only
					framesep=0pt, % separation between frame and text
					rulecolor=\color{Gray},
					firstline=1,
					lastline=68,
				}
				\VerbatimInput{irace-acotsp-stdout.txt}%
				\caption{Sample text output of \irace.}\label{fig:output}
		\end{figure}}

		First, \irace gives the user a warning informing that it has found a file with
		the default scenario filename and it will use it. Then, general information
		about the selected \irace options is printed:
		\begin{itemize}
			\item \parameter{nbIterations} indicates the minimum number of iterations \irace has calculated
			for the scenario. Depending on the development of the tuning the final iterations that
			are executed can be more.
			\item \parameter{minNbSurvival} indicates the minimum number of alive configurations that
			are required to continue a race. When less configurations are alive the race is
			stopped and a new iteration begins.
			\item \code{nbParameters} is the number of parameters of the scenario.
			\item \parameter{seed} is the number that was used to initialize the random number generator in \irace.
			\item \code{confidence level} is the confidence level of the statistical test.
			\item \code{budget} is the total number of evaluations available for the tuning.
			\item \code{time budget} is the  maximum execution time available for the tuning.
			\item \parameter{mu} is a value used for calculating the minimum number of iterations.
			\item \parameter{deterministic} indicates if the target algorithm is assumed to be deterministic.
		\end{itemize}

		At each iteration, information about the progress of the execution is printed as follows:
		\begin{itemize}
			\item \code{experimentsUsedSoFar} is the number of experiments from the total budget
			that have been used up to the current iteration.
			\item \code{timeUsed} is the execution time used so far in the experiments. Only available when reported
			in the \parameter{targetRunner} (activate it with the \parameter{maxTime} option).
			\item \code{remainingBudget} is the number of experiments that have not been used yet.
			\item \code{timeEstimate} estimation of the mean execution time. This is used to calculate 
			the remaining budget when \parameter{maxTime} is used.
			\item \code{currentBudget} is the number of evaluations \irace has allocated to the
			current iteration.
			\item \parameter{nbConfigurations} is the number of configurations \irace will use in the
			current iteration. In the first iteration, this number of configurations include
			the initial configurations provided; in later iterations, it includes the elite
			configurations from the previous iterations.
		\end{itemize}

		After the iteration information, a table shows the progress of the iteration execution.
		Each row of the table gives information about the execution of an instance in the race.
		The first column contains a symbol that describes the results of the
		statistical test:
		%
		\begin{description}
			\item \code{|x|} No statistical test was performed for this instance. The options \parameter{firstTest}
			and \parameter{eachTest} control on which instances the statistical test is performed.
			\item \code{|-|} Statistical test performed and configurations have been discarded. The column \code{Alive} gives an indication of how
			many configurations have been discarded.
			\item \code{|=|} Statistical test performed and no configurations have been discarded.
			This means \irace needs to evaluate more instances to identify the best configurations.
			\item \code{|!|} This indicator exists only for the elitist version of \irace. It indicates that
			the statistical test was performed and some elite configurations appear to show bad performance and
			could be discarded but they are kept because of the elitist rules. See option \parameter{elitist} in \autoref{sec:irace options} for more information.
		\end{description}

		Other columns have the following meaning:
		\begin{description}
			\item \code{Instance}: Index of \code{(instance, seed)} pair executed. This number
			corresponds to the index of the list found in \code{state$.irace$instancesList}. See \autoref{sec:output r}
			for more information. This is different from the instance ID passed to \parameter{targetRunner}.
			\item \code{Bound}: Only when \parameter{capping} is enabled. Execution time used as bound
			for the execution of new candidate configurations. 
			\item \code{Alive}: Number of configurations that have not been
			discarded after the statistical test was performed.
			\item \code{Best}: ID of the best configuration according to the instances seen
			so far in this race (i.e., not including previous iterations).
			\item \code{Mean best}: Mean of the best configuration across the instances seen so
			far in this race.
			\item  \code{Exp so far}: Number of experiments
			performed so far.
			\item \code{W time}:  Wall-clock time spent on this instance.
			\item \code{rho}, \code{KenW}, and \code{Qvar}: Spearman's rank correlation
			coefficient rho, Kendall's concordance coefficient W, and a variance measure
			described in \cite{SchHoo2012quanti}, respectively, of the configurations
			across the instances evaluated so far in this iteration.  These measures
			evaluate how consistent is the performance of the configurations across the
			instances.  Values close to 1 for \code{rho} and \code{KenW} and values close
			to 0 for \code{Qvar} indicate that the scenario is highly homogeneous. For
			heterogeneous scenarios, we provide advice in \autoref{sec:het}.
		\end{description}

		Finally, \irace outputs the best configuration found and a list of the elite
		configurations. The elite configurations are configurations that did not show
		statistically significant difference during the race; they are ordered
		according to their mean performance on the executed instances.

		\subsection[R data file (logFile)]{\aR data file (\code{logFile})}\label{sec:output r}

		The \aR data file created by \irace (by default as \code{irace.Rdata}, see option \parameter{logFile}) contains an object called
		\code{iraceResults}. You can load this file in the \aR console with:

		<<load_rdata, prompt=FALSE, eval=FALSE>>=
		load("irace-acotsp.Rdata")
		@

		The \code{iraceResults} object is a list, and the elements of a list can be
		accessed in \aR by using the \code{\$} or \code{[[]]} operators:

		<<show_version, prompt=TRUE, eval=TRUE, comment="">>=
		iraceResults$irace.version
		iraceResults[["irace.version"]]
		@

		\noindent%
		The \code{iraceResults} list contains the following elements:

		\begin{itemize}
			\item \code{scenario}: The scenario \aR object containing the \irace options
			used for the execution. See \autoref{sec:irace options} and the help
			of the \pkg{irace} package; open an \aR console and type: \code{?defaultScenario}.
			See \autoref{sec:irace options} for more information.

			\item \code{parameters}: The parameters \aR object containing the description of
			the target algorithm parameters. See \autoref{sec:target parameters}.

			\item \code{allConfigurations}: The target algorithm configurations generated by
			\irace. This object is a \code{data frame}, each row is a candidate
			configuration; the first column (\code{.ID.}) indicates the internal identifier of the
			configuration; the final column (\code{.PARENT.}) is the identifier
			of the configuration from which the current configuration was sampled; and the remaining columns correspond to the parameter values; each column is named as
			the parameter name specified in the parameter object.

			<<show_configurations, prompt=TRUE, eval=TRUE, comment="">>=
			head(iraceResults$allConfigurations)
			@

			\item \code{allElites}: A list that contains one element per iteration. Each element contains
			the internal identifier of the elite candidate configurations of the corresponding iteration
			(identifiers correspond to \code{allConfigurations$.ID.}).

			<<show_idelites, prompt=TRUE, eval=TRUE, comment="">>=
			print(iraceResults$allElites)
			@

			The configurations are ordered by mean performance, that is, the ID of the best
			configuration corresponds to the first ID. To obtain the values of the
			parameters of all elite configurations found by \irace use:

			<<get_elites, prompt=TRUE, eval=TRUE, comment="">>=
			getFinalElites(logFile = "irace-acotsp.Rdata", n = 0)
			@

			\item \code{iterationElites}: A vector containing the best candidate configuration ID of each iteration. The best configuration found corresponds to the last one
			of this vector.
			<<show_iditelites, prompt=TRUE, eval=TRUE, comment="">>=
			print(iraceResults$iterationElites)
			@

			One can obtain the full configuration with:

			<<get_elite, prompt=TRUE, eval=TRUE, comment="">>=
			last <- length(iraceResults$iterationElites)
			id <- iraceResults$iterationElites[last]
			getConfigurationById(logFile = "irace-acotsp.Rdata", ids = id)
			@

			\item \code{rejectedConfigurations}: A vector containing the rejected configurations IDs. These 
			correspond to configurations that produced failed executions and were ignored by \irace during the
			configuration process. See \autoref{sec:reject} to enable the detection of such configurations.

			\item \code{experiments}: A matrix with configurations as columns and instances
			as rows. Column names correspond to the internal identifier of the configuration
			(\code{allConfigurations$.ID.}). The results of a particular
			configuration can be obtained using:

			<<get_experiments, prompt=TRUE, eval=TRUE, comment="">>=
			# As an example, we use the best configuration found
			best.config <- getFinalElites(iraceResults = iraceResults, n = 1)
			id <- best.config$.ID.
			# Obtain the configurations using the identifier
			# of the best configuration
			all.exp <- iraceResults$experiments[,as.character(id)]
			all.exp[!is.na(all.exp)]
			@

			When a configuration was not executed on an instance, its value is \code{NA}. A
			configuration may not be executed on an instance because: 1) it was not created
			yet when the instance was used, or 2) it was
			discarded by the statistical test and not executed on subsequent instances, or
			3) the race terminated before this instance was considered.

			Row names correspond to the identifier of the \code{(instance, seed)} pairs
			defined in \code{state\$.irace\$instancesList}. The instance and seed used
			for a particular experiment can be obtained with:\sloppy

			<<get_instance_seed, prompt=TRUE, eval=TRUE, comment="">>=
			# As an example, we get seed and instance of the experiments
			# of the best candidate.
			# Get index of the instances
			pair.id <- names(all.exp[!is.na(all.exp)])
			index <- iraceResults$state$.irace$instancesList[pair.id,"instance"]
			# Obtain the instance names
			iraceResults$scenario$instances[index]
			# Get the seeds
			iraceResults$state$.irace$instancesList[index,"seed"]
			@
			\item \code{experimentLog}: A matrix with columns \code{iteration, instance, configuration}. This matrix contains the log of all the experiments that \irace
			performs during its execution. The \code{instance} column refers to the index of the
			\code{state$.irace$instancesList} data frame. When \parameter{capping} is enabled a column \code{bound} is added to log the execution bound applied 
			for each execution.

			\item \code{softRestart}: A logical vector that indicates if a soft restart was
			performed on each iteration. If \code{FALSE}, then no soft restart was performed.
			See option \parameter{softRestart} in \autoref{sec:irace options}.

			\item \code{state}: A list that contains the state of \irace, the recovery
			(\autoref{sec:recovery}) is done using the information contained in this object. The
			probabilistic model of the last elite configurations can be found here by doing:

			<<get_model, prompt=TRUE, eval=TRUE, comment="">>=
			# As an example, we get the model probabilities for the
			# localsearch parameter.
			iraceResults$state$model["localsearch"]
			# The order of the probabilities corresponds to:
			iraceResults$parameters$domain$localsearch
			@

			The example shows a list that has one element per elite configuration (ID as
			element name). In this case, \code{localsearch} is a categorical parameter and 
			it has a probability for each of its values.

			\item \code{testing}: A list that contains the testing results.
			The list contains the following elements:
			\begin{itemize}
				\item \code{experiments}: Matrix of experiments in the same format as the
				\code{iraceResults$experiments} matrix. The column names indicate the candidate
				configuration identifier and the row names contain the name of the instances.
				<<get_test_exp, prompt=TRUE, eval=TRUE, comment="">>=
				# Get the results of the testing
				iraceResults$testing$experiments
				@

				\item \code{seeds}: The seeds used for the experiments, each seed corresponds to
				each instance in the rows of the test \code{experiments} matrix.
				<<get_test_seeds, prompt=TRUE, eval=TRUE, comment="">>=
				# Get the seeds used for testing
				iraceResults$testing$seeds
				@

				In the example, instance \code{1000-1.tsp} is executed with seed \Sexpr{iraceResults$testing$seeds[1]}.

			\end{itemize}

		\end{itemize}

		\subsection{Analysis of results}\label{sec:analysis}

		The final configurations returned by \irace are the elites of the final race.
		They are reported in decreasing order of performance, that is, the best
		configuration is reported first.

		If testing is performed, you can further analyze the resulting best
		configurations by performing statistical tests in \aR or just plotting the
		results:

		<<plot_test, fig.pos="tbp", fig.align="center", fig.height = 4, fig.width = 8, out.width='0.85\\textwidth', fig.cap="Boxplot of the testing results of the best configurations.", prompt=TRUE, eval=TRUE, comment="">>=
		results <- iraceResults$testing$experiments
		# Wilcoxon paired test
		conf <- gl(ncol(results), # number of configurations
		nrow(results), # number of instances
		labels = colnames(results))
		pairwise.wilcox.test (as.vector(results), conf, paired = TRUE, p.adj = "bonf")
		# Plot the results
		configurationsBoxplot (results, ylab = "Solution cost")
		@

		%%FIXME: should we add something like this?
		%The Kendall concordance coefficient (\code{W}) and the Spearman's rho can be
		%applied over data that has the characteristics of the data obtained in the testing,
		%that is a full matrix where all configurations are executed in all instances. \code{W}
		%can show if the configurations tested have an homogeneous performance on the used instances
		%set. If evidence of an heterogeneous scenario found we recommend to make some adjustments
		%in the \irace options as described in \autoref{sec:}

		%<<conc, prompt=TRUE, eval=TRUE, comment="">>=
		%irace:::concordance(iraceResults$testing$experiments)
		%@
		%

		During the tuning, \irace iteratively updates the sampling models of the
		parameters to focus on the best regions of the parameter search space. The
		frequency of the sampled configurations can provide insights on the parameter
		search space. We provide a function for plotting the
		frequency of the sampling of a set of configurations. For more information on this function, please see the \aR help, type in the
		\aR console: \code{?parameterFrequency}. The following example
		plots the frequency of the parameters sampled during one \irace run:

		<<freq, fig.pos="tbp", fig.cap="Parameters sampling frequency.", out.width="\\textwidth",prompt=TRUE, eval=TRUE, comment="">>=
		parameterFrequency(iraceResults$allConfigurations, iraceResults$parameters)
		@

		By using parallel coordinates plots, it is possible to analyze how the parameters
		interact with each other. For more information on this function, please see the \aR help, type
		in the \aR console: (\code{?parallelCoordinatesPlot}).
		The following example shows how to create a parallel
		coordinate plot of the configurations in the last two iterations of \irace.

		<<parcord, fig.pos="tbp", fig.align="center", out.width="0.7\\textwidth", fig.cap="Parallel coordinate plots of the parameters of the configurations in the last two iterations of a run of \\irace.", prompt=FALSE, eval=TRUE, comment="">>=
		# Get last iteration number
		last <- length(iraceResults$iterationElites)
		# Get configurations in the last two iterations
		conf <- getConfigurationByIteration(iraceResults = iraceResults,
		iterations = c(last - 1, last))
		parallelCoordinatesPlot (conf, iraceResults$parameters,
		param_names = c("algorithm", "alpha", "beta", "rho", "q0"),
		hierarchy = FALSE)
		@

		%%% FIXME: This plot is completely misleading. We should either calculate the
		%%% performance over the whole training set, or find a way to plot the
		%%% performance estimated by irace as a confidence interval to account for
		%%% different number of instances. Or simply plot a heatmap of instances x
		%%% best-of-each iteration.
		\hide{It is also possible to plot the performance evolution of the best-so-far configuration over the number of experiments as follows:
			%% <<trainEvo, fig.pos="tbp", fig.align="center", out.width='0.75\\textwidth', fig.cap="Training set performance of the best-so-far configuration over number of experiments.", prompt=FALSE, eval=TRUE, comment="">>=
			# Get number of iterations
			iters <- unique(iraceResults$experimentLog[, "iteration"])
			# Get number of experiments (runs of target-runner) up to each iteration
			fes <- cumsum(table(iraceResults$experimentLog[,"iteration"]))
			# Get the mean value of all experiments executed up to each iteration
			# for the best configuration of that iteration.
			values <- sapply(iters, function(k) {
				instances <- as.character(
				unique(iraceResults$experimentLog[
				iraceResults$experimentLog[, "iteration"] == k,
				"instance"]))
				return(mean(iraceResults$experiments[
				instances,
				as.character(iraceResults$iterationElites[k])]))})
			plot(fes, values, type="s", xlab = "Number of runs of the target algorithm",
			ylab = "Estimated mean value over whole training set")
			points(fes, values)
		}

		It is also possible to plot the performance on the test set of the best-so-far configuration over the number of experiments as follows:
		<<testEvo, fig.pos="tbp", fig.align="center", out.width='0.75\\textwidth', fig.cap="Testing set performance of the best-so-far configuration over number of experiments. Label of each point is the configuration ID.", prompt=FALSE, eval=TRUE, comment="">>=
		# Get number of iterations
		iters <- unique(iraceResults$experimentLog[, "iteration"])
		# Get number of experiments (runs of target-runner) up to each iteration
		fes <- cumsum(table(iraceResults$experimentLog[,"iteration"]))
		# Get the mean value of all experiments executed up to each iteration
		# for the best configuration of that iteration.
		elites <- as.character(iraceResults$iterationElites)
		values <- colMeans(iraceResults$testing$experiments[, elites])
		plot(fes, values, type = "s",
		xlab = "Number of runs of the target algorithm",
		ylab = "Mean value over testing set")
		points(fes, values)
		text(fes, values, elites, pos = 1)
		@

		The \irace package also provides an implementation of the ablation
		method~\cite{FawHoos2015ablation}. See \autoref{sec:ablation}.

		%%
		%%
		%%
		%% Advanced options
		%%
		%%
		%%

		\section{Advanced topics}

		\subsection{Tuning budget}\label{sec:budget}

		Before setting the budget for a run of \irace, please consider the number of
		parameters that need to be tuned, available processing power and available
		time. The optimal budget depends on the difficulty of the tuning scenario, the
		size of the parameter space and the heterogeneity of the instances. Typical
		values range from $1\,000$ to $100\,000$ runs of the target algorithm, although
		smaller and larger values are also possible. Currently, \irace does not detect
		whether the given budget allows generating all possible configurations. In such
		a case, the use of \emph{iterated} racing is unnecessary: One can simply
		perform a single race of all configurations (see FAQ in
		\autoref{faq:allconfs}).

		\Irace provides two options for setting the total tuning budget (\parameter{maxExperiments} and 
		\parameter{maxTime}). The option \parameter{maxExperiments} 
		limits the number of executions of \parameter{targetRunner} performed by \irace. The option \parameter{maxTime} limits the 
		total time of the \parameter{targetRunner} executions. When this latter option is used, \parameter{targetRunner} 
		must return the evaluation cost together with the execution time (\code{"cost time"}).

		\begin{xwarningbox}
			When the goal is to minimize the computation time of an algorithm, and you wish to use \parameter{maxTime} as the tuning budget, \parameter{targetRunner} must return the time also as the evaluation cost, that is, return the time two times as \code{"time time"}.
		\end{xwarningbox}

		\begin{xwarningbox}
			When using \parameter{targetEvaluator} and using \parameter{maxTime} as tuning budget, \parameter{targetRunner} just returns the time (\code{"time"}) and \parameter{targetEvaluator} returns the cost.
		\end{xwarningbox}

		When using \parameter{maxTime}, \irace estimates the execution time of each \parameter{targetRunner} execution 
		before the configuration. The amount of budget used for the estimation is set with the option \parameter{budgetEstimation}
		(default is $2\%$). The obtained estimation is adjusted after each iteration using the obtained results and it is used to estimate the number of experiments that can be executed. Internally, \irace uses the number of remaining experiments to 
		adjust the number of configurations tested in each race.

		\subsection{Multi-objective tuning}\label{sec:multi objective}

		Currently, \irace only optimizes one cost value at a time, which can be
		solution cost, computation time or any other objective that is returned to
		\irace by the \parameter{targetRunner}. If the target algorithm is
		multi-objective, it will typically return not a single cost value, but a set of
		objective vectors (typically, a Pareto front). For tuning such a target
		algorithm with \irace, there are two alternatives. If the algorithm returns a
		single vector of objective values, they can be aggregated into one single number
		by using, for example, a weighted sum. Otherwise, if the target algorithm
		returns a set of objective vectors, a unary quality metric (\eg~the
		hypervolume) may be used to evaluate the quality of the set.\footnote{An implementation is publicly available at \url{http://lopez-ibanez.eu/hypervolume}~\cite{FonPaqLop06:hypervolume}}

		% The first option is simple, it requires to devise a formula that can aggregate
		% the objectives in a way that balances the importance of all of them. This might
		% not be an easy task in some scenarios, and therefore using a more adequate
		% indicator to evaluate the performance of a multi-objective optimizer, such as
		% the hypervolume, is strongly advised.

		%For using the hypervolume as evaluation \irace needs to postpone the evaluation of the
		%configurations in an instance until all the executions have been completed. Once
		%the executions are finalized, the evaluation starts by obtaining reference points for 
		%each objective from the configuration results. These points are used to define the 
		%Pareto front and the hypervolume of each configuration
		%is calculated and the tuning process continues normally.

		The use of aggregation or quality metrics often requires normalizing the
		different objectives. If normalization bounds are known a priori for each
		instance, normalized values can be computed by \parameter{targetRunner}.
		Otherwise, the bounds may be dynamically computed while running \irace, by
		using \parameter{targetEvaluator}. In this case, \parameter{targetRunner} will
		save the output of the algorithm, then the first call
		to \parameter{targetEvaluator} will examine the output produced by all calls
		to \parameter{targetRunner} for the same instance, update the normalization
		bounds and return the normalized output. Subsequent calls
		to \parameter{targetEvaluator} for the same instance will simply return the
		normalized output.
		%
		A similar approach can be used to dynamically compute the reference points or
		sets often required by unary quality metrics.

		For more information about defining a \parameter{targetEvaluator}, see
		\autoref{sec:evaluator}. Examples of tuning a multi-objective target algorithm
		using the hypervolume can be found in the examples at
		\IRACEHOME{/examples/hypervolume} and \IRACEHOME{/examples/moaco}.

		\subsection{Tuning for minimizing computation time}
		\label{sec:capping}

		When using \irace for tuning algorithms that report computation time to
		reach a target, \parameter{targetRunner} should return the execution time of a configuration instead of solution cost.

		%Even though \irace can be used for
		%minimizing computation time, \irace may itself require more time to do so in
		%its current version than other methods, such as
		%\code{ParamILS}\footnote{\url{http://www.cs.ubc.ca/labs/beta/Projects/ParamILS/}}
		%or \code{SMAC}\footnote{\url{http://www.cs.ubc.ca/labs/beta/Projects/SMAC/}},
		%since it does not make use of techniques, such as ``adaptive capping'', that
		%avoid long runs of the target algorithm.
		%We are currently extending \irace with an adaptive capping mechanism.

		Starting from version $3.0$,  \irace includes an elitist racing procedure that 
		implements an \textbf{adaptive capping mechanism} \citep{PerLopHooStu2017:lion}. 
		Adaptive capping \citep{HutHooLeyStu2009jair} is a configuration technique 
		that avoids the execution of long runs of the target algorithm, focusing the 
		configuration budget in the evaluation of the best configurations found. This is 
		done by bounding the execution time of each configuration based on the best 
		performing candidate configurations.

		To use adaptive capping, the \parameter{capping} option must be enabled and the 
		\parameter{elitist} irace option must be selected. When evaluating candidate 
		configurations on an instance, \irace calculates an execution bound based on the 
		execution times of the elite configurations.  The \parameter{boundType} option defines 
		how the performance of the elite configurations  is defined to obtain the execution bound. 
		The default value of \parameter{boundType} calculates the performance ($p_i^s$) of each 
		elite configuration ($s$) as the mean execution time of the  instances already executed 
		in the race and the currently executed instance ($i$). The \parameter{cappingType} option 
		specifies the measure used to obtain the elite configurations bound. By default, the execution bound 
		is calculated as the median of the execution times of the elite configurations:
		%
		\begin{equation}
			b_i= \text{Median}_{\theta_s \in \Celite} \{ p_i^s\} 
		\end{equation}

		The execution bound for new configurations ($j$) is calculated by multiplying the elite configurations
		bound by the number of instances ($i$) in the execution list and subtracting the mean execution time 
		of the instances executed by the candidate:
		%
		\begin{equation}
			k_i^{'j} = b_{i} \cdot i + \bmin - p^{j}_{i-1} \cdot (i-1)
		\end{equation}

		A small constant \bmin is added to account for time measurements errors. 
		These settings are also used to apply a dominance elimination criterion together with the
		statistical test elimination. The domination criterion is defined as:
		%
		\begin{equation}\label{eq:dom}
			b_i + \bmin  < p_i^{j}
		\end{equation}

		When elite configurations dominate new configurations, these are eliminated from the race.

		\begin{xwarningbox}
			The default statistical test when \parameter{capping} is enabled is \code{t-test}. 
			This test is more appropriate to configure algorithms for optimizing runtime (see \autoref{sec:stat test}).
		\end{xwarningbox}

		The execution bound is constantly adjusted by \irace based on the best configurations 
		times, nevertheless, a maximum execution time (\bmax) is never  exceeded. This maximum 
		execution time must be defined in the configuration scenario when \parameter{capping} 
		is enabled. To specify the maximum execution bound for the target runner executions 
		use the \parameter{boundMax} option. The final execution bound ($k_i^j$) is calculated by:
		%
		\begin{equation}\label{eq:cap}
			k_i^j = \begin{cases}
				\bmax &\text{if $k_i^{'j} > \bmax$,} \\
				\min \{b_i, \bmax\}& \text{if $k_i^{'j} \leq 0$,}\\
				k_i^{'j}&\text{otherwise;}
			\end{cases}
		\end{equation}

		Additionally, the \parameter{boundDigits} option defines the precision of the time bound
		provided by \irace, the default setting is 0.

		Timed out executions occur when the maximum execution bound
		(\parameter{boundMax}) is reached and the algorithm has not achieved successful
		termination or a defined quality goal.  In this case, it is a common practice
		to apply a penalty known as PARX, in which timeouts are penalized by
		multiplying \parameter{boundMax} by a constant $X$. The constant $X$ may be set
		using the \parameter{boundPar} option. Bounded executions are executions that
		do not achieve successful termination or a defined quality goal in the
		execution bound ($k_i^j$) set by \irace, which is smaller than
		\parameter{boundMax}. The \parameter{boundAsTimeout} option replaces the
		evaluation of bounded executions by the \parameter{boundMax} value. More
		details about the implementation of adaptive capping can be found in
		\citet{PerLopHooStu2017:lion}.

		\begin{xwarningbox}
			Note that bounded executions are not timed out executions and thus, they will not be penalized by PARX. 
		\end{xwarningbox}

		\begin{xwarningbox}
			Penalized evaluations of timed out and bounded executions are only used for
			the elimination tests and the comparison between the quality of
			configurations. To calculate execution bounds and computation budget
			consumed, \irace uses only unpenalized execution times. The unpenalized
			execution time must be provided by the target runner or target evaluator as
			described in \autoref{sec:runner} and \autoref{sec:evaluator} .
		\end{xwarningbox}

		\subsection{Hyper-pararameter optimization of machine learning methods}

		The \irace package can also be used for model selection and hyper-parameter optimization 
		of machine learning methods. For such a task, we recommend the \pkg{mlr} package~\citep{R:mlr}. 
		The following webpage documents how to use \irace for this purpose: \url{https://mlr-org.github.io/mlr-tutorial/devel/html/advanced_tune/index.html}

		\subsection{Heterogeneous scenarios}
		\label{sec:het}

		We classify a scenario as homogeneous when the target algorithm has a
		consistent performance regarding the instances; roughly speaking, good
		configurations tend to perform well and bad configurations tend to perform
		poorly on all instances of the problem. By contrast, in heterogeneous
		scenarios, the target algorithm has an inconsistent performance on different
		instances, that is, some configurations perform well for a subset of the
		instances, while they perform poorly for a different subset.

		When facing a heterogeneous scenario, the first question should be whether the
		objective of tuning is to find configurations that perform reasonably well over
		all instances, even if they are not the best ones in any of them. If this is
		not the case, then it would be better to partition instances into more similar
		subsets and execute \irace separately on each subset. This will lead to a
		portfolio of algorithm configurations, one for each subset, and algorithm
		selection techniques can be used to select the best configuration from the
		portfolio when facing a new instance.

		If finding an overall good configuration for all the instances is the
		objective, then we recommend that instances are randomly sampled
		(option \parameter{sampleInstances}), unless one can provide the instances in a
		particular order that does not bias the tuning towards any subset.  For
		example, let's assume a heterogeneous scenario with two classes of instances.  If
		training instances are not sampled and the first ten instances belong to only
		one class, the tuning will be biased towards configurations that perform good
		for those instances.  An optimal order would not ever present consecutively two
		instances of the same type.

		In addition, it may be useful to increase the number of instances executed
		before doing a statistical test in order to see more instance classes before
		discarding configurations. The option \parameter{elitistNewInstances} in elitist
		\irace (option \parameter{elitist}) can be used to increase the number of new
		instances executed in each iteration, \eg \code{--elitist-new-instances 5} (default
		value is 1). For the non-elitist \irace, the option \parameter{firstTest} may be used for the same purpose, \eg \code{--first-test 10} (default value is 5).
		\MANUEL{And both at the same time?}

		While executing \irace, the homogeneity of the scenario can be observed by
		examining the values of Spearman's rank correlation coefficient and Kendall's
		concordance coefficient in the text output of \irace.  See \autoref{sec:output
			text} for more information.

		\subsection{Choosing the statistical test}
		\label{sec:stat test}

		The statistical test used in \irace identifies statistically bad performing
		configurations that can be discarded from the race in order to save
		budget. Different statistical tests use different criteria to compare the
		cost of the configurations, which has an effect on the tuning results.

		\Irace provides two types of statistical tests
		(option \parameter{testType}). Each test has different characteristics that are
		beneficial for different goals:
		%
		\begin{itemize}
			\item Friedman test (\code{F-test}): This test uses the ranking of the
			configurations to analyze the differences between their performance. This
			makes the test suitable for scenarios where the scale of the performance
			metric is not as important to assess configurations as their relative
			ranking.  This test is also indicated when the distribution of the mean
			performances deviates greatly from a normal distribution. For example, the
			ranges of the performance metric on different instances may be completely
			difference and comparing the performance of different configurations using
			the mean over multiple instances may be deceiving.  We recommend to use the
			\code{F-test} (default when \parameter{capping} is not enabled) when tuning
			for solution cost and whenever the best performing algorithm should be among
			the best in as many instances as possible.

			\item Student's t-test (\code{t-test}): This test uses the mean performance of
			the configurations to analyze the differences between the
			configurations.\footnote{The t-test does not require that the performance
				values follow a normal distribution, only that the distribution of sample
				means does. In practice, the t-test is robust despite large deviations from
				the assumptions.} This makes the test suitable for scenarios where the
			differences between values obtained for different instances are relevant to
			assess good configurations.  We recommend using t-test, in particular, when
			the target algorithm is minimizing computation time and, in general, whenever
			the best configurations should obtain the best average solution cost.
		\end{itemize}

		The confidence level of the tests may be adjusted by using the
		option \parameter{confidence}. Increasing the value of \parameter{confidence}
		leads to a more strict statistical test.  Keep in mind that a stricter test
		will require more budget to identify which configurations perform worse. A less
		strict test discards configurations faster by requiring less evidence against
		them and, therefore, it is more likely to discard good configurations.

		\subsection{Complex parameter space constraints}

		Some parameters may have complex dependencies. Ideally, parameters should be
		defined in the way that is more likely to help the search performed by
		\irace.  For example, when tuning a branch and bound algorithm, one may have the
		following parameters:

		\begin{itemize}
			\item branching (\code{b}) that takes values in \code{\{0,1,2,3\}}, where 0 indicates no branching will be used
			and the rest are different types of branching.
			\item stabilization (\code{s}) that takes values in \code{\{0,1,2,3,4,5,6,7,8,9,10\}}, of which for \code{b=0}
			only \code{\{0,1,2,3,4,5\}} are relevant.
		\end{itemize}

		In this case, it is not possible to describe the parameter space by defining
		only two parameters for \irace. An extra parameter must be introduced as
		follows:

		\begin{center}
			\begin{minipage}{0.8\linewidth}
				\begin{CodeInput}
					# name    label   type   range                    condition
					b         "-b "   c      (0,1,2,3)
					s1        "-s "   c      (0,1,2,3,4,5)            | b == "0"
					s2        "-s "   c      (0,1,2,3,4,5,6,7,8,9,10) | b != "0"
				\end{CodeInput}
			\end{minipage}
		\end{center}

		Parameters whose values depend on the value of other parameters may also
		require using extra parameters or changing the parameters and processing them
		in \parameter{targetRunner}. For example, given the following parameters:
		%
		\begin{itemize}
			\item Population size (\code{p}) takes the integer values $[1,100]$.
			\item Selection size (\code{s}) takes the same values but no more than the
			population size, that is $[1,$\code{p}$]$.
		\end{itemize}

		In this case, it is possible to describe the parameters \code{p} and \code{s}
		using surrogate parameters for \irace that represent a ratio of the original
		interval as follows:

		\begin{center}
			\begin{minipage}{0.8\linewidth}
				\begin{CodeInput}
					# name    label   type   range
					p        "-p "    i      (1,100)
					s_f      "-s "   r      (0.0,1.0)
				\end{CodeInput}
			\end{minipage}
		\end{center}
		%
		and \parameter{targetRunner} must calculate the actual value of \code{s} as $\min(\max(\text{round}(\text{\code{s\_f}}\cdot \text{\code{p}}, 1)), 100)$. For
		example, if the parameter \code{p}  has value 50 and the surrogate parameter \code{s\_f} has value 0.3, then \code{s} will have value 15.

		The processing within \parameter{targetRunner} can also split and join
		parameters. For example, assume the following parameters:

		\begin{center}
			\begin{minipage}{0.8\linewidth}
				\begin{CodeInput}
					# name    label   type   range
					m         "-m "   i      (1,250)
					e         "-e "   r      (0.0,2.0)
				\end{CodeInput}
			\end{minipage}
		\end{center}

		These parameters could be used to define a value
		$\texttt{m}\cdot 10^\texttt{e}$ for another parameter
		(\code{--strength}) not known by \irace. Then, \parameter{targetRunner} takes care of parsing
		\code{-m} and \code{-e}, computing the strength value and passing the parameter
		\code{--strength} together with its value to the target algorithm.

		More complex parameter space constraints may be implemented by means of the
		\parameter{repairConfiguration} function (\autoref{sec:repairconf}).

		\subsection{Unreliable target algorithms and immediate rejection}\label{sec:reject}

		There are some situations in which the target algorithm may fail to execute
		correctly.  By default, \irace stops as soon as a call
		to \parameter{targetRunner} or \parameter{targetEvaluator} fails, which helps to
		detect bugs in the target algorithm.  Sometimes the failure cannot be fixed
		because it is due to system problems, network issues, memory limits, bugs for
		which no fix is available, or fixing them is impossible because there is no
		access to the source code.

		In those cases, if the failure is caused by random errors or transient system
		problems, one may wish to ignore the error and try again the same call in the
		hope that it succeeds. The option \parameter{targetRunnerRetries} indicates the
		number of times a \parameter{targetRunner} execution is repeated if it
		fails. Use this option only if you know additional repetitions could be
		successful.

		If the target algorithm consistently fails for a particular set of
		configurations, these configurations may be declared as forbidden
		(\parameter{forbiddenFile}) so that \irace avoids them.  On the other hand, if
		the configurations that cause the problem are unknown,
		the \parameter{targetRunner} should return \code{Inf} so that \irace
		immediately rejects the failing configuration. This immediate rejection should
		be used with care according to the goals of the tuning. For example, a
		configuration that crashes on a particular instance, \eg by running out of
		memory, might still be considered acceptable if it gives very good results on
		other instances. The configurations which were rejected during the execution of \irace are saved
		in the Rdata output file (see \autoref{sec:output r}).
		\begin{xwarningbox}
			If the configuration budget is specified in total execution time (\parameter{maxTime} option), 
			immediate rejected executions must provide the cost and time (which must be \code{Inf 0}). 
			Nevertheless, rejected configurations will be  excluded from the execution time estimation 
			and the execution bound calculation.
		\end{xwarningbox}

		\subsection{Ablation Analysis}\label{sec:ablation}

		The ablation method~\cite{FawHoos2015ablation} takes two configurations (source
		and target) and generates a sequence of configurations that differ between each
		other just in one parameter, where parameter values in source are replaced by
		values from target. The sequence can be seen as a ``path'' from the source to
		the target configuration.  This can be used to find new better ``intermediate''
		configurations or to analyse the impact of the parameters in the performance.
		To perform ablation use the \code{ablation} function and specify the IDs of the
		source and target configurations.  By default, the source is taken as the first
		configuration evaluated by \irace and the target as the best overall
		configuration found. The argument \code{ab.params} can be used to specify a
		subset of the parameters considered in the ablation. The option
		\parameter{firstTest} defines how many instances are selected for the
		evaluation of configurations, if a different number of instances is required it
		must be specified in the argument \code{n.instance}. If a PDF filename is
		provided (\code{pdf.file}), a plot will be produced from the ablation results
		(Fig.~\ref{fig:testAb}).

		<<ablation, prompt=FALSE, eval=FALSE>>=
		ablation(iraceLogFile = "irace.Rdata", 
		src = 1, target = 60, pdf.file = "plot-ablation.pdf")
		@

		<<testAb, fig.pos="htb!", fig.align="center", out.width="0.75\\textwidth", fig.cap="Example of plot generated by \\code{ablation()}.", prompt=FALSE, eval=TRUE, echo=FALSE>>=
		plotAblation(abLogFile = "log-ablation.Rdata")
		@

		The function returns a list containing the following elements:
		%
		\begin{description}
			\item \code{configurations}: A dataframe of configurations tested during ablation.
			\item \code{instances}: The instances used for the ablation.
			\item \code{scenario}: Scenario options provided by the user.
			\item \code{trajectory}: Best configuration IDs at each step of the ablation.
			\item \code{best}: Best overall configuration found.
		\end{description}

		\subsection{Postselection race}\label{sec:postselection}

		After the configuration process is finished it is possible perform a
		postselection race by specifying the \irace option \parameter{postselection}
		with value larger than $0$.  This option will perform a post-selection race of
		the set of best configurations of each iteration. The budget assigned for this
		race is obtained using the \parameter{postselection} option which defines a
		percentage of the \irace configuration budget. This budget is not considered in
		the total configuration budget that is, these evaluations are extra
		computation.

		The execution of the postselection race add an element (\code{psrace.log}) to 
		the \code{iraceResults} list saved in the \irace log file. The postselection log
		consists of a list with the following elements:
		\begin{description}
			\item \code{configurations}: Configurations used in the postselection race.
			\item \code{instances}: Instances used in the in the postselection race.
			\item \code{maxExperiments}: Configuration budget assigned for the postselection race.
			\item \code{experiments}:  Matrix of experiments in the same format as the
			\code{iraceResults$experiments} matrix. The column names indicate the candidate
			configuration identifier and the row names contain the name of the instances.
			\item \code{elites}: Elite configurations obtained in the postselection race.
		\end{description}

		Optionally, it is possible to perform a postselection race with all 
		elite configurations of the iterations or selecting a set of configurations
		from \code{iraceResults$allConfigurations}.
		<<postsel, prompt=FALSE, eval=FALSE>>=
		# Execute all elite configurations in the iterations
		psRace(iraceLogFile="irace.Rdata", elites=TRUE)
		# Execute a set of configurations IDs providing budget
		psRace(iraceLogFile="irace.Rdata", 
		conf.ids=c(34, 87, 102, 172, 293), 
		max.experiments=500)
		@

		\subsection[Parameter importance analysis using PyImp]{Parameter importance analysis using \PyImp}

		The \PyImp\footnote{\url{https://github.com/automl/ParameterImportance}} tool developed by the AutoML group\footnote{\url{https://www.automl.org/}}
		supports various parameter importance analysis methods using surrogate models. 
		Given a performance dataset of an algorithm configuration scenario, a Random Forest is built to predict performance of all algorithm configurations. 
		Parameter importance analyses are then applied on the prediction model. 
		The model serves as a surrogate for the original target algorithm, so that the algorithm does not need to be executed during the analyses.
		Three analysis methods are supported, namely fANOVA~\citep{HutHooLey2014icml} (functional analysis of variance), forward selection~\cite{HutHooLey2013lion},
		and ablation analysis with surrogates~\cite{BieLinEggFraFawHoo2017}. 
		Note that the \irace package directly supports ablation (without surrogate models) analysis with and without racing (\autoref{sec:ablation}).
		Although ablation analysis without surrogates may be more time-consuming, results of the surrogate version may be less accurate than the non-surrogate one.

		The \code{.Rdata} dataset generated by \irace can be used as input for \PyImp. 
		We provide a script to translate an \code{irace.Rdata} file into the input format required by \PyImp.
		The script is available in the \irace package, and can be accessed either through the \aR console (function \code{irace2pyimp}), or via command line (\IRACEHOME{/bin/irace2pyimp}).

		To see the usage of the executable, please run: \code{irace2pyimp --help}. 
		For more information on the \aR function \code{irace2pyimp}, type in the \aR console: \code{?irace2pyimp}.

		Given as input an \code{irace.Rdata} file, the script will generate the following output files:
		\begin{itemize}
			\item \code{params.pcs}: a text file containing the parameter space definition.
			\item \code{runhistory.json}: a JSON file containing the list of algorithm configurations evaluated during the tuning and the performance data obtained.
			\item \code{traj_aclib2.json}: a JSON file containing the best configurations after each iteration of \irace. The last configuration will be used as the target configuration in ablation analysis.
			\item \code{scenario.txt}: a text file containing the definition of the tuning scenario.
			\item \code{instances.txt}: a text file containing the list of instances.
			\item \code{features.csv}: a .csv file containing instance features. If no instance features are provided, the index of each instance will be used as a feature.
		\end{itemize}

		\PyImp can then be called using the files listed above as input. Several examples on how to use the script and to call \PyImp can be found at \IRACEHOME{/inst/examples/irace2pyimp/}.

		%\emph{\PyImp installation note:} At the time when this document is written (December 2019), the latest \PyImp release version (1.0.6) has some bugs. 
		%Please check on \url{https://pypi.org/project/PyImp/#history} to see if \PyImp version > 1.0.6 has been released. 
		%If it has, you can easily install \PyImp using \code{pip install pyimp}. 
		%Otherwise, please install \PyImp from source (\url{https://github.com/automl/ParameterImportance}), branch \code{development}. Installation instructions can be found at \url{https://automl.github.io/ParameterImportance/installation.html}.
		%%
		%% irace options
		%%

		\section{List of command-line and scenario options} \label{sec:irace options}

		Most \irace options can be specified in the command line using a flag or in the
		\irace scenario file using the option name (or setting their value in the
		\code{scenario} list passed to the various \aR functions exported by the
		package). This section describes the various \irace options that can be
		specified by the user in this way.

		\begin{xwarningbox}
			Relative filesystem paths (e.g., \path{../scenario/}) given in the
			command-line are relative to the current working directory (the directory at
			which \irace is invoked). However, paths given in
			the scenario file are relative to the directory containing
			the scenario file. See also Table~\ref{tab:paths}.
		\end{xwarningbox}

		\input{section/irace-options}

		%%
		%%
		%%
		%% FAQ
		%%
		%%
		%%

		\section{FAQ (Frequently Asked Questions)}

		\subsection{Is \irace minimizing or maximizing the output of my algorithm?}

		By default, \irace considers that the value returned by \parameter{targetRunner} (or by
		\parameter{targetEvaluator}, if used) should be \underline{\textbf{minimized}}. In case of a maximization
		problem, one can simply multiply the value by -1 before returning it to
		irace. This is done, for example, when maximizing the hypervolume (see the last
		lines in \IRACEHOME{/examples/hypervolume/target-evaluator}).

		\subsection[Is it possible to configure a MATLAB algorithm with irace?]
		{Is it possible to configure a \MATLAB algorithm with \irace ?}

		Definitely. There are three main ways to achieve this:
		\begin{enumerate}
			\item Edit the \parameter{targetRunner} script to call \MATLAB in a non-interactive
			way. See the \MATLAB documentation, or the following links.\footnote{\url{http://stackoverflow.com/questions/1518072/suppress-start-message-of-matlab}}\footnote{\url{http://stackoverflow.com/questions/4611195/how-to-call-matlab-from-command-line-and-print-to-stdout-before-exiting}} %
			You would need to pass the parameter received by \parameter{targetRunner} to your \MATLAB script.\footnote{\url{https://www.mathworks.com/matlabcentral/answers/97204-how-can-i-pass-input-parameters-when-running-matlab-in-batch-mode-in-windows}}\footnote{\url{https://stackoverflow.com/questions/3335505/how-can-i-pass-command-line-arguments-to-a-standalone-matlab-executable-running}} 
			%
			There is a minimal example in \IRACEHOME{/examples/matlab/}.

			\item Call \MATLAB code directly from \aR using the
			\pkg{matlabr} package (\url{https://cran.r-project.org/package=matlabr}). This
			is a better option if you are experienced in \aR. Define \parameter{targetRunner} as
			an \aR function instead of a path to a script. The function should
			call your \MATLAB code with appropriate parameters.

			\item Another possibility is calling \MATLAB directly from a different programming language and write \parameter{targetRunner} in that programming language,  for example, in \proglang{Python} (see examples in \IRACEHOME{/examples/target-runner-python/}).\footnote{\url{https://www.mathworks.com/help/matlab/matlab_external/call-matlab-functions-from-python.html}\\
				\url{https://www.mathworks.com/help/matlab/matlab_external/call-user-script-and-function-from-python.html}}
		\end{enumerate}

		\subsection{My program works perfectly on its own, but not when running under \irace. Is irace broken?}\label{faq:valgrind}

		Every time this was reported, it was a difficult-to-reproduce bug, i.e., a \href{https://en.wikipedia.org/wiki/Heisenbug}{Heisenbug}, in
		the program (target algorithm), not in \irace.
		%
		To detect such bugs, we recommend that you use, within
		\parameter{targetRunner}, a memory debugger (e.g.,
		\href{http://valgrind.org/}{\code{valgrind}}) to run your program. For example,
		if your program is executed by \parameter{targetRunner} as:
		% 
		\begin{lstlisting}[style=BashInputStyle]
			${EXE} ${FIXED_PARAMS} -i ${INSTANCE} ${CONFIG_PARAMS} 1> ${STDOUT} 2> ${STDERR}
		\end{lstlisting}
		%
		then replace that line with:
		%
		\begin{lstlisting}[style=BashInputStyle]
			valgrind --error-exitcode=1 ${EXE} ${FIXED_PARAMS} -i ${INSTANCE} \
			${CONFIG_PARAMS} 1> ${STDOUT} 2> ${STDERR}
		\end{lstlisting}

		If there are bugs in your program, they will appear in
		\code{\${STDERR}}, thus do not delete those files. Memory debuggers will
		significantly slowdown your code, so use them only as a means to find what is
		wrong with your target algorithm. Once you have fixed the bugs, you should
		remove the use of \code{valgrind}.

		\subsection{\irace seems to run forever without any progress, is this a bug?}\label{faq:infloop}

		Every time this problem was reported, the issue was in the target algorithm and
		not in \irace. Some ideas for debugging this problem:
		\begin{itemize}
			\item Check that the target algorithm is really not running nor paused nor
			sleeping nor waiting for input-output.
			\item Use \parameter{debugLevel}\code{=3} to see how  \irace calls \code{target-runner}, run the same command outside \irace and verify that it terminates.
			\item Add some output to your algorithm that reports at the very end the runtime and exit code. Verify that this output is printed when \irace calls your algorithm. 
			\item In \code{target-runner}, print something to a log file \emph{after} calling your target algorithm. Verify that this output appears in the log file when \irace is running.

			\item Set a maximum timeout when calling your target algorithm from \code{target-runner} (see FAQ~\ref{faq:timeout}).

		\end{itemize}

		\subsection{My program may be buggy and run into an infinite loop. Is it possible to set a maximum timeout?}\label{faq:timeout}

		We are not aware of any way to achieve this using \aR. However, in
		GNU/Linux, it is easy to implement by using the \code{timeout} command\footnote{\url{http://man7.org/linux/man-pages/man1/timeout.1.html}} in
		\code{targetRunner} when invoking your program.

		\subsection{When using the mpi option, \irace is aborted with an error message indicating that a
			function is not defined. How to fix this?}\label{sec:mpi-error}

		\pkg{Rmpi} does not work the same way when called from within a package and when called from a script or 
		interactively. When \irace creates the slave nodes, the slaves will load a copy of \irace automatically. 
		If the slave nodes are on different machines, they must have \irace installed. If \irace is not installed
		system-wide, \aR needs to be able to find \irace on the slave nodes. This is usually done by setting \code{R\_LIBS}, 
		\code{.libPaths()} or by loading \irace using \code{library()} or \code{require()} with the argument ``\code{lib.loc}''. 
		The settings on the master are not applied to the slave nodes automatically, thus the slave 
		nodes may need their own settings. After spawning the slaves, it is too late to modify 
		those settings, thus modifying the shell variable \code{R\_LIBS} seems the only valid 
		way to tell the slaves where to find \irace.

		If the path is set correctly and the problem persists, please check these instructions:
		% 
		\begin{enumerate}[leftmargin=*,widest=3]
			\item Test that \irace and \pkg{Rmpi} work. Run \irace on a single machine (submit node), without calling \code{qsub}, \code{mpirun} or a 
			similar wrapper around \irace or \aR.
			\item Test loading \irace on the slave nodes. However,
			jobs submitted by \code{qsub}/\code{mpirun} may load \aR packages using a different mechanism from the way it happens if you log directly into the node (e.g., with \code{ssh}). Thus, you need to write a little \aR program such as:
			<<faq3, eval=FALSE>>=
			library(Rmpi)
			mpi.spawn.Rslaves(nslaves = 10)
			paths <- mpi.applyLB(1:10, function(x) {
				library(irace); return(path.package("irace")) })
			print(paths)
			@
			Submit this program to the cluster like you would submit \irace (using \code{qsub}, \code{mpirun} or whatever program is used to submit jobs to the cluster).

			\item In the script \code{bin/parallel-irace-mpi}, the function \code{irace\_main()} creates an MPI job for our cluster. 
			You may need to speak with the admin of your cluster and ask them how to best submit a job for MPI. There may be 
			some particular settings that you need. \pkg{Rmpi} normally creates  log files; but \irace suppresses those files unless $\texttt{debugLevel} > 0$.
		\end{enumerate}

		Please contact us (\autoref{sec:contact}) if you have further problems.

		\subsection[Error: 4 arguments passed to {.Internal(nchar)} which requires 3]{Error: 4 arguments passed to \code{.Internal(nchar)} which requires 3}

		This is a bug in \aR 3.2.0 on Windows. The solution is to update your version of
		\aR.

		\subsection{Warning: \code{In read.table(filename,}  \code{header = TRUE,} \code{colClasses = "character",} \code{: incomplete final line found by} \ldots}

		This is a warning given by \aR when the last line of an input file does not
		finish with the newline character. The warning is harmless and can be
		ignored. If you want to suppress it, just open the file and press the
		\code{ENTER} key at the end of the last line of the file.

		\subsection{How are relative filesystem paths interpreted by \irace?}\label{faq:relpaths}

		The answer depends on where the path appears. Relative paths may appear as the
		argument of command-line options, as the value of options given in the scenario
		file, or within various scripts, functions or instance files. Table
		\ref{tab:paths} summarizes how paths are translated from relative to absolute.

		\begin{table}[h!]
			\centering
			\caption{Translation of relative to absolute filesystem paths.}
			\label{tab:paths}
			\begin{tabular}{rl}
				\toprule
				\bf    Relative path appears as \ldots  & \bf\ldots is relative to \ldots \\
				\midrule
				a string within \parameter{trainInstancesFile} & \parameter{trainInstancesDir}\\
				a string within \parameter{testInstancesFile} & \parameter{testInstancesDir}\\
				code within \parameter{targetRunner} or \parameter{targetEvaluator} & \parameter{execDir}\\
				the value of \parameter{logFile} or \code{--log-file} & \parameter{execDir}\\
				the value of other options in the scenario file  &  the directory containing the scenario file\\
				the value of other command-line options & invocation (working) directory of \irace\\
				\bottomrule    
			\end{tabular}
		\end{table}

		\subsection{My parameter space is small enough that \irace could generate all possible configurations; however, \irace generates repeated configurations and/or does not generate some of them. Is this a bug?}\label{faq:allconfs}

		Currently, \irace does not try to detect whether all possible configurations
		can be evaluated for the given budget, thus, the initial random sampling
		performed by \irace may generate repeated configurations and/or never generate
		some configurations, which is not ideal. The ideal approach in such cases is to
		provide all configurations explicitly to \irace (\autoref{sec:initial}) and
		execute a single race (\parameter{nbIterations}\code{=1}) with exactly the
		number of configurations provided (e.g.,
		\parameter{nbConfigurations}\code{=10}). A future version of \irace will
		automatically detect this case and switch to basic racing without having to set additional options.

		%%
		%%
		%%
		%% Contact info
		%%
		%%
		%%

		%FIXME: complete this section
		%\section{Known problems}

		\section{Resources and contact information} \label{sec:contact}

		More information about the package can be found on the \irace webpage:
		\begin{center} \url{http://iridia.ulb.ac.be/irace/} \end{center}

		For questions and suggestions please contact the development team through 
		the \irace package Google group:
		\begin{center}
			\url{https://groups.google.com/d/forum/irace-package}
		\end{center} 
		or by sending an email to:
		\begin{center}
			\href{mailto:irace-package@googlegroups.com}{irace-package@googlegroups.com}
		\end{center}

		%% MANUEL: I think we should use just one channel. If the Google group is the winner, that should be it.
		%the mailing list:
		%% LESLIE I agree.
		%\begin{center} \href{mailto:irace@iridia.ulb.ac.be}{irace@iridia.ulb.ac.be} \end{center}

		\section{Acknowledgements}

		We would like to thank all the people that directly or indirectly have
		collaborated in the development and improvement of \irace:
		%
		\newlist{acklist}{itemize*}{1}
		\setlist[acklist]{label={}, before=\unskip{}, afterlabel=\unskip{ }, itemjoin={{,}}, itemjoin*={, and }, after={{.}}}
		\begin{acklist}
			\item Prasanna Balaprakash
			\item Zhi (Eric) Yuan
			\item Franco Mascia
			\item Alberto Franzin
			\item Anthony Antoun
			\item Esteban Diaz Leiva %MATLAB example
			\item Federico Caselli % bug reports with fixes
			\item Pablo Valledor Pellicer % testing of --parallel under Windows
			\item André de Souza Andrade
			\item Nguyen Dang (\texttt{nttd@st-andrews.ac.uk}) % implemented \code{irace2pyimp}
		\end{acklist}

		\bibliographystyle{abbrvnat}
		\ifthenelse {\boolean{Release}}{%
			\bibliography{irace-package}%
		}{%
			\bibliography{optbib/abbrev,optbib/journals,optbib/authors,optbib/biblio,optbib/crossref}%
		}

		\newpage

		\begin{appendices}

			\section{Installing R} \label{sec:installation}

			This section gives a quick \aR installation guide that will work in
			most cases. The official instructions are available at
			\url{https://cran.r-project.org/doc/manuals/r-release/R-admin.html}

			\subsection{GNU/Linux}
			You should install \aR from your package manager. On a Debian/Ubuntu system
			it will be something like:
			\begin{lstlisting}[style=BashInputStyle]
				sudo apt-get install r-base
			\end{lstlisting}

			Once \aR is installed, you can launch \aR from the Terminal and from the \aR prompt
			install the \irace package (see \autoref{sec:irace install}).

			\subsection{OS X}

			You can install \aR directly from a CRAN mirror.\footnote{\url{https://cran.r-project.org/bin/macosx/}} %
			Alternatively, if you use
			homebrew, you can just brew the \aR formula from the science tap (unfortunately
			it does not come already bottled so you need to have Xcode\footnote{Xcode download
				webpage: \url{https://developer.apple.com/xcode/download/}} installed to compile it):
			%<<R_OS_install,engine='bash',eval=FALSE>>=
			\begin{lstlisting}[style=BashInputStyle]
				brew tap homebrew/science
				brew install r
			\end{lstlisting}
			%@

			Once \aR is installed, you can launch \aR from the Terminal (or from your Applications),
			and from the \aR prompt install the \irace package (see \autoref{sec:irace install}).

			\subsection{Windows}

			You can install \aR from a CRAN mirror.\footnote{\url{https://cran.r-project.org/bin/windows/}} %
			We recommend that you
			install \aR on a filesystem path without spaces, special characters or long
			names, such as \path{C:\R}. Once \aR is installed, you can launch the \aR
			console and install the \irace package from it (see \autoref{sec:irace
				install}).

			\section{targetRunner troubleshooting checklist} \label{sec:check list}

			If the \parameter{targetRunner} script fails to return the output expected by
			\irace, it can be sometimes difficult to diagnose where the problem lies. The
			more descriptive errors provided by your script, the easier it will be to debug
			it. If \parameter{targetRunner} enters an infinite loop, irace will wait
			indefinitely (see FAQ in \autoref{faq:timeout}).  If you are using temporary
			files to redirect the output of your algorithm, check that these files are
			properly created. We recommend to follow the structure of the example file
			(\code{target-runner}) provided in \IRACEHOME{/templates}. The following
			error examples are based on that example file.

			In case of failure of \parameter{targetRunner}, \irace will print an error on
			its output describing which execution of \parameter{targetRunner} failed.
			%
			Follow this checklist to detect where the problem is:

			\begin{enumerate}[leftmargin=*,widest=9]

				\item Make sure that your \code{targetRunner} script or program is at the specified location. If you see this error:
				\begin{CodeInput}
					Error: == irace == target runner '~/tuning/target-runner' does not exist
				\end{CodeInput}
				%
				it means that \irace cannot find the \code{target-runner} file. Check that the file is at
				the path specified by the error.

				\item Make sure that your \code{targetRunner} script is an executable file and the user running \irace has permission to execute it. The following errors:
				%
				\begin{CodeInput}
					Error: == irace == target runner '~/tuning/target-runner' is a directory,
					not a file
				\end{CodeInput}
				or
				\begin{CodeInput}
					Error: == irace == target runner '~/tuning/target-runner' is not executable
				\end{CodeInput}
				%
				mean that your \code{targetRunner} is not an executable file. In the first
				case, the script is a folder and therefore there must be a problem with the
				name of the script. In the second case, you must make the file executable,
				which in GNU/Linux can be done by:
				%
				\begin{lstlisting}[style=BashInputStyle]
					chmod +x ~/tuning/target-runner
				\end{lstlisting}

				\item If your \code{targetRunner} script calls another program, make sure it is at  the location described in the script
				(variable \code{EXE} in the examples and templates). A typical output for such an error is:
				%
				\begin{CodeInput}
					Error: == irace == running command ''~/tuning/target-runner' 1 8 676651103 
					~/tuning/Instances/1000-16.tsp --ras --localsearch 2 --alpha 4.03 --beta 1.89
					--rho  0.02 --ants 37 --nnls 48 --dlb 0 --rasranks 15 2>\&1' had status 1  
					== irace == The call to target.runner.default was:
					~/tuning/target-runner 1 8 676651103 ~/tuning/Instances/1000-16.tsp --ras 
					--localsearch 2 --alpha 4.03 --beta 1.89 --rho  0.02 --ants 37 --nnls 48 
					--dlb 0 --rasranks 15 
					== irace == The output was:
					Tue May  3 19:00:37 UTC 2016: error: ~/bin/acotsp: not found or not executable
					(pwd: ~/tuning/acotsp-arena)
				\end{CodeInput}
				%
				You may test your script by copying the command line shown in the error and
				executing \code{target-runner} directly on the execution directory (\parameter{execDir}). In this case, the command line is:
				%
				\begin{lstlisting}[style=BashInputStyle]
					~/tuning/target-runner 1 8 676651103 ~/tuning/Instances/1000-16.tsp  --ras \
					--localsearch 2 --alpha 4.03 --beta 1.89 --rho  0.02 --ants 37 --nnls 48 \
					--dlb 0 --rasranks 15
				\end{lstlisting}

				This executes the \code{targetRunner} script as \irace does. The output of this script must be only one number.

				\item Check that your \code{targetRunner} script is actually returning one
				number as output. For example:
				% 
				\begin{CodeInput}
					Error: == irace == The output of '~/tuning/target-runner 1 25 365157769
					~/tuning/Instances/1000-31.tsp --ras --localsearch 1 --alpha 0.26 --beta
					6.95 --rho 0.69 --ants 56 --nnls 10 --dlb 0 --rasranks 7' is not numeric! 
					== irace == The output was:
					Solution: 24479793
				\end{CodeInput}
				% FIXME: Duplicated above.
				%% For testing your script, copy the command-line of \code{target-runner} and execute it directly on the execution directory (\parameter{execDir}):
				%% %
				%% \begin{lstlisting}[style=BashInputStyle]
					%% ~/tuning/target-runner 1 25 365157769 ~/tuning/Instances/1000-31.tsp  --ras \
					%%  --localsearch 1 --alpha 0.26 --beta 6.95 --rho  0.69 --ants 56 \
					%%  --nnls 10 --dlb 0 --rasranks 7
					%% \end{lstlisting}
				%% This executes the \code{targetRunner} script as \irace does. 
				In the example above, the output of \code{target-runner} is ``\code{Solution:
					24479793}'', which is not a number. If \code{target-runner} is parsing the
				output of the target algorithm, you need to verify that the code only parses
				the solution cost value.

				\item Check that your \code{targetRunner} script is creating the output files
				for your algorithm.  If you see an error as:
				%
				\begin{CodeInput}
					== irace == The output was: Tue May  3 19:41:40 UTC 2016: 
					error: c1-9.stdout: No such file or directory
				\end{CodeInput}
				The output file of the execution of your algorithm has not been
				created (check permissions) or has been deleted before the result can be read.

				\item Other errors can produce the following output:
				%
				\begin{CodeInput}
					== irace == The output was: Tue May  3 19:49:06 UTC 2016: 
					error: c1-23.stdout: Output is not a number
				\end{CodeInput}
				%
				This might be because your \code{targetRunner} script is not executing your
				algorithm correctly. To further investigate this issue, comment out the line
				that eliminates the temporary files that saves the output of your
				algorithm. Similar to this one
				%
				\begin{lstlisting}[style=BashInputStyle]
					rm -f "${STDOUT}" "${STDERR}"
				\end{lstlisting}
				%
				Execute directly the \code{targetRunner} command-line that is provided in the
				error message, look in your execution directory for the files that are
				created. Check the \code{.stderr} file for errors and the \code{.stdout} file
				to see the output that your algorithm produces.

				\item Some command within \code{targetRunner} may not be working correctly. In
				that case, you must debug the commands individually exactly as \irace executes them. In
				order to find where the problem is, print the commands to a log file before
				executing them. For example:
				% 
				\begin{lstlisting}[style=BashInputStyle]
					echo "$EXE ${FIXED_PARAMS} -i $INSTANCE ${CONFIG_PARAMS}" >> ${STDERR}.log
					$EXE ${FIXED_PARAMS} -i $INSTANCE ${CONFIG_PARAMS} 1> ${STDOUT} 2> ${STDERR}
				\end{lstlisting}
				%
				then look at the \code{${STDERR}.log} file corresponding to the
				\code{targetRunner} call that failed and execute/debug the last command
				there.

				\item If the language of your operating system, the \code{target-runner} or the
				target algorithm is not English, \irace may not be able to recognize the
				numbers generated by \code{target-runner}. We recommend that you run \irace,
				the \code{target-runner} and the target algorithm under an English locale (or
				make sure that their languages and number format are compatible).

				\item It is possible that
				\href{https://en.wikipedia.org/wiki/Heisenbug}{transient bugs} in the target
				algorithm are only visible when running within \irace, and
				all commands within \code{targetRunner} appear to work fine when executed directly in the
				command-line outside \irace. See FAQ in \autoref{faq:valgrind}) for suggestions on how to detect such bugs.

				\item If your \code{targetRunner} script works when running irace with \parameter{parallel}\code{=0} but it fails when using higher number of cores, this may be due to any number of reasons:
				\begin{itemize}
					\item If you submit jobs through a queuing system, the running environment
					when using the queuing system may not be the same as when you launch \irace
					yourself. The queuing system may also send the job to different machines
					depending on the number of CPUs requested. One way to test this is to
					submit the failing execution of \code{targetRunner} to the queuing system,
					and specifically to any problematic machine.

					\item When using MPI, some calls to \code{targetRunner} may run on different
					computers than the one running the master \irace process. See FAQ in
					\autoref{sec:mpi-error}.

					\item Does \code{targetRunner} read or create intermediate files? These files
					may cause a race condition when two calls to \code{targetRunner} happen at
					the same time. You have to make sure that parallel runs of
					\code{targetRunner} do not interfere with each other's files.

					\item Maybe these files consume too much memory or fill the filesystem when
					there are simultaneous \code{targetRunner} calls? Moreover, queuing systems
					have stricter limits for computing nodes than for the submit/host node.

					\item Does the machine or the queuing system impose any limits on number of
					processes or CPU/memory/filesystem usage per job? Such limits may only
					trigger when more than one process is executed in parallel, killing the
					\code{targetRunner} process before it has a chance to print anything
					useful. In that case, \irace may not detect the the program finished
					unexpectedly, only that the expected output was not printed.
				\end{itemize}

			\end{enumerate}

			\section{targetEvaluator troubleshooting checklist} \label{sec:evaluator check list}

			Even if \parameter{targetRunner} appears to work, the use of \parameter{targetEvaluator} may lead to other problems. The same checklist of \code{targetRunner} can be followed here. In addition, we list here other potential problems unique to \code{targetEvaluator}:

			\begin{enumerate}
				\item If \parameter{targetEvaluator} fails only in the second or later
				iteration, this may because output files or data generated by a previous call
				to \parameter{targetRunner} are missing. Elite configurations are never
				re-executed on the same instance and seed pair, that is, \irace will call
				only once \parameter{targetRunner} for each pair of configuration ID and
				instance ID. However, \parameter{targetEvaluator} is always re-executed,
				which takes into account any updated information (normalization bounds,
				reference sets/points, best-known values, etc.). Thus, any files or data
				generated by \parameter{targetRunner} for a given configuration must remain
				available to \parameter{targetEvaluator} as long as that configuration is
				alive. The list of alive configurations is passed
				to \parameter{targetEvaluator}, which may decide then which data to keep or
				remove.
			\end{enumerate}

			\section{Glossary}

			\begin{description}
				\item[Parameter tuning:] Process of searching good settings for the parameters
				of an algorithm under a particular tuning scenario (instances, execution
				time, etc.).
				\item[Scenario:] Settings that define an instance of the tuning problem. These
				settings include the algorithm to be tuned (target), budget for the execution
				of the target algorithm (execution time, evaluations, iterations, etc.), set
				of problem instances and all the information that is required to perform the
				tuning.
				\item[Target algorithm:] Algorithm whose parameters will be tuned.
				\item[Target parameter:] Parameter of the target algorithm that will be tuned.
				\item[\irace option:] Configurable option of \irace.
				\item[Elite configurations:] Best configurations found so far by \irace. New
				configurations for the next iteration of \irace are sampled from the
				probabilistic models associated to the elite configurations.  All elite
				configurations are also included in the next iteration.
				\item[\hypertarget{irace_home}{\path{$IRACE_HOME}:}] The filesystem path where \irace is
				installed. You can find this information by opening an \aR console and
				executing:

				<<R_irace_home2, prompt=FALSE, eval=FALSE>>=
				system.file(package = "irace")
				@

			\end{description}

			\section{NEWS}

			\RecustomVerbatimCommand{\VerbatimInput}{VerbatimInput}%
			{fontsize=\footnotesize,
				%
				frame=lines,  % top and bottom rule only
				framesep=1em, % separation between frame and text
				rulecolor=\color{Gray},
				%
				label=\fbox{\color{Black}NEWS},
				labelposition=topline,
				%
				% commandchars=\|\(\), % escape character and argument delimiters for
				%                      % commands within the verbatim
				% commentchar=*        % comment character
			}

			%% NEWS is latin1 but this file is utf8
			\begingroup\inputencoding{utf8}%
			\VerbatimInput{NEWS.txt}%
			\endgroup

		\end{appendices}

	\end{document}

	%%% Local Variables:
	%%% TeX-master: "irace-package.Rnw"
	%%% End:

	% LocalWords: iteratively

### 2023-02-18 tsp_python/main.py

@@ -1,64 +1,64 @@
 
-# import sys
-from time import time
+import sys
+# from time import time
 
-from brkga_mp_ipr.enums import Sense
-from brkga_mp_ipr.types_io import load_configuration
-from brkga_mp_ipr.algorithm import BrkgaMpIpr
+# from brkga_mp_ipr.enums import Sense
+# from brkga_mp_ipr.types_io import load_configuration
+# from brkga_mp_ipr.algorithm import BrkgaMpIpr
 
 from tsp_instance import TSPInstance
-from tsp_decoder import TSPDecoder
-import greedy_tour as gt
+# from tsp_decoder import TSPDecoder
+# import greedy_tour as gt
 
 from anneal import SA
-from grasp import GRASP
+# from grasp import GRASP
 import common
 
 
-def brkga(instance_file):
-    # if len(sys.argv) < 4:
-    #     print("Usage: python main_minimal.py <seed> <config-file> "
-    #           "<num-generations> <tsp-instance-file>")
-    #     sys.exit(1)
+# def brkga(instance_file):
+#     # if len(sys.argv) < 4:
+#     #     print("Usage: python main_minimal.py <seed> <config-file> "
+#     #           "<num-generations> <tsp-instance-file>")
+#     #     sys.exit(1)
 
-    # seed = int(sys.argv[1])
-    # config_file = sys.argv[2]
-    # num_generations = int(sys.argv[3])
-    # instance_file = sys.argv[4]
+#     # seed = int(sys.argv[1])
+#     # config_file = sys.argv[2]
+#     # num_generations = int(sys.argv[3])
+#     # instance_file = sys.argv[4]
 
-    seed            = 1
-    config_file     = "config.conf"
-    num_generations = 50
-    # instance_file   = "instances/burma14.dat"    
+#     seed            = 1
+#     config_file     = "config.conf"
+#     num_generations = 50
+#     # instance_file   = "instances/burma14.dat"    
 
-    print(f"BRKGA reading data from {instance_file}")
-    instance = TSPInstance(instance_file)
+#     print(f"BRKGA reading data from {instance_file}")
+#     instance = TSPInstance(instance_file)
 
-    print(f"BRKGA  reading parameters from {config_file}")
-    brkga_params, _ = load_configuration(config_file)
+#     print(f"BRKGA  reading parameters from {config_file}")
[…]

### 2023-02-27 latex/comp_or.tex
Running iRace

@@
 
-%The testing experiments are performed on a 64-bit, 16GiB, 3.2GHz, 8-core processor, 2 threads per core, AMD FX-8320E, with {\it Linux Ubuntu 22.04} as the operational system and {\it Python 3.10.4}\/ as the programming language.
-
 The final experiments are performed on a 64-bit, 16GiB, 3.6GHz, 8-core AMD® Fx-8320e processor, 2 threads per core, with {\it Linux Ubuntu 20} as the operational system and {\it Python 3.8.2} as the programming language.
+
+A relevant problem with meta-heuristics is the parameter's adjustment to extract the best possible efficiency in the optimization.
+The bulk of the tools to fine-tune these parameters utilize a statistical backdrop that enables them to make accurate predictions regarding the differences in candidate configurations' performance.
+Inadequacies in the design of the statistical experiment may therefore lead to erroneous results from the statistical tests and, subsequently, cause the approach to produce inaccurate comparisons of candidate configurations.
+That is why all parameters were set using the $iRace$ package \citep{LopezIbanezManuel2016}.
+Supplying $iRace$ with a range for $volThreshold$ as [0.80 to 0.99], {\it iRace} yielded ....., by training on the uneven instances numbers.
 
@@
 
+	Another important contribution is that we employed the {\it iRace} package to fine tune the meta-heuristics parameters, guaranteeing the best possible performance for the instances under training and testing.
+
 \section*{Acknowledgments}

### 2023-02-27 common.py

@@ -4,6 +4,8 @@ import math
 import numpy as np
 import multiprocessing as mp
 import os.path
+from time import time
+import random
 
 CITIES = ["GRU", "GIG", "SSA", "CNF", "CWB", "BSB", "REC"]
 
@@ -153,8 +155,17 @@ def fillPallet(pallet, items, k, nodeTorque, solDict, cfg, threshold, itemsDict,
             counter += 1
     return counter
 
-def loadDistances():
-    fname =  f"./params/distances.txt"      
+def loadDistances(fname):
+
+    fname = os.path.abspath(fname)
+
+    output_string = ""
+    str_list = fname.split("iRace/")
+    for element in str_list:
+        output_string += element
+
+    fname = output_string
+    
     with open(fname, 'r') as f:
         distances = [ [ float(num) for num in line.split(' ') ] for line in f ] 
     return distances 
@@ -301,8 +312,17 @@ def loadNodeItems(scenario, instance, node, unatended, surplus): # unatended, fu
     """
     Load this node to unnatended items attributes
     """
-    dirname = f"./{surplus}/scenario_{scenario}/instance_{instance}"
-    fname = f"{dirname}/items.txt"
+
+    fname = f"{surplus}/scenario_{scenario}/instance_{instance}/items.txt"
+
+    fname = os.path.abspath(fname)
+
+    output_string = ""
+    str_list = fname.split("iRace/")
+    for element in str_list:
+        output_string += element
+
+    fname = output_string
 
     reader = open(fname, "r")
     lines = reader.readlines() 
@@ -542,6 +562,52 @@ def getTourValue(tour, cfg, pallets, cons):
         tour.cost += abs(epsilom) * (tour.legsCosts[k+1] / 20)
 
     return float(sTourAccum)/tour.cost
+
+
+# for testing only
+def testingGetCons(N, rndm):
+    """
+    Testing consolidated from cons_0_0.txt
+    """
+    cons = []
+
+    lines = [
+        [ 2620,  406,  4.121, 0, 2],
+        [2595, 2436, 13.690, 0, 2],
+        [3282, 3081, 13.689, 0, 2],
+        [4217, 3856, 13.674, 0, 2],
+        [2558, 3180, 13.692, 0, 1],
+        [3526, 3630, 13.691, 0, 1],
+        [4348, 3509, 13.697, 0, 1]
+    ]
+    # lines = [
[…]

### 2023-02-27 mpShims.py

@@ -120,8 +120,6 @@ def Solve(pallets, items, cfg, k, threshold, secBreak, mode, nodeTorque, solDict
     else:
         mode = "Serial"
 
-    print(f"\n{mode} Shims for ACLP+RPDP")
-
     surplus = 1.0 + 3*(1.0-threshold)
 
     lock  = mp.Lock()
@@ -161,9 +159,135 @@ def Solve(pallets, items, cfg, k, threshold, secBreak, mode, nodeTorque, solDict
             optcgcons.minCGdev(pallets, k, nodeTorque, cfg)
             getBestShims(      pallets[i], items, k, nodeTorque, solDict, cfg, surplus,   itemsDict, lock, tipo)
             counter += common.fillPallet( pallets[i], items, k, nodeTorque, solDict, cfg, 1.0, itemsDict, lock) 
-        print(f"---> {counter} items inserted by the local search.")
 
 
 if __name__ == "__main__":
 
-    print("----- Please execute module main -----")
+    import sys
+    seed         = sys.argv[1]
+    instance     = sys.argv[2] # /home/celio/Projects/ACLP_RPDP_P/data20/scenario_2/instance_3/items.txt'
+    volThreshold = float(sys.argv[3])
+
+    values = instance.split('/')
+
+    lenv = len(values)
+
+    # /home/celio/Projects/ACLP_RPDP_P/data20/scenario_2/instance_3/items.txt'
+    #                                  data20/scenario_2/instance_3/items.txt'
+
+    if lenv == 4:
+        surplus   = values[0]
+        scenario = int(values[1][-1:])
+        instance = int(values[2][-1:]) 
+
+    if lenv == 9:
+        surplus   = values[5]
+        scenario = int(values[6][-1:])
+        instance = int(values[7][-1:])         
+
+    # method = "mpShims" # best volThreshold = 
+    method = "Shims"  # best volThreshold: data20 (0.8512, )
+
+    secBreak = 1.8
+
+    tipo = "FFD"
+
+    # --- distances and costs matrix ---
+    dists = common.loadDistances("params/distances.txt")
+
+    costs = [[0.0 for _ in dists] for _ in dists]
+
+    cfg = common.Config(scenario)                                      
+
+    for i, cols in enumerate(dists):
+        for j, dist in enumerate(cols):
+            costs[i][j] = cfg.kmCost*dist
+
+    pallets, rampDistCG = common.loadPallets(cfg)
+    lock = mp.Lock
+
+    # pallets capacities
[…]

### 2023-02-27 latex/comp_or.tex
Paper update

@@
 That is why all parameters were set using the $iRace$ package \citep{LopezIbanezManuel2016}.
-Supplying $iRace$ with a range for $volThreshold$ as [0.80 to 0.99], {\it iRace} yielded ....., by training on the uneven instances numbers.
+Supplying $iRace$ with a range for $volThreshold$ as [0.80 to 0.99], {\it iRace} yielded 0.8542 for volume surplus 20\%, by training on the uneven instances numbers.
 

### 2023-02-28 latex/PO_suportabilidade.tex
PO suportabilidade

% !TeX spellcheck = pt_BR
\documentclass{article}
\usepackage[utf8]{inputenc}

\title{Pesquisa Operacional em Suportabilidade}
\author{Antonio Celio Pereira de Mesquita}
\date{ Abril 2023}

\begin{document}

\maketitle
\section{Introdução}

\section{Organização científica da manutenção}

\subsection{Decisões sobre internalizar ou não a manutenção de 3° nível}\

O processo de tomada de decisão sobre internalizar ou não a manutenção de de 3º nível normalmente envolve a ponderação dos custos e benefícios associados a ambas as opções. Internalizar a manutenção pode ser uma opção mais econômica, pois elimina a necessidade de contratar e pagar pessoal de manutenção externo. Por outro lado, pode ser um processo mais demorado e complexo, pois requer treinamento e gerenciamento da equipe para garantir a qualidade do trabalho. Além disso, pode ser difícil obter peças e componentes com o mesmo custo de fornecedores de manutenção externos. Em última análise, a decisão deve ser baseada em uma modelagem e otimização dos custos, benefícios e riscos associados.
Normalmente, esta análise é realizada pelo AeroLogLab por meio da plataforma OPUS 10.

\subsection{Análise de dados para diagnóstico ou prognóstico}\

A análise de dados para diagnóstico ou prognóstico para planejamento de manutenção de componentes de aeronaves normalmente envolve a coleta e análise de uma variedade de fontes de dados, como registros de voo e manutenção, previsões meteorológicas, aeronaves e sistemas e outras fontes de informações relacionadas à aeronave. Esses dados são usados para identificar possíveis problemas que podem afetar o desempenho e a segurança da aeronave, bem como para avaliar a condição de componentes e sistemas.
Esse tipo de análise é frequentemente usado para antecipar e planejar futuras tarefas de manutenção, como manutenção preventiva, reparos e substituições. Além disso, a análise de dados pode ser usada para monitorar o desempenho de aeronaves e identificar possíveis tendências ou padrões na manutenção de aeronaves.
Ainda não houve trabalho científico capitaneado pelo AeroLogLab sobre este assunto por falta de candidato ao programa de pós-graduação com experiência sobre este tema.

\subsection{Avaliação do impacto de atrasos e cancelamentos em rotas operacionais}\

Os atrasos e cancelamentos em rotas operacionais de aeronaves comerciais envolve a análise de diversos fatores, como o custo direto e indireto, o tempo perdido para os passageiros, a satisfação dos clientes, a segurança e a eficiência operacional.
O custo direto inclui prejuízos financeiros como a perda de receita, custos de combustível, compensações aos passageiros, aumento de horas extras para os tripulantes e outras despesas. O custo indireto inclui os efeitos sobre a imagem da companhia aérea, a fidelidade dos passageiros e a reputação da companhia aérea. Além disso, a análise do impacto também pode incluir fatores como o número de atrasos e cancelamentos, a duração dos atrasos e o tipo de aeronave.
Ainda não houve trabalho científico capitaneado pelo AeroLogLab sobre este assunto por falta de candidato ao programa de pós-graduação com experiência sobre este tema.

\subsection{Plano de manutenção resiliente com otimização e aprendizado de máquina}

Um plano de manutenção resiliente com otimização e aprendizado de máquina é uma abordagem proativa para o planejamento de manutenção que usa análise preditiva e ciência de dados para identificar possíveis problemas e otimizar os cronogramas de manutenção.
Esse tipo de plano usa algoritmos de aprendizado de máquina para analisar leituras de sensores, dados históricos e outras informações para prever quando o equipamento provavelmente precisará de manutenção e otimizar o tempo das tarefas de manutenção.
Essa abordagem pode melhorar a eficiência e a confiabilidade, além de reduzir o tempo de inatividade e os custos. Além disso, ajuda a identificar possíveis problemas antes que se tornem grandes, evitando assim reparos caros ou paradas não planejadas.

a Equação \ref{eq:prev} estabelece o custo da manutenção preventiva.

\begin{equation}\label{eq:prev}
prev_j =  pmc_j + pmoc_j,\ for\ j \in \{1, 2, 3, \ldots, m\}
\end{equation}

A Equação \ref{eq:corr} estabelece o custo da manutenção corretiva.

\begin{equation}\label{eq:corr}
corr_j =  cmc_j + cmoc_j,\ for\ j \in \{1, 2, 3, \ldots, m\}
\end{equation}

A Equação  \ref{eq:dist} estabelece a distância em termos temporais entre o limite de horas de voo e o momento da parada da aeronave.

\begin{equation}\label{eq:dist}
dist_{ij} = \lceil{\frac{stop_i}{max_j}}\rceil{}-\frac{stop_i}{max_j} 
\end{equation}

A Equação  \ref{eq:prob} determina a probabilidade de falha.

\begin{equation}\label{eq:prob}
prob_{ij} = 1 - e^{(-\lambda{}_j * stop_i)}
\end{equation}

As tarefas de preparação não podem ser duplicadas. As tarefas executam preparações comuns comente uma vez, a fim de não desperdiçar recursos.
$K_i$ é o conjunto de preparações do pacote "i". A restrição que as mantêm únicas é o tipo de dados "set" da linguagem {\it Python}, que não permite elementos duplicados.

A Equação  \ref{eq:min} estabelece a Função Objetivo.

\begin{equation} \label{eq:min}
minimize\ =\ \sum_{i=1}^{m} \sum_{j=1}^{n} X_{ij} \times ( prev_j + dist_{ij}*prev_j + prob_{ij}*corr_j )
\end{equation}

\subsection{Problema de planejamento de voos e manutenção, do inglês flight maintenance planning problem (FMPP)}

O Problema de Planejamento de Voos e Manutenção é um problema de otimização que visa encontrar a melhor rota para o voo de um avião e a manutenção necessária para mantê-lo em condições seguras. Ele envolve determinar a rota ótima para o voo, bem como o calendário de manutenção apropriado para a aeronave antes de cada voo.

Está relacionado ao problema de programação de voos, mas tem um foco maior na manutenção e na segurança operacional. É importante para garantir que os voos sejam executados com segurança e eficiência, além de reduzir custos de manutenção. Este problema envolve as seguintes variáveis:

1. Tempo de voo
2. Custo de combustível
3. Tempo de manutenção
4. Custo de manutenção
5. Restrições de segurança
6. Restrições de custo
7. Restrições de tempo
8. Restrições de rota

Ainda não houve trabalho científico capitaneado pelo AeroLogLab sobre este assunto por falta de candidato ao programa de pós-graduação com experiência sobre este tema.

\subsection{Problema de roteamento e manutenção de aeronaves para frotas compartilhadas, incluindo informações de prognóstico de saúde das aeronaves}

Os problemas de roteamento e manutenção de aeronaves são considerações críticas para frotas compartilhadas de aeronaves. Esses problemas afetam a segurança e a confiabilidade da aeronave e podem ter consequências financeiras significativas para a companhia aérea ou empresa que possui e opera a frota.

O roteamento de aeronave refere-se ao processo de determinação da rota ideal para uma aeronave viajar, levando em consideração fatores como consumo de combustível, condições climáticas e congestionamento do tráfego aéreo. Esse processo envolve a análise de grandes quantidades de dados, incluindo dados históricos de voos, previsões meteorológicas e informações de tráfego aéreo em tempo real. O objetivo do roteamento de aeronaves é minimizar o tempo de voo, o consumo de combustível e os custos operacionais gerais, garantindo segurança e confiabilidade.

Os problemas de manutenção, por outro lado, referem-se aos problemas que podem surgir durante a operação de uma aeronave, como falhas mecânicas, problemas no motor e outros problemas técnicos. A manutenção é fundamental para a segurança e confiabilidade da aeronave, e é essencial manter a frota em boas condições para evitar paradas inesperadas e reparos caros.

Uma maneira de resolver problemas de roteamento e manutenção de aeronaves é por meio do uso de sistemas de prognóstico de integridade de aeronaves. Esses sistemas usam sensores e análise de dados para monitorar a integridade e o desempenho dos componentes da aeronave em tempo real, permitindo que as equipes de manutenção detectem e resolvam possíveis problemas antes que se tornem problemas sérios. Os sistemas de manutenção preditiva também podem usar algoritmos de aprendizado de máquina para analisar dados históricos e prever quando a manutenção é necessária, ajudando a minimizar o tempo de inatividade e reduzir os custos.

Em resumo, os problemas de roteamento e manutenção de aeronaves são considerações críticas para frotas compartilhadas de aeronaves, e o gerenciamento eficaz dessas questões é essencial para garantir a segurança, confiabilidade e lucratividade da frota. O uso de sistemas de prognóstico de integridade de aeronaves pode ajudar a resolver esses problemas, fornecendo monitoramento em tempo real e recursos de manutenção preditiva.

\subsection{Otimização combinatória das tarefas de manutenção de aeronaves}

Conhecido como um problema de programação ou {\it scheduling}, tarefas são agrupadas em pacotes temporais, algumas tarefas eventualmente poderão ser colocadas como {\it out-of-phase} (fora dos pacotes) e, dentro de cada pacote, as tarefas são agrupadas por zona, por qualificação do mecânico e por precedência entre tarefas.
O objetivo global é a eficiência traduzida como minimização de custos e do tempo de parada, bem como minimização dos custos de oportunidade.

\subsection{O problema da otimização do leiaute de uma instalação para manutenção de aeronaves}

A otimização do layout de uma instalação para manutenção de aeronaves é uma parte importante do processo geral de manutenção de aeronaves. Este processo envolve a otimização do espaço dentro da instalação para garantir que todos os equipamentos e pessoal necessários estejam localizados nas posições mais eficientes para maximizar a produtividade e minimizar os custos.
Isso pode envolver reorganizar o layout da instalação, como mover estações de trabalho e equipamentos, adicionar ou remover paredes e reconfigurar o fluxo de pessoal e materiais. Além disso, as instalações devem ser projetadas para reduzir riscos, como perigos elétricos e de incêndio, além de permitir acesso fácil a ferramentas e peças. O layout ideal de uma instalação para manutenção de aeronaves também inclui levar em consideração a ergonomia e a segurança do pessoal, fornecendo espaço adequado para movimentação e equipamentos de segurança.
Sob o ponto de vista da otimização, o objetivo seria minimizar os percursos de movimentação dos técnicos, conferindo agilidade e eficiência. As ferramentas mais utilizadas tenderiam a ficar mais acessíveis e as pouco utilizadas em posições mais recuadas.
Ainda não houve trabalho científico capitaneado pelo AeroLogLab sobre este assunto por falta de candidato ao programa de pós-graduação com experiência sobre este tema.

\subsection{Otimização da localização de instalações de manutenção}

A otimização da localização de instalações de manutenção de aeronaves é a análise de fatores para determinar o melhor local para a instalação de uma instalação de manutenção de aeronaves. Estes fatores incluem fatores como a proximidade de outras instalações de manutenção, a quantidade de tráfego aéreo, as condições climáticas, a infraestrutura disponível, as qualificações de mecânicos disponíveis e outros fatores. Além disso, existem várias variáveis que devem ser consideradas, como custos, localização e segurança.
O objetivo é maximizar a eficiência global da companhia aérea.
Ainda não houve trabalho científico capitaneado pelo AeroLogLab sobre este assunto por falta de candidato ao programa de pós-graduação com experiência sobre este tema.

\section{Organização científica do suprimento}

\subsection{Otimização de estoques de peças sobressalentes}

A otimização de peças de reposição é um processo utilizado para minimizar a quantidade de peças de reposição que uma empresa precisa ter em seu estoque a fim de atender à demanda dos clientes, especialmente, no nosso caso, as oficinas de manutenção. Envolve analisar os vários tipos de peças estocadas e determinar quais peças são realmente necessárias e quais possuem baixa taxa de consumo. Eventualmente, compensa ser reativo e adquirir somente quando surge a necessidade. A não ser quando é sabido que determinados fornecimentos são demorados e o componente precisa ser mantido em estoque a fim de evitar disrupção nas operações.

A otimização do estoque de peças sobressalentes pode ajudar uma empresa a economizar dinheiro em custos de armazenamento, reduzir o desperdício e melhorar o atendimento ao cliente, tendo as peças necessárias prontamente disponíveis quando os clientes precisarem delas, seja oriunda do estoque ou de um fornecedor local.

\subsection{Simulação dos processos internos de suprimento e otimização de leiaute}

Os processos internos e a otimização de leiaute pode ser simulada e testada para determinar o melhor leiaute possível para um determinado espaço.
A simulação de processos internos e otimização de leiaute podem ajudar a reduzir custos, melhorar o atendimento ao cliente do armazém e aumentar a eficiência da cadeia de suprimentos.

\subsection{Otimização da localização de instalações de distribuição}

\subsection{Análise multicritério na seleção de fornecedores}

\subsection{Estimativas de abastecimento por meio de aprendizado de máquina}

\subsection{problema de substituição de múltiplos equipamentos para modernização de meia vida}

\subsection{problema de gerenciamento de LRU, To summarize, we discovered two problems. First, spare parts supply pays little attention to optimizing transportation routes. Second, the existing transportation route optimization methods are seldom considered from the standpoint of supportability}

\subsection{problema de Suporte Logístico Integrado envolvendo frotas, spare parts supply}

\subsection{problema da gestão de obsolescência}

\section{Desenvolvimento da suportabilidade}

\subsection{Análise multicritério dos elementos de suporte logístico}

\subsection{Otimização e simulação da suportabilidade de um novo produto}

\subsection{Análise e simulação da degradação por questões de suportabilidade}

\subsection{problema de gerenciamento dos desempenhos de suportabilidade}

\subsection{problema de alocação de aeronaves à luz do desempenho individualizado em termos de confiabilidade de cada aeronave}

\subsection{problema de imaturidade nos fatores nos fatores RAMS}

\subsection{optimizing reliability, maintainability and supportability under performance based logistics}

\subsection{Predict key Performance and Cost metrics for new system}

\subsection{Calculate Supportability for legacy and new systems}

\end{document}

### 2023-02-28 latex/PO_suportabilidade.tex
PO suportabilidade

@@
 
+A otimização da localização das instalações de distribuição é um aspecto importante do gerenciamento da cadeia de suprimentos.
+Ao localizar estrategicamente os centros de distribuição próximos ao ponto de demanda, as empresas podem reduzir os custos de transporte, reduzir os prazos de entrega e melhorar o atendimento ao cliente.
+
+A otimização da localização das instalações de distribuição geralmente é feita usando uma combinação de modelos analíticos, como sistemas de informações geográficas (GIS), para identificar os locais mais adequados e, em seguida, usar algoritmos de otimização para escolher a localização ideal.
+Esse processo pode ajudar a garantir que o centro de distribuição esteja localizado no local mais eficiente e econômico.
+
+Localizações não otimizadas pode acarretar ineficiência na suportabilidade de sistemas complexos.
+
 \subsection{Análise multicritério na seleção de fornecedores}
+
+%sudo apt-get install myspell-pt-br
 

### 2023-02-28 latex/PO_suportabilidade.tex
Chapter update

@@
 \usepackage[utf8]{inputenc}
+\usepackage[left]{lineno}
+\usepackage[T1]{fontenc}
+\usepackage{lmodern}
+\usepackage{multirow}
+\usepackage{lineno,hyperref}
+\modulolinenumbers[5]
+\usepackage[none]{hyphenat}
+\usepackage{url}
+\usepackage{booktabs}
+\usepackage{graphicx}
+\usepackage{mathtools}
+\usepackage{amssymb}
+\usepackage{amsthm}
+\usepackage{caption}
+\usepackage{subfigure}
+
+\begin{document}
 
@@
 
-\begin{document}
+\maketitle
 
-\maketitle
+	Este capítulo aborda os problemas inerentes à suportabilidade de sistemas, esboça modelos matemáticos e as possibilidades do uso das ferramentas de Pesquisa Operacional na sua solução.
+
 \section{Introdução}
+
+A capacidade de suporte de sistemas complexos refere-se à capacidade de um sistema tecnológico ou uma organização, de ser efetivamente mantido, atualizado, reparado e aprimorado ao longo do tempo.
+Um sistema complexo é aquele que é composto por muitas partes interconectadas que interagem de forma não linear, tornando difícil prever como o sistema se comportará em diferentes condições.
+Porém, um sistema complexo pode ser subdividido, para fins de análise, em subsistemas modeláveis e otimizáveis.
+
+A capacidade de suporte envolve vários aspectos, incluindo a capacidade de diagnosticar problemas, a disponibilidade dos recursos necessários (como pessoal, ferramentas e peças de reposição), a eficiência dos processos de manutenção e reparo, a capacidade de fazer alterações no sistema sem causar interrupções e a capacidade de se adaptar a requisitos e ambientes em mudança.
+
+Também é importante destacar que cada subsistema está exposto a riscos que precisam ser quantificados sob os pontos de vista da probabilidade de ocorrência de algum infortúnio, bem como do impacto que isto causaria ao sistema.
+
+Um sistema complexo e seus subsistemas devem ser projetados com a capacidade de suporte em mente desde a concepção.
+Isso pode envolver o projeto do sistema com componentes modulares que podem ser facilmente substituídos ou atualizados, utilizando-se de ferramentas de diagnóstico eficazes, garantindo que os recursos necessários estejam disponíveis e projetando processos de manutenção e reparo que sejam eficientes.
+Além disso, o sistema deve ser continuamente monitorado e avaliado para identificar possíveis problemas, riscos e possibilidades de melhoria.
+
+Melhores detalhes poderão ser encontrados em {\it Marques, Henrique Costa; Fernando Teixeira Mendes Abrahão; Guilherme Conceição Rocha.
+Mapping Challenges Through the ILS Approach.
+DOI: 10.4018/978-1-7998-3904-0.ch012}
+
+\subsection{Como a pesquisa operacional pode ser aplicada à suportabilidade}
+
+A pesquisa operacional (PO) é um campo de estudo que usa modelagem matemática, técnicas de otimização e métodos analíticos para resolver problemas complexos em negócios, engenharia e outros campos.
+O objetivo da pesquisa operacional é encontrar soluções ótimas para problemas que envolvem múltiplas variáveis, restrições e objetivos.
+
+A PO pode ser usada para mitigar problemas de capacidade de suporte em sistemas complexos, fornecendo uma estrutura para analisar e melhorar a capacidade de suporte do sistema.
+Especificamente, a PO pode ser usada para:
+
+\begin{itemize}
+
+\item Prever a probabilidade e o impacto dos problemas de suporte: ao criar modelos que simulam o comportamento do sistema e identificam possíveis gargalos ou pontos de falha, a PO pode ajudar a prever onde os problemas de suporte são mais prováveis de ocorrer e qual seria o impacto desses problemas.
+
+\item Otimizar a alocação de recursos: a PO pode ajudar a otimizar a alocação de recursos (como pessoal, ferramentas e peças sobressalentes) para garantir que os problemas de capacidade de suporte sejam tratados de forma eficiente.
+
+\item Projetar processos eficazes de manutenção e reparo: a PO pode ser usada para projetar processos de manutenção e reparo que sejam eficientes e minimizem a interrupção do sistema.
+
+\item Avaliar o impacto das mudanças: A PO pode ser usada para avaliar o impacto das mudanças propostas no sistema (como atualizações ou modificações) na capacidade de suporte e para identificar possíveis problemas antes que eles ocorram.
+
+\end{itemize}
+
+No geral, a PO fornece um poderoso conjunto de ferramentas e técnicas para mitigar problemas de capacidade de suporte em sistemas complexos e pode ajudar a garantir que esses sistemas permaneçam confiáveis ao longo do tempo.
 
@@
 Em última análise, a decisão deve ser baseada em uma modelagem e otimização dos custos, benefícios e riscos associados.
-Normalmente, esta análise é realizada pelo AeroLogLab por meio da plataforma OPUS 10.
+Normalmente, esta análise é realizada pelo AeroLogLab por meio da plataforma OPUS 10 (www.systecongroup.com).
 
@@
 Esses dados são usados para identificar possíveis problemas que podem afetar o desempenho e a segurança da aeronave, bem como para avaliar a condição de componentes e sistemas.
+
 Esse tipo de análise é frequentemente usado para antecipar e planejar futuras tarefas de manutenção, como manutenção preventiva, reparos e substituições.
 Além disso, a análise de dados pode ser usada para monitorar o desempenho de aeronaves e identificar possíveis tendências ou padrões na manutenção de aeronaves.
-Ainda não houve trabalho científico capitaneado pelo AeroLogLab sobre este assunto por falta de candidato ao programa de pós-graduação com experiência sobre este tema.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.5\textwidth]{Images/analise_dados.jpg}
+	\small\textsuperscript{Origem: livre de direitos autorais em www.pexels.com}
+\end{figure}
+
+Informações relacionadas poderão ser encontradas em {\it Imbassahy, Dennys Wallace Duncan; Henrique Costa Marques, Guilherme Conceição Rocha, and Alberto Martinetti.
+Empowering Predictive Maintenance: A Hybrid Method to Diagnose Abnormal Situations.
+Applied Sciences 10.19 DOI: 10.4018/978-1-7998-3904-0.ch012 1-27}
 
@@
 Os atrasos e cancelamentos em rotas operacionais de aeronaves comerciais envolve a análise de diversos fatores, como o custo direto e indireto, o tempo perdido para os passageiros, a satisfação dos clientes, a segurança e a eficiência operacional.
+
 O custo direto inclui prejuízos financeiros como a perda de receita, custos de combustível, compensações aos passageiros, aumento de horas extras para os tripulantes e outras despesas.
@@
 Além disso, a análise do impacto também pode incluir fatores como o número de atrasos e cancelamentos, a duração dos atrasos e o tipo de aeronave.
-Ainda não houve trabalho científico capitaneado pelo AeroLogLab sobre este assunto por falta de candidato ao programa de pós-graduação com experiência sobre este tema.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.5\textwidth]{Images/flight_delay.jpg}
+	\small\textsuperscript{Origem: livre de direitos autorais em www.pixabay.com}
+\end{figure}
+
+Um modelo matemático e uma ferramenta de PO pertinente podem auxiliar no planejamento, levando em conta os riscos de atraso ou cancelamento, com o objetivo de antever esses riscos e facilitar o planejamento.
 
@@
 Esse tipo de plano usa algoritmos de aprendizado de máquina para analisar leituras de sensores, dados históricos e outras informações para prever quando o equipamento provavelmente precisará de manutenção e otimizar o tempo das tarefas de manutenção.
+
 Essa abordagem pode melhorar a eficiência e a confiabilidade, além de reduzir o tempo de inatividade e os custos.
 Além disso, ajuda a identificar possíveis problemas antes que se tornem grandes, evitando assim reparos caros ou paradas não planejadas.
+
+Conhecido como um problema de programação ou {\it scheduling}, tarefas são agrupadas em pacotes temporais, algumas tarefas eventualmente poderão ser colocadas como {\it out-of-phase} (fora dos pacotes) e, dentro de cada pacote, as tarefas são agrupadas por zona, por qualificação do mecânico e por precedência entre tarefas.
+O objetivo global é a eficiência traduzida como minimização de custos e do tempo de parada, bem como minimização dos custos de oportunidade.
 
@@
 As tarefas de preparação não podem ser duplicadas. As tarefas executam preparações comuns comente uma vez, a fim de não desperdiçar recursos.
-$K_i$ é o conjunto de preparações do pacote "i". A restrição que as mantêm únicas é o tipo de dados "set" da linguagem {\it Python}, que não permite elementos duplicados.
-
-A Equação  \ref{eq:min} estabelece a Função Objetivo.
+$K_i$ é o conjunto de preparações do pacote $i$. A restrição que as mantêm únicas é o tipo de dados $set$" da linguagem {\it Python}, que não permite elementos duplicados.
 
@@
 
-\subsection{Problema de planejamento de voos e manutenção, do inglês flight maintenance planning problem (FMPP)}
+A Equação \ref{eq:min} estabelece a Função Objetivo, que visa minimizar custos.
 
-O Problema de Planejamento de Voos e Manutenção é um problema de otimização que visa encontrar a melhor rota para o voo de um avião e a manutenção necessária para mantê-lo em condições seguras.
+\subsection{Problema de planejamento de voos e manutenção}
+
+Do inglês {\it flight maintenance planning problem} (FMPP), podem planejados pelas companhias aéreas por meio da divisão em grupos de problemas elementares: Desenvolvimento das rotas, Desenho do cronograma, Designação de frota, Rotas de manutenção e Pareamento de tripulação.
+
+As abordagens que são usadas para resolver esses sub-problemas podem ser divididas em dois grupos básicos: abordagens que são usadas apenas para resolver um dos problemas mencionados (abordagens de propósito único) e procedimentos que resolvem mais de um problema básicos ao mesmo tempo (abordagens integradas).
+
+O Problema de Planejamento de Voos e Manutenção é uma abordagem integrada de otimização que visa encontrar a melhor rota para o voo de um avião e a manutenção necessária para mantê-lo em condições seguras.
 Ele envolve determinar a rota ótima para o voo, bem como o calendário de manutenção apropriado para a aeronave antes de cada voo.
@@
 
-1. Tempo de voo
-2. Custo de combustível
-3. Tempo de manutenção
-4. Custo de manutenção
-5. Restrições de segurança
-6. Restrições de custo
-7. Restrições de tempo
-8. Restrições de rota
-
-Ainda não houve trabalho científico capitaneado pelo AeroLogLab sobre este assunto por falta de candidato ao programa de pós-graduação com experiência sobre este tema.
+\begin{minipage}[h]{.98\linewidth}
+	\begin{minipage}[t]{0.33\linewidth}
+		\begin{tabular}{l}\hline
+		Tempo de voo\\
+		Custo de combustível\\
+		Consumo por hora de voo\\
+		Tempo de manutenção\\
+		Custo de manutenção\\
+		Custo de oportunidade\\
+		Restrições de segurança\\
+		Restrições de custo\\
+		Restrições de tempo\\
+		Restrições de rota\\
+		\end{tabular}
+	\end{minipage}
+	\hfill
+	\begin{minipage}[t]{0.53\linewidth}
+		\vspace{-2ex}
+		\includegraphics[width=1.0\linewidth]{Images/programacao_voo.jpg}
+		\small\textsuperscript{Origem: livre de direitos autorais em www.pexels.com}
+	\end{minipage}
+\end{minipage}
 
@@
 
-Os problemas de manutenção, por outro lado, referem-se aos problemas que podem surgir durante a operação de uma aeronave, como falhas mecânicas, problemas no motor e outros problemas técnicos.
-A manutenção é fundamental para a segurança e confiabilidade da aeronave, e é essencial manter a frota em boas condições para evitar paradas inesperadas e reparos caros.
-
 Uma maneira de resolver problemas de roteamento e manutenção de aeronaves é por meio do uso de sistemas de prognóstico de integridade de aeronaves.
@@
 
-\subsection{Otimização combinatória das tarefas de manutenção de aeronaves}
-
-Conhecido como um problema de programação ou {\it scheduling}, tarefas são agrupadas em pacotes temporais, algumas tarefas eventualmente poderão ser colocadas como {\it out-of-phase} (fora dos pacotes) e, dentro de cada pacote, as tarefas são agrupadas por zona, por qualificação do mecânico e por precedência entre tarefas.
-O objetivo global é a eficiência traduzida como minimização de custos e do tempo de parada, bem como minimização dos custos de oportunidade.
+Para detalhes sobre este assunto consulte {\it Pereira Barreto, Eduardo Afonso.
+Modelling the aircraft maintenance routing problem for fractional fleets with the inclusion of prognostics and health monitoring information.
+São José dos Campos.
+Instituto Tecnológico de Aeronáutica (Aeronautical Institute of Technology).}
 
@@
 
-A otimização do layout de uma instalação para manutenção de aeronaves é uma parte importante do processo geral de manutenção de aeronaves.
+A otimização do leiaute de uma instalação para manutenção de aeronaves é uma parte importante do processo geral de manutenção de aeronaves.
 Este processo envolve a otimização do espaço dentro da instalação para garantir que todos os equipamentos e pessoal necessários estejam localizados nas posições mais eficientes para maximizar a produtividade e minimizar os custos.
-Isso pode envolver reorganizar o layout da instalação, como mover estações de trabalho e equipamentos, adicionar ou remover paredes e reconfigurar o fluxo de pessoal e materiais.
+
+Isso pode envolver reorganizar o leiaute da instalação, como mover estações de trabalho e equipamentos, adicionar ou remover paredes e reconfigurar o fluxo de pessoal e materiais.
 Além disso, as instalações devem ser projetadas para reduzir riscos, como perigos elétricos e de incêndio, além de permitir acesso fácil a ferramentas e peças.
-O layout ideal de uma instalação para manutenção de aeronaves também inclui levar em consideração a ergonomia e a segurança do pessoal, fornecendo espaço adequado para movimentação e equipamentos de segurança.
+
+O leiaute ideal de uma instalação para manutenção de aeronaves também inclui levar em consideração a ergonomia e a segurança do pessoal, fornecendo espaço adequado para movimentação e equipamentos de segurança.
+
 Sob o ponto de vista da otimização, o objetivo seria minimizar os percursos de movimentação dos técnicos, conferindo agilidade e eficiência.
 As ferramentas mais utilizadas tenderiam a ficar mais acessíveis e as pouco utilizadas em posições mais recuadas.
-Ainda não houve trabalho científico capitaneado pelo AeroLogLab sobre este assunto por falta de candidato ao programa de pós-graduação com experiência sobre este tema.
+
+O modelo matemático geralmente utilizado é o de uma árvore geradora mínima ({\it minimum spanning tree}).
+
+Dada um grafo (G, $\phi$), pode-se definir o peso de um subgráfo $H \subset G$ como:
+
+\begin{equation} \label{eq:fo1}
+\phi(H) \gets \sum_{e \in E(H)} \phi (e)
+\end{equation}
+
+$E(H)$ é o subconjunto de arestas $e$ de $H$.
+
+Em um grafo conectado $G$, uma árvore geradora mínima $T$ é uma árvore com valor mínimo.
+
+\begin{equation} \label{eq:fo2}
+min_T\ H(T) \gets \sum_{e \in E(H)} \phi (e) \times X_{ij}
+\end{equation}
+
+Considere $X_{ij}$ igual a 1 se a aresta (i,j) está na árvore, 0 caso contrário.
 
@@
 
-A otimização da localização de instalações de manutenção de aeronaves é a análise de fatores para determinar o melhor local para a instalação de uma instalação de manutenção de aeronaves.
-Estes fatores incluem fatores como a proximidade de outras instalações de manutenção, a quantidade de tráfego aéreo, as condições climáticas, a infraestrutura disponível, as qualificações de mecânicos disponíveis e outros fatores.
+A otimização da localização de instalações de manutenção de aeronaves é a análise de fatores para determinar o melhor local para a instalação de um hangar de manutenção.
+
+Estes fatores incluem a proximidade de outras instalações de manutenção, a quantidade de tráfego aéreo, as condições climáticas, a infraestrutura disponível, as qualificações de mecânicos disponíveis e outros fatores.
 Além disso, existem várias variáveis que devem ser consideradas, como custos, localização e segurança.
-O objetivo é maximizar a eficiência global da companhia aérea.
-Ainda não houve trabalho científico capitaneado pelo AeroLogLab sobre este assunto por falta de candidato ao programa de pós-graduação com experiência sobre este tema.
+
+O objetivo é maximizar a eficiência global da companhia aérea, minimizando custos.
+
+Um modelo simplicado pode ser semelhante a este:
+
+\begin{equation} \label{eq:fo3}
+minimize \sum_{i \in M} \sum_{j \in N} c_{ij} \times X_{ij} + \sum_{i \in M} f_i \times z_i
+\end{equation}
+
+onde:
+
+$X_{ij}$ é a fração da demanda do cliente $j$ fornecida pela instalação $i$
+
+$z_i$ = 1 se a instalaçãoo estiver aberta, 0 caso contrário
+
+$c_{ij}$ é o custo do serviço e distribuiçãoo para fornecer aos clientes a partir da instalação $i$
+
+$f_i$ é o custo fixo de construir a instalação $i$
+
+$M$, $N$ são os conjuntos de possíveis locais para as instalações e a localização dos clientes.
 
@@
 A otimização do estoque de peças sobressalentes pode ajudar uma empresa a economizar dinheiro em custos de armazenamento, reduzir o desperdício e melhorar o atendimento ao cliente, tendo as peças necessárias prontamente disponíveis quando os clientes precisarem delas, seja oriunda do estoque ou de um fornecedor local.
+
+Assumindo que a demanda é conhecida a partir dos dados do ano anterior e que seria constante para uma temporada inteira, mas diferente para diferentes temporadas.
+
+Um modelo simples que representa essa situação é o da quantidade de pedido econômico ({\it economic order quantity} - EOQ), às vezes também é chamado de modelo econômico de tamanho de lote.
+
+O modelo EOQ é a quantidade do pedido que minimiza os custos totais de manutenção e os custos de pedido.
+
+Exemplo de Quantidade de Pedido Econômico:
+
+O EOQ considera o momento do novo pedido, o custo incorrido para fazer um pedido e os custos para armazenar mercadorias.
+Se uma empresa está constantemente fazendo pequenos pedidos para manter um nível de estoque específico, os custos de pedidos são maiores, juntamente com a necessidade de espaço de armazenamento adicional.
+
+\begin{equation} \label{eq:eoq}
+	EOQ \gets \sqrt{\frac{2 \times S \times D}{H}}
+\end{equation}
+
+Onde $S$ é o custo por pedido, $D$ a quantidade vendida por ano e $H$ o custo de manutenção do estoque.
+
+Por exemplo, considere um fornecedor de componentes eletronicos que oferece uma linha de fones aeronáuticos.
+Ele vende 1.000 fones por ano.
+Custa à empresa \$5 por ano para manter um único fone em estoque, e o custo fixo para fazer um pedido é de \$2.
+
+A fórmula EOQ é a raiz quadrada de (2 x 1.000 fones x custo do pedido de \$2) / (custo de manutenção de \$5), ou 28,3 com arredondamento.
+O tamanho ideal do pedido para minimizar custos e atender a demanda dos clientes é de pouco mais de 28 fones.
+
+https://www.investopedia.com/ask/answers/052715/how-economic-order-quantity-model-used-inventory-management.asp
 

### 2023-03-01 latex/PO_suportabilidade.tex
PO em suportabilidade

@@
 
-https://www.investopedia.com/ask/answers/052715/how-economic-order-quantity-model-used-inventory-management.asp
-
 \subsection{Simulação dos processos internos de suprimento e otimização de leiaute}
@@
 Os processos internos e a otimização de leiaute pode ser simulada e testada para determinar o melhor leiaute possível para um determinado espaço.
+
 A simulação de processos internos e otimização de leiaute podem ajudar a reduzir custos, melhorar o atendimento ao cliente do armazém e aumentar a eficiência da cadeia de suprimentos.
+
+Sob o ponto de vista da otimização, o objetivo seria minimizar os percursos de movimentação dos suprimentistas, conferindo agilidade e eficiência.
+Os componentes mais fornecidos tenderiam a ficar mais acessíveis e os pouco fornecidos em armários mais recuados e prateleiras mais altas.
+
+O modelo matemático geralmente utilizado é o de uma árvore geradora mínima ({\it minimum spanning tree}).
+
+Dada um grafo (G, $\phi$), pode-se definir o peso de um subgráfo $H \subset G$ como:
+
+\begin{equation} \label{eq:fo1}
+	\phi(H) \gets \sum_{e \in E(H)} \phi (e)
+\end{equation}
+
+$E(H)$ é o subconjunto de arestas $e$ de $H$.
+
+Em um grafo conectado $G$, uma árvore geradora mínima $T$ é uma árvore com valor mínimo.
+
+\begin{equation} \label{eq:fo2}
+	min_T\ H(T) \gets \sum_{e \in E(H)} \phi (e) \times X_{ij}
+\end{equation}
+
+Considere $X_{ij}$ igual a 1 se a aresta (i,j) está na árvore, 0 caso contrário. As arestas são os percursos entre o ponto de entrega das mercadorias e os armários ou prateleiras.
 
@@
 
-%sudo apt-get install myspell-pt-br
+As análises multicritério geralmente utilizam os métodos {\it analytic network process} (ANP) e {\it analytic hierarchy process} (AHP).
+
+O ANP é uma forma mais geral do AHP usado na análise de decisão multicritério.
+
+O AHP estrutura um problema de decisão em uma hierarquia com um objetivo, critérios de decisão e alternativas, enquanto o ANP o estrutura como uma rede.
+Ambos então usam um sistema de comparações pareadas para medir os pesos dos componentes da estrutura e, finalmente, classificar as alternativas na decisão.
+
+Por exemplo:
+
+A fim de se contratar um engenheiro, pode-se elencar duas expertises: experiência e habilidades em gestão.
+
+A experência pode ser técnica ou de vendas, e as habilidades de gestão podem ser: da produção, de equipes, de finanças e de manutenção.
+
+Supondo-se que haja três candidatos, compara-se pares de competências, pontuando-se de 0 a 9 para um candidato ou para o outro, sendo que, no final, uma operação entre matrizes define graus de acordo com todas as avaliações par-a-par realizadas por um ou por mais de um avaliador.
+
+Um exemplo prático pode ser encontrado em:
+
+\begin{itemize}
+\item tool.aerologlab.ita.br/mcdm2
+\end{itemize}
+
+No caso de uma seleção de fornecedores, será preciso definir o objetivo, os critérios e subcritérios de avaliação, bem como os fornecedores candidatos.
+
+Detalhes do funcionamento podem ser encontrados em:
+
+\begin{itemize}
+\item Saaty, Thomas L. (1996). Decision Making with Dependence and Feedback: The Analytic Network Process. Pittsburgh, Pennsylvania: RWS Publications. ISBN 0-9620317-9-8.
+
+\item Saaty, Thomas L.
+(2005).
+Theory and Applications of the Analytic Network Process: Decision Making with Benefits, Opportunities, Costs and Risks.
+Pittsburgh, Pennsylvania: RWS Publications.
+ISBN 1-888603-06-2.
+\end{itemize}
 
 \subsection{Estimativas de abastecimento por meio de aprendizado de máquina}
+
+{\it Machine Learning} (ML) – ou aprendizado da máquina – está ligado às máquinas que possuem capacidade de aprendizagem própria para realizar ações ou tomar decisões para atividades através de técnicas específicas, {\it Big Data} e inteligência artificial ({\it IA}).
+{\it Mitchell, Tom (1997).
+Machine Learning.
+New York: McGraw Hill.
+ISBN 0-07-042807-7.
+OCLC 36417892}
+
+O aprendizado pode ocorrer pela supervisão de um programador que faz a inserção de parâmetros no sistema, ou de forma interativa, no qual os algoritmos adaptam-se aos dados, permitindo cruzar informações que façam previsões baseadas na análise de dados, sem que tenham sido programadas especificamente para isso.
+
+{\bf Importância}: A importância do ML no controle de estoque é projetar a demanda com maior segurança e precisão, prevendo fornecimentos futuros sem grandes riscos e viabilizando pedidos de compras automaticamente.
+Além disso, o ML pode prever perdas e solucionar problemas que comprometem os resultados, como rupturas, custos elevados com carregamentos e baixo giro dos produtos em suas respectivas curvas.
+
+{\bf Vantagens}: A aprendizagem de máquina influencia as atividades mais complexas da logística e dos armazéns, proporcionando maior precisão e transparência nos dados.
+
+{\bf Desempenho da entrega}: A eficiência na gestão de entregas e os custos de frete podem ser aperfeiçoados com a utilização do ML.
+Com parâmetros e uma extensa base de dados, a escolha dos métodos de transporte é mais precisa, diminuindo os riscos para toda a cadeia de suprimentos.
+
+{\bf Previsão da demanda}: Com o aprendizado de máquina, a previsão da demanda se torna mais definida, auxiliando no planejamento estratégico de vendas e logística e permitindo que as operações trabalhem com baixos níveis de estoque.
+
+{\bf Visibilidade de toda a cadeia de suprimento}: O ML, juntamente à telemática (Internet das Coisas - loT), permite que toda a logística seja monitorada em tempo real, aumentando os resultados das entregas e a comunicação entre setores.
 

### 2023-03-01 latex/PO_suportabilidade.tex
Suportabilidade e PO
PO update

@@
 % !TeX spellcheck = pt_BR
-\documentclass{article}
-\usepackage[utf8]{inputenc}
-\usepackage[left]{lineno}
-\usepackage[T1]{fontenc}
-\usepackage{lmodern}
+\documentclass{amsart}
+
+\usepackage{graphicx}
+\graphicspath{ {.} }
+
+\setkeys{Gin}{width=0.8\linewidth,keepaspectratio=true}
+\usepackage{natbib}
+
+\usepackage{color}
+\usepackage{url}
+\usepackage{epsfig}
+
+\usepackage{algorithm,algpseudocode}
+
+\newcommand{\specialcell}[2][c]{%
+	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+
+\newcommand{\Comentar}[1]{\State {\cmmt{#1}}}
+\newcommand{\Break}{\State \textbf{break}}
+\renewcommand{\Return}{\State \textbf{return}~}
+\renewcommand{\algorithmicensure}{\textbf{Parameters:}}
+
+\usepackage{longtable}
+\newcommand{\elite}{\mathcal{E}}
+\usepackage{rotating}
+\usepackage{subfigure}
+\usepackage{lscape}
 \usepackage{multirow}
-\usepackage{lineno,hyperref}
-\modulolinenumbers[5]
-\usepackage[none]{hyphenat}
-\usepackage{url}
+\usepackage{adjustbox}
+\usepackage{comment}
+\usepackage{setspace}
+\usepackage{caption}
+\captionsetup{labelfont=bf}
+\captionsetup{skip=0pt} % vertical distance from the table or figure
+\usepackage[margin=0.9in]{geometry}
+\usepackage[table]{xcolor}
 \usepackage{booktabs}
-\usepackage{graphicx}
-\usepackage{mathtools}
-\usepackage{amssymb}
-\usepackage{amsthm}
-\usepackage{caption}
-\usepackage{subfigure}
 
@@
 	\author{Antonio Celio Pereira de Mesquita}
-	\date{ Abril 2023}
+	\date{ Abril 2023 \\ celio.mesquita@gp.ita.br}
 
@@
 	\centering
-	\includegraphics[width=0.5\textwidth]{Images/analise_dados.jpg}
+	\includegraphics[width=0.5\textwidth]{Images/analise_dados.jpg}\\
 	\small\textsuperscript{Origem: livre de direitos autorais em www.pexels.com}
@@
 	\centering
-	\includegraphics[width=0.5\textwidth]{Images/flight_delay.jpg}
+	\includegraphics[width=0.5\textwidth]{Images/flight_delay.jpg}\\
 	\small\textsuperscript{Origem: livre de direitos autorais em www.pixabay.com}
@@
 
+{\it Machine Learning} (ML) – ou aprendizado da máquina – está ligado às máquinas que possuem capacidade de aprendizagem própria para realizar ações ou tomar decisões para atividades através de técnicas específicas, {\it Big Data} e inteligência artificial ({\it IA}).
+{\it Mitchell, Tom (1997).
+Machine Learning.
+New York: McGraw Hill.
+ISBN 0-07-042807-7.
+OCLC 36417892}
+
+O aprendizado pode ocorrer pela supervisão de um programador que faz a inserção de parâmetros no sistema, ou de forma interativa, no qual os algoritmos adaptam-se aos dados, permitindo cruzar informações que façam previsões baseadas na análise de dados, sem que tenham sido programadas especificamente para isso.
+
 Um plano de manutenção resiliente com otimização e aprendizado de máquina é uma abordagem proativa para o planejamento de manutenção que usa análise preditiva e ciência de dados para identificar possíveis problemas e otimizar os cronogramas de manutenção.
+
 Esse tipo de plano usa algoritmos de aprendizado de máquina para analisar leituras de sensores, dados históricos e outras informações para prever quando o equipamento provavelmente precisará de manutenção e otimizar o tempo das tarefas de manutenção.
@@
 
+\vspace{0.5cm}
+
 \begin{minipage}[h]{.98\linewidth}
-	\begin{minipage}[t]{0.33\linewidth}
-		\begin{tabular}{l}\hline
+	\begin{minipage}[t]{0.25\linewidth}
+		\begin{tabular}{l}
 		Tempo de voo\\
@@
 	\hfill
-	\begin{minipage}[t]{0.53\linewidth}
+	\begin{minipage}[t]{0.72\linewidth}
 		\vspace{-2ex}
-		\includegraphics[width=1.0\linewidth]{Images/programacao_voo.jpg}
+		\includegraphics[width=0.8\linewidth]{Images/programacao_voo.jpg}\\
 		\small\textsuperscript{Origem: livre de direitos autorais em www.pexels.com}
@@
 
+\subsection{Estimativas de abastecimento por meio de aprendizado de máquina}
+
+{\it Machine Learning} (ML) – ou aprendizado da máquina – está ligado às máquinas que possuem capacidade de aprendizagem própria para realizar ações ou tomar decisões para atividades através de técnicas específicas, {\it Big Data} e inteligência artificial ({\it IA}).
+{\it Mitchell, Tom (1997).
+Machine Learning.
+New York: McGraw Hill.
+ISBN 0-07-042807-7.
+OCLC 36417892}
+
+O aprendizado pode ocorrer pela supervisão de um programador que faz a inserção de parâmetros no sistema, ou de forma interativa, no qual os algoritmos adaptam-se aos dados, permitindo cruzar informações que façam previsões baseadas na análise de dados, sem que tenham sido programadas especificamente para isso.
+
+{\bf Importância}: A importância do ML no controle de estoque é projetar a demanda com maior segurança e precisão, prevendo fornecimentos futuros sem grandes riscos e viabilizando pedidos de compras automaticamente.
+Além disso, o ML pode prever perdas e solucionar problemas que comprometem os resultados, como rupturas, custos elevados com carregamentos e baixo giro dos produtos em suas respectivas curvas.
+
+{\bf Vantagens}: A aprendizagem de máquina influencia as atividades mais complexas da logística e dos armazéns, proporcionando maior precisão e transparência nos dados.
+
+{\bf Desempenho da entrega}: A eficiência na gestão de entregas e os custos de frete podem ser aperfeiçoados com a utilização do ML.
+Com parâmetros e uma extensa base de dados, a escolha dos métodos de transporte é mais precisa, diminuindo os riscos para toda a cadeia de suprimentos.
+
+{\bf Previsão da demanda}: Com o aprendizado de máquina, a previsão da demanda se torna mais definida, auxiliando no planejamento estratégico de vendas e logística e permitindo que as operações trabalhem com baixos níveis de estoque.
+
+{\bf Visibilidade de toda a cadeia de suprimento}: O ML, juntamente à telemática (Internet das Coisas - loT), permite que toda a logística seja monitorada em tempo real, aumentando os resultados das entregas e a comunicação entre setores.
+
 \subsection{Simulação dos processos internos de suprimento e otimização de leiaute}
@@
 
-\subsection{Estimativas de abastecimento por meio de aprendizado de máquina}
-
-{\it Machine Learning} (ML) – ou aprendizado da máquina – está ligado às máquinas que possuem capacidade de aprendizagem própria para realizar ações ou tomar decisões para atividades através de técnicas específicas, {\it Big Data} e inteligência artificial ({\it IA}).
-{\it Mitchell, Tom (1997).
-Machine Learning.
-New York: McGraw Hill.
-ISBN 0-07-042807-7.
-OCLC 36417892}
-
-O aprendizado pode ocorrer pela supervisão de um programador que faz a inserção de parâmetros no sistema, ou de forma interativa, no qual os algoritmos adaptam-se aos dados, permitindo cruzar informações que façam previsões baseadas na análise de dados, sem que tenham sido programadas especificamente para isso.
-
-{\bf Importância}: A importância do ML no controle de estoque é projetar a demanda com maior segurança e precisão, prevendo fornecimentos futuros sem grandes riscos e viabilizando pedidos de compras automaticamente.
-Além disso, o ML pode prever perdas e solucionar problemas que comprometem os resultados, como rupturas, custos elevados com carregamentos e baixo giro dos produtos em suas respectivas curvas.
-
-{\bf Vantagens}: A aprendizagem de máquina influencia as atividades mais complexas da logística e dos armazéns, proporcionando maior precisão e transparência nos dados.
-
-{\bf Desempenho da entrega}: A eficiência na gestão de entregas e os custos de frete podem ser aperfeiçoados com a utilização do ML.
-Com parâmetros e uma extensa base de dados, a escolha dos métodos de transporte é mais precisa, diminuindo os riscos para toda a cadeia de suprimentos.
-
-{\bf Previsão da demanda}: Com o aprendizado de máquina, a previsão da demanda se torna mais definida, auxiliando no planejamento estratégico de vendas e logística e permitindo que as operações trabalhem com baixos níveis de estoque.
-
-{\bf Visibilidade de toda a cadeia de suprimento}: O ML, juntamente à telemática (Internet das Coisas - loT), permite que toda a logística seja monitorada em tempo real, aumentando os resultados das entregas e a comunicação entre setores.
-
 \subsection{problema de substituição de múltiplos equipamentos para modernização de meia vida}

### 2023-03-01 latex/PO_suportabilidade.tex
PO suportabilidade

@@
 
-\subsection{Como a pesquisa operacional pode ser aplicada à suportabilidade}
+\subsection{Como a pesquisa operacional pode ser aplicada à suportabilidade}\
 
@@
 
-\subsection{Plano de manutenção resiliente com otimização e aprendizado de máquina}
+\subsection{Plano de manutenção resiliente com otimização e aprendizado de máquina}\
 
@@
 
-\subsection{Problema de planejamento de voos e manutenção}
+\subsection{Problema de planejamento de voos e manutenção}\
 
@@
 
-\subsection{Problema de roteamento e manutenção de aeronaves para frotas compartilhadas, incluindo informações de prognóstico de saúde das aeronaves}
+\subsection{Problema de roteamento e manutenção de aeronaves para frotas compartilhadas, incluindo informações de prognóstico de saúde das aeronaves}\
 
@@
 
-\subsection{O problema da otimização do leiaute de uma instalação para manutenção de aeronaves}
+\subsection{O problema da otimização do leiaute de uma instalação para manutenção de aeronaves}\
 
@@
 
-\subsection{Otimização da localização de instalações de manutenção}
+\subsection{Otimização da localização de instalações de manutenção}\
 
@@
 
-\subsection{Otimização de estoques de peças sobressalentes}
+\subsection{Otimização de estoques de peças sobressalentes}\
 
@@
 
-\subsection{Estimativas de abastecimento por meio de aprendizado de máquina}
+\subsection{Estimativas de abastecimento por meio de aprendizado de máquina}\
 
@@
 
-\subsection{Simulação dos processos internos de suprimento e otimização de leiaute}
+\subsection{Simulação dos processos internos de suprimento e otimização de leiaute}\
 
@@
 
-\subsection{Otimização da localização de instalações de distribuição}
+\subsection{Otimização da localização de instalações de distribuição}\
 
@@
 
-\subsection{Análise multicritério na seleção de fornecedores}
+\subsection{Análise multicritério na seleção de fornecedores}\
 
@@
 
-\subsection{problema de substituição de múltiplos equipamentos para modernização de meia vida}
+\subsection{Problema de roteirização como fator impactante da suportabilidade}\
 
-\subsection{problema de gerenciamento de LRU, To summarize, we discovered two problems.
-First, spare parts supply pays little attention to optimizing transportation routes.
-Second, the existing transportation route optimization methods are seldom considered from the standpoint of supportability}
+O fornecimento de peças de reposição muitas vezes dá pouca atenção à otimização das rotas de transporte porque muitas vezes é visto como uma preocupação secundária e não como uma prioridade.
+Além disso, os métodos existentes de otimização de rotas de transporte nem sempre são considerados do ponto de vista da capacidade de suporte.
+Como resultado, muitas organizações simplesmente optam pela rota mais direta, que muitas vezes não leva em consideração outros fatores, como recursos disponíveis, custo e tempo.
 
-\subsection{problema de Suporte Logístico Integrado envolvendo frotas, spare parts supply}
+Adicionalmente, a complexidade do processo e a falta de recursos e expertise também podem contribuir para a falta de atenção dada à suportabilidade na otimização das rotas de transporte.
 
-\subsection{problema da gestão de obsolescência}
+Parte desse processo de distribuição pode ser terceirizado ou realizado por meio de frota própria, a depender da estratégia da organização.
 
-\section{Desenvolvimento da suportabilidade}
+De qualquer forma, faz-se mister que as demandas de cada cliente sejam atendidas, trecho a trecho, de possíveis rotas de coleta e entrega de componentes e equipamentos.
 
-\subsection{Análise multicritério dos elementos de suporte logístico}
+O problema, então, consiste em realizar rotas de coleta e entrega simultâneas com restrições de capacidade em volume,  peso e passeio do centro de gravidade do veículo (aeronave, caminhão ou navio).
+Além disso, restrições de destino, a fim de que componentes e equipamentos sejam entregues ao cliente correto.
 
-\subsection{Otimização e simulação da suportabilidade de um novo produto}
+\subsection{problema da gestão de obsolescência}\
 
-\subsection{Análise e simulação da degradação por questões de suportabilidade}
+O gerenciamento de obsolescência de componentes envolve identificar e gerenciar os riscos associados a componentes e peças obsoletos ou em vias de faltarem fornecedores.
+Isso inclui monitorar os  níveis de consumo e estoque de componentes e peças, prever a demanda futura, planejar fornecedores alternativos e desenvolver estratégias para substituir componentes obsoletos.
 
-\subsection{problema de gerenciamento dos desempenhos de suportabilidade}
+Destaca-se, ainda, que as organizações devam considerar o desenvolvimento de relacionamentos com fornecedores que possam ajudar a reduzir os riscos de obsolescência, seja por meio de engenharia reversa ou de modelos atualizados que superem as funcionalidades dos antigos.
 
-\subsection{problema de alocação de aeronaves à luz do desempenho individualizado em termos de confiabilidade de cada aeronave}
+\section{Desenvolvimento da suportabilidade}\
 
-\subsection{problema de imaturidade nos fatores nos fatores RAMS}
+\subsection{Otimização e simulação da suportabilidade de um novo produto}\
 
-\subsection{optimizing reliability, maintainability and supportability under performance based logistics}
+\subsection{Análise e simulação da degradação por questões de suportabilidade}\
 
-\subsection{Predict key Performance and Cost metrics for new system}
+\subsection{problema de gerenciamento dos desempenhos de suportabilidade}\
 
-\subsection{Calculate Supportability for legacy and new systems}
+\subsection{problema de alocação de aeronaves à luz do desempenho individualizado em termos de confiabilidade de cada aeronave}\
+
+\subsection{problema de imaturidade nos fatores nos fatores RAMS}\
+
+\subsection{Garantia da confiabilidade, da capacidade de manutenção e da capacidade de suporte sob logística baseada em desempenho}\
+
+\subsection{Previsão das principais métricas de desempenho e custo para um novo sistema}\
 

### 2023-03-02 latex/PO_suportabilidade.tex
Papers update

@@
 
-A capacidade de suporte de sistemas complexos refere-se à capacidade de um sistema tecnológico ou uma organização, de ser efetivamente mantido, atualizado, reparado e aprimorado ao longo do tempo.
-Um sistema complexo é aquele que é composto por muitas partes interconectadas que interagem de forma não linear, tornando difícil prever como o sistema se comportará em diferentes condições.
+A capacidade de suporte de sistemas complexos refere-se à capacidade de um sistema tecnológico ou uma organização, de ser efetivamente mantido, atualizado, reparado e aprimorado temporalmente.
+Um sistema complexo é aquele sendo composto por muitas partes interconectadas que interagem de forma não linear, dificultando prever como o sistema se comportará em diferentes condições.
 Porém, um sistema complexo pode ser subdividido, para fins de análise, em subsistemas modeláveis e otimizáveis.
 
-A capacidade de suporte envolve vários aspectos, incluindo a capacidade de diagnosticar problemas, a disponibilidade dos recursos necessários (como pessoal, ferramentas e peças de reposição), a eficiência dos processos de manutenção e reparo, a capacidade de fazer alterações no sistema sem causar interrupções e a capacidade de se adaptar a requisitos e ambientes em mudança.
+A capacidade de suporte envolve vários aspectos, incluindo a aptidão de diagnosticar problemas, a disponibilidade dos recursos necessários (como pessoal, ferramentas e peças de reposição), a eficiência dos processos de manutenção e reparo, a propriedade de fazer alterações no sistema sem causar interrupções e também de se adaptar a requisitos e ambientes em mudança.
 
@@
 
-\item Prever a probabilidade e o impacto dos problemas de suporte: ao criar modelos que simulam o comportamento do sistema e identificam possíveis gargalos ou pontos de falha, a PO pode ajudar a prever onde os problemas de suporte são mais prováveis de ocorrer e qual seria o impacto desses problemas.
+\item Prever a probabilidade e o impacto dos problemas de suporte: ao criar modelos que simulam o comportamento do sistema e identificam possíveis gargalos ou pontos de falha, a PO pode ajudar a prever onde os problemas de suporte são prováveis de ocorrer e qual seria o impacto dessas questões.
 
@@
 
-No geral, a PO fornece um poderoso conjunto de ferramentas e técnicas para mitigar problemas de capacidade de suporte em sistemas complexos e pode ajudar a garantir que esses sistemas permaneçam confiáveis ao longo do tempo.
+No geral, a PO fornece um poderoso conjunto de ferramentas e técnicas para mitigar problemas de capacidade de suporte em sistemas complexos e pode ajudar a garantir que esses sistemas permaneçam confiáveis.
 
@@
 
-O processo de tomada de decisão sobre internalizar ou não a manutenção de de 3º nível normalmente envolve a ponderação dos custos e benefícios associados a ambas as opções.
+O processo de tomada de decisão sobre internalizar ou não a manutenção de 3º nível normalmente envolve a ponderação dos custos e benefícios associados a ambas as opções.
 Internalizar a manutenção pode ser uma opção mais econômica, pois elimina a necessidade de contratar e pagar pessoal de manutenção externo.
@@
 Além disso, pode ser difícil obter peças e componentes com o mesmo custo de fornecedores de manutenção externos.
-Em última análise, a decisão deve ser baseada em uma modelagem e otimização dos custos, benefícios e riscos associados.
+No final das contas, a decisão deve ser baseada em uma modelagem e otimização dos custos, benefícios e riscos associados.
 Normalmente, esta análise é realizada pelo AeroLogLab por meio da plataforma OPUS 10 (www.systecongroup.com).
@@
 
-A análise de dados para diagnóstico ou prognóstico para planejamento de manutenção de componentes de aeronaves normalmente envolve a coleta e análise de uma variedade de fontes de dados, como registros de voo e manutenção, previsões meteorológicas, aeronaves e sistemas e outras fontes de informações relacionadas à aeronave.
+A análise de dados para diagnóstico ou prognóstico para planejamento de manutenção de componentes de aeronaves normalmente envolve a coleta e análise de uma variedade de fontes de dados, como registros de voo e manutenção, previsões meteorológicas, aeronaves, sistemas e outras fontes de informações relacionadas à aeronave.
 Esses dados são usados para identificar possíveis problemas que podem afetar o desempenho e a segurança da aeronave, bem como para avaliar a condição de componentes e sistemas.
 
-Esse tipo de análise é frequentemente usado para antecipar e planejar futuras tarefas de manutenção, como manutenção preventiva, reparos e substituições.
+Essa análise é frequentemente usada para antecipar e planejar futuras tarefas de manutenção, como manutenção preventiva, reparos e substituições.
 Além disso, a análise de dados pode ser usada para monitorar o desempenho de aeronaves e identificar possíveis tendências ou padrões na manutenção de aeronaves.
@@
 O custo indireto inclui os efeitos sobre a imagem da companhia aérea, a fidelidade dos passageiros e a reputação da companhia aérea.
-Além disso, a análise do impacto também pode incluir fatores como o número de atrasos e cancelamentos, a duração dos atrasos e o tipo de aeronave.
+Além disso, a análise do impacto também pode incluir fatores como o número de atrasos e cancelamentos, a duração dos atrasos e a categoria de aeronave.
 
@@
 
-O aprendizado pode ocorrer pela supervisão de um programador que faz a inserção de parâmetros no sistema, ou de forma interativa, no qual os algoritmos adaptam-se aos dados, permitindo cruzar informações que façam previsões baseadas na análise de dados, sem que tenham sido programadas especificamente para isso.
+O aprendizado pode ocorrer pela supervisão de um programador que faz a inserção de parâmetros no sistema, ou de forma interativa, onde os algoritmos adaptam-se aos dados, permitindo cruzar informações que façam previsões baseadas na análise de dados, sem terem sido programadas especificamente para isso.
 
-Um plano de manutenção resiliente com otimização e aprendizado de máquina é uma abordagem proativa para o planejamento de manutenção que usa análise preditiva e ciência de dados para identificar possíveis problemas e otimizar os cronogramas de manutenção.
+Um plano de manutenção resiliente com otimização e aprendizado de máquina é uma abordagem proativa que usa análise preditiva e ciência de dados a fim de identificar possíveis problemas e otimizar os cronogramas de manutenção.
 
-Esse tipo de plano usa algoritmos de aprendizado de máquina para analisar leituras de sensores, dados históricos e outras informações para prever quando o equipamento provavelmente precisará de manutenção e otimizar o tempo das tarefas de manutenção.
+Essa categoria de plano usa algoritmos de aprendizado de máquina para analisar leituras de sensores, dados históricos e outras informações para prever quando o equipamento provavelmente precisará de manutenção e otimizar o tempo das tarefas de manutenção.
 
@@
 As tarefas de preparação não podem ser duplicadas. As tarefas executam preparações comuns comente uma vez, a fim de não desperdiçar recursos.
-$K_i$ é o conjunto de preparações do pacote $i$. A restrição que as mantêm únicas é o tipo de dados $set$" da linguagem {\it Python}, que não permite elementos duplicados.
+$K_i$ é o conjunto de preparações do pacote $i$.
 
@@
 
-Do inglês {\it flight maintenance planning problem} (FMPP), podem planejados pelas companhias aéreas por meio da divisão em grupos de problemas elementares: Desenvolvimento das rotas, Desenho do cronograma, Designação de frota, Rotas de manutenção e Pareamento de tripulação.
+Do inglês {\it flight maintenance planning problem} (FMPP), podem planejados pelas companhias aéreas por meio da divisão em grupos de problemas elementares: desenvolvimento das rotas, desenho do cronograma, designação de frota, rotas de manutenção e Pareamento de tripulação.
 
-As abordagens que são usadas para resolver esses sub-problemas podem ser divididas em dois grupos básicos: abordagens que são usadas apenas para resolver um dos problemas mencionados (abordagens de propósito único) e procedimentos que resolvem mais de um problema básicos ao mesmo tempo (abordagens integradas).
+As abordagens usadas para resolver esses sub-problemas podem ser divididas em dois grupos básicos: abordagens usadas apenas para resolver um dos problemas mencionados (abordagens de propósito único) e procedimentos que resolvem mais de um problema básicos em simultâneo (abordagens integradas).
 
@@
 
-O roteamento de aeronave refere-se ao processo de determinação da rota ideal para uma aeronave viajar, levando em consideração fatores como consumo de combustível, condições climáticas e congestionamento do tráfego aéreo.
-Esse processo envolve a análise de grandes quantidades de dados, incluindo dados históricos de voos, previsões meteorológicas e informações de tráfego aéreo em tempo real.
+O roteamento de aeronave refere-se ao processo de determinação da rota ideal para uma aeronave viajar, considerando, fatores como consumo de combustível, condições climáticas e congestionamento do tráfego aéreo.
+Esse processo envolve a análise de abundantes de dados, incluindo dados históricos de voos, previsões meteorológicas e informações de tráfego aéreo em tempo real.
 O objetivo do roteamento de aeronaves é minimizar o tempo de voo, o consumo de combustível e os custos operacionais gerais, garantindo segurança e confiabilidade.
@@
 A otimização do leiaute de uma instalação para manutenção de aeronaves é uma parte importante do processo geral de manutenção de aeronaves.
-Este processo envolve a otimização do espaço dentro da instalação para garantir que todos os equipamentos e pessoal necessários estejam localizados nas posições mais eficientes para maximizar a produtividade e minimizar os custos.
+Este processo envolve a otimização do espaço na instalação para garantir que todos os equipamentos e pessoal necessários estejam localizados nas posições mais eficientes para maximizar a produtividade e minimizar os custos.
 
@@
 
-O leiaute ideal de uma instalação para manutenção de aeronaves também inclui levar em consideração a ergonomia e a segurança do pessoal, fornecendo espaço adequado para movimentação e equipamentos de segurança.
+O leiaute ideal de uma instalação para manutenção de aeronaves também inclui considerar a ergonomia e a segurança do pessoal, fornecendo espaço adequado para movimentação e equipamentos de segurança.
 
-Sob o ponto de vista da otimização, o objetivo seria minimizar os percursos de movimentação dos técnicos, conferindo agilidade e eficiência.
+Do ponto de vista da otimização, o objetivo seria minimizar os percursos de movimentação dos técnicos, conferindo agilidade e eficiência.
 As ferramentas mais utilizadas tenderiam a ficar mais acessíveis e as pouco utilizadas em posições mais recuadas.
@@
 O EOQ considera o momento do novo pedido, o custo incorrido para fazer um pedido e os custos para armazenar mercadorias.
-Se uma empresa está constantemente fazendo pequenos pedidos para manter um nível de estoque específico, os custos de pedidos são maiores, juntamente com a necessidade de espaço de armazenamento adicional.
+Se uma empresa está constantemente fazendo pequenos pedidos para manter um nível de estoque específico, os custos de pedidos são maiores, com a necessidade de espaço de armazenamento adicional.
 
@@
 
-Por exemplo, considere um fornecedor de componentes eletronicos que oferece uma linha de fones aeronáuticos.
+Por exemplo, considere um fornecedor de componentes eletrônicos que oferece uma linha de fones aeronáuticos.
 Ele vende 1.000 fones por ano.
@@
 
-O aprendizado pode ocorrer pela supervisão de um programador que faz a inserção de parâmetros no sistema, ou de forma interativa, no qual os algoritmos adaptam-se aos dados, permitindo cruzar informações que façam previsões baseadas na análise de dados, sem que tenham sido programadas especificamente para isso.
+O aprendizado pode ocorrer pela supervisão de um programador que faz a inserção de parâmetros no sistema, ou interativamente, no qual os algoritmos adaptam-se aos dados, permitindo cruzar informações que façam previsões baseadas na análise de dados, sem que tenham sido programadas especificamente para isso.
 
@@
 
-Sob o ponto de vista da otimização, o objetivo seria minimizar os percursos de movimentação dos suprimentistas, conferindo agilidade e eficiência.
+Do ponto de vista da otimização, o objetivo seria minimizar os percursos de movimentação dos suprimentistas, conferindo agilidade e eficiência.
 Os componentes mais fornecidos tenderiam a ficar mais acessíveis e os pouco fornecidos em armários mais recuados e prateleiras mais altas.
@@
 
-A fim de se contratar um engenheiro, pode-se elencar duas expertises: experiência e habilidades em gestão.
+A fim de se contratar um engenheiro, pode-se elencar duas competências: experiência e habilidades em gestão.
 
-A experência pode ser técnica ou de vendas, e as habilidades de gestão podem ser: da produção, de equipes, de finanças e de manutenção.
+A experiência pode ser técnica ou de vendas, e as habilidades de gestão podem ser: da produção, de equipes, de finanças e de manutenção.
 
@@
 Além disso, os métodos existentes de otimização de rotas de transporte nem sempre são considerados do ponto de vista da capacidade de suporte.
-Como resultado, muitas organizações simplesmente optam pela rota mais direta, que muitas vezes não leva em consideração outros fatores, como recursos disponíveis, custo e tempo.
+Como resultado, muitas organizações simplesmente optam pela rota mais direta, que muitas vezes não considera outros fatores, como recursos disponíveis, custo e tempo.
 
-Adicionalmente, a complexidade do processo e a falta de recursos e expertise também podem contribuir para a falta de atenção dada à suportabilidade na otimização das rotas de transporte.
+Adicionalmente, a complexidade do processo e a falta de recursos e competência também podem contribuir para a falta de atenção dada à suportabilidade na otimização das rotas de transporte.
 
@@
 
-Destaca-se, ainda, que as organizações devam considerar o desenvolvimento de relacionamentos com fornecedores que possam ajudar a reduzir os riscos de obsolescência, seja por meio de engenharia reversa ou de modelos atualizados que superem as funcionalidades dos antigos.
+Destaca-se, ainda, que as organizações devam considerar o desenvolvimento de relacionamentos com fornecedores que ajudem a reduzir os riscos de obsolescência, seja por meio de engenharia reversa ou de modelos atualizados que superem as funcionalidades dos antigos.
 

### 2023-03-02 latex/comp_or.tex
Papers update

--- 2023-02-27 latex/comp_or.tex
@@
 That is why all parameters were set using the $iRace$ package \citep{LopezIbanezManuel2016}.
-Supplying $iRace$ with a range for $volThreshold$ as [0.80 to 0.99], {\it iRace} yielded ....., by training on the uneven instances numbers.
+Supplying $iRace$ with a range for $volThreshold$ as [0.80 to 0.99], {\it iRace} yielded 0.92, by training on the uneven instances numbers.
+The values tested for 20\%, 50\%, and 100\% volume surpluses were very close, considering the confidence level offered by {\it iRace}.
 
@@
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 9.66        &  6.37       &  9.58       &  10.87      &   12.24     & 11.94        &  0.82\\%
+&\multirow{2}{*}{ $Shims$}                       & $f$                & 9.69        &  6.37       &  9.58       &  11.07      &   12.24     & 11.94        &  0.82\\%
 &                                                & {\it run-time (s)} &  < 1        &  3          &    5        &   11        &   27        & 342          &  5.3\\%389

### 2023-03-02 latex/PO_suportabilidade.tex
PO suportabilidade

@@
 
-\subsection{Como a pesquisa operacional pode ser aplicada à suportabilidade}\
+\subsection{Aplicações da Pesquisa Operacional no desenho da suportabilidade}\
 
 A pesquisa operacional (PO) é um campo de estudo que usa modelagem matemática, técnicas de otimização e métodos analíticos para resolver problemas complexos em negócios, engenharia e outros campos.
-O objetivo da pesquisa operacional é encontrar soluções ótimas para problemas que envolvem múltiplas variáveis, restrições e objetivos.
+O objetivo da PO é encontrar soluções ótimas ou muito próximas do ótimo em curto tempo para problemas que envolvem múltiplas variáveis, restrições e objetivos.
 
@@
 
-\subsection{problema da gestão de obsolescência}\
+\subsection{O problema da gestão de obsolescência}\
 

### 2023-03-02 latex/PO_suportabilidade.tex
PO update

@@
 % !TeX spellcheck = pt_BR
-\documentclass{amsart}
+
+\documentclass{abntex2}
+\usepackage{abntex2cite}
+
+\usepackage[brazil]{babel}
+%sudo apt-get install texlive-lang-portuguese
 
@@
 
-\setkeys{Gin}{width=0.8\linewidth,keepaspectratio=true}
+%\setkeys{Gin}{width=0.8\linewidth,keepaspectratio=true}
 \usepackage{natbib}
@@
 \usepackage{rotating}
-\usepackage{subfigure}
+%\usepackage{subfigure}
 \usepackage{lscape}
@@
 
+%\usepackage{hyperref}
+
 \begin{document}
@@
 	\title{Pesquisa Operacional em Suportabilidade}
-	\author{Antonio Celio Pereira de Mesquita}
-	\date{ Abril 2023 \\ celio.mesquita@gp.ita.br}
+	\author{Antonio Celio Pereira de Mesquita\\ AeroLogLab-ITA\\  \small{celio.mesquita@gp.ita.br} }
+	\date{} % deixado vazio de propósito a fim de excluir a data
 
 \maketitle
+
+\begin{sloppypar}
 
@@
 
-A capacidade de suporte de sistemas complexos refere-se à capacidade de um sistema tecnológico ou uma organização, de ser efetivamente mantido, atualizado, reparado e aprimorado temporalmente.
+A capacidade de suporte de sistemas complexos refere-se à aptidão de um sistema tecnológico ou uma organização, de ser efetivamente mantido, atualizado, reparado e aprimorado temporalmente.
 Um sistema complexo é aquele sendo composto por muitas partes interconectadas que interagem de forma não linear, dificultando prever como o sistema se comportará em diferentes condições.
@@
 
-Melhores detalhes poderão ser encontrados em {\it Marques, Henrique Costa; Fernando Teixeira Mendes Abrahão; Guilherme Conceição Rocha.
-Mapping Challenges Through the ILS Approach.
-DOI: 10.4018/978-1-7998-3904-0.ch012}
+Melhores detalhes poderão ser encontrados em \cite{MarquesAbrahaoGuilherme:2020}.
+
+\subsection{O problema da programação da manutenção preventiva}
+
+O problema de programação da manutenção preventiva (PPMP) pode ser definido como o problema relacionado com a programação da manutenção vivenciado por um sistema que possua uma frota de veículos (aviões, caminhões, maquinas, tratores, trens, etc.) sujeita a múltiplas restrições.
+O PPMP envolve dois tipos básicos de programação de atividades:
+
+\begin{itemize}
+
+\item[I] Programação das tarefas de manutenção nas oficinas, onde os problemas são similares aos problemas de sequenciamento de tarefas (do inglês \textit{scheduling}), que envolvem, tipicamente, a alocação de recursos limitados a tarefas ao longo do tempo.
+Neste caso, a solução do problema está em encontrar uma sequência para todas as fases necessárias a uma revisão (testes, inspeções, delineamento das atividades, desmontagem, limpeza, reparos, substituições, montagem, pintura, testes, etc.), dadas limitações de recursos das oficinas.
+
+\item[II] Programação dos recolhimentos das aeronaves para as atividades de manutenção preventiva necessárias, segundo as restrições verificadas no sistema.
+
+\end{itemize}
+
+A solução, neste segundo caso, pouco ou nada tem a ver com a programação do que acontece nas oficinas, mas sim, com a ordenação dos recolhimentos para as atividades de manutenção preventiva necessários a uma frota de aeronaves ao longo de um período previsto de seu ciclo de vida.
+Este foi o foco do trabalho de  \cite{Abrahao:2006}.
 
@@
 No final das contas, a decisão deve ser baseada em uma modelagem e otimização dos custos, benefícios e riscos associados.
-Normalmente, esta análise é realizada pelo AeroLogLab por meio da plataforma OPUS 10 (www.systecongroup.com).
+Normalmente, esta análise é realizada pelo AeroLogLab (\url{www.aerologlab.ita.br}) por meio da plataforma OPUS 10 (\url{www.systecongroup.com}).
 
@@
 
-Informações relacionadas poderão ser encontradas em {\it Imbassahy, Dennys Wallace Duncan; Henrique Costa Marques, Guilherme Conceição Rocha, and Alberto Martinetti.
-Empowering Predictive Maintenance: A Hybrid Method to Diagnose Abnormal Situations.
-Applied Sciences 10.19 DOI: 10.4018/978-1-7998-3904-0.ch012 1-27}
+Informações relacionadas poderão ser encontradas em \cite{DuncanMarquesGuilhermeMartinetti:2005}.
 
@@
 
-{\it Machine Learning} (ML) – ou aprendizado da máquina – está ligado às máquinas que possuem capacidade de aprendizagem própria para realizar ações ou tomar decisões para atividades através de técnicas específicas, {\it Big Data} e inteligência artificial ({\it IA}).
-{\it Mitchell, Tom (1997).
-Machine Learning.
-New York: McGraw Hill.
-ISBN 0-07-042807-7.
-OCLC 36417892}
+\textit{Machine Learning} (ML) – ou aprendizado da máquina – está ligado às máquinas que possuem capacidade de aprendizagem própria para realizar ações ou tomar decisões para atividades através de técnicas específicas, \textit{Big Data} e inteligência artificial (\textit{IA}), \cite{Mitchell:1997}.
 
@@
 
-Conhecido como um problema de programação ou {\it scheduling}, tarefas são agrupadas em pacotes temporais, algumas tarefas eventualmente poderão ser colocadas como {\it out-of-phase} (fora dos pacotes) e, dentro de cada pacote, as tarefas são agrupadas por zona, por qualificação do mecânico e por precedência entre tarefas.
+Conhecido como um problema de programação ou \textit{scheduling}, tarefas são agrupadas em pacotes temporais, algumas tarefas eventualmente poderão ser colocadas como \textit{out-of-phase} (fora dos pacotes) e, dentro de cada pacote, as tarefas são agrupadas por zona, por qualificação do mecânico e por precedência entre tarefas.
 O objetivo global é a eficiência traduzida como minimização de custos e do tempo de parada, bem como minimização dos custos de oportunidade.
@@
 
-Do inglês {\it flight maintenance planning problem} (FMPP), podem planejados pelas companhias aéreas por meio da divisão em grupos de problemas elementares: desenvolvimento das rotas, desenho do cronograma, designação de frota, rotas de manutenção e Pareamento de tripulação.
+Do inglês \textit{flight maintenance planning problem} (FMPP), podem planejados pelas companhias aéreas por meio da divisão em grupos de problemas elementares: desenvolvimento das rotas, desenho do cronograma, designação de frota, rotas de manutenção e Pareamento de tripulação.
 
@@
 		\vspace{-2ex}
-		\includegraphics[width=0.8\linewidth]{Images/programacao_voo.jpg}\\
+		\includegraphics[width=0.7\linewidth]{Images/programacao_voo.jpg}\\
 		\small\textsuperscript{Origem: livre de direitos autorais em www.pexels.com}
@@
 
-Para detalhes sobre este assunto consulte {\it Pereira Barreto, Eduardo Afonso.
-Modelling the aircraft maintenance routing problem for fractional fleets with the inclusion of prognostics and health monitoring information.
-São José dos Campos.
-Instituto Tecnológico de Aeronáutica (Aeronautical Institute of Technology).}
+Para detalhes sobre este assunto consulte \cite{Eduardo:2022a}.
 
@@
 Do ponto de vista da otimização, o objetivo seria minimizar os percursos de movimentação dos técnicos, conferindo agilidade e eficiência.
-As ferramentas mais utilizadas tenderiam a ficar mais acessíveis e as pouco utilizadas em posições mais recuadas.
+As ferramentas mais utilizadas tenderiam a ficar mais acessíveis e as pouco empregadas em posições mais recuadas.
 
-O modelo matemático geralmente utilizado é o de uma árvore geradora mínima ({\it minimum spanning tree}).
+O modelo matemático geralmente utilizado é o de uma árvore geradora mínima (\textit{ minimum spanning tree}).
 
@@
 
-Um modelo simplicado pode ser semelhante a este:
+Um modelo simplificado pode ser semelhante a este:
 
@@
 
-$z_i$ = 1 se a instalaçãoo estiver aberta, 0 caso contrário
+$z_i$ = 1 se a instalação estiver aberta, 0 caso contrário
 
-$c_{ij}$ é o custo do serviço e distribuiçãoo para fornecer aos clientes a partir da instalação $i$
+$c_{ij}$ é o custo do serviço e distribuição para fornecer aos clientes a partir da instalação $i$
 
@@
 
-Um modelo simples que representa essa situação é o da quantidade de pedido econômico ({\it economic order quantity} - EOQ), às vezes também é chamado de modelo econômico de tamanho de lote.
+Um modelo simples que representa essa situação é o da quantidade de pedido econômico (\textit{ economic order quantity} - EOQ), às vezes também é chamado de modelo econômico de tamanho de lote.
 
@@
 
-{\bf Importância}: A importância do ML no controle de estoque é projetar a demanda com maior segurança e precisão, prevendo fornecimentos futuros sem grandes riscos e viabilizando pedidos de compras automaticamente.
+\textbf{Importância}: A importância do ML no controle de estoque é projetar a demanda com maior segurança e precisão, prevendo fornecimentos futuros sem grandes riscos e viabilizando pedidos de compras automaticamente.
 Além disso, o ML pode prever perdas e solucionar problemas que comprometem os resultados, como rupturas, custos elevados com carregamentos e baixo giro dos produtos em suas respectivas curvas.
 
-{\bf Vantagens}: A aprendizagem de máquina influencia as atividades mais complexas da logística e dos armazéns, proporcionando maior precisão e transparência nos dados.
+\textbf{Vantagens}: A aprendizagem de máquina influencia as atividades mais complexas da logística e dos armazéns, proporcionando maior precisão e transparência nos dados.
 
-{\bf Desempenho da entrega}: A eficiência na gestão de entregas e os custos de frete podem ser aperfeiçoados com a utilização do ML.
+\textbf{Desempenho da entrega}: A eficiência na gestão de entregas e os custos de frete podem ser aperfeiçoados com a utilização do ML.
 Com parâmetros e uma extensa base de dados, a escolha dos métodos de transporte é mais precisa, diminuindo os riscos para toda a cadeia de suprimentos.
 
-{\bf Previsão da demanda}: Com o aprendizado de máquina, a previsão da demanda se torna mais definida, auxiliando no planejamento estratégico de vendas e logística e permitindo que as operações trabalhem com baixos níveis de estoque.
+\textbf{Previsão da demanda}: Com o aprendizado de máquina, a previsão da demanda se torna mais definida, auxiliando no planejamento estratégico de vendas e logística e permitindo que as operações trabalhem com baixos níveis de estoque.
 
-{\bf Visibilidade de toda a cadeia de suprimento}: O ML, juntamente à telemática (Internet das Coisas - loT), permite que toda a logística seja monitorada em tempo real, aumentando os resultados das entregas e a comunicação entre setores.
+\textbf{Visibilidade de toda a cadeia de suprimento}: O ML, juntamente à telemática (Internet das Coisas - loT), permite que toda a logística seja monitorada em tempo real, aumentando os resultados das entregas e a comunicação entre setores.
 
@@
 
-O modelo matemático geralmente utilizado é o de uma árvore geradora mínima ({\it minimum spanning tree}).
+O modelo matemático geralmente utilizado é o de uma árvore geradora mínima (\textit{minimum spanning tree}).
 
-Dada um grafo (G, $\phi$), pode-se definir o peso de um subgráfo $H \subset G$ como:
+Dada um grafo (G, $\phi$), pode-se definir o peso de um subgrafo $H \subset G$ como:
 
-\begin{equation} \label{eq:fo1}
+\begin{equation} \label{eq:fo4}
 	\phi(H) \gets \sum_{e \in E(H)} \phi (e)
@@
 
-Em um grafo conectado $G$, uma árvore geradora mínima $T$ é uma árvore com valor mínimo.
+Em um grafo conectado $G$, uma árvore geradora mínima $T$ é um grafo com valor mínimo.
 
-\begin{equation} \label{eq:fo2}
+\begin{equation} \label{eq:fo5}
 	min_T\ H(T) \gets \sum_{e \in E(H)} \phi (e) \times X_{ij}
@@
 A otimização da localização das instalações de distribuição é um aspecto importante do gerenciamento da cadeia de suprimentos.
-Ao localizar estrategicamente os centros de distribuição próximos ao ponto de demanda, as empresas podem reduzir os custos de transporte, reduzir os prazos de entrega e melhorar o atendimento ao cliente.
+Ao localizar estrategicamente os centros de distribuição próximos ao ponto de demanda, as empresas podem reduzir os custos de transporte, minimizar os prazos de entrega e melhorar o atendimento ao cliente.
 
-A otimização da localização das instalações de distribuição geralmente é feita usando uma combinação de modelos analíticos, como sistemas de informações geográficas (GIS), para identificar os locais mais adequados e, em seguida, usar algoritmos de otimização para escolher a localização ideal.
+A otimização da localização das instalações de distribuição geralmente é feita usando uma combinação de modelos analíticos, como sistemas de informações geográficas, para identificar os locais mais adequados e, em seguida, usar algoritmos de otimização para escolher a localização ideal.
 Esse processo pode ajudar a garantir que o centro de distribuição esteja localizado no local mais eficiente e econômico.
@@
 
-As análises multicritério geralmente utilizam os métodos {\it analytic network process} (ANP) e {\it analytic hierarchy process} (AHP).
+As análises multicritério geralmente utilizam os métodos \textit{analytic network process} (ANP) e \textit{analytic hierarchy process} (AHP).
 
@@
 
-Um exemplo prático pode ser encontrado em:
-
-\begin{itemize}
-\item tool.aerologlab.ita.br/mcdm2
-\end{itemize}
+Um exemplo prático pode ser encontrado em \url{tool.aerologlab.ita.br/mcdm2}.
 
 No caso de uma seleção de fornecedores, será preciso definir o objetivo, os critérios e subcritérios de avaliação, bem como os fornecedores candidatos.
-
-Detalhes do funcionamento podem ser encontrados em:
-
-\begin{itemize}
-\item Saaty, Thomas L. (1996). Decision Making with Dependence and Feedback: The Analytic Network Process. Pittsburgh, Pennsylvania: RWS Publications. ISBN 0-9620317-9-8.
-
-\item Saaty, Thomas L.
-(2005).
-Theory and Applications of the Analytic Network Process: Decision Making with Benefits, Opportunities, Costs and Risks.
-Pittsburgh, Pennsylvania: RWS Publications.
-ISBN 1-888603-06-2.
-\end{itemize}
+Detalhes do funcionamento podem ser encontrados em \cite{Saaty:1996} e \cite{Saaty:2005}.
 
@@
 
+%\bibliographystyle{elsarticle-harv}
+
+\bibliography{livro}
+
+\end{sloppypar}
+
 \end{document}

### 2023-03-02 latex/abnt.tex
PO update

\documentclass[a4paper,12pt]{article}
\usepackage{verbatim}
\usepackage[portuges]{babel}

\newcommand{\ii}{\'{\i}}
\newcommand{\ca}{\c{c}\~ao}
\newcommand{\co}{\c{c}\~oes}
\newcommand{\enf}{\em}
%\newcommand{\identacao}{\\[1mm] \hspace*{\parindent}}
\newcommand{\identacao}{\\[1mm] \hspace*{3cm}}
\hyphenation{pri-mei-ros}

\begin{document}

	\title{Estilo Bib\TeX\ com as Normas da ABNT}
	\author{Fernando Gon\c{c}alves Pilotto} 
	\date{11 de Abril de 2003}
	\maketitle

	O estilo \verb+abnt+ \'e uma modifica{\ca} do estilo \verb+unsrt+ que
	incorpora as normas da Associa{\ca} Brasileira de Normas T\'ecnicas (ABNT).

	Neste texto explicamos como usar o Bib\TeX\ e o estilo \verb+abnt+.
	Se houver algum problema, meu e-mail \'e
	\verb+pilotto@if.ufrgs.br+.

	\tableofcontents

	\newpage

	\section{Estilos de bibliografia}

	Um estilo de bibliografia \'e um conjunto de regras que determinam
	a ordena{\ca} e o modo como devem ser escritos os elementos 
	de uma refer\^encia. Por exemplo, uma cita{\ca} de artigo no estilo
	da {\it Physical Review} \'e escrita como
	\begin{center}
		R. P. Feynman, Physical Review 76 (1949) 749.
	\end{center}
	A mesma refer\^encia no estilo da ABNT \'e escrita na forma
	\begin{center}
		\parbox{0.8\textwidth}{
			FEYNMAN, R. P. The theory of positrons. {\it Physical Review}, \\
			New York, v. 76, n. 1, p. 749-759, May 1949.}
	\end{center}
	O estilo de bibliografia varia bastante com o tipo de documento que
	estamos produzindo.
	Cada peri\'odico tem o seu estilo, que em geral \'e bem 
	parecido com o da {\it Physical Review}. Editores de livros
	tamb\'em gostam de criar estilos pr\'oprios. Para documentos
	publicados no Brasil, o correto \'e usar as normas da ABNT.

	\section{O \LaTeX\ e a bibliografia}

	No \LaTeX\ as refer\^encias bibliogr\'aficas s\~ao formatadas 
	atrav\'es do ambiente \verb+thebibliography+.
	\'E bom enfatizar que o \LaTeX\ faz somente
	a {\it formata{\ca}} do texto: 
	isto \'e, determina as quebras de linha, o espa\c{c}o
	entre as palavras, o ajuste do texto com as margens, o espa\c{c}o 
	entre as linhas, etc. Para produzir a refer\^encia ao artigo
	de Feynman segundo o estilo da {\it Physical Review},
	devemos escrever
	\begin{verbatim}
		\bibitem{Feynman} R. P. Feynman, Physical Review 76 (1949) 749.
	\end{verbatim}
	Para produzir a mesma refer\^encia segundo o estilo da ABNT
	devemos escre\-ver
	\begin{verbatim}
		\bibitem{Feynman} FEYNMAN, R. P. The theory of positrons.
		{\it Physical Review}, \\ 
		New York, v. 76, n. 1, p. 749-759, May 1949.
	\end{verbatim}
	Vemos que \'e muito trabalhoso modificar o estilo bibliogr\'afico
	se estivermos trabalhando com o \LaTeX . Isto por que a mudan\c{c}a
	de estilo requer n\~ao somente uma formata{\ca} diferente do texto
	como tamb\'em a {\it manipula{\ca}} do texto. Por exemplo, o nome
	do autor no estilo da ABNT deve aparecer com o sobrenome em letras
	mai\'usculas e os primeiros nomes devem ser abreviados.

	\section{O que \'e o Bib\TeX\ }

	O Bib\TeX\ \'e uma complementa{\ca} do \LaTeX\ que permite um melhor
	gerenciamento de refer\^encias bibliogr\'aficas, possibilitando
	a manipula{\ca} do texto.
	Para us\'a-lo,
	os dados bibliogr\'aficos devem ser armazenados de forma padro\-ni\-zada
	em um arquivo separado com a termina{\ca} ``\verb+.bib+''.
	O Bib\TeX\ processar\'a este arquivo, produzindo as refer\^encias
	de acordo com o estilo biblio\-gr\'afico que escolhermos.

	A refer\^encia ao artigo de Feynman, por exemplo, pode ser escrita como
	\begin{verbatim}
		@ARTICLE{Feynman,
			author="Richard P. Feynman", 
			title="The theory of positrons",
			journal="Physical Review",
			address="New York",
			year=1949,
			month=may,
			volume=76,
			number=1,
			pages="749-759"}
	\end{verbatim}
	Aqui est\'a toda a informa{\ca} que precisamos para esta
	refer\^encia. Depois de escolhermos um estilo bibliogr\'afico,
	o Bib\TeX\ pode determinar quais as informa{\co} que ele
	vai utilizar (o t{\ii}tulo do artigo, por exemplo,  n\~ao \'e
	relevante no estilo da {\it Physical Review}) e como vai tratar
	cada uma delas. A forma final da refer\^encia ao artigo de Feynman
	segundo o estilo da ABNT pode ser vista na refer\^encia 
	\cite{Feynman}.

	O Bib\TeX\ pode fazer ainda mais. No estilo ABNT n\~ao \'e
	necess\'ario informar a cidade onde o peri\'odico foi
	publicado, pois isto \'e feito automaticamente
	(ver se{\ca} \ref{sec_abnt}).

	\section{Como usar o Bib\TeX\ }

	O usu\'ario do Bib\TeX\ tem apenas dois comandos \`a disposi{\ca}:
	o comando \verb+\bibliographystyle{+ {\it filename}\verb+}+, 
	que define o estilo bibliogr\'afico,
	e o comando \verb+\bibliography{+{\it filename1, filename2,}\verb+...}+, que
	indica em quais arquivos est\~ao os dados bibliogr\'aficos. Voc\^e pode
	ter mais de um arquivo: por exemplo, um arquivo com trabalhos publicados
	por voc\^e ou pelo seu grupo, e outro arquivo com trabalhos publicados
	por outras pessoas.

	\newpage

	Para usar o Bib\TeX{,} voc\^e deve incluir no final do arquivo 
	``\verb+.tex+'' as linhas
	\begin{verbatim}
		\bibliographystyle{abnt}
		\bibliography{mybibliography}
	\end{verbatim}
	sendo \verb+mybibliography.bib+
	o arquivo em formato Bib\TeX\ que cont\'em os dados de bibliografia.
	Lembre de colocar o arquivo \verb+abnt.bst+ em um caminho onde
	o \LaTeX\ encontre-o (por exemplo, o mesmo diret\'orio onde est\'a
	o arquivo ``\verb+.tex+''). A seguir, d\^e os comandos
	\begin{center}
		\begin{tabular}{l}
			\verb+latex +{\it filename}\verb+.tex+ \\
			\verb+bibtex +{\it filename} \\
			\verb+latex +{\it filename}\verb+.tex+ \\
			\verb+latex +{\it filename}\verb+.tex+
		\end{tabular}
	\end{center}
	Observe que somente as refer\^encias citadas no
	arquivo {\it filename}\verb+.tex+ aparecer\~ao no documento final
	({\it filename}\verb+.dvi+). Por exemplo, se o arquivo 
	\verb+mybibli+ \verb+ography.bib+ cont\'em 500 refer\^encias e somente
	12 foram citadas em {\it filename} \verb+.tex+, ent\~ao somente estas
	12 aparecer\~ao em {\it filename}\verb+.dvi+.

	\section{Como o Bib\TeX\ comunica-se com o \LaTeX }

	O Bib\TeX\ e o \LaTeX\ s\~ao dois programas separados.
	O funcionamento em conjunto acontece da seguinte maneira:
	\begin{enumerate}
		\item 
		Ao processar o arquivo {\it filename}\verb+.tex+,
		o \LaTeX\ ignora o comando \verb+\bibli+ \verb+ographystyle{...}+,
		que para ele n\~ao tem sentido. Pela presen\c{c}a do comando
		\verb+\bibliography{...}+, ele recebe a instru{\ca} de processar a
		bibliografia no arquivo {\it filename}\verb+.bbl+. 
		\item
		Ao ser rodado, o Bib\TeX\, procura no arquivo
		{\it filename}\verb+.tex+ pelos comandos \verb+\bibliographystyle{...}+
		e \verb+\bibliography{...}+. Um define o estilo bibliogr\'afico, o outro
		indica em qual arquivo est\~ao os dados bibliogr\'aficos.
		A seguir, o Bib\TeX\ busca no arquivo
		{\it filename}\verb+.aux+ as refer\^encias que foram citadas,
		e procura estas refer\^encias no(s) arquivo(s) ``\verb+.bib+'' que
		aparecem no comando \verb+\bibliography{...}+.
		Por fim, estas refer\^encias s\~ao formatadas de acordo com o estilo que
		voc\^e escolheu e escritas no arquivo {\it filename}\verb+.bbl+.
	\end{enumerate}

	Se voc\^e olhar o arquivo {\it filename}\verb+.bbl+, 
	vai ver que ele come\c{c}a com o comando \verb+\begin{thebibliography}+.
		Este \'e o ambiente de bibliografia usual do \LaTeX , e \'e
		o que voc\^e digitaria se n\~ao usasse o Bib\TeX .

		Agora podemos entender o porqu\^e da lista de comandos na se{\ca}
		anterior. O primeiro comando, \verb+latex +{\it filename}\verb+.tex+,
		gera o arquivo {\it filename}\verb+.aux+, que cont\'em as refer\^encias
		citadas. O segundo comando, \verb+bibtex +{\it filename},
		gera o arquivo {\it filename}\verb+.bbl+, que cont\'em o ambiente
		\verb+\begin{thebibliography}+. O terceiro e o quarto comando,
			\verb+latex +{\it filename}\verb+.tex+, relacionam as cita{\co}
			com as refer\^encias e colocam elas em ordem.

			\section{Algumas vantagens do estilo ABNT}
			\label{sec_abnt}

			Uma das principais vantagens do estilo \verb+abnt+ \'e que n\~ao
			precisamos informar a cidade onde os peri\'odicos foram publicados,
			isto \'e feito automaticamente. Por exemplo, as revistas 
			{\em Physical Review} foram publicadas em Nova York at\'e 1983, e depois de
			1984 foram publicadas em Woodbury. Voc\^e n\~ao precisa
			saber disso, pois o \verb+abnt+ sabe.

			Voc\^e tamb\'em n\~ao precisa digitar \verb+p.+, \verb+v.+, \verb+n.+,
			etc. antes do n\'umero de p\'aginas, do volume e do n\'umero
			do peri\'odico. Isto faz parte do estilo \verb+abnt+, e
			\'e feito automaticamente. Na se{\ca} \ref{entradas} temos as
			descri{\co} de todas as entradas e do que \'e feito
			automaticamente e do que deve ser feito manualmente.

			Outra vantagem \'e que voc\^e pode usar o mesmo arquivo
			de bibliografias
			\verb+mybibliography.bib+ para os seus artigos e para a sua 
			disserta{\ca} ou tese. Para os artigos, basta utilisar um outro
			estilo de bibliografia, e as refer\^encias aparecer\~ao na maneira
			``usual''.

			\section{Instru{\co} para o arquivo .bib}
			\label{entradas}

			O arquivo ``\verb+.bib+'' tem um formato padronizado.
			Isto \'e necess\'ario para que o Bib\TeX\ possa manipular
			v\'arios tipos de refer\^encias.

			O tipo de refer\^encia define o que se chama de {\it entrada}:
			para artigos em peri\'odicos, usa-se a entrada \verb+@Article+, para
			livros, usa-se a entrada \verb+@Book+, para teses de doutorado,
			usa-se a entrada \verb+@Phdthesis+, etc.

			Cada entrada possui {\it campos},
			como por exemplo \verb+author+, \verb+journal+, etc., onde podemos
			colocar os dados espec{\ii}ficos de cada refer\^encia.
			Alguns campos s\~ao obrigat\'orios, outros s\~ao opcionais.

			Leia com aten{\ca} a descri{\ca} de cada entrada. 
			Olhe as refer\^encias para ver como cada entrada \'e impressa.

			\subsection{Entradas}

			\subsection*{Article}

			A entrada {\enf Article} \'e usada para referenciar um artigo publicado
			em um peri\'odico.
			Um exemplo \'e
			\begin{verbatim}
				@ARTICLE{Nogami83,
					author={Y. Nogami and Akira Suzuky}, 
					title={Divergence Disease of the Pion-Baryon
						Interaction in Quark-Based Models},
					journal={Progress of Theoretical Physics},
					address={Kyoto},
					month=apr,
					year=1983,
					volume=69,
					number=4,
					pages={1184-1194}}
			\end{verbatim}
			Esta \'e a refer\^encia \cite{Nogami83}.
			Os campos obrigat\'orios s\~ao \verb+author+, \verb+title+, \verb+journal+, 
			\verb+address+, \verb+month+, \verb+year+, \verb+volume+, \verb+number+ e
			\verb+pages+. Se o peri\'odico citado est\'a inclu{\ii}do na lista
			da se{\ca} \ref{revistas}, o campo \verb+address+ n\~ao \'e
			necess\'ario, pois o preenchimento ser\'a autom\'atico.
			O campo opcional \'e \verb+note+.

			\subsection*{Book}

			A entrada {\enf Book} \'e usada para referenciar um livro.
			Um exemplo \'e
			\begin{verbatim}
				@BOOK{Muta87,
					author={Taizo Muta},
					title={Foundations of Quantum Chromodynamics},
					subtitle={An Introduction to Perturbative Methods
						in Gauge Theories},
					publisher={World Scientific},
					address={Singapore},
					year=1987}
			\end{verbatim}
			Esta \'e a refer\^encia \cite{Muta87}. Se o livro for
			uma colet\^anea de v\'arios textos, cita-se o nome dos editores, 
			como por exemplo em
			\begin{verbatim}
				@BOOK{Abramowitz,
					editor={M. ABRAMOWITZ and I. A. STEGUN},
					title={Handbook of mathematical functions},
					publisher={National Bureau of Standards},
					address={Washington},
					year=1964}
			\end{verbatim}
			Esta \'e a refer\^encia \cite{Abramowitz}.
			Os campos obrigat\'orios s\~ao \verb+author+ ou \verb+editor+,
			\verb+title+, \verb+publisher+, \verb+address+ e \verb+year+.
			Os campos opcionais s\~ao \verb+subtitle+, \verb+volume+, \verb+number+,
			\verb+series+, \verb+edition+ e \verb+note+.

			\begin{comment}
				\subsection*{Booklet}

				A entrada {\enf booklet} \'e usada para referenciar livros publicados
				sem aux{\ii}lio de editora.
				Sua forma geral \'e \cite{}
				\begin{verbatim}
				\end{verbatim}

				\subsection*{Conference}

				\'E o mesmo que a entrada {\enf Inproceedings}.
			\end{comment}

			\subsection*{Inbook}

			A entrada {\enf Inbook} \'e usada para referenciar uma parte de um livro,
			que pode ser um cap{\ii}tulo ou algumas p\'aginas. Ela \'e quase
			id\^entica \`a entrada {\enf Book}.
			\begin{verbatim}
				@BOOK{Muta87,
					author={Taizo Muta},
					title={Foundations of Quantum Chromodynamics},
					subtitle={An Introduction to Perturbative Methods
						in Gauge Theories},
					publisher={World Scientific},
					address={Singapore},
					year=1987,
					chapter=2}
			\end{verbatim}
			Esta \'e a refer\^encia \cite{Muta87in}.
			Os campos obrigat\'orios e opcionais s\~ao os mesmos da entrada
			{\enf Book}, exceto pela obrigatoriedade de um campo
			\verb+chapter+ ou \verb+pages+.

			\subsection*{Incollection}

			A entrada {\enf Incollection} \'e usada para referenciar 
			uma parte (em geral um cap{\ii}tulo) de uma colet\^anea.
			\begin{verbatim}
				@INCOLLECTION{Jaffe79,
					author={R. L. JAFFE},
					title={The bag},
					editor={A. Zichichi},
					booktitle={Pointlike structures inside and outside hadrons},
					publisher={Plenum Press},
					address={New York},
					year=1982,
					pages={99-146}}
			\end{verbatim}
			Esta \'e a refer\^encia \cite{Jaffe79}.
			Os campos obrigat\'orios s\~ao \verb+author+, \verb+title+, \verb+editor+,
			\verb+booktitle+, \verb+publisher+, \verb+address+, \verb+year+ e
			\verb+chapter+ ou \verb+pages+.
			Os campos opcionais s\~ao \verb+subtitle+, \verb+volume+, \verb+number+,
			\verb+series+, \verb+edition+ e \verb+note+.

			\subsection*{Inproceedings}

			A entrada {\enf Inproceedings} \'e usada para referenciar um artigo em um 
			proceedings de uma confer\^encia.
			\begin{verbatim}
				@INPROCEEDINGS{Inproceedings,
					author={H. Leutwyler},
					title={Principles of Chiral Perturbation Theory},
					conference={Workshop on Hadron Physics},
					cnumber=4,
					cyear=1994,
					caddress={Gramado},
					ctitle={Topics on the Structure and Interaction
						of Hadronic Systems},
					editor={Victoria E. Herscovitz and Cesar A. Z. Vasconcellos
						and Erasmo Ferreira},
					address={Singapore},
					publisher={World Scientific},
					year=1995,
					pages={1-46}}
			\end{verbatim}
			Esta \'e a refer\^encia \cite{Inproceedings}.

			\subsection*{Internet}

			A entrada {\enf Internet} \'e usada para referenciar artigos ou
			documentos dispon{\ii}\-veis na internet.
			\begin{verbatim}
				@INTERNET{Steane97,
					author={Andrew Steane},
					title={Quantum Computing},
					internetaddress={quant-ph/9708022 v2},
					day=24,
					month=sep,
					year=1997}
			\end{verbatim}
			Esta \'e a refer\^encia \cite{Steane97}.
			Os campos obrigat\'orios s\~ao \verb+author+,
			\verb+title+, \verb+interne+\-\verb+taddress+, 
			\verb+day+, \verb+month+ e \verb+year+.
			O campo opcional \'e \verb+note+.

			\begin{comment}
				\subsection*{Manual}

				A entrada {\enf Manual} \'e usada para referenciar documenta{\ca}
				t\'ecnica.
				Sua forma geral \'e \cite{}
				\begin{verbatim}
				\end{verbatim}
			\end{comment}

			\subsection*{Mastersthesis}

			A entrada {\enf Mastersthesis} \'e usada para referenciar uma
			disserta{\ca} de mes\-tra\-do.
			\begin{verbatim}
				@MASTERSTHESIS{Master,
					author={Jo{\~a}o Ningu\'em},
					title={Como Passar o Tempo Durante Dois Anos},
					address={Porto Alegre},
					school={Universidade Federal do Rio Grande do Sul,
						Instituto de F\'{\i}sica,
						Curso de P\'os-Gradua\c{c}\~ao em F\'{\i}sica},
					year=1995}
			\end{verbatim}
			Esta \'e a refer\^encia \cite{Master}.
			Os campos obrigat\'orios s\~ao \verb+author+,
			\verb+title+, \verb+address+, \verb+school+ e \verb+year+.
			O campo opcional \'e \verb+note+.

			\begin{comment}
				\subsection*{Misc}

				A entrada {\enf Misc} \'e usada para referenciar algo que n\~ao se 
				enquadre nos moldes das outras entradas.
				Sua forma geral \'e \cite{Misc}
				\begin{verbatim}
				\end{verbatim}
			\end{comment}

			\subsection*{Phdthesis}

			A entrada {\enf Phdthesis} \'e usada para referenciar teses de doutorado.
			\begin{verbatim}
				@PHDTHESIS{Phd,
					author={Ningu{\'e}m~da~Silva, Jo{\~a}o},
					title={Como Passar o Tempo Durante Quatro Anos},
					address={Porto Alegre},
					school={Universidade Federal do Rio Grande do Sul,
						Instituto de F\'{\i}sica,
						Curso de P\'os-Gradua\c{c}\~ao em F\'{\i}sica},
					year=1999}
			\end{verbatim}
			Esta \'e a refer\^encia \cite{Phd}.
			Os campos obrigat\'orios s\~ao \verb+author+,
			\verb+title+, \verb+address+, \verb+school+ e \verb+year+.
			O campo opcional \'e \verb+note+.

			\subsection*{Proceedings}

			A entrada {\enf Proceedings} \'e usada para referenciar os proceedings de uma 
			confer\^encia.
			Sua forma geral \'e \cite{Proceedings}
			\begin{verbatim}
				@PROCEEDINGS{Proceedings,
					conference={Workshop on Hadron Physics},
					cnumber=4,
					cyear=1994,
					caddress={Gramado},
					title={Topics on the Structure and Interaction
						of Hadronic Systems},
					editor={Victoria E. Herscovitz and Cesar A. Z. Vasconcellos
						and Erasmo Ferreira},
					address={Singapore},
					publisher={World Scientific},
					year=1995}
			\end{verbatim}

			\begin{comment}
				\subsection*{Techreport}

				A entrada {\enf Techreport} \'e usada para referenciar um artigo
				publicado por uma institui{\ca}.
				Sua forma geral \'e \cite{Techreport}
				\begin{verbatim}
				\end{verbatim}

				\subsection*{Unpublished}

				A entrada {\enf Unpublished} \'e usada para referenciar um artigo
				ainda n\~ao publicado.
				Sua forma geral \'e \cite{Unpublished}
				\begin{verbatim}
				\end{verbatim}
			\end{comment}

			\subsection{Campos}

			O campo ``\verb+author+'' \'e
			o mais complicado. Os autores devem ser separados por
			``\verb+and+''. O primeiro nome de cada autor pode ser escrito por extenso
			ou abreviado. O sobrenome pode ser escrito na forma usual, com
			a primeira letra em mai\'usculo e as outras em min\'usculo.
			Nomes que incluam ``Jr.'' no final devem ser escritos entre chaves, como
			``\verb+C. G. {CALLAN Jr.}+''. 
			Colocando ``\verb+and OTHERS+'' produzir\'a
			a substitui{\ca} dos nomes seguintes por ``et. al''.
			Olhe os exemplos de entradas para mais possibilidades.

			O campo \verb+title+ \'e formatado automaticamente com a primeira 
			letra da primeira palavra em mai\'usculo, todas as outras em min\'usculo. 
			Para que nomes pr\'oprios sejam formatados
			corretamente, deve-se proteger a letra mai\'uscula usando
			chaves, como por exemplo em
			\begin{center}
				\verb+title={Monopolos de {D}irac}+
			\end{center}

			O campo \verb+journal+ aceita as abrevia{\co} listadas na
			se{\ca} \ref{revistas}.

			O campo \verb+month+ aceita as abrevia{\co} listadas na
			se{\ca} \ref{sec_mes}.

			\section{Usando outros estilos}

			O arquivo \verb+mybibliography.bib+, o qual cont\'em os dados 
			bibliogr\'aficos, pode ser usado com outros estilos de bibliografia
			sem gerar incompatibilidades. A \'unica inconveni\^encia \'e a
			refer\^encia \`as p\'aginas de um artigo. Usualmente coloca-se somente
			a p\'agina inicial, mas de acordo com as regras da ABNT, a p\'agina
			final tamb\'em deve ser referenciada.

			\section{Abrevia{\co} de revistas}
			\label{revistas}

			O campo \verb+journal+ aceita as abrevia{\co}
			\identacao
			{\verb+actapola+} = {Acta Physica Polonica A}
			\identacao
			{\verb+actapolb+} = {Acta Physica Polonica B}
			\identacao
			{\verb+advnuclp+} = {Advances in Nuclear Physics}
			\identacao
			{\verb+amjp+} = {American Journal of Physics}
			\identacao
			{\verb+annphys+} = {Annals of Physics}
			\identacao
			{\verb+arnpc+} = {Annual Review of Nuclear and Particle Science}
			\identacao
			{\verb+canjphys+} = {Canadian Journal of Physics}
			\identacao
			{\verb+commnuclp+} = {Comments on Nuclear and Particle Physics}
			\identacao
			{\verb+npa+} = {Nuclear Physics A}
			\identacao
			{\verb+npb+} = {Nuclear Physics B}
			\identacao
			{\verb+npc+} = {Nuclear Physics C}
			\identacao
			{\verb+npd+} = {Nuclear Physics D}
			\identacao
			{\verb+pla+} = {Physics Letters A}
			\identacao
			{\verb+plb+} = {Physics Letters B}
			\identacao
			{\verb+physrep+} = {Physics Reports}
			\identacao
			{\verb+pr+} = {Physical Review}
			\identacao
			{\verb+pra+} = {Physical Review A}
			\identacao
			{\verb+prb+} = {Physical Review B}
			\identacao
			{\verb+prc+} = {Physical Review C}
			\identacao
			{\verb+prd+} = {Physical Review D}
			\identacao
			{\verb+pre+} = {Physical Review E}
			\identacao
			{\verb+progtheophys+} = {Progress of Theoretical Physics}
			\begin{comment}
				\identacao
				{\verb+zfpa+} = {Zeitschrift f\"ur Physik A}
				\identacao
				{\verb+zfpb+} = {Zeitschrift f\"ur Physik B}
				\identacao
				{\verb+zfpc+} = {Zeitschrift f\"ur Physik C}
			\end{comment}

			\section{Abrevia{\co} dos meses}
			\label{sec_mes}

			O campo \verb+month+ aceita as abrevia{\co} para os meses em ingl\^es:
			\begin{center}
				\begin{tabular}{l@{\hspace{1.2cm}}l}
					\verb+jan = "Jan."+  & \verb+jul = "July"+ \\
					\verb+feb = "Feb."+  & \verb+aug = "Aug."+ \\
					\verb+mar = "Mar."+  & \verb+sep = "Sept."+ \\
					\verb+apr = "Apr."+  & \verb+oct = "Oct."+ \\
					\verb+may = "May"+   & \verb+nov = "Nov."+ \\
					\verb+jun = "June"+  & \verb+dec = "Dec."+
				\end{tabular}
			\end{center}
			Para revistas em outras l{\ii}nguas, deve-se usar a abrevia{\ca}
			correspondente.

			%\appendix
			%\addcontentsline{toc}{chapter}{REFER\^ENCIAS}

			%\bibliographystyle{unsrt}
			\bibliographystyle{abnt}
			\bibliography{abntdoc}
			%\bibliography{periodicos_extenso,abntdoc}
			%\bibliography{macros,abntdoc}

		\end{document}

### 2023-03-03 latex/comp_or.tex
Paper update

@@
 	\renewcommand{\arraystretch}{1.1} % alturas das linhas \specialcell{{\bf Parallel}\\{\bf heuristics}}
-	\begin{tabular}{r|cc|cc|cc|ccc|c}
+	\begin{tabular}{r|ccc|cc|cc|ccc|c}
 		\toprule
-		                            & {\bf APP}  & {\bf WBP}  &  {\bf SPDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} & \specialcell{{\bf Parallel}\\{\bf heuristics}}\\
-		\midrule
-		\cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$&  $.$ \\
-		\cite{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$&  $.$ \\
-		\cite{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ &  $.$\\
-		\cite{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$&  $.$ \\
-		\cite{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ &  $.$\\
-		\cite{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$&  $.$ \\	
-		\cite{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$&  $.$ \\
-		\cite{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$&  $.$ \\
-		\cite{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$&  $.$ \\
-		\cite{MesquitaCunha2011}   & $.$ & $.$    & $\bigstar$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ &  $.$\\		
-		\cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$&  $.$ \\
-
-		\cite{kaspi2019} & $.$ & $.$  & $.$ & $\bigstar$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$&  $.$ \\
-
-		\cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$&  $.$ \\
-		\cite{Vancroonemburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$&  $.$ \\
-		\cite{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$&  $.$ \\
-		\cite{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$&  $.$ \\
-		\cite{PaquaySchynsLimbourg2016,PaquayLimbourgSchynsOliveira2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$&  $.$ \\
-		\cite{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$&  $.$ \\
-		\cite{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$&  $.$  \\
-		\cite{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$&  $.$  \\
-		\cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$&  $.$ \\
-
-		\cite{MesquitaSanches2023}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$&  $.$  \\
-
-		{\bf This work}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$&  $\bigstar$  \\
+		                            & {\bf APP}  & {\bf WBP}   & {\bf MLU} &{\bf SPDP}&{\bf TSP} & {\bf 2D} & {\bf 3D} & {\bf Heu} & {\bf Int}  & {\bf Lin} & \specialcell{{\bf Parallel}\\{\bf heuristics}}\\
+		\midrule
+		\cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $\bigstar$& $.$        &  $.$       &  $.$ \\
+		\cite{Brosh1981}            & $.$        & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $.$       & $.$        &  $\bigstar$&  $.$ \\
+		\cite{Kevin1992}            & $.$        & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $.$       & $\bigstar$ &  $.$       &  $.$\\
+		\cite{Heidelberg1998}       & $.$        & $\bigstar$  & $.$       & $.$      & $.$      &$\bigstar$& $.$      & $\bigstar$& $.$        &  $.$       &  $.$ \\
+		\cite{MongeauBes2003}       & $\bigstar$ & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $.$       & $\bigstar$ &  $.$       &  $.$\\
+		\cite{fok2004optimizing}    & $.$        & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $.$       & $\bigstar$ &  $.$       &  $.$ \\	
+		\cite{Chan2006}             & $\bigstar$ & $.$         & $.$       & $.$      & $.$      & $.$      &$\bigstar$& $\bigstar$& $.$        &  $.$       &  $.$ \\
+		\cite{KaluznyBohdanL2009Oalb}& $.$       & $\bigstar$  & $.$       & $.$      & $.$      &$\bigstar$& $.$      & $.$       & $\bigstar$ &  $.$       &  $.$ \\
+		\cite{Verstichel2011}       & $.$        & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $.$       & $\bigstar$ &  $.$       &  $.$ \\
+		\cite{MesquitaCunha2011}    & $.$        & $.$         & $.$       &$\bigstar$& $.$      & $.$      & $.$      & $\bigstar$& $.$        &  $.$       &  $.$ \\		
+		\cite{Limbourg2012}         & $.$        & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $.$       & $\bigstar$ &  $.$       &  $.$ \\
+
+		\cite{kaspi2019}            & $.$        & $.$         & $.$       & $.$      &$\bigstar$& $.$      & $.$      &$\bigstar$ & $.$        &  $.$       &  $.$ \\
+
+		\cite{RoesenerHall2014}     & $\bigstar$ & $\bigstar$  & $.$       & $.$      & $.$      & $.$      &$\bigstar$& $.$       & $\bigstar$ &  $.$       &  $.$ \\
+		\cite{Vancroonemburg2014}   & $\bigstar$ & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $.$       & $\bigstar$ &  $.$       &  $.$ \\
+		\cite{LurkinSchyns2015}     & $.$        & $\bigstar$  & $\bigstar$&$\bigstar$& $.$      & $.$      & $.$      & $.$       & $\bigstar$ &  $.$       &  $.$ \\
+		\cite{RoesenerBarnes2016}   & $.$        & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $\bigstar$& $.$        &  $.$       &  $.$ \\
+
+		\cite{PaquaySchynsLimbourg2016,PaquayLimbourgSchynsOliveira2018}&$\bigstar$&$.$&$.$&$.$&$.$&$.$     &$\bigstar$&$\bigstar$ & $\bigstar$&$\bigstar$  &  $.$ \\
+
+		\cite{YangLiuGao2018}       & $.$        & $\bigstar$  & $.$       & $.$      & $.$      &$\bigstar$& $.$      & $\bigstar$ & $.$       & $.$        &  $.$ \\
+		\cite{wong2020}             & $\bigstar$ & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $.$        & $\bigstar$& $.$        &  $.$  \\
+		\cite{eugene2021}           & $\bigstar$ & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $.$        & $\bigstar$& $.$        &  $.$  \\
+		\cite{zhao2021}             & $.$        & $\bigstar$  & $.$       & $.$      & $.$      & $.$      & $.$      & $.$        & $\bigstar$&  $.$       &  $.$ \\
+
+		\cite{MesquitaSanches2023}  & $\bigstar$ & $\bigstar$  & $.$       &$\bigstar$&$\bigstar$& $.$      & $.$      & $\bigstar$ & $\bigstar$&  $.$       &  $.$  \\
+
+		{\bf This work}             & $\bigstar$ & $\bigstar$  &$\bigstar$ &$\bigstar$&$\bigstar$& $.$      & $.$      & $\bigstar$ & $\bigstar$&  $.$       &  $\bigstar$  \\
 		\bottomrule 
@@
 Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a pre-defined flight plan restricted to two legs.
-It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modelled through a container sequencing problem.
+It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading (MLU) costs at the intermediate node was modelled through a container sequencing problem.
 Referring directly to this work, \cite[p.

### 2023-03-05 latex/comp_or.tex
Linear relaxation

@@
 
-We converted Shims into a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it surpassed all Shims results and some of the integer programming results, all methods under a time limit to guarantee that a distribution plan is yielded in less than one hour.
+We converted Shims into a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it surpassed all Shims results.
 
@@
 		& The total weight must be less than the aircraft payload or the total pallet capacity, whichever is the lowest.\\	
-		& The pallets destined for the next node of the tour should be put as near as possible to the cargo ramp door (our inclusion).\\		
+		& {\bf The pallets destined for the next node of the tour should be put as near as possible to the cargo ramp door (our inclusion)}.\\		
 		\midrule
@@
 			\State $G_2 \gets SolveNode(method, \pi(k), G_1, timeLim)$ \label{tour:node}
-			\State $s, \epsilon \gets ScoreAndDeviation(\pi(k), G_2)$ \label{tour:analyse}
+
+{\color{blue} \State	$G_3 \gets minRampDist(G_2)$  \label{tour:minRampDist} }
+
+			\State $s, \epsilon \gets ScoreAndDeviation(\pi(k), G_3)$ \label{tour:analyse}
 			\State $score \gets score + s$ \label{tour:score2}
@@
 The destinations of each pallet are defined proportionally to the volume of items, regarding the pallets with consolidated items.
+
+In line \ref{tour:minRampDist}, $minRampDist$ minimizes the distance of the next node destined pallets, keeping the CG in its operational range.
+This precedure is implemented in integer programming as presented in Subsection \ref{improvement}, according to equations \ref{eq:distobj} to \ref{eq:disttorque}.
 
@@
 		{\it Data structure:}     & Items, pallets and edges are arrays of \emph{Python} class objects, which are initialized with the same data loaded from text files. \\
-
-		{\it Stopping criterion:} & All methods have the same time limit of $1.8$ seconds per node.\\
 
@@
 
-\multirow{7}{*}{$1.2$}&\multirow{2}{*}{ $Gurobi$}& $f$                & {\bf 12.00} & {\bf 8.38}  & {\bf 12.19} & {\bf 13.38} & {\bf 14.42} & 13.25        & {\bf 0.99} \\% 7412
-&                                                & {\it run-time (s)} &  8          &   14        &  25         &  95         &    217      &  1719        &  1.0       \\%2078
+\multirow{7}{*}{$1.2$}&\multirow{1}{*}{ $Gurobi$}& $f$                &{\bf 11.22}   &{\bf 7.88}   & 10.25      & 11.02       & 11.88        & 12.92       & 0.90 \\% 6484    7222
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 9.69        &  6.37       &  9.58       &  11.07      &   12.24     & 11.94        &  0.82\\%
-&                                                & {\it run-time (s)} &  < 1        &  3          &    5        &   11        &   27        & 342          &  5.3\\%389
+&\multirow{2}{*}{ $Shims$}                       & $f$                & 9.66         &  6.38       &  9.55       &  11.12      &  12.37      &   12.03      & 0.85 \\%6111
+&                                                & {\it run-time (s)} &  < 1         &  5          &    11       &   43        &  128        &  835        &  1.0 \\%1023
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & 11.90       &  7.61       &  10.86      & 13.13       & 13.75       & {\bf 13.75}  &  0.96\\%
-&                                                & {\it run-time (s)} &  < 1        &  < 1        &    3        &  9          &  26         & 156          &  {\bf 10.6}\\%
-
-\midrule
-
-\multirow{7}{*}{$1.5$}&\multirow{2}{*}{ $Gurobi$}& $f$                & 16.49       & {\bf 11.60} & {\bf 16.68} & {\bf 18.33} & {\bf 20.10} & 17.94        & {\bf 0.99}\\%10209
-&                                                & {\it run-time (s)} &  12         &   25        &  45         &  169        &   437       &  2377        & 1.0\\%3065
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & 11.03        & 7.43        &{\bf 12.58}  &{\bf 13.08}  &{\bf 14.02}  & {\bf 13.44}  &  {\bf 0.99}\\%7158
+
+&                                                & {\it run-time (s)} &  < 1         &  2          &    4        &  15         &  47         & 295          &  {\bf 2.8}\\%364
+
+\midrule
+
+\multirow{7}{*}{$1.2$}&\multirow{1}{*}{ $Gurobi$}& $f$                &{\it 0}   &{\it 0}   &{\it 0}  &{\it 0}  &{\it 0}   &{\it 0}  & {\it 1.00} \\% 6607
 
@@
 \midrule
-\multirow{7}{*}{$2.0$}&\multirow{2}{*}{ $Gurobi$}& $f$                & {\bf 24.04} & {\bf 17.33} & {\bf 24.62} & {\bf 26.36} & {\bf 28.88} & {\bf 25.89}  & {\bf 1.00}\\%14712
-&                                                & {\it run-time (s)} &  14         &   34        &  62         &  236        &  609        & 3431         & 1.0\\%4836
+\multirow{7}{*}{$1.2$}&\multirow{1}{*}{ $Gurobi$}& $f$                &{\it 0}   &{\it 0}   &{\it 0}  &{\it 0}  &{\it 0}   &{\it 0}  & {\it 1.00} \\% 6607
 
@@
 
-For the MIP solver in some scenarios, less than 50\% of the results were optimal, and the remaining were suboptimal due to the time limit.
-This explains that {\it mpShims} surpassed some of the MIP solver results.
-
-In \cite{MesquitaSanches2023}, {\it Shims} surpassed the MIP solver, on average, because the time limit they used (0.7s) was sufficient to guarantee a worst-case solution in less than 30min.
-This decision caused most of the MIP solver results to be poor, making {\it Shims} outstanding as it was the unique method to perform well with all scenarios and instances.
-
-In this work, we used as a time limit the worst {\it Shims} run-time, which was 1.8s.
-This made the MIP solver yield optimal or near-optimal solutions.
-This is good for the reader to have an idea of the real performance differences among the methods.
-
-On average, {\it mpShims} is 14 times faster than the MIP solver, and it reached 96\% of {\it Gurobi} $f$ values, on average, which is good for a tool of operational use.
-The use of parallelism is another important contribution of this work.
-It has improved {\it Shims} results and also its speed.
+The upper bound is the objective function value of an LP relaxation of the problem, proving that no integer feasible solution of larger objective value than this value can exist.
+
+In an attempt to find the upper bound for $f$, we solved all scenarios and instances with {\it Gurobi} in a linear relaxed model.
+We applied a time limit of 60s in the expectation that it would be enough for {\it Gurobi} to determine the upper bounds.
+This is good for the reader to have an idea of the real performance differences between serial and parallel {\it Shims}.
+
+Most of the {\it mpShims} $f$ results are superior to the upper bound found because they are not really the upper bound.
+As {\it Gurobi} runtime was limited, it did not have enough time to update the upper bound.
+But it is a good reference to compare {\it Shims} and {\it mpShims} performances.
+
+As to the speedup, on average, {\it mpShims} was 2.8 times faster than {\it Shims}.
+Theoretically, it could be more, but due to the burden of sharing the CG deviation and the list of items among the processes for reading and writing, the runtime management to hinder race condition limited {\it mpShims} performance.
 
 \subsection{Yet another improvement}
+\label{improvement}
 
@@
 
-Let $Z^{\pi(k)}_{iq}$ be the set of decision variables that relates pallet $i$ to consolidated $q$ in node $\pi(k)$.
+Let $Z^{\pi(k)}_{iq}$ be the set of binary decision variables that relates pallet $i$ to consolidated $q$ in node $\pi(k)$.
 
@@
 Column {\bf Improvement} is the percentage of improvements in distances minimization.
+
+% \ref{eq:distobj} to \ref{eq:disttorque}
 
@@
 
+We added an important contribution by favoring ground operations reducing in 29\%, in average, the distances pallets will be moved on disembarking operations.
+
 Another important contribution is that we employed the {\it iRace} package to fine tune the meta-heuristics parameters, guaranteeing the best possible performance for the instances under training and testing.

### 2023-03-07 latex/comp_or.tex
Shims and mpShims 50 run
scenarios = [6] # infeasible solutions with Shims............
First version of second paper

@@
 
-We also contributed by carrying out multiple experiments with the Shims heuristic on synthetic data based on real data from the {\it Brazilian Air Force} transportation history and a new procedure to minimize the distance from the next node destined pallet to the ramp door.
-
-We converted Shims into a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it surpassed all Shims results.
+We also contributed by carrying out multiple experiments with the {\it Shims} heuristic on synthetic data based on real data from the {\it Brazilian Air Force} transportation history and a new procedure to minimize the distance from the next node destined pallet to the ramp door.
+
+We converted {\it Shims} into a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it surpassed all {\it Shims} results.
 
@@
 \begin{keyword}
-Air Cargo \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Vehicle Routing \sep Parallel computing
+Air Cargo \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Parallel computing \sep iRace
 \end{keyword}
@@
 Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a pre-defined flight plan restricted to two legs.
-It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading (MLU) costs at the intermediate node was modelled through a container sequencing problem.
+It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs ({\bf MLU}) at the intermediate node was modelled through a container sequencing problem.
 Referring directly to this work, \cite[p.
@@
 		\includegraphics[scale=0.25]{Images/shims.png}
-		\captionof{figure}{Shims of various thicknesses}
+		\captionof{figure}{{\it Shims} of various thicknesses}
 		\small\textsuperscript{Source: www.mscdirect.com/product/details/70475967}
@@
 
-$3^{st}$ - Shims construction and best Shims selection
+$3^{st}$ - {\it Shims} construction and best {\it Shims} selection
 
@@
 
-\subsubsection{Shims first phase}
+\subsubsection{{\it Shims} first phase}
 
@@
 
-\subsubsection{Shims second phase: improvement on minimizing CG deviation}
+\subsubsection{{\it Shims} second phase: improvement on minimizing CG deviation}
 
@@
 
-\subsubsection{Shims third and fourth phases}
+\subsubsection{{\it Shims} third and fourth phases}
 
@@
 
-Algorithm \ref{alg:getBestShims} describes the procedure do get the best Shims, the best subset of edges to fill each pallet gap.
+Algorithm \ref{alg:getBestShims} describes the procedure do get the best {\it Shims}, the best subset of edges to fill each pallet gap.
 This algorithm follows the logic of the {\it First-Fit Decreasing} algorithm as described by \cite{JohnsonGarey1985}.
@@
 The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
-	\item {\bf Speed-up}: ratio of the sums of the worst runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
+	\item {\bf Speedup}: ratio of the sums of the worst runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
 \end{itemize}
@@
 \toprule
-&                          &                     &\multicolumn{6}{c}{\bf Scenarios}                                                                        &{\bf Normalized}\\
+{\bf Volume}     &  {\bf Resolution}   &                              &\multicolumn{6}{c}{\bf Scenarios}                                                   &{\bf Normalized}\\
                                                 \cmidrule{4-9}		
-{\it surplus}         & {$method$}               & {\bf Results}      &{\bf 1}      &{\bf 2}      &{\bf 3}      &{\bf 4}      &{\bf 5}       &{\bf 6}      &{\bf Speed-up} \\
+{\it surplus}    & {\it method}        & {\bf Results}                &{\bf 1}      &{\bf 2}      &{\bf 3}      &{\bf 4}      &{\bf 5}       &{\bf 6}      &{\bf Speedup} \\
 \toprule
 
-\multirow{7}{*}{$1.2$}&\multirow{1}{*}{ $Gurobi$}& $f$                &{\bf 11.22}   &{\bf 7.88}   & 10.25      & 11.02       & 11.88        & 12.92       & 0.90 \\% 6484    7222
+%\specialcell{$Volume$\\$surplus$}
+
+\multirow{6}{*}{$1.2$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 11.22}   &{\it 7.88}   &{\it 10.25} &{\it 13.34}  &{\it 14.30}   &{\it 12.92}       & 1.00 \\% 6845
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 9.66         &  6.38       &  9.55       &  11.12      &  12.37      &   12.03      & 0.85 \\%6111
-&                                                & {\it run-time (s)} &  < 1         &  5          &    11       &   43        &  128        &  835        &  1.0 \\%1023
+&\multirow{2}{*}{ $Shims$}                       & $f$                & 9.66         &  6.38       &  9.55       &  11.12      &  12.37      &   12.03      & 0.89 \\% 6111
+&                                                & {\it run-time (s)} &  < 1         &  5          &    11       &   43        &  128        &  835         &  1.0 \\% 1023
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & 11.03        & 7.43        &{\bf 12.58}  &{\bf 13.08}  &{\bf 14.02}  & {\bf 13.44}  &  {\bf 0.99}\\%7158
-
-&                                                & {\it run-time (s)} &  < 1         &  2          &    4        &  15         &  47         & 295          &  {\bf 2.8}\\%364
-
-\midrule
-
-\multirow{7}{*}{$1.2$}&\multirow{1}{*}{ $Gurobi$}& $f$                &{\it 0}   &{\it 0}   &{\it 0}  &{\it 0}  &{\it 0}   &{\it 0}  & {\it 1.00} \\% 6607
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & 11.03        & 7.43        & 10.05       & 13.08       & 14.02       &  12.66       & 0.99\\% 6827
+
+&                                                & {\it run-time (s)} &  < 1         &  2          &    4        &  15         &  47         & 295          &  {\bf 2.8}\\% 364
+
+\midrule
+
+\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 15.17}   &{\it 10.98}  &{\it 17.80}  &{\it 16.92}  &{\it 19.63}   &{\it 18.62}  & {\it 1.00} \\% 9912
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 13.71       &  9.06       &  13.54      &  15.21      &   17.09     &  17.15       & 0.84\\%
-&                                                & {\it run-time (s)} &  < 1        &  3          &    6        &   24        &   70        &  419         & 5.9\\%
+&\multirow{2}{*}{ $Shims$}                       & $f$                & 13.70        &  9.07       &  13.70      &  15.43      &   17.19     &  17.24       & 0.87 \\%8633
+&                                                & {\it run-time (s)} &  < 1         &  4          &    7        &   26        &   81        &  520         & 1.0\\% 639
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & {\bf 16.85} & 10.65       & {\bf 16.68} & 17.05       & 19.72       & {\bf 18.53}  & 0.98\\%
-&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   10        &   28        &   173        & {\bf 14.1}\\%
-
-\midrule
-\multirow{7}{*}{$1.2$}&\multirow{1}{*}{ $Gurobi$}& $f$                &{\it 0}   &{\it 0}   &{\it 0}  &{\it 0}  &{\it 0}   &{\it 0}  & {\it 1.00} \\% 6607
+&\multirow{2}{*}{ $mpShims$}                     & $f$                & 14.82        & 10.45       & 17.45       &{\bf 16.92}  & 19.27       & 18.30        & 0.98\\%9721
+&                                                & {\it run-time (s)} &  < 1         &  2          &    3        &    12       &   38        &   236        & {\bf 2.2}\\% 292
+
+\midrule
+\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 23.11}   &{\it 16.74}  &{\it 23.63}  &{\it 24.33}  &{\it 26.22}  &{\it 25.80}   & {\it 1.00} \\% 13983
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 20.42       &  13.89      &  20.36      &  22.71      &   25.13     & 24.69        & 0.86\\%
-&                                                & {\it run-time (s)} &  < 1        &  4          &    7        &   30        &   85        &  501         & 7.7\\%
+&\multirow{2}{*}{ $Shims$}                       & $f$                &   20.42      &   13.90     &  20.30      &  22.96      &   25.28     &   24.81      & 0.91\\% 12767
+&                                                & {\it run-time (s)} &  2           &  9          &    17       &   68        &   206       &  1216        & 1.0\\% 1518
 
 \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & 23.12       & 14.89       & 22.08       & 23.96       & 26.98       & 25.29        &  0.93\\%
-&                                                & {\it run-time (s)} &  < 1        &  2          &    3        &   12        &   33        &  203         & {\bf 19.0}\\%
+&\multirow{2}{*}{ $mpShims$}                     & $f$                &  22.66       &  15.41      &  23.17      &  23.37      &  23,91      &  25.25       &  0.96\\% 13377
+&                                                & {\it run-time (s)} &  2           &  3          &    5        &   20        &   64        &  398         & {\bf 2.8}\\% 544
 
@@
 
-The upper bound is the objective function value of an LP relaxation of the problem, proving that no integer feasible solution of larger objective value than this value can exist.
-
-In an attempt to find the upper bound for $f$, we solved all scenarios and instances with {\it Gurobi} in a linear relaxed model.
-We applied a time limit of 60s in the expectation that it would be enough for {\it Gurobi} to determine the upper bounds.
+The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater than this.
+
+We solved all scenarios and instances with Gurobi in a relaxed linear model, {\it Relaxed Gurobi}.
 This is good for the reader to have an idea of the real performance differences between serial and parallel {\it Shims}.
 
-Most of the {\it mpShims} $f$ results are superior to the upper bound found because they are not really the upper bound.
-As {\it Gurobi} runtime was limited, it did not have enough time to update the upper bound.
-But it is a good reference to compare {\it Shims} and {\it mpShims} performances.
-
-As to the speedup, on average, {\it mpShims} was 2.8 times faster than {\it Shims}.
-Theoretically, it could be more, but due to the burden of sharing the CG deviation and the list of items among the processes for reading and writing, the runtime management to hinder race condition limited {\it mpShims} performance.
+We were lucky that {\it mpShims} touched the edge of the polyhedron that bounds the feasible region, it found the optimal solution in volume $surplus = 1.5$, scenario {\bf 4}.
+
+{\it mpShims} had a quality performance very close to the {\it Relaxed Gurobi} upper bound solution, and, as to the {\bf Speedup}, on average, mpShims was 2.6 times faster than {\it Shims}.
+Theoretically, it could be more, but due to the burden of sharing the CG deviation and the list of items among the processes (pallets) for reading and writing, the runtime {\it race condition} management limited {\it mpShims} performance.
+
+{\it When threads or processes attempt to simultaneously access a shared resource, and the accesses can result in an error, we often say the program has a race condition, because the threads or processes are in a "race" to carry out an operation.} \cite[p.
+53]{Pacheco:2020}.
 
@@
 
-xxxxx
-
-We added an important contribution by favoring ground operations reducing in 29\%, in average, the distances pallets will be moved on disembarking operations.
-
-Another important contribution is that we employed the {\it iRace} package to fine tune the meta-heuristics parameters, guaranteeing the best possible performance for the instances under training and testing.
+In this work we dealt with a hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem} considering standardized pallets in fixed positions, obeying the center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
+
+We converted \cite{MesquitaSanches2023} {\it Shims} into a process-based parallel computing heuristic, which we called {\it Multiprocessing Shims} or {\it mpShims}, that quickly finds good solutions for a wide range of problem sizes, an essential addition as it surpassed all {\it Shims} results, our first contribution.
+
+We also carryed out 126 experiments with each method on synthetic data based on real data from the {\it Brazilian Air Force} transportation history.
+
+We added another important contribution by favoring ground operations reducing in 29\%, in average, the distances pallets will be moved on disembarking operations, our second contribution.
+
+Another important contribution is that we employed the {\it iRace} package \citep{LopezIbanezManuel2016} to fine tune the meta-heuristics parameters, guaranteeing the best possible performance for the instances under training and testing.
 

### 2023-03-07 main.py

@@ -255,8 +255,8 @@ if __name__ == "__main__":
     # plot = True
     plot = False
 
-    # scenarios = [1,2,3,4,5,6]
-    scenarios = [6] # infeasible solutions with Shims............
+    scenarios = [1,2,3,4,5,6]
+    # scenarios = [6] # infeasible solutions with Shims............
 
     # surplus   = "data20"
     # surplus   = "data50"
@@ -266,8 +266,8 @@ if __name__ == "__main__":
     
     # methods = ["Shims", "mpShims"]
     # methods = ["GRB"]
-    methods = ["Shims"]
-    # methods = ["mpShims"]
+    # methods = ["Shims"]
+    methods = ["mpShims"]
 
     tipo = "KP"    # data50 13.70    13      4.3     1.01    -0.62   0.0     154.1 & 96.3 & 37.5
     # tipo = "FFD" # data50 13.54    5       1.6     1.01    -0.63   0.0     153.7 & 102.7 & 33.2

### 2023-03-07 latex/comp_or.tex
Second paper update

@@
 &\multirow{2}{*}{ $Shims$}                       & $f$                & 9.66         &  6.38       &  9.55       &  11.12      &  12.37      &   12.03      & 0.89 \\% 6111
-&                                                & {\it run-time (s)} &  < 1         &  5          &    11       &   43        &  128        &  835         &  1.0 \\% 1023
+&                                                & {\it run time (s)} &  < 1         &  5          &    11       &   43        &  128        &  835         &  1.0 \\% 1023
 
@@
 
-&                                                & {\it run-time (s)} &  < 1         &  2          &    4        &  15         &  47         & 295          &  {\bf 2.8}\\% 364
+&                                                & {\it run time (s)} &  < 1         &  2          &    4        &  15         &  47         & 295          &  {\bf 2.8}\\% 364
 
@@
 &\multirow{2}{*}{ $Shims$}                       & $f$                & 13.70        &  9.07       &  13.70      &  15.43      &   17.19     &  17.24       & 0.87 \\%8633
-&                                                & {\it run-time (s)} &  < 1         &  4          &    7        &   26        &   81        &  520         & 1.0\\% 639
+&                                                & {\it run time (s)} &  < 1         &  4          &    7        &   26        &   81        &  520         & 1.0\\% 639
 
@@
 &\multirow{2}{*}{ $mpShims$}                     & $f$                & 14.82        & 10.45       & 17.45       &{\bf 16.92}  & 19.27       & 18.30        & 0.98\\%9721
-&                                                & {\it run-time (s)} &  < 1         &  2          &    3        &    12       &   38        &   236        & {\bf 2.2}\\% 292
+&                                                & {\it run time (s)} &  < 1         &  2          &    3        &    12       &   38        &   236        & {\bf 2.2}\\% 292
 
@@
 &\multirow{2}{*}{ $Shims$}                       & $f$                &   20.42      &   13.90     &  20.30      &  22.96      &   25.28     &   24.81      & 0.91\\% 12767
-&                                                & {\it run-time (s)} &  2           &  9          &    17       &   68        &   206       &  1216        & 1.0\\% 1518
+&                                                & {\it run time (s)} &  2           &  9          &    17       &   68        &   206       &  1216        & 1.0\\% 1518
 
@@
 &\multirow{2}{*}{ $mpShims$}                     & $f$                &  22.66       &  15.41      &  23.17      &  23.37      &  23,91      &  25.25       &  0.96\\% 13377
-&                                                & {\it run-time (s)} &  2           &  3          &    5        &   20        &   64        &  398         & {\bf 2.8}\\% 544
+&                                                & {\it run time (s)} &  2           &  3          &    5        &   20        &   64        &  398         & {\bf 2.8}\\% 544
 
@@
 {\it mpShims} had a quality performance very close to the {\it Relaxed Gurobi} upper bound solution, and, as to the {\bf Speedup}, on average, mpShims was 2.6 times faster than {\it Shims}.
-Theoretically, it could be more, but due to the burden of sharing the CG deviation and the list of items among the processes (pallets) for reading and writing, the runtime {\it race condition} management limited {\it mpShims} performance.
+Theoretically, it could be more, but due to the burden of sharing the CG deviation and the list of items among the processes (pallets) for reading and writing, the run-time {\it race condition} management limited {\it mpShims} performance.
 
@@
 
-Another important contribution is that we employed the {\it iRace} package \citep{LopezIbanezManuel2016} to fine tune the meta-heuristics parameters, guaranteeing the best possible performance for the instances under training and testing.
+Another important contribution is that we employed the {\it iRace} package \citep{LopezIbanezManuel2016} to fine tune the {\it mpShims} parameters, guaranteeing the best possible performance for the instances under training and testing.
 

### 2023-03-07 latex/comp_or.tex
Second paper update

@@
 
-Below, we present the implementation for $SolveNode$, the new parallel heuristic that we propose called {\it Multi-process Shims (mpShims)}.
+Below, we present the implementation for $SolveNode$, the new parallel heuristic that we propose called {\it Multiprocessing Shims (mpShims)}.
 
@@
 
-\subsection{Multi-process Shims - mpShims}
+\subsection{Multiprocessing Shims - mpShims}
 
@@
 
-We ran Algorithm \ref{alg:main} in the 6 scenarios described in Table \ref{tab:scenarios}, considering 3 methods for node-by-node resolution: the {\it Gurobi} MIP solver, {\it Shims}, and {\it mpShims} (Algorithm \ref{alg:mpshims}).
+We ran Algorithm \ref{alg:main} in the 6 scenarios described in Table \ref{tab:scenarios}, considering 3 methods for node-by-node resolution: the {\it Gurobi} MIP solver, {\it Shims}, and {\it mpShims} (Algorithm \ref{alg:mpShims}).
 
 In generating the items in each node, we consider 3 values for the parameter ${\it surplus}$.
-The results obtained for the function $f$, with the corresponding runtime in seconds, are shown in Tables \ref{tab:results} (${\it surplus} = 1.2$, ${\it surplus} = 1.5$, and ${\it surplus} = 2.0$).
+The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:results} (${\it surplus} = 1.2$, ${\it surplus} = 1.5$, and ${\it surplus} = 2.0$).
 
@@
 
-The average tour values were presented for $f$ and, for the runtime, the worst result obtained.
+The average tour values were presented for $f$ and, for the run time, the worst result obtained.
 To facilitate the comparison between the methods, we added a last column in these tables, where two values are indicated:

### 2023-03-14 optcgcons.py
OptCGcons and Shims updated

@@ -146,8 +146,6 @@ def minRampDist(pallets, k, tour, rampDistCG, cfg, nodeTorque):
 
         for i in PalletsRange:
 
-            nodeTorque.value += 140 * pallets[i].D 
-
             for j in ConsRange:
 
                 if X[i][j].x >= 0.99:
@@ -158,7 +156,7 @@ def minRampDist(pallets, k, tour, rampDistCG, cfg, nodeTorque):
                     pallets[i].PCV     = cons[j].V
                     pallets[i].PCS     = cons[j].S
 
-                    nodeTorque.value += cons[j].W * pallets[i].D                     
+                    nodeTorque.value += (140 + cons[j].W) * pallets[i].D                     
 
 
 if __name__ == "__main__":

### 2023-03-14 latex/comp_or.tex
OptCGcons and Shims updated
Paper updated

@@
 
-{\color{blue}
 Figure \ref{fig:whip} represents the $n_{\pi(k)}$\/ possible edges $e_{ij}^{\pi(k)}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi(k)}$.
@@
 
-Although the second phase solves a knapsack problem, an algorithm for solving a bin packing problem was used (the First Fit Decreasing - FFD), and the most profitable bin (shims) was selected.
-This method was proved to be faster than a dynamic program solution for the knapsack problem.
+Although the second phase solves a knapsack problem, an algorithm for solving a Bin Packing Problem was used (the First Fit Decreasing - FFD), and the most profitable bin (Shims) was selected.
+This method was proved to be faster than a Dynamic Program solution for the Knapsack Problem.
 
@@
 
-$1^{st}$ - Greedy phase with torque surplus
-
-$2^{st}$ - Minimization of CG deviation
-
-$3^{st}$ - {\it Shims} construction and best {\it Shims} selection
-
-All executed in a multi-process runtime, accessing concurrently, reading and writing, the list of items and the current value of the CG deviation.
-
-As to the multi-process runtime, we considered each pallet as a process.
-}
+$1^{st}$ - A Greedy phase to partially build pallets.
+
+$2^{st}$ - Minimization of CG deviation.
+
+$3^{st}$ - {\it Shims} set construction and best {\it Shims} selection
+
+All executed in a multi-process execution, accessing concurrently, reading and writing, the list of items and the current value of the CG deviation.
+As to the multi-process execution, we considered each pallet as a process.
 
@@
 
-Line \ref{mpShims:minCGdev} calls the $minCGdev$ method which updates the $nodeTorque$ and $G$. $minCGdev$ minimizes the CG deviation and repairs the greedy solution if it is infeasible.
+Line \ref{mpShims:minCGdev} calls the $minCGdev$ integer programming method (Subsection \ref{subsection:second_phase}) which updates the $nodeTorque$ and $G$.
+$minCGdev$ minimizes the CG deviation.
+We raise the hypothesis that, in sequential mode, where one pallet is built at a time, torque constraints may be more restrictive than volume or weight constraints, which could stop the palletization process earlier, fostering a solution that may be not as good as it could be.
+So, we decided to balance cargo before the next {\it Shims} phase to allow more room for pallet fulfillment.
+This will be confirmed (or not) in Section \ref{sec6}.
 
@@
 \subsubsection{{\it Shims} second phase: improvement on minimizing CG deviation}
-
-Between the greedy and the {\it Shims} selection phases we inserted, in this work, a method to minimize the CG deviation, as stated by the model that follows.
+\label{subsection:second_phase}
+
+Between the greedy and the {\it Shims} selection phases we inserted, in this work, an integer programming method to minimize the CG deviation, as stated by the model that follows.
 
@@
 
+\vspace{2.0mm}
+
+\begin{table}[H]
+	\centering
+	\caption{Volume surpluses, methods and scenarios results}  \label{tab:results2}
+	\footnotesize
+	%\scriptsize
+	\begin{tabular}{crcccccccc}
+		\toprule
+		{\bf Volume}     &  {\bf Resolution}   &                              &\multicolumn{6}{c}{\bf Scenarios}                                                   &{\bf Normalized}\\
+		\cmidrule{4-9}		
+		{\it surplus}    & {\it method}        & {\bf Results}                &{\bf 1}      &{\bf 2}      &{\bf 3}      &{\bf 4}      &{\bf 5}       &{\bf 6}      &{\bf Speedup} \\
+		\toprule
+
+		%\specialcell{$Volume$\\$surplus$}
+
+		\multirow{6}{*}{$1.2$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 12.02}   &{\it 8.74}   &{\it 12.56} &{\it 14.34}  &{\it 14.79}   &{\it 13.49}       & 1.00 \\%
+
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $Shims$}                       & $f$                &  10.34       & 8.50        & 12.42      &  15.06      &  15.97       &  15.88       &   \\% 
+		&                                                & {\it run time (s)} &  < 1         &  3          &  5         &   20        &  58          &  352         &  1.0 \\%
+
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $mpShims$}                     & $f$                &  11.76       &  8.57       & 12.28      &  14.02      &  14.50       & 13.23        & \\%
+
+		&                                                & {\it run time (s)} &  < 1         &  2          &  3         &   47        &  33          & 193         &  {\bf 2.8}\\%
+
+		\midrule
+
+		\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 15.17}   &{\it 10.98}  &{\it 17.80}  &{\it 16.92}  &{\it 19.63}   &{\it 18.62}  & {\it 1.00} \\% 9912
+
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $Shims$}                       & $f$                &         &         &       &        &        &         &  \\%8633
+		&                                                & {\it run time (s)} &  < 1         &  4          &    7        &   26        &   81        &  520         & 1.0\\% 639
+
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $mpShims$}                     & $f$                &         &        &        &  &        &         & \\%9721
+		&                                                & {\it run time (s)} &  < 1         &  2          &    3        &    12       &   38        &   236        & {\bf 2.2}\\% 292
+
+		\midrule
+		\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 23.11}   &{\it 16.74}  &{\it 23.63}  &{\it 24.33}  &{\it 26.22}  &{\it 25.80}   & {\it 1.00} \\% 13983
+
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $Shims$}                       & $f$                &         &        &        &        &        &         & \\% 12767
+		&                                                & {\it run time (s)} &  2           &  9          &    17       &   68        &   206       &  1216        & 1.0\\% 1518
+
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $mpShims$}                     & $f$                &         &        &        &        &        &         &  \\% 13377
+		&                                                & {\it run time (s)} &  2           &  3          &    5        &   20        &   64        &  398         & {\bf 2.8}\\% 544
+
+		\bottomrule
+
+	\end{tabular}
+	%\textsuperscript{asdf sd .}	
+	\normalsize
+\end{table}
+
 The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater than this.
@@
 {\it mpShims} had a quality performance very close to the {\it Relaxed Gurobi} upper bound solution, and, as to the {\bf Speedup}, on average, mpShims was 2.6 times faster than {\it Shims}.
-Theoretically, it could be more, but due to the burden of sharing the CG deviation and the list of items among the processes (pallets) for reading and writing, the run-time {\it race condition} management limited {\it mpShims} performance.
+Theoretically, it could be more, but due to the burden of sharing the CG deviation and the list of items among the processes (pallets) for reading and writing, the execution {\it race condition} management limited {\it mpShims} performance.
 
@@
 53]{Pacheco:2020}.
+
+We confirmed the hypothesis raised in line \ref{mpShims:minCGdev} of Algorithm \ref{alg:mpShims} that, in sequential mode, where one pallet is built at a time, torque constraints could be more restrictive than volume or weight constraints, which could stop the palletization process earlier, fostering a solution that may be not as good as it could be.
+This integer programming procedure inclusion really improved {\it Shims} results.
 
@@
 \midrule
-\multirow{6}{*}{$1.2$}& 1 &  33  &  16 & 52\\
-&                       2 & 194  & 154 & 21\\
-&                       3 & 149  & 119 & 20 \\
-&                       4 & 135  & 103 & 23 \\
-&                       5 & 121  &  84 & 31 \\
-&                       6 & 105  &  71 & 33 \\
-\midrule
-\multirow{6}{*}{$1.5$}& 1 &  32 &  18 & 44 \\
-&                       2 & 198 & 135 & 32 \\
-&                       3 & 156 & 128 & 18 \\
-&                       4 & 137 & 106 & 22\\
-&                       5 & 120 &  87 & 28 \\
-&                       6 & 106 &  77 & 27 \\
-\midrule
-\multirow{6}{*}{$2.0$}& 1 &  33 &  20 & 38 \\
-&                       2 & 199 & 142 & 28 \\
-&                       3 & 156 & 122 & 22 \\
-&                       4 & 138 & 106 & 23 \\
-&                       5 & 123 &  84 & 31 \\
-&                       6 & 109 &  75 & 31 \\
+\multirow{6}{*}{$1.2$}& 1 &  32 &  32 & 0.0 \\
+&                       2 & 181 & 181 & 0.0 \\
+&                       3 & 149 & 149 & 0.0\\
+&                       4 & 131 & 131 & 0.0  \\
+&                       5 & 113 & 113 & 0.0  \\
+&                       6 & 105 & 105 & 0.0 \\
+\midrule
+\multirow{6}{*}{$1.5$}& 1 &  &  &  \\
+&                       2 &  &  &  \\
+&                       3 &  &  &  \\
+&                       4 & 123 & 92 & 25.3  \\
+&                       5 &  &  &  \\
+&                       6 &  &  &  \\
+\midrule
+\multirow{6}{*}{$2.0$}& 1 &  &  &  \\
+&                       2 &  &  &  \\
+&                       3 &  &  &  \\
+&                       4 &  &  &  \\
+&                       5 &  &  &  \\
+&                       6 &  &  &  \\
 \midrule

### 2023-03-14 latex/comp_or.tex
minCGdev optimized

@@
 
-\multirow{6}{*}{$1.2$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 11.22}   &{\it 7.88}   &{\it 10.25} &{\it 13.34}  &{\it 14.30}   &{\it 12.92}       & 1.00 \\% 6845
+		\multirow{6}{*}{$1.2$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 12.23}&{\it 8.67} &{\it 12.67}&{\it 15.36}&{\it 16.29}&{\it 16.20}&  \specialcell{{\bf 1.00}\\-} \\%  8142
 
 		\cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 9.66         &  6.38       &  9.55       &  11.12      &  12.37      &   12.03      & 0.89 \\% 6111
-&                                                & {\it run time (s)} &  < 1         &  5          &    11       &   43        &  128        &  835         &  1.0 \\% 1023
+		&\multirow{2}{*}{ $Shims$}                       & $f$                &     10.34 &     8.50  &     12.42 &     15.06 &     15.97 &     15.88 &  0.96 \\% 7817
+		&                                                & {\it run time (s)} &     < 1   &     3     &     5     &     20    &     58    &     352   &  1.0 \\% 439
 
 		\cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & 11.03        & 7.43        & 10.05       & 13.08       & 14.02       &  12.66       & 0.99\\% 6827
-
-&                                                & {\it run time (s)} &  < 1         &  2          &    4        &  15         &  47         & 295          &  {\bf 2.8}\\% 364
-
-\midrule
-
-\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 15.17}   &{\it 10.98}  &{\it 17.80}  &{\it 16.92}  &{\it 19.63}   &{\it 18.62}  & {\it 1.00} \\% 9912
+		&\multirow{2}{*}{ $mpShims$}                     & $f$                &     11.73 &     7.57  &     12.05 &     12.95 &     13.94 &     13.62 & 0.88 \\% 7186
+
+		&                                                & {\it run time (s)} &     < 1   &     2     &     3     &     16    &     33    &     193   &  {\bf 1.8}\\% 248
+
+		\midrule
+
+		\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it }&{\it  } &{\it  }&{\it  }&{\it  }&{\it  }&  \specialcell{{\bf 1.00}\\-} \\%
 
         \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                & 13.70        &  9.07       &  13.70      &  15.43      &   17.19     &  17.24       & 0.87 \\%8633
-&                                                & {\it run time (s)} &  < 1         &  4          &    7        &   26        &   81        &  520         & 1.0\\% 639
+        &\multirow{2}{*}{ $Shims$}                       & $f$                &     14.68 &   12.21   &  17.67    &  21.01    &  22.64    &       &    \\% 
+        &                                                & {\it run time (s)} &     < 1   &     2     &     3     &     12    &     33    &        &  1.0 \\%
 
         \cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                & 14.82        & 10.45       & 17.45       &{\bf 16.92}  & 19.27       & 18.30        & 0.98\\%9721
-&                                                & {\it run time (s)} &  < 1         &  2          &    3        &    12       &   38        &   236        & {\bf 2.2}\\% 292
-
-\midrule
-\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 23.11}   &{\it 16.74}  &{\it 23.63}  &{\it 24.33}  &{\it 26.22}  &{\it 25.80}   & {\it 1.00} \\% 13983
-
-\cmidrule{2-10}		                       
-&\multirow{2}{*}{ $Shims$}                       & $f$                &   20.42      &   13.90     &  20.30      &  22.96      &   25.28     &   24.81      & 0.91\\% 12767
-&                                                & {\it run time (s)} &  2           &  9          &    17       &   68        &   206       &  1216        & 1.0\\% 1518
-
-\cmidrule{2-10}		                       
-&\multirow{2}{*}{ $mpShims$}                     & $f$                &  22.66       &  15.41      &  23.17      &  23.37      &  23,91      &  25.25       &  0.96\\% 13377
-&                                                & {\it run time (s)} &  2           &  3          &    5        &   20        &   64        &  398         & {\bf 2.8}\\% 544
+        &\multirow{2}{*}{ $mpShims$}                     & $f$                &       &        &       &       &       &       &   \\% 7186
+        &                                                & {\it run time (s)} &     < 1   &     2     &     3     &     16    &     33    &     193   &  {\bf 1.8}\\% 248
+
+\midrule
 
@@
 
-\vspace{2.0mm}
-
-\begin{table}[H]
-	\centering
-	\caption{Volume surpluses, methods and scenarios results}  \label{tab:results2}
-	\footnotesize
-	%\scriptsize
-	\begin{tabular}{crcccccccc}
-		\toprule
-		{\bf Volume}     &  {\bf Resolution}   &                              &\multicolumn{6}{c}{\bf Scenarios}                                                   &{\bf Normalized}\\
-		\cmidrule{4-9}		
-		{\it surplus}    & {\it method}        & {\bf Results}                &{\bf 1}      &{\bf 2}      &{\bf 3}      &{\bf 4}      &{\bf 5}       &{\bf 6}      &{\bf Speedup} \\
-		\toprule
-
-		%\specialcell{$Volume$\\$surplus$}
-
-		\multirow{6}{*}{$1.2$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 12.02}   &{\it 8.74}   &{\it 12.56} &{\it 14.34}  &{\it 14.79}   &{\it 13.49}       & 1.00 \\%
-
-		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}                       & $f$                &  10.34       & 8.50        & 12.42      &  15.06      &  15.97       &  15.88       &   \\% 
-		&                                                & {\it run time (s)} &  < 1         &  3          &  5         &   20        &  58          &  352         &  1.0 \\%
-
-		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}                     & $f$                &  11.76       &  8.57       & 12.28      &  14.02      &  14.50       & 13.23        & \\%
-
-		&                                                & {\it run time (s)} &  < 1         &  2          &  3         &   47        &  33          & 193         &  {\bf 2.8}\\%
-
-		\midrule
-
-		\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 15.17}   &{\it 10.98}  &{\it 17.80}  &{\it 16.92}  &{\it 19.63}   &{\it 18.62}  & {\it 1.00} \\% 9912
-
-		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}                       & $f$                &         &         &       &        &        &         &  \\%8633
-		&                                                & {\it run time (s)} &  < 1         &  4          &    7        &   26        &   81        &  520         & 1.0\\% 639
-
-		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}                     & $f$                &         &        &        &  &        &         & \\%9721
-		&                                                & {\it run time (s)} &  < 1         &  2          &    3        &    12       &   38        &   236        & {\bf 2.2}\\% 292
-
-		\midrule
-		\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 23.11}   &{\it 16.74}  &{\it 23.63}  &{\it 24.33}  &{\it 26.22}  &{\it 25.80}   & {\it 1.00} \\% 13983
-
-		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}                       & $f$                &         &        &        &        &        &         & \\% 12767
-		&                                                & {\it run time (s)} &  2           &  9          &    17       &   68        &   206       &  1216        & 1.0\\% 1518
-
-		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}                     & $f$                &         &        &        &        &        &         &  \\% 13377
-		&                                                & {\it run time (s)} &  2           &  3          &    5        &   20        &   64        &  398         & {\bf 2.8}\\% 544
-
-		\bottomrule
-
-	\end{tabular}
-	%\textsuperscript{asdf sd .}	
-	\normalsize
-\end{table}
-
 The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater than this.
@@
 
-We were lucky that {\it mpShims} touched the edge of the polyhedron that bounds the feasible region, it found the optimal solution in volume $surplus = 1.5$, scenario {\bf 4}.
-
-{\it mpShims} had a quality performance very close to the {\it Relaxed Gurobi} upper bound solution, and, as to the {\bf Speedup}, on average, mpShims was 2.6 times faster than {\it Shims}.
+{\it Shims} had a quality performance very close to the {\it Relaxed Gurobi} upper bound solution, and, as to the {\bf Speedup}, on average, {\it mpShims} was 1.8 times faster than {\it Shims}.
 Theoretically, it could be more, but due to the burden of sharing the CG deviation and the list of items among the processes (pallets) for reading and writing, the execution {\it race condition} management limited {\it mpShims} performance.
@@
 
-\subsection{Yet another improvement}
+After this inclusion {\it Shims} results surpassed that of {\it mpShims}. In multi-processing mode ({\it mpShims}) this inclusion (line \ref{mpShims:minCGdev}) made a negligible difference.
+
+\subsection{Another possible improvement}
 \label{improvement}
@@
 
-Columns {\bf Before} and {\bf After} on Table \ref{tab:distRampDoor} refer to the application of the integer programming solver according to equations \ref{eq:distobj} to \ref{eq:disttorque}.
-Column {\bf Improvement} is the percentage of improvements in distances minimization.
-
-% \ref{eq:distobj} to \ref{eq:disttorque}
-
-\vspace{2.0mm}
-
-\begin{table}[H]
-\centering
-\caption{Minimization of pallets distances from the ramp door}  \label{tab:distRampDoor}
-\footnotesize
-%\scriptsize
-\begin{tabular}{ccccc}
-\toprule
-
-{\it surplus} & {\bf Scenario} & {\bf Before (m)} & {\bf After (m)}  & {\bf Improvement (\%)}  \\
-\midrule
-\multirow{6}{*}{$1.2$}& 1 &  32 &  32 & 0.0 \\
-&                       2 & 181 & 181 & 0.0 \\
-&                       3 & 149 & 149 & 0.0\\
-&                       4 & 131 & 131 & 0.0  \\
-&                       5 & 113 & 113 & 0.0  \\
-&                       6 & 105 & 105 & 0.0 \\
-\midrule
-\multirow{6}{*}{$1.5$}& 1 &  &  &  \\
-&                       2 &  &  &  \\
-&                       3 &  &  &  \\
-&                       4 & 123 & 92 & 25.3  \\
-&                       5 &  &  &  \\
-&                       6 &  &  &  \\
-\midrule
-\multirow{6}{*}{$2.0$}& 1 &  &  &  \\
-&                       2 &  &  &  \\
-&                       3 &  &  &  \\
-&                       4 &  &  &  \\
-&                       5 &  &  &  \\
-&                       6 &  &  &  \\
-\midrule
-&                         &        & {\bf Average} & {\bf 29}\\
-\bottomrule	
-\end{tabular}
-%\textsuperscript{asdf sd .}	
-\normalsize
-\end{table}
-
-This last procedure favored ground operations as it reduced in 29\% the distances pallets will be moved on disembarking operations.
-This last step is not part of {\it mpShims} and may be applied as a post improvement for any method.
-This is another important contribution of this work.
+There was some expectation on this post-improvement method as it could have minimized the distances of travel of the pallets to the ramp door, but unfortunately, there was no alteration to these distances.
 
@@
 
-We added another important contribution by favoring ground operations reducing in 29\%, in average, the distances pallets will be moved on disembarking operations, our second contribution.
-
 Another important contribution is that we employed the {\it iRace} package \citep{LopezIbanezManuel2016} to fine tune the {\it mpShims} parameters, guaranteeing the best possible performance for the instances under training and testing.

### 2023-03-14 main.py

@@ -102,7 +102,6 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
                                 vNodeAccum += c.V
 
                             tour.score += c.S
-                            nodeVol    += c.V
 
         # set pallets destinations with items and consolidated to be delivered
         if k < k0: # except when the current node is the base on returning
@@ -195,7 +194,7 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
         epsilon = torque/cfg.maxTorque
 
         if method == "GRB":
-            tour.score += ObjBound # Gurobi linear relaxation
+            tour.score += max(ObjBound, nodeScore) # Gurobi linear relaxation
         else:
             tour.score += nodeScore
 
@@ -265,17 +264,17 @@ if __name__ == "__main__":
     plot = False
 
     scenarios = [1,2,3,4,5,6]
-    # scenarios = [2] # infeasible solutions with Shims............
+    # scenarios = [6] # infeasible solutions with Shims............
 
-    surplus   = "data20"
-    # surplus   = "data50"
+    # surplus   = "data20"
+    surplus   = "data50"
     # surplus   = "data100"
 
-    # methods = ["Shims","mpShims","GRB"]  
+    methods = ["Shims","mpShims","GRB"]  
     
     # methods = ["Shims", "mpShims"]
     # methods = ["GRB"]
-    methods = ["Shims"]
+    # methods = ["Shims"]
     # methods = ["mpShims"]
 
     # tipo = "KP"
@@ -423,4 +422,5 @@ if __name__ == "__main__":
                 print(f"Before:\t{beforeDict['Before']:.1f}") 
                 print(f"After:\t{afterDict['After']:.1f}")
                 print(f"% of optima: {numOptDict['numOpt']:.2f}")
+                print(f"{method}")
 

### 2023-03-14 mipGRB.py

@@ -9,6 +9,12 @@ from gurobipy import GRB
 # import multiprocessing as mp
 # import numpy as np
 
+def mycallback(model, where):
+    if where == GRB.Callback.MIPNODE:
+        status = model.cbGet(GRB.Callback.MIPNODE_STATUS)
+        if status == GRB.OPTIMAL:
+            print(model.cbGetNodeRel(model._vars))
+
 def Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict):
 
     # itemsDict to control items inclusion feasibility
@@ -20,18 +26,17 @@ def Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict):
     N = len(items)
     M = len(pallets)
 
-    palletWeights = [0 for _ in pallets]
 
     set_M = range( M ) # i, pallets
     set_N = range( N ) # j, items
 
     # initialize a model
     mod = gp.Model()
-    # mod = mod.relax()
+    mod = mod.relax()
     
     # decision matrix for which items will be put in which pallet in node "k"         
-    X = [ [ mod.addVar(name=f"X[{i}],[{j}]", vtype=GRB.BINARY) for j in set_N ] for i in set_M ]  
-    # X = [ [ mod.addVar(name=f"X[{i}],[{j}]", vtype=GRB.CONTINUOUS) for j in set_N ] for i in set_M ]   
+    # X = [ [ mod.addVar(name=f"X[{i}],[{j}]", vtype=GRB.BINARY) for j in set_N ] for i in set_M ]  
+    X = [ [ mod.addVar(name=f"X[{i}],[{j}]", vtype=GRB.CONTINUOUS) for j in set_N ] for i in set_M ]   
 
     mod.setObjective(sum( X[i][j] * items[j].S for i in set_M for j in set_N ))
 
@@ -45,6 +50,10 @@ def Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict):
             sum(X[i][j] for i in set_M ) <= 1
         )
 
+    palletWeights = [0 for _ in pallets]
+    itemsWeights  = [0 for _ in pallets]
+    itemsVolumes  = [0 for _ in pallets]
+
     for i in set_M:
 
         # items must be grouped in a pallet with the same destination
@@ -57,17 +66,19 @@ def Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict):
             )
 
[…]

### 2023-03-15 latex/comp_or.tex
mpShims still worst
Paper COR updated

@@
 
-{\color{blue} Each node, except the base, inherits information from the previous nodes (pallets that must remain on board), and must solved taking into account the remaining not visited nodes.
-One more reason for a node-by-node approach.}
+Each node, except the base, inherits information from the previous nodes (pallets that must remain on board), and must be solved taking into account the remaining not visited nodes.
+One more reason for a node-by-node approach.
 
@@
 
-\subsubsection{{\it Shims} third and fourth phases}
+\subsubsection{{\it Shims} third phase}
 
@@
 
-Algorithm \ref{alg:getBestShims} describes the procedure do get the best {\it Shims}, the best subset of edges to fill each pallet gap.
-This algorithm follows the logic of the {\it First-Fit Decreasing} algorithm as described by \cite{JohnsonGarey1985}.
+Algorithm \ref{alg:getBestShims} describes the procedure to get the top {\it Shims}, the best subset of edges to fill each pallet gap.
+This algorithm follows the logic of the {\it First-Fit Decreasing} method as described by \cite{JohnsonGarey1985}.
+Although it does not aim to minimize the number of bins, its mechanism helped build the set of {\it Shims} to allow the best to be chosen.
 
 Line \ref{empty_shims} creates the first empty shims.
-Line \ref{empty_set} creates the first set of shims from where the best shims will be chosen.
+Line \ref{empty_set} produce the first set of shims from where the best shims will be chosen.
 
@@
 		&                                                & {\it run time (s)} &     < 1   &     3     &     5     &     20    &     58    &     352   &  1.0 \\% 439			
-
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}                     & $f$                &     11.73 &     7.57  &     12.05 &     12.95 &     13.94 &     13.62 & 0.88 \\% 7186
-
-		&                                                & {\it run time (s)} &     < 1   &     2     &     3     &     16    &     33    &     193   &  {\bf 1.8}\\% 248
-
-		\midrule
-
-		\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it }&{\it  } &{\it  }&{\it  }&{\it  }&{\it  }&  \specialcell{{\bf 1.00}\\-} \\%
+		&\multirow{2}{*}{ $mpShims$}                     & $f$                &   12.96   &   7.53    &  12.28    &   13.20   &  13.71    &   13.44   &  0.90\\% 7278
+		&                                                & {\it run time (s)} &     < 1   &     2     &     2     &     8    &     22    &     120   &  {\bf 2.8}\\%
+
+		\midrule
+
+		\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 17.01}&{\it 12.45}&{\it 18.02}&{\it 22.03}&{\it 23.09}&{\it 22.35}&  \specialcell{{\bf 1.00}\\-} \\%11495
 
         \cmidrule{2-10}		                       
-        &\multirow{2}{*}{ $Shims$}                       & $f$                &     14.68 &   12.21   &  17.67    &  21.01    &  22.64    &       &    \\% 
-        &                                                & {\it run time (s)} &     < 1   &     2     &     3     &     12    &     33    &        &  1.0 \\%
+        &\multirow{2}{*}{ $Shims$}                       & $f$                &     14.68 &     12.21 &     17.67 &  21.01    &  22.64    &   21.91   &  0.96  \\%  11012
+        &                                                & {\it run time (s)} &     < 1   &     2     &     3     &     12    &     33    &   209     &  1.0 \\% 260
 
         \cmidrule{2-10}		                       
-        &\multirow{2}{*}{ $mpShims$}                     & $f$                &       &        &       &       &       &       &   \\% 7186
-        &                                                & {\it run time (s)} &     < 1   &     2     &     3     &     16    &     33    &     193   &  {\bf 1.8}\\% 248
-
-\midrule
+        &\multirow{2}{*}{ $mpShims$}                     & $f$                &    16.12  &     11.81 &     17.48 &  17.66    &   19.81   &  17.61    &  0.87 \\% 9966
+        &                                                & {\it run time (s)} &     < 1   &     1     &     2     &     7     &   20      &  120      &  {\bf 1.7}\\% 151
+
+		\midrule
+		\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 24.56}&{\it 17.47}&{\it 25.19}&{\it 27.94}&{\it  }&{\it  }&  \specialcell{{\bf 1.00}\\-} \\%11495
+
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $Shims$}                       & $f$                &    21.31  &  17.35    &   24.24   &   27.39    &  29.31     &  27.06     &     \\%  11012
+		&                                                & {\it run time (s)} &     < 1   &     2     &     4     &     14     &     35     &   220     &  1.0 \\% 260
+
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $mpShims$}                     & $f$                &    24.33  &   17.26   &   24.70   &  24.81     &   27.71    &  24.93        &    \\% 9966
+		&                                                & {\it run time (s)} &     < 1   &     2     &     3     &     9      &   26       &  151      &  {\bf 1.7}\\% 151
 
@@
 
-We solved all scenarios and instances with Gurobi in a relaxed linear model, {\it Relaxed Gurobi}.
+We solved all scenarios and instances with {\it Gurobi} in a relaxed linear model, {\it Relaxed Gurobi}.
 This is good for the reader to have an idea of the real performance differences between serial and parallel {\it Shims}.
 
-{\it Shims} had a quality performance very close to the {\it Relaxed Gurobi} upper bound solution, and, as to the {\bf Speedup}, on average, {\it mpShims} was 1.8 times faster than {\it Shims}.
+{\it Shims} had a quality performance very close to the {\it Relaxed Gurobi} upper bound solution, and, as to the {\bf Speedup}, on average, {\it mpShims} was ....
+times faster than {\it Shims}.
 Theoretically, it could be more, but due to the burden of sharing the CG deviation and the list of items among the processes (pallets) for reading and writing, the execution {\it race condition} management limited {\it mpShims} performance.
@@
 We confirmed the hypothesis raised in line \ref{mpShims:minCGdev} of Algorithm \ref{alg:mpShims} that, in sequential mode, where one pallet is built at a time, torque constraints could be more restrictive than volume or weight constraints, which could stop the palletization process earlier, fostering a solution that may be not as good as it could be.
-This integer programming procedure inclusion really improved {\it Shims} results.
+This integer programming procedure inclusion really improved all {\it Shims} results.
 
@@
 
-There was some expectation on this post-improvement method as it could have minimized the distances of travel of the pallets to the ramp door, but unfortunately, there was no alteration to these distances.
+There was some expectation on this post-improvement method as it could have minimized the distances of travel of the pallets to the ramp door, but unfortunately, after running all scenarios and problem instances, there was no improvement in minimizing these distances.
+
+This proves that the solutions provided by {\it Shims} and {\it mpShims} are tight enough to hinder this intended improvement.
 

### 2023-03-15 main.py

@@ -267,15 +267,15 @@ if __name__ == "__main__":
     # scenarios = [2] # infeasible solutions with Shims............
 
     # surplus   = "data20"
-    surplus   = "data50"
-    # surplus   = "data100"
+    # surplus   = "data50"
+    surplus   = "data100"
 
-    # methods = ["Shims","mpShims","GRB"]  
+    methods = ["Shims","mpShims","GRB"]  
     
     # methods = ["Shims", "mpShims"]
     # methods = ["GRB"]
     # methods = ["Shims"]
-    methods = ["mpShims"]
+    # methods = ["mpShims"]
 
     # tipo = "KP"
     tipo = "FFD"

### 2023-03-16 latex/comp_or.tex
Paper Compo OR update.

@@
 They develop a heuristic that can be executed on a simple handheld computer (like a laptop or a tablet) and that provides a solution quickly enough to keep this cargo handling time under an hour.
-This work seeks to improve their results in terms of quality and performance by the use of computer parallelism and by minimizing the distance between the next node destined pallets and the cargo door.
+
+In this work, we raise three hypotheses for improvements to the work of \cite{MesquitaSanches2023}:
+(1) Computer parallelism could improve {\it Shims}'s performance.
+(2) The {\it Shims}' results may have been inferior to that; it could be that a sequential approach to pallet building may lead to a premature stop due to the torque constraints.
+(3) The distances to move pallets on unloading in the next node could be minimized to benefit unloading costs.
+
+So, this work seeks to improve their results in terms of quality and performance by using computer parallelism, CG balancing between {\it Shims}' phases, and minimizing the distance between the next node-destined pallets and the cargo door.
 
@@
 		\midrule
-		\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 24.56}&{\it 17.47}&{\it 25.19}&{\it 27.94}&{\it  }&{\it  }&  \specialcell{{\bf 1.00}\\-} \\%11495
+		\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 24.56}&{\it 17.47}&{\it 25.19}&{\it 27.94}&{\it 29.90}&{\it 27.60 }&  \specialcell{{\bf 1.00}\\-} \\%15266
 
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}                       & $f$                &    21.31  &  17.35    &   24.24   &   27.39    &  29.31     &  27.06     &     \\%  11012
+		&\multirow{2}{*}{ $Shims$}                       & $f$                &    21.31  &  17.35    &   24.24   &   27.39    &  29.31     &  27.06     &   0.96  \\%  14666
 		&                                                & {\it run time (s)} &     < 1   &     2     &     4     &     14     &     35     &   220     &  1.0 \\% 260
@@
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}                     & $f$                &    24.33  &   17.26   &   24.70   &  24.81     &   27.71    &  24.93        &    \\% 9966
+		&\multirow{2}{*}{ $mpShims$}                     & $f$                &    24.33  &   17.26   &   24.70   &  24.81     &   27.71    &  24.93        &  0.94  \\% 14374
 		&                                                & {\it run time (s)} &     < 1   &     2     &     3     &     9      &   26       &  151      &  {\bf 1.7}\\% 151
@@
 
-The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater than this.
+The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.
 
@@
 
-{\it Shims} had a quality performance very close to the {\it Relaxed Gurobi} upper bound solution, and, as to the {\bf Speedup}, on average, {\it mpShims} was ....
-times faster than {\it Shims}.
-Theoretically, it could be more, but due to the burden of sharing the CG deviation and the list of items among the processes (pallets) for reading and writing, the execution {\it race condition} management limited {\it mpShims} performance.
+{\it Shims} had a quality performance very close to the {\it Relaxed Gurobi} upper bound solution, and, as to the {\bf Speedup}, on average, {\it mpShims} was 2 times faster than {\it Shims}.
+Theoretically, it could be more, but due to the burden of sharing, for reading and writing, the CG deviation and the list of items among the processes (pallets), the {\it race condition} management limited {\it mpShims} performance.
 
@@
 
-Let $\pi$ be a tour in the set of permutations of nodes.
+So, to confirm or not Hypothesis (3), that the distances to move pallets on unloading in the next node could be minimized to benefit unloading costs. We modeled this improvement as follows.
 
@@
 
-There was some expectation on this post-improvement method as it could have minimized the distances of travel of the pallets to the ramp door, but unfortunately, after running all scenarios and problem instances, there was no improvement in minimizing these distances.
+There was some expectation on this post-improvement method as it could have minimized the distances of travel of the pallets through the ramp door, but unfortunately, after running all scenarios and problem instances, there was no improvement in minimizing these distances.
 
 This proves that the solutions provided by {\it Shims} and {\it mpShims} are tight enough to hinder this intended improvement.
+
+\subsection{Discussion summary}
+
+In this work, we raised three hypotheses for improvements to the work of \cite{MesquitaSanches2023}:
+
+(1) Computer parallelism could improve {\it Shims}'s performance: {\it mpShims} was twice as faster as {\it Shims}. This hypothesis was confirmed.
+
+(2) The {\it Shims}' results may have been inferior in that of \cite{MesquitaSanches2023}; it could be that a sequential approach to pallet building may have lead to a premature stop due to the torque constraints.
+This hypothesis was confirmed as the results were improved by 5\% on average with the inclusion of a CG balancing integer programming procedure.
+
+(3) The distances to move pallets on unloading in the next node could be minimized to benefit unloading costs.
+This hypothesis was not confirmed.
+We implemented an integer programming procedure to minimize these distances, but the results proved that any change in {\it Shims} or {\it mpShims} outcomes could turn the solutions unfeasible due to the CG balance constraints.
+
+A new question is raised about whether it is worth using parallelism as, although {\it mpShims} was twice as fast as {\it Shims}, the results of {\it Shims} were still better than those of {\it mpShims}.
+May this question be solved through the real-world application of these methods.
 
@@
 
-We converted \cite{MesquitaSanches2023} {\it Shims} into a process-based parallel computing heuristic, which we called {\it Multiprocessing Shims} or {\it mpShims}, that quickly finds good solutions for a wide range of problem sizes, an essential addition as it surpassed all {\it Shims} results, our first contribution.
-
-We also carryed out 126 experiments with each method on synthetic data based on real data from the {\it Brazilian Air Force} transportation history.
-
-Another important contribution is that we employed the {\it iRace} package \citep{LopezIbanezManuel2016} to fine tune the {\it mpShims} parameters, guaranteeing the best possible performance for the instances under training and testing.
+We converted \cite{MesquitaSanches2023} {\it Shims} into a process-based parallel computing heuristic, which we called {\it Multiprocessing Shims} or {\it mpShims}, that quickly finds good solutions for a wide range of problem sizes, our first contribution.
+
+We also carried out 126 experiments with each method on synthetic data based on real data from the {\it Brazilian Air Force} transportation history.
+
+Other important contributions are that we checked three hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure added to {\it Shims} algorithm enhanced it's results; but the integer programming procedure to minimize pallets' travel in the cargo bay did not improve the results.
+
+And another important contribution is that we employed the {\it iRace} package \citep{LopezIbanezManuel2016} to fine tune the {\it mpShims} volume threshold parameter, guaranteeing the best possible performance for the instances under training and testing.
 

### 2023-03-21 latex/comp_or.tex
Paper update

@@
 
-		{\bf This work}             & $\bigstar$ & $\bigstar$  &$\bigstar$ &$\bigstar$&$\bigstar$& $.$      & $.$      & $\bigstar$ & $\bigstar$&  $.$       &  $\bigstar$  \\
+		{\bf This work}             & $\bigstar$ & $\bigstar$  &$\bigstar$ &$\bigstar$&$\bigstar$& $.$      & $\bigstar$      & $\bigstar$ & $\bigstar$&  $.$       &  $\bigstar$  \\
 		\bottomrule 
@@
 
-In this work, we raise three hypotheses for improvements to the work of \cite{MesquitaSanches2023}:
+In this work, we raise four hypotheses for improvements to the work of \cite{MesquitaSanches2023}:
 (1) Computer parallelism could improve {\it Shims}'s performance.
 (2) The {\it Shims}' results may have been inferior to that; it could be that a sequential approach to pallet building may lead to a premature stop due to the torque constraints.
-(3) The distances to move pallets on unloading in the next node could be minimized to benefit unloading costs.
-
-So, this work seeks to improve their results in terms of quality and performance by using computer parallelism, CG balancing between {\it Shims}' phases, and minimizing the distance between the next node-destined pallets and the cargo door.
+(3) The distances to move pallets on unloading in the next node could be minimized to benefit ground handling costs.
+(4) A post-optimization procedure for 3D packing running in parallel mode would keep the run time adequate for operational use.
+
+So, this work seeks to improve their results in terms of quality and performance by using computer parallelism, CG balancing between {\it Shims}' phases, minimizing the distance between the next node-destined pallets and the cargo door, and performing a 3D packing procedure on each pallet.
 
 These improvements in their heuristics are meant to reduce the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets, and picking up and delivering at each node.
-To the best of our knowledge, that is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP and TSP has been addressed, and the first time ACLP+RPDP is solved with computer parallelism.
+To the best of our knowledge, that is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP, TSP, and 3D packing has been addressed, and the first time ACLP+RPDP is solved with computer parallelism.
 
@@
 
-	\item At each node of the flight plan, the items to be allocated are characterized by weight, volume, scores, and previously known destinations, but do not have dimensions.
-We leave the consideration of 2D or 3D items for future work.
-
-	\item We also disregarded {\it hazardous} items, which eventually could be treated as high score items and other specific constraints.
+	\item We disregarded {\it hazardous} items, which eventually could be treated as high score items and other specific constraints.
 
@@
 		& {\bf The pallets destined for the next node of the tour should be put as near as possible to the cargo ramp door (our inclusion)}.\\
+		& {\bf The items allocated to each pallet should be submitted to a 3D packing procedure (our inclusion)}.\\				
 		\midrule
@@
 {\color{blue}   \State	$G_3 \gets minRampDist(G_2)$  \label{tour:minRampDist} }
-
-			\State $s, \epsilon \gets ScoreAndDeviation(\pi(k), G_3)$ \label{tour:analyse}
+{\color{purple} \State	$G_4 \gets 3Dpacking(G_3)$  \label{tour:3dpacking} }
+
+			\State $s, \epsilon \gets ScoreAndDeviation(\pi(k), G_4)$ \label{tour:analyse}
 			\State $score \gets score + s$ \label{tour:score2}
@@
 
-		\State Let $G(M,\ Q_{\pi(k)},\ E_{\pi(k)})$ \label{mpShims:g1} \Comment{the initial solution solution}
+		\State Let $G(M,\ Q_{\pi(k)},\ E_{\pi(k)})$ \label{mpShims:g1} \Comment{the initial solution}
 
@@
 The best score shims edges will be returned.
+
+\subsection{3D packing}
+
+We implemented a post-optimization procedure for three-dimensional (3D) packing according to the instructions on \cite{DubeKanavathy2006} and on \cite{PaquayLimbourgSchynsOliveira2018}, to pack the items selected for each pallet.
+
+Our implementation was also based on \cite{RuizRibeiro}, taking advantage of the already implemented Python classes and auxiliary procedures.
+
+The heuristic solutions consider that, theoretically, all items selected for the pallets would fit on that pallet.
+But, due to the variability in item dimensions, an actual fit is not possible, causing some unfit items to be excluded from the previous heuristic's solutions (line \ref{3dpacking:unfit}).
+
+Another important contribution is that we implemented parallel 3D packing with each pallet as a process.
+
+\begin{algorithm}[H]
+	\caption{Parallel 3D packing procedure}  \label{alg:3dpacking}
+
+	\begin{algorithmic}[1]
+
+		\Procedure{$3Dpacking$}{${N_{\pi(k)}, \pi(k)}, limit, G$}
+
+		\State Let $G(M,\ Q_{\pi(k)},\ E_{\pi(k)})$ \label{3dpacking:g1} \Comment{the initial solution}
+
+		\State $proc \gets \{p_1,p_2,..., p_m\}$  \Comment{each pallet has its own process}
+
+		\For{$i \gets 1$ to $m$}
+		\State $proc_i \gets 3Dpacker(p_i,\ N_{\pi(k)},\ \pi(k),\ G )$   \label{3dpacking:greedy}
+		\State $proc_i.start$
+		\EndFor
+
+		\For{$i \gets 1$ to $m$}
+		\State $proc_i.join$   \Comment{wait for the packing process to finish.}
+		\State $E_{\pi(k)} \gets E_{\pi(k)} \setminus proc_i.unfit$  \label{3dpacking:unfit} \Comment{exclude the unfit item from solution}
+		\EndFor
+
+		\Return $G(M,\ N_{\pi(k)} \cup Q_{\pi(k)},\ E_{\pi(k)})$ \label{3dpacking:return}
+
+		\EndProcedure
+
+	\end{algorithmic}
+\end{algorithm}
 
@@
 
-In this work, we raised three hypotheses for improvements to the work of \cite{MesquitaSanches2023}:
+In this work, we raised four hypotheses for improvements to the work of \cite{MesquitaSanches2023}:
 
@@
 
+(4) The post-optimization procedure for 3-D packing generated an unfit item percentage of approximately 20\%.
+This solution phase was a parallel 3-D packing with each pallet as a process, which decreased this phase's run time by an $n$ factor where $n$ is the number of pallets.
+This was possible because each packing process is completely independent; there is no multiprocessing shared memory.
+It is also important to highlight that, after the items were excluded, the CG deviation was kept within its operational and feasible range.
+
 \section{Conclusions}
@@
 
-We converted \cite{MesquitaSanches2023} {\it Shims} into a process-based parallel computing heuristic, which we called {\it Multiprocessing Shims} or {\it mpShims}, that quickly finds good solutions for a wide range of problem sizes, our first contribution.
+We converted {\it Shims} into a process-based parallel computing heuristic, which we called {\it Multiprocessing Shims} or {\it mpShims}, that quickly finds good solutions for a wide range of problem sizes, our first contribution.
 
@@
 
-Other important contributions are that we checked three hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure added to {\it Shims} algorithm enhanced it's results; but the integer programming procedure to minimize pallets' travel in the cargo bay did not improve the results.
+Other important contributions are that we checked four hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure added to {\it Shims} algorithm enhanced it's results; but the integer programming procedure to minimize pallets' travel in the cargo bay did not improve the results; and a post-optimization procedure for 3-D packing was turned possible for operational use due to a process-based parallel computing heuristic.
 

### 2023-03-21 py3dbp/main.py

@@ -1,0 +1,218 @@
+from .constants import RotationType, Axis
+from .auxiliary_methods import intersect, set_to_decimal
+
+DEFAULT_NUMBER_OF_DECIMALS = 3
+START_POSITION = [0, 0, 0]
+
+
+class Item:
+    def __init__(self, name, width, height, depth, weight, id):
+        self.name = name
+        self.ID = id
+        self.width = width
+        self.height = height
+        self.depth = depth
+        self.weight = weight
+        self.rotation_type = 0
+        self.position = START_POSITION
+        self.number_of_decimals = DEFAULT_NUMBER_OF_DECIMALS
+
+    def format_numbers(self, number_of_decimals):
+        self.width = set_to_decimal(self.width, number_of_decimals)
+        self.height = set_to_decimal(self.height, number_of_decimals)
+        self.depth = set_to_decimal(self.depth, number_of_decimals)
+        self.weight = set_to_decimal(self.weight, number_of_decimals)
+        self.number_of_decimals = number_of_decimals
+
+    def string(self):
+        return "%s(%sx%sx%s, weight: %s) pos(%s) rt(%s) vol(%s)" % (
+            self.name, self.width, self.height, self.depth, self.weight,
+            self.position, self.rotation_type, self.get_volume()
+        )
+
+    def get_volume(self):
+        return set_to_decimal(
+            self.width * self.height * self.depth, self.number_of_decimals
+        )
+
+    def get_dimension(self):
+        if self.rotation_type == RotationType.RT_WHD:
+            dimension = [self.width, self.height, self.depth]
+        elif self.rotation_type == RotationType.RT_HWD:
+            dimension = [self.height, self.width, self.depth]
+        elif self.rotation_type == RotationType.RT_HDW:
+            dimension = [self.height, self.depth, self.width]
+        elif self.rotation_type == RotationType.RT_DHW:
+            dimension = [self.depth, self.height, self.width]
+        elif self.rotation_type == RotationType.RT_DWH:
+            dimension = [self.depth, self.width, self.height]
+        elif self.rotation_type == RotationType.RT_WDH:
+            dimension = [self.width, self.depth, self.height]
+        else:
+            dimension = []
+
+        return dimension
+
+
+class Bin:
+    def __init__(self, name, width, height, depth, max_weight, id):
+        self.name = name
+        self.ID = id
+        self.width = width
+        self.height = height
+        self.depth = depth
[…]

### 2023-03-21 latex/comp_or.tex
Comp OR paper update

@@
 We converted {\it Shims} into a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it surpassed all {\it Shims} results.
+We also solved a three-dimensional packing problem in the item-pallet allocations yielded by {\it Shims}.
 
@@
 \begin{keyword}
-Air Cargo \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Parallel computing \sep iRace
+Air Cargo \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Parallel computing \sep 3D packing
 \end{keyword}
@@
 
-In this work, we raise four hypotheses for improvements to the work of \cite{MesquitaSanches2023}:
+In this work, we raise four hypotheses for improvements to the work of \cite{MesquitaSanches2023}: \label{hypotheses}
 (1) Computer parallelism could improve {\it Shims}'s performance.
@@
 
+The final experiments are performed on a 64-bit, 16GiB, 3.4GHz, 8-core Intel® i7-3770 CPU, 2 threads per core, with {\it Linux Ubuntu 22.04 LTS} as the operational system and {\it Python 3.10.6} as the programming language.
+
 \subsection{Results obtained}
-
-The final experiments are performed on a 64-bit, 16GiB, 3.6GHz, 8-core AMD® Fx-8320e processor, 2 threads per core, with {\it Linux Ubuntu 20} as the operational system and {\it Python 3.8.2} as the programming language.
 
@@
 Inadequacies in the design of the statistical experiment may therefore lead to erroneous results from the statistical tests and, subsequently, cause the approach to produce inaccurate comparisons of candidate configurations.
-That is why all parameters were set using the $iRace$ package \citep{LopezIbanezManuel2016}.
+That is why all parameters were set using the $iRace$ package \cite{LopezIbanezManuel2016}.
 Supplying $iRace$ with a range for $volThreshold$ as [0.80 to 0.99], {\it iRace} yielded 0.92, by training on the uneven instances numbers.
@@
 
-According to \cite[p.271]{Breshears2009}, a {\it Process} is the operating system’s spawned and controlled entity that encapsulates an executing application.
+According to \citet[p.271]{Breshears2009}, a {\it Process} is the operating system’s spawned and controlled entity that encapsulates an executing application.
 A process has two main jobs: the first is to hold the application's resources, and the second is to carry out the application's instructions.
@@
 It is known that working concurrently opens up synchronization issues.
-But the {\it Multi-processing} package (mp) offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
+But the {\it Multi-processing} package ({\it mp}) offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
 Because of this, the {\it Multi-processing} module lets the programmer take full advantage of the fact that a machine has more than one processor, normally capable of 2 threads each.
@@
 Essentially, the law states that while a process can be decomposed into steps that may then be run in parallel, the time taken for the whole process will be significantly limited by the steps that remain serialized.
+In our case, the reading and writing process on the shared torque variable and the list of items to be embarked.
 
@@
 
-We confirmed the hypothesis raised in line \ref{mpShims:minCGdev} of Algorithm \ref{alg:mpShims} that, in sequential mode, where one pallet is built at a time, torque constraints could be more restrictive than volume or weight constraints, which could stop the palletization process earlier, fostering a solution that may be not as good as it could be.
+We confirmed the hypothesis raised in \ref{hypotheses} and in line \ref{mpShims:minCGdev} of Algorithm \ref{alg:mpShims} that, in sequential mode, where one pallet is built at a time, torque constraints could be more restrictive than volume or weight constraints, which could stop the palletization process earlier, yielding a solution that may be not as good as it could be.
 This integer programming procedure inclusion really improved all {\it Shims} results.
@@
 
-\subsection{Another possible improvement}
+\subsection{An attempt to minimize ground handling time}
 \label{improvement}
@@
 
-In this work, we raised four hypotheses for improvements to the work of \cite{MesquitaSanches2023}:
+In this work, we raised four hypotheses for improvements to the work of \cite{MesquitaSanches2023}: \label{hypotheses}
 

### 2023-03-21 common.py

@@ -64,9 +64,9 @@ class Item(object):
             # generate dimensions randomly
             w = random.uniform(0.5, 1.5) * v ** 1/3 # width
             l = random.uniform(0.5, 1.5) * math.sqrt(v/w) # length
-            self.w = w  # width
-            self.l = l # length
-            self.h = v /(w * l) # height
+            self.w = 0.95*w  # width
+            self.l = 0.95*l # length
+            self.h = 0.95*v /(w * l) # height
 
         self.ID = id
         self.P  = p  # -1 if an item, -2 if a consollidated, or pallet ID.
@@ -90,7 +90,7 @@ class Pallet(object):
         self.PCS = 0.
         self.w = 2.15 # width 84"
         self.l = 2.65 # length 104"
-        self.h = 1.7 * self.V /(self.w * self.l) # height
+        self.h = 1.5 * self.V /(self.w * self.l) # height
 
     def reset(self, numNodes):
         self.Dest = np.full(numNodes, -1)

### 2023-03-22 latex/comp_or.tex
Paper Comp OR update

--- 2023-03-21 latex/comp_or.tex
@@
 
-We converted {\it Shims} into a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it surpassed all {\it Shims} results.
+We converted {\it Shims} into a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes.
+In addition, we applied a 3D packing procedure to guarantee that only the items that fit on each pallet are present in the solution.
 
@@
 (3) The distances to move pallets on unloading in the next node could be minimized to benefit ground handling costs.
-(4) A post-optimization procedure for 3D packing running in parallel mode would keep the run time adequate for operational use.
+(4) A post-optimization procedure for 3D packing running in parallel mode would keep the running time adequate for operational use.
 
@@
 As can be seen, except for \cite{MesquitaSanches2023}, the remaining works do not address air cargo palletization and load balancing with route optimization in a multi-leg flight plan, and none of them employ parallelization in their algorithms.
-This is the objective of our work: to reshape \cite{MesquitaSanches2023} solution process and algorithms to accommodate parallel features that can improve solution quality and performance, and also include a new feature that is to minimize the distances between the next node's destined pallets and the cargo ramp door.
+This is the objective of our work: to reshape \cite{MesquitaSanches2023} solution process and algorithms to accommodate parallel features that can improve solution quality and performance and also include a new feature that is to minimize the distances between the next node's destined pallets and the cargo ramp door.
+We also intend to apply a 3D packing procedure to each pallet to guarantee that only the items that fit on each pallet are present in the solution.
 
@@
 
+{\color{blue}
 In line \ref{tour:minRampDist}, $minRampDist$ minimizes the distance of the next node destined pallets, keeping the CG in its operational range.
-This precedure is implemented in integer programming as presented in Subsection \ref{improvement}, according to equations \ref{eq:distobj} to \ref{eq:disttorque}.
+This procedure is implemented in integer programming as presented in Subsection \ref{improvement}, according to equations \ref{eq:distobj} to \ref{eq:disttorque}.
+}
+
+{\color{purple}
+In line \ref{tour:3dpacking}, we reshape the solution by selecting the items that actually fit into the pallet, as the previous result ($G_3$) is an upper bound.
+i.e., although the sum of volumes is less than the volumetric capacity of the pallet, it is not guaranteed that all items will fit on the pallet.
+}
 
@@
 		\EndIf
-		\EndFor
-
-		\Return $G(V_{\pi(k)}, E^N_{\pi(k)} \cup E^Q_{\pi(k)}),\ \eta_1$ \Comment{return the partial solution and the last item indexes array }
+		\EndFor \Comment{return the partial solution and the last item indexes array }
+
+		\Return $G(V_{\pi(k)}, E^N_{\pi(k)} \cup E^Q_{\pi(k)}),\ \eta_1$
 
@@
 
-Our implementation was also based on \cite{RuizRibeiro}, taking advantage of the already implemented Python classes and auxiliary procedures.
+Let $p_i.H$ be the maximum height for each pallet to be built, which depends on the size of the aircraft and the pallet position along the cargo bay; each {\it 463l} pallet has a width $p_i.W = 2.15m$ and a length of $p_i.L = 2.65m$.
+
+\begin{equation} \label{eq:3dof}
+	\mbox{minimize}\ \Bigg [ p_i.W \times p_i.H \times p_i.L - \sum_{j \in p_i} t_j.w \times t_j.h \times t_j.l \ \Bigg ]
+\end{equation}
+
+\begin{equation} \label{eq:3dh}
+	s.t.: \sum_{j \in p_i} t_j.h \leq p_i.H 
+\end{equation}
+
+\begin{equation} \label{eq:3dw}
+	s.t.: \sum_{j \in p_i} t_j.w \leq p_i.W 
+\end{equation}
+
+\begin{equation} \label{eq:3dl}
+	s.t.: \sum_{j \in p_i} t_j.l \leq p_i.L 
+\end{equation}
+
+Our implementation was also based on \cite{RuizRibeiro} (who implemented the algorithm described in \cite{DubeKanavathy2006}), taking advantage of the already implemented Python classes and auxiliary procedures, but making the necessary adjustments for parallel processing.
 
@@
 Another important contribution is that we implemented parallel 3D packing with each pallet as a process.
+
+Bin packing being an NP-Hard problem suggests that an exhaustive search for the optimal solution is in general computationally intractable and that there is thus no known computationally feasible optimal solution method for the problem.
+So other means to obtain a solution must be found. Normally, heuristic solution methods are used.
+
+The two primary bin packing approximate algorithms that we used were the {\it First Fit Decreasing - FFD} and the {\it Best Fit - BF}.
+They were chosen over other approximate algorithms because they have a faster running time and produce solutions that are much closer to the optimal solution than most other heuristic algorithms.
+
+The running time for BF and FFD are $O(n\ Log\ n)$, excluding the running time for sorting.
+
+The $3Dpacker$ function specification referred in Algorithm \ref{alg:3dpacking} is well described in \citet{DubeKanavathy2006}.
+This method returns the list of packed items, as well the list of unfit items to be excluded from solution.
 
@@
 		\State $proc_i.join$   \Comment{wait for the packing process to finish.}
-		\State $E_{\pi(k)} \gets E_{\pi(k)} \setminus proc_i.unfit$  \label{3dpacking:unfit} \Comment{exclude the unfit item from solution}
+		\State $E_{\pi(k)} \gets E_{\pi(k)} \setminus proc_i.unfit$  \label{3dpacking:unfit} \Comment{exclude the unfit items from solution}
 		\EndFor
@@
 In generating the items in each node, we consider 3 values for the parameter ${\it surplus}$.
-The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:results} (${\it surplus} = 1.2$, ${\it surplus} = 1.5$, and ${\it surplus} = 2.0$).
+The results obtained for the function $f$, with the corresponding running time in seconds, are shown in Tables \ref{tab:results} (${\it surplus} = 1.2$, ${\it surplus} = 1.5$, and ${\it surplus} = 2.0$).
 
@@
 
-The average tour values were presented for $f$ and, for the run time, the worst result obtained.
+The average tour values were presented for $f$ and, for the running time, the worst result obtained.
 To facilitate the comparison between the methods, we added a last column in these tables, where two values are indicated:
@@
 (4) The post-optimization procedure for 3-D packing generated an unfit item percentage of approximately 20\%.
-This solution phase was a parallel 3-D packing with each pallet as a process, which decreased this phase's run time by an $n$ factor where $n$ is the number of pallets.
+This solution phase was a parallel 3-D packing with each pallet as a process, which decreased this phase's running time by approximately an $n-1$ factor where $n$ is the number of pallets.
 This was possible because each packing process is completely independent; there is no multiprocessing shared memory.
@@
 
-Other important contributions are that we checked four hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure added to {\it Shims} algorithm enhanced it's results; but the integer programming procedure to minimize pallets' travel in the cargo bay did not improve the results; and a post-optimization procedure for 3-D packing was turned possible for operational use due to a process-based parallel computing heuristic.
+Other important contributions are that we checked four hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure for CG deviation minimization added to {\it Shims} algorithm enhanced it's results; but the integer programming procedure to minimize pallets' travel in the cargo bay did not improve the results; and the post-optimization procedure for 3-D packing was turned possible for operational use due to the approximation algorithms chosen and a process-based parallel computing, having each pallet as a process.
 

### 2023-03-22 latex/comp_or.tex
Paper Comp OR update

@@
 		& The total weight must be less than the aircraft payload or the total pallet capacity, whichever is the lowest.\\	
-		& {\bf The pallets destined for the next node of the tour should be put as near as possible to the cargo ramp door (our inclusion)}.\\
-		& {\bf The items allocated to each pallet should be submitted to a 3D packing procedure (our inclusion)}.\\				
+		& {\bf The pallets destined for the next node of the tour should be put as near as possible to the cargo ramp door (this work)}.\\
+		& {\bf The items allocated to each pallet should be submitted to a 3D packing procedure (this work)}.\\				
 		\midrule

### 2023-03-22 latex/comp_or.tex
Paper Comp OR updated

@@
 
-Let $p_i.H$ be the maximum height for each pallet to be built, which depends on the size of the aircraft and the pallet position along the cargo bay; each {\it 463l} pallet has a width $p_i.W = 2.15m$ and a length of $p_i.L = 2.65m$.
+Let $p_i.H$ be the maximum height for each pallet to be built, which depends on the size of the aircraft and the pallet position along the cargo bay; each {\it 463l} pallet has a width $p_i.W = 2.15m$ and a depth of $p_i.D = 2.65m$.
 
 \begin{equation} \label{eq:3dof}
-	\mbox{minimize}\ \Bigg [ p_i.W \times p_i.H \times p_i.L - \sum_{j \in p_i} t_j.w \times t_j.h \times t_j.l \ \Bigg ]
+	\mbox{minimize}\ \Bigg [ p_i.W \times p_i.H \times p_i.D - \sum_{j \in p_i} t_j.w \times t_j.h \times t_j.d \ \Bigg ]
 \end{equation}
@@
 \begin{equation} \label{eq:3dl}
-	s.t.: \sum_{j \in p_i} t_j.l \leq p_i.L 
+	s.t.: \sum_{j \in p_i} t_j.d \leq p_i.D 
 \end{equation}

### 2023-03-22 common.py

@@ -57,16 +57,16 @@ class Item(object):
     def __init__(self, id, p, w, s, v, frm, to):
 
         self.w = 0.  # width
-        self.l = 0. # length
+        self.d = 0. # length
         self.h = 0. # height
 
         if v > 0:
             # generate dimensions randomly
             w = random.uniform(0.5, 1.5) * v ** 1/3 # width
-            l = random.uniform(0.5, 1.5) * math.sqrt(v/w) # length
+            d = random.uniform(0.5, 1.5) * math.sqrt(v/w) # length
             self.w = 0.95*w  # width
-            self.l = 0.95*l # length
-            self.h = 0.95*v /(w * l) # height
+            self.d = 0.95*d  # depth
+            self.h = 0.95*v /(w * d) # height
 
         self.ID = id
         self.P  = p  # -1 if an item, -2 if a consollidated, or pallet ID.
@@ -89,8 +89,8 @@ class Pallet(object):
         self.PCV = 0.
         self.PCS = 0.
         self.w = 2.15 # width 84"
-        self.l = 2.65 # length 104"
-        self.h = 1.5 * self.V /(self.w * self.l) # height
+        self.d = 2.65 # depth 104"
+        self.h = 1.5 * self.V /(self.w * self.d) # height
 
     def reset(self, numNodes):
         self.Dest = np.full(numNodes, -1)
@@ -152,7 +152,7 @@ def loadPallets(cfg):
     dists = [8.39,6.25,4.5,2.1,-0.3,-2.7,-5.1] # distances of pallets centroids to the center of gravity
 
     if cfg.size == "larger":
-        vol = 14.8 # ramp                                                                        forward
+        vol = 14.8
         dists = [14.89,14.89,11.47,11.47,8.77,8.77,4.40,4.40,0.00,0.00,-4.40,-4.40,-8.77,-8.77,-13.17,-13.17,-17.57,-17.57]
    
     pallets = []

### 2023-03-22 main.py

@@ -175,12 +175,14 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
         for i, row in enumerate(Y):
 
             packers[i] = Packer()
-            packers[i].add_bin( Bin(f'pallet{i}', pallets[i].w, pallets[i].h, pallets[i].l, pallets[i].W, i) )
+            packers[i].add_bin( Bin(f'pallet{i}', pallets[i].w, pallets[i].h, pallets[i].d, pallets[i].W, i) )
             
+            pos = 0
             for j, X_ij in enumerate(row):
-                if X_ij:
-                    packers[i].add_item(Item(f'item{j}', items[j].w, items[j].h, items[j].l, items[j].W, j))
+                if X_ij == 1:
+                    packers[i].add_item(Item(f'item{j}', items[j].w, items[j].h, items[j].d, items[j].W, j))
                     counter1 += 1
+                    pos += 1
             
             procs[i] = mp.Process( target=packers[i].pack() )
             procs[i].start()
@@ -189,7 +191,6 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
         for i, proc in enumerate(procs):
             proc.join()
             for bin in packers[i].bins:
-                i = bin.ID
                 for item in bin.unfitted_items:
                     j = item.ID
                     Y[i][j] = 0
@@ -304,8 +305,8 @@ if __name__ == "__main__":
     # scenarios = [1,2,3,4,5,6]
     scenarios = [1] # infeasible solutions with Shims............
 
-    surplus   = "data20"
-    # surplus   = "data50"
+    # surplus   = "data20"
+    surplus   = "data50"
     # surplus   = "data100"
 
     # methods = ["Shims","mpShims","GRB"]  

### 2023-03-23 latex/comp_or.tex
Paper Comp OR update

--- 2023-03-22 latex/comp_or.tex
@@
 
+\usepackage{siunitx}
+
 %----- end celio
@@
 By leveraging parallelism, handheld computers can divide tasks into numerous smaller sections and execute these parts simultaneously, so utilizing multiple cores of the processor and offering a substantial performance improvement.
-This enables for more complicated applications and speedier performance, making mobile computers more competent than ever before.
+This enables for more complicated applications and speedier execution, making mobile computers more competent than ever before.
 
@@
 
-We also contributed by carrying out multiple experiments with the {\it Shims} heuristic on synthetic data based on real data from the {\it Brazilian Air Force} transportation history and a new procedure to minimize the distance from the next node destined pallet to the ramp door.
+We also contributed by carrying out multiple experiments with the multi-processing heuristics on synthetic data based on real data from the {\it Brazilian Air Force} transportation history and a new procedure to minimize the distance from the next node destined pallet to the ramp door.
 
 We converted {\it Shims} into a process-based parallel computing heuristic that quickly finds good solutions for a wide range of problem sizes.
-In addition, we applied a 3D packing procedure to guarantee that only the items that fit on each pallet are present in the solution.
+We also solved a three-dimensional packing problem in the item-pallet allocations yielded by {\it Shims}, which can guarantee that the palletization plan will include sets of items that actually fit into the pallets.
 
@@
 Air cargo transport involves several sub-problems that are difficult to solve.
-Recently, \cite{MesquitaSanches2023} modeled and solved the {\it Air Cargo Load Planning with Routing, Pickup and Delivery} (ACLP+RPDP) composed by four sub-problems: {\it Build-up Scheduling Problem}, {\it Air Palletization Problem} (APP), {\it Weight and Balance Problem} (WBP), and a special case of the {\it Traveling Salesman Problem}, similar to the proposed by \cite{kaspi2019}, where the profit per unit time is defined as the profit gained after completing the tour divided by the total time (cost in our case) required to complete the tour.
+Recently, \cite{MesquitaSanches2023} modeled and solved the {\it Air Cargo Load Planning with Routing, Pickup and Delivery} (ACLP+RPDP) composed by four sub-problems: {\it Build-up Scheduling Problem}, {\it Air Palletization Problem} (APP), {\it Weight and Balance Problem} (WBP), and a special case of the {\it Traveling Salesman Problem}, similar to the proposed by \cite{kaspi2019}, where the profit per unit time is defined as the gain after completing the tour divided by the total time (cost in our case) required to complete the tour.
 
@@
 Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a pre-defined flight plan restricted to two legs.
-It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs ({\bf MLU}) at the intermediate node was modelled through a container sequencing problem.
+It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs ({\bf MLU}) at the intermediate node was modeled through a container sequencing problem.
 Referring directly to this work, \cite[p.
@@
 \end{table}
+\label{hypotheses}
 
@@
 (2) The {\it Shims}' results may have been inferior to that; it could be that a sequential approach to pallet building may lead to a premature stop due to the torque constraints.
-(3) The distances to move pallets on unloading in the next node could be minimized to benefit ground handling costs.
-(4) A post-optimization procedure for 3D packing running in parallel mode would keep the running time adequate for operational use.
+(3) A procedure for 3D packing running in parallel mode would keep the run time adequate for operational use.
+(4) The distances to move pallets on unloading in the next node could be minimized to benefit ground handling costs.
 
@@
 \cite{Chan2006} carried out a case study with heterogeneous pallets.
-In order to minimise the total cost of shipping, they developed a 3D packing heuristic, with a loading plan for each pallet.
+In order to minimize the total cost of shipping, they developed a 3D packing heuristic, with a loading plan for each pallet.
 Although the authors do not consider load balancing or positioning of pallets in the cargo hold, this method is relevant in commercial and industrial applications, where cargo items tend to be less dense.
@@
 \cite{PaquaySchynsLimbourg2016} presented a mathematical modeling to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
-Its objective is to maximise the volume used in containers, considering load balancing constraints, the presence of fragile items and the possibility of rotating these boxes.
+Its objective is to maximize the volume used in containers, considering load balancing constraints, the presence of fragile items and the possibility of rotating these boxes.
 \cite{PaquayLimbourgSchynsOliveira2018} developed some heuristics to solve this problem.
 
-\cite{YangLiuGao2018} modelled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing in order to minimise fuel consumption.
+\cite{YangLiuGao2018} modeled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing in order to minimize fuel consumption.
 
@@
 
-\cite{zhao2021} proposed a new modelling for WBP based on mixed integer programming.
+\cite{zhao2021} proposed a new model for WBP based on mixed integer programming.
 Instead of focusing on the center of gravity (CG) deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
@@
 They also create a new heuristic called {\it Shims} which is fast and may be run in a handheld computer.
-They did not use computer parallelism to improve performance, nor tryed to minimize the distances from the next nodes destined pallets to the cargo ramp door.
+They did not use multi-core parallelism to improve performance, nor tried to minimize the distances from the next nodes destined pallets to the cargo ramp door, nor tried to perform a 3D packing to keep in solution only items that fit in the pallets usable volume.
 
 As can be seen, except for \cite{MesquitaSanches2023}, the remaining works do not address air cargo palletization and load balancing with route optimization in a multi-leg flight plan, and none of them employ parallelization in their algorithms.
-This is the objective of our work: to reshape \cite{MesquitaSanches2023} solution process and algorithms to accommodate parallel features that can improve solution quality and performance and also include a new feature that is to minimize the distances between the next node's destined pallets and the cargo ramp door.
-We also intend to apply a 3D packing procedure to each pallet to guarantee that only the items that fit on each pallet are present in the solution.
+This is the objective of our work: to reshape \cite{MesquitaSanches2023} solution process and algorithms to accommodate parallel features that can improve solution quality and performance, include a new feature that is to minimize the distances between the next node's destined pallets and the cargo ramp door, and also include a 3D packing procedure to guarantee that only items that fit into the pallets usable volume are planned to palletization.
 
@@
 It is the primary air cargo pallet for more than 70 Air Forces and many air transport companies.
-This pallet has a capacity of $4500 kg$ and $14.8 m^3$, is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$.
+This pallet has a capacity of $4500 kg$ and $14.8 m^3$ (depending on the aircraft ceiling height), is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$.
 For more information, see {\tt www.463LPallet.com}.
@@
 A pallet which has not yet reached its destination may receive more items, although it is known that these operations of removing restraining nets increase handling time and the risk of improper delivery.
-We do not consider oversized cargo in this work, but only cargoes that fit on these pallets.
-
-	\item Pallets with destinations set to the next node should be put as near as possible to the cargo ramp door.
-
-	\item Finally, as we are interested in minimizing fuel costs while keeping the CG in its operational range, we disregarded some lower ones as handling costs.
+We do not consider over-sized cargo in this work, but only cargoes that fit on these pallets.
 
@@
 
-\section{The mathematical modelling}
+\section{The mathematical modeling}
 \label{sec4}
 
-Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modelling of ACLP+RPDP.
+Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modeling of ACLP+RPDP.
 
@@
 
-Let $S_K = \{s: \{1, \dots, K\} \rightarrow \{1, \dots, K\} \}$ be the set of $K!$ permutations, which correspond to all possible tours (or itineraries) that have $l_0$ as origin and end, passing through the other $K$ nodes.
-{\color{blue} $k$ is the node index and $\pi(k)$ is the node in position $k$ of tour $\pi$}.
-
-Let $M = \{p_1, p_2, \ldots, p_m \}$ the set of $m$ pallets.
-Each pallet $p_i$, $1 \leq i \leq m$, has weight capacity $p_i.w$, volume capacity $p_i.v$, pallet destinations $p_i.to[{\pi(k)}]$, $0 \leq k \leq K$, and distance to the CG of aircraft $p_i.d$.
-$p_i.to[{\pi(k)}]$ denotes that pallet $p_i$ may assume a different destination in each node ${\pi(k)}$.
-
-Let $N_{\pi(k)} = \{t^{\pi(k)}_1, t^{\pi(k)}_2, \ldots, t^{\pi(k)}_{n_{\pi(k)}} \}$ be the set of $n_{\pi(k)}$ items to be loaded in node $l_{\pi(k)}$, $0 \leq k \leq K$.
-Each item $t^{\pi(k)}_j$, $1 \leq j \leq n_{\pi(k)}$, has score $t^{\pi(k)}_j.s$, weight $t^{\pi(k)}_j.w$, volume $t^{\pi(k)}_j.v$, and destination $t^{\pi(k)}_j.to \in L_{\pi(k)}$.
-Let $N = \bigcup_{0 \leq k \leq K} N_{\pi(k)}$ be the set of items of all nodes along a tour.
-
-Let $Q_{\pi(k)} = \{a^{\pi(k)}_1, a^{\pi(k)}_2, \ldots, a^{\pi(k)}_{m_{\pi(k)}} \}$ be the set of consolidated items loaded in $m_k \leq m$ pallets when the aircraft arrives at node $l_{\pi(k)}$, with $0 \leq k \leq K$.
-$a^{\pi(k)}_i$, $1 \leq i \leq m_{\pi(k)}$, is the group of picked-up items that were allocated on pallet $p_i$ in some of the previous nodes.
-$a^{\pi(k)}_i$ has total weight $a^{\pi(k)}_i.w$, total volume $a^{\pi(k)}_i.v$, and destination $a^{\pi(k)}_i.to \in L_{\pi(k)} \cup \{l_{\pi(k)}\}$.
-If $a^{\pi(k)}_i.to = l_{\pi(k)}$, then $a^{\pi(k)}_i$ is unloaded, and $p_i$ will be available for reloading; otherwise, $a^{\pi(k)}_i$ remains on the aircraft, eventually in another pallet, and with items of $N_{\pi(k)}$ having the same destination.
-
-Let $X_{ij}^{\pi(k)}$ and $Y_{iq}^{\pi(k)}$ be binary variables, where $0 \leq k \leq K$, $1 \leq j \leq n_{\pi(k)}$, $1 \leq i \leq m$ and $1 \leq q \leq m_{\pi(k)}$.
-$X_{ij}^{\pi(k)} = 1$ if $t_j^{\pi(k)}$ is assigned to $p_i$ in node $l_{\pi(k)}$, and 0 otherwise.
-$Y_{iq}^{\pi(k)} = 1$ if $a_q^{\pi(k)}$ is assigned to $p_i$ in node $l_{\pi(k)}$, and 0 otherwise.
-By definition, $Y_{iq}^0=0$.
-Allocations of items or consolidated items to pallets in node $l_{\pi(k)}$ can be seen as a bipartite graph $G_{\pi(k)}(V_{\pi(k)}, E_{\pi(k)})$, where $V_{\pi(k)} = M \cup N_{\pi(k)} \cup Q_{\pi(k)}$, $E_{\pi(k)} = E^N_{\pi(k)} \cup E^Q_{\pi(k)}$, $(p_i, t_j^{\pi(k)}) \in E^N_{\pi(k)}$ if $X_{ij}^{\pi(k)} = 1$, and $(p_i, a_q^{\pi(k)}) \in E^Q_{\pi(k)}$ if $Y_{iq}^{\pi(k)} = 1$.
+Let $S_K = \{s: \{1, \dots, K\} \rightarrow \{1, \dots, K\} \}$ be the set of $K!$ permutations {\color{blue} indexed by $\pi$}, which correspond to all possible tours (or itineraries) that have $l_0$ as origin and end, passing through the other $K$ nodes.
+
+{\color{blue} $k$ is the node index and $\pi(k)$ is the node in position $k$ of tour $\pi$. $\tau^{\pi(k)}$ is the total torque applied by the loaded pallets to the aircraft in this node. }
+
+Let $M^{\pi(k)} = \{p_1, p_2, \ldots, p_m \}$ the set of $m$ pallets.
+Each pallet $p_i$, $1 \leq i \leq m$, has weight capacity $p_i.w$, volume capacity $p_i.v$, pallet destinations $p_i.to^{\pi(k)}$, $0 \leq k \leq K$, and distance to the CG of aircraft $p_i.d$.
+$p_i.to^{\pi(k)}$ denotes that pallet $p_i$ may assume a different destination in each node ${\pi(k)}$.
+
+Let $N^{\pi(k)} = \{t^{\pi(k)}_1, t^{\pi(k)}_2, \ldots, t^{\pi(k)}_{n^{\pi(k)}} \}$ be the set of $n^{\pi(k)}$ items to be loaded in node $l^{\pi(k)}$, $0 \leq k \leq K$.
+Each item $t^{\pi(k)}_j$, $1 \leq j \leq n^{\pi(k)}$, has score $t^{\pi(k)}_j.s$, weight $t^{\pi(k)}_j.w$, volume $t^{\pi(k)}_j.v$, and destination $t^{\pi(k)}_j.to \in L^{\pi(k)}$.
+Let $N = \bigcup_{0 \leq k \leq K} N^{\pi(k)}$ be the set of items of all nodes along a tour.
+
+Let $Q^{\pi(k)} = \{a^{\pi(k)}_1, a^{\pi(k)}_2, \ldots, a^{\pi(k)}_{m} \}$ be the set of consolidated items loaded in $m_k \leq m$ pallets when the aircraft arrives at node $l^{\pi(k)}$, with $0 \leq k \leq K$.
+$a^{\pi(k)}_i$, $1 \leq i \leq m$, is the group of picked-up items that were allocated on pallet $p_i$ in some of the previous nodes.
+$a^{\pi(k)}_i$ has total weight $a^{\pi(k)}_i.w$, total volume $a^{\pi(k)}_i.v$, and destination $a^{\pi(k)}_i.to \in L^{\pi(k)} \cup \{l^{\pi(k)}\}$.
+If $a^{\pi(k)}_i.to = l^{\pi(k)}$, then $a^{\pi(k)}_i$ is unloaded, and $p_i$ will be available for reloading; otherwise, $a^{\pi(k)}_i$ remains on the aircraft, eventually in another pallet, and with items of $N^{\pi(k)}$ having the same destination.
+
+Let $X_{ij}^{\pi(k)}$ and $Y_{iq}^{\pi(k)}$ be binary variables, where $0 \leq k \leq K$, $1 \leq j \leq n^{\pi(k)}$, $1 \leq i \leq m$ and $1 \leq q \leq m$.
+$X_{ij}^{\pi(k)} = 1$ if $t_j^{\pi(k)}$ is assigned to $p_i$ in node $l^{\pi(k)}$, and 0 otherwise.
+
+$Y_{iq}^{\pi(k)} = 1$ if $a_q^{\pi(k)}$ is assigned to $p_i$ in node $l^{\pi(k)}$, and 0 otherwise. By definition, $Y_{iq}^0=0$.
+
+Allocations of items or consolidated to pallets in node $l^{\pi(k)}$ can be seen as a bipartite graph $G^{\pi(k)}(V^{\pi(k)}, E^{\pi(k)})$,
+where $V^{\pi(k)} = M^{\pi(k)} \cup N^{\pi(k)} \cup Q^{\pi(k)}$, $E^{\pi(k)} = E_N^{\pi(k)} \cup E_Q^{\pi(k)}$,
+$(p_i, t_j^{\pi(k)}) \in E_N^{\pi(k)}$
+if $X_{ij}^{\pi(k)} = 1$, and $(p_i, a_q^{\pi(k)}) \in E_Q^{\pi(k)}$ if $Y_{iq}^{\pi(k)} = 1$.
 
@@
 \begin{equation} \label{eq:scores}
-	\tilde{s} = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.s
+	\tilde{s} = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.s
 \end{equation}
@@
 \begin{equation} \label{eq:costs}
-	\tilde{c} = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} [ c_{\pi(k), \pi(k+1)}\times(1+c_g\times|\epsilon_{\pi(k)}|) ] + c_{\pi(K),0}\times(1+c_g\times|\epsilon_K|)
+	\tilde{c} = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} [ c^{\pi(k), \pi(k+1)}\times(1+c_g\times|\epsilon^{\pi(k)}|) ] + c^{\pi(K),0}\times(1+c_g\times|\epsilon_K|)
 \end{equation}
@@
 \begin{equation} \label{eq:tau}
-\tau_{\pi(k)} = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w +  \sum_{q=1}^{m} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w)];\ k \in \{0, 1, \ldots, K\}
+\tau^{\pi(k)} = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w +  \sum_{q=1}^{m} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w)];\ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:eps}
-\epsilon_{\pi(k)} = \frac{\tau_{\pi(k)}}{maxW \times limit^{CG}_{long}};\ k \in \{0, 1, \ldots, K\}
+\epsilon^{\pi(k)} = \frac{\tau^{\pi(k)}}{maxW \times limit^{CG}_{long}};\ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp11}
-	L_0 = L; L_{\pi(k)} = L_{{\pi(k-1)}} - \{l_{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
+	L_0 = L; L^{\pi(k)} = L^{\pi(k-1)} - \{l^{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp12}
-	X_{ij}^{\pi(k)} = 0 \ \mbox{if} \ t_j^{\pi(k)}.to \notin L_{\pi(k)}; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_{\pi(k)}\}; \ k \in \{1,2, \ldots, K\}
+	X_{ij}^{\pi(k)} = 0 \ \mbox{if} \ t_j^{\pi(k)}.to \notin L^{\pi(k)}; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi(k)}\}; \ k \in \{1,2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp9}
-	Y_{iq}^{\pi(k)} = 0 \ \mbox{if} \ a_i^{\pi(k)}.to \notin L_{\pi(k)}; \ i \in \{1, 2, \ldots, m\}; \ q \in \{1, 2, \ldots, m_{\pi(k)}\}; \ k \in \{ 1, 2, \ldots, K\}
+	Y_{iq}^{\pi(k)} = 0 \ \mbox{if} \ a_i^{\pi(k)}.to \notin L^{\pi(k)}; \ i \in \{1, 2, \ldots, m\}; \ q \in \{1, 2, \ldots, m\}; \ k \in \{ 1, 2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:cons2}
-	a_i^{\pi(k+1)}.w = \sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_{\pi(k)}} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w;  \ i \in \{1, 2, \ldots, m_{\pi(k)}\}; k \in \{0, 1, \ldots, K-1\}
+	a_i^{\pi(k+1)}.w = \sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w;  \ i \in \{1, 2, \ldots, m\}; k \in \{0, 1, \ldots, K-1\}
 \end{equation}
@@
 \begin{equation} \label{eq:cons3}
-	a_i^{\pi(k+1)}.v = \sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.v + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.v;  \ i \in \{1, 2, \ldots, m_{\pi(k)}\}; k \in \{0, 1, \ldots, K-1\}
+	a_i^{\pi(k+1)}.v = \sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.v + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.v;  \ i \in \{1, 2, \ldots, m\}; k \in \{0, 1, \ldots, K-1\}
 \end{equation}
@@
 \begin{equation} \label{eq:cons5}
-	a_i^{\pi(k+1)}.to = t^{\pi(k)}_j.to  \ \mbox{if} \ X^{\pi(k)}_{ij} = 1  \ \mbox{and} \ t^{\pi(k)}_j.to \in L_{\pi(k+1)};  \ i \in \{1, 2, \ldots, m_{\pi(k)}\}; \ j \in \{1, 2, \ldots, n_{\pi(k)}\}; \ k \in \{0,1, \ldots, K-1\}
+	a_i^{\pi(k+1)}.to = t^{\pi(k)}_j.to  \ \mbox{if} \ X^{\pi(k)}_{ij} = 1  \ \mbox{and} \ t^{\pi(k)}_j.to \in L_{\pi(k+1)};  \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi(k)}\}; \ k \in \{0,1, \ldots, K-1\}
 \end{equation}
@@
 \begin{equation} \label{eq:LatIt}
-	LatIt_{\pi(k)} = \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi(k)}} ( X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w \times (i\%2) - X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w \times (i+1)\%2 ); \ k \in \{0, 1, \ldots, K\}
+	LatIt^{\pi(k)} = \sum_{i=1}^{m} \sum_{j=1}^{n^{\pi(k)}} ( X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w \times (i\%2) - X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w \times (i+1)\%2 ); \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:LatCons}
-	LatCons_{\pi(k)} =  \sum_{i=1}^{m} \sum_{q=1}^{m_{\pi(k)}}  ( Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w \times (i\%2) - Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w \times (i+1)\%2); \ k \in \{0, 1, \ldots, K\}
+	LatCons^{\pi(k)} =  \sum_{i=1}^{m} \sum_{q=1}^{m}  ( Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w \times (i\%2) - Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w \times (i+1)\%2); \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:torqlat}
-	s.t.: d^{CG}_{pallet} \times | LatIt_{\pi(k)} + LatCons_{\pi(k)} | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: d^{CG}_{pallet} \times | LatIt^{\pi(k)} + LatCons^{\pi(k)} | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:torqlong}
-	s.t.: |\tau_{\pi(k)}| \leq maxW \times limit^{CG}_{long};\ k \in \{0, 1, \ldots, K\}
+	s.t.: |\tau^{\pi(k)}| \leq maxW \times limit^{CG}_{long};\ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:payload}
-	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^k.w ) \leq maxW; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^k.w ) \leq maxW; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app2}
-	s.t.: \sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m_{\pi(k)}\}; \ {\pi(k)} \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m\}; \ {\pi(k)} \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app3}
-	s.t.: \sum_{j=1}^{n_{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.v + \sum_{q=1}^{m_{\pi(k)}} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m_{\pi(k)}\}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.v + \sum_{q=1}^{m} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m\}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app4}
-	s.t.: \sum_{i=1}^{m} X_{ij}^{\pi(k)} \leq 1; \ j \in \{1, 2, \ldots, n_{\pi(k)}\}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{i=1}^{m} X_{ij}^{\pi(k)} \leq 1; \ j \in \{1, 2, \ldots, n^{\pi(k)}\}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app5}
-	s.t.:  Y_{iq}^{\pi(k)} = 1 \ \mbox{if} \ a^{\pi(k)}_q.to \in L_{\pi(k)}; \ q \in \{1, 2, \ldots, m_{\pi(k)}\}; \ k \in \{0, 1, \ldots, K\}
+	s.t.:  Y_{iq}^{\pi(k)} = 1 \ \mbox{if} \ a^{\pi(k)}_q.to \in L^{\pi(k)}; \ q \in \{1, 2, \ldots, m\}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp8}
-	s.t.: p_i.to[{\pi(k)}] = t^{\pi(k)}_j.to\ \mbox{if} \ X_{ij}^{\pi(k)} = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
+	s.t.: p_i.to^{\pi(k)} = t^{\pi(k)}_j.to\ \mbox{if} \ X_{ij}^{\pi(k)} = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp2}
-	s.t.:  p_i.to[{\pi(k)}] = a^{\pi(k)}_q.to\ \mbox{if} \ Y_{iq}^{\pi(k)} = 1; \ i \in \{1, 2, \ldots, m\};\ q \in \{1, 2, \ldots, m_{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
+	s.t.:  p_i.to^{\pi(k)} = a^{\pi(k)}_q.to\ \mbox{if} \ Y_{iq}^{\pi(k)} = 1; \ i \in \{1, 2, \ldots, m\};\ q \in \{1, 2, \ldots, m\}; \ k \in \{1, 2, \ldots, K\}
 \end{equation}
@@
 The objective of this problem is to find a permutation $\pi \in S_K$\/ that maximizes the function $f_\pi(\tilde{s},\tilde{c})$ \ref{eq:maxf}.
-In this way, the flight plan will be $l_0, l_{\pi(1)}, \ldots, l_{\pi(K)}, l_0$.
+In this way, the flight plan will be $l_0, l_{\pi(1)}, \ldots, l^{\pi(K)}, l_0$.
 $\tilde{s}$\/ is the total score of transported items \ref{eq:scores} and  $\tilde{c}$\/ is the total cost of fuel consumed \ref{eq:costs}.
@@
 
-The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $l_{\pi(k)}$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n_{\pi(k)}} t^{\pi(k)}_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.v$).
+The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $l^{\pi(k)}$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n^{\pi(k)}} t^{\pi(k)}_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.v$).
 This parameter is passed to $ItemsGeneration$ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section.
@@
 The loop of lines \ref{main:loop1}-\ref{main:loop2} goes through all permutations $\pi$, where the node-by-node resolutions are performed by $SolveTour$, whose result is stored in $f_{\pi}$.
-The best result among all $K!$\/ tours will be the answer for $scenario$, $volume$\/ and $method$\/ (line \ref{main:f}).
+The best outcome among all $K!$\/ tours will be the answer for $scenario$, $volume$\/ and $method$\/ (line \ref{main:f}).
 
@@
 				\State $L_k \gets L_k - \pi(k)$  \label{tour:lk1}			
-				\State $Q_{\pi(k)}, E^Q_{\pi(k)}, M \gets UpdateConsolidated(\pi(k))$ \label{tour:dest}			
-				\State Let $G_1(M \cup N_{\pi(k)} \cup Q_{\pi(k)}, E^Q_{\pi(k)})$ \label{tour:g12}
+				\State $Q^{\pi(k)}, E_Q^{\pi(k)}, M^{\pi(k)} \gets UpdateConsolidated(\pi(k))$ \label{tour:dest}			
+				\State Let $G_1(M^{\pi(k)} \cup N^{\pi(k)} \cup Q^{\pi(k)}, E_Q^{\pi(k)})$ \label{tour:g12}
 			\EndIf  \label{tour:lk2}	
-			\State $M \gets SetPalletsDestination( \pi(k) )$ \label{tour:dest2}		
+			\State $M^{\pi(k)} \gets SetPalletsDestination( \pi(k) )$ \label{tour:dest2}		
 			\State $G_2 \gets SolveNode(method, \pi(k), G_1, timeLim)$ \label{tour:node}
 
-{\color{blue}   \State	$G_3 \gets minRampDist(G_2)$  \label{tour:minRampDist} }
-{\color{purple} \State	$G_4 \gets 3Dpacking(G_3)$  \label{tour:3dpacking} }
+{\color{purple} \State	$G_3 \gets 3Dpacking(G_2)$  \label{tour:3dpacking} }
+
+{\color{blue}   \State	$G_4 \gets minRampDist(G_3)$  \label{tour:minRampDist} }
 
@@
 
-{\color{blue}
 In line \ref{tour:minRampDist}, $minRampDist$ minimizes the distance of the next node destined pallets, keeping the CG in its operational range.
 This procedure is implemented in integer programming as presented in Subsection \ref{improvement}, according to equations \ref{eq:distobj} to \ref{eq:disttorque}.
-}
-
-{\color{purple}
-In line \ref{tour:3dpacking}, we reshape the solution by selecting the items that actually fit into the pallet, as the previous result ($G_3$) is an upper bound.
-i.e., although the sum of volumes is less than the volumetric capacity of the pallet, it is not guaranteed that all items will fit on the pallet.
-}
 
@@
 
-In this subsection we present implementations of $SolveNode(...)$, where $method$\/ corresponds to a heuristic, $k$\/ is the index of the current node $l_{\pi(k)}$ and $G$\/ is the allocation graph of the consolidated items that remain on board at $l_{\pi(k)}$.
+In this subsection we present implementations of $SolveNode(...)$, where $method$\/ corresponds to a heuristic, $k$\/ is the index of the current node $l^{\pi(k)}$ and $G$\/ is the allocation graph of the consolidated items that remain on board at $l^{\pi(k)}$.
 
@@
 
-		{\it Common procedures:}  & All methods have the same procedures for: solution printing, pallets destinations and initial parameters setting, solution integrity checking, and the same procedures for selecting and inserting edges.\\
+		{\it Common procedures:}  & All methods have the same procedures for: solution printing, pallets destinations and initial parameters setting, result integrity checking, and the same procedures for selecting and inserting edges.\\
 
@@
 
-The selection of edges for $E^N_{\pi(k)}$\/ uses the {\it edge attractiveness}\/ $\theta^{\pi(k)}_{ij}$, Equation \ref{eq:edge}, which can be understood as the tendency to allocate the item $t^{\pi(k)}_j$\/ to the pallet $p_i$.
-It is directly proportional to the score, and inversely proportional to the volume and the torque of each item.
+The selection of edges for $E_N^{\pi(k)}$\/ uses the {\it edge attractiveness}\/ $\theta^{\pi(k)}_{ij}$, Equation \ref{eq:edge}, which can be understood as the tendency to allocate the item $t^{\pi(k)}_j$\/ to the pallet $p_i$.
+It is directly proportional to the score, and inversely relative to the volume and torque of each item.
 
@@
 
-In the pseudo-codes presented below, given an allocation graph $G$\/ for the node $l_{\pi(k)}$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation({\pi(k)},G) $, described in \cite{MesquitaSanches2023}.
+In the pseudo-codes presented below, given an allocation graph $G$\/ for the node $l^{\pi(k)}$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation({\pi(k)},G) $, described in \cite{MesquitaSanches2023}.
 
@@
 
-		\captionof{figure}{$n_{\pi(k)}$\/ edges $e_{ij}^{\pi(k)}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi(k)}$\/ in non-ascending order}
+		\captionof{figure}{$n^{\pi(k)}$\/ edges $e_{ij}^{\pi(k)}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi(k)}$\/ in non-ascending order}
 		\label{fig:whip}		
@@
 
-Figure \ref{fig:whip} represents the $n_{\pi(k)}$\/ possible edges $e_{ij}^{\pi(k)}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi(k)}$.
-\emph{Shims}\/ starts with a greedy solution, stopping at the edge with index in $\eta_1[i]$ close to the local optimum (first phase).
-Then, considering even the edge with the index $\eta_2[i]$, it elaborates different possible complements for this pallet and selects the best of these complements (second phase).
-
-Although the second phase solves a knapsack problem, an algorithm for solving a Bin Packing Problem was used (the First Fit Decreasing - FFD), and the most profitable bin (Shims) was selected.
-This method was proved to be faster than a Dynamic Program solution for the Knapsack Problem.
+Figure \ref{fig:whip} represents the $n^{\pi(k)}$\/ possible edges $e_{ij}^{\pi(k)}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi(k)}$.
+\emph{mpShims}\/ starts with a greedy solution for each pallet, stopping at the edge with index in $\eta_1[i]$ close to the local optimum (first phase).
+Then, performs an integer programming procedure to minimize the CG deviation (second phase), and finally, considering even the edge with the index $\eta_2[i]$, it elaborates different possible complements for this pallet and selects the best of these complements (third phase).
+
+Although the third phase solves a {\it Knapsack Problem}, a faster algorithm for solving a {\it Bin Packing Problem} was used ({\it the First Fit Decreasing - FFD}), and the most profitable bin ({\it Shims}) was selected.
+This method was proved to be faster than a {\it Dynamic Programming} solution for the {\it Knapsack Problem}.
 
@@
 
-All executed in a multi-process execution, accessing concurrently, reading and writing, the list of items and the current value of the CG deviation.
-As to the multi-process execution, we considered each pallet as a process.
+All executed in a multi-process run-time, accessing concurrently, reading and writing the list of items and the current value of the CG deviation.
+As to the multi-process run-time, we considered each pallet as a parallel process, taking advantage of nowadays multi-core computers.
 
@@
 
-		\Procedure{$mpShims$}{${N_{\pi(k)}, \pi(k)}, limit, G$}
-
-		\State Let $G(M,\ Q_{\pi(k)},\ E_{\pi(k)})$ \label{mpShims:g1} \Comment{the initial solution}
+		\Procedure{$mpShims$}{${N^{\pi(k)}, \pi(k)}, limit, G$}
+
+		\State Let $G(M^{\pi(k)},\ Q^{\pi(k)},\ E^{\pi(k)})$ \label{mpShims:g1} \Comment{the initial solution}
 
@@
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets Greedy(p_i,\ N_{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ limit,\ lock,\ ts )$   \label{mpShims:greedy}
+		\State $proc_i \gets Greedy(p_i,\ N^{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ limit,\ lock,\ ts )$   \label{mpShims:greedy}
 		\State $proc_i.start$
@@
 
-		\State $minCGdev(M,\ \pi(k),\ nodeTorque)$  \label{mpShims:minCGdev} \Comment{minimize the CG deviation.}
+		\State $minCGdev(M^{\pi(k)},\ \pi(k),\ nodeTorque)$  \label{mpShims:minCGdev} \Comment{minimize the CG deviation.}
 
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets getBestShims(p_i,\ N_{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ surplus,\ lock,\ ts )$ \label{mpShims:getBestShims}
+		\State $proc_i \gets getBestShims(p_i,\ N^{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ surplus,\ lock,\ ts )$ \label{mpShims:getBestShims}
 		\State $proc_i.start$
@@
 
-		\Return $G(M,\ N_{\pi(k)} \cup Q_{\pi(k)},\ E_{\pi(k)})$ \label{mpShims:return}
+		\Return $G(M^{\pi(k)},\ N^{\pi(k)} \cup Q^{\pi(k)},\ E^{\pi(k)})$ \label{mpShims:return}
 
@@
 
-Line \ref{mpShims:greedy} calls the greedy method which updates $p_i$, $N_{\pi(k)}$, $nodeTorque$, and $G$.
+Line \ref{mpShims:greedy} calls the greedy method which updates $p_i$, $N^{\pi(k)}$, $nodeTorque$, and $G$.
 
@@
 
-Line \ref{mpShims:getBestShims} calls the $getBestShims$ method which updates $p_i$, $N_{\pi(k)}$, $nodeTorque$, and $G$.
+Line \ref{mpShims:getBestShims} calls the $getBestShims$ method which updates $p_i$, $N^{\pi(k)}$, $nodeTorque$, and $G$.
 
@@
 
-Algorithm \ref{alg:greedy} generates a greedy allocation of the items available in node $l_{\pi(k)}$, according to the non-ascending order of $\theta^{\pi(k)}_{ij}$, and considering the consolidated items already shipped ($E^Q_{\pi(k)}$).
+Algorithm \ref{alg:greedy} generates a greedy allocation of the items available in node ${\pi(k)}$, according to the non-ascending order of $\theta^{\pi(k)}_{ij}$, and considering the consolidated items already shipped ($E_Q^{\pi(k)}$).
 Edges are included in this allocation as long as they respect feasibility constraints.
@@
 
-		\Procedure{$Greedy$}{${\pi(k)}, G, limit$}
-
-		\State Let $G(V_{\pi(k)}, E^Q_{\pi(k)})$
+		\Procedure{$Greedy$}{${\pi(k)},\ G,\ limit,\ ts$}
+
+		\State Let $G(V^{\pi(k)}, E_Q^{\pi(k)})$
 
@@
 
-		\State $E^N_{\pi(k)} \gets \varnothing$
+		\State $E_N^{\pi(k)} \gets \varnothing$
 
 		\For{$q \gets 1$ to $m$}
-		\If{$(p_i, a_q^{\pi(k)}) \in E^Q_{\pi(k)}$} 
-		\State $volume[q] \gets volume[q] + a_q^{\pi(k)}.v$ 
+		\If{$(p_i, a_q^{\pi(k)}) \in E_Q^{\pi(k)}$} 
+		\State $volume_q \gets volume_q + a_q^{\pi(k)}.v$ 
 		\EndIf		
@@
 
+		\State $\tau_{max} \gets maxW \times limit^{CG}_{long}$
+
 		\For {each $e_{ij}^{\pi(k)}$ in non-ascending order of $\theta^{\pi(k)}_{ij}$}
-		\If{($E^N_{\pi(k)} \cup \{e_{ij}^{\pi(k)}\}$ is feasible) {\bf and} ($volume[i] \leq p_i.v \times limit$)} 
-		\State $E^N_{\pi(k)} \gets E^N_{\pi(k)} \cup \{e_{ij}^{\pi(k)}\}$ 
-		\State $volume[i] \gets volume[i] + t_j^{\pi(k)}.v$
-		\State $\eta_1[i] \gets \eta_1[i] + 1$ 	
+
+		\State $\Delta\tau \gets t_j.w \times p_i.d$
+
+		\State $\tau_{new} \gets | \tau^{\pi(k)} + \Delta\tau |$
+
+		\If{ ($E_N^{\pi(k)} \cup \{e_{ij}^{\pi(k)}\}$ is feasible) {\bf and} ($volume_i \leq p_i.v \times limit$) {\bf and} ($\tau_{new} \leq \tau_{max} \times ts$ )} \label{greedy:ts}
+
+			\State $E_N^{\pi(k)} \gets E_N^{\pi(k)} \cup \{e_{ij}^{\pi(k)}\}$ 
+			\State $volume_i \gets volume_i + t_j^{\pi(k)}.v$
+			\State $\eta_1^i \gets \eta_1^i + 1$ 
+			\State $\tau^{\pi(k)} \gets \tau^{\pi(k)} + \Delta\tau$
+
 		\EndIf
@@
 
-		\Return $G(V_{\pi(k)}, E^N_{\pi(k)} \cup E^Q_{\pi(k)}),\ \eta_1$
+		\Return $G(V^{\pi(k)}, E_N^{\pi(k)} \cup E_Q^{\pi(k)}),\ \eta_1$
 
@@
 The set $\eta_1$ is returned together with the greedy solution.
+
+Line \ref{greedy:ts} presents the variable $ts$ (torque $\tau$ surplus), that is an extra-torque limit to allow each greedy pallet construction to, temporally, exceed the torque limit, that will be turned feasible again by the next {\it mpShims}' phase, the CG minimization phase.
+$ts$ value will be adjusted empirically to produce feasible solutions in all problem instances.
 
@@
 		\State $b \gets 1$
-		\State $shims[b] \gets \{\}$  \label{empty_shims}
-		\State $Set \gets Set \cup \{shims[b]\}$ \label{empty_set}
-
-		\For{$x \gets \eta_1[i]$ to $\eta_2[i]$} \label{edges_indexes}
+		\State $shims_b \gets \{\}$  \label{empty_shims}
+		\State $Set \gets Set \cup \{shims_b\}$ \label{empty_set}
+
+		\For{$x \gets \eta_1^i$ to $\eta_2^i$} \label{edges_indexes}
 
 		\State $NewShims \gets$ {\bf True} \label{new_shims}
-		\State $e_{ij}^k \gets E[i][x]$
+		\State $e_{ij}^k \gets E_x^i$
 
@@
 
-		\If { $e_{ij}^k \not\in (E^N_k \cup shims)$ {\bf and} $e_{ij}^k$ is feasible  {\bf and}  $(t_j^k.v + volume) \leq slack[i]$}
+		\If { $e_{ij}^k \not\in (E^N_k \cup shims)$ {\bf and} $e_{ij}^k$ is feasible  {\bf and}  $(t_j^k.v + volume) \leq slack^i$}
 
@@
 		\State $b \gets b + 1$
-		\State $shims[b] \gets \{\}$
-		\State $shims[b] \gets shims[b] \cup \{e_{ij}^k\}$
-		\State $Set \gets Set \cup \{shims[b]\}$
+		\State $shims_b \gets \{\}$
+		\State $shims_b \gets shims_b \cup \{e_{ij}^k\}$
+		\State $Set \gets Set \cup \{shims_b\}$
 		\EndIf
@@
 
-We implemented a post-optimization procedure for three-dimensional (3D) packing according to the instructions on \cite{DubeKanavathy2006} and on \cite{PaquayLimbourgSchynsOliveira2018}, to pack the items selected for each pallet.
-
-Let $p_i.H$ be the maximum height for each pallet to be built, which depends on the size of the aircraft and the pallet position along the cargo bay; each {\it 463l} pallet has a width $p_i.W = 2.15m$ and a length of $p_i.L = 2.65m$.
-
-\begin{equation} \label{eq:3dof}
-	\mbox{minimize}\ \Bigg [ p_i.W \times p_i.H \times p_i.L - \sum_{j \in p_i} t_j.w \times t_j.h \times t_j.l \ \Bigg ]
-\end{equation}
-
-\begin{equation} \label{eq:3dh}
-	s.t.: \sum_{j \in p_i} t_j.h \leq p_i.H 
-\end{equation}
-
-\begin{equation} \label{eq:3dw}
-	s.t.: \sum_{j \in p_i} t_j.w \leq p_i.W 
-\end{equation}
-
-\begin{equation} \label{eq:3dl}
-	s.t.: \sum_{j \in p_i} t_j.l \leq p_i.L 
-\end{equation}
-
-Our implementation was also based on \cite{RuizRibeiro} (who implemented the algorithm described in \cite{DubeKanavathy2006}), taking advantage of the already implemented Python classes and auxiliary procedures, but making the necessary adjustments for parallel processing.
+We implemented a procedure for three-dimensional (3D) packing according to the instructions on \cite{DubeKanavathy2006}, on \cite{PaquaySchynsLimbourg2016}, and on \cite{PaquayLimbourgSchynsOliveira2018}, to pack the items selected for each pallet.
+
+Our implementation was also based on \cite{RuizRibeiro}, taking advantage of the already implemented Python classes and auxiliary procedures, and the needed improvements to multi-core parallelism.
 
@@
 
-Bin packing being an NP-Hard problem suggests that an exhaustive search for the optimal solution is in general computationally intractable and that there is thus no known computationally feasible optimal solution method for the problem.
-So other means to obtain a solution must be found. Normally, heuristic solution methods are used.
-
-The two primary bin packing approximate algorithms that we used were the {\it First Fit Decreasing - FFD} and the {\it Best Fit - BF}.
-They were chosen over other approximate algorithms because they have a faster running time and produce solutions that are much closer to the optimal solution than most other heuristic algorithms.
-
-The running time for BF and FFD are $O(n\ Log\ n)$, excluding the running time for sorting.
-
-The $3Dpacker$ function specification referred in Algorithm \ref{alg:3dpacking} is well described in \citet{DubeKanavathy2006}.
-This method returns the list of packed items, as well the list of unfit items to be excluded from solution.
-
 \begin{algorithm}[H]
@@
 
-		\Procedure{$3Dpacking$}{${N_{\pi(k)}, \pi(k)}, limit, G$}
-
-		\State Let $G(M,\ Q_{\pi(k)},\ E_{\pi(k)})$ \label{3dpacking:g1} \Comment{the initial solution}
+		\Procedure{$mp3Dpacking$}{${N^{\pi(k)}, \pi(k)}, limit, G$} \Comment{multi-processing 3D packing}
+
+		\State Let $G(M^{\pi(k)},\ Q^{\pi(k)},\ E^{\pi(k)})$ \label{3dpacking:g1} \Comment{the initial solution}
 
@@
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets 3Dpacker(p_i,\ N_{\pi(k)},\ \pi(k),\ G )$   \label{3dpacking:greedy}
+		\State $proc_i \gets 3Dpacker(p_i,\ N^{\pi(k)}_i,\ \pi(k),\ G )$   \label{3dpacking:greedy}
 		\State $proc_i.start$
@@
 		\State $proc_i.join$   \Comment{wait for the packing process to finish.}
-		\State $E_{\pi(k)} \gets E_{\pi(k)} \setminus proc_i.unfit$  \label{3dpacking:unfit} \Comment{exclude the unfit items from solution}
+		\State $E^{\pi(k)}_i \gets E^{\pi(k)}_i \setminus proc_i.unfit$  \label{3dpacking:unfit} \Comment{exclude the unfit items from solution}
 		\EndFor
 
-		\Return $G(M,\ N_{\pi(k)} \cup Q_{\pi(k)},\ E_{\pi(k)})$ \label{3dpacking:return}
+		\Return $G(M^{\pi(k)},\ N^{\pi(k)} \cup Q^{\pi(k)},\ E^{\pi(k)})$ \label{3dpacking:return}
 
@@
 In generating the items in each node, we consider 3 values for the parameter ${\it surplus}$.
-The results obtained for the function $f$, with the corresponding running time in seconds, are shown in Tables \ref{tab:results} (${\it surplus} = 1.2$, ${\it surplus} = 1.5$, and ${\it surplus} = 2.0$).
+The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:results} (${\it surplus} = 1.2$, ${\it surplus} = 1.5$, and ${\it surplus} = 2.0$).
 
@@
 
-The average tour values were presented for $f$ and, for the running time, the worst result obtained.
+The average tour values were presented for $f$ and, for the run time, the worst result obtained.
 To facilitate the comparison between the methods, we added a last column in these tables, where two values are indicated:
@@
 
-So, to confirm or not Hypothesis (3), that the distances to move pallets on unloading in the next node could be minimized to benefit unloading costs. We modeled this improvement as follows.
+So, to confirm or not Hypothesis (4), that the distances to move pallets on unloading in the next node could be minimized to benefit unloading costs. We modeled this improvement as follows.
 
@@
 \begin{equation} \label{eq:distobj}
-	\mbox{minimize} \sum_{i=1}^{m} \sum_{q=1}^{m} Z^{\pi(k)}_{iq} \times ( maxD - p_i.d) ,\ \mbox{{\bf if}}\ a^{\pi(k)}_q.to = \pi(k+1) \}
+	\mbox{minimize} \sum_{i=1}^m \sum_{q=1}^m Z^{\pi(k)}_{iq} \times ( maxD - p_i.d) ,\ \mbox{{\bf if}}\ a^{\pi(k)}_q.to = \pi(k+1) \}
 \end{equation}
@@
 \begin{equation} \label{eq:distcons}
-	s.t.: \sum_{i=1}^{m} Z^{\pi(k)}_{iq} = 1;\ q \in \{1,2,\ldots,m\}
+	s.t.: \sum_{i=1}^m Z^{\pi(k)}_{iq} = 1;\ q \in \{1,2,\ldots,m\}
 \end{equation}
@@
 
-There was some expectation on this post-improvement method as it could have minimized the distances of travel of the pallets through the ramp door, but unfortunately, after running all scenarios and problem instances, there was no improvement in minimizing these distances.
-
-This proves that the solutions provided by {\it Shims} and {\it mpShims} are tight enough to hinder this intended improvement.
+\vspace{2.0mm}
+\begin{table}[!ht]
+\centering
+\caption{Distances minimization}  \label{tab:ground}
+\footnotesize
+%\scriptsize
+\begin{tabular}{cc S[table-format=-1.1] S[table-format=-1.1] S[table-format=-1.3]}
+\toprule
+{\bf Volume}            &                & {\bf Before} & {\bf After} & {\bf Improvement} \\
+{\bf Surplus}           & {\bf Scenario} & {\bf (m)}    & {\bf (m)}   & {\bf (\%)} \\		
+\midrule
+\multirow{6}{*}{ $1.2$} & $1$ &  31.0 &  29.0 & -5.9 \\
+                        & $2$ & 189.7 & 189.4 & -0.4 \\
+                        & $3$ & 152.0 & 152.0 & 0.0 \\
+                        & $4$ & 134.0 & 130.7 & -2.4 \\ 
+                        & $5$ & 115.7 & 115.7 & 0.0\\    
+                        & $6$ &  &  & 0.0\\                                                                  
+\midrule
+\multirow{6}{*}{ $1.5$} & $1$ &  .0 &  .0 & 5.9 \\
+						& $2$ & .7 & .4 & 0.4 \\
+						& $3$ & .0 & .0 & 0.0 \\
+						& $4$ & .0 & .7 & 2.4 \\ 
+						& $5$ & .7 & .7 & 0.0\\    
+						& $6$ &  &  & 0.0\\                                                                  
+\midrule
+\multirow{6}{*}{ $2.0$} & $1$ &  .0 &  .0 & 5.9 \\
+						& $2$ & .7 & .4 & 0.4 \\
+						& $3$ & .0 & .0 & 0.0 \\
+						& $4$ & .0 & .7 & 2.4 \\ 
+						& $5$ & .7 & .7 & 0.0\\    
+						& $6$ &  &  & 0.0\\
+
+\bottomrule
+\end{tabular}
+\end{table}
+
+Table \ref{tab:ground} presents the distances minimization of the next node destined pallets to the ramp door. Columns {\bf Before} and {\bf After} refer to the application of the current method.
+
+%There was some expectation on this post-improvement method as it could have minimized the distances of travel of the pallets through the ramp door, but unfortunately, after running all scenarios and problem instances, there was no improvement in minimizing these distances.
+
+%This proves that the solutions provided by {\it Shims} and {\it mpShims} are tight enough to hinder this intended improvement.
 
@@
 
-(3) The distances to move pallets on unloading in the next node could be minimized to benefit unloading costs.
-This hypothesis was not confirmed.
+(3) The procedure for 3-D packing generated an unfit item percentage of approximately 20\%.
+This solution phase was a parallel 3-D packing with each pallet as a process, which decreased this phase's run time by approximate a $n-1$ factor where $n$ is the number of pallets.
+This was possible because each packing process is completely independent; there is no multiprocessing shared memory.
+It is also important to highlight that, after the items were excluded, the CG deviation was kept within its operational and feasible range.
+
+(4) The distances to move pallets on unloading in the next node could be minimized to benefit ground handling costs.
+%This hypothesis was not confirmed.
 We implemented an integer programming procedure to minimize these distances, but the results proved that any change in {\it Shims} or {\it mpShims} outcomes could turn the solutions unfeasible due to the CG balance constraints.
@@
 
-(4) The post-optimization procedure for 3-D packing generated an unfit item percentage of approximately 20\%.
-This solution phase was a parallel 3-D packing with each pallet as a process, which decreased this phase's running time by approximately an $n-1$ factor where $n$ is the number of pallets.
-This was possible because each packing process is completely independent; there is no multiprocessing shared memory.
-It is also important to highlight that, after the items were excluded, the CG deviation was kept within its operational and feasible range.
-
 \section{Conclusions}
@@
 
-Other important contributions are that we checked four hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure for CG deviation minimization added to {\it Shims} algorithm enhanced it's results; but the integer programming procedure to minimize pallets' travel in the cargo bay did not improve the results; and the post-optimization procedure for 3-D packing was turned possible for operational use due to the approximation algorithms chosen and a process-based parallel computing, having each pallet as a process.
-
-And another important contribution is that we employed the {\it iRace} package \citep{LopezIbanezManuel2016} to fine tune the {\it mpShims} volume threshold parameter, guaranteeing the best possible performance for the instances under training and testing.
+Other important contributions are that we checked four hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure added to {\it Shims} algorithm enhanced it's results; but the integer programming procedure to minimize pallets' travel in the cargo bay did not improve the results; and a procedure for 3-D packing was turned possible for operational use due to a process-based parallel computing heuristic.
+
+And another important contribution is that we employed the {\it iRace} package to fine tune the {\it mpShims} volume threshold and torque surplus parameters, guaranteeing the best possible performance for the instances under training and testing.
 

### 2023-03-23 common.py

@@ -98,6 +98,17 @@ class Pallet(object):
         self.PCV = 0.
         self.PCS = 0.
 
+    def popItem(self, item, nodeTorque, solDict, N, itemsDict): # pop an item from this pallet
+        self.PCW -= item.W
+        self.PCV -= item.V
+        self.PCS -= item.S
+
+        nodeTorque.value -= float(item.W) * float(self.D)
+        i = self.ID
+        j = item.ID 
+        solDict["solMatrix"][N*i+j] = 0 # solution array
+        itemsDict["mpItems"][j]     = 0 # to control items inclusion
+
     def putItem(self, item, nodeTorque, solDict, N, itemsDict, lock): # put an item in this pallet
         self.PCW += item.W
         self.PCV += item.V

### 2023-03-23 main.py

@@ -146,10 +146,44 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
             numOptDict["numOpt"] += 1
 
 
+        nodeElapsed = time.perf_counter() - startNodeTime
+
+        Y = np.reshape(solDict["solMatrix"], (-1, N)) # N number of items (columns)
+       
         if method != "GRB":
 
-            # print(f"Torque before minRampDist: {nodeTorque.value/cfg.maxTorque:.2f}")
+            # begin ---- parallel solving the 3D packing for each pallet  
+            procs   = [None for _ in pallets]
+            packers = [None for _ in pallets]
+            counter1 = 0
+            for i, row in enumerate(Y):
 
+                packers[i] = Packer()
+                packers[i].add_bin( Bin(f'pallet{i}', pallets[i].w, pallets[i].h, pallets[i].l, pallets[i].W, i) )
+                
+                for j, X_ij in enumerate(row):
+                    if X_ij:
+                        packers[i].add_item(Item(f'item{j}', items[j].w, items[j].h, items[j].l, items[j].W, j))
+                        counter1 += 1
+                
+                procs[i] = mp.Process( target=packers[i].pack() )
+                procs[i].start()
+
+            counter2 = 0
+            for i, proc in enumerate(procs):
+                proc.join()
+                for bin in packers[i].bins:
+                    i = bin.ID
+                    for item in bin.unfitted_items:
+                        j = item.ID
+                        Y[i][j] = 0
+                        counter2 += 1
+                        pallets[i].popItem(items[j], nodeTorque, solDict, N, itemsDict)
+
+            print(f"{100*counter2/counter1:.1f}% unfit items excluded from solution!")
+            # end ---- parallel solving the 3D packing for each pallet         
+
+            # begin minRampDist
             for p in pallets:
                 if p.Dest[k] == next.ID:
                     beforeDict['Before'] += rampDistCG - p.D # distance from the pallet to the ramp door
@@ -159,44 +193,7 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
             for p in pallets:
                 if p.Dest[k] == next.ID:
                     afterDict['After'] += rampDistCG - p.D # distance from the pallet to the ramp door
-
[…]

### 2023-03-23 main.py
Paper Comp OR update

@@ -159,11 +159,11 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
             for i, row in enumerate(Y):
 
                 packers[i] = Packer()
-                packers[i].add_bin( Bin(f'pallet{i}', pallets[i].w, pallets[i].h, pallets[i].l, pallets[i].W, i) )
+                packers[i].add_bin( Bin(f'pallet{i}', pallets[i].w, pallets[i].h, pallets[i].d, pallets[i].W, i) )
                 
                 for j, X_ij in enumerate(row):
                     if X_ij:
-                        packers[i].add_item(Item(f'item{j}', items[j].w, items[j].h, items[j].l, items[j].W, j))
+                        packers[i].add_item(Item(f'item{j}', items[j].w, items[j].h, items[j].d, items[j].W, j))
                         counter1 += 1
                 
                 procs[i] = mp.Process( target=packers[i].pack() )
@@ -299,7 +299,7 @@ if __name__ == "__main__":
     plot = False
 
     # scenarios = [1,2,3,4,5,6]
-    scenarios = [5,6] # infeasible solutions with Shims............
+    scenarios = [6] # infeasible solutions with Shims............
 
     surplus   = "data20"
     # surplus   = "data50"

### 2023-03-24 latex/comp_or.tex
Paper Comp OR update

@@
 
-\subsection{mpShims}
-
 \begin{algorithm}[H]
@@
 
-\subsubsection{{\it Shims} first phase}
+\subsubsection{{\it mpShims} first phase}
 
@@
 
-\subsubsection{{\it Shims} second phase: improvement on minimizing CG deviation}
+\subsubsection{{\it mpShims} second phase}
 \label{subsection:second_phase}
 
-Between the greedy and the {\it Shims} selection phases we inserted, in this work, an integer programming method to minimize the CG deviation, as stated by the model that follows.
+Between the greedy and the {\it Shims} selection phases we inserted, in this work, an integer programming method to minimize the CG deviation, as stated by the integer programming model that follows.
 
@@
 
-\subsubsection{{\it Shims} third phase}
+\subsubsection{{\it mpShims} third phase}
 
@@
 
-\subsection{3D packing}
-
-We implemented a procedure for three-dimensional (3D) packing according to the instructions on \cite{DubeKanavathy2006}, on \cite{PaquaySchynsLimbourg2016}, and on \cite{PaquayLimbourgSchynsOliveira2018}, to pack the items selected for each pallet.
-
-Our implementation was also based on \cite{RuizRibeiro}, taking advantage of the already implemented Python classes and auxiliary procedures, and the needed improvements to multi-core parallelism.
-
-The heuristic solutions consider that, theoretically, all items selected for the pallets would fit on that pallet.
-But, due to the variability in item dimensions, an actual fit is not possible, causing some unfit items to be excluded from the previous heuristic's solutions (line \ref{3dpacking:unfit}).
-
-Another important contribution is that we implemented parallel 3D packing with each pallet as a process.
+\subsection{Multiprocessing 3D packing}
+
+According to \cite{BrandtStefan2019}, {\it Technically the task to solve is a three-dimensional Container Loading Problem (CLP) with a multitude of constraints.
+The CLP alone is known to be NP-hard and extremely hard to solve even for instances of moderate size.
+The typical number of items per transport segment ranges between 300 and 800 items.}
+
+So, to confirm or not Hypothesis (3), we decided to adopt the model for three-dimensional (3D) packing based in the instructions on \cite{DubeKanavathy2006}, \cite{li2014} and on \cite{PaquayLimbourgSchynsOliveira2018}, to pack the items selected for each pallet.
+
+The main logic of the 3D packing procedure is based on an approximate algorithm like follows:
+
+From a list of items, items are sorted from the biggest to the smallest and placed in such an order on a pallet.Each item has 1 to 6 orientations to choose from when it can be placed.
+The orientation procedure can select the best direction type among possible orientations.
+
+A pivot point is used to determine item's position. The pivot is an (x, y, z) coordinate which represents a point in the pallet at which an attempt to pack an item will be made.
+
+The back lower left corner of the item will be placed at the pivot.
+If the item cannot be packed at the pivot position then it is rotated until it can be packed at the pivot point or until we have tried all 6 possible rotation types.
+
+If, after rotating it, the item still cannot be packed at the pivot point, then we move on to packing another item and add the unpacked item to a list of items that will be packed after an attempt to pack the remaining items is made.
+
+The first pivot in the empty pallet is always (0,0,0). When one item can be placed into multiple optimal pivot point, the placement selection module can help make a choice.
+
+Let $n$ be the number of items and $m$ the number of pallets.
+
+Let $A_{j}\ =\ 1$ if item $j$ is assigned to the pallet, $A_{j}\ =\ 0$ otherwise, for $j \in \{1,\ 2,\ 3,\ ...,\ n\}$.
+
+Let $X_{ij}\ =\ 1$ if item $j$ was already assigned to pallet $i$ by {\it mpShims}, $X_{ij}\ =\ 0$ otherwise, for $i \in \{1,\ 2,\ 3,\ ...,\ m\}$.
+
+Let $d_j,\ w_j,\ h_j$ be parameters indicating the depth, width, and height of an item. 
+Let $D_i,\ W_i,\ H_i$ be parameters indicating the depth, width, and height of the pallet.
+
+Let $x_j,\ y_j,\ z_j$ be real type variables for coordinates of item's left-bottom-behind corner.
+
+Let $o_j^1,\ o_j^2,\ o_j^3,\ o_j^4,\ o_j^5,\ o_j^6$ be binary variables, indicating the possible orientations for item $j$.
+One of the values equal to 1 if item $j$ has orientation in $\{1,\ 2,\ 3,\ 4,\ 5,\ 6\}$, respectively, otherwise equal to 0.
+
+\begin{equation} \label{eq:3dpacking1}
+	\mbox{minimize} \Big [ D_i \times W_i \times H_i - \sum_{j \gets 1}^{n} X_{ij} \times A_{j} \times d_j \times w_j \times h_j \Big ]
+\end{equation}
+
+\begin{equation} \label{eq:3dpacking2}
+	\sum_{j \gets 1}^{n} A_{j} \times d_j  \leq X_{ij} \times D_i,\ i \in \{1,\ 2,\ 3,\ ...,\ m\}
+\end{equation}
+
+\begin{equation} \label{eq:3dpacking3}
+	\sum_{j \gets 1}^{n} A_{j} \times w_j  \leq X_{ij} \times W_i,\ i \in \{1,\ 2,\ 3,\ ...,\ m\}
+\end{equation}
+
+\begin{equation} \label{eq:3dpacking4}
+	\sum_{j \gets 1}^{n} A_{j} \times h_j  \leq X_{ij} \times H_i,\ i \in \{1,\ 2,\ 3,\ ...,\ m\}
+\end{equation}
+
+Algorithm \ref{alg:3dpacking} presents the multiprocessing solution method for 3D packing to be applied after {\it mpShims}, to guarantee that the item to be palletized actually fit into the pallet.
 
@@
 
-		\Procedure{$mp3Dpacking$}{${N^{\pi(k)}, \pi(k)}, limit, G$} \Comment{multi-processing 3D packing}
-
-		\State Let $G(M^{\pi(k)},\ Q^{\pi(k)},\ E^{\pi(k)})$ \label{3dpacking:g1} \Comment{the initial solution}
-
-		\State $proc \gets \{p_1,p_2,..., p_m\}$  \Comment{each pallet has its own process}
+		\Procedure{$mp3Dpacking$}{${\pi(k)},\ G$}
+
+		\State Let $G(M^{\pi(k)},\ N^{\pi(k)} \cup Q^{\pi(k)},\ E^{\pi(k)})$ \label{3dpacking:1}
+
+		\State $proc \gets \{p_1,p_2,..., p_m\}$ \label{3dpacking:2}
 
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets 3Dpacker(p_i,\ N^{\pi(k)}_i,\ \pi(k),\ G )$   \label{3dpacking:greedy}
+		\State $proc_i \gets 3Dpacker(p_i,\ N^{\pi(k)}_i,\ \pi(k),\ G )$   \label{3dpacking:3}
 		\State $proc_i.start$
@@
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i.join$   \Comment{wait for the packing process to finish.}
-		\State $E^{\pi(k)}_i \gets E^{\pi(k)}_i \setminus proc_i.unfit$  \label{3dpacking:unfit} \Comment{exclude the unfit items from solution}
+		\State $proc_i.join$ \label{3dpacking:4}
+		\State $E^{\pi(k)}_i \gets E^{\pi(k)}_i \setminus proc_i.unfit$  \label{3dpacking:5} 
 		\EndFor
 
-		\Return $G(M^{\pi(k)},\ N^{\pi(k)} \cup Q^{\pi(k)},\ E^{\pi(k)})$ \label{3dpacking:return}
+		\Return $G(M^{\pi(k)},\ N^{\pi(k)} \cup Q^{\pi(k)},\ E^{\pi(k)})$
 
@@
 \end{algorithm}
+
+According to line \ref{3dpacking:1}, this method receives an initial solution produced by {\it mpShims}, and according to line \ref{3dpacking:2}, each pallet has its own process.
+
+In line \ref{3dpacking:3}, $N^{\pi(k)}_i$ is the set of items previously assigned to pallet $i$.
+
+In line \ref{3dpacking:4}, the algorithm waits for the packing process to finish.
+
+Line \ref{3dpacking:5} excludes the unfit items from solution.
+
+\subsection{Minimization of the distance of a next node destined pallet to the ramp door }
+
+\cite{LurkinSchyns2015} considered an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem.
+But, as both aircraft dealt with in this work have only the ramp door, we tried to minimize the distance of travel of pallets inside the cargo bay, favoring unloading in the next node, by a post optimization of pallets positions to favor this distance minimization.
+
+So, to confirm or not Hypothesis (4), that the distances to move pallets on unloading in the next node could be minimized to benefit unloading costs.
+We modeled this improvement as an integer programming model.
+
+Let $maxD$ be the distance from the last ramp pallet to the center of gravity.
+
+Let $Q^{\pi(k)}  = \{ a^{\pi(k)}_1, a^{\pi(k)}_2, \ldots, a^{\pi(k)}_m \}$ be the set of consolidated items assembled on the set of $m$ pallets in node $\pi(k)$.
+
+Let $Z^{\pi(k)}_{iq}$ be the set of binary decision variables that relates pallet $i$ to consolidated $q$ in node $\pi(k)$.
+
+\begin{equation} \label{eq:distobj}
+	\mbox{minimize} \sum_{i=1}^m \sum_{q=1}^m Z^{\pi(k)}_{iq} \times ( maxD - p_i.d) ,\ \mbox{{\bf if}}\ a^{\pi(k)}_q.to = \pi(k+1) \}
+\end{equation}
+
+\begin{equation} \label{eq:distcons}
+	s.t.: \sum_{i=1}^m Z^{\pi(k)}_{iq} = 1;\ q \in \{1,2,\ldots,m\}
+\end{equation}
+
+\begin{equation} \label{eq:distpallets}
+	s.t.: \sum_{q=1}^{m} Z^{\pi(k)}_{iq}=1;\ i \in \{1,2,\ldots,m\}
+\end{equation}
+
+\begin{equation} \label{eq:distpalletw}
+	p_i.w = 140 + a^{\pi(k)}_q.w ;\ i \in \{1,2,\ldots,m\};\ q \in \{1,2,\ldots,m\}
+\end{equation}
+
+\begin{equation} \label{eq:disttorque}
+	s.t.:\ | \sum\ p_i.d * p_i.w\ | \leq maxW \times limit^{CG}_{long} ;\ i \in \{1,2,\ldots,m\}
+\end{equation}
+
+Equation \ref{eq:distobj} minimizes the sum of distances of the consolidated in node $\pi(k)$ destined to the next node $\pi(k+1)$; Equation \ref{eq:distcons} states that each consolidated will be assigned to exactly one pallet; Equation \ref{eq:distpallets} states that each pallet will receive one consolidated; Equation \ref{eq:distpalletw} calculates the total weight of each pallet; and Equation \ref{eq:disttorque} guarantees that the torque limit will not be exceeded.
 
@@
 		&\multirow{2}{*}{ $Shims$}                       & $f$                &     10.34 &     8.50  &     12.42 &     15.06 &     15.97 &     15.88 &  0.96 \\% 7817
-		&                                                & {\it run time (s)} &     < 1   &     3     &     5     &     20    &     58    &     352   &  1.0 \\% 439			
+		&                                                & {\it run time (s)} &     < 1   &     3     &     5     &     20    &     58    &     352   &  1.00 \\% 439			
 		\cmidrule{2-10}		                       
 		&\multirow{2}{*}{ $mpShims$}                     & $f$                &   12.96   &   7.53    &  12.28    &   13.20   &  13.71    &   13.44   &  0.90\\% 7278
-		&                                                & {\it run time (s)} &     < 1   &     2     &     2     &     8    &     22    &     120   &  {\bf 2.8}\\%
+		&                                                & {\it run time (s)} &     < 1   &     2     &     2     &     8    &     22    &     120   &  {\bf 2.80}\\%
 
@@
         &\multirow{2}{*}{ $Shims$}                       & $f$                &     14.68 &     12.21 &     17.67 &  21.01    &  22.64    &   21.91   &  0.96  \\%  11012
-        &                                                & {\it run time (s)} &     < 1   &     2     &     3     &     12    &     33    &   209     &  1.0 \\% 260
+        &                                                & {\it run time (s)} &     < 1   &     2     &     3     &     12    &     33    &   209     &  1.00 \\% 260
 
@@
         &\multirow{2}{*}{ $mpShims$}                     & $f$                &    16.12  &     11.81 &     17.48 &  17.66    &   19.81   &  17.61    &  0.87 \\% 9966
-        &                                                & {\it run time (s)} &     < 1   &     1     &     2     &     7     &   20      &  120      &  {\bf 1.7}\\% 151
+        &                                                & {\it run time (s)} &     < 1   &     1     &     2     &     7     &   20      &  120      &  {\bf 1.70}\\% 151
 
@@
 		&\multirow{2}{*}{ $Shims$}                       & $f$                &    21.31  &  17.35    &   24.24   &   27.39    &  29.31     &  27.06     &   0.96  \\%  14666
-		&                                                & {\it run time (s)} &     < 1   &     2     &     4     &     14     &     35     &   220     &  1.0 \\% 260
+		&                                                & {\it run time (s)} &     < 1   &     2     &     4     &     14     &     35     &   220     &  1.00 \\% 260
 
@@
 		&\multirow{2}{*}{ $mpShims$}                     & $f$                &    24.33  &   17.26   &   24.70   &  24.81     &   27.71    &  24.93        &  0.94  \\% 14374
-		&                                                & {\it run time (s)} &     < 1   &     2     &     3     &     9      &   26       &  151      &  {\bf 1.7}\\% 151
+		&                                                & {\it run time (s)} &     < 1   &     2     &     3     &     9      &   26       &  151      &  {\bf 1.70}\\% 151
 
@@
 
-\subsection{An attempt to minimize ground handling time}
-\label{improvement}
-
-\cite{LurkinSchyns2015} considered an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem.
-But, as both aircrafts dealt with in this work have only the ramp door, we tried to minimize the distance of travel of pallets inside the cargo bay, favoring unloading in the next node, by a post optimization of pallets positions to favor this distance minimization.
-
-So, to confirm or not Hypothesis (4), that the distances to move pallets on unloading in the next node could be minimized to benefit unloading costs. We modeled this improvement as follows.
-
-Let $maxD$ be the distance from the last ramp pallet to the center of gravity.
-
-Let $Q^{\pi(k)}  = \{ a^{\pi(k)}_1, a^{\pi(k)}_2, \ldots, a^{\pi(k)}_m \}$ be the set of consolidated items assembled on the set of $m$ pallets in node $\pi(k)$.
-
-Let $Z^{\pi(k)}_{iq}$ be the set of binary decision variables that relates pallet $i$ to consolidated $q$ in node $\pi(k)$.
-
-\begin{equation} \label{eq:distobj}
-	\mbox{minimize} \sum_{i=1}^m \sum_{q=1}^m Z^{\pi(k)}_{iq} \times ( maxD - p_i.d) ,\ \mbox{{\bf if}}\ a^{\pi(k)}_q.to = \pi(k+1) \}
-\end{equation}
-
-\begin{equation} \label{eq:distcons}
-	s.t.: \sum_{i=1}^m Z^{\pi(k)}_{iq} = 1;\ q \in \{1,2,\ldots,m\}
-\end{equation}
-
-\begin{equation} \label{eq:distpallets}
-	s.t.: \sum_{q=1}^{m} Z^{\pi(k)}_{iq}=1;\ i \in \{1,2,\ldots,m\}
-\end{equation}
-
-\begin{equation} \label{eq:distpalletw}
-	p_i.w = 140 + a^{\pi(k)}_q.w ;\ i \in \{1,2,\ldots,m\};\ q \in \{1,2,\ldots,m\}
-\end{equation}
-
-\begin{equation} \label{eq:disttorque}
-	s.t.:\ | \sum\ p_i.d * p_i.w\ | \leq maxW \times limit^{CG}_{long} ;\ i \in \{1,2,\ldots,m\}
-\end{equation}
-
-Equation \ref{eq:distobj} minimizes the sum of distances of the consolidated in node $\pi(k)$ destined to the next node $\pi(k+1)$; Equation \ref{eq:distcons} states that each consolidated will be assigned to exactly one pallet; Equation \ref{eq:distpallets} states that each pallet will receive one consolidated; Equation \ref{eq:distpalletw} calculates the total weight of each pallet; and Equation \ref{eq:disttorque} guarantees that the torque limit will not be exceeded.
-
-As the number of pallets was small (7 or 18) for an integer program solver, the increased time was negligible (less than 1s).
+\subsection{Results of the 3D packing}
+
+%\begin{table}[!h]
+%	\begin{minipage}{0.5\linewidth}
+%		left
+%	\end{minipage}\hfill % these two lines must be close to each other
+%	\begin{minipage}{0.5\linewidth}
+%		right		
+%	\end{minipage}
+%\end{table}
 
@@
 \centering
-\caption{Distances minimization}  \label{tab:ground}
-\footnotesize
-%\scriptsize
-\begin{tabular}{cc S[table-format=-1.1] S[table-format=-1.1] S[table-format=-1.3]}
+\caption{Parallel 3D packing time increment}  \label{tab:3d}
+\scriptsize
+\begin{tabular}{c S[table-format=-1.1] S[table-format=-1.1] S[table-format=-1.1]}
 \toprule
-{\bf Volume}            &                & {\bf Before} & {\bf After} & {\bf Improvement} \\
-{\bf Surplus}           & {\bf Scenario} & {\bf (m)}    & {\bf (m)}   & {\bf (\%)} \\		
-\midrule
-\multirow{6}{*}{ $1.2$} & $1$ &  31.0 &  29.0 & -5.9 \\
-                        & $2$ & 189.7 & 189.4 & -0.4 \\
-                        & $3$ & 152.0 & 152.0 & 0.0 \\
-                        & $4$ & 134.0 & 130.7 & -2.4 \\ 
-                        & $5$ & 115.7 & 115.7 & 0.0\\    
-                        & $6$ &  &  & 0.0\\                                                                  
-\midrule
-\multirow{6}{*}{ $1.5$} & $1$ &  .0 &  .0 & 5.9 \\
-						& $2$ & .7 & .4 & 0.4 \\
-						& $3$ & .0 & .0 & 0.0 \\
-						& $4$ & .0 & .7 & 2.4 \\ 
-						& $5$ & .7 & .7 & 0.0\\    
-						& $6$ &  &  & 0.0\\                                                                  
-\midrule
-\multirow{6}{*}{ $2.0$} & $1$ &  .0 &  .0 & 5.9 \\
-						& $2$ & .7 & .4 & 0.4 \\
-						& $3$ & .0 & .0 & 0.0 \\
-						& $4$ & .0 & .7 & 2.4 \\ 
-						& $5$ & .7 & .7 & 0.0\\    
-						& $6$ &  &  & 0.0\\
-
+&   \multicolumn{3}{c}{\bf Average tour time - 3D packing}  \\
+\cmidrule{2-4}
+& {\bf without} & {\bf with} & {\bf increment} \\
+{\bf Scenario} & {\bf (s)}    & {\bf (s)}   & {\bf (times)} \\		
+\midrule
+$1$ &   1 &   15 & 15 \\
+$2$ &   5 &   76 & 15 \\
+$3$ &   8 &   97 & 12 \\
+$4$ &  35 &  330 &  9 \\ 
+$5$ &  98 &  740 &  8 \\    
+$6$ & 220 & 1500 &  7 \\                                                                  
 \bottomrule
@@
 
+According to Table \ref{tab:3d}, the time increase with the 3D packing inclusion was approximately 10 times.
+About 40 min to solve a complete tour, against 4 min without the 3D packing procedure, considering the worst (6) scenario.
+i.e., with the 3D packing procedure running in multiprocessing mode, this complete solution method for {\it ACLP+RPDP} is still viable for operational use.
+
+\subsection{Results of the minimization of the pallet handling distances in the cargo bay}
+\label{improvement}
+
+As the number of pallets was small (7 or 18) for an integer program solver, the increased time was negligible (less than 1s).
+
+\vspace{2.0mm}
+\begin{table}[!ht]
+\centering
+\caption{Average distances minimization}  \label{tab:ground}
+%\footnotesize
+\scriptsize
+\begin{tabular}{c S[table-format=-1.1] S[table-format=-1.1] S[table-format=-1.3]}
+\toprule
+& {\bf Before} & {\bf After} & {\bf Improvement} \\
+{\bf Scenario} & {\bf (m)}    & {\bf (m)}   & {\bf (\%)} \\		
+\midrule
+$1$ &  32.7 & 28.7 & -12.3 \\
+$2$ & 193.3 & 188.9 & -2.3 \\
+$3$ & 152.0 & 152.0 &  0.0 \\
+$4$ & 134.0 & 130.7 & -2.4 \\ 
+$5$ & 118.0 & 116.1 & -1.6 \\    
+$6$ & 106.0 & 104.1 & -1.8 \\                                                                  
+\bottomrule
+\end{tabular}
+\end{table}
+
 Table \ref{tab:ground} presents the distances minimization of the next node destined pallets to the ramp door. Columns {\bf Before} and {\bf After} refer to the application of the current method.
@@
 
-(3) The procedure for 3-D packing generated an unfit item percentage of approximately 20\%.
-This solution phase was a parallel 3-D packing with each pallet as a process, which decreased this phase's run time by approximate a $n-1$ factor where $n$ is the number of pallets.
+(3) The procedure for 3-D packing generated an unfit item percentage of approximately 17\%, in average. 
+We are conscious that this unfit rate could be smaller if we had chosen \cite{PaquaySchynsLimbourg2016} or \cite{PaquayLimbourgSchynsOliveira2018} approaches, but as we are concerned with an operational application where the running time is crucial, we stick to an approximate approach.
+
+{\it mpShims} consider that, theoretically, all items selected for the pallets would fit.
+But, due to the variability in item dimensions, an actual complete fit is not possible, causing some unfit items to be excluded from the previous solution (line \ref{3dpacking:unfit}).
+
+This solution phase was a parallel 3-D packing with each pallet as a process, which decreased this phase's run time by approximate a $n-1$ factor where $n$ is the number of processes.
 This was possible because each packing process is completely independent; there is no multiprocessing shared memory.

### 2023-03-24 mpShims.py

@@ -30,7 +30,7 @@ class Shims(object):
         if item.To != self.Pallet.Dest[k]:
             return False
 
-        if self.Pallet.PCV + self.SCV + item.V > self.Pallet.V:
+        if self.Pallet.PCV + self.SCV + item.V > self.Pallet.V * 1.1:
             return False
 
         if self.Pallet.PCW + self.SCW + item.W > self.Pallet.W:
@@ -80,7 +80,7 @@ def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict,
 
             for i in indexes:
                 item = whip[i]
-                if pallet.isFeasible(item, 1.0, k, nodeTorque,  cfg, itemsDict, lock):
+                if pallet.isFeasible(item, 1.1, k, nodeTorque,  cfg, itemsDict, lock):
                     pallet.putItem(item, nodeTorque, solDict, N, itemsDict, lock)
 
     if tipo == "FFD":
@@ -109,7 +109,7 @@ def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict,
                 bestIndex = i
         # put the best Shim in the solution
         for item in Set[bestIndex].Items:
-            if item != None and pallet.isFeasible(item, 1.0, k, nodeTorque,  cfg, itemsDict, lock):
+            if item != None and pallet.isFeasible(item, 1.1, k, nodeTorque,  cfg, itemsDict, lock):
                 pallet.putItem(item, nodeTorque, solDict, N, itemsDict, lock)
 
 
@@ -132,14 +132,14 @@ def Solve(pallets, items, cfg, k, threshold, secBreak, mode, nodeTorque, solDict
         # parallel greedy phase
         for i, _ in enumerate(pallets):
             procs[i] = mp.Process( target=common.fillPallet, args=( pallets[i], items, k,\
-                 nodeTorque, solDict, cfg, threshold, itemsDict, lock, 1.0) ) # torque surplus
+                 nodeTorque, solDict, cfg, threshold, itemsDict, lock, 1.1) ) # torque surplus
             time.sleep(0.001)
             procs[i].start()
         
         for proc in procs:
             proc.join()
 
-        # optcgcons.minCGdev(pallets, k, nodeTorque, cfg) # made no significant difference
+        optcgcons.minCGdev(pallets, k, nodeTorque, cfg) # made no significant difference
 
         # parallel shims phase
[…]

### 2023-03-24 py3Djanet/packer.py

@@ -1,0 +1,461 @@
+#!/usr/bin/env python
+# coding: utf-8
+
+from py3Djanet.constants import Axis
+from py3Djanet.methods import DEFAULT_NUMBER_OF_DECIMALS, START_POSITION
+
+class Packer:
+    def __init__(self):
+        self.bins = [] 
+        self.unplaced_items = []
+        self.placed_items = []
+        self.unfit_items = []
+        self.total_items = 0
+        self.total_used_bins = 0 # not used yet
+        self.used_bins = [] # not used yet
+    
+    def add_bin(self, bin):
+        return self.bins.append(bin)
+    
+    def add_item(self, item): 
+        """Add unplaced items into bin's unplaced_items list.
+        Args:
+            item: an unplaced item.
+        Returns:
+            The unplaced item is added into bin's unplaced_items list."""
+        
+        self.total_items += 1
+        return self.unplaced_items.append(item) 
+    
+    def pivot_dict(self, bin, item):
+        """For each item to be placed into a certain bin, obtain a corresponding comparison parameter of each optional pivot that the item can be placed.
+        Args:
+            bin: a bin in bin list that a certain item will be placed into.
+            item: an unplaced item in item list.
+        Returns:
+            a pivot_dict contain all optional pivot point and their comparison parameter of the item.
+            a empty dict may be returned if the item couldn't be placed into the bin.
+        """
+        
+        pivot_dict = {}
+        can_put = False
+        
+        for axis in range(0, 3): 
+            items_in_bin = bin.items 
+            items_in_bin_temp = items_in_bin[:] 
+            
+            n = 0
+            while n < len(items_in_bin):
+                pivot = [0, 0, 0] 
+                
+                if axis == Axis.LENGTH: # axis = 0/ x-axis
+                    ib = items_in_bin[n]
+                    pivot = [ib.position[0] + ib.get_dimension()[0],
+                            ib.position[1],
+                            ib.position[2]]
+                    try_put_item = bin.can_hold_item_with_rotation(item, pivot) 
+                    
+                    if try_put_item: 
+                        can_put = True
+                        q = 0
+                        q = 0
+                        ib_neigh_x_axis = []
+                        ib_neigh_y_axis = []
+                        ib_neigh_z_axis = []
+                        right_neighbor = False
+                        front_neighbor = False
+                        above_neighbor = False
+                        
+                        while q < len(items_in_bin_temp):
+                            if items_in_bin_temp[q] == items_in_bin[n]: 
+                                q += 1 
[…]

### 2023-03-25 latex/comp_or.tex
Paper CompOR update

@@
 
-We also contributed by carrying out multiple experiments with the multi-processing heuristics on synthetic data based on real data from the {\it Brazilian Air Force} transportation history and a new procedure to minimize the distance from the next node destined pallet to the ramp door.
+We also contributed by carrying out multiple experiments with the multiprocessing heuristics on synthetic data based on real data from the {\it Brazilian Air Force} transportation history and a new procedure to minimize the distance from the next node destined pallet to the ramp door.
 
@@
 
-{\color{blue} $k$ is the node index and $\pi(k)$ is the node in position $k$ of tour $\pi$. $\tau^{\pi(k)}$ is the total torque applied by the loaded pallets to the aircraft in this node. }
-
-Let $M^{\pi(k)} = \{p_1, p_2, \ldots, p_m \}$ the set of $m$ pallets.
-Each pallet $p_i$, $1 \leq i \leq m$, has weight capacity $p_i.w$, volume capacity $p_i.v$, pallet destinations $p_i.to^{\pi(k)}$, $0 \leq k \leq K$, and distance to the CG of aircraft $p_i.d$.
-$p_i.to^{\pi(k)}$ denotes that pallet $p_i$ may assume a different destination in each node ${\pi(k)}$.
-
-Let $N^{\pi(k)} = \{t^{\pi(k)}_1, t^{\pi(k)}_2, \ldots, t^{\pi(k)}_{n^{\pi(k)}} \}$ be the set of $n^{\pi(k)}$ items to be loaded in node $l^{\pi(k)}$, $0 \leq k \leq K$.
-Each item $t^{\pi(k)}_j$, $1 \leq j \leq n^{\pi(k)}$, has score $t^{\pi(k)}_j.s$, weight $t^{\pi(k)}_j.w$, volume $t^{\pi(k)}_j.v$, and destination $t^{\pi(k)}_j.to \in L^{\pi(k)}$.
-Let $N = \bigcup_{0 \leq k \leq K} N^{\pi(k)}$ be the set of items of all nodes along a tour.
-
-Let $Q^{\pi(k)} = \{a^{\pi(k)}_1, a^{\pi(k)}_2, \ldots, a^{\pi(k)}_{m} \}$ be the set of consolidated items loaded in $m_k \leq m$ pallets when the aircraft arrives at node $l^{\pi(k)}$, with $0 \leq k \leq K$.
-$a^{\pi(k)}_i$, $1 \leq i \leq m$, is the group of picked-up items that were allocated on pallet $p_i$ in some of the previous nodes.
-$a^{\pi(k)}_i$ has total weight $a^{\pi(k)}_i.w$, total volume $a^{\pi(k)}_i.v$, and destination $a^{\pi(k)}_i.to \in L^{\pi(k)} \cup \{l^{\pi(k)}\}$.
-If $a^{\pi(k)}_i.to = l^{\pi(k)}$, then $a^{\pi(k)}_i$ is unloaded, and $p_i$ will be available for reloading; otherwise, $a^{\pi(k)}_i$ remains on the aircraft, eventually in another pallet, and with items of $N^{\pi(k)}$ having the same destination.
-
-Let $X_{ij}^{\pi(k)}$ and $Y_{iq}^{\pi(k)}$ be binary variables, where $0 \leq k \leq K$, $1 \leq j \leq n^{\pi(k)}$, $1 \leq i \leq m$ and $1 \leq q \leq m$.
-$X_{ij}^{\pi(k)} = 1$ if $t_j^{\pi(k)}$ is assigned to $p_i$ in node $l^{\pi(k)}$, and 0 otherwise.
-
-$Y_{iq}^{\pi(k)} = 1$ if $a_q^{\pi(k)}$ is assigned to $p_i$ in node $l^{\pi(k)}$, and 0 otherwise. By definition, $Y_{iq}^0=0$.
-
-Allocations of items or consolidated to pallets in node $l^{\pi(k)}$ can be seen as a bipartite graph $G^{\pi(k)}(V^{\pi(k)}, E^{\pi(k)})$,
-where $V^{\pi(k)} = M^{\pi(k)} \cup N^{\pi(k)} \cup Q^{\pi(k)}$, $E^{\pi(k)} = E_N^{\pi(k)} \cup E_Q^{\pi(k)}$,
-$(p_i, t_j^{\pi(k)}) \in E_N^{\pi(k)}$
-if $X_{ij}^{\pi(k)} = 1$, and $(p_i, a_q^{\pi(k)}) \in E_Q^{\pi(k)}$ if $Y_{iq}^{\pi(k)} = 1$.
+{\color{blue} $k$ is the node index and $\pi_k$ is the node in position $k$ of tour $\pi$. $\tau^{\pi_k}$ is the total torque applied by the loaded pallets to the aircraft in this node. }
+
+Let $M^{\pi_k} = \{p_1, p_2, \ldots, p_m \}$ the set of $m$ pallets.
+Each pallet $p_i$, $1 \leq i \leq m$, has weight capacity $p_i.w$, volume capacity $p_i.v$, pallet destinations $p_i.to^{\pi_k}$, $0 \leq k \leq K$, and distance to the CG of aircraft $p_i.d$.
+$p_i.to^{\pi_k}$ denotes that pallet $p_i$ may assume a different destination in each node ${\pi_k}$.
+
+Let $N^{\pi_k} = \{t^{\pi_k}_1, t^{\pi_k}_2, \ldots, t^{\pi_k}_{n^{\pi_k}} \}$ be the set of $n^{\pi_k}$ items to be loaded in node $\pi_k$, $0 \leq k \leq K$.
+Each item $t^{\pi_k}_j$, $1 \leq j \leq n^{\pi_k}$, has score $t^{\pi_k}_j.s$, weight $t^{\pi_k}_j.w$, volume $t^{\pi_k}_j.v$, and destination $t^{\pi_k}_j.to \in L^{\pi_k}$.
+Let $N = \bigcup_{0 \leq k \leq K} N^{\pi_k}$ be the set of items of all nodes along a tour.
+
+Let $Q^{\pi_k} = \{a^{\pi_k}_1, a^{\pi_k}_2, \ldots, a^{\pi_k}_{m} \}$ be the set of consolidated items loaded in $m_k \leq m$ pallets when the aircraft arrives at node $\pi_k$, with $0 \leq k \leq K$.
+$a^{\pi_k}_i$, $1 \leq i \leq m$, is the group of picked-up items that were allocated on pallet $p_i$ in some of the previous nodes.
+$a^{\pi_k}_i$ has total weight $a^{\pi_k}_i.w$, total volume $a^{\pi_k}_i.v$, and destination $a^{\pi_k}_i.to \in L^{\pi_k} \cup \{l^{\pi_k}\}$.
+If $a^{\pi_k}_i.to = l^{\pi_k}$, then $a^{\pi_k}_i$ is unloaded, and $p_i$ will be available for reloading; otherwise, $a^{\pi_k}_i$ remains on the aircraft, eventually in another pallet, and with items of $N^{\pi_k}$ having the same destination.
+
+Let $X_{ij}^{\pi_k}$ and $Y_{iq}^{\pi_k}$ be binary variables, where $0 \leq k \leq K$, $1 \leq j \leq n^{\pi_k}$, $1 \leq i \leq m$ and $1 \leq q \leq m$.
+$X_{ij}^{\pi_k} = 1$ if $t_j^{\pi_k}$ is assigned to $p_i$ in node $\pi_k$, and 0 otherwise.
+
+$Y_{iq}^{\pi_k} = 1$ if $a_q^{\pi_k}$ is assigned to $p_i$ in node $\pi_k$, and 0 otherwise. By definition, $Y_{iq}^0=0$.
+
+Allocations of items or consolidated to pallets in node $\pi_k$ can be seen as a bipartite graph $G^{\pi_k}(V^{\pi_k}, E^{\pi_k})$,
+where $V^{\pi_k} = M^{\pi_k} \cup N^{\pi_k} \cup Q^{\pi_k}$, $E^{\pi_k} = E_N^{\pi_k} \cup E_Q^{\pi_k}$,
+$(p_i, t_j^{\pi_k}) \in E_N^{\pi_k}$
+if $X_{ij}^{\pi_k} = 1$, and $(p_i, a_q^{\pi_k}) \in E_Q^{\pi_k}$ if $Y_{iq}^{\pi_k} = 1$.
 
@@
 \begin{equation} \label{eq:scores}
-	\tilde{s} = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.s
+	\tilde{s} = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.s
 \end{equation}
@@
 \begin{equation} \label{eq:costs}
-	\tilde{c} = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} [ c^{\pi(k), \pi(k+1)}\times(1+c_g\times|\epsilon^{\pi(k)}|) ] + c^{\pi(K),0}\times(1+c_g\times|\epsilon_K|)
+	\tilde{c} = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} [ c^{\pi(k), \pi(k+1)}\times(1+c_g\times|\epsilon^{\pi_k}|) ] + c^{\pi(K),0}\times(1+c_g\times|\epsilon_K|)
 \end{equation}
@@
 \begin{equation} \label{eq:tau}
-\tau^{\pi(k)} = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w +  \sum_{q=1}^{m} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w)];\ k \in \{0, 1, \ldots, K\}
+\tau^{\pi_k} = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.w +  \sum_{q=1}^{m} Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w)];\ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:eps}
-\epsilon^{\pi(k)} = \frac{\tau^{\pi(k)}}{maxW \times limit^{CG}_{long}};\ k \in \{0, 1, \ldots, K\}
+\epsilon^{\pi_k} = \frac{\tau^{\pi_k}}{maxW \times limit^{CG}_{long}};\ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp11}
-	L_0 = L; L^{\pi(k)} = L^{\pi(k-1)} - \{l^{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
+	L_0 = L; L^{\pi_k} = L^{\pi(k-1)} - \{l^{\pi_k}\}; \ k \in \{1, 2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp12}
-	X_{ij}^{\pi(k)} = 0 \ \mbox{if} \ t_j^{\pi(k)}.to \notin L^{\pi(k)}; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi(k)}\}; \ k \in \{1,2, \ldots, K\}
+	X_{ij}^{\pi_k} = 0 \ \mbox{if} \ t_j^{\pi_k}.to \notin L^{\pi_k}; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi_k}\}; \ k \in \{1,2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp9}
-	Y_{iq}^{\pi(k)} = 0 \ \mbox{if} \ a_i^{\pi(k)}.to \notin L^{\pi(k)}; \ i \in \{1, 2, \ldots, m\}; \ q \in \{1, 2, \ldots, m\}; \ k \in \{ 1, 2, \ldots, K\}
+	Y_{iq}^{\pi_k} = 0 \ \mbox{if} \ a_i^{\pi_k}.to \notin L^{\pi_k}; \ i \in \{1, 2, \ldots, m\}; \ q \in \{1, 2, \ldots, m\}; \ k \in \{ 1, 2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:cons2}
-	a_i^{\pi(k+1)}.w = \sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w;  \ i \in \{1, 2, \ldots, m\}; k \in \{0, 1, \ldots, K-1\}
+	a_i^{\pi(k+1)}.w = \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.w + \sum_{q=1}^{m} Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w;  \ i \in \{1, 2, \ldots, m\}; k \in \{0, 1, \ldots, K-1\}
 \end{equation}
@@
 \begin{equation} \label{eq:cons3}
-	a_i^{\pi(k+1)}.v = \sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.v + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.v;  \ i \in \{1, 2, \ldots, m\}; k \in \{0, 1, \ldots, K-1\}
+	a_i^{\pi(k+1)}.v = \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.v + \sum_{q=1}^{m_k} Y_{iq}^{\pi_k} \times a_q^{\pi_k}.v;  \ i \in \{1, 2, \ldots, m\}; k \in \{0, 1, \ldots, K-1\}
 \end{equation}
@@
 \begin{equation} \label{eq:cons5}
-	a_i^{\pi(k+1)}.to = t^{\pi(k)}_j.to  \ \mbox{if} \ X^{\pi(k)}_{ij} = 1  \ \mbox{and} \ t^{\pi(k)}_j.to \in L_{\pi(k+1)};  \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi(k)}\}; \ k \in \{0,1, \ldots, K-1\}
+	a_i^{\pi(k+1)}.to = t^{\pi_k}_j.to  \ \mbox{if} \ X^{\pi_k}_{ij} = 1  \ \mbox{and} \ t^{\pi_k}_j.to \in L_{\pi(k+1)};  \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi_k}\}; \ k \in \{0,1, \ldots, K-1\}
 \end{equation}
@@
 \begin{equation} \label{eq:LatIt}
-	LatIt^{\pi(k)} = \sum_{i=1}^{m} \sum_{j=1}^{n^{\pi(k)}} ( X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w \times (i\%2) - X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w \times (i+1)\%2 ); \ k \in \{0, 1, \ldots, K\}
+	LatIt^{\pi_k} = \sum_{i=1}^{m} \sum_{j=1}^{n^{\pi_k}} ( X_{ij}^{\pi_k} \times t_j^{\pi_k}.w \times (i\%2) - X_{ij}^{\pi_k} \times t_j^{\pi_k}.w \times (i+1)\%2 ); \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:LatCons}
-	LatCons^{\pi(k)} =  \sum_{i=1}^{m} \sum_{q=1}^{m}  ( Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w \times (i\%2) - Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w \times (i+1)\%2); \ k \in \{0, 1, \ldots, K\}
+	LatCons^{\pi_k} =  \sum_{i=1}^{m} \sum_{q=1}^{m}  ( Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w \times (i\%2) - Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w \times (i+1)\%2); \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:torqlat}
-	s.t.: d^{CG}_{pallet} \times | LatIt^{\pi(k)} + LatCons^{\pi(k)} | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: d^{CG}_{pallet} \times | LatIt^{\pi_k} + LatCons^{\pi_k} | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:torqlong}
-	s.t.: |\tau^{\pi(k)}| \leq maxW \times limit^{CG}_{long};\ k \in \{0, 1, \ldots, K\}
+	s.t.: |\tau^{\pi_k}| \leq maxW \times limit^{CG}_{long};\ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:payload}
-	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^k.w ) \leq maxW; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi_k} \times a_q^k.w ) \leq maxW; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app2}
-	s.t.: \sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m\}; \ {\pi(k)} \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m\}; \ {\pi_k} \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app3}
-	s.t.: \sum_{j=1}^{n^{\pi(k)}} X_{ij}^{\pi(k)} \times t_j^{\pi(k)}.v + \sum_{q=1}^{m} Y_{iq}^{\pi(k)} \times a_q^{\pi(k)}.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m\}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.v + \sum_{q=1}^{m} Y_{iq}^{\pi_k} \times a_q^{\pi_k}.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m\}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app4}
-	s.t.: \sum_{i=1}^{m} X_{ij}^{\pi(k)} \leq 1; \ j \in \{1, 2, \ldots, n^{\pi(k)}\}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{i=1}^{m} X_{ij}^{\pi_k} \leq 1; \ j \in \{1, 2, \ldots, n^{\pi_k}\}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app5}
-	s.t.:  Y_{iq}^{\pi(k)} = 1 \ \mbox{if} \ a^{\pi(k)}_q.to \in L^{\pi(k)}; \ q \in \{1, 2, \ldots, m\}; \ k \in \{0, 1, \ldots, K\}
+	s.t.:  Y_{iq}^{\pi_k} = 1 \ \mbox{if} \ a^{\pi_k}_q.to \in L^{\pi_k}; \ q \in \{1, 2, \ldots, m\}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp8}
-	s.t.: p_i.to^{\pi(k)} = t^{\pi(k)}_j.to\ \mbox{if} \ X_{ij}^{\pi(k)} = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
+	s.t.: p_i.to^{\pi_k} = t^{\pi_k}_j.to\ \mbox{if} \ X_{ij}^{\pi_k} = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi_k}\}; \ k \in \{1, 2, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:pdp2}
-	s.t.:  p_i.to^{\pi(k)} = a^{\pi(k)}_q.to\ \mbox{if} \ Y_{iq}^{\pi(k)} = 1; \ i \in \{1, 2, \ldots, m\};\ q \in \{1, 2, \ldots, m\}; \ k \in \{1, 2, \ldots, K\}
+	s.t.:  p_i.to^{\pi_k} = a^{\pi_k}_q.to\ \mbox{if} \ Y_{iq}^{\pi_k} = 1; \ i \in \{1, 2, \ldots, m\};\ q \in \{1, 2, \ldots, m\}; \ k \in \{1, 2, \ldots, K\}
 \end{equation}
@@
 The objective of this problem is to find a permutation $\pi \in S_K$\/ that maximizes the function $f_\pi(\tilde{s},\tilde{c})$ \ref{eq:maxf}.
-In this way, the flight plan will be $l_0, l_{\pi(1)}, \ldots, l^{\pi(K)}, l_0$.
+In this way, the flight plan will be $l_0, l_{\pi(1)}, \ldots, l^{\pi_k}, l_0$.
 $\tilde{s}$\/ is the total score of transported items \ref{eq:scores} and  $\tilde{c}$\/ is the total cost of fuel consumed \ref{eq:costs}.
@@
 We could have used another criterion, but it was observed in the experiments that volume is more constrictive in airlift.
-Once the destinations of the pallets are defined, we will use serial and Multi-processing heuristics to find the best possible node-by-node solutions.
+Once the destinations of the pallets are defined, we will use serial and multiprocessing heuristics to find the best possible node-by-node solutions.
 This strategy is summarized in Algorithm \ref{alg:main}, retrieved from \cite{MesquitaSanches2023}.
@@
 
-The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $l^{\pi(k)}$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n^{\pi(k)}} t^{\pi(k)}_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.v$).
+The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $\pi_k$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n^{\pi_k}} t^{\pi_k}_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.v$).
 This parameter is passed to $ItemsGeneration$ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section.
@@
 
-As we mentioned in the previous section, all tours start and end at $l_0$\/ (lines \ref{tour:pi1}-\ref{tour:pi2}).
+As we mentioned in the previous section, all tours start and end at $\pi_0$\/ (lines \ref{tour:pi1}-\ref{tour:pi2}).
 After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ flights (lines \ref{tour:loop1}-\ref{tour:loop2}).
 Initially we set pallets destination as $-1$\/ (line \ref{tour:-1}).
-When the aircraft is at node $l_0$, the initial graph $G_1$\/ is empty because it has no consolidated items \ref{tour:g11}.
-Otherwise, the set $L_k$\/ of remaining nodes when departing from $\pi(k)$ is updated (line \ref{tour:lk1}), and $UpdateConsolidated$\/ (line \ref{tour:dest}) returns the set of consolidated items that have not yet reached their destinations and remain on board, rearranging them on the pallets to minimize CG deviation.
+When the aircraft is at node $\pi_0$, the initial graph $G_1$\/ is empty because it has no consolidated items \ref{tour:g11}.
+Otherwise, the set $L_k$\/ of remaining nodes when departing from $\pi_k$ is updated (line \ref{tour:lk1}), and $UpdateConsolidated$\/ (line \ref{tour:dest}) returns the set of consolidated items that have not yet reached their destinations and remain on board, rearranging them on the pallets to minimize CG deviation.
 This allocation is stored in graph $G_1$\/ (line \ref{tour:g12}).
-
-In the context of this work, we know that $m>K$, once the aircraft has $7$\/ or $18$\/ pallets and $K\leq 6$, allowing there to be at least one pallet for each node to be visited.
-$SetPalletsDestination$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands caused by the items to be embarked in the current node, without changing the pallets destination with consolidated items.
-
-Finally, $SolveNode$\/ includes the edges corresponding to the items shipped at the current node, returning the graph $G_2$\/ (line \ref{tour:node}).
-The score and the CG deviation of this graph are calculated (line \ref{tour:analyse}) and accumulated (lines \ref{tour:score2}-\ref{tour:cost2}), allowing the final result of this tour (line \ref{tour:f}).
 
@@
 
-		\State $\pi(0) \gets 0$ \label{tour:pi1}
+		\State $\pi[0] \gets 0$ \label{tour:pi1}
 		\State $\pi(K+1) \gets 0$ \label{tour:pi2}
@@
 				\State $L_k \gets L_k - \pi(k)$  \label{tour:lk1}			
-				\State $Q^{\pi(k)}, E_Q^{\pi(k)}, M^{\pi(k)} \gets UpdateConsolidated(\pi(k))$ \label{tour:dest}			
-				\State Let $G_1(M^{\pi(k)} \cup N^{\pi(k)} \cup Q^{\pi(k)}, E_Q^{\pi(k)})$ \label{tour:g12}
+				\State $Q^{\pi_k}, E_Q^{\pi_k}, M^{\pi_k} \gets UpdateConsolidated(\pi(k))$ \label{tour:dest}			
+				\State Let $G_1(M^{\pi_k} \cup N^{\pi_k} \cup Q^{\pi_k}, E_Q^{\pi_k})$ \label{tour:g12}
 			\EndIf  \label{tour:lk2}	
-			\State $M^{\pi(k)} \gets SetPalletsDestination( \pi(k) )$ \label{tour:dest2}		
+			\State $M^{\pi_k} \gets SetPalletsDestination( \pi(k) )$ \label{tour:dest2}		
 			\State $G_2 \gets SolveNode(method, \pi(k), G_1, timeLim)$ \label{tour:node}
@@
 
+In the context of this work, we know that $m>K$, once the aircraft has $7$\/ or $18$\/ pallets and $K\leq 6$, allowing there to be at least one pallet for each node to be visited.
+$SetPalletsDestination$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands caused by the items to be embarked in the current node, without changing the pallets destination with consolidated items.
+
+Finally, $SolveNode$\/ includes the edges corresponding to the items shipped at the current node, returning the graph $G_2$\/ (line \ref{tour:node}).
+The score and the CG deviation of this graph are calculated (line \ref{tour:analyse}) and accumulated (lines \ref{tour:score2}-\ref{tour:cost2}), allowing the final result of this tour (line \ref{tour:f}).
+
 $UpdateConsolidated$ finds the best allocation for the consolidated items that remain on board. Initially, the set $Q$ is created, with the consolidated items that did not reach their destination.
@@
 Then $MinCGDeviation$ is run through a MIP solver to relocate the consolidated items on the pallets minimizing torque and ensuring that they all remain on board, one on each pallet.
-As there are few variables, the MIP solver returns an allocation $E^Q_k$ very quickly.
+As there are few variables, the MIP solver returns an allocation $E_Q^{\pi_k}$ very quickly.
 
@@
 
-In this subsection we present implementations of $SolveNode(...)$, where $method$\/ corresponds to a heuristic, $k$\/ is the index of the current node $l^{\pi(k)}$ and $G$\/ is the allocation graph of the consolidated items that remain on board at $l^{\pi(k)}$.
+In this subsection we present implementations of $SolveNode(...)$, where $method$\/ corresponds to a heuristic, $k$\/ is the index of the current node $\pi_k$ and $G$\/ is the allocation graph of the consolidated items that remain on board at $\pi_k$.
 
@@
 
-The selection of edges for $E_N^{\pi(k)}$\/ uses the {\it edge attractiveness}\/ $\theta^{\pi(k)}_{ij}$, Equation \ref{eq:edge}, which can be understood as the tendency to allocate the item $t^{\pi(k)}_j$\/ to the pallet $p_i$.
+The selection of edges for $E_N^{\pi_k}$\/ uses the {\it edge attractiveness}\/ $\theta^{\pi_k}_{ij}$, Equation \ref{eq:edge}, which can be understood as the tendency to allocate the item $t^{\pi_k}_j$\/ to the pallet $p_i$.
 It is directly proportional to the score, and inversely relative to the volume and torque of each item.
@@
 \begin{equation} \label{eq:edge}
-	\theta^{\pi(k)}_{ij}= \frac{t^{\pi(k)}_j.s}{t^{\pi(k)}_j.v \times 3000}\times(1-\frac{t^{\pi(k)}_j.w\times|p_i.d|}{\max_s\{t^{\pi(k)}_s.w\}\times\max_q\{|p_q.d|\}});\ i \in \{1,2,\ldots,m\},\ j \in \{1,2,\ldots,n_k\}
+	\theta^{\pi_k}_{ij}= \frac{t^{\pi_k}_j.s}{t^{\pi_k}_j.v \times 3000}\times(1-\frac{t^{\pi_k}_j.w\times|p_i.d|}{\max_s\{t^{\pi_k}_s.w\}\times\max_q\{|p_q.d|\}});\ i \in \{1,2,\ldots,m\},\ j \in \{1,2,\ldots,n_k\}
 \end{equation}
@@
 
-In the pseudo-codes presented below, given an allocation graph $G$\/ for the node $l^{\pi(k)}$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation({\pi(k)},G) $, described in \cite{MesquitaSanches2023}.
+In the pseudo-codes presented below, given an allocation graph $G$\/ for the node $\pi_k$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation({\pi_k},G) $, described in \cite{MesquitaSanches2023}.
 
@@
 
-According to \cite[p.226]{manfrin2006}, there are five topologies for multi-process interchange of information: {\it fully-connected}, where the master process broadcasts to all remaining child processes; {\it replace-worst}, where the best-so-far solution process broadcasts only to the current worst solution process; {\it hypercube}, where processes are connected as a hypercube, and a vertex process broadcasts only to the connected vertices; {\it ring}, in which one process only sends a message to the next process connected to it; and {\it parallel independent runs (PIR)}, in which there are no communication costs and the best solution is chosen among all processes.
+According to \cite[p.226]{manfrin2006}, there are five topologies for multiprocessing interchange of information: {\it fully-connected}, where the master process broadcasts to all remaining child processes; {\it replace-worst}, where the best-so-far solution process broadcasts only to the current worst solution process; {\it hypercube}, where processes are connected as a hypercube, and a vertex process broadcasts only to the connected vertices; {\it ring}, in which one process only sends a message to the next process connected to it; and {\it parallel independent runs (PIR)}, in which there are no communication costs and the best solution is chosen among all processes.
 They concluded that the PIR topology presented, in average, the smallest distance from the optimal solutions.
@@
 			\node at (7, 0.3) {$\eta_2[i]$};
-			\node at (0.3, 5) {$\theta_{ij}^{\pi(k)}$};
+			\node at (0.3, 5) {$\theta_{ij}^{\pi_k}$};
 			\draw[->] (.6, .2) -- coordinate (y axis mid) (0.6, 5.3);
-			\node at (12, 0.3) {$e_{ij}^{\pi(k)}$};
+			\node at (12, 0.3) {$e_{ij}^{\pi_k}$};
 			\draw[smooth, domain = 0.09:2, color=black, thick] plot (.3+1/\x,{4.2+log2(\x)});
@@
 
-		\captionof{figure}{$n^{\pi(k)}$\/ edges $e_{ij}^{\pi(k)}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi(k)}$\/ in non-ascending order}
+		\captionof{figure}{$n^{\pi_k}$\/ edges $e_{ij}^{\pi_k}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi_k}$\/ in non-ascending order}
 		\label{fig:whip}		
@@
 
-Figure \ref{fig:whip} represents the $n^{\pi(k)}$\/ possible edges $e_{ij}^{\pi(k)}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi(k)}$.
-\emph{mpShims}\/ starts with a greedy solution for each pallet, stopping at the edge with index in $\eta_1[i]$ close to the local optimum (first phase).
-Then, performs an integer programming procedure to minimize the CG deviation (second phase), and finally, considering even the edge with the index $\eta_2[i]$, it elaborates different possible complements for this pallet and selects the best of these complements (third phase).
+Figure \ref{fig:whip} represents the $n^{\pi_k}$\/ possible edges $e_{ij}^{\pi_k}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi_k}$.
+\emph{mpShims}\/ starts with a greedy solution for each pallet, stopping at the edge with index in $\eta_1^i$ close to the local optimum (first phase).
+Then, performs an integer programming procedure to minimize the CG deviation (second phase), and finally, considering even the edge with the index $\eta_2^i$, it elaborates different possible complements for this pallet and selects the best of these complements (third phase).
 
@@
 $3^{st}$ - {\it Shims} set construction and best {\it Shims} selection
-
-All executed in a multi-process run-time, accessing concurrently, reading and writing the list of items and the current value of the CG deviation.
-As to the multi-process run-time, we considered each pallet as a parallel process, taking advantage of nowadays multi-core computers.
+ess run-time, we considered each pallet as a parallel process, taking advantage of nowadays multi-core computers.
 
@@
 
-		\Procedure{$mpShims$}{${N^{\pi(k)}, \pi(k)}, limit, G$}
-
-		\State Let $G(M^{\pi(k)},\ Q^{\pi(k)},\ E^{\pi(k)})$ \label{mpShims:g1} \Comment{the initial solution}
+		\Procedure{$mpShims$}{${N^{\pi_k}, \pi(k)}, limit, G$}
+
+		\State Let $G(M^{\pi_k},\ Q^{\pi_k},\ E^{\pi_k})$ \label{mpShims:g1} \Comment{the initial solution}
 
@@
 
-		\State $lock \gets$ multi-processing lock  \label{mpShims:lock} \Comment{avoid race condition}
+		\State $lock \gets$ multiprocessing lock  \label{mpShims:lock} \Comment{avoid race condition}
 
@@
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets Greedy(p_i,\ N^{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ limit,\ lock,\ ts )$   \label{mpShims:greedy}
+		\State $proc_i \gets Greedy(p_i,\ N^{\pi_k},\ \pi(k),\ nodeTorque,\ G,\ limit,\ lock,\ ts )$   \label{mpShims:greedy}
 		\State $proc_i.start$
@@
 
-		\State $minCGdev(M^{\pi(k)},\ \pi(k),\ nodeTorque)$  \label{mpShims:minCGdev} \Comment{minimize the CG deviation.}
+		\State $minCGdev(M^{\pi_k},\ \pi(k),\ nodeTorque)$  \label{mpShims:minCGdev} \Comment{minimize the CG deviation.}
 
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets getBestShims(p_i,\ N^{\pi(k)},\ \pi(k),\ nodeTorque,\ G,\ surplus,\ lock,\ ts )$ \label{mpShims:getBestShims}
+		\State $proc_i \gets getBestShims(p_i,\ N^{\pi_k},\ \pi(k),\ nodeTorque,\ G,\ surplus,\ lock,\ ts )$ \label{mpShims:getBestShims}
 		\State $proc_i.start$
@@
 
-		\Return $G(M^{\pi(k)},\ N^{\pi(k)} \cup Q^{\pi(k)},\ E^{\pi(k)})$ \label{mpShims:return}
+		\Return $G(M^{\pi_k},\ N^{\pi_k} \cup Q^{\pi_k},\ E^{\pi_k})$ \label{mpShims:return}
 
@@
 
+All phases being executed in a multiprocessing run-time, accessing concurrently, reading and writing the list of items and the current value of the CG deviation.
+
+Follow the comments on Algorithm \ref{alg:mpShims}:
+
 Line \ref{mpShims:g1} the initial solution comes with some consolidated on board, a bipartite graph between pallets and consolidated.
@@
 
-Line \ref{mpShims:greedy} calls the greedy method which updates $p_i$, $N^{\pi(k)}$, $nodeTorque$, and $G$.
+Line \ref{mpShims:greedy} calls the greedy method which updates $p_i$, $N^{\pi_k}$, $nodeTorque$, and $G$.
 
@@
 
-Line \ref{mpShims:getBestShims} calls the $getBestShims$ method which updates $p_i$, $N^{\pi(k)}$, $nodeTorque$, and $G$.
+Line \ref{mpShims:getBestShims} calls the $getBestShims$ method which updates $p_i$, $N^{\pi_k}$, $nodeTorque$, and $G$.
 
@@
 
-Algorithm \ref{alg:greedy} generates a greedy allocation of the items available in node ${\pi(k)}$, according to the non-ascending order of $\theta^{\pi(k)}_{ij}$, and considering the consolidated items already shipped ($E_Q^{\pi(k)}$).
+Algorithm \ref{alg:greedy} generates a greedy allocation of the items available in node ${\pi_k}$, according to the non-ascending order of $\theta^{\pi_k}_{ij}$, and considering the consolidated items already shipped ($E_Q^{\pi_k}$).
 Edges are included in this allocation as long as they respect feasibility constraints.
@@
 
-		\Procedure{$Greedy$}{${\pi(k)},\ G,\ limit,\ ts$}
-
-		\State Let $G(V^{\pi(k)}, E_Q^{\pi(k)})$
-
-		\State $\eta_1 \gets \{0\} \times m$ \label{greedy:eta}
-		\State $volume \gets \{0\} \times m$
-
-		\State $E_N^{\pi(k)} \gets \varnothing$
-
-		\For{$q \gets 1$ to $m$}
-		\If{$(p_i, a_q^{\pi(k)}) \in E_Q^{\pi(k)}$} 
-		\State $volume_q \gets volume_q + a_q^{\pi(k)}.v$ 
+		\Procedure{$Greedy$}{${\pi_k},\ G,\ limit,\ ts$}
+
+		\State Let $G(V^{\pi_k}, E_Q^{\pi_k})$
+
+		\State $\eta_1 \gets \{0\ |\ 1\ to\ m \}$ \label{greedy:eta}
+		\State $volume \gets \{0\ |\ 1\ to\ m \}$
+
+		\State $E_N^{\pi_k} \gets \varnothing$
+
+		\State $\tau^{\pi_k} \gets 0$	 \Comment{aircraft torque in node $\pi_k$}
+
+		\For{$q \gets 1$ {\bf to} $m$}
+		\If{$(p_i, a_q^{\pi_k}) \in E_Q^{\pi_k}$} 
+		\State $volume_q \gets volume_q + a_q^{\pi_k}.v$ 
 		\EndIf		
@@
 
-		\State $\tau_{max} \gets maxW \times limit^{CG}_{long}$
-
-		\For {each $e_{ij}^{\pi(k)}$ in non-ascending order of $\theta^{\pi(k)}_{ij}$}
+		\State $\tau_{max} \gets maxW \times limit^{CG}_{long}$ \Comment{maximum aircraft torque}
+
+		\For {each $e_{ij}^{\pi_k}$ in non-ascending order of $\theta^{\pi_k}_{ij}$}
 
@@
 
-		\State $\tau_{new} \gets | \tau^{\pi(k)} + \Delta\tau |$
-
-		\If{ ($E_N^{\pi(k)} \cup \{e_{ij}^{\pi(k)}\}$ is feasible) {\bf and} ($volume_i \leq p_i.v \times limit$) {\bf and} ($\tau_{new} \leq \tau_{max} \times ts$ )} \label{greedy:ts}
-
-			\State $E_N^{\pi(k)} \gets E_N^{\pi(k)} \cup \{e_{ij}^{\pi(k)}\}$ 
-			\State $volume_i \gets volume_i + t_j^{\pi(k)}.v$
+		\State $\tau_{new} \gets | \tau^{\pi_k} + \Delta\tau |$
+
+		\If{ ($E_N^{\pi_k} \cup \{e_{ij}^{\pi_k}\}$ is feasible) {\bf and} ($volume_i \leq p_i.v \times limit$) {\bf and} ($\tau_{new} \leq \tau_{max} \times ts$ )} \label{greedy:ts}
+
+			\State $E_N^{\pi_k} \gets E_N^{\pi_k} \cup \{e_{ij}^{\pi_k}\}$ 
+			\State $volume_i \gets volume_i + t_j^{\pi_k}.v$
 			\State $\eta_1^i \gets \eta_1^i + 1$ 
-			\State $\tau^{\pi(k)} \gets \tau^{\pi(k)} + \Delta\tau$
+			\State $\tau^{\pi_k} \gets \tau^{\pi_k} + \Delta\tau$
 
@@
 
-		\Return $G(V^{\pi(k)}, E_N^{\pi(k)} \cup E_Q^{\pi(k)}),\ \eta_1$
+		\Return $G(V^{\pi_k}, E_N^{\pi_k} \cup E_Q^{\pi_k}),\ \eta_1$
 
@@
 Line \ref{greedy:ts} presents the variable $ts$ (torque $\tau$ surplus), that is an extra-torque limit to allow each greedy pallet construction to, temporally, exceed the torque limit, that will be turned feasible again by the next {\it mpShims}' phase, the CG minimization phase.
-$ts$ value will be adjusted empirically to produce feasible solutions in all problem instances.
+$ts$ value will be adjusted empirically, by means of the {\it iRace} tool (\cite{LopezIbanezManuel2016}), to produce feasible solutions in all problem instances.
 
@@
 
-Let $\pi$ be a tour in the set of permutations of nodes.
-
-Let $Q^{\pi(k)}  = \{ a^{\pi(k)}_1, a^{\pi(k)}_2, \ldots, a^{\pi(k)}_m \}$ be the set of consolidated items assembled on the set of $m$ pallets in node $\pi(k)$.
+Let $Q^{\pi_k}  = \{ a^{\pi_k}_1, a^{\pi_k}_2, \ldots, a^{\pi_k}_m \}$ be the set of consolidated items assembled on the set of $m$ pallets in node $\pi_k$.
 There is a bi-univocal relation between pallets and consolidated.
 
-Let $H^{\pi(k)}_{iq}$ be the set of decision variables that relates pallet $i$ to consolidated $q$ in node $\pi(k)$.
+Let $H^{\pi_k}_{iq}$ be the set of decision variables that relates pallet $i$ to consolidated $q$ in node $\pi_k$.
 
 \begin{equation} \label{eq:distobjshims}
-	\mbox{minimize}\ \Bigg |  \sum_{i=1}^{m} \sum_{q=1}^{m} H^{\pi(k)}_{iq} \times  (140 + a^{\pi(k)}_q.w) * p_i.d\ \Bigg | 
+	\mbox{minimize}\ \Bigg |  \sum_{i=1}^{m} \sum_{q=1}^{m} H^{\pi_k}_{iq} \times  (140 + a^{\pi_k}_q.w) \times p_i.d\ \Bigg | 
 \end{equation}
@@
 \begin{equation} \label{eq:distconsshims}
-	s.t.: \sum_{i=1}^{m} H^{\pi(k)}_{iq} = 1;\ q \in \{1,2,\ldots,m\}
+	s.t.: \sum_{i=1}^{m} H^{\pi_k}_{iq} = 1;\ q \in \{1,2,\ldots,m\}
 \end{equation}
@@
 \begin{equation} \label{eq:distpalletsshims}
-	s.t.: \sum_{q=1}^{m} H^{\pi(k)}_{iq}=1;\ i \in \{1,2,\ldots,m\}
-\end{equation}
-
-Equation \ref{eq:distobjshims} minimizes the aircraft absolute torque in node $\pi(k)$, where 140 is the weight of an empty pallet, $a^{\pi(k)}_q.w$ the weight of the consolidated, and $p_i.d$ the distance between the pallet centroid to the aircraft CG; Equation \ref{eq:distconsshims} states that each consolidated will be assigned to exactly one pallet; and Equation \ref{eq:distpalletsshims} states that each pallet will receive one consolidated.
+	s.t.: \sum_{q=1}^{m} H^{\pi_k}_{iq}=1;\ i \in \{1,2,\ldots,m\}
+\end{equation}
+
+Equation \ref{eq:distobjshims} minimizes the aircraft absolute torque in node $\pi_k$, where 140 is the weight of an empty pallet, $a^{\pi_k}_q.w$ the weight of the consolidated, and $p_i.d$ the distance between the pallet centroid to the aircraft CG; Equation \ref{eq:distconsshims} states that each consolidated will be assigned to exactly one pallet; and Equation \ref{eq:distpalletsshims} states that each pallet will receive one consolidated.
 
@@
 		\State $NewShims \gets$ {\bf True} \label{new_shims}
-		\State $e_{ij}^k \gets E_x^i$
+		\State $e_{ij}^{\pi_k} \gets E_x^i$
 
@@
 
-		\If { $e_{ij}^k \not\in (E^N_k \cup shims)$ {\bf and} $e_{ij}^k$ is feasible  {\bf and}  $(t_j^k.v + volume) \leq slack^i$}
-
-		\State $shims \gets shims \cup \{e_{ij}^k\}$
+		\If { $e_{ij}^{\pi_k} \not\in (E_N^{\pi_k} \cup shims)$ {\bf and} $e_{ij}^{\pi_k}$ is feasible  {\bf and}  $(t_j^k.v + volume) \leq slack^i$}
+
+		\State $shims \gets shims \cup \{e_{ij}^{\pi_k}\}$
 		\State $volume \gets volume + t_j^k.v$
@@
 		\State $shims_b \gets \{\}$
-		\State $shims_b \gets shims_b \cup \{e_{ij}^k\}$
+		\State $shims_b \gets shims_b \cup \{e_{ij}^{\pi_k}\}$
 		\State $Set \gets Set \cup \{shims_b\}$
@@
 
-		\Procedure{$mp3Dpacking$}{${\pi(k)},\ G$}
-
-		\State Let $G(M^{\pi(k)},\ N^{\pi(k)} \cup Q^{\pi(k)},\ E^{\pi(k)})$ \label{3dpacking:1}
+		\Procedure{$mp3Dpacking$}{${\pi_k},\ G$}
+
+		\State Let $G(M^{\pi_k},\ N^{\pi_k} \cup Q^{\pi_k},\ E^{\pi_k})$ \label{3dpacking:1}
 
@@
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets 3Dpacker(p_i,\ N^{\pi(k)}_i,\ \pi(k),\ G )$   \label{3dpacking:3}
+		\State $proc_i \gets 3Dpacker(p_i,\ N^{\pi_k}_i,\ \pi(k),\ G )$   \label{3dpacking:3}
 		\State $proc_i.start$
@@
 		\State $proc_i.join$ \label{3dpacking:4}
-		\State $E^{\pi(k)}_i \gets E^{\pi(k)}_i \setminus proc_i.unfit$  \label{3dpacking:5} 
+		\State $E^{\pi_k}_i \gets E^{\pi_k}_i \setminus proc_i.unfit$  \label{3dpacking:5} 
 		\EndFor
 
-		\Return $G(M^{\pi(k)},\ N^{\pi(k)} \cup Q^{\pi(k)},\ E^{\pi(k)})$
+		\Return $G(M^{\pi_k},\ N^{\pi_k} \cup Q^{\pi_k},\ E^{\pi_k})$
 
@@
 
-In line \ref{3dpacking:3}, $N^{\pi(k)}_i$ is the set of items previously assigned to pallet $i$.
+In line \ref{3dpacking:3}, $N^{\pi_k}_i$ is the set of items previously assigned to pallet $i$.
 
@@
 
-Let $Q^{\pi(k)}  = \{ a^{\pi(k)}_1, a^{\pi(k)}_2, \ldots, a^{\pi(k)}_m \}$ be the set of consolidated items assembled on the set of $m$ pallets in node $\pi(k)$.
-
-Let $Z^{\pi(k)}_{iq}$ be the set of binary decision variables that relates pallet $i$ to consolidated $q$ in node $\pi(k)$.
+Let $Q^{\pi_k}  = \{ a^{\pi_k}_1, a^{\pi_k}_2, \ldots, a^{\pi_k}_m \}$ be the set of consolidated items assembled on the set of $m$ pallets in node $\pi_k$.
+
+Let $Z^{\pi_k}_{iq}$ be the set of binary decision variables that relates pallet $i$ to consolidated $q$ in node $\pi_k$.
 
 \begin{equation} \label{eq:distobj}
-	\mbox{minimize} \sum_{i=1}^m \sum_{q=1}^m Z^{\pi(k)}_{iq} \times ( maxD - p_i.d) ,\ \mbox{{\bf if}}\ a^{\pi(k)}_q.to = \pi(k+1) \}
+	\mbox{minimize} \sum_{i=1}^m \sum_{q=1}^m Z^{\pi_k}_{iq} \times ( maxD - p_i.d) ,\ \mbox{{\bf if}}\ a^{\pi_k}_q.to = \pi(k+1) \}
 \end{equation}
@@
 \begin{equation} \label{eq:distcons}
-	s.t.: \sum_{i=1}^m Z^{\pi(k)}_{iq} = 1;\ q \in \{1,2,\ldots,m\}
+	s.t.: \sum_{i=1}^m Z^{\pi_k}_{iq} = 1;\ q \in \{1,2,\ldots,m\}
 \end{equation}
@@
 \begin{equation} \label{eq:distpallets}
-	s.t.: \sum_{q=1}^{m} Z^{\pi(k)}_{iq}=1;\ i \in \{1,2,\ldots,m\}
+	s.t.: \sum_{q=1}^{m} Z^{\pi_k}_{iq}=1;\ i \in \{1,2,\ldots,m\}
 \end{equation}
@@
 \begin{equation} \label{eq:distpalletw}
-	p_i.w = 140 + a^{\pi(k)}_q.w ;\ i \in \{1,2,\ldots,m\};\ q \in \{1,2,\ldots,m\}
+	p_i.w = 140 + a^{\pi_k}_q.w ;\ i \in \{1,2,\ldots,m\};\ q \in \{1,2,\ldots,m\}
 \end{equation}
@@
 
-Equation \ref{eq:distobj} minimizes the sum of distances of the consolidated in node $\pi(k)$ destined to the next node $\pi(k+1)$; Equation \ref{eq:distcons} states that each consolidated will be assigned to exactly one pallet; Equation \ref{eq:distpallets} states that each pallet will receive one consolidated; Equation \ref{eq:distpalletw} calculates the total weight of each pallet; and Equation \ref{eq:disttorque} guarantees that the torque limit will not be exceeded.
+Equation \ref{eq:distobj} minimizes the sum of distances of the consolidated in node $\pi_k$ destined to the next node $\pi(k+1)$; Equation \ref{eq:distcons} states that each consolidated will be assigned to exactly one pallet; Equation \ref{eq:distpallets} states that each pallet will receive one consolidated; Equation \ref{eq:distpalletw} calculates the total weight of each pallet; and Equation \ref{eq:disttorque} guarantees that the torque limit will not be exceeded.
 
@@
 
-For parallelization, we used the {\it Python Multi-processing} package, which implements {\it process-based parallelism} or {\it parallel shared memory algorithm}.
-According to \cite{multiprocessing}, {\it Multi-processing} is a package that supports spawning processes using an API similar to the {\it Python Threading} package.
+For parallelization, we used the {\it Python Multiprocessing} package, which implements {\it process-based parallelism} or {\it parallel shared memory algorithm}.
+According to \cite{multiprocessing}, {\it Multiprocessing} is a package that supports spawning processes using an API similar to the {\it Python Threading} package.
 
@@
 It is known that working concurrently opens up synchronization issues.
-But the {\it Multi-processing} package ({\it mp}) offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
-Because of this, the {\it Multi-processing} module lets the programmer take full advantage of the fact that a machine has more than one processor, normally capable of 2 threads each.
+But the {\it Multiprocessing} package ({\it mp}) offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
+Because of this, the {\it Multiprocessing} module lets the programmer take full advantage of the fact that a machine has more than one processor, normally capable of 2 threads each.
 
@@
 
-After this inclusion {\it Shims} results surpassed that of {\it mpShims}. In multi-processing mode ({\it mpShims}) this inclusion (line \ref{mpShims:minCGdev}) made a negligible difference.
+After this inclusion {\it Shims} results surpassed that of {\it mpShims}. In multiprocessing mode ({\it mpShims}) this inclusion (line \ref{mpShims:minCGdev}) made a negligible difference.
 
@@
 $5$ &  98 &  740 &  8 \\    
-$6$ & 220 & 1500 &  7 \\                                                                  
+$6$ & 591 & 3709 &  6 \\                                                                  
 \bottomrule
@@
 According to Table \ref{tab:3d}, the time increase with the 3D packing inclusion was approximately 10 times.
-About 40 min to solve a complete tour, against 4 min without the 3D packing procedure, considering the worst (6) scenario.
+Considering the worst (6) scenario, about 62 min to solve a complete tour, against 10 min without the 3D packing procedure.
 i.e., with the 3D packing procedure running in multiprocessing mode, this complete solution method for {\it ACLP+RPDP} is still viable for operational use.
@@
 $5$ & 118.0 & 116.1 & -1.6 \\    
-$6$ & 106.0 & 104.1 & -1.8 \\                                                                  
+$6$ & 105.2 & 104.8 & -0.4 \\                                                                  
 \bottomrule
@@
 
-%There was some expectation on this post-improvement method as it could have minimized the distances of travel of the pallets through the ramp door, but unfortunately, after running all scenarios and problem instances, there was no improvement in minimizing these distances.
-
-%This proves that the solutions provided by {\it Shims} and {\it mpShims} are tight enough to hinder this intended improvement.
-
 \subsection{Discussion summary}
@@
 (3) The procedure for 3-D packing generated an unfit item percentage of approximately 17\%, in average. 
-We are conscious that this unfit rate could be smaller if we had chosen \cite{PaquaySchynsLimbourg2016} or \cite{PaquayLimbourgSchynsOliveira2018} approaches, but as we are concerned with an operational application where the running time is crucial, we stick to an approximate approach.
+We are conscious that this unfit rate could be smaller if we had chosen \cite{PaquayLimbourgSchynsOliveira2018} approaches, but as we are concerned with an operational application where the running time is crucial, we stick to an approximate approach.
 
 {\it mpShims} consider that, theoretically, all items selected for the pallets would fit.
-But, due to the variability in item dimensions, an actual complete fit is not possible, causing some unfit items to be excluded from the previous solution (line \ref{3dpacking:unfit}).
+But, due to the variability in item dimensions, an actual complete fit is not possible, causing some unfit items to be excluded from the previous solution.
 
@@
 (4) The distances to move pallets on unloading in the next node could be minimized to benefit ground handling costs.
-%This hypothesis was not confirmed.
-We implemented an integer programming procedure to minimize these distances, but the results proved that any change in {\it Shims} or {\it mpShims} outcomes could turn the solutions unfeasible due to the CG balance constraints.
+This hypothesis was confirmed by the results on Table \ref{tab:ground}, although in Scenario 3 this procedure could not improve in any of the volume surplus instances (1.2, 1.5, and 2.0).
 

### 2023-03-25 main.py

@@ -11,11 +11,11 @@ import optcgcons
 import mipGRB
 from plots import TTT
 
-# from py3Djanet.packer import Packer
-# from py3Djanet.methods import Item
-# from py3Djanet.bin import Bin
+from py3Djanet.packer import Packer
+from py3Djanet.methods import Item
+from py3Djanet.bin import Bin
 
-from py3Druiz import Packer, Item, Bin
+# from py3Druiz import Packer, Item, Bin
 
 
 def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus, tipo, numOptDict, rampDistCG, afterDict, beforeDict):
@@ -306,8 +306,8 @@ if __name__ == "__main__":
     # plot = True
     plot = False
 
-    scenarios = [1,2,3,4,5,6]
-    # scenarios = [3] # infeasible solutions with Shims............
+    # scenarios = [1,2,3,4,5,6]
+    scenarios = [6] # infeasible solutions with Shims............
 
     surplus   = "data20"
     # surplus   = "data50"

### 2023-03-25 latex/comp_or.tex
Paper CompOR update

@@
 	\item in the larger aircraft, the CG of the payload must be at a maximum lateral distance of $limit^{CG}_{lat}$ from the CG of the aircraft;
-	\item in the larger aircraft, pallets are distributed in two identical rows (with odd and even indices, respectively), and their centroids are at a distance $d^{CG}_{pallet}$ from the center-line of the aircraft.
+	\item in the larger aircraft, pallets are distributed in two identical rows (with odd and even indices, respectively), and their centroids are at a distance $p_i.D^{lat}$ from the center-line of the aircraft.
 \end{itemize}
@@
 
-\section{The mathematical modeling}
+\section{The mathematical model}
 \label{sec4}
@@
 
-Let $L = \{ l_0, l_1, \ldots, l_K \}$ be the set of $K+1$ nodes (or destinations), where $l_0$ is the origin and end of a flight plan.
-Let $d(l_i,l_j)$ be the distance from $l_i$ to $l_j$, where $0 \leq i,j \leq K$.
-By definition, $d(l_i,l_i)=0$.
-Let $L_k$ be the set of remaining nodes when the aircraft is in $l_k$, $0 \leq k \leq K$.
-Therefore, $L_0=L$ and $L_K = \{ l_0 \}$.
-
 Let $C=\{c_{ij}\}$ be the cost matrix of flights, where $c_{ij} = c_d*d(l_i,l_j), 0 \leq i,j \leq K$.
@@
 
-{\color{blue} $k$ is the node index and $\pi_k$ is the node in position $k$ of tour $\pi$. $\tau^{\pi_k}$ is the total torque applied by the loaded pallets to the aircraft in this node. }
+{\color{blue}
+$k$ is the node index and $\pi_k$ is the node in position $k$ of tour $\pi$. $\tau^{\pi_k}$ is the total torque applied by the loaded pallets to the aircraft in this node.
+
+Let $L = \{ \pi_0, \pi_1, \ldots, \pi_K \}$ be the set of $K+1$ nodes, where $\pi_0$ is the origin and end of a flight path.
+Let $d(\pi_i,\ \pi_j)$ be the distance from $\pi_i$ to $\pi_j$, where $0 \leq i,\ j \leq K$.
+By definition, $d(\pi_i,\ \pi_i)\ =\ 0$.
+Let $L^{\pi_k}$ be the set of remaining nodes when the aircraft is in $\pi_k$, $0 \leq k \leq K$. Therefore, $L^{\pi_0}\ =\ L$ and $L^{\pi_K}\ =\ \{ \pi_0 \}$.
+}
 
 Let $M^{\pi_k} = \{p_1, p_2, \ldots, p_m \}$ the set of $m$ pallets.
-Each pallet $p_i$, $1 \leq i \leq m$, has weight capacity $p_i.w$, volume capacity $p_i.v$, pallet destinations $p_i.to^{\pi_k}$, $0 \leq k \leq K$, and distance to the CG of aircraft $p_i.d$.
+Each pallet $p_i$, $1 \leq i \leq m$, has weight capacity $p_i.W$, volume capacity $p_i.V$, pallet destinations $p_i.to^{\pi_k}$, $0 \leq k \leq K$, and distance to the CG of aircraft $p_i.D$.
 $p_i.to^{\pi_k}$ denotes that pallet $p_i$ may assume a different destination in each node ${\pi_k}$.
+Attributes for constant values are in uppercase.
 
@@
 \begin{equation} \label{eq:costs}
-	\tilde{c} = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} [ c^{\pi(k), \pi(k+1)}\times(1+c_g\times|\epsilon^{\pi_k}|) ] + c^{\pi(K),0}\times(1+c_g\times|\epsilon_K|)
+	\tilde{c} = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} \Big [ c^{\pi(k), \pi(k+1)}\times(1+c_g\times|\epsilon^{\pi_k}|) \Big ] + c^{\pi(K),0}\times(1+c_g\times|\epsilon_K|)
 \end{equation}
@@
 \begin{equation} \label{eq:maxW}
-	maxW = \min(Payload, \sum_{i=1}^{m}p_i.w)
+	maxW = \min(Payload, \sum_{i=1}^{m}p_i.W)
 \end{equation}
@@
 \begin{equation} \label{eq:tau}
-\tau^{\pi_k} = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.w +  \sum_{q=1}^{m} Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w)];\ k \in \{0, 1, \ldots, K\}
+\tau^{\pi_k} = \sum_{i=1}^{m} \Big [ p_i.D \times (\sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.w +  \sum_{q=1}^{m} Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w) \Big ];\ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \end{equation}
-
+{\color{blue}
 \begin{equation} \label{eq:pdp11}
-	L_0 = L; L^{\pi_k} = L^{\pi(k-1)} - \{l^{\pi_k}\}; \ k \in \{1, 2, \ldots, K\}
-\end{equation}
-
+	L^{\pi_0} = L;\ L^{\pi_k} = L^{\pi_{k-1}} - \{\pi_k\}; \ k \in \{1, 2, \ldots, K\}
+\end{equation}
+}
 \begin{equation} \label{eq:pdp13}
@@
 \begin{equation} \label{eq:LatIt}
-	LatIt^{\pi_k} = \sum_{i=1}^{m} \sum_{j=1}^{n^{\pi_k}} ( X_{ij}^{\pi_k} \times t_j^{\pi_k}.w \times (i\%2) - X_{ij}^{\pi_k} \times t_j^{\pi_k}.w \times (i+1)\%2 ); \ k \in \{0, 1, \ldots, K\}
+	LatIt^{\pi_k} = \sum_{i=1}^{m} \sum_{j=1}^{n^{\pi_k}} \Big ( X_{ij}^{\pi_k} \times t_j^{\pi_k}.w \times (i\%2) - X_{ij}^{\pi_k} \times t_j^{\pi_k}.w \times (i+1)\%2 \Big ); \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:LatCons}
-	LatCons^{\pi_k} =  \sum_{i=1}^{m} \sum_{q=1}^{m}  ( Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w \times (i\%2) - Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w \times (i+1)\%2); \ k \in \{0, 1, \ldots, K\}
+	LatCons^{\pi_k} =  \sum_{i=1}^{m} \sum_{q=1}^{m} \Big ( Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w \times (i\%2) - Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w \times (i+1)\%2 \Big ); \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:torqlat}
-	s.t.: d^{CG}_{pallet} \times | LatIt^{\pi_k} + LatCons^{\pi_k} | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: p_i.D^{lat} \times \Big | LatIt^{\pi_k} + LatCons^{\pi_k} \Big | \leq  \sum_{i=1}^{m}p_i.W \times limit^{CG}_{lat}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:torqlong}
-	s.t.: |\tau^{\pi_k}| \leq maxW \times limit^{CG}_{long};\ k \in \{0, 1, \ldots, K\}
+	s.t.: \Big | \tau^{\pi_k} \Big | \leq maxW \times limit^{CG}_{long};\ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:payload}
-	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi_k} \times a_q^k.w ) \leq maxW; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{i=1}^{m} \Big (\sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi_k} \times a_q^k.w \Big ) \leq maxW; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app2}
-	s.t.: \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m\}; \ {\pi_k} \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.w + \sum_{q=1}^{m_k} Y_{iq}^{\pi_k} \times a_q^{\pi_k}.w  \leq p_i.W; \ i \in \{1, 2, \ldots, m\}; \ {\pi_k} \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 \begin{equation} \label{eq:app3}
-	s.t.: \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.v + \sum_{q=1}^{m} Y_{iq}^{\pi_k} \times a_q^{\pi_k}.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m\}; \ k \in \{0, 1, \ldots, K\}
+	s.t.: \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times t_j^{\pi_k}.v + \sum_{q=1}^{m} Y_{iq}^{\pi_k} \times a_q^{\pi_k}.v  \leq\ p_i.V; \ i \in \{1, 2, \ldots, m\}; \ k \in \{0, 1, \ldots, K\}
 \end{equation}
@@
 The objective of this problem is to find a permutation $\pi \in S_K$\/ that maximizes the function $f_\pi(\tilde{s},\tilde{c})$ \ref{eq:maxf}.
-In this way, the flight plan will be $l_0, l_{\pi(1)}, \ldots, l^{\pi_k}, l_0$.
+In this way, the flight path will be $\pi_0, \pi_1, \ldots, \pi_K, \pi_0$.
 $\tilde{s}$\/ is the total score of transported items \ref{eq:scores} and  $\tilde{c}$\/ is the total cost of fuel consumed \ref{eq:costs}.
@@
 
-The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $\pi_k$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n^{\pi_k}} t^{\pi_k}_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.v$).
+The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $\pi_k$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n^{\pi_k}} t^{\pi_k}_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.V$).
 This parameter is passed to $ItemsGeneration$ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section.
@@
 \begin{equation} \label{eq:edge}
-	\theta^{\pi_k}_{ij}= \frac{t^{\pi_k}_j.s}{t^{\pi_k}_j.v \times 3000}\times(1-\frac{t^{\pi_k}_j.w\times|p_i.d|}{\max_s\{t^{\pi_k}_s.w\}\times\max_q\{|p_q.d|\}});\ i \in \{1,2,\ldots,m\},\ j \in \{1,2,\ldots,n_k\}
+	\theta^{\pi_k}_{ij}= \frac{t^{\pi_k}_j.s}{t^{\pi_k}_j.v \times 3000}\times(1-\frac{t^{\pi_k}_j.w\times|p_i.D|}{\max_s\{t^{\pi_k}_s.w\}\times\max_q\{|p_q.d|\}});\ i \in \{1,2,\ldots,m\},\ j \in \{1,2,\ldots,n_k\}
 \end{equation}
@@
 Edges are included in this allocation as long as they respect feasibility constraints.
-Furthermore, the volume of each pallet $p_i$\/ cannot exceed $p_i.v \times limit$, where $ 0 < limit \leq 1$\/ is a given parameter.
+Furthermore, the volume of each pallet $p_i$\/ cannot exceed $p_i.V \times limit$, where $ 0 < limit \leq 1$\/ is a given parameter.
 
@@
 
-		\State $\Delta\tau \gets t_j.w \times p_i.d$
+		\State $\Delta\tau \gets t_j.w \times p_i.D$
 
@@
 
-		\If{ ($E_N^{\pi_k} \cup \{e_{ij}^{\pi_k}\}$ is feasible) {\bf and} ($volume_i \leq p_i.v \times limit$) {\bf and} ($\tau_{new} \leq \tau_{max} \times ts$ )} \label{greedy:ts}
+		\If{ ($E_N^{\pi_k} \cup \{e_{ij}^{\pi_k}\}$ is feasible) {\bf and} ($volume_i \leq p_i.V \times limit$) {\bf and} ($\tau_{new} \leq \tau_{max} \times ts$ )} \label{greedy:ts}
 
@@
 \begin{equation} \label{eq:distobjshims}
-	\mbox{minimize}\ \Bigg |  \sum_{i=1}^{m} \sum_{q=1}^{m} H^{\pi_k}_{iq} \times  (140 + a^{\pi_k}_q.w) \times p_i.d\ \Bigg | 
+	\mbox{minimize}\ \Bigg |  \sum_{i=1}^{m} \sum_{q=1}^{m} H^{\pi_k}_{iq} \times  (140 + a^{\pi_k}_q.w) \times p_i.D\ \Bigg | 
 \end{equation}
@@
 
-Equation \ref{eq:distobjshims} minimizes the aircraft absolute torque in node $\pi_k$, where 140 is the weight of an empty pallet, $a^{\pi_k}_q.w$ the weight of the consolidated, and $p_i.d$ the distance between the pallet centroid to the aircraft CG; Equation \ref{eq:distconsshims} states that each consolidated will be assigned to exactly one pallet; and Equation \ref{eq:distpalletsshims} states that each pallet will receive one consolidated.
+Equation \ref{eq:distobjshims} minimizes the aircraft absolute torque in node $\pi_k$, where 140 is the weight of an empty pallet, $a^{\pi_k}_q.w$ the weight of the consolidated, and $p_i.D$ the distance between the pallet centroid to the aircraft CG; Equation \ref{eq:distconsshims} states that each consolidated will be assigned to exactly one pallet; and Equation \ref{eq:distpalletsshims} states that each pallet will receive one consolidated.
 
@@
 \begin{equation} \label{eq:distobj}
-	\mbox{minimize} \sum_{i=1}^m \sum_{q=1}^m Z^{\pi_k}_{iq} \times ( maxD - p_i.d) ,\ \mbox{{\bf if}}\ a^{\pi_k}_q.to = \pi(k+1) \}
+	\mbox{minimize} \sum_{i=1}^m \sum_{q=1}^m Z^{\pi_k}_{iq} \times ( maxD - p_i.D) ,\ \mbox{{\bf if}}\ a^{\pi_k}_q.to = \pi(k+1) \}
 \end{equation}
@@
 \begin{equation} \label{eq:distpalletw}
-	p_i.w = 140 + a^{\pi_k}_q.w ;\ i \in \{1,2,\ldots,m\};\ q \in \{1,2,\ldots,m\}
+	p_i.W = 140 + a^{\pi_k}_q.w ;\ i \in \{1,2,\ldots,m\};\ q \in \{1,2,\ldots,m\}
 \end{equation}
@@
 \begin{equation} \label{eq:disttorque}
-	s.t.:\ | \sum\ p_i.d * p_i.w\ | \leq maxW \times limit^{CG}_{long} ;\ i \in \{1,2,\ldots,m\}
+	s.t.:\ | \sum\ p_i.D * p_i.W\ | \leq maxW \times limit^{CG}_{long} ;\ i \in \{1,2,\ldots,m\}
 \end{equation}

### 2023-03-27 latex/comp_or.tex
Paper Comp OR update

@@
 
-\subsection{Results obtained}
-
 A relevant problem with meta-heuristics is the parameter's adjustment to extract the best possible efficiency in the optimization.
@@
 
-According to \citet[p.271]{Breshears2009}, a {\it Process} is the operating system’s spawned and controlled entity that encapsulates an executing application.
+According to \cite[p.271]{Breshears2009}, a {\it Process} is the operating system’s spawned and controlled entity that encapsulates an executing application.
 A process has two main jobs: the first is to hold the application's resources, and the second is to carry out the application's instructions.
@@
 
-We ran Algorithm \ref{alg:main} in the 6 scenarios described in Table \ref{tab:scenarios}, considering 3 methods for node-by-node resolution: the {\it Gurobi} MIP solver, {\it Shims}, and {\it mpShims} (Algorithm \ref{alg:mpShims}).
+\subsection{Results obtained}
+
+We ran Algorithm \ref{alg:main} in the 6 scenarios described in Table \ref{tab:scenarios}, considering 3 methods for node-by-node resolution: the state-of-the-art {\it Gurobi} MIP solver, {\it Shims}, and {\it mpShims} (Algorithm \ref{alg:mpShims}).
 
@@
 The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
-	\item {\bf Speedup}: ratio of the sums of the worst runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
+	\item {\bf Speedup}: ratio of the sums of the worst run times of all scenarios and the sum of the method run times in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
 \end{itemize}
@@
 	\centering
-	\caption{Volume surpluses, methods and scenarios results}  \label{tab:results}
+	\caption{Volume surpluses, methods and scenarios results {\bf before} 3D packing}  \label{tab:results}
 	\footnotesize
@@
 
-		\multirow{6}{*}{$1.2$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 12.23}&{\it 8.67} &{\it 12.67}&{\it 15.36}&{\it 16.29}&{\it 16.20}&  \specialcell{{\bf 1.00}\\-} \\%  8142
-
+		\multirow{6}{*}{$1.2$}&$Relaxed$                 & $f$         &{\it 12.23}&{\it 8.67} &{\it 12.67}&{\it 15.36}&{\it 16.29}&{\it 16.20}&  \specialcell{{\bf 1.00}\\-} \\%  8142
+		&                      $MIP$                     & {\it run time (s)} &     3     &     13    &     20    &     78   &      198  &     1103  &  1.00 \\% 439							
 		\cmidrule{2-10}		                       
@@
 
-		\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 17.01}&{\it 12.45}&{\it 18.02}&{\it 22.03}&{\it 23.09}&{\it 22.35}&  \specialcell{{\bf 1.00}\\-} \\%11495
+		\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$MIP solver$}& $f$         &{\it 17.01}&{\it 12.45}&{\it 18.02}&{\it 22.03}&{\it 23.09}&{\it 22.35}&  \specialcell{{\bf 1.00}\\-} \\%11495
 
@@
 		\midrule
-		\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$Gurobi$}& $f$         &{\it 24.56}&{\it 17.47}&{\it 25.19}&{\it 27.94}&{\it 29.90}&{\it 27.60 }&  \specialcell{{\bf 1.00}\\-} \\%15266
+		\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$MIP solver$}& $f$         &{\it 24.56}&{\it 17.47}&{\it 25.19}&{\it 27.94}&{\it 29.90}&{\it 27.60 }&  \specialcell{{\bf 1.00}\\-} \\%15266
 
@@
 
+Table \ref{tab:results2} presents the results of the MIP solver compared to {\it Shims} and {\it mpShims}, after the post-optimization with the multiprocessing 3D packing.
+
+\vspace{2.0mm}
+
+\begin{table}[H]
+	\centering
+	\caption{Volume surpluses, methods and scenarios results {\bf after} 3D packing}  \label{tab:results2}
+	\footnotesize
+
+\color{blue}{
+
+	%\scriptsize
+	\begin{tabular}{crcccccccc}
+		\toprule
+		{\bf Volume}     &  {\bf Resolution}   &                              &\multicolumn{6}{c}{\bf Scenarios}                                                   &{\bf Normalized}\\
+		\cmidrule{4-9}		
+		{\it surplus}    & {\it method}        & {\bf Results}                &{\bf 1}    &{\bf 2}    &{\bf 3}    &{\bf 4}    &{\bf 5}    &{\bf 6}    &{\bf Speedup} \\
+		\toprule
+
+		%\specialcell{$Volume$\\$surplus$}
+
+		\multirow{6}{*}{$1.2$}&\specialcell{$Relaxed$\\$MIP solver$}& $f$            &{\it 11.47}&{\it 7.24} &{\it 10.78}&{\it 12.06}&{\it 12.97}&{\it 12.44}&  \specialcell{{\bf 1.00}\\-} \\%  8142
+		&                                                & {\it run time (s)} &     6     &     25    &     41    &     158   &      401  &     2161  &  1.00 \\% 439					
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $Shims$}                       & $f$                &    9.05   &   5.11    &   9.75    &   10.96   &   12.86   &   13.25   &  0.96 \\% 7817
+		&                                                & {\it run time (s)} &     16    &     68    &     94    &    323    &    719    &    3549   &  1.00 \\% 439			
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $mpShims$}                     & $f$                &    12.14  &   7.29    &   10.36   &   12.63   &   14.04   &   13.92   &  0.90\\% 7278
+		&                                                & {\it run time (s)} &     9     &     36    &     54    &     192   &     427   &    2118   &  {\bf 2.80}\\%
+
+		\midrule
+
+		\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$MIP solver$}& $f$     &{\it 15.83}&{\it 10.08}&{\it 15.14}&{\it 16.87}&{\it }&{\it }&  \specialcell{{\bf 1.00}\\-} \\%11495  
+		&                                                & {\it run time (s)} &    10     &     43    &     72    &     262   &     58    &     352   &  1.00 \\% 439					
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $Shims$}                       & $f$                &   11.55   &   7.61   &   12.28   &   15.00   &   17.55   &   17.57   &  0.96  \\%  11012
+		&                                                & {\it run time (s)} &    25     &    107   &     156   &    498    &    1217   &   5743     &  1.00 \\% 260
+
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $mpShims$}                     & $f$                &   17.05  &   10.01   &   15.01  &   16.75   &   20.09   &      &  0.87 \\% 9966
+		&                                                & {\it run time (s)} &    18    &    64     &     101  &    351    &   828     &  120      &  {\bf 1.70}\\% 151
+
+		\midrule
+		\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$MIP solver$}& $f$            &{\it }&{\it }&{\it }&{\it }&{\it }&{\it  }&  \specialcell{{\bf 1.00}\\-} \\%15266  
+		&                                                & {\it run time (s)} &     < 1   &     3     &     5     &     20    &     58    &     352   &  1.00 \\% 439					
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $Shims$}                       & $f$                &      &      &      &       &       &       &   0.96  \\%  14666
+		&                                                & {\it run time (s)} &     < 1   &     2     &     4     &     14     &     35     &   220     &  1.00 \\% 260
+
+		\cmidrule{2-10}		                       
+		&\multirow{2}{*}{ $mpShims$}                     & $f$                &      &      &      &       &       &          &  0.94  \\% 14374
+		&                                                & {\it run time (s)} &     < 1   &     2     &     3     &     9      &   26       &  151      &  {\bf 1.70}\\% 151
+
+		\bottomrule
+
+	\end{tabular}
+}
+	%\textsuperscript{asdf sd .}	
+	\normalsize
+\end{table}
+
 The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.
 
-We solved all scenarios and instances with {\it Gurobi} in a relaxed linear model, {\it Relaxed Gurobi}.
+We solved all scenarios and instances with a MIP solver in a relaxed linear model.
 This is good for the reader to have an idea of the real performance differences between serial and parallel {\it Shims}.
 
-{\it Shims} had a quality performance very close to the {\it Relaxed Gurobi} upper bound solution, and, as to the {\bf Speedup}, on average, {\it mpShims} was 2 times faster than {\it Shims}.
+{\it Shims} had a quality performance very close to the {\it Relaxed MIP solver} upper bound solution, and, as to the {\bf Speedup}, on average, {\it mpShims} was 2 times faster than {\it Shims}.
 Theoretically, it could be more, but due to the burden of sharing, for reading and writing, the CG deviation and the list of items among the processes (pallets), the {\it race condition} management limited {\it mpShims} performance.
@@
 \midrule
-$1$ &   1 &   15 & 15 \\
-$2$ &   5 &   76 & 15 \\
-$3$ &   8 &   97 & 12 \\
-$4$ &  35 &  330 &  9 \\ 
-$5$ &  98 &  740 &  8 \\    
-$6$ & 591 & 3709 &  6 \\                                                                  
+$1$ &   1 &   15 & 15 \\ % 1	 9   , 3	 6, 1	 25	
+$2$ &   5 &   76 & 15 \\ % 2	 36  , 13	 25
+$3$ &   8 &   97 & 12 \\ % 3	 54  , 20	 41	
+$4$ &  35 &  330 &  9 \\ % 12	 192 , 78	 158
+$5$ &  98 &  740 &  8 \\ % 34	 427 , 198	 401
+$6$ & 591 & 3709 &  6 \\ % 202	 2118, 1103	 2161                                                       
 \bottomrule
@@
 \midrule
-$1$ &  32.7 & 28.7 & -12.3 \\
-$2$ & 193.3 & 188.9 & -2.3 \\
-$3$ & 152.0 & 152.0 &  0.0 \\
-$4$ & 134.0 & 130.7 & -2.4 \\ 
-$5$ & 118.0 & 116.1 & -1.6 \\    
-$6$ & 105.2 & 104.8 & -0.4 \\                                                                  
+$1$ &  32.7 & 28.7 & -12.3 \\ % CBC 32.1 & 31.3 & -2.4 Shims 34.0 & 29.5 & -13.1, 31.5 & 29.9 & -5.1
+$2$ & 193.3 & 188.9 & -2.3 \\ % mpShims 212.0 & 188.9 & -10.9
+$3$ & 165.1 & 153.3 & -7.1 \\ % 
+$4$ & 134.0 & 130.7 & -2.4 \\ % CBC 137.1 & 132.0 & -3.7 mpShim 143.4 & 136.6 & -4.7
+$5$ & 118.0 & 116.1 & -1.6 \\ % CBC 124.9 & 116.3 & -6.9 mpShim 126.7 & 122.0 & -3.8 
+$6$ & 105.2 & 104.8 & -0.4 \\ % CBC 117.4 & 107.6 & -8.3 mpShim 115.3 & 110.4 & -4.2                                                                
 \bottomrule

### 2023-03-27 latex/comp_or.tex
paper CompOR updtate

@@
 
-In line \ref{tour:minRampDist}, $minRampDist$ minimizes the distance of the next node destined pallets, keeping the CG in its operational range.
-This procedure is implemented in integer programming as presented in Subsection \ref{improvement}, according to equations \ref{eq:distobj} to \ref{eq:disttorque}.
-
-$ScoreAndDeviation$ evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.
+{\color{purple}
+In line \ref{tour:3dpacking}, $3Dpacking(G)$ packs the items previously assigned to a pallet, which is feasible in terms of volume, weight, and torque but still not guaranteed to fit into the pallet.
+In this procedure, some unfit items are excluded from the solution.
+}
+
+{\color{blue}
+In line \ref{tour:minRampDist}, $minRampDist(G)$ minimizes the distance of the next node destined pallets, keeping the CG in its operational range.
+This procedure is implemented in integer programming.
+This method does not affect palletization or the objective function value.
+}
+
+$ScoreAndDeviation(\pi(k), G_4)$ evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.
 It consists of a loop that goes through all the pallets, accumulating the scores and the torques of the shipped items, allowing the final calculation of the CG deviation.
@@
 
+In this work, we had to select the {\it fully-connected} topology because the list of items to embark and the aircraft torque must be accessible for reading and writing by all child processes.
+
 \subsection{Multiprocessing Shims - mpShims}
 
-Finally, we present a new heuristic designed specifically for ACLP+RPDP, which we named \emph{mpShims}.
+Finally, we present a new multiprocessing heuristic designed specifically for ACLP+RPDP, which we named \emph{mpShims}.
 Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}).
@@
 			\draw[->] (.2, .6) -- coordinate (x axis mid) (12, .6);
-			\node at (5, 0.3) {$\eta_1[i]$};
+			\node at (5, 0.3) {$\eta_1^{\ i}$};
 			\node at (5, 2.5) {|};
@@
 			\node at (7, 2.5) {|};
-			\node at (7, 0.3) {$\eta_2[i]$};
+			\node at (7, 0.3) {$\eta_2^{\ i}$};
 			\node at (0.3, 5) {$\theta_{ij}^{\pi_k}$};
@@
 Figure \ref{fig:whip} represents the $n^{\pi_k}$\/ possible edges $e_{ij}^{\pi_k}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi_k}$.
-\emph{mpShims}\/ starts with a greedy solution for each pallet, stopping at the edge with index in $\eta_1^i$ close to the local optimum (first phase).
-Then, performs an integer programming procedure to minimize the CG deviation (second phase), and finally, considering even the edge with the index $\eta_2^i$, it elaborates different possible complements for this pallet and selects the best of these complements (third phase).
+\emph{mpShims}\/ starts with a greedy solution for each pallet, stopping at the edge with index in $\eta_1^{\ i}$ close to the local optimum (first phase).
+Then, performs an integer programming procedure to minimize the CG deviation (second phase), and finally, considering even the edge with the index $\eta_2^{\ i}$, it elaborates different possible complements for this pallet and selects the best of these complements (third phase).
 
 Although the third phase solves a {\it Knapsack Problem}, a faster algorithm for solving a {\it Bin Packing Problem} was used ({\it the First Fit Decreasing - FFD}), and the most profitable bin ({\it Shims}) was selected.
-This method was proved to be faster than a {\it Dynamic Programming} solution for the {\it Knapsack Problem}.
-
-	In this work, we included a new phase after the first: a minimization of the CG deviation. An integer programming procedure that helps improve the overall {\it mpShims} results.
 
@@
 
-$3^{st}$ - {\it Shims} set construction and best {\it Shims} selection
-ess run-time, we considered each pallet as a parallel process, taking advantage of nowadays multi-core computers.
+$3^{st}$ - {\it Shims} set construction and best {\it Shims} selection.
+
+In the first and third phases, we considered each pallet as a parallel process, taking advantage of nowadays multi-core computers.
 
@@
 
-		\State $ts \gets 2.0$ \Comment{torque surplus} \label{mpShims:ts}
+		\State $ts \gets 1.2$ \Comment{torque surplus} \label{mpShims:ts}
 
@@
 Line \ref{mpShims:ts} defines the torque surplus ($ts$).
-For the greedy phase, the solution may generate torque infeasible solutions with doubled torque ($2.0$), which will be repaired by the next phase, the minimization of the CG deviation.
+For the $Greedy$ phase, the solution may generate temporary torque unfeasible solutions.
+We raised the Hypothesis (2) that, in sequential mode, where one pallet is built at a time, torque constraints may be more restrictive than volume or weight constraints, which could stop the palletization process earlier, fostering a solution that may be not as good as it could be.
+So, we decided to extend torque limit and balance cargo before the next {\it Shims} phase to allow more room for pallet fulfillment.
 
@@
 
-Line \ref{mpShims:minCGdev} calls the $minCGdev$ integer programming method (Subsection \ref{subsection:second_phase}) which updates the $nodeTorque$ and $G$.
-$minCGdev$ minimizes the CG deviation.
-We raise the hypothesis that, in sequential mode, where one pallet is built at a time, torque constraints may be more restrictive than volume or weight constraints, which could stop the palletization process earlier, fostering a solution that may be not as good as it could be.
-So, we decided to balance cargo before the next {\it Shims} phase to allow more room for pallet fulfillment.
-This will be confirmed (or not) in Section \ref{sec6}.
-
-Line \ref{mpShims:getBestShims} calls the $getBestShims$ method which updates $p_i$, $N^{\pi_k}$, $nodeTorque$, and $G$.
-
-Line \ref{mpShims:join2} wait for all $getBestShims$ processes to finish.
+Line \ref{mpShims:minCGdev} calls the $minCGdev()$ integer programming method (Subsection \ref{subsection:second_phase}) which updates the $nodeTorque$ and $G$.
+$minCGdev()$ minimizes the CG deviation, bringing it back to its operational range.
+A proper value for $ts$ must be empirically found to guarantee this intent (the {\it iRace} tool (\cite{LopezIbanezManuel2016}) is good for this value determination).
+
+Line \ref{mpShims:getBestShims} calls the $getBestShims()$ method which updates $p_i$, $N^{\pi_k}$, $nodeTorque$, and $G$.
+
+Line \ref{mpShims:join2} wait for all $getBestShims()$ processes to finish.
 
@@
 
-\subsubsection{{\it mpShims} first phase}
+\subsubsection{The {\it mpShims} first phase}
 
@@
 			\State $volume_i \gets volume_i + t_j^{\pi_k}.v$
-			\State $\eta_1^i \gets \eta_1^i + 1$ 
+			\State $\eta_1^{\ i} \gets \eta_1^{\ i} + 1$ 
 			\State $\tau^{\pi_k} \gets \tau^{\pi_k} + \Delta\tau$
@@
 
-Line \ref{greedy:ts} presents the variable $ts$ (torque $\tau$ surplus), that is an extra-torque limit to allow each greedy pallet construction to, temporally, exceed the torque limit, that will be turned feasible again by the next {\it mpShims}' phase, the CG minimization phase.
-$ts$ value will be adjusted empirically, by means of the {\it iRace} tool (\cite{LopezIbanezManuel2016}), to produce feasible solutions in all problem instances.
-
-\subsubsection{{\it mpShims} second phase}
+Line \ref{greedy:ts} presents the variable $ts$ (torque surplus), that is an extra-torque limit to allow each greedy pallet construction to, temporally, exceed the torque limit, that will be turned feasible again by the next {\it mpShims}' phase, the CG minimization phase.
+
+\subsubsection{The {\it mpShims} second phase}
 \label{subsection:second_phase}
@@
 
-Equation \ref{eq:distobjshims} minimizes the aircraft absolute torque in node $\pi_k$, where 140 is the weight of an empty pallet, $a^{\pi_k}_q.w$ the weight of the consolidated, and $p_i.D$ the distance between the pallet centroid to the aircraft CG; Equation \ref{eq:distconsshims} states that each consolidated will be assigned to exactly one pallet; and Equation \ref{eq:distpalletsshims} states that each pallet will receive one consolidated.
-
-\subsubsection{{\it mpShims} third phase}
+Equation \ref{eq:distobjshims} minimizes the aircraft absolute torque in node $\pi_k$, where 140kg is the weight of an empty pallet, $a^{\pi_k}_q.w$ the weight of the consolidated, and $p_i.D$ the distance between the pallet centroid to the aircraft CG; Equation \ref{eq:distconsshims} states that each consolidated will be assigned to exactly one pallet; and Equation \ref{eq:distpalletsshims} states that each pallet will receive one consolidated.
+
+\subsubsection{The {\it mpShims} third phase}
 
@@
 
-		\For{$x \gets \eta_1^i$ to $\eta_2^i$} \label{edges_indexes}
+		\For{$x \gets \eta_1^{\ i}$ to $\eta_2^{\ i}$} \label{edges_indexes}
 
 		\State $NewShims \gets$ {\bf True} \label{new_shims}
-		\State $e_{ij}^{\pi_k} \gets E_x^i$
+		\State $e_{ij}^{\pi_k} \gets E_x^{\ i}$
 
@@
 
-From a list of items, items are sorted from the biggest to the smallest and placed in such an order on a pallet.Each item has 1 to 6 orientations to choose from when it can be placed.
+\begin{itemize}
+	\item[1] From a list of items assigned to a pallet, items are sorted from the biggest to the smallest and placed in such an order on a pallet.
+Each item has 1 to 6 orientations to choose from when it can be placed.
 The orientation procedure can select the best direction type among possible orientations.
-
-A pivot point is used to determine item's position. The pivot is an (x, y, z) coordinate which represents a point in the pallet at which an attempt to pack an item will be made.
-
-The back lower left corner of the item will be placed at the pivot.
+	\item[2] A pivot point is used to determine item's position. The pivot is an (x, y, z) coordinate which represents a point in the pallet at which an attempt to pack an item will be made.
+	\item[3] The back lower left corner of the item will be placed at the pivot.
 If the item cannot be packed at the pivot position then it is rotated until it can be packed at the pivot point or until we have tried all 6 possible rotation types.
-
-If, after rotating it, the item still cannot be packed at the pivot point, then we move on to packing another item and add the unpacked item to a list of items that will be packed after an attempt to pack the remaining items is made.
-
-The first pivot in the empty pallet is always (0,0,0). When one item can be placed into multiple optimal pivot point, the placement selection module can help make a choice.
-
-Let $n$ be the number of items and $m$ the number of pallets.
-
-Let $A_{j}\ =\ 1$ if item $j$ is assigned to the pallet, $A_{j}\ =\ 0$ otherwise, for $j \in \{1,\ 2,\ 3,\ ...,\ n\}$.
-
-Let $X_{ij}\ =\ 1$ if item $j$ was already assigned to pallet $i$ by {\it mpShims}, $X_{ij}\ =\ 0$ otherwise, for $i \in \{1,\ 2,\ 3,\ ...,\ m\}$.
-
-Let $d_j,\ w_j,\ h_j$ be parameters indicating the depth, width, and height of an item. 
-Let $D_i,\ W_i,\ H_i$ be parameters indicating the depth, width, and height of the pallet.
+	\item[4] If, after rotating it, the item still cannot be packed at the pivot point, then we move on to packing another item and add the unpacked item to a list of items that will be packed after an attempt to pack the remaining items is made.
+	\item[5] The first pivot in the empty pallet is always (0, 0, 0). When one item can be placed into multiple optimal pivot point, the placement selection module can help make a choice.
+\end{itemize}
+
+Let $n^i$ be the number of items assigned to pallet $i$ and $m$ the number of pallets.
+
+Let $A_{j} = 1$ if item $t_j$ is assigned to the pallet, $A_{j}\ =\ 0$ otherwise, for $j \in \{1,\ 2,\ 3,\ ...,\ n^i\}$.
+
+Let $X_{ij} = 1$ if item $j$ was already assigned to pallet $i$ by {\it mpShims}, $X_{ij}\ =\ 0$ otherwise.
+
+Let $t_j.d,\ t_j.wh,\ t_j.h$ be parameters indicating the depth, width, and height of an item.
+
+Let $p_i.Dh,\ p_i.Wh,\ p_i.H$ be parameters indicating the depth, width, and height of the pallet.
+We included letter $h$ at the end of parameters $p_i.Dh$ and $p_i.Wh$ to turn them different from $p_i.D$ (distance from CG) and $p_i.W$ (weight limit).
 
@@
 
-Let $o_j^1,\ o_j^2,\ o_j^3,\ o_j^4,\ o_j^5,\ o_j^6$ be binary variables, indicating the possible orientations for item $j$.
-One of the values equal to 1 if item $j$ has orientation in $\{1,\ 2,\ 3,\ 4,\ 5,\ 6\}$, respectively, otherwise equal to 0.
+Let $t_j.o^1,\ t_j.o^2,\ t_j.o^3,\ t_j.o^4,\ t_j.o^5,\ t_j.o^6$ be binary variables, indicating the possible orientations for item $t_j$.
 
 \begin{equation} \label{eq:3dpacking1}
-	\mbox{minimize} \Big [ D_i \times W_i \times H_i - \sum_{j \gets 1}^{n} X_{ij} \times A_{j} \times d_j \times w_j \times h_j \Big ]
+	\mbox{minimize} \Big [ p_i.Dh \times p_i.Wh \times p_i.H - \sum_{j \gets 1}^{n^{\ i}} X_{ij} \times A_{j} \times t_j.d \times t_j.wth \times t_j.h \Big ]
 \end{equation}
@@
 \begin{equation} \label{eq:3dpacking2}
-	\sum_{j \gets 1}^{n} A_{j} \times d_j  \leq X_{ij} \times D_i,\ i \in \{1,\ 2,\ 3,\ ...,\ m\}
+	\sum_{j \gets 1}^{n^{\ i}} A_{j} \times t_j.d  \leq X_{ij} \times p_i.D\ i \in \{1,\ 2,\ 3,\ ...,\ m\}
 \end{equation}
@@
 \begin{equation} \label{eq:3dpacking3}
-	\sum_{j \gets 1}^{n} A_{j} \times w_j  \leq X_{ij} \times W_i,\ i \in \{1,\ 2,\ 3,\ ...,\ m\}
+	\sum_{j \gets 1}^{n^{\ i}} A_{j} \times t_j.wh  \leq X_{ij} \times p_i.Wh,\ i \in \{1,\ 2,\ 3,\ ...,\ m\}
 \end{equation}
@@
 \begin{equation} \label{eq:3dpacking4}
-	\sum_{j \gets 1}^{n} A_{j} \times h_j  \leq X_{ij} \times H_i,\ i \in \{1,\ 2,\ 3,\ ...,\ m\}
-\end{equation}
+	\sum_{j \gets 1}^{n^{\ i}} A_{j} \times t_j.h  \leq X_{ij} \times p_i.H,\ i \in \{1,\ 2,\ 3,\ ...,\ m\}
+\end{equation}
+
+\begin{equation} \label{eq:3dpacking5}
+t_j.o^1 + t_j.o^2 + t_j.o^3 + t_j.o^4 + t_j.o^5 + t_j.o^6 = 1
+\end{equation}
+
+Equation \ref{eq:3dpacking1} minimizes the volume slack; equations \ref{eq:3dpacking2}, \ref{eq:3dpacking3}, and \ref{eq:3dpacking4} constrains the packing to the three dimensions; and \ref{eq:3dpacking5} guarantees that only one orientation is chosen for the item.
 
@@
 
-In line \ref{3dpacking:3}, $N^{\pi_k}_i$ is the set of items previously assigned to pallet $i$.
+In line \ref{3dpacking:3}, $N^{\pi_k}_i$ is the set of items previously assigned to pallet $i$ and the method $3Dpacker()$ conforms to equations \ref{eq:3dpacking1}, \ref{eq:3dpacking2}, \ref{eq:3dpacking3}, \ref{eq:3dpacking4}, and \ref{eq:3dpacking5}.
 
@@
 \begin{equation} \label{eq:distpalletw}
-	p_i.W = 140 + a^{\pi_k}_q.w ;\ i \in \{1,2,\ldots,m\};\ q \in \{1,2,\ldots,m\}
+	p_i.w = 140 + a^{\pi_k}_q.w ;\ i \in \{1,2,\ldots,m\};\ q \in \{1,2,\ldots,m\}
 \end{equation}
@@
 \begin{equation} \label{eq:disttorque}
-	s.t.:\ | \sum\ p_i.D * p_i.W\ | \leq maxW \times limit^{CG}_{long} ;\ i \in \{1,2,\ldots,m\}
-\end{equation}
-
-Equation \ref{eq:distobj} minimizes the sum of distances of the consolidated in node $\pi_k$ destined to the next node $\pi(k+1)$; Equation \ref{eq:distcons} states that each consolidated will be assigned to exactly one pallet; Equation \ref{eq:distpallets} states that each pallet will receive one consolidated; Equation \ref{eq:distpalletw} calculates the total weight of each pallet; and Equation \ref{eq:disttorque} guarantees that the torque limit will not be exceeded.
+	s.t.:\ | \sum\ p_i.D * p_i.w\ | \leq maxW \times limit^{CG}_{long} ;\ i \in \{1,2,\ldots,m\}
+\end{equation}
+
+Equation \ref{eq:distobj} minimizes the sum of distances of the consolidated in node $\pi_k$ destined to the next node $\pi(k+1)$;
+
+Equation \ref{eq:distcons} states that each consolidated will be assigned to exactly one pallet;
+
+Equation \ref{eq:distpallets} states that each pallet will receive one consolidated;
+
+Equation \ref{eq:distpalletw} calculates the total weight on each pallet (note that it is different from $p_i.W$, the capacity of the pallet); and
+
+Equation \ref{eq:disttorque} guarantees that the torque limit will not be exceeded.
 
@@
 The final experiments are performed on a 64-bit, 16GiB, 3.4GHz, 8-core Intel® i7-3770 CPU, 2 threads per core, with {\it Linux Ubuntu 22.04 LTS} as the operational system and {\it Python 3.10.6} as the programming language.
+
+\subsection{Parameters tuning}
 
@@
 That is why all parameters were set using the $iRace$ package \cite{LopezIbanezManuel2016}.
-Supplying $iRace$ with a range for $volThreshold$ as [0.80 to 0.99], {\it iRace} yielded 0.92, by training on the uneven instances numbers.
+Supplying $iRace$ with a range for $volThreshold$ as $[0.80,\ 0.99]$, {\it iRace} yielded 0.92, by training on the uneven instances numbers.
 The values tested for 20\%, 50\%, and 100\% volume surpluses were very close, considering the confidence level offered by {\it iRace}.
 
-For parallelization, we used the {\it Python Multiprocessing} package, which implements {\it process-based parallelism} or {\it parallel shared memory algorithm}.
-According to \cite{multiprocessing}, {\it Multiprocessing} is a package that supports spawning processes using an API similar to the {\it Python Threading} package.
-
-According to \cite[p.271]{Breshears2009}, a {\it Process} is the operating system’s spawned and controlled entity that encapsulates an executing application.
-A process has two main jobs: the first is to hold the application's resources, and the second is to carry out the application's instructions.
-
-It is known that working concurrently opens up synchronization issues.
-But the {\it Multiprocessing} package ({\it mp}) offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
-Because of this, the {\it Multiprocessing} module lets the programmer take full advantage of the fact that a machine has more than one processor, normally capable of 2 threads each.
+For the {\it torque surplus - ts}, we input {\it iRace} with the range $[1.0,\ 2.0]$ and this tool chose $ts = 1.7$ as a value that brings torque of the $Greedy$ first phase to a feasible region.
+
+\subsection{Multiprocessing}
 
@@
 In our case, the reading and writing process on the shared torque variable and the list of items to be embarked.
+
+According to \cite[p.271]{Breshears2009}, a {\it Process} is the operating system’s spawned and controlled entity that encapsulates an executing application.
+
+It is known that working concurrently opens up synchronization issues.
+But the {\it Multiprocessing} package ({\it mp}) offers both local and remote concurrency, effectively side-stepping the global interpreter lock by using sub-processes instead of threads.
+Because of this, the {\it Multiprocessing} module lets the programmer take full advantage of the fact that a machine has more than one processor, normally capable of 2 threads each.
+
+As in the larger aircraft the number of pallets is bigger than the number of threads available in a common handheld computer, for parallelization, we used the {\it Python Multiprocessing} package, which implements {\it process-based parallelism} or {\it parallel shared memory algorithm}.
+According to \cite{multiprocessing}, {\it Multiprocessing} is a package that supports spawning processes using an API similar to the {\it Python Threading} package.
 
@@
 \midrule
-$1$ &  32.7 & 28.7 & -12.3 \\ % CBC 32.1 & 31.3 & -2.4 Shims 34.0 & 29.5 & -13.1, 31.5 & 29.9 & -5.1
-$2$ & 193.3 & 188.9 & -2.3 \\ % mpShims 212.0 & 188.9 & -10.9
+$1$ &  32.7 &  28.7 & -7.3 \\ % CBC 32.1 & 31.3 & -2.4 Shims 34.0 & 29.5 & -13.1, 31.5 & 29.9 & -5.1
+$2$ & 193.3 & 188.9 & -6.3 \\ % mpShims 212.0 & 188.9 & -10.9
 $3$ & 165.1 & 153.3 & -7.1 \\ % 
-$4$ & 134.0 & 130.7 & -2.4 \\ % CBC 137.1 & 132.0 & -3.7 mpShim 143.4 & 136.6 & -4.7
-$5$ & 118.0 & 116.1 & -1.6 \\ % CBC 124.9 & 116.3 & -6.9 mpShim 126.7 & 122.0 & -3.8 
-$6$ & 105.2 & 104.8 & -0.4 \\ % CBC 117.4 & 107.6 & -8.3 mpShim 115.3 & 110.4 & -4.2                                                                
+$4$ & 137.1 & 132.0 & -3.7 \\ % CBC 137.1 & 132.0 & -3.7 mpShim 143.4 & 136.6 & -4.7
+$5$ & 126.7 & 122.0 & -3.8 \\ % CBC 124.9 & 116.3 & -6.9 mpShim 126.7 & 122.0 & -3.8 
+$6$ & 115.3 & 110.4 & -4.2 \\ % CBC 117.4 & 107.6 & -8.3 mpShim 115.3 & 110.4 & -4.2                                                                
 \bottomrule
@@
 
-Table \ref{tab:ground} presents the distances minimization of the next node destined pallets to the ramp door. Columns {\bf Before} and {\bf After} refer to the application of the current method.
+Table \ref{tab:ground} presents the distances minimization of the next node destined pallets to the ramp door.
+Columns {\bf Before} and {\bf After} refer to the application of the current method.
+In average, the distances from the pallets and the last ramp door position was diminished by 5\% along a tour.
 
@@
 (4) The distances to move pallets on unloading in the next node could be minimized to benefit ground handling costs.
-This hypothesis was confirmed by the results on Table \ref{tab:ground}, although in Scenario 3 this procedure could not improve in any of the volume surplus instances (1.2, 1.5, and 2.0).
-
-A new question is raised about whether it is worth using parallelism as, although {\it mpShims} was twice as fast as {\it Shims}, the results of {\it Shims} were still better than those of {\it mpShims}.
-May this question be solved through the real-world application of these methods.
+This hypothesis was confirmed by the results on Table \ref{tab:ground}.
 
@@
 
-Other important contributions are that we checked four hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure added to {\it Shims} algorithm enhanced it's results; but the integer programming procedure to minimize pallets' travel in the cargo bay did not improve the results; and a procedure for 3-D packing was turned possible for operational use due to a process-based parallel computing heuristic.
+Other important contributions are that we verified four hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure added to {\it Shims} algorithm enhanced it's results; a procedure for 3-D packing was turned possible for operational use due to a process-based parallel computing heuristic, and the integer programming procedure to minimize pallets' travel in the cargo bay was successful.
 

### 2023-03-27 latex/comp_or.tex
Paper CompOR update

@@
 
-The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $\pi_k$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n^{\pi_k}} t^{\pi_k}_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.V$).
-This parameter is passed to $ItemsGeneration$ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section.
-
-$method$\/ corresponds to one of the heuristics that we will present in subsection \ref{methods}.
-The loop of lines \ref{main:loop1}-\ref{main:loop2} goes through all permutations $\pi$, where the node-by-node resolutions are performed by $SolveTour$, whose result is stored in $f_{\pi}$.
-The best outcome among all $K!$\/ tours will be the answer for $scenario$, $volume$\/ and $method$\/ (line \ref{main:f}).
+Argument $method$ corresponds to one of the heuristics that we will present in subsection \ref{methods}.
+
+The loop of lines \ref{main:loop1}-\ref{main:loop2} goes through all permutations $\pi$, where the node-by-node resolutions are performed by $SolveTour()$, whose result is stored in $f_{\pi}$.
+
+The best outcome among all $K!$ tours will be the answer for $scenario$, $volume$ and $method$ (line \ref{main:f}).
 
@@
 
-In addition to the set of nodes, pallets, costs and items, $SolveTour$, described in Algorithm \ref{alg:tour}, receives the parameter $method$, which corresponds to a heuristic for solving the node-by-node problems, and the parameter $\pi$, which is a permutation that defines the order of visits in this tour.
+In addition to the set of nodes, pallets, costs and items, $SolveTour$, described in Algorithm \ref{alg:tour}, receives the parameter $method$, which corresponds to a heuristic for solving the node-by-node sub-problems, and the parameter $\pi$, which is a permutation that defines the order of visits in this tour.
 
 As we mentioned in the previous section, all tours start and end at $\pi_0$\/ (lines \ref{tour:pi1}-\ref{tour:pi2}).
-After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ flights (lines \ref{tour:loop1}-\ref{tour:loop2}).
+After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ tours (lines \ref{tour:loop1}-\ref{tour:loop2}).
 Initially we set pallets destination as $-1$\/ (line \ref{tour:-1}).
 When the aircraft is at node $\pi_0$, the initial graph $G_1$\/ is empty because it has no consolidated items \ref{tour:g11}.
-Otherwise, the set $L_k$\/ of remaining nodes when departing from $\pi_k$ is updated (line \ref{tour:lk1}), and $UpdateConsolidated$\/ (line \ref{tour:dest}) returns the set of consolidated items that have not yet reached their destinations and remain on board, rearranging them on the pallets to minimize CG deviation.
+Otherwise, the set $L^{\pi_k}$\/ of remaining nodes when departing from $\pi_k$ is updated (line \ref{tour:lk1}), and $UpdateConsolidated$\/ (line \ref{tour:dest}) returns the set of consolidated items that have not yet reached their destinations and remain on board, rearranging them on the pallets to minimize CG deviation.
 This allocation is stored in graph $G_1$\/ (line \ref{tour:g12}).
@@
 			\EndIf  \label{tour:lk2}	
-			\State $M^{\pi_k} \gets SetPalletsDestination( \pi(k) )$ \label{tour:dest2}		
+			\State $M^{\pi_k} \gets SetPalletsDestinations( \pi(k) )$ \label{tour:dest2}		
 			\State $G_2 \gets SolveNode(method, \pi(k), G_1, timeLim)$ \label{tour:node}
 
-{\color{purple} \State	$G_3 \gets 3Dpacking(G_2)$  \label{tour:3dpacking} }
+{\color{purple} \State	$G_3 \gets mp3Dpacking(G_2)$  \label{tour:3dpacking} }
 
@@
 In the context of this work, we know that $m>K$, once the aircraft has $7$\/ or $18$\/ pallets and $K\leq 6$, allowing there to be at least one pallet for each node to be visited.
-$SetPalletsDestination$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands caused by the items to be embarked in the current node, without changing the pallets destination with consolidated items.
+$SetPalletsDestinations$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands caused by the items to be embarked in the current node, without changing the pallets destination with consolidated items.
 
@@
 
-$SetPalletsDestination$ sets pallets destination not yet defined.
+$SetPalletsDestinations$ sets pallets destination not yet defined.
 The total volume of items and the number of pallets with consolidated items destined for each node.
@@
 {\color{purple}
-In line \ref{tour:3dpacking}, $3Dpacking(G)$ packs the items previously assigned to a pallet, which is feasible in terms of volume, weight, and torque but still not guaranteed to fit into the pallet.
+In line \ref{tour:3dpacking}, $mp3Dpacking(G)$ packs the items previously assigned to a pallet, which is feasible in terms of volume, weight, and torque but still not guaranteed to fit into the pallet.
 In this procedure, some unfit items are excluded from the solution.
@@
 
-$UpdateConsolidated$, $MinCGDeviation$, $SetPalletsDestination$, and $ScoreAndDeviation$ are thoroughly described by \cite{MesquitaSanches2023}.
+$UpdateConsolidated$, $MinCGDeviation$, $SetPalletsDestinations$, and $ScoreAndDeviation$ are thoroughly described by \cite{MesquitaSanches2023}.
 
@@
 
-According to \cite[p.226]{manfrin2006}, there are five topologies for multiprocessing interchange of information: {\it fully-connected}, where the master process broadcasts to all remaining child processes; {\it replace-worst}, where the best-so-far solution process broadcasts only to the current worst solution process; {\it hypercube}, where processes are connected as a hypercube, and a vertex process broadcasts only to the connected vertices; {\it ring}, in which one process only sends a message to the next process connected to it; and {\it parallel independent runs (PIR)}, in which there are no communication costs and the best solution is chosen among all processes.
-They concluded that the PIR topology presented, in average, the smallest distance from the optimal solutions.
-
-In this work, we had to select the {\it fully-connected} topology because the list of items to embark and the aircraft torque must be accessible for reading and writing by all child processes.
+According to \cite[p.226]{manfrin2006}, there are five topologies for multiprocessing interchange of information: {\it fully-connected}, where the master process broadcasts to all remaining child processes; {\it replace-worst}, where the best-so-far solution process broadcasts only to the current worst solution process; {\it hypercube}, where processes are connected as a hypercube, and a vertex process broadcasts only to the connected vertices; {\it ring}, in which one process only sends a message to the next process connected to it; and {\it parallel independent runs}, in which there are no communication costs and the best solution is chosen among all processes.
+
+In this work, we had to select the {\it fully-connected} topology for the {\it Greedy} heuristics and for {\it mpShims} because the list of items to embark and the aircraft torque must be accessible for reading and writing by all child processes.
+This decision require an effective race condition management.
+
+{\it When threads or processes attempt to simultaneously access a shared resource, and the accesses can result in an error, we often say the program has a race condition, because the threads or processes are in a "race" to carry out an operation.} \cite[p.
+53]{Pacheco:2020}.
+
+For the {\it mp3Dpacking} we used {\it parallel independent runs}, because there is no memory sharing.
 
@@
 
-In the first and third phases, we considered each pallet as a parallel process, taking advantage of nowadays multi-core computers.
-
 \begin{algorithm}[H]
@@
 
-All phases being executed in a multiprocessing run-time, accessing concurrently, reading and writing the list of items and the current value of the CG deviation.
+In the first and third phases, we considered each pallet as a parallel process, taking advantage of nowadays multi-core computers, being executed in a multiprocessing run-time, accessing concurrently, reading and writing the list of items and the current value of the CG deviation.
 
@@
 
-The main logic of the 3D packing procedure is based on an approximate algorithm like follows:
+The main logic of the 3D packing procedure ($3Dpacker$) is based on an approximate algorithm like follows:
 
@@
 	\item[1] From a list of items assigned to a pallet, items are sorted from the biggest to the smallest and placed in such an order on a pallet.
-Each item has 1 to 6 orientations to choose from when it can be placed.
+Each item has 1 to 6 orientations to choose from in the moment of placement.
 The orientation procedure can select the best direction type among possible orientations.
@@
 	\item[3] The back lower left corner of the item will be placed at the pivot.
-If the item cannot be packed at the pivot position then it is rotated until it can be packed at the pivot point or until we have tried all 6 possible rotation types.
+If the item cannot be packed at the pivot position then it is rotated until it can be packed at the pivot point or until we have tried all 6 possible rotation types (first fit).
 	\item[4] If, after rotating it, the item still cannot be packed at the pivot point, then we move on to packing another item and add the unpacked item to a list of items that will be packed after an attempt to pack the remaining items is made.
-	\item[5] The first pivot in the empty pallet is always (0, 0, 0). When one item can be placed into multiple optimal pivot point, the placement selection module can help make a choice.
+	\item[5] The first pivot in the empty pallet is always (0, 0, 0). When one item can be placed into multiple optimal pivot point, the placement selection module can help make a choice (best fit).
 \end{itemize}
@@
 Let $p_i.Dh,\ p_i.Wh,\ p_i.H$ be parameters indicating the depth, width, and height of the pallet.
-We included letter $h$ at the end of parameters $p_i.Dh$ and $p_i.Wh$ to turn them different from $p_i.D$ (distance from CG) and $p_i.W$ (weight limit).
+We included letter $h$ at the end of parameters $p_i.Dh$ and $p_i.Wh$ to make them different from $p_i.D$ (distance from CG) and $p_i.W$ (weight limit).
 
@@
 \begin{equation} \label{eq:3dpacking1}
-	\mbox{minimize} \Big [ p_i.Dh \times p_i.Wh \times p_i.H - \sum_{j \gets 1}^{n^{\ i}} X_{ij} \times A_{j} \times t_j.d \times t_j.wth \times t_j.h \Big ]
+	\mbox{minimize} \Big [ p_i.Dh \times p_i.Wh \times p_i.H - \sum_{j \gets 1}^{n^{\ i}} X_{ij} \times A_{j} \times t_j.d \times t_j.wh \times t_j.h \Big ]
 \end{equation}
@@
 \begin{equation} \label{eq:3dpacking2}
-	\sum_{j \gets 1}^{n^{\ i}} A_{j} \times t_j.d  \leq X_{ij} \times p_i.D\ i \in \{1,\ 2,\ 3,\ ...,\ m\}
+	\sum_{j \gets 1}^{n^{\ i}} A_{j} \times t_j.d  \leq X_{ij} \times p_i.Dh\ i \in \{1,\ 2,\ 3,\ ...,\ m\}
 \end{equation}
@@
 \begin{equation} \label{eq:3dpacking5}
-t_j.o^1 + t_j.o^2 + t_j.o^3 + t_j.o^4 + t_j.o^5 + t_j.o^6 = 1
-\end{equation}
-
-Equation \ref{eq:3dpacking1} minimizes the volume slack; equations \ref{eq:3dpacking2}, \ref{eq:3dpacking3}, and \ref{eq:3dpacking4} constrains the packing to the three dimensions; and \ref{eq:3dpacking5} guarantees that only one orientation is chosen for the item.
+t_j.o^1 + t_j.o^2 + t_j.o^3 + t_j.o^4 + t_j.o^5 + t_j.o^6 \leq 1
+\end{equation}
+
+Equation \ref{eq:3dpacking1} minimizes the volume slack; equations \ref{eq:3dpacking2}, \ref{eq:3dpacking3}, and \ref{eq:3dpacking4} constrains the packing to the three dimensions; and \ref{eq:3dpacking5} guarantees that at most one orientation is chosen for the item.
 
@@
 \begin{algorithm}[H]
-	\caption{Parallel 3D packing procedure}  \label{alg:3dpacking}
+	\caption{Multiprocessing 3D packing procedure}  \label{alg:3dpacking}
 
@@
 
-\subsection{Minimization of the distance of a next node destined pallet to the ramp door }
+\subsection{Minimization of the unloading costs}
 
@@
 Equation \ref{eq:disttorque} guarantees that the torque limit will not be exceeded.
+
+The instruction referred in Algorithm \ref{alg:tour}, line \ref{tour:minRampDist} implements this model.
 
@@
 
-\subsection{Results obtained}
+\subsection{Results before 3D packing}
 
@@
 		\toprule
-		{\bf Volume}     &  {\bf Resolution}   &                              &\multicolumn{6}{c}{\bf Scenarios}                                                   &{\bf Normalized}\\
+		{\bf Volume}     &{\bf Resolution}&                              &\multicolumn{6}{c}{\bf Scenarios}                                                   &{\bf Normalized}\\
 		\cmidrule{4-9}		
-		{\it surplus}    & {\it method}        & {\bf Results}                &{\bf 1}    &{\bf 2}    &{\bf 3}    &{\bf 4}    &{\bf 5}    &{\bf 6}    &{\bf Speedup} \\
+		{\it surplus}    &{\it method}    & {\bf Results}                &{\bf 1}    &{\bf 2}    &{\bf 3}    &{\bf 4}    &{\bf 5}    &{\bf 6}    &{\bf Speedup} \\
 		\toprule	
-
-		%\specialcell{$Volume$\\$surplus$}
-
-		\multirow{6}{*}{$1.2$}&$Relaxed$                 & $f$         &{\it 12.23}&{\it 8.67} &{\it 12.67}&{\it 15.36}&{\it 16.29}&{\it 16.20}&  \specialcell{{\bf 1.00}\\-} \\%  8142
+		\multirow{6}{*}{$1.2$}&$Relaxed$  & $f$                &{\it 12.23}&{\it 8.67} &{\it 12.67}&{\it 15.36}&{\it 16.29}&{\it 16.20}&  \specialcell{{\bf 1.00}} \\%  8142
 		&                      $MIP$      & {\it run time (s)} &     3     &     13    &     20    &     78   &      198  &     1103  &  1.00 \\% 439							
@@
 		&                                 & {\it run time (s)} &     < 1   &     2     &     2     &     8    &     22    &     120   &  {\bf 2.80}\\%
-
-		\midrule
-
-		\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$MIP solver$}& $f$         &{\it 17.01}&{\it 12.45}&{\it 18.02}&{\it 22.03}&{\it 23.09}&{\it 22.35}&  \specialcell{{\bf 1.00}\\-} \\%11495
-
+		\midrule		
+		\multirow{6}{*}{$1.5$}&$Relaxed$  & $f$                &{\it 17.01}&{\it 12.45}&{\it 18.02}&{\it 22.03}&{\it 23.09}&{\it 22.35}&  \specialcell{{\bf 1.00}} \\%11495  
+		&                       $MIP$     & {\it run time (s)} &     6     &     25    &     41    &     158   &      401  &     2161  &  1.00\\%					
         \cmidrule{2-10}		                       
@@
         &                                 & {\it run time (s)} &     < 1   &     2     &     3     &     12    &     33    &   209     &  1.00 \\% 260
-
         \cmidrule{2-10}		                      
@@
         &                                 & {\it run time (s)} &     < 1   &     1     &     2     &     7     &   20      &  120      &  {\bf 1.70}\\% 151
-
-		\midrule
-		\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$MIP solver$}& $f$         &{\it 24.56}&{\it 17.47}&{\it 25.19}&{\it 27.94}&{\it 29.90}&{\it 27.60 }&  \specialcell{{\bf 1.00}\\-} \\%15266
-
+		\midrule
+		\multirow{6}{*}{$2.0$}&$Relaxed$  & $f$                &{\it 24.56}&{\it 17.47}&{\it 25.19}&{\it 27.94}&{\it 29.90}&{\it 27.60 }&  \specialcell{{\bf 1.00}} \\%15266  
+		&                       $MIP$     & {\it run time (s)} &     6     &     25    &     41    &     158   &      401  &     2161  & 1.00\\%							
 		\cmidrule{2-10}		                       
@@
 		&                                 & {\it run time (s)} &     < 1   &     2     &     4     &     14     &     35     &   220     &  1.00 \\% 260
-
 		\cmidrule{2-10}		                       
@@
 		&                                 & {\it run time (s)} &     < 1   &     2     &     3     &     9      &   26       &  151      &  {\bf 1.70}\\% 151
-
 		\bottomrule	
-
 	\end{tabular}
-	%\textsuperscript{asdf sd .}	
 	\normalsize
@@
 
+The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.
+
+We solved all scenarios and instances with a MIP solver in a relaxed linear model.
+This is good for the reader to have an idea of the real performance differences between serial and parallel {\it Shims}.
+
+{\it Shims} had a quality performance very close to the {\it Relaxed MIP solver} upper bound solution, and, as to the {\bf Speedup}, on average, {\it mpShims} was 2 times faster than {\it Shims}.
+Theoretically, it could be more, but due to the burden of sharing, for reading and writing, the CG deviation and the list of items among the processes (pallets), the {\it race condition} management limited {\it mpShims} performance.
+
+We confirmed the hypothesis raised in \ref{hypotheses} and in line \ref{mpShims:minCGdev} of Algorithm \ref{alg:mpShims} that, in sequential mode, where one pallet is built at a time, torque constraints could be more restrictive than volume or weight constraints, which could stop the palletization process earlier, yielding a solution that may be not as good as it could be.
+This integer programming procedure inclusion noticeably improved all {\it Shims} results.
+
+After this inclusion {\it Shims} results surpassed that of {\it mpShims}. In multiprocessing mode ({\it mpShims}) this inclusion (line \ref{mpShims:minCGdev}) made a negligible difference.
+
+\subsection{Results after 3D packing}
+
 \vspace{2.0mm}
-
 \begin{table}[H]
 	\centering
+\color{blue}{	
 	\caption{Volume surpluses, methods and scenarios results {\bf after} 3D packing}  \label{tab:results2}
+}
 	\footnotesize
-
-\color{blue}{
-
-	%\scriptsize
 	\begin{tabular}{crcccccccc}
 		\toprule
-		{\bf Volume}     &  {\bf Resolution}   &                              &\multicolumn{6}{c}{\bf Scenarios}                                                   &{\bf Normalized}\\
+		{\bf Volume}     &{\bf Resolution}&                    &\multicolumn{6}{c}{\bf Scenarios}                                                   &{\bf Normalized}\\
 		\cmidrule{4-9}		
-		{\it surplus}    & {\it method}        & {\bf Results}                &{\bf 1}    &{\bf 2}    &{\bf 3}    &{\bf 4}    &{\bf 5}    &{\bf 6}    &{\bf Speedup} \\
+		{\it surplus}    &{\it method}    & {\bf Results}      &{\bf 1}    &{\bf 2}    &{\bf 3}    &{\bf 4}    &{\bf 5}    &{\bf 6}    &{\bf Speedup} \\
 		\toprule
-
-		%\specialcell{$Volume$\\$surplus$}
-
-		\multirow{6}{*}{$1.2$}&\specialcell{$Relaxed$\\$MIP solver$}& $f$            &{\it 11.47}&{\it 7.24} &{\it 10.78}&{\it 12.06}&{\it 12.97}&{\it 12.44}&  \specialcell{{\bf 1.00}\\-} \\%  8142
-		&                                                & {\it run time (s)} &     6     &     25    &     41    &     158   &      401  &     2161  &  1.00 \\% 439					
+		\multirow{6}{*}{$1.2$}&$Relaxed$  & $f$                &{\it 11.47}&{\it 7.24} &{\it 10.78}&{\it 12.06}&{\it 12.97}&{\it 12.44}&  \specialcell{{\bf 1.00}} \\%
+		&                       $MIP$     & {\it run time (s)} &     6     &     25    &     41    &     158   &      401  &     2161  &  1.00\\%					
 		\cmidrule{2-10}		                       
@@
 		&                                 & {\it run time (s)} &     9     &     36    &     54    &     192   &     427   &    2118   &  {\bf 2.80}\\%		
-
-		\midrule
-
-		\multirow{6}{*}{$1.5$}&\specialcell{$Relaxed$\\$MIP solver$}& $f$     &{\it 15.83}&{\it 10.08}&{\it 15.14}&{\it 16.87}&{\it }&{\it }&  \specialcell{{\bf 1.00}\\-} \\%11495  
-		&                                                & {\it run time (s)} &    10     &     43    &     72    &     262   &     58    &     352   &  1.00 \\% 439					
+		\midrule		
+		\multirow{6}{*}{$1.5$}&$Relaxed$  & $f$                &{\it 15.83}&{\it 10.08}&{\it 15.14}&{\it 16.87}&{\it }&{\it }&  \specialcell{{\bf 1.00}} \\%11495  
+		&                      $MIP$      & {\it run time (s)} &    10     &     43    &     72    &     262   &     58    &     352   &  1.00 \\% 439					
 		\cmidrule{2-10}		                       
@@
 		&                                 & {\it run time (s)} &    25     &    107   &     156   &    498    &    1217   &   5743     &  1.00 \\% 260
-
 		\cmidrule{2-10}		                       
@@
 		&                                 & {\it run time (s)} &    18     &    64     &     101  &    351    &   828     &  120      &  {\bf 1.70}\\% 151
-
-		\midrule
-		\multirow{6}{*}{$2.0$}&\specialcell{$Relaxed$\\$MIP solver$}& $f$            &{\it }&{\it }&{\it }&{\it }&{\it }&{\it  }&  \specialcell{{\bf 1.00}\\-} \\%15266  
-		&                                                & {\it run time (s)} &     < 1   &     3     &     5     &     20    &     58    &     352   &  1.00 \\% 439					
+		\midrule
+		\multirow{6}{*}{$2.0$}&$Relaxed$  & $f$                &{\it 23.73}&{\it }&{\it }&{\it }&{\it }&{\it  }&  \specialcell{{\bf 1.00}} \\%15266  
+		&                       $MIP$     & {\it run time (s)} &    14     &     3     &     5     &     20    &     58    &     352   &  1.00 \\% 439					
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}                       & $f$                &      &      &      &       &       &       &   0.96  \\%  14666
-		&                                                & {\it run time (s)} &     < 1   &     2     &     4     &     14     &     35     &   220     &  1.00 \\% 260
-
+		&\multirow{2}{*}{ $Shims$}        & $f$                &   16.28   &      &      &       &       &       &   0.96  \\%  14666
+		&                                 & {\it run time (s)} &    37     &     2     &     4     &     14     &     35     &   220     &  1.00 \\% 260
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}                     & $f$                &      &      &      &       &       &          &  0.94  \\% 14374
-		&                                                & {\it run time (s)} &     < 1   &     2     &     3     &     9      &   26       &  151      &  {\bf 1.70}\\% 151
-
+		&\multirow{2}{*}{ $mpShims$}      & $f$                &   24.52   &      &      &       &       &          &  0.94  \\% 14374
+		&                                 & {\it run time (s)} &    27     &     2     &     3     &     9      &   26       &  151      &  {\bf 1.70}\\% 151
 		\bottomrule	
-
 	\end{tabular}
-}
-	%\textsuperscript{asdf sd .}	
 	\normalsize
@@
 
-The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.
-
-We solved all scenarios and instances with a MIP solver in a relaxed linear model.
-This is good for the reader to have an idea of the real performance differences between serial and parallel {\it Shims}.
-
-{\it Shims} had a quality performance very close to the {\it Relaxed MIP solver} upper bound solution, and, as to the {\bf Speedup}, on average, {\it mpShims} was 2 times faster than {\it Shims}.
-Theoretically, it could be more, but due to the burden of sharing, for reading and writing, the CG deviation and the list of items among the processes (pallets), the {\it race condition} management limited {\it mpShims} performance.
-
-{\it When threads or processes attempt to simultaneously access a shared resource, and the accesses can result in an error, we often say the program has a race condition, because the threads or processes are in a "race" to carry out an operation.} \cite[p.
-53]{Pacheco:2020}.
-
-We confirmed the hypothesis raised in \ref{hypotheses} and in line \ref{mpShims:minCGdev} of Algorithm \ref{alg:mpShims} that, in sequential mode, where one pallet is built at a time, torque constraints could be more restrictive than volume or weight constraints, which could stop the palletization process earlier, yielding a solution that may be not as good as it could be.
-This integer programming procedure inclusion really improved all {\it Shims} results.
-
-After this inclusion {\it Shims} results surpassed that of {\it mpShims}. In multiprocessing mode ({\it mpShims}) this inclusion (line \ref{mpShims:minCGdev}) made a negligible difference.
-
-\subsection{Results of the 3D packing}
-
-%\begin{table}[!h]
-%	\begin{minipage}{0.5\linewidth}
-%		left
-%	\end{minipage}\hfill % these two lines must be close to each other
-%	\begin{minipage}{0.5\linewidth}
-%		right		
-%	\end{minipage}
-%\end{table}
-
-\vspace{2.0mm}
-\begin{table}[!ht]
-\centering
-\caption{Parallel 3D packing time increment}  \label{tab:3d}
-\scriptsize
-\begin{tabular}{c S[table-format=-1.1] S[table-format=-1.1] S[table-format=-1.1]}
-\toprule
-&   \multicolumn{3}{c}{\bf Average tour time - 3D packing}  \\
-\cmidrule{2-4}
-& {\bf without} & {\bf with} & {\bf increment} \\
-{\bf Scenario} & {\bf (s)}    & {\bf (s)}   & {\bf (times)} \\		
-\midrule
-$1$ &   1 &   15 & 15 \\ % 1	 9   , 3	 6, 1	 25	
-$2$ &   5 &   76 & 15 \\ % 2	 36  , 13	 25
-$3$ &   8 &   97 & 12 \\ % 3	 54  , 20	 41	
-$4$ &  35 &  330 &  9 \\ % 12	 192 , 78	 158
-$5$ &  98 &  740 &  8 \\ % 34	 427 , 198	 401
-$6$ & 591 & 3709 &  6 \\ % 202	 2118, 1103	 2161                                                       
-\bottomrule
-\end{tabular}
-\end{table}
-
-According to Table \ref{tab:3d}, the time increase with the 3D packing inclusion was approximately 10 times.
-Considering the worst (6) scenario, about 62 min to solve a complete tour, against 10 min without the 3D packing procedure.
-i.e., with the 3D packing procedure running in multiprocessing mode, this complete solution method for {\it ACLP+RPDP} is still viable for operational use.
-
-\subsection{Results of the minimization of the pallet handling distances in the cargo bay}
+\subsection{Results on the minimization of the unloading costs}
 \label{improvement}
@@
 \midrule
-$1$ &  32.7 &  28.7 & -7.3 \\ % CBC 32.1 & 31.3 & -2.4 Shims 34.0 & 29.5 & -13.1, 31.5 & 29.9 & -5.1
-$2$ & 193.3 & 188.9 & -6.3 \\ % mpShims 212.0 & 188.9 & -10.9
-$3$ & 165.1 & 153.3 & -7.1 \\ % 
-$4$ & 137.1 & 132.0 & -3.7 \\ % CBC 137.1 & 132.0 & -3.7 mpShim 143.4 & 136.6 & -4.7
-$5$ & 126.7 & 122.0 & -3.8 \\ % CBC 124.9 & 116.3 & -6.9 mpShim 126.7 & 122.0 & -3.8 
-$6$ & 115.3 & 110.4 & -4.2 \\ % CBC 117.4 & 107.6 & -8.3 mpShim 115.3 & 110.4 & -4.2                                                                
+$1$ &  32.7 &  28.7 & -7.3 \\
+$2$ & 193.3 & 188.9 & -6.3 \\
+$3$ & 165.1 & 153.3 & -7.1 \\
+$4$ & 137.1 & 132.0 & -3.7 \\
+$5$ & 126.7 & 122.0 & -3.8 \\
+$6$ & 115.3 & 110.4 & -4.2 \\
 \bottomrule

### 2023-03-28 latex/comp_or.tex
Paper CompOR update

@@
 
-\subsection{Results before 3D packing}
+\subsection{Results {\bf before} 3D packing}
 
@@
 In each $scenario$, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.
+
+Note that, in Table \ref{tab:results} all $Relaxed\ MIP\ f$ value are italicized to indicate that these values may not be optimal or feasible as the decision variables are not integers.
 
@@
 		\toprule	
-		\multirow{6}{*}{$1.2$}&$Relaxed$  & $f$                &{\it 12.23}&{\it 8.67} &{\it 12.67}&{\it 15.36}&{\it 16.29}&{\it 16.20}&  \specialcell{{\bf 1.00}} \\%  8142
-		&                      $MIP$      & {\it run time (s)} &     3     &     13    &     20    &     78   &      198  &     1103  &  1.00 \\% 439							
+		\multirow{7}{*}{$1.2$}&$Relaxed$  & $f$                &{\it 12.23}&{\it 8.67} &{\it 12.67}&{\it 15.36}&{\it 16.29}&{\it 16.20}& - \\%  
+		&                      $MIP$      & {\it run time (s)} &     3     &     13    &     20    &     78    &      198  &     1103  & - \\% 							
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}        & $f$                &     10.34 &     8.50  &     12.42 &     15.06 &     15.97 &     15.88 &  0.96 \\% 7817
+		&\multirow{2}{*}{ $Shims$}        & $f$                &     10.34 &{\bf 8.50} &{\bf 12.42}&{\bf 15.06}&{\bf 15.97}&{\bf 15.88}&{\bf 0.97} \\%   best 8079
 		&                                 & {\it run time (s)} &     < 1   &     3     &     5     &     20    &     58    &     352   &  1.00 \\% 	439		
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}      & $f$                &   12.96   &   7.53    &  12.28    &   13.20   &  13.71    &   13.44   &  0.90\\% 7278
-		&                                 & {\it run time (s)} &     < 1   &     2     &     2     &     8    &     22    &     120   &  {\bf 2.80}\\%
-		\midrule		
-		\multirow{6}{*}{$1.5$}&$Relaxed$  & $f$                &{\it 17.01}&{\it 12.45}&{\it 18.02}&{\it 22.03}&{\it 23.09}&{\it 22.35}&  \specialcell{{\bf 1.00}} \\%11495  
-		&                       $MIP$     & {\it run time (s)} &     6     &     25    &     41    &     158   &      401  &     2161  &  1.00\\%					
+		&\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 12.96}&   7.53    &  12.28    &   13.20   &  13.71    &   13.44   &  0.91\\% 
+		&                                 & {\it run time (s)} &     < 1   &     2     &     2     &     8     &     22    &     120   &  {\bf 2.80}\\% 155
+		\toprule		
+		\multirow{7}{*}{$1.5$}&$Relaxed$  & $f$                &{\it 17.01}&{\it 12.45}&{\it 18.02}&{\it 22.03}&{\it 23.09}&{\it 22.35}& - \\%  
+		&                       $MIP$     & {\it run time (s)} &     4     &     20    &     32    &     115   &     296   &     1685  &  -\\%					
         \cmidrule{2-10}		                       
-        &\multirow{2}{*}{ $Shims$}        & $f$                &     14.68 &     12.21 &     17.67 &  21.01    &  22.64    &   21.91   &  0.96  \\%  11012
-        &                                 & {\it run time (s)} &     < 1   &     2     &     3     &     12    &     33    &   209     &  1.00 \\% 260
+        &\multirow{2}{*}{ $Shims$}        & $f$                &     14.68 &{\bf 12.21}&{\bf 17.67}&{\bf 21.01}&{\bf 22.64}&{\bf 21.91}&{\bf 0.99}  \\%  best 11156
+        &                                 & {\it run time (s)} &     < 1   &     2     &     3     &     12    &     33    &   209     &  1.00 \\% 
         \cmidrule{2-10}		                      
-        &\multirow{2}{*}{ $mpShims$}      & $f$                &    16.12  &     11.81 &     17.48 &  17.66    &   19.81   &  17.61    &  0.87 \\% 9966
-        &                                 & {\it run time (s)} &     < 1   &     1     &     2     &     7     &   20      &  120      &  {\bf 1.70}\\% 151
-		\midrule
-		\multirow{6}{*}{$2.0$}&$Relaxed$  & $f$                &{\it 24.56}&{\it 17.47}&{\it 25.19}&{\it 27.94}&{\it 29.90}&{\it 27.60 }&  \specialcell{{\bf 1.00}} \\%15266  
-		&                       $MIP$     & {\it run time (s)} &     6     &     25    &     41    &     158   &      401  &     2161  & 1.00\\%							
+        &\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 16.12}&     11.81 &     17.48 &  17.66    &   19.81   &  17.61    &  0.90 \\% 
+        &                                 & {\it run time (s)} &     < 1   &     1     &     2     &     7     &   20      &  120      &  {\bf 1.70}\\% 
+		\toprule
+		\multirow{7}{*}{$2.0$}&$Relaxed$  & $f$                &{\it 24.56}&{\it 17.47}&{\it 25.19}&{\it 27.94}&{\it 29.90}&{\it 27.60}& -\\%  
+		&                       $MIP$     & {\it run time (s)} &     4     &     23    &    34     &   119     &     317      &           & -\\%							
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}        & $f$                &    21.31  &  17.35    &   24.24   &   27.39    &  29.31     &  27.06     &   0.96  \\%  14666
-		&                                 & {\it run time (s)} &     < 1   &     2     &     4     &     14     &     35     &   220     &  1.00 \\% 260
+		&\multirow{2}{*}{ $Shims$}        & $f$                &    21.31  &{\bf 17.35}&   24.24   &{\bf 27.39}&{\bf 29.31}&{\bf 27.06}&{\bf 0.98}  \\%  best 15014
+		&                                 & {\it run time (s)} &     < 1   &     2     &     4     &     14    &     35    &   220     &  1.00 \\% 
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}      & $f$                &    24.33  &   17.26   &   24.70   &  24.81     &   27.71    &  24.93        &  0.94  \\% 14374
-		&                                 & {\it run time (s)} &     < 1   &     2     &     3     &     9      &   26       &  151      &  {\bf 1.70}\\% 151
+		&\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 24.33}&   17.26   &{\bf 24.70}&   24.81   &   27.71   &  24.93    &  0.96  \\% 
+		&                                 & {\it run time (s)} &     < 1   &     2     &     3     &     9     &   26      &  151      &  {\bf 1.70}\\% 
 		\bottomrule	
@@
 
-Table \ref{tab:results2} presents the results of the MIP solver compared to {\it Shims} and {\it mpShims}, after the post-optimization with the multiprocessing 3D packing.
-
 The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.
@@
 
-\subsection{Results after 3D packing}
+\subsection{Results {\bf after} 3D packing}
+
+Table \ref{tab:results2} presents the results of the MIP solver compared to {\it Shims} and {\it mpShims}, after the post-optimization with the multiprocessing 3D packing.
+
+Although the relaxed MIP solver results are not always feasible, the multiprocessing 3D packing turns them feasible regarding dimensions and weight.
+So, in Table \ref{tab:results2}, we considered it's results in the performance comparisons.
 
@@
 		\toprule
-		\multirow{6}{*}{$1.2$}&$Relaxed$  & $f$                &{\it 11.47}&{\it 7.24} &{\it 10.78}&{\it 12.06}&{\it 12.97}&{\it 12.44}&  \specialcell{{\bf 1.00}} \\%
-		&                       $MIP$     & {\it run time (s)} &     6     &     25    &     41    &     158   &      401  &     2161  &  1.00\\%					
+		\multirow{7}{*}{$1.2$}&$Relaxed$  & $f$                &   11.47   &    7.24   &{\bf 10.78}&     12.06 &     12.97 &    12.44  &     0.95 \\%
+		&                       $MIP$     & {\it run time (s)} &     6     &     25    &     41    &     158   &      401  &     2161  &{\bf 1.70}\\% 2792					
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}        & $f$                &    9.05   &   5.11    &   9.75    &   10.96   &   12.86   &   13.25   &  0.96 \\% 7817
-		&                                 & {\it run time (s)} &     16    &     68    &     94    &    323    &    719    &    3549   &  1.00 \\% 439			
+		&\multirow{2}{*}{ $Shims$}        & $f$                &    9.05   &   5.11    &   9.75    &   10.96   &   12.86   &   13.25   &     0.86 \\% 
+		&                                 & {\it run time (s)} &     16    &     68    &     94    &    323    &    719    &    3549   &     1.00 \\% 	4769		
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}      & $f$                &    12.14  &   7.29    &   10.36   &   12.63   &   14.04   &   13.92   &  0.90\\% 7278
-		&                                 & {\it run time (s)} &     9     &     36    &     54    &     192   &     427   &    2118   &  {\bf 2.80}\\%		
-		\midrule		
-		\multirow{6}{*}{$1.5$}&$Relaxed$  & $f$                &{\it 15.83}&{\it 10.08}&{\it 15.14}&{\it 16.87}&{\it }&{\it }&  \specialcell{{\bf 1.00}} \\%11495  
-		&                      $MIP$      & {\it run time (s)} &    10     &     43    &     72    &     262   &     58    &     352   &  1.00 \\% 439					
+		&\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 12.14}&{\bf 7.29} &   10.36   &{\bf 12.63}&{\bf 14.04}&{\bf 13.92}&{\bf 0.99}\\% best 7080
+		&                                 & {\it run time (s)} &     9     &     36    &     54    &     192   &     427   &    2118   &{\bf 1.70}\\%	2836	
+		\toprule		
+		\multirow{7}{*}{$1.5$}&$Relaxed$  & $f$                &     15.83 &{\bf 10.08}&{\bf 15.14}&{\bf 16.87}&     18.58 &     17.52 &     0.97 \\%  
+		&                      $MIP$      & {\it run time (s)} &    10     &     43    &     72    &     262   &     696   &     3741  &{\bf 1.60} \\% 4824					
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}        & $f$                &   11.55   &   7.61   &   12.28   &   15.00   &   17.55   &   17.57   &  0.96  \\%  11012
-		&                                 & {\it run time (s)} &    25     &    107   &     156   &    498    &    1217   &   5743     &  1.00 \\% 260
+		&\multirow{2}{*}{ $Shims$}        & $f$                &   11.55   &   7.61    &   12.28   &   15.00   &   17.55   &   17.57   &  0.84  \\%  
+		&                                 & {\it run time (s)} &    25     &    107    &     156   &    498    &    1217   &   5743    &  1.00 \\% 7746
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}      & $f$                &   17.05   &   10.01   &   15.01  &   16.75   &   20.09   &      &  0.87 \\% 9966
-		&                                 & {\it run time (s)} &    18     &    64     &     101  &    351    &   828     &  120      &  {\bf 1.70}\\% 151
-		\midrule
-		\multirow{6}{*}{$2.0$}&$Relaxed$  & $f$                &{\it 23.73}&{\it }&{\it }&{\it }&{\it }&{\it  }&  \specialcell{{\bf 1.00}} \\%15266  
-		&                       $MIP$     & {\it run time (s)} &    14     &     3     &     5     &     20    &     58    &     352   &  1.00 \\% 439					
+		&\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 17.05}&   10.01   &   15.01   &   16.75   &{\bf 20.09}&{\bf 17.85}&{\bf 1.00} \\%   best 9708
+		&                                 & {\it run time (s)} &    18     &    64     &     101   &    351    &   828     &  3879     &     1.50\\% 5421
+		\toprule
+		\multirow{7}{*}{$2.0$}&$Relaxed$  & $f$                &     23.73 &{\bf 15.23}&   21.96   &   24.22   &   26.69   &           &  {\bf 1.00} \\%  
+		&                       $MIP$     & {\it run time (s)} &    14     &     59    &     95    &    348    &   880     &         &  1.00 \\% 					
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}        & $f$                &   16.28   &      &      &       &       &       &   0.96  \\%  14666
-		&                                 & {\it run time (s)} &    37     &     2     &     4     &     14     &     35     &   220     &  1.00 \\% 260
+		&\multirow{2}{*}{ $Shims$}        & $f$                &   16.28   &   10.93   &   18.16   &   22.35   &   25.12   &   24.53    &   0.96  \\%  
+		&                                 & {\it run time (s)} &    37     &    141    &   196     &     653   &   1453    &   6851     &  1.00 \\% 
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}      & $f$                &   24.52   &      &      &       &       &          &  0.94  \\% 14374
-		&                                 & {\it run time (s)} &    27     &     2     &     3     &     9      &   26       &  151      &  {\bf 1.70}\\% 151
+		&\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 24.52}&   15.18   &{\bf 21.99}&{\bf 24.61}&{\bf 26.94}&   24.83    &  0.94  \\% 
+		&                                 & {\it run time (s)} &    27     &    87     &   146     &    496    &   1080    &   5160     &  {\bf 1.70}\\% 
 		\bottomrule	
@@
 
-(3) The procedure for 3-D packing generated an unfit item percentage of approximately 17\%, in average. 
-We are conscious that this unfit rate could be smaller if we had chosen \cite{PaquayLimbourgSchynsOliveira2018} approaches, but as we are concerned with an operational application where the running time is crucial, we stick to an approximate approach.
+(3) The procedure for 3-D packing generated an unfit item rate of, on average, 17\% after {\it Shims} solutions, 13\% after {\it mpShims} solutions, and 9\% after the MIP solver solutions. 
+We are conscious that this unfit rate could be smaller if we had chosen \cite{PaquayLimbourgSchynsOliveira2018} approach, but as we are concerned with an operational application where the running time is crucial, we stick to faster approximate approaches like the {\it First Fit Decreasing} (first orientation that fits) and the {\it Best Fit} (best pivot point if more than one is available).
 
@@
 
-This solution phase was a parallel 3-D packing with each pallet as a process, which decreased this phase's run time by approximate a $n-1$ factor where $n$ is the number of processes.
+This solution phase was a multiprocessing 3-D packing with each pallet as a process, which decreased this phase's run time by approximate a $n-1$ factor where $n$ is the number of processes.
 This was possible because each packing process is completely independent; there is no multiprocessing shared memory.
@@
 
+Another important discovery is that the relaxed version of the MIP solver followed by the multiprocessing 3D packing yielded efficient solutions, even a little bit faster than {\it mpShims} (Table \ref{tab:results2}), but not surpassing {\it mpShims}' quality results.
+
 \section{Conclusions}

### 2023-03-29 latex/comp_or.tex
Paper CompOR update
Finished Comp OR paper

@@
 The CLP alone is known to be NP-hard and extremely hard to solve even for instances of moderate size.
-The typical number of items per transport segment ranges between 300 and 800 items.}
+The typical number of items per transport segment ranges between 300 and 800 items.} This work also adopts a similar range as items quantities, from 300 to 1000 items per node.
 
@@
 		\multirow{7}{*}{$2.0$}&$Relaxed$  & $f$                &{\it 24.56}&{\it 17.47}&{\it 25.19}&{\it 27.94}&{\it 29.90}&{\it 27.60}& -\\%  
-		&                       $MIP$     & {\it run time (s)} &     4     &     23    &    34     &   119     &     317      &           & -\\%							
+		&                       $MIP$     & {\it run time (s)} &     4     &     23    &    34     &   119     &     317      &   1762        & -\\%							
 		\cmidrule{2-10}		                       
@@
 
-Table \ref{tab:results2} presents the results of the MIP solver compared to {\it Shims} and {\it mpShims}, after the post-optimization with the multiprocessing 3D packing.
-
-Although the relaxed MIP solver results are not always feasible, the multiprocessing 3D packing turns them feasible regarding dimensions and weight.
+Table \ref{tab:results2} presents the results of the {\it Relaxed MIP solver} compared to {\it Shims} and {\it mpShims}, after the post-optimization with the multiprocessing 3D packing.
+
+Although the {\it Relaxed MIP solver} results are not always feasible, the multiprocessing 3D packing turns them feasible regarding dimensions and weight.
 So, in Table \ref{tab:results2}, we considered it's results in the performance comparisons.
@@
 	\centering
-\color{blue}{	
 	\caption{Volume surpluses, methods and scenarios results {\bf after} 3D packing}  \label{tab:results2}
-}
 	\footnotesize
@@
 		\multirow{7}{*}{$1.2$}&$Relaxed$  & $f$                &   11.47   &    7.24   &{\bf 10.78}&     12.06 &     12.97 &    12.44  &     0.95 \\%
-		&                       $MIP$     & {\it run time (s)} &     6     &     25    &     41    &     158   &      401  &     2161  &{\bf 1.70}\\% 2792					
+		&                       $MIP$     & {\it run time (s)} &     12     &     51    &     81    &     258   &      601  &     3161  &{\bf 1.15}\\% 	4164				
 		\cmidrule{2-10}		                       
@@
 		\multirow{7}{*}{$1.5$}&$Relaxed$  & $f$                &     15.83 &{\bf 10.08}&{\bf 15.14}&{\bf 16.87}&     18.58 &     17.52 &     0.97 \\%  
-		&                      $MIP$      & {\it run time (s)} &    10     &     43    &     72    &     262   &     696   &     3741  &{\bf 1.60} \\% 4824					
+		&                      $MIP$      & {\it run time (s)} &    20     &     83    &     142   &     362   &     996   &     4741  &{\bf 1.22} \\% 6344					
 		\cmidrule{2-10}		                       
@@
 		&\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 17.05}&   10.01   &   15.01   &   16.75   &{\bf 20.09}&{\bf 17.85}&{\bf 1.00} \\%   best 9708
-		&                                 & {\it run time (s)} &    18     &    64     &     101   &    351    &   828     &  3879     &     1.50\\% 5421
+		&                                 & {\it run time (s)} &    18     &    64     &     101   &    351    &   828     &  3879     &     1.50  \\% 5421
 		\toprule
-		\multirow{7}{*}{$2.0$}&$Relaxed$  & $f$                &     23.73 &{\bf 15.23}&   21.96   &   24.22   &   26.69   &           &  {\bf 1.00} \\%  
-		&                       $MIP$     & {\it run time (s)} &    14     &     59    &     95    &    348    &   880     &         &  1.00 \\% 					
+		\multirow{7}{*}{$2.0$}&$Relaxed$  & $f$                &     23.73 &{\bf 15.23}&   21.96   &   24.22   &   26.69   &    24.29  &     0.99  \\%  
+		&                       $MIP$     & {\it run time (s)} &    28     &     99    &     185   &    648    &   1680    &   8774    &     1.00  \\% 11414					
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}        & $f$                &   16.28   &   10.93   &   18.16   &   22.35   &   25.12   &   24.53    &   0.96  \\%  
-		&                                 & {\it run time (s)} &    37     &    141    &   196     &     653   &   1453    &   6851     &  1.00 \\% 
+		&\multirow{2}{*}{ $Shims$}        & $f$                &   16.28   &   10.93   &   18.16   &   22.35   &   25.12   &   24.53   &     0.85  \\%  
+		&                                 & {\it run time (s)} &    37     &    141    &   196     &     653   &   1453    &   6851    &     1.22  \\% 9331
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 24.52}&   15.18   &{\bf 21.99}&{\bf 24.61}&{\bf 26.94}&   24.83    &  0.94  \\% 
-		&                                 & {\it run time (s)} &    27     &    87     &   146     &    496    &   1080    &   5160     &  {\bf 1.70}\\% 
+		&\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 24.52}&   15.18   &{\bf 21.99}&{\bf 24.61}&{\bf 26.94}&{\bf 24.83}&{\bf 1.00} \\% best 13812
+		&                                 & {\it run time (s)} &    27     &    87     &   146     &    496    &   1080    &   5160    &     1.63  \\% 6996
 		\bottomrule	
@@
 \end{table}
+
+The average value of $f$ obtained by {\it mpShims} was 0.997, followed by the {\it Relaxed MIP} with 0.971.  {\it Shims} obtained 0.851, on average.
+
+As to the performance, the average {\bf Speedup} obtained by {\it mpShims} was 1.61, followed by the {\it Relaxed MIP} with 1.12. {\it Shims} reached 1.07, on average.
+
+So, in terms of quality and speed, {\it mpShims} with the {\it Multiprocessing 3D packing} was the best method to solve the ACLP-RPDP.
 

### 2023-03-29 latex/comp_or.tex
Finished Comp OR paper

@@
 Columns {\bf Before} and {\bf After} refer to the application of the current method.
-In average, the distances from the pallets and the last ramp door position was diminished by 5\% along a tour.
+On average, the distances from the pallets and the last ramp door position was diminished by 5\% along a tour.
 
@@
 
-Another important discovery is that the relaxed version of the MIP solver followed by the multiprocessing 3D packing yielded efficient solutions, even a little bit faster than {\it mpShims} (Table \ref{tab:results2}), but not surpassing {\it mpShims}' quality results.
-
 \section{Conclusions}

### 2023-03-29 latex/comp_or.tex
Comp OR paper update

@@
 
-		\State $ts \gets 1.2$ \Comment{torque surplus} \label{mpShims:ts}
+		\State $ts \gets 1.7$ \Comment{torque surplus} \label{mpShims:ts}
 
@@
 So, we decided to extend torque limit and balance cargo before the next {\it Shims} phase to allow more room for pallet fulfillment.
+The value $1.7$ was determined by the {\it iRace} tool developed by \cite{LopezIbanezManuel2016}.
 
@@
 $minCGdev()$ minimizes the CG deviation, bringing it back to its operational range.
-A proper value for $ts$ must be empirically found to guarantee this intent (the {\it iRace} tool (\cite{LopezIbanezManuel2016}) is good for this value determination).
 
@@
 
-For each $method$, $scenario$ and $surplus$, 7 different instances were generated. Therefore, 126 tests (6 scenarios $\times$ 3 volumes $\times$ 7 instances) were performed for each method.
-
 The average tour values were presented for $f$ and, for the run time, the worst result obtained.
@@
 The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
-	\item {\bf Speedup}: ratio of the sums of the worst run times of all scenarios and the sum of the method run times in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
+	\item {\bf Speedup}: ratio of the sums of the worst run times of all scenarios and the sum of the method run times in all scenarios. The method with the highest {\bf Speedup}\/ is the fastest.
 \end{itemize}
 
-In each $scenario$, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.
+In each $scenario$, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speedup}\/ values.
 
@@
 
-(3) The procedure for 3-D packing generated an unfit item rate of, on average, 17\% after {\it Shims} solutions, 13\% after {\it mpShims} solutions, and 9\% after the MIP solver solutions. 
+(3) The procedure for 3D packing generated an unfit item rate of, on average, 17\% after {\it Shims} solutions, 13\% after {\it mpShims} solutions, and 9\% after the MIP solver solutions. 
 We are conscious that this unfit rate could be smaller if we had chosen \cite{PaquayLimbourgSchynsOliveira2018} approach, but as we are concerned with an operational application where the running time is crucial, we stick to faster approximate approaches like the {\it First Fit Decreasing} (first orientation that fits) and the {\it Best Fit} (best pivot point if more than one is available).
@@
 
-This solution phase was a multiprocessing 3-D packing with each pallet as a process, which decreased this phase's run time by approximate a $n-1$ factor where $n$ is the number of processes.
+This solution phase was a multiprocessing 3D packing with each pallet as a process, which decreased this phase's run time by approximate a $n-1$ factor where $n$ is the number of processes.
 This was possible because each packing process is completely independent; there is no multiprocessing shared memory.
@@
 
+It is important to  emphasize that, as the {\it Multiprocessing 3D packing} is a post-optimization procedure, it benefited all three resolution methods.
+
 (4) The distances to move pallets on unloading in the next node could be minimized to benefit ground handling costs.
@@
 
+Another important fact is the inversion of quality results between {\it Shims} and {\it mpShims}.
+In Table \ref{tab:results}, {\it Shims} was better, but, in Table \ref{tab:results2} (with the inclusion of the {\it Multiprocessing 3D packing}), {\it mpShims} surpassed {\it Shims}' results.
+The rationale behind this is that .....
+
 \section{Conclusions}
@@
 
-Other important contributions are that we verified four hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure added to {\it Shims} algorithm enhanced it's results; a procedure for 3-D packing was turned possible for operational use due to a process-based parallel computing heuristic, and the integer programming procedure to minimize pallets' travel in the cargo bay was successful.
+Other important contributions are that we verified four hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure added to {\it Shims} algorithm enhanced it's results; a procedure for 3D packing was turned possible for operational use due to a process-based parallel computing heuristic, and the integer programming procedure to minimize pallets' travel in the cargo bay was successful.
 

### 2023-03-29 latex/comp_or.tex
Comp OR paper update

@@
 
-In this work, we explore computer parallelism in solving the problem involving scheduling the loading and routing of an airplane in a tour of simultaneous pickup and delivery at intermediate hubs while taking into account a utility score, weight and balance rules, and fuel usage.
+In this work, we explore process-based parallel computing in solving the problem involving scheduling the loading and routing of an airplane in a tour of simultaneous pickup and delivery at intermediate hubs while taking into account a utility score, weight and balance rules, and fuel usage.
 
@@
 (2) The {\it Shims}' results may have been inferior to that; it could be that a sequential approach to pallet building may lead to a premature stop due to the torque constraints.
-(3) A procedure for 3D packing running in parallel mode would keep the run time adequate for operational use.
+(3) A procedure for 3D packing running in process-based parallel computing mode would keep the run time adequate for operational use.
 (4) The distances to move pallets on unloading in the next node could be minimized to benefit ground handling costs.
@@
 These improvements in their heuristics are meant to reduce the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets, and picking up and delivering at each node.
-To the best of our knowledge, that is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP, TSP, and 3D packing has been addressed, and the first time ACLP+RPDP is solved with computer parallelism.
+To the best of our knowledge, that is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP, TSP, and 3D packing has been addressed, and the first time ACLP+RPDP is solved with process-based parallelism.
 
@@
 
-Below, we present the implementation for $SolveNode$, the new parallel heuristic that we propose called {\it Multiprocessing Shims (mpShims)}.
+Below, we present the implementation for $SolveNode$, the new process-based parallel heuristic that we propose called {\it Multiprocessing Shims (mpShims)}.
 
@@
 We solved all scenarios and instances with a MIP solver in a relaxed linear model.
-This is good for the reader to have an idea of the real performance differences between serial and parallel {\it Shims}.
+This is good for the reader to have an idea of the real performance differences between serial and process-based parallel {\it Shims}.
 
@@
 In Table \ref{tab:results}, {\it Shims} was better, but, in Table \ref{tab:results2} (with the inclusion of the {\it Multiprocessing 3D packing}), {\it mpShims} surpassed {\it Shims}' results.
-The rationale behind this is that .....
+
+The rationale behind this is that, in sequential mode ({\it Shims}), items with smaller volumes (higher edge attractiveness $\eta_{ij}^{\pi_k}$, Equation \ref{eq:edge}) tend to be assigned first to central pallets, and items with bigger volumes tend to be assigned to the ramp door or to the forward-positioned pallets; these bigger items are more difficult for the 3D packer actions.
+On {\it mpShims}, as all pallets are built in parallel, the size distribution is more uniform, with a better size diversity to be allocated to each pallet, which is less difficult for the 3D packer actions due to the variety of item dimensions for packing.
+This explains why {\it mpShims} exceeded {\it Shims}' results after the {\it Multiprocessing 3D packing}.
 
@@
 
-Other important contributions are that we verified four hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure added to {\it Shims} algorithm enhanced it's results; a procedure for 3D packing was turned possible for operational use due to a process-based parallel computing heuristic, and the integer programming procedure to minimize pallets' travel in the cargo bay was successful.
+Other important contributions are that we verified four hypotheses: Computer parallelism improved {\it Shims} performance; the intermediate integer programming procedure added to {\it Shims} algorithm enhanced it's results; a 3D packing method was turned possible for operational use due to a process-based parallel computing heuristic, and the integer programming procedure to minimize pallets' travel in the cargo bay was successful.
 

### 2023-03-29 latex/comp_or.tex
Comp OR paper update

@@
 A pallet which has not yet reached its destination may receive more items, although it is known that these operations of removing restraining nets increase handling time and the risk of improper delivery.
-We do not consider over-sized cargo in this work, but only cargoes that fit on these pallets.
+
+	\item We do not consider over-sized cargo in this work, but only cargoes that fit on these pallets.
 
@@
 		\midrule
+\multicolumn{2}{l}{--- from \cite{MesquitaSanches2023} ---}\\		
 s.t.    & Along a tour, the set of unvisited nodes is updated. \\
@@
 		& The total weight must be less than the aircraft payload or the total pallet capacity, whichever is the lowest.\\
-		& {\bf The pallets destined for the next node of the tour should be put as near as possible to the cargo ramp door (our inclusion)}.\\
-		& {\bf The items allocated to each pallet should be submitted to a 3D packing procedure (our inclusion)}.\\				
+\multicolumn{2}{l}{--- our inclusion ---}\\
+s.t.    & Pallets destined for the next node should be put as near as possible to the ramp door.\\
+		& Items allocated to the pallet must fit on the pallet before consolidation.\\				
 		\midrule
@@
 		\State $proc_i \gets Greedy(p_i,\ N^{\pi_k},\ \pi(k),\ nodeTorque,\ G,\ limit,\ lock,\ ts )$   \label{mpShims:greedy}
-		\State $proc_i.start$
+		\State $proc_i.fork$
 		\EndFor
@@
 		\State $proc_i \gets getBestShims(p_i,\ N^{\pi_k},\ \pi(k),\ nodeTorque,\ G,\ surplus,\ lock,\ ts )$ \label{mpShims:getBestShims}
-		\State $proc_i.start$
+		\State $proc_i.fork$
 		\EndFor
@@
 
-Let $n^i$ be the number of items assigned to pallet $i$ and $m$ the number of pallets.
+The algorithm $3Dpacker$ is implemented according to the mathematical model that follows:
+
+Let $n^i$ be the number of items assigned to pallet $i$ and $m$ the amount of pallets.
 
@@
 
-Let $X_{ij} = 1$ if item $j$ was already assigned to pallet $i$ by {\it mpShims}, $X_{ij}\ =\ 0$ otherwise.
+Let $X_{ij} = 1$ if item $j$ was already assigned to pallet $i$ by the previous resolution, $X_{ij}\ =\ 0$ otherwise.
 
@@
 
-Algorithm \ref{alg:3dpacking} presents the multiprocessing solution method for 3D packing to be applied after {\it mpShims}, to guarantee that the item to be palletized actually fit into the pallet.
+Algorithm \ref{alg:3dpacking} presents the multiprocessing solution method for 3D packing to be applied to an existent solution, to guarantee that the item to be palletized actually fit into the pallet.
 
@@
 		\State $proc_i \gets 3Dpacker(p_i,\ N^{\pi_k}_i,\ \pi(k),\ G )$   \label{3dpacking:3}
-		\State $proc_i.start$
+		\State $proc_i.fork$ \Comment{pallet $p_i$ packing processes is forked}
 		\EndFor
@@
 
-According to line \ref{3dpacking:1}, this method receives an initial solution produced by {\it mpShims}, and according to line \ref{3dpacking:2}, each pallet has its own process.
+According to line \ref{3dpacking:1}, this method receives an initial solution produced by a previous resolution, and according to line \ref{3dpacking:2}, each pallet has its own process.
 
@@
 As in the larger aircraft the number of pallets is bigger than the number of threads available in a common handheld computer, for parallelization, we used the {\it Python Multiprocessing} package, which implements {\it process-based parallelism} or {\it parallel shared memory algorithm}.
-According to \cite{multiprocessing}, {\it Multiprocessing} is a package that supports spawning processes using an API similar to the {\it Python Threading} package.
+More details on {\it process-based parallelism} may be found in  \cite{multiprocessing}.
 
@@
 	\centering
-	\caption{Volume surpluses, methods and scenarios results {\bf after} 3D packing}  \label{tab:results2}
+	\caption{Volume surpluses, methods and scenarios results {\bf after} 3D packing (3Dp)}  \label{tab:results2}
 	\footnotesize
@@
 		\multirow{7}{*}{$1.2$}&$Relaxed$  & $f$                &   11.47   &    7.24   &{\bf 10.78}&     12.06 &     12.97 &    12.44  &     0.95 \\%
-		&                       $MIP$     & {\it run time (s)} &     12     &     51    &     81    &     258   &      601  &     3161  &{\bf 1.15}\\% 	4164				
+		&                       $MIP+3Dp$     & {\it run time (s)} &     12     &     51    &     81    &     258   &      601  &     3161  &{\bf 1.15}\\% 	4164				
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}        & $f$                &    9.05   &   5.11    &   9.75    &   10.96   &   12.86   &   13.25   &     0.86 \\% 
+		&\multirow{2}{*}{ $Shims+3Dp$}        & $f$                &    9.05   &   5.11    &   9.75    &   10.96   &   12.86   &   13.25   &     0.86 \\% 
 		&                                 & {\it run time (s)} &     16    &     68    &     94    &    323    &    719    &    3549   &     1.00 \\% 	4769		
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 12.14}&{\bf 7.29} &   10.36   &{\bf 12.63}&{\bf 14.04}&{\bf 13.92}&{\bf 0.99}\\% best 7080
+		&\multirow{2}{*}{ $mpShims+3Dp$}      & $f$                &{\bf 12.14}&{\bf 7.29} &   10.36   &{\bf 12.63}&{\bf 14.04}&{\bf 13.92}&{\bf 0.99}\\% best 7080
 		&                                 & {\it run time (s)} &     9     &     36    &     54    &     192   &     427   &    2118   &{\bf 1.70}\\%	2836	
@@
 		\multirow{7}{*}{$1.5$}&$Relaxed$  & $f$                &     15.83 &{\bf 10.08}&{\bf 15.14}&{\bf 16.87}&     18.58 &     17.52 &     0.97 \\%  
-		&                      $MIP$      & {\it run time (s)} &    20     &     83    &     142   &     362   &     996   &     4741  &{\bf 1.22} \\% 6344					
+		&                      $MIP+3Dp$      & {\it run time (s)} &    20     &     83    &     142   &     362   &     996   &     4741  &{\bf 1.22} \\% 6344					
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}        & $f$                &   11.55   &   7.61    &   12.28   &   15.00   &   17.55   &   17.57   &     0.84  \\%  
+		&\multirow{2}{*}{ $Shims+3Dp$}        & $f$                &   11.55   &   7.61    &   12.28   &   15.00   &   17.55   &   17.57   &     0.84  \\%  
 		&                                 & {\it run time (s)} &    25     &    107    &     156   &    498    &    1217   &   5743    &     1.00  \\% 7746
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 17.05}&   10.01   &   15.01   &   16.75   &{\bf 20.09}&{\bf 17.85}&{\bf 1.00} \\%   best 9708
+		&\multirow{2}{*}{ $mpShims+3Dp$}      & $f$                &{\bf 17.05}&   10.01   &   15.01   &   16.75   &{\bf 20.09}&{\bf 17.85}&{\bf 1.00} \\%   best 9708
 		&                                 & {\it run time (s)} &    18     &    64     &     101   &    351    &   828     &  3879     &     1.50  \\% 5421
@@
 		\multirow{7}{*}{$2.0$}&$Relaxed$  & $f$                &     23.73 &{\bf 15.23}&   21.96   &   24.22   &   26.69   &    24.29  &     0.99  \\%  
-		&                       $MIP$     & {\it run time (s)} &    28     &     99    &     185   &    648    &   1680    &   8774    &     1.00  \\% 11414					
+		&                       $MIP+3Dp$     & {\it run time (s)} &    28     &     99    &     185   &    648    &   1680    &   8774    &     1.00  \\% 11414					
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $Shims$}        & $f$                &   16.28   &   10.93   &   18.16   &   22.35   &   25.12   &   24.53   &     0.85  \\%  
+		&\multirow{2}{*}{ $Shims+3Dp$}        & $f$                &   16.28   &   10.93   &   18.16   &   22.35   &   25.12   &   24.53   &     0.85  \\%  
 		&                                 & {\it run time (s)} &    37     &    141    &   196     &     653   &   1453    &   6851    &     1.22  \\% 9331
 		\cmidrule{2-10}		                       
-		&\multirow{2}{*}{ $mpShims$}      & $f$                &{\bf 24.52}&   15.18   &{\bf 21.99}&{\bf 24.61}&{\bf 26.94}&{\bf 24.83}&{\bf 1.00} \\% best 13812
+		&\multirow{2}{*}{ $mpShims+3Dp$}      & $f$                &{\bf 24.52}&   15.18   &{\bf 21.99}&{\bf 24.61}&{\bf 26.94}&{\bf 24.83}&{\bf 1.00} \\% best 13812
 		&                                 & {\it run time (s)} &    27     &    87     &   146     &    496    &   1080    &   5160    &     1.63  \\% 6996
@@
 
-(3) The procedure for 3D packing generated an unfit item rate of, on average, 17\% after {\it Shims} solutions, 13\% after {\it mpShims} solutions, and 9\% after the MIP solver solutions. 
+(3) The procedure for 3D packing generated an unfit item rate of, on average, 13\%.
 We are conscious that this unfit rate could be smaller if we had chosen \cite{PaquayLimbourgSchynsOliveira2018} approach, but as we are concerned with an operational application where the running time is crucial, we stick to faster approximate approaches like the {\it First Fit Decreasing} (first orientation that fits) and the {\it Best Fit} (best pivot point if more than one is available).

### 2023-03-30 latex/comp_or.tex
Comp OR paper update

@@
 \begin{equation} \label{eq:costs}
-	\tilde{c} = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} \Big [ c^{\pi(k), \pi(k+1)}\times(1+c_g\times|\epsilon^{\pi_k}|) \Big ] + c^{\pi(K),0}\times(1+c_g\times|\epsilon_K|)
+	\tilde{c} = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} \Big [ c^{\pi_k, \pi(k+1)}\times(1+c_g\times|\epsilon^{\pi_k}|) \Big ] + c^{\pi_k,0}\times(1+c_g\times|\epsilon_K|)
 \end{equation}
@@
 
-		\State $\pi[0] \gets 0$ \label{tour:pi1}
-		\State $\pi(K+1) \gets 0$ \label{tour:pi2}
+		\State $\pi_0 \gets 0$ \label{tour:pi1}
+		\State $\pi_{K+1} \gets 0$ \label{tour:pi2}
 		\State $score \gets 0$ \label{tour:score}
@@
 			\For{$i \gets 1$ to $m$}
-				\State $p_i.to[k] \gets -1$ \label{tour:-1} \Comment{reset this pallet destination}
+				\State $p_i.to^{\pi_k} \gets -1$ \label{tour:-1} \Comment{reset this pallet destination}
 			\EndFor	
 			\If {$k = 0$}
-				\State $L_0 \gets L$
+				\State $L^{\pi_0} \gets L$
 				\State Let $G_1(M \cup N_0, \varnothing)$ \label{tour:g11}
 			\Else
-				\State $L_k \gets L_k - \pi(k)$  \label{tour:lk1}			
-				\State $Q^{\pi_k}, E_Q^{\pi_k}, M^{\pi_k} \gets UpdateConsolidated(\pi(k))$ \label{tour:dest}			
+				\State $L^{\pi_k} \gets L^{\pi_k} - \{\pi_k\}$  \label{tour:lk1}			
+				\State $Q^{\pi_k}, E_Q^{\pi_k}, M^{\pi_k} \gets UpdateConsolidated(\pi_k)$ \label{tour:dest}			
 				\State Let $G_1(M^{\pi_k} \cup N^{\pi_k} \cup Q^{\pi_k}, E_Q^{\pi_k})$ \label{tour:g12}
 			\EndIf  \label{tour:lk2}	
-			\State $M^{\pi_k} \gets SetPalletsDestinations( \pi(k) )$ \label{tour:dest2}		
-			\State $G_2 \gets SolveNode(method, \pi(k), G_1, timeLim)$ \label{tour:node}
+			\State $M^{\pi_k} \gets SetPalletsDestinations( \pi_k )$ \label{tour:dest2}		
+			\State $G_2 \gets SolveNode(method, \pi_k, G_1, timeLim)$ \label{tour:node}
 
@@
 
-			\State $s, \epsilon \gets ScoreAndDeviation(\pi(k), G_4)$ \label{tour:analyse}
+			\State $s, \epsilon \gets ScoreAndDeviation(\pi_k, G_4)$ \label{tour:analyse}
 			\State $score \gets score + s$ \label{tour:score2}
-			\State $cost \gets cost + c_{\pi(k),\pi(k+1)} * (1 + c_g * |\epsilon|)$ \label{tour:cost2} 
+			\State $cost \gets cost + c_{\pi_k, \pi_{k+1}} * (1 + c_g * |\epsilon|)$ \label{tour:cost2} 
 		\EndFor  \label{tour:loop2}
@@
 
-$ScoreAndDeviation(\pi(k), G_4)$ evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.
+$ScoreAndDeviation(\pi_k, G_4)$ evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.
 It consists of a loop that goes through all the pallets, accumulating the scores and the torques of the shipped items, allowing the final calculation of the CG deviation.
@@
 
-		\Procedure{$mpShims$}{${N^{\pi_k}, \pi(k)}, limit, G$}
+		\Procedure{$mpShims$}{${N^{\pi_k}, \pi_k}, limit, G$}
 
@@
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets Greedy(p_i,\ N^{\pi_k},\ \pi(k),\ nodeTorque,\ G,\ limit,\ lock,\ ts )$   \label{mpShims:greedy}
+		\State $proc_i \gets Greedy(p_i,\ N^{\pi_k},\ \pi_k,\ nodeTorque,\ G,\ limit,\ lock,\ ts )$   \label{mpShims:greedy}
 		\State $proc_i.fork$
@@
 
-		\State $minCGdev(M^{\pi_k},\ \pi(k),\ nodeTorque)$  \label{mpShims:minCGdev} \Comment{minimize the CG deviation.}
+		\State $minCGdev(M^{\pi_k},\ \pi_k,\ nodeTorque)$  \label{mpShims:minCGdev} \Comment{minimize the CG deviation.}
 
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets getBestShims(p_i,\ N^{\pi_k},\ \pi(k),\ nodeTorque,\ G,\ surplus,\ lock,\ ts )$ \label{mpShims:getBestShims}
+		\State $proc_i \gets getBestShims(p_i,\ N^{\pi_k},\ \pi_k,\ nodeTorque,\ G,\ surplus,\ lock,\ ts )$ \label{mpShims:getBestShims}
 		\State $proc_i.fork$
@@
 
-		\Procedure{$getBestShims$}{$i,\ \eta_1,\ \eta_2,\ E,\ k,\ slack$}
+		\Procedure{$getBestShims$}{$i,\ \eta_1,\ \eta_2,\ E,\ \pi_k,\ slack$}
 
@@
 
-		\If { $e_{ij}^{\pi_k} \not\in (E_N^{\pi_k} \cup shims)$ {\bf and} $e_{ij}^{\pi_k}$ is feasible  {\bf and}  $(t_j^k.v + volume) \leq slack^i$}
+		\If { $e_{ij}^{\pi_k} \not\in (E_N^{\pi_k} \cup shims)$ {\bf and} $e_{ij}^{\pi_k}$ is feasible  {\bf and}  $(t_j^{\pi_k}.v + volume) \leq slack^i$}
 
 		\State $shims \gets shims \cup \{e_{ij}^{\pi_k}\}$
-		\State $volume \gets volume + t_j^k.v$
+		\State $volume \gets volume + t_j^{\pi_k}.v$
 		\State $NewShims \gets$ {\bf False} \label{new_shims_false}
@@
 
-		\State $sh_w \gets shims$, where $shims \in Set$ and $\sum_{e_{ab}^k \in shims} t_b^k.w$ is maximum  \label{best_weight}		
-		\State $sh_v \gets shims$, where $shims \in Set$ and $\sum_{e_{ab}^k \in shims} t_b^k.v$ is maximum \label{best_volume}
-		\State $sh_{best} \gets shims$, where $shims \in \{sh_w, sh_v\}$ and $\sum_{e_{ab}^k \in x} t_b^k.s$ is maximum \label{best_score}
+		\State $sh_w \gets shims$, where $shims \in Set$ and $\sum_{e_{ab}^{\pi_k} \in shims} t_b^{\pi_k}.w$ is maximum  \label{best_weight}		
+		\State $sh_v \gets shims$, where $shims \in Set$ and $\sum_{e_{ab}^{\pi_k} \in shims} t_b^{\pi_k}.v$ is maximum \label{best_volume}
+		\State $sh_{best} \gets shims$, where $shims \in \{sh_w, sh_v\}$ and $\sum_{e_{ab}^{\pi_k} \in x} t_b^{\pi_k}.s$ is maximum \label{best_score}
 
@@
 		\For{$i \gets 1$ to $m$}
-		\State $proc_i \gets 3Dpacker(p_i,\ N^{\pi_k}_i,\ \pi(k),\ G )$   \label{3dpacking:3}
+		\State $proc_i \gets 3Dpacker(p_i,\ N^{\pi_k}_i,\ \pi_k,\ G )$   \label{3dpacking:3}
 		\State $proc_i.fork$ \Comment{pallet $p_i$ packing processes is forked}
@@
 
-Equation \ref{eq:distobj} minimizes the sum of distances of the consolidated in node $\pi_k$ destined to the next node $\pi(k+1)$;
+Equation \ref{eq:distobj} minimizes the sum of distances of the consolidated in node $\pi_k$ destined to the next node $\pi_{k+1}$;
 

### 2023-03-30 py3Djanet/example.py

@@ -1,7 +1,7 @@
 #!/usr/bin/env python
 # coding: utf-8
 
-from packer import Packer
+from py3Djanet.packer import Packer
 from methods import Item
 from bin import Bin
 

### 2023-05-28 main.py
New results

@@ -145,8 +145,8 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
         if method == "GRB":
             modStatus, ObjBound = mipGRB.Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict) 
 
-        if method == "CBC":
-            modStatus, ObjBound = mipCBC.Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict) 
+        if method == "CBC": # new formulation
+            modStatus, ObjBound = mipCBC.Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict, len(tour.nodes)) 
 
         if modStatus == 2: # 2: optimal
             numOptDict["numOpt"] += 1
@@ -157,50 +157,50 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
         Y = np.reshape(solDict["solMatrix"], (-1, N)) # N number of items (columns)
        
         # begin ---- parallel solving the 3D packing for each pallet  
-        procs   = [None for _ in pallets]
-        packers = [None for _ in pallets]
-        counter1 = 0
-        for i, row in enumerate(Y):
+        # procs   = [None for _ in pallets]
+        # packers = [None for _ in pallets]
+        # counter1 = 0
+        # for i, row in enumerate(Y):
 
-            packers[i] = Packer()
-            packers[i].add_bin( Bin(f'pallet{i}', pallets[i].w, pallets[i].h, pallets[i].d, pallets[i].W, i) )
+        #     packers[i] = Packer()
+        #     packers[i].add_bin( Bin(f'pallet{i}', pallets[i].w, pallets[i].h, pallets[i].d, pallets[i].W, i) )
             
-            for j, X_ij in enumerate(row):
-                if X_ij:
-                    packers[i].add_item(Item(f'item{j}', items[j].w, items[j].h, items[j].d, 0, j))
-                    counter1 += 1
+        #     for j, X_ij in enumerate(row):
+        #         if X_ij:
+        #             packers[i].add_item(Item(f'item{j}', items[j].w, items[j].h, items[j].d, 0, j))
+        #             counter1 += 1
             
-            procs[i] = mp.Process( target=packers[i].pack() )
-            procs[i].start()
+        #     procs[i] = mp.Process( target=packers[i].pack() )
+        #     procs[i].start()
 
[…]

### 2023-05-28 mipCBC.py

@@ -1,8 +1,8 @@
 
 # include the MILP Solver: https://www.python-mip.com/
-from mip import Model, xsum, maximize, BINARY, CBC, CONTINUOUS
+from mip import Model, xsum, maximize, BINARY, CBC, CONTINUOUS, INTEGER
 
-def Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict):
+def Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict, K):
 
     # itemsDict to control items inclusion feasibility
 
@@ -19,9 +19,15 @@ def Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict):
 
     mod = Model(solver_name=CBC)
     mod.verbose = 0 # hide messages
-    
+    mod.max_seconds = secBreak
+
+    # Z = mod.add_var(name=f"Z", var_type=INTEGER)
+   
     # decision matrix for which items will be put in which pallet in node "k"         
-    X = [ [ mod.add_var(name=f"X[{i}],[{j}]", var_type=CONTINUOUS) for j in set_N ] for i in set_M ]   
+    # X = [ [ mod.add_var(name=f"X[{i}],[{j}]", var_type=CONTINUOUS) for j in set_N ] for i in set_M ]   
+    X = [ [ mod.add_var(name=f"X[{i}],[{j}]", var_type=BINARY) for j in set_N ] for i in set_M ] 
+
+    Z = [ [ mod.add_var(name=f"Z[{a}],[{b}]", var_type=INTEGER) for b in set_N ] for a in set_N ] 
 
     mod.objective = maximize( xsum( X[i][j] * items[j].S for i in set_M for j in set_N ) )
 
@@ -48,11 +54,37 @@ def Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict):
                 X[i][j] <= X[i][j] * ( items[j].To - pallets[i].Dest[k] + 1 )
             )
 
+
+# As long as variables X and Y you want to multiply are binaries, you might introduce a new binary variable Z and
+# add constraints: Z <= X, Z <= Y and Z >= X + Y - 1. That way, Z = X * Y. You can then use Z in your equations.
+
+        # for a in set_N:
+        #     for b in set_N:
+
+        #         if a != b:
+
+        #             mod.add_constr(
+        #                 Z[a][b] <= X[i][a]
+        #             )
[…]

### 2023-06-03 CAIE/modelo.tex
CAIE paper transferred to parallel folder

% !TeX spellcheck = en_US
%\documentclass[preprint,authoryear]{elsarticle}
\documentclass[preprint]{elsarticle}
% sudo apt-get install -y texlive-publishers

\usepackage[left]{lineno}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{multirow}

\usepackage{lineno,hyperref}
\modulolinenumbers[5]

\usepackage[none]{hyphenat}

\usepackage{url}
\usepackage{booktabs}

\usepackage{tikz}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsthm}

\usepackage{caption}
\captionsetup{labelfont=bf}
\captionsetup{skip=0pt}

\usepackage{pgfplots}
\pgfplotsset{compat=newest}
\usepackage{algorithm}

\usepackage[noend]{algpseudocode}

\usepackage[margin=0.9in]{geometry}

\usepackage[normalem]{ulem}
%\usepackage{xcolor}

\newcommand{\Comentar}[1]{\State {\cmmt{#1}}}
\newcommand{\Break}{\State \textbf{break}}
\renewcommand{\Return}{\State \textbf{return}~}
\renewcommand{\algorithmicensure}{\textbf{Parameters:}}

\renewcommand\algorithmicthen{}
\renewcommand\algorithmicdo{}

\algblock{ForEach}{EndFor}
\algblock{ForDownTo}{EndFor}
\algblock{ForTo}{EndFor}

\newcommand{\Block}[1]{\State #1 \{}
\newcommand{\EndBlock}{\State \}}

\usepackage{scrextend}
\newcommand{\boldm}[1] {\mathversion{bold}#1\mathversion{normal}}
\newcommand{\round}[1]{\ensuremath{\lfloor#1\rceil}}
\usepackage{setspace}
\usepackage{array}
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}
\hyphenation{cost-ef-fec-tive-ness}

\newcommand{\specialcell}[2][c]{%
	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

%\journal{Computers \& Operations Research}

\begin{document}

\begin{frontmatter}

\title{Air cargo load and route planning in pickup and delivery operations}

\author{A.C.P.~Mesquita}
\ead{celio@ita.br}

\author{C.A.A.~Sanches\corref{cor1}}
\ead{alonso@ita.br}
\cortext[cor1]{Corresponding author.}

\address {Instituto Tecnol\'{o}gico de Aeron\'{a}utica - DCTA/ITA/IEC\\
Pra\c{c}a Mal. Eduardo Gomes, 50\\
S\~{a}o Jos\'{e} dos Campos - SP - 12.228-900 - Brazil}

\end{frontmatter}

\section{The mathematical modeling}
\label{sec4}

Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modeling of ACLP+RPDP, which is one of the contributions of our work.

The flight plan will be carried out by a {\bf single} aircraft, which has predefined values of $W_{max}$, $limit^{CG}_{long}$, $limit^{CG}_{lat }$, $d^{CG}_{pallet}$, $c_d$ and $c_g$. This aircraft has a set $M = \{p_1, p_2, \ldots, p_m \}$\/ of $m$\/ pallets, where each pallet $p_i$, $1 \leq i \leq m$, has weight capacity $wp_i$, volume capacity $vp_i$, and distance $dp_i$\/ to the CG of aircraft. Table ?? shows the data of the aircraft considered in this work.

Let $L = \{ l_0, l_1, \ldots, l_K \}$ be the set of $K+1$\/ nodes (or destinations), where $l_0$\/ is the origin and the end of a tour. Let $d(l_i,l_j)$\/ be the distance from $l_i$\/ to $l_j$, where $0 \leq i,j \leq K$. By definition, $d(l_i,l_i)=0$. Let $C=\{c_{ij}\}$ be the cost matrix of the aircraft, where $c_{ij} = c_d*d(l_i,l_j), 0 \leq i,j \leq K$.

Let $S_K = \{s: \{1, \dots, K\} \rightarrow \{1, \dots, K\} \}$\/ be the set of the $K!$\/ permutations of the nodes, which correspond to all possible tours (or itineraries) that have $l_0$\/ as origin and end, passing through the others $K$\/ nodes.

The objective of ACLP+RPDP is to find the permutation $\pi \in S_K$, with the corresponding allocation of items on the pallets at each node, that maximizes the function $\tilde{s}_\pi/\tilde{c}_\pi$ \ref{eq1}, where $\tilde{s}_\pi$\/ is the total score of transported items, and  $\tilde{c}_\pi$\/ is the total cost of fuel consumed. In this way, the tour plan will be $l_0, l_{\pi(1)}, \ldots, l_{\pi(K)}, l_0$. Later, we will describe the calculations of $\tilde{s}_\pi$\/ and $\tilde{c}_\pi$.

\begin{equation} \label{eq1}
	\max_{\pi \in S_K} \tilde{s}_\pi/\tilde{c}_\pi
\end{equation}

Let us now describe the input and output data at each node $l_k$\/ of the flight plan, where $0 \leq k \leq K$. Let $L_k$ \ref{eq2} be the set of remaining nodes when the aircraft is in $l_k$, $0 \leq k \leq K$.

\begin{equation} \label{eq2}
	L_0 = L; L_{k} = L_{k-1} - \{l_{\pi(k)}\}; \ k \in \{1, 2, \ldots, K\}
\end{equation}

Let $N_k = \{t^k_1, t^k_2, \ldots, t^k_{n_k} \}$ be the input set of $n_k$\/ items to be loaded in node $l_k$, $0 \leq k \leq K$. Each item $t^k_j$, $1 \leq j \leq n_k$, has score $st^k_j$, weight $wt^k_j$, volume $vt^k_j$, and destination $lt^k_j \in L_k$. Let $N = \bigcup_{0 \leq k \leq K} N_k$\/ be the input set of items of all nodes along a tour.

Let $X_{ij}^k$\/ be output binary variables, where $0 \leq k \leq K$, $1 \leq i \leq m$\/ and $1 \leq j \leq n_k$. $X_{ij}^k = 1$\/ if $t_j^k$\/ is assigned to $p_i$\/ in node $l_k$, and 0 otherwise \ref{eq3}.

\begin{equation} \label{eq3}
	X_{ij}^k = 0;  \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}; \ k \in \{1,2, \ldots, K\}; \ lt_j^k \notin L_k
\end{equation}

In this way, we can consider $\tilde{s}_\pi$ as the sum of the scores of the items loaded on the aircraft throughout the flight plan \ref{eq4}.

\begin{equation} \label{eq4}
	\tilde{s}_\pi = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n_k} X_{ij}^k \times st_j^k
\end{equation}

Let $Q_k = \{a^k_1, a^k_2, \ldots, a^k_{m_k} \}$\/ be the input set of $m_k \leq m$\/ consolidated items at node $l_k$, $0 \leq k \leq K$. $a^k_i$, $1 \leq i \leq m_k$, is a package of items allocated in some of the previous nodes, with total weight $wa^k_i$, total volume $va^k_i$, and destination $la^k_i \in L_k$. By definition, $m_0=0$ \ref{eq5}. Consolidated items that were destined to node $l_k$\/ are unloaded, that is, they are not considered in $Q_k$.

\begin{equation} \label{eq5}
	m_0 = 0
\end{equation}

Let $Y_{iq}^k$\/ be output binary variables, where $0 \leq k \leq K$, $1 \leq i \leq m$\/ and $1 \leq q \leq m_k$. $Y_{iq}^k = 1$\/ if $a_q^k$\/ is assigned to $p_i$\/ in node $l_k$, and 0 otherwise \ref{eq6}.

\begin{equation} \label{eq6}
	Y_{iq}^k = 0; \ i \in \{1, 2, \ldots, m\}; \ q \in \{1, 2, \ldots, m_k\}; \ k \in \{ 1, 2, \ldots, K\}; \  la_i^k \notin L_k
\end{equation}

Therefore, allocations of items and consolidated items to pallets in node $l_k$ can be seen as a bipartite graph $G_k(V_k, E_k)$, where $V_k = M \cup N_k \cup Q_k$, $E_k = E^N_k \cup E^Q_k$, $(p_i, t_j^k) \in E^N_k$ if $X_{ij}^k = 1$, and $(p_i, a_q^k) \in E^Q_k$ if $Y_{iq}^k = 1$.

Through the output binary variables $X_{ij}^k$\/ and $Y_{iq}^k$, we can calculate the lateral \ref{eq7}\ref{eq8} and the longitudinal \ref{eq9} torques, and the CG deviation \ref{eq10}, caused by shipped items at node $l_k$.

\begin{equation} \label{eq7}
	LatIt_k = \sum_{i=1}^{m} \sum_{j=1}^{n_k} ( X_{ij}^k \times wt_j^k \times (i\%2) - X_{ij}^k \times wt_j^k \times (i+1)\%2 ); \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq8}
	LatCons_k =  \sum_{i=1}^{m} \sum_{q=1}^{m_k}  ( Y_{iq}^k \times wa_q^k \times (i\%2) - Y_{iq}^k \times wa_q^k \times (i+1)\%2); \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq9}
	\tau_k = \sum_{i=1}^{m}[ dp_i \times (\sum_{j=1}^{n_k} X_{ij}^k \times wt_j^k +  \sum_{q=1}^{m} Y_{iq}^k \times wa_q^k)];\ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq10}
	\epsilon_k = \frac{\tau_k}{W_{max} \times limit^{CG}_{long}};\ k \in \{0, 1, \ldots, K\}
\end{equation}

With the CG deviations, the distances covered and the cost matrix of the aircraft, we calculate the total fuel consumption $\tilde{c}_\pi$ \ref{eq11} along the flight plan.

\begin{equation} \label{eq11}
	\tilde{c}_\pi = c_{0,\pi(1)}\times(1+c_g\times|\epsilon_0|) + \sum_{k=1}^{K-1} [ c_{\pi(k), \pi(k+1)}\times(1+c_g\times|\epsilon_{\pi(k)}|) ] + c_{\pi(K),0}\times(1+c_g\times|\epsilon_{\pi(K)}|)
\end{equation}

Finally, we can consider the constraints on each node $l_k$:

\begin{itemize}
	\item The latitudinal \ref{eq12} and the longitudinal \ref{eq13} torques must be within the limits of the aircraft;
	\item The items allocated to each pallet cannot exceed its weight \ref{eq14} and volume \ref{eq15} limits;
	\item At most, each item is associated with a single pallet \ref{eq16};
	\item Consolidated items that have not yet reached their destination must remain on board \ref{eq17};
	\item Items allocated on the same pallet must have the same destinations. In this case, we need two constraints: \ref{eq18} and \ref{eq19}, where $B$\/ is a big value. If $X_{ij}^k = X_{iq}^k = 1$, both constraints require that $lt_j^k = lt_q^k$; otherwise these constraints have no effect.
	\item If there is a consolidated item on the pallet, it must also have the same destination as the other items. Similarly, we use two constraints: \ref{eq20} and \ref{eq21}.
\end{itemize}

\begin{equation} \label{eq12}
	s.t.: d^{CG}_{pallet} \times | LatIt_k + LatCons_k | \leq  W_{max} \times limit^{CG}_{lat}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq13}
	s.t.: |\tau_k| \leq W_{max} \times limit^{CG}_{long};\ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq14}
	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times wt_j^k + \sum_{q=1}^{m_k} Y_{iq}^k \times wa_q^k  \leq wp_i; \ i \in \{1, 2, \ldots, m_k\}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq15}
	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times vt_j^k + \sum_{q=1}^{m_k} Y_{iq}^k \times va_q^k  \leq\ vp_i; \ i \in \{1, 2, \ldots, m_k\}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq16}
	s.t.: \sum_{i=1}^{m} X_{ij}^k \leq 1; \ j \in \{1, 2, \ldots, n_k\}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq17}
	s.t.:  Y_{iq}^k = 1; \ q \in \{1, 2, \ldots, m_k\}; \ k \in \{0, 1, \ldots, K\};  \ la^k_q \in L_k
\end{equation}

\begin{equation} \label{eq18}
	s.t.: lt_j^k - lt_q^k \geq -B \times (1-X_{ij}^k \times X_{iq}^k); \ j,q \in \{1, 2, \ldots, n_k\}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq19}
	s.t.: lt_j^k - lt_q^k \leq B\times (1-X_{ij}^k \times X_{iq}^k); \ j,q \in \{1, 2, \ldots, n_k\}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq20}
	s.t.: lt_j^k - la_q^k \geq -B \times (1-X_{ij}^k \times Y_{iq}^k); \ j \in \{1, 2, \ldots, n_k\}; \ q \in \{1, 2, \ldots, m_k\}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq21}
	s.t.: lt_j^k - la_q^k \leq B\times (1-X_{ij}^k \times Y_{iq}^k); \ j \in \{1, 2, \ldots, n_k\}; \ q \in \{1, 2, \ldots, m_k\}; \ k \in \{0, 1, \ldots, K\}
\end{equation}

Thus, ACLP+RPDP is defined through the equations \ref{eq1} to \ref{eq11}, subject to constraints \ref{eq12} to \ref{eq21}.

\end{document}

### 2023-06-03 CAIE/old.tex
CAIE paper transferred to parallel folder

--- 2022-10-17 comp_or.tex
@@
-% !TeX spellcheck = en_GB
-\documentclass[preprint,authoryear]{elsarticle}
+% !TeX spellcheck = en_US
+%\documentclass[preprint,authoryear]{elsarticle}
+\documentclass[preprint]{elsarticle}
+% sudo apt-get install -y texlive-publishers
 
 \usepackage[left]{lineno}
-\usepackage[utf8]{inputenc}
+\usepackage[latin1]{inputenc}
 \usepackage[T1]{fontenc}
@@
 \captionsetup{labelfont=bf}
-\captionsetup{skip=0pt} % vertical distance from the table or figure
+\captionsetup{skip=0pt}
 
@@
 \usepackage[normalem]{ulem}
-\usepackage{xcolor}
+%\usepackage{xcolor}
 
@@
 
-%----- begin celio
 \usepackage{scrextend}
@@
 \definecolor{Gray}{gray}{0.9}
+\hyphenation{cost-ef-fec-tive-ness}
 
@@
 
-%----- end celio
-
-\journal{European Journal of Operational Research}
+\journal{Computers \& Operations Research}
 
@@
 
-In the scenario of aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading for rapid take-off and mission accomplishment, especially in times of crisis, public calamity, client contract short lead times, or any external pressure for immediate take-off.
-Also, there is no technological assistance to help the load and trip planners with the huge number of demands for transport in each hub.
-This enables other risks such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
-We contributed by modelling and solving a new problem of planning the loading and routing of an aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
-This new hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modelled using standardised pallets in fixed positions, obeying the centre of gravity constraints, delivering each item to its destination, and minimising fuel consumption costs.
-We also contributed by carrying out multiple experiments with a commercial solver and four well-known meta-heuristics on synthetic data based on real data from the {\it Brazilian Air Force}\/ transportation history.
-These challenging benchmark instances are made publicly available.
-We also created a heuristic that quickly finds good solutions for a wide range of problem sizes, an essential contribution as it was the unique method that managed to solve all testing scenarios.
-
-% Reescrever
+In aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading, for fast take-off, and mission accomplishment, especially in times of crisis, public calamity, short deadlines for contracting customers, or any external pressure for immediate take-off.
+In addition, there is no technological assistance to help the load and trip planners with the huge number of demands for transport in each hub.
+This enables other risks, such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
+We defined and solved the problem of planning the loading and routing of an aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
+This NP-hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modeled using standardized pallets in fixed positions, obeying center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
+We performed multiple experiments with a commercial solver and well-known meta-heuristics on data based on the transport history of the {\it Brazilian Air Force}, and also designed a new method that quickly finds good solutions for a wide range of problem sizes, which solved all test scenarios within an acceptable operational runtime.
 
@@
 \begin{keyword}
-Air Cargo \sep Load Planning \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Vehicle Routing
+Load Planning \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Vehicle Routing
 \end{keyword}
@@
 Air cargo transport involves several sub-problems that are difficult to solve.
-Recently, \cite[p.
+Recently, Brandt and Nickel \cite[p.
 401]{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP) and {\it Weight and Balance Problem} (WBP).
-Several aspects were considered in this modelling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc.
+Several aspects were considered in this modeling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc.
 They also presented a comprehensive bibliographic survey of solving methods that had been developed in different situations.
@@
 However, there are still other important challenges in air cargo transport that go beyond the definition of the ACLPP, especially with regard to the flight itinerary and the loading and unloading at each destination (or node) of this travel plan.
-In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Pickup and Delivery Problem} (PDP), and the search for the lowest cost route, which is the well-known {\it Travelling Salesman Problem} (TSP).
-
-Considering air cargo transport, Table \ref{tab:sa} lists the main works in the literature and the corresponding sub-problems addressed.
+In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Pickup and Delivery Problem} (PDP), and the search for the lowest cost route, which is the well-known {\it Traveling Salesman Problem} (TSP).
+
+Considering air cargo transport, Table \ref{tab:sa} lists the main works in the literature in chronological order, and the corresponding sub-problems addressed.
 We also indicate whether the dimensions of the items were taken into account ({\bf 3D} or {\bf 2D}) and which solution method was used: heuristics ({\bf Heu}), integers ({\bf Int}), or linear programming ({\bf Lin}).
@@
 	\scriptsize
-	\renewcommand{\arraystretch}{1.1} % alturas das linhas
+	\renewcommand{\arraystretch}{1.1} 
 	\begin{tabular}{r|cc|cc|cc|ccc}
 		\toprule
-		                            & {\bf APP}  & {\bf WBP}  &  {\bf PDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} \\
+		 Work & {\bf APP}  & {\bf WBP}  &  {\bf PDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} \\
 		\midrule
-		\cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$ \\
-		\cite{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$ \\
-		\cite{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$ \\
-		\cite{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
-		\cite{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$ \\
-		\cite{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$ \\
-		\cite{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{MesquitaCunha2011}   & $.$ & $.$    & $\bigstar$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\		
-		\cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{Vancroonemburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
-		\cite{PaquaySchynsLimbourg2016,PaquayLimbourgSchynsOliveira2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$ \\
-		\cite{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$ \\
-		\cite{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$  \\
-		\cite{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$  \\
-		\cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
-		{\bf This work}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$  \\
+		Larsen and Mikkelsen \cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$ \\
+		Brosh \cite{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$ \\
+		Ng \cite{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ \\
+		Heidelberg {\it et al.} \cite{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$ \\
+		Mongeau and Bes \cite{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
+		Fok and Chun \cite{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
+		Chan {\it et al.} \cite{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$ \\
+		Kaluzny and Shaw \cite{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$ \\
+		Verstichel {\it et al.} \cite{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
+		Mesquita and Cunha \cite{MesquitaCunha2011}   & $.$ & $.$    & $\bigstar$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\		
+		Limbourg {\it et al.} \cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
+		Roesener and Barnes \cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
+		Vancroonenburg {\it et al.} \cite{Vancroonenburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
+		Lurkin and Schyns \cite{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
+		Roesener and Barnes \cite{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
+		Paquay {\it et al} \cite{PaquaySchynsLimbourg2016,PaquayLimbourgSchynsOliveira2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$ \\
+		Chenguang {\it el al.} \cite{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$ \\
+		Wong and Ling \cite{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$  \\
+		Wong {\it et al.} \cite{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$  \\
+		Zhao {\it et al.} \cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
+		{\bf This article}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$  \\
 		\bottomrule 
@@
 
-As can be seen, so far \cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem.
-Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a pre-defined flight plan restricted to two legs.
-It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modelled through a container sequencing problem.
-Referring directly to this work, \cite[p.
+As can be seen, so far Lurkin and Schyns \cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem.
+Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a pre-defined tour plan restricted to two legs.
+It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem.
+Referring directly to this work, Brandt and Nickel \cite[p.
 409]{BrandtStefan2019} comment: {\it However, not even these sub-problems are acceptably solved for real-world problem sizes or the models omit some practically relevant constraints}.
@@
 Due to its dimensions, this country has the largest air market on the continent with $2,499$\/ registered airports, of which $1,911$\/ are private and $588$\/ are public.
-Although it is an immense distribution network, airlift missions consider 3 to 5 nodes per flight plan.
+Although it is an immense distribution network, airlift missions consider 3 to 5 nodes per tour plan.
 Throughout this work, we address routes with up to 7 nodes, as can be seen in Table \ref{tab:costs} and Figure \ref{fig:nodes}.
@@
 
-	\end{minipage}\hfill % these two lines must be close to each other
+	\end{minipage}\hfill 
 	\begin{minipage}{0.45\linewidth}
@@
 		\centering
-
 		\footnotesize
-
 		\newcolumntype{Y}{>{\centering\arraybackslash}p{0.09\textwidth}}
@@
 
-	\end{minipage}\hfill % these two lines must be close to each other
+	\end{minipage}\hfill 
 	\begin{minipage}{0.50\linewidth}
@@
 		\includegraphics[scale=0.25]{Images/nodes.png}
-		\captionof{figure}{A route between Brazilian airports}
+		\captionof{figure}{A route with some Brazilian airports}
 		\label{fig:nodes}		
@@
 
-In the {\it Brazilian Air Force}\/ missions, hundreds of items can be carried at each node, where the objectives are to prioritise the transport of the most important items and minimise the cost of fuel along the route.
-As standardised pallets are used with predefined positions on the aircraft, it is possible to carry out loading and unloading at each node in around two hours.
+In the {\it Brazilian Air Force}\/ missions, hundreds of items can be carried at each node, where the objectives are to prioritise the transport of the most important items and minimize the cost of fuel along the route.
+As standardized pallets are used with predefined positions on the aircraft, it is possible to carry out loading and unloading at each node in around two hours.
 However, there is no technological assistance that guarantees the achievement of these objectives.
 
-This work proposes a method that attends these objectives: a pallet building and arrangement plan with a routing option that maximises the benefit-cost ratio for the smooth execution of pickup and delivery transport missions.
-We develop a heuristic that can be executed on a simple handheld computer (like a laptop or a tablet) and that provides a solution quickly enough to keep this cargo handling time under an hour.
-This heuristic will reduce the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets, and picking up and delivering at each node.
+This work proposes a method that attends these objectives: a pallet building and arrangement plan with a routing option that maximizes the benefit-cost ratio for the smooth execution of pickup and delivery transport missions.
+We develop a method that can be executed on a simple handheld computer (like a laptop or a tablet) and that provides a solution quickly enough to keep this cargo handling time under an hour.
+This method will reduce the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets, and picking up and delivering at each node.
 To the best of our knowledge, this is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP and TSP has been addressed.
@@
 
-This article is organised into six more sections.
+This article is organized into six more sections.
 In Section \ref{sec2}, we give a brief review of the literature.
-In Section \ref{sec3}, we present the problem context and assumptions and, in Section \ref{sec4}, the mathematical model and how we dealt with its issues.
+In Section \ref{sec3}, we present the problem context and assumptions, and in Section \ref{sec4}, we describe the mathematical model and how we dealt with its issues.
 In Section \ref{sec5}, we describe the elaborate algorithms, whose results are presented in Section \ref{sec6}.
@@
 
-\cite{LarsenMikkelsen1979} developed an interactive procedure for loading 14 types of Boeing 747 into a two-leg flight plan.
+Larsen and Mikkelsen \cite{LarsenMikkelsen1979} developed an interactive procedure for loading 14 types of Boeing 747 into a two-leg transportation plan.
 Seven types of items were considered to be allocated in 17 to 42 positions.
-With non-linear programming and heuristics, they present a solution that minimises positioning changes in the intermediate node, optimising the load balancing in the aircraft.
-
-\cite{Brosh1981} addressed the problem of planning the allocation of cargo on an aircraft.
+With non-linear programming and heuristics, they present a solution that minimizes positioning changes in the intermediate node, optimizing the load balancing in the aircraft.
+
+Brosh \cite{Brosh1981} addressed the problem of planning the allocation of cargo on an aircraft.
 Considering volume, weight and structural constraints, the author finds the optimal load layout through a fractional programming problem.
 
-\cite{Kevin1992} developed a multi-criteria optimization approach to load the {\it C-130} aircraft of the {\it Canadian Air Force}.
+Ng \cite{Kevin1992} developed a multi-criteria optimization approach to load the {\it C-130} aircraft of the {\it Canadian Air Force}.
 Based on integer programming, this model provides timely planning and improves airlift support for combat operations, solving WBP with pallets in fixed positions, and considering 20 different items.
 
-\cite{Heidelberg1998} developed a heuristic for 2D packing in air loading, comparing it with methods for solving the {\it Bin Packing Problem}.
-Authors conclude that the classical algorithms are inadequate in this context, because they ignore the aircraft balancing constraints.
-
-%\cite{Gueret2003} proposed a method for loading military aircraft (specifically, the french military) for airlift operations.
-They used a two-phased solution method consisting of heuristics that quickly compute suboptimal initial solutions for a 2-dimensional bin packing and a local search algorithm to improve upon the initial solution, while ignoring the aircraft center of gravity.
-
-\cite{MongeauBes2003} presented a method based on linear integer programming to solve the problem of choosing and positioning containers on the {\it Airbus 340-300}.
-Safety and stability restrictions were considered, with the objective of minimising fuel consumption.
-
-\cite{fok2004optimizing} developed a web-based application to make efficient use of space and load balancing for an air cargo company.
+Heidelberg {\it et al.} \cite{Heidelberg1998} developed a heuristic for 2D packing in air loading, comparing it with methods for solving the {\it Bin Packing Problem}.
+These authors conclude that the classical algorithms are inadequate in this context, because they ignore the aircraft balancing constraints.
+
+Mongeau and Bes \cite{MongeauBes2003} presented a method based on linear integer programming to solve the problem of choosing and positioning containers in the {\it Airbus 340-300}.
+Safety and stability constraints were considered, with the objective of minimizing fuel consumption.
+
+Fok and Chun \cite{fok2004optimizing} developed a web-based application to make efficient use of space and load balancing for an air cargo company.
 Based on an analysis of historical data, an operational load planning with mathematical optimization is obtained.
-This container load planning is usually done roughly 2 hours before departure, when all cargo details are in place.
-
-\cite{Chan2006} carried out a case study with heterogeneous pallets.
-In order to minimise the total cost of shipping, they developed a 3D packing heuristic, with a loading plan for each pallet.
+This container load planning is usually done in roughly 2 hours before departure, when all cargo details are in place.
+
+Chan {\it et al.} \cite{Chan2006} carried out a case study with heterogeneous pallets.
+In order to minimize the total cost of shipping, they developed a 3D packing heuristic, with a loading plan for each pallet.
 Although the authors do not consider load balancing or positioning of pallets in the cargo hold, this method is relevant in commercial and industrial applications, where cargo items tend to be less dense.
 
-\cite{KaluznyBohdanL2009Oalb} developed a mixed integer linear programming model to arrange a set of items in a military context that optimises the load balance.
-
-\cite{Verstichel2011} solved WBP by selecting the most profitable subset of containers to be loaded onto an aircraft using mixed-integer programming.
+Kaluzny and Shaw \cite{KaluznyBohdanL2009Oalb} developed a mixed integer linear programming model to arrange a set of items in a military context that optimizes the load balance.
+
+Verstichel {\it et al.} \cite{Verstichel2011} solved WBP by selecting the most profitable subset of containers to be loaded into an aircraft using mixed-integer programming.
 Experimental results on real-life data showed significant improvements compared to those obtained manually by an experienced planner.
 
-\cite{MesquitaCunha2011} presented a heuristic for a real problem of the {\it Brazilian Air Force}, which consists of defining transport routes with simultaneous collection and delivery from a central distribution terminal.
-
-\cite{Limbourg2012} developed a mixed-integer program for optimally rearranging a set of pallets into a compartmentalised cargo aircraft, specifically the {\it Boeing 747}.
-
-\cite{RoesenerHall2014} solved APP and WBP as an integer programming problem, which also allows items to be loaded into pallets according to a specific orientation (e.g., this side up).
-
-\cite{Vancroonemburg2014} presented a mixed integer linear programming model that selects the most profitable pallets, satisfying safety and load balancing constraints on the {\it Boeing 747-400}.
-Using a solver, authors solved real problems in less than an hour.
-
-As already mentioned, \cite{LurkinSchyns2015} was the first work that simultaneously modelled WBP and PDP in air cargo transport.
+Mesquita and Cunha \cite{MesquitaCunha2011} presented a heuristic for a real problem of the {\it Brazilian Air Force}, which consists of defining transport routes with simultaneous collection and delivery from a central distribution terminal.
+
+Limbourg {\it et al.} \cite{Limbourg2012} developed a mixed-integer program for optimally rearranging a set of pallets into a compartmentalized cargo aircraft, specifically the {\it Boeing 747}.
+
+Roesener and Hall \cite{RoesenerHall2014} solved APP and WBP as an integer programming problem, which also allows items to be loaded onto pallets according to a specific orientation (e.g., this side up).
+
+Vancroonenbur {\it et al} \cite{Vancroonenburg2014} presented a mixed integer linear programming model that selects the most profitable pallets, satisfying safety and load balancing constraints on the {\it Boeing 747-400}.
+Using a solver, these authors solved real problems in less than an hour.
+
+As already mentioned, Lurkin and Schyns \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and PDP in air cargo transport.
 The authors demonstrated that this problem is NP-hard and performed some experiments with real data, noting that their model offers better results than those obtained manually.
 
-\cite{RoesenerBarnes2016} proposed a heuristic to solve the {\it Dynamic Airlift Loading Problem} (DALP).
-Given a set of palletized cargo items that require transport between two nodes in a time frame, the objective of this problem is to select an efficient subset of aircraft, partition the pallets into aircraft loads and assign them to allowable positions on those aircraft.
-
-\cite{PaquaySchynsLimbourg2016} presented a mathematical modelling to optimise the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
-Its objective is to maximise the volume used in containers, considering load balancing restrictions, the presence of fragile items and the possibility of rotating these boxes.
-\cite{PaquayLimbourgSchynsOliveira2018} developed some heuristics to solve this problem.
-
-\cite{YangLiuGao2018} modelled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing in order to minimise fuel consumption.
-
-\cite{wong2020}A developed a mathematical model and a tool based on mixed integer programming for optimising cargo in aircraft with different pallet configurations.
-Balance restrictions and the presence of dangerous items were considered.
-\cite{eugene2021} integrated this tool to a digital simulation model, with a visualisation and validation system, based on sensors that alert about load deviations.
-
-\cite{zhao2021} proposed a new modelling for WBP based on mixed integer programming.
-Instead of focusing on the centre of gravity (CG) deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
-
-As can be seen, none of these works address air cargo palletization and load balancing with route optimization in a multi-leg flight plan.
-This is the objective of our work: to model and elaborate heuristics for a real problem that simultaneously involves 4 intractable sub-problems: APP, WBP, PDP and TSP.
+Roesener and Barnes \cite{RoesenerBarnes2016} proposed a heuristic to solve the {\it Dynamic Airlift Loading Problem} (DALP).
+Given a set of palletized cargo items that require transport between two nodes in a time frame, the objective of this problem is to select an efficient subset of aircraft, partition the pallets into aircraft loads and assign them to allowable positions in those aircraft.
+
+Paquay {\it et al.} \cite{PaquaySchynsLimbourg2016} presented a mathematical modeling to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
+Its objective is to maximize the volume used in containers, considering load balancing restrictions, the presence of fragile items and the possibility of rotating these boxes.
+Paquay {\it et al.} \cite{PaquayLimbourgSchynsOliveira2018} developed some heuristics to solve this problem.
+
+Chenguang {\it et al.} \cite{YangLiuGao2018} modeled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing to minimize fuel consumption.
+
+Wong and Ling \cite{wong2020}A developed a mathematical model and a tool based on mixed integer programming for optimizing cargo in aircraft with different pallet configurations.
+Balance constraints and the presence of dangerous items were considered.
+Wong {\it et al.} \cite{eugene2021} integrated this tool to a digital simulation model, with a visualization and validation system, based on sensors that alert about load deviations.
+
+Zhao {\it et al.} \cite{zhao2021} proposed a new modeling for WBP based on mixed integer programming.
+Instead of focusing on the center of gravity (CG) deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
+
+As can be seen, none of these works address air cargo palletization and load balancing with route optimization in a multi-leg transportation plan.
+This is the objective of our work: to model and elaborate a solution for a real problem that simultaneously involves 4 intractable sub-problems: APP, WBP, PDP and TSP.
 
@@
 
-As we are dealing with an extremely complex and diverse problem, we decided to establish some simplifying characteristics:
+As we are dealing with an extremely complex and diverse problem, we decided to establish some operational premises:
 
@@
 
-	\item At each node of the flight plan, the items to be allocated are characterised by weight, volume, scores, and previously known destinations, but do not have dimensions.
-We leave the consideration of 2D or 3D items for future work.
-
-	\item We also disregarded {\it hazardous} items, which eventually could be treated as high score items and other specific constraints.
+	\item At each node of the tour plan, the items to be allocated are characterized by weight, volume, scores, and previously known destinations, but the dimensions are unknown.
+We leave the consideration of 2D or 3D items to a future work.
 
@@
 
-	\item Finally, as we are interested in minimising fuel costs, we disregarded some lower ones as {\it handling operation costs}.
+	\item Finally, as we are interested in minimizing fuel costs, we disregarded some lower ones as {\it handling operation costs}. Other costs like this may be included in future work.
 
@@
 	\caption{Smaller aircraft layout}
-	%\small\textsuperscript{Source: the author}	
 	\label{fig:smaller}
@@
 	\caption{Aircraft longitudinal cut showing red lines as pallets}
-	%\small\textsuperscript{Source: the author}
 	\label{fig:lateral}
@@
 In both aircraft, as the ramps have an inclination of 25 degrees, we made the necessary corrections in {\it CGx}, {\it Weight} and {\it Volume limits} of the corresponding pallets.
-The monetary costs of both aircraft are also indicated: per unit of distance in flights between legs ($c_d$) and per deviation in the CG ($c_g$).
+The monetary costs of both aircraft are also indicated: per unit of distance in flights legs ($c_d$) and per deviation in the CG ($c_g$).
 It is important to consider that $c_g$ tends to zero as the aircraft attitude tends to level.
@@
 	\item in the larger aircraft, the CG of the payload must be at a maximum lateral distance of $limit^{CG}_{lat}$ from the CG of the aircraft;
-	\item in the larger aircraft, pallets are distributed in two identical rows (with odd and even indices, respectively), and their centroids are at a distance $d^{CG}_{pallet}$ from the centre-line of the aircraft.
+	\item in the larger aircraft, pallets are distributed in two identical rows (with odd and even indices, respectively), and their centroids are at a distance $d^{CG}_{pallet}$ from the center-line of the aircraft.
 \end{itemize}
 
-\subsection{Problem Summary}
+\subsection{Problem summary}
 
@@
 		& Consolidated items can only be included in the same pallet if their destinations are the same.\\
-		& Only items destined for the remaining nodes can be loaded.  \\
+		& Only items destined to the remaining nodes can be loaded.  \\
 		& The lateral and longitudinal torques must be within the operational range of the aircraft.\\
-		& Weight and volume limitations of pallets must be respected.\\
+		& Weight and volume limitations of the pallets must be respected.\\
 		& The total weight must be less than the aircraft payload or the total pallet capacity, whichever is the lowest.\\	
@@
 
-\section{The mathematical modelling}
+\section{The mathematical modeling}
 \label{sec4}
 
-Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modelling of ACLP+RPDP, which is one of the contributions of our work.
-
-Let $L = \{ l_0, l_1, \ldots, l_K \}$ be the set of $K+1$ nodes (or destinations), where $l_0$ is the origin and end of a flight plan.
+Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modeling of ACLP+RPDP, which is one of the contributions of our work.
+
+Let $L = \{ l_0, l_1, \ldots, l_K \}$ be the set of $K+1$ nodes (or destinations), where $l_0$ is the origin and end of a tour.
 Let $d(l_i,l_j)$ be the distance from $l_i$ to $l_j$, where $0 \leq i,j \leq K$.
@@
 
-The objective of this problem is to find a permutation $\pi \in S_K$\/ that maximises the function $f_\pi(\tilde{s},\tilde{c})$ \ref{eq:maxf}.
-In this way, the flight plan will be $l_0, l_{\pi(1)}, \ldots, l_{\pi(K)}, l_0$.
+The objective of this problem is to find a permutation $\pi \in S_K$\/ that maximizes the function $f_\pi(\tilde{s},\tilde{c})$ \ref{eq:maxf}.
+In this way, the tour plan will be $l_0, l_{\pi(1)}, \ldots, l_{\pi(K)}, l_0$.
 $\tilde{s}$\/ is the total score of transported items \ref{eq:scores} and  $\tilde{c}$\/ is the total cost of fuel consumed \ref{eq:costs}.
@@
 
-For each step of the flight plan, the set of unvisited nodes is updated \ref{eq:pdp11}.
-Although there are no items consolidated at the beginning of the flight plan, we defined these variables for ease of notation \ref{eq:pdp13}.
-Items destined outside the rest of the flight plan will not be loaded (\ref{eq:pdp12} and \ref{eq:pdp9}).
+For each step of the tour, the set of unvisited nodes is updated \ref{eq:pdp11}.
+Although there are no items consolidated at the beginning of the tour, we defined these variables for ease of notation \ref{eq:pdp13}.
+Items destined outside the rest of the tour will not be loaded (\ref{eq:pdp12} and \ref{eq:pdp9}).
 
@@
 
-Once the assumptions of this work and the mathematical modelling of the problem are presented, it is easy to see that ACLP+RPDP is NP-hard.
-In a similar way to \cite[p.
+Once the assumptions of this work and the mathematical modeling of the problem are presented, it is easy to see that ACLP+RPDP is NP-hard.
+In a similar way to Lurkin and Schyns \cite[p.
 6]{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in $l_0$, and no items in $l_1$.
@@
 
-Real cases are more complex as they have hundreds of different items in each node and involve three intractable sub-problems: APP, WBP and PDP.
-Through the mathematical modelling presented in the previous section, we verify that {\it Mixed-Integer Programming}\/ (MIP) is not able to solve these cases in feasible time.
-Thus, it is necessary to adopt some strategy to find a viable solution, not necessarily optimal, that seeks to maximise the objective function $f$.
+Real cases are more complex as they have hundreds of different items in each node and involve four intractable sub-problems: APP, WBP, PDP and TSP.
+Through the mathematical modeling presented in the previous section, we verify that {\it Mixed-Integer Programming}\/ (MIP) is not able to solve these cases in feasible time.
+Thus, it is necessary to adopt some strategy to find a viable solution, not necessarily optimal, that seeks to maximize the objective function $f$.
 
@@
 \begin{algorithm}[H]
-	\caption{$ACLP+RPDP (scenario,volume)$}  \label{alg:main}
+	\caption{Main procedure}  \label{alg:main}
 	\begin{algorithmic}[1]
-		\State Let $L, M, C$ be according to $scenario$ \label{main:LMC}
-		\State $N \gets ItemsGeneration(scenario,volume)$ \label{main:items}
-		\For {each $method$}
-			\For {each $\pi \in S_K$} \label{main:loop1}
-				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method )$ \label{main:method}
+
+		\Procedure{$ACLP+RPDP$}{$scenario,\ volume$}
+
+		\State Let $L, M, C$ be according to $scenario$ \label{main:LMC} \Comment{Nodes, pallets and costs}
+		\State $N \gets ItemsGeneration(scenario,\ volume)$ \label{main:items} \Comment{Items available for shipment}
+		\For {each $method$} \Comment{$method$ is a MIP solver or a heuristic}
+			\For {each $\pi \in S_K$} \label{main:loop1} \Comment{$\pi$ is a permutation of the nodes}
+				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method )$ \Comment{A solution to the permutation $\pi$} \label{main:method}
 			\EndFor \label{main:loop2}
-			\State $answer[scenario,volume,method] \gets \max f$ \label{main:f}
+			\State $answer[scenario,volume,method] \gets \max f$ \label{main:f} \Comment{Best result among all permutations}
 		\EndFor
-		\State $ {\bf return}\ answer$
+		\Return $answer$
+
+		\EndProcedure
+
 	\end{algorithmic}
@@
 
-The other parameter $volume$\/ is a value greater than 1, which corresponds, at each node $l_k$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n_k} t^k_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.v$).
+The parameter $volume$\/ is a value greater than 1, which corresponds, at each node $l_k$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n_k} t^k_j.v$) and the load capacity of the pallets ($\sum_{i=1}^{m} p_i.v$).
 This parameter is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).
 
-$method$\/ corresponds to a MIP solver or one of the heuristics that we will present in subsection \ref{methods}.
+$method$\/ corresponds to a MIP solver or a heuristic that we will present in subsection \ref{methods}.
 The loop of lines \ref{main:loop1}-\ref{main:loop2} goes through all permutations $\pi$, where the node-by-node resolutions are performed by $SolveTour$, whose result is stored in $f_{\pi}$.
@@
 
-Next, we will present two subsections: in the first we explain how $SolveTour$ is executed, while in the second we will present the heuristics developed for node-by-node resolutions.
+Next, we will present two subsections: in the first we explain how $SolveTour$ is executed, while in the second we will present the heuristic developed for node-by-node resolutions.
 
@@
 In the context of this work, we know that $m>K$, once the aircraft has $7$\/ or $18$\/ pallets and $K\leq 6$, allowing there to be at least one pallet for each node to be visited.
-$SetPalletsDestination$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node, without changing the pallets destination with consolidated items.
+$SetPalletsDestinations$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node, without changing the pallets destination with consolidated items.
 
@@
 \begin{algorithm}[H]
-	\caption{$SolveTour(\pi, L, M, C, N, method)$}  \label{alg:tour}
-
+	\caption{Procedure to solve tour $\pi$ with $method$ (MPI solver or heuristic)}  \label{alg:tour}
 	\begin{algorithmic}[1]
 
-		\State $\pi(0) \gets 0$ \label{tour:pi1}
-		\State $\pi(K+1) \gets 0$ \label{tour:pi2}
+		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method$}
+
+		\State $\pi(0) \gets 0$ \label{tour:pi1} \Comment{The first node is the base}
+		\State $\pi(K+1) \gets 0$ \label{tour:pi2} \Comment{The last node is also the base}
 		\State $score \gets 0$ \label{tour:score}
@@
 			\For{$i \gets 1$ to $m$}
-				\State $p_i.to[k] \gets -1$ \label{tour:-1}
+				\State $p_i.to[k] \gets -1$ \label{tour:-1} \Comment{Pallet destination unset}
 			\EndFor	
@@
 			\Else
-				\State $L_k \gets L_k - \pi(k)$  \label{tour:lk1}			
+				\State $L_k \gets L_k - \pi(k)$  \label{tour:lk1}	\Comment{$\pi(k)$ is the node $k$ from tour $\pi$}		
 				\State $Q_{\pi(k)}, E^Q_{\pi(k)}, M \gets UpdateConsolidated(\pi(k))$ \label{tour:dest}			
@@
 			\EndIf  \label{tour:lk2}	
-			\State $M \gets SetPalletsDestination( \pi(k) )$ \label{tour:dest2}		
+			\State $M \gets SetPalletsDestinations( \pi(k) )$ \label{tour:dest2}		
 			\State $G_2 \gets SolveNode(method, \pi(k), G_1)$ \label{tour:node}
@@
 \begin{algorithm}[H]
-	\caption{$UpdateConsolidated(k)$}  \label{alg:cons}
+	\caption{Procedure to update consolidated items that remain boarded on node $k$}  \label{alg:cons}
 	\begin{algorithmic}[1]
+
+		\Procedure{$UpdateConsolidated$}{$k$}
 
@@
 		\EndFor \label{cons:Yend}
-		\State ${\bf return}\ Q, E^Q_k, M$ 
+		\Return $Q, E^Q_k, M$
+		\EndProcedure
 	\end{algorithmic}
@@
 
-$SetPalletsDestination$, which set pallets destination not yet defined, is described in Algorithm \ref{alg:dest}.
+$SetPalletsDestinations$, which set pallets destination not yet defined, is described in Algorithm \ref{alg:dest}.
 The vectors $vol[0..K]$\/ and $PalCons[0..K]$\/ store the total volume of items and the number of pallets with consolidated items destined for each node, respectively (lines \ref{dest:vector1}-\ref{dest:vector2}).
@@
 \begin{algorithm}[H]
-	\caption{$SetPalletsDestination(k)$}  \label{alg:dest}
+	\caption{Procedure to set pallets destinations based on the items to be embarked on node $k$}  \label{alg:dest}
 	\begin{algorithmic}[1]
+
+		\Procedure{$SetPalletsDestinations$}{$k$}
 
@@
 		\State $total \gets 0$ 
-
 		\For{$j \gets 1$ to $n_k$}
@@
 			\If {$vol[x] \neq 0$}
-				\State $quant \gets \max \{1, \lfloor{ m \times vol[x]/total}\rfloor \}-PalCons[x]$ 
+				\State $quant \gets \max \{1, \lfloor{ m \times vol[x]/total}\rfloor \}-PalCons[x]$  \Comment{Number of pallets needed}
 				\State $np \gets 0$
@@
 					\EndIf
-					\If{$p_i.to[k] = -1$}
-						\State $p_i.to[x] \gets x$
+					\If{$p_i.to[k] = -1$}              \Comment{Pallet destination in node $k$ was unset}
+						\State $p_i.to[k] \gets x$
 						\State $np \gets np + 1$
@@
 				\EndFor
-
 			\EndIf 
@@
 			\If{$p_i.to[k] = -1$}
-				\State $p_i.to[x] \gets max$
+				\State $p_i.to[k] \gets max$
 			\EndIf
 		\EndFor \label{dest:max2}
-		\State ${\bf return}\ M$
-
+		\Return $M$
+		\EndProcedure
 	\end{algorithmic}
@@
 \begin{algorithm}[H]
-	\caption{$ScoreAndDeviation(k, G)$}  \label{alg:eval}
-
+	\caption{Procedure to calculate the solution score and the relative torque deviation on node $k$}  \label{alg:eval}
 	\begin{algorithmic}[1]
+
+		\Procedure{$ScoreAndDeviation$}{$k,\ G$}
+
 		\State Let $G(V_k, E^N_k \cup E^Q_k)$
@@
 		\State $\epsilon \gets \epsilon/maxTorque$ \label{eval:torque2}
-		\State ${\bf return}\ s, \epsilon$	
+		\Return $s,\ \epsilon$
+		\EndProcedure
+
 	\end{algorithmic}
@@
 
-In this subsection we present implementations of $SolveNode(method,k, G)$, where $method$\/ corresponds to a MIP solver or a heuristic, $k$\/ is the index of the current node and $G$\/ is the allocation graph of the consolidated items that remain on board at node $l_k$.
-
-An important issue on meta-heuristics is that they are known for the hard work in tuning parameters.
-For this, when we use the word {\it empirically}\/ throughout the text, we want to say that we accomplished many tests with the hardest instances to fit these values.
-This does not guarantee best results for all instances, but, on average, overall results are acceptable.
-Another important observation is that our heuristics were designed based on traditional ideas from the literature \cite{NiarFreville1997,Fidanova2006,Laabadi2018,Alonso2019,Zhan2020} and on the experiences learned during tests.
-
-To guarantee that performance comparisons are made correctly, our implementations use the same elements described below.
-
-\newcolumntype{C}{>{\raggedright}p{0.18\textwidth}}
-\newcolumntype{D}{>{\raggedright\arraybackslash}p{0.82\textwidth}}
-
-\bgroup
-\def\arraystretch{1.2}
-\begin{table}[H]
-	\centering
-	\small
-	\begin{tabular}{CD}
-
-		{\it Data structure:}       & Items, pallets and edges are arrays of \emph{python} class objects, which are initialized with the same data loaded from text files. \\
-
-		{\it Common procedures:}    & All heuristics use the same procedures for selecting and inserting edges, and printing solutions.  \\
-
-		{\it Stopping criterion:} & All heuristics have the same time limit of $0.7$ seconds per node.\\
-
-	\end{tabular}
-	\normalsize
-\end{table}
-\egroup
+In this subsection we present two implementations of $SolveNode(method,\ k,\ G)$, where  $k$\/ is the index of the current node and $G$\/ is the allocation graph of the consolidated items that remain on board at node $l_k$.
+These implementations correspond to the two possible values of the $method$\/ parameter: $MPI$\/ (a solver) or $Shims$\/ (a heuristic).
+
+\subsubsection{MIP solver}
+\label{solver}
+
+To find a viable solution for ACLP+RPDP, the strategy adopted in $SolveTour$\/ previously defines the values of some variables.
+Concretely, all permutations between the nodes are tested, the set of nodes to be visited is updated, the consolidated items that remain on board are reallocated to minimize the CG deviation, and the pallets destination is determined according to the volume of items available for shipment.
+In this way, the mathematical modeling for $SolveNode(MIP,\ k,\ G)$\/ becomes simpler, which finds an allocation of available items in node $l_k$\/ using previously defined values of $L_k$, $p_i.to[k]$, $Y^k_{iq}$\/ and $a^k_q$.
+
+This simplified modeling is described in the equations \ref{eq:newf} to \ref{eq:lastnew}, which maximizes the score of the $t^k_j$\/ items to be shipped, maintaining load balancing.
+The variables $X^k_{ij}$\/ define the set of edges $E^N_k$, which will be added to the graph $G$.
+
+\begin{equation} \label{eq:newf}
+	\max f_s = \sum_{i=1}^{m} \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.s
+\end{equation}
+
+\begin{equation} 
+	maxW = \min(Payload, \sum_{i=1}^{m}p_i.w)
+\end{equation}
+
+\begin{equation} 
+	\epsilon_k = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w +  \sum_{q=1}^{m} Y_{iq}^k \times a_q^k.w)]/(maxW \times limit^{CG}_{long})
+\end{equation}
+
+\begin{equation}
+	X_{ij}^k = 0 \ \mbox{if} \ t_j^k.to \notin L_k; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}
+\end{equation}
+
+\begin{equation} 
+	LatIt_k = \sum_{i=1}^{m} \sum_{j=1}^{n_k} ( X_{ij}^k \times t_j^k.w \times (i\%2) - X_{ij}^k \times t_j^k.w \times (i+1)\%2 )
+\end{equation}
+
+\begin{equation} 
+	LatCons_k =  \sum_{i=1}^{m} \sum_{q=1}^{m_k}  ( Y_{iq}^k \times a_q^k.w \times (i\%2) - Y_{iq}^k \times a_q^k.w \times (i+1)\%2)
+\end{equation}
+
+\begin{equation} 
+	s.t.: d^{CG}_{pallet} \times | LatIt_k + LatCons_k | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}
+\end{equation}
+
+\begin{equation}
+	s.t.: |\tau_k| \leq maxW \times limit^{CG}_{long}
+\end{equation}
+
+\begin{equation} 
+	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w ) \leq maxW
+\end{equation}
+
+\begin{equation} 
+	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m_k\}
+\end{equation}
+
+\begin{equation} 
+	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.v + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m_k\}
+\end{equation}
+
+\begin{equation} 
+	s.t.: \sum_{i=1}^{m} X_{ij}^k \leq 1; \ j \in \{1, 2, \ldots, n_k\}
+\end{equation}
+
+\begin{equation} \label{eq:lastnew}
+	s.t.: p_i.to[k] = t^k_j.to\ \mbox{if} \ X_{ij}^k = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}
+\end{equation}
+
+\subsubsection{Shims heuristic}
+
+Our goal is to find a heuristic that offers a good-quality solution for the node-by-node problem.
+As we are testing all $K!$\/ tours, its runtime is a crucial requirement.
+Taking this into account, we perform a series of implementations based on known meta-heuristics: {\it Ant Colony Optimization} (ACO) \cite{Dorigo1992, DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \cite{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \cite{Glover1986} and {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \cite{FeoResende1989}.
+We also considered several ideas from the literature \cite{NiarFreville1997,Fidanova2006,Laabadi2018,Alonso2019,Zhan2020} and we were careful to use the same data structures and procedures in all implementations.
+The decision for a simple and fast heuristic is due to the need to obtain, in less than an hour, a complete solution for ACLP+RPDP, allowing its use in an operational environment.
+
+However, when we established the runtime limit of $0.7$\/ seconds per node, the heuristic that presented better solutions was none of the previous ones.
+In this subsection, we will present a new heuristic for the node-by-node problem, called $Shims$.
+Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}).
+This strategy is based on a practical observation: usually, subsets of smaller and lighter items are saved for later adjustments to the remaining clearances.
 
@@
 
-We use four types of random selections:
-\begin{itemize}
-	\item $RandomReal(r_1,r_2)$: randomly selects a real number in $[r_1,r_2]$, where $r_1$ and $r_2$ are real numbers;
-	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$, where $i_1$ and $i_2$ are integer numbers;
-	\item $RandomChoice(set)$: randomly selects an element from $set$;
-	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
-\end{itemize}
-
-We also developed Algorithm \ref{alg:greedy}, which generates a greedy allocation of the items available in node $l_k$, according to the non-ascending order of $\theta^k_{ij}$, and considering the consolidated items already shipped.
-Edges are included in this allocation as long as they respect feasibility constraints.
-Furthermore, the volume of each pallet $p_i$\/ cannot exceed $p_i.v \times limit$, where $ 0 < limit \leq 1$\/ is a parameter.
-
-\begin{algorithm}[H]
-	\caption{ $Greedy(k, G, limit)$}  \label{alg:greedy}
-
-	\begin{algorithmic}[1]
-		\State Let $G(V_k, E^Q_k)$
-		\State $E^N_k \gets \varnothing$ 			
-		\For{$i \gets 1$ to $m$}
-			\State $pv[i] \gets 0$ 		
-		\EndFor
-		\For{$q \gets 1$ to $m_k$}
-			\If{$(p_i, a_q^k) \in E^Q_k$} 
-				\State $pv[i] \gets pv[i] + a_q^k.v$ 
-			\EndIf		
-		\EndFor		
-		\For {each $e_{ij}^k$ in non-ascending order of $\theta^k_{ij}$}
-			\If{($E^N_k \cup \{e_{ij}^k\}$ is feasible) {\bf and} ($pv[i] \leq p_i.v \times limit$)} 
-				\State $E^N_k \gets E^N_k \cup \{e_{ij}^k\}$ 
-				\State $pv[i] \gets pv[i] + t_j^k.v$ 	
-			\EndIf
-		\EndFor
-		\State ${\bf return}\ G(V_k, E^N_k \cup E^Q_k)$
-
-	\end{algorithmic}
-\end{algorithm}
-
-To compare allocation graphs generated by the heuristics, we choice the one that offer a higher associated score, since the CG deviation is limited by the restrictions.
-In the pseudocodes presented below, given an allocation graph $G$\/ for the node $l_k$, $f_s(G)$\/ is equivalent to the score $s$\/ returned by $ScoreAndDeviation(k,G) $, described in Algorithm \ref{alg:eval}.
-
-Below, we present six implementations for $SolveNode$: the first using a MIP solver, and five with developed heuristics: {\it Ant Colony Optimization}\/ (ACO), {\it Noising Method Optimization}\/ (NMO), {\it Tabu Search}\/ (TS), {\it Greedy Randomized Adaptive Search Procedure}\/ (GRASP) and {\it Shims}.
-
-\subsubsection{MIP solver}
-\label{solver}
-
-In order to find a viable solution for ACLP+RPDP, the strategy adopted in $SolveTour$\/ previously defines the values of some variables.
-Concretely, all permutations between the nodes are tested, the set of nodes to be visited is updated, the consolidated items that remain on board are reallocated to minimize the CG deviation, and the pallets destination is determined according to the volume of items available for shipment.
-In this way, the mathematical modelling for $SolveNode(MIP,k,G)$\/ becomes simpler, which finds an allocation of available items in node $l_k$\/ using previously defined values of $L_k$, $p_i.to[k]$, $Y^k_{iq}$\/ and $a^k_q$.
-
-This simplified modelling is described in the equations \ref{eq:newf} to \ref{eq:lastnew}, which maximizes the score of the $t^k_j$\/ items to be shipped, maintaining load balancing.
-The variables $X^k_{ij}$\/ define the set of edges $E^N_k$, which will be added to the graph $G$.
-
-\begin{equation} \label{eq:newf}
-	\max f_s = \sum_{i=1}^{m} \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.s
-\end{equation}
-
-\begin{equation} 
-	maxW = \min(Payload, \sum_{i=1}^{m}p_i.w)
-\end{equation}
-
-\begin{equation} 
-	\epsilon_k = \sum_{i=1}^{m}[ p_i.d \times (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w +  \sum_{q=1}^{m} Y_{iq}^k \times a_q^k.w)]/(maxW \times limit^{CG}_{long})
-\end{equation}
-
-\begin{equation}
-	X_{ij}^k = 0 \ \mbox{if} \ t_j^k.to \notin L_k; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}
-\end{equation}
-
-\begin{equation} 
-	LatIt_k = \sum_{i=1}^{m} \sum_{j=1}^{n_k} ( X_{ij}^k \times t_j^k.w \times (i\%2) - X_{ij}^k \times t_j^k.w \times (i+1)\%2 )
-\end{equation}
-
-\begin{equation} 
-	LatCons_k =  \sum_{i=1}^{m} \sum_{q=1}^{m_k}  ( Y_{iq}^k \times a_q^k.w \times (i\%2) - Y_{iq}^k \times a_q^k.w \times (i+1)\%2)
-\end{equation}
-
-\begin{equation} 
-	s.t.: d^{CG}_{pallet} \times | LatIt_k + LatCons_k | \leq  \sum_{i=1}^{m}p_i.w \times limit^{CG}_{lat}
-\end{equation}
-
-\begin{equation}
-	s.t.: |\tau_k| \leq maxW \times limit^{CG}_{long}
-\end{equation}
-
-\begin{equation} 
-	s.t.: \sum_{i=1}^{m} (\sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w ) \leq maxW
-\end{equation}
-
-\begin{equation} 
-	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.w + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.w  \leq p_i.w; \ i \in \{1, 2, \ldots, m_k\}
-\end{equation}
-
-\begin{equation} 
-	s.t.: \sum_{j=1}^{n_k} X_{ij}^k \times t_j^k.v + \sum_{q=1}^{m_k} Y_{iq}^k \times a_q^k.v  \leq\ p_i.v; \ i \in \{1, 2, \ldots, m_k\}
-\end{equation}
-
-\begin{equation} 
-	s.t.: \sum_{i=1}^{m} X_{ij}^k \leq 1; \ j \in \{1, 2, \ldots, n_k\}
-\end{equation}
-
-\begin{equation} \label{eq:lastnew}
-	s.t.: p_i.to[k] = t^k_j.to\ \mbox{if} \ X_{ij}^k = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n_k\}
-\end{equation}
-
-\subsubsection{Ant Colony Optimization (ACO)}
-
-In ACO \cite{Dorigo1992}, a population of $A$\/ ants performs independent search, where each ant $a$\/ finds its own allocation graph $G^a$.
-Each edge $e^k_{ij}=(p_i,t^k_j)$\/ has a general attractiveness $ga^k_{ij} = (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$, where $\tau^k_{ij}$\/ is the pheromone of this edge, and $\alpha$\/ and $\beta$\/ are constants.
-If $e^k_{ij}$\/ is included in $G^a$, ant $a$\/ increases the pheromone $\tau^k_{ij}$.
-Algorithm \ref{alg:aco} presents this heuristic.
-
-\begin{algorithm}[H]
-
-	\caption{ $SolveNode(ACO, k, G)$ } \label{alg:aco}
-	\begin{algorithmic}[1]
-		\State Let $G(V_k, E^Q_k)$
-		\State $G^*(V_k, E^{N*}_k \cup E^Q_k) \gets Greedy(k, G, 1)$  \label{aco:greedy}
-		\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^*(V_k,E^{N*}_k \cup E^Q_k)$
-		\State $A \gets 4$ \label{aco:ants}
-		\State $ga^k_{ij}   \gets 0$, $1 \leq i \leq m$, $1 \leq j \leq n_k$  \label{aco:init1}
-		\State $\tau^k_{ij} \gets 1$, $1 \leq i \leq m$, $1 \leq j \leq n_k$   \label{aco:init2}
-		\State $\alpha    \gets 1.0$ \label{aco:alpha}
-		\State $\beta     \gets 4.0$ \label{aco:beta}
-		\State $\rho      \gets 0.2$ \label{aco:rho}
-		\State $stagnant \gets 0$
-		\While{($stagnant < 3$) {\bf and} ($runtime < 0.7s$)} \label{aco:while}
-			\For{$a \gets 1$ to $A$} \label{aco:g1}
-				\State $G^a(V_k, E^{Na}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$
-				\State $E \gets \{ (p_i,t^k_j) \mid (p_i,t^k_j) \not\in E^{Na}_k \}$
-				\While{$|E| > 0$}
-					\State $e \gets Roulette(E)$ biased through $ga^k_{ij}$  \label{aco:roulette}
-					\State $E \gets E - \{e\}$		
-					\If{$E^{Na}_k \cup \{e\}$ is feasible}
-						\State $ E^{Na}_k \gets E^{Na}_k \cup \{e\} $ \label{aco:ant2}
-					\EndIf	
-				\EndWhile	
-				\If{$f_s(G^{a}) > f_s(G^{+})$} \label{aco:gen1}
-					\State $G^+(V_k, E^{N+}_k \cup E^Q_k) \gets G^a(V_k, E^{Na}_k \cup E^Q_k)$  \label{aco:gen2}
-						\For{each $(p_i,t^k_j) \in E^{N+}_k$}
-							\State $\tau^k_{ij} \gets \tau^k_{ij} - \Delta\tau^k_{ij}$ \label{aco:phero1}
-							\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{aco:ga1}
-						\EndFor
-				\EndIf 
-			\EndFor \label{aco:g2}
-			\If{$f_s(G^+) > f_s(G^*)$} \label{aco:best1}
-				\State $stagnant \gets 0$
-				\State $G^*(V_k, E^{N*}_k \cup E^Q_k) \gets G^+(V_k, E^{N+}_k \cup E^Q_k)$
-				\For{each $(p_i,t^k_j) \in E^{N*}_k$}
-					\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{aco:rho1}
-					\State $\tau^k_{ij} \gets \tau^k_{ij} + \Delta\tau^k_{ij}$ \label{aco:phero2}
-					\State $ga^k_{ij} \gets (\tau^k_{ij})^\alpha + (\theta^k_{ij})^\beta$ \label{aco:ga2}
-				\EndFor			
-			\Else
-				\State $stagnant \gets stagnant + 1$ \label{aco:best2}
-			\EndIf
-			\For{all $\tau^k_{ij}$} 
-				\State $\tau^k_{ij} \gets (1-\rho)\tau^k_{ij}$ \label{aco:rho2}
-			\EndFor	
-		\EndWhile	
-		\State ${\bf return}$ $G^*(V_k, E^{N*}_k \cup E^Q_k)$		
-	\end{algorithmic}
-\end{algorithm}
-
-A greedy solution is obtained in line \ref{aco:greedy}, being initially stored in $G^*$\/ and $G^+$.
-The number $A$\/ of ants was empirically set to 4 (line \ref{aco:ants}), yielding solutions with a good relation quality/time.
-The $m \times n_k$\/ matrices $\tau^k_{ij}$ and $ga^k_{ij}$\/ (lines \ref{aco:init1}-\ref{aco:init2}) are used and updated by all ants.
-
-In each generation, $A$\/ ants are created (lines \ref{aco:g1}-\ref{aco:g2}).
-The number of generations depends on the time limit or stagnation in improving results (line \ref{aco:while}).
-Each ant $a$\/ elaborates its allocation graph $G^a$, considering the current values of $ga^k_{ij}$.
-To enforce solution diversification, if $(p_i,t^k_j)$ is included in $G^a$\/ then pheromone $\tau^k_{ij}$\/ is decreased (line \ref{aco:phero1}).
-Similarly, the values of $ga^k_{ij}$\/ are also updated (line \ref{aco:ga1}), where the exponents $\alpha$\/ and $\beta$\/ were chosen empirically: $\alpha$ was set to 1.0 and $\beta$\/ to $4.0$\/ (lines \ref{aco:alpha}-\ref{aco:beta}).
-For further explanation, see \cite{DorigoManiezzoColorni1996}.
-
-The best solution of each generation is stored in $G^+$\/ (lines \ref{aco:gen1}-\ref{aco:gen2}), and the update of $G^*$\/ and the stopping criterion are performed on lines \ref{aco:best1}-\ref{aco:best2}.
-Each edge $(p_i,t^k_j)$\/ in some solution has its corresponding pheromone updated in a cumulative way (line \ref{aco:phero2}), that is, more pheromone will be given to the most relevant edges, where the increment is defined in \ref{eq:delta_tau}.
-The same with $ga^k_{ij}$\/ (line \ref{aco:ga2}).
-
-\begin{equation} \label{eq:delta_tau}
-	\Delta\tau^k_{ij} = \frac{ ga^k_{ij}\times t^k_j.s }{ A \times t^k_j.v}
-\end{equation}
-
-Pheromone $\tau^k_{ij}$\/ evaporates (lines \ref{aco:rho1} and \ref{aco:rho2}), where $\rho$\/ is a evaporation rate, which can be seen as a mechanism that delays premature convergence towards a sub-optimal solution.
-$\rho$\/ was empirically set to $0.2$\/ (line \ref{aco:rho}).
-
-\subsubsection{Shims}
-
-Finally, we present a new heuristic designed specifically for ACLP+RPDP, which we named \emph{Shims}.
-Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}).
-This strategy is based on a practical observation: usually, subsets of smaller and lighter items are saved for later adjustments to the remaining clearances.
-
 \begin{table}[H]
@@
 
-	\end{minipage}\hfill % these two lines must be close to each other
+	\end{minipage}\hfill 
 	\begin{minipage}{0.34\linewidth}
@@
 
-	\end{minipage}\hfill % these two lines must be close to each other
+	\end{minipage}\hfill
 	\begin{minipage}{0.58\linewidth}
@@
 
-%Examples:
-%$\to$ for a small problem if limit is 0.85 for the first phase of the solution process, remains 0.15 for second and third phases.
-%$\to$ for a big problem if it is 0.98 for the first phase, it is 0.02 for the remaining phases.
-
 \begin{algorithm}[H]
-	\caption{ $SolveNode(Shims, k, G)$}  \label{alg:shims}
+	\caption{$Shims$\/ heuristic on node $k$}  \label{alg:shims}
 	\begin{algorithmic}[1]
+
+		\Procedure{$SolveNode$}{$Shims,\ k,\ G$}
+
 		\State Let $G(V_k, E^Q_k)$
@@
 		\Until{($\eta_2[i] = n_k$) {\bf or} ($ pv[i] \geq (1 + 3*gap)*p_i.v)$)} \label{shims:phase1b}
-
 		\State $Set \gets \varnothing$  \label{shims:phase2a}
@@
 		\EndFor   \label{shims:phase2b}
-
 		\State $sh_w \gets S$, where $S \in Set$ and $\sum_{e_{ab}^k \in S} t_b^k.w$ is maximum \label{shims:phase3a}
@@
 		\EndFor
-		\State ${\bf return}$ $G(V_k, E^N_k \cup E^Q_k)$
+		\Return $G(V_k, E^N_k \cup E^Q_k)$
+		\EndProcedure
 	\end{algorithmic}
@@
 
-This section is composed of two parts: the generation of test instances and the results obtained in our implementations.
+This section is composed of two parts: the generation of test instances and the results obtained in our implementation.
 
@@
 
-As we are dealing with a new problem, which until now had not been modelled in the literature, we have to create our own benchmarks.
-For this, we based on the characteristics of real airlifts carried out by the {\em Brazilian Air Force}, as described below.
+As we are dealing with a new problem, which until now had not been modeled in the literature, we have to create our own benchmarks.
+For this, we based on the characteristics of real airlifts carried out by the {\it Brazilian Air Force}, as described below.
 
 In the military airlift carried out in Brazil from 2008 to 2010, 23\% of the items weighed between $10kg$ and $20kg$, 22\% from $21kg$ to $40kg$, 24\% from $41kg$ to $80kg$, 23\% from $81kg$ to $200kg$, and 8\% between $201kg$ and $340kg$.
-These five groups of items are described in Table \ref{tab:weights}.
+These five groups of items are described in Table \ref{tab:weights}, where $P$\/ represents the group probability.
 On the other hand, the average density of these items is approximately $246 kg/m^3$.
@@
 			\toprule
-			$item$ & $P$ & $low$ & $high$  \\		
+			$Group$ & $P$ & $low$ & $high$  \\		
 			\midrule
-			1              & 0.23           & 10  & 20\\
-			2              & 0.22           & 21  & 40\\
+			1              & 0.23           & 10  & 20  \\
+			2              & 0.22           & 21  & 40  \\
 			3              & 0.24           & 41  & 80  \\		
-			4              & 0.23           & 81  & 200\\
+			4              & 0.23           & 81  & 200 \\
 			5              & 0.08           & 201 & 340 \\
@@
 
+In the generation of test instances, we use three types of random selections:
+\begin{itemize}
+	\item $RandomReal(r_1,r_2)$: randomly selects a real number in $[r_1,r_2]$, where $r_1$ and $r_2$ are real numbers;
+	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$, where $i_1$ and $i_2$ are integer numbers;
+	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
+\end{itemize}
+
 $ItemsGeneration$, which generates $N$, is described in Algorithm \ref{alg:itemsgen}.
@@
 \begin{algorithm}[H]
-	\caption{$ItemsGeneration(scenario,volume)$}  \label{alg:itemsgen}
+	\caption{Procedure to generate items}  \label{alg:itemsgen}
 	\begin{algorithmic}[1]
+
+		\Procedure{$ItemsGeneration$}{$scenario,\ volume$}
+
 		\State Let $L$ and $M$ be according to $scenario$ \label{ig:LM}
@@
 				\Until{$t_j^k.to \neq k$}
-				\State $x = Roulette(item)$ biased through $P$ \label{ig:weight1}	
+				\State $x = Roulette(item)$ biased through $P$ \label{ig:weight1} \Comment{From Table \ref{tab:weights}}	
 				\State $t_j^k.w \gets RandomReal(low(x), high(x))$        \label{ig:weight2}		
@@
 		\State $N \gets \bigcup_{0 \leq k \leq K} N_k$ 
-		\State ${\bf return}$ $N$
+		\Return $N$
+		\EndProcedure
 	\end{algorithmic}
@@
 
+We ran Algorithm \ref{alg:main} in the 6 scenarios described in Table \ref{tab:scenarios}, considering 6 methods for node-by-node resolution: Gurobi (see \ref{solver}), ACO, NMO, TS, GRASP, and Shims (Algorithm \ref{alg:shims}).
+In generating the items in each node, we consider 3 values for the parameter $volume$.
+The results obtained for the function $f$, with the corresponding runtime in seconds, are shown in Tables \ref{tab:20} ($volume = 1.2$), \ref{tab:50} ($volume = 1.5$) and \ref{tab:100} ($volume = 2.0$).
+For each $method$, $scenario$\/ and $volume$, 7 different instances were generated.
+Therefore, 126 tests (6 scenarios $\times$\/ 3 volumes $\times$\/ 7 instances) were performed for each method.
+
+The average values were presented for $f$\/ and, for the runtime, the worst result obtained.
+To facilitate the comparison between the methods, we added a last column in these tables, where two values are indicated:
+\begin{itemize}
+\item {\bf Normalized}: value between 0 and 1, which corresponds to the ratio between the sum of $f$\/ values obtained by the method in all scenarios and the sum of the best values obtained among all methods in all scenarios.
+The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
+\item {\bf Speed-up}: ratio of the sums of the worst runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
+\end{itemize}
+
+In each $scenario$, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.
+
 \begin{table}[H]
 	\centering
-	\caption{xxx} 
+	\caption{Methods and scenarios with $volume =1.2$}  \label{tab:20}
 	\footnotesize
-
-\begin{tabular}{c c c c} 
+	\begin{tabular}{c|c|cccccc|c}
 		\toprule
-	\rowcolor{Gray}\multicolumn{4}{l}{Node 0: GRU solution} \\ 
-	$Pallet$ & $From:To$ & $Kg$ & $m^3$ \\ 
-	\midrule 
-	1 & GRU:GIG & 2504 & 10.7 \\ 
-	2 & GRU:GIG & 2405 & 10.4 \\ 
-	3 & GRU:GIG & 3710 & 15.2 \\ 
-	4 & GRU:SSA & 3386 & 14.3 \\ 
-	5 & GRU:SSA & 3835 & 15.6 \\ 
-	6 & GRU:SSA & 3124 & 12.6 \\ 
-	7 & GRU:SSA & 3518 & 13.8 \\ 
-	\midrule 
-	Score: 8911.0 & Weight: 0.86 & Volume: 1.10 & Torque: 2.24 \\ 
+		{$method$}                        & {\bf Results}    & {\bf 1}     &  {\bf 2}    & {\bf 3}      & {\bf 4}      & {\bf 5}      & {\bf 6}     & \specialcell{{\bf Normalized}\\{\bf Speed-up}}  
+		\\
 		\toprule
-	\rowcolor{Gray}\multicolumn{4}{l}{Node 1: GIG solution} \\ 
-	$Pallet$ & $From:To$ & $Kg$ & $m^3$ \\ 
-	\midrule 
-	1 & GIG:SSA & 2585 & 12.2 \\ 
-	2 & GIG:SSA & 3603 & 17.0 \\ 
-	3 & GIG:SSA & 2881 & 12.8 \\ 
-	4 & GIG:SSA & 3039 & 13.7 \\ 
-	5 & GIG:GRU & 2075 &  8.9 \\ 
-	6 & GIG:GRU & 3478 & 14.5 \\ 
-	7 & GIG:GRU & 1632 &  7.4 \\ 
-	\midrule 
-	Score: 3546.0 & Weight: 0.74 & Volume: 1.03 & Torque: 3.13 \\ 
-	\toprule 
-	\rowcolor{Gray}\multicolumn{4}{l}{Node 2: SSA solution} \\ 
-	$Pallet$ & $From:To$ & $Kg$ & $m^3$ \\ 
-	\midrule 
-	1 & SSA:GRU & 3027 & 13.2 \\ 
-	2 & SSA:GRU & 1287 &  6.2 \\ 
-	3 & SSA:GRU & 3044 & 15.0 \\ 
-	4 & SSA:GRU & 1939 &  9.1 \\ 
-	5 & SSA:GRU & 2010 &  8.9 \\ 
-	6 & SSA:GRU & 2686 & 11.8 \\ 
-	7 & SSA:GRU & 1410 &  5.8 \\ 
-	\midrule 
-	Score: 3632.0 & Weight: 0.59 & Volume: 0.83 & Torque: 2.50 \\ 
-	\bottomrule 
-	\multicolumn{4}{l}{Tour 0\ \ \ \ \ S/C: 4.291\ \ \ \ \ \ \ \ \ \ elapsed: 0.05s\ \ \ \ \ method:\ Greedy} \\ 
+		\multirow{2}{*}{\textbf{Gurobi}}  & $f$      & {\bf 10.76} & 6.10  & {\bf 10.02}& {\bf 10.82} & {\bf 12.13} & {\bf 12.10} & {\bf 0.998} \\
+		&{\bf runtime (s)} & 6         & 14     & 51        & 243       & 1,417     & 9,720     &  1.000 \\ 
+		\midrule[.1pt]
+		\multirow{2}{*}{\textbf{TS}}      & $f$     & 8.88       & 6.56     & 8.49       & 9.19       & 11.25      & 11.48      &  0.900 \\
+		&{\bf runtime (s)}& 4         & 4       & 17        & 86        & 516       & 3,600     &  2.709 \\ 
+		\midrule[.1pt]		
+		\multirow{2}{*}{\textbf{GRASP}}   & $f$   & 9.30       & 6.76     & 8.85       & 9.65       & 11.75      & 11.54      &  0.932 \\ 
+		&{\bf runtime (s)}& $<1$         & 4       & 17        & 80        & 472       & 3,265     &  2.984 \\ 
+		\midrule[.1pt]
+		\multirow{2}{*}{\textbf{ACO}}     & $f$      & 9.43       & {\bf 6.77}& 8.85      & 9.65       & 11.75      & 11.54      &  0.934 \\ 
+		&{\bf runtime (s)}& 3         & 4        & 17       & 86        & 513       & 3,596     &  2.714 \\ 
+		\midrule[.1pt]	
+		\multirow{2}{*}{\textbf{NMO}}     & $f$     & 9.15       & 6.74      & 8.80      & 9.61       & 11.76      & 11.50      &  0.928 \\
+		&{\bf runtime (s)}& $< 1$         & 4        & 17       & 86        & 508       & 3,480     &  2.796 \\ 
+		\midrule[.1pt]	
+		\multirow{2}{*}{\textbf{Shims}}   & $f$     & 10.18       & {\bf 6.77} & 8.92     & 9.60       & 11.78      & 11.61      &  0.948       \\ 
+		&{\bf runtime (s)}& $< 1$         & 2         & 8       & 31        & 166       & 1,087     &  {\bf 8.856} \\ 
 		\bottomrule
 	\end{tabular}
-
 	\normalsize
-
 \end{table}
+
+\begin{table}[H]
+	\centering
+	\caption{Methods and scenarios with $volume =1.5$}  \label{tab:50}
+	\footnotesize
+	\begin{tabular}{c|c|cccccc|c}
+		\toprule
+		{$method$}                        & {\bf Results}   & {\bf 1}     &  {\bf 2}    & {\bf 3}      & {\bf 4}      & {\bf 5}      & {\bf 6}     & \specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
+		\toprule
+		\multirow{2}{*}{\textbf{Gurobi}}  & $f$     & {\bf 15.90}& 4.756       & 8.84        & 11.94       & 16.21       &{\bf 16.47} & 0.881  \\	 
+		&{\bf runtime (s)}& 6         & 18        & 68         & 326        & 1,906      & 12,960    & 1.000  \\	 
+		\midrule[.1pt]
+		\multirow{2}{*}{\textbf{TS}}      & $f$     & 12.40      & 8.97       & 11.44       & 12.92       & 15.20       & 15.39      &  0.907 \\		
+		&{\bf runtime (s)}& 4         & 4         & 18         & 88         & 525        & 3,660     &  3.555  \\		
+		\midrule[.1pt]		
+		\multirow{2}{*}{\textbf{GRASP}}   & $f$    & 13.12      & 9.36       & 12.34       & 13.63       & 16.18       & 15.44      &  0.951 \\ 
+		&{\bf runtime (s)}& $< 1$         & 4         & 17         & 86         & 515        & 3,564     &  3.651  \\
+		\midrule[.1pt]	
+		\multirow{2}{*}{\textbf{ACO}}     & $f$      & 13.17      & 9.36       & 12.34       & 13.63       & 16.18       & 15.44      &  0.952 \\
+		&{\bf runtime (s)}& 4         & 5         & 19         & 90         & 539        & 3,720     &  3.492  \\	
+		\midrule[.1pt]	
+		\multirow{2}{*}{\textbf{NMO}}     & $f$     & 12.94      & 9.36       & 12.33       & 13.60       & 16.16       & 15.43      &  0.948 \\ 
+		&{\bf runtime (s)}& 2         & 5         & 19         & 91         & 544        & 3,780     & 3.442  \\
+		\midrule[.1pt]	
+		\multirow{2}{*}{\textbf{Shims}}   & $f$     & 14.15      & {\bf 9.39} & {\bf 12.42} & {\bf 13.72} & {\bf 16.28} & 15.49      &  {\bf 0.968} \\	
+		&{\bf runtime (s)}& $< 1$         & 3         & 11         & 48         & 265        & 1,720     &  {\bf 7.470}  \\
+		\bottomrule
+	\end{tabular}
+	\normalsize
+\end{table}
+
+\begin{table}[H]
+	\centering
+	\caption{Methods and scenarios with $volume = 2.0$}  \label{tab:100}
+	\footnotesize
+	\begin{tabular}{c|c|cccccc|c}
+		\toprule
+		{$method$}                        & {\bf Results}    & {\bf 1}     &  {\bf 2}   & {\bf 3}    & {\bf 4}    & {\bf 5}     & {\bf 6}     & \specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
+		\toprule
+		\multirow{2}{*}{\textbf{Gurobi}}  & $f$     &{\bf 22.45} & 4.92      & 6.58      & 6.34       & 7.24       & 8.15      &  0.457 \\ 
+		&{\bf runtime (s)}& 8         & 29       & 106      & 493      & 2,833     & 19,380      &  1.000  \\ 
+		\midrule[.1pt]
+		\multirow{2}{*}{\textbf{TS}}      & $f$    & 18.87      & 13.72     & 17.53     & 19.28     & 21.78      & 22.35      &  0.929 \\		
+		&{\bf runtime (s)}& 4         & 5        & 19       & 93       & 556       & 3,900     &  4.992  \\	
+		\midrule[.1pt]		
+		\multirow{2}{*}{\textbf{GRASP}}   & $f$    & 19.36      & 13.99     & 19.39     & 19.86     & 23.50      & 22.57      &  0.973 \\
+		&{\bf runtime (s)}& 2         & 6        & 21       & 97       & 571       & 3,960     &  4.906  \\	
+		\midrule[.1pt]
+		\multirow{2}{*}{\textbf{ACO}}     & $f$    & 19.36      & 13.99     & 19.39     & 19.86     & 23.50      & 22.57      &  0.973 \\
+		&{\bf runtime (s)}& 4         & 6        & 23       & 105      & 597       & 4,080     &  4.745  \\	
+		\midrule[.1pt]	
+		\multirow{2}{*}{\textbf{NMO}}     & $f$    & 18.85      & 13.99     & 19.39     & 19.85     & 23.50      & 22.56      & 0.969 \\	
+		&{\bf runtime (s)}& 4         & 9        & 27       & 120      & 681       & 4,620     & 4.184  \\	
+		\midrule[.1pt]	
+		\multirow{2}{*}{\textbf{Shims}}   & $f$     & 20.46      &{\bf 14.00} & {\bf 19.44} & {\bf 19.94} & {\bf 23.55} & {\bf 22.58} & {\bf 0.984} \\	
+		&{\bf runtime (s)}& $< 1$         & 7        & 20       & 84       & 450       & 2,918     & {\bf 6.570}  \\
+		\bottomrule
+	\end{tabular}
+	\normalsize
+\end{table}
+
+In Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, $Gurobi$\/ was the slowest method and $Shims$\/ was the fastest.
+This was expected, as $Gurobi$\/ is a MIP solver, which seeks an optimal solution within the established constraints, usually reaching its time limit on each node, while {\it Shims} is a fast heuristic.
+This way, $Gurobi$\/ even demands prohibitive times (more than 5 hours in scenario 6 of Table \ref{tab:100}), while the maximum time required by $Shims$\/ did not reach one hour.
+
+Considering the {\bf Normalized}\/ values, in Table \ref{tab:20} $Gurobi$\/ was the best ($0.998$), followed closely by $Shims$\/ ($0.948$).
+In Tables \ref{tab:50} and \ref{tab:100}, $Shims$ was the best ($0.968$\/ and $0.984$).
+As expected, the performance of $Gurobi$\/ drops sharply as the volume of shipments at the nodes increases: $0.881$\/ in Table \ref{tab:50} and $0.457$\/ in Table \ref{tab:100}.
+
+It is worth noting that the other heuristics always obtained {\bf Normalized}\/ values greater than or equal to $0.9$, and were always faster than $Gurobi$.
 
@@
 
-xxxxx
+In this work, we modeled and solved a real air transport problem, named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem}\/ (ACLP+RPDP).
+For the first time in the literature, an NP-hard problem that involves {\it simultaneously}\/ pallet assembly, load balancing and route planning is addressed, where the cost-effectiveness of transport is maximized.
+We adopted some simplifications that are not critical, but that allowed the unprecedented resolution of this problem considering more than two nodes.
+
+We consider that, in practical cases, the number $K$\/ of nodes, excluding the base, is small ($K \leq 6$), each of them with hundreds of items to be shipped.
+Thus, considering real aircraft models, we developed some node-by-node heuristics, which allow us to enumerate all $K!$\/ tours and choose the best.
+The complete process can be executed quickly on a simple handheld computer, offering good results and reducing stress for the transport planners.
+As validation, we performed tests in several scenarios with real data from {\it Brazilian Air Force}.
+
+In less than one hour, this development can establish a near optimal distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximizing the total score and minimizing fuel consumption for the airlift.
+The output, which includes the tour plan and the pallet building and arrangement plan, is an essential part of airlift: it improves flight safety, makes ground operations more efficient, and makes sure that each item gets to its right destination.
+
+Our main contribution was the mathematical modeling of this complex problem, which involves four NP-hard subproblems, and a complete process that offers a fast and good-quality solution.
+As future work, we intend to include dimensions in the items, and consider the use of multiple aircraft or load sequencing in an aircraft with two doors.
+
+The simple heuristic $Shims$\/ presented the best node-by-node results in a restricted time interval.
+Without it, it would not be possible to test all routes: after all, the greater the number of nodes, the shorter the runtime limit for node-by-node resolution.
+We also show that, as the shipment volume increases, the exact node-by-node resolution becomes unfeasible, indicating the importance of a fast and good-quality heuristic.
+
+Finally, by focusing on node-by-node resolution, the result of this work is not exclusive to aircraft and airports: it can be adapted, for example, to ships and ports, or vehicles and warehouses, or freight cars and railways.
+In these cases, it would be necessary to make some changes in the modeling: for example, modify the load balancing constraints, and consider the available space in vehicles or freight cars instead of pallets.
 
@@
 
-This research was partially supported by \textit{S\~{a}o Paulo Research Foundation} (FAPESP, grant 2016/01860-1).
-
-%\bibliographystyle{elsarticle-num} % descomentar antes de submeter
-\bibliographystyle{elsarticle-harv} % comentar antes de submeter
-
-%\section*{References}
-
-\bibliography{references}
+This research was partially supported by \textit{São Paulo Research Foundation} (FAPESP, grant 2016/01860-1).
+
+\begin{thebibliography}{00}
+
+\bibitem{LarsenMikkelsen1979} O. Larsen and G. Mikkelsen, An interactive system for the loading of cargo aircraft, {\it European Journal of Operational Research}, Vol. 4 (6), pp. 367-373, 1980.
+
+\bibitem{Brosh1981} I. Brosh, Optimal cargo allocation on board a plane: a sequential linear programming approach, {\it European Journal of Operational Research}, Vol. 8 (1), pp. 40-46, 1981.
+
+\bibitem{JohnsonGarey1985} D.S. Johnson and M.R. Garey, A 71/60 theorem for bin packing, {\it Journal of Complexity}, Vol. 1 (1), pp. 65-106, 1985.
+
+\bibitem{Glover1986} F. Glover, Future paths for integer programming and links to artificial intelligence, {\it Computers and Operations Research}, Vol. 13, pp. 533-549, 1986.
+
+\bibitem{FeoResende1989} T.A. Feo and M.G.C. Resende, A probabilistic heuristic for a computationally difficult set covering problem, {\it Operations Research Letters}, Vol. 8 (2), pp. 67-71, 1989.
+
+\bibitem{Kevin1992} K.Y.K. Ng, A multicriteria optimization approach to aircraft loading, {\it Operations Research}, Vol. 40 (6), pp. 1200-1205, 1992.
+
+\bibitem{Dorigo1992} M. Dorigo, {\it Optimization, Learning and Natural Algorithms}, PhD Thesis, Politecnico di Milano, 1992.
+
+\bibitem{CharonHudry1993} I. Charon and O. Hudry, The noising method: a new method for combinatorial optimization, {\it Operations Research Letters}, Vol. 14 (3), pp. 133-137, 1993.
+
+\bibitem{DorigoManiezzoColorni1996} M.
+Dorigo, V.
+Maniezzo and A.
+Colorni, The ant system: optimization by a colony of cooperating agents, {\it IEEE Transactions on Systems, Man, and Cybernetics}, Vol.
+26 (1), pp.
+29-41, 1996.
+
+\bibitem{NiarFreville1997} S.
+Niar and A.
+Freville, A parallel tabu search algorithm for the 0-1 multidimensional knapsack problem, Proceedings of {\it 11th International Parallel Processing Symposium}, 1997.
+
+\bibitem{Heidelberg1998} K.R. Heidelberg, G.S. Parnell and J.E. Ames, Automated air load planning, {\it Naval Research Logistics}, Vol. 45 (8), pp. 751-768, 1998.
+
+\bibitem{CharonHudry2001} I. Charon and O. Hudry, The noising methods: a generalization of some metaheuristics, {\it European Journal of Operational Research}, Vol. 135 (1), pp. 86-101, 2001.
+
+\bibitem{MongeauBes2003} M. Mongeau and C. Bes, Optimization of aircraft container loading, {\it IEEE Transaction on Aerospace and Electronic Systems}, Vol. 39 (1), pp. 140-150, 2003.
+
+\bibitem{fok2004optimizing} K.
+Fok and A.
+Chun, Optimizing air cargo load planning and analysis, Proceedings of {\it The International Conference on Computing, Communications and Control Technologies}, 2004.
+
+\bibitem{Fidanova2006} S. Fidanova, Ant colony optimization and multiple knapsack problem and model bias, {\it Numerical Analysis and Its Applications}, Springer Berlin Heidelberg, pp. 280-287, 2005.
+
+\bibitem{Chan2006} F.T.S.
+Chan, R.
+Bhagwat, N.
+Kumar.
+M.K.
+Tiwari and P.
+Lam, Development of a decision support system for air-cargo pallets loading problem: a case study, {\it Expert Systems with Applications}, Vol.
+31 (3), pp.
+472-485, 2006.
+
+\bibitem{KaluznyBohdanL2009Oalb} B.L. Kaluzny and R.H.A.D. Shaw, Optimal aircraft load balancing, {\it International Transactions in Operational Research}, Vol. 16 (6), pp. 767-787, 2009.
+
+\bibitem{MesquitaCunha2011} A.C.P.
+Mesquita and C.B.
+Cunha, An integrated heuristic based on the Scatter Search metaheuristic for vehicle routing problems with simultaneous delivery and pickup in the context of the Brazilian Air Force, {\it Transportes}, Vol.
+19, pp.
+33-42, 2011.
+
+\bibitem{Verstichel2011} J.
+Verstichel, W.
+Vancroonenburg, W.
+Souffriau and G.V.
+Berghe, A mixed integer programming approach to the aircraft weight and balance problem, {\it Procedia Social and Behavioral Sciences}, Vol.
+20, pp.
+1051-1059, 2011.
+
+\bibitem{Limbourg2012} S. Limbourg, M. Schyns and G. Laporte, Automatic aircraft cargo load planning, {\it Journal of the Operational Research Society}, Vol. 63 (9), pp. 1271-1283, 2012.
+
+\bibitem{RoesenerHall2014} A.G.
+Roesener and S.
+Hall, A nonlinear integer programming formulation for the airlift loading problem with insufficient aircraft, {\it Journal of Nonlinear Analysis and Optimization: Theory and Applications}, Vol.
+5 (1), pp.
+125-141, 2014.
+
+\bibitem{Vancroonenburg2014} W.
+Vancroonenburg, J.
+Verstichel, K.
+Tavernier and G.V.
+Berghe, Automatic air cargo selection and weight balancing: a mixed integer programming approach, {\it Transportation Research Part E: Logistics and Transportation Review}, Vol.
+65, pp.
+70-83, 2014.
+
+\bibitem{LurkinSchyns2015} V. Lurkin and M. Schyns, The airline container loading problem with pickup and delivery, {\it European Journal of Operational Research}, Vol. 244 (3), pp. 955-965, 2015.
+
+\bibitem{RoesenerBarnes2016} A.G. Roesener and J.W. Barnes, An advanced tabu search approach to the dynamic airlift loading problem, {\it Logistics Research}, Vol. 9 (1), pp. 1-18, 2016.
+
+\bibitem{PaquaySchynsLimbourg2016} C.
+Paquay, M.
+Schyns and S.
+Limbourg, A mixed integer programming formulation for the three-dimensional bin packing problem deriving from an air cargo application, {\it International Transactions in Operational Research}, Vol.
+23 (1-2), pp.
+187-213, 2016.
+
+\bibitem{PaquayLimbourgSchynsOliveira2018} C.
+Paquay, S.
+Limbourg, M.
+Schyns and J.F.
+Oliveira, MIP-based constructive heuristics for the three-dimensional bin packing problem with transportation constraints, {\it International Journal of Production Research},
+Vol. 56 (4), pp. 1581-1592, 2018.
+
+\bibitem{Laabadi2018} S.
+Laabadi, M.
+Naimi, H.
+El Amri and B.
+Achchab, The 0/1 multidimensional knapsack problem and its variants: a survey of practical models and heuristic approaches, {\it American Journal of Operations Research}, Vol.
+8, pp.
+395-439, 2018.
+
+\bibitem{YangLiuGao2018} Y. Chenguang, L. Hu and G. Yuan, Load planning of transport aircraft based on hybrid genetic algorithm, {\it MATEC Web of Conferences}, Vol. 179, pp. 1-6, 2018.
+
+\bibitem{Alonso2019} M.T.
+Alonso, R.
+Alvarez-Valdes and F.
+Parreno, A GRASP algorithm for multi-container loading problems with practical constraints, {\it A Quarterly Journal of Operations Research}, Vol.
+18, pp.
+49-72, 2019.
+
+\bibitem{BrandtStefan2019} F.
+Brandt and S.
+Nickel, The air cargo load planning problem - a consolidated problem definition and literature review on related problems, {\it European Journal of Operational Research}, Vol.
+275 (2), pp.
+399-410, 2019.
+
+\bibitem{wong2020} E.Y.C.
+Wong and K.K.T.
+Ling, A mixed integer programming approach to air cargo load planning with multiple aircraft configurations and dangerous goods, Proceedings of {\it 7th International Conference on Frontiers of Industrial Engineering}, 2020.
+
+\bibitem{Zhan2020} S. Zhan, L. Wang, Z. Zhang and Y. Zhong, Noising methods with hybrid greedy repair operator for 0-1 knapsack problem, {\it Memetic Computing}, Vol. 12, pp. 37-50, 2020.
+
+\bibitem{eugene2021} E.Y.C.
+Wong, D.Y.
+Mo and S.
+So, Closed-loop digital twin system for air cargo load planning operations, {\it International Journal of Computer Integrated Manufacturing}, Vol.
+34 (7-8), pp.
+801-813, 2021.
+
+\bibitem{zhao2021} X.
+Zhao, Y.
+Yuan, Y.
+Dong and R.
+Zhao, Optimization approach to the aircraft weight and balance problem with the centre of gravity envelope constraints, {\it IET Intelligent Transport Systems}, Vol.
+15 (10), pp.
+1269-1286, 2021.
+
+\end{thebibliography}
 

### 2023-06-03 CAIE/paper.tex
CAIE paper transferred to parallel folder

% !TeX spellcheck = en_US
%\documentclass[preprint,authoryear]{elsarticle}
\documentclass[preprint]{elsarticle}
% sudo apt-get install -y texlive-publishers

\usepackage[left]{lineno}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{multirow}

\usepackage{lineno,hyperref}
\modulolinenumbers[5]

\usepackage[none]{hyphenat}

\usepackage{url}
\usepackage{booktabs}

\usepackage{tikz}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsthm}

\usepackage{caption}
\captionsetup{labelfont=bf}
\captionsetup{skip=0pt}

\usepackage{pgfplots}
\pgfplotsset{compat=newest}
\usepackage{algorithm}

\usepackage[noend]{algpseudocode}

\usepackage[margin=0.9in]{geometry}

\usepackage[normalem]{ulem}
%\usepackage{xcolor}

\newcommand{\Comentar}[1]{\State {\cmmt{#1}}}
\newcommand{\Break}{\State \textbf{break}}
\renewcommand{\Return}{\State \textbf{return}~}
\renewcommand{\algorithmicensure}{\textbf{Parameters:}}

\renewcommand\algorithmicthen{}
\renewcommand\algorithmicdo{}

\algblock{ForEach}{EndFor}
\algblock{ForDownTo}{EndFor}
\algblock{ForTo}{EndFor}

\newcommand{\Block}[1]{\State #1 \{}
\newcommand{\EndBlock}{\State \}}

\usepackage{scrextend}
\newcommand{\boldm}[1] {\mathversion{bold}#1\mathversion{normal}}
\newcommand{\round}[1]{\ensuremath{\lfloor#1\rceil}}
\usepackage{setspace}
\usepackage{array}
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}
\hyphenation{cost-ef-fec-tive-ness}

\newcommand{\specialcell}[2][c]{%
	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\journal{Computers \& Industrial Engineering}

\begin{document}

\begin{frontmatter}

\title{Air cargo load and route planning in pickup and delivery operations}

\author{A.C.P.~Mesquita}
\ead{celio@ita.br}

\author{C.A.A.~Sanches\corref{cor1}}
\ead{alonso@ita.br}
\cortext[cor1]{Corresponding author.}

\address {Instituto Tecnol\'{o}gico de Aeron\'{a}utica - DCTA/ITA/IEC\\
Pra\c{c}a Mal. Eduardo Gomes, 50\\
S\~{a}o Jos\'{e} dos Campos - SP - 12.228-900 - Brazil}

\begin{abstract}

In aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading, for fast take-off, and mission accomplishment, especially in times of crisis, disaster relief missions, short deadlines for contracting customers, or any external pressure for immediate take-off. In addition, there are no commercially available systems that can assist the load and trip planners with the pallet building and aircraft loading and balancing demands for transport at each hub. This enables other risks, such as improper delivery, excessive fuel burn, and longer than necessary turn-around time. We defined and solved the problem of planning the loading and routing of a single aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs. This NP-hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modeled using standardized pallets in fixed positions, obeying center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs. We performed multiple experiments with a commercial solver and four well-known meta-heuristics on data based on the transport history of the {\it Brazilian Air Force}. We also designed a new heuristic that quickly finds good solutions for a wide range of problem sizes: an essential contribution, which solved all test scenarios within an acceptable operational run time.

\end{abstract}

\begin{keyword}
Load Planning \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Vehicle Routing
\end{keyword}

\end{frontmatter}

\label{sec1}
\section{Introduction}

Air cargo transport involves several sub-problems that are difficult to solve. Recently, Brandt and Nickel \cite[p. 401]{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP) and {\it Weight and Balance Problem} (WBP). Several aspects were considered in this modeling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc. They also presented a comprehensive bibliographic survey of solving methods that had been developed in different situations.

However, there are still other important challenges in air cargo transport that go beyond the definition of the ACLPP, especially with regard to the flight itinerary and the loading and unloading at each destination (or node) of this travel plan. In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Pickup and Delivery Problem} (PDP), and the search for the lowest cost route, which is the well-known {\it Traveling Salesman Problem} (TSP).

Considering air cargo transport, Table \ref{tab:sa} lists the main works in the literature and the corresponding sub-problems addressed. We also indicate whether the dimensions of the items were taken into account ({\bf 3D} or {\bf 2D}) and which solution method was used: heuristic search methods ({\bf Heu}), integer programming ({\bf Int}), or linear programming ({\bf Lin}).

\begin{table}[H]
	\centering
	\caption{Air cargo transport: literature, problems and features}  \label{tab:sa}
	\scriptsize
	\renewcommand{\arraystretch}{1.1} 
	\begin{tabular}{r|cccc|ccccc}
		\toprule
		 Work & {\bf APP}  & {\bf WBP}  &  {\bf PDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} \\
		\midrule
		Larsen and Mikkelsen \cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$ \\
		Brosh \cite{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$ \\
		Ng \cite{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ \\
		Heidelberg {\it et al.} \cite{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$ \\
		Mongeau and Bes \cite{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		Fok and Chun \cite{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
		Chan {\it et al.} \cite{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$ \\
		Kaluzny and Shaw \cite{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$ \\
		Verstichel {\it et al.} \cite{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		Mesquita and Cunha \cite{MesquitaCunha2011}   & $.$ & $.$    & $\bigstar$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\		
		Limbourg {\it et al.} \cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		Roesener and Hall \cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
		Vancroonenburg {\it et al.} \cite{Vancroonenburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		Lurkin and Schyns \cite{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		Roesener and Barnes \cite{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
		Paquay {\it et al} \cite{PaquaySchynsLimbourg2016,PaquayLimbourgSchynsOliveira2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$ \\
		Chenguang {\it el al.} \cite{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$ \\
		Wong and Ling \cite{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$  \\
		Wong {\it et al.} \cite{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$  \\
		Zhao {\it et al.} \cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
		{\bf This article}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$  \\
		\bottomrule 
	\end{tabular}
	\normalsize 
\end{table}

As can be seen, so far Lurkin and Schyns \cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem. Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a predefined tour plan restricted to two legs. It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem. Referring directly to this work, Brandt and Nickel \cite[p. 409]{BrandtStefan2019} comment: {\it However, not even these sub-problems are acceptably solved for real-world problem sizes or the models omit some practically relevant constraints}.

There are real situations that are much more complex. In this work, we consider a practical case in Brazil, which is the largest economy in Latin America. Due to its dimensions, this country has the largest air market on the continent with $2,499$\/ registered airports, of which $1,911$\/ are private and $588$\/ are public. Although it is an immense distribution network, the {\it Brazilian Air Force}\/ missions have always considered 3 to 5 nodes per flight plan.

It is important to emphasize that this data is not an imposed limitation, but a historical fact that we will explore in our solution method. Throughout this work we approach routes with up to 7 nodes, as can be seen in Table \ref{tab:costs} and in Figure \ref{fig:nodes}. Although there are many other airports of interest to {\it Brazilian Air Force}, these 7 nodes were chosen due to their high demand and short transport times. The other Brazilian airports tend to have a lower demand, which is usually met, in a less expensive way, by cabotage, rail or road transport.

\begin{table}[H]

	\begin{minipage}{0.05\linewidth}

	\end{minipage}\hfill 
	\begin{minipage}{0.45\linewidth}

		\caption{Brazilian airports distances ($km$)}  \label{tab:costs}
		\centering
		\footnotesize
		\newcolumntype{Y}{>{\centering\arraybackslash}p{0.09\textwidth}}
		\newcolumntype{X}{>{\centering}p{0.09\textwidth}}

		\begin{tabular}{X X X X X X X Y}
			\toprule
			Node & $l_0$ & $l_1$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ & $l_6$ \\
			IATA*   & GRU   & GIG   & SSA   & CNF   & CWB   & BSB   & REC \\	
			\midrule	
			GRU     & 0	    &343	&1,439   &504    &358    &866    &2,114\\
			GIG	    & 343	&0	    &1,218   &371    &677    &935    &1,876\\
			SSA	    & 1,439	&1,218	&0	    &938    &1,788   &1,062   &676\\
			CNF	    & 504	&371	&938	&0	    &851    &606    &1,613\\
			CWB	    & 358	&677	&1,788	&851	&0	    &1,084   &2,462\\
			BSB	    & 866	&935	&1,062	&606	&1,084	&0	    &1,658\\
			REC	    & 2,114	&1,876	&676	&1,613	&2,462	&1,658	&0\\
			\bottomrule
			\multicolumn{8}{c}{*International Air Transport Association}\\
			\multicolumn{8}{c}{\small\textsuperscript{Source: www.airportdistancecalculator.com}}\\
		\end{tabular}
		\normalsize

	\end{minipage}\hfill 
	\begin{minipage}{0.50\linewidth}
		\centering
		\includegraphics[scale=0.25]{Images/nodes.png}
		\captionof{figure}{A route with 7 Brazilian airports}
		\label{fig:nodes}		
	\end{minipage}
\end{table}

Cargo handling at modern airports usually involves powerful equipment such as dollies, motorized conveyors, elevator transfer vehicles, caster decking, scissor lifts, electrical controls, hydraulically adjustable heights, forklift trucks, and plenty of room in the airport cargo terminal (see Figure \ref{fig:handling}). These facilities, together with the use of standardized pallets with predefined positions on the aircraft, allow loading and unloading at each node to be carried out in about 30 minutes.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.2]{Images/handling.png}
	\captionof{figure}{Cargo handling equipment}
	\label{fig:handling}
	\small\textsuperscript{Source: NARA \& DVIDS Public Domain Archive}	
\end{figure}

However, as reported by Fok and Chun, {\it load planning (...) is usually done in roughly 2 hours before departure, when all the details of the cargo are present.} \cite{fok2004optimizing}. We may infer that this is because there is no commercial software to solve this problem. In addition, there is still the necessary time to plan the route, in order to minimize fuel consumption and ensure the transport of priority items. In these circumstances, it is of great importance to find a method that speeds up the load planning and the definition of the flight itinerary.

Our work proposes a method that prioritizes the transport of the most relevant items at each node and the fuel economy along the tour. We developed a heuristic that can be run on a simple handheld computer (such as a laptop or a tablet) that previously defines a flight plan, and provides a quick solution for cargo handling. Consequently, this method reduces the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets (regarding their positions), and a pick up and delivery plan to each node.

To the best of our knowledge, this is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP and TSP has been addressed. This new problem is named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem} (ACLP+RPDP).

This article is organized into six more sections. In Section \ref{sec2}, we give a brief review of the literature. In Section \ref{sec3}, we present the problem context and assumptions, and in Section \ref{sec4}, we describe the mathematical model and how we dealt with its issues. In Section \ref{sec5}, we describe the elaborate algorithms, whose results are presented in Section \ref{sec6}. Finally, our conclusions are in Section \ref{sec7}.

\section{Related literature}
\label{sec2}

In this section, we briefly describe the characteristics of the main works related to air cargo transport, following the chronological order of Table \ref{tab:sa}.

Larsen and Mikkelsen \cite{LarsenMikkelsen1979} developed an interactive procedure for loading 14 types of Boeing 747 into a two-leg transportation plan. Seven types of items were considered to be allocated in 17 to 42 positions. With non-linear programming and heuristics, they present a solution that minimizes positioning changes in the intermediate node, optimizing the load balancing in the aircraft.

Brosh \cite{Brosh1981} addressed the problem of planning the allocation of cargo on an aircraft. Considering volume, weight and structural constraints, the author finds the optimal load layout through a fractional programming problem.

Ng \cite{Kevin1992} developed a multi-criteria optimization approach to load the {\it C-130} aircraft of the {\it Canadian Air Force}. Based on integer programming, this model provides timely planning and improves airlift support for combat operations, solving WBP with pallets in fixed positions, and considering 20 different items.

Heidelberg {\it et al.} \cite{Heidelberg1998} developed a heuristic for 2D packing in air loading, comparing it with methods for solving the {\it Bin Packing Problem}. These authors conclude that the classical algorithms are inadequate in this context, because they ignore the aircraft balancing constraints.

Mongeau and Bes \cite{MongeauBes2003} presented a method based on linear integer programming to solve the problem of choosing and positioning containers in the {\it Airbus 340-300}. Safety and stability constraints were considered, with the objective of minimizing fuel consumption.

Fok and Chun \cite{fok2004optimizing} developed a web-based application to make efficient use of space and load balancing for an air cargo company. Based on an analysis of historical data, an operational load planning with mathematical optimization is obtained. This container load planning is usually done in roughly 2 hours before departure, when all cargo details are in place.

Chan {\it et al.} \cite{Chan2006} carried out a case study with heterogeneous pallets. To minimize the total cost of shipping, they developed a 3D packing heuristic, with a loading plan for each pallet. Although the authors do not consider load balancing or positioning of pallets in the cargo hold, this method is relevant in commercial and industrial applications, where cargo items tend to be less dense.

Kaluzny and Shaw \cite{KaluznyBohdanL2009Oalb} developed a mixed integer linear programming model to arrange a set of items in a military context that optimizes the load balance. The objective function can be chosen to minimize the deviation of the center of gravity (CG) from the target position or to maximize the function of the items loaded. This approach does not palletize items, which are arranged in the cargo bay by a 2D packing procedure.

Verstichel {\it et al.} \cite{Verstichel2011} solved WBP by selecting the most profitable subset of containers to be loaded into an aircraft by the use of mixed-integer programming. Experimental results on real-life data showed significant improvements compared to those obtained manually by an experienced planner.

Mesquita and Cunha \cite{MesquitaCunha2011} presented a heuristic for a real problem of the {\it Brazilian Air Force}, which consists of defining transport routes with simultaneous collection and delivery from a central distribution terminal.

Limbourg {\it et al.} \cite{Limbourg2012} developed a mixed-integer program for optimally rearranging a set of pallets into a compartmentalized cargo aircraft, specifically the {\it Boeing 747}.

Roesener and Hall \cite{RoesenerHall2014} solved APP and WBP as an integer programming problem, where the items are selected for shipment based on a utility score, then assigned to pallets, which will be loaded into an aircraft in a specific pallet position.
The pallets are then packed in a manner to optimize both the pallet and aircraft characteristics, such as item utility, pallet occupancy, and aircraft center of balance.
This method groups items by destinations to be served by multiple aircraft.

Vancroonenburg {\it et al} \cite{Vancroonenburg2014} presented a mixed integer linear programming model that selects the most profitable pallets, satisfying safety and load balancing constraints on the {\it Boeing 747-400}. Using a solver, these authors solved real problems in less than an hour.

As already mentioned, Lurkin and Schyns \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and PDP in air cargo transport. The authors demonstrated that this problem is NP-hard and performed some experiments with real data, noting that their model offers better results than those obtained manually.

Roesener and Barnes \cite{RoesenerBarnes2016} proposed a heuristic to solve the {\it Dynamic Airlift Loading Problem} (DALP). Given a set of palletized cargo items that require transport between two nodes in a time frame, the objective of this problem is to select an efficient subset of aircraft, partition the pallets into aircraft loads and assign them to allowable positions in those aircraft.

Paquay {\it et al.} \cite{PaquaySchynsLimbourg2016} presented a mathematical modeling to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format. Its objective is to maximize the volume used in containers, considering load balancing constraints, the presence of fragile items and the possibility of rotating these boxes. Paquay {\it et al.} \cite{PaquayLimbourgSchynsOliveira2018} developed some heuristics to solve this problem.

Chenguang {\it et al.} \cite{YangLiuGao2018} modeled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing to minimize fuel consumption.

Wong and Ling \cite{wong2020}A developed a mathematical model and a tool based on mixed integer programming for optimizing cargo in aircraft with different pallet configurations. Balance constraints and the presence of dangerous items were considered. Wong {\it et al.} \cite{eugene2021} integrated this tool to a digital simulation model, with a visualization and validation system, based on sensors that alert about load deviations.

Zhao {\it et al.} \cite{zhao2021} proposed a new modeling for WBP based on mixed integer programming. Instead of focusing on the CG deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.

As can be seen, none of these works address air cargo palletization and load balancing with route optimization in a multi-leg transportation plan for a single aircraft. This is the objective of our work: to model and elaborate heuristics for a real problem that simultaneously involves 4 intractable sub-problems: APP, WBP, PDP and TSP.

\section{Problem context and assumptions}
\label{sec3}

In this section, we describe the context of the problem addressed in this work, as well as the assumptions considered.

\subsection{Operational premises}

As we are dealing with an extremely complex and diverse problem, we decided to establish some simplifying characteristics:

\begin{itemize}

	\item At each node of the tour, the items to be allocated are characterized by weight, volume, scores, and previously known destinations. We leave the consideration of 2D or 3D items to a future work.

	\item We considered a unique pallet type: the {\it 463L Master Pallet}, a common size platform for bundling and moving air cargo. It is the primary air cargo pallet for more than 70 Air Forces and many air transport companies. This pallet has a capacity of $4,500 kg$ and $13.7 m^3$, which may be limited by its position along the cargo bay. It is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$. For more information, see {\tt www.463LPallet.com}.

	\item All items allocated on a pallet must have the same destination. A pallet which has not yet reached its destination may receive more items, although it is known that these operations of removing restraining nets increase handling time and the risk of improper delivery. We do not consider oversized cargo in this work, but only cargo items that fit on these pallets.

	\item Finally, as we are interested in minimizing fuel costs, we disregarded others costs not directly associated with aircraft flight, such as handling.

\end{itemize}

Throughout this text, we call {\it Packed Contents}\/ (PC) a set of items of same destination stacked on a pallet and covered with a restraining net (see Figure \ref{fig:larger2}). It is considered unique, having the same attributes of its components, whose values are the sum of individual scores, weights and volumes. To ensure accuracy in pickup and delivery operations, a PC must remain on board until its destination.

\subsection{Aircraft parameters and load balancing}

We consider real scenarios with an aircraft with payload of $75,000 kg$, with enough fuel for a range of $4,400km$. The aircraft layout is presented in Figure \ref{fig:larger}, where the pallets positions are identified by $p_i$, $1\leq i \leq 18$.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.4]{Images/consolidated.jpg}
	\caption{A PC on a 463L pallet inside a {\it Boeing C-17}}
	\small\textsuperscript{Source: From Wikimedia Commons, the free media repository}
	\label{fig:larger2}
\end{figure}

\begin{figure}[!h]
	\centering

	\begin{tikzpicture}[scale=1.2, samples=100]

		\filldraw[fill=blue!5!white, draw=black] (0, 0) rectangle (10.23, 2.6);

		\filldraw[fill=gray!20!white, draw=black] (0.20, 1.40) rectangle node{$p_{17}$} (1.43, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (0.20, 0.20) rectangle node{$p_{18}$} (1.43, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (1.63, 1.40) rectangle node{$p_{15}$} (2.86, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (1.63, 0.20) rectangle node{$p_{16}$} (2.86, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (3.06, 1.40) rectangle node{$p_{13}$} (4.29, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (3.06, 0.20) rectangle node{$p_{14}$} (4.29, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (4.49, 1.40) rectangle node{$p_{11}$} (5.72, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (4.49, 0.20) rectangle node{$p_{12}$} (5.72, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (5.92, 1.40) rectangle node{$p_{9}$} (7.15, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (5.92, 0.20) rectangle node{$p_{10}$} (7.15, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (7.35, 1.40) rectangle node{$p_{7}$} (8.58, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (7.35, 0.20) rectangle node{$p_{8}$} (8.58, 1.20);

		\filldraw[fill=gray!20!white, draw=black] (8.78, 1.40) rectangle node{$p_{5}$} (10.1, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (8.78, 0.20) rectangle node{$p_{6}$} (10.1, 1.20);

		\filldraw[fill=blue!5!white, draw=black] (10.23, 0) rectangle (13.27, 2.6);
		\filldraw[fill=gray!20!white, draw=black] (10.41, 1.40) rectangle node{$p_{3}$} (11.64, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (10.41, 0.20) rectangle node{$p_{4}$} (11.64, 1.20);
		\filldraw[fill=gray!20!white, draw=black] (11.84, 1.40) rectangle node{$p_{1}$} (13.07, 2.40);
		\filldraw[fill=gray!20!white, draw=black] (11.84, 0.20) rectangle node{$p_{2}$} (13.07, 1.20);

	\end{tikzpicture}

	\begin{tikzpicture}[scale=1.2, samples=100]
		\draw[black, thick][<-]  (1, 2.77) node[anchor=east, font=\fontsize{7}{3.5}\selectfont]{Forward} -- (9, 2.77) node[anchor=west, font=\fontsize{7}{3.5}\selectfont]{Ramp} ;
	\end{tikzpicture}

	\caption{Aircraft layout} \label{fig:larger}
\end{figure}

The torque applied to the aircraft must keep its CG in the operational range, which corresponds to a fixed percentage of the {\it Mean Aerodynamic Chord} \footnote{Chord is the distance between the leading and trailing edges of the wing, measured parallel to the normal airflow over the wing. The average length of the chord is known as the {\it Mean Aerodynamic Chord} (MAC).} which is considered $1.17m$ for the aircraft of this work (see Figure \ref{fig:lateral}).

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.22]{Images/lateral.png}
	\caption{Aircraft longitudinal cut, where red lines are pallets positions}
	\label{fig:lateral}
\end{figure}

Table \ref{tab:larger} shows the parameters of this aircraft. $p_i$\/ are pallets, $1\leq i \leq 18$, with weight $W_i$\/ and $V_i$\/ volume limits, respectively. $D_i^{long}$\/ and $D_i^{lat}$\/ are, respectively, the longitudinal and latitudinal distances of each pallet centroids to the CG of aircraft along both axes. These distances will be used in the calculation of the torque referring to the items allocated on each pallet. In this aircraft, as the ramp have an inclination of $25°$, we made the necessary corrections in $D_i^{long}$, $W_i$\/ and $V_i$\/ of the corresponding pallets.

\begin{table}[H]
	\centering
	\caption{Aircraft parameters}  \label{tab:larger}
	\footnotesize
	\begin{tabular}{c | c c c c c c c c c}
		\toprule
		& \multicolumn{3}{c}{$Payload$: 75,000kg} & \multicolumn{3}{c}{$limit^{CG}_{long}$: $1.170m$} &
		\multicolumn{3}{c}{$limit^{CG}_{lat}$: $0.19m$} \\
		\midrule
		\multirow{2}{*}{{\boldm{$p_i$}}}  & $p_{17}$ & $p_{15}$ & $p_{13}$ & $p_{11}$ & $p_{9}$ & $p_{7}$ & $p_{5}$ & $p_{3}$ & $p_{1}$ \\
		& $p_{18}$ & $p_{16}$ & $p_{14}$ & $p_{12}$ & $p_{10}$ & $p_{8}$ & $p_{6}$ & $p_{4}$ & $p_{2}$ \\
		\midrule 
		\multirow{2}{*}{\boldm{$D_i^{long}$} ($m$)} & -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\
		& -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\			
		\midrule 
		\multirow{2}{*}{\boldm{$D_i^{lat}$} ($m$)}  & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 \\
		& -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 \\	
		\midrule
		{\boldm{$W_i$}} ($kg$)      &   4,500   &    4,500  &   4,500   &  4,500    & 4,500     & 4,500     & 4,500     & 3,000    & 3,000   \\
		{\boldm{$V_i$}} ($m^3$)   &   14.8   &   14.8   &  14.8    &  14.8    & 14.8     & 14.8     & 14.8     & 10.0    & 7.0 \\	
		\midrule

		\textbf{Fuel cost}  & \multicolumn{9}{c}{ $c_d$ = US\$ $4.90/km$ }	 \\

		\midrule

		\textbf{Fuel consumption rate}  & \multicolumn{9}{c}{ $c_g = 5\%$} \\

		\midrule

		\textbf{Maximum weight}  & \multicolumn{9}{c}{ $W_{max} = \sum_i W_i = 75,000kg$} \\

		\bottomrule
	\end{tabular}
	\normalsize 
\end{table}

This aircraft spends $c_d$\/ dollars per kilometer flown, and can carry up to $W_{max}$\/ of cargo distributed on the pallets. The {\it fuel consumption rate}\/ $c_g$\/ is the percentage limit of cost increase  due to the CG deviation on the longitudinal axis of the aircraft. $c_g$\/ depends on the characteristics of the aircraft, and is an arbitrated value for this work that should be better evaluated in future research. It is important to consider that the $c_g$\/ tends to zero as the aircraft attitude tends to level. As the CG deviation varies from $0$\/ to $limit^{CG}_{long}$, the {\it fuel cost}\/ increase varies from $0$\/ to $c_g$.

\vspace{3mm}
We also make the following assumptions:
\begin{itemize}
	\item on each pallet, the items are distributed in such a way that their CG coincides with the centroid of the pallet;
	\item the CG of the total load must be at a maximum longitudinal distance $limit^{CG}_{long}$\/ from the CG of the aircraft;
	\item the CG of the total load must be at a maximum lateral distance $limit^{CG}_{lat}$\/ from the CG of the aircraft;
	\item the pallets are distributed in two identical rows (with odd and even indices, respectively), and the centroid of $p_i$\/ is at distance $D^{lat}_{i}$\/ from the center-line of the aircraft.
\end{itemize}

\section{The mathematical modeling}
\label{sec4}

Given the assumptions and parameters described in the previous section, we are ready to present the mathematical modeling of ACLP+RPDP.

ACLP+RPDP has the objective function \ref{eq1}, and the calculus equations \ref{eq:scores} to \ref{eq:LatPacked} subject to constraints \ref{eq:torqlat} to \ref{eq21}, which will be described below.

\subsection{Problem structure and input data}

In the definitions below and in the tests of Section \ref{sec6}, we use the values assigned in Tables \ref{tab:costs} and \ref{tab:larger}. To simplify the notation throughout this text, $p_i$\/ will be called pallet $i$, and $l_k$\/ will be called node $k$.

Let $L = \{ 0, 1, \ldots, K \}$ be the set with the $K+1$\/ nodes, and let $L_k$\/ be the set of remaining nodes when the aircraft is in the node $k$, $0 \leq k \leq K$. Therefore, $L_0 = L$.

Let $d(a,b)$ be the distance from node $a$\/ to node $b$, where $0 \leq a, b \leq K$. By definition, $d(a,a) = 0, \forall a$.

Let $C=\left[c_{a,b}\right]$\/ be the cost matrix of flights, where $c_{a,b} = c_d*d(a,b), 0 \leq a,b \leq K$.

Let $M = \{1, 2, \ldots, m \}$\/ the set of $m$\/ empty pallets assigned to specific positions within the aircraft. In Table \ref{tab:larger}, $m=18$. Each pallet $i$, $1 \leq i \leq m$, has weight capacity $W_i$, volume capacity $V_i$, longitudinal distance to the CG of aircraft $D^{long}_i$, and lateral distance to the center line of aircraft $D^{lat}_i$.

Let $N_k = \{1, \ldots, n_k \}$\/ be the set of $n_k$\/ items $t_j^k$\/ available for loading on the node $k$, $1 \leq j \leq n_k$, $0 \leq k \leq K$. For ease of notation, $t_j^k$\/ will be called item $j$\/ of the node $k$, with score $s_j$, weight $w_j$, volume $v_j$, and destination $to_j \in L_k$. Let $N = \bigcup_{0 \leq k \leq K} N_k$ be the set of items of all nodes along a tour.

Let $Q_k = \{1, \ldots, m_k \}$\/ be the set of $m_k \leq m$\/ packed contents $a^k_q$\/ at the node $k$, $1 \leq q \leq m_k$, $0 \leq k \leq K$. For ease of notation, $a^k_q$\/ will be called packed content $q$\/ of the node $k$, with total weight $w_q$, total volume $v_q$, and destination $to_q \in L_k$. By definition, $m_0 = 0$, and therefore $Q_0 = \varnothing$. Packets contents that were destined to node $k$\/ are unloaded when the aircraft arrives at this node, that is, they are not considered in $Q_k$.

\subsection{The decision variables}

Let $X_{ij}^k$\/ and $Y_{iq}^k$\/ be binary variables, where $1 \leq i \leq m$, $1 \leq j \leq n_k$, $1 \leq q \leq m_k$\/ and $0 \leq k \leq K$.

$X_{ij}^k = 1$\/ if the item $j$\/ of the node $k$\/ is assigned to the pallet $i$, and 0 otherwise.

$Y_{iq}^k = 1$\/ if the packed content $q$\/ of the node $k$\/ is assigned to the pallet $i$, and 0 otherwise.

\subsection{The allocation graph}

Allocations of items or packed contents to the pallets in the node $k$\/ can be seen as a bipartite graph $G_k(V_k, E_k)$, where:

\begin{itemize}

\item $V_k = M \cup N_k \cup Q_k$;

\item $E_k = E_{N_k} \cup E_{Q_k}$;

\item $(i, j) \in E_{N_k}$\/ if $X_{ij}^k = 1$, where $i$\/ is a pallet and $j$\/ is a item of the node $k$;

\item $(i, q) \in E_{Q_k}$\/ if $Y_{iq}^k = 1$, where $i$\/ is a pallet and $q$\/ is a packed content of the node $k$.

\end{itemize}

\subsection{The objective function}

Let $S_K = \{s: \{1, \dots, K\} \rightarrow \{1, \dots, K\} \}$\/ be the set of $K!$ permutations $\pi$, which correspond to all possible tours (or itineraries) that have node $0$\/ as origin and end, passing through the other $K$\/ nodes. Let $\pi_k$\/ be $k^{th}$\/ node of the tour $\pi$, $1 \leq k \leq K$. In this way, the tour $\pi$\/ is described as $\{0,\pi_1,\ldots,\pi_K, 0\}$, {\color{blue}$K > 1$}. For ease of notation, we can define $\pi_0 = \pi_{K+1} = 0$.

We consider K > 1 because the case K = 1 has a single and trivial solution for the tour.

The objective of ACLP+RPDP is to find the permutation $\pi \in S_K$, with the corresponding allocation of items on the pallets at each node, that maximizes the function $f_\pi = \tilde{s}_\pi/\tilde{c}_\pi$ \ref{eq1}, where $\tilde{s}_\pi$\/ is the total score of transported items, and $\tilde{c}_\pi$\/ is the total cost of fuel consumed.

\begin{equation} \label{eq1}
	\max_{\pi \in S_K} f_\pi = \tilde{s}_\pi/\tilde{c}_\pi
\end{equation}

\subsection{The calculus equations}

$\tilde{s}_\pi$\/ \ref{eq:scores} is the sum of the scores of the items loaded on the aircraft throughout the tour $\pi$.

\begin{equation} \label{eq:scores}
	\tilde{s}_\pi = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times s_j
\end{equation}

$\tau_{\pi_k}$\/ \ref{eq:tau} is the longitudinal torque applied by the loaded pallets, in the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.

\begin{equation} \label{eq:tau}
	\tau_{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times \Big ( \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{long}; \ k \in \{0, \ldots, K\}
\end{equation}

In this way, we can calculate the cost $\tilde{c}_\pi$\/ \ref{eq:costs}, referring to the total fuel consumed in the tour $\pi$\/ due to the distances traveled and the CG longitudinal deviations in each pallet.

\begin{equation} \label{eq:costs}
	\tilde{c}_{\pi} = \sum_{k=0}^{K} \Big [ c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|) \Big ] 
\end{equation}

The sets of nodes not yet visited are defined in \ref{eq:pdp11} and \ref{eq:departing}.

\begin{equation} \label{eq:pdp11}
	L_0 = L
\end{equation}

\begin{equation} \label{eq:departing}
	L_{\pi_k} = L_{\pi_{k-1}} - \{\pi_k\}; \ k \in \{1, \ldots, K\}
\end{equation}

At the beginning of the tour $\pi$, there are no packed contents \ref{eq:pdp13}.

\begin{equation} \label{eq:pdp13}
	m_0 = 0
\end{equation}

$\epsilon_{\pi_k}^t$\/ \ref{eq:LatItem} and $\epsilon_{\pi_k}^a$\/ \ref{eq:LatPacked} are the lateral torque applied by the loaded pallets, in the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft. For ease of notation, $\epsilon_{\pi_k}^t$\/ corresponds to the items, and $\epsilon_{\pi_k}^a$\/ to the packed contents.

\begin{equation} \label{eq:LatItem}
	\epsilon_{\pi_k}^t = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{j=1}^{n_{\pi_k}} \Big ( X_{ij}^{\pi_k} \times w_j \times (i\%2) - X_{ij}^{\pi_k} \times w_j \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}; \ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:LatPacked}
	\epsilon_{\pi_k}^a = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{q=1}^{m_{\pi_k}} \Big ( Y_{iq}^{\pi_k} \times w_q \times (i\%2) - Y_{iq}^{\pi_k} \times w_q \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}; \ k \in \{0, \ldots, K\}
\end{equation}

\subsection{The constraints}

Finally, we can consider the constraints on each node $\pi_k$:

\begin{itemize}
	\item The longitudinal \ref{eq:torqlong} and the lateral \ref{eq:torqlat} torques must be within the limits of the aircraft.
	\item The items allocated to each pallet cannot exceed its weight \ref{eq:app2} and volume \ref{eq:app3} limits.
	\item At most, each item is associated with a single pallet \ref{eq:app4};
	\item Packet contents that have not yet reached their destination must remain on board \ref{eq:app5}.
	\item Items allocated on the same pallet must have the same destinations. In this case, we need two constraints: \ref{eq18} and \ref{eq19}. If $X_{ia}^{\pi_k} = X_{ib}^{\pi_k} = 1$, both constraints require that $to_a = to_b$; otherwise these constraints have no effect.
	\item If there is a packed content on the pallet, it must also have the same destination as the other items. Similarly, we use two constraints: \ref{eq20} and \ref{eq21}.
\end{itemize}

\begin{equation} \label{eq:torqlong}
	| \tau_{\pi_k} | \leq 1;\ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:torqlat}
	| \epsilon_{\pi_k}^t + \epsilon_{\pi_k}^a| \leq  1; \ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:app2}
	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}; \ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:app3}
	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}; \ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:app4}
	\sum_{i=1}^{m} X_{ij}^{\pi_k} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:app5}
	\sum_{i=1}^{m} Y_{iq}^{\pi_k} = 1;\ to_q \in L_{\pi_k}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq18}
	to_a - to_b \geq -K \times (1-X_{ia}^{\pi_k} \times X_{ib}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ a,b \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq19}
	to_a - to_b \leq K \times (1-X_{ia}^{\pi_k} \times X_{ib}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ a,b \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq20}
	to_j - to_q \geq -K \times (1-X_{ij}^{\pi_k} \times Y_{iq}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq21}
	to_j - to_q \leq  K \times (1-X_{ij}^{\pi_k} \times Y_{iq}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
\end{equation}

Constraints \ref{eq18} to \ref{eq21} are not within standard {\it Mixed-Integer Programming}\/ (MIP), and therefore need to be handled in a different way. In the next section, we will explain the strategy adopted to solve ACLP+RPDP.

\section{Solution strategy}
\label{sec5}

Once the assumptions of this work and the mathematical modeling of the problem are presented, it is possible to see that ACLP+RPDP is {\it NP-hard}. In a similar way to Lurkin and Schyns \cite[p. 6]{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$\/ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in node 0, and no items in node 1. Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is {\it NP-complete}.

Throughout our research, we have thoughtfully described the ACLP+RPDP model in standard MIP format and found that no MIP solver can handle its practical cases in a feasible time. Thus, as ACLP+RPDP is highly complex, involving four intractable subproblems (APP, WBP, PDP and TSP), our strategy will be to focus only on real cases and to develop quick node-by-node solutions, not necessarily optimal, that allow to build a complete tour.

The ACLP+RPDP solution defines a tour and a corresponding loading and unloading plan at each node. In this way, the loading and unloading time will be limited only by the use of the equipment at each node. Therefore, there will only be waiting time at the base node, corresponding to the runtime of ACLP+RPDP algorithm. For this reason, our method will also be parameterized by the maximum time that can be waited until the start of the load in the base node.

On the other hand, we know that transport aircraft generally have a few dozen pallets, the flight plan has less than $6$\/ nodes, and each node has hundreds of items to be shipped. Under these circumstances, we can adopt two more strategies:

\begin{itemize}

\item Considering the number of destinations less than the number of pallets ($K < m$), we can preset the destinations of the pallets in each node. In this way, we will reserve a number of pallets proportional to the volume demanded by each destination at the shipping node. We could have used another criterion, but it was observed in the experiments that volume is more constrictive in airlift.

\item If the number of nodes is small and we have quick node-by-node solutions, we have the possibility to check the solution corresponding to the tour with the shortest total distance, and even test all possible tours, selecting the one that provides the best value for the objective function.

\end{itemize}

Our complete strategy is summarized in Algorithm \ref{alg:main}.

\begin{algorithm}[H]
	\caption{ACLP+RPDP solving}  \label{alg:main}
	\begin{algorithmic}[1]

		\Procedure{$ACLP+RPDP$}{$scenario, surplus, time$}

		\State Let aircraft and $M$\/ be according to Table \ref{tab:larger}  \label{main:M} \Comment{Aircraft parameters}

		\State Let $K$, $L$\/ and $C$ be according to $scenario$\/ and Table \ref{tab:costs} \label{main:KLC} \Comment{Nodes, distances and costs}

		\State Let $\pi_{TSP}$\/ be the tour with the shortest distance traveled

		\State $N \gets ItemsGeneration(scenario, surplus)$ \label{main:items} \Comment{Items available for shipment}

		\For {each $method$} \Comment{$method$ is a MIP solver or a heuristic}

			\State $f_{\pi_{TSP}} \gets SolveTour(\pi_{TSP}, L, M, C, N, method, time/(K+1))$ \label{main:f1}
			\State $answer1[scenario,surplus,method] \gets f_{\pi_{TSP}}$ \label{main:a1} \Comment{Results corresponding to $\pi_{TSP}$}
			\For {each $\pi \in S_K$} \label{main:loop1} \Comment{$\pi$ is a permutation of nodes (a tour)}
				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, time/K!)$ \Comment{$f_{\pi}$\/ is a solution to the tour $\pi$} \label{main:f2}
			\EndFor \label{main:loop2}
			\State $answer2[scenario,surplus,method] \gets \max f_{\pi}$ \label{main:a2} \Comment{Best result among all permutations}
		\EndFor

		\Return $answer$

		\EndProcedure

	\end{algorithmic}
\end{algorithm}

The aircraft parameters are in Table \ref{tab:larger} (line \ref{main:M}). In this algorithm, we use five values for $scenario$, according to Table \ref{tab:scenarios}, which defines $K$\/ and the set $L$\/ of nodes (line \ref{main:KLC}).

\vspace{2.0mm}
\begin{table}[H]
	\centering
	\caption{Testing scenarios}  \label{tab:scenarios}
	\begin{tabular}{c c c }
		\toprule
		{\bf Scenario} & {$K$} & {$L$} \\		
		\midrule
		1 & 2    & \{$1,2$\}           \\
		2 & 3    & \{$1,2,3$\}         \\
		3 & 4    & \{$1,2,3,4$\}       \\
		4 & 5    & \{$1,2,3,4,5$\}     \\
		5 & 6    & \{$1,2,3,4,5,6$\}   \\
		\bottomrule
	\end{tabular}
\end{table}

$surplus$\/ is a value in $\{1.2, 1.5, 2.0\}$, which corresponds, at each node $k$, to the ratio between the sum of the volumes of the items  and the load capacity of the pallets ($surplus = \sum_{j=1}^{n_k} v_j$/$\sum_{i=1}^{m} V_i$). This parameter allows us to verify the different behavior of each $method$, according to $scenario$\/ and the quantity of items available for shipment. It is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).

$time$\/ is a runtime limit, which will be distributed among the node-by-node solutions (lines \ref{main:f1} and \ref{main:f2}). Each tour consists of $K+1$\/ trips, and the total number of tours is $K!$.

$method$\/ corresponds to a MIP solver or a heuristic that we will present in subsection \ref{methods}, where $SolveTour$\/ is the node-by-node solution.

The results corresponding to the shortest tour are stored in $answer1$\/ (line \ref{main:a1}), and those obtained by testing all tours are stored in $answer2$\/ (line \ref{main:a2}).

Next, we will present two subsections: in the first we explain how $SolveTour$\/ is executed, presetting the destinations of the pallets. In the second we will present the heuristic developed for node-by-node solutions.

\subsection{SolveTour algorithm}
\label{tour}

As we commented in the previous subsection, we will adopt the strategy of presetting the destinations of each pallet throughout the tour. This is feasible in practical cases, where $K<m$. For this, each pallet $i$\/ also has a field $T^k_i$, $0\leq k \leq K$, which stores its next destination after being loaded in node $k$. For this reason, $T^k_i \in L_k$, $1 \leq i \leq m$, $0\leq k \leq K$.

$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $time$\/ is the runtime limit that will be distributed among the $K+1$\/ legs of the tour.

\begin{algorithm}[H]
	\caption{Procedure to solve tour $\pi$ with $method$ (MIP solver or heuristic)}  \label{alg:tour}
	\begin{algorithmic}[1]

		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method$, $time$}

		\State $\pi_0     \gets 0$ \label{tour:pi1} \Comment{Base $0$\/ is the first node}
		\State $\pi_{K+1} \gets 0$ \label{tour:pi2} \Comment{Base $0$\/ is the last node}

		\State $score \gets 0$ \label{tour:score}
		\State $cost  \gets 0$ \label{tour:cost}

		\For {$k \gets 0$ to $K$} \label{tour:loop1}	
		\State $L_{\pi_k} \gets L - \{\pi_0,\pi_1,\ldots,\pi_k\}$  \label{tour:lk1}	\Comment{$\pi_k$ is the current node}	
		\For{$i \gets 1$ to $m$}
		\State $T_i^{\pi_k} \gets -1$ \label{tour:-1} \Comment{Unset the pallet destinations}
		\EndFor

		\If {$k = 0$}
		\State Let $G_1(M \cup N_0, \varnothing)$ \label{tour:g11}
		\Else

		\State $E_{Q_{\pi_k}}, M \gets UpdatePacked(M, Q_{\pi_k}, \pi_k)$ \label{tour:dest}			
		\State Let $G_1(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$ \label{tour:g12}
		\EndIf  \label{tour:lk2}

		\State $M \gets SetPalletsDestinations(M, \pi_k )$ \label{tour:dest2}

		\State $G_2 \gets SolveNode(method, \pi_k, G_1, time/(K+1))$ \label{tour:node}

		\State $s, \tau \gets ScoreAndDeviation(\pi_k, G_2)$ \label{tour:analyse}

		\State $score \gets score + s$ \label{tour:score2}
		\State $cost  \gets cost  + c_{\pi_k,\pi_{k+1}} \times (1 + c_g \times |\tau|)$ \label{tour:cost2}

		\EndFor  \label{tour:loop2}

		\Return $score / cost$ \label{tour:f}

		\EndProcedure

	\end{algorithmic}
\end{algorithm}

As we mentioned in the previous section, all tours start and end at the base $0$\/ (lines \ref{tour:pi1}-\ref{tour:pi2}). After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ flights (lines \ref{tour:loop1}-\ref{tour:loop2}). Initially, the set $L_{\pi_k}$\/ of remaining nodes is updated (line \ref{tour:lk1}), and the pallet destinations are unset (line \ref{tour:-1}).

When the aircraft is at the base, the initial graph $G_1$\/ is empty because it has no packed content \ref{tour:g11}. Otherwise, $UpdatePacked$\/ (line \ref{tour:dest}) returns the set of packed contents that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation. This allocation is stored in graph $G_1$\/ (line \ref{tour:g12}).

$SetPalletsDestinations$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node, without changing the pallets destination with packed contents.

Finally, $SolveNode$\/ includes the edges corresponding to the items shipped at the current node, returning the graph $G_2$\/ (line \ref{tour:node}). The score and the CG deviation of $G_2$\/ are calculated (line \ref{tour:analyse}) and accumulated (lines \ref{tour:score2}-\ref{tour:cost2}), allowing the final result of this tour (line \ref{tour:f}).

$UpdatePacked$, described in Algorithm \ref{alg:cons}, finds the best packed-pallet allocation, in terms of CG deviation, for the packed contents that remain on board.

\begin{algorithm}[H]
	\caption{Procedure to update the PC that remain boarded on node $\pi_k$}  \label{alg:cons}
	\begin{algorithmic}[1]

		\Procedure{$UpdatePacked$}{${M, Q_{\pi_k}, \pi_k}$}

		\State $E_{Q_{\pi_k}} \gets MinCGDeviation(E_{Q_{\pi_k}})$ \label{cons:minCG}

		\For{$i \gets 1$ to $m$} \label{cons:Ybegin}

		\For{$q \gets 1$ to $m_{\pi_k}$}

		\State $T_i^{\pi_k} \gets -1$

		\If{$(i, q) \in E_{Q_{\pi_k}}$} 
		\State $T_i^{\pi_k} \gets to_q$   \Comment{Set the pallet destinations}
		\EndIf
		\EndFor		
		\EndFor \label{cons:Yend}

		\Return $E_{Q_{\pi_k}}, M$

		\EndProcedure

	\end{algorithmic}
\end{algorithm}

$MinCGDeviation$\/ (line \ref{cons:minCG}) is run through a MIP solver with the equations \ref{eq:torque}, \ref{eq:embarked} and \ref {eq:one}, to relocate the packed contents on the pallets, minimizing torque and ensuring that they all remain on board, one packed content on each pallet. As there are few variables, the MIP solver returns an allocation $E_{Q_{\pi_k}}$\/ in less than 30 milliseconds. Finally, the destination of each pallet with a packed content is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).

\vspace{3mm}
$MinCGDeviation$ objective function:

\begin{equation} \label{eq:torque}
	\mbox{minimize}\ \Big | \sum_{i=1}^{m} \sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \times w_q \times D_i^{long}  \Big |
\end{equation}

\vspace{3mm}
s.t.:

\begin{equation} \label{eq:embarked}
	\sum_{i=1}^{m} Y^k_{iq} = 1;\ q \in \{1,\ldots,m_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq:one}
	\sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \leq 1;\ i \in \{1,\ldots,m\}
\end{equation}

$SetPalletsDestinations$, which sets the pallets destination not yet defined, is described in Algorithm \ref{alg:dest}.

\begin{algorithm}[H]
	\caption{Procedure to set pallets destination based on the items to be embarked on node $\pi_k$}  \label{alg:dest}
	\begin{algorithmic}[1]

		\Procedure{$SetPalletsDestinations$}{$M, \pi_k$}

			\For{$x \in L_{\pi_k}$}		
				\State $vol_x \gets  0$  \label{dest:vector1}
			\EndFor

			\State $max \gets 0$

			\State $total \gets 0$

			\For{$j \gets 1$ to $n_{\pi_k}$}

				\If {$to_j \in L_{\pi_k}$}

					\State $vol_{to_j} \gets vol_{to_j} + v_j$ 
					\State $total \gets total + v_j$

					\If {$vol_{to_j} > vol_{max}$}

						\State $max \gets to_j$ \label{dest:max} \Comment{$max$\/ is the destination with maximum volume demand}
					\EndIf

				\EndIf

			\EndFor

			\For{$x \in L_{\pi_k}$} \label{dest:propor1}

				\If {$vol_{x} \neq 0$}

					\State $needed \gets \max \{1, \lfloor{ (m-m_{\pi_k}) \times vol_{x}/total}\rfloor \}$  \Comment{Number of pallets to node $x$}

					\State $np \gets 0$

					\For{$i \gets 1$ to $m$}

						\If{$np < needed$ {\bf and} $T_i^{\pi_k} = -1$}
							\State $T_i^{\pi_k} \gets x$
							\State $np \gets np + 1$
						\EndIf

					\EndFor
				\EndIf

			\EndFor \label{dest:propor2}

			\For{$i \gets 1$ to $m$} \label{dest:max1}
				\If{$T_i^{\pi_k} \gets -1$}  \Comment{Remaining pallets}
					\State $T_i^{\pi_k} \gets max$
				\EndIf
			\EndFor \label{dest:max2}

			\Return $M$

		\EndProcedure

	\end{algorithmic}
\end{algorithm}

$vol$\/ stores the demand volume of items destined to the non-visited nodes (line \ref{dest:vector1}). The destination of empty pallets is defined proportionally to the volume of items to be embarked (lines \ref{dest:propor1}-\ref{dest:propor2}). The destination with the maximum volume defines any remaining pallets (lines \ref{dest:max1}-\ref{dest:max2}).

$ScoreAndDeviation$\/ is described in Algorithm \ref{alg:eval}, which evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.

\begin{algorithm}[H]
	\caption{Procedure to calculate the score and the relative torque deviation of graph $G$\/ in node $\pi_k$}  \label{alg:eval}
	\begin{algorithmic}[1]

		\Procedure{$ScoreAndDeviation$}{$\pi_k, G$}

		\State Let $G(V_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$

		\State $s \gets 0$
		\For{$i \gets 1$ to $m$}
			\State $\tau_i \gets 0$
		\EndFor
		\For{$i \gets 1$ to $m$}	\label{eval:loop1}

			\For{$j \gets 1$ to $n_{\pi_k}$}

				\If{$X_{ij}^{\pi_k} = 1$}
					\State $s \gets s + s_j$ \label{eval:score1}
					\State $\tau_i \gets \tau_i + w_j \times D_i^{long}$ \label{eval:eps1}
				\EndIf
			\EndFor

			\For{$q \gets 1$ to $m_{\pi_k}$}
				\If{$Y_{iq}^{\pi_k} = 1$}
					\State $s \gets s + s_q$   \label{eval:score2}
					\State $\tau_i \gets \tau_i + w_q \times D_i^{long}$  \label{eval:eps2}
				\EndIf
			\EndFor	   \label{eval:loop2}
		\EndFor

		\State $\tau \gets  \sum_{1}^{m} \tau_i / (W_{max} \times limit^{CG}_{long})$ \label{eval:torque}

		\Return $s,\ \tau$
		\EndProcedure

	\end{algorithmic}
\end{algorithm}

This algorithm consists of a loop that goes through all the pallets (lines \ref{eval:loop1}-\ref{eval:loop2}), accumulating the scores (lines \ref{eval:score1} and \ref{eval:score2}) and the torques (lines \ref{eval:eps1} and \ref{eval:eps2}) of the shipped items, allowing the final calculation of the CG deviation (line \ref{eval:torque}).

\subsection{Node-by-node solutions}
\label{methods}

In this subsection we present two implementations of $SolveNode(method,\ \pi_k,\ G)$, where  $\pi_k$\/ is the current node and $G$\/ is the allocation graph of the PC that remain on board at node $\pi_k$. These implementations correspond to the two possible values of the $method$\/ parameter: $MIP$\/ ({\it Gurobi}) or $Shims$\/ (a heuristic).

\subsubsection{MIP solver}
\label{solver}

To find a viable solution for ACLP+RPDP, the strategy adopted in $SolveTour$\/ previously defines the values of some variables. Concretely, all permutations between the nodes are tested, the set of nodes to be visited is updated, the PC that remain on board are reallocated to minimize the CG deviation, and the pallets destination is determined according to the volume of items available for shipment. In this way, the mathematical modeling for $SolveNode(MIP,\ \pi_k,\ G)$\/ becomes simpler, which finds an allocation of available items in node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $to_i^{\pi_k}$, and $a^{\pi_k}_q$.

This simplified modeling is described in the equations \ref{eq:obj2} to \ref{eq:if2}, which maximizes the score of the $n_{\pi_k}$ items to be shipped, maintaining load balancing.

The variables $X_{ij}$ and $Y_{iq}$ define the sets of edges $E_{N_{\pi_k}}$ and $E_{Q_{\pi_k}}$, respectively, which will be added to the graph $G$.

\vspace{3mm}
\textbf{Objective function:}

\begin{equation} \label{eq:obj2}
	maximize\ \ \tilde{s} / \tilde{c}
\end{equation}

\vspace{3mm}
\textbf{Where:}

\begin{equation} \label{eq:newf}
	\tilde{s} = \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij} \times s_j
\end{equation}

\begin{equation} 
	W_{max} = \sum_{i=1}^{m} W_i
\end{equation}

\begin{equation} 
	\tau^{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q) \Big ] \Big / W_{max} \times limit^{CG}_{long}
\end{equation}

\vspace{3mm}
In Equation \ref{eq:costs2}, the first parcel denotes the fuel cost from the base until the last node, and the second the fuel cost in the last leg, from the last node back to the base. Note that $c_g$ is the fuel consumption potential increase due to the loaded aircraft CG deviation in the node ($\tau_{\pi_k}$).

\begin{equation} \label{eq:costs2}
	\tilde{c} = \sum_{k=0}^{K} \Big [ c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|) \Big ]
\end{equation}

\vspace{3mm}
\textbf{s.t.:}

{\color{blue} It was observed in the experiments that the lateral torque was always in the order of magnitude -3. So we did not consider these constraints.}

\begin{equation}
	|\tau_{\pi_k}| \leq 1
\end{equation}

\begin{equation} 
	\sum_{i=1}^{m} (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q ) \leq W_{max}
\end{equation}

\begin{equation} 
	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}
\end{equation}

\begin{equation} 
	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}
\end{equation}

\begin{equation} 
	\sum_{i=1}^{m} X_{ij} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq:if1}
	to_i^{\pi_k} = to_j;\ X_{ij} = 1; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq:if2}
	X_{ij}       = 0;\ to_j \notin L_{\pi_k}; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
\end{equation}

\subsubsection{Shims heuristic}

Our goal is to find a heuristic that offers a good-quality solution for the node-by-node problem. As we are testing all $K!$\/ tours, its run time is a crucial requirement. Taking this into account, we perform a series of implementations based on known meta-heuristics: {\it Ant Colony Optimization} (ACO) \cite{Dorigo1992, DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \cite{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \cite{Glover1986} and {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \cite{FeoResende1989}.

{\color{blue}
These meta-heuristics were implemented in their standard forms. We did not use any specialized features to allow for the exploitation of the problem structure.
}

We also considered several ideas from the literature \cite{NiarFreville1997,Fidanova2006,Laabadi2018,Alonso2019,Zhan2020}, and we were careful to use the same data structures and procedures in all implementations. The decision for a simple and fast heuristic is due to the need to obtain, in less than an hour {\color{blue} (because the {\it Brazilian Air Force}'s transportation teams usually performs cargo planning in more than 2 hours)}, a complete solution for ACLP+RPDP, allowing its use in an operational environment.

However, when we established the run time limit of $0.7$\/ seconds per node, the heuristic that presented better solutions was none of the previous ones. In this subsection, we will present a new heuristic for the node-by-node problem, called $Shims$. Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}). This strategy is based on a practical observation: usually, subsets of smaller and lighter items are saved for later adjustments to the remaining {\color{blue}available space}.

The selection of edges for $E_{ N_{\pi_k} }$ uses the {\it edge attractiveness}\/ $\theta^{\pi_k}_{ij}$\/ \ref{eq:edge}, which can be understood as the tendency to allocate the item $j$ to pallet $i$. It is directly proportional to the score, and inversely proportional to the volume and the torque of each item.

\begin{equation} \label{eq:edge}
	\theta^{\pi_k}_{ij} = \frac{ s_j }{ v_j }\ \times \Big ( 1 - \frac{ w_j \times |D_i^{long}| }{ w_{max} \times\ | D_{max}^{long} | } \Big );\ i \in \{1,\ldots,m\},\ j \in \{1,\ldots,n_{\pi_k}\}
\end{equation}

{\color{blue}The denominator $w_{max} \times\ | D_{max}^{long} | $ represents the heaviest item put on the most distant pallet.}

\begin{table}[H]

	\begin{minipage}{0.08\linewidth}

	\end{minipage}\hfill % these two lines must be close to each other
	\begin{minipage}{0.34\linewidth}

		\includegraphics[scale=0.25]{Images/shims.png}
		\captionof{figure}{{\it Shims} of various thicknesses}
		\small\textsuperscript{Source: www.mscdirect.com/product/details/70475967}
		\label{fig:shims}

	\end{minipage}\hfill % these two lines must be close to each other
	\begin{minipage}{0.58\linewidth}

		\begin{tikzpicture}[scale=0.75, samples=100]
			\filldraw[fill=blue!3!white, draw=black] (0, 0) rectangle (12.5, 5.5);
			\draw[->] (.2, .6) -- coordinate (x axis mid) (12, .6);
			\node at (5, 0.3) {$\eta_1^{\ i}$};
			\node at (5, 2.5) {|};
			\node at (6, 2.5) {shims};
			\node at (7, 2.5) {|};
			\node at (7, 0.3) {$\eta_2^{\ i}$};
			\node at (0.3, 5) {$\theta_{ij}^{\pi_k}$};
			\draw[->] (.6, .2) -- coordinate (y axis mid) (0.6, 5.3);
			\node at (12, 0.3) {$e_{ij}^{\pi_k}$};
			\draw[smooth, domain = 0.09:2, color=black, thick] plot (.3+1/\x,{4.2+log2(\x)});
			\draw[->, dashed] (5, 0.6)--(5, 2.0);
			\draw[->, dashed] (7, 0.6)--(7, 1.5);
		\end{tikzpicture}

		\captionof{figure}{$n_{\pi_k}$\/ edges $e_{ij}^{\pi_k}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi_k}$\/ in non-ascending order}
		\label{fig:whip}		
	\end{minipage}
\end{table}

Figure \ref{fig:whip} represents the $n_{\pi_k}$\/ possible edges $e_{ij}^{\pi_k}$\/ of $i$\/ sorted by $\theta_{ij}^{\pi_k}$. \emph{Shims}\/ starts with a greedy solution, stopping at the edge with index $\eta_1^i$ (first phase). Then, considering even the edge with the index $\eta_2^i$, it elaborates different possible complements for this pallet (second phase) and selects the best of these complements (third phase).

In this heuristic, described in Algorithm \ref{alg:shims}, pallets $i$\/ are considered in non-descending order of $|D_i^{long}|$ (line \ref{shims:pallets}). For each pallet, the $n_{\pi_k}$\/ possible edges $e_{ ij}^{\pi_k}$\/ are considered in non-increasing order of $\theta_{ij}^{\pi_k}$\/ (line \ref{shims:possible}) in three phases: the greedy phase (until line \ref{shims:endfirst}), the composition phase of the shims (lines \ref{shims:beginsecond}-\ref{shims:endsecond}) and the selection phase of the best shim (lines \ref{shims:beginthird}-\ref{shims:endthird}).

{\color{blue}
	In the beginning of the first phase (line \ref{shims:initQ}), the set $Q_{\pi_k}$ represents the PC whose destination is not $\pi_k$, the current node, thus remaining on board.

	It is important to remember that $Q_{\pi_k}$ and $M$ were modified by the procedures $UpdatePacked(M, Q_{\pi_k}, \pi_k)$ and $SetPalletsDestinations(M, \pi_k)$, having their positions and destinations reassigned.

	Until line \ref{shims:endfirst}, a greedy and partial solution for pallet $i$ is constructed by edges inclusion, until a certain volume limit ($V_i \times limit$) is reached.
	The value of the {\it limit} (0.92) was empirically determined by the {\it iRace} tool developed by \cite{LopezIbanezManuel2016}.

	In the second phase (lines \ref{shims:beginsecond} to \ref{shims:endsecond}), a set of shims named $Set$\/ is created, where each {\it Shim} is formed by a set of edges in the range $[\eta_1^i,\eta_2^i]$, whose total volume is limited by $slack_i$. In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it First-Fit Decreasing}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \cite{JohnsonGarey1985}. Basically, shims are created by accumulating the following edges, taking $slack_i$\/ as a limit.

	In the third phase (lines \ref{shims:beginthird}-\ref{shims:endthird}), the best {\it Shim} in the $Set$\/ is chosen. Initially, two shims are found: $sh_w$\/ with larger weight and $sh_v$\/ with larger volume. Among the two, the {\it Shim} with the highest score will be chosen and its edges are inserted into $E_{N_{\pi_k}}$.
}

\begin{algorithm}[H]
	\caption{$Shims$\/ heuristic on node $\pi_k$}  \label{alg:shims}
	\begin{algorithmic}[1]

		\Procedure{$SolveNode$}{$Shims,\ \pi_k,\ G,\ time$}

		\State $T_{begin} \gets$ current system time

		\State $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$ \Comment{a graph with pallets positions ($M$), PC ($Q_{\pi_k}$), and edges ($E_{Q_{\pi_k}}$)}  \label{shims:initQ}

		\State Sort $M$ by $|D_i|$ in non-descending order \label{shims:pallets}

		\State $E_{N_{\pi_k}} \gets \{\ \}$  \Comment{an empty set of pallet-item edges}

		\State Let $E_{ij}$ be an array with $m \times n_{\pi_k}$ edges $(i,\ j)$ sorted by $\theta_{ij}^{\pi_k}$ in non-ascending order \label{shims:possible}

		\State $\tau^{\pi_k} \gets 0$	 \Comment{aircraft torque in node $\pi_k$}	
		\State $\tau_{max} \gets W_{max} \times limit^{CG}_{long}$ \Comment{maximum aircraft torque}

		\State $volume \gets \{0\ |\ [1,\ldots,m] \}$
		\State $slack  \gets \{0\ |\ [1,\ldots,m] \}$

		\State $limit \gets 0.92$ \Comment{0.92 was obtained by the {\it iRace} package \cite{LopezIbanezManuel2016}}

		\For{$i \in \{1,\ldots,m\}$}

			\For{$q \in \{1,\ldots,m_{\pi_k}\}$}

				\If{$(i, q) \in E_{Q_{\pi_k}}$} 
					\State $volume_i \gets volume_i + v_q$ 
				\EndIf		
			\EndFor

			\For {$e_{ij} \in E_{ij}$}

				\State $\Delta\tau \gets w_j \times D_i$

				\State $\tau_{new} \gets | \tau^{\pi_k} + \Delta\tau |$

				\If{ ($E_{N_{\pi_k}} \cup \{e_{ij}\}$ is feasible) {\bf and} ($volume_i \leq V_i \times limit$) {\bf and} ($\tau_{new} \leq \tau_{max}$ ) }

					\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup \{e_{ij}\}$

					\State $volume_i \gets volume_i + v_j$

					\State $\eta_1^{\ i} \gets \eta_1^{\ i} + 1$

					\State $\tau^{\pi_k} \gets \tau^{\pi_k} + \Delta\tau$

				\EndIf

			\EndFor			 \Comment{end of the first phase} \label{shims:endfirst}

			\State $slack_i \gets V_i - volume_i$  \label{shims:beginsecond}
			\State $\eta_2^{\ i} \gets \eta_1^{\ i}$

			\Repeat
				\State $\eta_2^{\ i} \gets \eta_2^{\ i} + 1$  \label{shims:eta2}

				\State $e_{ij} \gets E_{i,\eta_2^{\ i}}$

				\State $volume_i \gets volume_i + v_j$

			\Until{($\eta_2^{\ i} = n_{\pi_k}$) {\bf or} ($ volume_i \geq (1 + 3 \times (1-limit)) \times V_i)$)} \label{shims:endsecond} \Comment{end of the second phase}

			\State $volume \gets 0$	\label{shims:beginthird}

			\State $volume \gets 0$, $b \gets 1$ $shims_b \gets \{\}$, $Set \gets Set \cup \{shims_b\}$

			\State $shims_b \gets \{\}$  \label{empty_shims}

			\State $Set \gets Set \cup \{shims_b\}$ \label{empty_set}

			\For{$x \in \{\eta_1^{\ i},\ldots,\eta_2^{\ i} \}$} \label{edges_indexes}

			{\color{blue}

				\If{ current system time - $T_{begin} > time$}
				\State {\bf break}
				\EndIf

			}

				\State $NewShims \gets$ {\bf True} \label{new_shims}
				\State $e_{ij} \gets E_x^{\ i}$

				\For{$shims \in Set$} \label{shims_set}

					\If { $e_{ij} \not\in (E_{N_{\pi_k}} \cup shims)$ {\bf and} $e_{ij}$ is feasible  {\bf and}  $(v_j + volume) \leq slack_i$}

						\State $shims \gets shims \cup \{\ e_{ij}\ \}$
						\State $volume \gets volume + v_j$
						\State $NewShims \gets$ {\bf False} \label{new_shims_false}

						\State {\bf break}
					\EndIf

				\EndFor

				\If{$NewShims$} \label{new_shims2}
					\State $volume \gets 0$
					\State $b \gets b + 1$
					\State $shims_b \gets \{\ \}$
					\State $shims_b \gets shims_b \cup \{e_{ij}\}$
					\State $Set \gets Set \cup \{\ shims_b\ \}$
				\EndIf

			\EndFor

			\State $sh_w \gets shims$, where $shims \in Set$ and $\sum_{e_{ab} \in shims} w_b$ is maximum  \label{best_weight}		
			\State $sh_v \gets shims$, where $shims \in Set$ and $\sum_{e_{ab} \in shims} v_b$ is maximum \label{best_volume}
			\State $sh_{best} \gets shims$, where $shims \in \{sh_w, sh_v\}$ and $\sum_{e_{ab} \in x} s_b$ is maximum \label{best_score}

			\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup sh_{best}$ 	\label{shims:endthird}   \Comment{end of the third phase}

		\EndFor

		\Return $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$

		\EndProcedure
	\end{algorithmic}
\end{algorithm}

\section{Implementation and results}
\label{sec6}

This section is composed of two parts: the generation of test instances and the results obtained in our implementation.

\subsection{Instances generation}
\label{items}

As we are dealing with a new problem, which until now had not been modeled in the literature, we have to create our own benchmarks. For this, we based on the characteristics of real airlifts carried out by the {\it Brazilian Air Force}, as described below.

In the military airlift carried out in Brazil from 2008 to 2010, 23\% of the items weighed between $10kg$ and $20kg$, 22\% from $21kg$ to $40kg$, 24\% from $41kg$ to $80kg$, 23\% from $81kg$ to $200kg$, and 8\% between $201kg$ and $340kg$. These five groups of items are described in Table \ref{tab:weights}, where $P$\/ represents the group probability. On the other hand, the average density of these items is approximately $246 kg/m^3$.

\begin{table}[H]
	\centering
	\caption{Items weight distribution ($kg$) in Brazil}  \label{tab:weights}
	\begin{tabular}{c c c c }
		\toprule
		$Group$ & $P$ & $low$ & $high$  \\		
		\midrule
		1              & 0.23           & 10  & 20  \\
		2              & 0.22           & 21  & 40  \\
		3              & 0.24           & 41  & 80  \\		
		4              & 0.23           & 81  & 200 \\
		5              & 0.08           & 201 & 340 \\
		\bottomrule
	\end{tabular}
\end{table}

In the generation of test instances, we use three types of random selections:
\begin{itemize}
	\item $RandomReal(r_1,r_2)$: randomly selects a real number in $[r_1,r_2]$, where $r_1$ and $r_2$ are real numbers;
	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$, where $i_1$ and $i_2$ are integer numbers;
	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
\end{itemize}

\begin{algorithm}[H]
	\caption{Procedure to generate items}  \label{alg:itemsgen}
	\begin{algorithmic}[1]

		\Procedure{$ItemsGeneration$}{$scenario, surplus$}

		\State Let $L$ and $M$ be according to $scenario$ \label{ig:LM}
		\State $limit \gets surplus \times \sum_{i=1}^{m} V_i$ \label{ig:extended}

		\State Let $t_j^k$ be an item $j$ in node $k$.

		\For{$k \in \{0,\ldots,K\}$}

			\State $N_k \gets \varnothing$
			\State $j \gets 0$
			\State $vol \gets 0$	\label{ig:totals}	
			\While{$vol < limit$}

				\State $j \gets j+1$

				\Repeat
					\State $to_j \gets RandomInt(0, K)$ \label{ig:dest}
				\Until{$to_j \neq k$}

				\State $x = Roulette(item)$ biased through $P$ \label{ig:weight1} \Comment{From Table \ref{tab:weights}}

				\State $w_j \gets RandomReal(low(x), high(x))$        \label{ig:weight2}

				\State $s_j \gets \round{100 \times (1 - \log_{10}(RandomInt(1, 9)))} $ \label{ig:score}

				\State $v_j \gets w_j / RandomReal(148, 344)$ \label{ig:volume}

				\State $vol \gets vol + v_j$

				\State $N_k \gets N_k \cup \{t_j^k\}$ 
			\EndWhile

		\EndFor

		\State $N \gets \bigcup_{0 \leq k \leq K} N_k$

		\Return $N$

		\EndProcedure
	\end{algorithmic}
\end{algorithm}

$ItemsGeneration$, which generates $N$, is described in Algorithm \ref{alg:itemsgen}. The parameter $scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and the parameter $surplus$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}). To avoid simply loading all items, we use $surplus > 1$ {\color{blue}(1.2, 1.5, and 2.0). This represents more test instances for each scenario.}

For each generated $t^k_j$\/ item, its destination is randomly selected (line \ref{ig:dest}), its weight has a distribution according to Table \ref{tab:weights} (lines \ref{ig:weight1}-\ref{ig:weight2}), its score varies $100$\/ (highest) and $5$\/ (lowest) according to a logarithmic scale (line \ref{ig:score}, and its volume is randomly defined from the density, where we allow a variation of 40\% more or less than the average density of $246 kg/m^3$\/ (line \ref{ig:volume}).

{\color{blue}

\subsection{Results obtained}

In the tests performed, we used a 64-bit, 16GB, 3.6GHz, eight-core processor with {\it Linux Ubuntu 22.04.1 LTS 64-bit} as the operational system and {\it Python 3.10.4}\/ as the programming language. We also used the well-known solver {\it Gurobi}\/ (www.gurobi.com), version 9.5.2.

We ran Algorithm \ref{alg:main} in the 5 scenarios described in Table \ref{tab:scenarios}, considering 6 methods for node-by-node solution: Gurobi (see \ref{solver}), ACO, NMO, TS, GRASP, and Shims (Algorithm \ref{alg:shims}). In generating the items in each node, we consider 3 values for the parameter $surplus$. The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:20} ($surplus = 1.2$), \ref{tab:50} ($surplus = 1.5$) and \ref{tab:100} ($surplus = 2.0$). For each $method$, $scenario$\/ and $surplus$, 7 different instances were generated. Therefore, 105 tests (5 scenarios $\times$\/ 3 volumes $\times$\/ 7 instances) were performed for each method.

The average values were presented for $f_{\pi}$\/ and, for the run time, the worst result obtained. To facilitate the comparison between the methods, we added a last column in these tables, where two values are indicated:
\begin{itemize}
	\item {\bf Normalized}: value between 0 and 1, which corresponds to the ratio between the sum of $f$\/ values obtained by the method in all scenarios and the sum of the best values obtained among all methods in all scenarios. The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
	\item {\bf Speed-up}: ratio of the sums of the worst runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
\end{itemize}

In each $scenario$\/, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.

{\color{red}

As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's RAM (16 GiB) too early, we decided to execute a relaxed model with the decision variables set to CONTINUOUS (a Gurobi parameter) to obtain the upper bound for the objective function. This approach is much faster than using binary decision variables.

The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.

We solved all scenarios and instances with Gurobi in a relaxed linear model without a time limit. Although the solutions are not guaranteed to be feasible, it is good for the reader to have an idea of the possible upper bound for the metaheuristic solutions.

Note that, in Table \ref{tab:20} all $Gurobi\ f$\/ values are italicized to indicate that these values may not be optimal or feasible as the decision variables are not integer.

}

\begin{table}[H]
\centering
\caption{Methods and scenarios with $surplus =1.2$}  \label{tab:20}
\footnotesize
\begin{tabular}{c|c|ccccc|c}
\toprule
{$method$}      & {\bf Results}&{\bf 1}&{\bf 2} & {\bf 3}      & {\bf 4}      & {\bf 5}  & \specialcell{{\bf Normalized}\\{\bf Speed-up}}  \\
\midrule[.1pt]
\multirow{2}{*}{{\it Gurobi}}     & Upper bound for $f$      & {\it 7.91} & {\it 11.92} & {\it 13.12} & {\it 14.29} & {\it 13.34} & -  \\ %51,84
&{\bf run time (s)}                          & 14         & 19          & 73          & 188         & 1,097      & -  \\ 
\toprule
\multirow{2}{*}{\textbf{TS}}      & $f$      & 5.37       &             &             &             &            &   \\ % 5.49	 901
&{\bf run time (s)}                          & 659        &             &             &             &            &   \\ 
\midrule[.1pt]		
\multirow{2}{*}{\textbf{GRASP}}   & $f$      &            &             &             &             &            &   \\ 
&{\bf run time (s)}                          &            &             &             &             &            &   \\ 
\midrule[.1pt]
\multirow{2}{*}{\textbf{ACO}}     & $f$      & 5.25       & 5.75        & 5.06        & 4.84        &  4.20      &   \\ % 4.20	 1133
&{\bf run time (s)}                          & 313        & 326         & 303         & 256         &  1,133     &   \\ 
\midrule[.1pt]	
\multirow{2}{*}{\textbf{NMO}}     & $f$      &            &             &             &             &            &   \\
&{\bf run time (s)}                          &            &             &             &             &            &   \\ 
\midrule[.1pt]	
\multirow{2}{*}{\textbf{Shims}}   & $f$      & 7.91       & 11.55       & 12.62       &  13.82      &  14.85          &   \\ % 14.85	 563
&{\bf run time (s)}                          & 4          &  8          &  33         &  94         &  563          &   \\

\bottomrule
\end{tabular}
\normalsize
\end{table}

\begin{table}[H]
	\centering
	\caption{Methods and scenarios with $surplus =1.5$}  \label{tab:50}
	\footnotesize
	\begin{tabular}{c|c|ccccc|c}
		\toprule
		{$method$}                        & {\bf Results}   & {\bf 1}     &  {\bf 2}    & {\bf 3}      & {\bf 4}      & {\bf 5}    & \specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
		\toprule
		\multirow{2}{*}{\textbf{Gurobi}}  & $f$     & 4.756       & 8.84        & 11.94       & 16.21       &{\bf 16.47} & 0.852  \\	% 6828 
		&{\bf run time (s)}                         & 18        & 68         & 326        & 1,906      & 12,960    & 1.000  \\	 
		\midrule[.1pt]
		\multirow{2}{*}{\textbf{TS}}      & $f$     & 8.97       & 11.44       & 12.92       & 15.20       & 15.39      &  0.936 \\		
		&{\bf run time (s)}                         & 4         & 18         & 88         & 525        & 3,660     &  3.555  \\		
		\midrule[.1pt]		
		\multirow{2}{*}{\textbf{GRASP}}   & $f$     & 9.36       & 12.34       & 13.63       & 16.18       & 15.44      &  0.981 \\ 
		&{\bf run time (s)}                         & 4         & 17         & 86         & 515        & 3,564     &  3.651  \\
		\midrule[.1pt]	
		\multirow{2}{*}{\textbf{ACO}}     & $f$     & 9.36       & 12.34       & 13.63       & 16.18       & 15.44      &  0.981 \\
		&{\bf run time (s)}                         & 5         & 19         & 90         & 539        & 3,720     &  3.492  \\	
		\midrule[.1pt]	
		\multirow{2}{*}{\textbf{NMO}}     & $f$     & 9.36       & 12.33       & 13.60       & 16.16       & 15.43      &  0.979 \\ 
		&{\bf run time (s)}                         & 5         & 19         & 91         & 544        & 3,780     & 3.442  \\
		\midrule[.1pt]	
		\multirow{2}{*}{\textbf{Shims}}   & $f$     & {\bf 9.39} & {\bf 12.42} & {\bf 13.72} & {\bf 16.28} & 15.49      &  {\bf 0.986} \\	
		&{\bf run time (s)}                         & 3         & 11         & 48         & 265        & 1,720     &  {\bf 7.470}  \\
		\bottomrule
	\end{tabular}
	\normalsize
\end{table}

\begin{table}[H]
	\centering
	\caption{Methods and scenarios with $surplus = 2.0$}  \label{tab:100}
	\footnotesize
	\begin{tabular}{c|c|ccccc|c}
		\toprule
		{$method$}                        & {\bf Results}    & {\bf 1}     &  {\bf 2}   & {\bf 3}    & {\bf 4}    & {\bf 5}      & \specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
		\toprule
		\multirow{2}{*}{\textbf{Gurobi}}  & $f$              & 4.92      & 6.58      & 6.34       & 7.24       & 8.15      &  0.334 \\ % 9951
		                                  &{\bf run time (s)}& 29       & 106      & 493      & 2,833     & 19,380      &  1.000  \\ 
		\midrule[.1pt]
		\multirow{2}{*}{\textbf{TS}}      & $f$              & 13.72     & 17.53     & 19.28     & 21.78      & 22.35      &  0.951 \\		
		&{\bf run time (s)}                                  & 5        & 19       & 93       & 556       & 3,900     &  4.992  \\	
		\midrule[.1pt]		
		\multirow{2}{*}{\textbf{GRASP}}   & $f$        & 13.99     & 19.39     & 19.86     & 23.50      & 22.57      &  0.998 \\
		&{\bf run time (s)}                            & 6        & 21       & 97       & 571       & 3,960     &  4.906  \\	
		\midrule[.1pt]
		\multirow{2}{*}{\textbf{ACO}}     & $f$        & 13.99     & 19.39     & 19.86     & 23.50      & 22.57      &  0.998 \\
		&{\bf run time (s)}                            & 6        & 23       & 105      & 597       & 4,080     &  4.745  \\	
		\midrule[.1pt]	
		\multirow{2}{*}{\textbf{NMO}}     & $f$        & 13.99     & 19.39     & 19.85     & 23.50      & 22.56      & 0.998 \\	
		&{\bf run time (s)}                            & 9        & 27       & 120      & 681       & 4,620     & 4.184  \\	
		\midrule[.1pt]	
		\multirow{2}{*}{\textbf{Shims}}   & $f$        &{\bf 14.00} & {\bf 19.44} & {\bf 19.94} & {\bf 23.55} & {\bf 22.58} & {\bf 1.000} \\	
		&{\bf run time (s)}                            & 7        & 20       & 84       & 450       & 2,918     & {\bf 6.570}  \\
		\bottomrule
	\end{tabular}
	\normalsize
\end{table}

{\color{blue}
It is important to note that the best tour found in each test was seldom the tour with the lowest cost (the minimal cost tour without considering the CG deviation), but any of the first 25\% of the best tours. This indicates that our strategy of enumerating and solving all tours was assertive, although it is possible to devise a more proactive strategy to discard most of the higher-cost tours and improve performance. This must be one of the next steps in this research.

If a posterior application of this method requires 8, 9, or more nodes, it is possible to use this 25\% as a threshold to discard the higher-cost tours and keep the run time within an operational acceptable limit for the client.
}

In Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, $Gurobi$\/ was the slowest method and $Shims$\/ was the fastest. This was expected, as $Gurobi$\/ is a MIP solver, which seeks an optimal solution within the established constraints, usually reaching its time limit on each node, while {\it Shims} is a fast heuristic. This way, $Gurobi$\/ even demands prohibitive times (more than 5 hours in scenario 6 of Table \ref{tab:100}), while the maximum time required by $Shims$\/ did not reach one hour.

Considering the {\bf Normalized}\/ values, in Table \ref{tab:20} $Gurobi$\/ was the best ($0.997$), followed closely by $Shims$\/ ($0.939$). In Tables \ref{tab:50} and \ref{tab:100}, $Shims$ was the best ($0.986$\/ and $1.000$). As expected, the performance of $Gurobi$\/ drops sharply as the volume of shipments at the nodes increases: $0.852$\/ in Table \ref{tab:50} and $0.334$\/ in Table \ref{tab:100}.

It is worth noting that the other heuristics always obtained {\bf Normalized}\/ values greater than or equal to $0.9$, and were always faster than $Gurobi$.

It is not possible to state how far these results are from the optimal solutions, but we can report that all occupancy rates in terms of volume were above 97\%, and in terms of weight above 79\%, which denotes that the results achieved are good, considering that the CG deviation constraints keeps these values from going higher.

\subsubsection{Proposed strategy assessment}

Another important test is to compare the best tour score/cost ratio with the least expensive tour score/cost ratio. This is important to see how far our solution strategy is from the simplest form, just solving a TSP.

For this achievement, we ran {\it Shims} in 5 scenarios with 7 instances each, and 3 volume surpluses, collecting the averages of the best tour result and the least cost tour result.

\begin{table}[H]
\centering
\caption{Shims best and least cost tour results}  \label{tab:20least}
\footnotesize
\begin{tabular}{c|c|ccccc|cc}
\toprule
{$surplus$}              & {\bf Tour} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4}  & {\bf 5}   & {\bf Sum} & {\bf Improvement}\\

\midrule[.1pt]	
\multirow{2}{*}{$1.2$}   & Least cost    & 6.23  & 11.56   & 12.88   & 13.54    & 12.47  & 56.68 &       \\ 
& Best                                   & 7.96  & 13.04   & 14.30   & 15.56    & 15.28  & 66.14 & 16.7\%  \\ 
\midrule[.1pt]	
\multirow{2}{*}{$1.5$}   & Least cost    & 8.44  & 16.61   & 18.18   & 19.68    & 17.64  & 80.55 &   \\ % 17.64	 20.94	
& Best                                   & 11.47 & 18.25   & 20.61   & 22.22    & 20.94  & 93.49 & 16.1\%  \\ 
\midrule[.1pt]	
\multirow{2}{*}{$2.0$}   & Least cost    & 13.38 & 22.46   & 25.97   & 26.23    & 22.51  & 110.55 &   \\ %22.51	 29.09
& Best                                   & 17.93 & 25.59   & 28.61   & 31.28    & 29.09  & 132.50 & 19.9\%  \\

\midrule[.1pt]
 \multicolumn{8}{r}{{\bf Average improvement}} & 17.6\% \\
\bottomrule
\end{tabular}
\normalsize
\end{table}

As the selection of the best tour, considering what was collected and delivered in each node, and considering the cost increase due to CG deviation, was 17.6\% better than the option for a simple TSP solution (\textit{Least cost} results), this confirms that our solution strategy was adequate.

}

\section{Conclusions}
\label{sec7}

In this work, we modeled and solved a real air transport problem, named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem}\/ (ACLP+RPDP). For the first time in the literature, a NP-hard problem that involves {\it simultaneously}\/ pallet assembly, load balancing and route planning is addressed, where the cost-effectiveness of transport is maximized. We adopted some simplifications that are not critical, but that allowed the unprecedented solution of this problem considering more than two nodes.

We consider that, in practical cases, the number $K$\/ of nodes, excluding the base, is small ($K \leq 6$), each of them with hundreds of items to be shipped. Thus, considering real aircraft models, we developed some node-by-node heuristics, which allow us to enumerate all $K!$\/ tours and choose the best. The complete process can be executed quickly on a simple handheld computer, offering good results and reducing stress for the transport planners. As validation, we performed tests in several scenarios with real data from the {\it Brazilian Air Force}.

In less than one hour, this development can establish a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximizing the total score and minimizing fuel consumption for the airlift. The output, which includes the tour plan and the pallet building and arrangement plan, is an essential part of airlift: it improves flight safety, makes ground operations more efficient, and makes sure that each item gets to its right destination.

Our main contributions were the mathematical modeling of this complex problem, which involves four NP-hard sub-problems, and a complete process that offers a fast and good-quality solution.

{\color{blue}
As this is ongoing research, some improvements are possible, which are the next steps of the development:

(1) This work addresses all possible tours, but we plan bounding schemes to allow for discarding most of the higher-cost tours. This is expected to reduce the running time, favoring the discovery of better search mechanisms.

(2) We are also working on computer parallelism to accelerate the algorithms.

(3) In the event that these gains in performance are achieved, we strongly believe that they will permit the application of a 3-dimensional packing procedure to maximize the occupancy rate among the selected items.
}

The simple heuristic $Shims$\/ presented a solution with the best node-by-node results in a restricted time interval. Without it, it would not be possible to test all routes: after all, the greater the number of nodes, the shorter the run time limit for node-by-node solution. We also show that, as the shipment volume increases, the exact node-by-node solution becomes unfeasible, indicating the importance of a fast and good-quality heuristic.

Finally, by focusing on node-by-node solution, the method of this work is not exclusive to aircraft and airports: it can be adapted, for example, to ships and ports, or vehicles and warehouses, or freight cars and railways. In these cases, it would be necessary to make some changes in the modeling: for example, modify the load balancing constraints, and consider the available space in vehicles or freight cars instead of pallets.

\section*{Acknowledgments}

This research was partially supported by \textit{São Paulo Research Foundation} (FAPESP, grant 2016/01860-1).

\begin{thebibliography}{00}

\bibitem{LarsenMikkelsen1979} O. Larsen and G. Mikkelsen, An interactive system for the loading of cargo aircraft, {\it European Journal of Operational Research}, Vol. 4 (6), pp. 367-373, 1980.

\bibitem{Brosh1981} Israel Brosh, Optimal cargo allocation on board a plane: a sequential linear programming approach, {\it European Journal of Operational Research}, Vol. 8 (1), pp. 40-46, 1981.

\bibitem{JohnsonGarey1985} D.S. Johnson and M.R. Garey, A 71/60 theorem for bin packing, {\it Journal of Complexity}, Vol. 1 (1), pp. 65-106, 1985.

\bibitem{Glover1986} Fred Glover, Future paths for integer programming and links to artificial intelligence, {\it Computers and Operations Research}, Vol. 13, pp. 533-549, 1986.

\bibitem{FeoResende1989} Thomas A. Feo and Mauricio G.C. Resende, A probabilistic heuristic for a computationally difficult set covering problem, {\it Operations Research Letters}, Vol. 8 (2), pp. 67-71, 1989.

\bibitem{Kevin1992} Kevin Y.K. Ng, A multicriteria optimization approach to aircraft loading, {\it Operations Research}, Vol. 40 (6), pp. 1200-1205, 1992.

\bibitem{Dorigo1992} Marco Dorigo, {\it Optimization, Learning and Natural Algorithms}, PhD Thesis, Politecnico di Milano, 1992.

\bibitem{CharonHudry1993} I. Charon and O. Hudry, The noising method: a new method for combinatorial optimization, {\it Operations Research Letters}, Vol. 14 (3), pp. 133-137, 1993.

\bibitem{DorigoManiezzoColorni1996} M. Dorigo, V. Maniezzo and A. Colorni, The ant system: optimization by a colony of cooperating agents, {\it IEEE Transactions on Systems, Man, and Cybernetics}, Vol. 26 (1), pp. 29-41, 1996.

\bibitem{NiarFreville1997} S. Niar and A. Freville, A parallel tabu search algorithm for the 0-1 multidimensional knapsack problem, Proceedings of {\it 11th International Parallel Processing Symposium}, 1997.

\bibitem{Heidelberg1998} K.R. Heidelberg, G.S. Parnell and J.E. Ames, Automated air load planning, {\it Naval Research Logistics}, Vol. 45 (8), pp. 751-768, 1998.

\bibitem{CharonHudry2001} I. Charon and O. Hudry, The noising methods: a generalization of some metaheuristics, {\it European Journal of Operational Research}, Vol. 135 (1), pp. 86-101, 2001.

\bibitem{MongeauBes2003} M. Mongeau and C. Bes, Optimization of aircraft container loading, {\it IEEE Transaction on Aerospace and Electronic Systems}, Vol. 39 (1), pp. 140-150, 2003.

\bibitem{fok2004optimizing} M.K.K. Fok and A.H.W. Chun, Optimizing air cargo load planning and analysis, Proceedings of {\it The International Conference on Computing, Communications and Control Technologies}, 2004.

\bibitem{Fidanova2006} S. Fidanova, Ant colony optimization and multiple knapsack problem and model bias, {\it Numerical Analysis and Its Applications}, Springer Berlin Heidelberg, pp. 280-287, 2005.

\bibitem{Chan2006} F.T.S. Chan, R. Bhagwat, N. Kumar. M.K. Tiwari and P. Lam, Development of a decision support system for air-cargo pallets loading problem: a case study, {\it Expert Systems with Applications}, Vol. 31 (3), pp. 472-485, 2006.

\bibitem{KaluznyBohdanL2009Oalb} B.L. Kaluzny and R.H.A.D. Shaw, Optimal aircraft load balancing, {\it International Transactions in Operational Research}, Vol. 16 (6), pp. 767-787, 2009.

\bibitem{MesquitaCunha2011} A.C.P. Mesquita and C.B. Cunha, An integrated heuristic based on the Scatter Search metaheuristic for vehicle routing problems with simultaneous delivery and pickup in the context of the Brazilian Air Force, {\it Transportes}, Vol. 19, pp. 33-42, 2011.

\bibitem{Verstichel2011} J. Verstichel, W. Vancroonenburg, W. Souffriau and G.V. Berghe, A mixed integer programming approach to the aircraft weight and balance problem, {\it Procedia Social and Behavioral Sciences}, Vol. 20, pp. 1051-1059, 2011.

\bibitem{Limbourg2012} S. Limbourg, M. Schyns and G. Laporte, Automatic aircraft cargo load planning, {\it Journal of the Operational Research Society}, Vol. 63 (9), pp. 1271-1283, 2012.

\bibitem{RoesenerHall2014} A.G. Roesener and S. Hall, A nonlinear integer programming formulation for the airlift loading problem with insufficient aircraft, {\it Journal of Nonlinear Analysis and Optimization: Theory and Applications}, Vol. 5 (1), pp. 125-141, 2014.

\bibitem{Vancroonenburg2014} W. Vancroonenburg, J. Verstichel, K. Tavernier and G.V. Berghe, Automatic air cargo selection and weight balancing: a mixed integer programming approach, {\it Transportation Research Part E: Logistics and Transportation Review}, Vol. 65, pp. 70-83, 2014.

\bibitem{LurkinSchyns2015} V. Lurkin and M. Schyns, The airline container loading problem with pickup and delivery, {\it European Journal of Operational Research}, Vol. 244 (3), pp. 955-965, 2015.

\bibitem{RoesenerBarnes2016} A.G. Roesener and J.W. Barnes, An advanced tabu search approach to the dynamic airlift loading problem, {\it Logistics Research}, Vol. 9 (1), pp. 1-18, 2016.

\bibitem{PaquaySchynsLimbourg2016} C. Paquay, M. Schyns and S. Limbourg, A mixed integer programming formulation for the three-dimensional bin packing problem deriving from an air cargo application, {\it International Transactions in Operational Research}, Vol. 23 (1-2), pp. 187-213, 2016.

\bibitem{PaquayLimbourgSchynsOliveira2018} C. Paquay, S. Limbourg, M. Schyns and J.F. Oliveira, MIP-based constructive heuristics for the three-dimensional bin packing problem with transportation constraints, {\it International Journal of Production Research},
Vol. 56 (4), pp. 1581-1592, 2018.

\bibitem{Laabadi2018} S. Laabadi, M. Naimi, H. El Amri and B. Achchab, The 0/1 multidimensional knapsack problem and its variants: a survey of practical models and heuristic approaches, {\it American Journal of Operations Research}, Vol. 8, pp. 395-439, 2018.

\bibitem{YangLiuGao2018} Y. Chenguang, L. Hu and G. Yuan, Load planning of transport aircraft based on hybrid genetic algorithm, {\it MATEC Web of Conferences}, Vol. 179, pp. 1-6, 2018.

\bibitem{Alonso2019} M.T. Alonso, R. Alvarez-Valdes and F. Parreno, A GRASP algorithm for multi-container loading problems with practical constraints, {\it A Quarterly Journal of Operations Research}, Vol. 18, pp. 49-72, 2019.

\bibitem{BrandtStefan2019} F. Brandt and S. Nickel, The air cargo load planning problem - a consolidated problem definition and literature review on related problems, {\it European Journal of Operational Research}, Vol. 275 (2), pp. 399-410, 2019.

\bibitem{wong2020} E.Y.C. Wong and K.K.T. Ling, A mixed integer programming approach to air cargo load planning with multiple aircraft configurations and dangerous goods, Proceedings of {\it 7th International Conference on Frontiers of Industrial Engineering}, 2020.

\bibitem{Zhan2020} S. Zhan, L. Wang, Z. Zhang and Y. Zhong, Noising methods with hybrid greedy repair operator for 0-1 knapsack problem, {\it Memetic Computing}, Vol. 12, pp. 37-50, 2020.

\bibitem{eugene2021} E.Y.C. Wong, D.Y. Mo and S. So, Closed-loop digital twin system for air cargo load planning operations, {\it International Journal of Computer Integrated Manufacturing}, Vol. 34 (7-8), pp. 801-813, 2021.

\bibitem{zhao2021} X. Zhao, Y. Yuan, Y. Dong and R. Zhao, Optimization approach to the aircraft weight and balance problem with the centre of gravity envelope constraints, {\it IET Intelligent Transport Systems}, Vol. 15 (10), pp. 1269-1286, 2021.

\bibitem{LopezIbanezManuel2016} López-Ibáñez, Manuel and Dubois-Lacoste, Jérémie and Pérez Cáceres, Leslie and Birattari, Mauro and Stützle, Thomas. 2016. The irace package: Iterated racing for automatic algorithm configuration. Elsevier. pp. 43-58. Volume 3.

\end{thebibliography}

\end{document}

{\color{blue}
	Until scenario 4, solving all tours and selecting the best is fast enough for operational use. For scenarios 4 and 5 (which are not typical in the Brazilian Air Force), we plan bounding schemes to allow for the discarding of most of the higher-cost tours. This is expected to reduce the running time for scenarios 4 and 5, favoring the discovery of better search mechanisms as well as further applications that may require more nodes per tour.	
}

{\color{blue}

	Equations \ref{eq18} and \ref{eq19} were set to group items with the same destinations. But, as they mean a combination of $n_{\pi_k}$ items 2 by 2, this may easily lead to huge numbers, $60,000$ to $320,000$ possible combinations, due to the number of items in each node, which together with the other constraints make a MIP solution very hard.

	To work around this, we decided to create the destination attribute for the pallets ($TO_i^{\pi_k}$), to reduce the number of combinations and enable the solution using heuristics. $TO^{\pi_k}_i$ denotes that pallet $i$ may assume a different destination in each node ${\pi_k}$.

	On subsequent nodes, the PC can be allocated with other items of same destination \ref{eq:cons5}.

	\begin{equation} \label{eq:cons5}
		TO_i^{\pi_{k+1}} = to_j;\ X^{\pi_k}_{ij} = 1  \ \mbox{and}\ to_j \in L_{\pi_{k+1}};  \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K-1\}
	\end{equation}

	\vspace{3mm}
	At each node, an item (\ref{eq:pdp8}) and the already packed contents (\ref{eq:pdp2}) must only be allocated to a pallet if the destinations are the same.

	\begin{equation} \label{eq:pdp8}
		TO_i^{\pi_k} = to_j;\ X_{ij}^{\pi_k} = 1; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
	\end{equation}

	\begin{equation} \label{eq:pdp2}
		TO_i^{\pi_k} = to_q;\ Y_{iq}^{\pi_k} = 1; \ i \in \{1, \ldots, m\};\ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
	\end{equation}

}

{\color{blue}Another possibility to make this approach more general is to devise some bounding schemes to allow discarding most higher-cost tours and allowing an increase in the number of nodes without compromising performance.}

### 2023-06-03 CAIE/paper_old.tex
CAIE paper transferred to parallel folder

--- 2023-06-03 CAIE/paper.tex
@@
 In aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading, for fast take-off, and mission accomplishment, especially in times of crisis, disaster relief missions, short deadlines for contracting customers, or any external pressure for immediate take-off.
-In addition, there are no commercially available systems that can assist the load and trip planners with the pallet building and aircraft loading and balancing demands for transport at each hub.
-This enables other risks, such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
-We defined and solved the problem of planning the loading and routing of a single aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
+{\color{blue} In addition, there are no commercially available systems that can assist the load and trip planners with the pallet building and aircraft loading and balancing demands for transport at each hub.} This enables other risks, such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
+We defined and solved the problem of planning the loading and routing of an aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
 This NP-hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modeled using standardized pallets in fixed positions, obeying center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
@@
 Considering air cargo transport, Table \ref{tab:sa} lists the main works in the literature and the corresponding sub-problems addressed.
-We also indicate whether the dimensions of the items were taken into account ({\bf 3D} or {\bf 2D}) and which solution method was used: heuristic search methods ({\bf Heu}), integer programming ({\bf Int}), or linear programming ({\bf Lin}).
+We also indicate whether the dimensions of the items were taken into account ({\bf 3D} or {\bf 2D}) and which solution method {\color{blue}was used: heuristic search methods ({\bf Heu}), integer programming ({\bf Int}), or linear programming ({\bf Lin}).}
 
@@
 		Limbourg {\it et al.} \cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		Roesener and Hall \cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
+		{\color{blue}Roesener and Hall} \cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
 		Vancroonenburg {\it et al.} \cite{Vancroonenburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
@@
 
-As can be seen, so far Lurkin and Schyns \cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem.
-Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a predefined tour plan restricted to two legs.
+As can be seen, so far Virginie Lurkin and Michaël Schyns \cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem.
+Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a pre-defined tour plan restricted to two legs.
 It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem.
@@
 Due to its dimensions, this country has the largest air market on the continent with $2,499$\/ registered airports, of which $1,911$\/ are private and $588$\/ are public.
-Although it is an immense distribution network, the {\it Brazilian Air Force}\/ missions have always considered 3 to 5 nodes per flight plan.
-
-It is important to emphasize that this data is not an imposed limitation, but a historical fact that we will explore in our solution method.
-Throughout this work we approach routes with up to 7 nodes, as can be seen in Table \ref{tab:costs} and in Figure \ref{fig:nodes}.
-Although there are many other airports of interest to {\it Brazilian Air Force}, these 7 nodes were chosen due to their high demand and short transport times.
-The other Brazilian airports tend to have a lower demand, which is usually met, in a less expensive way, by cabotage, rail or road transport.
+Although it is an immense distribution network, airlift missions consider 3 to 5 nodes per tour plan.
+Throughout this work, we address routes with up to 7 nodes, as can be seen in Table \ref{tab:costs} and Figure \ref{fig:nodes}.
+
+{\color{blue} 
+	This 7-node limitation is more than enough for the case of the {\it Brazilian Air Force}.
+In case some extension is needed for this range, a more suitable TSP solution method may be adopted, as full enumeration may make it impractical for other clients' real-world scenarios.
+}
 
@@
 		\includegraphics[scale=0.25]{Images/nodes.png}
-		\captionof{figure}{A route with 7 Brazilian airports}
+		\captionof{figure}{A route with some Brazilian airports}
 		\label{fig:nodes}		
@@
 
-Cargo handling at modern airports usually involves powerful equipment such as dollies, motorized conveyors, elevator transfer vehicles, caster decking, scissor lifts, electrical controls, hydraulically adjustable heights, forklift trucks, and plenty of room in the airport cargo terminal (see Figure \ref{fig:handling}).
-These facilities, together with the use of standardized pallets with predefined positions on the aircraft, allow loading and unloading at each node to be carried out in about 30 minutes.
-
-\begin{figure}[H]
+{\color{blue}
+
+	Although there are many other airports in the Brazilian Air Force's interest, these seven were chosen due to their huge demands and short lead times for transportation, which are worthy of aerial transportation.
+The other airports may have a lower transportation demand, which is frequently less costly to address by cabotage, railways, or road transport as well.
+}
+
+In the {\it Brazilian Air Force}\/ missions, hundreds of items can be carried at each node, where the objectives are to prioritize the transport of the most important items and minimize the cost of fuel along the route.
+As standardized pallets are used with predefined positions on the aircraft, it is possible to carry out loading and unloading at each node in around 30 minutes{\color{blue}, if the pallet's position and contents are already planned}.
+
+{\color{blue}
+As to the cargo item arrangement on the pallets as well as the route planning, which normally takes two hours, which is the same time reported by Kelly Fok and Andy Chun \cite{fok2004optimizing} "the container load planning is usually done in roughly 2 hours before departure, when all cargo details are in place.", it is essential to find a way to shorten this time to improve the efficiency of cargo transportation.
+However, there are no commercially available systems or software packages that can simultaneously solve these problems.
+}
+
+\begin{table}[H]
+
+	\begin{minipage}{0.05\linewidth}
+
+	\end{minipage}\hfill 
+	\begin{minipage}{0.40\linewidth}
+
+	{\color{blue}	
+Cargo handling at modern airports usually involves powerful equipment like dollies, powered conveyors, elevating transfer vehicles, caster decking, scissor lifts, electrical controls, hydraulically adjustable heights, forklift trucks, and plenty of room in the airport cargo terminal.
+
+\vspace{3mm}
+These facilities can help with a complete pallet re-arrangement in the cargo bay, if necessary, in less than 30 minutes if the pallet position contents are already planned.
+}
+
+	\end{minipage}\hfill 
+	\begin{minipage}{0.55\linewidth}
 		\centering
@@
 		\label{fig:handling}
+
 		\small\textsuperscript{Source: NARA \& DVIDS Public Domain Archive}	
-\end{figure}
-
-However, as reported by Fok and Chun, {\it load planning (...) is usually done in roughly 2 hours before departure, when all the details of the cargo are present.} \cite{fok2004optimizing}.
-We may infer that this is because there is no commercial software to solve this problem.
-In addition, there is still the necessary time to plan the route, in order to minimize fuel consumption and ensure the transport of priority items.
-In these circumstances, it is of great importance to find a method that speeds up the load planning and the definition of the flight itinerary.
-
-Our work proposes a method that prioritizes the transport of the most relevant items at each node and the fuel economy along the tour.
-We developed a heuristic that can be run on a simple handheld computer (such as a laptop or a tablet) that previously defines a flight plan, and provides a quick solution for cargo handling.
-Consequently, this method reduces the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets (regarding their positions), and a pick up and delivery plan to each node.
+	\end{minipage}
+\end{table}
+
+This work proposes a method that {\color{blue}addresses} these objectives: a pallet building and arrangement plan with a routing option that maximizes the benefit-cost ratio for the smooth execution of pickup and delivery transport missions.
+
+We develop a heuristic that can be executed on a simple handheld computer (like a laptop or a tablet) and that provides a solution quickly enough to keep this cargo handling time {\color{blue}under 30 minutes.} This heuristic {\color{blue} solution certainly will} reduce the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets ({\color{blue}regarding their positions), and a pick up and delivery plan to each node.}
 
@@
 
-Larsen and Mikkelsen \cite{LarsenMikkelsen1979} developed an interactive procedure for loading 14 types of Boeing 747 into a two-leg transportation plan.
+Ole Larsen and Gert Mikkelsen \cite{LarsenMikkelsen1979} developed an interactive procedure for loading 14 types of Boeing 747 into a two-leg transportation plan.
 Seven types of items were considered to be allocated in 17 to 42 positions.
@@
 
-Brosh \cite{Brosh1981} addressed the problem of planning the allocation of cargo on an aircraft.
+Israel Brosh \cite{Brosh1981} addressed the problem of planning the allocation of cargo on an aircraft.
 Considering volume, weight and structural constraints, the author finds the optimal load layout through a fractional programming problem.
 
-Ng \cite{Kevin1992} developed a multi-criteria optimization approach to load the {\it C-130} aircraft of the {\it Canadian Air Force}.
+Kevin Ng \cite{Kevin1992} developed a multi-criteria optimization approach to load the {\it C-130} aircraft of the {\it Canadian Air Force}.
 Based on integer programming, this model provides timely planning and improves airlift support for combat operations, solving WBP with pallets in fixed positions, and considering 20 different items.
@@
 
-Mongeau and Bes \cite{MongeauBes2003} presented a method based on linear integer programming to solve the problem of choosing and positioning containers in the {\it Airbus 340-300}.
+Marcel Mongeau and Christian Bes \cite{MongeauBes2003} presented a method based on linear integer programming to solve the problem of choosing and positioning containers in the {\it Airbus 340-300}.
 Safety and stability constraints were considered, with the objective of minimizing fuel consumption.
 
-Fok and Chun \cite{fok2004optimizing} developed a web-based application to make efficient use of space and load balancing for an air cargo company.
+Kelly Fok and Andy Chun \cite{fok2004optimizing} developed a web-based application to make efficient use of space and load balancing for an air cargo company.
 Based on an analysis of historical data, an operational load planning with mathematical optimization is obtained.
@@
 
-Kaluzny and Shaw \cite{KaluznyBohdanL2009Oalb} developed a mixed integer linear programming model to arrange a set of items in a military context that optimizes the load balance.
-The objective function can be chosen to minimize the deviation of the center of gravity (CG) from the target position or to maximize the function of the items loaded.
+Bohdan Kaluzny and David Shaw \cite{KaluznyBohdanL2009Oalb} developed a mixed integer linear programming model to arrange a set of items in a military context that optimizes the load balance.
+{\color{blue}The objective function can be chosen to minimize the deviation of the center of gravity from the target position or to maximize the function of the items loaded.
 This approach does not palletize items, which are arranged in the cargo bay by a 2D packing procedure.
+}
 
@@
 Limbourg {\it et al.} \cite{Limbourg2012} developed a mixed-integer program for optimally rearranging a set of pallets into a compartmentalized cargo aircraft, specifically the {\it Boeing 747}.
-
-Roesener and Hall \cite{RoesenerHall2014} solved APP and WBP as an integer programming problem, where the items are selected for shipment based on a utility score, then assigned to pallets, which will be loaded into an aircraft in a specific pallet position.
+{\color{blue}These authors approach regarding CG deviation is different from ours because we simply position the CG within its permitted range with the sole objective of maximizing the transportation value.
+As the longitudinal stability is handled by fly-by-wire-equipped aircraft with automatic longitudinal trimming, it is expected to increase our proposed method's ability to offer a solution to carry additional freight.
+}
+
+August Roesener and Shane Hall \cite{RoesenerHall2014} solved APP and WBP as an integer programming problem {\color{blue} where the items are selected for shipment based on a utility score, then assigned to pallets, which will be loaded into an aircraft in a specific pallet position.
 The pallets are then packed in a manner to optimize both the pallet and aircraft characteristics, such as item utility, pallet occupancy, and aircraft center of balance.
-This method groups items by destinations to be served by multiple aircraft.
+This paper groups items by destinations to be served by multiple aircraft with multiple destinations.
+
+Differently from these authors' work, we plan to meet all transportation demands in a set of nodes with one aircraft in a simultaneous delivery and pickup tour.
+}
 
@@
 
-As already mentioned, Lurkin and Schyns \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and PDP in air cargo transport.
+As already mentioned, Virginie Lurkin and Michaël Schyns \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and PDP in air cargo transport.
 The authors demonstrated that this problem is NP-hard and performed some experiments with real data, noting that their model offers better results than those obtained manually.
@@
 Paquay {\it et al.} \cite{PaquayLimbourgSchynsOliveira2018} developed some heuristics to solve this problem.
+{\color{blue} We also plan to apply these authors' technique to the ACLP part of our problem if we manage to keep the running time within one hour.}
 
 Chenguang {\it et al.} \cite{YangLiuGao2018} modeled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing to minimize fuel consumption.
+{\color{blue}We also plan to select the best tour that minimizes fuel consumption through cargo balancing, as it is known (see Mongeau and Bes \cite{MongeauBes2003} for further details) that cargo unbalancing may lead to extra fuel consumption.}
 
@@
 Zhao {\it et al.} \cite{zhao2021} proposed a new modeling for WBP based on mixed integer programming.
-Instead of focusing on the CG deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
-
-As can be seen, none of these works address air cargo palletization and load balancing with route optimization in a multi-leg transportation plan for a single aircraft.
+Instead of focusing on the center of gravity (CG) deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
+
+As can be seen, none of these works address air cargo palletization and load balancing with route optimization in a multi-leg transportation plan.
 This is the objective of our work: to model and elaborate heuristics for a real problem that simultaneously involves 4 intractable sub-problems: APP, WBP, PDP and TSP.
@@
 
-	\item At each node of the tour, the items to be allocated are characterized by weight, volume, scores, and previously known destinations.
-We leave the consideration of 2D or 3D items to a future work.
+	\item At each node of the tour, the items to be allocated are characterized by weight, volume {\color{blue}(the product of their dimensions, as most of them have a parallelepiped shape)}, scores, and previously known destinations.
+{\color{blue}But, as the dimensions are unknown, we leave the consideration of 3D packing to the next step of this work if we manage to keep the worst running time within one hour.}
 
@@
 It is the primary air cargo pallet for more than 70 Air Forces and many air transport companies.
-This pallet has a capacity of $4,500 kg$ and $13.7 m^3$, which may be limited by its position along the cargo bay.
-It is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$.
+This pallet has a capacity of $4,500 kg$ and $13.7 m^3$ {\color{blue}(which may be limited by its position along the cargo bay)}, is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$.
 For more information, see {\tt www.463LPallet.com}.
@@
 A pallet which has not yet reached its destination may receive more items, although it is known that these operations of removing restraining nets increase handling time and the risk of improper delivery.
-We do not consider oversized cargo in this work, but only cargo items that fit on these pallets.
-
-	\item Finally, as we are interested in minimizing fuel costs, we disregarded others costs not directly associated with aircraft flight, such as handling.
+We do not consider oversized cargo in this work, but only {\color{blue}cargo items} that fit on these pallets.
+
+	\item Finally, as we are interested in minimizing fuel costs, {\color{blue}we disregarded some fuel costs not directly associated with aircraft flight, such as handling.}
 
@@
 
-Throughout this text, we call {\it Packed Contents}\/ (PC) a set of items of same destination stacked on a pallet and covered with a restraining net (see Figure \ref{fig:larger2}).
+Throughout this text, we call a {\color{blue} {\it Packed Contents (PC)} } a set of items of the same destination stacked on a pallet and covered with a restraining net.
 It is considered unique, having the same attributes of its components, whose values are the sum of individual scores, weights and volumes.
+See Figure \ref{fig:larger2}.
 To ensure accuracy in pickup and delivery operations, a PC must remain on board until its destination.
-
-\subsection{Aircraft parameters and load balancing}
-
-We consider real scenarios with an aircraft with payload of $75,000 kg$, with enough fuel for a range of $4,400km$.
-The aircraft layout is presented in Figure \ref{fig:larger}, where the pallets positions are identified by $p_i$, $1\leq i \leq 18$.
 
@@
 	\includegraphics[scale=0.4]{Images/consolidated.jpg}
-	\caption{A PC on a 463L pallet inside a {\it Boeing C-17}}
+	\caption{Packed Contents on a 463L pallet inside a {\it Boeing C-17}}
 	\small\textsuperscript{Source: From Wikimedia Commons, the free media repository}
@@
 \end{figure}
+
+\subsection{Aircraft and load balancing}
+
+We consider real scenarios with an aircraft with payload of 75,000 kg (with enough fuel for a range of 4,400km).
+The aircraft layout is presented in Figure \ref{fig:larger}, where the pallets positions are identified by $p_i$.
 
@@
 The torque applied to the aircraft must keep its CG in the operational range, which corresponds to a fixed percentage of the {\it Mean Aerodynamic Chord} \footnote{Chord is the distance between the leading and trailing edges of the wing, measured parallel to the normal airflow over the wing.
-The average length of the chord is known as the {\it Mean Aerodynamic Chord} (MAC).} which is considered $1.17m$ for the aircraft of this work (see Figure \ref{fig:lateral}).
+The average length of the chord is known as the {\it Mean Aerodynamic Chord} (MAC).} which is considered $1.17m$ for the aircraft of this work.
+See Figure \ref{fig:lateral}.
 
@@
 	\includegraphics[scale=0.22]{Images/lateral.png}
-	\caption{Aircraft longitudinal cut, where red lines are pallets positions}
+	\caption{Aircraft longitudinal cut showing red lines as pallets positions}
 	\label{fig:lateral}
@@
 
-Table \ref{tab:larger} shows the parameters of this aircraft.
-$p_i$\/ are pallets, $1\leq i \leq 18$, with weight $W_i$\/ and $V_i$\/ volume limits, respectively.
-$D_i^{long}$\/ and $D_i^{lat}$\/ are, respectively, the longitudinal and latitudinal distances of each pallet centroids to the CG of aircraft along both axes.
-These distances will be used in the calculation of the torque referring to the items allocated on each pallet.
-In this aircraft, as the ramp have an inclination of $25°$, we made the necessary corrections in $D_i^{long}$, $W_i$\/ and $V_i$\/ of the corresponding pallets.
+Table \ref{tab:larger} shows the parameters. {\it $D_i^{long}$} and {\it $D_i^{lat}$} refer to the relative distances of pallet centroids (in meters) in relation to the CG of aircraft along both axes.
+
+{\color{blue}
+
+%Note that since the smaller aircraft has a single row of pallets positions, the y-axis passes through the geometric center of each pallet position.
+Since the value is 0 for each position, $D_i^{lat}$ is not provided for the smaller aircraft.
+
+Throughout this work, when we refer to the word "torque", we mean item or PC weight x pallet centroid distance to the CG ($D_i^{long}$ or $D_i^{lat}$), considering the angle between these dimensions to always be 90°.
+In both aircraft, as the ramps have an inclination of 25°, we made the necessary corrections in $D_i^{long}$, {\it Weight} and {\it Volume limits} of the corresponding pallets positions (see Table \ref{tab:larger}).
+
+}
 
@@
 		\toprule
-		& \multicolumn{3}{c}{$Payload$: 75,000kg} & \multicolumn{3}{c}{$limit^{CG}_{long}$: $1.170m$} &
+		\textbf{Limits}& \multicolumn{3}{c}{$Payload$: 75,000kg} & \multicolumn{3}{c}{$limit^{CG}_{long}$: $1.170m$} &
 		\multicolumn{3}{c}{$limit^{CG}_{lat}$: $0.19m$} \\
 		\midrule
-		\multirow{2}{*}{{\boldm{$p_i$}}}  & $p_{17}$ & $p_{15}$ & $p_{13}$ & $p_{11}$ & $p_{9}$ & $p_{7}$ & $p_{5}$ & $p_{3}$ & $p_{1}$ \\
+		\multirow{2}{*}{{\bf Pallets} ($p_i$)}  & $p_{17}$ & $p_{15}$ & $p_{13}$ & $p_{11}$ & $p_{9}$ & $p_{7}$ & $p_{5}$ & $p_{3}$ & $p_{1}$ \\
 		& $p_{18}$ & $p_{16}$ & $p_{14}$ & $p_{12}$ & $p_{10}$ & $p_{8}$ & $p_{6}$ & $p_{4}$ & $p_{2}$ \\
@@
 		\midrule
-		{\boldm{$W_i$}} ($kg$)      &   4,500   &    4,500  &   4,500   &  4,500    & 4,500     & 4,500     & 4,500     & 3,000    & 3,000   \\
-		{\boldm{$V_i$}} ($m^3$)   &   14.8   &   14.8   &  14.8    &  14.8    & 14.8     & 14.8     & 14.8     & 10.0    & 7.0 \\	
+		{\bf Weight limits} ($W_i,\ kg$)      &   4,500   &    4,500  &   4,500   &  4,500    & 4,500     & 4,500     & 4,500     & 4,000    & 3,000   \\
+		{\bf Volume limits} ($m^3$)   &   14.8   &   14.8   &  14.8    &  14.8    & 14.8     & 14.8     & 14.8     & 10.0    & 7.0 \\	
 		\midrule
 
-		\textbf{Fuel cost}  & \multicolumn{9}{c}{ $c_d$ = US\$ $4.90/km$ }	 \\
+		\textbf{Fuel cost rate}  & \multicolumn{9}{c}{ $c_d$ = US\$ $4.900/km$ }	 \\
 
@@
 
-		\textbf{Fuel consumption rate}  & \multicolumn{9}{c}{ $c_g = 5\%$} \\
+		\textbf{Fuel consumption limit rate}  & \multicolumn{9}{c}{ $c_g$ = $5\%$} \\
 
@@
 
-		\textbf{Maximum weight}  & \multicolumn{9}{c}{ $W_{max} = \sum_i W_i = 75,000kg$} \\
+		\textbf{Maximum weight capacity}  & \multicolumn{9}{c}{ $W_{max}$ = $\sum_{i=1}^m W_i$} \\
 
@@
 
-This aircraft spends $c_d$\/ dollars per kilometer flown, and can carry up to $W_{max}$\/ of cargo distributed on the pallets.
-The {\it fuel consumption rate}\/ $c_g$\/ is the percentage limit of cost increase  due to the CG deviation on the longitudinal axis of the aircraft.
-$c_g$\/ depends on the characteristics of the aircraft, and is an arbitrated value for this work that should be better evaluated in future research.
-It is important to consider that the $c_g$\/ tends to zero as the aircraft attitude tends to level.
-As the CG deviation varies from $0$\/ to $limit^{CG}_{long}$, the {\it fuel cost}\/ increase varies from $0$\/ to $c_g$.
+{\color{blue}
+
+The {\it Fuel consumption limit rate} in Table \ref{tab:larger} is the percentage limit of cost increase  due to the CG deviation on the longitudinal axis of the aircraft (this is an arbitrated value for this work that should be better evaluated in future research, as it depends on the characteristics of the aircraft.).
+
+It is important to consider that the range $[0,\ c_g]$ tends to zero as the aircraft attitude tends to level.
+As the CG deviation varies from 0 to the longitudinal CG limit, the leg cost increase rate varies from 0 to $c_g$.
+}
 
@@
 	\item on each pallet, the items are distributed in such a way that their CG coincides with the centroid of the pallet;
-	\item the CG of the total load must be at a maximum longitudinal distance $limit^{CG}_{long}$\/ from the CG of the aircraft;
-	\item the CG of the total load must be at a maximum lateral distance $limit^{CG}_{lat}$\/ from the CG of the aircraft;
-	\item the pallets are distributed in two identical rows (with odd and even indices, respectively), and the centroid of $p_i$\/ is at distance $D^{lat}_{i}$\/ from the center-line of the aircraft.
+	\item the CG of the payload must be at a maximum longitudinal distance of $limit^{CG}_{long}$ from the CG of the aircraft;
+	\item the CG of the payload must be at a maximum lateral distance of $limit^{CG}_{lat}$ from the CG of the aircraft;
+	\item pallets positions are distributed in two identical rows (with odd and even indices, respectively), and their centroids are at a distance $D^{lat}_{i}$ (where $i$ is the pallet index) from the center-line of the aircraft.
 \end{itemize}
+
+\subsection{Problem Summary}
+
+\newcolumntype{M}{>{\raggedright}p{0.03\textwidth}}
+\newcolumntype{T}{>{\raggedright\arraybackslash}p{0.9\textwidth}}
+
+\bgroup
+\def\arraystretch{1.2}
+\begin{table}[H]
+	\centering
+	\small
+	\begin{tabular}{MT}
+		&Informally, ACLP+RPDP can be summarised as follows:\\
+		\midrule
+		max &  (items score sum) / (tour cost) of picked up and delivered items at each node on a tour  \\
+		\midrule
+s.t.    & Along a tour, the set of unvisited nodes is updated. \\
+		& In each node, an item may be included in at most one pallet.\\		
+		& In each node, {\it Packed Contents (PC)} are composed of items with the same destination. \\
+		& Weight, volume and score of a PC are the corresponding sum of its components.\\
+		& A PC remains on board until its destination.\\	
+		& Items and a PC can only be included in the same pallet if their destinations are the same.\\
+		& Only items destined to the remaining nodes can be loaded.  \\
+		& The lateral and longitudinal torques must be within the operational range of the aircraft.\\
+		& Weight and volume limitations of the pallets must be respected.\\
+		& The total weight must be less than the aircraft payload or the total pallet capacity, whichever is the lowest.\\	
+		\midrule
+	\end{tabular}
+	\normalsize
+\end{table}
+\egroup
 
@@
 
-Given the assumptions and parameters described in the previous section, we are ready to present the mathematical modeling of ACLP+RPDP.
-
-ACLP+RPDP has the objective function \ref{eq1}, and the calculus equations \ref{eq:scores} to \ref{eq:LatPacked} subject to constraints \ref{eq:torqlat} to \ref{eq21}, which will be described below.
-
-\subsection{Problem structure and input data}
-
-In the definitions below and in the tests of Section \ref{sec6}, we use the values assigned in Tables \ref{tab:costs} and \ref{tab:larger}.
-To simplify the notation throughout this text, $p_i$\/ will be called pallet $i$, and $l_k$\/ will be called node $k$.
-
-Let $L = \{ 0, 1, \ldots, K \}$ be the set with the $K+1$\/ nodes, and let $L_k$\/ be the set of remaining nodes when the aircraft is in the node $k$, $0 \leq k \leq K$. Therefore, $L_0 = L$.
-
-Let $d(a,b)$ be the distance from node $a$\/ to node $b$, where $0 \leq a, b \leq K$. By definition, $d(a,a) = 0, \forall a$.
-
-Let $C=\left[c_{a,b}\right]$\/ be the cost matrix of flights, where $c_{a,b} = c_d*d(a,b), 0 \leq a,b \leq K$.
-
-Let $M = \{1, 2, \ldots, m \}$\/ the set of $m$\/ empty pallets assigned to specific positions within the aircraft.
-In Table \ref{tab:larger}, $m=18$.
-Each pallet $i$, $1 \leq i \leq m$, has weight capacity $W_i$, volume capacity $V_i$, longitudinal distance to the CG of aircraft $D^{long}_i$, and lateral distance to the center line of aircraft $D^{lat}_i$.
-
-Let $N_k = \{1, \ldots, n_k \}$\/ be the set of $n_k$\/ items $t_j^k$\/ available for loading on the node $k$, $1 \leq j \leq n_k$, $0 \leq k \leq K$.
-For ease of notation, $t_j^k$\/ will be called item $j$\/ of the node $k$, with score $s_j$, weight $w_j$, volume $v_j$, and destination $to_j \in L_k$.
-Let $N = \bigcup_{0 \leq k \leq K} N_k$ be the set of items of all nodes along a tour.
-
-Let $Q_k = \{1, \ldots, m_k \}$\/ be the set of $m_k \leq m$\/ packed contents $a^k_q$\/ at the node $k$, $1 \leq q \leq m_k$, $0 \leq k \leq K$.
-For ease of notation, $a^k_q$\/ will be called packed content $q$\/ of the node $k$, with total weight $w_q$, total volume $v_q$, and destination $to_q \in L_k$.
-By definition, $m_0 = 0$, and therefore $Q_0 = \varnothing$.
-Packets contents that were destined to node $k$\/ are unloaded when the aircraft arrives at this node, that is, they are not considered in $Q_k$.
-
-\subsection{The decision variables}
-
-Let $X_{ij}^k$\/ and $Y_{iq}^k$\/ be binary variables, where $1 \leq i \leq m$, $1 \leq j \leq n_k$, $1 \leq q \leq m_k$\/ and $0 \leq k \leq K$.
-
-$X_{ij}^k = 1$\/ if the item $j$\/ of the node $k$\/ is assigned to the pallet $i$, and 0 otherwise.
-
-$Y_{iq}^k = 1$\/ if the packed content $q$\/ of the node $k$\/ is assigned to the pallet $i$, and 0 otherwise.
-
-\subsection{The allocation graph}
-
-Allocations of items or packed contents to the pallets in the node $k$\/ can be seen as a bipartite graph $G_k(V_k, E_k)$, where:
-
-\begin{itemize}
-
-\item $V_k = M \cup N_k \cup Q_k$;
-
-\item $E_k = E_{N_k} \cup E_{Q_k}$;
-
-\item $(i, j) \in E_{N_k}$\/ if $X_{ij}^k = 1$, where $i$\/ is a pallet and $j$\/ is a item of the node $k$;
-
-\item $(i, q) \in E_{Q_k}$\/ if $Y_{iq}^k = 1$, where $i$\/ is a pallet and $q$\/ is a packed content of the node $k$.
-
-\end{itemize}
-
-\subsection{The objective function}
-
-Let $S_K = \{s: \{1, \dots, K\} \rightarrow \{1, \dots, K\} \}$\/ be the set of $K!$ permutations $\pi$, which correspond to all possible tours (or itineraries) that have node $0$\/ as origin and end, passing through the other $K$\/ nodes.
-Let $\pi_k$\/ be $k^{th}$\/ node of the tour $\pi$, $1 \leq k \leq K$.
-In this way, the tour $\pi$\/ is described as $\{0,\pi_1,\ldots,\pi_K, 0\}$, {\color{blue}$K > 1$}.
-For ease of notation, we can define $\pi_0 = \pi_{K+1} = 0$.
-
-We consider K > 1 because the case K = 1 has a single and trivial solution for the tour.
-
-The objective of ACLP+RPDP is to find the permutation $\pi \in S_K$, with the corresponding allocation of items on the pallets at each node, that maximizes the function $f_\pi = \tilde{s}_\pi/\tilde{c}_\pi$ \ref{eq1}, where $\tilde{s}_\pi$\/ is the total score of transported items, and $\tilde{c}_\pi$\/ is the total cost of fuel consumed.
+Given the assumptions, scenarios and parameters described in the previous section, we are ready to present the mathematical modeling of the ACLP+RPDP.
+
+\vspace{3mm}
+\textbf{Problem structure and input data:}
+
+Let $C=\{c_{ab}\}$ be the cost matrix of flights, where $c_{ab} = c_d*d(l_a,l_b), 0 \leq a,b \leq \mbox{K}$.
+
+Let $S_K = \{s: \{1, \dots, \mbox{K}\} \rightarrow \{1, \dots, \mbox{K}\} \}$ be the set of $\mbox{K}!$ permutations indexed by $\pi$, which correspond to all possible tours (or itineraries) that have node $0$ as origin and end, passing through the other $\mbox{K}$ nodes.
+
+Note that $\pi_k$ is the node in position $k$ of tour $\pi$, and $\pi_0$ is always node 0, the base.
+
+$\tau^{\pi_k}$ is the total torque applied by the loaded pallets to the aircraft in this node.
+
+Let $L = \{ 0, \pi_1, \ldots, \pi_{K} \}$ be a tour with $\mbox{K}+1$ nodes.
+
+Let $d(\pi_a,\ \pi_b)$ be the distance from $\pi_a$ to $\pi_b$, where $0 \leq a,\ b \leq \mbox{K}$. By definition, $d(\pi_a,\ \pi_a)\ =\ 0$.
+
+Let $L^{\pi_k}$ be the set of remaining nodes when the aircraft is in $\pi_k$, $0 \leq k \leq \mbox{K}$. Therefore, $L^{\pi_0}\ =\ L$ and $L^{\pi_K}\ =\ \{ 0 \}$.
+
+Let $M^{\pi_k} = \{1, 2, \ldots, m \}$ the set of $m$ empty pallets assigned to specific positions within an aircraft.
+Each pallet $i$, has weight capacity $W_i$, volume capacity $V_i$, pallet destinations $to^{\pi_k}_i$, $0 \leq k \leq K$, the longitudinal distance to the CG of aircraft $D^{long}_i$, and the lateral distance to the center line of aircraft $D^{lat}_i$.
+Note that all pallets and aircraft attributes are in uppercase.
+
+Let $N^{\pi_k} = \{1, 2, \ldots, n^{\pi_k} \}$ be the set of $n^{\pi_k}$ items $t_j^{{\pi_k}}$ existent in node $\pi_k$, $0 \leq k \leq K$.
+Each item $j$, has score $s_j$, weight $w_j$, volume $v_j$, and destination $to_j \in L^{\pi_k}$.
+Let $N = \bigcup_{0 \leq k \leq K} N^{\pi_k}$ be the set of items of all nodes along a tour.
+
+Let $Q^{\pi_k} = \{1, 2, \ldots, m^{\pi_k} \}$ be the set of packed contents (PC) in $m^{\pi_k} \leq m$ pallets that will remain on board when the aircraft arrives at node $\pi_k$, with $0 \leq k \leq K$.
+$a^{\pi_k}_q$, $1 \leq q \leq m^{\pi_k}$, is the set of picked-up items that were allocated on a pallet in some of the previous nodes.
+Note that $Q^0 = \{\ \}$ as there is no PC departing from the base.
+
+Note that all items and PC attributes are in lowercase.
+
+$a^{\pi_k}_q$ has total weight $w_q$, total volume $v_q$, and destination $to_q^{\pi_k} \in L^{\pi_k} \cup \{\pi_k\}$.
+If $to_q^{\pi_k} = \pi_k$, then $a^{\pi_k}_q$ is unloaded, and the pallet will be available to receive other packet items; otherwise, $a^{\pi_k}_q$ remains on the aircraft, eventually in another pallet, and with some items of $N^{\pi_k}$ having the same destination.
+
+\vspace{3mm}
+\textbf{The decision variables:}
+
+Let $X_{ij}^{\pi_k}$ and $Y_{iq}^{\pi_k}$ be binary variables, where $0 \leq k \leq \mbox{K}$, $1 \leq j \leq n^{\pi_k}$, $1 \leq i \leq m$ and $1 \leq q \leq m^{\pi_k}$.
+
+$X_{ij}^{\pi_k} = 1$ if item $t_j^{{\pi_k}}$ is assigned to pallet $i$ in node $\pi_k$, and 0 otherwise.
+
+$Y_{iq}^{\pi_k} = 1$ if a PC $a_q^{{\pi_k}}$ is assigned to pallet $i$ in node $\pi_k$, and 0 otherwise. By definition, $Y_{iq}^0=0$.
+
+\vspace{3mm}
+\textbf{The structure of a solution:}
+
+Allocations of items or PC to pallets position in the node $\pi_k$ can be seen as a bipartite graph $G^{\pi_k}(V^{\pi_k}, E^{\pi_k})$,
+where:
+
+$V^{\pi_k} = M^{\pi_k} \cup N^{\pi_k} \cup Q^{\pi_k}$,
+
+$E^{\pi_k} = E^M_{N^{\pi_k}} \cup E^M_{Q^{\pi_k}}$,
+
+$(i, j) \in E^M_{N^{\pi_k}}$ if $X_{ij}^{\pi_k} = 1$, and
+
+$(i, q) \in E^M_{Q^{\pi_k}}$ if $Y_{iq}^{\pi_k} = 1$.
+
+\vspace{3mm}
+\textbf{The objective function:}
+
+The objective of ACLP+RPDP is to find the permutation $\pi \in S_K$, with the corresponding allocation of items on the pallets at each node, that maximizes the function $\tilde{s}_\pi/\tilde{c}_\pi$ \ref{eq1}, where $\tilde{s}_\pi$\/ is the total score of transported items, and  $\tilde{c}_\pi$\/ is the total cost of fuel consumed.
+In this way, the tour is described as $\{0,\ {\pi(1)},\ \ldots,\ {\pi(K)},\ 0\}$.
 
 \begin{equation} \label{eq1}
-	\max_{\pi \in S_K} f_\pi = \tilde{s}_\pi/\tilde{c}_\pi
-\end{equation}
-
-\subsection{The calculus equations}
-
-$\tilde{s}_\pi$\/ \ref{eq:scores} is the sum of the scores of the items loaded on the aircraft throughout the tour $\pi$.
+	\max_{\pi \in S_K} \tilde{s}_\pi/\tilde{c}_\pi
+\end{equation}
+
+\vspace{3mm}
+\textbf{Where:}
+\vspace{3mm}
 
 \begin{equation} \label{eq:scores}
-	\tilde{s}_\pi = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times s_j
-\end{equation}
-
-$\tau_{\pi_k}$\/ \ref{eq:tau} is the longitudinal torque applied by the loaded pallets, in the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.
+	\tilde{s}_\pi = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times s_j
+\end{equation}
+
+\vspace{3mm}
+In Equation \ref{eq:tau}, $\tau^{\pi_k}$ is the total torque applied to the aircraft in node ${\pi_k}$.
+This is the sum of all pallets torque (pallet centroid distance to the CG times the total weight packed on the pallet).
 
 \begin{equation} \label{eq:tau}
-	\tau_{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times \Big ( \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{long}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-In this way, we can calculate the cost $\tilde{c}_\pi$\/ \ref{eq:costs}, referring to the total fuel consumed in the tour $\pi$\/ due to the distances traveled and the CG longitudinal deviations in each pallet.
+	\tau^{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times (\sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times w_j +  \sum_{q=1}^{m^{\pi_k}} Y_{iq}^{\pi_k} \times w_q) \Big ];\ k \in \{0,1,2, \ldots, \mbox{K}\}
+\end{equation}
+
+\vspace{3mm}
+In Equation \ref{eq:eps}, $\epsilon^{\pi_k}$ the torque applied to the aircraft in node ${\pi_k}$ divided by the maximum torque possible.
+
+\begin{equation} \label{eq:eps}
+	\epsilon^{\pi_k} = \frac{\tau^{\pi_k}}{W_{max} \times limit^{CG}_{long}};\ k \in \{0,1,2, \ldots, \mbox{K}\}
+\end{equation}
+
+\vspace{3mm}
+In the tour cost calculation in Equation \ref{eq:costs}, the first parcel denotes the fuel cost from the base until the last node, and the second the fuel cost in the last leg, from the last node back to the base.
 
 \begin{equation} \label{eq:costs}
-	\tilde{c}_{\pi} = \sum_{k=0}^{K} \Big [ c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|) \Big ] 
-\end{equation}
-
-The sets of nodes not yet visited are defined in \ref{eq:pdp11} and \ref{eq:departing}.
+	\tilde{c} = \sum_{k=0}^{K-1} \Big [ c^{\pi_k, \pi(k+1)}\times(1+c_g\times|\epsilon^{\pi_k}|) \Big ] + c^{\pi_K,0}\times(1+c_g\times|\epsilon^{\pi_K}|)
+\end{equation}
+
+\vspace{3mm}
+Equations \ref{eq:pdp11} and \ref{eq:departing} state the set of unvisited nodes departing from the base ($0$) and $\pi_k$.
 
 \begin{equation} \label{eq:pdp11}
-	L_0 = L
+	L^{0} = L
 \end{equation}
@@
 \begin{equation} \label{eq:departing}
-	L_{\pi_k} = L_{\pi_{k-1}} - \{\pi_k\}; \ k \in \{1, \ldots, K\}
-\end{equation}
-
-At the beginning of the tour $\pi$, there are no packed contents \ref{eq:pdp13}.
+	L^{\pi_k} = L^{\pi_{k-1}} - \{\pi_k\}; \ k \in \{1,2, \ldots, \mbox{K}\}
+\end{equation}
+
+\vspace{3mm}	
+Although there are no items packed at the beginning of the transportation plan, we defined these variables for ease of notation \ref{eq:pdp13}.
 
 \begin{equation} \label{eq:pdp13}
-	m_0 = 0
-\end{equation}
-
-$\epsilon_{\pi_k}^t$\/ \ref{eq:LatItem} and $\epsilon_{\pi_k}^a$\/ \ref{eq:LatPacked} are the lateral torque applied by the loaded pallets, in the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.
-For ease of notation, $\epsilon_{\pi_k}^t$\/ corresponds to the items, and $\epsilon_{\pi_k}^a$\/ to the packed contents.
+	Y^0_{iq} = 0; w_q^0 = 0; v_q^0 = 0; to_q^0 = -1;\ i,q \in \{1, 2, \ldots, m\}
+\end{equation}
+
+\vspace{3mm}
+PC appear when there are items on the pallets that will not be unloaded on the next node.
+The PC's total weight (\ref{eq:cons2}) and volume (\ref{eq:cons3}) correspond to all the items that were on the pallet, since all these items have the same destination.
+
+\begin{equation} \label{eq:cons2}
+	w_q^{\pi(k+1)} = \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times w_j + \sum_{q=1}^{m^{\pi_k}} Y_{iq}^{\pi_k} \times w_q;  \ q \in \{1, 2, \ldots, m^{\pi_k}\}; k \in \{0,1,2, \ldots, \mbox{K-1}\}
+\end{equation}
+
+\begin{equation} \label{eq:cons3}
+	v_q^{\pi(k+1)} = \sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times v_j + \sum_{q=1}^{m^{\pi_k}} Y_{iq}^{\pi_k} \times v_q;  \ q \in \{1, 2, \ldots, m^{\pi_k}\}; k \in \{0,1,2, \ldots, \mbox{K-1}\}
+\end{equation}
+
+\vspace{3mm}
+Equations \ref{eq:LatItem} and \ref{eq:LatPacked} calculate the weight differences between the aircraft sides to be used ahead in the constraint of the CG lateral deviation through Equation \ref{eq:torqlat}.
 
 \begin{equation} \label{eq:LatItem}
-	\epsilon_{\pi_k}^t = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{j=1}^{n_{\pi_k}} \Big ( X_{ij}^{\pi_k} \times w_j \times (i\%2) - X_{ij}^{\pi_k} \times w_j \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}; \ k \in \{0, \ldots, K\}
+	w^{\pi_k}_{items}   = \sum_{i=1}^{m} \sum_{j=1}^{n^{\pi_k}}  \Big ( X_{ij}^{\pi_k} \times w_j \times (i\%2) - X_{ij}^{\pi_k} \times w_j \times (i+1)\%2 \Big ); \ k \in \{0,1,2, \ldots, \mbox{K}\}
 \end{equation}
@@
 \begin{equation} \label{eq:LatPacked}
-	\epsilon_{\pi_k}^a = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{q=1}^{m_{\pi_k}} \Big ( Y_{iq}^{\pi_k} \times w_q \times (i\%2) - Y_{iq}^{\pi_k} \times w_q \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\subsection{The constraints}
-
-Finally, we can consider the constraints on each node $\pi_k$:
-
-\begin{itemize}
-	\item The longitudinal \ref{eq:torqlong} and the lateral \ref{eq:torqlat} torques must be within the limits of the aircraft.
-	\item The items allocated to each pallet cannot exceed its weight \ref{eq:app2} and volume \ref{eq:app3} limits.
-	\item At most, each item is associated with a single pallet \ref{eq:app4};
-	\item Packet contents that have not yet reached their destination must remain on board \ref{eq:app5}.
-	\item Items allocated on the same pallet must have the same destinations.
-In this case, we need two constraints: \ref{eq18} and \ref{eq19}.
-If $X_{ia}^{\pi_k} = X_{ib}^{\pi_k} = 1$, both constraints require that $to_a = to_b$; otherwise these constraints have no effect.
-	\item If there is a packed content on the pallet, it must also have the same destination as the other items. Similarly, we use two constraints: \ref{eq20} and \ref{eq21}.
-\end{itemize}
+	w^{\pi_k}_{packed} =  \sum_{i=1}^{m} \sum_{q=1}^{m^{\pi_k}} \Big ( Y_{iq}^{\pi_k} \times w_q \times (i\%2) - Y_{iq}^{\pi_k} \times w_q \times (i+1)\%2 \Big ); \ k \in \{0,1,2, \ldots, \mbox{K}\}
+\end{equation}
+
+\vspace{3mm}
+\textbf{Subject to:}
+\vspace{3mm}
+
+Load only items and keep boarded PC destined to the unvisited nodes (\ref{eq:pdp12} and \ref{eq:pdp9}).
+
+\begin{equation} \label{eq:pdp12}
+	X_{ij}^{\pi_k} = 0;\ to_j \notin L^{\pi_k}; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi_k}\}; \ k \in \{0,1,2, \ldots, \mbox{K}\}
+\end{equation}
+
+\begin{equation} \label{eq:pdp9}
+	Y_{iq}^{\pi_k} = 0;\ to_q \notin L^{\pi_k}; \ i \in \{1, 2, \ldots, m\}; \ q \in \{1, 2, \ldots, m^{\pi_k}; \ k \in \{0,1,2, \ldots, \mbox{K}\}
+\end{equation}
+
+\vspace{3mm}
+The Equation \ref{eq:torqlat} is the lateral torque constraint for the aircraft, while the Equation \ref{eq:torqlong} is the longitudinal torque constraint.
+
+\begin{equation} \label{eq:torqlat}
+	D_i^{lat} \times \Big | w^{\pi_k}_{items} + w^{\pi_k}_{packed} \Big | \leq  W_{max} \times limit^{CG}_{lat}; \ k \in \{0,1,2, \ldots, \mbox{K}\}
+\end{equation}
 
 \begin{equation} \label{eq:torqlong}
-	| \tau_{\pi_k} | \leq 1;\ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:torqlat}
-	| \epsilon_{\pi_k}^t + \epsilon_{\pi_k}^a| \leq  1; \ k \in \{0, \ldots, K\}
+	\Big | \tau^{\pi_k} \Big | \leq W_{max} \times limit^{CG}_{long};\ k \in \{0,1,2, \ldots, \mbox{K}\}
+\end{equation}
+
+\vspace{3mm}
+The weight limitation of the aircraft must be respected \ref{eq:payload}.
+The sum of weights \ref{eq:app2} and volumes \ref{eq:app3} in each pallet must not exceed its capacity.
+Each item is associated with a pallet at most \ref{eq:app4}.
+
+\begin{equation} \label{eq:payload}
+	\sum_{i=1}^{m^{\pi_k}} \Big (\sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times w_j + \sum_{q=1}^{m^{\pi_k}} Y_{iq}^{\pi_k} \times w_q \Big ) \leq W_{max}; \ k \in \{0,1,2, \ldots, \mbox{K}\}
 \end{equation}
@@
 \begin{equation} \label{eq:app2}
-	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}; \ k \in \{0, \ldots, K\}
+	\sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times w_j + \sum_{q=1}^{m^{\pi_k}} Y_{iq}^{\pi_k} \times w_q  \leq W_i; \ i \in \{1, 2, \ldots, m\}; \ k \in \{0,1,2, \ldots, \mbox{K}\}
 \end{equation}
@@
 \begin{equation} \label{eq:app3}
-	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}; \ k \in \{0, \ldots, K\}
+	\sum_{j=1}^{n^{\pi_k}} X_{ij}^{\pi_k} \times v_j + \sum_{q=1}^{m^{\pi_k}} Y_{iq}^{\pi_k} \times v_q  \leq\ V_i; \ i \in \{1, 2, \ldots, m\}; \ k \in \{0,1,2, \ldots, \mbox{K}\}
 \end{equation}
@@
 \begin{equation} \label{eq:app4}
-	\sum_{i=1}^{m} X_{ij}^{\pi_k} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
+	\sum_{i=1}^{m} X_{ij}^{\pi_k} \leq 1; \ j \in \{1, 2, \ldots, n^{\pi_k}\}; \ k \in \{0,1,2, \ldots, \mbox{K}\}
+\end{equation}
+
+\vspace{3mm}
+Equations \ref{eq18} and \ref{eq19} were set to group items with the same destinations.
+
+\begin{equation} \label{eq18}
+	to_a - to_b \geq -K \times (1-X_{ia}^{\pi_k} \times X_{ib}^{\pi_k}); \ a,b \in \{1, 2, \ldots, n^{\pi_k}\}; \ k \in \{0, 1, \ldots, K\}
+\end{equation}
+
+\begin{equation} \label{eq19}
+	to_a - to_b \leq K \times (1-X_{ia}^{\pi_k} \times X_{ib}^{\pi_k}); \ a,b \in \{1, 2, \ldots, n^{\pi_k}\}; \ k \in \{0, 1, \ldots, K\}
+\end{equation}
+
+\vspace{3mm}
+A PC is allocated to a pallet if its destination in in the not visited nodes set.
+\begin{equation} \label{eq17}
+	Y_{iq}^{\pi_k} = 1; \ q \in \{1, 2, \ldots, m^{\pi_k}\}; \ k \in \{1, \ldots, K\};  \ to_q \in L^{\pi_k}
+\end{equation}
+
+\vspace{3mm}
+Equations \ref{eq20} and \ref{eq21} were set to group items and PC with the same destinations.
+
+\begin{equation} \label{eq20}
+	to_j - to_q \geq -K \times (1-X_{ij}^{\pi_k} \times Y_{iq}^k); \ j \in \{1, 2, \ldots, n^{\pi_k}\}; \ q \in \{1, 2, \ldots, m^{\pi_k}\}; \ k \in \{0, 1, \ldots, K\}
+\end{equation}
+
+\begin{equation} \label{eq21}
+	to_j - to_q \leq  K \times (1-X_{ij}^{\pi_k} \times Y_{iq}^k); \ j \in \{1, 2, \ldots, n^{\pi_k}\}; \ q \in \{1, 2, \ldots, m^{\pi_k}\}; \ k \in \{0, 1, \ldots, K\}
+\end{equation}
+
+\vspace{3mm}
+The PC must remain on board \ref{eq:app5} until their destinations.
 
 \begin{equation} \label{eq:app5}
-	\sum_{i=1}^{m} Y_{iq}^{\pi_k} = 1;\ to_q \in L_{\pi_k}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq18}
-	to_a - to_b \geq -K \times (1-X_{ia}^{\pi_k} \times X_{ib}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ a,b \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq19}
-	to_a - to_b \leq K \times (1-X_{ia}^{\pi_k} \times X_{ib}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ a,b \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq20}
-	to_j - to_q \geq -K \times (1-X_{ij}^{\pi_k} \times Y_{iq}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq21}
-	to_j - to_q \leq  K \times (1-X_{ij}^{\pi_k} \times Y_{iq}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-Constraints \ref{eq18} to \ref{eq21} are not within standard {\it Mixed-Integer Programming}\/ (MIP), and therefore need to be handled in a different way.
-In the next section, we will explain the strategy adopted to solve ACLP+RPDP.
+	Y_{iq}^{\pi_k} = 1;\ to_q \in L^{\pi_k}; \ q \in \{1, 2, \ldots, m^{\pi_k}\}; \ k \in \{0,1,2, \ldots, \mbox{K}\}
+\end{equation}
+
+Until this point, our research problem has been thoroughly described, and it is certainly very difficult since it combines several sub-problems, which are difficult to solve in their own right.
+Then, we develop an innovative process to efficiently solve it.
 
@@
 
-Once the assumptions of this work and the mathematical modeling of the problem are presented, it is possible to see that ACLP+RPDP is {\it NP-hard}.
+Once the assumptions of this work and the mathematical modeling of the problem are presented, it is easy to see that ACLP+RPDP is NP-hard.
 In a similar way to Lurkin and Schyns \cite[p.
-6]{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$\/ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in node 0, and no items in node 1.
-Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is {\it NP-complete}.
-
-Throughout our research, we have thoughtfully described the ACLP+RPDP model in standard MIP format and found that no MIP solver can handle its practical cases in a feasible time.
-Thus, as ACLP+RPDP is highly complex, involving four intractable subproblems (APP, WBP, PDP and TSP), our strategy will be to focus only on real cases and to develop quick node-by-node solutions, not necessarily optimal, that allow to build a complete tour.
-
-The ACLP+RPDP solution defines a tour and a corresponding loading and unloading plan at each node.
-In this way, the loading and unloading time will be limited only by the use of the equipment at each node.
-Therefore, there will only be waiting time at the base node, corresponding to the runtime of ACLP+RPDP algorithm.
-For this reason, our method will also be parameterized by the maximum time that can be waited until the start of the load in the base node.
-
-On the other hand, we know that transport aircraft generally have a few dozen pallets, the flight plan has less than $6$\/ nodes, and each node has hundreds of items to be shipped.
-Under these circumstances, we can adopt two more strategies:
-
-\begin{itemize}
-
-\item Considering the number of destinations less than the number of pallets ($K < m$), we can preset the destinations of the pallets in each node.
-In this way, we will reserve a number of pallets proportional to the volume demanded by each destination at the shipping node.
-We could have used another criterion, but it was observed in the experiments that volume is more constrictive in airlift.
-
-\item If the number of nodes is small and we have quick node-by-node solutions, we have the possibility to check the solution corresponding to the tour with the shortest total distance, and even test all possible tours, selecting the one that provides the best value for the objective function.
-
-\end{itemize}
-
-Our complete strategy is summarized in Algorithm \ref{alg:main}.
+6]{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in node 0, and no items in node 1.
+Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is NP-complete.
+
+{\color{blue}
+
+Equations \ref{eq18} and \ref{eq19} were set to group items with the same destinations.
+But, as they mean a combination of $n^{\pi_k}$ items 2 by 2, this may easily lead to huge numbers, 60,000 to 320,000 possible combinations, due to the number of items in each node, which together with the other constraints make a MIP solution very hard.
+
+To work around this, we decided to create the destination attribute for the pallets ($TO_i^{\pi_k}$), to reduce the number of combinations and enable the solution using heuristics.
+$TO^{\pi_k}_i$ denotes that pallet $i$ may assume a different destination in each node ${\pi_k}$.
+
+On subsequent nodes, the PC can be allocated with other items of same destination \ref{eq:cons5}.
+
+\begin{equation} \label{eq:cons5}
+	TO_i^{\pi(k+1)} = to_j;\ X^{\pi_k}_{ij} = 1  \ \mbox{and}\ to_j \in L^{\pi_{k+1}};  \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi_k}\}; \ k \in \{0,1,2, \ldots, \mbox{K-1}\}
+\end{equation}
+
+\vspace{3mm}
+At each node, an item (\ref{eq:pdp8}) and the already packed contents (\ref{eq:pdp2}) must only be allocated to a pallet if the destinations are the same.
+
+\begin{equation} \label{eq:pdp8}
+	TO_i^{\pi_k} = to_j;\ X_{ij}^{\pi_k} = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi_k}\}; \ k \in \{0,1,2, \ldots, \mbox{K}\}
+\end{equation}
+
+\begin{equation} \label{eq:pdp2}
+	TO_i^{\pi_k} = to_q;\ Y_{iq}^{\pi_k} = 1; \ i \in \{1, 2, \ldots, m\};\ q \in \{1, 2, \ldots, m^{\pi_k}\}; \ k \in \{0,1,2, \ldots, \mbox{K}\}
+\end{equation}
+
+}
+
+Real cases are more complex as they have hundreds of different items in each node and involve four intractable sub-problems: APP, WBP, PDP and TSP.
+Through the mathematical modeling presented in the previous section, we verify that {\it Mixed-Integer Programming}\/ (MIP) is not able to solve these cases in feasible time.
+Thus, it is necessary to adopt some strategy to find a viable solution, not necessarily optimal, that seeks to maximize the objective function $f_{\pi}$.
+
+Our strategy is based on the fact that, in real cases, $K$\/ is usually small.
+Specifically, we will consider $K \leq 6$\/ throughout this work, which is a higher value than usual in {\it Brazilian Air Force}'s missions.
+As a result, if we have fast node-by-node solutions that allow us to construct a complete tour, we will be able to test all possible $K!$\/ tours and thus select the one that provides the best value for the $f_{\pi}$\/ function.
+
+{\color{blue}Another possibility to make this approach more general is to devise some bounding schemes to allow discarding most higher-cost tours and allowing an increase in the number of nodes without compromising performance.}
+
+For now, our tactic will be, at each shipping node, to predefine the destinations of the pallets at that node.
+In this way, we will reserve a number of pallets proportional to the volume demanded by each destination at the shipping node (we could have used another criterion, but it was observed in the experiments that volume is more constrictive in airlift).
+Once the destinations of the pallets are defined, we will use heuristics and a MIP solver to find the best possible node-by-node solution. This strategy is summarized in Algorithm \ref{alg:main}.
 
 \begin{algorithm}[H]
-	\caption{ACLP+RPDP solving}  \label{alg:main}
+	\caption{Solves ACLP+RPDP for a scenario with certain volume surplus (1.2, 1.5, or 2.0)}  \label{alg:main}
 	\begin{algorithmic}[1]
 
-		\Procedure{$ACLP+RPDP$}{$scenario, surplus, time$}
-
-		\State Let aircraft and $M$\/ be according to Table \ref{tab:larger}  \label{main:M} \Comment{Aircraft parameters}
-
-		\State Let $K$, $L$\/ and $C$ be according to $scenario$\/ and Table \ref{tab:costs} \label{main:KLC} \Comment{Nodes, distances and costs}
-
-		\State Let $\pi_{TSP}$\/ be the tour with the shortest distance traveled
-
-		\State $N \gets ItemsGeneration(scenario, surplus)$ \label{main:items} \Comment{Items available for shipment}
+		\Procedure{$ACLP+RPDP$}{$scenario,\ surplus,\ timeLim$}
+
+		\State Let $L, M, C$ be according to $scenario$ \label{main:LMC} \Comment{Nodes, pallets and costs}
+
+		\State $N \gets ItemsGeneration(scenario,\ volume)$ \label{main:items} \Comment{Items available for shipment}
 
 		\For {each $method$} \Comment{$method$ is a MIP solver or a heuristic}
-
-			\State $f_{\pi_{TSP}} \gets SolveTour(\pi_{TSP}, L, M, C, N, method, time/(K+1))$ \label{main:f1}
-			\State $answer1[scenario,surplus,method] \gets f_{\pi_{TSP}}$ \label{main:a1} \Comment{Results corresponding to $\pi_{TSP}$}
 			\For {each $\pi \in S_K$} \label{main:loop1} \Comment{$\pi$ is a permutation of nodes (a tour)}
-				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, time/K!)$ \Comment{$f_{\pi}$\/ is a solution to the tour $\pi$} \label{main:f2}
+				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method,\ timeLim )$ \Comment{a solution to the tour $\pi$} \label{main:method}
 			\EndFor \label{main:loop2}
-			\State $answer2[scenario,surplus,method] \gets \max f_{\pi}$ \label{main:a2} \Comment{Best result among all permutations}
+			\State $answer[scenario,surplus,method] \gets \max f_{\pi}$ \label{main:f} \Comment{Best result among all permutations}
 		\EndFor
@@
 
-The aircraft parameters are in Table \ref{tab:larger} (line \ref{main:M}).
-In this algorithm, we use five values for $scenario$, according to Table \ref{tab:scenarios}, which defines $K$\/ and the set $L$\/ of nodes (line \ref{main:KLC}).
+{\color{blue}
+	Until scenario 4, solving all tours and selecting the best is fast enough for operational use.
+For scenarios 4 and 5 (which are not typical in the Brazilian Air Force), we plan bounding schemes to allow for the discarding of most of the higher-cost tours.
+This is expected to reduce the running time for scenarios 4 and 5, favoring the discovery of better search mechanisms as well as further applications that may require more nodes per tour.
+}
+
+In this algorithm, there are six values for the $scenario$\/ parameter, according to Table \ref{tab:scenarios}, which defines $K$, the sets of nodes, the aircraft, the pallets and the costs from Tables \ref{tab:smaller} or \ref{tab:larger} that will be used (line \ref{main:LMC}).
+
+The parameter $volume$\/ is a value in (1.2, 1.5, 2.0), which corresponds, at each node $\pi_k$, to the ratio between the sum of the volumes of the items ($\sum_{j=1}^{n^{\pi_k}} v_j$) and the load capacity of the pallets ($\sum_{i=1}^{m} V_i$).
+This parameter is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).
+
+The parameter $method$\/ corresponds to a MIP solver or a heuristic that we will present in subsection \ref{methods}.
+The loop of lines \ref{main:loop1}-\ref{main:loop2} goes through all permutations, where the node-by-node solutions are performed by $SolveTour$, whose result is stored in $f_{\pi}$.
+The best result among all $K!$\/ tours will be the answer for $scenario$, $volume$\/ and $method$\/ (line \ref{main:f}).
 
@@
 		\toprule
-		{\bf Scenario} & {$K$} & {$L$} \\		
+		{\bf Scenario} & {$K$} & {$L^{0}$} \\		
 		\midrule
-		1 & 2    & \{$1,2$\}           \\
-		2 & 3    & \{$1,2,3$\}         \\
-		3 & 4    & \{$1,2,3,4$\}       \\
-		4 & 5    & \{$1,2,3,4,5$\}     \\
-		5 & 6    & \{$1,2,3,4,5,6$\}   \\
+		1 & 2    & \{$\pi_1,\ \pi_2,\ 0$\}                                 \\
+		2 & 3    & \{$\pi_1,\ \pi_2,\ \pi_3,\ 0$\}                         \\
+		3 & 4    & \{$\pi_1,\ \pi_2,\ \pi_3,\ \pi_4,\ 0$\}                 \\
+		4 & 5    & \{$\pi_1,\ \pi_2,\ \pi_3,\ \pi_4,\ \pi_5,\ 0$\}         \\
+		5 & 6    & \{$\pi_1,\ \pi_2,\ \pi_3,\ \pi_4,\ \pi_5,\ \pi_6,\ 0$\} \\
 		\bottomrule
@@
 
-$surplus$\/ is a value in $\{1.2, 1.5, 2.0\}$, which corresponds, at each node $k$, to the ratio between the sum of the volumes of the items  and the load capacity of the pallets ($surplus = \sum_{j=1}^{n_k} v_j$/$\sum_{i=1}^{m} V_i$).
-This parameter allows us to verify the different behavior of each $method$, according to $scenario$\/ and the quantity of items available for shipment.
-It is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).
-
-$time$\/ is a runtime limit, which will be distributed among the node-by-node solutions (lines \ref{main:f1} and \ref{main:f2}).
-Each tour consists of $K+1$\/ trips, and the total number of tours is $K!$.
-
-$method$\/ corresponds to a MIP solver or a heuristic that we will present in subsection \ref{methods}, where $SolveTour$\/ is the node-by-node solution.
-
-The results corresponding to the shortest tour are stored in $answer1$\/ (line \ref{main:a1}), and those obtained by testing all tours are stored in $answer2$\/ (line \ref{main:a2}).
-
-Next, we will present two subsections: in the first we explain how $SolveTour$\/ is executed, presetting the destinations of the pallets.
-In the second we will present the heuristic developed for node-by-node solutions.
+{\color{blue} Column $L^{0}$ in Table \ref{tab:scenarios} represents the set of non-visited nodes staying in the base. }
+
+Next, we will present two subsections: in the first we explain how $SolveTour$ is executed, while in the second we will present the heuristic developed for node-by-node resolutions.
 
@@
 
-As we commented in the previous subsection, we will adopt the strategy of presetting the destinations of each pallet throughout the tour.
-This is feasible in practical cases, where $K<m$.
-For this, each pallet $i$\/ also has a field $T^k_i$, $0\leq k \leq K$, which stores its next destination after being loaded in node $k$.
-For this reason, $T^k_i \in L_k$, $1 \leq i \leq m$, $0\leq k \leq K$.
-
-$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $time$\/ is the runtime limit that will be distributed among the $K+1$\/ legs of the tour.
+In addition to the set of nodes, pallets, costs and items, $SolveTour$, described in Algorithm \ref{alg:tour}, receives the parameter $method$, which corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and the parameter $\pi$, which is a {\color{blue} permutation of nodes (excluding the base)} that defines the order of visits in this tour.
+
+As we mentioned in the previous section, all tours start and end at the base\/ (lines \ref{tour:pi1}-\ref{tour:pi2}).
+After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ flights (lines \ref{tour:loop1}-\ref{tour:loop2}).
+Initially we set pallets destination as $\mbox{-1}$\/ (line \ref{tour:-1}).
+
+When the aircraft is at the base, the initial graph $G_1$\/ is empty because it has no PC \ref{tour:g11}.
+Otherwise, the set $L^{\pi_k}$\/ of remaining nodes is updated (line \ref{tour:lk1}), and $UpdatePacked$\/ (line \ref{tour:dest}) returns the set of PC that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation.
+This allocation is stored in graph $G_1$\/ (line \ref{tour:g12}).
+
+In the context of this work, we know that $m>K$, once the aircraft has $18$\/ pallets {\color{blue}} and $K\leq6$, allowing there to be at least one pallet for each node to be visited.
+$SetPalletsDestinations$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node, without changing the pallets destination with PC (packed contents).
+
+Finally, $SolveNode$\/ includes the edges corresponding to the items shipped at the current node, returning the graph $G_2$\/ (line \ref{tour:node}).
+The score and the CG deviation of this graph are calculated (line \ref{tour:analyse}) and accumulated (lines \ref{tour:score2}-\ref{tour:cost2}), allowing the final result of this tour (line \ref{tour:f}).
 
@@
 
-		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method$, $time$}
-
-		\State $\pi_0     \gets 0$ \label{tour:pi1} \Comment{Base $0$\/ is the first node}
-		\State $\pi_{K+1} \gets 0$ \label{tour:pi2} \Comment{Base $0$\/ is the last node}
+		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method$}
+
+		\State $\pi_0     \gets 0$ \label{tour:pi1} \Comment{The first node is the base}
+		\State $\pi_{K+1} \gets 0$ \label{tour:pi2} \Comment{The last node is also the base}
 
@@
 
-		\For {$k \gets 0$ to $K$} \label{tour:loop1}	
-		\State $L_{\pi_k} \gets L - \{\pi_0,\pi_1,\ldots,\pi_k\}$  \label{tour:lk1}	\Comment{$\pi_k$ is the current node}	
-		\For{$i \gets 1$ to $m$}
-		\State $T_i^{\pi_k} \gets -1$ \label{tour:-1} \Comment{Unset the pallet destinations}
+		\For {$k \in \{0,1,2,...,K\}$} \label{tour:loop1}
+
+			\For{$i \in \{1,2,3,...,m\}$}
+				\State $to_i^{\pi_k} \gets -1$ \label{tour:-1} \Comment{unset the pallet destination}
 			\EndFor
@@
 			\If {$k = 0$}
-		\State Let $G_1(M \cup N_0, \varnothing)$ \label{tour:g11}
+				\State $L^0 \gets L - \{0\}$
+				\State Let $G_1(M \cup N^0, \varnothing)$ \label{tour:g11}
 			\Else
-
-		\State $E_{Q_{\pi_k}}, M \gets UpdatePacked(M, Q_{\pi_k}, \pi_k)$ \label{tour:dest}			
-		\State Let $G_1(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$ \label{tour:g12}
+				\State $L^{\pi_k} \gets L - \{0,...,\pi_k\}$  \label{tour:lk1}	\Comment{$\pi_k$ is the current node}		
+				\State $E^M_{Q^{\pi_k}}, M \gets UpdatePacked(M, Q^{\pi_k}, \pi_k)$ \label{tour:dest}			
+				\State Let $G_1(M \cup N^{\pi_k} \cup Q^{\pi_k}, E^M_{Q^{\pi_k}})$ \label{tour:g12}
 			\EndIf  \label{tour:lk2}
@@
 
-		\State $G_2 \gets SolveNode(method, \pi_k, G_1, time/(K+1))$ \label{tour:node}
-
-		\State $s, \tau \gets ScoreAndDeviation(\pi_k, G_2)$ \label{tour:analyse}
+			\State $G_2 \gets SolveNode(method, \pi_k, G_1)$ \label{tour:node}
+
+			\State $s, \epsilon \gets ScoreAndDeviation(\pi_k, G_2)$ \label{tour:analyse}
 
 			\State $score \gets score + s$ \label{tour:score2}
-		\State $cost  \gets cost  + c_{\pi_k,\pi_{k+1}} \times (1 + c_g \times |\tau|)$ \label{tour:cost2}
+			\State $cost  \gets cost  + c_{\pi_k,\pi_{k+1}} \times (1 + c_g \times |\epsilon|)$ \label{tour:cost2}
 
@@
 
-As we mentioned in the previous section, all tours start and end at the base $0$\/ (lines \ref{tour:pi1}-\ref{tour:pi2}).
-After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ flights (lines \ref{tour:loop1}-\ref{tour:loop2}).
-Initially, the set $L_{\pi_k}$\/ of remaining nodes is updated (line \ref{tour:lk1}), and the pallet destinations are unset (line \ref{tour:-1}).
-
-When the aircraft is at the base, the initial graph $G_1$\/ is empty because it has no packed content \ref{tour:g11}.
-Otherwise, $UpdatePacked$\/ (line \ref{tour:dest}) returns the set of packed contents that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation.
-This allocation is stored in graph $G_1$\/ (line \ref{tour:g12}).
-
-$SetPalletsDestinations$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node, without changing the pallets destination with packed contents.
-
-Finally, $SolveNode$\/ includes the edges corresponding to the items shipped at the current node, returning the graph $G_2$\/ (line \ref{tour:node}).
-The score and the CG deviation of $G_2$\/ are calculated (line \ref{tour:analyse}) and accumulated (lines \ref{tour:score2}-\ref{tour:cost2}), allowing the final result of this tour (line \ref{tour:f}).
-
-$UpdatePacked$, described in Algorithm \ref{alg:cons}, finds the best packed-pallet allocation, in terms of CG deviation, for the packed contents that remain on board.
+{\color{blue}
+
+$UpdatePacked$, described in Algorithm \ref{alg:cons}, finds the best packet-pallet allocation, in terms of CG deviation, for the PC that remain on board.
+Then $MinCGDeviation$\/ (line \ref{cons:minCG}) is run through a MIP solver ({\it Gurobi}) with the equations \ref{eq:torque}, \ref{eq:embarked} and \ref {eq:one}, to relocate the PC on the pallets, minimizing torque and ensuring that they all remain on board, one PC on each pallet.
+As there are few variables, the MIP solver returns an allocation $E^M_{Q^{\pi_k}}$ in less than 30 milliseconds.
+Finally, the destination of each pallet with a PC is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).
+
+}
 
@@
 
-		\Procedure{$UpdatePacked$}{${M, Q_{\pi_k}, \pi_k}$}
-
-		\State $E_{Q_{\pi_k}} \gets MinCGDeviation(E_{Q_{\pi_k}})$ \label{cons:minCG}
-
-		\For{$i \gets 1$ to $m$} \label{cons:Ybegin}
-
-		\For{$q \gets 1$ to $m_{\pi_k}$}
-
-		\State $T_i^{\pi_k} \gets -1$
-
-		\If{$(i, q) \in E_{Q_{\pi_k}}$} 
-		\State $T_i^{\pi_k} \gets to_q$   \Comment{Set the pallet destinations}
+		\Procedure{$UpdatePacked$}{${M, Q^{\pi_k}, \pi_k}$}
+
+		\State $E^M_{Q^{\pi_k}} \gets MinCGDeviation(E^M_{Q^{\pi_k}})$ \label{cons:minCG}
+
+		\For{$i \in \{ 1, 2, \ldots, m \}$} \label{cons:Ybegin}
+
+			\For{$q  \in \{ 1, 2, \ldots, m^{\pi_k} \}$}
+
+				\State $to_i^{\pi_k} \gets -1$
+
+				\If{$(i, q) \in E^M_{Q^{\pi_k}}$} 
+					\State $to_i^{\pi_k} \gets to_q$   \Comment{set the pallet destination.}
 				\EndIf
@@
 
-		\Return $E_{Q_{\pi_k}}, M$
+		\Return $E^M_{Q^{\pi_k}}, M$
 
@@
 
-$MinCGDeviation$\/ (line \ref{cons:minCG}) is run through a MIP solver with the equations \ref{eq:torque}, \ref{eq:embarked} and \ref {eq:one}, to relocate the packed contents on the pallets, minimizing torque and ensuring that they all remain on board, one packed content on each pallet.
-As there are few variables, the MIP solver returns an allocation $E_{Q_{\pi_k}}$\/ in less than 30 milliseconds.
-Finally, the destination of each pallet with a packed content is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).
-
 \vspace{3mm}
@@
 \begin{equation} \label{eq:torque}
-	\mbox{minimize}\ \Big | \sum_{i=1}^{m} \sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \times w_q \times D_i^{long}  \Big |
+	\mbox{minimize}\ \Big | \sum_{i=1}^{m} \sum_{q=1}^{m^{\pi_k}} Y^k_{iq} \times w_q \times D_i^{long}  \Big |
 \end{equation}
@@
 \begin{equation} \label{eq:embarked}
-	\sum_{i=1}^{m} Y^k_{iq} = 1;\ q \in \{1,\ldots,m_{\pi_k}\}
+	\sum_{i=1}^{m} Y^k_{iq} = 1;\ q \in \{1,2,\ldots,m^{\pi_k}\}
 \end{equation}
@@
 \begin{equation} \label{eq:one}
-	\sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \leq 1;\ i \in \{1,\ldots,m\}
-\end{equation}
-
-$SetPalletsDestinations$, which sets the pallets destination not yet defined, is described in Algorithm \ref{alg:dest}.
+	\sum_{q=1}^{m^{\pi_k}} Y^k_{iq} \leq 1;\ i \in \{1,2,\ldots,m\}
+\end{equation}
+
+$SetPalletsDestinations$, {\color{blue}which sets pallets' destinations not yet defined}, is described in Algorithm \ref{alg:dest}.
+
+The set $volumes$ stores the demand volume of items destined to the non-visited nodes ($L^{\pi_k}$) (line \ref{dest:vector1}).
+
+The destinations of each pallet are defined proportionally to the volume of items, regarding the pallets with a PC (lines \ref{dest:propor1}-\ref{dest:propor2}).
+
+The node with the maximum volume demand defines the destination of any remaining pallets (lines \ref{dest:max1}-\ref{dest:max2}).
 
 \begin{algorithm}[H]
-	\caption{Procedure to set pallets destination based on the items to be embarked on node $\pi_k$}  \label{alg:dest}
+	\caption{Procedure to set pallets destinations based on the items to be embarked on node $k$}  \label{alg:dest}
 	\begin{algorithmic}[1]
@@
 
-			\For{$x \in L_{\pi_k}$}		
-				\State $vol_x \gets  0$  \label{dest:vector1}
+			\State $volumes \gets \{ 0\ |\ L^{\pi_k} \}$  \label{dest:vector1}
+
+			\State $max \gets 0$ \label{dest:max} \Comment{destination with maximum volume demand.}
+
+			\State $total \gets 0$
+
+			\For{$j \in \{ 1,2,3,...,n^{\pi_k} \}$}
+
+				\If {$to_j \in L^{\pi_k}$}
+
+					\State $volumes_{to_j} \gets volumes_{to_j} + v_j$ 
+					\State $total \gets total + v_j$
+
+					\If {$volumes_{to_j} > volumes_{max}$}
+
+						\State $max \gets to_j$ 
+					\EndIf
+
+				\EndIf
+
 			\EndFor
 
-			\State $max \gets 0$
-
-			\State $total \gets 0$
-
-			\For{$j \gets 1$ to $n_{\pi_k}$}
-
-				\If {$to_j \in L_{\pi_k}$}
-
-					\State $vol_{to_j} \gets vol_{to_j} + v_j$ 
-					\State $total \gets total + v_j$
-
-					\If {$vol_{to_j} > vol_{max}$}
-
-						\State $max \gets to_j$ \label{dest:max} \Comment{$max$\/ is the destination with maximum volume demand}
-					\EndIf
-
-				\EndIf
-
-			\EndFor
-
-			\For{$x \in L_{\pi_k}$} \label{dest:propor1}
-
-				\If {$vol_{x} \neq 0$}
-
-					\State $needed \gets \max \{1, \lfloor{ (m-m_{\pi_k}) \times vol_{x}/total}\rfloor \}$  \Comment{Number of pallets to node $x$}
+			\For{$x \in L^{\pi_k}$} \label{dest:propor1}
+
+				\If {$volumes^{x} \neq 0$}
+
+					\State $needed \gets \max \{1, \lfloor{ m \times volumes_{x}/total}\rfloor \}$  \Comment{number of pallets needed}
 
@@
 
-					\For{$i \gets 1$ to $m$}
-
-						\If{$np < needed$ {\bf and} $T_i^{\pi_k} = -1$}
-							\State $T_i^{\pi_k} \gets x$
+					\For{$i \in \{1,2,3,...,m\}$}
+
+						\If{$np < needed$ {\bf and} $to_i^{\pi_k} = -1$}              \Comment{if the pallet destination in node $k$ is unset}
+							\State $to_i^{\pi_k} \gets x$
 							\State $np \gets np + 1$
@@
 
-			\For{$i \gets 1$ to $m$} \label{dest:max1}
-				\If{$T_i^{\pi_k} \gets -1$}  \Comment{Remaining pallets}
-					\State $T_i^{\pi_k} \gets max$
+			\For{$i \in \{1,2,3,...,m\}$} \label{dest:max1}
+				\If{$to_i^{\pi_k} \gets -1$}
+					\State $to_i^{\pi_k} \gets max$
 				\EndIf
@@
 
-$vol$\/ stores the demand volume of items destined to the non-visited nodes (line \ref{dest:vector1}).
-The destination of empty pallets is defined proportionally to the volume of items to be embarked (lines \ref{dest:propor1}-\ref{dest:propor2}).
-The destination with the maximum volume defines any remaining pallets (lines \ref{dest:max1}-\ref{dest:max2}).
-
 $ScoreAndDeviation$\/ is described in Algorithm \ref{alg:eval}, which evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.
+It consists of a loop that goes through all the pallets (lines \ref{eval:loop1}-\ref{eval:loop2}), accumulating the scores (lines \ref{eval:score1}-\ref{eval:score2} ) and the torques (lines \ref{eval:eps1}-\ref{eval:eps2}) of the shipped items, allowing the final calculation of the CG deviation (lines \ref{eval:torque1}-\ref{eval:torque2}).
 
 \begin{algorithm}[H]
-	\caption{Procedure to calculate the score and the relative torque deviation of graph $G$\/ in node $\pi_k$}  \label{alg:eval}
+	\caption{Procedure to calculate the solution score and the relative torque deviation in node $\pi_k$}  \label{alg:eval}
 	\begin{algorithmic}[1]
 
-		\Procedure{$ScoreAndDeviation$}{$\pi_k, G$}
-
-		\State Let $G(V_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$
+		\Procedure{$ScoreAndDeviation$}{$\pi_k,\ G$}
+
+		\State Let $G(M \cup N^{\pi_k} \cup Q^{\pi_k}, E^M_{Q^{\pi_k}} \cup E^M_{N^{\pi_k}})$
+
+		\State Let $X_{ij}^{\pi_k} = 1$ if $(i, j) \in E^M_{N^{\pi_k}}$, $1 \leq i \leq m$, $1 \leq j \leq n^{\pi_k}$
+		\State Let $Y_{iq}^{\pi_k} = 1$ if $(i, q) \in E^M_{Q^{\pi_k}}$, $1 \leq i \leq m$, $1 \leq q \leq m^{\pi_k}$
 
 		\State $s \gets 0$
-		\For{$i \gets 1$ to $m$}
-			\State $\tau_i \gets 0$
-		\EndFor
-		\For{$i \gets 1$ to $m$}	\label{eval:loop1}
-
-			\For{$j \gets 1$ to $n_{\pi_k}$}
+		\State $\tau_i \gets \{\ 0\ |\ [1,2,3,...,m]\}$
+
+		\For{$i \in \{1,2,3,...,m\}$}	\label{eval:loop1}
+
+			\For{$j \in \{1,2,3,...,n^{\pi_k}\}$}
 
@@
 
-			\For{$q \gets 1$ to $m_{\pi_k}$}
+			\For{$q \in \{1,2,3,...,m^{\pi_k}\}$}
 				\If{$Y_{iq}^{\pi_k} = 1$}
@@
 
-		\State $\tau \gets  \sum_{1}^{m} \tau_i / (W_{max} \times limit^{CG}_{long})$ \label{eval:torque}
-
-		\Return $s,\ \tau$
+		\State $W_{max} \gets \min(Payload,\ \sum_{1}^{m} W_i)$ \label{eval:torque1}
+
+		\State $\tau_{max} \gets W_{max} \times limit^{CG}_{long}$
+
+		\State $\epsilon \gets  \sum_{1}^{m} \tau_i / \tau_{max}$ \label{eval:torque2}
+
+		\Return $s,\ \epsilon$
 		\EndProcedure
@@
 
-This algorithm consists of a loop that goes through all the pallets (lines \ref{eval:loop1}-\ref{eval:loop2}), accumulating the scores (lines \ref{eval:score1} and \ref{eval:score2}) and the torques (lines \ref{eval:eps1} and \ref{eval:eps2}) of the shipped items, allowing the final calculation of the CG deviation (line \ref{eval:torque}).
-
 \subsection{Node-by-node solutions}
@@
 Concretely, all permutations between the nodes are tested, the set of nodes to be visited is updated, the PC that remain on board are reallocated to minimize the CG deviation, and the pallets destination is determined according to the volume of items available for shipment.
-In this way, the mathematical modeling for $SolveNode(MIP,\ \pi_k,\ G)$\/ becomes simpler, which finds an allocation of available items in node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $to_i^{\pi_k}$, and $a^{\pi_k}_q$.
-
-This simplified modeling is described in the equations \ref{eq:obj2} to \ref{eq:if2}, which maximizes the score of the $n_{\pi_k}$ items to be shipped, maintaining load balancing.
-
-The variables $X_{ij}$ and $Y_{iq}$ define the sets of edges $E_{N_{\pi_k}}$ and $E_{Q_{\pi_k}}$, respectively, which will be added to the graph $G$.
+In this way, the mathematical modeling for $SolveNode(MIP,\ \pi_k,\ G)$\/ becomes simpler, which finds an allocation of available items in node $\pi_k$\/ using previously defined values of $L^{\pi_k}$, $to_i^{\pi_k}$, and $a^{\pi_k}_q$.
+
+This simplified modeling is described in the equations \ref{eq:obj2} to \ref{eq:if2}, which maximizes the score of the $n^{\pi_k}$ items to be shipped, maintaining load balancing.
+
+The variables $X_{ij}$ and $Y_{iq}$ define the sets of edges $E^M_{N^{\pi_k}}$ and $E^M_{Q^{\pi_k}}$, respectively, which will be added to the graph $G$.
 
@@
 \begin{equation} \label{eq:newf}
-	\tilde{s} = \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij} \times s_j
+	\tilde{s} = \sum_{i=1}^{m} \sum_{j=1}^{n^{\pi_k}} X_{ij} \times s_j
 \end{equation}
@@
 \begin{equation} 
-	\tau^{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q) \Big ] \Big / W_{max} \times limit^{CG}_{long}
+	\tau^{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times (\sum_{j=1}^{n^{\pi_k}} X_{ij} \times w_j +  \sum_{q=1}^{m^{\pi_k}} Y_{iq} \times w_q) \Big ] 
+\end{equation}
+
+\begin{equation} 
+	\epsilon^{\pi_k} = \tau^{\pi_k} \Big / (W_{max} \times limit^{CG}_{long})
 \end{equation}
@@
 In Equation \ref{eq:costs2}, the first parcel denotes the fuel cost from the base until the last node, and the second the fuel cost in the last leg, from the last node back to the base.
-Note that $c_g$ is the fuel consumption potential increase due to the loaded aircraft CG deviation in the node ($\tau_{\pi_k}$).
+Note that $c_g$ is the fuel consumption potential increase due to the loaded aircraft CG deviation in the node ($\epsilon^{\pi_k}$).
 
 \begin{equation} \label{eq:costs2}
-	\tilde{c} = \sum_{k=0}^{K} \Big [ c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|) \Big ]
+	\tilde{c} = \sum_{k=0}^{K-1} \Big [ c^{\pi_k, \pi(k+1)}\times(1+c_g\times|\epsilon^{\pi_k}|) \Big ] + c^{\pi_K,0}\times(1+c_g\times|\epsilon^{\pi_K}|)
 \end{equation}
@@
 \begin{equation}
-	|\tau_{\pi_k}| \leq 1
+	|\tau^{\pi_k}| \leq W_{max} \times limit^{CG}_{long}
 \end{equation}
@@
 \begin{equation} 
-	\sum_{i=1}^{m} (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q ) \leq W_{max}
+	\sum_{i=1}^{m} (\sum_{j=1}^{n^{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m^{\pi_k}} Y_{iq} \times w_q ) \leq W_{max}
 \end{equation}
@@
 \begin{equation} 
-	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}
+	\sum_{j=1}^{n^{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m^{\pi_k}} Y_{iq} \times w_q  \leq W_i; \ i \in \{1, 2, \ldots, m\}
 \end{equation}
@@
 \begin{equation} 
-	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}
+	\sum_{j=1}^{n^{\pi_k}} X_{ij} \times v_j + \sum_{q=1}^{m^{\pi_k}} Y_{iq} \times v_q  \leq\ V_i; \ i \in \{1, 2, \ldots, m\}
 \end{equation}
@@
 \begin{equation} 
-	\sum_{i=1}^{m} X_{ij} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}
+	\sum_{i=1}^{m} X_{ij} \leq 1; \ j \in \{1, 2, \ldots, n^{\pi_k}\}
 \end{equation}
@@
 \begin{equation} \label{eq:if1}
-	to_i^{\pi_k} = to_j;\ X_{ij} = 1; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
+	to_i^{\pi_k} = to_j;\ X_{ij} = 1; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi_k}\}
 \end{equation}
@@
 \begin{equation} \label{eq:if2}
-	X_{ij}       = 0;\ to_j \notin L_{\pi_k}; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
+	X_{ij}       = 0;\ to_j \notin L^{\pi_k}; \ i \in \{1, 2, \ldots, m\}; \ j \in \{1, 2, \ldots, n^{\pi_k}\}
 \end{equation}
@@
 
-The selection of edges for $E_{ N_{\pi_k} }$ uses the {\it edge attractiveness}\/ $\theta^{\pi_k}_{ij}$\/ \ref{eq:edge}, which can be understood as the tendency to allocate the item $j$ to pallet $i$.
+The selection of edges for $E_M^{ N^{\pi_k} }$ uses the {\it edge attractiveness}\/ $\theta^{\pi_k}_{ij}$\/ \ref{eq:edge}, which can be understood as the tendency to allocate the item $j$ to pallet $i$.
 It is directly proportional to the score, and inversely proportional to the volume and the torque of each item.
@@
 \begin{equation} \label{eq:edge}
-	\theta^{\pi_k}_{ij} = \frac{ s_j }{ v_j }\ \times \Big ( 1 - \frac{ w_j \times |D_i^{long}| }{ w_{max} \times\ | D_{max}^{long} | } \Big );\ i \in \{1,\ldots,m\},\ j \in \{1,\ldots,n_{\pi_k}\}
+	\theta^{\pi_k}_{ij} = \frac{ s_j }{ v_j }\ \times \Big ( 1 - \frac{ w_j \times |D_i^{long}| }{ w_{max} \times\ | D_{max}^{long} | } \Big );\ i \in \{1,2,\ldots,m\},\ j \in \{1,2,\ldots,n^{\pi_k}\}
 \end{equation}
@@
 
-		\captionof{figure}{$n_{\pi_k}$\/ edges $e_{ij}^{\pi_k}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi_k}$\/ in non-ascending order}
+		\captionof{figure}{$n^{\pi_k}$\/ edges $e_{ij}^{\pi_k}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi_k}$\/ in non-ascending order}
 		\label{fig:whip}		
@@
 
-Figure \ref{fig:whip} represents the $n_{\pi_k}$\/ possible edges $e_{ij}^{\pi_k}$\/ of $i$\/ sorted by $\theta_{ij}^{\pi_k}$.
+Figure \ref{fig:whip} represents the $n^{\pi_k}$\/ possible edges $e_{ij}^{\pi_k}$\/ of $i$\/ sorted by $\theta_{ij}^{\pi_k}$.
 \emph{Shims}\/ starts with a greedy solution, stopping at the edge with index $\eta_1^i$ (first phase).
@@
 In this heuristic, described in Algorithm \ref{alg:shims}, pallets $i$\/ are considered in non-descending order of $|D_i^{long}|$ (line \ref{shims:pallets}).
-For each pallet, the $n_{\pi_k}$\/ possible edges $e_{ ij}^{\pi_k}$\/ are considered in non-increasing order of $\theta_{ij}^{\pi_k}$\/ (line \ref{shims:possible}) in three phases: the greedy phase (until line \ref{shims:endfirst}), the composition phase of the shims (lines \ref{shims:beginsecond}-\ref{shims:endsecond}) and the selection phase of the best shim (lines \ref{shims:beginthird}-\ref{shims:endthird}).
+For each pallet, the $n^{\pi_k}$\/ possible edges $e_{ ij}^{\pi_k}$\/ are considered in non-increasing order of $\theta_{ij}^{\pi_k}$\/ (line \ref{shims:possible}) in three phases: the greedy phase (until line \ref{shims:endfirst}), the composition phase of the shims (lines \ref{shims:beginsecond}-\ref{shims:endsecond}) and the selection phase of the best shim (lines \ref{shims:beginthird}-\ref{shims:endthird}).
 
 {\color{blue}
-	In the beginning of the first phase (line \ref{shims:initQ}), the set $Q_{\pi_k}$ represents the PC whose destination is not $\pi_k$, the current node, thus remaining on board.
-
-	It is important to remember that $Q_{\pi_k}$ and $M$ were modified by the procedures $UpdatePacked(M, Q_{\pi_k}, \pi_k)$ and $SetPalletsDestinations(M, \pi_k)$, having their positions and destinations reassigned.
+	In the beginning of the first phase (line \ref{shims:initQ}), the set $Q^{\pi_k}$ represents the PC whose destination is not $\pi_k$, the current node, thus remaining on board.
+
+	It is important to remember that $Q^{\pi_k}$ and $M$ were modified by the procedures $UpdatePacked(M, Q^{\pi_k}, \pi_k)$ and $SetPalletsDestinations(M, \pi_k)$, having their positions and destinations reassigned.
 
@@
 Initially, two shims are found: $sh_w$\/ with larger weight and $sh_v$\/ with larger volume.
-Among the two, the {\it Shim} with the highest score will be chosen and its edges are inserted into $E_{N_{\pi_k}}$.
+Among the two, the {\it Shim} with the highest score will be chosen and its edges are inserted into $E^M_{N^{\pi_k}}$.
 }
@@
 
-		\Procedure{$SolveNode$}{$Shims,\ \pi_k,\ G,\ time$}
-
-		\State $T_{begin} \gets$ current system time
-
-		\State $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$ \Comment{a graph with pallets positions ($M$), PC ($Q_{\pi_k}$), and edges ($E_{Q_{\pi_k}}$)}  \label{shims:initQ}
+		\Procedure{$SolveNode$}{$Shims,\ \pi_k,\ G$}
+
+		\State $G(M \cup N^{\pi_k} \cup Q^{\pi_k}, E^M_{Q^{\pi_k}})$ \Comment{a graph with pallets positions ($M$), PC ($Q^{\pi_k}$), and edges ($E^M_{Q^{\pi_k}}$)}  \label{shims:initQ}
 
@@
 
-		\State $E_{N_{\pi_k}} \gets \{\ \}$  \Comment{an empty set of pallet-item edges}
-
-		\State Let $E_{ij}$ be an array with $m \times n_{\pi_k}$ edges $(i,\ j)$ sorted by $\theta_{ij}^{\pi_k}$ in non-ascending order \label{shims:possible}
+		\State $E^M_{N^{\pi_k}} \gets \{\ \}$  \Comment{an empty set of pallet-item edges}
+
+		\State Let $E_{ij}$ be an array with $m \times n^{\pi_k}$ edges $(i,\ j)$ sorted by $\theta_{ij}^{\pi_k}$ in non-ascending order \label{shims:possible}
 
@@
 
-		\State $volume \gets \{0\ |\ [1,\ldots,m] \}$
-		\State $slack  \gets \{0\ |\ [1,\ldots,m] \}$
+		\State $volume \gets \{0\ |\ [1,2,3,...,m] \}$
+		\State $slack  \gets \{0\ |\ [1,2,3,...,m] \}$
 
@@
 
-		\For{$i \in \{1,\ldots,m\}$}
-
-			\For{$q \in \{1,\ldots,m_{\pi_k}\}$}
-
-				\If{$(i, q) \in E_{Q_{\pi_k}}$} 
+		\For{$i \in \{1,2,3,...,m\}$}
+
+			\For{$q \in \{1,2,3,...,m^{\pi_k}\}$}
+
+				\If{$(i, q) \in E^M_{Q^{\pi_k}}$} 
 					\State $volume_i \gets volume_i + v_q$ 
@@
 
-				\If{ ($E_{N_{\pi_k}} \cup \{e_{ij}\}$ is feasible) {\bf and} ($volume_i \leq V_i \times limit$) {\bf and} ($\tau_{new} \leq \tau_{max}$ ) }
-
-					\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup \{e_{ij}\}$
+				\If{ ($E^M_{N^{\pi_k}} \cup \{e_{ij}\}$ is feasible) {\bf and} ($volume_i \leq V_i \times limit$) {\bf and} ($\tau_{new} \leq \tau_{max}$ )}
+
+					\State $E^M_{N^{\pi_k}} \gets E^M_{N^{\pi_k}} \cup \{e_{ij}\}$
 
@@
 
-			\Until{($\eta_2^{\ i} = n_{\pi_k}$) {\bf or} ($ volume_i \geq (1 + 3 \times (1-limit)) \times V_i)$)} \label{shims:endsecond} \Comment{end of the second phase}
+			\Until{($\eta_2^{\ i} = n^{\pi_k}$) {\bf or} ($ volume_i \geq (1 + 3 \times (1-limit)) \times V_i)$)} \label{shims:endsecond} \Comment{end of the second phase}
 
@@
 
-			\For{$x \in \{\eta_1^{\ i},\ldots,\eta_2^{\ i} \}$} \label{edges_indexes}
-
-			{\color{blue}
-
-				\If{ current system time - $T_{begin} > time$}
-				\State {\bf break}
-				\EndIf
-
-			}
+			\For{$x \in \{\eta_1^{\ i},...,\eta_2^{\ i} \}$} \label{edges_indexes}
 
@@
 
-					\If { $e_{ij} \not\in (E_{N_{\pi_k}} \cup shims)$ {\bf and} $e_{ij}$ is feasible  {\bf and}  $(v_j + volume) \leq slack_i$}
+					\If { $e_{ij} \not\in (E^M_{N^{\pi_k}} \cup shims)$ {\bf and} $e_{ij}$ is feasible  {\bf and}  $(v_j + volume) \leq slack_i$}
 
@@
 
-			\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup sh_{best}$ 	\label{shims:endthird}   \Comment{end of the third phase}
+			\State $E^M_{N^{\pi_k}} \gets E^M_{N^{\pi_k}} \cup sh_{best}$ 	\label{shims:endthird}   \Comment{end of the third phase}
 
@@
 
-		\Return $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$
+		\Return $G(M \cup N^{\pi_k} \cup Q^{\pi_k}, E^M_{Q^{\pi_k}} \cup E^M_{N^{\pi_k}})$
 
@@
 
-		\Procedure{$ItemsGeneration$}{$scenario, surplus$}
+		\Procedure{$ItemsGeneration$}{$scenario,\ volume$}
 
 		\State Let $L$ and $M$ be according to $scenario$ \label{ig:LM}
-		\State $limit \gets surplus \times \sum_{i=1}^{m} V_i$ \label{ig:extended}
+		\State $limit \gets volume \times \sum_{i=1}^{m} V_i$ \label{ig:extended}
 
@@
 
-		\For{$k \in \{0,\ldots,K\}$}
-
-			\State $N_k \gets \varnothing$
+		\For{$k \in \{0,1,2,...,K\}$}
+
+			\State $N_k \gets \{\ \}$
 			\State $j \gets 0$
@@
 $ItemsGeneration$, which generates $N$, is described in Algorithm \ref{alg:itemsgen}.
-The parameter $scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and the parameter $surplus$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}).
-To avoid simply loading all items, we use $surplus > 1$ {\color{blue}(1.2, 1.5, and 2.0).
+The parameter $scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and the parameter $volume$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}).
+To avoid simply loading all items, we use $volume > 1$ {\color{blue}(1.2, 1.5, and 2.0).
 This represents more test instances for each scenario.}
@@
 We ran Algorithm \ref{alg:main} in the 5 scenarios described in Table \ref{tab:scenarios}, considering 6 methods for node-by-node solution: Gurobi (see \ref{solver}), ACO, NMO, TS, GRASP, and Shims (Algorithm \ref{alg:shims}).
-In generating the items in each node, we consider 3 values for the parameter $surplus$.
-The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:20} ($surplus = 1.2$), \ref{tab:50} ($surplus = 1.5$) and \ref{tab:100} ($surplus = 2.0$).
-For each $method$, $scenario$\/ and $surplus$, 7 different instances were generated.
+In generating the items in each node, we consider 3 values for the parameter $volume$.
+The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:20} ($volume = 1.2$), \ref{tab:50} ($volume = 1.5$) and \ref{tab:100} ($volume = 2.0$).
+For each $method$, $scenario$\/ and $volume$, 7 different instances were generated.
 Therefore, 105 tests (5 scenarios $\times$\/ 3 volumes $\times$\/ 7 instances) were performed for each method.
@@
 
-In each $scenario$\/, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.
-
-{\color{red}
-
-As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's RAM (16 GiB) too early, we decided to execute a relaxed model with the decision variables set to CONTINUOUS (a Gurobi parameter) to obtain the upper bound for the objective function.
-This approach is much faster than using binary decision variables.
-
-The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.
-
-We solved all scenarios and instances with Gurobi in a relaxed linear model without a time limit.
-Although the solutions are not guaranteed to be feasible, it is good for the reader to have an idea of the possible upper bound for the metaheuristic solutions.
-
-Note that, in Table \ref{tab:20} all $Gurobi\ f$\/ values are italicized to indicate that these values may not be optimal or feasible as the decision variables are not integer.
-
-}
+In each $scenario$, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.
 
@@
 	\centering
-\caption{Methods and scenarios with $surplus =1.2$}  \label{tab:20}
+	\caption{Methods and scenarios with $volume =1.2$}  \label{tab:20}
 	\footnotesize
@@
 		\toprule
-{$method$}      & {\bf Results}&{\bf 1}&{\bf 2} & {\bf 3}      & {\bf 4}      & {\bf 5}  & \specialcell{{\bf Normalized}\\{\bf Speed-up}}  \\
+		{$method$}                        & {\bf Results}    & {\bf 1}     &  {\bf 2}    & {\bf 3}      & {\bf 4}      & {\bf 5}  & \specialcell{{\bf Normalized}\\{\bf Speed-up}}  % 14.40	 184
+		\\
+		\toprule
+		\multirow{2}{*}{\textbf{Gurobi}}  & $f$      & 8.03   & {\bf 12.02}& {\bf 13.15} & {\bf 14.40} & {\bf 0.00} & {\bf 0.997} \\ %51,84
+		&{\bf run time (s)}                          & 14     & 20         & 73          & 184         & 1,044       &  1.000 \\ 
 		\midrule[.1pt]
-\multirow{2}{*}{{\it Gurobi}}     & Upper bound for $f$      & {\it 7.91} & {\it 11.92} & {\it 13.12} & {\it 14.29} & {\it 13.34} & -  \\ %51,84
-&{\bf run time (s)}                          & 14         & 19          & 73          & 188         & 1,097      & -  \\ 
-\toprule
-\multirow{2}{*}{\textbf{TS}}      & $f$      & 5.37       &             &             &             &            &   \\ % 5.49	 901
-&{\bf run time (s)}                          & 659        &             &             &             &            &   \\ 
+		\multirow{2}{*}{\textbf{TS}}      & $f$      & 6.56     & 8.49       & 9.19       & 11.25      & 11.48      &  0.906 \\
+		&{\bf run time (s)}                          & 4       & 17        & 86        & 516       & 3,600     &  2.709 \\ 
 		\midrule[.1pt]		
-\multirow{2}{*}{\textbf{GRASP}}   & $f$      &            &             &             &             &            &   \\ 
-&{\bf run time (s)}                          &            &             &             &             &            &   \\ 
+		\multirow{2}{*}{\textbf{GRASP}}   & $f$      & 6.76     & 8.85       & 9.65       & 11.75      & 11.54      &  0.937 \\ 
+		&{\bf run time (s)}                          & 4       & 17        & 80        & 472       & 3,265     &  2.984 \\ 
 		\midrule[.1pt]
-\multirow{2}{*}{\textbf{ACO}}     & $f$      & 5.25       & 5.75        & 5.06        & 4.84        &  4.20      &   \\ % 4.20	 1133
-&{\bf run time (s)}                          & 313        & 326         & 303         & 256         &  1,133     &   \\ 
+		\multirow{2}{*}{\textbf{ACO}}     & $f$      & {\bf 6.77}& 8.85      & 9.65       & 11.75      & 11.54      &  0.937 \\ 
+		&{\bf run time (s)}                          & 4        & 17       & 86        & 513       & 3,596     &  2.714 \\ 
 		\midrule[.1pt]	
-\multirow{2}{*}{\textbf{NMO}}     & $f$      &            &             &             &             &            &   \\
-&{\bf run time (s)}                          &            &             &             &             &            &   \\ 
+		\multirow{2}{*}{\textbf{NMO}}     & $f$      & 6.74      & 8.80      & 9.61       & 11.76      & 11.50      &  0.934 \\
+		&{\bf run time (s)}                          & 4        & 17       & 86        & 508       & 3,480     &  2.796 \\ 
 		\midrule[.1pt]	
-\multirow{2}{*}{\textbf{Shims}}   & $f$      & 7.91       & 11.55       & 12.62       &  13.82      &  14.85          &   \\ % 14.85	 563
-&{\bf run time (s)}                          & 4          &  8          &  33         &  94         &  563          &   \\
-
+		\multirow{2}{*}{\textbf{Shims}}   & $f$      & {\bf 6.77} & 8.92     & 9.60       & 11.78      & 11.61      &  0.939       \\ 
+		&{\bf run time (s)}                          & 2         & 8       & 31        & 166       & 1,087     &  {\bf 8.856} \\ 
 		\bottomrule
@@
 	\centering
-	\caption{Methods and scenarios with $surplus =1.5$}  \label{tab:50}
+	\caption{Methods and scenarios with $volume =1.5$}  \label{tab:50}
 	\footnotesize
@@
 	\centering
-	\caption{Methods and scenarios with $surplus = 2.0$}  \label{tab:100}
+	\caption{Methods and scenarios with $volume = 2.0$}  \label{tab:100}
 	\footnotesize
@@
 
-Our main contributions were the mathematical modeling of this complex problem, which involves four NP-hard sub-problems, and a complete process that offers a fast and good-quality solution.
+Our main contributions were the mathematical modeling of this complex problem, which involves four NP-hard subproblems, and a complete process that offers a fast and good-quality solution.
 
@@
 
-The simple heuristic $Shims$\/ presented a solution with the best node-by-node results in a restricted time interval.
+The simple heuristic $Shims$\/ presented a resolution with the best node-by-node results in a restricted time interval.
 Without it, it would not be possible to test all routes: after all, the greater the number of nodes, the shorter the run time limit for node-by-node solution.
@@
 
-Finally, by focusing on node-by-node solution, the method of this work is not exclusive to aircraft and airports: it can be adapted, for example, to ships and ports, or vehicles and warehouses, or freight cars and railways.
+Finally, by focusing on node-by-node solution, the resolution of this work is not exclusive to aircraft and airports: it can be adapted, for example, to ships and ports, or vehicles and warehouses, or freight cars and railways.
 In these cases, it would be necessary to make some changes in the modeling: for example, modify the load balancing constraints, and consider the available space in vehicles or freight cars instead of pallets.
@@
 
-\bibitem{MongeauBes2003} M. Mongeau and C. Bes, Optimization of aircraft container loading, {\it IEEE Transaction on Aerospace and Electronic Systems}, Vol. 39 (1), pp. 140-150, 2003.
-
-\bibitem{fok2004optimizing} M.K.K.
-Fok and A.H.W.
-Chun, Optimizing air cargo load planning and analysis, Proceedings of {\it The International Conference on Computing, Communications and Control Technologies}, 2004.
+\bibitem{MongeauBes2003} Marcel Mongeau and Christian Bes, Optimization of aircraft container loading, {\it IEEE Transaction on Aerospace and Electronic Systems}, Vol. 39 (1), pp. 140-150, 2003.
+
+\bibitem{fok2004optimizing} Kelly Fok and Andy Chun, Optimizing air cargo load planning and analysis, Proceedings of {\it The International Conference on Computing, Communications and Control Technologies}, 2004.
 
@@
 \end{document}
-
-{\color{blue}
-	Until scenario 4, solving all tours and selecting the best is fast enough for operational use.
-For scenarios 4 and 5 (which are not typical in the Brazilian Air Force), we plan bounding schemes to allow for the discarding of most of the higher-cost tours.
-This is expected to reduce the running time for scenarios 4 and 5, favoring the discovery of better search mechanisms as well as further applications that may require more nodes per tour.
-}
-
-{\color{blue}
-
-	Equations \ref{eq18} and \ref{eq19} were set to group items with the same destinations.
-But, as they mean a combination of $n_{\pi_k}$ items 2 by 2, this may easily lead to huge numbers, $60,000$ to $320,000$ possible combinations, due to the number of items in each node, which together with the other constraints make a MIP solution very hard.
-
-	To work around this, we decided to create the destination attribute for the pallets ($TO_i^{\pi_k}$), to reduce the number of combinations and enable the solution using heuristics.
-$TO^{\pi_k}_i$ denotes that pallet $i$ may assume a different destination in each node ${\pi_k}$.
-
-	On subsequent nodes, the PC can be allocated with other items of same destination \ref{eq:cons5}.
-
-	\begin{equation} \label{eq:cons5}
-		TO_i^{\pi_{k+1}} = to_j;\ X^{\pi_k}_{ij} = 1  \ \mbox{and}\ to_j \in L_{\pi_{k+1}};  \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K-1\}
-	\end{equation}
-
-	\vspace{3mm}
-	At each node, an item (\ref{eq:pdp8}) and the already packed contents (\ref{eq:pdp2}) must only be allocated to a pallet if the destinations are the same.
-
-	\begin{equation} \label{eq:pdp8}
-		TO_i^{\pi_k} = to_j;\ X_{ij}^{\pi_k} = 1; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-	\end{equation}
-
-	\begin{equation} \label{eq:pdp2}
-		TO_i^{\pi_k} = to_q;\ Y_{iq}^{\pi_k} = 1; \ i \in \{1, \ldots, m\};\ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-	\end{equation}
-
-}
-
-{\color{blue}Another possibility to make this approach more general is to devise some bounding schemes to allow discarding most higher-cost tours and allowing an increase in the number of nodes without compromising performance.}

### 2023-06-03 CAIE/results.tex
CAIE paper transferred to parallel folder

--- 2023-06-03 CAIE/paper.tex
@@
 
-\begin{frontmatter}
-
-\title{Air cargo load and route planning in pickup and delivery operations}
-
-\author{A.C.P.~Mesquita}
-\ead{celio@ita.br}
-
-\author{C.A.A.~Sanches\corref{cor1}}
-\ead{alonso@ita.br}
-\cortext[cor1]{Corresponding author.}
-
-\address {Instituto Tecnol\'{o}gico de Aeron\'{a}utica - DCTA/ITA/IEC\\
-Pra\c{c}a Mal. Eduardo Gomes, 50\\
-S\~{a}o Jos\'{e} dos Campos - SP - 12.228-900 - Brazil}
-
-\begin{abstract}
-
-In aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading, for fast take-off, and mission accomplishment, especially in times of crisis, disaster relief missions, short deadlines for contracting customers, or any external pressure for immediate take-off.
-In addition, there are no commercially available systems that can assist the load and trip planners with the pallet building and aircraft loading and balancing demands for transport at each hub.
-This enables other risks, such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
-We defined and solved the problem of planning the loading and routing of a single aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
-This NP-hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modeled using standardized pallets in fixed positions, obeying center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
-We performed multiple experiments with a commercial solver and four well-known meta-heuristics on data based on the transport history of the {\it Brazilian Air Force}.
-We also designed a new heuristic that quickly finds good solutions for a wide range of problem sizes: an essential contribution, which solved all test scenarios within an acceptable operational run time.
-
-\end{abstract}
-
-\begin{keyword}
-Load Planning \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Vehicle Routing
-\end{keyword}
-
-\end{frontmatter}
-
-\label{sec1}
-\section{Introduction}
-
-Air cargo transport involves several sub-problems that are difficult to solve.
-Recently, Brandt and Nickel \cite[p.
-401]{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP) and {\it Weight and Balance Problem} (WBP).
-Several aspects were considered in this modeling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc.
-They also presented a comprehensive bibliographic survey of solving methods that had been developed in different situations.
-
-However, there are still other important challenges in air cargo transport that go beyond the definition of the ACLPP, especially with regard to the flight itinerary and the loading and unloading at each destination (or node) of this travel plan.
-In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Pickup and Delivery Problem} (PDP), and the search for the lowest cost route, which is the well-known {\it Traveling Salesman Problem} (TSP).
-
-Considering air cargo transport, Table \ref{tab:sa} lists the main works in the literature and the corresponding sub-problems addressed.
-We also indicate whether the dimensions of the items were taken into account ({\bf 3D} or {\bf 2D}) and which solution method was used: heuristic search methods ({\bf Heu}), integer programming ({\bf Int}), or linear programming ({\bf Lin}).
-
-\begin{table}[H]
-	\centering
-	\caption{Air cargo transport: literature, problems and features}  \label{tab:sa}
-	\scriptsize
-	\renewcommand{\arraystretch}{1.1} 
-	\begin{tabular}{r|cccc|ccccc}
-		\toprule
-		 Work & {\bf APP}  & {\bf WBP}  &  {\bf PDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} \\
-		\midrule
-		Larsen and Mikkelsen \cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$ \\
-		Brosh \cite{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$ \\
-		Ng \cite{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ \\
-		Heidelberg {\it et al.} \cite{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$ \\
-		Mongeau and Bes \cite{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		Fok and Chun \cite{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
-		Chan {\it et al.} \cite{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$ \\
-		Kaluzny and Shaw \cite{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$ \\
-		Verstichel {\it et al.} \cite{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		Mesquita and Cunha \cite{MesquitaCunha2011}   & $.$ & $.$    & $\bigstar$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\		
-		Limbourg {\it et al.} \cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		Roesener and Hall \cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
-		Vancroonenburg {\it et al.} \cite{Vancroonenburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		Lurkin and Schyns \cite{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		Roesener and Barnes \cite{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
-		Paquay {\it et al} \cite{PaquaySchynsLimbourg2016,PaquayLimbourgSchynsOliveira2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$ \\
-		Chenguang {\it el al.} \cite{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$ \\
-		Wong and Ling \cite{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$  \\
-		Wong {\it et al.} \cite{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$  \\
-		Zhao {\it et al.} \cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
-		{\bf This article}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$  \\
-		\bottomrule 
-	\end{tabular}
-	\normalsize 
-\end{table}
-
-As can be seen, so far Lurkin and Schyns \cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem.
-Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a predefined tour plan restricted to two legs.
-It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem.
-Referring directly to this work, Brandt and Nickel \cite[p.
-409]{BrandtStefan2019} comment: {\it However, not even these sub-problems are acceptably solved for real-world problem sizes or the models omit some practically relevant constraints}.
-
-There are real situations that are much more complex.
-In this work, we consider a practical case in Brazil, which is the largest economy in Latin America.
-Due to its dimensions, this country has the largest air market on the continent with $2,499$\/ registered airports, of which $1,911$\/ are private and $588$\/ are public.
-Although it is an immense distribution network, the {\it Brazilian Air Force}\/ missions have always considered 3 to 5 nodes per flight plan.
-
-It is important to emphasize that this data is not an imposed limitation, but a historical fact that we will explore in our solution method.
-Throughout this work we approach routes with up to 7 nodes, as can be seen in Table \ref{tab:costs} and in Figure \ref{fig:nodes}.
-Although there are many other airports of interest to {\it Brazilian Air Force}, these 7 nodes were chosen due to their high demand and short transport times.
-The other Brazilian airports tend to have a lower demand, which is usually met, in a less expensive way, by cabotage, rail or road transport.
-
-\begin{table}[H]
-
-	\begin{minipage}{0.05\linewidth}
-
-	\end{minipage}\hfill 
-	\begin{minipage}{0.45\linewidth}
-
-		\caption{Brazilian airports distances ($km$)}  \label{tab:costs}
-		\centering
-		\footnotesize
-		\newcolumntype{Y}{>{\centering\arraybackslash}p{0.09\textwidth}}
-		\newcolumntype{X}{>{\centering}p{0.09\textwidth}}
-
-		\begin{tabular}{X X X X X X X Y}
-			\toprule
-			Node & $l_0$ & $l_1$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ & $l_6$ \\
-			IATA*   & GRU   & GIG   & SSA   & CNF   & CWB   & BSB   & REC \\	
-			\midrule	
-			GRU     & 0	    &343	&1,439   &504    &358    &866    &2,114\\
-			GIG	    & 343	&0	    &1,218   &371    &677    &935    &1,876\\
-			SSA	    & 1,439	&1,218	&0	    &938    &1,788   &1,062   &676\\
-			CNF	    & 504	&371	&938	&0	    &851    &606    &1,613\\
-			CWB	    & 358	&677	&1,788	&851	&0	    &1,084   &2,462\\
-			BSB	    & 866	&935	&1,062	&606	&1,084	&0	    &1,658\\
-			REC	    & 2,114	&1,876	&676	&1,613	&2,462	&1,658	&0\\
-			\bottomrule
-			\multicolumn{8}{c}{*International Air Transport Association}\\
-			\multicolumn{8}{c}{\small\textsuperscript{Source: www.airportdistancecalculator.com}}\\
-		\end{tabular}
-		\normalsize
-
-	\end{minipage}\hfill 
-	\begin{minipage}{0.50\linewidth}
-		\centering
-		\includegraphics[scale=0.25]{Images/nodes.png}
-		\captionof{figure}{A route with 7 Brazilian airports}
-		\label{fig:nodes}		
-	\end{minipage}
-\end{table}
-
-Cargo handling at modern airports usually involves powerful equipment such as dollies, motorized conveyors, elevator transfer vehicles, caster decking, scissor lifts, electrical controls, hydraulically adjustable heights, forklift trucks, and plenty of room in the airport cargo terminal (see Figure \ref{fig:handling}).
-These facilities, together with the use of standardized pallets with predefined positions on the aircraft, allow loading and unloading at each node to be carried out in about 30 minutes.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=0.2]{Images/handling.png}
-	\captionof{figure}{Cargo handling equipment}
-	\label{fig:handling}
-	\small\textsuperscript{Source: NARA \& DVIDS Public Domain Archive}	
-\end{figure}
-
-However, as reported by Fok and Chun, {\it load planning (...) is usually done in roughly 2 hours before departure, when all the details of the cargo are present.} \cite{fok2004optimizing}.
-We may infer that this is because there is no commercial software to solve this problem.
-In addition, there is still the necessary time to plan the route, in order to minimize fuel consumption and ensure the transport of priority items.
-In these circumstances, it is of great importance to find a method that speeds up the load planning and the definition of the flight itinerary.
-
-Our work proposes a method that prioritizes the transport of the most relevant items at each node and the fuel economy along the tour.
-We developed a heuristic that can be run on a simple handheld computer (such as a laptop or a tablet) that previously defines a flight plan, and provides a quick solution for cargo handling.
-Consequently, this method reduces the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets (regarding their positions), and a pick up and delivery plan to each node.
-
-To the best of our knowledge, this is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP and TSP has been addressed.
-This new problem is named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem} (ACLP+RPDP).
-
-This article is organized into six more sections.
-In Section \ref{sec2}, we give a brief review of the literature.
-In Section \ref{sec3}, we present the problem context and assumptions, and in Section \ref{sec4}, we describe the mathematical model and how we dealt with its issues.
-In Section \ref{sec5}, we describe the elaborate algorithms, whose results are presented in Section \ref{sec6}.
-Finally, our conclusions are in Section \ref{sec7}.
-
-\section{Related literature}
-\label{sec2}
-
-In this section, we briefly describe the characteristics of the main works related to air cargo transport, following the chronological order of Table \ref{tab:sa}.
-
-Larsen and Mikkelsen \cite{LarsenMikkelsen1979} developed an interactive procedure for loading 14 types of Boeing 747 into a two-leg transportation plan.
-Seven types of items were considered to be allocated in 17 to 42 positions.
-With non-linear programming and heuristics, they present a solution that minimizes positioning changes in the intermediate node, optimizing the load balancing in the aircraft.
-
-Brosh \cite{Brosh1981} addressed the problem of planning the allocation of cargo on an aircraft.
-Considering volume, weight and structural constraints, the author finds the optimal load layout through a fractional programming problem.
-
-Ng \cite{Kevin1992} developed a multi-criteria optimization approach to load the {\it C-130} aircraft of the {\it Canadian Air Force}.
-Based on integer programming, this model provides timely planning and improves airlift support for combat operations, solving WBP with pallets in fixed positions, and considering 20 different items.
-
-Heidelberg {\it et al.} \cite{Heidelberg1998} developed a heuristic for 2D packing in air loading, comparing it with methods for solving the {\it Bin Packing Problem}.
-These authors conclude that the classical algorithms are inadequate in this context, because they ignore the aircraft balancing constraints.
-
-Mongeau and Bes \cite{MongeauBes2003} presented a method based on linear integer programming to solve the problem of choosing and positioning containers in the {\it Airbus 340-300}.
-Safety and stability constraints were considered, with the objective of minimizing fuel consumption.
-
-Fok and Chun \cite{fok2004optimizing} developed a web-based application to make efficient use of space and load balancing for an air cargo company.
-Based on an analysis of historical data, an operational load planning with mathematical optimization is obtained.
-This container load planning is usually done in roughly 2 hours before departure, when all cargo details are in place.
-
-Chan {\it et al.} \cite{Chan2006} carried out a case study with heterogeneous pallets.
-To minimize the total cost of shipping, they developed a 3D packing heuristic, with a loading plan for each pallet.
-Although the authors do not consider load balancing or positioning of pallets in the cargo hold, this method is relevant in commercial and industrial applications, where cargo items tend to be less dense.
-
-Kaluzny and Shaw \cite{KaluznyBohdanL2009Oalb} developed a mixed integer linear programming model to arrange a set of items in a military context that optimizes the load balance.
-The objective function can be chosen to minimize the deviation of the center of gravity (CG) from the target position or to maximize the function of the items loaded.
-This approach does not palletize items, which are arranged in the cargo bay by a 2D packing procedure.
-
-Verstichel {\it et al.} \cite{Verstichel2011} solved WBP by selecting the most profitable subset of containers to be loaded into an aircraft by the use of mixed-integer programming.
-Experimental results on real-life data showed significant improvements compared to those obtained manually by an experienced planner.
-
-Mesquita and Cunha \cite{MesquitaCunha2011} presented a heuristic for a real problem of the {\it Brazilian Air Force}, which consists of defining transport routes with simultaneous collection and delivery from a central distribution terminal.
-
-Limbourg {\it et al.} \cite{Limbourg2012} developed a mixed-integer program for optimally rearranging a set of pallets into a compartmentalized cargo aircraft, specifically the {\it Boeing 747}.
-
-Roesener and Hall \cite{RoesenerHall2014} solved APP and WBP as an integer programming problem, where the items are selected for shipment based on a utility score, then assigned to pallets, which will be loaded into an aircraft in a specific pallet position.
-The pallets are then packed in a manner to optimize both the pallet and aircraft characteristics, such as item utility, pallet occupancy, and aircraft center of balance.
-This method groups items by destinations to be served by multiple aircraft.
-
-Vancroonenburg {\it et al} \cite{Vancroonenburg2014} presented a mixed integer linear programming model that selects the most profitable pallets, satisfying safety and load balancing constraints on the {\it Boeing 747-400}.
-Using a solver, these authors solved real problems in less than an hour.
-
-As already mentioned, Lurkin and Schyns \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and PDP in air cargo transport.
-The authors demonstrated that this problem is NP-hard and performed some experiments with real data, noting that their model offers better results than those obtained manually.
-
-Roesener and Barnes \cite{RoesenerBarnes2016} proposed a heuristic to solve the {\it Dynamic Airlift Loading Problem} (DALP).
-Given a set of palletized cargo items that require transport between two nodes in a time frame, the objective of this problem is to select an efficient subset of aircraft, partition the pallets into aircraft loads and assign them to allowable positions in those aircraft.
-
-Paquay {\it et al.} \cite{PaquaySchynsLimbourg2016} presented a mathematical modeling to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
-Its objective is to maximize the volume used in containers, considering load balancing constraints, the presence of fragile items and the possibility of rotating these boxes.
-Paquay {\it et al.} \cite{PaquayLimbourgSchynsOliveira2018} developed some heuristics to solve this problem.
-
-Chenguang {\it et al.} \cite{YangLiuGao2018} modeled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing to minimize fuel consumption.
-
-Wong and Ling \cite{wong2020}A developed a mathematical model and a tool based on mixed integer programming for optimizing cargo in aircraft with different pallet configurations.
-Balance constraints and the presence of dangerous items were considered.
-Wong {\it et al.} \cite{eugene2021} integrated this tool to a digital simulation model, with a visualization and validation system, based on sensors that alert about load deviations.
-
-Zhao {\it et al.} \cite{zhao2021} proposed a new modeling for WBP based on mixed integer programming.
-Instead of focusing on the CG deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
-
-As can be seen, none of these works address air cargo palletization and load balancing with route optimization in a multi-leg transportation plan for a single aircraft.
-This is the objective of our work: to model and elaborate heuristics for a real problem that simultaneously involves 4 intractable sub-problems: APP, WBP, PDP and TSP.
-
-\section{Problem context and assumptions}
-\label{sec3}
-
-In this section, we describe the context of the problem addressed in this work, as well as the assumptions considered.
-
-\subsection{Operational premises}
-
-As we are dealing with an extremely complex and diverse problem, we decided to establish some simplifying characteristics:
-
-\begin{itemize}
-
-	\item At each node of the tour, the items to be allocated are characterized by weight, volume, scores, and previously known destinations.
-We leave the consideration of 2D or 3D items to a future work.
-
-	\item We considered a unique pallet type: the {\it 463L Master Pallet}, a common size platform for bundling and moving air cargo.
-It is the primary air cargo pallet for more than 70 Air Forces and many air transport companies.
-This pallet has a capacity of $4,500 kg$ and $13.7 m^3$, which may be limited by its position along the cargo bay.
-It is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$.
-For more information, see {\tt www.463LPallet.com}.
-
-	\item All items allocated on a pallet must have the same destination.
-A pallet which has not yet reached its destination may receive more items, although it is known that these operations of removing restraining nets increase handling time and the risk of improper delivery.
-We do not consider oversized cargo in this work, but only cargo items that fit on these pallets.
-
-	\item Finally, as we are interested in minimizing fuel costs, we disregarded others costs not directly associated with aircraft flight, such as handling.
-
-\end{itemize}
-
-Throughout this text, we call {\it Packed Contents}\/ (PC) a set of items of same destination stacked on a pallet and covered with a restraining net (see Figure \ref{fig:larger2}).
-It is considered unique, having the same attributes of its components, whose values are the sum of individual scores, weights and volumes.
-To ensure accuracy in pickup and delivery operations, a PC must remain on board until its destination.
-
-\subsection{Aircraft parameters and load balancing}
-
-We consider real scenarios with an aircraft with payload of $75,000 kg$, with enough fuel for a range of $4,400km$.
-The aircraft layout is presented in Figure \ref{fig:larger}, where the pallets positions are identified by $p_i$, $1\leq i \leq 18$.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=0.4]{Images/consolidated.jpg}
-	\caption{A PC on a 463L pallet inside a {\it Boeing C-17}}
-	\small\textsuperscript{Source: From Wikimedia Commons, the free media repository}
-	\label{fig:larger2}
-\end{figure}
-
-\begin{figure}[!h]
-	\centering
-
-	\begin{tikzpicture}[scale=1.2, samples=100]
-
-		\filldraw[fill=blue!5!white, draw=black] (0, 0) rectangle (10.23, 2.6);
-
-		\filldraw[fill=gray!20!white, draw=black] (0.20, 1.40) rectangle node{$p_{17}$} (1.43, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (0.20, 0.20) rectangle node{$p_{18}$} (1.43, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (1.63, 1.40) rectangle node{$p_{15}$} (2.86, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (1.63, 0.20) rectangle node{$p_{16}$} (2.86, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (3.06, 1.40) rectangle node{$p_{13}$} (4.29, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (3.06, 0.20) rectangle node{$p_{14}$} (4.29, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (4.49, 1.40) rectangle node{$p_{11}$} (5.72, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (4.49, 0.20) rectangle node{$p_{12}$} (5.72, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (5.92, 1.40) rectangle node{$p_{9}$} (7.15, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (5.92, 0.20) rectangle node{$p_{10}$} (7.15, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (7.35, 1.40) rectangle node{$p_{7}$} (8.58, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (7.35, 0.20) rectangle node{$p_{8}$} (8.58, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (8.78, 1.40) rectangle node{$p_{5}$} (10.1, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (8.78, 0.20) rectangle node{$p_{6}$} (10.1, 1.20);
-
-		\filldraw[fill=blue!5!white, draw=black] (10.23, 0) rectangle (13.27, 2.6);
-		\filldraw[fill=gray!20!white, draw=black] (10.41, 1.40) rectangle node{$p_{3}$} (11.64, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (10.41, 0.20) rectangle node{$p_{4}$} (11.64, 1.20);
-		\filldraw[fill=gray!20!white, draw=black] (11.84, 1.40) rectangle node{$p_{1}$} (13.07, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (11.84, 0.20) rectangle node{$p_{2}$} (13.07, 1.20);
-
-	\end{tikzpicture}
-
-	\begin{tikzpicture}[scale=1.2, samples=100]
-		\draw[black, thick][<-]  (1, 2.77) node[anchor=east, font=\fontsize{7}{3.5}\selectfont]{Forward} -- (9, 2.77) node[anchor=west, font=\fontsize{7}{3.5}\selectfont]{Ramp} ;
-	\end{tikzpicture}
-
-	\caption{Aircraft layout} \label{fig:larger}
-\end{figure}
-
-The torque applied to the aircraft must keep its CG in the operational range, which corresponds to a fixed percentage of the {\it Mean Aerodynamic Chord} \footnote{Chord is the distance between the leading and trailing edges of the wing, measured parallel to the normal airflow over the wing.
-The average length of the chord is known as the {\it Mean Aerodynamic Chord} (MAC).} which is considered $1.17m$ for the aircraft of this work (see Figure \ref{fig:lateral}).
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=0.22]{Images/lateral.png}
-	\caption{Aircraft longitudinal cut, where red lines are pallets positions}
-	\label{fig:lateral}
-\end{figure}
-
-Table \ref{tab:larger} shows the parameters of this aircraft.
-$p_i$\/ are pallets, $1\leq i \leq 18$, with weight $W_i$\/ and $V_i$\/ volume limits, respectively.
-$D_i^{long}$\/ and $D_i^{lat}$\/ are, respectively, the longitudinal and latitudinal distances of each pallet centroids to the CG of aircraft along both axes.
-These distances will be used in the calculation of the torque referring to the items allocated on each pallet.
-In this aircraft, as the ramp have an inclination of $25°$, we made the necessary corrections in $D_i^{long}$, $W_i$\/ and $V_i$\/ of the corresponding pallets.
-
-\begin{table}[H]
-	\centering
-	\caption{Aircraft parameters}  \label{tab:larger}
-	\footnotesize
-	\begin{tabular}{c | c c c c c c c c c}
-		\toprule
-		& \multicolumn{3}{c}{$Payload$: 75,000kg} & \multicolumn{3}{c}{$limit^{CG}_{long}$: $1.170m$} &
-		\multicolumn{3}{c}{$limit^{CG}_{lat}$: $0.19m$} \\
-		\midrule
-		\multirow{2}{*}{{\boldm{$p_i$}}}  & $p_{17}$ & $p_{15}$ & $p_{13}$ & $p_{11}$ & $p_{9}$ & $p_{7}$ & $p_{5}$ & $p_{3}$ & $p_{1}$ \\
-		& $p_{18}$ & $p_{16}$ & $p_{14}$ & $p_{12}$ & $p_{10}$ & $p_{8}$ & $p_{6}$ & $p_{4}$ & $p_{2}$ \\
-		\midrule 
-		\multirow{2}{*}{\boldm{$D_i^{long}$} ($m$)} & -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\
-		& -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\			
-		\midrule 
-		\multirow{2}{*}{\boldm{$D_i^{lat}$} ($m$)}  & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 \\
-		& -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 \\	
-		\midrule
-		{\boldm{$W_i$}} ($kg$)      &   4,500   &    4,500  &   4,500   &  4,500    & 4,500     & 4,500     & 4,500     & 3,000    & 3,000   \\
-		{\boldm{$V_i$}} ($m^3$)   &   14.8   &   14.8   &  14.8    &  14.8    & 14.8     & 14.8     & 14.8     & 10.0    & 7.0 \\	
-		\midrule
-
-		\textbf{Fuel cost}  & \multicolumn{9}{c}{ $c_d$ = US\$ $4.90/km$ }	 \\
-
-		\midrule
-
-		\textbf{Fuel consumption rate}  & \multicolumn{9}{c}{ $c_g = 5\%$} \\
-
-		\midrule
-
-		\textbf{Maximum weight}  & \multicolumn{9}{c}{ $W_{max} = \sum_i W_i = 75,000kg$} \\
-
-		\bottomrule
-	\end{tabular}
-	\normalsize 
-\end{table}
-
-This aircraft spends $c_d$\/ dollars per kilometer flown, and can carry up to $W_{max}$\/ of cargo distributed on the pallets.
-The {\it fuel consumption rate}\/ $c_g$\/ is the percentage limit of cost increase  due to the CG deviation on the longitudinal axis of the aircraft.
-$c_g$\/ depends on the characteristics of the aircraft, and is an arbitrated value for this work that should be better evaluated in future research.
-It is important to consider that the $c_g$\/ tends to zero as the aircraft attitude tends to level.
-As the CG deviation varies from $0$\/ to $limit^{CG}_{long}$, the {\it fuel cost}\/ increase varies from $0$\/ to $c_g$.
-
-\vspace{3mm}
-We also make the following assumptions:
-\begin{itemize}
-	\item on each pallet, the items are distributed in such a way that their CG coincides with the centroid of the pallet;
-	\item the CG of the total load must be at a maximum longitudinal distance $limit^{CG}_{long}$\/ from the CG of the aircraft;
-	\item the CG of the total load must be at a maximum lateral distance $limit^{CG}_{lat}$\/ from the CG of the aircraft;
-	\item the pallets are distributed in two identical rows (with odd and even indices, respectively), and the centroid of $p_i$\/ is at distance $D^{lat}_{i}$\/ from the center-line of the aircraft.
-\end{itemize}
-
-\section{The mathematical modeling}
-\label{sec4}
-
-Given the assumptions and parameters described in the previous section, we are ready to present the mathematical modeling of ACLP+RPDP.
-
-ACLP+RPDP has the objective function \ref{eq1}, and the calculus equations \ref{eq:scores} to \ref{eq:LatPacked} subject to constraints \ref{eq:torqlat} to \ref{eq21}, which will be described below.
-
-\subsection{Problem structure and input data}
-
-In the definitions below and in the tests of Section \ref{sec6}, we use the values assigned in Tables \ref{tab:costs} and \ref{tab:larger}.
-To simplify the notation throughout this text, $p_i$\/ will be called pallet $i$, and $l_k$\/ will be called node $k$.
-
-Let $L = \{ 0, 1, \ldots, K \}$ be the set with the $K+1$\/ nodes, and let $L_k$\/ be the set of remaining nodes when the aircraft is in the node $k$, $0 \leq k \leq K$. Therefore, $L_0 = L$.
-
-Let $d(a,b)$ be the distance from node $a$\/ to node $b$, where $0 \leq a, b \leq K$. By definition, $d(a,a) = 0, \forall a$.
-
-Let $C=\left[c_{a,b}\right]$\/ be the cost matrix of flights, where $c_{a,b} = c_d*d(a,b), 0 \leq a,b \leq K$.
-
-Let $M = \{1, 2, \ldots, m \}$\/ the set of $m$\/ empty pallets assigned to specific positions within the aircraft.
-In Table \ref{tab:larger}, $m=18$.
-Each pallet $i$, $1 \leq i \leq m$, has weight capacity $W_i$, volume capacity $V_i$, longitudinal distance to the CG of aircraft $D^{long}_i$, and lateral distance to the center line of aircraft $D^{lat}_i$.
-
-Let $N_k = \{1, \ldots, n_k \}$\/ be the set of $n_k$\/ items $t_j^k$\/ available for loading on the node $k$, $1 \leq j \leq n_k$, $0 \leq k \leq K$.
-For ease of notation, $t_j^k$\/ will be called item $j$\/ of the node $k$, with score $s_j$, weight $w_j$, volume $v_j$, and destination $to_j \in L_k$.
-Let $N = \bigcup_{0 \leq k \leq K} N_k$ be the set of items of all nodes along a tour.
-
-Let $Q_k = \{1, \ldots, m_k \}$\/ be the set of $m_k \leq m$\/ packed contents $a^k_q$\/ at the node $k$, $1 \leq q \leq m_k$, $0 \leq k \leq K$.
-For ease of notation, $a^k_q$\/ will be called packed content $q$\/ of the node $k$, with total weight $w_q$, total volume $v_q$, and destination $to_q \in L_k$.
-By definition, $m_0 = 0$, and therefore $Q_0 = \varnothing$.
-Packets contents that were destined to node $k$\/ are unloaded when the aircraft arrives at this node, that is, they are not considered in $Q_k$.
-
-\subsection{The decision variables}
-
-Let $X_{ij}^k$\/ and $Y_{iq}^k$\/ be binary variables, where $1 \leq i \leq m$, $1 \leq j \leq n_k$, $1 \leq q \leq m_k$\/ and $0 \leq k \leq K$.
-
-$X_{ij}^k = 1$\/ if the item $j$\/ of the node $k$\/ is assigned to the pallet $i$, and 0 otherwise.
-
-$Y_{iq}^k = 1$\/ if the packed content $q$\/ of the node $k$\/ is assigned to the pallet $i$, and 0 otherwise.
-
-\subsection{The allocation graph}
-
-Allocations of items or packed contents to the pallets in the node $k$\/ can be seen as a bipartite graph $G_k(V_k, E_k)$, where:
-
-\begin{itemize}
-
-\item $V_k = M \cup N_k \cup Q_k$;
-
-\item $E_k = E_{N_k} \cup E_{Q_k}$;
-
-\item $(i, j) \in E_{N_k}$\/ if $X_{ij}^k = 1$, where $i$\/ is a pallet and $j$\/ is a item of the node $k$;
-
-\item $(i, q) \in E_{Q_k}$\/ if $Y_{iq}^k = 1$, where $i$\/ is a pallet and $q$\/ is a packed content of the node $k$.
-
-\end{itemize}
-
-\subsection{The objective function}
-
-Let $S_K = \{s: \{1, \dots, K\} \rightarrow \{1, \dots, K\} \}$\/ be the set of $K!$ permutations $\pi$, which correspond to all possible tours (or itineraries) that have node $0$\/ as origin and end, passing through the other $K$\/ nodes.
-Let $\pi_k$\/ be $k^{th}$\/ node of the tour $\pi$, $1 \leq k \leq K$.
-In this way, the tour $\pi$\/ is described as $\{0,\pi_1,\ldots,\pi_K, 0\}$, {\color{blue}$K > 1$}.
-For ease of notation, we can define $\pi_0 = \pi_{K+1} = 0$.
-
-We consider K > 1 because the case K = 1 has a single and trivial solution for the tour.
-
-The objective of ACLP+RPDP is to find the permutation $\pi \in S_K$, with the corresponding allocation of items on the pallets at each node, that maximizes the function $f_\pi = \tilde{s}_\pi/\tilde{c}_\pi$ \ref{eq1}, where $\tilde{s}_\pi$\/ is the total score of transported items, and $\tilde{c}_\pi$\/ is the total cost of fuel consumed.
-
-\begin{equation} \label{eq1}
-	\max_{\pi \in S_K} f_\pi = \tilde{s}_\pi/\tilde{c}_\pi
-\end{equation}
-
-\subsection{The calculus equations}
-
-$\tilde{s}_\pi$\/ \ref{eq:scores} is the sum of the scores of the items loaded on the aircraft throughout the tour $\pi$.
-
-\begin{equation} \label{eq:scores}
-	\tilde{s}_\pi = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times s_j
-\end{equation}
-
-$\tau_{\pi_k}$\/ \ref{eq:tau} is the longitudinal torque applied by the loaded pallets, in the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.
-
-\begin{equation} \label{eq:tau}
-	\tau_{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times \Big ( \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{long}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-In this way, we can calculate the cost $\tilde{c}_\pi$\/ \ref{eq:costs}, referring to the total fuel consumed in the tour $\pi$\/ due to the distances traveled and the CG longitudinal deviations in each pallet.
-
-\begin{equation} \label{eq:costs}
-	\tilde{c}_{\pi} = \sum_{k=0}^{K} \Big [ c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|) \Big ] 
-\end{equation}
-
-The sets of nodes not yet visited are defined in \ref{eq:pdp11} and \ref{eq:departing}.
-
-\begin{equation} \label{eq:pdp11}
-	L_0 = L
-\end{equation}
-
-\begin{equation} \label{eq:departing}
-	L_{\pi_k} = L_{\pi_{k-1}} - \{\pi_k\}; \ k \in \{1, \ldots, K\}
-\end{equation}
-
-At the beginning of the tour $\pi$, there are no packed contents \ref{eq:pdp13}.
-
-\begin{equation} \label{eq:pdp13}
-	m_0 = 0
-\end{equation}
-
-$\epsilon_{\pi_k}^t$\/ \ref{eq:LatItem} and $\epsilon_{\pi_k}^a$\/ \ref{eq:LatPacked} are the lateral torque applied by the loaded pallets, in the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.
-For ease of notation, $\epsilon_{\pi_k}^t$\/ corresponds to the items, and $\epsilon_{\pi_k}^a$\/ to the packed contents.
-
-\begin{equation} \label{eq:LatItem}
-	\epsilon_{\pi_k}^t = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{j=1}^{n_{\pi_k}} \Big ( X_{ij}^{\pi_k} \times w_j \times (i\%2) - X_{ij}^{\pi_k} \times w_j \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:LatPacked}
-	\epsilon_{\pi_k}^a = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{q=1}^{m_{\pi_k}} \Big ( Y_{iq}^{\pi_k} \times w_q \times (i\%2) - Y_{iq}^{\pi_k} \times w_q \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\subsection{The constraints}
-
-Finally, we can consider the constraints on each node $\pi_k$:
-
-\begin{itemize}
-	\item The longitudinal \ref{eq:torqlong} and the lateral \ref{eq:torqlat} torques must be within the limits of the aircraft.
-	\item The items allocated to each pallet cannot exceed its weight \ref{eq:app2} and volume \ref{eq:app3} limits.
-	\item At most, each item is associated with a single pallet \ref{eq:app4};
-	\item Packet contents that have not yet reached their destination must remain on board \ref{eq:app5}.
-	\item Items allocated on the same pallet must have the same destinations.
-In this case, we need two constraints: \ref{eq18} and \ref{eq19}.
-If $X_{ia}^{\pi_k} = X_{ib}^{\pi_k} = 1$, both constraints require that $to_a = to_b$; otherwise these constraints have no effect.
-	\item If there is a packed content on the pallet, it must also have the same destination as the other items. Similarly, we use two constraints: \ref{eq20} and \ref{eq21}.
-\end{itemize}
-
-\begin{equation} \label{eq:torqlong}
-	| \tau_{\pi_k} | \leq 1;\ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:torqlat}
-	| \epsilon_{\pi_k}^t + \epsilon_{\pi_k}^a| \leq  1; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:app2}
-	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:app3}
-	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:app4}
-	\sum_{i=1}^{m} X_{ij}^{\pi_k} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:app5}
-	\sum_{i=1}^{m} Y_{iq}^{\pi_k} = 1;\ to_q \in L_{\pi_k}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq18}
-	to_a - to_b \geq -K \times (1-X_{ia}^{\pi_k} \times X_{ib}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ a,b \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq19}
-	to_a - to_b \leq K \times (1-X_{ia}^{\pi_k} \times X_{ib}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ a,b \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq20}
-	to_j - to_q \geq -K \times (1-X_{ij}^{\pi_k} \times Y_{iq}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq21}
-	to_j - to_q \leq  K \times (1-X_{ij}^{\pi_k} \times Y_{iq}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-Constraints \ref{eq18} to \ref{eq21} are not within standard {\it Mixed-Integer Programming}\/ (MIP), and therefore need to be handled in a different way.
-In the next section, we will explain the strategy adopted to solve ACLP+RPDP.
-
-\section{Solution strategy}
-\label{sec5}
-
-Once the assumptions of this work and the mathematical modeling of the problem are presented, it is possible to see that ACLP+RPDP is {\it NP-hard}.
-In a similar way to Lurkin and Schyns \cite[p.
-6]{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$\/ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in node 0, and no items in node 1.
-Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is {\it NP-complete}.
-
-Throughout our research, we have thoughtfully described the ACLP+RPDP model in standard MIP format and found that no MIP solver can handle its practical cases in a feasible time.
-Thus, as ACLP+RPDP is highly complex, involving four intractable subproblems (APP, WBP, PDP and TSP), our strategy will be to focus only on real cases and to develop quick node-by-node solutions, not necessarily optimal, that allow to build a complete tour.
-
-The ACLP+RPDP solution defines a tour and a corresponding loading and unloading plan at each node.
-In this way, the loading and unloading time will be limited only by the use of the equipment at each node.
-Therefore, there will only be waiting time at the base node, corresponding to the runtime of ACLP+RPDP algorithm.
-For this reason, our method will also be parameterized by the maximum time that can be waited until the start of the load in the base node.
-
-On the other hand, we know that transport aircraft generally have a few dozen pallets, the flight plan has less than $6$\/ nodes, and each node has hundreds of items to be shipped.
-Under these circumstances, we can adopt two more strategies:
-
-\begin{itemize}
-
-\item Considering the number of destinations less than the number of pallets ($K < m$), we can preset the destinations of the pallets in each node.
-In this way, we will reserve a number of pallets proportional to the volume demanded by each destination at the shipping node.
-We could have used another criterion, but it was observed in the experiments that volume is more constrictive in airlift.
-
-\item If the number of nodes is small and we have quick node-by-node solutions, we have the possibility to check the solution corresponding to the tour with the shortest total distance, and even test all possible tours, selecting the one that provides the best value for the objective function.
-
-\end{itemize}
-
-Our complete strategy is summarized in Algorithm \ref{alg:main}.
-
-\begin{algorithm}[H]
-	\caption{ACLP+RPDP solving}  \label{alg:main}
-	\begin{algorithmic}[1]
-
-		\Procedure{$ACLP+RPDP$}{$scenario, surplus, time$}
-
-		\State Let aircraft and $M$\/ be according to Table \ref{tab:larger}  \label{main:M} \Comment{Aircraft parameters}
-
-		\State Let $K$, $L$\/ and $C$ be according to $scenario$\/ and Table \ref{tab:costs} \label{main:KLC} \Comment{Nodes, distances and costs}
-
-		\State Let $\pi_{TSP}$\/ be the tour with the shortest distance traveled
-
-		\State $N \gets ItemsGeneration(scenario, surplus)$ \label{main:items} \Comment{Items available for shipment}
-
-		\For {each $method$} \Comment{$method$ is a MIP solver or a heuristic}
-
-			\State $f_{\pi_{TSP}} \gets SolveTour(\pi_{TSP}, L, M, C, N, method, time/(K+1))$ \label{main:f1}
-			\State $answer1[scenario,surplus,method] \gets f_{\pi_{TSP}}$ \label{main:a1} \Comment{Results corresponding to $\pi_{TSP}$}
-			\For {each $\pi \in S_K$} \label{main:loop1} \Comment{$\pi$ is a permutation of nodes (a tour)}
-				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, time/K!)$ \Comment{$f_{\pi}$\/ is a solution to the tour $\pi$} \label{main:f2}
-			\EndFor \label{main:loop2}
-			\State $answer2[scenario,surplus,method] \gets \max f_{\pi}$ \label{main:a2} \Comment{Best result among all permutations}
-		\EndFor
-
-		\Return $answer$
-
-		\EndProcedure
-
-	\end{algorithmic}
-\end{algorithm}
-
-The aircraft parameters are in Table \ref{tab:larger} (line \ref{main:M}).
-In this algorithm, we use five values for $scenario$, according to Table \ref{tab:scenarios}, which defines $K$\/ and the set $L$\/ of nodes (line \ref{main:KLC}).
-
-\vspace{2.0mm}
-\begin{table}[H]
-	\centering
-	\caption{Testing scenarios}  \label{tab:scenarios}
-	\begin{tabular}{c c c }
-		\toprule
-		{\bf Scenario} & {$K$} & {$L$} \\		
-		\midrule
-		1 & 2    & \{$1,2$\}           \\
-		2 & 3    & \{$1,2,3$\}         \\
-		3 & 4    & \{$1,2,3,4$\}       \\
-		4 & 5    & \{$1,2,3,4,5$\}     \\
-		5 & 6    & \{$1,2,3,4,5,6$\}   \\
-		\bottomrule
-	\end{tabular}
-\end{table}
-
-$surplus$\/ is a value in $\{1.2, 1.5, 2.0\}$, which corresponds, at each node $k$, to the ratio between the sum of the volumes of the items  and the load capacity of the pallets ($surplus = \sum_{j=1}^{n_k} v_j$/$\sum_{i=1}^{m} V_i$).
-This parameter allows us to verify the different behavior of each $method$, according to $scenario$\/ and the quantity of items available for shipment.
-It is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).
-
-$time$\/ is a runtime limit, which will be distributed among the node-by-node solutions (lines \ref{main:f1} and \ref{main:f2}).
-Each tour consists of $K+1$\/ trips, and the total number of tours is $K!$.
-
-$method$\/ corresponds to a MIP solver or a heuristic that we will present in subsection \ref{methods}, where $SolveTour$\/ is the node-by-node solution.
-
-The results corresponding to the shortest tour are stored in $answer1$\/ (line \ref{main:a1}), and those obtained by testing all tours are stored in $answer2$\/ (line \ref{main:a2}).
-
-Next, we will present two subsections: in the first we explain how $SolveTour$\/ is executed, presetting the destinations of the pallets.
-In the second we will present the heuristic developed for node-by-node solutions.
-
-\subsection{SolveTour algorithm}
-\label{tour}
-
-As we commented in the previous subsection, we will adopt the strategy of presetting the destinations of each pallet throughout the tour.
-This is feasible in practical cases, where $K<m$.
-For this, each pallet $i$\/ also has a field $T^k_i$, $0\leq k \leq K$, which stores its next destination after being loaded in node $k$.
-For this reason, $T^k_i \in L_k$, $1 \leq i \leq m$, $0\leq k \leq K$.
-
-$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $time$\/ is the runtime limit that will be distributed among the $K+1$\/ legs of the tour.
-
-\begin{algorithm}[H]
-	\caption{Procedure to solve tour $\pi$ with $method$ (MIP solver or heuristic)}  \label{alg:tour}
-	\begin{algorithmic}[1]
-
-		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method$, $time$}
-
-		\State $\pi_0     \gets 0$ \label{tour:pi1} \Comment{Base $0$\/ is the first node}
-		\State $\pi_{K+1} \gets 0$ \label{tour:pi2} \Comment{Base $0$\/ is the last node}
-
-		\State $score \gets 0$ \label{tour:score}
-		\State $cost  \gets 0$ \label{tour:cost}
-
-		\For {$k \gets 0$ to $K$} \label{tour:loop1}	
-		\State $L_{\pi_k} \gets L - \{\pi_0,\pi_1,\ldots,\pi_k\}$  \label{tour:lk1}	\Comment{$\pi_k$ is the current node}	
-		\For{$i \gets 1$ to $m$}
-		\State $T_i^{\pi_k} \gets -1$ \label{tour:-1} \Comment{Unset the pallet destinations}
-		\EndFor
-
-		\If {$k = 0$}
-		\State Let $G_1(M \cup N_0, \varnothing)$ \label{tour:g11}
-		\Else
-
-		\State $E_{Q_{\pi_k}}, M \gets UpdatePacked(M, Q_{\pi_k}, \pi_k)$ \label{tour:dest}			
-		\State Let $G_1(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$ \label{tour:g12}
-		\EndIf  \label{tour:lk2}
-
-		\State $M \gets SetPalletsDestinations(M, \pi_k )$ \label{tour:dest2}
-
-		\State $G_2 \gets SolveNode(method, \pi_k, G_1, time/(K+1))$ \label{tour:node}
-
-		\State $s, \tau \gets ScoreAndDeviation(\pi_k, G_2)$ \label{tour:analyse}
-
-		\State $score \gets score + s$ \label{tour:score2}
-		\State $cost  \gets cost  + c_{\pi_k,\pi_{k+1}} \times (1 + c_g \times |\tau|)$ \label{tour:cost2}
-
-		\EndFor  \label{tour:loop2}
-
-		\Return $score / cost$ \label{tour:f}
-
-		\EndProcedure
-
-	\end{algorithmic}
-\end{algorithm}
-
-As we mentioned in the previous section, all tours start and end at the base $0$\/ (lines \ref{tour:pi1}-\ref{tour:pi2}).
-After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ flights (lines \ref{tour:loop1}-\ref{tour:loop2}).
-Initially, the set $L_{\pi_k}$\/ of remaining nodes is updated (line \ref{tour:lk1}), and the pallet destinations are unset (line \ref{tour:-1}).
-
-When the aircraft is at the base, the initial graph $G_1$\/ is empty because it has no packed content \ref{tour:g11}.
-Otherwise, $UpdatePacked$\/ (line \ref{tour:dest}) returns the set of packed contents that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation.
-This allocation is stored in graph $G_1$\/ (line \ref{tour:g12}).
-
-$SetPalletsDestinations$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node, without changing the pallets destination with packed contents.
-
-Finally, $SolveNode$\/ includes the edges corresponding to the items shipped at the current node, returning the graph $G_2$\/ (line \ref{tour:node}).
-The score and the CG deviation of $G_2$\/ are calculated (line \ref{tour:analyse}) and accumulated (lines \ref{tour:score2}-\ref{tour:cost2}), allowing the final result of this tour (line \ref{tour:f}).
-
-$UpdatePacked$, described in Algorithm \ref{alg:cons}, finds the best packed-pallet allocation, in terms of CG deviation, for the packed contents that remain on board.
-
-\begin{algorithm}[H]
-	\caption{Procedure to update the PC that remain boarded on node $\pi_k$}  \label{alg:cons}
-	\begin{algorithmic}[1]
-
-		\Procedure{$UpdatePacked$}{${M, Q_{\pi_k}, \pi_k}$}
-
-		\State $E_{Q_{\pi_k}} \gets MinCGDeviation(E_{Q_{\pi_k}})$ \label{cons:minCG}
-
-		\For{$i \gets 1$ to $m$} \label{cons:Ybegin}
-
-		\For{$q \gets 1$ to $m_{\pi_k}$}
-
-		\State $T_i^{\pi_k} \gets -1$
-
-		\If{$(i, q) \in E_{Q_{\pi_k}}$} 
-		\State $T_i^{\pi_k} \gets to_q$   \Comment{Set the pallet destinations}
-		\EndIf
-		\EndFor		
-		\EndFor \label{cons:Yend}
-
-		\Return $E_{Q_{\pi_k}}, M$
-
-		\EndProcedure
-
-	\end{algorithmic}
-\end{algorithm}
-
-$MinCGDeviation$\/ (line \ref{cons:minCG}) is run through a MIP solver with the equations \ref{eq:torque}, \ref{eq:embarked} and \ref {eq:one}, to relocate the packed contents on the pallets, minimizing torque and ensuring that they all remain on board, one packed content on each pallet.
-As there are few variables, the MIP solver returns an allocation $E_{Q_{\pi_k}}$\/ in less than 30 milliseconds.
-Finally, the destination of each pallet with a packed content is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).
-
-\vspace{3mm}
-$MinCGDeviation$ objective function:
-
-\begin{equation} \label{eq:torque}
-	\mbox{minimize}\ \Big | \sum_{i=1}^{m} \sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \times w_q \times D_i^{long}  \Big |
-\end{equation}
-
-\vspace{3mm}
-s.t.:
-
-\begin{equation} \label{eq:embarked}
-	\sum_{i=1}^{m} Y^k_{iq} = 1;\ q \in \{1,\ldots,m_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:one}
-	\sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \leq 1;\ i \in \{1,\ldots,m\}
-\end{equation}
-
-$SetPalletsDestinations$, which sets the pallets destination not yet defined, is described in Algorithm \ref{alg:dest}.
-
-\begin{algorithm}[H]
-	\caption{Procedure to set pallets destination based on the items to be embarked on node $\pi_k$}  \label{alg:dest}
-	\begin{algorithmic}[1]
-
-		\Procedure{$SetPalletsDestinations$}{$M, \pi_k$}
-
-			\For{$x \in L_{\pi_k}$}		
-				\State $vol_x \gets  0$  \label{dest:vector1}
-			\EndFor
-
-			\State $max \gets 0$
-
-			\State $total \gets 0$
-
-			\For{$j \gets 1$ to $n_{\pi_k}$}
-
-				\If {$to_j \in L_{\pi_k}$}
-
-					\State $vol_{to_j} \gets vol_{to_j} + v_j$ 
-					\State $total \gets total + v_j$
-
-					\If {$vol_{to_j} > vol_{max}$}
-
-						\State $max \gets to_j$ \label{dest:max} \Comment{$max$\/ is the destination with maximum volume demand}
-					\EndIf
-
-				\EndIf
-
-			\EndFor
-
-			\For{$x \in L_{\pi_k}$} \label{dest:propor1}
-
-				\If {$vol_{x} \neq 0$}
-
-					\State $needed \gets \max \{1, \lfloor{ (m-m_{\pi_k}) \times vol_{x}/total}\rfloor \}$  \Comment{Number of pallets to node $x$}
-
-					\State $np \gets 0$
-
-					\For{$i \gets 1$ to $m$}
-
-						\If{$np < needed$ {\bf and} $T_i^{\pi_k} = -1$}
-							\State $T_i^{\pi_k} \gets x$
-							\State $np \gets np + 1$
-						\EndIf
-
-					\EndFor
-				\EndIf
-
-			\EndFor \label{dest:propor2}
-
-			\For{$i \gets 1$ to $m$} \label{dest:max1}
-				\If{$T_i^{\pi_k} \gets -1$}  \Comment{Remaining pallets}
-					\State $T_i^{\pi_k} \gets max$
-				\EndIf
-			\EndFor \label{dest:max2}
-
-			\Return $M$
-
-		\EndProcedure
-
-	\end{algorithmic}
-\end{algorithm}
-
-$vol$\/ stores the demand volume of items destined to the non-visited nodes (line \ref{dest:vector1}).
-The destination of empty pallets is defined proportionally to the volume of items to be embarked (lines \ref{dest:propor1}-\ref{dest:propor2}).
-The destination with the maximum volume defines any remaining pallets (lines \ref{dest:max1}-\ref{dest:max2}).
-
-$ScoreAndDeviation$\/ is described in Algorithm \ref{alg:eval}, which evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.
-
-\begin{algorithm}[H]
-	\caption{Procedure to calculate the score and the relative torque deviation of graph $G$\/ in node $\pi_k$}  \label{alg:eval}
-	\begin{algorithmic}[1]
-
-		\Procedure{$ScoreAndDeviation$}{$\pi_k, G$}
-
-		\State Let $G(V_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$
-
-		\State $s \gets 0$
-		\For{$i \gets 1$ to $m$}
-			\State $\tau_i \gets 0$
-		\EndFor
-		\For{$i \gets 1$ to $m$}	\label{eval:loop1}
-
-			\For{$j \gets 1$ to $n_{\pi_k}$}
-
-				\If{$X_{ij}^{\pi_k} = 1$}
-					\State $s \gets s + s_j$ \label{eval:score1}
-					\State $\tau_i \gets \tau_i + w_j \times D_i^{long}$ \label{eval:eps1}
-				\EndIf
-			\EndFor
-
-			\For{$q \gets 1$ to $m_{\pi_k}$}
-				\If{$Y_{iq}^{\pi_k} = 1$}
-					\State $s \gets s + s_q$   \label{eval:score2}
-					\State $\tau_i \gets \tau_i + w_q \times D_i^{long}$  \label{eval:eps2}
-				\EndIf
-			\EndFor	   \label{eval:loop2}
-		\EndFor
-
-		\State $\tau \gets  \sum_{1}^{m} \tau_i / (W_{max} \times limit^{CG}_{long})$ \label{eval:torque}
-
-		\Return $s,\ \tau$
-		\EndProcedure
-
-	\end{algorithmic}
-\end{algorithm}
-
-This algorithm consists of a loop that goes through all the pallets (lines \ref{eval:loop1}-\ref{eval:loop2}), accumulating the scores (lines \ref{eval:score1} and \ref{eval:score2}) and the torques (lines \ref{eval:eps1} and \ref{eval:eps2}) of the shipped items, allowing the final calculation of the CG deviation (line \ref{eval:torque}).
-
-\subsection{Node-by-node solutions}
-\label{methods}
-
-In this subsection we present two implementations of $SolveNode(method,\ \pi_k,\ G)$, where  $\pi_k$\/ is the current node and $G$\/ is the allocation graph of the PC that remain on board at node $\pi_k$.
-These implementations correspond to the two possible values of the $method$\/ parameter: $MIP$\/ ({\it Gurobi}) or $Shims$\/ (a heuristic).
-
-\subsubsection{MIP solver}
-\label{solver}
-
-To find a viable solution for ACLP+RPDP, the strategy adopted in $SolveTour$\/ previously defines the values of some variables.
-Concretely, all permutations between the nodes are tested, the set of nodes to be visited is updated, the PC that remain on board are reallocated to minimize the CG deviation, and the pallets destination is determined according to the volume of items available for shipment.
-In this way, the mathematical modeling for $SolveNode(MIP,\ \pi_k,\ G)$\/ becomes simpler, which finds an allocation of available items in node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $to_i^{\pi_k}$, and $a^{\pi_k}_q$.
-
-This simplified modeling is described in the equations \ref{eq:obj2} to \ref{eq:if2}, which maximizes the score of the $n_{\pi_k}$ items to be shipped, maintaining load balancing.
-
-The variables $X_{ij}$ and $Y_{iq}$ define the sets of edges $E_{N_{\pi_k}}$ and $E_{Q_{\pi_k}}$, respectively, which will be added to the graph $G$.
-
-\vspace{3mm}
-\textbf{Objective function:}
-
-\begin{equation} \label{eq:obj2}
-	maximize\ \ \tilde{s} / \tilde{c}
-\end{equation}
-
-\vspace{3mm}
-\textbf{Where:}
-
-\begin{equation} \label{eq:newf}
-	\tilde{s} = \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij} \times s_j
-\end{equation}
-
-\begin{equation} 
-	W_{max} = \sum_{i=1}^{m} W_i
-\end{equation}
-
-\begin{equation} 
-	\tau^{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q) \Big ] \Big / W_{max} \times limit^{CG}_{long}
-\end{equation}
-
-\vspace{3mm}
-In Equation \ref{eq:costs2}, the first parcel denotes the fuel cost from the base until the last node, and the second the fuel cost in the last leg, from the last node back to the base.
-Note that $c_g$ is the fuel consumption potential increase due to the loaded aircraft CG deviation in the node ($\tau_{\pi_k}$).
-
-\begin{equation} \label{eq:costs2}
-	\tilde{c} = \sum_{k=0}^{K} \Big [ c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|) \Big ]
-\end{equation}
-
-\vspace{3mm}
-\textbf{s.t.:}
-
-{\color{blue} It was observed in the experiments that the lateral torque was always in the order of magnitude -3. So we did not consider these constraints.}
-
-\begin{equation}
-	|\tau_{\pi_k}| \leq 1
-\end{equation}
-
-\begin{equation} 
-	\sum_{i=1}^{m} (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q ) \leq W_{max}
-\end{equation}
-
-\begin{equation} 
-	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}
-\end{equation}
-
-\begin{equation} 
-	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}
-\end{equation}
-
-\begin{equation} 
-	\sum_{i=1}^{m} X_{ij} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:if1}
-	to_i^{\pi_k} = to_j;\ X_{ij} = 1; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:if2}
-	X_{ij}       = 0;\ to_j \notin L_{\pi_k}; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\subsubsection{Shims heuristic}
-
-Our goal is to find a heuristic that offers a good-quality solution for the node-by-node problem.
-As we are testing all $K!$\/ tours, its run time is a crucial requirement.
-Taking this into account, we perform a series of implementations based on known meta-heuristics: {\it Ant Colony Optimization} (ACO) \cite{Dorigo1992, DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \cite{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \cite{Glover1986} and {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \cite{FeoResende1989}.
-
-{\color{blue}
-These meta-heuristics were implemented in their standard forms. We did not use any specialized features to allow for the exploitation of the problem structure.
-}
-
-We also considered several ideas from the literature \cite{NiarFreville1997,Fidanova2006,Laabadi2018,Alonso2019,Zhan2020}, and we were careful to use the same data structures and procedures in all implementations.
-The decision for a simple and fast heuristic is due to the need to obtain, in less than an hour {\color{blue} (because the {\it Brazilian Air Force}'s transportation teams usually performs cargo planning in more than 2 hours)}, a complete solution for ACLP+RPDP, allowing its use in an operational environment.
-
-However, when we established the run time limit of $0.7$\/ seconds per node, the heuristic that presented better solutions was none of the previous ones.
-In this subsection, we will present a new heuristic for the node-by-node problem, called $Shims$.
-Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}).
-This strategy is based on a practical observation: usually, subsets of smaller and lighter items are saved for later adjustments to the remaining {\color{blue}available space}.
-
-The selection of edges for $E_{ N_{\pi_k} }$ uses the {\it edge attractiveness}\/ $\theta^{\pi_k}_{ij}$\/ \ref{eq:edge}, which can be understood as the tendency to allocate the item $j$ to pallet $i$.
-It is directly proportional to the score, and inversely proportional to the volume and the torque of each item.
-
-\begin{equation} \label{eq:edge}
-	\theta^{\pi_k}_{ij} = \frac{ s_j }{ v_j }\ \times \Big ( 1 - \frac{ w_j \times |D_i^{long}| }{ w_{max} \times\ | D_{max}^{long} | } \Big );\ i \in \{1,\ldots,m\},\ j \in \{1,\ldots,n_{\pi_k}\}
-\end{equation}
-
-{\color{blue}The denominator $w_{max} \times\ | D_{max}^{long} | $ represents the heaviest item put on the most distant pallet.}
-
-\begin{table}[H]
-
-	\begin{minipage}{0.08\linewidth}
-
-	\end{minipage}\hfill % these two lines must be close to each other
-	\begin{minipage}{0.34\linewidth}
-
-		\includegraphics[scale=0.25]{Images/shims.png}
-		\captionof{figure}{{\it Shims} of various thicknesses}
-		\small\textsuperscript{Source: www.mscdirect.com/product/details/70475967}
-		\label{fig:shims}
-
-	\end{minipage}\hfill % these two lines must be close to each other
-	\begin{minipage}{0.58\linewidth}
-
-		\begin{tikzpicture}[scale=0.75, samples=100]
-			\filldraw[fill=blue!3!white, draw=black] (0, 0) rectangle (12.5, 5.5);
-			\draw[->] (.2, .6) -- coordinate (x axis mid) (12, .6);
-			\node at (5, 0.3) {$\eta_1^{\ i}$};
-			\node at (5, 2.5) {|};
-			\node at (6, 2.5) {shims};
-			\node at (7, 2.5) {|};
-			\node at (7, 0.3) {$\eta_2^{\ i}$};
-			\node at (0.3, 5) {$\theta_{ij}^{\pi_k}$};
-			\draw[->] (.6, .2) -- coordinate (y axis mid) (0.6, 5.3);
-			\node at (12, 0.3) {$e_{ij}^{\pi_k}$};
-			\draw[smooth, domain = 0.09:2, color=black, thick] plot (.3+1/\x,{4.2+log2(\x)});
-			\draw[->, dashed] (5, 0.6)--(5, 2.0);
-			\draw[->, dashed] (7, 0.6)--(7, 1.5);
-		\end{tikzpicture}
-
-		\captionof{figure}{$n_{\pi_k}$\/ edges $e_{ij}^{\pi_k}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi_k}$\/ in non-ascending order}
-		\label{fig:whip}		
-	\end{minipage}
-\end{table}
-
-Figure \ref{fig:whip} represents the $n_{\pi_k}$\/ possible edges $e_{ij}^{\pi_k}$\/ of $i$\/ sorted by $\theta_{ij}^{\pi_k}$.
-\emph{Shims}\/ starts with a greedy solution, stopping at the edge with index $\eta_1^i$ (first phase).
-Then, considering even the edge with the index $\eta_2^i$, it elaborates different possible complements for this pallet (second phase) and selects the best of these complements (third phase).
-
-In this heuristic, described in Algorithm \ref{alg:shims}, pallets $i$\/ are considered in non-descending order of $|D_i^{long}|$ (line \ref{shims:pallets}).
-For each pallet, the $n_{\pi_k}$\/ possible edges $e_{ ij}^{\pi_k}$\/ are considered in non-increasing order of $\theta_{ij}^{\pi_k}$\/ (line \ref{shims:possible}) in three phases: the greedy phase (until line \ref{shims:endfirst}), the composition phase of the shims (lines \ref{shims:beginsecond}-\ref{shims:endsecond}) and the selection phase of the best shim (lines \ref{shims:beginthird}-\ref{shims:endthird}).
-
-{\color{blue}
-	In the beginning of the first phase (line \ref{shims:initQ}), the set $Q_{\pi_k}$ represents the PC whose destination is not $\pi_k$, the current node, thus remaining on board.
-
-	It is important to remember that $Q_{\pi_k}$ and $M$ were modified by the procedures $UpdatePacked(M, Q_{\pi_k}, \pi_k)$ and $SetPalletsDestinations(M, \pi_k)$, having their positions and destinations reassigned.
-
-	Until line \ref{shims:endfirst}, a greedy and partial solution for pallet $i$ is constructed by edges inclusion, until a certain volume limit ($V_i \times limit$) is reached.
-	The value of the {\it limit} (0.92) was empirically determined by the {\it iRace} tool developed by \cite{LopezIbanezManuel2016}.
-
-	In the second phase (lines \ref{shims:beginsecond} to \ref{shims:endsecond}), a set of shims named $Set$\/ is created, where each {\it Shim} is formed by a set of edges in the range $[\eta_1^i,\eta_2^i]$, whose total volume is limited by $slack_i$.
-In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it First-Fit Decreasing}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \cite{JohnsonGarey1985}.
-Basically, shims are created by accumulating the following edges, taking $slack_i$\/ as a limit.
-
-	In the third phase (lines \ref{shims:beginthird}-\ref{shims:endthird}), the best {\it Shim} in the $Set$\/ is chosen.
-Initially, two shims are found: $sh_w$\/ with larger weight and $sh_v$\/ with larger volume.
-Among the two, the {\it Shim} with the highest score will be chosen and its edges are inserted into $E_{N_{\pi_k}}$.
-}
-
-\begin{algorithm}[H]
-	\caption{$Shims$\/ heuristic on node $\pi_k$}  \label{alg:shims}
-	\begin{algorithmic}[1]
-
-		\Procedure{$SolveNode$}{$Shims,\ \pi_k,\ G,\ time$}
-
-		\State $T_{begin} \gets$ current system time
-
-		\State $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$ \Comment{a graph with pallets positions ($M$), PC ($Q_{\pi_k}$), and edges ($E_{Q_{\pi_k}}$)}  \label{shims:initQ}
-
-		\State Sort $M$ by $|D_i|$ in non-descending order \label{shims:pallets}
-
-		\State $E_{N_{\pi_k}} \gets \{\ \}$  \Comment{an empty set of pallet-item edges}
-
-		\State Let $E_{ij}$ be an array with $m \times n_{\pi_k}$ edges $(i,\ j)$ sorted by $\theta_{ij}^{\pi_k}$ in non-ascending order \label{shims:possible}
-
-		\State $\tau^{\pi_k} \gets 0$	 \Comment{aircraft torque in node $\pi_k$}	
-		\State $\tau_{max} \gets W_{max} \times limit^{CG}_{long}$ \Comment{maximum aircraft torque}
-
-		\State $volume \gets \{0\ |\ [1,\ldots,m] \}$
-		\State $slack  \gets \{0\ |\ [1,\ldots,m] \}$
-
-		\State $limit \gets 0.92$ \Comment{0.92 was obtained by the {\it iRace} package \cite{LopezIbanezManuel2016}}
-
-		\For{$i \in \{1,\ldots,m\}$}
-
-			\For{$q \in \{1,\ldots,m_{\pi_k}\}$}
-
-				\If{$(i, q) \in E_{Q_{\pi_k}}$} 
-					\State $volume_i \gets volume_i + v_q$ 
-				\EndIf		
-			\EndFor
-
-			\For {$e_{ij} \in E_{ij}$}
-
-				\State $\Delta\tau \gets w_j \times D_i$
-
-				\State $\tau_{new} \gets | \tau^{\pi_k} + \Delta\tau |$
-
-				\If{ ($E_{N_{\pi_k}} \cup \{e_{ij}\}$ is feasible) {\bf and} ($volume_i \leq V_i \times limit$) {\bf and} ($\tau_{new} \leq \tau_{max}$ ) }
-
-					\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup \{e_{ij}\}$
-
-					\State $volume_i \gets volume_i + v_j$
-
-					\State $\eta_1^{\ i} \gets \eta_1^{\ i} + 1$
-
-					\State $\tau^{\pi_k} \gets \tau^{\pi_k} + \Delta\tau$
-
-				\EndIf
-
-			\EndFor			 \Comment{end of the first phase} \label{shims:endfirst}
-
-			\State $slack_i \gets V_i - volume_i$  \label{shims:beginsecond}
-			\State $\eta_2^{\ i} \gets \eta_1^{\ i}$
-
-			\Repeat
-				\State $\eta_2^{\ i} \gets \eta_2^{\ i} + 1$  \label{shims:eta2}
-
-				\State $e_{ij} \gets E_{i,\eta_2^{\ i}}$
-
-				\State $volume_i \gets volume_i + v_j$
-
-			\Until{($\eta_2^{\ i} = n_{\pi_k}$) {\bf or} ($ volume_i \geq (1 + 3 \times (1-limit)) \times V_i)$)} \label{shims:endsecond} \Comment{end of the second phase}
-
-			\State $volume \gets 0$	\label{shims:beginthird}
-
-			\State $volume \gets 0$, $b \gets 1$ $shims_b \gets \{\}$, $Set \gets Set \cup \{shims_b\}$
-
-			\State $shims_b \gets \{\}$  \label{empty_shims}
-
-			\State $Set \gets Set \cup \{shims_b\}$ \label{empty_set}
-
-			\For{$x \in \{\eta_1^{\ i},\ldots,\eta_2^{\ i} \}$} \label{edges_indexes}
-
-			{\color{blue}
-
-				\If{ current system time - $T_{begin} > time$}
-				\State {\bf break}
-				\EndIf
-
-			}
-
-				\State $NewShims \gets$ {\bf True} \label{new_shims}
-				\State $e_{ij} \gets E_x^{\ i}$
-
-				\For{$shims \in Set$} \label{shims_set}
-
-					\If { $e_{ij} \not\in (E_{N_{\pi_k}} \cup shims)$ {\bf and} $e_{ij}$ is feasible  {\bf and}  $(v_j + volume) \leq slack_i$}
-
-						\State $shims \gets shims \cup \{\ e_{ij}\ \}$
-						\State $volume \gets volume + v_j$
-						\State $NewShims \gets$ {\bf False} \label{new_shims_false}
-
-						\State {\bf break}
-					\EndIf
-
-				\EndFor
-
-				\If{$NewShims$} \label{new_shims2}
-					\State $volume \gets 0$
-					\State $b \gets b + 1$
-					\State $shims_b \gets \{\ \}$
-					\State $shims_b \gets shims_b \cup \{e_{ij}\}$
-					\State $Set \gets Set \cup \{\ shims_b\ \}$
-				\EndIf
-
-			\EndFor
-
-			\State $sh_w \gets shims$, where $shims \in Set$ and $\sum_{e_{ab} \in shims} w_b$ is maximum  \label{best_weight}		
-			\State $sh_v \gets shims$, where $shims \in Set$ and $\sum_{e_{ab} \in shims} v_b$ is maximum \label{best_volume}
-			\State $sh_{best} \gets shims$, where $shims \in \{sh_w, sh_v\}$ and $\sum_{e_{ab} \in x} s_b$ is maximum \label{best_score}
-
-			\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup sh_{best}$ 	\label{shims:endthird}   \Comment{end of the third phase}
-
-		\EndFor
-
-		\Return $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$
-
-		\EndProcedure
-	\end{algorithmic}
-\end{algorithm}
-
-\section{Implementation and results}
-\label{sec6}
-
-This section is composed of two parts: the generation of test instances and the results obtained in our implementation.
-
-\subsection{Instances generation}
-\label{items}
-
-As we are dealing with a new problem, which until now had not been modeled in the literature, we have to create our own benchmarks.
-For this, we based on the characteristics of real airlifts carried out by the {\it Brazilian Air Force}, as described below.
-
-In the military airlift carried out in Brazil from 2008 to 2010, 23\% of the items weighed between $10kg$ and $20kg$, 22\% from $21kg$ to $40kg$, 24\% from $41kg$ to $80kg$, 23\% from $81kg$ to $200kg$, and 8\% between $201kg$ and $340kg$.
-These five groups of items are described in Table \ref{tab:weights}, where $P$\/ represents the group probability.
-On the other hand, the average density of these items is approximately $246 kg/m^3$.
-
-\begin{table}[H]
-	\centering
-	\caption{Items weight distribution ($kg$) in Brazil}  \label{tab:weights}
-	\begin{tabular}{c c c c }
-		\toprule
-		$Group$ & $P$ & $low$ & $high$  \\		
-		\midrule
-		1              & 0.23           & 10  & 20  \\
-		2              & 0.22           & 21  & 40  \\
-		3              & 0.24           & 41  & 80  \\		
-		4              & 0.23           & 81  & 200 \\
-		5              & 0.08           & 201 & 340 \\
-		\bottomrule
-	\end{tabular}
-\end{table}
-
-In the generation of test instances, we use three types of random selections:
-\begin{itemize}
-	\item $RandomReal(r_1,r_2)$: randomly selects a real number in $[r_1,r_2]$, where $r_1$ and $r_2$ are real numbers;
-	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$, where $i_1$ and $i_2$ are integer numbers;
-	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
-\end{itemize}
-
-\begin{algorithm}[H]
-	\caption{Procedure to generate items}  \label{alg:itemsgen}
-	\begin{algorithmic}[1]
-
-		\Procedure{$ItemsGeneration$}{$scenario, surplus$}
-
-		\State Let $L$ and $M$ be according to $scenario$ \label{ig:LM}
-		\State $limit \gets surplus \times \sum_{i=1}^{m} V_i$ \label{ig:extended}
-
-		\State Let $t_j^k$ be an item $j$ in node $k$.
-
-		\For{$k \in \{0,\ldots,K\}$}
-
-			\State $N_k \gets \varnothing$
-			\State $j \gets 0$
-			\State $vol \gets 0$	\label{ig:totals}	
-			\While{$vol < limit$}
-
-				\State $j \gets j+1$
-
-				\Repeat
-					\State $to_j \gets RandomInt(0, K)$ \label{ig:dest}
-				\Until{$to_j \neq k$}
-
-				\State $x = Roulette(item)$ biased through $P$ \label{ig:weight1} \Comment{From Table \ref{tab:weights}}
-
-				\State $w_j \gets RandomReal(low(x), high(x))$        \label{ig:weight2}
-
-				\State $s_j \gets \round{100 \times (1 - \log_{10}(RandomInt(1, 9)))} $ \label{ig:score}
-
-				\State $v_j \gets w_j / RandomReal(148, 344)$ \label{ig:volume}
-
-				\State $vol \gets vol + v_j$
-
-				\State $N_k \gets N_k \cup \{t_j^k\}$ 
-			\EndWhile
-
-		\EndFor
-
-		\State $N \gets \bigcup_{0 \leq k \leq K} N_k$
-
-		\Return $N$
-
-		\EndProcedure
-	\end{algorithmic}
-\end{algorithm}
-
-$ItemsGeneration$, which generates $N$, is described in Algorithm \ref{alg:itemsgen}.
-The parameter $scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and the parameter $surplus$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}).
-To avoid simply loading all items, we use $surplus > 1$ {\color{blue}(1.2, 1.5, and 2.0).
-This represents more test instances for each scenario.}
-
-For each generated $t^k_j$\/ item, its destination is randomly selected (line \ref{ig:dest}), its weight has a distribution according to Table \ref{tab:weights} (lines \ref{ig:weight1}-\ref{ig:weight2}), its score varies $100$\/ (highest) and $5$\/ (lowest) according to a logarithmic scale (line \ref{ig:score}, and its volume is randomly defined from the density, where we allow a variation of 40\% more or less than the average density of $246 kg/m^3$\/ (line \ref{ig:volume}).
-
 {\color{blue}
@@
 \subsection{Results obtained}
-
 In the tests performed, we used a 64-bit, 16GB, 3.6GHz, eight-core processor with {\it Linux Ubuntu 22.04.1 LTS 64-bit} as the operational system and {\it Python 3.10.4}\/ as the programming language.
@@
 		\toprule
-{$method$}      & {\bf Results}&{\bf 1}&{\bf 2} & {\bf 3}      & {\bf 4}      & {\bf 5}  & \specialcell{{\bf Normalized}\\{\bf Speed-up}}  \\
+		{$method$}                        & {\bf Results}    & {\bf 1}     &  {\bf 2}    & {\bf 3}      & {\bf 4}      & {\bf 5}  & \specialcell{{\bf Normalized}\\{\bf Speed-up}}  
+		\\
+		\toprule
+		\multirow{2}{*}{\textbf{Gurobi}}  & $f$      & 6.10   & {\bf 10.02}& {\bf 10.82} & {\bf 12.13} & {\bf 12.10} & {\bf 0.997} \\ %51,84
+		&{\bf run time (s)}                          & 14     & 51         & 243         & 1,417       & 9,720       &  1.000 \\ 
 		\midrule[.1pt]
-\multirow{2}{*}{{\it Gurobi}}     & Upper bound for $f$      & {\it 7.91} & {\it 11.92} & {\it 13.12} & {\it 14.29} & {\it 13.34} & -  \\ %51,84
-&{\bf run time (s)}                          & 14         & 19          & 73          & 188         & 1,097      & -  \\ 
-\toprule
-\multirow{2}{*}{\textbf{TS}}      & $f$      & 5.37       &             &             &             &            &   \\ % 5.49	 901
-&{\bf run time (s)}                          & 659        &             &             &             &            &   \\ 
+		\multirow{2}{*}{\textbf{TS}}      & $f$      & 6.56     & 8.49       & 9.19       & 11.25      & 11.48      &  0.906 \\
+		&{\bf run time (s)}                          & 4       & 17        & 86        & 516       & 3,600     &  2.709 \\ 
 		\midrule[.1pt]		
-\multirow{2}{*}{\textbf{GRASP}}   & $f$      &            &             &             &             &            &   \\ 
-&{\bf run time (s)}                          &            &             &             &             &            &   \\ 
+		\multirow{2}{*}{\textbf{GRASP}}   & $f$      & 6.76     & 8.85       & 9.65       & 11.75      & 11.54      &  0.937 \\ 
+		&{\bf run time (s)}                          & 4       & 17        & 80        & 472       & 3,265     &  2.984 \\ 
 		\midrule[.1pt]
-\multirow{2}{*}{\textbf{ACO}}     & $f$      & 5.25       & 5.75        & 5.06        & 4.84        &  4.20      &   \\ % 4.20	 1133
-&{\bf run time (s)}                          & 313        & 326         & 303         & 256         &  1,133     &   \\ 
+		\multirow{2}{*}{\textbf{ACO}}     & $f$      & {\bf 6.77}& 8.85      & 9.65       & 11.75      & 11.54      &  0.937 \\ 
+		&{\bf run time (s)}                          & 4        & 17       & 86        & 513       & 3,596     &  2.714 \\ 
 		\midrule[.1pt]	
-\multirow{2}{*}{\textbf{NMO}}     & $f$      &            &             &             &             &            &   \\
-&{\bf run time (s)}                          &            &             &             &             &            &   \\ 
+		\multirow{2}{*}{\textbf{NMO}}     & $f$      & 6.74      & 8.80      & 9.61       & 11.76      & 11.50      &  0.934 \\
+		&{\bf run time (s)}                          & 4        & 17       & 86        & 508       & 3,480     &  2.796 \\ 
 		\midrule[.1pt]	
-\multirow{2}{*}{\textbf{Shims}}   & $f$      & 7.91       & 11.55       & 12.62       &  13.82      &  14.85          &   \\ % 14.85	 563
-&{\bf run time (s)}                          & 4          &  8          &  33         &  94         &  563          &   \\
-
+		\multirow{2}{*}{\textbf{Shims}}   & $f$      & {\bf 6.77} & 8.92     & 9.60       & 11.78      & 11.61      &  0.939       \\ 
+		&{\bf run time (s)}                          & 2         & 8       & 31        & 166       & 1,087     &  {\bf 8.856} \\ 
 		\bottomrule
@@
 
-\section{Conclusions}
-\label{sec7}
-
-In this work, we modeled and solved a real air transport problem, named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem}\/ (ACLP+RPDP).
-For the first time in the literature, a NP-hard problem that involves {\it simultaneously}\/ pallet assembly, load balancing and route planning is addressed, where the cost-effectiveness of transport is maximized.
-We adopted some simplifications that are not critical, but that allowed the unprecedented solution of this problem considering more than two nodes.
-
-We consider that, in practical cases, the number $K$\/ of nodes, excluding the base, is small ($K \leq 6$), each of them with hundreds of items to be shipped.
-Thus, considering real aircraft models, we developed some node-by-node heuristics, which allow us to enumerate all $K!$\/ tours and choose the best.
-The complete process can be executed quickly on a simple handheld computer, offering good results and reducing stress for the transport planners.
-As validation, we performed tests in several scenarios with real data from the {\it Brazilian Air Force}.
-
-In less than one hour, this development can establish a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximizing the total score and minimizing fuel consumption for the airlift.
-The output, which includes the tour plan and the pallet building and arrangement plan, is an essential part of airlift: it improves flight safety, makes ground operations more efficient, and makes sure that each item gets to its right destination.
-
-Our main contributions were the mathematical modeling of this complex problem, which involves four NP-hard sub-problems, and a complete process that offers a fast and good-quality solution.
-
-{\color{blue}
-As this is ongoing research, some improvements are possible, which are the next steps of the development:
-
-(1) This work addresses all possible tours, but we plan bounding schemes to allow for discarding most of the higher-cost tours.
-This is expected to reduce the running time, favoring the discovery of better search mechanisms.
-
-(2) We are also working on computer parallelism to accelerate the algorithms.
-
-(3) In the event that these gains in performance are achieved, we strongly believe that they will permit the application of a 3-dimensional packing procedure to maximize the occupancy rate among the selected items.
-}
-
-The simple heuristic $Shims$\/ presented a solution with the best node-by-node results in a restricted time interval.
-Without it, it would not be possible to test all routes: after all, the greater the number of nodes, the shorter the run time limit for node-by-node solution.
-We also show that, as the shipment volume increases, the exact node-by-node solution becomes unfeasible, indicating the importance of a fast and good-quality heuristic.
-
-Finally, by focusing on node-by-node solution, the method of this work is not exclusive to aircraft and airports: it can be adapted, for example, to ships and ports, or vehicles and warehouses, or freight cars and railways.
-In these cases, it would be necessary to make some changes in the modeling: for example, modify the load balancing constraints, and consider the available space in vehicles or freight cars instead of pallets.
-
-\section*{Acknowledgments}
-
-This research was partially supported by \textit{São Paulo Research Foundation} (FAPESP, grant 2016/01860-1).
-
-\begin{thebibliography}{00}
-
-\bibitem{LarsenMikkelsen1979} O. Larsen and G. Mikkelsen, An interactive system for the loading of cargo aircraft, {\it European Journal of Operational Research}, Vol. 4 (6), pp. 367-373, 1980.
-
-\bibitem{Brosh1981} Israel Brosh, Optimal cargo allocation on board a plane: a sequential linear programming approach, {\it European Journal of Operational Research}, Vol. 8 (1), pp. 40-46, 1981.
-
-\bibitem{JohnsonGarey1985} D.S. Johnson and M.R. Garey, A 71/60 theorem for bin packing, {\it Journal of Complexity}, Vol. 1 (1), pp. 65-106, 1985.
-
-\bibitem{Glover1986} Fred Glover, Future paths for integer programming and links to artificial intelligence, {\it Computers and Operations Research}, Vol. 13, pp. 533-549, 1986.
-
-\bibitem{FeoResende1989} Thomas A.
-Feo and Mauricio G.C.
-Resende, A probabilistic heuristic for a computationally difficult set covering problem, {\it Operations Research Letters}, Vol.
-8 (2), pp.
-67-71, 1989.
-
-\bibitem{Kevin1992} Kevin Y.K. Ng, A multicriteria optimization approach to aircraft loading, {\it Operations Research}, Vol. 40 (6), pp. 1200-1205, 1992.
-
-\bibitem{Dorigo1992} Marco Dorigo, {\it Optimization, Learning and Natural Algorithms}, PhD Thesis, Politecnico di Milano, 1992.
-
-\bibitem{CharonHudry1993} I. Charon and O. Hudry, The noising method: a new method for combinatorial optimization, {\it Operations Research Letters}, Vol. 14 (3), pp. 133-137, 1993.
-
-\bibitem{DorigoManiezzoColorni1996} M.
-Dorigo, V.
-Maniezzo and A.
-Colorni, The ant system: optimization by a colony of cooperating agents, {\it IEEE Transactions on Systems, Man, and Cybernetics}, Vol.
-26 (1), pp.
-29-41, 1996.
-
-\bibitem{NiarFreville1997} S.
-Niar and A.
-Freville, A parallel tabu search algorithm for the 0-1 multidimensional knapsack problem, Proceedings of {\it 11th International Parallel Processing Symposium}, 1997.
-
-\bibitem{Heidelberg1998} K.R. Heidelberg, G.S. Parnell and J.E. Ames, Automated air load planning, {\it Naval Research Logistics}, Vol. 45 (8), pp. 751-768, 1998.
-
-\bibitem{CharonHudry2001} I. Charon and O. Hudry, The noising methods: a generalization of some metaheuristics, {\it European Journal of Operational Research}, Vol. 135 (1), pp. 86-101, 2001.
-
-\bibitem{MongeauBes2003} M. Mongeau and C. Bes, Optimization of aircraft container loading, {\it IEEE Transaction on Aerospace and Electronic Systems}, Vol. 39 (1), pp. 140-150, 2003.
-
-\bibitem{fok2004optimizing} M.K.K.
-Fok and A.H.W.
-Chun, Optimizing air cargo load planning and analysis, Proceedings of {\it The International Conference on Computing, Communications and Control Technologies}, 2004.
-
-\bibitem{Fidanova2006} S. Fidanova, Ant colony optimization and multiple knapsack problem and model bias, {\it Numerical Analysis and Its Applications}, Springer Berlin Heidelberg, pp. 280-287, 2005.
-
-\bibitem{Chan2006} F.T.S.
-Chan, R.
-Bhagwat, N.
-Kumar.
-M.K.
-Tiwari and P.
-Lam, Development of a decision support system for air-cargo pallets loading problem: a case study, {\it Expert Systems with Applications}, Vol.
-31 (3), pp.
-472-485, 2006.
-
-\bibitem{KaluznyBohdanL2009Oalb} B.L. Kaluzny and R.H.A.D. Shaw, Optimal aircraft load balancing, {\it International Transactions in Operational Research}, Vol. 16 (6), pp. 767-787, 2009.
-
-\bibitem{MesquitaCunha2011} A.C.P.
-Mesquita and C.B.
-Cunha, An integrated heuristic based on the Scatter Search metaheuristic for vehicle routing problems with simultaneous delivery and pickup in the context of the Brazilian Air Force, {\it Transportes}, Vol.
-19, pp.
-33-42, 2011.
-
-\bibitem{Verstichel2011} J.
-Verstichel, W.
-Vancroonenburg, W.
-Souffriau and G.V.
-Berghe, A mixed integer programming approach to the aircraft weight and balance problem, {\it Procedia Social and Behavioral Sciences}, Vol.
-20, pp.
-1051-1059, 2011.
-
-\bibitem{Limbourg2012} S. Limbourg, M. Schyns and G. Laporte, Automatic aircraft cargo load planning, {\it Journal of the Operational Research Society}, Vol. 63 (9), pp. 1271-1283, 2012.
-
-\bibitem{RoesenerHall2014} A.G.
-Roesener and S.
-Hall, A nonlinear integer programming formulation for the airlift loading problem with insufficient aircraft, {\it Journal of Nonlinear Analysis and Optimization: Theory and Applications}, Vol.
-5 (1), pp.
-125-141, 2014.
-
-\bibitem{Vancroonenburg2014} W.
-Vancroonenburg, J.
-Verstichel, K.
-Tavernier and G.V.
-Berghe, Automatic air cargo selection and weight balancing: a mixed integer programming approach, {\it Transportation Research Part E: Logistics and Transportation Review}, Vol.
-65, pp.
-70-83, 2014.
-
-\bibitem{LurkinSchyns2015} V. Lurkin and M. Schyns, The airline container loading problem with pickup and delivery, {\it European Journal of Operational Research}, Vol. 244 (3), pp. 955-965, 2015.
-
-\bibitem{RoesenerBarnes2016} A.G. Roesener and J.W. Barnes, An advanced tabu search approach to the dynamic airlift loading problem, {\it Logistics Research}, Vol. 9 (1), pp. 1-18, 2016.
-
-\bibitem{PaquaySchynsLimbourg2016} C.
-Paquay, M.
-Schyns and S.
-Limbourg, A mixed integer programming formulation for the three-dimensional bin packing problem deriving from an air cargo application, {\it International Transactions in Operational Research}, Vol.
-23 (1-2), pp.
-187-213, 2016.
-
-\bibitem{PaquayLimbourgSchynsOliveira2018} C.
-Paquay, S.
-Limbourg, M.
-Schyns and J.F.
-Oliveira, MIP-based constructive heuristics for the three-dimensional bin packing problem with transportation constraints, {\it International Journal of Production Research},
-Vol. 56 (4), pp. 1581-1592, 2018.
-
-\bibitem{Laabadi2018} S.
-Laabadi, M.
-Naimi, H.
-El Amri and B.
-Achchab, The 0/1 multidimensional knapsack problem and its variants: a survey of practical models and heuristic approaches, {\it American Journal of Operations Research}, Vol.
-8, pp.
-395-439, 2018.
-
-\bibitem{YangLiuGao2018} Y. Chenguang, L. Hu and G. Yuan, Load planning of transport aircraft based on hybrid genetic algorithm, {\it MATEC Web of Conferences}, Vol. 179, pp. 1-6, 2018.
-
-\bibitem{Alonso2019} M.T.
-Alonso, R.
-Alvarez-Valdes and F.
-Parreno, A GRASP algorithm for multi-container loading problems with practical constraints, {\it A Quarterly Journal of Operations Research}, Vol.
-18, pp.
-49-72, 2019.
-
-\bibitem{BrandtStefan2019} F.
-Brandt and S.
-Nickel, The air cargo load planning problem - a consolidated problem definition and literature review on related problems, {\it European Journal of Operational Research}, Vol.
-275 (2), pp.
-399-410, 2019.
-
-\bibitem{wong2020} E.Y.C.
-Wong and K.K.T.
-Ling, A mixed integer programming approach to air cargo load planning with multiple aircraft configurations and dangerous goods, Proceedings of {\it 7th International Conference on Frontiers of Industrial Engineering}, 2020.
-
-\bibitem{Zhan2020} S. Zhan, L. Wang, Z. Zhang and Y. Zhong, Noising methods with hybrid greedy repair operator for 0-1 knapsack problem, {\it Memetic Computing}, Vol. 12, pp. 37-50, 2020.
-
-\bibitem{eugene2021} E.Y.C.
-Wong, D.Y.
-Mo and S.
-So, Closed-loop digital twin system for air cargo load planning operations, {\it International Journal of Computer Integrated Manufacturing}, Vol.
-34 (7-8), pp.
-801-813, 2021.
-
-\bibitem{zhao2021} X.
-Zhao, Y.
-Yuan, Y.
-Dong and R.
-Zhao, Optimization approach to the aircraft weight and balance problem with the centre of gravity envelope constraints, {\it IET Intelligent Transport Systems}, Vol.
-15 (10), pp.
-1269-1286, 2021.
-
-\bibitem{LopezIbanezManuel2016} López-Ibáñez, Manuel and Dubois-Lacoste, Jérémie and Pérez Cáceres, Leslie and Birattari, Mauro and Stützle, Thomas.
-2016.
-The irace package: Iterated racing for automatic algorithm configuration.
-Elsevier.
-pp.
-43-58.
-Volume 3.
-
-\end{thebibliography}
-
 \end{document}
-
-{\color{blue}
-	Until scenario 4, solving all tours and selecting the best is fast enough for operational use.
-For scenarios 4 and 5 (which are not typical in the Brazilian Air Force), we plan bounding schemes to allow for the discarding of most of the higher-cost tours.
-This is expected to reduce the running time for scenarios 4 and 5, favoring the discovery of better search mechanisms as well as further applications that may require more nodes per tour.
-}
-
-{\color{blue}
-
-	Equations \ref{eq18} and \ref{eq19} were set to group items with the same destinations.
-But, as they mean a combination of $n_{\pi_k}$ items 2 by 2, this may easily lead to huge numbers, $60,000$ to $320,000$ possible combinations, due to the number of items in each node, which together with the other constraints make a MIP solution very hard.
-
-	To work around this, we decided to create the destination attribute for the pallets ($TO_i^{\pi_k}$), to reduce the number of combinations and enable the solution using heuristics.
-$TO^{\pi_k}_i$ denotes that pallet $i$ may assume a different destination in each node ${\pi_k}$.
-
-	On subsequent nodes, the PC can be allocated with other items of same destination \ref{eq:cons5}.
-
-	\begin{equation} \label{eq:cons5}
-		TO_i^{\pi_{k+1}} = to_j;\ X^{\pi_k}_{ij} = 1  \ \mbox{and}\ to_j \in L_{\pi_{k+1}};  \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K-1\}
-	\end{equation}
-
-	\vspace{3mm}
-	At each node, an item (\ref{eq:pdp8}) and the already packed contents (\ref{eq:pdp2}) must only be allocated to a pallet if the destinations are the same.
-
-	\begin{equation} \label{eq:pdp8}
-		TO_i^{\pi_k} = to_j;\ X_{ij}^{\pi_k} = 1; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-	\end{equation}
-
-	\begin{equation} \label{eq:pdp2}
-		TO_i^{\pi_k} = to_q;\ Y_{iq}^{\pi_k} = 1; \ i \in \{1, \ldots, m\};\ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-	\end{equation}
-
-}
-
-{\color{blue}Another possibility to make this approach more general is to devise some bounding schemes to allow discarding most higher-cost tours and allowing an increase in the number of nodes without compromising performance.}

### 2023-06-03 common.py

@@ -154,6 +154,30 @@ class Pallet(object):
 
         return feasible
 
+def copyPallets(pallets):
+    array = [None for _ in pallets]
+    for i, p in enumerate(pallets):
+        array[i] = Pallet(p.ID, p.D, p.V, p.W, 1)
+        array[i].Dest = p.Dest
+        array[i].PCW  = p.PCW 
+        array[i].PCV  = p.PCV
+        array[i].PCS  = p.PCS
+    return array
+
+def copySolDict(solDict):
+    N_M = len(solDict["solMatrix"])
+    solMatrix = mp.Array('i', [0 for _ in np.arange(N_M)] ) 
+    for pos, v in enumerate(solDict["solMatrix"]):
+        solMatrix[pos] = v
+    return dict(solMatrix=solMatrix)
+
+def copyItemsDict(itemsDict):
+    N = len(itemsDict["mpItems"])
+    mpItems = mp.Array('i', [0 for _ in np.arange(N)] ) 
+    for pos, v in enumerate(itemsDict["mpItems"]):
+        mpItems[pos] = v
+    return dict(mpItems=mpItems)
+
 def loadPallets(cfg):
     """
     Load pallets attributes based on aircraft size
@@ -184,6 +208,17 @@ def fillPallet(pallet, items, k, nodeTorque, solDict, cfg, threshold, itemsDict,
             pallet.putItem(  item,               nodeTorque, solDict,      N, itemsDict, lock)
             counter += 1
     return counter
+
+def transform(pallet, items, k, nodeTorque, solDict, cfg, threshold, itemsDict, lock, torqueSurplus=1.0):
+
+    N = len(items)
+
+    for item in items:
+
+        if pallet.isFeasible(item, threshold, k, nodeTorque,   cfg,           itemsDict, lock, torqueSurplus):
+
+            pallet.putItem( item, nodeTorque, solDict, N, itemsDict, lock)
+
 
 def loadDistances(fname):
 

### 2023-06-03 grasp.py

@@ -1,0 +1,111 @@
+
+import common
+from time import time
+import math
+import numpy as np
+from multiprocessing import Queue, Process
+import common
+
+RNG = np.random.default_rng()
+
+def pickFirstEdge(vector):
+    e = vector[0]
+    vector.pop(0)
+    return e
+
+def pickRandomEdge(vector):
+    n = len(vector)
+    i = math.floor(RNG.random()*n)
+    e = vector[i]
+    vector.pop(i)
+    return e
+
+# RCL the Restricted Candidates List
+class RCL(object):
+    def __init__(self, size, nbhood): # nbhood is received sorted by eta
+        self.size = size
+        self.list = []
+        if len(nbhood) >= size:
+            for _ in range(size):
+                # greedilly assemble the RCL
+                e = pickFirstEdge(nbhood)
+                self.list.append(e)
+
+
+def Solve( pallets, items, startTime, cfg, k):  # items include kept on board
+
+    print("\nGRASP for ACLP+RPDP")
+
+    Gbest = common.Solution(pallets, items, 1.0, cfg, k, False) # False: greedy solution
+
+    # initialize the edges between pallets and items
+    Ek = [None] * len(pallets) * len(items)
+    id = 0
+    for p in pallets:
+        for it in items:
+            Ek[id] = common.Edge(id, p, it, cfg)
+            id += 1    
+
+    Ek.sort(key=lambda  x: x.Heuristic, reverse=True)
+
+    L_rcl = math.floor( len(Ek) / (len(Gbest.Edges)) )
+
+    Etemp  = []
+    Eprime = [] 
+
+    stagnant = 0
+    while stagnant < 3 and (time() - startTime) < common.SEC_BREAK:
+
+        Glocal = Gbest
+
+        Eprime = Ek
+        rcl = RCL(L_rcl, Eprime) # edges are drawed from Eprime
+
+        while 1:
+
+            if ((time() - startTime) > common.SEC_BREAK):
+                break
+
+            while 1:
+
+                if ((time() - startTime) > common.SEC_BREAK):
+                    break
+
+                if len(rcl.list) < L_rcl:
+                    break
+
+                ce = pickRandomEdge(rcl.list)
+
+                for e in rcl.list:
+                    Etemp.append(e)
+
+                if Glocal.isFeasible(ce, 1.0, cfg, k): # Check all constraints
+                    Glocal.AppendEdge(ce)
[…]

### 2023-06-03 main.py

@@ -10,6 +10,8 @@ import mpACO
 import optcgcons
 import mipGRB
 import mipCBC
+import tabu
+
 from plots import TTT
 
 # from py3Djanet import Packer, Item, Bin
@@ -140,13 +142,16 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
             mpACO.Solve(  pallets, items, cfg, k, volThreshold, secBreak, "p", nodeTorque, solDict, itemsDict) 
 
         if method == "ACO":       
-            mpACO.Solve(  pallets, items, cfg, k, volThreshold, secBreak, "s", nodeTorque, solDict, itemsDict) 
+            mpACO.Solve(  pallets, items, cfg, k, volThreshold, secBreak, "s", nodeTorque, solDict, itemsDict)
+
+        if method == "TS":       
+            tabu.Solve(  pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict)             
 
         if method == "GRB":
             modStatus, ObjBound = mipGRB.Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict) 
 
         if method == "CBC": # new formulation
-            modStatus, ObjBound = mipCBC.Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict, len(tour.nodes)) 
+            modStatus, ObjBound = mipCBC.Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict) 
 
         if modStatus == 2: # 2: optimal
             numOptDict["numOpt"] += 1
@@ -308,19 +313,20 @@ if __name__ == "__main__":
     plot = False
 
     scenarios = [2,3,4,5,6]
-    # scenarios = [6] # 
+    # scenarios = [2] # 
 
-    # surplus   = "data20"
+    surplus   = "data20"
     # surplus   = "data50"
-    surplus   = "data100"
+    # surplus   = "data100"
 
-    # methods = ["Shims","mpShims","CBC"]  
-    
-    # methods = ["Shims", "mpShims"]
     # methods = ["GRB"]
     # methods = ["CBC"]
-    methods = ["Shims"]
+    # methods = ["Shims"]
     # methods = ["mpShims"]
+    # methods = ["mpACO"]
+    # methods = ["ACO"]
+    methods = ["TS"]
+
 
     # tipo = "KP"
     tipo = "FFD"
[…]

### 2023-06-03 noise.py

@@ -1,0 +1,76 @@
+
+import common
+from time import time
+import math
+import numpy as np
+from multiprocessing import Queue, Process
+import common
+
+RNG = np.random.default_rng()
+
+# ProbAccept is the Noising Method acceptance method.
+# This feature prevents the method from becoming stuck at a local optima.
+def ProbAccept(newScore, oldScore, r):
+	delta = float(newScore-oldScore) / float(oldScore)
+	return delta + (2*r*RNG.random() - r)  > 0 
+
+def Solve( pallets, items, startTime, cfg, k):  # items include kept on board
+
+    print("\nNoising Method Optimization for ACLP+RPDP")
+
+    Gbest  = common.Solution(pallets, items, 1.0, cfg, k, False) # False is Greedy
+    Gprime = common.Solution(pallets, items, 0.7, cfg, k, False)
+
+    r_max = 1 - (Gbest.S - Gprime.S)/Gbest.S # maximum initial noise
+
+    numTrials = math.ceil(len(Gbest.Nbhood)/50)
+
+    numIter = math.ceil(len(Gbest.Edges)/30)
+
+    print(numTrials,numIter,math.ceil(numTrials/numIter))
+
+    step = r_max/(numTrials/numIter-1)
+    r = r_max
+
+    trial = 0
+    while trial < numTrials:
+
+        # restart: copy the best so far local iteration
+        Glocal = Gbest
+        
+        for _ in range(numIter):
+
+            trial += 1
+
+            if ((time() - startTime) > common.SEC_BREAK):
+                trial = numTrials
+                break
+
+            Gprime = Glocal
+
+            oldScore = Gprime.S
+
+            if Gprime.Transform(k, cfg) and ProbAccept(Gprime.S, oldScore, r):
+                Glocal = Gprime
+            
+        for ce in Glocal.Nbhood: # try to improve
+            if Glocal.isFeasible(ce, 1.0, cfg, k): # Check all constraints
+                Glocal.AppendEdge(ce)
+                Glocal.Nbhood.remove(ce)
+
+        if Glocal.S > Gbest.S:
+            Gbest = Glocal
+
+        r -= step
+
+    # decision matrix for which items will be put in which pallet
+    X = [[0 for _ in np.arange(len(items))] for _ in np.arange(len(pallets))]
+    for e in Gbest.Edges:
+        X[e.Pallet.ID][e.Item.ID] = 1
+
+    return X
+
+
+if __name__ == "__main__":
+
+    print("----- Please execute the main py file -------------")

### 2023-06-04 CAIE/paper.tex
Meta-heuristics created

@@
 In this way, the loading and unloading time will be limited only by the use of the equipment at each node.
-Therefore, there will only be waiting time at the base node, corresponding to the runtime of ACLP+RPDP algorithm.
+Therefore, there will only be waiting time at the base node, corresponding to the run time of ACLP+RPDP algorithm.
 For this reason, our method will also be parameterized by the maximum time that can be waited until the start of the load in the base node.
@@
 
-$time$\/ is a runtime limit, which will be distributed among the node-by-node solutions (lines \ref{main:f1} and \ref{main:f2}).
+$time$\/ is a run time limit, which will be distributed among the node-by-node solutions (lines \ref{main:f1} and \ref{main:f2}).
 Each tour consists of $K+1$\/ trips, and the total number of tours is $K!$.
@@
 
-$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $time$\/ is the runtime limit that will be distributed among the $K+1$\/ legs of the tour.
+$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $time$\/ is the run time limit that will be distributed among the $K+1$\/ legs of the tour.
 
@@
 The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
-	\item {\bf Speed-up}: ratio of the sums of the worst runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
+	\item {\bf Speed-up}: ratio of the sums of the worst run times of all scenarios and the sum of the method run times in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
+
+	\item {\bf run time (s)} is the average time to solve a problem instance.
 \end{itemize}
@@
 
-The upper bound is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.
+The upper bound (UB) is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.
 
@@
 \toprule
-{$method$}      & {\bf Results}&{\bf 1}&{\bf 2} & {\bf 3}      & {\bf 4}      & {\bf 5}  & \specialcell{{\bf Normalized}\\{\bf Speed-up}}  \\
-\midrule[.1pt]
-\multirow{2}{*}{{\it Gurobi}}     & Upper bound for $f$      & {\it 7.91} & {\it 11.92} & {\it 13.12} & {\it 14.29} & {\it 13.34} & -  \\ %51,84
-&{\bf run time (s)}                          & 14         & 19          & 73          & 188         & 1,097      & -  \\ 
+$method$& {\bf Scenarios}                 &{\bf 1}&{\bf 2} & {\bf 3}      & {\bf 4}      & {\bf 5}  & \specialcell{{\bf Normalized}\\{\bf Speed-up}}  \\
+\midrule[.1pt]
+{\it Gurobi}     & UB for $f$           & {\it 7.91} & {\it 11.92} & {\it 13.12} & {\it 14.29} & {\it 13.34} & -  \\ %51,84
 \toprule
-\multirow{2}{*}{\textbf{TS}}      & $f$      & 5.37       &             &             &             &            &   \\ % 5.49	 901
-&{\bf run time (s)}                          & 659        &             &             &             &            &   \\ 
-\midrule[.1pt]		
-\multirow{2}{*}{\textbf{GRASP}}   & $f$      &            &             &             &             &            &   \\ 
-&{\bf run time (s)}                          &            &             &             &             &            &   \\ 
-\midrule[.1pt]
-\multirow{2}{*}{\textbf{ACO}}     & $f$      & 5.25       & 5.75        & 5.06        & 4.84        &  4.20      &   \\ % 4.20	 1133
-&{\bf run time (s)}                          & 313        & 326         & 303         & 256         &  1,133     &   \\ 
-\midrule[.1pt]	
-\multirow{2}{*}{\textbf{NMO}}     & $f$      &            &             &             &             &            &   \\
-&{\bf run time (s)}                          &            &             &             &             &            &   \\ 
-\midrule[.1pt]	
-\multirow{2}{*}{\textbf{Shims}}   & $f$      & 7.91       & 11.55       & 12.62       &  13.82      &  14.85          &   \\ % 14.85	 563
-&{\bf run time (s)}                          & 4          &  8          &  33         &  94         &  563          &   \\
+\multirow{2}{*}{\it TS }      & $f$     & 5.37       & 5.86        & 5.25        & 4.62        & 3.60       &   \\ %  3.60	 86
+&{\bf run time (s)}                          & 659        & 361         & 251         & 105         & 86         &   \\ 
+\midrule[.1pt]		
+\multirow{2}{*}{\it GRASP }   & $f$     & 8.35       & 11.94       & 12.89       & 14.10       & 12.98      &   \\ % 8.35	 97	
+&{\bf run time (s)}                          & 97         & 156         & 222         & 109         & 105        &   \\ 
+\midrule[.1pt]
+\multirow{2}{*}{\it ACO}     & $f$      & 5.25       & 5.75        & 5.06        & 4.84        &  4.20      &   \\ % 4.20	 1133
+&{\bf run time (s)}                          & 313        & 326         & 303         & 256         &  453       &   \\ 
+\midrule[.1pt]	
+\multirow{2}{*}{\it NMO}     & $f$      & 8.62       &             &             &             &            &   \\ 
+&{\bf run time (s)}                          & 901        &             &             &             &            &   \\ 
+\midrule[.1pt]	
+\multirow{2}{*}{\it Shims}   & $f$      & 8.43       & 12.77       & 14.47       &  16.28      &  15.47          &   \\
+&{\bf run time (s)}                     & 4          &  8          &  33         &  93         &  570          &   \\
 

### 2023-06-04 main.py

@@ -11,6 +11,8 @@ import optcgcons
 import mipGRB
 import mipCBC
 import tabu
+import grasp
+import noise2
 
 from plots import TTT
 
@@ -147,6 +149,12 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
         if method == "TS":       
             tabu.Solve(  pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict)             
 
+        if method == "GRASP":       
+            grasp.Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict) 
+        
+        if method == "NMO":       
+            noise2.Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict) 
+
         if method == "GRB":
             modStatus, ObjBound = mipGRB.Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict) 
 
@@ -250,8 +258,9 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
         tour.AvgVol    += nodeVol
         tour.AvgTorque += epsilon
 
-        print(f"----- node {node.ICAO},", end='')
-        print(f" score {tour.score:.0f}, cost {tour.cost:.0f}, vol {nodeVol:.2f}, epsilon {epsilon:.2f} -----")
+        print(f"\tnode {node.ICAO},", end='')
+        print(f" score {tour.score:.0f}, cost {tour.cost:.0f}, vol {nodeVol:.2f}, epsilon {epsilon:.2f}")
+        
 
         if writeConsFile:
 
@@ -312,8 +321,13 @@ if __name__ == "__main__":
     # plot = True
     plot = False
 
+    testing = False
+    # testing = True
+
     scenarios = [2,3,4,5,6]
-    # scenarios = [2] # 
+
+    if testing:
+        scenarios = [2]
 
     surplus   = "data20"
     # surplus   = "data50"
@@ -321,11 +335,13 @@ if __name__ == "__main__":
 
     # methods = ["GRB"]
     # methods = ["CBC"]
-    # methods = ["Shims"]
+    methods = ["Shims"]
     # methods = ["mpShims"]
     # methods = ["mpACO"]
     # methods = ["ACO"]
-    methods = ["TS"]
+    # methods = ["TS"]
+    # methods = ["GRASP"]
+    # methods = ["NMO"]
 
 
[…]

### 2023-06-04 mpShims.py

@@ -30,7 +30,7 @@ class Shims(object):
         if item.To != self.Pallet.Dest[k]:
             return False
 
-        if self.Pallet.PCV + self.SCV + item.V > self.Pallet.V * 1.02:
+        if self.Pallet.PCV + self.SCV + item.V > self.Pallet.V:
             return False
 
         if self.Pallet.PCW + self.SCW + item.W > self.Pallet.W:
@@ -45,22 +45,21 @@ class Shims(object):
         return ret
 
 # create a set of shims for this pallet and selects the best shims
-def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict, lock, tipo):
-
-    maxVol = pallet.V * surplus
+def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict, lock, tipo, startTime, secBreak):
 
     vol = 0.
     whip = []
     N = len(items)
     i = pallet.ID
    
+   # create the whip
     with lock:   
         for item in items:
             j = item.ID
             if itemsDict["mpItems"][j] == 0:
                 vol += item.V
                 whip.append(item)
-                if vol > maxVol:
+                if vol > pallet.V * surplus:
                     break
 
     # create the first shim
@@ -80,7 +79,7 @@ def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict,
 
             for i in indexes:
                 item = whip[i]
-                if pallet.isFeasible(item, 1.1, k, nodeTorque,  cfg, itemsDict, lock):
+                if pallet.isFeasible(item, 1.0, k, nodeTorque,  cfg, itemsDict, lock):
                     pallet.putItem(item, nodeTorque, solDict, N, itemsDict, lock)
 
     if tipo == "FFD":
@@ -88,6 +87,10 @@ def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict,
         whip.sort(key=lambda x: abs(x.V), reverse=True)
 
         for w, item in enumerate(whip):
+
+            if ((time.perf_counter() - startTime) > secBreak):
+                break 
+
             newShims = True
             for sh in Set:
                 if sh.isFeasible(item, k, nodeTorque, cfg, lock):
@@ -109,11 +112,13 @@ def getBestShims(pallet, items, k, nodeTorque, solDict, cfg, surplus, itemsDict,
                 bestIndex = i
         # put the best Shim in the solution
[…]

### 2023-06-04 noise2.py

@@ -1,0 +1,153 @@
+
+import common
+import time
+import math
+import numpy as np
+import multiprocessing as mp
+import common
+# import random
+
+RNG = np.random.default_rng()
+
+def Transform(iterPallets, iterItemsDict, iterSolDict, iterScore, N, items, nodeTorque, k, cfg, lock):
+
+    for i, _ in enumerate(iterPallets):
+
+        arr0 = [j for j, a in enumerate(iterItemsDict["mpItems"])                if a == 0]
+        arr1 = [j for j, a in enumerate(iterSolDict["solMatrix"][N*i:N*i+(N-1)]) if a == 1]
+
+        if len(arr0) > 0 and len(arr1) > 0:
+            # id0 = np.random.choice(arr0) # not in solution
+            # id1 = np.random.choice(arr1) # in this pallet
+
+            id0 = RNG.choice(arr0) # not in solution
+            id1 = RNG.choice(arr1) # in this pallet            
+
+            iterPallets[i].popItem(items[id1], nodeTorque, iterSolDict, N, iterItemsDict)
+            
+            if iterPallets[i].isFeasible(items[id0], 1.0, k, nodeTorque, cfg, iterItemsDict, lock, torqueSurplus=1.0):
+                iterPallets[i].putItem(items[id0], nodeTorque, iterSolDict, N, iterItemsDict, lock)
+            else:
+                iterPallets[i].putItem(items[id1], nodeTorque, iterSolDict, N, iterItemsDict, lock)
+
+
+        iterScore += iterPallets[i].PCS
+
+
+
+# ProbAccept is the Noising Method acceptance method.
+# This feature prevents the method from becoming stuck at a local optima.
+def ProbAccept(newScore, oldScore, r):
+	delta = float(newScore-oldScore) / float(oldScore)
+	return delta + (2*r*RNG.random() - r)  > 0 
+
+
+def Solve(pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict):
+
+    startTime = time.perf_counter()
+
+    N = len(items)
+    M = len(pallets)
+
+    print("\nNoising Method Optimization for ACLP+RPDP\n")        
+
+    lock  = mp.Lock() # for use in parallel mode
+
+    print(f"{N} items  {M} pallets")
+
+    initScore = 0.0 # G
+
+    for i, _ in enumerate(pallets):               
[…]

### 2023-06-05 CAIE/paper.tex
NMO numTrials adjusted
MIPgap

@@
 
-			\State $f_{\pi_{TSP}} \gets SolveTour(\pi_{TSP}, L, M, C, N, method, time/(K+1))$ \label{main:f1}
-			\State $answer1[scenario,surplus,method] \gets f_{\pi_{TSP}}$ \label{main:a1} \Comment{Results corresponding to $\pi_{TSP}$}
+			\State {\color{blue} $f_{\pi_{dir}} \gets SolveTour(\pi_{dir}, L, M, C, N, method, time)$ } \label{main:fdirect}
+
+		    \State {\color{blue} $f_{\pi_{rev}} \gets SolveTour(\pi_{rev}, L, M, C, N, method, time)$ } \label{main:freverse}
+
+			\State {\color{blue} $answer1[scenario,surplus,method] \gets max( f_{\pi_{dir}}, f_{\pi_{rev}}  )$ } \label{main:a1}
+
 			\For {each $\pi \in S_K$} \label{main:loop1} \Comment{$\pi$ is a permutation of nodes (a tour)}
-				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, time/K!)$ \Comment{$f_{\pi}$\/ is a solution to the tour $\pi$} \label{main:f2}
+
+				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, time/(K+1)!)$ \Comment{$f_{\pi}$\/ is a solution to the tour $\pi$} \label{main:f2}
+
 			\EndFor \label{main:loop2}
@@
 \end{algorithm}
+
+{\color{blue} In line \ref{main:fdirect}, $SolveTour$ solves the lowest-cost tour, and in line 8, its reverse. }
 
@@
 
-{\color{red}
+{\color{blue}
 
@@
 
+{\color{red}
+Let $X_{ij}^k$\/ and $Y_{iq}^k$\/ be CONTINUOUS variables, where $1 \leq i \leq m$, $1 \leq j \leq n_k$, $1 \leq q \leq m_k$\/ and $0 \leq k \leq K$.
+}
 The upper bound (UB) is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.
 
-We solved all scenarios and instances with Gurobi in a relaxed linear model without a time limit.
-Although the solutions are not guaranteed to be feasible, it is good for the reader to have an idea of the possible upper bound for the metaheuristic solutions.
-
-Note that, in Table \ref{tab:20} all $Gurobi\ f$\/ values are italicized to indicate that these values may not be optimal or feasible as the decision variables are not integer.
+We solved all scenarios and instances with Gurobi in a {\it MIPGap} parameter equal to 1\% to control the minimal quality of the returned solution, without a time limit.
+
+Note that, in Table \ref{tab:20} all $Gurobi\ f$\/ values are italicized to indicate that these values are at most 1\% distant from the optimal solution.
 
 }
+
+\begin{table}[H]
+\centering
+\caption{Methods and scenarios with $surplus =1.2$ and the least cost tours criteria}  \label{tab:20leastcost}
+\footnotesize
+\begin{tabular}{c|c|cccc}
+\toprule
+$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4}  \\
+\toprule
+\multirow{2}{*}{\it Gurobi* } & $f$ & 7.90    & 11.46   & 12.82   & 13.40     \\
+&               {\bf run time (s)}  & 10      & 8       & 11      & 11      \\ 
+\midrule[.1pt]	
+\multirow{2}{*}{\it Shims}    & $f$ & 7.83    & 11.34   & 12.77   & 13.20    \\
+&         {\bf run time (s)}        & 4       & 5       &  6      & 7        \\ 
+\bottomrule
+\multicolumn{6}{c}{*integer relaxed variables}
+\end{tabular}
+\normalsize
+\end{table}
 
@@
 \midrule[.1pt]
-{\it Gurobi}     & UB for $f$           & {\it 7.91} & {\it 11.92} & {\it 13.12} & {\it 14.29} & {\it 13.34} & -  \\ %51,84
+{\it Gurobi}     & UB for $f$           & {\it 8.91} & {\it 12.92} & {\it 15.12} & {\it 16.29} & {\it 15.34} & -  \\ %51,84
 \toprule
-\multirow{2}{*}{\it TS }      & $f$     & 5.37       & 5.86        & 5.25        & 4.62        & 3.60       &   \\ %  3.60	 86
-&{\bf run time (s)}                          & 659        & 361         & 251         & 105         & 86         &   \\ 
+
 \midrule[.1pt]	
-\multirow{2}{*}{\it GRASP }   & $f$     & 8.35       & 11.94       & 12.89       & 14.10       & 12.98      &   \\ % 8.35	 97	
-&{\bf run time (s)}                          & 97         & 156         & 222         & 109         & 105        &   \\ 
-\midrule[.1pt]
-\multirow{2}{*}{\it ACO}     & $f$      & 5.25       & 5.75        & 5.06        & 4.84        &  4.20      &   \\ % 4.20	 1133
-&{\bf run time (s)}                          & 313        & 326         & 303         & 256         &  453       &   \\ 
-\midrule[.1pt]	
-\multirow{2}{*}{\it NMO}     & $f$      & 8.62       &             &             &             &            &   \\ 
-&{\bf run time (s)}                          & 901        &             &             &             &            &   \\ 
-\midrule[.1pt]	
-\multirow{2}{*}{\it Shims}   & $f$      & 8.43       & 12.77       & 14.47       &  16.28      &  15.47          &   \\
-&{\bf run time (s)}                     & 4          &  8          &  33         &  93         &  570          &   \\
+\multirow{2}{*}{\it Shims}   & $f$      & 8.43       & 12.77       & 14.47       & 16.28       & 15.15          &   \\
+&{\bf run time (s)}                     & 4          &  8          &  33         & 93          & 560          &   \\
 
@@
 
-\begin{table}[H]
-	\centering
-	\caption{Methods and scenarios with $surplus =1.5$}  \label{tab:50}
-	\footnotesize
-	\begin{tabular}{c|c|ccccc|c}
-		\toprule
-		{$method$}                        & {\bf Results}   & {\bf 1}     &  {\bf 2}    & {\bf 3}      & {\bf 4}      & {\bf 5}    & \specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
-		\toprule
-		\multirow{2}{*}{\textbf{Gurobi}}  & $f$     & 4.756       & 8.84        & 11.94       & 16.21       &{\bf 16.47} & 0.852  \\	% 6828 
-		&{\bf run time (s)}                         & 18        & 68         & 326        & 1,906      & 12,960    & 1.000  \\	 
-		\midrule[.1pt]
-		\multirow{2}{*}{\textbf{TS}}      & $f$     & 8.97       & 11.44       & 12.92       & 15.20       & 15.39      &  0.936 \\		
-		&{\bf run time (s)}                         & 4         & 18         & 88         & 525        & 3,660     &  3.555  \\		
-		\midrule[.1pt]		
-		\multirow{2}{*}{\textbf{GRASP}}   & $f$     & 9.36       & 12.34       & 13.63       & 16.18       & 15.44      &  0.981 \\ 
-		&{\bf run time (s)}                         & 4         & 17         & 86         & 515        & 3,564     &  3.651  \\
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{ACO}}     & $f$     & 9.36       & 12.34       & 13.63       & 16.18       & 15.44      &  0.981 \\
-		&{\bf run time (s)}                         & 5         & 19         & 90         & 539        & 3,720     &  3.492  \\	
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{NMO}}     & $f$     & 9.36       & 12.33       & 13.60       & 16.16       & 15.43      &  0.979 \\ 
-		&{\bf run time (s)}                         & 5         & 19         & 91         & 544        & 3,780     & 3.442  \\
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{Shims}}   & $f$     & {\bf 9.39} & {\bf 12.42} & {\bf 13.72} & {\bf 16.28} & 15.49      &  {\bf 0.986} \\	
-		&{\bf run time (s)}                         & 3         & 11         & 48         & 265        & 1,720     &  {\bf 7.470}  \\
-		\bottomrule
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\begin{table}[H]
-	\centering
-	\caption{Methods and scenarios with $surplus = 2.0$}  \label{tab:100}
-	\footnotesize
-	\begin{tabular}{c|c|ccccc|c}
-		\toprule
-		{$method$}                        & {\bf Results}    & {\bf 1}     &  {\bf 2}   & {\bf 3}    & {\bf 4}    & {\bf 5}      & \specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
-		\toprule
-		\multirow{2}{*}{\textbf{Gurobi}}  & $f$              & 4.92      & 6.58      & 6.34       & 7.24       & 8.15      &  0.334 \\ % 9951
-		                                  &{\bf run time (s)}& 29       & 106      & 493      & 2,833     & 19,380      &  1.000  \\ 
-		\midrule[.1pt]
-		\multirow{2}{*}{\textbf{TS}}      & $f$              & 13.72     & 17.53     & 19.28     & 21.78      & 22.35      &  0.951 \\		
-		&{\bf run time (s)}                                  & 5        & 19       & 93       & 556       & 3,900     &  4.992  \\	
-		\midrule[.1pt]		
-		\multirow{2}{*}{\textbf{GRASP}}   & $f$        & 13.99     & 19.39     & 19.86     & 23.50      & 22.57      &  0.998 \\
-		&{\bf run time (s)}                            & 6        & 21       & 97       & 571       & 3,960     &  4.906  \\	
-		\midrule[.1pt]
-		\multirow{2}{*}{\textbf{ACO}}     & $f$        & 13.99     & 19.39     & 19.86     & 23.50      & 22.57      &  0.998 \\
-		&{\bf run time (s)}                            & 6        & 23       & 105      & 597       & 4,080     &  4.745  \\	
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{NMO}}     & $f$        & 13.99     & 19.39     & 19.85     & 23.50      & 22.56      & 0.998 \\	
-		&{\bf run time (s)}                            & 9        & 27       & 120      & 681       & 4,620     & 4.184  \\	
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{Shims}}   & $f$        &{\bf 14.00} & {\bf 19.44} & {\bf 19.94} & {\bf 23.55} & {\bf 22.58} & {\bf 1.000} \\	
-		&{\bf run time (s)}                            & 7        & 20       & 84       & 450       & 2,918     & {\bf 6.570}  \\
-		\bottomrule
-	\end{tabular}
-	\normalsize
-\end{table}
-
 {\color{blue}
@@
 }
-
-In Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, $Gurobi$\/ was the slowest method and $Shims$\/ was the fastest.
-This was expected, as $Gurobi$\/ is a MIP solver, which seeks an optimal solution within the established constraints, usually reaching its time limit on each node, while {\it Shims} is a fast heuristic.
-This way, $Gurobi$\/ even demands prohibitive times (more than 5 hours in scenario 6 of Table \ref{tab:100}), while the maximum time required by $Shims$\/ did not reach one hour.
-
-Considering the {\bf Normalized}\/ values, in Table \ref{tab:20} $Gurobi$\/ was the best ($0.997$), followed closely by $Shims$\/ ($0.939$).
-In Tables \ref{tab:50} and \ref{tab:100}, $Shims$ was the best ($0.986$\/ and $1.000$).
-As expected, the performance of $Gurobi$\/ drops sharply as the volume of shipments at the nodes increases: $0.852$\/ in Table \ref{tab:50} and $0.334$\/ in Table \ref{tab:100}.
 

### 2023-06-05 mipGRB.py

@@ -67,9 +67,9 @@ def Solve( pallets, items, cfg, k, secBreak, nodeTorque, solDict, itemsDict):
 
         # Pallet weight constraint                      PCW: pallet current weight
         itemsWeights[i] = sum(X[i][j] * items[j].W  for j in set_N)
-        # mod.addConstr(
-        #     itemsWeights[i] + pallets[i].PCW <= pallets[i].W
-        # )
+        mod.addConstr(
+            itemsWeights[i] + pallets[i].PCW <= pallets[i].W
+        )
 
         # Pallet volume constraint
         itemsVolumes[i] = sum(X[i][j] * items[j].V  for j in set_N)

### 2023-06-06 CAIE/paper.tex
Table 1.5

@@
 	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+
+\usepackage{textcomp}
+\usepackage{graphicx} 
+\setlength{\fboxsep}{0.005pt}
+\newcommand{\tmpframe}[1]{\fbox{#1}}
 
@@
 	\caption{Testing scenarios}  \label{tab:scenarios}
-	\begin{tabular}{c c c }
+	\begin{tabular}{c c c c}
 		\toprule
-		{\bf Scenario} & {$K$} & {$L$} \\		
+		{\bf Scenario} & {$K$} & {$L$} & {Shortest tours} \\		
 		\midrule
-		1 & 2    & \{$1,2$\}           \\
-		2 & 3    & \{$1,2,3$\}         \\
-		3 & 4    & \{$1,2,3,4$\}       \\
-		4 & 5    & \{$1,2,3,4,5$\}     \\
-		5 & 6    & \{$1,2,3,4,5,6$\}   \\
+		1 & 2    & \{$1,2$\}           & 0 1 2 0 \\
+		2 & 3    & \{$1,2,3$\}         & 0 1 2 3 0 \\
+		3 & 4    & \{$1,2,3,4$\}       & 0 4 1 2 3 0 \\
+		4 & 5    & \{$1,2,3,4,5$\}     & 0 4 1 2 5 3 0 \\
+		5 & 6    & \{$1,2,3,4,5,6$\}   & 0 4 1 2 6 5 3 0 \\
 		\bottomrule
@@
 
-As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's RAM (16 GiB) too early, we decided to execute a relaxed model with the decision variables set to CONTINUOUS (a Gurobi parameter) to obtain the upper bound for the objective function.
-This approach is much faster than using binary decision variables.
-
-{\color{red}
-Let $X_{ij}^k$\/ and $Y_{iq}^k$\/ be CONTINUOUS variables, where $1 \leq i \leq m$, $1 \leq j \leq n_k$, $1 \leq q \leq m_k$\/ and $0 \leq k \leq K$.
-}
-The upper bound (UB) is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.
-
-We solved all scenarios and instances with Gurobi in a {\it MIPGap} parameter equal to 1\% to control the minimal quality of the returned solution, without a time limit.
-
-Note that, in Table \ref{tab:20} all $Gurobi\ f$\/ values are italicized to indicate that these values are at most 1\% distant from the optimal solution.
+As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's RAM (16 GiB) too early, we decided to set the Gurobi {\it MIPgap} parameter to 1\% to control the minimal quality of the returned solutions and abbreviate the solution time.
+
+More details on the Gurobi {\it MIPgap} refer to \cite{mipgap}.
 
@@
 \centering
-\caption{Methods and scenarios with $surplus =1.2$ and the least cost tours criteria}  \label{tab:20leastcost}
+\caption{Solutions with $surplus =1.2$\/ with the direct and reverse least cost tours.}  \label{tab:20}
 \footnotesize
-\begin{tabular}{c|c|cccc}
+\begin{tabular}{c|c|ccccc|c}
 \toprule
-$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4}  \\
+$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
 \toprule
-\multirow{2}{*}{\it Gurobi* } & $f$ & 7.90    & 11.46   & 12.82   & 13.40     \\
-&               {\bf run time (s)}  & 10      & 8       & 11      & 11      \\ 
+\multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\ % 5927
+&               {\bf run time (s)}  & 17      & 17       & 16      & 17   & 26        &  1.0 \\ % 93
 \midrule[.1pt]	
-\multirow{2}{*}{\it Shims}    & $f$ & 7.83    & 11.34   & 12.77   & 13.20    \\
-&         {\bf run time (s)}        & 4       & 5       &  6      & 7        \\ 
+\multirow{2}{*}{\it Shims}    & $f$ & 7.83    & 11.34   & 12.77   & 13.20 & 12.22     &  0.97 \\ % 5736
+&         {\bf run time (s)}        & 4       & 5       &  6      & 7     & 10        &  2.9  \\ % 32
+
 \bottomrule
-\multicolumn{6}{c}{*integer relaxed variables}
+\multicolumn{8}{c}{Average $f$ and run time values of K! tours.}\\
+\toprule
+
+\multirow{2}{*}{\it Gurobi* } & $f$ & 8.51    & 12.25   & 13.32   & 14.61  & x     &  \\ % 5927
+&               {\bf run time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\ % 93
+\midrule[.1pt]	
+\multirow{2}{*}{\it Shims}    & $f$ & 8.11    & 12.18   & 13.25   & 14.57  & 14.33     &  0.97 \\ % 5736
+&         {\bf run time (s)}        & 4       & 8       & 33      & 95     & 572       &  2.9  \\ % 32
+\bottomrule
+\multicolumn{8}{c}{*{\it MIPgap} parameter set to 1\%. The $f$ values are \textpm 1\% distant from the optimal solution}\\
+
 \end{tabular}
@@
 
+97\% of 16GiB was the Read-Only Memory (RAM) state before we aborted the Gurobi execution on solving scenario 5 in Table \ref{tab:20}.
+
+\begin{figure}[H]
+	\centering
+	\tmpframe{\includegraphics[scale=0.6]{Images/memory.png}}
+	\captionof{figure}{RAM usage on solving scenario 5 by Gurobi (Table \ref{tab:20}).}
+	\label{fig:memory}
+\end{figure}
+
 \begin{table}[H]
 	\centering
-\caption{Methods and scenarios with $surplus =1.2$}  \label{tab:20}
+	\caption{Solutions with $surplus=1.5$\/ with the direct and reverse least cost tours.}  \label{tab:50}
 	\footnotesize
@@
 		\toprule
-$method$& {\bf Scenarios}                 &{\bf 1}&{\bf 2} & {\bf 3}      & {\bf 4}      & {\bf 5}  & \specialcell{{\bf Normalized}\\{\bf Speed-up}}  \\
+		$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		\toprule
+		\multirow{2}{*}{\it Gurobi* } & $f$ &     &    &    &   &   &  \\ % 
+		&               {\bf run time (s)}  &     &    &    &   &   &   \\ % 
 		\midrule[.1pt]	
-{\it Gurobi}     & UB for $f$           & {\it 8.91} & {\it 12.92} & {\it 15.12} & {\it 16.29} & {\it 15.34} & -  \\ %51,84
+		\multirow{2}{*}{\it Shims}    & $f$ &     &    &    &   &   &   \\ % 
+		&         {\bf run time (s)}        &     &    &    &   &   &    \\ %
+
+		\bottomrule
+		\multicolumn{8}{c}{Average $f$ and run time values of K! tours.}\\
 		\toprule
 
+		\multirow{2}{*}{\it Gurobi* } & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   &  \\ % 
+		&               {\bf run time (s)}  & 53    & 61    & 193   & 470   & 2241    &   \\ % 
 		\midrule[.1pt]	
-\multirow{2}{*}{\it Shims}   & $f$      & 8.43       & 12.77       & 14.47       & 16.28       & 15.15          &   \\
-&{\bf run time (s)}                     & 4          &  8          &  33         & 93          & 560          &   \\
-
+		\multirow{2}{*}{\it Shims}    & $f$ & 11.33 & 16.97 & 17.56 & 20.43 & 17.94   &   \\ % 
+		&         {\bf run time (s)}        & 4     & 8     & 34    & 96    & 591     &   \\ % 
 		\bottomrule
+		\multicolumn{8}{c}{*{\it MIPgap} parameter set to 1\%. The $f$ values are \textpm 1\% distant from the optimal solution}\\
+
 	\end{tabular}
@@
 For this achievement, we ran {\it Shims} in 5 scenarios with 7 instances each, and 3 volume surpluses, collecting the averages of the best tour result and the least cost tour result.
+
+ESTA TABELA DEVERÁ REFEITA DEPOIS QUE AS ANTERIORES ESTIVEREM PRONTAS.
 
@@
 
+\bibitem{mipgap} What is the MIPGap? \url{https://support.gurobi.com/hc/en-us/articles/8265539575953-What-is-the-MIPGap-}, 2023.
+
 \end{thebibliography}

### 2023-06-06 main.py

@@ -328,21 +328,22 @@ if __name__ == "__main__":
 
 
     # testing    = True
-    leastCost  = True
+    # leastCost  = True
     # plot = True 
 
-    scenarios = [2,3,4,5]
+    scenarios = [2,3,4,5,6]
+    # scenarios = [6]
 
     if testing:
         scenarios = [2]
 
-    surplus   = "data20"
-    # surplus   = "data50"
+    # surplus   = "data20"
+    surplus   = "data50"
     # surplus   = "data100"
 
     # methods = ["GRB"]
-    methods = ["CBC"]
-    # methods = ["Shims"]
+    # methods = ["CBC"]
+    methods = ["Shims"]
     # methods = ["mpShims"]
     # methods = ["mpACO"]
     # methods = ["ACO"]
@@ -509,6 +510,7 @@ if __name__ == "__main__":
                 str = f"{leastSC/numInst:.2f}\t {instanceSC/numInst:.2f}\t {avgTime:.0f}\t {avgTime2:.0f}\t {worstTime:.1f}\t {bestAV:.2f}\t {bestAT:.2f}\t {numOptDict['numOpt']:.1f}\t {beforeDict['value']:.1f} & {afterDict['value']:.1f} & {percent:.1f}\n"
                 # instances average
                 writeAvgResults(method, scenario, str, surplus)
+
                 print(f"\n{str}")
                 print(f"{surplus}")
                 print(f"{len(tours)} tours")

### 2023-06-06 CAIE/paper.tex
Editing results

@@
 
-\begin{equation} \label{eq:if1}
-	to_i^{\pi_k} = to_j;\ X_{ij} = 1; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
 \begin{equation} \label{eq:if2}
 	X_{ij}       = 0;\ to_j \notin L_{\pi_k}; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
+\end{equation}
+
+\begin{equation} \label{eq:ifA}
+  X_{ij} \leq X_{ij} \times (T_i^{\pi_k} - to_j + 1); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
+\end{equation}
+
+\begin{equation} \label{eq:ifB}
+	X_{ij} \leq X_{ij} \times (to_j - T_i^{\pi_k} + 1 ); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
 \end{equation}
@@
 
-\multirow{2}{*}{\it Gurobi* } & $f$ & 8.51    & 12.25   & 13.32   & 14.61  & x     &  \\ % 5927
-&               {\bf run time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\ % 93
+\multirow{2}{*}{\it Gurobi* } & $f$ & 8.51    & 12.25   & 13.32   & 14.61  & x        & 1.00 \\ % 4869
+&               {\bf run time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\ % 479
 \midrule[.1pt]	
-\multirow{2}{*}{\it Shims}    & $f$ & 8.11    & 12.18   & 13.25   & 14.57  & 14.33     &  0.97 \\ % 5736
-&         {\bf run time (s)}        & 4       & 8       & 33      & 95     & 572       &  2.9  \\ % 32
+\multirow{2}{*}{\it Shims}    & $f$ & 8.11    & 12.18   & 13.25   & 14.57  & 14.33    &  0.99 \\ % 4811
+&         {\bf run time (s)}        & 4       & 8       & 33      & 95     & 572      &  3.4  \\ % 140
 \bottomrule
@@
 		\toprule
-		\multirow{2}{*}{\it Gurobi* } & $f$ &     &    &    &   &   &  \\ % 
-		&               {\bf run time (s)}  &     &    &    &   &   &   \\ % 
+		\multirow{2}{*}{\it Gurobi* } & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\ % 8236
+		&               {\bf run time (s)}  & 38    & 30    & 27    & 29    & 81    & 1.00  \\ % 205
 		\midrule[.1pt]	
-		\multirow{2}{*}{\it Shims}    & $f$ &     &    &    &   &   &   \\ % 
-		&         {\bf run time (s)}        &     &    &    &   &   &    \\ %
+		\multirow{2}{*}{\it Shims}    & $f$ & 10.89 & 16.65 & 17.80 & 18.69 & 16.45 & 0.98  \\ % 8048
+		&         {\bf run time (s)}        & 4     & 5     & 6     & 7     & 8     & 7.1   \\ % 29
 
@@
 
-		\multirow{2}{*}{\it Gurobi* } & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   &  \\ % 
+		\multirow{2}{*}{\it Gurobi* } & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ % 8602
+		&               {\bf run time (s)}  & 53    & 61    & 193   & 470   & 2241    & 1.0  \\ % 3018
+		\midrule[.1pt]	
+		\multirow{2}{*}{\it Shims}    & $f$ & 11.33 & 16.97 & 17.56 & 20.43 & 17.94   & 0.98  \\ % 8423
+		&         {\bf run time (s)}        & 4     & 8     & 34    & 96    & 591     & 4.1 \\ % 733
+		\bottomrule
+		\multicolumn{8}{c}{*{\it MIPgap} parameter set to 1\%. The $f$ values are \textpm 1\% distant from the optimal solution}\\
+
+	\end{tabular}
+	\normalsize
+\end{table}
+
+\begin{table}[H]
+	\centering
+	\caption{Solutions with $surplus=2.0$\/ with the direct and reverse least cost tours.}  \label{tab:100}
+	\footnotesize
+	\begin{tabular}{c|c|ccccc|c}
+		\toprule
+		$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		\toprule
+		\multirow{2}{*}{\it Gurobi* } & $f$ &  &  &  &  &  &  \\ % 
+		&               {\bf run time (s)}  & 38    & 30    & 27    & 29    & 81    &   \\ % 
+		\midrule[.1pt]	
+		\multirow{2}{*}{\it Shims}    & $f$ &  &  &  &  &  &   \\ % 
+		&         {\bf run time (s)}        & 4     & 5     & 6     & 7     & 8     &    \\ %
+
+		\bottomrule
+		\multicolumn{8}{c}{Average $f$ and run time values of K! tours.}\\
+		\toprule
+
+		\multirow{2}{*}{\it Gurobi* } & $f$ &  &  &  &  &    &  \\ % 
 		&               {\bf run time (s)}  & 53    & 61    & 193   & 470   & 2241    &   \\ % 
 		\midrule[.1pt]	
-		\multirow{2}{*}{\it Shims}    & $f$ & 11.33 & 16.97 & 17.56 & 20.43 & 17.94   &   \\ % 
+		\multirow{2}{*}{\it Shims}    & $f$ &  &  &  &  &    &   \\ % 
 		&         {\bf run time (s)}        & 4     & 8     & 34    & 96    & 591     &   \\ % 

### 2023-06-06 CAIE/results.tex
Editing results

@@
 
+\usepackage{textcomp}
+\usepackage{graphicx} 
+\setlength{\fboxsep}{0.005pt}
+\newcommand{\tmpframe}[1]{\fbox{#1}}
+
 \journal{Computers \& Industrial Engineering}
@@
 
-{\color{blue}
+\subsection{Results obtained}
 
-\subsection{Results obtained}
 In the tests performed, we used a 64-bit, 16GB, 3.6GHz, eight-core processor with {\it Linux Ubuntu 22.04.1 LTS 64-bit} as the operational system and {\it Python 3.10.4}\/ as the programming language.
@@
 We ran Algorithm \ref{alg:main} in the 5 scenarios described in Table \ref{tab:scenarios}, considering 6 methods for node-by-node solution: Gurobi (see \ref{solver}), ACO, NMO, TS, GRASP, and Shims (Algorithm \ref{alg:shims}).
-In generating the items in each node, we consider 3 values for the parameter $volume$.
-The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:20} ($volume = 1.2$), \ref{tab:50} ($volume = 1.5$) and \ref{tab:100} ($volume = 2.0$).
-For each $method$, $scenario$\/ and $volume$, 7 different instances were generated.
+In generating the items in each node, we consider 3 values for the parameter $surplus$.
+The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:20} ($surplus = 1.2$), \ref{tab:50} ($surplus = 1.5$) and \ref{tab:100} ($surplus = 2.0$).
+For each $method$, $scenario$\/ and $surplus$, 7 different instances were generated.
 Therefore, 105 tests (5 scenarios $\times$\/ 3 volumes $\times$\/ 7 instances) were performed for each method.
@@
 The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
-	\item {\bf Speed-up}: ratio of the sums of the worst runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
+	\item {\bf Speed-up}: ratio of the sums of the worst run times of all scenarios and the sum of the method run times in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
+
+	\item {\bf run time (s)} is the average time to solve a problem instance.
 \end{itemize}
 
-In each $scenario$, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.
+In each $scenario$\/, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.
+
+As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's Read-Access Memory (RAM) (16 GiB) too early, we decided to set the Gurobi {\it MIPgap} parameter to 1\% to control the minimal quality of the returned solutions and abbreviate the solution time.
+
+For more details on the Gurobi {\it MIPgap}, refer to \cite{mipgap}.
 
@@
 \centering
-	\caption{Methods and scenarios with $volume =1.2$}  \label{tab:20}
+\caption{Solutions with volume $surplus =1.2$.}  \label{tab:20}
 \footnotesize
-	\begin{tabular}{c|c|ccccc|c}
+\begin{tabular}{ccccccccc}
 \toprule
-		{$method$}                        & {\bf Results}    & {\bf 1}     &  {\bf 2}    & {\bf 3}      & {\bf 4}      & {\bf 5}  & \specialcell{{\bf Normalized}\\{\bf Speed-up}}  
-		\\
+{\bf Tours}&$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
 \toprule
-		\multirow{2}{*}{\textbf{Gurobi}}  & $f$      & 6.10   & {\bf 10.02}& {\bf 10.82} & {\bf 12.13} & {\bf 12.10} & {\bf 0.997} \\ %51,84
-		&{\bf run time (s)}                          & 14     & 51         & 243         & 1,417       & 9,720       &  1.000 \\ 
-		\midrule[.1pt]
-		\multirow{2}{*}{\textbf{TS}}      & $f$      & 6.56     & 8.49       & 9.19       & 11.25      & 11.48      &  0.906 \\
-		&{\bf run time (s)}                          & 4       & 17        & 86        & 516       & 3,600     &  2.709 \\ 
-		\midrule[.1pt]		
-		\multirow{2}{*}{\textbf{GRASP}}   & $f$      & 6.76     & 8.85       & 9.65       & 11.75      & 11.54      &  0.937 \\ 
-		&{\bf run time (s)}                          & 4       & 17        & 80        & 472       & 3,265     &  2.984 \\ 
-		\midrule[.1pt]
-		\multirow{2}{*}{\textbf{ACO}}     & $f$      & {\bf 6.77}& 8.85      & 9.65       & 11.75      & 11.54      &  0.937 \\ 
-		&{\bf run time (s)}                          & 4        & 17       & 86        & 513       & 3,596     &  2.714 \\ 
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{NMO}}     & $f$      & 6.74      & 8.80      & 9.61       & 11.76      & 11.50      &  0.934 \\
-		&{\bf run time (s)}                          & 4        & 17       & 86        & 508       & 3,480     &  2.796 \\ 
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{Shims}}   & $f$      & {\bf 6.77} & 8.92     & 9.60       & 11.78      & 11.61      &  0.939       \\ 
-		&{\bf run time (s)}                          & 2         & 8       & 31        & 166       & 1,087     &  {\bf 8.856} \\ 
+\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}\\ \pi_{TSP_2}$ }} &\multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\ % 5927
+&&               {\bf run time (s)}  & 17      & 17       & 16      & 17   & 26        &  1.0 \\ % 93
+\cmidrule(lr){2-9}	
+&\multirow{2}{*}{\it Shims}    & $f$ & 7.83    & 11.34   & 12.77   & 13.20 & 12.22     &  0.97 \\ % 5736
+&&         {\bf run time (s)}        & 4       & 5       &  6      & 7     & 10        &  2.9  \\ % 32
+
+\midrule
+
+\multirow{4}{*}{\it $S_{K}$ }&\multirow{2}{*}{\it Gurobi* } & $f$ & 8.51    & 12.25   & 13.32   & 14.61  & x        & 1.00 \\ % 4869
+&&               {\bf run time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\ % 479
+\cmidrule(lr){2-9}	
+&\multirow{2}{*}{\it Shims}    & $f$ & 8.11    & 12.18   & 13.25   & 14.57  & 14.33    &  0.99 \\ % 4811
+&&         {\bf run time (s)}        & 4       & 8       & 33      & 95     & 572      &  3.4  \\ % 140
 \bottomrule
+\multicolumn{9}{c}{*{\it MIPgap} parameter set to 1\%. The $f$ values are \textpm 1\% distant from the optimal solution}\\
+
+\end{tabular}
+\normalsize
+\end{table}
+
+97\% of 16GiB was the RAM state before we aborted the Gurobi execution on solving scenario 5 in Table \ref{tab:20}.
+
+\begin{table}[H]
+	\centering
+	\caption{Solutions with volume $surplus=1.5$.}  \label{tab:50}
+	\footnotesize
+	\begin{tabular}{ccccccccc}
+		\toprule
+		{\bf Tours} & $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		\toprule
+		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}\\ \pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\ % 8236
+		&&               {\bf run time (s)}  & 38    & 30    & 27    & 29    & 81    & 1.00  \\ % 205
+		\cmidrule(lr){2-9}	
+		&\multirow{2}{*}{\it Shims}    & $f$ & 10.89 & 16.65 & 17.80 & 18.69 & 16.45 & 0.98  \\ % 8048
+		&&         {\bf run time (s)}        & 4     & 5     & 6     & 7     & 8     & 7.1   \\ % 29
+
+		\midrule
+
+		\multirow{4}{*}{\it $S_{K}$ } & \multirow{2}{*}{\it Gurobi* } & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ % 8602
+		&&               {\bf run time (s)}  & 53    & 61    & 193   & 470   & 2241    & 1.0  \\ % 3018
+		\cmidrule(lr){2-9}	
+		&\multirow{2}{*}{\it Shims}    & $f$ & 11.33 & 16.97 & 17.56 & 20.43 & 17.94   & 0.98  \\ % 8423
+		&&         {\bf run time (s)}        & 4     & 8     & 34    & 96    & 591     & 4.1 \\ % 733
+		\bottomrule
+
 	\end{tabular}
@@
 	\centering
-	\caption{Methods and scenarios with $volume =1.5$}  \label{tab:50}
+	\caption{Solutions with volume $surplus=2.0$.}  \label{tab:100}
 	\footnotesize
-	\begin{tabular}{c|c|ccccc|c}
+	\begin{tabular}{ccccccccc}
 		\toprule
-		{$method$}                        & {\bf Results}   & {\bf 1}     &  {\bf 2}    & {\bf 3}      & {\bf 4}      & {\bf 5}    & \specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
+		{\bf Tours} & $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
 		\toprule
-		\multirow{2}{*}{\textbf{Gurobi}}  & $f$     & 4.756       & 8.84        & 11.94       & 16.21       &{\bf 16.47} & 0.852  \\	% 6828 
-		&{\bf run time (s)}                         & 18        & 68         & 326        & 1,906      & 12,960    & 1.000  \\	 
-		\midrule[.1pt]
-		\multirow{2}{*}{\textbf{TS}}      & $f$     & 8.97       & 11.44       & 12.92       & 15.20       & 15.39      &  0.936 \\		
-		&{\bf run time (s)}                         & 4         & 18         & 88         & 525        & 3,660     &  3.555  \\		
-		\midrule[.1pt]		
-		\multirow{2}{*}{\textbf{GRASP}}   & $f$     & 9.36       & 12.34       & 13.63       & 16.18       & 15.44      &  0.981 \\ 
-		&{\bf run time (s)}                         & 4         & 17         & 86         & 515        & 3,564     &  3.651  \\
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{ACO}}     & $f$     & 9.36       & 12.34       & 13.63       & 16.18       & 15.44      &  0.981 \\
-		&{\bf run time (s)}                         & 5         & 19         & 90         & 539        & 3,720     &  3.492  \\	
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{NMO}}     & $f$     & 9.36       & 12.33       & 13.60       & 16.16       & 15.43      &  0.979 \\ 
-		&{\bf run time (s)}                         & 5         & 19         & 91         & 544        & 3,780     & 3.442  \\
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{Shims}}   & $f$     & {\bf 9.39} & {\bf 12.42} & {\bf 13.72} & {\bf 16.28} & 15.49      &  {\bf 0.986} \\	
-		&{\bf run time (s)}                         & 3         & 11         & 48         & 265        & 1,720     &  {\bf 7.470}  \\
+		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}\\ \pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ &  &  &  &  &  &  \\ % 
+		&&               {\bf run time (s)}  & 38    & 30    & 27    & 29    & 81    &   \\ % 
+		\cmidrule(lr){2-9}	
+		& \multirow{2}{*}{\it Shims}    & $f$ &  &  &  &  &  &   \\ % 
+		&&         {\bf run time (s)}        & 4     & 5     & 6     & 7     & 8     &    \\ %
+
+		\midrule
+
+		\multirow{4}{*}{\it $S_{K}$ } & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &    &  \\ % 
+		&&               {\bf run time (s)}  & 126   & 103   & 266   & 574   &     &   \\ % 
+		\cmidrule(lr){2-9}	
+		&\multirow{2}{*}{\it Shims}    & $f$ &  &  &  &  &    &   \\ % 
+		&&         {\bf run time (s)}        & 4     & 8     & 34    & 96    & 591     &   \\ % 
 		\bottomrule
+
 	\end{tabular}
@@
 \centering
-	\caption{Methods and scenarios with $volume = 2.0$}  \label{tab:100}
+\caption{Solutions with volume $surplus=1.2$ with different time limitations for $S_K$ tours.}  \label{tab:20time}
 \footnotesize
-	\begin{tabular}{c|c|ccccc|c}
+\begin{tabular}{ccccccccc}
 \toprule
-		{$method$}                        & {\bf Results}    & {\bf 1}     &  {\bf 2}   & {\bf 3}    & {\bf 4}    & {\bf 5}      & \specialcell{{\bf Normalized}\\{\bf Speed-up}} \\
+  $method$  & {\bf \specialcell{Time\\limit (s)}}        & {\bf Scenarios}                   & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
 \toprule
-		\multirow{2}{*}{\textbf{Gurobi}}  & $f$              & 4.92      & 6.58      & 6.34       & 7.24       & 8.15      &  0.334 \\ % 9951
-		                                  &{\bf run time (s)}& 29       & 106      & 493      & 2,833     & 19,380      &  1.000  \\ 
-		\midrule[.1pt]
-		\multirow{2}{*}{\textbf{TS}}      & $f$              & 13.72     & 17.53     & 19.28     & 21.78      & 22.35      &  0.951 \\		
-		&{\bf run time (s)}                                  & 5        & 19       & 93       & 556       & 3,900     &  4.992  \\	
-		\midrule[.1pt]		
-		\multirow{2}{*}{\textbf{GRASP}}   & $f$        & 13.99     & 19.39     & 19.86     & 23.50      & 22.57      &  0.998 \\
-		&{\bf run time (s)}                            & 6        & 21       & 97       & 571       & 3,960     &  4.906  \\	
-		\midrule[.1pt]
-		\multirow{2}{*}{\textbf{ACO}}     & $f$        & 13.99     & 19.39     & 19.86     & 23.50      & 22.57      &  0.998 \\
-		&{\bf run time (s)}                            & 6        & 23       & 105      & 597       & 4,080     &  4.745  \\	
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{NMO}}     & $f$        & 13.99     & 19.39     & 19.85     & 23.50      & 22.56      & 0.998 \\	
-		&{\bf run time (s)}                            & 9        & 27       & 120      & 681       & 4,620     & 4.184  \\	
-		\midrule[.1pt]	
-		\multirow{2}{*}{\textbf{Shims}}   & $f$        &{\bf 14.00} & {\bf 19.44} & {\bf 19.94} & {\bf 23.55} & {\bf 22.58} & {\bf 1.000} \\	
-		&{\bf run time (s)}                            & 7        & 20       & 84       & 450       & 2,918     & {\bf 6.570}  \\
+
+\multirow{6}{*}{\it Gurobi* }& \multirow{2}{*}{1200} & $f$ &    &    &    &    &    &  \\ % 
+                                                           &        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 
+\cmidrule(lr){2-9}                              
+                                                           & \multirow{2}{*}{2400}& $f$                &    &    &    &    &    &  \\ % 
+                                                           &        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 
+\cmidrule(lr){2-9}                              		
+                                                           & \multirow{2}{*}{3600}& $f$                &    &    &    &    &    &  \\ % 
+                                                           &        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 		
+\midrule
+                              \multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$                &    &    &    &    &    &   \\ % 
+                                                           &        & {\bf run time (s)} &    &    &    &    &    &   \\ % 
+\cmidrule(lr){2-9}                              
+                                                           & \multirow{2}{*}{2400}& $f$                &    &    &    &    &    &  \\ % 
+                                                           &        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 	
+\cmidrule(lr){2-9}                              
+                                                           & \multirow{2}{*}{3600} & $f$                &    &    &    &    &    &  \\ % 
+                                                           &        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 	                                        
 \bottomrule
+
 \end{tabular}
@@
 
-{\color{blue}
 It is important to note that the best tour found in each test was seldom the tour with the lowest cost (the minimal cost tour without considering the CG deviation), but any of the first 25\% of the best tours.
@@
 If a posterior application of this method requires 8, 9, or more nodes, it is possible to use this 25\% as a threshold to discard the higher-cost tours and keep the run time within an operational acceptable limit for the client.
-}
-
-In Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, $Gurobi$\/ was the slowest method and $Shims$\/ was the fastest.
-This was expected, as $Gurobi$\/ is a MIP solver, which seeks an optimal solution within the established constraints, usually reaching its time limit on each node, while {\it Shims} is a fast heuristic.
-This way, $Gurobi$\/ even demands prohibitive times (more than 5 hours in scenario 6 of Table \ref{tab:100}), while the maximum time required by $Shims$\/ did not reach one hour.
-
-Considering the {\bf Normalized}\/ values, in Table \ref{tab:20} $Gurobi$\/ was the best ($0.997$), followed closely by $Shims$\/ ($0.939$).
-In Tables \ref{tab:50} and \ref{tab:100}, $Shims$ was the best ($0.986$\/ and $1.000$).
-As expected, the performance of $Gurobi$\/ drops sharply as the volume of shipments at the nodes increases: $0.852$\/ in Table \ref{tab:50} and $0.334$\/ in Table \ref{tab:100}.
 
@@
 \footnotesize
-\begin{tabular}{c|c|ccccc|cc}
+\begin{tabular}{ccccccccc}
 \toprule
@@
 \midrule[.1pt]	
-\multirow{2}{*}{$1.2$}   & Least cost    & 6.23  & 11.56   & 12.88   & 13.54    & 12.47  & 56.68 &       \\ 
-& Best                                   & 7.96  & 13.04   & 14.30   & 15.56    & 15.28  & 66.14 & 16.7\%  \\ 
+\multirow{2}{*}{$1.2$}   & Least cost    & 7.83    & 11.34 & 12.77 & 13.20 & 12.22  & 57.36 &       \\ 
+& Best                                   & 8.11    & 12.18 & 13.25 & 14.57 & 14.33  & 62.44 & 8.9\%  \\ 
 \midrule[.1pt]	
-\multirow{2}{*}{$1.5$}   & Least cost    & 8.44  & 16.61   & 18.18   & 19.68    & 17.64  & 80.55 &   \\ % 17.64	 20.94	
-& Best                                   & 11.47 & 18.25   & 20.61   & 22.22    & 20.94  & 93.49 & 16.1\%  \\ 
+\multirow{2}{*}{$1.5$}   & Least cost    & 10.89   & 16.65 & 17.80 & 18.69 & 16.45  & 80.48 & \\ %	
+& Best                                   & 11.33   & 16.97 & 17.56 & 20.43 & 17.94  & 84.23 & 4.7\%  \\ 
 \midrule[.1pt]	
-\multirow{2}{*}{$2.0$}   & Least cost    & 13.38 & 22.46   & 25.97   & 26.23    & 22.51  & 110.55 &   \\ %22.51	 29.09
-& Best                                   & 17.93 & 25.59   & 28.61   & 31.28    & 29.09  & 132.50 & 19.9\%  \\
+\multirow{2}{*}{$2.0$}   & Least cost    &  &    &    &     &   &  &   \\ %22.51	 29.09
+& Best                                   &  &    &    &     &   &  & \%  \\
 
@@
 
-}
+\begin{thebibliography}{00}
+
+	\bibitem{mipgap} Matthias Miltenberger. What is the MIPGap? \url{https://support.gurobi.com/hc/en-us/articles/8265539575953-What-is-the-MIPGap-}, 2023.
+
+\end{thebibliography}
 

### 2023-06-07 CAIE/results.tex
Gurobi 2400 1.2 done

@@
 
+\section{Introduction}
+\section{Related literature}
+\section{Problem context and assumptions}
+\section{The mathematical modeling}
+\section{Solution strategy}
+\section{Implementation and results}
+\label{sec6}
+
 \subsection{Results obtained}
@@
 
-We ran Algorithm \ref{alg:main} in the 5 scenarios described in Table \ref{tab:scenarios}, considering 6 methods for node-by-node solution: Gurobi (see \ref{solver}), ACO, NMO, TS, GRASP, and Shims (Algorithm \ref{alg:shims}).
+We ran Algorithm \ref{alg:main} in the 5 scenarios described in Table \ref{tab:scenarios}, considering 6 methods for node-by-node solution: Gurobi (see \ref{solver}), ACO, NMO, TS, GRASP, and {\it Shims} (Algorithm \ref{alg:shims}).
 In generating the items in each node, we consider 3 values for the parameter $surplus$.
@@
 	\item {\bf Speed-up}: ratio of the sums of the worst run times of all scenarios and the sum of the method run times in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
-
-	\item {\bf run time (s)} is the average time to solve a problem instance.
 \end{itemize}
@@
 In each $scenario$\/, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.
+{\bf run time (s)} is the average time to solve a problem instance.
 
-As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's Read-Access Memory (RAM) (16 GiB) too early, we decided to set the Gurobi {\it MIPgap} parameter to 1\% to control the minimal quality of the returned solutions and abbreviate the solution time.
+{\color{blue}
 
+As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's Read-Access Memory (RAM, 16 GiB) too early, we decided to set the Gurobi {\it MIPgap} parameter to 1\% to control the minimal quality of the returned solutions and abbreviate the solution time.
 For more details on the Gurobi {\it MIPgap}, refer to \cite{mipgap}.
+
+This decision hinders the MIP solver from reaching optimal solutions in favor of the possibility of result comparisons with the {\it Shims} heuristic.
+
+}
 
@@
 \centering
-\caption{Solutions with volume $surplus =1.2$.}  \label{tab:20}
+\caption{Solutions with volume {\color{red}$surplus =1.2$} and 3600s as time limit.}  \label{tab:20}
 \footnotesize
@@
 \toprule
-\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}\\ \pi_{TSP_2}$ }} &\multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\ % 5927
+\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} &\multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\ % 5927
 &&               {\bf run time (s)}  & 17      & 17       & 16      & 17   & 26        &  1.0 \\ % 93
@@
 
+{\color{blue}
+
 97\% of 16GiB was the RAM state before we aborted the Gurobi execution on solving scenario 5 in Table \ref{tab:20}.
+
+}
 
@@
 	\centering
-	\caption{Solutions with volume $surplus=1.5$.}  \label{tab:50}
+	\caption{Solutions with volume {\color{blue}$surplus=1.5$} and 3600s as time limit.}  \label{tab:50}
 	\footnotesize
@@
 		\toprule
-		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}\\ \pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\ % 8236
+		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\ % 8236
 		&&               {\bf run time (s)}  & 38    & 30    & 27    & 29    & 81    & 1.00  \\ % 205
@@
 	\centering
-	\caption{Solutions with volume $surplus=2.0$.}  \label{tab:100}
+	\caption{Solutions with volume {\color{violet}$surplus=2.0$} and 3600s as time limit.}  \label{tab:100}
 	\footnotesize
@@
 		\toprule
-		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}\\ \pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ &  &  &  &  &  &  \\ % 
-		&&               {\bf run time (s)}  & 38    & 30    & 27    & 29    & 81    &   \\ % 
+		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 24.39 & 26.46 & 27.09 & 24.00 & 1.00 \\ % 119.69
+		&&                                                                               {\bf run time (s)}  & 162   & 95    & 74    & 68    & 67    & 1.0  \\ % 466
 		\cmidrule(lr){2-9}	
-		& \multirow{2}{*}{\it Shims}    & $f$ &  &  &  &  &  &   \\ % 
-		&&         {\bf run time (s)}        & 4     & 5     & 6     & 7     & 8     &    \\ %
+		&                                                                   \multirow{2}{*}{\it Shims} & $f$ & 16.48 & 24.06 & 26.15 & 26.25 & 23.54 &  0.97 \\ % 116,48
+		&&                                                                                {\bf run time (s)} & 5     & 6     & 8     & 10    & 11    &  11.7  \\ % 40
 
@@
 		                                 \multirow{4}{*}{\it $S_{K}$ } & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &     &  \\ % 
-		&&               {\bf run time (s)}  & 126   & 103   & 266   & 574   &     &   \\ % 
+		&&                                                                                {\bf run time (s)} & 151   & 123   & 319   & 689   &     &   \\ % 
 		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{\it Shims}    & $f$ &  &  &  &  &    &   \\ % 
-		&&         {\bf run time (s)}        & 4     & 8     & 34    & 96    & 591     &   \\ % 
+		&                                                                \multirow{2}{*}{\it Shims}    & $f$ & 16.35 & 24.30 & 26.24 & 28.65 & 28.74   &   \\ % 
+		&&                                                                                {\bf run time (s)} & 5     & 9     & 37    & 108   & 625     &   \\ % 
+		\bottomrule
+
+	\end{tabular}
+	\normalsize
+\end{table}
+
+\subsubsection{Time limits comparisons}
+
+\begin{table}[H]
+\centering
+\caption{Solutions with volume {\color{red}$surplus=1.2$} with different time limitations for $S_K$ tours.}  \label{tab:20time}
+\footnotesize
+\begin{tabular}{ccccccccc}
+\toprule
+  $method$  & {\bf \specialcell{Time\\limit (s)}} & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}\\
+\toprule
+
+\multirow{6}{*}{\it Gurobi* }& \multirow{2}{*}{1200} & $f$ & 8.53 &    &    &    & x   &  \\ % 
+                             &        &  {\bf run time (s)}& 29   &       &       &       & x   &   \\ % 
+\cmidrule(lr){2-9}                              
+                             & \multirow{2}{*}{2400}& $f$  & 8.53 & 12.25 & 13.31 & 14.59  & 13.59   &  \\ % 
+                             &        &  {\bf run time (s)}& 29   & 40    & 129   & 313    & 1460   &   \\ % 
+\cmidrule(lr){2-9}                              		
+                             & \multirow{2}{*}{3600}& $f$  & 8.51 & 12.25 & 13.32 & 14.61 & x   & 1.00 \\ % 4869
+                             &        &  {\bf run time (s)}& 25   & 36    & 121   & 297   & x   & 1.0 \\ % 479		
+\midrule
+\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ &    &    &    &    &    &   \\ % 
+                             &        & {\bf run time (s)} &    &    &    &    &    &   \\ % 
+\cmidrule(lr){2-9}                              
+                             & \multirow{2}{*}{2400}& $f$  &    &    &    &    &    &  \\ % 
+                             &        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 	
+\cmidrule(lr){2-9}                              
+                             & \multirow{2}{*}{3600} & $f$ & 8.11 & 12.18 & 13.25 & 14.57 & 14.33 & 0.99 \\ % 4811 
+                             &        &  {\bf run time (s)}& 4    & 8     & 33    & 95    & 572   & 3.4  \\ % 140 	                                        
 \bottomrule
@@
 	\centering
-\caption{Solutions with volume $surplus=1.2$ with different time limitations for $S_K$ tours.}  \label{tab:20time}
+	\caption{Solutions with volume {\color{blue}$surplus=1.5$} with different time limitations for $S_K$ tours.}  \label{tab:50time}
 	\footnotesize
@@
 		\toprule
-  $method$  & {\bf \specialcell{Time\\limit (s)}}        & {\bf Scenarios}                   & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		$method$  & {\bf \specialcell{Time\\limit (s)}} & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}\\
 		\toprule
@@
 		\cmidrule(lr){2-9}                              		
-                                                           & \multirow{2}{*}{3600}& $f$                &    &    &    &    &    &  \\ % 
-                                                           &        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 		
+		& \multirow{2}{*}{3600}& $f$  & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ % 8602
+		&        &  {\bf run time (s)}& 53    & 61    & 193   & 470   & 2241    & 1.0  \\ % 3018		
 		\midrule
@@
 		\cmidrule(lr){2-9}                              
-                                                           & \multirow{2}{*}{3600} & $f$                &    &    &    &    &    &  \\ % 
-                                                           &        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 	                                        
+		& \multirow{2}{*}{3600} & $f$ & 11.33 & 16.97 & 17.56 & 20.43 & 17.94   & 0.98  \\ % 8423 
+		&        &  {\bf run time (s)}& 4     & 8     & 34    & 96    & 591     & 4.1 \\ % 733 	                                        
 		\bottomrule
@@
 
-It is important to note that the best tour found in each test was seldom the tour with the lowest cost (the minimal cost tour without considering the CG deviation), but any of the first 25\% of the best tours.
-This indicates that our strategy of enumerating and solving all tours was assertive, although it is possible to devise a more proactive strategy to discard most of the higher-cost tours and improve performance.
-This must be one of the next steps in this research.
+\begin{table}[H]
+	\centering
+	\caption{Solutions with volume {\color{violet}$surplus=2.0$} with different time limitations for $S_K$ tours.}  \label{tab:100time}
+	\footnotesize
+	\begin{tabular}{ccccccccc}
+		\toprule
+		$method$  & {\bf \specialcell{Time\\limit (s)}} & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}\\
+		\toprule
 
-If a posterior application of this method requires 8, 9, or more nodes, it is possible to use this 25\% as a threshold to discard the higher-cost tours and keep the run time within an operational acceptable limit for the client.
+		\multirow{6}{*}{\it Gurobi* }& \multirow{2}{*}{1200} & $f$ &    &    &    &    &    &  \\ % 
+		&        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 
+		\cmidrule(lr){2-9}                              
+		& \multirow{2}{*}{2400}& $f$  &    &    &    &    &    &  \\ % 
+		&        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 
+		\cmidrule(lr){2-9}                              		
+		& \multirow{2}{*}{3600}& $f$  & 17.75 & 25.24 & 26.46 & 29.11 &    &  \\ % 
+		&        &  {\bf run time (s)}& 126   & 103   & 266   & 574   &     &   \\ % 		
+		\midrule
+		\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ &    &    &    &    &    &   \\ % 
+		&        & {\bf run time (s)} &    &    &    &    &    &   \\ % 
+		\cmidrule(lr){2-9}                              
+		& \multirow{2}{*}{2400}& $f$  &    &    &    &    &    &  \\ % 
+		&        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 	
+		\cmidrule(lr){2-9}                              
+		& \multirow{2}{*}{3600} & $f$ & 8.11    & 12.18   & 13.25   & 14.57  & 14.33    &  0.99 \\ % 4811 
+		&        &  {\bf run time (s)}& 4       & 8       & 33      & 95     & 572      &  3.4  \\ % 140 	                                        
+		\bottomrule
 
-It is worth noting that the other heuristics always obtained {\bf Normalized}\/ values greater than or equal to $0.9$, and were always faster than $Gurobi$.
+	\end{tabular}
+	\normalsize
+\end{table}
 
-It is not possible to state how far these results are from the optimal solutions, but we can report that all occupancy rates in terms of volume were above 97\%, and in terms of weight above 79\%, which denotes that the results achieved are good, considering that the CG deviation constraints keeps these values from going higher.
+{\color{blue}
 
@@
 \centering
-\caption{Shims best and least cost tour results}  \label{tab:20least}
+\caption{{\it Shims} best and least cost tour results}  \label{tab:20least}
 \footnotesize
@@
 \midrule[.1pt]	
-\multirow{2}{*}{$1.5$}   & Least cost    & 10.89   & 16.65 & 17.80 & 18.69 & 16.45  & 80.48 & \\ %	
+\multirow{2}{*}{$1.5$}   & Least cost    & 10.89   & 16.65 & 17.80 & 18.69 & 16.45  & 80.48 & \\ 	
                          & Best          & 11.33   & 16.97 & 17.56 & 20.43 & 17.94  & 84.23 & 4.7\%  \\ 
 \midrule[.1pt]	
-\multirow{2}{*}{$2.0$}   & Least cost    &  &    &    &     &   &  &   \\ %22.51	 29.09
-& Best                                   &  &    &    &     &   &  & \%  \\
-
+\multirow{2}{*}{$2.0$}   & Least cost    &         & 24.06 & 26.15 & 26.25 & 23.54  &       &  \\ 
+                         & Best          &         & 24.30 & 26.24 & 28.65 & 28.74  &       &  \\ 		                         
 \midrule[.1pt]
- \multicolumn{8}{r}{{\bf Average improvement}} & 17.6\% \\
+ \multicolumn{8}{r}{{\bf Average improvement}}                                              & \% \\
 \bottomrule
@@
 
-As the selection of the best tour, considering what was collected and delivered in each node, and considering the cost increase due to CG deviation, was 17.6\% better than the option for a simple TSP solution (\textit{Least cost} results), this confirms that our solution strategy was adequate.
+As the selection of the best tour, considering what was collected and delivered in each node, and considering the cost increase due to CG deviation, was better than the option for a simple TSP solution (\textit{Least cost} results), this confirms that our solution strategy was adequate.
+
+It is important to report that the best tour found in each test was seldom the tour with the lowest cost (the minimal cost tour without considering the CG deviation), but any of the first 25\% of the best tours.
+This indicates that our strategy of enumerating and solving all tours was assertive, although it is possible to devise a more proactive strategy to discard most of the higher-cost tours and improve performance.
+This must be one of the next steps in this research.
+
+If a posterior application of this method requires 8, 9, or more nodes, it is possible to use this 25\% as a threshold to discard the higher-cost tours and keep the run time within an operational acceptable limit for the client.
+
+\subsubsection{Final considerations on the results}
+
+Considering that the computer used in the experiments has a standard RAM consumption when idle of 2.5 GiB and all Gurobi executions consumed from 12 to 15 GiB of RAM, the Gurobi executions consumed 10 to 12.5 GiB in total.
+
+All {\it Shims} executions consumed at most 1.5 GiB, or 3 to 4 GiB in total.
+
+These details are important for a future application to adequately dimension the computer to be used.
+
+In the beginning of this work, we devised, as one of the objectives, a method that may be run on a handheld computer.
+If this is the case, the use of a MIP solver may not be appropriate, as most handheld computers come with 4 GB of RAM.
+
+}
 
@@
 
-	\bibitem{mipgap} Matthias Miltenberger. What is the MIPGap? \url{https://support.gurobi.com/hc/en-us/articles/8265539575953-What-is-the-MIPGap-}, 2023.
+	\bibitem{mipgap} Matthias Miltenberger. {\it What is the MIPGap?} \url{https://support.gurobi.com/hc/en-us/articles/8265539575953-What-is-the-MIPGap-}, 2023.
 

### 2023-06-07 CAIE/results.tex
Experiments continues

@@
 In each $scenario$\/, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.
+
+{\color{blue}
 {\bf run time (s)} is the average time to solve a problem instance.
 
-{\color{blue}
-
 As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's Read-Access Memory (RAM, 16 GiB) too early, we decided to set the Gurobi {\it MIPgap} parameter to 1\% to control the minimal quality of the returned solutions and abbreviate the solution time.
+From now on, we will refer to this Gurobi-specific model as {\it Gurobi* }.
 For more details on the Gurobi {\it MIPgap}, refer to \cite{mipgap}.
@@
 
-97\% of 16GiB was the RAM state before we aborted the Gurobi execution on solving scenario 5 in Table \ref{tab:20}.
+97\% of 16GiB was the RAM state before we aborted the {\it Gurobi*} execution on solving scenario 5 in Table \ref{tab:20}.
 
@@
 		\cmidrule(lr){2-9}	
-		&                                                                   \multirow{2}{*}{\it Shims} & $f$ & 16.48 & 24.06 & 26.15 & 26.25 & 23.54 &  0.97 \\ % 116,48
+		&                                                                   \multirow{2}{*}{\it Shims} & $f$ & 16.73 & 24.06 & 26.15 & 26.25 & 23.54 &  0.97 \\ % 116,48
 		&&                                                                                {\bf run time (s)} & 5     & 6     & 8     & 10    & 11    &  11.7  \\ % 40
@@
 
-		                                 \multirow{4}{*}{\it $S_{K}$ } & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &     &  \\ % 
-		&&                                                                                {\bf run time (s)} & 151   & 123   & 319   & 689   &     &   \\ % 
+		                                 \multirow{4}{*}{\it $S_{K}$ } & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &  x   & 1.00 \\ % 9856
+		&&                                                                                {\bf run time (s)} & 151   & 123   & 319   & 689   &  x   & 1.0  \\ % 1282
 		\cmidrule(lr){2-9}	
-		&                                                                \multirow{2}{*}{\it Shims}    & $f$ & 16.35 & 24.30 & 26.24 & 28.65 & 28.74   &   \\ % 
-		&&                                                                                {\bf run time (s)} & 5     & 9     & 37    & 108   & 625     &   \\ % 
-		\bottomrule
-
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\subsubsection{Time limits comparisons}
-
-\begin{table}[H]
-\centering
-\caption{Solutions with volume {\color{red}$surplus=1.2$} with different time limitations for $S_K$ tours.}  \label{tab:20time}
-\footnotesize
-\begin{tabular}{ccccccccc}
-\toprule
-  $method$  & {\bf \specialcell{Time\\limit (s)}} & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}\\
-\toprule
-
-\multirow{6}{*}{\it Gurobi* }& \multirow{2}{*}{1200} & $f$ & 8.53 &    &    &    & x   &  \\ % 
-                             &        &  {\bf run time (s)}& 29   &       &       &       & x   &   \\ % 
-\cmidrule(lr){2-9}                              
-                             & \multirow{2}{*}{2400}& $f$  & 8.53 & 12.25 & 13.31 & 14.59  & 13.59   &  \\ % 
-                             &        &  {\bf run time (s)}& 29   & 40    & 129   & 313    & 1460   &   \\ % 
-\cmidrule(lr){2-9}                              		
-                             & \multirow{2}{*}{3600}& $f$  & 8.51 & 12.25 & 13.32 & 14.61 & x   & 1.00 \\ % 4869
-                             &        &  {\bf run time (s)}& 25   & 36    & 121   & 297   & x   & 1.0 \\ % 479		
-\midrule
-\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ &    &    &    &    &    &   \\ % 
-                             &        & {\bf run time (s)} &    &    &    &    &    &   \\ % 
-\cmidrule(lr){2-9}                              
-                             & \multirow{2}{*}{2400}& $f$  &    &    &    &    &    &  \\ % 
-                             &        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 	
-\cmidrule(lr){2-9}                              
-                             & \multirow{2}{*}{3600} & $f$ & 8.11 & 12.18 & 13.25 & 14.57 & 14.33 & 0.99 \\ % 4811 
-                             &        &  {\bf run time (s)}& 4    & 8     & 33    & 95    & 572   & 3.4  \\ % 140 	                                        
-\bottomrule
-
-\end{tabular}
-\normalsize
-\end{table}
-
-\begin{table}[H]
-	\centering
-	\caption{Solutions with volume {\color{blue}$surplus=1.5$} with different time limitations for $S_K$ tours.}  \label{tab:50time}
-	\footnotesize
-	\begin{tabular}{ccccccccc}
-		\toprule
-		$method$  & {\bf \specialcell{Time\\limit (s)}} & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}\\
-		\toprule
-
-		\multirow{6}{*}{\it Gurobi* }& \multirow{2}{*}{1200} & $f$ &    &    &    &    &    &  \\ % 
-		&        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$  &    &    &    &    &    &  \\ % 
-		&        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 
-		\cmidrule(lr){2-9}                              		
-		& \multirow{2}{*}{3600}& $f$  & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ % 8602
-		&        &  {\bf run time (s)}& 53    & 61    & 193   & 470   & 2241    & 1.0  \\ % 3018		
-		\midrule
-		\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ &    &    &    &    &    &   \\ % 
-		&        & {\bf run time (s)} &    &    &    &    &    &   \\ % 
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$  &    &    &    &    &    &  \\ % 
-		&        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 	
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{3600} & $f$ & 11.33 & 16.97 & 17.56 & 20.43 & 17.94   & 0.98  \\ % 8423 
-		&        &  {\bf run time (s)}& 4     & 8     & 34    & 96    & 591     & 4.1 \\ % 733 	                                        
-		\bottomrule
-
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\begin{table}[H]
-	\centering
-	\caption{Solutions with volume {\color{violet}$surplus=2.0$} with different time limitations for $S_K$ tours.}  \label{tab:100time}
-	\footnotesize
-	\begin{tabular}{ccccccccc}
-		\toprule
-		$method$  & {\bf \specialcell{Time\\limit (s)}} & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}\\
-		\toprule
-
-		\multirow{6}{*}{\it Gurobi* }& \multirow{2}{*}{1200} & $f$ &    &    &    &    &    &  \\ % 
-		&        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$  &    &    &    &    &    &  \\ % 
-		&        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 
-		\cmidrule(lr){2-9}                              		
-		& \multirow{2}{*}{3600}& $f$  & 17.75 & 25.24 & 26.46 & 29.11 &    &  \\ % 
-		&        &  {\bf run time (s)}& 126   & 103   & 266   & 574   &     &   \\ % 		
-		\midrule
-		\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ &    &    &    &    &    &   \\ % 
-		&        & {\bf run time (s)} &    &    &    &    &    &   \\ % 
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$  &    &    &    &    &    &  \\ % 
-		&        &  {\bf run time (s)}&    &    &    &    &    &   \\ % 	
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{3600} & $f$ & 8.11    & 12.18   & 13.25   & 14.57  & 14.33    &  0.99 \\ % 4811 
-		&        &  {\bf run time (s)}& 4       & 8       & 33      & 95     & 572      &  3.4  \\ % 140 	                                        
+		&                                                                \multirow{2}{*}{\it Shims}    & $f$ & 16.88 & 24.30 & 26.24 & 28.65 & 28.74 & 0.97  \\ % 9607
+		&&                                                                                {\bf run time (s)} & 5     & 9     & 37    & 108   & 625   & 8.1  \\ % 159
 		\bottomrule
@@
 {\color{blue}
+{\it Gurobi*} could not solve scenario 5 in Table \ref{tab:100} because $3600/K! = 0.71s$, not enough time for {\it Gurobi*} to find a feasible node solution.
+}
 
-\subsubsection{Proposed strategy assessment}
+\subsubsection{Time limits comparisons}
+
+\begin{table}[H]
+	\centering
+	\caption{Solutions with different time limitations for $S_K$ tours.}  \label{tab:100time}
+	\footnotesize
+	\begin{tabular}{ccc|cc|cc|cc}
+		\toprule
+		&             \multicolumn{2}{r}{\bf Volume $surplus$} &\multicolumn{2}{c}{$1.2$}&\multicolumn{2}{c}{$1.5$}&\multicolumn{2}{c}{$2.0$} \\
+		\midrule
+$method$& {\bf \specialcell{Time\\limit (s)}} & {\bf Scenarios}    & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} \\
+		\toprule
+
+		\multirow{6}{*}{\it Gurobi* }& \multirow{2}{*}{1200} & $f$ & 14.67   & x       &  20.64  &  x      & 27.97   & x \\ % 
+		&        &  {\bf run time (s)}                             & 217     & x       &  304    &  x      & 579     & x \\ % 
+		\cmidrule(lr){2-9}                              
+		& \multirow{2}{*}{2400}& $f$                               & 14.59   & 13.59   &  20.60  &  x      & 29.10   & x \\ % 
+		&        &  {\bf run time (s)}                             & 313     & 1460    &  327    &  x      & 543     & x \\ % 
+		\cmidrule(lr){2-9}                              		
+		& \multirow{2}{*}{3600}& $f$                               & 14.61   & x       &  20.59  & 18.36   & 29.11   & x \\ % 
+		&        &  {\bf run time (s)}                             & 297     & x       &  470    & 2241    & 689     & x \\ % 		
+		\midrule
+		\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ & 15.14   &         &         &         &         &   \\ % 
+		&        & {\bf run time (s)}                              & 73      &         &         &         &         &   \\ % 
+		\cmidrule(lr){2-9}                              
+		& \multirow{2}{*}{2400}& $f$                               &         &         &         &         &         &   \\ % 
+		&        &  {\bf run time (s)}                             &         &         &         &         &         &   \\ % 	
+		\cmidrule(lr){2-9}                              
+		& \multirow{2}{*}{3600} & $f$                              & 14.57   & 14.33   &  20.43  & 17.94   & 28.65   & 28.74 \\ %  
+		&        &  {\bf run time (s)}                             &  95     &  572    &  96     & 591     & 108     &  625  \\ %  	                                        
+		\bottomrule
+
+	\end{tabular}
+	\normalsize
+\end{table}
+
+\begin{figure}[H]
+\centering
+\begin{tikzpicture}
+	\begin{axis}[
+		area style,
+		]
+		\addplot+[ybar interval,mark=no] plot coordinates { 
+			(0,.3)
+			(1,.5)
+			(2,.6)
+			(3,.5)
+			(4,.25)
+			(5,.1)
+			(6,.5)
+			(7,.25)
+			(8,.1)
+		};
+	\end{axis}
+\end{tikzpicture}
+\captionof{figure}{ESTE GRÁFICO É APENAS UM MODELO}
+\end{figure}
+
+{\color{blue}
+
+\subsubsection{Proposed K! strategy assessment}
 
@@
 {$surplus$}              & {\bf Tour} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4}  & {\bf 5}   & {\bf Sum} & {\bf Improvement}\\
+\toprule
 
+\multirow{2}{*}{$1.2$}   & Least cost    & 7.83    & 11.34 & 12.77 & 13.20 & 12.22  & 57.36 &       \\ 
+                         & Best in $K!$  & 8.11    & 12.18 & 13.25 & 14.57 & 14.33  & 62.44 & 8.9\%  \\ 
 \midrule[.1pt]	
-\multirow{2}{*}{$1.2$}   & Least cost    & 7.83    & 11.34 & 12.77 & 13.20 & 12.22  & 57.36 &       \\ 
-                         & Best          & 8.11    & 12.18 & 13.25 & 14.57 & 14.33  & 62.44 & 8.9\%  \\ 
+\multirow{2}{*}{$1.5$}   & Least cost    & 10.89   & 16.65 & {\color{red}17.80} & 18.69 & 16.45  & 80.48 & \\ 	
+                         & Best in $K!$  & 11.33   & 16.97 & 17.56 & 20.43 & 17.94  & 84.23 & 4.8\%  \\ 
 \midrule[.1pt]	
-\multirow{2}{*}{$1.5$}   & Least cost    & 10.89   & 16.65 & 17.80 & 18.69 & 16.45  & 80.48 & \\ 	
-                         & Best          & 11.33   & 16.97 & 17.56 & 20.43 & 17.94  & 84.23 & 4.7\%  \\ 
+\multirow{2}{*}{$2.0$}   & Least cost    & 16.73   & 24.06 & 26.15 & 26.25 & 23.54  & 116.73&  \\ 
+                         & Best in $K!$  & 16.88   & 24.30 & 26.24 & 28.65 & 28.74  & 124.81& 6.7\%   \\ 		                         
 \midrule[.1pt]
-\multirow{2}{*}{$2.0$}   & Least cost    &         & 24.06 & 26.15 & 26.25 & 23.54  &       &  \\ 
-                         & Best          &         & 24.30 & 26.24 & 28.65 & 28.74  &       &  \\ 		                         
-\midrule[.1pt]
- \multicolumn{8}{r}{{\bf Average improvement}}                                              & \% \\
+\multicolumn{8}{r}{{\bf K! strategy advantage}}                                              & {\bf 6.8}\% \\
+
 \bottomrule
@@
 
-Considering that the computer used in the experiments has a standard RAM consumption when idle of 2.5 GiB and all Gurobi executions consumed from 12 to 15 GiB of RAM, the Gurobi executions consumed 10 to 12.5 GiB in total.
+Most of the {\it Gurobi*} executions consumed over 12 GiB in total, and considering that the computer used in the experiments has a standard RAM consumption when idle of 3.5 GiB (with the Python Interface Development Environment (IDE), a PDF viewer, a system monitor, and a {\it LaTeX} IDE simultaneously open), the actual RAM consumption of {\it Gurobi*} was over 8.5 GiB.
 
-All {\it Shims} executions consumed at most 1.5 GiB, or 3 to 4 GiB in total.
-
+All {\it Shims} executions consumed at most 1.5 GiB, requiring 5 GiB of total memory.
 These details are important for a future application to adequately dimension the computer to be used.
-
-In the beginning of this work, we devised, as one of the objectives, a method that may be run on a handheld computer.
-If this is the case, the use of a MIP solver may not be appropriate, as most handheld computers come with 4 GB of RAM.
 

### 2023-06-08 CAIE/cie.tex
cie file updated and experimentations almost ended

--- 2023-06-05 CAIE/paper.tex
@@
 	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+
+\usepackage{textcomp}
 
@@
 As can be seen, so far Lurkin and Schyns \cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem.
-Although it is innovative, strong simplifications were imposed by the authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a predefined tour plan restricted to two legs.
+Although it is innovative, strong simplifications were imposed by these authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a predefined tour plan restricted to only two legs.
 It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem.
@@
 However, as reported by Fok and Chun, {\it load planning (...) is usually done in roughly 2 hours before departure, when all the details of the cargo are present.} \cite{fok2004optimizing}.
-We may infer that this is because there is no commercial software to solve this problem.
+We can infer that this is because adequate commercial software does not exist.
 In addition, there is still the necessary time to plan the route, in order to minimize fuel consumption and ensure the transport of priority items.
@@
 
-Throughout this text, we call {\it Packed Contents}\/ (PC) a set of items of same destination stacked on a pallet and covered with a restraining net (see Figure \ref{fig:larger2}).
-It is considered unique, having the same attributes of its components, whose values are the sum of individual scores, weights and volumes.
-To ensure accuracy in pickup and delivery operations, a PC must remain on board until its destination.
+Throughout this text, we call {\it packed content}\/ a set of items of same destination stacked on a pallet and covered with a restraining net (see Figure \ref{fig:larger2}).
+It is considered a single item, having the same attributes of its components, whose values are the sum of individual scores, weights and volumes.
+To ensure accuracy in pickup and delivery operations, a packed content must remain on board until its destination.
 
@@
 
-We consider real scenarios with an aircraft with payload of $75,000 kg$, with enough fuel for a range of $4,400km$.
+We consider real scenarios with an single aircraft with payload of $75,000 kg$, with enough fuel for a range of $4,400km$.
 The aircraft layout is presented in Figure \ref{fig:larger}, where the pallets positions are identified by $p_i$, $1\leq i \leq 18$.
@@
 	\includegraphics[scale=0.4]{Images/consolidated.jpg}
-	\caption{A PC on a 463L pallet inside a {\it Boeing C-17}}
+	\caption{A packed content on a 463L pallet inside a {\it Boeing C-17}}
 	\small\textsuperscript{Source: From Wikimedia Commons, the free media repository}
@@
 $p_i$\/ are pallets, $1\leq i \leq 18$, with weight $W_i$\/ and $V_i$\/ volume limits, respectively.
-$D_i^{long}$\/ and $D_i^{lat}$\/ are, respectively, the longitudinal and latitudinal distances of each pallet centroids to the CG of aircraft along both axes.
+$D_i^{long}$\/ and $D_i^{lat}$\/ are, respectively, the longitudinal and lateral distances of each pallet centroids to the CG of aircraft along both axes.
 These distances will be used in the calculation of the torque referring to the items allocated on each pallet.
-In this aircraft, as the ramp have an inclination of $25°$, we made the necessary corrections in $D_i^{long}$, $W_i$\/ and $V_i$\/ of the corresponding pallets.
+In this aircraft, as the ramp have an inclination of $25^{\circ}$, we made the necessary corrections in $D_i^{long}$, $W_i$\/ and $V_i$\/ of the corresponding pallets.
 
@@
 The {\it fuel consumption rate}\/ $c_g$\/ is the percentage limit of cost increase  due to the CG deviation on the longitudinal axis of the aircraft.
-$c_g$\/ depends on the characteristics of the aircraft, and is an arbitrated value for this work that should be better evaluated in future research.
+$c_g$\/ depends on the characteristics of the aircraft, and is an arbitrated value that should be better evaluated in future research.
 It is important to consider that the $c_g$\/ tends to zero as the aircraft attitude tends to level.
@@
 Let $\pi_k$\/ be $k^{th}$\/ node of the tour $\pi$, $1 \leq k \leq K$.
-In this way, the tour $\pi$\/ is described as $\{0,\pi_1,\ldots,\pi_K, 0\}$, {\color{blue}$K > 1$}.
+In this way, the tour $\pi$\/ is described as $\{0,\pi_1,\ldots,\pi_K, 0\}$.
 For ease of notation, we can define $\pi_0 = \pi_{K+1} = 0$.
-
-We consider K > 1 because the case K = 1 has a single and trivial solution for the tour.
 
@@
 
-$\tilde{s}_\pi$\/ \ref{eq:scores} is the sum of the scores of the items loaded on the aircraft throughout the tour $\pi$.
+We have the following calculation equations:
+
+\begin{itemize}
+	\item $\tilde{s}_\pi$\/ \ref{eq:scores} is the sum of the scores of the items loaded on the aircraft throughout the tour $\pi$.
+	\item $\tau_{\pi_k}$\/ \ref{eq:tau} is the longitudinal torque applied by the loaded pallets at the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.
+	\item $\tilde{c}_\pi$\/ \ref{eq:costs} is the cost of to the total fuel consumed in the tour $\pi$\/ due to the distances traveled and the CG longitudinal deviations \footnote{In our experiments, we found that the magnitude of the lateral torque was always very small, and therefore we decided to ignore it in the fuel consumption.} in each pallet.
+	\item The sets of nodes not yet visited along the tour $\pi$\/ are defined in \ref{eq:pdp11} and \ref{eq:departing}.
+	\item At the beginning of the tour $\pi$, there are no packed contents \ref{eq:pdp13}. 
+	\item $\epsilon_{\pi_k}^t$\/ \ref{eq:LatItem} and $\epsilon_{\pi_k}^a$\/ \ref{eq:LatPacked} are the lateral torque applied by the loaded pallets at the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.
+For ease of notation, $\epsilon_{\pi_k}^t$\/ corresponds to the items, and $\epsilon_{\pi_k}^a$\/ to the packed contents.
+\end{itemize}
 
@@
 
-$\tau_{\pi_k}$\/ \ref{eq:tau} is the longitudinal torque applied by the loaded pallets, in the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.
-
 \begin{equation} \label{eq:tau}
@@
 
-In this way, we can calculate the cost $\tilde{c}_\pi$\/ \ref{eq:costs}, referring to the total fuel consumed in the tour $\pi$\/ due to the distances traveled and the CG longitudinal deviations in each pallet.
-
 \begin{equation} \label{eq:costs}
@@
 
-The sets of nodes not yet visited are defined in \ref{eq:pdp11} and \ref{eq:departing}.
-
 \begin{equation} \label{eq:pdp11}
@@
 
-At the beginning of the tour $\pi$, there are no packed contents \ref{eq:pdp13}.
-
 \begin{equation} \label{eq:pdp13}
@@
 
-$\epsilon_{\pi_k}^t$\/ \ref{eq:LatItem} and $\epsilon_{\pi_k}^a$\/ \ref{eq:LatPacked} are the lateral torque applied by the loaded pallets, in the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.
-For ease of notation, $\epsilon_{\pi_k}^t$\/ corresponds to the items, and $\epsilon_{\pi_k}^a$\/ to the packed contents.
-
 \begin{equation} \label{eq:LatItem}
@@
 
-Finally, we can consider the constraints on each node $\pi_k$:
+Finally, we can consider the constraints at each node $\pi_k$:
 
@@
 	\item At most, each item is associated with a single pallet \ref{eq:app4};
-	\item Packet contents that have not yet reached their destination must remain on board \ref{eq:app5}.
+	\item Packed contents that have not yet reached their destination must remain on board \ref{eq:app5}.
 	\item Items allocated on the same pallet must have the same destinations.
@@
 Throughout our research, we have thoughtfully described the ACLP+RPDP model in standard MIP format and found that no MIP solver can handle its practical cases in a feasible time.
-Thus, as ACLP+RPDP is highly complex, involving four intractable subproblems (APP, WBP, PDP and TSP), our strategy will be to focus only on real cases and to develop quick node-by-node solutions, not necessarily optimal, that allow to build a complete tour.
+Thus, as ACLP+RPDP is highly complex, involving four intractable subproblems (APP, WBP, PDP and TSP), our strategy will be to focus only on {\it real cases}\/ and to develop quick node-by-node solutions, not necessarily optimal, that allow to build a complete tour.
 
@@
 On the other hand, we know that transport aircraft generally have a few dozen pallets, the flight plan has less than $6$\/ nodes, and each node has hundreds of items to be shipped.
+We also know that missions with fewer nodes are more frequent than longer missions.
 Under these circumstances, we can adopt two more strategies:
@@
 
-\item Considering the number of destinations less than the number of pallets ($K < m$), we can preset the destinations of the pallets in each node.
+\item We will consider that the number of destinations is greater than 1, to avoid the case of a single and trivial tour, but less than the number of pallets, that is, $1 < K < m$.
+So we can preset the destinations of the pallets at each node.
 In this way, we will reserve a number of pallets proportional to the volume demanded by each destination at the shipping node.
@@
 
-\item If the number of nodes is small and we have quick node-by-node solutions, we have the possibility to check the solution corresponding to the tour with the shortest total distance, and even test all possible tours, selecting the one that provides the best value for the objective function.
+\item If we have quick node-by-node solutions, as the number of nodes is small we have the possibility to check the solution corresponding to the tour with the shortest total distance, and even test all possible tours, selecting the one that provides the best value for the objective function.
 
@@
 \begin{algorithm}[H]
-	\caption{ACLP+RPDP solving}  \label{alg:main}
+	\caption{Solving ACLP+RPDP}  \label{alg:main}
 	\begin{algorithmic}[1]
@@
 
-		\State Let aircraft and $M$\/ be according to Table \ref{tab:larger}  \label{main:M} \Comment{Aircraft parameters}
-
-		\State Let $K$, $L$\/ and $C$ be according to $scenario$\/ and Table \ref{tab:costs} \label{main:KLC} \Comment{Nodes, distances and costs}
-
-		\State Let $\pi_{TSP}$\/ be the tour with the shortest distance traveled
+		\State Let aircraft and $M$  \label{main:M} \Comment{Aircraft parameters from Table \ref{tab:larger}}
+
+		\State Let $K$, $L$\/ and $C$ be according to $scenario$ \label{main:KLC} \Comment{Nodes and costs from Tables \ref{tab:costs} and \ref{tab:scenarios}}
+
+		\State Let $\pi_{TSP1}$\/ and $\pi_{TSP2}$  \Comment{The shortest tours from Tables \ref{tab:costs} and \ref{tab:scenarios}}
 
@@
 
-			\State {\color{blue} $f_{\pi_{dir}} \gets SolveTour(\pi_{dir}, L, M, C, N, method, time)$ } \label{main:fdirect}
-
-		    \State {\color{blue} $f_{\pi_{rev}} \gets SolveTour(\pi_{rev}, L, M, C, N, method, time)$ } \label{main:freverse}
-
-			\State {\color{blue} $answer1[scenario,surplus,method] \gets max( f_{\pi_{dir}}, f_{\pi_{rev}}  )$ } \label{main:a1}
+			\State $f_1 \gets SolveTour(\pi_{TSP1}, L, M, C, N, method, time)$  \label{main:f1}
+
+		    \State $f_2 \gets SolveTour(\pi_{TSP2}, L, M, C, N, method, time)$  \label{main:f2}
+
+			\State $answer1[scenario,surplus,method] \gets max( f_1, f_2  )$  \label{main:a1} \Comment{Best result between the shortest tours}
 
@@
 
-				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, time/(K+1)!)$ \Comment{$f_{\pi}$\/ is a solution to the tour $\pi$} \label{main:f2}
+				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, time/K!)$ \Comment{$f_{\pi}$\/ is a solution to the tour $\pi$} \label{main:fpi}
 
@@
 
-		\Return $answer$
+		\Return $answer1, answer2$
 
@@
 
-{\color{blue} In line \ref{main:fdirect}, $SolveTour$ solves the lowest-cost tour, and in line 8, its reverse. }
-
-The aircraft parameters are in Table \ref{tab:larger} (line \ref{main:M}).
-In this algorithm, we use five values for $scenario$, according to Table \ref{tab:scenarios}, which defines $K$\/ and the set $L$\/ of nodes (line \ref{main:KLC}).
+The aircraft parameters and the set $M$\/ of pallets are in Table \ref{tab:larger}.
+In this algorithm, we use five values for $scenario$, according to Tables \ref{tab:costs} and \ref{tab:scenarios}, which define the number $K$ of destinations, the set $L$\/ of nodes, the costs $C$\/, and the shortest tours $\pi_{TSP1}$\/ and $\pi_{TSP2}$.
 
@@
 	\caption{Testing scenarios}  \label{tab:scenarios}
-	\begin{tabular}{c c c }
+	\begin{tabular}{c c c c c}
 		\toprule
-		{\bf Scenario} & {$K$} & {$L$} \\		
+		{\bf Scenario} & $K$ & $L$ & $\pi_{TSP1}$ & $\pi_{TSP2}$ \\		
 		\midrule
-		1 & 2    & \{$1,2$\}           \\
-		2 & 3    & \{$1,2,3$\}         \\
-		3 & 4    & \{$1,2,3,4$\}       \\
-		4 & 5    & \{$1,2,3,4,5$\}     \\
-		5 & 6    & \{$1,2,3,4,5,6$\}   \\
+		1 & 2    & \{$0,1,2$\}           & 0 1 2 0  & 0 2 1 0 \\
+		2 & 3    & \{$0,1,2,3$\}         & 0 1 2 3 0 & 0 3 2 1 0\\
+		3 & 4    & \{$0,1,2,3,4$\}       & 0 4 1 2 3 0 & 0 3 2 1 4 0 \\
+		4 & 5    & \{$0,1,2,3,4,5$\}     & 0 4 1 2 5 3 0 & 0 3 5 2 1 4 0 \\
+		5 & 6    & \{$0,1,2,3,4,5,6$\}   & 0 4 1 2 6 5 3 0 & 0 3 5 6 2 1 4 0 \\
 		\bottomrule
@@
 
-$time$\/ is a run time limit, which will be distributed among the node-by-node solutions (lines \ref{main:f1} and \ref{main:f2}).
-Each tour consists of $K+1$\/ trips, and the total number of tours is $K!$.
-
-$method$\/ corresponds to a MIP solver or a heuristic that we will present in subsection \ref{methods}, where $SolveTour$\/ is the node-by-node solution.
-
-The results corresponding to the shortest tour are stored in $answer1$\/ (line \ref{main:a1}), and those obtained by testing all tours are stored in $answer2$\/ (line \ref{main:a2}).
+$time$\/ is a run time limit, which will be distributed among the tours (lines \ref{main:f1}, \ref{main:f2} and \ref{main:fpi}).
+$method$\/ corresponds to a MIP solver or a heuristic to the node-by-node solution $SolveTour$, which will be presented in subsection \ref{methods}.
+
+The best results corresponding to the shortest tours are stored in $answer1$\/ (line \ref{main:a1}), and those obtained by testing all $K!$\/ tours are stored in $answer2$\/ (line \ref{main:a2}).
 
 Next, we will present two subsections: in the first we explain how $SolveTour$\/ is executed, presetting the destinations of the pallets.
-In the second we will present the heuristic developed for node-by-node solutions.
+In the second we will present the heuristics developed for node-by-node solutions.
 
@@
 As we commented in the previous subsection, we will adopt the strategy of presetting the destinations of each pallet throughout the tour.
-This is feasible in practical cases, where $K<m$.
-For this, each pallet $i$\/ also has a field $T^k_i$, $0\leq k \leq K$, which stores its next destination after being loaded in node $k$.
+This is feasible in practical cases, where $1<K<m$.
+For this, each pallet $i$\/ also has a field $T^k_i$, $0\leq k \leq K$, which stores its next destination after being loaded at node $k$.
 For this reason, $T^k_i \in L_k$, $1 \leq i \leq m$, $0\leq k \leq K$.
 
-$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $time$\/ is the run time limit that will be distributed among the $K+1$\/ legs of the tour.
+$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $time$\/ is the runtime limit that will be distributed among the $K+1$\/ legs of the tour.
 
 \begin{algorithm}[H]
-	\caption{Procedure to solve tour $\pi$ with $method$ (MIP solver or heuristic)}  \label{alg:tour}
+	\caption{Solving the tour $\pi$ with $method$}  \label{alg:tour}
 	\begin{algorithmic}[1]
@@
 		\State $L_{\pi_k} \gets L - \{\pi_0,\pi_1,\ldots,\pi_k\}$  \label{tour:lk1}	\Comment{$\pi_k$ is the current node}	
-		\For{$i \gets 1$ to $m$}
-		\State $T_i^{\pi_k} \gets -1$ \label{tour:-1} \Comment{Unset the pallet destinations}
-		\EndFor
+		\State $T_i^{\pi_k} \gets -1$, $1 \leq i \leq m$ \label{tour:-1} \Comment{Unset the pallet destinations}
 
@@
 
-		\State $G_2 \gets SolveNode(method, \pi_k, G_1, time/(K+1))$ \label{tour:node}
+		\State $G_2 \gets SolveNode(method, \pi_k, G_1, time/(K+1))$ \Comment{Each tour has $K+1$\/ legs} \label{tour:node}
 
@@
 
-When the aircraft is at the base, the initial graph $G_1$\/ is empty because it has no packed content \ref{tour:g11}.
+When the aircraft is at the base, the initial graph $G_1$\/ is empty and there is no packed contents \ref{tour:g11}.
 Otherwise, $UpdatePacked$\/ (line \ref{tour:dest}) returns the set of packed contents that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation.
@@
 \begin{algorithm}[H]
-	\caption{Procedure to update the PC that remain boarded on node $\pi_k$}  \label{alg:cons}
+	\caption{Updating the packed contents that remain boarded at the node $\pi_k$}  \label{alg:cons}
 	\begin{algorithmic}[1]
@@
 
-		\State $E_{Q_{\pi_k}} \gets MinCGDeviation(E_{Q_{\pi_k}})$ \label{cons:minCG}
+		\State $E_{Q_{\pi_k}} \gets MinCGDeviation(E_{Q_{\pi_k}})$ \Comment{Using a MIP solver} \label{cons:minCG}
 
@@
 
-$MinCGDeviation$\/ (line \ref{cons:minCG}) is run through a MIP solver with the equations \ref{eq:torque}, \ref{eq:embarked} and \ref {eq:one}, to relocate the packed contents on the pallets, minimizing torque and ensuring that they all remain on board, one packed content on each pallet.
-As there are few variables, the MIP solver returns an allocation $E_{Q_{\pi_k}}$\/ in less than 30 milliseconds.
+$MinCGDeviation$\/ (line \ref{cons:minCG}) relocates the packed contents on the pallets, minimizing torque and ensuring that they all remain on board, one packed content on each pallet.
+It is run through a MIP solver with objective function \ref{eq:torque} and constraints \ref{eq:embarked} and \ref{eq:one}.
+As there are few variables, $E_{Q_{\pi_k}}$\/ is obtained in less than $30$\/ milliseconds.
 Finally, the destination of each pallet with a packed content is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).
 
-\vspace{3mm}
-$MinCGDeviation$ objective function:
-
 \begin{equation} \label{eq:torque}
-	\mbox{minimize}\ \Big | \sum_{i=1}^{m} \sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \times w_q \times D_i^{long}  \Big |
-\end{equation}
-
-\vspace{3mm}
-s.t.:
+	\min f =  \Big | \sum_{i=1}^{m} \sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \times w_q \times D_i^{long}  \Big |
+\end{equation}
 
@@
 \begin{algorithm}[H]
-	\caption{Procedure to set pallets destination based on the items to be embarked on node $\pi_k$}  \label{alg:dest}
+	\caption{Setting pallets destination based on the items to be embarked at the node $\pi_k$}  \label{alg:dest}
 	\begin{algorithmic}[1]
@@
 
-			\For{$x \in L_{\pi_k}$}		
-				\State $vol_x \gets  0$  \label{dest:vector1}
-			\EndFor
+			\State $vol_x \gets  0$, $x \in L_{\pi_k}$  \label{dest:vector1}
 
@@
 
-						\If{$np < needed$ {\bf and} $T_i^{\pi_k} = -1$}
+						\If{($np < needed$) {\bf and} ($T_i^{\pi_k} = -1$)}
 							\State $T_i^{\pi_k} \gets x$
@@
 
-$ScoreAndDeviation$\/ is described in Algorithm \ref{alg:eval}, which evaluates the allocation graph generated by $SolveNode$, returning the corresponding score and CG deviation.
+$ScoreAndDeviation$, described in Algorithm \ref{alg:eval}, evaluates the allocation graph $G$\/ generated by $SolveNode$\/ at node $\pi_k$\/ and returns the corresponding score and CG deviation.
 
 \begin{algorithm}[H]
-	\caption{Procedure to calculate the score and the relative torque deviation of graph $G$\/ in node $\pi_k$}  \label{alg:eval}
+	\caption{Calculating the score and the relative torque deviation of the graph $G$\/ at the node $\pi_k$}  \label{alg:eval}
 	\begin{algorithmic}[1]
@@
 		\State $s \gets 0$
-		\For{$i \gets 1$ to $m$}
-			\State $\tau_i \gets 0$
-		\EndFor
+		\State $\tau_i \gets 0$, $1 \leq i \leq m$
 		\For{$i \gets 1$ to $m$}	\label{eval:loop1}
@@
 
-		\State $\tau \gets  \sum_{1}^{m} \tau_i / (W_{max} \times limit^{CG}_{long})$ \label{eval:torque}
+		\State $\tau \gets  \sum_{i=1}^{m} \tau_i / (W_{max} \times limit^{CG}_{long})$ \label{eval:torque}
 
@@
 
-In this subsection we present two implementations of $SolveNode(method,\ \pi_k,\ G)$, where  $\pi_k$\/ is the current node and $G$\/ is the allocation graph of the PC that remain on board at node $\pi_k$.
-These implementations correspond to the two possible values of the $method$\/ parameter: $MIP$\/ ({\it Gurobi}) or $Shims$\/ (a heuristic).
-
-\subsubsection{MIP solver}
+In this subsection we present two implementations of $SolveNode$\/ algorithm: with a MIP solver and a heuristic.
+
+\subsubsection{With a MIP solver}
 \label{solver}
 
-To find a viable solution for ACLP+RPDP, the strategy adopted in $SolveTour$\/ previously defines the values of some variables.
-Concretely, all permutations between the nodes are tested, the set of nodes to be visited is updated, the PC that remain on board are reallocated to minimize the CG deviation, and the pallets destination is determined according to the volume of items available for shipment.
-In this way, the mathematical modeling for $SolveNode(MIP,\ \pi_k,\ G)$\/ becomes simpler, which finds an allocation of available items in node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $to_i^{\pi_k}$, and $a^{\pi_k}_q$.
-
-This simplified modeling is described in the equations \ref{eq:obj2} to \ref{eq:if2}, which maximizes the score of the $n_{\pi_k}$ items to be shipped, maintaining load balancing.
-
-The variables $X_{ij}$ and $Y_{iq}$ define the sets of edges $E_{N_{\pi_k}}$ and $E_{Q_{\pi_k}}$, respectively, which will be added to the graph $G$.
-
-\vspace{3mm}
-\textbf{Objective function:}
+Our strategy adopted in $SolveTour$\/ previously defines the values of some variables: the set of nodes to be visited is updated, the packed contents that remain on board are reallocated to minimize the CG deviation, and the pallets destinations are determined according to the volume of items available for shipment.
+
+In this way, the mathematical modeling for $SolveNode(MIP,\pi_k,G, time)$\/ becomes simpler, which finds an allocation of available items at the node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$.
+Thus, we use a MIP solver with runtime limit $time$\/ at the node $\pi_k$\/ to maximize the objective function \ref{eq:obj2}, with the calculus equations \ref{eq:newf} to \ref{eq:costs2}, subject to the constraints \ref{eq:tau2} to \ref{eq:ifB}.
+The binary variables $X_{ij}$\/ and $Y_{iq}$\/ define the sets of edges $E_{N_{\pi_k}}$\/ and $E_{Q_{\pi_k}}$, respectively, included in the graph $G$.
 
 \begin{equation} \label{eq:obj2}
-	maximize\ \ \tilde{s} / \tilde{c}
-\end{equation}
-
-\vspace{3mm}
-\textbf{Where:}
+	\max f= \tilde{s} / \tilde{c}
+\end{equation}
 
@@
 \begin{equation} 
-	W_{max} = \sum_{i=1}^{m} W_i
-\end{equation}
-
-\begin{equation} 
-	\tau^{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q) \Big ] \Big / W_{max} \times limit^{CG}_{long}
-\end{equation}
-
-\vspace{3mm}
-In Equation \ref{eq:costs2}, the first parcel denotes the fuel cost from the base until the last node, and the second the fuel cost in the last leg, from the last node back to the base.
-Note that $c_g$ is the fuel consumption potential increase due to the loaded aircraft CG deviation in the node ($\tau_{\pi_k}$).
+	\tau_{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q) \Big ] \Big / W_{max} \times limit^{CG}_{long}
+\end{equation}
 
 \begin{equation} \label{eq:costs2}
-	\tilde{c} = \sum_{k=0}^{K} \Big [ c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|) \Big ]
-\end{equation}
-
-\vspace{3mm}
-\textbf{s.t.:}
-
-{\color{blue} It was observed in the experiments that the lateral torque was always in the order of magnitude -3. So we did not consider these constraints.}
-
-\begin{equation}
+	\tilde{c} =  c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|)
+\end{equation}
+
+\begin{equation} \label{eq:tau2}
 	|\tau_{\pi_k}| \leq 1
-\end{equation}
-
-\begin{equation} 
-	\sum_{i=1}^{m} (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q ) \leq W_{max}
 \end{equation}
@@
 
-\begin{equation} \label{eq:if1}
-	to_i^{\pi_k} = to_j;\ X_{ij} = 1; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
 \begin{equation} \label{eq:if2}
@@
 
-\subsubsection{Shims heuristic}
-
-Our goal is to find a heuristic that offers a good-quality solution for the node-by-node problem.
-As we are testing all $K!$\/ tours, its run time is a crucial requirement.
-Taking this into account, we perform a series of implementations based on known meta-heuristics: {\it Ant Colony Optimization} (ACO) \cite{Dorigo1992, DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \cite{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \cite{Glover1986} and {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \cite{FeoResende1989}.
-
-{\color{blue}
-These meta-heuristics were implemented in their standard forms. We did not use any specialized features to allow for the exploitation of the problem structure.
-}
-
-We also considered several ideas from the literature \cite{NiarFreville1997,Fidanova2006,Laabadi2018,Alonso2019,Zhan2020}, and we were careful to use the same data structures and procedures in all implementations.
-The decision for a simple and fast heuristic is due to the need to obtain, in less than an hour {\color{blue} (because the {\it Brazilian Air Force}'s transportation teams usually performs cargo planning in more than 2 hours)}, a complete solution for ACLP+RPDP, allowing its use in an operational environment.
-
-However, when we established the run time limit of $0.7$\/ seconds per node, the heuristic that presented better solutions was none of the previous ones.
-In this subsection, we will present a new heuristic for the node-by-node problem, called $Shims$.
+\begin{equation} \label{eq:ifA}
+	X_{ij} \leq X_{ij} \times (T_i^{\pi_k} - to_j + 1); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
+\end{equation}
+
+\begin{equation} \label{eq:ifB}
+	X_{ij} \leq X_{ij} \times (to_j - T_i^{\pi_k} + 1 ); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
+\end{equation}
+
+The constraints \ref{eq:ifA} and \ref{eq:ifB} are equivalent to $X_{ij} =1$\/ if $to_j = T_i^{\pi_k}$, and $X_{ij} =0$\/ otherwise .
+
+\subsubsection{With Shims heuristic}
+
+One of the main objectives of this work was to find a quick heuristic that offers a good-quality solution for the node-by-node problem.
+Taking this into account, we design algorithms based on known meta-heuristics: {\it Ant Colony Optimization} (ACO) \cite{Dorigo1992, DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \cite{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \cite{Glover1986} and {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \cite{FeoResende1989}.
+We considered several ideas from the literature \cite{NiarFreville1997,Fidanova2006,Laabadi2018,Alonso2019,Zhan2020}, and we were careful to use the same data structures and procedures in all implementations.
+
+However, the heuristic that presented better solutions was none of the previous ones. In this subsection, we will present a new heuristic for the node-by-node problem, called $Shims$.
+
 Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}).
-This strategy is based on a practical observation: usually, subsets of smaller and lighter items are saved for later adjustments to the remaining {\color{blue}available space}.
-
-The selection of edges for $E_{ N_{\pi_k} }$ uses the {\it edge attractiveness}\/ $\theta^{\pi_k}_{ij}$\/ \ref{eq:edge}, which can be understood as the tendency to allocate the item $j$ to pallet $i$.
-It is directly proportional to the score, and inversely proportional to the volume and the torque of each item.
+This strategy is based on a practical observation: usually, subsets of smaller and lighter items are saved for later adjustments to the remaining available space.
+
+The selection of edges for $E_{ N_{\pi_k} }$ uses the {\it edge attractiveness}\/ $\theta_{ij}$\/, Equation \ref{eq:edge}, which can be understood as the tendency to allocate the item $j$\/ to the pallet $i$\/ at the node $\pi_k$.
+It is directly proportional to the score, and inversely to the volume and the torque of each item.
 
 \begin{equation} \label{eq:edge}
-	\theta^{\pi_k}_{ij} = \frac{ s_j }{ v_j }\ \times \Big ( 1 - \frac{ w_j \times |D_i^{long}| }{ w_{max} \times\ | D_{max}^{long} | } \Big );\ i \in \{1,\ldots,m\},\ j \in \{1,\ldots,n_{\pi_k}\}
-\end{equation}
-
-{\color{blue}The denominator $w_{max} \times\ | D_{max}^{long} | $ represents the heaviest item put on the most distant pallet.}
+	\theta_{ij} = \frac{ s_j }{ v_j }\ \times \Big ( 1 - \frac{ w_j \times |D_i^{long}| }{ \max w_j \times\  \max | D_i^{long} | } \Big );\ i \in \{1,\ldots,m\},\ j \in \{1,\ldots,n_{\pi_k}\}
+\end{equation}
 
@@
 	\begin{minipage}{0.58\linewidth}
-
+		\scriptsize
 		\begin{tikzpicture}[scale=0.75, samples=100]
@@
 			\draw[->] (.2, .6) -- coordinate (x axis mid) (12, .6);
-			\node at (5, 0.3) {$\eta_1^{\ i}$};
+			\node at (5, 0.3) {$\eta_1$};
 			\node at (5, 2.5) {|};
-			\node at (6, 2.5) {shims};
+			\node at (6, 2.5) {\small{shims}};
 			\node at (7, 2.5) {|};
-			\node at (7, 0.3) {$\eta_2^{\ i}$};
-			\node at (0.3, 5) {$\theta_{ij}^{\pi_k}$};
+			\node at (7, 0.3) {$\eta_2$};
+			\node at (0.3, 5) {$\theta_{ij}$};
 			\draw[->] (.6, .2) -- coordinate (y axis mid) (0.6, 5.3);
-			\node at (12, 0.3) {$e_{ij}^{\pi_k}$};
+			\node at (12, 0.3) {$e_{ij}$};
 			\draw[smooth, domain = 0.09:2, color=black, thick] plot (.3+1/\x,{4.2+log2(\x)});
@@
 
-		\captionof{figure}{$n_{\pi_k}$\/ edges $e_{ij}^{\pi_k}$\/ of $p_i$\/ sorted by $\theta_{ij}^{\pi_k}$\/ in non-ascending order}
+		\captionof{figure}{$n_{\pi_k}$\/ possible edges $e_{ij}$\/ sorted by $\theta_{ij}$\/ in non-ascending order}
 		\label{fig:whip}		
@@
 
-Figure \ref{fig:whip} represents the $n_{\pi_k}$\/ possible edges $e_{ij}^{\pi_k}$\/ of $i$\/ sorted by $\theta_{ij}^{\pi_k}$.
-\emph{Shims}\/ starts with a greedy solution, stopping at the edge with index $\eta_1^i$ (first phase).
-Then, considering even the edge with the index $\eta_2^i$, it elaborates different possible complements for this pallet (second phase) and selects the best of these complements (third phase).
-
-In this heuristic, described in Algorithm \ref{alg:shims}, pallets $i$\/ are considered in non-descending order of $|D_i^{long}|$ (line \ref{shims:pallets}).
-For each pallet, the $n_{\pi_k}$\/ possible edges $e_{ ij}^{\pi_k}$\/ are considered in non-increasing order of $\theta_{ij}^{\pi_k}$\/ (line \ref{shims:possible}) in three phases: the greedy phase (until line \ref{shims:endfirst}), the composition phase of the shims (lines \ref{shims:beginsecond}-\ref{shims:endsecond}) and the selection phase of the best shim (lines \ref{shims:beginthird}-\ref{shims:endthird}).
-
-{\color{blue}
-	In the beginning of the first phase (line \ref{shims:initQ}), the set $Q_{\pi_k}$ represents the PC whose destination is not $\pi_k$, the current node, thus remaining on board.
-
-	It is important to remember that $Q_{\pi_k}$ and $M$ were modified by the procedures $UpdatePacked(M, Q_{\pi_k}, \pi_k)$ and $SetPalletsDestinations(M, \pi_k)$, having their positions and destinations reassigned.
-
-	Until line \ref{shims:endfirst}, a greedy and partial solution for pallet $i$ is constructed by edges inclusion, until a certain volume limit ($V_i \times limit$) is reached.
-	The value of the {\it limit} (0.92) was empirically determined by the {\it iRace} tool developed by \cite{LopezIbanezManuel2016}.
-
-	In the second phase (lines \ref{shims:beginsecond} to \ref{shims:endsecond}), a set of shims named $Set$\/ is created, where each {\it Shim} is formed by a set of edges in the range $[\eta_1^i,\eta_2^i]$, whose total volume is limited by $slack_i$.
-In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it First-Fit Decreasing}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \cite{JohnsonGarey1985}.
-Basically, shims are created by accumulating the following edges, taking $slack_i$\/ as a limit.
-
-	In the third phase (lines \ref{shims:beginthird}-\ref{shims:endthird}), the best {\it Shim} in the $Set$\/ is chosen.
-Initially, two shims are found: $sh_w$\/ with larger weight and $sh_v$\/ with larger volume.
-Among the two, the {\it Shim} with the highest score will be chosen and its edges are inserted into $E_{N_{\pi_k}}$.
-}
+Considering only the items that can be shipped at the node $\pi_k$, Figure \ref{fig:whip} represents $n_{\pi_k}$\/ possible edges $e_{ij}$\/ of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order.
+Initially, $Shims$\/ builds a greedy solution for the pallet $i$\/ selecting edges up to index $\eta_1$ (first phase).
+Then, with the edges between $\eta_1$\/ and $\eta_2$, it elaborates different possible complements (second phase), including later the best ones in the same pallet (third phase).
+$Shims$\/ is described in Algorithm \ref{alg:shims}.
 
 \begin{algorithm}[H]
-	\caption{$Shims$\/ heuristic on node $\pi_k$}  \label{alg:shims}
+	\caption{$Shims$\/ heuristic at the node $\pi_k$}  \label{alg:shims}
 	\begin{algorithmic}[1]
 
-		\Procedure{$SolveNode$}{$Shims,\ \pi_k,\ G,\ time$}
+		\Procedure{$SolveNode$}{$Shims, \pi_k, G, time$}
 
@@
 
-		\State $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$ \Comment{a graph with pallets positions ($M$), PC ($Q_{\pi_k}$), and edges ($E_{Q_{\pi_k}}$)}  \label{shims:initQ}
-
-		\State Sort $M$ by $|D_i|$ in non-descending order \label{shims:pallets}
-
-		\State $E_{N_{\pi_k}} \gets \{\ \}$  \Comment{an empty set of pallet-item edges}
-
-		\State Let $E_{ij}$ be an array with $m \times n_{\pi_k}$ edges $(i,\ j)$ sorted by $\theta_{ij}^{\pi_k}$ in non-ascending order \label{shims:possible}
-
-		\State $\tau^{\pi_k} \gets 0$	 \Comment{aircraft torque in node $\pi_k$}	
-		\State $\tau_{max} \gets W_{max} \times limit^{CG}_{long}$ \Comment{maximum aircraft torque}
-
-		\State $volume \gets \{0\ |\ [1,\ldots,m] \}$
-		\State $slack  \gets \{0\ |\ [1,\ldots,m] \}$
-
-		\State $limit \gets 0.92$ \Comment{0.92 was obtained by the {\it iRace} package \cite{LopezIbanezManuel2016}}
-
-		\For{$i \in \{1,\ldots,m\}$}
-
-			\For{$q \in \{1,\ldots,m_{\pi_k}\}$}
-
-				\If{$(i, q) \in E_{Q_{\pi_k}}$} 
-					\State $volume_i \gets volume_i + v_q$ 
+		\State Let $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$   \label{shims:initQ}
+
+		\State Sort $M$ by $|D_i^{long}|$ in non-descending order \label{shims:pallets}
+
+		\State $E_{N_{\pi_k}} \gets \varnothing$  \Comment{An empty set of pallet-item edges}
+
+		\State $\tau_{max} \gets W_{max} \times limit^{CG}_{long}$ \Comment{Maximum aircraft torque}
+
+		\State $limit \gets 0.92$  \label{shims:limit}
+
+		\For{$i \gets 1$ to $m$}  \label{shims:eta1_a}
+
+			\State $\tau_{\pi_k} \gets \sum_{(i,q)\in E_{Q_{\pi_k}}} w_q \times D_i^{long}$ \Comment{Torque due to the packed contents}	
+			\State $vol_i \gets \sum_{(i,q)\in E_{Q_{\pi_k}}} v_q$ \Comment{Volume at the pallet $i$\/ due to the packed contents}
+
+			\State Let $E$\/ be an array of $n_{\pi_k}$ possibles edges of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order \label{shims:possible}
+			\State $\eta_1 \gets 1$  
+			\Repeat
+				\State $e_{ij} \gets E_{\eta_1}$ \Comment{$e_{ij}$\/ is the possible allocation of the item $j$\/ to the pallet $i$}
+
+				\If{ ($E_{N_{\pi_k}} \cup \{e_{ij}\}$ is feasible) {\bf and} ($vol_i \leq V_i \times limit$) {\bf and} ($| \tau_{\pi_k} + w_j \times D_i^{long} | \leq W_{max} \times limit^{CG}_{long}$)}
+
+					\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup \{e_{ij}\}$
+
+					\State $vol_i \gets vol_i + v_j$
+
+					\State $\tau_{\pi_k} \gets \tau_{\pi_k} + w_j \times D_i^{long}$
+
+					\State $\eta_1 \gets \eta_1 + 1$
+
 				\EndIf
-			\EndFor
-
-			\For {$e_{ij} \in E_{ij}$}
-
-				\State $\Delta\tau \gets w_j \times D_i$
-
-				\State $\tau_{new} \gets | \tau^{\pi_k} + \Delta\tau |$
-
-				\If{ ($E_{N_{\pi_k}} \cup \{e_{ij}\}$ is feasible) {\bf and} ($volume_i \leq V_i \times limit$) {\bf and} ($\tau_{new} \leq \tau_{max}$ ) }
-
-					\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup \{e_{ij}\}$
-
-					\State $volume_i \gets volume_i + v_j$
-
-					\State $\eta_1^{\ i} \gets \eta_1^{\ i} + 1$
-
-					\State $\tau^{\pi_k} \gets \tau^{\pi_k} + \Delta\tau$
-
+			\Until ($vol_i > V_i \times limit$) {\bf or} ($\eta_1>n_{\pi_k}$)		 \Comment{End of the first phase} \label{shims:endfirst}
+
+			\State $slack_i \gets V_i - vol_i$  \label{shims:beginsecond}
+			\State $\eta_2 \gets \eta_1$  \label{shims:eta2a}
+			\While{($\eta_2 \leq n_{\pi_k}$) {\bf and} ($ vol_i < (1 + 3 \times (1-limit)) \times V_i)$)}
+
+					\State $e_{ij} \gets E_{\eta_2}$
+
+					\State $vol_i \gets vol_i + v_j$
+					\State $\eta_2 \gets \eta_2 + 1$  \label{shims:eta2} \label{shims:endsecond} \Comment{End of the second phase}
+			\EndWhile
+
+			\State $vol \gets 0$, $b \gets 1$, $shims_b \gets \varnothing$, $Set \gets \{ shims_b \}$  \label{shims:beginthird}
+
+			\For{$x \gets \eta_1$ to $\eta_2$} \label{edges_indexes}
+
+				\If{ $T_{current} - T_{begin} > time$}
+				\State {\bf break}  \Comment{Time limit exceeded}
 				\EndIf
 
-			\EndFor			 \Comment{end of the first phase} \label{shims:endfirst}
-
-			\State $slack_i \gets V_i - volume_i$  \label{shims:beginsecond}
-			\State $\eta_2^{\ i} \gets \eta_1^{\ i}$
-
-			\Repeat
-				\State $\eta_2^{\ i} \gets \eta_2^{\ i} + 1$  \label{shims:eta2}
-
-				\State $e_{ij} \gets E_{i,\eta_2^{\ i}}$
-
-				\State $volume_i \gets volume_i + v_j$
-
-			\Until{($\eta_2^{\ i} = n_{\pi_k}$) {\bf or} ($ volume_i \geq (1 + 3 \times (1-limit)) \times V_i)$)} \label{shims:endsecond} \Comment{end of the second phase}
-
-			\State $volume \gets 0$	\label{shims:beginthird}
-
-			\State $volume \gets 0$, $b \gets 1$ $shims_b \gets \{\}$, $Set \gets Set \cup \{shims_b\}$
-
-			\State $shims_b \gets \{\}$  \label{empty_shims}
-
-			\State $Set \gets Set \cup \{shims_b\}$ \label{empty_set}
-
-			\For{$x \in \{\eta_1^{\ i},\ldots,\eta_2^{\ i} \}$} \label{edges_indexes}
-
-			{\color{blue}
-
-				\If{ current system time - $T_{begin} > time$}
+				\State $NewShims \gets$ {\bf True} \label{new_shims}
+				\State $e_{ij} \gets E_x$
+
+				\For{$shims \in Set$} \label{shims_set}
+
+					\If {($e_{ij} \not\in (E_{N_{\pi_k}} \cup shims))$ {\bf and} ($e_{ij}$ is feasible)  {\bf and} ($(v_j + vol) \leq slack_i$)}
+
+						\State $shims \gets shims \cup \{e_{ij}\}$
+						\State $vol \gets vol + v_j$
+						\State $NewShims \gets$ {\bf False} \label{new_shims_false}
+
 						\State {\bf break}  
@@
 
-			}
-
-				\State $NewShims \gets$ {\bf True} \label{new_shims}
-				\State $e_{ij} \gets E_x^{\ i}$
-
-				\For{$shims \in Set$} \label{shims_set}
-
-					\If { $e_{ij} \not\in (E_{N_{\pi_k}} \cup shims)$ {\bf and} $e_{ij}$ is feasible  {\bf and}  $(v_j + volume) \leq slack_i$}
-
-						\State $shims \gets shims \cup \{\ e_{ij}\ \}$
-						\State $volume \gets volume + v_j$
-						\State $NewShims \gets$ {\bf False} \label{new_shims_false}
-
-						\State {\bf break}
-					\EndIf
-
 				\EndFor
@@
 				\If{$NewShims$} \label{new_shims2}
-					\State $volume \gets 0$
+					\State $vol \gets 0$
 					\State $b \gets b + 1$
@@
 
-			\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup sh_{best}$ 	\label{shims:endthird}   \Comment{end of the third phase}
+			\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup sh_{best}$ 	\label{shims:endthird}   \Comment{End of the third phase}
 
@@
 \end{algorithm}
+
+Initially, $Q_{\pi_k}$\/ (line \ref{shims:initQ}) corresponds to the packed contents that remain on board.
+It is important to remember that $E_{Q_{\pi_k}}$\/ and $M$\/ were modified by $UpdatePacked(M, Q_{\pi_k}, \pi_k)$ and $SetPalletsDestinations(M, \pi_k)$.
+Then, the pallets $i$\/ are considered in non-descending order of $|D_i^{long}|$.
+
+For each pallet $i$, its $n_{\pi_k}$\/ possible edges $e_{ij}$\/ are considered in non-increasing order of $\theta_{ij}$:
+
+\begin{itemize}
+
+	\item In {\it greedy phase}\/ (lines \ref{shims:pallets}-\ref{shims:endfirst}), a partial solution for each pallet $i$\/ is constructed by edges inclusion, until its volume $V_i \times limit$\/ is reached.
+The value $limit=0.92$\/ (line \ref{shims:limit}) was empirically determined by the {\it iRace} tool \cite{LopezIbanezManuel2016}.
+From lines \ref{shims:limit} to \ref{shims:endfirst}, as items are inserted, the value of $\eta_1$ is incremented and stops when the pallet volume reaches 92\% of its capacity.
+When it stops, this is the initial value of $\eta_2$, line \ref{shims:eta2a}.
+	$\eta_2$, then it is incremented until the volume exceeds the pallet capacity with a factor of 1 + ( 2 x [1 - limit]).
+	This factor of was obtained empirically, through many test with different problem instances.
+Higher values can improve the solution, but at the expense of performance.
+So we tried to keep it as low as possible, a trade-off between quality and performance.
+
+	\item In {\it composition phase}\/ (lines \ref{shims:beginsecond}-\ref{shims:endsecond}), a set of shims named $Set$\/ is created for each pallet $i$, where each shim is formed by a set of edges in the range $[\eta_1,\eta_2]$, whose total volume is limited by $slack_i$.
+After the $\eta_2$ index, there may be items with little probability of being in the optimal solution because their low value of $\theta^{\pi_k}_{ij}$, being not considered for inclusion.
+In line \ref{shims:beginthird} the initial $Shim$ and $Set$ are created.
+In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it First-Fit Decreasing}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \cite{JohnsonGarey1985}.
+Basically, shims are created by accumulating the following edges, taking $slack_i$\/ as a limit.
+
+	\item In {\it selection phase}\/ (lines \ref{shims:beginthird}-\ref{shims:endthird}), the best shim in $Set$\/ is chosen.
+Initially, two shims are found: $sh_w$\/ with larger weight, and $sh_v$\/ with larger volume.
+Between the two, the one with the highest score will be chosen and its edges are inserted into $E_{N_{\pi_k}}$.
+
+\end{itemize}
 
@@
 \begin{algorithm}[H]
-	\caption{Procedure to generate items}  \label{alg:itemsgen}
+	\caption{Generating items}  \label{alg:itemsgen}
 	\begin{algorithmic}[1]
@@
 		\Procedure{$ItemsGeneration$}{$scenario, surplus$}
-
-		\State Let $L$ and $M$ be according to $scenario$ \label{ig:LM}
+		\State Let $L$\/ and $M$  \Comment{From Tables \ref{tab:larger} and \ref{tab:scenarios}} \label{ig:LM}
 		\State $limit \gets surplus \times \sum_{i=1}^{m} V_i$ \label{ig:extended}
 
-		\State Let $t_j^k$ be an item $j$ in node $k$.
-
-		\For{$k \in \{0,\ldots,K\}$}
+		\For{$k \gets 0$ to $K$}
 
@@
 				\State $j \gets j+1$
-
+				\State Let $t_j^k$\/ be the item $j$\/ at the node $k$
 				\Repeat
@@
 The parameter $scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and the parameter $surplus$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}).
-To avoid simply loading all items, we use $surplus > 1$ {\color{blue}(1.2, 1.5, and 2.0).
-This represents more test instances for each scenario.}
+To avoid simply loading all items, we use $surplus \in \{1.2, 1.5, 2.0\}$.
+This also represents more instances for tests in each scenario.
 
@@
 
-We ran Algorithm \ref{alg:main} in the 5 scenarios described in Table \ref{tab:scenarios}, considering 6 methods for node-by-node solution: Gurobi (see \ref{solver}), ACO, NMO, TS, GRASP, and Shims (Algorithm \ref{alg:shims}).
+We ran Algorithm \ref{alg:main} in the 5 scenarios described in Table \ref{tab:scenarios}, considering 6 methods for node-by-node solution: Gurobi (see \ref{solver}), ACO, NMO, TS, GRASP, and {\it Shims} (Algorithm \ref{alg:shims}).
 In generating the items in each node, we consider 3 values for the parameter $surplus$.
-The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:20} ($surplus = 1.2$), \ref{tab:50} ($surplus = 1.5$) and \ref{tab:100} ($surplus = 2.0$).
+3 values for the time limit (1200s, 2400s, and 3600s).
+
+The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:20} ($surplus = 1.2$), \ref{tab:50} ($surplus = 1.5$), \ref{tab:100} ($surplus = 2.0$), and in Table \ref{tab:100time} (different time limit comparisons).
+
 For each $method$, $scenario$\/ and $surplus$, 7 different instances were generated.
-Therefore, 105 tests (5 scenarios $\times$\/ 3 volumes $\times$\/ 7 instances) were performed for each method.
-
-The average values were presented for $f_{\pi}$\/ and, for the run time, the worst result obtained.
+Therefore, 315 tests (5 scenarios $\times$\/ 3 volumes $\times$\/ 3 time limits $\times$\/ 7 instances) were performed for each method.
+
+The average values were presented for $f_{\pi}$\/ and, for the {\bf run time (s)}, the average result obtained.
 To facilitate the comparison between the methods, we added a last column in these tables, where two values are indicated:
@@
 The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
-	\item {\bf Speed-up}: ratio of the sums of the worst run times of all scenarios and the sum of the method run times in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
-
-	\item {\bf run time (s)} is the average time to solve a problem instance.
+	\item {\bf Speed-up}: ratio of the sums of the run times of all scenarios and the sum of the method run times in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
 \end{itemize}
 
-In each $scenario$\/, we indicate in bold the best value of $f$\/ found. In each table, we also indicate in bold the best {\bf Normalized}\/ and {\bf Speed-up}\/ values.
-
-{\color{blue}
-
-As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's RAM (16 GiB) too early, we decided to execute a relaxed model with the decision variables set to CONTINUOUS (a Gurobi parameter) to obtain the upper bound for the objective function.
-This approach is much faster than using binary decision variables.
-
-{\color{red}
-Let $X_{ij}^k$\/ and $Y_{iq}^k$\/ be CONTINUOUS variables, where $1 \leq i \leq m$, $1 \leq j \leq n_k$, $1 \leq q \leq m_k$\/ and $0 \leq k \leq K$.
-}
-The upper bound (UB) is the objective function value of a linear programming relaxation of the problem, demonstrating that no integer viable solution can have an objective value greater (in maximization) than this.
-
-We solved all scenarios and instances with Gurobi in a {\it MIPGap} parameter equal to 1\% to control the minimal quality of the returned solution, without a time limit.
-
-Note that, in Table \ref{tab:20} all $Gurobi\ f$\/ values are italicized to indicate that these values are at most 1\% distant from the optimal solution.
-
-}
+	As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's Read-Access Memory (RAM, 16 GiB) too early, we decided to set the Gurobi {\it MIPgap} parameter to 1\% to control the minimal quality of the returned solutions and abbreviate the solution time.
+	From now on, we will refer to this Gurobi-specific model as {\it Gurobi* }.
+	For more details on the Gurobi {\it MIPgap}, refer to \cite{mipgap}.
+
+	This decision hinders the MIP solver from reaching optimal solutions in favor of the possibility of result comparisons with the {\it Shims} heuristic.
+
+	The parameter $\pi_{TSP_1}$ represents the least cost tour (through an optimal TSP solution), and $\pi_{TSP_2}$, the opposite direction of travel, beginning and ending in the base.
+The values of $f$ are the best in these to solutions.
 
@@
 	\centering
-\caption{Methods and scenarios with $surplus =1.2$ and the least cost tours criteria}  \label{tab:20leastcost}
+	\caption{Solutions with volume {\it surplus} = {\bf 1.2} and 3600s as time limit.}  \label{tab:20}
 	\footnotesize
-\begin{tabular}{c|c|cccc}
+	\begin{tabular}{ccccccccc}
 		\toprule
-$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4}  \\
+		{\bf Tours}&$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
 		\toprule
-\multirow{2}{*}{\it Gurobi* } & $f$ & 7.90    & 11.46   & 12.82   & 13.40     \\
-&               {\bf run time (s)}  & 10      & 8       & 11      & 11      \\ 
-\midrule[.1pt]	
-\multirow{2}{*}{\it Shims}    & $f$ & 7.83    & 11.34   & 12.77   & 13.20    \\
-&         {\bf run time (s)}        & 4       & 5       &  6      & 7        \\ 
+		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} &\multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\ % 5927
+		&                                                               &               {\bf run time (s)}  & 17      & 17      & 16      & 17    & 26        &  1.0 \\ % 93
+		\cmidrule(lr){2-9}	
+		&                                                               \multirow{2}{*}{\it Shims}    & $f$ & 8.54    & 11.63   & 12.97   & 13.43 & 12.36     &  0.99 \\ % 5893
+		&                                                               &         {\bf run time (s)}        & 1       & 1       & 2       & 2     & 2         &  11.6 \\ % 8
+
+		\midrule
+
+		\multirow{4}{*}{\it $S_{K}$ }&                                  \multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 12.25   & 13.32   & 14.61  & x        & 1.00 \\ % 4869
+		&                                                               &               {\bf run time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\ % 479
+		\cmidrule(lr){2-9}	
+		&                                                               \multirow{2}{*}{\it Shims}    & $f$ & 8.54    & 12.26   & 13.23   & 14.53  & 13.42    & 0.99  \\ % 4813
+		&                                                               &         {\bf run time (s)}        & 1       & 2       & 6       & 17     & 106      & 18.4   \\ % 132
 		\bottomrule
-\multicolumn{6}{c}{*integer relaxed variables}
+		\multicolumn{9}{c}{*{\it MIPgap} parameter set to 1\%. The $f$ values are \textpm 1\% distant from the optimal solution}\\
+
 	\end{tabular}
@@
 
+{\color{blue}
+
+	97\% of 16GiB was the RAM state before we aborted the {\it Gurobi*} execution on solving scenario 5 in Table \ref{tab:20}.
+
+}
+
 \begin{table}[H]
 	\centering
-\caption{Methods and scenarios with $surplus =1.2$}  \label{tab:20}
+	\caption{Solutions with volume {\it surplus} = {\bf 1.5} and 3600s as time limit.}  \label{tab:50}
 	\footnotesize
-\begin{tabular}{c|c|ccccc|c}
+	\begin{tabular}{ccccccccc}
 		\toprule
-$method$& {\bf Scenarios}                 &{\bf 1}&{\bf 2} & {\bf 3}      & {\bf 4}      & {\bf 5}  & \specialcell{{\bf Normalized}\\{\bf Speed-up}}  \\
+		{\bf Tours} & $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		\toprule
+		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\ % 8236
+		&&               {\bf run time (s)}                                                                  & 38    & 30    & 27    & 29    & 81    & 1.00  \\ % 205
+		\cmidrule(lr){2-9}	
+		&\multirow{2}{*}{\it Shims}                                                                    & $f$ & 11.83 & 16.74 & 18.01 & 18.89 & 16.91 & 1.00  \\ % 
+		&&         {\bf run time (s)}                                                                        & 1     & 2     & 2     & 3     & 3     & 20.5  \\ %
+
+		\midrule
+
+		\multirow{4}{*}{\it $S_{K}$ } & \multirow{2}{*}{\it Gurobi* }                                  & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ % 8602
+		&&               {\bf run time (s)}                                                                  & 53    & 61    & 193   & 470   & 2241    & 1.0  \\ % 3018
+		\cmidrule(lr){2-9}	
+		&\multirow{2}{*}{\it Shims}                                                                    & $f$ & 11.83 & 17.00 & 18.21 & 20.45 & 18.00   & 0.99  \\ % 8549
+		&&         {\bf run time (s)}                                                                        & 1     & 2     & 9     & 26    & 157     & 15.5 \\ % 195
+		\bottomrule
+
+	\end{tabular}
+	\normalsize
+\end{table}
+
+\begin{table}[H]
+	\centering
+	\caption{Solutions with volume {\it surplus} = {\bf 2.0} and 3600s as time limit.}  \label{tab:100}
+	\footnotesize
+	\begin{tabular}{ccccccccc}
+		\toprule
+		{\bf Tours} & $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		\toprule
+		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 24.39 & 26.46 & 27.09 & 24.00 & 1.00 \\ % 11969
+		&&                                                                               {\bf run time (s)}  & 162   & 95    & 74    & 68    & 67    & 1.0  \\ % 466
+		\cmidrule(lr){2-9}	
+		&                                                                   \multirow{2}{*}{\it Shims} & $f$ & 17.78 & 24.40 & 26.36 & 27.01 & 23.94 & 0.99  \\ % 11949
+		&&                                                                                {\bf run time (s)} & 2     & 3     & 3     & 4     & 5     & 27.4  \\ % 17
+
+		\midrule
+
+		\multirow{4}{*}{\it $S_{K}$ } & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &  x   & 1.00 \\ % 9856
+		&&                                                                                {\bf run time (s)} & 151   & 123   & 319   & 689   &  x   & 1.0  \\ % 1282
+		\cmidrule(lr){2-9}	
+		&                                                                \multirow{2}{*}{\it Shims}    & $f$ & 17.78 & 25.25 & 26.38 & 28.98 & 26.08 & 1.00  \\ % 9839
+		&&                                                                                {\bf run time (s)} & 2     & 3     & 13    & 37    & 229   & 23.3  \\ % 284
+		\bottomrule
+
+	\end{tabular}
+	\normalsize
+\end{table}
+
+{\color{blue}
+	{\it Gurobi*} could not solve scenario 5 in Table \ref{tab:100} because $3600/K! = 0.71s$ is not enough time for {\it Gurobi*} to find a feasible node solution.
+}
+
+\subsubsection{Time limits comparisons}
+
+\begin{table}[H]
+	\centering
+	\caption{Solutions with different time limitations for $S_K$ tours.}  \label{tab:100time}
+	\footnotesize
+	\begin{tabular}{ccc|cc|cc|cc}
+		\toprule
+		&             \multicolumn{2}{r}{\bf Volume $surplus$} &\multicolumn{2}{c}{$1.2$}&\multicolumn{2}{c}{$1.5$}&\multicolumn{2}{c}{$2.0$} \\
+		\midrule
+		$method$& {\bf \specialcell{Time\\limit (s)}} & {\bf Scenarios}    & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} \\
+		\toprule
+
+		\multirow{6}{*}{\it Gurobi* }& \multirow{2}{*}{1200} & $f$ & 14.67   & x       &  20.64  &  x      & 27.97   & x \\ % 
+		&        &  {\bf run time (s)}                             & 217     & x       &  304    &  x      & 579     & x \\ % 
+		\cmidrule(lr){2-9}                              
+		& \multirow{2}{*}{2400}& $f$                               & 14.59   &         &  20.60  &  x      & 29.10   & x \\ % 
+		&        &  {\bf run time (s)}                             & 313     &         &  327    &  x      & 543     & x \\ % 
+		\cmidrule(lr){2-9}                              		
+		& \multirow{2}{*}{3600}& $f$                               & 14.61   & x       &  20.59  &         & 29.11   & x \\ % 
+		&        &  {\bf run time (s)}                             & 297     & x       &  470    &         & 689     & x \\ % 		
+		\midrule
+		\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
+		&        & {\bf run time (s)}                              & 18      & 109     &  26     & 160     & 37      & 203   \\ % 
+		\cmidrule(lr){2-9}                              
+		& \multirow{2}{*}{2400}& $f$                               & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
+		&        &  {\bf run time (s)}                             & 18      & 107     &  25     & 159     & 37      & 192   \\ % 	
+		\cmidrule(lr){2-9}                              
+		& \multirow{2}{*}{3600} & $f$                              & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ %  
+		&        &  {\bf run time (s)}                             & 17      & 106     &  26     & 157     & 37      & 181   \\ %  
+		\bottomrule
+
+		&                                       & {\bf Normalized} & 0.99    & x       & 0.99    & x       & 1.00    & x\\
+		&                                       & {\bf Speed-up}   & 15.6    & x       & 14.3    & x       & 16.3    & x\\
+
+		\bottomrule
+
+	\end{tabular}
+	\normalsize
+\end{table}
+
+\subsubsection{Proposed K! strategy assessment}
+
+	Another important test is to compare the best tour score/cost ratio with the least expensive tour score/cost ratio.
+This is important to see how far our solution strategy is from the simplest form, just solving a TSP.
+
+	For this achievement, we ran {\it Shims} in 5 scenarios with 7 instances each, and 3 volume surpluses, collecting the averages of the best tour result and the least cost tour result.
+
+	\begin{table}[H]
+		\centering
+		\caption{{\it Shims} best and least cost tour results}  \label{tab:20least}
+		\footnotesize
+		\begin{tabular}{ccccccccc}
+			\toprule
+			{$surplus$}              & {\bf Tour} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4}  & {\bf 5}   & {\bf Sum} & {\bf Improvement}\\
+			\toprule
+
+			\multirow{2}{*}{$1.2$}   & Least cost    & 8.54    & 11.63   & 12.97   & 13.43  & 12.36 & 58.93 &       \\ 
+			& Best in $K!$  & 8.54    & 12.26   & 13.23   & 14.53  & 13.42 & 61.98 & 5.2\%  \\ 
 			\midrule[.1pt]	
-{\it Gurobi}     & UB for $f$           & {\it 8.91} & {\it 12.92} & {\it 15.12} & {\it 16.29} & {\it 15.34} & -  \\ %51,84
-\toprule
-
+			\multirow{2}{*}{$1.5$}   & Least cost    & 10.89   & 11.83   & 16.74   & 18.01  & 18.89 & 76.36 & \\ 	
+			& Best in $K!$  & 11.33   & 11.83   & 17.00   & 18.21  & 20.45 & 78.82 &  3.2\%  \\ 
 			\midrule[.1pt]	
-\multirow{2}{*}{\it Shims}   & $f$      & 8.43       & 12.77       & 14.47       & 16.28       & 15.15          &   \\
-&{\bf run time (s)}                     & 4          &  8          &  33         & 93          & 560          &   \\
+			\multirow{2}{*}{$2.0$}   & Least cost    & 17.78   & 24.40   & 26.36   & 27.01  & 23.94 & 119.49 &  \\ 
+			& Best in $K!$  & 17.78   & 25.25   & 26.38   & 28.98  & 26.08 & 124.47 & 4.2\%   \\ 		                         
+			\midrule[.1pt]
+			\multicolumn{8}{r}{{\bf K! strategy advantage}}                                              & {\bf 4.2}\% \\
 
@@
 
-{\color{blue}
-It is important to note that the best tour found in each test was seldom the tour with the lowest cost (the minimal cost tour without considering the CG deviation), but any of the first 25\% of the best tours.
+	As the selection of the best tour, considering what was collected and delivered in each node, and considering the cost increase due to CG deviation, was better than the option for a simple TSP solution (\textit{Least cost} results), this confirms that our solution strategy was adequate.
+
+	It is important to report that the best tour found in each test was seldom the tour with the lowest cost (the minimal cost tour without considering the CG deviation), but any of the first 25\% of the best tours.
 This indicates that our strategy of enumerating and solving all tours was assertive, although it is possible to devise a more proactive strategy to discard most of the higher-cost tours and improve performance.
@@
 	If a posterior application of this method requires 8, 9, or more nodes, it is possible to use this 25\% as a threshold to discard the higher-cost tours and keep the run time within an operational acceptable limit for the client.
-}
-
-It is worth noting that the other heuristics always obtained {\bf Normalized}\/ values greater than or equal to $0.9$, and were always faster than $Gurobi$.
-
-It is not possible to state how far these results are from the optimal solutions, but we can report that all occupancy rates in terms of volume were above 97\%, and in terms of weight above 79\%, which denotes that the results achieved are good, considering that the CG deviation constraints keeps these values from going higher.
-
-\subsubsection{Proposed strategy assessment}
-
-Another important test is to compare the best tour score/cost ratio with the least expensive tour score/cost ratio.
-This is important to see how far our solution strategy is from the simplest form, just solving a TSP.
-
-For this achievement, we ran {\it Shims} in 5 scenarios with 7 instances each, and 3 volume surpluses, collecting the averages of the best tour result and the least cost tour result.
-
-\begin{table}[H]
-\centering
-\caption{Shims best and least cost tour results}  \label{tab:20least}
-\footnotesize
-\begin{tabular}{c|c|ccccc|cc}
-\toprule
-{$surplus$}              & {\bf Tour} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4}  & {\bf 5}   & {\bf Sum} & {\bf Improvement}\\
-
-\midrule[.1pt]	
-\multirow{2}{*}{$1.2$}   & Least cost    & 6.23  & 11.56   & 12.88   & 13.54    & 12.47  & 56.68 &       \\ 
-& Best                                   & 7.96  & 13.04   & 14.30   & 15.56    & 15.28  & 66.14 & 16.7\%  \\ 
-\midrule[.1pt]	
-\multirow{2}{*}{$1.5$}   & Least cost    & 8.44  & 16.61   & 18.18   & 19.68    & 17.64  & 80.55 &   \\ % 17.64	 20.94	
-& Best                                   & 11.47 & 18.25   & 20.61   & 22.22    & 20.94  & 93.49 & 16.1\%  \\ 
-\midrule[.1pt]	
-\multirow{2}{*}{$2.0$}   & Least cost    & 13.38 & 22.46   & 25.97   & 26.23    & 22.51  & 110.55 &   \\ %22.51	 29.09
-& Best                                   & 17.93 & 25.59   & 28.61   & 31.28    & 29.09  & 132.50 & 19.9\%  \\
-
-\midrule[.1pt]
- \multicolumn{8}{r}{{\bf Average improvement}} & 17.6\% \\
-\bottomrule
-\end{tabular}
-\normalsize
-\end{table}
-
-As the selection of the best tour, considering what was collected and delivered in each node, and considering the cost increase due to CG deviation, was 17.6\% better than the option for a simple TSP solution (\textit{Least cost} results), this confirms that our solution strategy was adequate.
+
+	\subsubsection{Final considerations on the results}
+
+	Most of the {\it Gurobi*} executions consumed over 12 GiB in total, and considering that the computer used in the experiments has a standard RAM consumption when idle of 3.5 GiB (with the Python Interface Development Environment (IDE), a PDF viewer, a system monitor, and a {\it LaTeX} IDE simultaneously open), the actual RAM consumption of {\it Gurobi*} was over 8.5 GiB.
+
+	All {\it Shims} executions consumed at most 1.5 GiB, requiring 5 GiB of total memory.
+	These details are important for a future application to adequately dimension the computer to be used.
 
@@
 In this work, we modeled and solved a real air transport problem, named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem}\/ (ACLP+RPDP).
-For the first time in the literature, a NP-hard problem that involves {\it simultaneously}\/ pallet assembly, load balancing and route planning is addressed, where the cost-effectiveness of transport is maximized.
+For the first time in the literature, a {\it NP-hard}\/ problem that involves {\it simultaneously}\/ pallet assembly, load balancing and route planning is addressed, where the cost-effectiveness of transport is maximized.
 We adopted some simplifications that are not critical, but that allowed the unprecedented solution of this problem considering more than two nodes.
@@
 We consider that, in practical cases, the number $K$\/ of nodes, excluding the base, is small ($K \leq 6$), each of them with hundreds of items to be shipped.
-Thus, considering real aircraft models, we developed some node-by-node heuristics, which allow us to enumerate all $K!$\/ tours and choose the best.
+Thus, considering real aircraft model, we developed some node-by-node heuristics, which allow us to enumerate all $K!$\/ tours and choose the best.
 The complete process can be executed quickly on a simple handheld computer, offering good results and reducing stress for the transport planners.
@@
 
-In less than one hour, this development can establish a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximizing the total score and minimizing fuel consumption for the airlift.
+{\color{blue} In less than 20 minutes}, this development can establish a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximizing the total score and minimizing fuel consumption for the airlift.
 The output, which includes the tour plan and the pallet building and arrangement plan, is an essential part of airlift: it improves flight safety, makes ground operations more efficient, and makes sure that each item gets to its right destination.
 
-Our main contributions were the mathematical modeling of this complex problem, which involves four NP-hard sub-problems, and a complete process that offers a fast and good-quality solution.
-
 {\color{blue}
+
+Our main contributions were the mathematical modeling (1) of this complex problem, which involves four {\it NP-hard}\/ sub-problems, a complete process (2), and a new heuristic (3) that offer fast and good-quality solutions to the ACLP+RPDP.
+
 As this is ongoing research, some improvements are possible, which are the next steps of the development:
 
-(1) This work addresses all possible tours, but we plan bounding schemes to allow for discarding most of the higher-cost tours.
-This is expected to reduce the running time, favoring the discovery of better search mechanisms.
-
-(2) We are also working on computer parallelism to accelerate the algorithms.
-
-(3) In the event that these gains in performance are achieved, we strongly believe that they will permit the application of a 3-dimensional packing procedure to maximize the occupancy rate among the selected items.
-}
-
-The simple heuristic $Shims$\/ presented a solution with the best node-by-node results in a restricted time interval.
+(1) This work addresses all possible tours, but we plan bounding schemes to allow for discarding most of the higher-cost tours. This is expected to reduce the running time.
+
+(2) We are also working on computer parallelism to accelerate {\it Shims}.
+
+(3) In the event that these gains in performance are achieved, we strongly believe that they will permit the application of a 3-dimensional packing procedure to maximize the occupancy rate of the items selected by {\it Shims}.
+
+The $Shims$\/ heuristic presented a solution with the best node-by-node results in a restricted time interval.
 Without it, it would not be possible to test all routes: after all, the greater the number of nodes, the shorter the run time limit for node-by-node solution.
@@
 In these cases, it would be necessary to make some changes in the modeling: for example, modify the load balancing constraints, and consider the available space in vehicles or freight cars instead of pallets.
+
+}
 
@@
 
-\bibitem{LopezIbanezManuel2016} López-Ibáñez, Manuel and Dubois-Lacoste, Jérémie and Pérez Cáceres, Leslie and Birattari, Mauro and Stützle, Thomas.
-2016.
-The irace package: Iterated racing for automatic algorithm configuration.
-Elsevier.
-pp.
-43-58.
-Volume 3.
+\bibitem{LopezIbanezManuel2016} M.
+López-Ibáñez, J.
+Dubois-Lacoste, L.P.
+Cáceres, M.
+Birattari and T.
+Stützle, The irace package: Iterated racing for automatic algorithm configuration, {\it Operations Research Perspectives}, Vol.
+3, pp.
+43-58, 2016.
+
+	\bibitem{mipgap} Matthias Miltenberger. {\it What is the MIPGap?} \url{https://support.gurobi.com/hc/en-us/articles/8265539575953-What-is-the-MIPGap-}, 2023.
 

### 2023-06-08 CAIE/results.tex
cie file updated and experimentations almost ended

@@
 \multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} &\multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\ % 5927
-&&               {\bf run time (s)}  & 17      & 17       & 16      & 17   & 26        &  1.0 \\ % 93
+&                                                               &               {\bf run time (s)}  & 17      & 17      & 16      & 17    & 26        &  1.0 \\ % 93
 \cmidrule(lr){2-9}	
-&\multirow{2}{*}{\it Shims}    & $f$ & 7.83    & 11.34   & 12.77   & 13.20 & 12.22     &  0.97 \\ % 5736
-&&         {\bf run time (s)}        & 4       & 5       &  6      & 7     & 10        &  2.9  \\ % 32
+&                                                               \multirow{2}{*}{\it Shims}    & $f$ & 8.54    & 11.63   & 12.97   & 13.43 & 12.36     &  0.99 \\ % 5893
+&                                                               &         {\bf run time (s)}        & 1       & 1       & 2       & 2     & 2         &  11.6 \\ % 8
 
@@
 
-\multirow{4}{*}{\it $S_{K}$ }&\multirow{2}{*}{\it Gurobi* } & $f$ & 8.51    & 12.25   & 13.32   & 14.61  & x        & 1.00 \\ % 4869
-&&               {\bf run time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\ % 479
+\multirow{4}{*}{\it $S_{K}$ }&                                  \multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 12.25   & 13.32   & 14.61  & x        & 1.00 \\ % 4869
+&                                                               &               {\bf run time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\ % 479
 \cmidrule(lr){2-9}	
-&\multirow{2}{*}{\it Shims}    & $f$ & 8.11    & 12.18   & 13.25   & 14.57  & 14.33    &  0.99 \\ % 4811
-&&         {\bf run time (s)}        & 4       & 8       & 33      & 95     & 572      &  3.4  \\ % 140
+&                                                               \multirow{2}{*}{\it Shims}    & $f$ & 8.54    & 12.26   & 13.23   & 14.53  & 13.42    & 0.99  \\ % 4813
+&                                                               &         {\bf run time (s)}        & 1       & 2       & 6       & 17     & 106      & 18.4   \\ % 132
 \bottomrule
@@
 		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{\it Shims}    & $f$ & 10.89 & 16.65 & 17.80 & 18.69 & 16.45 & 0.98  \\ % 8048
-		&&         {\bf run time (s)}        & 4     & 5     & 6     & 7     & 8     & 7.1   \\ % 29
+		&\multirow{2}{*}{\it Shims}                                                                    & $f$ & 11.83 & 16.74 & 18.01 & 18.89 & 16.91 & 1.00  \\ % 
+		&&         {\bf run time (s)}                                                                        & 1     & 2     & 2     & 3     & 3     & 20.5  \\ %
 
@@
 		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{\it Shims}    & $f$ & 11.33 & 16.97 & 17.56 & 20.43 & 17.94   & 0.98  \\ % 8423
-		&&         {\bf run time (s)}        & 4     & 8     & 34    & 96    & 591     & 4.1 \\ % 733
+		&\multirow{2}{*}{\it Shims}                                                                    & $f$ & 11.83 & 17.00 & 18.21 & 20.45 & 18.00   & 0.99  \\ % 8549
+		&&         {\bf run time (s)}                                                                        & 1     & 2     & 9     & 26    & 157     & 15.5 \\ % 195
 		\bottomrule
@@
 		\toprule
-		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 24.39 & 26.46 & 27.09 & 24.00 & 1.00 \\ % 119.69
+		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 24.39 & 26.46 & 27.09 & 24.00 & 1.00 \\ % 11969
 		&&                                                                               {\bf run time (s)}  & 162   & 95    & 74    & 68    & 67    & 1.0  \\ % 466
 		\cmidrule(lr){2-9}	
-		&                                                                   \multirow{2}{*}{\it Shims} & $f$ & 16.73 & 24.06 & 26.15 & 26.25 & 23.54 &  0.97 \\ % 116,48
-		&&                                                                                {\bf run time (s)} & 5     & 6     & 8     & 10    & 11    &  11.7  \\ % 40
+		&                                                                   \multirow{2}{*}{\it Shims} & $f$ & 17.78 & 24.40 & 26.36 & 27.01 & 23.94 & 0.99  \\ % 11949
+		&&                                                                                {\bf run time (s)} & 2     & 3     & 3     & 4     & 5     & 27.4  \\ % 17
 
@@
 		\cmidrule(lr){2-9}	
-		&                                                                \multirow{2}{*}{\it Shims}    & $f$ & 16.88 & 24.30 & 26.24 & 28.65 & 28.74 & 0.97  \\ % 9607
-		&&                                                                                {\bf run time (s)} & 5     & 9     & 37    & 108   & 625   & 8.1  \\ % 159
+		&                                                                \multirow{2}{*}{\it Shims}    & $f$ & 17.78 & 25.25 & 26.38 & 28.98 & 26.08 & 1.00  \\ % 9839
+		&&                                                                                {\bf run time (s)} & 2     & 3     & 13    & 37    & 229   & 23.3  \\ % 284
 		\bottomrule
@@
 {\color{blue}
-{\it Gurobi*} could not solve scenario 5 in Table \ref{tab:100} because $3600/K! = 0.71s$, not enough time for {\it Gurobi*} to find a feasible node solution.
+{\it Gurobi*} could not solve scenario 5 in Table \ref{tab:100} because $3600/K! = 0.71s$ is not enough time for {\it Gurobi*} to find a feasible node solution.
 }
@@
 		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$                               & 14.59   & 13.59   &  20.60  &  x      & 29.10   & x \\ % 
-		&        &  {\bf run time (s)}                             & 313     & 1460    &  327    &  x      & 543     & x \\ % 
+		& \multirow{2}{*}{2400}& $f$                               & 14.59   &         &  20.60  &  x      & 29.10   & x \\ % 
+		&        &  {\bf run time (s)}                             & 313     &         &  327    &  x      & 543     & x \\ % 
 		\cmidrule(lr){2-9}                              		
-		& \multirow{2}{*}{3600}& $f$                               & 14.61   & x       &  20.59  & 18.36   & 29.11   & x \\ % 
-		&        &  {\bf run time (s)}                             & 297     & x       &  470    & 2241    & 689     & x \\ % 		
+		& \multirow{2}{*}{3600}& $f$                               & 14.61   & x       &  20.59  &         & 29.11   & x \\ % 
+		&        &  {\bf run time (s)}                             & 297     & x       &  470    &         & 689     & x \\ % 		
 		\midrule
-		\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ & 15.14   &         &         &         &         &   \\ % 
-		&        & {\bf run time (s)}                              & 73      &         &         &         &         &   \\ % 
+		\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
+		&        & {\bf run time (s)}                              & 18      & 109     &  26     & 160     & 37      & 203   \\ % 
 		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$                               &         &         &         &         &         &   \\ % 
-		&        &  {\bf run time (s)}                             &         &         &         &         &         &   \\ % 	
+		& \multirow{2}{*}{2400}& $f$                               & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
+		&        &  {\bf run time (s)}                             & 18      & 107     &  25     & 159     & 37      & 192   \\ % 	
 		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{3600} & $f$                              & 14.57   & 14.33   &  20.43  & 17.94   & 28.65   & 28.74 \\ %  
-		&        &  {\bf run time (s)}                             &  95     &  572    &  96     & 591     & 108     &  625  \\ %  	                                        
+		& \multirow{2}{*}{3600} & $f$                              & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ %  
+		&        &  {\bf run time (s)}                             & 17      & 106     &  26     & 157     & 37      & 181   \\ %  
+		\bottomrule
+
+		&                                       & {\bf Normalized} & 0.99    & x       & 0.99    & x       & 1.00    & x\\
+		&                                       & {\bf Speed-up}   & 15.6    & x       & 14.3    & x       & 16.3    & x\\
+
 		\bottomrule
@@
 \end{table}
-
-\begin{figure}[H]
-\centering
-\begin{tikzpicture}
-	\begin{axis}[
-		area style,
-		]
-		\addplot+[ybar interval,mark=no] plot coordinates { 
-			(0,.3)
-			(1,.5)
-			(2,.6)
-			(3,.5)
-			(4,.25)
-			(5,.1)
-			(6,.5)
-			(7,.25)
-			(8,.1)
-		};
-	\end{axis}
-\end{tikzpicture}
-\captionof{figure}{ESTE GRÁFICO É APENAS UM MODELO}
-\end{figure}
 
@@
 
-\multirow{2}{*}{$1.2$}   & Least cost    & 7.83    & 11.34 & 12.77 & 13.20 & 12.22  & 57.36 &       \\ 
-                         & Best in $K!$  & 8.11    & 12.18 & 13.25 & 14.57 & 14.33  & 62.44 & 8.9\%  \\ 
+\multirow{2}{*}{$1.2$}   & Least cost    & 8.54    & 11.63   & 12.97   & 13.43  & 12.36 & 58.93 &       \\ 
+                         & Best in $K!$  & 8.54    & 12.26   & 13.23   & 14.53  & 13.42 & 61.98 & 5.2\%  \\ 
 \midrule[.1pt]	
-\multirow{2}{*}{$1.5$}   & Least cost    & 10.89   & 16.65 & {\color{red}17.80} & 18.69 & 16.45  & 80.48 & \\ 	
-                         & Best in $K!$  & 11.33   & 16.97 & 17.56 & 20.43 & 17.94  & 84.23 & 4.8\%  \\ 
+\multirow{2}{*}{$1.5$}   & Least cost    & 10.89   & 11.83   & 16.74   & 18.01  & 18.89 & 76.36 & \\ 	
+                         & Best in $K!$  & 11.33   & 11.83   & 17.00   & 18.21  & 20.45 & 78.82 &  3.2\%  \\ 
 \midrule[.1pt]	
-\multirow{2}{*}{$2.0$}   & Least cost    & 16.73   & 24.06 & 26.15 & 26.25 & 23.54  & 116.73&  \\ 
-                         & Best in $K!$  & 16.88   & 24.30 & 26.24 & 28.65 & 28.74  & 124.81& 6.7\%   \\ 		                         
+\multirow{2}{*}{$2.0$}   & Least cost    & 17.78   & 24.40   & 26.36   & 27.01  & 23.94 & 119.49 &  \\ 
+                         & Best in $K!$  & 17.78   & 25.25   & 26.38   & 28.98  & 26.08 & 124.47 & 4.2\%   \\ 		                         
 \midrule[.1pt]
-\multicolumn{8}{r}{{\bf K! strategy advantage}}                                              & {\bf 6.8}\% \\
+\multicolumn{8}{r}{{\bf K! strategy advantage}}                                              & {\bf 4.2}\% \\
 

### 2023-06-08 CAIE/shims.tex
cie file updated and experimentations almost ended

--- 2023-06-08 CAIE/cie.tex
@@
 
-\begin{frontmatter}
-
-\title{Air cargo load and route planning in pickup and delivery operations}
-
-\author{A.C.P.~Mesquita}
-\ead{celio@ita.br}
-
-\author{C.A.A.~Sanches\corref{cor1}}
-\ead{alonso@ita.br}
-\cortext[cor1]{Corresponding author.}
-
-\address {Instituto Tecnol\'{o}gico de Aeron\'{a}utica - DCTA/ITA/IEC\\
-Pra\c{c}a Mal. Eduardo Gomes, 50\\
-S\~{a}o Jos\'{e} dos Campos - SP - 12.228-900 - Brazil}
-
-\begin{abstract}
-
-In aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading, for fast take-off, and mission accomplishment, especially in times of crisis, disaster relief missions, short deadlines for contracting customers, or any external pressure for immediate take-off.
-In addition, there are no commercially available systems that can assist the load and trip planners with the pallet building and aircraft loading and balancing demands for transport at each hub.
-This enables other risks, such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
-We defined and solved the problem of planning the loading and routing of a single aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
-This NP-hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modeled using standardized pallets in fixed positions, obeying center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
-We performed multiple experiments with a commercial solver and four well-known meta-heuristics on data based on the transport history of the {\it Brazilian Air Force}.
-We also designed a new heuristic that quickly finds good solutions for a wide range of problem sizes: an essential contribution, which solved all test scenarios within an acceptable operational run time.
-
-\end{abstract}
-
-\begin{keyword}
-Load Planning \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Vehicle Routing
-\end{keyword}
-
-\end{frontmatter}
-
-\label{sec1}
-\section{Introduction}
-
-Air cargo transport involves several sub-problems that are difficult to solve.
-Recently, Brandt and Nickel \cite[p.
-401]{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP) and {\it Weight and Balance Problem} (WBP).
-Several aspects were considered in this modeling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc.
-They also presented a comprehensive bibliographic survey of solving methods that had been developed in different situations.
-
-However, there are still other important challenges in air cargo transport that go beyond the definition of the ACLPP, especially with regard to the flight itinerary and the loading and unloading at each destination (or node) of this travel plan.
-In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Pickup and Delivery Problem} (PDP), and the search for the lowest cost route, which is the well-known {\it Traveling Salesman Problem} (TSP).
-
-Considering air cargo transport, Table \ref{tab:sa} lists the main works in the literature and the corresponding sub-problems addressed.
-We also indicate whether the dimensions of the items were taken into account ({\bf 3D} or {\bf 2D}) and which solution method was used: heuristic search methods ({\bf Heu}), integer programming ({\bf Int}), or linear programming ({\bf Lin}).
-
-\begin{table}[H]
-	\centering
-	\caption{Air cargo transport: literature, problems and features}  \label{tab:sa}
-	\scriptsize
-	\renewcommand{\arraystretch}{1.1} 
-	\begin{tabular}{r|cccc|ccccc}
-		\toprule
-		 Work & {\bf APP}  & {\bf WBP}  &  {\bf PDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} \\
-		\midrule
-		Larsen and Mikkelsen \cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$ \\
-		Brosh \cite{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$ \\
-		Ng \cite{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ \\
-		Heidelberg {\it et al.} \cite{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$ \\
-		Mongeau and Bes \cite{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		Fok and Chun \cite{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
-		Chan {\it et al.} \cite{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$ \\
-		Kaluzny and Shaw \cite{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$ \\
-		Verstichel {\it et al.} \cite{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		Mesquita and Cunha \cite{MesquitaCunha2011}   & $.$ & $.$    & $\bigstar$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\		
-		Limbourg {\it et al.} \cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		Roesener and Hall \cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
-		Vancroonenburg {\it et al.} \cite{Vancroonenburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		Lurkin and Schyns \cite{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		Roesener and Barnes \cite{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
-		Paquay {\it et al} \cite{PaquaySchynsLimbourg2016,PaquayLimbourgSchynsOliveira2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$ \\
-		Chenguang {\it el al.} \cite{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$ \\
-		Wong and Ling \cite{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$  \\
-		Wong {\it et al.} \cite{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$  \\
-		Zhao {\it et al.} \cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
-		{\bf This article}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$  \\
-		\bottomrule 
-	\end{tabular}
-	\normalsize 
-\end{table}
-
-As can be seen, so far Lurkin and Schyns \cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem.
-Although it is innovative, strong simplifications were imposed by these authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a predefined tour plan restricted to only two legs.
-It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem.
-Referring directly to this work, Brandt and Nickel \cite[p.
-409]{BrandtStefan2019} comment: {\it However, not even these sub-problems are acceptably solved for real-world problem sizes or the models omit some practically relevant constraints}.
-
-There are real situations that are much more complex.
-In this work, we consider a practical case in Brazil, which is the largest economy in Latin America.
-Due to its dimensions, this country has the largest air market on the continent with $2,499$\/ registered airports, of which $1,911$\/ are private and $588$\/ are public.
-Although it is an immense distribution network, the {\it Brazilian Air Force}\/ missions have always considered 3 to 5 nodes per flight plan.
-
-It is important to emphasize that this data is not an imposed limitation, but a historical fact that we will explore in our solution method.
-Throughout this work we approach routes with up to 7 nodes, as can be seen in Table \ref{tab:costs} and in Figure \ref{fig:nodes}.
-Although there are many other airports of interest to {\it Brazilian Air Force}, these 7 nodes were chosen due to their high demand and short transport times.
-The other Brazilian airports tend to have a lower demand, which is usually met, in a less expensive way, by cabotage, rail or road transport.
-
-\begin{table}[H]
-
-	\begin{minipage}{0.05\linewidth}
-
-	\end{minipage}\hfill 
-	\begin{minipage}{0.45\linewidth}
-
-		\caption{Brazilian airports distances ($km$)}  \label{tab:costs}
-		\centering
-		\footnotesize
-		\newcolumntype{Y}{>{\centering\arraybackslash}p{0.09\textwidth}}
-		\newcolumntype{X}{>{\centering}p{0.09\textwidth}}
-
-		\begin{tabular}{X X X X X X X Y}
-			\toprule
-			Node & $l_0$ & $l_1$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ & $l_6$ \\
-			IATA*   & GRU   & GIG   & SSA   & CNF   & CWB   & BSB   & REC \\	
-			\midrule	
-			GRU     & 0	    &343	&1,439   &504    &358    &866    &2,114\\
-			GIG	    & 343	&0	    &1,218   &371    &677    &935    &1,876\\
-			SSA	    & 1,439	&1,218	&0	    &938    &1,788   &1,062   &676\\
-			CNF	    & 504	&371	&938	&0	    &851    &606    &1,613\\
-			CWB	    & 358	&677	&1,788	&851	&0	    &1,084   &2,462\\
-			BSB	    & 866	&935	&1,062	&606	&1,084	&0	    &1,658\\
-			REC	    & 2,114	&1,876	&676	&1,613	&2,462	&1,658	&0\\
-			\bottomrule
-			\multicolumn{8}{c}{*International Air Transport Association}\\
-			\multicolumn{8}{c}{\small\textsuperscript{Source: www.airportdistancecalculator.com}}\\
-		\end{tabular}
-		\normalsize
-
-	\end{minipage}\hfill 
-	\begin{minipage}{0.50\linewidth}
-		\centering
-		\includegraphics[scale=0.25]{Images/nodes.png}
-		\captionof{figure}{A route with 7 Brazilian airports}
-		\label{fig:nodes}		
-	\end{minipage}
-\end{table}
-
-Cargo handling at modern airports usually involves powerful equipment such as dollies, motorized conveyors, elevator transfer vehicles, caster decking, scissor lifts, electrical controls, hydraulically adjustable heights, forklift trucks, and plenty of room in the airport cargo terminal (see Figure \ref{fig:handling}).
-These facilities, together with the use of standardized pallets with predefined positions on the aircraft, allow loading and unloading at each node to be carried out in about 30 minutes.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=0.2]{Images/handling.png}
-	\captionof{figure}{Cargo handling equipment}
-	\label{fig:handling}
-	\small\textsuperscript{Source: NARA \& DVIDS Public Domain Archive}	
-\end{figure}
-
-However, as reported by Fok and Chun, {\it load planning (...) is usually done in roughly 2 hours before departure, when all the details of the cargo are present.} \cite{fok2004optimizing}.
-We can infer that this is because adequate commercial software does not exist.
-In addition, there is still the necessary time to plan the route, in order to minimize fuel consumption and ensure the transport of priority items.
-In these circumstances, it is of great importance to find a method that speeds up the load planning and the definition of the flight itinerary.
-
-Our work proposes a method that prioritizes the transport of the most relevant items at each node and the fuel economy along the tour.
-We developed a heuristic that can be run on a simple handheld computer (such as a laptop or a tablet) that previously defines a flight plan, and provides a quick solution for cargo handling.
-Consequently, this method reduces the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets (regarding their positions), and a pick up and delivery plan to each node.
-
-To the best of our knowledge, this is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP and TSP has been addressed.
-This new problem is named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem} (ACLP+RPDP).
-
-This article is organized into six more sections.
-In Section \ref{sec2}, we give a brief review of the literature.
-In Section \ref{sec3}, we present the problem context and assumptions, and in Section \ref{sec4}, we describe the mathematical model and how we dealt with its issues.
-In Section \ref{sec5}, we describe the elaborate algorithms, whose results are presented in Section \ref{sec6}.
-Finally, our conclusions are in Section \ref{sec7}.
-
-\section{Related literature}
-\label{sec2}
-
-In this section, we briefly describe the characteristics of the main works related to air cargo transport, following the chronological order of Table \ref{tab:sa}.
-
-Larsen and Mikkelsen \cite{LarsenMikkelsen1979} developed an interactive procedure for loading 14 types of Boeing 747 into a two-leg transportation plan.
-Seven types of items were considered to be allocated in 17 to 42 positions.
-With non-linear programming and heuristics, they present a solution that minimizes positioning changes in the intermediate node, optimizing the load balancing in the aircraft.
-
-Brosh \cite{Brosh1981} addressed the problem of planning the allocation of cargo on an aircraft.
-Considering volume, weight and structural constraints, the author finds the optimal load layout through a fractional programming problem.
-
-Ng \cite{Kevin1992} developed a multi-criteria optimization approach to load the {\it C-130} aircraft of the {\it Canadian Air Force}.
-Based on integer programming, this model provides timely planning and improves airlift support for combat operations, solving WBP with pallets in fixed positions, and considering 20 different items.
-
-Heidelberg {\it et al.} \cite{Heidelberg1998} developed a heuristic for 2D packing in air loading, comparing it with methods for solving the {\it Bin Packing Problem}.
-These authors conclude that the classical algorithms are inadequate in this context, because they ignore the aircraft balancing constraints.
-
-Mongeau and Bes \cite{MongeauBes2003} presented a method based on linear integer programming to solve the problem of choosing and positioning containers in the {\it Airbus 340-300}.
-Safety and stability constraints were considered, with the objective of minimizing fuel consumption.
-
-Fok and Chun \cite{fok2004optimizing} developed a web-based application to make efficient use of space and load balancing for an air cargo company.
-Based on an analysis of historical data, an operational load planning with mathematical optimization is obtained.
-This container load planning is usually done in roughly 2 hours before departure, when all cargo details are in place.
-
-Chan {\it et al.} \cite{Chan2006} carried out a case study with heterogeneous pallets.
-To minimize the total cost of shipping, they developed a 3D packing heuristic, with a loading plan for each pallet.
-Although the authors do not consider load balancing or positioning of pallets in the cargo hold, this method is relevant in commercial and industrial applications, where cargo items tend to be less dense.
-
-Kaluzny and Shaw \cite{KaluznyBohdanL2009Oalb} developed a mixed integer linear programming model to arrange a set of items in a military context that optimizes the load balance.
-The objective function can be chosen to minimize the deviation of the center of gravity (CG) from the target position or to maximize the function of the items loaded.
-This approach does not palletize items, which are arranged in the cargo bay by a 2D packing procedure.
-
-Verstichel {\it et al.} \cite{Verstichel2011} solved WBP by selecting the most profitable subset of containers to be loaded into an aircraft by the use of mixed-integer programming.
-Experimental results on real-life data showed significant improvements compared to those obtained manually by an experienced planner.
-
-Mesquita and Cunha \cite{MesquitaCunha2011} presented a heuristic for a real problem of the {\it Brazilian Air Force}, which consists of defining transport routes with simultaneous collection and delivery from a central distribution terminal.
-
-Limbourg {\it et al.} \cite{Limbourg2012} developed a mixed-integer program for optimally rearranging a set of pallets into a compartmentalized cargo aircraft, specifically the {\it Boeing 747}.
-
-Roesener and Hall \cite{RoesenerHall2014} solved APP and WBP as an integer programming problem, where the items are selected for shipment based on a utility score, then assigned to pallets, which will be loaded into an aircraft in a specific pallet position.
-The pallets are then packed in a manner to optimize both the pallet and aircraft characteristics, such as item utility, pallet occupancy, and aircraft center of balance.
-This method groups items by destinations to be served by multiple aircraft.
-
-Vancroonenburg {\it et al} \cite{Vancroonenburg2014} presented a mixed integer linear programming model that selects the most profitable pallets, satisfying safety and load balancing constraints on the {\it Boeing 747-400}.
-Using a solver, these authors solved real problems in less than an hour.
-
-As already mentioned, Lurkin and Schyns \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and PDP in air cargo transport.
-The authors demonstrated that this problem is NP-hard and performed some experiments with real data, noting that their model offers better results than those obtained manually.
-
-Roesener and Barnes \cite{RoesenerBarnes2016} proposed a heuristic to solve the {\it Dynamic Airlift Loading Problem} (DALP).
-Given a set of palletized cargo items that require transport between two nodes in a time frame, the objective of this problem is to select an efficient subset of aircraft, partition the pallets into aircraft loads and assign them to allowable positions in those aircraft.
-
-Paquay {\it et al.} \cite{PaquaySchynsLimbourg2016} presented a mathematical modeling to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
-Its objective is to maximize the volume used in containers, considering load balancing constraints, the presence of fragile items and the possibility of rotating these boxes.
-Paquay {\it et al.} \cite{PaquayLimbourgSchynsOliveira2018} developed some heuristics to solve this problem.
-
-Chenguang {\it et al.} \cite{YangLiuGao2018} modeled the air transport problem as a 2D packing problem, and presented a heuristic for its optimization in several aircraft, considering load balancing to minimize fuel consumption.
-
-Wong and Ling \cite{wong2020}A developed a mathematical model and a tool based on mixed integer programming for optimizing cargo in aircraft with different pallet configurations.
-Balance constraints and the presence of dangerous items were considered.
-Wong {\it et al.} \cite{eugene2021} integrated this tool to a digital simulation model, with a visualization and validation system, based on sensors that alert about load deviations.
-
-Zhao {\it et al.} \cite{zhao2021} proposed a new modeling for WBP based on mixed integer programming.
-Instead of focusing on the CG deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
-
-As can be seen, none of these works address air cargo palletization and load balancing with route optimization in a multi-leg transportation plan for a single aircraft.
-This is the objective of our work: to model and elaborate heuristics for a real problem that simultaneously involves 4 intractable sub-problems: APP, WBP, PDP and TSP.
-
-\section{Problem context and assumptions}
-\label{sec3}
-
-In this section, we describe the context of the problem addressed in this work, as well as the assumptions considered.
-
-\subsection{Operational premises}
-
-As we are dealing with an extremely complex and diverse problem, we decided to establish some simplifying characteristics:
-
-\begin{itemize}
-
-	\item At each node of the tour, the items to be allocated are characterized by weight, volume, scores, and previously known destinations.
-We leave the consideration of 2D or 3D items to a future work.
-
-	\item We considered a unique pallet type: the {\it 463L Master Pallet}, a common size platform for bundling and moving air cargo.
-It is the primary air cargo pallet for more than 70 Air Forces and many air transport companies.
-This pallet has a capacity of $4,500 kg$ and $13.7 m^3$, which may be limited by its position along the cargo bay.
-It is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$.
-For more information, see {\tt www.463LPallet.com}.
-
-	\item All items allocated on a pallet must have the same destination.
-A pallet which has not yet reached its destination may receive more items, although it is known that these operations of removing restraining nets increase handling time and the risk of improper delivery.
-We do not consider oversized cargo in this work, but only cargo items that fit on these pallets.
-
-	\item Finally, as we are interested in minimizing fuel costs, we disregarded others costs not directly associated with aircraft flight, such as handling.
-
-\end{itemize}
-
-Throughout this text, we call {\it packed content}\/ a set of items of same destination stacked on a pallet and covered with a restraining net (see Figure \ref{fig:larger2}).
-It is considered a single item, having the same attributes of its components, whose values are the sum of individual scores, weights and volumes.
-To ensure accuracy in pickup and delivery operations, a packed content must remain on board until its destination.
-
-\subsection{Aircraft parameters and load balancing}
-
-We consider real scenarios with an single aircraft with payload of $75,000 kg$, with enough fuel for a range of $4,400km$.
-The aircraft layout is presented in Figure \ref{fig:larger}, where the pallets positions are identified by $p_i$, $1\leq i \leq 18$.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=0.4]{Images/consolidated.jpg}
-	\caption{A packed content on a 463L pallet inside a {\it Boeing C-17}}
-	\small\textsuperscript{Source: From Wikimedia Commons, the free media repository}
-	\label{fig:larger2}
-\end{figure}
-
-\begin{figure}[!h]
-	\centering
-
-	\begin{tikzpicture}[scale=1.2, samples=100]
-
-		\filldraw[fill=blue!5!white, draw=black] (0, 0) rectangle (10.23, 2.6);
-
-		\filldraw[fill=gray!20!white, draw=black] (0.20, 1.40) rectangle node{$p_{17}$} (1.43, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (0.20, 0.20) rectangle node{$p_{18}$} (1.43, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (1.63, 1.40) rectangle node{$p_{15}$} (2.86, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (1.63, 0.20) rectangle node{$p_{16}$} (2.86, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (3.06, 1.40) rectangle node{$p_{13}$} (4.29, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (3.06, 0.20) rectangle node{$p_{14}$} (4.29, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (4.49, 1.40) rectangle node{$p_{11}$} (5.72, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (4.49, 0.20) rectangle node{$p_{12}$} (5.72, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (5.92, 1.40) rectangle node{$p_{9}$} (7.15, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (5.92, 0.20) rectangle node{$p_{10}$} (7.15, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (7.35, 1.40) rectangle node{$p_{7}$} (8.58, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (7.35, 0.20) rectangle node{$p_{8}$} (8.58, 1.20);
-
-		\filldraw[fill=gray!20!white, draw=black] (8.78, 1.40) rectangle node{$p_{5}$} (10.1, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (8.78, 0.20) rectangle node{$p_{6}$} (10.1, 1.20);
-
-		\filldraw[fill=blue!5!white, draw=black] (10.23, 0) rectangle (13.27, 2.6);
-		\filldraw[fill=gray!20!white, draw=black] (10.41, 1.40) rectangle node{$p_{3}$} (11.64, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (10.41, 0.20) rectangle node{$p_{4}$} (11.64, 1.20);
-		\filldraw[fill=gray!20!white, draw=black] (11.84, 1.40) rectangle node{$p_{1}$} (13.07, 2.40);
-		\filldraw[fill=gray!20!white, draw=black] (11.84, 0.20) rectangle node{$p_{2}$} (13.07, 1.20);
-
-	\end{tikzpicture}
-
-	\begin{tikzpicture}[scale=1.2, samples=100]
-		\draw[black, thick][<-]  (1, 2.77) node[anchor=east, font=\fontsize{7}{3.5}\selectfont]{Forward} -- (9, 2.77) node[anchor=west, font=\fontsize{7}{3.5}\selectfont]{Ramp} ;
-	\end{tikzpicture}
-
-	\caption{Aircraft layout} \label{fig:larger}
-\end{figure}
-
-The torque applied to the aircraft must keep its CG in the operational range, which corresponds to a fixed percentage of the {\it Mean Aerodynamic Chord} \footnote{Chord is the distance between the leading and trailing edges of the wing, measured parallel to the normal airflow over the wing.
-The average length of the chord is known as the {\it Mean Aerodynamic Chord} (MAC).} which is considered $1.17m$ for the aircraft of this work (see Figure \ref{fig:lateral}).
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=0.22]{Images/lateral.png}
-	\caption{Aircraft longitudinal cut, where red lines are pallets positions}
-	\label{fig:lateral}
-\end{figure}
-
-Table \ref{tab:larger} shows the parameters of this aircraft.
-$p_i$\/ are pallets, $1\leq i \leq 18$, with weight $W_i$\/ and $V_i$\/ volume limits, respectively.
-$D_i^{long}$\/ and $D_i^{lat}$\/ are, respectively, the longitudinal and lateral distances of each pallet centroids to the CG of aircraft along both axes.
-These distances will be used in the calculation of the torque referring to the items allocated on each pallet.
-In this aircraft, as the ramp have an inclination of $25^{\circ}$, we made the necessary corrections in $D_i^{long}$, $W_i$\/ and $V_i$\/ of the corresponding pallets.
-
-\begin{table}[H]
-	\centering
-	\caption{Aircraft parameters}  \label{tab:larger}
-	\footnotesize
-	\begin{tabular}{c | c c c c c c c c c}
-		\toprule
-		& \multicolumn{3}{c}{$Payload$: 75,000kg} & \multicolumn{3}{c}{$limit^{CG}_{long}$: $1.170m$} &
-		\multicolumn{3}{c}{$limit^{CG}_{lat}$: $0.19m$} \\
-		\midrule
-		\multirow{2}{*}{{\boldm{$p_i$}}}  & $p_{17}$ & $p_{15}$ & $p_{13}$ & $p_{11}$ & $p_{9}$ & $p_{7}$ & $p_{5}$ & $p_{3}$ & $p_{1}$ \\
-		& $p_{18}$ & $p_{16}$ & $p_{14}$ & $p_{12}$ & $p_{10}$ & $p_{8}$ & $p_{6}$ & $p_{4}$ & $p_{2}$ \\
-		\midrule 
-		\multirow{2}{*}{\boldm{$D_i^{long}$} ($m$)} & -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\
-		& -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\			
-		\midrule 
-		\multirow{2}{*}{\boldm{$D_i^{lat}$} ($m$)}  & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 \\
-		& -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 \\	
-		\midrule
-		{\boldm{$W_i$}} ($kg$)      &   4,500   &    4,500  &   4,500   &  4,500    & 4,500     & 4,500     & 4,500     & 3,000    & 3,000   \\
-		{\boldm{$V_i$}} ($m^3$)   &   14.8   &   14.8   &  14.8    &  14.8    & 14.8     & 14.8     & 14.8     & 10.0    & 7.0 \\	
-		\midrule
-
-		\textbf{Fuel cost}  & \multicolumn{9}{c}{ $c_d$ = US\$ $4.90/km$ }	 \\
-
-		\midrule
-
-		\textbf{Fuel consumption rate}  & \multicolumn{9}{c}{ $c_g = 5\%$} \\
-
-		\midrule
-
-		\textbf{Maximum weight}  & \multicolumn{9}{c}{ $W_{max} = \sum_i W_i = 75,000kg$} \\
-
-		\bottomrule
-	\end{tabular}
-	\normalsize 
-\end{table}
-
-This aircraft spends $c_d$\/ dollars per kilometer flown, and can carry up to $W_{max}$\/ of cargo distributed on the pallets.
-The {\it fuel consumption rate}\/ $c_g$\/ is the percentage limit of cost increase  due to the CG deviation on the longitudinal axis of the aircraft.
-$c_g$\/ depends on the characteristics of the aircraft, and is an arbitrated value that should be better evaluated in future research.
-It is important to consider that the $c_g$\/ tends to zero as the aircraft attitude tends to level.
-As the CG deviation varies from $0$\/ to $limit^{CG}_{long}$, the {\it fuel cost}\/ increase varies from $0$\/ to $c_g$.
-
-\vspace{3mm}
-We also make the following assumptions:
-\begin{itemize}
-	\item on each pallet, the items are distributed in such a way that their CG coincides with the centroid of the pallet;
-	\item the CG of the total load must be at a maximum longitudinal distance $limit^{CG}_{long}$\/ from the CG of the aircraft;
-	\item the CG of the total load must be at a maximum lateral distance $limit^{CG}_{lat}$\/ from the CG of the aircraft;
-	\item the pallets are distributed in two identical rows (with odd and even indices, respectively), and the centroid of $p_i$\/ is at distance $D^{lat}_{i}$\/ from the center-line of the aircraft.
-\end{itemize}
-
-\section{The mathematical modeling}
-\label{sec4}
-
-Given the assumptions and parameters described in the previous section, we are ready to present the mathematical modeling of ACLP+RPDP.
-
-ACLP+RPDP has the objective function \ref{eq1}, and the calculus equations \ref{eq:scores} to \ref{eq:LatPacked} subject to constraints \ref{eq:torqlat} to \ref{eq21}, which will be described below.
-
-\subsection{Problem structure and input data}
-
-In the definitions below and in the tests of Section \ref{sec6}, we use the values assigned in Tables \ref{tab:costs} and \ref{tab:larger}.
-To simplify the notation throughout this text, $p_i$\/ will be called pallet $i$, and $l_k$\/ will be called node $k$.
-
-Let $L = \{ 0, 1, \ldots, K \}$ be the set with the $K+1$\/ nodes, and let $L_k$\/ be the set of remaining nodes when the aircraft is in the node $k$, $0 \leq k \leq K$. Therefore, $L_0 = L$.
-
-Let $d(a,b)$ be the distance from node $a$\/ to node $b$, where $0 \leq a, b \leq K$. By definition, $d(a,a) = 0, \forall a$.
-
-Let $C=\left[c_{a,b}\right]$\/ be the cost matrix of flights, where $c_{a,b} = c_d*d(a,b), 0 \leq a,b \leq K$.
-
-Let $M = \{1, 2, \ldots, m \}$\/ the set of $m$\/ empty pallets assigned to specific positions within the aircraft.
-In Table \ref{tab:larger}, $m=18$.
-Each pallet $i$, $1 \leq i \leq m$, has weight capacity $W_i$, volume capacity $V_i$, longitudinal distance to the CG of aircraft $D^{long}_i$, and lateral distance to the center line of aircraft $D^{lat}_i$.
-
-Let $N_k = \{1, \ldots, n_k \}$\/ be the set of $n_k$\/ items $t_j^k$\/ available for loading on the node $k$, $1 \leq j \leq n_k$, $0 \leq k \leq K$.
-For ease of notation, $t_j^k$\/ will be called item $j$\/ of the node $k$, with score $s_j$, weight $w_j$, volume $v_j$, and destination $to_j \in L_k$.
-Let $N = \bigcup_{0 \leq k \leq K} N_k$ be the set of items of all nodes along a tour.
-
-Let $Q_k = \{1, \ldots, m_k \}$\/ be the set of $m_k \leq m$\/ packed contents $a^k_q$\/ at the node $k$, $1 \leq q \leq m_k$, $0 \leq k \leq K$.
-For ease of notation, $a^k_q$\/ will be called packed content $q$\/ of the node $k$, with total weight $w_q$, total volume $v_q$, and destination $to_q \in L_k$.
-By definition, $m_0 = 0$, and therefore $Q_0 = \varnothing$.
-Packets contents that were destined to node $k$\/ are unloaded when the aircraft arrives at this node, that is, they are not considered in $Q_k$.
-
-\subsection{The decision variables}
-
-Let $X_{ij}^k$\/ and $Y_{iq}^k$\/ be binary variables, where $1 \leq i \leq m$, $1 \leq j \leq n_k$, $1 \leq q \leq m_k$\/ and $0 \leq k \leq K$.
-
-$X_{ij}^k = 1$\/ if the item $j$\/ of the node $k$\/ is assigned to the pallet $i$, and 0 otherwise.
-
-$Y_{iq}^k = 1$\/ if the packed content $q$\/ of the node $k$\/ is assigned to the pallet $i$, and 0 otherwise.
-
-\subsection{The allocation graph}
-
-Allocations of items or packed contents to the pallets in the node $k$\/ can be seen as a bipartite graph $G_k(V_k, E_k)$, where:
-
-\begin{itemize}
-
-\item $V_k = M \cup N_k \cup Q_k$;
-
-\item $E_k = E_{N_k} \cup E_{Q_k}$;
-
-\item $(i, j) \in E_{N_k}$\/ if $X_{ij}^k = 1$, where $i$\/ is a pallet and $j$\/ is a item of the node $k$;
-
-\item $(i, q) \in E_{Q_k}$\/ if $Y_{iq}^k = 1$, where $i$\/ is a pallet and $q$\/ is a packed content of the node $k$.
-
-\end{itemize}
-
-\subsection{The objective function}
-
-Let $S_K = \{s: \{1, \dots, K\} \rightarrow \{1, \dots, K\} \}$\/ be the set of $K!$ permutations $\pi$, which correspond to all possible tours (or itineraries) that have node $0$\/ as origin and end, passing through the other $K$\/ nodes.
-Let $\pi_k$\/ be $k^{th}$\/ node of the tour $\pi$, $1 \leq k \leq K$.
-In this way, the tour $\pi$\/ is described as $\{0,\pi_1,\ldots,\pi_K, 0\}$.
-For ease of notation, we can define $\pi_0 = \pi_{K+1} = 0$.
-
-The objective of ACLP+RPDP is to find the permutation $\pi \in S_K$, with the corresponding allocation of items on the pallets at each node, that maximizes the function $f_\pi = \tilde{s}_\pi/\tilde{c}_\pi$ \ref{eq1}, where $\tilde{s}_\pi$\/ is the total score of transported items, and $\tilde{c}_\pi$\/ is the total cost of fuel consumed.
-
-\begin{equation} \label{eq1}
-	\max_{\pi \in S_K} f_\pi = \tilde{s}_\pi/\tilde{c}_\pi
-\end{equation}
-
-\subsection{The calculus equations}
-
-We have the following calculation equations:
-
-\begin{itemize}
-	\item $\tilde{s}_\pi$\/ \ref{eq:scores} is the sum of the scores of the items loaded on the aircraft throughout the tour $\pi$.
-	\item $\tau_{\pi_k}$\/ \ref{eq:tau} is the longitudinal torque applied by the loaded pallets at the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.
-	\item $\tilde{c}_\pi$\/ \ref{eq:costs} is the cost of to the total fuel consumed in the tour $\pi$\/ due to the distances traveled and the CG longitudinal deviations \footnote{In our experiments, we found that the magnitude of the lateral torque was always very small, and therefore we decided to ignore it in the fuel consumption.} in each pallet.
-	\item The sets of nodes not yet visited along the tour $\pi$\/ are defined in \ref{eq:pdp11} and \ref{eq:departing}.
-	\item At the beginning of the tour $\pi$, there are no packed contents \ref{eq:pdp13}. 
-	\item $\epsilon_{\pi_k}^t$\/ \ref{eq:LatItem} and $\epsilon_{\pi_k}^a$\/ \ref{eq:LatPacked} are the lateral torque applied by the loaded pallets at the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.
-For ease of notation, $\epsilon_{\pi_k}^t$\/ corresponds to the items, and $\epsilon_{\pi_k}^a$\/ to the packed contents.
-\end{itemize}
-
-\begin{equation} \label{eq:scores}
-	\tilde{s}_\pi = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times s_j
-\end{equation}
-
-\begin{equation} \label{eq:tau}
-	\tau_{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times \Big ( \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{long}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:costs}
-	\tilde{c}_{\pi} = \sum_{k=0}^{K} \Big [ c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|) \Big ] 
-\end{equation}
-
-\begin{equation} \label{eq:pdp11}
-	L_0 = L
-\end{equation}
-
-\begin{equation} \label{eq:departing}
-	L_{\pi_k} = L_{\pi_{k-1}} - \{\pi_k\}; \ k \in \{1, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:pdp13}
-	m_0 = 0
-\end{equation}
-
-\begin{equation} \label{eq:LatItem}
-	\epsilon_{\pi_k}^t = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{j=1}^{n_{\pi_k}} \Big ( X_{ij}^{\pi_k} \times w_j \times (i\%2) - X_{ij}^{\pi_k} \times w_j \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:LatPacked}
-	\epsilon_{\pi_k}^a = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{q=1}^{m_{\pi_k}} \Big ( Y_{iq}^{\pi_k} \times w_q \times (i\%2) - Y_{iq}^{\pi_k} \times w_q \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\subsection{The constraints}
-
-Finally, we can consider the constraints at each node $\pi_k$:
-
-\begin{itemize}
-	\item The longitudinal \ref{eq:torqlong} and the lateral \ref{eq:torqlat} torques must be within the limits of the aircraft.
-	\item The items allocated to each pallet cannot exceed its weight \ref{eq:app2} and volume \ref{eq:app3} limits.
-	\item At most, each item is associated with a single pallet \ref{eq:app4};
-	\item Packed contents that have not yet reached their destination must remain on board \ref{eq:app5}.
-	\item Items allocated on the same pallet must have the same destinations.
-In this case, we need two constraints: \ref{eq18} and \ref{eq19}.
-If $X_{ia}^{\pi_k} = X_{ib}^{\pi_k} = 1$, both constraints require that $to_a = to_b$; otherwise these constraints have no effect.
-	\item If there is a packed content on the pallet, it must also have the same destination as the other items. Similarly, we use two constraints: \ref{eq20} and \ref{eq21}.
-\end{itemize}
-
-\begin{equation} \label{eq:torqlong}
-	| \tau_{\pi_k} | \leq 1;\ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:torqlat}
-	| \epsilon_{\pi_k}^t + \epsilon_{\pi_k}^a| \leq  1; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:app2}
-	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:app3}
-	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:app4}
-	\sum_{i=1}^{m} X_{ij}^{\pi_k} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:app5}
-	\sum_{i=1}^{m} Y_{iq}^{\pi_k} = 1;\ to_q \in L_{\pi_k}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq18}
-	to_a - to_b \geq -K \times (1-X_{ia}^{\pi_k} \times X_{ib}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ a,b \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq19}
-	to_a - to_b \leq K \times (1-X_{ia}^{\pi_k} \times X_{ib}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ a,b \in \{1, \ldots, n_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq20}
-	to_j - to_q \geq -K \times (1-X_{ij}^{\pi_k} \times Y_{iq}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq21}
-	to_j - to_q \leq  K \times (1-X_{ij}^{\pi_k} \times Y_{iq}^{\pi_k}); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}; \ q \in \{1, \ldots, m_{\pi_k}\}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-Constraints \ref{eq18} to \ref{eq21} are not within standard {\it Mixed-Integer Programming}\/ (MIP), and therefore need to be handled in a different way.
-In the next section, we will explain the strategy adopted to solve ACLP+RPDP.
-
-\section{Solution strategy}
-\label{sec5}
-
-Once the assumptions of this work and the mathematical modeling of the problem are presented, it is possible to see that ACLP+RPDP is {\it NP-hard}.
-In a similar way to Lurkin and Schyns \cite[p.
-6]{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$\/ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in node 0, and no items in node 1.
-Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is {\it NP-complete}.
-
-Throughout our research, we have thoughtfully described the ACLP+RPDP model in standard MIP format and found that no MIP solver can handle its practical cases in a feasible time.
-Thus, as ACLP+RPDP is highly complex, involving four intractable subproblems (APP, WBP, PDP and TSP), our strategy will be to focus only on {\it real cases}\/ and to develop quick node-by-node solutions, not necessarily optimal, that allow to build a complete tour.
-
-The ACLP+RPDP solution defines a tour and a corresponding loading and unloading plan at each node.
-In this way, the loading and unloading time will be limited only by the use of the equipment at each node.
-Therefore, there will only be waiting time at the base node, corresponding to the run time of ACLP+RPDP algorithm.
-For this reason, our method will also be parameterized by the maximum time that can be waited until the start of the load in the base node.
-
-On the other hand, we know that transport aircraft generally have a few dozen pallets, the flight plan has less than $6$\/ nodes, and each node has hundreds of items to be shipped.
-We also know that missions with fewer nodes are more frequent than longer missions.
-Under these circumstances, we can adopt two more strategies:
-
-\begin{itemize}
-
-\item We will consider that the number of destinations is greater than 1, to avoid the case of a single and trivial tour, but less than the number of pallets, that is, $1 < K < m$.
-So we can preset the destinations of the pallets at each node.
-In this way, we will reserve a number of pallets proportional to the volume demanded by each destination at the shipping node.
-We could have used another criterion, but it was observed in the experiments that volume is more constrictive in airlift.
-
-\item If we have quick node-by-node solutions, as the number of nodes is small we have the possibility to check the solution corresponding to the tour with the shortest total distance, and even test all possible tours, selecting the one that provides the best value for the objective function.
-
-\end{itemize}
-
-Our complete strategy is summarized in Algorithm \ref{alg:main}.
-
-\begin{algorithm}[H]
-	\caption{Solving ACLP+RPDP}  \label{alg:main}
-	\begin{algorithmic}[1]
-
-		\Procedure{$ACLP+RPDP$}{$scenario, surplus, time$}
-
-		\State Let aircraft and $M$  \label{main:M} \Comment{Aircraft parameters from Table \ref{tab:larger}}
-
-		\State Let $K$, $L$\/ and $C$ be according to $scenario$ \label{main:KLC} \Comment{Nodes and costs from Tables \ref{tab:costs} and \ref{tab:scenarios}}
-
-		\State Let $\pi_{TSP1}$\/ and $\pi_{TSP2}$  \Comment{The shortest tours from Tables \ref{tab:costs} and \ref{tab:scenarios}}
-
-		\State $N \gets ItemsGeneration(scenario, surplus)$ \label{main:items} \Comment{Items available for shipment}
-
-		\For {each $method$} \Comment{$method$ is a MIP solver or a heuristic}
-
-			\State $f_1 \gets SolveTour(\pi_{TSP1}, L, M, C, N, method, time)$  \label{main:f1}
-
-		    \State $f_2 \gets SolveTour(\pi_{TSP2}, L, M, C, N, method, time)$  \label{main:f2}
-
-			\State $answer1[scenario,surplus,method] \gets max( f_1, f_2  )$  \label{main:a1} \Comment{Best result between the shortest tours}
-
-			\For {each $\pi \in S_K$} \label{main:loop1} \Comment{$\pi$ is a permutation of nodes (a tour)}
-
-				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, time/K!)$ \Comment{$f_{\pi}$\/ is a solution to the tour $\pi$} \label{main:fpi}
-
-			\EndFor \label{main:loop2}
-			\State $answer2[scenario,surplus,method] \gets \max f_{\pi}$ \label{main:a2} \Comment{Best result among all permutations}
-		\EndFor
-
-		\Return $answer1, answer2$
-
-		\EndProcedure
-
-	\end{algorithmic}
-\end{algorithm}
-
-The aircraft parameters and the set $M$\/ of pallets are in Table \ref{tab:larger}.
-In this algorithm, we use five values for $scenario$, according to Tables \ref{tab:costs} and \ref{tab:scenarios}, which define the number $K$ of destinations, the set $L$\/ of nodes, the costs $C$\/, and the shortest tours $\pi_{TSP1}$\/ and $\pi_{TSP2}$.
-
-\vspace{2.0mm}
-\begin{table}[H]
-	\centering
-	\caption{Testing scenarios}  \label{tab:scenarios}
-	\begin{tabular}{c c c c c}
-		\toprule
-		{\bf Scenario} & $K$ & $L$ & $\pi_{TSP1}$ & $\pi_{TSP2}$ \\		
-		\midrule
-		1 & 2    & \{$0,1,2$\}           & 0 1 2 0  & 0 2 1 0 \\
-		2 & 3    & \{$0,1,2,3$\}         & 0 1 2 3 0 & 0 3 2 1 0\\
-		3 & 4    & \{$0,1,2,3,4$\}       & 0 4 1 2 3 0 & 0 3 2 1 4 0 \\
-		4 & 5    & \{$0,1,2,3,4,5$\}     & 0 4 1 2 5 3 0 & 0 3 5 2 1 4 0 \\
-		5 & 6    & \{$0,1,2,3,4,5,6$\}   & 0 4 1 2 6 5 3 0 & 0 3 5 6 2 1 4 0 \\
-		\bottomrule
-	\end{tabular}
-\end{table}
-
-$surplus$\/ is a value in $\{1.2, 1.5, 2.0\}$, which corresponds, at each node $k$, to the ratio between the sum of the volumes of the items  and the load capacity of the pallets ($surplus = \sum_{j=1}^{n_k} v_j$/$\sum_{i=1}^{m} V_i$).
-This parameter allows us to verify the different behavior of each $method$, according to $scenario$\/ and the quantity of items available for shipment.
-It is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).
-
-$time$\/ is a run time limit, which will be distributed among the tours (lines \ref{main:f1}, \ref{main:f2} and \ref{main:fpi}).
-$method$\/ corresponds to a MIP solver or a heuristic to the node-by-node solution $SolveTour$, which will be presented in subsection \ref{methods}.
-
-The best results corresponding to the shortest tours are stored in $answer1$\/ (line \ref{main:a1}), and those obtained by testing all $K!$\/ tours are stored in $answer2$\/ (line \ref{main:a2}).
-
-Next, we will present two subsections: in the first we explain how $SolveTour$\/ is executed, presetting the destinations of the pallets.
-In the second we will present the heuristics developed for node-by-node solutions.
-
-\subsection{SolveTour algorithm}
-\label{tour}
-
-As we commented in the previous subsection, we will adopt the strategy of presetting the destinations of each pallet throughout the tour.
-This is feasible in practical cases, where $1<K<m$.
-For this, each pallet $i$\/ also has a field $T^k_i$, $0\leq k \leq K$, which stores its next destination after being loaded at node $k$.
-For this reason, $T^k_i \in L_k$, $1 \leq i \leq m$, $0\leq k \leq K$.
-
-$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $time$\/ is the runtime limit that will be distributed among the $K+1$\/ legs of the tour.
-
-\begin{algorithm}[H]
-	\caption{Solving the tour $\pi$ with $method$}  \label{alg:tour}
-	\begin{algorithmic}[1]
-
-		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method$, $time$}
-
-		\State $\pi_0     \gets 0$ \label{tour:pi1} \Comment{Base $0$\/ is the first node}
-		\State $\pi_{K+1} \gets 0$ \label{tour:pi2} \Comment{Base $0$\/ is the last node}
-
-		\State $score \gets 0$ \label{tour:score}
-		\State $cost  \gets 0$ \label{tour:cost}
-
-		\For {$k \gets 0$ to $K$} \label{tour:loop1}	
-		\State $L_{\pi_k} \gets L - \{\pi_0,\pi_1,\ldots,\pi_k\}$  \label{tour:lk1}	\Comment{$\pi_k$ is the current node}	
-		\State $T_i^{\pi_k} \gets -1$, $1 \leq i \leq m$ \label{tour:-1} \Comment{Unset the pallet destinations}
-
-		\If {$k = 0$}
-		\State Let $G_1(M \cup N_0, \varnothing)$ \label{tour:g11}
-		\Else
-
-		\State $E_{Q_{\pi_k}}, M \gets UpdatePacked(M, Q_{\pi_k}, \pi_k)$ \label{tour:dest}			
-		\State Let $G_1(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$ \label{tour:g12}
-		\EndIf  \label{tour:lk2}
-
-		\State $M \gets SetPalletsDestinations(M, \pi_k )$ \label{tour:dest2}
-
-		\State $G_2 \gets SolveNode(method, \pi_k, G_1, time/(K+1))$ \Comment{Each tour has $K+1$\/ legs} \label{tour:node}
-
-		\State $s, \tau \gets ScoreAndDeviation(\pi_k, G_2)$ \label{tour:analyse}
-
-		\State $score \gets score + s$ \label{tour:score2}
-		\State $cost  \gets cost  + c_{\pi_k,\pi_{k+1}} \times (1 + c_g \times |\tau|)$ \label{tour:cost2}
-
-		\EndFor  \label{tour:loop2}
-
-		\Return $score / cost$ \label{tour:f}
-
-		\EndProcedure
-
-	\end{algorithmic}
-\end{algorithm}
-
-As we mentioned in the previous section, all tours start and end at the base $0$\/ (lines \ref{tour:pi1}-\ref{tour:pi2}).
-After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ flights (lines \ref{tour:loop1}-\ref{tour:loop2}).
-Initially, the set $L_{\pi_k}$\/ of remaining nodes is updated (line \ref{tour:lk1}), and the pallet destinations are unset (line \ref{tour:-1}).
-
-When the aircraft is at the base, the initial graph $G_1$\/ is empty and there is no packed contents \ref{tour:g11}.
-Otherwise, $UpdatePacked$\/ (line \ref{tour:dest}) returns the set of packed contents that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation.
-This allocation is stored in graph $G_1$\/ (line \ref{tour:g12}).
-
-$SetPalletsDestinations$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node, without changing the pallets destination with packed contents.
-
-Finally, $SolveNode$\/ includes the edges corresponding to the items shipped at the current node, returning the graph $G_2$\/ (line \ref{tour:node}).
-The score and the CG deviation of $G_2$\/ are calculated (line \ref{tour:analyse}) and accumulated (lines \ref{tour:score2}-\ref{tour:cost2}), allowing the final result of this tour (line \ref{tour:f}).
-
-$UpdatePacked$, described in Algorithm \ref{alg:cons}, finds the best packed-pallet allocation, in terms of CG deviation, for the packed contents that remain on board.
-
-\begin{algorithm}[H]
-	\caption{Updating the packed contents that remain boarded at the node $\pi_k$}  \label{alg:cons}
-	\begin{algorithmic}[1]
-
-		\Procedure{$UpdatePacked$}{${M, Q_{\pi_k}, \pi_k}$}
-
-		\State $E_{Q_{\pi_k}} \gets MinCGDeviation(E_{Q_{\pi_k}})$ \Comment{Using a MIP solver} \label{cons:minCG}
-
-		\For{$i \gets 1$ to $m$} \label{cons:Ybegin}
-
-		\For{$q \gets 1$ to $m_{\pi_k}$}
-
-		\State $T_i^{\pi_k} \gets -1$
-
-		\If{$(i, q) \in E_{Q_{\pi_k}}$} 
-		\State $T_i^{\pi_k} \gets to_q$   \Comment{Set the pallet destinations}
-		\EndIf
-		\EndFor		
-		\EndFor \label{cons:Yend}
-
-		\Return $E_{Q_{\pi_k}}, M$
-
-		\EndProcedure
-
-	\end{algorithmic}
-\end{algorithm}
-
-$MinCGDeviation$\/ (line \ref{cons:minCG}) relocates the packed contents on the pallets, minimizing torque and ensuring that they all remain on board, one packed content on each pallet.
-It is run through a MIP solver with objective function \ref{eq:torque} and constraints \ref{eq:embarked} and \ref{eq:one}.
-As there are few variables, $E_{Q_{\pi_k}}$\/ is obtained in less than $30$\/ milliseconds.
-Finally, the destination of each pallet with a packed content is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).
-
-\begin{equation} \label{eq:torque}
-	\min f =  \Big | \sum_{i=1}^{m} \sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \times w_q \times D_i^{long}  \Big |
-\end{equation}
-
-\begin{equation} \label{eq:embarked}
-	\sum_{i=1}^{m} Y^k_{iq} = 1;\ q \in \{1,\ldots,m_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:one}
-	\sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \leq 1;\ i \in \{1,\ldots,m\}
-\end{equation}
-
-$SetPalletsDestinations$, which sets the pallets destination not yet defined, is described in Algorithm \ref{alg:dest}.
-
-\begin{algorithm}[H]
-	\caption{Setting pallets destination based on the items to be embarked at the node $\pi_k$}  \label{alg:dest}
-	\begin{algorithmic}[1]
-
-		\Procedure{$SetPalletsDestinations$}{$M, \pi_k$}
-
-			\State $vol_x \gets  0$, $x \in L_{\pi_k}$  \label{dest:vector1}
-
-			\State $max \gets 0$
-
-			\State $total \gets 0$
-
-			\For{$j \gets 1$ to $n_{\pi_k}$}
-
-				\If {$to_j \in L_{\pi_k}$}
-
-					\State $vol_{to_j} \gets vol_{to_j} + v_j$ 
-					\State $total \gets total + v_j$
-
-					\If {$vol_{to_j} > vol_{max}$}
-
-						\State $max \gets to_j$ \label{dest:max} \Comment{$max$\/ is the destination with maximum volume demand}
-					\EndIf
-
-				\EndIf
-
-			\EndFor
-
-			\For{$x \in L_{\pi_k}$} \label{dest:propor1}
-
-				\If {$vol_{x} \neq 0$}
-
-					\State $needed \gets \max \{1, \lfloor{ (m-m_{\pi_k}) \times vol_{x}/total}\rfloor \}$  \Comment{Number of pallets to node $x$}
-
-					\State $np \gets 0$
-
-					\For{$i \gets 1$ to $m$}
-
-						\If{($np < needed$) {\bf and} ($T_i^{\pi_k} = -1$)}
-							\State $T_i^{\pi_k} \gets x$
-							\State $np \gets np + 1$
-						\EndIf
-
-					\EndFor
-				\EndIf
-
-			\EndFor \label{dest:propor2}
-
-			\For{$i \gets 1$ to $m$} \label{dest:max1}
-				\If{$T_i^{\pi_k} \gets -1$}  \Comment{Remaining pallets}
-					\State $T_i^{\pi_k} \gets max$
-				\EndIf
-			\EndFor \label{dest:max2}
-
-			\Return $M$
-
-		\EndProcedure
-
-	\end{algorithmic}
-\end{algorithm}
-
-$vol$\/ stores the demand volume of items destined to the non-visited nodes (line \ref{dest:vector1}).
-The destination of empty pallets is defined proportionally to the volume of items to be embarked (lines \ref{dest:propor1}-\ref{dest:propor2}).
-The destination with the maximum volume defines any remaining pallets (lines \ref{dest:max1}-\ref{dest:max2}).
-
-$ScoreAndDeviation$, described in Algorithm \ref{alg:eval}, evaluates the allocation graph $G$\/ generated by $SolveNode$\/ at node $\pi_k$\/ and returns the corresponding score and CG deviation.
-
-\begin{algorithm}[H]
-	\caption{Calculating the score and the relative torque deviation of the graph $G$\/ at the node $\pi_k$}  \label{alg:eval}
-	\begin{algorithmic}[1]
-
-		\Procedure{$ScoreAndDeviation$}{$\pi_k, G$}
-
-		\State Let $G(V_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$
-
-		\State $s \gets 0$
-		\State $\tau_i \gets 0$, $1 \leq i \leq m$
-		\For{$i \gets 1$ to $m$}	\label{eval:loop1}
-
-			\For{$j \gets 1$ to $n_{\pi_k}$}
-
-				\If{$X_{ij}^{\pi_k} = 1$}
-					\State $s \gets s + s_j$ \label{eval:score1}
-					\State $\tau_i \gets \tau_i + w_j \times D_i^{long}$ \label{eval:eps1}
-				\EndIf
-			\EndFor
-
-			\For{$q \gets 1$ to $m_{\pi_k}$}
-				\If{$Y_{iq}^{\pi_k} = 1$}
-					\State $s \gets s + s_q$   \label{eval:score2}
-					\State $\tau_i \gets \tau_i + w_q \times D_i^{long}$  \label{eval:eps2}
-				\EndIf
-			\EndFor	   \label{eval:loop2}
-		\EndFor
-
-		\State $\tau \gets  \sum_{i=1}^{m} \tau_i / (W_{max} \times limit^{CG}_{long})$ \label{eval:torque}
-
-		\Return $s,\ \tau$
-		\EndProcedure
-
-	\end{algorithmic}
-\end{algorithm}
-
-This algorithm consists of a loop that goes through all the pallets (lines \ref{eval:loop1}-\ref{eval:loop2}), accumulating the scores (lines \ref{eval:score1} and \ref{eval:score2}) and the torques (lines \ref{eval:eps1} and \ref{eval:eps2}) of the shipped items, allowing the final calculation of the CG deviation (line \ref{eval:torque}).
-
-\subsection{Node-by-node solutions}
-\label{methods}
-
-In this subsection we present two implementations of $SolveNode$\/ algorithm: with a MIP solver and a heuristic.
-
-\subsubsection{With a MIP solver}
-\label{solver}
-
-Our strategy adopted in $SolveTour$\/ previously defines the values of some variables: the set of nodes to be visited is updated, the packed contents that remain on board are reallocated to minimize the CG deviation, and the pallets destinations are determined according to the volume of items available for shipment.
-
-In this way, the mathematical modeling for $SolveNode(MIP,\pi_k,G, time)$\/ becomes simpler, which finds an allocation of available items at the node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$.
-Thus, we use a MIP solver with runtime limit $time$\/ at the node $\pi_k$\/ to maximize the objective function \ref{eq:obj2}, with the calculus equations \ref{eq:newf} to \ref{eq:costs2}, subject to the constraints \ref{eq:tau2} to \ref{eq:ifB}.
-The binary variables $X_{ij}$\/ and $Y_{iq}$\/ define the sets of edges $E_{N_{\pi_k}}$\/ and $E_{Q_{\pi_k}}$, respectively, included in the graph $G$.
-
-\begin{equation} \label{eq:obj2}
-	\max f= \tilde{s} / \tilde{c}
-\end{equation}
-
-\begin{equation} \label{eq:newf}
-	\tilde{s} = \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij} \times s_j
-\end{equation}
-
-\begin{equation} 
-	\tau_{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q) \Big ] \Big / W_{max} \times limit^{CG}_{long}
-\end{equation}
-
-\begin{equation} \label{eq:costs2}
-	\tilde{c} =  c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|)
-\end{equation}
-
-\begin{equation} \label{eq:tau2}
-	|\tau_{\pi_k}| \leq 1
-\end{equation}
-
-\begin{equation} 
-	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}
-\end{equation}
-
-\begin{equation} 
-	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}
-\end{equation}
-
-\begin{equation} 
-	\sum_{i=1}^{m} X_{ij} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:if2}
-	X_{ij} = 0;\ to_j \notin L_{\pi_k}; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:ifA}
-	X_{ij} \leq X_{ij} \times (T_i^{\pi_k} - to_j + 1); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:ifB}
-	X_{ij} \leq X_{ij} \times (to_j - T_i^{\pi_k} + 1 ); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-The constraints \ref{eq:ifA} and \ref{eq:ifB} are equivalent to $X_{ij} =1$\/ if $to_j = T_i^{\pi_k}$, and $X_{ij} =0$\/ otherwise .
-
-\subsubsection{With Shims heuristic}
-
-One of the main objectives of this work was to find a quick heuristic that offers a good-quality solution for the node-by-node problem.
-Taking this into account, we design algorithms based on known meta-heuristics: {\it Ant Colony Optimization} (ACO) \cite{Dorigo1992, DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \cite{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \cite{Glover1986} and {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \cite{FeoResende1989}.
-We considered several ideas from the literature \cite{NiarFreville1997,Fidanova2006,Laabadi2018,Alonso2019,Zhan2020}, and we were careful to use the same data structures and procedures in all implementations.
-
-However, the heuristic that presented better solutions was none of the previous ones. In this subsection, we will present a new heuristic for the node-by-node problem, called $Shims$.
-
-Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}).
-This strategy is based on a practical observation: usually, subsets of smaller and lighter items are saved for later adjustments to the remaining available space.
-
-The selection of edges for $E_{ N_{\pi_k} }$ uses the {\it edge attractiveness}\/ $\theta_{ij}$\/, Equation \ref{eq:edge}, which can be understood as the tendency to allocate the item $j$\/ to the pallet $i$\/ at the node $\pi_k$.
-It is directly proportional to the score, and inversely to the volume and the torque of each item.
-
-\begin{equation} \label{eq:edge}
-	\theta_{ij} = \frac{ s_j }{ v_j }\ \times \Big ( 1 - \frac{ w_j \times |D_i^{long}| }{ \max w_j \times\  \max | D_i^{long} | } \Big );\ i \in \{1,\ldots,m\},\ j \in \{1,\ldots,n_{\pi_k}\}
-\end{equation}
-
-\begin{table}[H]
-
-	\begin{minipage}{0.08\linewidth}
-
-	\end{minipage}\hfill % these two lines must be close to each other
-	\begin{minipage}{0.34\linewidth}
-
-		\includegraphics[scale=0.25]{Images/shims.png}
-		\captionof{figure}{{\it Shims} of various thicknesses}
-		\small\textsuperscript{Source: www.mscdirect.com/product/details/70475967}
-		\label{fig:shims}
-
-	\end{minipage}\hfill % these two lines must be close to each other
-	\begin{minipage}{0.58\linewidth}
-		\scriptsize
-		\begin{tikzpicture}[scale=0.75, samples=100]
-			\filldraw[fill=blue!3!white, draw=black] (0, 0) rectangle (12.5, 5.5);
-			\draw[->] (.2, .6) -- coordinate (x axis mid) (12, .6);
-			\node at (5, 0.3) {$\eta_1$};
-			\node at (5, 2.5) {|};
-			\node at (6, 2.5) {\small{shims}};
-			\node at (7, 2.5) {|};
-			\node at (7, 0.3) {$\eta_2$};
-			\node at (0.3, 5) {$\theta_{ij}$};
-			\draw[->] (.6, .2) -- coordinate (y axis mid) (0.6, 5.3);
-			\node at (12, 0.3) {$e_{ij}$};
-			\draw[smooth, domain = 0.09:2, color=black, thick] plot (.3+1/\x,{4.2+log2(\x)});
-			\draw[->, dashed] (5, 0.6)--(5, 2.0);
-			\draw[->, dashed] (7, 0.6)--(7, 1.5);
-		\end{tikzpicture}
-
-		\captionof{figure}{$n_{\pi_k}$\/ possible edges $e_{ij}$\/ sorted by $\theta_{ij}$\/ in non-ascending order}
-		\label{fig:whip}		
-	\end{minipage}
-\end{table}
-
-Considering only the items that can be shipped at the node $\pi_k$, Figure \ref{fig:whip} represents $n_{\pi_k}$\/ possible edges $e_{ij}$\/ of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order.
-Initially, $Shims$\/ builds a greedy solution for the pallet $i$\/ selecting edges up to index $\eta_1$ (first phase).
-Then, with the edges between $\eta_1$\/ and $\eta_2$, it elaborates different possible complements (second phase), including later the best ones in the same pallet (third phase).
-$Shims$\/ is described in Algorithm \ref{alg:shims}.
-
 \begin{algorithm}[H]
@@
 
+		\State $\tau_{\pi_k} \gets 0$	 \Comment{Torque at the node $\pi_k$}	
 		\State $\tau_{max} \gets W_{max} \times limit^{CG}_{long}$ \Comment{Maximum aircraft torque}
+
+		\State $vol_i \gets 0$, $1\leq i \leq m$
 
@@
 
-			\State $\tau_{\pi_k} \gets \sum_{(i,q)\in E_{Q_{\pi_k}}} w_q \times D_i^{long}$ \Comment{Torque due to the packed contents}	
-			\State $vol_i \gets \sum_{(i,q)\in E_{Q_{\pi_k}}} v_q$ \Comment{Volume at the pallet $i$\/ due to the packed contents}
+			\For{$q \gets 1$ to $m_{\pi_k}$}
 
-			\State Let $E$\/ be an array of $n_{\pi_k}$ possibles edges of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order \label{shims:possible}
+				\If{$(i, q) \in E_{Q_{\pi_k}}$} 
+					\State $vol_i \gets vol_i + v_q$ 
+					\State $\tau_{\pi_k} \gets \tau_{\pi_k} + w_q \times D_i^{long}$
+				\EndIf	
+			\EndFor	
+			\State Let $E$\/ be an array of $n_{\pi_k}$ possibles edges of the pallet $i$\/ sorted by $\theta_{ij}^{\pi_k}$ in non-ascending order \label{shims:possible}
 			\State $\eta_1 \gets 1$  
 			\Repeat
-				\State $e_{ij} \gets E_{\eta_1}$ \Comment{$e_{ij}$\/ is the possible allocation of the item $j$\/ to the pallet $i$}
+				\State $e_{ij} \gets E_{\eta_1}$ \Comment{$e_{ij}$\/ is the allocation of the item $j$\/ to the pallet $i$}
 
@@
 
-			\State $vol \gets 0$, $b \gets 1$, $shims_b \gets \varnothing$, $Set \gets \{ shims_b \}$  \label{shims:beginthird}
+			\State $vol \gets 0$, $b \gets 1$, $shims_b \gets \varnothing$, {\color{blue} $Set \gets \{ shims_b \}$ } \label{shims:beginthird}
 
@@
 
-For each pallet $i$, its $n_{\pi_k}$\/ possible edges $e_{ij}$\/ are considered in non-increasing order of $\theta_{ij}$:
+For each pallet $i$, its $n_{\pi_k}$\/ possible edges $e_{ij}^{\pi_k}$\/ are considered in non-increasing order of $\theta_{ij}^{\pi_k}$:
+
+{\color{blue}
 
@@
 
-	\item In {\it greedy phase}\/ (lines \ref{shims:pallets}-\ref{shims:endfirst}), a partial solution for each pallet $i$\/ is constructed by edges inclusion, until its volume $V_i \times limit$\/ is reached.
+	\item In the {\it greedy phase}\/ (lines \ref{shims:pallets}-\ref{shims:endfirst}), a partial solution for each pallet $i$\/ is constructed by edges inclusion, until its volume $V_i \times limit$\/ is reached.
 The value $limit=0.92$\/ (line \ref{shims:limit}) was empirically determined by the {\it iRace} tool \cite{LopezIbanezManuel2016}.
+
 	From lines \ref{shims:limit} to \ref{shims:endfirst}, as items are inserted, the value of $\eta_1$ is incremented and stops when the pallet volume reaches 92\% of its capacity.
+
 	When it stops, this is the initial value of $\eta_2$, line \ref{shims:eta2a}.
@@
 	This factor of was obtained empirically, through many test with different problem instances.
-Higher values can improve the solution, but at the expense of performance.
-So we tried to keep it as low as possible, a trade-off between quality and performance.
+	Higher values can improve the solution, but at the expense of performance. So we tried to keep it as low as possible, a trade-off between quality and performance.
 
-	\item In {\it composition phase}\/ (lines \ref{shims:beginsecond}-\ref{shims:endsecond}), a set of shims named $Set$\/ is created for each pallet $i$, where each shim is formed by a set of edges in the range $[\eta_1,\eta_2]$, whose total volume is limited by $slack_i$.
+	\item In the {\it composition phase}\/ (lines \ref{shims:beginsecond}-\ref{shims:endsecond}), a set of shims named $Set$\/ is created for each pallet $i$, where each shim is formed by a set of edges in the range $[\eta_1,\eta_2]$, whose total volume is limited by $slack_i$.
+
 	After the $\eta_2$ index, there may be items with little probability of being in the optimal solution because their low value of $\theta^{\pi_k}_{ij}$, being not considered for inclusion.
+
 	In line \ref{shims:beginthird} the initial $Shim$ and $Set$ are created.
+
 	In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it First-Fit Decreasing}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \cite{JohnsonGarey1985}.
@@
 
-	\item In {\it selection phase}\/ (lines \ref{shims:beginthird}-\ref{shims:endthird}), the best shim in $Set$\/ is chosen.
+	\item In the {\it selection phase}\/ (lines \ref{shims:beginthird}-\ref{shims:endthird}), the best shim in $Set$\/ is chosen.
 Initially, two shims are found: $sh_w$\/ with larger weight, and $sh_v$\/ with larger volume.
@@
 
-\section{Implementation and results}
-\label{sec6}
-
-This section is composed of two parts: the generation of test instances and the results obtained in our implementation.
-
-\subsection{Instances generation}
-\label{items}
-
-As we are dealing with a new problem, which until now had not been modeled in the literature, we have to create our own benchmarks.
-For this, we based on the characteristics of real airlifts carried out by the {\it Brazilian Air Force}, as described below.
-
-In the military airlift carried out in Brazil from 2008 to 2010, 23\% of the items weighed between $10kg$ and $20kg$, 22\% from $21kg$ to $40kg$, 24\% from $41kg$ to $80kg$, 23\% from $81kg$ to $200kg$, and 8\% between $201kg$ and $340kg$.
-These five groups of items are described in Table \ref{tab:weights}, where $P$\/ represents the group probability.
-On the other hand, the average density of these items is approximately $246 kg/m^3$.
-
-\begin{table}[H]
-	\centering
-	\caption{Items weight distribution ($kg$) in Brazil}  \label{tab:weights}
-	\begin{tabular}{c c c c }
-		\toprule
-		$Group$ & $P$ & $low$ & $high$  \\		
-		\midrule
-		1              & 0.23           & 10  & 20  \\
-		2              & 0.22           & 21  & 40  \\
-		3              & 0.24           & 41  & 80  \\		
-		4              & 0.23           & 81  & 200 \\
-		5              & 0.08           & 201 & 340 \\
-		\bottomrule
-	\end{tabular}
-\end{table}
-
-In the generation of test instances, we use three types of random selections:
-\begin{itemize}
-	\item $RandomReal(r_1,r_2)$: randomly selects a real number in $[r_1,r_2]$, where $r_1$ and $r_2$ are real numbers;
-	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$, where $i_1$ and $i_2$ are integer numbers;
-	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
-\end{itemize}
-
-\begin{algorithm}[H]
-	\caption{Generating items}  \label{alg:itemsgen}
-	\begin{algorithmic}[1]
-
-		\Procedure{$ItemsGeneration$}{$scenario, surplus$}
-		\State Let $L$\/ and $M$  \Comment{From Tables \ref{tab:larger} and \ref{tab:scenarios}} \label{ig:LM}
-		\State $limit \gets surplus \times \sum_{i=1}^{m} V_i$ \label{ig:extended}
-
-		\For{$k \gets 0$ to $K$}
-
-			\State $N_k \gets \varnothing$
-			\State $j \gets 0$
-			\State $vol \gets 0$	\label{ig:totals}	
-			\While{$vol < limit$}
-
-				\State $j \gets j+1$
-				\State Let $t_j^k$\/ be the item $j$\/ at the node $k$
-				\Repeat
-					\State $to_j \gets RandomInt(0, K)$ \label{ig:dest}
-				\Until{$to_j \neq k$}
-
-				\State $x = Roulette(item)$ biased through $P$ \label{ig:weight1} \Comment{From Table \ref{tab:weights}}
-
-				\State $w_j \gets RandomReal(low(x), high(x))$        \label{ig:weight2}
-
-				\State $s_j \gets \round{100 \times (1 - \log_{10}(RandomInt(1, 9)))} $ \label{ig:score}
-
-				\State $v_j \gets w_j / RandomReal(148, 344)$ \label{ig:volume}
-
-				\State $vol \gets vol + v_j$
-
-				\State $N_k \gets N_k \cup \{t_j^k\}$ 
-			\EndWhile
-
-		\EndFor
-
-		\State $N \gets \bigcup_{0 \leq k \leq K} N_k$
-
-		\Return $N$
-
-		\EndProcedure
-	\end{algorithmic}
-\end{algorithm}
-
-$ItemsGeneration$, which generates $N$, is described in Algorithm \ref{alg:itemsgen}.
-The parameter $scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and the parameter $surplus$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}).
-To avoid simply loading all items, we use $surplus \in \{1.2, 1.5, 2.0\}$.
-This also represents more instances for tests in each scenario.
-
-For each generated $t^k_j$\/ item, its destination is randomly selected (line \ref{ig:dest}), its weight has a distribution according to Table \ref{tab:weights} (lines \ref{ig:weight1}-\ref{ig:weight2}), its score varies $100$\/ (highest) and $5$\/ (lowest) according to a logarithmic scale (line \ref{ig:score}, and its volume is randomly defined from the density, where we allow a variation of 40\% more or less than the average density of $246 kg/m^3$\/ (line \ref{ig:volume}).
-
-{\color{blue}
-
-\subsection{Results obtained}
-
-In the tests performed, we used a 64-bit, 16GB, 3.6GHz, eight-core processor with {\it Linux Ubuntu 22.04.1 LTS 64-bit} as the operational system and {\it Python 3.10.4}\/ as the programming language.
-We also used the well-known solver {\it Gurobi}\/ (www.gurobi.com), version 9.5.2.
-
-We ran Algorithm \ref{alg:main} in the 5 scenarios described in Table \ref{tab:scenarios}, considering 6 methods for node-by-node solution: Gurobi (see \ref{solver}), ACO, NMO, TS, GRASP, and {\it Shims} (Algorithm \ref{alg:shims}).
-In generating the items in each node, we consider 3 values for the parameter $surplus$.
-3 values for the time limit (1200s, 2400s, and 3600s).
-
-The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:20} ($surplus = 1.2$), \ref{tab:50} ($surplus = 1.5$), \ref{tab:100} ($surplus = 2.0$), and in Table \ref{tab:100time} (different time limit comparisons).
-
-For each $method$, $scenario$\/ and $surplus$, 7 different instances were generated.
-Therefore, 315 tests (5 scenarios $\times$\/ 3 volumes $\times$\/ 3 time limits $\times$\/ 7 instances) were performed for each method.
-
-The average values were presented for $f_{\pi}$\/ and, for the {\bf run time (s)}, the average result obtained.
-To facilitate the comparison between the methods, we added a last column in these tables, where two values are indicated:
-\begin{itemize}
-	\item {\bf Normalized}: value between 0 and 1, which corresponds to the ratio between the sum of $f$\/ values obtained by the method in all scenarios and the sum of the best values obtained among all methods in all scenarios.
-The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
-	\item {\bf Speed-up}: ratio of the sums of the run times of all scenarios and the sum of the method run times in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
-\end{itemize}
-
-	As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's Read-Access Memory (RAM, 16 GiB) too early, we decided to set the Gurobi {\it MIPgap} parameter to 1\% to control the minimal quality of the returned solutions and abbreviate the solution time.
-	From now on, we will refer to this Gurobi-specific model as {\it Gurobi* }.
-	For more details on the Gurobi {\it MIPgap}, refer to \cite{mipgap}.
-
-	This decision hinders the MIP solver from reaching optimal solutions in favor of the possibility of result comparisons with the {\it Shims} heuristic.
-
-	The parameter $\pi_{TSP_1}$ represents the least cost tour (through an optimal TSP solution), and $\pi_{TSP_2}$, the opposite direction of travel, beginning and ending in the base.
-The values of $f$ are the best in these to solutions.
-
-\begin{table}[H]
-	\centering
-	\caption{Solutions with volume {\it surplus} = {\bf 1.2} and 3600s as time limit.}  \label{tab:20}
-	\footnotesize
-	\begin{tabular}{ccccccccc}
-		\toprule
-		{\bf Tours}&$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
-		\toprule
-		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} &\multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\ % 5927
-		&                                                               &               {\bf run time (s)}  & 17      & 17      & 16      & 17    & 26        &  1.0 \\ % 93
-		\cmidrule(lr){2-9}	
-		&                                                               \multirow{2}{*}{\it Shims}    & $f$ & 8.54    & 11.63   & 12.97   & 13.43 & 12.36     &  0.99 \\ % 5893
-		&                                                               &         {\bf run time (s)}        & 1       & 1       & 2       & 2     & 2         &  11.6 \\ % 8
-
-		\midrule
-
-		\multirow{4}{*}{\it $S_{K}$ }&                                  \multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 12.25   & 13.32   & 14.61  & x        & 1.00 \\ % 4869
-		&                                                               &               {\bf run time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\ % 479
-		\cmidrule(lr){2-9}	
-		&                                                               \multirow{2}{*}{\it Shims}    & $f$ & 8.54    & 12.26   & 13.23   & 14.53  & 13.42    & 0.99  \\ % 4813
-		&                                                               &         {\bf run time (s)}        & 1       & 2       & 6       & 17     & 106      & 18.4   \\ % 132
-		\bottomrule
-		\multicolumn{9}{c}{*{\it MIPgap} parameter set to 1\%. The $f$ values are \textpm 1\% distant from the optimal solution}\\
-
-	\end{tabular}
-	\normalsize
-\end{table}
-
-{\color{blue}
-
-	97\% of 16GiB was the RAM state before we aborted the {\it Gurobi*} execution on solving scenario 5 in Table \ref{tab:20}.
-
 }
 
-\begin{table}[H]
-	\centering
-	\caption{Solutions with volume {\it surplus} = {\bf 1.5} and 3600s as time limit.}  \label{tab:50}
-	\footnotesize
-	\begin{tabular}{ccccccccc}
-		\toprule
-		{\bf Tours} & $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
-		\toprule
-		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\ % 8236
-		&&               {\bf run time (s)}                                                                  & 38    & 30    & 27    & 29    & 81    & 1.00  \\ % 205
-		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{\it Shims}                                                                    & $f$ & 11.83 & 16.74 & 18.01 & 18.89 & 16.91 & 1.00  \\ % 
-		&&         {\bf run time (s)}                                                                        & 1     & 2     & 2     & 3     & 3     & 20.5  \\ %
-
-		\midrule
-
-		\multirow{4}{*}{\it $S_{K}$ } & \multirow{2}{*}{\it Gurobi* }                                  & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ % 8602
-		&&               {\bf run time (s)}                                                                  & 53    & 61    & 193   & 470   & 2241    & 1.0  \\ % 3018
-		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{\it Shims}                                                                    & $f$ & 11.83 & 17.00 & 18.21 & 20.45 & 18.00   & 0.99  \\ % 8549
-		&&         {\bf run time (s)}                                                                        & 1     & 2     & 9     & 26    & 157     & 15.5 \\ % 195
-		\bottomrule
-
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\begin{table}[H]
-	\centering
-	\caption{Solutions with volume {\it surplus} = {\bf 2.0} and 3600s as time limit.}  \label{tab:100}
-	\footnotesize
-	\begin{tabular}{ccccccccc}
-		\toprule
-		{\bf Tours} & $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
-		\toprule
-		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 24.39 & 26.46 & 27.09 & 24.00 & 1.00 \\ % 11969
-		&&                                                                               {\bf run time (s)}  & 162   & 95    & 74    & 68    & 67    & 1.0  \\ % 466
-		\cmidrule(lr){2-9}	
-		&                                                                   \multirow{2}{*}{\it Shims} & $f$ & 17.78 & 24.40 & 26.36 & 27.01 & 23.94 & 0.99  \\ % 11949
-		&&                                                                                {\bf run time (s)} & 2     & 3     & 3     & 4     & 5     & 27.4  \\ % 17
-
-		\midrule
-
-		\multirow{4}{*}{\it $S_{K}$ } & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &  x   & 1.00 \\ % 9856
-		&&                                                                                {\bf run time (s)} & 151   & 123   & 319   & 689   &  x   & 1.0  \\ % 1282
-		\cmidrule(lr){2-9}	
-		&                                                                \multirow{2}{*}{\it Shims}    & $f$ & 17.78 & 25.25 & 26.38 & 28.98 & 26.08 & 1.00  \\ % 9839
-		&&                                                                                {\bf run time (s)} & 2     & 3     & 13    & 37    & 229   & 23.3  \\ % 284
-		\bottomrule
-
-	\end{tabular}
-	\normalsize
-\end{table}
-
-{\color{blue}
-	{\it Gurobi*} could not solve scenario 5 in Table \ref{tab:100} because $3600/K! = 0.71s$ is not enough time for {\it Gurobi*} to find a feasible node solution.
-}
-
-\subsubsection{Time limits comparisons}
-
-\begin{table}[H]
-	\centering
-	\caption{Solutions with different time limitations for $S_K$ tours.}  \label{tab:100time}
-	\footnotesize
-	\begin{tabular}{ccc|cc|cc|cc}
-		\toprule
-		&             \multicolumn{2}{r}{\bf Volume $surplus$} &\multicolumn{2}{c}{$1.2$}&\multicolumn{2}{c}{$1.5$}&\multicolumn{2}{c}{$2.0$} \\
-		\midrule
-		$method$& {\bf \specialcell{Time\\limit (s)}} & {\bf Scenarios}    & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} \\
-		\toprule
-
-		\multirow{6}{*}{\it Gurobi* }& \multirow{2}{*}{1200} & $f$ & 14.67   & x       &  20.64  &  x      & 27.97   & x \\ % 
-		&        &  {\bf run time (s)}                             & 217     & x       &  304    &  x      & 579     & x \\ % 
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$                               & 14.59   &         &  20.60  &  x      & 29.10   & x \\ % 
-		&        &  {\bf run time (s)}                             & 313     &         &  327    &  x      & 543     & x \\ % 
-		\cmidrule(lr){2-9}                              		
-		& \multirow{2}{*}{3600}& $f$                               & 14.61   & x       &  20.59  &         & 29.11   & x \\ % 
-		&        &  {\bf run time (s)}                             & 297     & x       &  470    &         & 689     & x \\ % 		
-		\midrule
-		\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
-		&        & {\bf run time (s)}                              & 18      & 109     &  26     & 160     & 37      & 203   \\ % 
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$                               & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
-		&        &  {\bf run time (s)}                             & 18      & 107     &  25     & 159     & 37      & 192   \\ % 	
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{3600} & $f$                              & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ %  
-		&        &  {\bf run time (s)}                             & 17      & 106     &  26     & 157     & 37      & 181   \\ %  
-		\bottomrule
-
-		&                                       & {\bf Normalized} & 0.99    & x       & 0.99    & x       & 1.00    & x\\
-		&                                       & {\bf Speed-up}   & 15.6    & x       & 14.3    & x       & 16.3    & x\\
-
-		\bottomrule
-
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\subsubsection{Proposed K! strategy assessment}
-
-	Another important test is to compare the best tour score/cost ratio with the least expensive tour score/cost ratio.
-This is important to see how far our solution strategy is from the simplest form, just solving a TSP.
-
-	For this achievement, we ran {\it Shims} in 5 scenarios with 7 instances each, and 3 volume surpluses, collecting the averages of the best tour result and the least cost tour result.
-
-	\begin{table}[H]
-		\centering
-		\caption{{\it Shims} best and least cost tour results}  \label{tab:20least}
-		\footnotesize
-		\begin{tabular}{ccccccccc}
-			\toprule
-			{$surplus$}              & {\bf Tour} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4}  & {\bf 5}   & {\bf Sum} & {\bf Improvement}\\
-			\toprule
-
-			\multirow{2}{*}{$1.2$}   & Least cost    & 8.54    & 11.63   & 12.97   & 13.43  & 12.36 & 58.93 &       \\ 
-			& Best in $K!$  & 8.54    & 12.26   & 13.23   & 14.53  & 13.42 & 61.98 & 5.2\%  \\ 
-			\midrule[.1pt]	
-			\multirow{2}{*}{$1.5$}   & Least cost    & 10.89   & 11.83   & 16.74   & 18.01  & 18.89 & 76.36 & \\ 	
-			& Best in $K!$  & 11.33   & 11.83   & 17.00   & 18.21  & 20.45 & 78.82 &  3.2\%  \\ 
-			\midrule[.1pt]	
-			\multirow{2}{*}{$2.0$}   & Least cost    & 17.78   & 24.40   & 26.36   & 27.01  & 23.94 & 119.49 &  \\ 
-			& Best in $K!$  & 17.78   & 25.25   & 26.38   & 28.98  & 26.08 & 124.47 & 4.2\%   \\ 		                         
-			\midrule[.1pt]
-			\multicolumn{8}{r}{{\bf K! strategy advantage}}                                              & {\bf 4.2}\% \\
-
-			\bottomrule
-		\end{tabular}
-		\normalsize
-	\end{table}
-
-	As the selection of the best tour, considering what was collected and delivered in each node, and considering the cost increase due to CG deviation, was better than the option for a simple TSP solution (\textit{Least cost} results), this confirms that our solution strategy was adequate.
-
-	It is important to report that the best tour found in each test was seldom the tour with the lowest cost (the minimal cost tour without considering the CG deviation), but any of the first 25\% of the best tours.
-This indicates that our strategy of enumerating and solving all tours was assertive, although it is possible to devise a more proactive strategy to discard most of the higher-cost tours and improve performance.
-This must be one of the next steps in this research.
-
-	If a posterior application of this method requires 8, 9, or more nodes, it is possible to use this 25\% as a threshold to discard the higher-cost tours and keep the run time within an operational acceptable limit for the client.
-
-	\subsubsection{Final considerations on the results}
-
-	Most of the {\it Gurobi*} executions consumed over 12 GiB in total, and considering that the computer used in the experiments has a standard RAM consumption when idle of 3.5 GiB (with the Python Interface Development Environment (IDE), a PDF viewer, a system monitor, and a {\it LaTeX} IDE simultaneously open), the actual RAM consumption of {\it Gurobi*} was over 8.5 GiB.
-
-	All {\it Shims} executions consumed at most 1.5 GiB, requiring 5 GiB of total memory.
-	These details are important for a future application to adequately dimension the computer to be used.
-
-}
-
-\section{Conclusions}
-\label{sec7}
-
-In this work, we modeled and solved a real air transport problem, named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem}\/ (ACLP+RPDP).
-For the first time in the literature, a {\it NP-hard}\/ problem that involves {\it simultaneously}\/ pallet assembly, load balancing and route planning is addressed, where the cost-effectiveness of transport is maximized.
-We adopted some simplifications that are not critical, but that allowed the unprecedented solution of this problem considering more than two nodes.
-
-We consider that, in practical cases, the number $K$\/ of nodes, excluding the base, is small ($K \leq 6$), each of them with hundreds of items to be shipped.
-Thus, considering real aircraft model, we developed some node-by-node heuristics, which allow us to enumerate all $K!$\/ tours and choose the best.
-The complete process can be executed quickly on a simple handheld computer, offering good results and reducing stress for the transport planners.
-As validation, we performed tests in several scenarios with real data from the {\it Brazilian Air Force}.
-
-{\color{blue} In less than 20 minutes}, this development can establish a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximizing the total score and minimizing fuel consumption for the airlift.
-The output, which includes the tour plan and the pallet building and arrangement plan, is an essential part of airlift: it improves flight safety, makes ground operations more efficient, and makes sure that each item gets to its right destination.
-
-{\color{blue}
-
-Our main contributions were the mathematical modeling (1) of this complex problem, which involves four {\it NP-hard}\/ sub-problems, a complete process (2), and a new heuristic (3) that offer fast and good-quality solutions to the ACLP+RPDP.
-
-As this is ongoing research, some improvements are possible, which are the next steps of the development:
-
-(1) This work addresses all possible tours, but we plan bounding schemes to allow for discarding most of the higher-cost tours. This is expected to reduce the running time.
-
-(2) We are also working on computer parallelism to accelerate {\it Shims}.
-
-(3) In the event that these gains in performance are achieved, we strongly believe that they will permit the application of a 3-dimensional packing procedure to maximize the occupancy rate of the items selected by {\it Shims}.
-
-The $Shims$\/ heuristic presented a solution with the best node-by-node results in a restricted time interval.
-Without it, it would not be possible to test all routes: after all, the greater the number of nodes, the shorter the run time limit for node-by-node solution.
-We also show that, as the shipment volume increases, the exact node-by-node solution becomes unfeasible, indicating the importance of a fast and good-quality heuristic.
-
-Finally, by focusing on node-by-node solution, the method of this work is not exclusive to aircraft and airports: it can be adapted, for example, to ships and ports, or vehicles and warehouses, or freight cars and railways.
-In these cases, it would be necessary to make some changes in the modeling: for example, modify the load balancing constraints, and consider the available space in vehicles or freight cars instead of pallets.
-
-}
-
-\section*{Acknowledgments}
-
-This research was partially supported by \textit{São Paulo Research Foundation} (FAPESP, grant 2016/01860-1).
-
-\begin{thebibliography}{00}
-
-\bibitem{LarsenMikkelsen1979} O. Larsen and G. Mikkelsen, An interactive system for the loading of cargo aircraft, {\it European Journal of Operational Research}, Vol. 4 (6), pp. 367-373, 1980.
-
-\bibitem{Brosh1981} Israel Brosh, Optimal cargo allocation on board a plane: a sequential linear programming approach, {\it European Journal of Operational Research}, Vol. 8 (1), pp. 40-46, 1981.
-
-\bibitem{JohnsonGarey1985} D.S. Johnson and M.R. Garey, A 71/60 theorem for bin packing, {\it Journal of Complexity}, Vol. 1 (1), pp. 65-106, 1985.
-
-\bibitem{Glover1986} Fred Glover, Future paths for integer programming and links to artificial intelligence, {\it Computers and Operations Research}, Vol. 13, pp. 533-549, 1986.
-
-\bibitem{FeoResende1989} Thomas A.
-Feo and Mauricio G.C.
-Resende, A probabilistic heuristic for a computationally difficult set covering problem, {\it Operations Research Letters}, Vol.
-8 (2), pp.
-67-71, 1989.
-
-\bibitem{Kevin1992} Kevin Y.K. Ng, A multicriteria optimization approach to aircraft loading, {\it Operations Research}, Vol. 40 (6), pp. 1200-1205, 1992.
-
-\bibitem{Dorigo1992} Marco Dorigo, {\it Optimization, Learning and Natural Algorithms}, PhD Thesis, Politecnico di Milano, 1992.
-
-\bibitem{CharonHudry1993} I. Charon and O. Hudry, The noising method: a new method for combinatorial optimization, {\it Operations Research Letters}, Vol. 14 (3), pp. 133-137, 1993.
-
-\bibitem{DorigoManiezzoColorni1996} M.
-Dorigo, V.
-Maniezzo and A.
-Colorni, The ant system: optimization by a colony of cooperating agents, {\it IEEE Transactions on Systems, Man, and Cybernetics}, Vol.
-26 (1), pp.
-29-41, 1996.
-
-\bibitem{NiarFreville1997} S.
-Niar and A.
-Freville, A parallel tabu search algorithm for the 0-1 multidimensional knapsack problem, Proceedings of {\it 11th International Parallel Processing Symposium}, 1997.
-
-\bibitem{Heidelberg1998} K.R. Heidelberg, G.S. Parnell and J.E. Ames, Automated air load planning, {\it Naval Research Logistics}, Vol. 45 (8), pp. 751-768, 1998.
-
-\bibitem{CharonHudry2001} I. Charon and O. Hudry, The noising methods: a generalization of some metaheuristics, {\it European Journal of Operational Research}, Vol. 135 (1), pp. 86-101, 2001.
-
-\bibitem{MongeauBes2003} M. Mongeau and C. Bes, Optimization of aircraft container loading, {\it IEEE Transaction on Aerospace and Electronic Systems}, Vol. 39 (1), pp. 140-150, 2003.
-
-\bibitem{fok2004optimizing} M.K.K.
-Fok and A.H.W.
-Chun, Optimizing air cargo load planning and analysis, Proceedings of {\it The International Conference on Computing, Communications and Control Technologies}, 2004.
-
-\bibitem{Fidanova2006} S. Fidanova, Ant colony optimization and multiple knapsack problem and model bias, {\it Numerical Analysis and Its Applications}, Springer Berlin Heidelberg, pp. 280-287, 2005.
-
-\bibitem{Chan2006} F.T.S.
-Chan, R.
-Bhagwat, N.
-Kumar.
-M.K.
-Tiwari and P.
-Lam, Development of a decision support system for air-cargo pallets loading problem: a case study, {\it Expert Systems with Applications}, Vol.
-31 (3), pp.
-472-485, 2006.
-
-\bibitem{KaluznyBohdanL2009Oalb} B.L. Kaluzny and R.H.A.D. Shaw, Optimal aircraft load balancing, {\it International Transactions in Operational Research}, Vol. 16 (6), pp. 767-787, 2009.
-
-\bibitem{MesquitaCunha2011} A.C.P.
-Mesquita and C.B.
-Cunha, An integrated heuristic based on the Scatter Search metaheuristic for vehicle routing problems with simultaneous delivery and pickup in the context of the Brazilian Air Force, {\it Transportes}, Vol.
-19, pp.
-33-42, 2011.
-
-\bibitem{Verstichel2011} J.
-Verstichel, W.
-Vancroonenburg, W.
-Souffriau and G.V.
-Berghe, A mixed integer programming approach to the aircraft weight and balance problem, {\it Procedia Social and Behavioral Sciences}, Vol.
-20, pp.
-1051-1059, 2011.
-
-\bibitem{Limbourg2012} S. Limbourg, M. Schyns and G. Laporte, Automatic aircraft cargo load planning, {\it Journal of the Operational Research Society}, Vol. 63 (9), pp. 1271-1283, 2012.
-
-\bibitem{RoesenerHall2014} A.G.
-Roesener and S.
-Hall, A nonlinear integer programming formulation for the airlift loading problem with insufficient aircraft, {\it Journal of Nonlinear Analysis and Optimization: Theory and Applications}, Vol.
-5 (1), pp.
-125-141, 2014.
-
-\bibitem{Vancroonenburg2014} W.
-Vancroonenburg, J.
-Verstichel, K.
-Tavernier and G.V.
-Berghe, Automatic air cargo selection and weight balancing: a mixed integer programming approach, {\it Transportation Research Part E: Logistics and Transportation Review}, Vol.
-65, pp.
-70-83, 2014.
-
-\bibitem{LurkinSchyns2015} V. Lurkin and M. Schyns, The airline container loading problem with pickup and delivery, {\it European Journal of Operational Research}, Vol. 244 (3), pp. 955-965, 2015.
-
-\bibitem{RoesenerBarnes2016} A.G. Roesener and J.W. Barnes, An advanced tabu search approach to the dynamic airlift loading problem, {\it Logistics Research}, Vol. 9 (1), pp. 1-18, 2016.
-
-\bibitem{PaquaySchynsLimbourg2016} C.
-Paquay, M.
-Schyns and S.
-Limbourg, A mixed integer programming formulation for the three-dimensional bin packing problem deriving from an air cargo application, {\it International Transactions in Operational Research}, Vol.
-23 (1-2), pp.
-187-213, 2016.
-
-\bibitem{PaquayLimbourgSchynsOliveira2018} C.
-Paquay, S.
-Limbourg, M.
-Schyns and J.F.
-Oliveira, MIP-based constructive heuristics for the three-dimensional bin packing problem with transportation constraints, {\it International Journal of Production Research},
-Vol. 56 (4), pp. 1581-1592, 2018.
-
-\bibitem{Laabadi2018} S.
-Laabadi, M.
-Naimi, H.
-El Amri and B.
-Achchab, The 0/1 multidimensional knapsack problem and its variants: a survey of practical models and heuristic approaches, {\it American Journal of Operations Research}, Vol.
-8, pp.
-395-439, 2018.
-
-\bibitem{YangLiuGao2018} Y. Chenguang, L. Hu and G. Yuan, Load planning of transport aircraft based on hybrid genetic algorithm, {\it MATEC Web of Conferences}, Vol. 179, pp. 1-6, 2018.
-
-\bibitem{Alonso2019} M.T.
-Alonso, R.
-Alvarez-Valdes and F.
-Parreno, A GRASP algorithm for multi-container loading problems with practical constraints, {\it A Quarterly Journal of Operations Research}, Vol.
-18, pp.
-49-72, 2019.
-
-\bibitem{BrandtStefan2019} F.
-Brandt and S.
-Nickel, The air cargo load planning problem - a consolidated problem definition and literature review on related problems, {\it European Journal of Operational Research}, Vol.
-275 (2), pp.
-399-410, 2019.
-
-\bibitem{wong2020} E.Y.C.
-Wong and K.K.T.
-Ling, A mixed integer programming approach to air cargo load planning with multiple aircraft configurations and dangerous goods, Proceedings of {\it 7th International Conference on Frontiers of Industrial Engineering}, 2020.
-
-\bibitem{Zhan2020} S. Zhan, L. Wang, Z. Zhang and Y. Zhong, Noising methods with hybrid greedy repair operator for 0-1 knapsack problem, {\it Memetic Computing}, Vol. 12, pp. 37-50, 2020.
-
-\bibitem{eugene2021} E.Y.C.
-Wong, D.Y.
-Mo and S.
-So, Closed-loop digital twin system for air cargo load planning operations, {\it International Journal of Computer Integrated Manufacturing}, Vol.
-34 (7-8), pp.
-801-813, 2021.
-
-\bibitem{zhao2021} X.
-Zhao, Y.
-Yuan, Y.
-Dong and R.
-Zhao, Optimization approach to the aircraft weight and balance problem with the centre of gravity envelope constraints, {\it IET Intelligent Transport Systems}, Vol.
-15 (10), pp.
-1269-1286, 2021.
-
-\bibitem{LopezIbanezManuel2016} M.
-López-Ibáñez, J.
-Dubois-Lacoste, L.P.
-Cáceres, M.
-Birattari and T.
-Stützle, The irace package: Iterated racing for automatic algorithm configuration, {\it Operations Research Perspectives}, Vol.
-3, pp.
-43-58, 2016.
-
-	\bibitem{mipgap} Matthias Miltenberger. {\it What is the MIPGap?} \url{https://support.gurobi.com/hc/en-us/articles/8265539575953-What-is-the-MIPGap-}, 2023.
-
-\end{thebibliography}
-
 \end{document}

### 2023-06-08 main.py

@@ -262,8 +262,10 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplus,
         tour.AvgVol    += nodeVol
         tour.AvgTorque += epsilon
 
-        print(f"\tnode {node.ICAO},", end='')
-        print(f" score {tour.score:.0f}, cost {tour.cost:.0f}, vol {nodeVol:.2f}, epsilon {epsilon:.2f}")
+        f = tour.score/tour.cost
+
+        print(f"\tnode {node.ICAO}")
+        print(f"f {f:.2f}  vol {nodeVol:.2f} epsilon {epsilon:.2f}")
         
 
         if writeConsFile:
@@ -326,9 +328,9 @@ if __name__ == "__main__":
     testing   = False
     leastCost = False
 
-    timeLimit = 1200
+    # timeLimit = 1200
     # timeLimit = 2400
-    # timeLimit = 3600
+    timeLimit = 3600
 
 
     # testing    = True
@@ -338,18 +340,18 @@ if __name__ == "__main__":
     # free -s 1 -h -c 3  memory
 
     # scenarios = [2,3,4,5,6]
-    scenarios = [5,6]
+    scenarios = [6]
 
     if testing:
         scenarios = [2]
 
-    surplus = "data20"  # 1.2
-    # surplus = "data50"  # 1.5
+    # surplus = "data20"  # 1.2
+    surplus = "data50"  # 1.5
     # surplus = "data100" # 2.0
 
     # methods = ["GRB"]
-    # methods = ["CBC"]
-    methods = ["Shims"]
+    methods = ["CBC"]
+    # methods = ["Shims"]
     # methods = ["mpShims"]
     # methods = ["mpACO"]
     # methods = ["ACO"]

### 2023-06-08 optcgcons.py

@@ -64,23 +64,20 @@ def minCGdev(pallets, k, nodeTorque, cfg):
     # mod.setParam('OutputFlag', 0)
     mod.verbose = 0 # hide messages
 
-    # X = [ [ mod.addVar(name=f"X[{i}],[{j}]", vtype=GRB.BINARY) for j in ConsRange ] for i in PalletsRange ] 
-    X = [ [ mod.add_var(name=f"X[{i}],[{j}]", var_type=BINARY) for j in ConsRange ] for i in PalletsRange ] 
+    Y = [ [ mod.add_var(name=f"X[{i}],[{q}]", var_type=BINARY) for q in ConsRange ] for i in PalletsRange ] 
 
-    torque1 = xsum( X[i][j] * ((140+cons[j].W) * pallets[i].D) for i in PalletsRange for j in ConsRange if pallets[i].D > 0 ) 
-    torque2 = xsum( X[i][j] * ((140+cons[j].W) * pallets[i].D) for i in PalletsRange for j in ConsRange if pallets[i].D < 0 )
+    torque1 = xsum( Y[i][q] * ((140+cons[q].W) * pallets[i].D) for i in PalletsRange for q in ConsRange if pallets[i].D > 0 ) 
+    torque2 = xsum( Y[i][q] * ((140+cons[q].W) * pallets[i].D) for i in PalletsRange for q in ConsRange if pallets[i].D < 0 )
 
-    # mod.setObjective(torque1-torque2)
-    # mod.ModelSense = GRB.MINIMIZE
     mod.objective = minimize( torque1-torque2 )
 
-    for j in ConsRange:
+    for q in ConsRange:
         mod.add_constr(
-            xsum(X[i][j] for i in PalletsRange) == 1
+            xsum(Y[i][q] for i in PalletsRange) <= 1
         )        
     for i in PalletsRange:                
         mod.add_constr(
-            xsum(X[i][j] for j in ConsRange) == 1
+            xsum(Y[i][q] for q in ConsRange) == 1
         )
 
     mod.optimize() 
@@ -89,17 +86,15 @@ def minCGdev(pallets, k, nodeTorque, cfg):
 
     for i, _ in enumerate(pallets):
 
-        for j in ConsRange:
-            if X[i][j].x >= 0.99:
-                cons[j].P = i # put the consolidated in the best position to minimize torque                        
-                pallets[i].Dest[k] = cons[j].To                
-                pallets[i].PCW     = cons[j].W
[…]

### 2023-06-08 CAIE/cie.tex
cie file updated

@@
 		\cmidrule(lr){2-9}                              		
-		& \multirow{2}{*}{3600}& $f$                               & 14.61   & x       &  20.59  &         & 29.11   & x \\ % 
-		&        &  {\bf run time (s)}                             & 297     & x       &  470    &         & 689     & x \\ % 		
+		& \multirow{2}{*}{3600}& $f$                               & 14.61   & x       &  20.59  & 18.36   & 29.11   & x \\ % 
+		&        &  {\bf run time (s)}                             & 297     & x       &  470    & 2241    & 689     & x \\ % 		
 		\midrule
@@
 
-		&                                       & {\bf Normalized} & 0.99    & x       & 0.99    & x       & 1.00    & x\\
-		&                                       & {\bf Speed-up}   & 15.6    & x       & 14.3    & x       & 16.3    & x\\
+		&                                       & {\bf Normalized} & 0.99    & x       & 0.99    & 0.98    & 1.00    & x\\
+		&                                       & {\bf Speed-up}   & 15.6    & x       & 14.3    & 14.3    & 16.3    & x\\
 
@@
 
-\subsubsection{Proposed K! strategy assessment}
+\subsubsection{Proposed $S_K$ strategy assessment}
 
@@
 
-	For this achievement, we ran {\it Shims} in 5 scenarios with 7 instances each, and 3 volume surpluses, collecting the averages of the best tour result and the least cost tour result.
+	For this achievement, we collected {\it Shims} results from tables \ref{tab:20}, \ref{tab:50}, and \ref{tab:100}, gathering the best tour results and the least cost results to assemble Table \ref{tab:20least}.
 
@@
 			\multirow{2}{*}{$1.2$}   & Least cost    & 8.54    & 11.63   & 12.97   & 13.43  & 12.36 & 58.93 &       \\ 
-			& Best in $K!$  & 8.54    & 12.26   & 13.23   & 14.53  & 13.42 & 61.98 & 5.2\%  \\ 
+			& Best in $S_K$  & 8.54    & 12.26   & 13.23   & 14.53  & 13.42 & 61.98 & 5.2\%  \\ 
 			\midrule[.1pt]	
 			\multirow{2}{*}{$1.5$}   & Least cost    & 10.89   & 11.83   & 16.74   & 18.01  & 18.89 & 76.36 & \\ 	
-			& Best in $K!$  & 11.33   & 11.83   & 17.00   & 18.21  & 20.45 & 78.82 &  3.2\%  \\ 
+			& Best in $S_K$  & 11.33   & 11.83   & 17.00   & 18.21  & 20.45 & 78.82 &  3.2\%  \\ 
 			\midrule[.1pt]	
 			\multirow{2}{*}{$2.0$}   & Least cost    & 17.78   & 24.40   & 26.36   & 27.01  & 23.94 & 119.49 &  \\ 
-			& Best in $K!$  & 17.78   & 25.25   & 26.38   & 28.98  & 26.08 & 124.47 & 4.2\%   \\ 		                         
+			& Best in $S_K$  & 17.78   & 25.25   & 26.38   & 28.98  & 26.08 & 124.47 & 4.2\%   \\ 		                         
 			\midrule[.1pt]
-			\multicolumn{8}{r}{{\bf K! strategy advantage}}                                              & {\bf 4.2}\% \\
+			\multicolumn{8}{r}{\boldm{$S_K$} {\bf strategy advantage}}                                              & {\bf 4.2}\% \\
 
@@
 
-	It is important to report that the best tour found in each test was seldom the tour with the lowest cost (the minimal cost tour without considering the CG deviation), but any of the first 25\% of the best tours.
+\subsubsection{Final considerations on the results}
+
+It is important to report that the best tour found in each test was seldom the tour with the lowest cost (the minimal cost tour without considering the CG deviation), but any of the best tours.
 This indicates that our strategy of enumerating and solving all tours was assertive, although it is possible to devise a more proactive strategy to discard most of the higher-cost tours and improve performance.
 This must be one of the next steps in this research.
-
-	If a posterior application of this method requires 8, 9, or more nodes, it is possible to use this 25\% as a threshold to discard the higher-cost tours and keep the run time within an operational acceptable limit for the client.
-
-	\subsubsection{Final considerations on the results}
+If effective, this strategy may permit posterior applications for eight or more nodes.
 
@@
 
-	All {\it Shims} executions consumed at most 1.5 GiB, requiring 5 GiB of total memory.
-	These details are important for a future application to adequately dimension the computer to be used.
+All {\it Shims} executions consumed at most 1.5 GiB, requiring 5 GiB of a computer total memory.
+These details are important for a future application to adequately dimension the machine to be used.
 

### 2023-06-09 CAIE/cie.tex
Experimentations ended
New experimentation to cie.tex
Alternative to table 9

@@
 	\centering
-	\caption{Solutions with different time limitations for $S_K$ tours.}  \label{tab:100time}
+	\caption{Solutions with different time limitations for $S_K$ tours.}  \label{tab:100time_2}
 	\footnotesize
-	\begin{tabular}{ccc|cc|cc|cc}
+	\begin{tabular}{ccc|ccc|ccc|ccc}
 		\toprule
-		&             \multicolumn{2}{r}{\bf Volume $surplus$} &\multicolumn{2}{c}{$1.2$}&\multicolumn{2}{c}{$1.5$}&\multicolumn{2}{c}{$2.0$} \\
+		&             \multicolumn{2}{r}{\bf Volume $surplus$} &\multicolumn{3}{c}{$1.2$}&\multicolumn{3}{c}{$1.5$}&\multicolumn{3}{c}{$2.0$} \\
 		\midrule
-		$method$& {\bf \specialcell{Time\\limit (s)}} & {\bf Scenarios}    & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} \\
+		$method$& {\bf limit (s)} & {\bf Scenarios}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 3}&{\bf 4}&{\bf 5} \\
 		\toprule
 
-		\multirow{6}{*}{\it Gurobi* }& \multirow{2}{*}{1200} & $f$ & 14.67   & x       &  20.64  &  x      & 27.97   & x \\ % 
-		&        &  {\bf run time (s)}                             & 217     & x       &  304    &  x      & 579     & x \\ % 
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$                               & 14.59   &         &  20.60  &  x      & 29.10   & x \\ % 
-		&        &  {\bf run time (s)}                             & 313     &         &  327    &  x      & 543     & x \\ % 
-		\cmidrule(lr){2-9}                              		
-		& \multirow{2}{*}{3600}& $f$                               & 14.61   & x       &  20.59  & 18.36   & 29.11   & x \\ % 
-		&        &  {\bf run time (s)}                             & 297     & x       &  470    & 2241    & 689     & x \\ % 		
+		\multirow{8}{*}{\it Gurobi } & \multirow{2}{*}{240*} & $f$ & 13.34 &  x      &  x      & 18.31 &  x      &  x      & x     & x       & x \\ % 
+&               &  {\bf run time (s)}                              & 123   &  x      &  x      & 181   &  x      &  x      & x     & x       & x \\ % 
+\cmidrule(lr){2-12}
+
+		&                              \multirow{2}{*}{1200} & $f$ & 13.31 & 14.67   & x       & 18.25 &  20.64  &  x      & 26.49 & 27.97   & x \\ % 
+		&        &  {\bf run time (s)}                             & 129   & 217     & x       & 190   &  304    &  x      & 384   & 579     & x \\ % 
+		\cmidrule(lr){2-12}                              
+		& \multirow{2}{*}{2400}& $f$                               & 13.30 & 14.59   & 13.59   & 18.25 &  20.60  &  x      & 26.46 & 29.10   & x \\ % 
+		&        &  {\bf run time (s)}                             & 127   & 313     & 1444    & 188   &  327    &  x      & 381   & 543     & x \\ % 
+		\cmidrule(lr){2-12}
+
+		& \multirow{2}{*}{3600}& $f$                               & 13.32 & 14.61   & x       & 18.25 &  20.59  & 18.36   & 26.46 & 29.11   & x \\ % 
+		&        &  {\bf run time (s)}                             & 121   & 297     & x       & 193   &  470    & 2241    & 319   & 689     & x \\ %
+
 		\midrule
-		\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
-		&        & {\bf run time (s)}                              & 18      & 109     &  26     & 160     & 37      & 203   \\ % 
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$                               & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
-		&        &  {\bf run time (s)}                             & 18      & 107     &  25     & 159     & 37      & 192   \\ % 	
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{3600} & $f$                              & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ %  
-		&        &  {\bf run time (s)}                             & 17      & 106     &  26     & 157     & 37      & 181   \\ %  
+
+		 \multirow{2}{*}{\it Shims} & \multirow{2}{*}{240*}  & $f$ & 13.23 & 14.53   & 13.42   & 18.21 &  20.45  & 18.00   & 26.38 & 28.98   & 26.08 \\ % 
+		&         & {\bf run time (s)}                             & 6     & 17      & 109     & 8     &  26     & 160     & 13    & 37      & 229   \\ %
+
 		\bottomrule
 
-		&                                       & {\bf Normalized} & 0.99    & x       & 0.99    & 0.98    & 1.00    & x\\
-		&                                       & {\bf Speed-up}   & 15.6    & x       & 14.3    & 14.3    & 16.3    & x\\
+		&                                       & {\bf Normalized} & 0.99  & 0.99    & 0.99    & 1.00  & 0.99    & 0.98    & 1.00  & 1.00    & x\\
+		&                                       & {\bf Speed-up}   & 20.9  & 15.6    & 13.5    & 23.8  & 14.3    & 14.3    & 29.0  & 16.3    & x\\
 
 		\bottomrule
-
+				\multicolumn{12}{c}{*4 minutes is the time {\it Shims} needs to solve a seven-node tour.}\\
 	\end{tabular}
@@
 \end{table}
+
+According to the results in Table \ref{tab:100time}, {\it Shims} was 15 times faster than {\it Gurobi* } to solve the ACLP-RPDP, and {\it Gurobi* } was not fully capable of solving scenario 5.
 

### 2023-06-09 main.py

@@ -329,8 +329,9 @@ if __name__ == "__main__":
     leastCost = False
 
     # timeLimit = 1200
-    timeLimit = 2400
+    # timeLimit = 2400
     # timeLimit = 3600
+    timeLimit = 240
 
 
     # testing    = True

### 2023-06-10 CAIE/CIE.tex
Paper finished.

--- 2023-06-08 CAIE/cie.tex
@@
 In aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading, for fast take-off, and mission accomplishment, especially in times of crisis, disaster relief missions, short deadlines for contracting customers, or any external pressure for immediate take-off.
-In addition, there are no commercially available systems that can assist the load and trip planners with the pallet building and aircraft loading and balancing demands for transport at each hub.
-This enables other risks, such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
+{\color{blue} In addition, there are no commercially available systems that can assist load and trip planners with pallet building and aircraft-balanced loading plans, with demands for transport at each hub.} This enables other risks, such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
 We defined and solved the problem of planning the loading and routing of a single aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
@@
 We performed multiple experiments with a commercial solver and four well-known meta-heuristics on data based on the transport history of the {\it Brazilian Air Force}.
-We also designed a new heuristic that quickly finds good solutions for a wide range of problem sizes: an essential contribution, which solved all test scenarios within an acceptable operational run time.
+{\color{blue}We also designed a new heuristic that quickly finds practical solutions for a wide range of problem sizes, an essential contribution that solved all test scenarios within a much shorter time than is operationally acceptable.}
 
@@
 
-\subsubsection{With Shims heuristic}
+\subsubsection{With the Shims heuristic}
 
@@
 			\State $\eta_2 \gets \eta_1$  \label{shims:eta2a}
-			\While{($\eta_2 \leq n_{\pi_k}$) {\bf and} ($ vol_i < (1 + 3 \times (1-limit)) \times V_i)$)}
+			\While{($\eta_2 \leq n_{\pi_k}$) {\bf and} ($ vol_i < (1 + 2 \times (1-limit)) \times V_i$)}
 
@@
 
-			\State $vol \gets 0$, $b \gets 1$, $shims_b \gets \varnothing$, $Set \gets \{ shims_b \}$  \label{shims:beginthird}
+			\State $vol \gets 0$; $b \gets 1$; $shims_b \gets \varnothing$; $Set \gets \{ shims_b \}$  \label{shims:beginthird}
 
@@
 				\If{$NewShims$} \label{new_shims2}
-					\State $vol \gets 0$
-					\State $b \gets b + 1$
-					\State $shims_b \gets \{\ \}$
-					\State $shims_b \gets shims_b \cup \{e_{ij}\}$
-					\State $Set \gets Set \cup \{\ shims_b\ \}$
+					\State $vol \gets 0$; $b \gets b + 1$;  $shims_b \gets \{e_{ij}\}$
+					\State $Set \gets Set \cup \{ shims_b \}$
 				\EndIf
@@
 
-			\State $sh_w \gets shims$, where $shims \in Set$ and $\sum_{e_{ab} \in shims} w_b$ is maximum  \label{best_weight}		
-			\State $sh_v \gets shims$, where $shims \in Set$ and $\sum_{e_{ab} \in shims} v_b$ is maximum \label{best_volume}
-			\State $sh_{best} \gets shims$, where $shims \in \{sh_w, sh_v\}$ and $\sum_{e_{ab} \in x} s_b$ is maximum \label{best_score}
-
+			\State $sh_w \gets shims$, where $shims \in Set$ and $\sum_{e_{ij} \in shims} w_j$\/ is maximum  \label{best_weight}		
+			\State $sh_v \gets shims$, where $shims \in Set$ and $\sum_{e_{ij} \in shims} v_j$\/ is maximum \label{best_volume}
+			\State $sh_{best} \gets shims$, where $shims \in \{sh_w, sh_v\}$\/ and $\sum_{e_{ij} \in shims} s_j$\/ is maximum \label{best_score}
 			\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup sh_{best}$ \label{shims:endthird}  \Comment{End of the third phase}
@@
 
-	\item In {\it greedy phase}\/ (lines \ref{shims:pallets}-\ref{shims:endfirst}), a partial solution for each pallet $i$\/ is constructed by edges inclusion, until its volume $V_i \times limit$\/ is reached.
-The value $limit=0.92$\/ (line \ref{shims:limit}) was empirically determined by the {\it iRace} tool \cite{LopezIbanezManuel2016}.
-From lines \ref{shims:limit} to \ref{shims:endfirst}, as items are inserted, the value of $\eta_1$ is incremented and stops when the pallet volume reaches 92\% of its capacity.
-When it stops, this is the initial value of $\eta_2$, line \ref{shims:eta2a}.
-	$\eta_2$, then it is incremented until the volume exceeds the pallet capacity with a factor of 1 + ( 2 x [1 - limit]).
-	This factor of was obtained empirically, through many test with different problem instances.
-Higher values can improve the solution, but at the expense of performance.
-So we tried to keep it as low as possible, a trade-off between quality and performance.
+	\item In {\it greedy phase}\/ (lines \ref{shims:pallets}-\ref{shims:endfirst}), a partial solution for each pallet $i$\/ is constructed by adding edges following $\theta_{ij}$\/ order.
+The $\eta_1$\/ index corresponds to the accumulated volume equal to $V_i \times limit$.
+In $\eta_2$\/ index, this same accumulated volume reaches $(1 + 2\times(1 - limit))\times V_i$.
+The size of this range $[\eta_1,\eta_2]$\/ was defined empirically, and the value $limit=0.92$\/ (line \ref{shims:limit}) was determined by the {\it iRace} tool \cite{LopezIbanezManuel2016}.
 
 	\item In {\it composition phase}\/ (lines \ref{shims:beginsecond}-\ref{shims:endsecond}), a set of shims named $Set$\/ is created for each pallet $i$, where each shim is formed by a set of edges in the range $[\eta_1,\eta_2]$, whose total volume is limited by $slack_i$.
-After the $\eta_2$ index, there may be items with little probability of being in the optimal solution because their low value of $\theta^{\pi_k}_{ij}$, being not considered for inclusion.
-In line \ref{shims:beginthird} the initial $Shim$ and $Set$ are created.
 In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it First-Fit Decreasing}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \cite{JohnsonGarey1985}.
@@
 	\centering
-	\caption{Items weight distribution ($kg$) in Brazil}  \label{tab:weights}
+	\caption{Items weight distribution}  \label{tab:weights}
 	\begin{tabular}{c c c c }
 		\toprule
-		$Group$ & $P$ & $low$ & $high$  \\		
+		$Group$ & $P$ & $low$ ($kg$) & $high$ ($kg$) \\		
 		\midrule
@@
 
+$ItemsGeneration$, which generates $N$, is described in Algorithm \ref{alg:itemsgen}.
+
 \begin{algorithm}[H]
@@
 
-$ItemsGeneration$, which generates $N$, is described in Algorithm \ref{alg:itemsgen}.
-The parameter $scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and the parameter $surplus$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}).
+$scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and $surplus$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}).
 To avoid simply loading all items, we use $surplus \in \{1.2, 1.5, 2.0\}$.
@@
 
-{\color{blue}
-
 \subsection{Results obtained}
@@
 In the tests performed, we used a 64-bit, 16GB, 3.6GHz, eight-core processor with {\it Linux Ubuntu 22.04.1 LTS 64-bit} as the operational system and {\it Python 3.10.4}\/ as the programming language.
-We also used the well-known solver {\it Gurobi}\/ (www.gurobi.com), version 9.5.2.
-
-We ran Algorithm \ref{alg:main} in the 5 scenarios described in Table \ref{tab:scenarios}, considering 6 methods for node-by-node solution: Gurobi (see \ref{solver}), ACO, NMO, TS, GRASP, and {\it Shims} (Algorithm \ref{alg:shims}).
-In generating the items in each node, we consider 3 values for the parameter $surplus$.
-3 values for the time limit (1200s, 2400s, and 3600s).
-
-The results obtained for the function $f$, with the corresponding run time in seconds, are shown in Tables \ref{tab:20} ($surplus = 1.2$), \ref{tab:50} ($surplus = 1.5$), \ref{tab:100} ($surplus = 2.0$), and in Table \ref{tab:100time} (different time limit comparisons).
-
-For each $method$, $scenario$\/ and $surplus$, 7 different instances were generated.
-Therefore, 315 tests (5 scenarios $\times$\/ 3 volumes $\times$\/ 3 time limits $\times$\/ 7 instances) were performed for each method.
-
-The average values were presented for $f_{\pi}$\/ and, for the {\bf run time (s)}, the average result obtained.
-To facilitate the comparison between the methods, we added a last column in these tables, where two values are indicated:
+We also used the well-known solver $Gurobi$\/ ({\tt www.gurobi.com}), version 9.5.2.
+
+We ran Algorithm \ref{alg:main} considering 5 values for $scenario$\/ from Table \ref{tab:scenarios}, 3 values for  $surplus$\/ from $\{1.2, 1.5, 2.0\}$, 3 values for $time$\/ from $\{1200s, 2400s, 3600s\}$, and 6 different $method$\/ for the node-by-node solution: $Gurobi$\/ (see \ref{solver}), ACO, NMO, TS, GRASP, and $Shims$\/ (Algorithm \ref{alg:shims}).
+
+In order for $Gurobi$\/ to be able to solve the largest possible number of tests without memory overflow, we set its parameter {\it MIPgap}\/ to 1\%.
+This shortens its runtime, in addition to ensuring that its objective function $f$\/ is at most 1\% of the optimal solution.
+For more details, see {\tt https://support.gurobi.com}.
+
+Among the heuristics, we will present only the results obtained by $Shims$, which had the best performance.
+For each $scenario$, $surplus$\/ and $time$, 7 different instances were generated.
+We will present the average of the objective function $f$\/ and the runtime of $Gurobi$\/ and $Shims$\/ in these instances.
+To facilitate the comparison between both, we added a last column in the tables, where two values are indicated:
 \begin{itemize}
-	\item {\bf Normalized}: value between 0 and 1, which corresponds to the ratio between the sum of $f$\/ values obtained by the method in all scenarios and the sum of the best values obtained among all methods in all scenarios.
+	\item {\bf Normalized}: value between 0 and 1, which corresponds to the ratio between the sum of $f$\/ values obtained by the method in all scenarios and the sum of the best values obtained by both methods in all scenarios.
 The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
-	\item {\bf Speed-up}: ratio of the sums of the run times of all scenarios and the sum of the method run times in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
+	\item {\bf Speed-up}: ratio of the sums of the runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
 \end{itemize}
 
-	As the MIP solver was not capable of solving all scenarios in polynomial time or reached the computer's Read-Access Memory (RAM, 16 GiB) too early, we decided to set the Gurobi {\it MIPgap} parameter to 1\% to control the minimal quality of the returned solutions and abbreviate the solution time.
-	From now on, we will refer to this Gurobi-specific model as {\it Gurobi* }.
-	For more details on the Gurobi {\it MIPgap}, refer to \cite{mipgap}.
-
-	This decision hinders the MIP solver from reaching optimal solutions in favor of the possibility of result comparisons with the {\it Shims} heuristic.
-
-	The parameter $\pi_{TSP_1}$ represents the least cost tour (through an optimal TSP solution), and $\pi_{TSP_2}$, the opposite direction of travel, beginning and ending in the base.
-The values of $f$ are the best in these to solutions.
+We also indicate the two adopted strategies: dedicating all the processing time to the two shortest tours, or distributing it among all the $K!$\/ tours.
+
+The results obtained with $time = 3600s$, which is the highest tested time limit, are in Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, with $surplus$\/ values $1.2$, $1.5$\/ and $2.0$, respectively.
+We indicate with an {\bf x} the cases where $Gurobi$\/ did not find a feasible solution within this time limit, or had to be aborted due to high RAM consumption.
 
@@
 	\centering
-	\caption{Solutions with volume {\it surplus} = {\bf 1.2} and 3600s as time limit.}  \label{tab:20}
+	\caption{Solutions with $surplus = 1.2$\/ and $time =  3600s$}  \label{tab:20}
 	\footnotesize
@@
 		\toprule
-		{\bf Tours}&$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		{\specialcell{{\bf Tested}\\{\bf tours}}}&$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
 		\toprule
-		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} &\multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\ % 5927
-		&                                                               &               {\bf run time (s)}  & 17      & 17      & 16      & 17    & 26        &  1.0 \\ % 93
+		\multirow{4}{*}{Shortest} &\multirow{2}{*}{\it Gurobi } & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\ % 5927
+		&                                                               &               {time (s)}  & 17      & 17      & 16      & 17    & 26        &  1.0 \\ % 93
 		\cmidrule(lr){2-9}	
 		&                                                               \multirow{2}{*}{\it Shims}    & $f$ & 8.54    & 11.63   & 12.97   & 13.43 & 12.36     &  0.99 \\ % 5893
-		&                                                               &         {\bf run time (s)}        & 1       & 1       & 2       & 2     & 2         &  11.6 \\ % 8
-
-		\midrule
-
-		\multirow{4}{*}{\it $S_{K}$ }&                                  \multirow{2}{*}{\it Gurobi* } & $f$ & 8.52    & 12.25   & 13.32   & 14.61  & x        & 1.00 \\ % 4869
-		&                                                               &               {\bf run time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\ % 479
+		&                                                               &         {time (s)}        & 1       & 1       & 2       & 2     & 2         &  11.6 \\ % 8
+
+		\midrule
+
+		\multirow{4}{*}{All $K!$}&                                  \multirow{2}{*}{\it Gurobi } & $f$ & 8.52    & 12.25   & 13.32   & 14.61  & x        & 1.00 \\ % 4869
+		&                                                               &               {time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\ % 479
 		\cmidrule(lr){2-9}	
 		&                                                               \multirow{2}{*}{\it Shims}    & $f$ & 8.54    & 12.26   & 13.23   & 14.53  & 13.42    & 0.99  \\ % 4813
-		&                                                               &         {\bf run time (s)}        & 1       & 2       & 6       & 17     & 106      & 18.4   \\ % 132
+		&                                                               &         {time (s)}        & 1       & 2       & 6       & 17     & 106      & 18.4   \\ % 132
 		\bottomrule
-		\multicolumn{9}{c}{*{\it MIPgap} parameter set to 1\%. The $f$ values are \textpm 1\% distant from the optimal solution}\\
 
@@
 
-{\color{blue}
-
-	97\% of 16GiB was the RAM state before we aborted the {\it Gurobi*} execution on solving scenario 5 in Table \ref{tab:20}.
-
-}
-
 \begin{table}[H]
 	\centering
-	\caption{Solutions with volume {\it surplus} = {\bf 1.5} and 3600s as time limit.}  \label{tab:50}
+	\caption{Solutions with $surplus = 1.5$\/ and $time =  3600s$}  \label{tab:50}
 	\footnotesize
@@
 		\toprule
-		{\bf Tours} & $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		{\specialcell{{\bf Tested}\\{\bf tours}}} & $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
 		\toprule
-		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\ % 8236
-		&&               {\bf run time (s)}                                                                  & 38    & 30    & 27    & 29    & 81    & 1.00  \\ % 205
+		\multirow{4}{*}{ Shortest } & \multirow{2}{*}{\it Gurobi } & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\ % 8236
+		&&               {time (s)}                                                                  & 38    & 30    & 27    & 29    & 81    & 1.00  \\ % 205
 		\cmidrule(lr){2-9}	
 		&\multirow{2}{*}{\it Shims}                                                                    & $f$ & 11.83 & 16.74 & 18.01 & 18.89 & 16.91 & 1.00  \\ % 
-		&&         {\bf run time (s)}                                                                        & 1     & 2     & 2     & 3     & 3     & 20.5  \\ %
-
-		\midrule
-
-		\multirow{4}{*}{\it $S_{K}$ } & \multirow{2}{*}{\it Gurobi* }                                  & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ % 8602
-		&&               {\bf run time (s)}                                                                  & 53    & 61    & 193   & 470   & 2241    & 1.0  \\ % 3018
+		&&         {time (s)}                                                                        & 1     & 2     & 2     & 3     & 3     & 20.5  \\ %
+
+		\midrule
+
+		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{\it Gurobi }                                  & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ % 8602
+		&&               {time (s)}                                                                  & 53    & 61    & 193   & 470   & 2241    & 1.0  \\ % 3018
 		\cmidrule(lr){2-9}	
 		&\multirow{2}{*}{\it Shims}                                                                    & $f$ & 11.83 & 17.00 & 18.21 & 20.45 & 18.00   & 0.99  \\ % 8549
-		&&         {\bf run time (s)}                                                                        & 1     & 2     & 9     & 26    & 157     & 15.5 \\ % 195
+		&&         {time (s)}                                                                        & 1     & 2     & 9     & 26    & 157     & 15.5 \\ % 195
 		\bottomrule
@@
 	\centering
-	\caption{Solutions with volume {\it surplus} = {\bf 2.0} and 3600s as time limit.}  \label{tab:100}
+	\caption{Solutions with $surplus = 2.0$\/ and $time =  3600s$}  \label{tab:100}
 	\footnotesize
@@
 		\toprule
-		{\bf Tours} & $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		{\specialcell{{\bf Tested}\\{\bf tours}}}& $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
 		\toprule
-		\multirow{4}{*}{ \specialcell{$\pi_{TSP_1}$\\ $\pi_{TSP_2}$ }} & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 24.39 & 26.46 & 27.09 & 24.00 & 1.00 \\ % 11969
-		&&                                                                               {\bf run time (s)}  & 162   & 95    & 74    & 68    & 67    & 1.0  \\ % 466
+		\multirow{4}{*}{ Shortest} & \multirow{2}{*}{\it Gurobi } & $f$ & 17.75 & 24.39 & 26.46 & 27.09 & 24.00 & 1.00 \\ % 11969
+		&&                                                                               {time (s)}  & 162   & 95    & 74    & 68    & 67    & 1.0  \\ % 466
 		\cmidrule(lr){2-9}	
 		&                                                                   \multirow{2}{*}{\it Shims} & $f$ & 17.78 & 24.40 & 26.36 & 27.01 & 23.94 & 0.99  \\ % 11949
-		&&                                                                                {\bf run time (s)} & 2     & 3     & 3     & 4     & 5     & 27.4  \\ % 17
-
-		\midrule
-
-		\multirow{4}{*}{\it $S_{K}$ } & \multirow{2}{*}{\it Gurobi* } & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &  x   & 1.00 \\ % 9856
-		&&                                                                                {\bf run time (s)} & 151   & 123   & 319   & 689   &  x   & 1.0  \\ % 1282
+		&&                                                                                {time (s)} & 2     & 3     & 3     & 4     & 5     & 27.4  \\ % 17
+
+		\midrule
+
+		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{\it Gurobi } & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &  x   & 1.00 \\ % 9856
+		&&                                                                                {time (s)} & 151   & 123   & 319   & 689   &  x   & 1.0  \\ % 1282
 		\cmidrule(lr){2-9}	
 		&                                                                \multirow{2}{*}{\it Shims}    & $f$ & 17.78 & 25.25 & 26.38 & 28.98 & 26.08 & 1.00  \\ % 9839
-		&&                                                                                {\bf run time (s)} & 2     & 3     & 13    & 37    & 229   & 23.3  \\ % 284
+		&&                                                                                {time (s)} & 2     & 3     & 13    & 37    & 229   & 23.3  \\ % 284
 		\bottomrule
@@
 
-{\color{blue}
-	{\it Gurobi*} could not solve scenario 5 in Table \ref{tab:100} because $3600/K! = 0.71s$ is not enough time for {\it Gurobi*} to find a feasible node solution.
-}
-
-\subsubsection{Time limits comparisons}
+From these data, we can make some conclusions:
+\begin{itemize}
+\item The strategy of testing all $K!$\/ often gives a better quality solution.
+\item $Gurobi$\/ fails in some cases, when $scenario=5$\/ and the strategy is to check all $K!$\/ tours.
+This occurs because the time limit per node is smaller, and there tends to be more packed contents on the aircraft, reducing the space for allocating items and making the solution difficult.
+\item When $Gurobi$\/ finishes, it finds the best solution, but the one obtained by $Shims$\/ reaches $99\%$\/ of that value.
+\item $Shims$\/ always finds a solution, being 11 to 27 times faster.
+\item All runtimes are well under the time limit because the solution on many nodes can be fast.
+Anyway, in all the tests performed, the maximum time spent by $Shims$\/ did not reach 4 minutes, that is, it is practically instantaneous.
+On the other hand, when $scenario=5$\/ and $surplus=1.5$, Gurobi spent almost 40 minutes.
+\end{itemize}
+
+Table \ref{tab:100time} shows the results obtained with the strategy of testing the $K!$\/ tours in scenarios 4 and 5, with different time limits.
+We can observe more cases where $Gurobi$\/ fails.
+When $Gurobi$ finishes, $Shims$\/ finds a solution of similar quality in much less time.
 
@@
 	\centering
-	\caption{Solutions with different time limitations for $S_K$ tours.}  \label{tab:100time}
+	\caption{Solutions in scenarios 4 and 5 testing all $K!$\/ tours}  \label{tab:100time}
 	\footnotesize
@@
 		\toprule
-		&             \multicolumn{2}{r}{\bf Volume $surplus$} &\multicolumn{2}{c}{$1.2$}&\multicolumn{2}{c}{$1.5$}&\multicolumn{2}{c}{$2.0$} \\
-		\midrule
-		$method$& {\bf \specialcell{Time\\limit (s)}} & {\bf Scenarios}    & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} \\
+		&             \multicolumn{2}{c}{$surplus$} &\multicolumn{2}{c}{$1.2$}&\multicolumn{2}{c}{$1.5$}&\multicolumn{2}{c}{$2.0$} \\
+		\midrule
+		$method$& {\specialcell{$time$ (s)}} & {\bf Scenarios}    & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} \\
 		\toprule
 
-		\multirow{6}{*}{\it Gurobi* }& \multirow{2}{*}{1200} & $f$ & 14.67   & x       &  20.64  &  x      & 27.97   & x \\ % 
-		&        &  {\bf run time (s)}                             & 217     & x       &  304    &  x      & 579     & x \\ % 
+		\multirow{6}{*}{\it Gurobi }& \multirow{2}{*}{1200} & $f$ & 14.67   & x       &  20.64  &  x      & 27.97   & x \\ % 
+		&        &  {time (s)}                             & 217     & x       &  304    &  x      & 579     & x \\ % 
 		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$                               & 14.59   &         &  20.60  &  x      & 29.10   & x \\ % 
-		&        &  {\bf run time (s)}                             & 313     &         &  327    &  x      & 543     & x \\ % 
+		& \multirow{2}{*}{2400}& $f$                               & 14.59   & 13.59   &  20.60  &  x      & 29.10   & x \\ % 
+		&        &  {time (s)}                             & 313     & 1444    &  327    &  x      & 543     & x \\ % 
 		\cmidrule(lr){2-9}                              		
 		& \multirow{2}{*}{3600}& $f$                               & 14.61   & x       &  20.59  & 18.36   & 29.11   & x \\ % 
-		&        &  {\bf run time (s)}                             & 297     & x       &  470    & 2241    & 689     & x \\ % 		
+		&        &  {time (s)}                             & 297     & x       &  470    & 2241    & 689     & x \\ % 		
 		\midrule
 		\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
-		&        & {\bf run time (s)}                              & 18      & 109     &  26     & 160     & 37      & 203   \\ % 
+		&        & {time (s)}                              & 18      & 109     &  26     & 160     & 37      & 203   \\ % 
 		\cmidrule(lr){2-9}                              
 		& \multirow{2}{*}{2400}& $f$                               & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
-		&        &  {\bf run time (s)}                             & 18      & 107     &  25     & 159     & 37      & 192   \\ % 	
+		&        &  {time (s)}                             & 18      & 107     &  25     & 159     & 37      & 192   \\ % 	
 		\cmidrule(lr){2-9}                              
 		& \multirow{2}{*}{3600} & $f$                              & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ %  
-		&        &  {\bf run time (s)}                             & 17      & 106     &  26     & 157     & 37      & 181   \\ %  
-		\bottomrule
-
-		&                                       & {\bf Normalized} & 0.99    & x       & 0.99    & 0.98    & 1.00    & x\\
-		&                                       & {\bf Speed-up}   & 15.6    & x       & 14.3    & 14.3    & 16.3    & x\\
+		&        &  {time (s)}                             & 17      & 106     &  26     & 157     & 37      & 181   \\
 
@@
 
-\subsubsection{Proposed $S_K$ strategy assessment}
-
-	Another important test is to compare the best tour score/cost ratio with the least expensive tour score/cost ratio.
-This is important to see how far our solution strategy is from the simplest form, just solving a TSP.
-
-	For this achievement, we collected {\it Shims} results from tables \ref{tab:20}, \ref{tab:50}, and \ref{tab:100}, gathering the best tour results and the least cost results to assemble Table \ref{tab:20least}.
-
 \begin{table}[H]
 	\centering
-		\caption{{\it Shims} best and least cost tour results}  \label{tab:20least}
-		\footnotesize
-		\begin{tabular}{ccccccccc}
+	\caption{Solutions with different time limitations for $S_K$ tours.}  \label{tab:100time_2}
+%\footnotesize
+%	\tiny
+\scriptsize	
+	\setlength{\tabcolsep}{3.5pt}
+
+	\begin{tabular}{ccc|ccccc|ccccc|ccccc}
 		\toprule
-			{$surplus$}              & {\bf Tour} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4}  & {\bf 5}   & {\bf Sum} & {\bf Improvement}\\
+		&             \multicolumn{2}{r}{\bf Volume $surplus$} &\multicolumn{5}{c}{$1.2$}&\multicolumn{5}{c}{$1.5$}&\multicolumn{5}{c}{$2.0$} \\
+		\midrule
+		$method$& \specialcell{\bf limit\\ (s)} & {\bf Scenarios}  &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5} \\
 		\toprule
 
-			\multirow{2}{*}{$1.2$}   & Least cost    & 8.54    & 11.63   & 12.97   & 13.43  & 12.36 & 58.93 &       \\ 
-			& Best in $S_K$  & 8.54    & 12.26   & 13.23   & 14.53  & 13.42 & 61.98 & 5.2\%  \\ 
-			\midrule[.1pt]	
-			\multirow{2}{*}{$1.5$}   & Least cost    & 10.89   & 11.83   & 16.74   & 18.01  & 18.89 & 76.36 & \\ 	
-			& Best in $S_K$  & 11.33   & 11.83   & 17.00   & 18.21  & 20.45 & 78.82 &  3.2\%  \\ 
-			\midrule[.1pt]	
-			\multirow{2}{*}{$2.0$}   & Least cost    & 17.78   & 24.40   & 26.36   & 27.01  & 23.94 & 119.49 &  \\ 
-			& Best in $S_K$  & 17.78   & 25.25   & 26.38   & 28.98  & 26.08 & 124.47 & 4.2\%   \\ 		                         
-			\midrule[.1pt]
-			\multicolumn{8}{r}{\boldm{$S_K$} {\bf strategy advantage}}                                              & {\bf 4.2}\% \\
+		\multirow{10}{*}{\it Gurobi } & \multirow{2}{*}{240*} & $f$ & 8.53 & 12.24 & 13.34 &  x      &  x      & 11.81 & 17.02 & 18.31 &  x      &  x      & 17.75 & 25.24 & x     & x       & x \\ % 
+		&               &  {\bf time (s)}                              & 41   & 43    & 123   &  x      &  x      & 50    & 62    & 181   &  x      &  x      & 166   & 140   & x     & x       & x \\ % 
+		\cmidrule(lr){2-18}
+
+		&                              \multirow{2}{*}{1200} & $f$ & 8.52 & 12.25 & 13.31 & 14.67   & x       & 11.81 & 17.02 & 18.25 &  20.64  &  x      & 17.75 & 25.24 & 26.49 & 27.97   & x \\ % 
+		&        &  {\bf time (s)}                             & 39   & 41    & 129   & 217     & x       & 52    & 61    & 190   &  304    &  x      & 161   & 139   & 384   & 579     & x \\ % 
+		\cmidrule(lr){2-18}                              
+		& \multirow{2}{*}{2400}& $f$                               & 8.53 & 12.25 & 13.30 & 14.59   & 13.59   & 11.80 & 17.02 & 18.25 &  20.60  &  x      & 17.75 & 25.24 & 26.46 & 29.10   & x \\ % 
+		&        &  {\bf time (s)}                             & 33   & 39    & 127   & 313     & 1444    & 51    & 62    & 188   &  327    &  x      & 158   & 127   & 381   & 543     & x \\ % 
+		\cmidrule(lr){2-18}
+
+		& \multirow{2}{*}{3600}& $f$                               & 8.52 & 12.25 & 13.32 & 14.61   & x       & 11.81 & 17.01 & 18.25 &  20.59  & 18.36   & 17.75 & 25.24 & 26.46 & 29.11   & x \\ % 
+		&        &  {\bf time (s)}                             & 25   & 36    & 121   & 297     & x       & 53    & 61    & 193   &  470    & 2241    & 151   & 123   & 319   & 689     & x \\ %
+
+		\midrule
+
+		\multirow{2}{*}{\it Shims} & \multirow{2}{*}{240*}  & $f$ & 8.54 & 12.26 & 13.23 & 14.53   & 13.42   & 11.83 & 17.00 & 18.21 &  20.45  & 18.00   & 17.78 & 25.25 & 26.38 & 28.98   & 26.08 \\ % 
+		&         & {\bf time (s)}                             & 1    & 2     & 6     & 17      & 109     & 1     & 2     & 8     &  26     & 160     & 2     & 3     & 13    & 37      & 229   \\ %
 
 		\bottomrule
+		\multicolumn{16}{l}{*4 minutes is the time {\it Shims} needs to solve a seven-node tour.}\\
 	\end{tabular}
@@
 
-	As the selection of the best tour, considering what was collected and delivered in each node, and considering the cost increase due to CG deviation, was better than the option for a simple TSP solution (\textit{Least cost} results), this confirms that our solution strategy was adequate.
-
-\subsubsection{Final considerations on the results}
-
-It is important to report that the best tour found in each test was seldom the tour with the lowest cost (the minimal cost tour without considering the CG deviation), but any of the best tours.
-This indicates that our strategy of enumerating and solving all tours was assertive, although it is possible to devise a more proactive strategy to discard most of the higher-cost tours and improve performance.
-This must be one of the next steps in this research.
-If effective, this strategy may permit posterior applications for eight or more nodes.
-
-Most of the {\it Gurobi*} executions consumed over 12 GiB in total, and considering that the computer used in the experiments has a standard RAM consumption when idle of 3.5 GiB (with the Python Interface Development Environment (IDE), a PDF viewer, a system monitor, and a {\it LaTeX} IDE simultaneously open), the actual RAM consumption of {\it Gurobi*} was over 8.5 GiB.
-
-All {\it Shims} executions consumed at most 1.5 GiB, requiring 5 GiB of a computer total memory.
-These details are important for a future application to adequately dimension the machine to be used.
-
-}
+Most of $Gurobi$\/ executions consumed over 12 GB of RAM memory.
+Considering that the computer used in the experiments has a standard RAM consumption when idle of 3.5 GB (with {\it Python Interface Development Environment}, a PDF viewer, a system monitor, and a \LaTeX\/ IDE simultaneously open), the actual RAM consumption of $Gurobi$\/ was over 8.5 GB.
+On the other hand, all $Shims$\/ executions consumed at most 1.5 GB of RAM memory.
 
@@
 We consider that, in practical cases, the number $K$\/ of nodes, excluding the base, is small ($K \leq 6$), each of them with hundreds of items to be shipped.
-Thus, considering real aircraft model, we developed some node-by-node heuristics, which allow us to enumerate all $K!$\/ tours and choose the best.
+Thus, considering a real aircraft, we have developed some node-by-node solutions, which are not necessarily optimal.
+This allows us to test two strategies: find a solution using the shortest tours, or enumerate all $K!$\/ tours and choose the best.
 The complete process can be executed quickly on a simple handheld computer, offering good results and reducing stress for the transport planners.
 As validation, we performed tests in several scenarios with real data from the {\it Brazilian Air Force}.
-
-{\color{blue} In less than 20 minutes}, this development can establish a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximizing the total score and minimizing fuel consumption for the airlift.
+At the moment, there is no {\color{blue}commercial} software available for this problem.
+
+We have developed a solution process that, in less than 4 minutes, can establish a flight plan for a single aircraft with a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximizing the total score and minimizing fuel consumption.
 The output, which includes the tour plan and the pallet building and arrangement plan, is an essential part of airlift: it improves flight safety, makes ground operations more efficient, and makes sure that each item gets to its right destination.
-
-{\color{blue}
-
-Our main contributions were the mathematical modeling (1) of this complex problem, which involves four {\it NP-hard}\/ sub-problems, a complete process (2), and a new heuristic (3) that offer fast and good-quality solutions to the ACLP+RPDP.
-
-As this is ongoing research, some improvements are possible, which are the next steps of the development:
-
-(1) This work addresses all possible tours, but we plan bounding schemes to allow for discarding most of the higher-cost tours. This is expected to reduce the running time.
-
-(2) We are also working on computer parallelism to accelerate {\it Shims}.
-
-(3) In the event that these gains in performance are achieved, we strongly believe that they will permit the application of a 3-dimensional packing procedure to maximize the occupancy rate of the items selected by {\it Shims}.
-
-The $Shims$\/ heuristic presented a solution with the best node-by-node results in a restricted time interval.
-Without it, it would not be possible to test all routes: after all, the greater the number of nodes, the shorter the run time limit for node-by-node solution.
-We also show that, as the shipment volume increases, the exact node-by-node solution becomes unfeasible, indicating the importance of a fast and good-quality heuristic.
-
-Finally, by focusing on node-by-node solution, the method of this work is not exclusive to aircraft and airports: it can be adapted, for example, to ships and ports, or vehicles and warehouses, or freight cars and railways.
-In these cases, it would be necessary to make some changes in the modeling: for example, modify the load balancing constraints, and consider the available space in vehicles or freight cars instead of pallets.
-
-}
+In this way, at each node of the tour, the time required is restricted exclusively to handling the airport equipment.
+
+Our main contributions were the mathematical modeling of ACLP+RPDP that involves four sub-problems {\it NP-hard}\/, a complete process to solve it on a simple handheld computer, and a new heuristic that offers fast node solutions with good quality.
+Finally, by focusing on the node-to-node solution, the method of this work is not exclusive to aircraft and airports: it can be adapted, for example, to ships and ports, or vehicles and warehouses, or wagons and railways, provided that their practical cases are similar to those considered here.
+In these situations, it would be necessary to make some changes in the modeling: for example, modify the load balancing constraints, and consider the available space in vehicles or wagons instead of pallets.
+
+As this is an ongoing research, we thought about some possible future improvements:
+\begin{itemize}
+\item consider a fleet of planes, rather than a single one;
+\item model three-dimensional items;
+\item discard the highest cost tours, to reduce the runtime;
+\item implement parallel algorithms in some steps of the solution.
+\end{itemize}
 
@@
 
-	\bibitem{mipgap} Matthias Miltenberger. {\it What is the MIPGap?} \url{https://support.gurobi.com/hc/en-us/articles/8265539575953-What-is-the-MIPGap-}, 2023.
-
 \end{thebibliography}

### 2023-06-10 main.py

@@ -341,7 +341,7 @@ if __name__ == "__main__":
     # free -s 1 -h -c 3  memory
 
     # scenarios = [2,3,4,5,6]
-    scenarios = [4]
+    scenarios = [2,3]
 
     if testing:
         scenarios = [2]
@@ -351,8 +351,8 @@ if __name__ == "__main__":
     surplus = "data100" # 2.0
 
     # methods = ["GRB"]
-    methods = ["CBC"]
-    # methods = ["Shims"]
+    # methods = ["CBC"]
+    methods = ["Shims"]
     # methods = ["mpShims"]
     # methods = ["mpACO"]
     # methods = ["ACO"]

### 2023-06-10 CAIE/cie_sub.tex
Submitted inclusion

--- 2023-06-10 CAIE/CIE.tex
@@
 % !TeX spellcheck = en_US
-%\documentclass[preprint,authoryear]{elsarticle}
 \documentclass[preprint]{elsarticle}
-% sudo apt-get install -y texlive-publishers
 
@@
 In aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading, for fast take-off, and mission accomplishment, especially in times of crisis, disaster relief missions, short deadlines for contracting customers, or any external pressure for immediate take-off.
-{\color{blue} In addition, there are no commercially available systems that can assist load and trip planners with pallet building and aircraft-balanced loading plans, with demands for transport at each hub.} This enables other risks, such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
+In addition, there are no commercially available systems that can assist load and trip planners with pallet building and aircraft-balanced loading plans, with demands for transport at each hub.
+This enables other risks, such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
 We defined and solved the problem of planning the loading and routing of a single aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
-This NP-hard problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modeled using standardized pallets in fixed positions, obeying center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
+This {\it NP-hard}\/ problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modeled using standardized pallets in fixed positions, obeying center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
 We performed multiple experiments with a commercial solver and four well-known meta-heuristics on data based on the transport history of the {\it Brazilian Air Force}.
-{\color{blue}We also designed a new heuristic that quickly finds practical solutions for a wide range of problem sizes, an essential contribution that solved all test scenarios within a much shorter time than is operationally acceptable.}
+We also designed a new heuristic that quickly finds practical solutions to a wide range of problem sizes, a key contribution that resolved all real scenarios in much less time than is operationally acceptable.
 
@@
 	\centering
-	\includegraphics[scale=0.4]{Images/consolidated.jpg}
+	\includegraphics[scale=0.4]{Images/packed.jpg}
 	\caption{A packed content on a 463L pallet inside a {\it Boeing C-17}}
@@
 In this way, the loading and unloading time will be limited only by the use of the equipment at each node.
-Therefore, there will only be waiting time at the base node, corresponding to the run time of ACLP+RPDP algorithm.
+Therefore, there will only be waiting time at the base node, corresponding to the runtime of ACLP+RPDP algorithm.
 For this reason, our method will also be parameterized by the maximum time that can be waited until the start of the load in the base node.
@@
 
-		\Procedure{$ACLP+RPDP$}{$scenario, surplus, time$}
+		\Procedure{$ACLP+RPDP$}{$scenario, surplus, tmax$}
 
@@
 
-			\State $f_1 \gets SolveTour(\pi_{TSP1}, L, M, C, N, method, time)$  \label{main:f1}
-
-		    \State $f_2 \gets SolveTour(\pi_{TSP2}, L, M, C, N, method, time)$  \label{main:f2}
+			\State $f_1 \gets SolveTour(\pi_{TSP1}, L, M, C, N, method, tmax)$  \label{main:f1}
+
+		    \State $f_2 \gets SolveTour(\pi_{TSP2}, L, M, C, N, method, tmax)$  \label{main:f2}
 
@@
 
-				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, time/K!)$ \Comment{$f_{\pi}$\/ is a solution to the tour $\pi$} \label{main:fpi}
+				\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, tmax/K!)$ \Comment{$f_{\pi}$\/ is a solution to the tour $\pi$} \label{main:fpi}
 
@@
 
-$time$\/ is a run time limit, which will be distributed among the tours (lines \ref{main:f1}, \ref{main:f2} and \ref{main:fpi}).
+$tmax$\/ is a runtime limit, which will be distributed among the tours (lines \ref{main:f1}, \ref{main:f2} and \ref{main:fpi}).
 $method$\/ corresponds to a MIP solver or a heuristic to the node-by-node solution $SolveTour$, which will be presented in subsection \ref{methods}.
@@
 
-$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $time$\/ is the runtime limit that will be distributed among the $K+1$\/ legs of the tour.
+$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $tmax$\/ is the runtime limit that will be distributed among the $K+1$\/ legs of the tour.
 
@@
 
-		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method$, $time$}
+		\Procedure{$SolveTour$}{$\pi, L, M, C, N, method$, $tmax$}
 
@@
 
-		\State $G_2 \gets SolveNode(method, \pi_k, G_1, time/(K+1))$ \Comment{Each tour has $K+1$\/ legs} \label{tour:node}
+		\State $G_2 \gets SolveNode(method, \pi_k, G_1, tmax/(K+1))$ \Comment{Each tour has $K+1$\/ legs} \label{tour:node}
 
@@
 
-In this way, the mathematical modeling for $SolveNode(MIP,\pi_k,G, time)$\/ becomes simpler, which finds an allocation of available items at the node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$.
-Thus, we use a MIP solver with runtime limit $time$\/ at the node $\pi_k$\/ to maximize the objective function \ref{eq:obj2}, with the calculus equations \ref{eq:newf} to \ref{eq:costs2}, subject to the constraints \ref{eq:tau2} to \ref{eq:ifB}.
+In this way, the mathematical modeling for $SolveNode(MIP,\pi_k,G, tmax)$\/ becomes simpler, which finds an allocation of available items at the node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$.
+Thus, we use a MIP solver with runtime limit $tmax$\/ at the node $\pi_k$\/ to maximize the objective function \ref{eq:obj2}, with the calculus equations \ref{eq:newf} to \ref{eq:costs2}, subject to the constraints \ref{eq:tau2} to \ref{eq:ifB}.
 The binary variables $X_{ij}$\/ and $Y_{iq}$\/ define the sets of edges $E_{N_{\pi_k}}$\/ and $E_{Q_{\pi_k}}$, respectively, included in the graph $G$.
@@
 
-	\end{minipage}\hfill % these two lines must be close to each other
+	\end{minipage}\hfill
 	\begin{minipage}{0.34\linewidth}
@@
 
-		\Procedure{$SolveNode$}{$Shims, \pi_k, G, time$}
+		\Procedure{$SolveNode$}{$Shims, \pi_k, G, tmax$}
 
@@
 
-				\If{ $T_{current} - T_{begin} > time$}
-				\State {\bf break}  \Comment{Time limit exceeded}
+				\If{ $T_{current} - T_{begin} > tmax$}
+				\State {\bf break}  \Comment{Runtime limit exceeded}
 				\EndIf
@@
 
-We ran Algorithm \ref{alg:main} considering 5 values for $scenario$\/ from Table \ref{tab:scenarios}, 3 values for  $surplus$\/ from $\{1.2, 1.5, 2.0\}$, 3 values for $time$\/ from $\{1200s, 2400s, 3600s\}$, and 6 different $method$\/ for the node-by-node solution: $Gurobi$\/ (see \ref{solver}), ACO, NMO, TS, GRASP, and $Shims$\/ (Algorithm \ref{alg:shims}).
+We ran Algorithm \ref{alg:main} considering 5 values for $scenario$\/ from Table \ref{tab:scenarios}, 3 values for  $surplus$\/ from $\{1.2, 1.5, 2.0\}$, 4 values for $tmax$\/ from $\{240s, 1200s, 2400s, 3600s\}$, and 6 different $method$\/ for the node-by-node solution: $Gurobi$\/ (see \ref{solver}), ACO, NMO, TS, GRASP, and $Shims$\/ (Algorithm \ref{alg:shims}).
 
@@
 Among the heuristics, we will present only the results obtained by $Shims$, which had the best performance.
-For each $scenario$, $surplus$\/ and $time$, 7 different instances were generated.
+For each $scenario$, $surplus$\/ and $tmax$, 7 different instances were generated.
 We will present the average of the objective function $f$\/ and the runtime of $Gurobi$\/ and $Shims$\/ in these instances.
@@
 
-The results obtained with $time = 3600s$, which is the highest tested time limit, are in Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, with $surplus$\/ values $1.2$, $1.5$\/ and $2.0$, respectively.
-We indicate with an {\bf x} the cases where $Gurobi$\/ did not find a feasible solution within this time limit, or had to be aborted due to high RAM consumption.
+The results obtained with $tmax = 3600s$, which is the highest tested runtime limit, are in Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, with $surplus$\/ values $1.2$, $1.5$\/ and $2.0$, respectively.
+We indicate with an {\bf x} the cases where $Gurobi$\/ did not find a feasible solution within this runtime limit, or had to be aborted due to high RAM consumption.
 
@@
 	\centering
-	\caption{Solutions with $surplus = 1.2$\/ and $time =  3600s$}  \label{tab:20}
+	\caption{Solutions with $surplus = 1.2$\/ and $tmax =  3600s$}  \label{tab:20}
 	\footnotesize
@@
 		\toprule
-		{\specialcell{{\bf Tested}\\{\bf tours}}}&$method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		{\specialcell{{\bf Tested}\\{\bf tours}}}&$method$          & $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
 		\toprule
-		\multirow{4}{*}{Shortest} &\multirow{2}{*}{\it Gurobi } & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\ % 5927
-		&                                                               &               {time (s)}  & 17      & 17      & 16      & 17    & 26        &  1.0 \\ % 93
+		\multirow{4}{*}{Shortest} &\multirow{2}{*}{$Gurobi$} & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\
+		&                                                               &               {time (s)}  & 17      & 17      & 16      & 17    & 26        &  1.0 \\
 		\cmidrule(lr){2-9}
-		&                                                               \multirow{2}{*}{\it Shims}    & $f$ & 8.54    & 11.63   & 12.97   & 13.43 & 12.36     &  0.99 \\ % 5893
-		&                                                               &         {time (s)}        & 1       & 1       & 2       & 2     & 2         &  11.6 \\ % 8
+		&                                                               \multirow{2}{*}{$Shims$}    & $f$ & 8.54    & 11.63   & 12.97   & 13.43 & 12.36     &  0.99 \\
+		&                                                               &         {time (s)}        & 1       & 1       & 2       & 2     & 2         &  11.6 \\
 
@@
 
-		\multirow{4}{*}{All $K!$}&                                  \multirow{2}{*}{\it Gurobi } & $f$ & 8.52    & 12.25   & 13.32   & 14.61  & x        & 1.00 \\ % 4869
-		&                                                               &               {time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\ % 479
+		\multirow{4}{*}{All $K!$}&                                  \multirow{2}{*}{$Gurobi$} & $f$ & 8.52    & 12.25   & 13.32   & 14.61  & x        & 1.00 \\
+		&                                                               &               {time (s)}  & 25      & 36      & 121     & 297    & x        &  1.0 \\
 		\cmidrule(lr){2-9}	
-		&                                                               \multirow{2}{*}{\it Shims}    & $f$ & 8.54    & 12.26   & 13.23   & 14.53  & 13.42    & 0.99  \\ % 4813
-		&                                                               &         {time (s)}        & 1       & 2       & 6       & 17     & 106      & 18.4   \\ % 132
+		&                                                               \multirow{2}{*}{$Shims$}    & $f$ & 8.54    & 12.26   & 13.23   & 14.53  & 13.42    & 0.99  \\
+		&                                                               &         {time (s)}        & 1       & 2       & 6       & 17     & 106      & 18.4   \\
 		\bottomrule
@@
 	\centering
-	\caption{Solutions with $surplus = 1.5$\/ and $time =  3600s$}  \label{tab:50}
+	\caption{Solutions with $surplus = 1.5$\/ and $tmax =  3600s$}  \label{tab:50}
 	\footnotesize
@@
 		\toprule
-		{\specialcell{{\bf Tested}\\{\bf tours}}} & $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		{\specialcell{{\bf Tested}\\{\bf tours}}} & $method$          & $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
 		\toprule
-		\multirow{4}{*}{ Shortest } & \multirow{2}{*}{\it Gurobi } & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\ % 8236
-		&&               {time (s)}                                                                  & 38    & 30    & 27    & 29    & 81    & 1.00  \\ % 205
+		\multirow{4}{*}{ Shortest } & \multirow{2}{*}{$Gurobi$} & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\
+		&&               {time (s)}                                                                  & 38    & 30    & 27    & 29    & 81    & 1.00  \\
 		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{\it Shims}                                                                    & $f$ & 11.83 & 16.74 & 18.01 & 18.89 & 16.91 & 1.00  \\ % 
-		&&         {time (s)}                                                                        & 1     & 2     & 2     & 3     & 3     & 20.5  \\ %
+		&\multirow{2}{*}{$Shims$}                                                                    & $f$ & 11.83 & 16.74 & 18.01 & 18.89 & 16.91 & 1.00  \\
+		&&         {time (s)}                                                                        & 1     & 2     & 2     & 3     & 3     & 20.5  \\
 
@@
 
-		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{\it Gurobi }                                  & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ % 8602
-		&&               {time (s)}                                                                  & 53    & 61    & 193   & 470   & 2241    & 1.0  \\ % 3018
+		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$}                                  & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ 
+		&&               {time (s)}                                                                  & 53    & 61    & 193   & 470   & 2241    & 1.0  \\
 		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{\it Shims}                                                                    & $f$ & 11.83 & 17.00 & 18.21 & 20.45 & 18.00   & 0.99  \\ % 8549
-		&&         {time (s)}                                                                        & 1     & 2     & 9     & 26    & 157     & 15.5 \\ % 195
+		&\multirow{2}{*}{$Shims$}                                                                    & $f$ & 11.83 & 17.00 & 18.21 & 20.45 & 18.00   & 0.99  \\ 
+		&&         {time (s)}                                                                        & 1     & 2     & 9     & 26    & 157     & 15.5 \\
 		\bottomrule
@@
 	\centering
-	\caption{Solutions with $surplus = 2.0$\/ and $time =  3600s$}  \label{tab:100}
+	\caption{Solutions with $surplus = 2.0$\/ and $tmax =  3600s$}  \label{tab:100}
 	\footnotesize
@@
 		\toprule
-		{\specialcell{{\bf Tested}\\{\bf tours}}}& $method$          & {\bf Scenarios} & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		{\specialcell{{\bf Tested}\\{\bf tours}}}& $method$          & $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
 		\toprule
-		\multirow{4}{*}{ Shortest} & \multirow{2}{*}{\it Gurobi } & $f$ & 17.75 & 24.39 & 26.46 & 27.09 & 24.00 & 1.00 \\ % 11969
-		&&                                                                               {time (s)}  & 162   & 95    & 74    & 68    & 67    & 1.0  \\ % 466
+		\multirow{4}{*}{Shortest} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.75 & 24.39 & 26.46 & 27.09 & 24.00 & 1.00 \\ 
+		&&                                                                               {time (s)}  & 162   & 95    & 74    & 68    & 67    & 1.0  \\
 		\cmidrule(lr){2-9}	
-		&                                                                   \multirow{2}{*}{\it Shims} & $f$ & 17.78 & 24.40 & 26.36 & 27.01 & 23.94 & 0.99  \\ % 11949
-		&&                                                                                {time (s)} & 2     & 3     & 3     & 4     & 5     & 27.4  \\ % 17
+		&                                                                   \multirow{2}{*}{$Shims$} & $f$ & 17.78 & 24.40 & 26.36 & 27.01 & 23.94 & 0.99  \\
+		&&                                                                                {time (s)} & 2     & 3     & 3     & 4     & 5     & 27.4  \\
 
@@
 
-		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{\it Gurobi } & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &  x   & 1.00 \\ % 9856
-		&&                                                                                {time (s)} & 151   & 123   & 319   & 689   &  x   & 1.0  \\ % 1282
+		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &  x   & 1.00 \\ 
+		&&                                                                                {time (s)} & 151   & 123   & 319   & 689   &  x   & 1.0  \\
 		\cmidrule(lr){2-9}	
-		&                                                                \multirow{2}{*}{\it Shims}    & $f$ & 17.78 & 25.25 & 26.38 & 28.98 & 26.08 & 1.00  \\ % 9839
-		&&                                                                                {time (s)} & 2     & 3     & 13    & 37    & 229   & 23.3  \\ % 284
+		&                                                                \multirow{2}{*}{$Shims$}    & $f$ & 17.78 & 25.25 & 26.38 & 28.98 & 26.08 & 1.00  \\
+		&&                                                                                {time (s)} & 2     & 3     & 13    & 37    & 229   & 23.3  \\
 		\bottomrule
@@
 \item $Gurobi$\/ fails in some cases, when $scenario=5$\/ and the strategy is to check all $K!$\/ tours.
-This occurs because the time limit per node is smaller, and there tends to be more packed contents on the aircraft, reducing the space for allocating items and making the solution difficult.
+This occurs because the runtime limit per node is smaller, and there tends to be more packed contents on the aircraft, reducing the space for allocating items and making the solution difficult.
 \item When $Gurobi$\/ finishes, it finds the best solution, but the one obtained by $Shims$\/ reaches $99\%$\/ of that value.
 \item $Shims$\/ always finds a solution, being 11 to 27 times faster.
-\item All runtimes are well under the time limit because the solution on many nodes can be fast.
+\item All runtimes are much lower than the limit because the solution on many nodes can be fast.
 Anyway, in all the tests performed, the maximum time spent by $Shims$\/ did not reach 4 minutes, that is, it is practically instantaneous.
@@
 
-Table \ref{tab:100time} shows the results obtained with the strategy of testing the $K!$\/ tours in scenarios 4 and 5, with different time limits.
-We can observe more cases where $Gurobi$\/ fails.
-When $Gurobi$ finishes, $Shims$\/ finds a solution of similar quality in much less time.
+Table \ref{tab:100time} shows the results obtained with the strategy of testing the $K!$\/ tours in all scenarios with different runtime limits.
+We can observe more cases where $Gurobi$\/ fails, even in smaller scenarios.
+When $Gurobi$\/ finishes, $Shims$\/ finds quickly a solution of similar quality ($98\%$\/ or better).
+In all cases, $Shims$\/ finds a good solution in less than 4 minutes.
 
@@
 	\centering
-	\caption{Solutions in scenarios 4 and 5 testing all $K!$\/ tours}  \label{tab:100time}
-	\footnotesize
-	\begin{tabular}{ccc|cc|cc|cc}
-		\toprule
-		&             \multicolumn{2}{c}{$surplus$} &\multicolumn{2}{c}{$1.2$}&\multicolumn{2}{c}{$1.5$}&\multicolumn{2}{c}{$2.0$} \\
-		\midrule
-		$method$& {\specialcell{$time$ (s)}} & {\bf Scenarios}    & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} & {\bf 4} & {\bf 5} \\
-		\toprule
-
-		\multirow{6}{*}{\it Gurobi }& \multirow{2}{*}{1200} & $f$ & 14.67   & x       &  20.64  &  x      & 27.97   & x \\ % 
-		&        &  {time (s)}                             & 217     & x       &  304    &  x      & 579     & x \\ % 
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$                               & 14.59   & 13.59   &  20.60  &  x      & 29.10   & x \\ % 
-		&        &  {time (s)}                             & 313     & 1444    &  327    &  x      & 543     & x \\ % 
-		\cmidrule(lr){2-9}                              		
-		& \multirow{2}{*}{3600}& $f$                               & 14.61   & x       &  20.59  & 18.36   & 29.11   & x \\ % 
-		&        &  {time (s)}                             & 297     & x       &  470    & 2241    & 689     & x \\ % 		
-		\midrule
-		\multirow{6}{*}{\it Shims}   & \multirow{2}{*}{1200} & $f$ & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
-		&        & {time (s)}                              & 18      & 109     &  26     & 160     & 37      & 203   \\ % 
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{2400}& $f$                               & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ % 
-		&        &  {time (s)}                             & 18      & 107     &  25     & 159     & 37      & 192   \\ % 	
-		\cmidrule(lr){2-9}                              
-		& \multirow{2}{*}{3600} & $f$                              & 14.53   & 13.42   &  20.45  & 18.00   & 28.98   & 26.08 \\ %  
-		&        &  {time (s)}                             & 17      & 106     &  26     & 157     & 37      & 181   \\
-
-		\bottomrule
-
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\begin{table}[H]
-	\centering
-	\caption{Solutions with different time limitations for $S_K$ tours.}  \label{tab:100time_2}
-%\footnotesize
-%	\tiny
+	\caption{Solutions testing all $K!$\/ tours with different runtime limits}  \label{tab:100time}
 	\scriptsize
 	\setlength{\tabcolsep}{3.5pt}
-
 	\begin{tabular}{ccc|ccccc|ccccc|ccccc}
 		\toprule
-		&             \multicolumn{2}{r}{\bf Volume $surplus$} &\multicolumn{5}{c}{$1.2$}&\multicolumn{5}{c}{$1.5$}&\multicolumn{5}{c}{$2.0$} \\
+		&             \multicolumn{2}{r}{$surplus$} &\multicolumn{5}{c}{\bf 1.2}&\multicolumn{5}{c}{\bf 1.5}&\multicolumn{5}{c}{\bf 2.0} \\
 		\midrule
-		$method$& \specialcell{\bf limit\\ (s)} & {\bf Scenarios}  &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5} \\
+		$method$& {$tmax$} & $scenario$  &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5} \\
 		\toprule
 
-		\multirow{10}{*}{\it Gurobi } & \multirow{2}{*}{240*} & $f$ & 8.53 & 12.24 & 13.34 &  x      &  x      & 11.81 & 17.02 & 18.31 &  x      &  x      & 17.75 & 25.24 & x     & x       & x \\ % 
-		&               &  {\bf time (s)}                              & 41   & 43    & 123   &  x      &  x      & 50    & 62    & 181   &  x      &  x      & 166   & 140   & x     & x       & x \\ % 
+		\multirow{10}{*}{$Gurobi$} & \multirow{2}{*}{240s} & $f$ & 8.53 & 12.24 & 13.34 &  x      &  x      & 11.81 & 17.02 & 18.31 &  x      &  x      & 17.75 & 25.24 & x     & x       & x \\ 
+		&               &  {time (s)}                              & 41   & 43    & 123   &  x      &  x      & 50    & 62    & 181   &  x      &  x      & 166   & 140   & x     & x       & x \\ 
 		\cmidrule(lr){2-18}
 
-		&                              \multirow{2}{*}{1200} & $f$ & 8.52 & 12.25 & 13.31 & 14.67   & x       & 11.81 & 17.02 & 18.25 &  20.64  &  x      & 17.75 & 25.24 & 26.49 & 27.97   & x \\ % 
-		&        &  {\bf time (s)}                             & 39   & 41    & 129   & 217     & x       & 52    & 61    & 190   &  304    &  x      & 161   & 139   & 384   & 579     & x \\ % 
+		&                              \multirow{2}{*}{1200s} & $f$ & 8.52 & 12.25 & 13.31 & 14.67   & x       & 11.81 & 17.02 & 18.25 &  20.64  &  x      & 17.75 & 25.24 & 26.49 & 27.97   & x \\ 
+		&        &  {time (s)}                             & 39   & 41    & 129   & 217     & x       & 52    & 61    & 190   &  304    &  x      & 161   & 139   & 384   & 579     & x \\ 
 		\cmidrule(lr){2-18}                              
-		& \multirow{2}{*}{2400}& $f$                               & 8.53 & 12.25 & 13.30 & 14.59   & 13.59   & 11.80 & 17.02 & 18.25 &  20.60  &  x      & 17.75 & 25.24 & 26.46 & 29.10   & x \\ % 
-		&        &  {\bf time (s)}                             & 33   & 39    & 127   & 313     & 1444    & 51    & 62    & 188   &  327    &  x      & 158   & 127   & 381   & 543     & x \\ % 
+		& \multirow{2}{*}{2400s}& $f$                               & 8.53 & 12.25 & 13.30 & 14.59   & 13.59   & 11.80 & 17.02 & 18.25 &  20.60  &  x      & 17.75 & 25.24 & 26.46 & 29.10   & x \\
+		&        &  {time (s)}                             & 33   & 39    & 127   & 313     & 1444    & 51    & 62    & 188   &  327    &  x      & 158   & 127   & 381   & 543     & x \\
 		\cmidrule(lr){2-18}
 
-		& \multirow{2}{*}{3600}& $f$                               & 8.52 & 12.25 & 13.32 & 14.61   & x       & 11.81 & 17.01 & 18.25 &  20.59  & 18.36   & 17.75 & 25.24 & 26.46 & 29.11   & x \\ % 
-		&        &  {\bf time (s)}                             & 25   & 36    & 121   & 297     & x       & 53    & 61    & 193   &  470    & 2241    & 151   & 123   & 319   & 689     & x \\ %
+		& \multirow{2}{*}{3600s}& $f$                               & 8.52 & 12.25 & 13.32 & 14.61   & x       & 11.81 & 17.01 & 18.25 &  20.59  & 18.36   & 17.75 & 25.24 & 26.46 & 29.11   & x \\ 
+		&        &  {time (s)}                             & 25   & 36    & 121   & 297     & x       & 53    & 61    & 193   &  470    & 2241    & 151   & 123   & 319   & 689     & x \\
 
@@
 
-		\multirow{2}{*}{\it Shims} & \multirow{2}{*}{240*}  & $f$ & 8.54 & 12.26 & 13.23 & 14.53   & 13.42   & 11.83 & 17.00 & 18.21 &  20.45  & 18.00   & 17.78 & 25.25 & 26.38 & 28.98   & 26.08 \\ % 
-		&         & {\bf time (s)}                             & 1    & 2     & 6     & 17      & 109     & 1     & 2     & 8     &  26     & 160     & 2     & 3     & 13    & 37      & 229   \\ %
+		\multirow{2}{*}{$Shims$} & \multirow{2}{*}{240s}  & $f$ & 8.54 & 12.26 & 13.23 & 14.53   & 13.42   & 11.83 & 17.00 & 18.21 &  20.45  & 18.00   & 17.78 & 25.25 & 26.38 & 28.98   & 26.08 \\ % 
+		&         & {time (s)}                             & 1    & 2     & 6     & 17      & 109     & 1     & 2     & 8     &  26     & 160     & 2     & 3     & 13    & 37      & 229   \\
 
 		\bottomrule
-		\multicolumn{16}{l}{*4 minutes is the time {\it Shims} needs to solve a seven-node tour.}\\
 	\end{tabular}
@@
 As validation, we performed tests in several scenarios with real data from the {\it Brazilian Air Force}.
-At the moment, there is no {\color{blue}commercial} software available for this problem.
+At the moment, there is no commercial software available for this problem.
 
@@
 \begin{itemize}
-\item consider a fleet of planes, rather than a single one;
+\item consider a aircraft fleet, rather than a single one;
 \item model three-dimensional items;

### 2023-07-19 CAIE/caie.tex
CAIE new submission

--- 2023-06-10 CAIE/cie_sub.tex
@@
 \usepackage{color, colortbl}
+\usepackage{setspace} \doublespacing
 \definecolor{Gray}{gray}{0.9}
@@
 
-\author{A.C.P.~Mesquita}
-\ead{celio@ita.br}
-
-\author{C.A.A.~Sanches\corref{cor1}}
-\ead{alonso@ita.br}
-\cortext[cor1]{Corresponding author.}
-
-\address {Instituto Tecnol\'{o}gico de Aeron\'{a}utica - DCTA/ITA/IEC\\
-Pra\c{c}a Mal. Eduardo Gomes, 50\\
-S\~{a}o Jos\'{e} dos Campos - SP - 12.228-900 - Brazil}
+%\author{A.C.P.~Mesquita}
+%\ead{celio@ita.br}
+
+%\author{C.A.A.~Sanches\corref{cor1}}
+%\ead{alonso@ita.br}
+%\cortext[cor1]{Corresponding author.}
+
+%\address {Instituto Tecnol\'{o}gico de Aeron\'{a}utica - DCTA/ITA/IEC\\
+%Pra\c{c}a Mal. Eduardo Gomes, 50\\
+%S\~{a}o Jos\'{e} dos Campos - SP - 12.228-900 - Brazil}
 
@@
 
-In aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading, for fast take-off, and mission accomplishment, especially in times of crisis, disaster relief missions, short deadlines for contracting customers, or any external pressure for immediate take-off.
+{\color{blue}In aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading, immediate take-off, and mission accomplishment, especially in disaster relief, short deadlines, or any pressure for immediate take-off.}
 In addition, there are no commercially available systems that can assist load and trip planners with pallet building and aircraft-balanced loading plans, with demands for transport at each hub.
@@
 
-Considering air cargo transport, Table \ref{tab:sa} lists the main works in the literature and the corresponding sub-problems addressed.
+{\color{blue} Lurkin and Schyns \cite{LurkinSchyns2015} show that this problem is NP-hard and therefore compare their results to current manually generated solutions vice optimal (or even near optimal) solutions.
+They have shown that the WBP is also NP-hard.}
+
+{\color{blue} Karp \cite{Karp1972} previously proved the 0-1 Knapsack problems to be NP-Hard, as the pallets may be seen as multiple knapsacks, the APP is also NP-hard. }
+
+{\color{blue} Feng {\it et al.} (2015) \cite{Feng2015} presented a comprehensive literature assessment of commercial air cargo operations.
+They stated that the airplane loading problem is one of the fundamental issues in combinatorial optimization and described it as an NP-hard problem.}
+
+There are real situations that are much more complex.
+In this work, we consider a practical case in Brazil, which is the largest economy in Latin America.
+Due to its dimensions, this country has the largest air market on the continent with $2,499$\/ registered airports, of which $1,911$\/ are private and $588$\/ are public.
+Although it is an immense distribution network, the {\it Brazilian Air Force}\/ missions have always considered 3 to 5 nodes per flight plan.
+
+It is important to emphasize that this data is not an imposed limitation, but a historical fact that we will explore in our solution method.
+Throughout this work we approach routes with up to 7 nodes, as can be seen in Table \ref{tab:costs} and in Figure \ref{fig:nodes}.
+Although there are many other airports of interest to {\it Brazilian Air Force}, these 7 nodes were chosen due to their high demand and short transport times.
+The other Brazilian airports tend to have a lower demand, which is usually met, in a less expensive way, by cabotage, rail or road transport.
+
+\begin{table}[H]
+
+	\begin{minipage}{0.05\linewidth}
+
+	\end{minipage}\hfill 
+	\begin{minipage}{0.45\linewidth}
+
+		\caption{Brazilian airports distances ($km$)}  \label{tab:costs}
+		\centering
+		\footnotesize
+		\newcolumntype{Y}{>{\centering\arraybackslash}p{0.09\textwidth}}
+		\newcolumntype{X}{>{\centering}p{0.09\textwidth}}
+
+		\begin{tabular}{X X X X X X X Y}
+			\toprule
+			Node & $l_0$ & $l_1$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ & $l_6$ \\
+			IATA*   & GRU   & GIG   & SSA   & CNF   & CWB   & BSB   & REC \\	
+			\midrule	
+			GRU     & 0	    &343	&1,439   &504    &358    &866    &2,114\\
+			GIG	    & 343	&0	    &1,218   &371    &677    &935    &1,876\\
+			SSA	    & 1,439	&1,218	&0	    &938    &1,788   &1,062   &676\\
+			CNF	    & 504	&371	&938	&0	    &851    &606    &1,613\\
+			CWB	    & 358	&677	&1,788	&851	&0	    &1,084   &2,462\\
+			BSB	    & 866	&935	&1,062	&606	&1,084	&0	    &1,658\\
+			REC	    & 2,114	&1,876	&676	&1,613	&2,462	&1,658	&0\\
+			\bottomrule
+			\multicolumn{8}{c}{*International Air Transport Association}\\
+			\multicolumn{8}{c}{\small\textsuperscript{Source: www.airportdistancecalculator.com}}\\
+		\end{tabular}
+		\normalsize
+
+	\end{minipage}\hfill 
+	\begin{minipage}{0.50\linewidth}
+		\centering
+		\includegraphics[scale=0.25]{Images/nodes.png}
+		\captionof{figure}{A route with 7 Brazilian airports}
+		\label{fig:nodes}
+	\end{minipage}
+\end{table}
+
+Cargo handling at modern airports usually involves powerful equipment such as dollies, motorized conveyors, elevator transfer vehicles, caster decking, scissor lifts, electrical controls, hydraulically adjustable heights, forklift trucks, and plenty of room in the airport cargo terminal.
+These facilities, together with the use of standardized pallets with predefined positions on the aircraft, allow loading and unloading at each node to be carried out in about 30 minutes.
+
+However, as reported by Fok and Chun, {\it load planning (...) is usually done in roughly 2 hours before departure, when all the details of the cargo are present.} \cite{fok2004optimizing}.
+We can infer that this is because adequate commercial software does not exist.
+In addition, there is still the necessary time to plan the route, in order to minimize fuel consumption and ensure the transport of priority items.
+In these circumstances, it is of great importance to find a method that speeds up the load planning and the definition of the flight itinerary.
+
+Our work proposes a method that prioritizes the transport of the most relevant items at each node and the fuel economy along the tour.
+We developed a heuristic that can be run on a simple handheld computer (such as a laptop or a tablet) that previously defines a flight plan, and provides a quick solution for cargo handling.
+Consequently, this method reduces the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets (regarding their positions), and a pick up and delivery plan to each node.
+
+To the best of our knowledge, this is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP and TSP has been addressed.
+This new problem is named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem} (ACLP+RPDP).
+
+This article is organized into six more sections.
+In Section \ref{sec2}, we give a brief review of the literature.
+In Section \ref{sec3}, we present the problem context and assumptions, and in Section \ref{sec4}, we describe the mathematical model and how we dealt with its issues.
+In Section \ref{sec5}, we describe the elaborate algorithms, whose results are presented in Section \ref{sec6}.
+Finally, our conclusions are in Section \ref{sec7}.
+
+\section{Related literature}
+\label{sec2}
+
+{\color{blue} In this section, we briefly describe the characteristics of the main works related to air cargo transport, following the chronological order of Table \ref{tab:sa}, that lists the main works in the literature and the corresponding sub-problems addressed.
 We also indicate whether the dimensions of the items were taken into account ({\bf 3D} or {\bf 2D}) and which solution method was used: heuristic search methods ({\bf Heu}), integer programming ({\bf Int}), or linear programming ({\bf Lin}).
+}
 
@@
 		Zhao {\it et al.} \cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
+
+		{\color{blue}		Zhao {\it et al.} } \cite{zhao2023} & $\bigstar$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$  \\
+
 		{\bf This article}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$  \\
@@
 
-There are real situations that are much more complex.
-In this work, we consider a practical case in Brazil, which is the largest economy in Latin America.
-Due to its dimensions, this country has the largest air market on the continent with $2,499$\/ registered airports, of which $1,911$\/ are private and $588$\/ are public.
-Although it is an immense distribution network, the {\it Brazilian Air Force}\/ missions have always considered 3 to 5 nodes per flight plan.
-
-It is important to emphasize that this data is not an imposed limitation, but a historical fact that we will explore in our solution method.
-Throughout this work we approach routes with up to 7 nodes, as can be seen in Table \ref{tab:costs} and in Figure \ref{fig:nodes}.
-Although there are many other airports of interest to {\it Brazilian Air Force}, these 7 nodes were chosen due to their high demand and short transport times.
-The other Brazilian airports tend to have a lower demand, which is usually met, in a less expensive way, by cabotage, rail or road transport.
-
-\begin{table}[H]
-
-	\begin{minipage}{0.05\linewidth}
-
-	\end{minipage}\hfill 
-	\begin{minipage}{0.45\linewidth}
-
-		\caption{Brazilian airports distances ($km$)}  \label{tab:costs}
-		\centering
-		\footnotesize
-		\newcolumntype{Y}{>{\centering\arraybackslash}p{0.09\textwidth}}
-		\newcolumntype{X}{>{\centering}p{0.09\textwidth}}
-
-		\begin{tabular}{X X X X X X X Y}
-			\toprule
-			Node & $l_0$ & $l_1$ & $l_2$ & $l_3$ & $l_4$ & $l_5$ & $l_6$ \\
-			IATA*   & GRU   & GIG   & SSA   & CNF   & CWB   & BSB   & REC \\	
-			\midrule	
-			GRU     & 0	    &343	&1,439   &504    &358    &866    &2,114\\
-			GIG	    & 343	&0	    &1,218   &371    &677    &935    &1,876\\
-			SSA	    & 1,439	&1,218	&0	    &938    &1,788   &1,062   &676\\
-			CNF	    & 504	&371	&938	&0	    &851    &606    &1,613\\
-			CWB	    & 358	&677	&1,788	&851	&0	    &1,084   &2,462\\
-			BSB	    & 866	&935	&1,062	&606	&1,084	&0	    &1,658\\
-			REC	    & 2,114	&1,876	&676	&1,613	&2,462	&1,658	&0\\
-			\bottomrule
-			\multicolumn{8}{c}{*International Air Transport Association}\\
-			\multicolumn{8}{c}{\small\textsuperscript{Source: www.airportdistancecalculator.com}}\\
-		\end{tabular}
-		\normalsize
-
-	\end{minipage}\hfill 
-	\begin{minipage}{0.50\linewidth}
-		\centering
-		\includegraphics[scale=0.25]{Images/nodes.png}
-		\captionof{figure}{A route with 7 Brazilian airports}
-		\label{fig:nodes}
-	\end{minipage}
-\end{table}
-
-Cargo handling at modern airports usually involves powerful equipment such as dollies, motorized conveyors, elevator transfer vehicles, caster decking, scissor lifts, electrical controls, hydraulically adjustable heights, forklift trucks, and plenty of room in the airport cargo terminal (see Figure \ref{fig:handling}).
-These facilities, together with the use of standardized pallets with predefined positions on the aircraft, allow loading and unloading at each node to be carried out in about 30 minutes.
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[scale=0.2]{Images/handling.png}
-	\captionof{figure}{Cargo handling equipment}
-	\label{fig:handling}
-	\small\textsuperscript{Source: NARA \& DVIDS Public Domain Archive}	
-\end{figure}
-
-However, as reported by Fok and Chun, {\it load planning (...) is usually done in roughly 2 hours before departure, when all the details of the cargo are present.} \cite{fok2004optimizing}.
-We can infer that this is because adequate commercial software does not exist.
-In addition, there is still the necessary time to plan the route, in order to minimize fuel consumption and ensure the transport of priority items.
-In these circumstances, it is of great importance to find a method that speeds up the load planning and the definition of the flight itinerary.
-
-Our work proposes a method that prioritizes the transport of the most relevant items at each node and the fuel economy along the tour.
-We developed a heuristic that can be run on a simple handheld computer (such as a laptop or a tablet) that previously defines a flight plan, and provides a quick solution for cargo handling.
-Consequently, this method reduces the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets (regarding their positions), and a pick up and delivery plan to each node.
-
-To the best of our knowledge, this is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP and TSP has been addressed.
-This new problem is named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem} (ACLP+RPDP).
-
-This article is organized into six more sections.
-In Section \ref{sec2}, we give a brief review of the literature.
-In Section \ref{sec3}, we present the problem context and assumptions, and in Section \ref{sec4}, we describe the mathematical model and how we dealt with its issues.
-In Section \ref{sec5}, we describe the elaborate algorithms, whose results are presented in Section \ref{sec6}.
-Finally, our conclusions are in Section \ref{sec7}.
-
-\section{Related literature}
-\label{sec2}
-
-In this section, we briefly describe the characteristics of the main works related to air cargo transport, following the chronological order of Table \ref{tab:sa}.
-
 Larsen and Mikkelsen \cite{LarsenMikkelsen1979} developed an interactive procedure for loading 14 types of Boeing 747 into a two-leg transportation plan.
@@
 Instead of focusing on the CG deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
+
+{\color{blue} Zhao {\it et al.} \cite{zhao2023}  proposed three integer linear programming models: a Bi-objective Optimization Model (BOM), a Combinatorial Optimization Model (COM), and an Improved Combinatorial Optimization Model (IOM).
+The objectives of their solutions were the maximum loading capacity and the lowest CG deviation from a specified target CG.
+Four scenarios with various conditional metrics for three models are solved for the B777F aircraft using Gurobi.
+The results of the computations demonstrated that the BOM has the fastest solution speed, but the CG deviation is the largest, and in several cases, the CG deviation results are unacceptable.
+The COM has the longest solution time, which is difficult to tolerate in practice.
+Despite taking a little longer to solve computationally than the BOM, the IOM offers the best optimization solution.}
 
@@
 
-Throughout this text, we call {\it packed content}\/ a set of items of same destination stacked on a pallet and covered with a restraining net (see Figure \ref{fig:larger2}).
-It is considered a single item, having the same attributes of its components, whose values are the sum of individual scores, weights and volumes.
-To ensure accuracy in pickup and delivery operations, a packed content must remain on board until its destination.
-
-\subsection{Aircraft parameters and load balancing}
-
-We consider real scenarios with an single aircraft with payload of $75,000 kg$, with enough fuel for a range of $4,400km$.
-The aircraft layout is presented in Figure \ref{fig:larger}, where the pallets positions are identified by $p_i$, $1\leq i \leq 18$.
-
 \begin{figure}[H]
 	\centering
-	\includegraphics[scale=0.4]{Images/packed.jpg}
+	\includegraphics[scale=0.3]{Images/packed.jpg}
 	\caption{A packed content on a 463L pallet inside a {\it Boeing C-17}}
@@
 \end{figure}
+
+Throughout this text, we call {\it packed content}\/ a set of items of same destination stacked on a pallet and covered with a restraining net (see Figure \ref{fig:larger2}).
+It is considered a single item, having the same attributes of its components, whose values are the sum of individual scores, weights and volumes.
+To ensure accuracy in pickup and delivery operations, a packed content must remain on board until its destination.
+
+\subsection{Aircraft parameters and load balancing}
+
+We consider real scenarios with an single aircraft with payload of $75,000 kg$, with enough fuel for a range of $4,400km$.
+The aircraft layout is presented in Figure \ref{fig:larger}, where the pallets positions are identified by $p_i$, $1\leq i \leq 18$.
 
@@
 By definition, $m_0 = 0$, and therefore $Q_0 = \varnothing$.
-Packets contents that were destined to node $k$\/ are unloaded when the aircraft arrives at this node, that is, they are not considered in $Q_k$.
+Packed contents that were destined to node $k$\/ are unloaded when the aircraft arrives at this node, that is, they are not considered in $Q_k$.
 
@@
 Once the assumptions of this work and the mathematical modeling of the problem are presented, it is possible to see that ACLP+RPDP is {\it NP-hard}.
-In a similar way to Lurkin and Schyns \cite[p.
+
+ {\color{red}  In a similar way to Lurkin and Schyns \cite[p.
 6]{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$\/ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in node 0, and no items in node 1.
-Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is {\it NP-complete}.
+Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is {\it NP-complete}.}
 
@@
 \begin{itemize}
-\item The strategy of testing all $K!$\/ often gives a better quality solution.
+\item The strategy of testing all $K!$\/ tours often gives a better quality solution.
 \item $Gurobi$\/ fails in some cases, when $scenario=5$\/ and the strategy is to check all $K!$\/ tours.
@@
 
-\section*{Acknowledgments}
-
-This research was partially supported by \textit{São Paulo Research Foundation} (FAPESP, grant 2016/01860-1).
+\section*{References}
 
 \begin{thebibliography}{00}
+
+\bibitem{Karp1972} R. Karp, {\it Reducibility among combinatorial problems. Complexity of Computer Computations.} (R.E. Miller and J.M. Thatcher, eds.), Plenum Press, 85-103 (1972).
 
@@
 
+\bibitem{Feng2015} Feng, B.; Li, Y.; Shen, Z.J.M. {\it Air cargo operations: Literature review and comparison with practices.} Transp. Res. Part C Emerg. Technol. 2015, 56, 263?280.
+
 \bibitem{RoesenerBarnes2016} A.G. Roesener and J.W. Barnes, An advanced tabu search approach to the dynamic airlift loading problem, {\it Logistics Research}, Vol. 9 (1), pp. 1-18, 2016.
@@
 1269-1286, 2021.
+
+\bibitem{zhao2023} Zhao, Xiangling, Yun Dong, and Lei Zuo. {\it A Combinatorial Optimization Approach for Air Cargo Palletization and Aircraft Loading.} Mathematics (Basel) 11.13 (2023): 2798. Web.
 

### 2023-07-20 CAIE/CIE.tex
mpShims tests on CIE.tex
May be discarded

@@
 \usepackage{textcomp}
+
+ \setlength{\fboxsep}{0.005pt}
+\newcommand{\tmpframe}[1]{\fbox{#1}}
 
@@
 
-Wong and Ling \cite{wong2020}A developed a mathematical model and a tool based on mixed integer programming for optimizing cargo in aircraft with different pallet configurations.
+Wong and Ling \cite{wong2020}A?developed a mathematical model and a tool based on mixed integer programming for optimizing cargo in aircraft with different pallet configurations.
 Balance constraints and the presence of dangerous items were considered.
@@
 
+		\multirow{2}{*}{\it mpShims} & \multirow{2}{*}{240*}  & $f$ & 8.32 & 12.14 & 13.19 & 14.60   & 13.52  & 11.57 & 16.86 & 18.21 & 20.52   & 18.06  & 17.58 & 25.21 & 26.46 & 29.11   & 26.14 \\ % 
+&         & {\bf time (s)}                                          & 1    & 2     & 5     & 13      & 71     & 1     & 2     & 6     & 15      & 79     & 2     & 2     & 8     & 21      & 102   \\ %
+
 		\bottomrule
@@
 \end{table}
+
+\begin{figure}[H]
+	\centering
+	\tmpframe{\includegraphics[scale=0.5]{Images/mpShimsResouces.png}}
+	\captionof{figure}{{\it mpShims} resources consumption}
+	\label{fig:mpShimsResouces}
+\end{figure}
 
@@
 
-This research was partially supported by \textit{São Paulo Research Foundation} (FAPESP, grant 2016/01860-1).
+This research was partially supported by \textit{S?o Paulo Research Foundation} (FAPESP, grant 2016/01860-1).
 
@@
 \bibitem{LopezIbanezManuel2016} M.
-López-Ibáñez, J.
+L?pez-Ib??ez, J.
 Dubois-Lacoste, L.P.
-Cáceres, M.
+C?ceres, M.
 Birattari and T.
-Stützle, The irace package: Iterated racing for automatic algorithm configuration, {\it Operations Research Perspectives}, Vol.
+St?tzle, The irace package: Iterated racing for automatic algorithm configuration, {\it Operations Research Perspectives}, Vol.
 3, pp.

### 2023-07-20 CAIE/caie.tex

@@
 
-{\color{blue}In aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading, immediate take-off, and mission accomplishment, especially in disaster relief, short deadlines, or any pressure for immediate take-off.}
+{\color{blue}In the aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of cargo unbalancing due to the urgency required for loading, immediate take-off, and mission accomplishment.
+It is especially critical for disaster relief, short deadlines, or any pressure for immediate takeoff.
 In addition, there are no commercially available systems that can assist load and trip planners with pallet building and aircraft-balanced loading plans, with demands for transport at each hub.
-This enables other risks, such as improper delivery, excessive fuel burn, and longer than necessary turn-around time.
-We defined and solved the problem of planning the loading and routing of a single aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery at intermediate hubs.
+This enables other risks, such as improper delivery, excessive fuel burn, and a longer than necessary turn-around time.
+We defined and solved the problem of planning the loading and routing of a single aircraft according to a utility score, weight and balance principles, and fuel consumption in a tour of simultaneous pickup and delivery.}
 This {\it NP-hard}\/ problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modeled using standardized pallets in fixed positions, obeying center of gravity constraints, delivering each item to its destination, and minimizing fuel consumption costs.
@@
 Recently, Brandt and Nickel \cite[p.
-401]{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP) and {\it Weight and Balance Problem} (WBP).
-Several aspects were considered in this modeling: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc.
+401]{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: the {\it Aircraft Configuration Problem} (ACP), the {\it Build-up Scheduling Problem} (BSP), the {\it Air Cargo Palletization Problem} (APP), and the {\it Weight and Balance Problem} (WBP).
+
+Several aspects were considered in these authors' model: characteristics of the items to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs), commonly called pallets; when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how the total cargo weight is balanced; etc.
 They also presented a comprehensive bibliographic survey of solving methods that had been developed in different situations.
@@
 However, there are still other important challenges in air cargo transport that go beyond the definition of the ACLPP, especially with regard to the flight itinerary and the loading and unloading at each destination (or node) of this travel plan.
-In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Pickup and Delivery Problem} (PDP), and the search for the lowest cost route, which is the well-known {\it Traveling Salesman Problem} (TSP).
+In this context, at least two more important sub-problems can be considered: pickup and delivery operations at each node, called {\it Pickup and Delivery Problem} (PDP), and the search for the lowest-cost route, which is the well-known {\it Traveling Salesman Problem} (TSP).
 
@@
 In this work, we consider a practical case in Brazil, which is the largest economy in Latin America.
-Due to its dimensions, this country has the largest air market on the continent with $2,499$\/ registered airports, of which $1,911$\/ are private and $588$\/ are public.
-Although it is an immense distribution network, the {\it Brazilian Air Force}\/ missions have always considered 3 to 5 nodes per flight plan.
-
-It is important to emphasize that this data is not an imposed limitation, but a historical fact that we will explore in our solution method.
-Throughout this work we approach routes with up to 7 nodes, as can be seen in Table \ref{tab:costs} and in Figure \ref{fig:nodes}.
-Although there are many other airports of interest to {\it Brazilian Air Force}, these 7 nodes were chosen due to their high demand and short transport times.
-The other Brazilian airports tend to have a lower demand, which is usually met, in a less expensive way, by cabotage, rail or road transport.
+Due to its dimensions, this country has the largest air market on the continent, with 2,499 registered airports, of which 1,911 are private and 588 are public.
+Although it is an immense distribution network, the {\it Brazilian Air Force} missions have always considered 3 to 5 nodes per flight plan.
+
+It is important to emphasize that this data is not an imposed limitation but a historical fact that we will explore in our solution method.
+Throughout this work, we approach routes with up to 7 nodes, as can be seen in Table \ref{tab:costs} and in Figure \ref{fig:nodes}.
+Although there are many other airports of interest to {\it Brazilian Air Force}, these seven nodes were chosen due to their high demand and short transport times.
+The other Brazilian airports tend to have lower demand, which is usually met in a less expensive way by cabotage, rail, or road transport.
 
@@
 
-However, as reported by Fok and Chun, {\it load planning (...) is usually done in roughly 2 hours before departure, when all the details of the cargo are present.} \cite{fok2004optimizing}.
+However, as reported by Fok and Chun, {\it load planning (...) is usually done roughly two hours before departure, when all the details of the cargo are present.} \cite{fok2004optimizing}.
 We can infer that this is because adequate commercial software does not exist.
-In addition, there is still the necessary time to plan the route, in order to minimize fuel consumption and ensure the transport of priority items.
-In these circumstances, it is of great importance to find a method that speeds up the load planning and the definition of the flight itinerary.
+In addition, there is still the necessary time to plan the route to minimize fuel consumption and ensure the transport of priority items.
+In these circumstances, it is of great importance to find a method that speeds up load planning and the definition of the flight itinerary.
 
 Our work proposes a method that prioritizes the transport of the most relevant items at each node and the fuel economy along the tour.
-We developed a heuristic that can be run on a simple handheld computer (such as a laptop or a tablet) that previously defines a flight plan, and provides a quick solution for cargo handling.
-Consequently, this method reduces the stress that transport planners are subjected to, because they have to deal with a lot of information in planning the aircraft route, assembling the pallets (regarding their positions), and a pick up and delivery plan to each node.
-
-To the best of our knowledge, this is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP and TSP has been addressed.
-This new problem is named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem} (ACLP+RPDP).
+We developed a heuristic that can be run on a simple handheld computer (such as a laptop or a tablet) that defines a flight plan and provides a quick solution for cargo handling.
+Consequently, this method reduces the stress that transport planners are subjected to because they have to deal with a lot of information in planning the aircraft route, assembling the pallets (regarding their positions), and a pick-up and delivery plan to each node.
+
+To the best of our knowledge, this is the first time that an air cargo transport problem that simultaneously involves APP, WBP, PDP, and TSP has been addressed.
+This new problem is named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem} (ACLP+RPDP).
 
@@
 
-{\color{blue} In this section, we briefly describe the characteristics of the main works related to air cargo transport, following the chronological order of Table \ref{tab:sa}, that lists the main works in the literature and the corresponding sub-problems addressed.
+{\color{blue} In this section, we briefly describe the characteristics of the main works related to air cargo transport, following the chronological order of Table \ref{tab:sa}, which lists the main works in the literature and the corresponding sub-problems addressed.
 We also indicate whether the dimensions of the items were taken into account ({\bf 3D} or {\bf 2D}) and which solution method was used: heuristic search methods ({\bf Heu}), integer programming ({\bf Int}), or linear programming ({\bf Lin}).
@@
 As can be seen, so far Lurkin and Schyns \cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem.
-Although it is innovative, strong simplifications were imposed by these authors: in relation to loading, APP was ignored; with regard to routing, it is assumed a predefined tour plan restricted to only two legs.
+Although it is innovative, strong simplifications were imposed by these authors: in relation to loading, APP was ignored; with regard to routing, it is assumed that a predefined tour plan is restricted to only two legs.
 It is important to note that these authors consider an aircraft with two doors, and the minimization of loading and unloading costs at the intermediate node was modeled through a container sequencing problem.
 Referring directly to this work, Brandt and Nickel \cite[p.
-409]{BrandtStefan2019} comment: {\it However, not even these sub-problems are acceptably solved for real-world problem sizes or the models omit some practically relevant constraints}.
+409]{BrandtStefan2019} comment: {\it However, not even these sub-problems are acceptably solved for real-world problem sizes, or the models omit some practically relevant constraints}.
 
 Larsen and Mikkelsen \cite{LarsenMikkelsen1979} developed an interactive procedure for loading 14 types of Boeing 747 into a two-leg transportation plan.
-Seven types of items were considered to be allocated in 17 to 42 positions.
+Seven types of items were considered to be allocated from 17 to 42 positions.
 With non-linear programming and heuristics, they present a solution that minimizes positioning changes in the intermediate node, optimizing the load balancing in the aircraft.
@@
 Brosh \cite{Brosh1981} addressed the problem of planning the allocation of cargo on an aircraft.
-Considering volume, weight and structural constraints, the author finds the optimal load layout through a fractional programming problem.
+Considering volume, weight, and structural constraints, the author finds the optimal load layout through a fractional programming problem.
 
 Ng \cite{Kevin1992} developed a multi-criteria optimization approach to load the {\it C-130} aircraft of the {\it Canadian Air Force}.
-Based on integer programming, this model provides timely planning and improves airlift support for combat operations, solving WBP with pallets in fixed positions, and considering 20 different items.
+Based on integer programming, this model provides timely planning and improves airlift support for combat operations by solving WBP with pallets in fixed positions and considering 20 different items.
 
 Heidelberg {\it et al.} \cite{Heidelberg1998} developed a heuristic for 2D packing in air loading, comparing it with methods for solving the {\it Bin Packing Problem}.
-These authors conclude that the classical algorithms are inadequate in this context, because they ignore the aircraft balancing constraints.
+These authors conclude that the classical algorithms are inadequate in this context because they ignore the aircraft balancing constraints.
 
@@
 Fok and Chun \cite{fok2004optimizing} developed a web-based application to make efficient use of space and load balancing for an air cargo company.
-Based on an analysis of historical data, an operational load planning with mathematical optimization is obtained.
-This container load planning is usually done in roughly 2 hours before departure, when all cargo details are in place.
+Based on an analysis of historical data, an operational load plan with mathematical optimization is obtained.
+This container load planning is usually done roughly 2 hours before departure, when all cargo details are in place.
 
 Chan {\it et al.} \cite{Chan2006} carried out a case study with heterogeneous pallets.
-To minimize the total cost of shipping, they developed a 3D packing heuristic, with a loading plan for each pallet.
-Although the authors do not consider load balancing or positioning of pallets in the cargo hold, this method is relevant in commercial and industrial applications, where cargo items tend to be less dense.
+To minimize the total cost of shipping, they developed a 3D packing heuristic with a loading plan for each pallet.
+Although the authors do not consider load balancing or the positioning of pallets in the cargo hold, this method is relevant in commercial and industrial applications, where cargo items tend to be less dense.
 
@@
 
-Verstichel {\it et al.} \cite{Verstichel2011} solved WBP by selecting the most profitable subset of containers to be loaded into an aircraft by the use of mixed-integer programming.
+Verstichel {\it et al.} \cite{Verstichel2011} solved WBP by selecting the most profitable subset of containers to be loaded into an aircraft by using mixed-integer programming.
 Experimental results on real-life data showed significant improvements compared to those obtained manually by an experienced planner.
@@
 
-Vancroonenburg {\it et al} \cite{Vancroonenburg2014} presented a mixed integer linear programming model that selects the most profitable pallets, satisfying safety and load balancing constraints on the {\it Boeing 747-400}.
+Vancroonenburg {\it et al} \cite{Vancroonenburg2014} presented a mixed integer linear programming model that selects the most profitable pallets while satisfying safety and load balancing constraints on the {\it Boeing 747-400}.
 Using a solver, these authors solved real problems in less than an hour.
 
-As already mentioned, Lurkin and Schyns \cite{LurkinSchyns2015} was the first work that simultaneously modeled WBP and PDP in air cargo transport.
+As already mentioned, Lurkin and Schyns \cite{LurkinSchyns2015} were the first to simultaneously model WBP and PDP in air cargo transport.
 The authors demonstrated that this problem is NP-hard and performed some experiments with real data, noting that their model offers better results than those obtained manually.
@@
 Roesener and Barnes \cite{RoesenerBarnes2016} proposed a heuristic to solve the {\it Dynamic Airlift Loading Problem} (DALP).
-Given a set of palletized cargo items that require transport between two nodes in a time frame, the objective of this problem is to select an efficient subset of aircraft, partition the pallets into aircraft loads and assign them to allowable positions in those aircraft.
-
-Paquay {\it et al.} \cite{PaquaySchynsLimbourg2016} presented a mathematical modeling to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepipeds format.
-Its objective is to maximize the volume used in containers, considering load balancing constraints, the presence of fragile items and the possibility of rotating these boxes.
+Given a set of palletized cargo items that require transport between two nodes in a given time frame, the objective of this problem is to select an efficient subset of aircraft, partition the pallets into aircraft loads, and assign them to allowable positions in those aircraft.
+
+Paquay {\it et al.} \cite{PaquaySchynsLimbourg2016} presented a mathematical model to optimize the loading of heterogeneous 3D boxes on pallets with a truncated parallelepiped format.
+Its objective is to maximize the volume used in containers, considering load balancing constraints, the presence of fragile items, and the possibility of rotating these boxes.
 Paquay {\it et al.} \cite{PaquayLimbourgSchynsOliveira2018} developed some heuristics to solve this problem.
@@
 Balance constraints and the presence of dangerous items were considered.
-Wong {\it et al.} \cite{eugene2021} integrated this tool to a digital simulation model, with a visualization and validation system, based on sensors that alert about load deviations.
-
-Zhao {\it et al.} \cite{zhao2021} proposed a new modeling for WBP based on mixed integer programming.
+Wong {\it et al.} \cite{eugene2021} integrated this tool into a digital simulation model with a visualization and validation system based on sensors that alert about load deviations.
+
+Zhao {\it et al.} \cite{zhao2021} proposed a new model for WBP based on mixed integer programming.
 Instead of focusing on the CG deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
@@
 {\color{blue} Zhao {\it et al.} \cite{zhao2023}  proposed three integer linear programming models: a Bi-objective Optimization Model (BOM), a Combinatorial Optimization Model (COM), and an Improved Combinatorial Optimization Model (IOM).
-The objectives of their solutions were the maximum loading capacity and the lowest CG deviation from a specified target CG.
-Four scenarios with various conditional metrics for three models are solved for the B777F aircraft using Gurobi.
-The results of the computations demonstrated that the BOM has the fastest solution speed, but the CG deviation is the largest, and in several cases, the CG deviation results are unacceptable.
-The COM has the longest solution time, which is difficult to tolerate in practice.
-Despite taking a little longer to solve computationally than the BOM, the IOM offers the best optimization solution.}
+The objectives of their solutions were the maximum loading possible and the lowest CG deviation from a specified target CG.
+Four scenarios with various metrics for three models are solved for the B777F aircraft using Gurobi.
+The results of the computations demonstrated that the BOM has the fastest solution speed, but the CG deviation is the worst, and in several cases, the CG deviation results are unacceptable.
+The COM has the longest solution time, which is difficult to tolerate in practice.}
 
 As can be seen, none of these works address air cargo palletization and load balancing with route optimization in a multi-leg transportation plan for a single aircraft.
-This is the objective of our work: to model and elaborate heuristics for a real problem that simultaneously involves 4 intractable sub-problems: APP, WBP, PDP and TSP.
+This is the objective of our work: to model and elaborate heuristics for a real problem that simultaneously involves four intractable sub-problems: APP, WBP, PDP, and TSP.
 
@@
 
-In this section, we describe the context of the problem addressed in this work, as well as the assumptions considered.
+In this section, we describe the context of the problem addressed in this work as well as the assumptions considered.
 
@@
 
-	\item At each node of the tour, the items to be allocated are characterized by weight, volume, scores, and previously known destinations.
-We leave the consideration of 2D or 3D items to a future work.
-
-	\item We considered a unique pallet type: the {\it 463L Master Pallet}, a common size platform for bundling and moving air cargo.
+	\item At each node of the tour, the items to be allocated are characterized by weight, volume, scores, and previously known destinations. We leave the consideration of 2D or 3D items for future work.
+
+	\item We considered a unique pallet type: the {\it 463L Master Pallet}, a common-size platform for bundling and moving air cargo.
 It is the primary air cargo pallet for more than 70 Air Forces and many air transport companies.
 This pallet has a capacity of $4,500 kg$ and $13.7 m^3$, which may be limited by its position along the cargo bay.
-It is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$.
+It is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weigh $140 kg$.
 For more information, see {\tt www.463LPallet.com}.
 
-	\item All items allocated on a pallet must have the same destination.
-A pallet which has not yet reached its destination may receive more items, although it is known that these operations of removing restraining nets increase handling time and the risk of improper delivery.
+	\item All items allocated to a pallet must have the same destination.
+A pallet that has not yet reached its destination may receive more items, although it is known that these operations of removing restraining nets increase handling time and the risk of improper delivery.
 We do not consider oversized cargo in this work, but only cargo items that fit on these pallets.
 
-	\item Finally, as we are interested in minimizing fuel costs, we disregarded others costs not directly associated with aircraft flight, such as handling.
+	\item Finally, as we are interested in minimizing fuel costs, we disregarded other costs not directly associated with aircraft flight, such as handling.
 
@@
 
-Throughout this text, we call {\it packed content}\/ a set of items of same destination stacked on a pallet and covered with a restraining net (see Figure \ref{fig:larger2}).
-It is considered a single item, having the same attributes of its components, whose values are the sum of individual scores, weights and volumes.
-To ensure accuracy in pickup and delivery operations, a packed content must remain on board until its destination.
+Throughout this text, we call {\it packed content}\/ a set of items of the same destination stacked on a pallet and covered with a restraining net (see Figure \ref{fig:larger2}).
+It is considered a single item, having the same attributes as its components, whose values are the sum of individual scores, weights, and volumes.
+To ensure accuracy in pickup and delivery operations, packed content must remain on board until its destination.
 
@@
 
-We consider real scenarios with an single aircraft with payload of $75,000 kg$, with enough fuel for a range of $4,400km$.
-The aircraft layout is presented in Figure \ref{fig:larger}, where the pallets positions are identified by $p_i$, $1\leq i \leq 18$.
+We consider real-world scenarios with a single aircraft with a payload of $75,000 kg$, and enough fuel for a range of $4,400km$.
+The aircraft layout is presented in Figure \ref{fig:larger}, where the pallet positions are identified by $p_i$, $1\leq i \leq 18$.
 
@@
 $D_i^{long}$\/ and $D_i^{lat}$\/ are, respectively, the longitudinal and lateral distances of each pallet centroids to the CG of aircraft along both axes.
-These distances will be used in the calculation of the torque referring to the items allocated on each pallet.
-In this aircraft, as the ramp have an inclination of $25^{\circ}$, we made the necessary corrections in $D_i^{long}$, $W_i$\/ and $V_i$\/ of the corresponding pallets.
+These distances will be used in the calculation of the torque, referring to the items allocated on each pallet.
+In this aircraft, as the ramp has an inclination of $25^{\circ}$, we made the necessary corrections in $D_i^{long}$, $W_i$\/ and $V_i$\/ of the corresponding pallets.
 
@@
 
-This aircraft spends $c_d$\/ dollars per kilometer flown, and can carry up to $W_{max}$\/ of cargo distributed on the pallets.
+This aircraft spends $c_d$\/ dollars per kilometer flown and can carry up to $W_{max}$\/ of cargo distributed on the pallets.
 The {\it fuel consumption rate}\/ $c_g$\/ is the percentage limit of cost increase due to the CG deviation on the longitudinal axis of the aircraft.
-$c_g$\/ depends on the characteristics of the aircraft, and is an arbitrated value that should be better evaluated in future research.
-It is important to consider that the $c_g$\/ tends to zero as the aircraft attitude tends to level.
+$c_g$\/ depends on the characteristics of the aircraft and is an arbitrated value that should be better evaluated in future research.
+It is important to consider that the $c_g$\/ tends to zero as the aircraft's attitude tends to be level.
 As the CG deviation varies from $0$\/ to $limit^{CG}_{long}$, the {\it fuel cost}\/ increase varies from $0$\/ to $c_g$.
@@
 \begin{itemize}
-	\item on each pallet, the items are distributed in such a way that their CG coincides with the centroid of the pallet;
-	\item the CG of the total load must be at a maximum longitudinal distance $limit^{CG}_{long}$\/ from the CG of the aircraft;
-	\item the CG of the total load must be at a maximum lateral distance $limit^{CG}_{lat}$\/ from the CG of the aircraft;
-	\item the pallets are distributed in two identical rows (with odd and even indices, respectively), and the centroid of $p_i$\/ is at distance $D^{lat}_{i}$\/ from the center-line of the aircraft.
+	\item On each pallet, the items are distributed in such a way that their CG coincides with the centroid of the pallet.
+	\item The CG of the total load must be at a maximum longitudinal distance of $limit^{CG}_{long}$\/ from the CG of the aircraft.
+	\item The CG of the total load must be at a maximum lateral distance of $limit^{CG}_{lat}$\/ from the CG of the aircraft.
+	\item The pallets are distributed in two identical rows (with odd and even indices, respectively), and the centroid of $p_i$\/ is at a distance $D^{lat}_{i}$\/ from the centerline of the aircraft.
 \end{itemize}
 
-\section{The mathematical modeling}
+\section{The mathematical model}
 \label{sec4}
 
-Given the assumptions and parameters described in the previous section, we are ready to present the mathematical modeling of ACLP+RPDP.
+Given the assumptions and parameters described in the previous section, we are ready to present the mathematical model of ACLP+RPDP.
 
@@
 
-\subsection{Problem structure and input data}
+\subsection{Problem structure}
 
@@
 	\item $\tau_{\pi_k}$\/ \ref{eq:tau} is the longitudinal torque applied by the loaded pallets at the node ${\pi_k}$, in proportion relative to the highest torque supported by the aircraft.
-	\item $\tilde{c}_\pi$\/ \ref{eq:costs} is the cost of to the total fuel consumed in the tour $\pi$\/ due to the distances traveled and the CG longitudinal deviations \footnote{In our experiments, we found that the magnitude of the lateral torque was always very small, and therefore we decided to ignore it in the fuel consumption.} in each pallet.
+	\item $\tilde{c}_\pi$\/ \ref{eq:costs} is the cost of to the total fuel consumed in the tour $\pi$\/ due to the distances traveled and the CG longitudinal deviations \footnote{In our experiments, we found that the magnitude of the lateral torque was always very small, so we decided to ignore it in the fuel consumption calculation in each pallet.}
 	\item The sets of nodes not yet visited along the tour $\pi$\/ are defined in \ref{eq:pdp11} and \ref{eq:departing}.
@@
 
-Once the assumptions of this work and the mathematical modeling of the problem are presented, it is possible to see that ACLP+RPDP is {\it NP-hard}.
-
- {\color{red}  In a similar way to Lurkin and Schyns \cite[p.
-6]{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$\/ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in node 0, and no items in node 1.
-Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is {\it NP-complete}.}
+Once the assumptions of this work and the mathematical model of the problem are presented, it is possible to see that ACLP+RPDP is {\it NP-hard}.
 
 Throughout our research, we have thoughtfully described the ACLP+RPDP model in standard MIP format and found that no MIP solver can handle its practical cases in a feasible time.
-Thus, as ACLP+RPDP is highly complex, involving four intractable subproblems (APP, WBP, PDP and TSP), our strategy will be to focus only on {\it real cases}\/ and to develop quick node-by-node solutions, not necessarily optimal, that allow to build a complete tour.
+Thus, as ACLP+RPDP is highly complex, involving four intractable subproblems (APP, WBP, PDP, and TSP), our strategy will be to focus only on {\it real cases}\/ and to develop quick node-by-node solutions, not necessarily optimal, that allow us to build a complete tour.
 
@@
 In this way, the loading and unloading time will be limited only by the use of the equipment at each node.
-Therefore, there will only be waiting time at the base node, corresponding to the runtime of ACLP+RPDP algorithm.
+Therefore, there will only be waiting time at the base node, corresponding to the runtime of the ACLP+RPDP algorithm.
 For this reason, our method will also be parameterized by the maximum time that can be waited until the start of the load in the base node.
 
-On the other hand, we know that transport aircraft generally have a few dozen pallets, the flight plan has less than $6$\/ nodes, and each node has hundreds of items to be shipped.
+On the other hand, we know that transport aircraft generally have a few dozen pallets, the flight plan has fewer than six nodes, and each node has hundreds of items to be shipped.
 We also know that missions with fewer nodes are more frequent than longer missions.
@@
 
-\item If we have quick node-by-node solutions, as the number of nodes is small we have the possibility to check the solution corresponding to the tour with the shortest total distance, and even test all possible tours, selecting the one that provides the best value for the objective function.
+\item If we have quick node-by-node solutions, as the number of nodes is small, we have the possibility to check the solution corresponding to the tour with the shortest total distance and even test all possible tours, selecting the one that provides the best value for the objective function.
 
@@
 
-		\State Let aircraft and $M$  \label{main:M} \Comment{Aircraft parameters from Table \ref{tab:larger}}
+		\State {\color{blue} Let $M$ be the set of pallets }  \label{main:M} \Comment{parameters in Table \ref{tab:larger}}
 
@@
 
-Next, we will present two subsections: in the first we explain how $SolveTour$\/ is executed, presetting the destinations of the pallets.
-In the second we will present the heuristics developed for node-by-node solutions.
+Next, we will present two subsections: in the first, we explain how $SolveTour$\/ is executed, presetting the destinations of the pallets.
+In the second, we will present the heuristics developed for node-by-node solutions.
 
@@
 As we commented in the previous subsection, we will adopt the strategy of presetting the destinations of each pallet throughout the tour.
-This is feasible in practical cases, where $1<K<m$.
+This is feasible in practical cases where $1<K<m$.
 For this, each pallet $i$\/ also has a field $T^k_i$, $0\leq k \leq K$, which stores its next destination after being loaded at node $k$.
@@
 
-When the aircraft is at the base, the initial graph $G_1$\/ is empty and there is no packed contents \ref{tour:g11}.
+When the aircraft is at the base, the initial graph $G_1$\/ is empty, and there are no packed contents \ref{tour:g11}.
 Otherwise, $UpdatePacked$\/ (line \ref{tour:dest}) returns the set of packed contents that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation.
@@
 
-$SetPalletsDestinations$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node, without changing the pallets destination with packed contents.
+$SetPalletsDestinations$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node without changing the pallet's destination with packed contents.
 
@@
 $MinCGDeviation$\/ (line \ref{cons:minCG}) relocates the packed contents on the pallets, minimizing torque and ensuring that they all remain on board, one packed content on each pallet.
-It is run through a MIP solver with objective function \ref{eq:torque} and constraints \ref{eq:embarked} and \ref{eq:one}.
+It is run through a MIP solver with the objective function \ref{eq:torque} and the constraints \ref{eq:embarked} and \ref{eq:one}.
 As there are few variables, $E_{Q_{\pi_k}}$\/ is obtained in less than $30$\/ milliseconds.
-Finally, the destination of each pallet with a packed content is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).
+Finally, the destination of each pallet with packed content is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).
 
@@
 
-$vol$\/ stores the demand volume of items destined to the non-visited nodes (line \ref{dest:vector1}).
+$vol$\/ stores the demand volume of items destined for the non-visited nodes (line \ref{dest:vector1}).
 The destination of empty pallets is defined proportionally to the volume of items to be embarked (lines \ref{dest:propor1}-\ref{dest:propor2}).
@@
 
-In this subsection we present two implementations of $SolveNode$\/ algorithm: with a MIP solver and a heuristic.
+In this subsection, we present two implementations of $SolveNode$\/ algorithm: a MIP solver and a heuristic.
 
@@
 
-Our strategy adopted in $SolveTour$\/ previously defines the values of some variables: the set of nodes to be visited is updated, the packed contents that remain on board are reallocated to minimize the CG deviation, and the pallets destinations are determined according to the volume of items available for shipment.
-
-In this way, the mathematical modeling for $SolveNode(MIP,\pi_k,G, tmax)$\/ becomes simpler, which finds an allocation of available items at the node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$.
-Thus, we use a MIP solver with runtime limit $tmax$\/ at the node $\pi_k$\/ to maximize the objective function \ref{eq:obj2}, with the calculus equations \ref{eq:newf} to \ref{eq:costs2}, subject to the constraints \ref{eq:tau2} to \ref{eq:ifB}.
-The binary variables $X_{ij}$\/ and $Y_{iq}$\/ define the sets of edges $E_{N_{\pi_k}}$\/ and $E_{Q_{\pi_k}}$, respectively, included in the graph $G$.
+Our strategy adopted in $SolveTour$\/ defines the values of some variables: the set of nodes to be visited is updated, the packed contents that remain on board are reallocated to minimize the CG deviation, and the pallet's destinations are determined according to the volume of items available for shipment.
+
+In this way, the mathematical model for $SolveNode(MIP,\pi_k,G, tmax)$\/ becomes simpler, which finds an allocation of available items at the node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$.
+Thus, we use a MIP solver with a runtime limit $tmax$\/ at the node $\pi_k$\/ to maximize the objective function \ref{eq:obj2 with the calculus equations \ref{eq:newf} to \ref{eq:costs2}, subject to the constraints \ref{eq:tau2} to \ref{eq:ifB}.
+The binary variables $X_{ij}$\/ and $Y_{iq}$\/ define the sets of edges $E_{N_{\pi_k}}$\/ and $E_{Q_{\pi_k}}$, respectively, included in graph $G$.
 
@@
 	\item In {\it selection phase}\/ (lines \ref{shims:beginthird}-\ref{shims:endthird}), the best shim in $Set$\/ is chosen.
-Initially, two shims are found: $sh_w$\/ with larger weight, and $sh_v$\/ with larger volume.
-Between the two, the one with the highest score will be chosen and its edges are inserted into $E_{N_{\pi_k}}$.
+Initially, two shims are found: $sh_w$\/ with larger weight and $sh_v$\/ with larger volume.
+Between the two, the one with the highest score will be chosen, and its edges will be inserted into $E_{N_{\pi_k}}$.
 
@@
 
-As we are dealing with a new problem, which until now had not been modeled in the literature, we have to create our own benchmarks.
-For this, we based on the characteristics of real airlifts carried out by the {\it Brazilian Air Force}, as described below.
+As we are dealing with a new problem that until now had not been modeled in the literature, we have to create our own benchmarks.
+For this, we based it on the characteristics of real airlifts carried out by the {\it Brazilian Air Force}, as described below.
 
@@
 
-In order for $Gurobi$\/ to be able to solve the largest possible number of tests without memory overflow, we set its parameter {\it MIPgap}\/ to 1\%.
+For Gurobi to be able to solve the largest possible number of tests without memory overflow, we set its parameter {\it MIPgap}\/ to 1\%.
 This shortens its runtime, in addition to ensuring that its objective function $f$\/ is at most 1\% of the optimal solution.
@@
 We will present the average of the objective function $f$\/ and the runtime of $Gurobi$\/ and $Shims$\/ in these instances.
-To facilitate the comparison between both, we added a last column in the tables, where two values are indicated:
+To facilitate the comparison between both, we added a last column in the tables where two values are indicated:
 \begin{itemize}
@@
 
-We also indicate the two adopted strategies: dedicating all the processing time to the two shortest tours, or distributing it among all the $K!$\/ tours.
+We also indicate the two adopted strategies: dedicating all the processing time to the two shortest tours or distributing it among all the $K!$\/ tours.
 
 The results obtained with $tmax = 3600s$, which is the highest tested runtime limit, are in Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, with $surplus$\/ values $1.2$, $1.5$\/ and $2.0$, respectively.
-We indicate with an {\bf x} the cases where $Gurobi$\/ did not find a feasible solution within this runtime limit, or had to be aborted due to high RAM consumption.
+We indicate with an {\bf x} the cases where $Gurobi$\/ did not find a feasible solution within this runtime limit or had to be aborted due to high RAM consumption.
 
@@
 
-From these data, we can make some conclusions:
+From these data, we can draw some conclusions:
 \begin{itemize}
-\item The strategy of testing all $K!$\/ tours often gives a better quality solution.
-\item $Gurobi$\/ fails in some cases, when $scenario=5$\/ and the strategy is to check all $K!$\/ tours.
-This occurs because the runtime limit per node is smaller, and there tends to be more packed contents on the aircraft, reducing the space for allocating items and making the solution difficult.
+\item The strategy of testing all $K!$\/ tours often gives a better-quality solution.
+\item $Gurobi$\/ fails in some cases when $scenario=5$\/ and the strategy is to check all $K!$\/ tours.
+This occurs because the runtime limit per node is smaller and there tend to be more packed contents on the aircraft, reducing the space for allocating items and making the solution difficult.
 \item When $Gurobi$\/ finishes, it finds the best solution, but the one obtained by $Shims$\/ reaches $99\%$\/ of that value.
@@
 \item All runtimes are much lower than the limit because the solution on many nodes can be fast.
-Anyway, in all the tests performed, the maximum time spent by $Shims$\/ did not reach 4 minutes, that is, it is practically instantaneous.
+Anyway, in all the tests performed, the maximum time spent by $Shims$\/ did not reach 4 minutes, i.e., it was practically instantaneous.
 On the other hand, when $scenario=5$\/ and $surplus=1.5$, Gurobi spent almost 40 minutes.
@@
 We can observe more cases where $Gurobi$\/ fails, even in smaller scenarios.
-When $Gurobi$\/ finishes, $Shims$\/ finds quickly a solution of similar quality ($98\%$\/ or better).
+When $Gurobi$\/ finishes, $Shims$\/ finds a solution of similar quality ($98\%$\/ or better).
 In all cases, $Shims$\/ finds a good solution in less than 4 minutes.
@@
 
-In this work, we modeled and solved a real air transport problem, named {\it Air Cargo Load Planning with Routing, Pickup and Delivery Problem}\/ (ACLP+RPDP).
-For the first time in the literature, a {\it NP-hard}\/ problem that involves {\it simultaneously}\/ pallet assembly, load balancing and route planning is addressed, where the cost-effectiveness of transport is maximized.
-We adopted some simplifications that are not critical, but that allowed the unprecedented solution of this problem considering more than two nodes.
+In this work, we modeled and solved a real air transport problem named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem}\/ (ACLP+RPDP).
+For the first time in the literature, a {\it NP-hard}\/ problem that involves {\it simultaneously}\/ pallet assembly, load balancing, and route planning simultaneously is addressed, where the cost-effectiveness of transport is maximized.
+We adopted some simplifications that are not critical but that allowed for an unprecedented solution to this problem considering more than two nodes.
 
 We consider that, in practical cases, the number $K$\/ of nodes, excluding the base, is small ($K \leq 6$), each of them with hundreds of items to be shipped.
-Thus, considering a real aircraft, we have developed some node-by-node solutions, which are not necessarily optimal.
+Thus, considering a real aircraft, we have developed some node-by-node solutions that are not necessarily optimal.
 This allows us to test two strategies: find a solution using the shortest tours, or enumerate all $K!$\/ tours and choose the best.
@@
 
-We have developed a solution process that, in less than 4 minutes, can establish a flight plan for a single aircraft with a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximizing the total score and minimizing fuel consumption.
+We have developed a solution process that, in less than 4 minutes, can establish a flight plan for a single aircraft with a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximizing the total score, and minimizing fuel consumption.
 The output, which includes the tour plan and the pallet building and arrangement plan, is an essential part of airlift: it improves flight safety, makes ground operations more efficient, and makes sure that each item gets to its right destination.
@@
 
-Our main contributions were the mathematical modeling of ACLP+RPDP that involves four sub-problems {\it NP-hard}\/, a complete process to solve it on a simple handheld computer, and a new heuristic that offers fast node solutions with good quality.
-Finally, by focusing on the node-to-node solution, the method of this work is not exclusive to aircraft and airports: it can be adapted, for example, to ships and ports, or vehicles and warehouses, or wagons and railways, provided that their practical cases are similar to those considered here.
-In these situations, it would be necessary to make some changes in the modeling: for example, modify the load balancing constraints, and consider the available space in vehicles or wagons instead of pallets.
-
-As this is an ongoing research, we thought about some possible future improvements:
+Our main contributions were the mathematical model of ACLP+RPDP that involves four sub-problems {\it NP-hard}\/, a complete process to solve it on a simple handheld computer, and a new heuristic that offers fast node solutions with good quality.
+Finally, by focusing on the node-to-node solution, the method of this work is not exclusive to aircraft and airports; it can be adapted, for example, to ships and ports, vehicles and warehouses, or wagons and railways, provided that their practical cases are similar to those considered here.
+In these situations, it would be necessary to make some changes in the model: for example, modify the load balancing constraints and consider the available space in vehicles or wagons instead of pallets.
+
+As this is ongoing research, we thought about some possible future improvements:
 \begin{itemize}
-\item consider a aircraft fleet, rather than a single one;
+\item consider an aircraft fleet rather than a single one;
 \item model three-dimensional items;
-\item discard the highest cost tours, to reduce the runtime;
+\item discard the highest-cost tours to reduce the runtime;
 \item implement parallel algorithms in some steps of the solution.

### 2023-07-20 CAIE/caie.tex
CAIE updated and sent to Carlos.
CAIE updated

@@
 
-In this way, the mathematical model for $SolveNode(MIP,\pi_k,G, tmax)$\/ becomes simpler, which finds an allocation of available items at the node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$.
-Thus, we use a MIP solver with a runtime limit $tmax$\/ at the node $\pi_k$\/ to maximize the objective function \ref{eq:obj2 with the calculus equations \ref{eq:newf} to \ref{eq:costs2}, subject to the constraints \ref{eq:tau2} to \ref{eq:ifB}.
+In this way, the mathematical model for $SolveNode(MIP,\pi_k,G, tmax)$\/ becomes simpler, which finds an allocation of available items at the node $\pi_k$\/ using previously defined values of
+$L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$.
+Thus, we use a MIP solver with a runtime limit $tmax$\/ at the node $\pi_k$ to maximize the objective function \ref{eq:obj2} with the calculus equations \ref{eq:newf} to \ref{eq:costs2}, subject to the constraints \ref{eq:tau2} to \ref{eq:ifB}.
 The binary variables $X_{ij}$\/ and $Y_{iq}$\/ define the sets of edges $E_{N_{\pi_k}}$\/ and $E_{Q_{\pi_k}}$, respectively, included in graph $G$.
@@
 
+{\color{blue} The dataset used in this study is available in \cite{MesquitaSanches2023}. }
+
 \section*{References}
@@
 
+\bibitem{MesquitaSanches2023} Mesquita, Antonio Celio, and Carlos Alberto Alonso Sanches. Air Cargo Load and Route Planning in Pickup and Delivery Operations (2022). Web.
+
 \end{thebibliography}

### 2024-01-25 latex/eswa.tex
iRace wtill not working

% !TeX spellcheck = en_GB
\documentclass[preprint,authoryear]{elsarticle}

\usepackage{natbib}
\bibliographystyle{elsarticle-harv}

\usepackage{algorithm}
\usepackage{algpseudocode}

\usepackage{amsmath,amsthm}
\usepackage{amssymb}

\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{multirow}

\usepackage[none]{hyphenat}

\usepackage{booktabs}

\usepackage{tikz}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}

\usepackage{caption}
\captionsetup{labelfont=bf}
\captionsetup{skip=0pt}

\usepackage{pgfplots}
\pgfplotsset{compat=newest}

\newcommand{\boldm}[1] {\mathversion{bold}#1\mathversion{normal}}
\newcommand{\round}[1]{\ensuremath{\lfloor#1\rceil}}

\usepackage{color, colortbl}

\definecolor{Gray}{gray}{0.9}
\hyphenation{cost-ef-fec-tive-ness}

\usepackage{setspace}

\newcommand{\specialcell}[2][c]{%
	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\usepackage[margin=2.3cm]{geometry}
\geometry{a4paper}

\journal{Expert Systems with Applications}

\begin{document}

\begin{frontmatter}

	\title{Air cargo load and route planning in pickup and delivery operations}

	\author{A.C.P.~Mesquita}
	\ead{celio@ita.br}

	\author{C.A.A.~Sanches\corref{cor1}}
	\ead{alonso@ita.br}
	\cortext[cor1]{Corresponding author.}

	\address {Instituto Tecnol\'{o}gico de Aeron\'{a}utica - DCTA/ITA/IEC\\
		Pra\c{c}a Mal. Eduardo Gomes, 50\\
		S\~{a}o Jos\'{e} dos Campos - SP - 12.228-900 - Brazil}

\begin{abstract}
In the aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of load imbalance due to the urgency required for loading, immediate take-off, and mission accomplishment. Transport planners deal with trip itineraries, prioritisation of items, building up pallets, and balanced loading, but there are no commercially available systems that can integrally assist in all these requirements. This enables other risks, such as improper delivery, excessive fuel burn, and possible safety issues due to cargo imbalance, as well as a longer than necessary turn-around time. This {\it NP-hard}\/ problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modelled using standardised pallets in fixed positions. We developed a strategy to solve this problem, considering historical transport data from some Brazilian hub networks, and performed several experiments with a commercial solver, four known meta-heuristics, and a new heuristic designed specifically for this problem. By using a portable computer, our strategy quickly found practical solutions to a wide range of real problems in much less than operationally acceptable time.
\end{abstract}

\begin{keyword}
	OR in Airlines \sep Load Planning \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Vehicle Routing
\end{keyword}

\end{frontmatter}

\section{Introduction}\label{sec:Introduction}

The aviation industry adapts during global crises to keep supply chains moving. Air cargo provided complex expertise and the ability to access diverse destinations, delivering essential goods such as medicines, vaccine supplies, testing kits and other necessities with exceptional speed. This mode of transportation has become a preferred choice for governments, corporations and global companies in urgent need of transportation solutions.

Air cargo services are specially designed for organisations that require customised transportation, handle sensitive goods, or serve remote locations with limited routes. Air carriers typically use high-capacity cargo planes for economies of scale. Many cargo airlines have worldwide networks spread across destinations around the world.

Recently, \citep{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP), and {\it Weight and Balance Problem} (WBP). Several aspects were considered: item characteristics to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs); when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how total cargo weight is balanced; etc.

However, it is crucial to highlight that there are still other important challenges in air cargo transport that go beyond the definition of ACLPP, especially with regard to routes, and pickup and delivery at each destination. In this context, at least two more important sub-problems can be considered: simultaneous pickup and delivery at each node, called the {\it Pickup and Delivery Problem} (PDP), and searching for the best benefit-cost route, which is a special case of the {\it Travelling Salesman Problem} (TSP).

The problem addressed in this work is the definition of the route of an aircraft that loads and unloads hundreds of items in up to seven hubs, dealing with weight, volume, and balance constraints, and maximising the benefit-cost ratio. This challenge has become even more acute during the COVID-19 pandemic due to the need for urgent medical supplies. Delays in shipping essential items such as respirators to critical areas highlighted the need for optimised solutions.

This problem is extremely complex, as it deals with different objectives: defining a route that visits all hubs; maximising the items transported along this route, prioritising the essential ones; making sure items reach the correct destinations; ensuring aircraft safety constraints; and saving fuel so that the flight is sustainable.

We have developed a heuristic process that can be run on a handheld computer, quickly providing a good solution to real instances of this problem. Solutions consist of flight itineraries, pickup and delivery plans, and the allocation of items onto pallets, ensuring the load-balancing constraints. Our method also reduces the stresses that transport planners are subject to, as they have to deal with extensive information in a short time frame.

To the best of our knowledge, this is the first time that an air cargo problem involving simultaneously APP, WBP, PDP, and TSP has been addressed. This new problem is named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem} (ACLP+RPDP). As we will describe in our mathematical modelling, these four sub-problems appear in an interconnected way in ACLP+RPDP and therefore cannot be solved independently.

As a real case study, we consider a crucial network for the {\it Brazilian Air Force}, as can be seen in Table \ref{tab:costs} and Figure \ref{fig:nodes}. Although there are other airports of interest, these nodes were chosen due to their high demand. Other Brazilian airports tend to have smaller transport requests, which are generally met less expensively by cabotage, rail, or road transport.

\begin{table}[htb]

	\begin{minipage}{0.58\linewidth}

		\caption{Brazilian airports distances ($km$)} \label{tab:costs}
		\centering
		\footnotesize

		\newcolumntype{C}{>{\centering\arraybackslash}p{0.07\textwidth}}
		\newcolumntype{R}{>{\raggedleft\arraybackslash}p{0.07\textwidth}}

		\begin{tabular}{C R R R R R R R}
			\toprule
			IATA*   & GRU   & GIG   & SSA   & CNF   & CWB   & BSB   & REC \\	
			\midrule	
			GRU     & 0	    &343	&1,439   &504    &358    &866    &2,114\\
			GIG	    & 343	&0	    &1,218   &371    &677    &935    &1,876\\
			SSA	    & 1,439	&1,218	&0	    &938    &1,788   &1,062   &676\\
			CNF	    & 504	&371	&938	&0	    &851    &606    &1,613\\
			CWB	    & 358	&677	&1,788	&851	&0	    &1,084   &2,462\\
			BSB	    & 866	&935	&1,062	&606	&1,084	&0	    &1,658\\
			REC	    & 2,114	&1,876	&676	&1,613	&2,462	&1,658	&0\\
			\bottomrule
			\multicolumn{8}{c}{*International Air Transport Association}\\
			\multicolumn{8}{c}{\small\textsuperscript{Source: www.airportdistancecalculator.com}}\\
		\end{tabular}
		\normalsize

	\end{minipage}\hfill 
	\begin{minipage}{0.42\linewidth}
		\centering
		\includegraphics[scale=0.17]{nodes.png}
		\captionof{figure}{A route with 7 airports}
		\label{fig:nodes}
	\end{minipage}
\end{table}

This article is organised into six more sections. In Section \ref{literature}, we make the literature review. In Section \ref{assumptions}, we present the context and requirements of ACLP+RPDP. In Section \ref{modelling}, we describe its mathematical modelling. In Section \ref{algorithms}, we describe the developed algorithms, whose results are presented in Section \ref{results}. Finally, our conclusions are in Section \ref{conclusions}.

\section{Literature review}
\label{literature}

The vast majority of operational research applied to air cargo is focused on challenges related to WBP, that is, the distribution of items on pallets to ensure load balancing. We can mention: \citep{LarsenMikkelsen1979}; \citep{Brosh1981}; \citep{Kevin1992}; \citep{Heidelberg1998}; \citep{fok2004optimizing}; \citep{KaluznyBohdanL2009Oalb}; \citep{Verstichel2011}; \citep{Limbourg2012}; \citep{RoesenerBarnes2016}; \citep{YangLiuGao2018}; \citep{zhao2021}; \citep{MiguelOptimalAP}.

Other authors have addressed pallet assembly (APP) on aircraft, possibly also considering load balancing (WBP): \citep{MongeauBes2003}; \citep{Chan2006}; \citep{RoesenerHall2014}; \citep{Vancroonemburg2014}; \citep{Paquay2016}; \citep{Paquay2018}; \citep{wong2020}; \citep{eugene2021}; \citep{zhao2023}.

In all these works, there is a great diversity of scenarios and solutions: some consider items in two dimensions, and others in three dimensions; some used integer programming, and others developed specific heuristics.

\citep{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem. The authors demonstrated that this problem is {\it NP-hard}. Although it is innovative, strong simplifications were imposed by these authors: in relation to loading, APP was ignored; regarding routing, it is assumed that a predefined tour plan is restricted to only two legs. Referring directly to this work, \citep{BrandtStefan2019} comment: {\it However, not even these sub-problems are acceptably solved for real-world problem sizes, or models omit some practically relevant constraints}.

Table \ref{tab:sa} lists this literature and the involved sub-problems. We also indicate whether the dimensions of the items were considered ({\bf 2D} or {\bf 3D}) and which solution method was used: heuristic search methods ({\bf Heu}), integer programming ({\bf Int}), or linear programming ({\bf Lin}).

\begin{table}[h]
	\centering
	\caption{Air cargo transport: literature, sub-problems and features}  \label{tab:sa}
	\footnotesize
	\begin{tabular}{r|cccc|ccccc}
		\toprule
		Work & {\bf APP}  & {\bf WBP}  &  {\bf PDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} \\
		\midrule
		\citep{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$ \\
		\citep{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$ \\
		\citep{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ \\
		\citep{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$ \\
		\citep{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		\citep{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
		\citep{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$ \\
		\citep{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$ \\
		\citep{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
		\citep{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		\citep{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
		\citep{Vancroonemburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		\citep{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
		\citep{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
		\citep{Paquay2016,Paquay2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$ \\
		\citep{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$ \\
		\citep{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$  \\
		\citep{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$  \\
		\citep{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
		\citep{zhao2023} & $\bigstar$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$  \\
		\citep{MiguelOptimalAP} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$  \\
		{\bf This article}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$  \\
		\bottomrule 
	\end{tabular}
	\normalsize 
\end{table}

As can be seen, none of these papers address air cargo palletization and load balancing with route optimisation in a multi-leg transport plan for a single aircraft. Our work is the first to address a real air transport problem in which APP, WBP, PDP and TSP arise in an interconnected way.

\section{Context and assumptions}
\label{assumptions}

In this section, we describe the context of the problem addressed in this work as well as the assumptions considered.

\subsection{Operational premises}

As we are dealing with an extremely complex and diverse problem, we decided to establish some simplifying characteristics:

\begin{itemize}
	\setlength\itemsep{-0.3em}

	\item At each node of the tour, the items to be allocated are characterized by weight, volume, scores, and previously known destinations. We leave the consideration of 2D or 3D items to a future work.

	\item We considered a unique pallet type: the {\it 463L Master Pallet}, a common size platform for bundling and moving air cargo. It is the primary air cargo pallet for more than 70 Air Forces and many air transport companies. This pallet has a capacity of $4,500 kg$ and $13.7 m^3$, which may be limited by its position along the cargo bay. It is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$. For more information, see {\tt www.463LPallet.com}.

	\item All items allocated on a pallet must have the same destination. A pallet which has not yet reached its destination may receive more items, although it is known that these operations of removing restraining nets increase handling time and the risk of improper delivery. We do not consider oversized cargo in this work, but only cargo items that fit on these pallets.

	\item Finally, as we are interested in minimizing fuel costs, we disregarded others costs not directly associated with aircraft flight, such as handling.

\end{itemize}

Throughout this text, we call {\it packed content}\/ a set of items of the same destination stacked on a pallet and covered with a restraining net. It is considered a single item, having the same attributes as its components, whose values are the sum of individual scores, weights, and volumes. To ensure accuracy in pickup and delivery operations, packed content must remain on board until its destination.

\subsection{Aircraft parameters and load balancing}

We consider real-world scenarios, where Table \ref{tab:larger} shows the aircraft parameters. $p_i$\/ are pallets, $1\leq i \leq 18$, whose weight and volume limits are $W_i$\/ and $V_i$, respectively. $D_i^{long}$\/ and $D_i^{lat}$\/ are, respectively, the longitudinal and lateral distances of each pallet centroids to the aircraft centre of gravity (CG) along both axes. These distances will be used in the calculation of the torque, referring to the items allocated on each pallet. In this aircraft, as the ramp has an inclination of $25^{\circ}$, we made the necessary corrections in $D_i^{long}$, $W_i$\/ and $V_i$\/ of the corresponding pallets ($p_1$, $p_2$, $p_3$, and $p_4$).

\begin{table}[h]
	\centering
	\caption{Aircraft parameters}  \label{tab:larger}
	\footnotesize
	\begin{tabular}{c | c c c c c c c c c}
		\toprule
		& \multicolumn{3}{c}{$Payload$: 75,000kg} & \multicolumn{3}{c}{$limit^{CG}_{long}$: $1.170m$} &
		\multicolumn{3}{c}{$limit^{CG}_{lat}$: $0.19m$} \\
		\midrule
		\multirow{2}{*}{{\boldm{$p_i$}}}  & $p_{17}$ & $p_{15}$ & $p_{13}$ & $p_{11}$ & $p_{9}$ & $p_{7}$ & $p_{5}$ & $p_{3}$ & $p_{1}$ \\
		& $p_{18}$ & $p_{16}$ & $p_{14}$ & $p_{12}$ & $p_{10}$ & $p_{8}$ & $p_{6}$ & $p_{4}$ & $p_{2}$ \\
		\midrule 
		\multirow{2}{*}{\boldm{$D_i^{long}$} ($m$)} & -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\
		& -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\			
		\midrule 
		\multirow{2}{*}{\boldm{$D_i^{lat}$} ($m$)}  & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 \\
		& -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 \\	
		\midrule
		{\boldm{$W_i$}} ($kg$)      &   4,500   &    4,500  &   4,500   &  4,500    & 4,500     & 4,500     & 4,500     & 3,000    & 3,000   \\
		{\boldm{$V_i$}} ($m^3$)   &   14.8   &   14.8   &  14.8    &  14.8    & 14.8     & 14.8     & 14.8     & 10.0    & 7.0 \\	
		\midrule	
		\textbf{Fuel cost}  & \multicolumn{9}{c}{ $c_d$ = US\$ $4.90/km$ }	 \\
		\midrule
		\textbf{Fuel consumption rate}  & \multicolumn{9}{c}{ $c_g = 5\%$} \\		
		\midrule
		\textbf{Maximum weight}  & \multicolumn{9}{c}{ $W_{max} = \sum_i W_i = 75,000kg$} \\
		\bottomrule
	\end{tabular}
	\normalsize 
\end{table}

This aircraft spends $c_d$\/ dollars per kilometre flown and can carry up to $W_{max}$\/ of cargo distributed on the pallets. The fuel penalty $c_g$\/ is the percentage of cost increase due to the CG deviation on the longitudinal axis, estimated at 5.0\%. It is important to consider that $c_g$\/ tends to zero as the aircraft attitude tends to be level. As the CG deviation varies from $0$\/ to $limit^{CG}_{long}$, this fuel penalty varies from $0$\/ to $c_g$.

The torque applied to the aircraft must keep its CG in the operational range, which corresponds to a fixed percentage of the {\it Mean Aerodynamic Chord} \footnote{Chord is the distance between the leading and trailing edges of the wing, measured parallel to the normal airflow over the wing. The average length of the chord is known as the {\it Mean Aerodynamic Chord} (MAC).} which is considered $1.17m$ for the aircraft of this work (see Figure \ref{fig:lateral}).

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.22]{lateral.png}
	\caption{Aircraft longitudinal cut, where red lines are pallets positions}
	\label{fig:lateral}
\end{figure}

We also make the following assumptions:

\begin{itemize}
	\setlength\itemsep{-0.3em}
	\item on each pallet, the items are distributed in such a way that their CG coincides with the centroid of the pallet, because builders are well-trained to do so;
	\item the CG of the total load must be at a maximum longitudinal distance of $limit^{CG}_{long}$\/ from the CG of the aircraft;
	\item the CG of the total load must be at a maximum lateral distance of $limit^{CG}_{lat}$\/ from the CG of the aircraft;
	\item the pallets are distributed in two identical rows (with odd and even indices, respectively), and the centroid of $p_i$\/ is at a distance $D^{lat}_{i}$\/ from the centreline of the aircraft;
	\item when there are items or packed contents in $p_i$, the common destination of this load will be assigned to variable $T_i$. 
\end{itemize}

\section{The mathematical modelling}
\label{modelling}

In this section, we present the mathematical modelling of ACLP+RPDP in Tables \ref{tab:struc}, \ref{tab:graph}, \ref{tab:functions}, and \ref{tab:constraints}, with their corresponding descriptions.

In Table \ref{tab:struc}, we describe the problem structure: nodes and their permutations, distances and associated costs, pallets characteristics, items available for shipment at each node, and packed contents shipped. The item $j$\/ in node $k$\/ has score $s_j$, weight $w_j$, volume $v_j$, and destination $to_j \in L_k$. Similarly, the packed content $q$, that remains on board at node $k$, has total weight $w_q$, total volume $v_q$, and destination $to_q \in L_k$. Packed contents that were destined to node $k$\/ are unloaded when the aircraft arrives there; that is, they are not considered in $Q_k$.

\begin{table}[ht]
	\centering
	\footnotesize
	\caption{Problem structure}  \label{tab:struc}
	\begin{tabular}{l l}
		\toprule
		\textbf{Notation}	& \textbf{Description} \\
		\midrule
		$L = \{ 0, 1, \ldots, K \}$ & Set of $K+1$\/ nodes of the tour, where node 0 is the base \\
		\midrule
		$\pi$               & A permutation between nodes 1, \ldots, $K$\\
		\midrule
		$S_K$               & Set of $K!$\/ permutations \\
		\midrule
		$\pi_k$             & The $k^{th}$\/ node of tour $\pi$, $1 \leq k \leq K$\\
		\midrule
		Tour $\pi$          & $\{0,\pi_1,\ldots, \pi_K, 0\}$ \\
		& For ease of notation, $\pi_0 = \pi_{K+1} = 0$\\
		\midrule
		$L_k$               & Set of remaining nodes of tour at node $k$, $0 \leq k \leq K$\\
		& By definition, $L_0 = L$\\
		\midrule
		$d(a,b)$            & Distance from node $a$\/ to node $b$, where $0 \leq a, b \leq K$ \\
		&   By definition, $d(a,a) = 0, \forall a$\\
		\midrule
		$C=\left[c_{a,b}\right]$    & Cost matrix of flights, where $c_{a,b} = c_d \times d(a,b)$\\
		\midrule
		$M = \{1, \ldots, m \}$  & Set of $m$\/ pallets in specific positions within the aircraft \\
		&  See Table \ref{tab:larger}, where $m=18$ \\
		\midrule
		$N_k = \{1, \ldots, n_k \}$ & Set of $n_k$\/ items available for loading at node $k$, $1 \leq j \leq n_k$, $0 \leq k \leq K$\\
		\midrule
		$N = \bigcup_{0 \leq k \leq K} N_k$ & Set of items in all nodes along a tour\\
		\midrule
		$Q_k = \{1, \ldots, m_k \}$ & Set of $m_k \leq m$\/ packed contents at node $k$, $1 \leq q \leq m_k$, $0 \leq k \leq K$ \\ 
		& By definition, $m_0 = 0$ and $Q_0 = \varnothing$\\
		\bottomrule
	\end{tabular}
	\normalsize
\end{table}

Table \ref{tab:graph} contains decision variables and the ACLP+RPDP allocation graph.

\begin{table}[H]
	\centering
	\footnotesize
	\caption{Decision variables and allocation graph}  \label{tab:graph}
	\begin{tabular}{l l}
		\toprule
		\textbf{Notation}	& \textbf{Description} \\
		\midrule
		$X_{ij}^{\pi_k}$\/ and $Y_{iq}^{\pi_k}$ & Binary variables, where $1 \leq i \leq m$, $1 \leq j \leq n_{\pi_k}$, $1 \leq q \leq m_{\pi_k}$\/ and $0 \leq k \leq K$\\
		\midrule
		$X_{ij}^{\pi_k} = 1$ & If item $j$\/ at node ${\pi_k}$\/ is assigned to pallet $i$, and 0 otherwise\\
		\midrule
		$Y_{iq}^{\pi_k} = 1$ & If packed content $q$\/ at node ${\pi_k}$\/ is assigned to pallet $i$, and 0 otherwise\\
		\midrule
		$T_i^{\pi_k} \in L_{\pi_k}$ &  Destination of items and packed contents assigned to pallet $i$\ at node ${\pi_k}$\\
		\midrule
		$G_{\pi_k}(V_{\pi_k}, E_{\pi_k})$ & Allocation graph at node ${\pi_k}$\\
		\midrule
		$V_{\pi_k} = M \cup N_{\pi_k} \cup Q_{\pi_k}$ & Allocation graph vertices at node ${\pi_k}$: pallets, items and packet contents\\
		\midrule
		$E_{N_{\pi_k}}$ & Allocation graph edges at node ${\pi_k}$, corresponding to shipped items\\
		\midrule
		$E_{Q_{\pi_k}}$ & Allocation graph edges at node ${\pi_k}$, corresponding to packed contents\\
		\midrule
		$E_{\pi_k} = E_{N_{\pi_k}} \cup E_{Q_{\pi_k}}$ & Allocation graph edges at node ${\pi_k}$\\
		\midrule
		$(i, j) \in E_{N_{\pi_k}}$ & If $X_{ij}^{\pi_k} = 1$, where $i$\/ is a pallet and $j$\/ is a item at node ${\pi_k}$ \\
		\midrule
		$(i, q) \in E_{Q_{\pi_k}}$ & If $Y_{iq}^{\pi_k} = 1$, where $i$\/ is a pallet and $q$\/ is a packed content at node ${\pi_k}$\\
		\bottomrule
	\end{tabular}
	\normalsize
\end{table}

The calculus functions of ACLP+RPDP are described in Table \ref{tab:functions}.

\begin{table}[H]
	\centering
	\footnotesize
	\caption{Calculus functions} \label{tab:functions}
	\begin{tabular}{ c l }
		\toprule
		\textbf{Function} & \textbf{Description} \\
		\midrule
		(\ref{eq:scores}) & Total score of transported items throughout tour $\pi$\\
		\midrule
		(\ref{eq:tau})    & Longitudinal torque applied by loaded pallets at node ${\pi_k}$\\
		\midrule
		(\ref{eq:costs})  & Total cost of fuel on tour $\pi$\/ (distances and CG longitudinal deviations) \\
		\midrule
		(\ref{eq:departing}) & Set of not visited nodes at node ${\pi_k}$\\
		\midrule
		(\ref{eq:LatItem}) & Lateral torque at node ${\pi_k}$ (shipped items) \\
		\midrule
		(\ref{eq:LatPacked}) & Lateral torque at node ${\pi_k}$ (packed contents)\\
		\midrule
		(\ref{eq1}) & Objective function of ACLP+RPDP\\
		\bottomrule
	\end{tabular}
	\normalsize
\end{table}

\begin{equation} \label{eq:scores}
	\tilde{s}_\pi = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times s_j
\end{equation}

\begin{equation} \label{eq:tau}
	\tau_{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times \Big ( \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{long}; \ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:costs}
	\tilde{c}_{\pi} = \sum_{k=0}^{K} \Big [ c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|) \Big ] 
\end{equation}

\begin{equation} \label{eq:departing}
	L_{\pi_k} = L_{\pi_{k-1}} - \{\pi_k\}; \ k \in \{1, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:LatItem}
	\epsilon_{\pi_k}^t = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{j=1}^{n_{\pi_k}} \Big ( X_{ij}^{\pi_k} \times w_j \times (i\%2) - X_{ij}^{\pi_k} \times w_j \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}
\end{equation}

\begin{equation} \label{eq:LatPacked}
	\epsilon_{\pi_k}^a = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{q=1}^{m_{\pi_k}} \Big ( Y_{iq}^{\pi_k} \times w_q \times (i\%2) - Y_{iq}^{\pi_k} \times w_q \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}
\end{equation}

\begin{equation} \label{eq1}
	\max_{\pi \in S_K} f_\pi = \tilde{s}_\pi/\tilde{c}_\pi
\end{equation}

Longitudinal (\ref{eq:tau}) and lateral torques (\ref{eq:LatItem}, \ref{eq:LatPacked}) are calculated in proportion to the highest torque supported by the aircraft. As there are two rows of pallets, one on each side of the centerline, we use the operator {\it modulo}\/ ($\%$) to calculate lateral torques. In our experiments, we found that the magnitude of these lateral torques was always minimal, so we decided to ignore them in the fuel consumption (\ref{eq:costs}). The objective of ACLP+RPDP (\ref{eq1}) is to find a permutation $\pi \in S_K$ with the corresponding allocation of items on pallets at each node that maximises the function $f_\pi = \tilde{s}_\pi/\tilde{c}_\pi$.

Finally, ACLP+RPDP constraints related to each node ${\pi_k}$\/ are described in Table \ref{tab:constraints}.

\begin{table}[H]
	\centering
	\footnotesize
	\caption{Constraints} \label{tab:constraints}
	\begin{tabular}{ c l }
		\toprule
		\textbf{Constraint} & \textbf{Description} \\
		\midrule
		(\ref{eq:torqlong}, \ref{eq:torqlat}) & Longitudinal and lateral torques must be within aircraft limits \\
		\midrule
		(\ref{eq:app2}, \ref{eq:app3}) & Items allocated to each pallet cannot exceed its weight and volume limits\\
		\midrule
		(\ref{eq:app4}) & At most, each item is associated with a single pallet \\
		\midrule
		(\ref{eq:app5}) & Packed contents that have not yet reached their destination must remain on board\\
		\midrule
		(\ref{eq22}, \ref{eq23}) & Items allocated on the same pallet must have the same destinations\\
		\midrule
		(\ref{eq24}, \ref{eq25}) & If there is a packed content on the pallet, it must also have the same destination as other items\\

		\bottomrule
	\end{tabular}
	\normalsize
\end{table}

\begin{equation} \label{eq:torqlong}
	| \tau_{\pi_k} | \leq 1;\ k \in \{0, \ldots, K\}
\end{equation}

\begin{equation} \label{eq:torqlat}
	| \epsilon_{\pi_k}^t + \epsilon_{\pi_k}^a| \leq  1
\end{equation}

\begin{equation} \label{eq:app2}
	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}
\end{equation}

\begin{equation} \label{eq:app3}
	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}
\end{equation}

\begin{equation} \label{eq:app4}
	\sum_{i=1}^{m} X_{ij}^{\pi_k} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq:app5}
	\sum_{i=1}^{m} Y_{iq}^{\pi_k} = 1;\ to_q \in L_{\pi_k}; \ q \in \{1, \ldots, m_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq22}
	X_{ij}^{\pi_k} <= X_{ij}^{\pi_k} \times (T_i^{\pi_k} - to_j + 1); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq23}
	X_{ij}^{\pi_k} <= X_{ij}^{\pi_k} \times (to_j - T_i^{\pi_k} + 1); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq24}
	Y_{iq}^{\pi_k} <= Y_{iq}^{\pi_k} \times (T_i^{\pi_k} - to_q + 1); \ i \in \{1, \ldots, m\}; \ q \in \{1, \ldots, m_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq25}
	Y_{iq}^{\pi_k} <= Y_{iq}^{\pi_k} \times (to_q - T_i^{\pi_k} + 1); \ i \in \{1, \ldots, m\}; \ q \in \{1, \ldots, m_{\pi_k}\}
\end{equation}

Once the assumptions and the mathematical modelling are presented, it is possible to see that ACLP+RPDP is {\it NP-hard}. In a similar way to \citep{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$\/ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in node 0, and no items in node 1. Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is {\it NP-complete}. For more details, see \citep[p.~6]{LurkinSchyns2015}.

\section{Solution strategies}
\label{algorithms}

Throughout our research, we have thoughtfully described ACLP+RPDP in standard MIP format and found that no solver can handle its practical cases in a feasible time. Thus, as ACLP+RPDP is highly complex, involving four intractable and interconnected sub-problems, we decided to focus only on {\it real cases}, developing quick node-by-node solutions, not necessarily optimal, but which would allow us to obtain a complete tour.

In practical cases, we know that a common aircraft has $m=18$\/ pallets, flight itineraries have $K \leq 6$\/ nodes plus the base, and each node has hundreds of items to be shipped. We also know that missions with fewer nodes are more frequent than longer ones. Under these circumstances, we can adopt some important strategies:

\begin{itemize}

	\item We consider that the number of destinations is smaller than the number of pallets, that is, $K < m$. We also avoid the trivial case where $K=1$. With this premise, we can preset the destinations of the pallets at each shipping node, reserving for each destination a number of pallets proportional to the volume available. We could have used another criterion, but it was observed in the experiments that the volume is more constrictive in airlift.

	\item As the number of nodes is small, we have the possibility to check all possible tours, selecting the one that provides the best value for the objective function. For this reason, a maximum runtime $tmax$\/ per tour is established. In the tests carried out in this work, $tmax \leq 3600s$.

	\item At each node, we will use a MIP solver or heuristics to check which method offers the best result within the runtime limit. As there are $K!$ tours, each of them with $K+1$\/ legs, each solution per node will have a runtime limit equal to $tmax/(K+1)!$.

	\item Just for speculation, we will especially test the two tours with the shortest distance, corresponding to the TSP solutions at the nodes considered. In these cases, the runtime limit per node will be $tmax/(K+1)$.

\end{itemize}

Our complete strategy is summarized in Algorithm \ref{alg:main}.

\begin{algorithm}[H]

	\caption{Solving the ACLP+RPDP}  \label{alg:main}

	\begin{algorithmic}[1]
			\State {\bf ACLP+RPDP} \,\, {\it in:}~$scenario, surplus, tmax$ \, \, {\it out:}~$answer1, answer2$

			\State Let $M$ be the set of pallets ({\it cfr.} Table \ref{tab:larger}) \label{main:M}

			\State Let $K$, $L$\/ and $C$ be according to the $scenario$ ({\it cfr.} Tables \ref{tab:costs}, \ref{tab:larger} and \ref{tab:scenarios}) \label{main:KLC}

			\State Let $\pi_{TSP1}$\/ and $\pi_{TSP2}$ be the shortest tours ({\it cfr.} Table \ref{tab:scenarios})  \label{main:shortest}

			\State $N \gets ItemsGeneration(scenario, surplus)$ \label{main:items}

			\For {each $method$}  \label{main:method}

				\State $f_1 \gets SolveTour(\pi_{TSP1}, L, M, C, N, method, tmax)$  \label{main:f1}

				\State $f_2 \gets SolveTour(\pi_{TSP2}, L, M, C, N, method, tmax)$  \label{main:f2}

				\State $answer1[scenario,surplus,method] \gets max( f_1, f_2  )$  \label{main:a1}

				\For {each $\pi \in S_K$} \label{main:loop1}

					\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, tmax/K!)$ \label{main:fpi}

				\EndFor \label{main:loop2}

				\State $answer2[scenario,surplus,method] \gets \max f_{\pi}$ \label{main:a2}
			\EndFor
	\end{algorithmic}

\end{algorithm}

In this algorithm, we use five values for $scenario$, according to Tables \ref{tab:costs} and \ref{tab:scenarios}, which define the number $K$ of destinations, the set $L$\/ of nodes, the costs $C$\/, and the shortest tours $\pi_{TSP1}$\/ and $\pi_{TSP2}$.

\begin{table}[h]
	\centering
	\caption{Testing scenarios}  \label{tab:scenarios}
	\begin{tabular}{c c c c c}
		\toprule
		{\bf Scenario} & $K$ & $L$ & $\pi_{TSP1}$ & $\pi_{TSP2}$ \\		
		\midrule
		1 & 2    & \{$0,1,2$\}           & 0 1 2 0  & 0 2 1 0 \\
		2 & 3    & \{$0,1,2,3$\}         & 0 1 2 3 0 & 0 3 2 1 0\\
		3 & 4    & \{$0,1,2,3,4$\}       & 0 4 1 2 3 0 & 0 3 2 1 4 0 \\
		4 & 5    & \{$0,1,2,3,4,5$\}     & 0 4 1 2 5 3 0 & 0 3 5 2 1 4 0 \\
		5 & 6    & \{$0,1,2,3,4,5,6$\}   & 0 4 1 2 6 5 3 0 & 0 3 5 6 2 1 4 0 \\
		\bottomrule
	\end{tabular}
\end{table}

$surplus$\/ is a value in $\{1.2, 1.5, 2.0\}$, which corresponds, at each node $k$, to the ratio between the sum of the volumes of the items  and the load capacity of the pallets ($surplus = \sum_{j=1}^{n_k} v_j$/$\sum_{i=1}^{m} V_i$). This parameter allows us to verify the different behaviour of each $method$, according to $scenario$\/ and the quantity of items available for shipment. It is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).

$tmax$\/ is the runtime limit, which will be distributed among the tours (lines \ref{main:f1}, \ref{main:f2} and \ref{main:fpi}). $method$\/ corresponds to a MIP solver or a heuristic to the node-by-node solution $SolveTour$, which will be presented in subsection \ref{methods}.

The best results corresponding to the shortest tours are stored in $answer1$\/ (line \ref{main:a1}), and those obtained by testing all $K!$\/ tours are stored in $answer2$\/ (line \ref{main:a2}).

Next, we will present two subsections: in the first we explain how $SolveTour$\/ is executed, presetting the destinations of the pallets. In the second we will present the heuristics developed for node-by-node solutions.

\subsection{SolveTour algorithm}
\label{tour}

As we commented in the previous subsection, we will adopt the strategy of presetting the destinations of each pallet throughout the tour. This is feasible in practical cases where $1<K<m$. For this, each pallet $i$\/ also has a field $T^k_i$, $0\leq k \leq K$, which stores its next destination after being loaded at node $k$. For this reason, $T^k_i \in L_k$, $1 \leq i \leq m$, $0\leq k \leq K$.

$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $tmax$\/ is the runtime limit that will be distributed among the $K+1$\/ legs of the tour.

\begin{algorithm}[H]
	\caption{Solving the tour $\pi$ with $method$}  \label{alg:tour}
	\begin{algorithmic}[1]

		\State {\bf SolveTour} \,\, {\it in: $\pi, L, M, C, N, method$, $tmax$} \, \, {\it out: $score / cost$}

		\State $\pi_0     \gets 0$ \label{tour:pi1} \Comment{all tours start and end at the base}
		\State $\pi_{K+1} \gets 0$ \label{tour:pi2}

		\State $score \gets 0$ \label{tour:score}
		\State $cost  \gets 0$ \label{tour:cost}

		\For {$k \gets 0$ to $K$} \label{tour:loop1}	
		\State $L_{\pi_k} \gets L - \{\pi_0,\pi_1,\ldots,\pi_k\}$  \label{tour:lk1}	 \Comment{the set of remaining nodes is updated}	
		\State $T_i^{\pi_k} \gets -1$, $1 \leq i \leq m$ \label{tour:-1}             \Comment{the pallet destination is unset}

		\If {$k = 0$}
			\State Let $G_1(M \cup N_0, \varnothing)$ \label{tour:g11}               \Comment{no packed contents at the base}
		\Else
			\State $E_{Q_{\pi_k}}, M \gets UpdatePacked(M, Q_{\pi_k}, \pi_k)$ \label{tour:dest}			
			\State Let $G_1(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$ \label{tour:g12}
		\EndIf  \label{tour:lk2}

		\State $M \gets SetPalletsDestinations(M, \pi_k )$ \label{tour:dest2}

		\State $G_2 \gets SolveNode(method, \pi_k, G_1, tmax/(K+1))$  \label{tour:SolveNode}

		\State $s, \tau \gets ScoreAndTorque(\pi_k, G_2)$ \label{tour:analyse}

		\State $score \gets score + s$ \label{tour:score2}
		\State $cost  \gets cost  + c_{\pi_k,\pi_{k+1}} \times (1 + c_g \times |\tau|)$ \label{tour:cost2}

		\EndFor  \label{tour:loop2}

	\end{algorithmic}
\end{algorithm}

As we mentioned in the previous section, all tours start and end at the base $0$\/ (lines \ref{tour:pi1}-\ref{tour:pi2}). After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ flights (lines \ref{tour:loop1}-\ref{tour:loop2}). Initially, the set $L_{\pi_k}$\/ of remaining nodes is updated (line \ref{tour:lk1}), and the pallet destinations are unset (line \ref{tour:-1}).

When the aircraft is at the base, the initial graph $G_1$\/ is empty, and there are no packed contents \ref{tour:g11}. Otherwise, $UpdatePacked$\/ (line \ref{tour:dest}) returns the set of packed contents that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation. This allocation is stored in graph $G_1$\/ (line \ref{tour:g12}).

$SetPalletsDestinations$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node without changing the pallet's destination with packed contents.

Finally, $SolveNode$\/ includes the edges corresponding to the items shipped at the current node, returning the graph $G_2$\/ (line \ref{tour:dest2}). The score and the CG deviation of $G_2$\/ are calculated (line \ref{tour:analyse}) and accumulated (lines \ref{tour:score2}-\ref{tour:cost2}), allowing the final result of this tour as output.

$UpdatePacked$, described in Algorithm \ref{alg:cons}, finds the best packed-pallet allocation, in terms of CG deviation, for the packed contents that remain on board.

\begin{algorithm}[H]
	\caption{Updating the packed contents that remain boarded at node $\pi_k$}  \label{alg:cons}
	\begin{algorithmic}[1]

		\State {\bf UpdatePacked} \,\, {\it in: ${M, Q_{\pi_k}, \pi_k}$} \, \, {\it out: $E_{Q_{\pi_k}}, M$}

		\State $E_{Q_{\pi_k}} \gets MinCGDeviation(E_{Q_{\pi_k}})$   \label{cons:mincg}

		\For{$i \gets 1$ to $m$} \label{cons:Ybegin}

		\For{$q \gets 1$ to $m_{\pi_k}$}

		\State $T_i^{\pi_k} \gets -1$

		\If{$(i, q) \in E_{Q_{\pi_k}}$} 
		\State $T_i^{\pi_k} \gets to_q$   \label{cons:dest} \Comment{reassign pallet destinations}
		\EndIf
		\EndFor		
		\EndFor \label{cons:Yend}

	\end{algorithmic}
\end{algorithm}

$MinCGDeviation$\/ (line \ref{cons:mincg}) relocates the packed contents on the pallets, minimizing torque and ensuring that they all remain on board, one packed content on each pallet. It is run through a MIP solver with the objective function (\ref{eq:torque}) and the constraints (\ref{eq:embarked}) and (\ref{eq:one}). As there are few variables, $E_{Q_{\pi_k}}$\/ is obtained in less than $30$\/ milliseconds. Finally, the destination of each pallet with packed content is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).

\begin{equation} \label{eq:torque}
	\min f =  \Big | \sum_{i=1}^{m} \sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \times w_q \times D_i^{long}  \Big |
\end{equation}

\begin{equation} \label{eq:embarked}
	\sum_{i=1}^{m} Y^k_{iq} = 1;\ q \in \{1,\ldots,m_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq:one}
	\sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \leq 1;\ i \in \{1,\ldots,m\}
\end{equation}

$SetPalletsDestinations$, which sets the pallets destination not yet defined, is described in Algorithm \ref{alg:dest}.

\begin{algorithm}[H]
	\caption{Setting pallets destination based on the items to be embarked at node $\pi_k$}  \label{alg:dest}
	\begin{algorithmic}[1]

		\State {\bf SetPalletsDestinations} \,\, {\it in: $M, \pi_k$} \, \, {\it out: $M$}

		\State $vol_x \gets  0$, $x \in L_{\pi_k}$  \label{dest:vector1}

		\State $max \gets 0$  \Comment{destination with maximum volume demand}

		\State $total \gets 0$

		\For{$j \gets 1$ to $n_{\pi_k}$}

		\If {$to_j \in L_{\pi_k}$}

		\State $vol_{to_j} \gets vol_{to_j} + v_j$ 
		\State $total \gets total + v_j$

		\If {$vol_{to_j} > vol_{max}$}

		\State $max \gets to_j$ \label{dest:max} 
		\EndIf

		\EndIf

		\EndFor

		\For{$x \in L_{\pi_k}$} \label{dest:propor1}

		\If {$vol_{x} \neq 0$}

		\State $needed \gets \max \{1, \lfloor{ (m-m_{\pi_k}) \times vol_{x}/total}\rfloor \}$  
		\label{dest:needed}
		\State $np \gets 0$

		\For{$i \gets 1$ to $m$}

		\If{($np < needed$) {\bf and} ($T_i^{\pi_k} = -1$)}  
		\State $T_i^{\pi_k} \gets x$
		\State $np \gets np + 1$ \Comment{number of necessary pallets to node $x$}
		\EndIf

		\EndFor
		\EndIf

		\EndFor \label{dest:propor2}

		\For{$i \gets 1$ to $m$} \label{dest:max1}
		\If{$T_i^{\pi_k} \gets -1$} 
		\State $T_i^{\pi_k} \gets max$  \Comment{any remaining pallet is assigned to the maximum demand destination}
		\EndIf
		\EndFor \label{dest:max2}

	\end{algorithmic}
\end{algorithm}

$vol$\/ stores the demand volume of items destined for the non-visited nodes (line \ref{dest:vector1}). The destination of empty pallets is defined proportionally to the volume of items to be embarked (lines \ref{dest:propor1}-\ref{dest:propor2}). $max$\/ is the destination with maximum volume demand (line \ref{dest:max}), and $needed$\/ is the number of necessary pallets to node $x$ (line \ref{dest:needed}). The destination with the maximum volume defines any remaining pallets (lines \ref{dest:max1}-\ref{dest:max2}).

$ScoreAndTorque$, described in Algorithm \ref{alg:eval}, evaluates the allocation graph $G$\/ generated by $SolveNode$\/ at node $\pi_k$\/ and returns the corresponding cargo score and aircraft torque.

\begin{algorithm}[H]
	\caption{Cargo score and aircraft torque}  \label{alg:eval}
	\begin{algorithmic}[1]

		\State {\bf ScoreAndTorque} \,\, {\it in: $\pi_k, G$} \, \, {\it out: $s, \tau$}

		\State Let $G(V_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$

		\State $s \gets 0$
		\State $\tau_i \gets 0$, $1 \leq i \leq m$
		\For{$i \gets 1$ to $m$} \label{eval:loop1}

		\For{$j \gets 1$ to $n_{\pi_k}$}

		\If{$X_{ij}^{\pi_k} = 1$}
		\State $s \gets s + s_j$ \label{eval:score1} \Comment{accumulates cargo score}
		\State $\tau_i \gets \tau_i + w_j \times D_i^{long}$  \label{eval:eps1} \Comment{accumulates aircraft torque}
		\EndIf
		\EndFor

		\For{$q \gets 1$ to $m_{\pi_k}$}
		\If{$Y_{iq}^{\pi_k} = 1$}
		\State $s \gets s + s_q$   \label{eval:score2} \Comment{accumulates cargo score}
		\State $\tau_i \gets \tau_i + w_q \times D_i^{long}$ \label{eval:eps2} \Comment{accumulates aircraft torque}
		\EndIf
		\EndFor	   
		\EndFor 		 \label{eval:loop2}
		\State $\tau \gets  \sum_{i=1}^{m} \tau_i / (W_{max} \times limit^{CG}_{long})$ \label{eval:torque} \Comment{final calculation of the aircraft torque}

	\end{algorithmic}
\end{algorithm}

Algorithm \ref{alg:eval}\/ consists of a loop that goes through all the pallets (lines \ref{eval:loop1}-\ref{eval:loop2}), accumulating the scores (lines \ref{eval:score1} and \ref{eval:score2}) and the torques (lines \ref{eval:eps1} and \ref{eval:eps2}) of the shipped items, allowing the final calculation of the aircraft torque (line \ref{eval:torque}).

\subsection{Node-by-node solutions}
\label{methods}

In this subsection, we present two implementations of $SolveNode$\/ algorithm: with a MIP solver and with heuristics.

\subsubsection{Node-by-node solutions with a MIP solver}
\label{solver}

Our strategy adopted in $SolveTour$\/ defines the values of some variables: the set of nodes to be visited is updated, the packed contents that remain on board are reallocated to minimize the CG deviation, and the pallet's destinations are determined according to the volume of items available for shipment.

In this way, the mathematical model for $SolveNode(MIP,\pi_k,G, tmax)$\/ becomes simpler, which finds an allocation of available items at the node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$. Thus, we use a MIP solver with a runtime limit $tmax$\/ at the node $\pi_k$ to maximize the objective function (\ref{eq:obj2}) with the calculus equations (\ref{eq:newf}) to (\ref{eq:costs2}), subject to the constraints (\ref{eq:tau2}) to (\ref{eq:ifB}). The binary variables $X_{ij}$\/ and $Y_{iq}$\/ define the sets of edges $E_{N_{\pi_k}}$\/ and $E_{Q_{\pi_k}}$, respectively, included in graph $G$.

\begin{equation} \label{eq:obj2}
	\max f= \tilde{s} / \tilde{c}
\end{equation}

\begin{equation} \label{eq:newf}
	\tilde{s} = \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij} \times s_j
\end{equation}

\begin{equation} 
	\tau_{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q) \Big ] \Big / W_{max} \times limit^{CG}_{long}
\end{equation}

\begin{equation} \label{eq:costs2}
	\tilde{c} =  c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|)
\end{equation}

\begin{equation} \label{eq:tau2}
	|\tau_{\pi_k}| \leq 1
\end{equation}

\begin{equation} 
	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}
\end{equation}

\begin{equation} 
	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}
\end{equation}

\begin{equation} 
	\sum_{i=1}^{m} X_{ij} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq:if2}
	X_{ij} = 0;\ to_j \notin L_{\pi_k}; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq:ifA}
	X_{ij} \leq X_{ij} \times (T_i^{\pi_k} - to_j + 1); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
\end{equation}

\begin{equation} \label{eq:ifB}
	X_{ij} \leq X_{ij} \times (to_j - T_i^{\pi_k} + 1 ); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
\end{equation}

The constraints (\ref{eq:ifA}) and (\ref{eq:ifB}) are equivalent to $X_{ij} =1$\/ if $to_j = T_i^{\pi_k}$, and $X_{ij} =0$\/ otherwise .

\subsubsection{Node-by-node solutions with heuristics}

One of the main objectives of this work was to find a quick heuristic that offers a good-quality solution for the node-by-node problem. Taking this into account, we design algorithms based on four known meta-heuristics: {\it Ant Colony Optimization} (ACO) \citep{Dorigo1992,DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \citep{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \citep{Glover1986}, and {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \citep{FeoResende1989}. We considered several ideas from the literature \citep{NiarFreville1997,Fidanova2006,Alonso2019, Zhan2020}, and we were careful to use the same data structures and procedures in all implementations to enforce fair results comparison.

However, the heuristic that presented better solutions was none of the previous ones. In this subsection, we will present a new heuristic for the node-by-node problem, called $Shims$. Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}). This strategy is based on a practical observation: usually, subsets of smaller and lighter items are saved for later adjustments to the remaining available space.

The selection of edges for $E_{ N_{\pi_k} }$ uses the {\it edge attractiveness}\/ $\theta_{ij}$\/ (\ref{eq:edge}), which can be understood as the tendency to allocate the item $j$\/ to the pallet $i$\/ at the node $\pi_k$. It is directly proportional to the score, and inversely to the volume and the torque of each item.

\begin{equation} \label{eq:edge}
	\theta_{ij} = \frac{ s_j }{ v_j }\ \times \Big ( 1 - \frac{ w_j \times |D_i^{long}| }{ \max w_j \times\  \max | D_i^{long} | } \Big );\ i \in \{1,\ldots,m\},\ j \in \{1,\ldots,n_{\pi_k}\}
\end{equation}

\begin{table}[H]

	\begin{minipage}{0.08\linewidth}

	\end{minipage}\hfill
	\begin{minipage}{0.34\linewidth}

		\includegraphics[scale=0.25]{shims.png}
		\captionof{figure}{{\it Shims} of various thicknesses}
		\small\textsuperscript{Source: www.mscdirect.com/product/details/70475967}
		\label{fig:shims}

	\end{minipage}\hfill
	\begin{minipage}{0.58\linewidth}
		\scriptsize
		\begin{tikzpicture}[scale=0.75, samples=100]
			\filldraw[fill=blue!3!white, draw=black] (0, 0) rectangle (12.5, 5.5);
			\draw[->] (.2, .6) -- coordinate (x axis mid) (12, .6);
			\node at (5, 0.3) {$\eta_1$};
			\node at (5, 2.5) {|};
			\node at (6, 2.5) {\small{shims}};
			\node at (7, 2.5) {|};
			\node at (7, 0.3) {$\eta_2$};
			\node at (0.3, 5) {$\theta_{ij}$};
			\draw[->] (.6, .2) -- coordinate (y axis mid) (0.6, 5.3);
			\node at (12, 0.3) {$e_{ij}$};
			\draw[smooth, domain = 0.09:2, color=black, thick] plot (.3+1/\x,{4.2+log2(\x)});
			\draw[->, dashed] (5, 0.6)--(5, 2.0);
			\draw[->, dashed] (7, 0.6)--(7, 1.5);
		\end{tikzpicture}

		\captionof{figure}{$n_{\pi_k}$\/ possible edges $e_{ij}$\/ sorted by $\theta_{ij}$\/ in non-ascending order}
		\label{fig:whip}		
	\end{minipage}
\end{table}

Considering only the items that can be shipped at the node $\pi_k$, Figure \ref{fig:whip} represents $n_{\pi_k}$\/ possible edges $e_{ij}$\/ of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order. Initially, $Shims$\/ builds a greedy solution for the pallet $i$\/ selecting edges up to index $\eta_1$ ({\it greedy phase}). Then, with the edges between $\eta_1$\/ and $\eta_2$, it elaborates different possible complements ({\it composition phase}), including later the best ones in the same pallet ({\it selection phase}). $Shims$\/ is depicted in Algorithm \ref{alg:shims}.

\begin{algorithm}[H]
	\caption{$Shims$\/ heuristic at node $\pi_k$}  \label{alg:shims}
	\begin{algorithmic}[1]

		\State {\bf SolveNode} \,\, {\it in: $Shims, \pi_k, G, tmax, {\color{blue}threshold_1, threshold_2}$} \, \, {\it out: $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$}

		\State $T_{begin} \gets$ current system time

		\State Let $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$   \label{shims:initQ}

		\State Sort $M$ by $|D_i^{long}|$ in non-descending order \label{shims:pallets}

		\State $E_{N_{\pi_k}} \gets \varnothing$

		\State $\tau_{max} \gets W_{max} \times limit^{CG}_{long}$

%		\State $limit \gets 0.92$  \label{shims:limit}

		\For{$i \gets 1$ to $m$}  \label{shims:eta1_a}

		\State $\tau_{\pi_k} \gets \sum_{(i,q)\in E_{Q_{\pi_k}}} w_q \times D_i^{long}$ 
		\State $vol_i \gets \sum_{(i,q)\in E_{Q_{\pi_k}}} v_q$

		\State Let $E$\/ be an array of $n_{\pi_k}$ possibles edges of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order \label{shims:possible}
		\State $\eta_1 \gets 1$  
		\Repeat
		\State $e_{ij} \gets E_{\eta_1}$

		\If{ ($E_{N_{\pi_k}} \cup \{e_{ij}\}$ is feasible) {\bf and} ($vol_i \leq V_i \times {\color{blue}threshold_1}$) {\bf and} ($| \tau_{\pi_k} + w_j \times D_i^{long} | \leq W_{max} \times limit^{CG}_{long}$)}

		\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup \{e_{ij}\}$

		\State $vol_i \gets vol_i + v_j$

		\State $\tau_{\pi_k} \gets \tau_{\pi_k} + w_j \times D_i^{long}$

		\State $\eta_1 \gets \eta_1 + 1$

		\EndIf
		\Until ($vol_i > V_i \times {\color{blue}threshold_1}$) {\bf or} ($\eta_1>n_{\pi_k}$) \label{shims:endfirst}

		\State $slack_i \gets V_i - vol_i$  \label{shims:beginsecond}
		\State $\eta_2 \gets \eta_1$  \label{shims:eta2a}
%		\While{($\eta_2 \leq n_{\pi_k}$) {\bf and} ($ vol_i < (1 + 2 \times (1-limit)) \times V_i$)}
{\color{blue}
		\While{($\eta_2 \leq n_{\pi_k}$) {\bf and} ($ vol_i < V_i \times {\color{blue}threshold_2}$)}

		\State $e_{ij} \gets E_{\eta_2}$

		\State $vol_i \gets vol_i + v_j$
		\State $\eta_2 \gets \eta_2 + 1$  \label{shims:eta2} 
		\EndWhile \label{shims:endsecond} 
	}
		\State $vol \gets 0$; $b \gets 1$; $shims_b \gets \varnothing$; $Set \gets \{ shims_b \}$  \label{shims:beginthird}

		\For{$x \gets \eta_1$ to $\eta_2$} \label{edges_indexes}

		\If{ $T_{current} - T_{begin} > tmax$}
		\State {\bf break}
		\EndIf

		\State $NewShims \gets$ {\bf True} \label{new_shims}
		\State $e_{ij} \gets E_x$

		\For{$shims \in Set$} \label{shims_set}

		\If {($e_{ij} \not\in (E_{N_{\pi_k}} \cup shims))$ {\bf and} ($e_{ij}$ is feasible)  {\bf and} ($(v_j + vol) \leq slack_i$)}

		\State $shims \gets shims \cup \{e_{ij}\}$
		\State $vol \gets vol + v_j$
		\State $NewShims \gets$ {\bf False} \label{new_shims_false}

		\State {\bf break}  
		\EndIf

		\EndFor

		\If{$NewShims$} \label{new_shims2}
		\State $vol \gets 0$; $b \gets b + 1$;  $shims_b \gets \{e_{ij}\}$
		\State $Set \gets Set \cup \{ shims_b \}$
		\EndIf

		\EndFor

		\State $sh_w \gets shims$, where $shims \in Set$ and $\sum_{e_{ij} \in shims} w_j$\/ is maximum  \label{best_weight}		
		\State $sh_v \gets shims$, where $shims \in Set$ and $\sum_{e_{ij} \in shims} v_j$\/ is maximum \label{best_volume}
		\State $sh_{best} \gets shims$, where $shims \in \{sh_w, sh_v\}$\/ and $\sum_{e_{ij} \in shims} s_j$\/ is maximum \label{best_score}
		\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup sh_{best}$ \label{shims:endthird}

		\EndFor

	\end{algorithmic}
\end{algorithm}

{\color{blue}The arguments $threshold_1$ and $threshold_2$ are the volume thresholds for $\eta_1$ and $\eta_2$ calculations, respectively.

The argument $tmax$ is the maximum time for Shims to solve a node. It is proportional to the sum of the candidate items volumes in each node, in relation to the sum of all nodes $tmax$, and the maximum time to solve the tour.

}

Initially, $Q_{\pi_k}$\/ (line \ref{shims:initQ}) corresponds to the packed contents that remain on board. It is important to remember that $E_{Q_{\pi_k}}$\/ and $M$\/ were modified by the procedure $UpdatePacked(M, Q_{\pi_k}, \pi_k)$ and the procedure $SetPalletsDestinations(M, \pi_k)$. Then, the pallets $i$\/ are considered in non-descending order of $|D_i^{long}|$.

For each pallet $i$, its $n_{\pi_k}$\/ possible edges $e_{ij}$\/ are considered in non-increasing order of $\theta_{ij}$:

\begin{itemize}

	\setlength\itemsep{-0.3em}

	\item In the {\it greedy phase}\/ (lines \ref{shims:pallets}-\ref{shims:endfirst}), a partial solution for each pallet $i$\/ is constructed by adding edges following $\theta_{ij}$\/ order.
	The $\eta_1$\/ index corresponds to the accumulated volume equal to $V_i \times {\color{blue}threshold_1}$. In $\eta_2$\/ index, this same accumulated volume reaches $V_i \times  {\color{blue}threshold_2}$. {\color{blue}The values of these thresholds were defined empirically by the {\it iRace}\/ tool \citep{LopezIbanezManuel2016}}.

	{\color{blue}
	iRace (Iterated Racing for Automatic Algorithm Configuration) streamlines the tuning of optimization algorithms by iteratively testing and eliminating weaker parameter configurations through statistical comparisons. It focuses computational resources on the most promising settings. This method significantly cuts down the time and effort needed to find optimal or near-optimal parameters, proving effective across various application domains.
}

	\item In the {\it composition phase}\/ (lines \ref{shims:beginsecond}-\ref{shims:endsecond}), a set of shims named $Set$\/ is created for each pallet $i$, where each shim is formed by a set of edges in the range $[\eta_1,\eta_2]$, whose total volume is limited by $slack_i$. In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it First-Fit Decreasing}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \citep{JohnsonGarey1985}. Basically, shims are created by accumulating the following edges, taking $slack_i$\/ as a limit.

	\item In the {\it selection phase}\/ (lines \ref{shims:beginthird}-\ref{shims:endthird}), the best shim in $Set$\/ is chosen. Initially, two shims are found: $sh_w$\/ with larger weight and $sh_v$\/ with larger volume. Between the two, the one with the highest score will be chosen, and its edges will be inserted into $E_{N_{\pi_k}}$.

\end{itemize}

\section{Implementation and results}
\label{results}

This section is composed of two parts: the generation of the test instances and the results obtained in our implementation.

\subsection{Instances generation}
\label{items}

As we are dealing with a new problem that until now had not been modelled in the literature, we have to create our own benchmarks. For this, we based it on the characteristics of real airlifts carried out by the {\it Brazilian Air Force}, as described below.

In the delivery of supplies carried out in Brazil from 2008 to 2010, 23\% of the items weighed between $10kg$ and $20kg$, 22\% from $21kg$ to $40kg$, 24\% from $41kg$ to $80kg$, 23\% from $81kg$ to $200kg$, and 8\% between $201kg$ and $340kg$. These five groups of items are described in Table \ref{tab:weights}, where $P$\/ represents the group probability. On the other hand, the average density of these items is approximately $246 kg/m^3$.

\begin{table}[h]
	\centering
	\caption{Items weight distribution}  \label{tab:weights}
	\begin{tabular}{c c c c }
		\toprule
		$x$ & $P$ & $low$ ($kg$) & $high$ ($kg$) \\		
		\midrule
		1              & 0.23           & 10  & 20  \\
		2              & 0.22           & 21  & 40  \\
		3              & 0.24           & 41  & 80  \\		
		4              & 0.23           & 81  & 200 \\
		5              & 0.08           & 201 & 340 \\
		\bottomrule
	\end{tabular}
\end{table}

In the generation of test instances, we use two types of random selections:
\begin{itemize}
	\setlength\itemsep{-0.3em}
	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$, where $i_1$ and $i_2$ are integer numbers;
	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
\end{itemize}

The procedure $ItemsGeneration$, which generates $N$ (all items to be moved among the nodes), is described in Algorithm \ref{alg:itemsgen}.

\begin{algorithm}[H]
	\caption{Generating items}  \label{alg:itemsgen}
	\begin{algorithmic}[1]
		\State {\bf ItemsGeneration} \,\, {\it in: $scenario, surplus$} \, \, {\it out: $N$}
		\State Let $L$ be the set of nodes and $M$ the set of pallets  \label{ig:LM}
		\State $limit \gets surplus \times \sum_{i=1}^{m} V_i$ \label{ig:extended}
		\For{$k \gets 0$ to $K$}
		\State $N_k \gets \varnothing$
		\State $j \gets 0$
		\State $vol \gets 0$	\label{ig:totals}	
		\While{$vol < limit$}
		\State $j \gets j+1$
		\State Let $t_j^k$\/ be the item $j$\/ at the node $k$
		\Repeat
		\State $to_j \gets RandomInt(0, K)$ \label{ig:dest}
		\Until{$to_j \neq k$}
		\State $x = Roulette()$ \Comment{biased through $P$ ({\it cfr.} Table \ref{tab:weights})} \label{ig:weight1} 
		\State $w_j \gets RandomInt(low(x), high(x))$        \label{ig:weight2}		
		\State $s_j \gets \round{100 \times (1 - \log_{10}(RandomInt(1, 9)))} $ \label{ig:score}
		\State $v_j \gets w_j / RandomInt(148, 344)$ \label{ig:volume}
		\State $vol \gets vol + v_j$ 
		\State $N_k \gets N_k \cup \{t_j^k\}$ 
		\EndWhile
		\EndFor
		\State $N \gets \bigcup_{0 \leq k \leq K} N_k$
	\end{algorithmic}
\end{algorithm}

$scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and the argument $surplus$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}). To avoid simply loading all items, we use $surplus \in \{1.2,\ 1.5,\ 2.0\}$\/. This also represents more instances for tests in each scenario.

For each generated $t^k_j$ item, its destination is randomly selected (line \ref{ig:dest}), its weight has a distribution according to Table \ref{tab:weights} (lines \ref{ig:weight1}-\ref{ig:weight2}), its score varies $100$\/ (highest) and $5$\/ (lowest) according to a logarithmic scale (line \ref{ig:score}), and its volume is randomly defined from the density, where we allow a variation of 40\% around the average density of $246 kg/m^3$\/ (line \ref{ig:volume}).

\subsection{Obtained solutions: quality and runtimes}

In the tests performed, we used a 64-bit, 16 GB, 3.6 GHz, eight-core processor with {\it Linux Ubuntu 22.04.1 LTS 64-bit} as the operational system and {\it Python 3.10.4}\/ as the programming language. We also used the well-known solver $Gurobi$\/ ({\tt www.gurobi.com}), version 9.5.2.

We ran Algorithm \ref{alg:main} considering the five scenarios from Table \ref{tab:scenarios}, three values for  $surplus$\/ from $\{1.2, 1.5, 2.0\}$, four values for $tmax$\/ from $\{240s, 1200s, 2400s, 3600s\}$, and six different $method$\/ for the node-by-node solution: $Gurobi$\/ (see \ref{solver}), ACO, NMO, TS, GRASP, and $Shims$\/ (Algorithm \ref{alg:shims}).

For $Gurobi$\/ to be able to solve the largest possible number of tests without memory overflow, we set its parameter {\it MIPgap}\/ to 1\%. This shortens its runtime, in addition to ensuring that its objective function $f$\/ is at most 1\% of the optimal solution. For more details, see {\tt www.support.gurobi.com}.

Among the heuristics, we will present only the results obtained by $Shims$, which had the best performance. For each $scenario$, $surplus$\/ and $tmax$, seven different instances were generated. We will present the average of the objective function $f$\/ and the runtime of $Gurobi$\/ and $Shims$\/ in these instances. To facilitate the comparison between both, we added a last column in the tables where two values are indicated:

\begin{itemize}
	\setlength\itemsep{-0.3em}
	\item {\bf Normalized}: value between 0 and 1, which corresponds to the ratio between the sum of $f$\/ values obtained by the method in all scenarios and the sum of the best values obtained by both methods in all scenarios. The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
	\item {\bf Speed-up}: ratio of the sums of the runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
\end{itemize}

We also indicate the two adopted strategies: dedicating all the processing time to the two shortest tours or distributing it among all $K!$\/ tours.

The results obtained with $tmax = 3600s$, which is the highest tested runtime limit, are in Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, with $surplus$\/ values of $1.2$, $1.5$\/ and $2.0$, respectively. We indicate with an {\bf x} the cases where $Gurobi$\/ did not find a feasible solution within this runtime limit or had to be aborted due to high random-access memory (RAM) usage.

\begin{table}[H]
	\centering
	\caption{Solutions with $surplus = 1.2$\/ and $tmax =  3600s$}  \label{tab:20}
	\footnotesize
	\begin{tabular}{ccccccccc}
		\toprule
		{\specialcell{{\bf Tested}\\{\bf tours}}}&$method$          & $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
		\toprule
		\multirow{4}{*}{Shortest} &\multirow{2}{*}{$Gurobi$} & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\
		&                                                               &               {time (s)}  & 17      & 17      & 16      & 17    & 26        &  1.0 \\
		\cmidrule(lr){2-9}
		&                                                               \multirow{2}{*}{$Shims$}    & $f$ & 8.54    & 11.63   & 12.97   & 13.43 & 12.36     &  0.99 \\
		&                                                               &         {time (s)}        & 1       & 1       & 2       & 2     & 2         &  11.6 \\

		\midrule

		\multirow{4}{*}{All $K!$}&                                  \multirow{2}{*}{$Gurobi$} & $f$ & 8.52    & 12.25   & 13.32   & 14.61  & \textbf{x}        & 1.00 \\
		&                                                               &               {time (s)}  & 25      & 36      & 121     & 297    & \textbf{x}        &  1.0 \\
		\cmidrule(lr){2-9}	
		&                                                               \multirow{2}{*}{$Shims$}    & $f$ & 8.54    & 12.26   & 13.23   & 14.53  & 13.42    & 0.99  \\
		&                                                               &         {time (s)}        & 1       & 2       & 6       & 17     & 106      & 18.4   \\
		\bottomrule

	\end{tabular}
	\normalsize
\end{table}

\begin{table}[H]
	\centering
	\caption{Solutions with $surplus = 1.5$\/ and $tmax =  3600s$}  \label{tab:50}
	\footnotesize
	\begin{tabular}{ccccccccc}
		\toprule
		{\specialcell{{\bf Tested}\\{\bf tours}}} & $method$          & $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
		\toprule
		\multirow{4}{*}{ Shortest } & \multirow{2}{*}{$Gurobi$} & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\
		&&               {time (s)}                                                                  & 38    & 30    & 27    & 29    & 81    & 1.00  \\
		\cmidrule(lr){2-9}	
		&\multirow{2}{*}{$Shims$}                                                                    & $f$ & 11.83 & 16.74 & 18.01 & 18.89 & 16.91 & 1.00  \\
		&&         {time (s)}                                                                        & 1     & 2     & 2     & 3     & 3     & 20.5  \\

		\midrule

		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$}                                  & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ 
		&&               {time (s)}                                                                  & 53    & 61    & 193   & 470   & 2241    & 1.0  \\
		\cmidrule(lr){2-9}	
		&\multirow{2}{*}{$Shims$}                                                                    & $f$ & 11.83 & 17.00 & 18.21 & 20.45 & 18.00   & 0.99  \\ 
		&&         {time (s)}                                                                        & 1     & 2     & 9     & 26    & 157     & 15.5 \\
		\bottomrule

	\end{tabular}
	\normalsize
\end{table}

\begin{table}[H]
	\centering
	\caption{Solutions with $surplus = 2.0$\/ and $tmax =  3600s$}  \label{tab:100}
	\footnotesize
	\begin{tabular}{ccccccccc}
		\toprule
		{\specialcell{{\bf Tested}\\{\bf tours}}}& $method$          & $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
		\toprule
		\multirow{4}{*}{Shortest} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.75 & 24.39 & 26.46 & 27.09 & 24.00 & 1.00 \\ 
		&&                                                                               {time (s)}  & 162   & 95    & 74    & 68    & 67    & 1.0  \\
		\cmidrule(lr){2-9}	
		&                                                                   \multirow{2}{*}{$Shims$} & $f$ & 17.78 & 24.40 & 26.36 & 27.01 & 23.94 & 0.99  \\
		&&                                                                                {time (s)} & 2     & 3     & 3     & 4     & 5     & 27.4  \\

		\midrule

		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &  \textbf{x}   & 1.00 \\ 
		&&                                                                                {time (s)} & 151   & 123   & 319   & 689   &  \textbf{x}   & 1.0  \\
		\cmidrule(lr){2-9}	
		&                                                                \multirow{2}{*}{$Shims$}    & $f$ & 17.78 & 25.25 & 26.38 & 28.98 & 26.08 & 1.00  \\
		&&                                                                                {time (s)} & 2     & 3     & 13    & 37    & 229   & 23.3  \\
		\bottomrule

	\end{tabular}
	\normalsize
\end{table}

From these data, we can draw some conclusions:

\begin{itemize}
	\setlength\itemsep{-0.3em}
	\item The strategy of testing all $K!$\/ tours often provide a better-quality solution, even with less time on each node. This shows that the four sub-problems are interconnected in such a way that it is not enough to solve them separately.
	\item $Gurobi$\/ fails in some cases when $scenario=5$\/ and the strategy is to check all $K!$\/ tours. This occurs because the runtime limit per node is smaller and there tend to be more packed contents on the aircraft, reducing the space for allocating items and making the solution difficult.
	\item When $Gurobi$\/ finishes, it finds the best solution, but the one obtained by $Shims$\/ reaches $99\%$\/ of that value.
	\item $Shims$\/ always finds a solution, being 11 to 27 times faster.
	\item All runtimes are much lower than the limit because the solution on many nodes can be fast. Anyway, in all the tests performed, the maximum time spent by $Shims$\/ did not reach 4 minutes. On the other hand, when $scenario=5$\/ and $surplus=1.5$, $Gurobi$\/ spent almost 40 minutes.
\end{itemize}

Table \ref{tab:100time} shows the results obtained with the strategy of testing the $K!$\/ tours in all scenarios with different $tmax$. We can observe more cases where $Gurobi$\/ fails, even in smaller scenarios. When $Gurobi$\/ finishes, $Shims$\/ finds a solution of similar quality ($98\%$\/ or better). In all cases, $Shims$\/ finds a good solution in less than 4 minutes.

\begin{table}[H]
	\centering
	\caption{Solutions testing all $K!$\/ tours with different runtime limits}  \label{tab:100time}
	\scriptsize
	\setlength{\tabcolsep}{3.5pt}
	\begin{tabular}{ccc|ccccc|ccccc|ccccc}
		\toprule
		&             \multicolumn{2}{r}{$surplus$} &\multicolumn{5}{c}{\bf 1.2}&\multicolumn{5}{c}{\bf 1.5}&\multicolumn{5}{c}{\bf 2.0} \\
		\midrule
		$method$& {$tmax$} & $scenario$  &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5} \\
		\toprule

		\multirow{10}{*}{$Gurobi$} & \multirow{2}{*}{240s} & $f$ & 8.53 & 12.24 & 13.34 &  x      &  x      & 11.81 & 17.02 & 18.31 &  x      &  x      & 17.75 & 25.24 & x     & x       & x \\ 
		&               &  {time (s)}                              & 41   & 43    & 123   &  x      &  x      & 50    & 62    & 181   &  x      &  x      & 166   & 140   & x     & x       & x \\ 
		\cmidrule(lr){2-18}

		&                              \multirow{2}{*}{1200s} & $f$ & 8.52 & 12.25 & 13.31 & 14.67   & x       & 11.81 & 17.02 & 18.25 &  20.64  &  x      & 17.75 & 25.24 & 26.49 & 27.97   & x \\ 
		&        &  {time (s)}                             & 39   & 41    & 129   & 217     & x       & 52    & 61    & 190   &  304    &  x      & 161   & 139   & 384   & 579     & x \\ 
		\cmidrule(lr){2-18}                              
		& \multirow{2}{*}{2400s}& $f$                               & 8.53 & 12.25 & 13.30 & 14.59   & 13.59   & 11.80 & 17.02 & 18.25 &  20.60  &  x      & 17.75 & 25.24 & 26.46 & 29.10   & x \\
		&        &  {time (s)}                             & 33   & 39    & 127   & 313     & 1444    & 51    & 62    & 188   &  327    &  x      & 158   & 127   & 381   & 543     & x \\
		\cmidrule(lr){2-18}

		& \multirow{2}{*}{3600s}& $f$                               & 8.52 & 12.25 & 13.32 & 14.61   & x       & 11.81 & 17.01 & 18.25 &  20.59  & 18.36   & 17.75 & 25.24 & 26.46 & 29.11   & x \\ 
		&        &  {time (s)}                             & 25   & 36    & 121   & 297     & x       & 53    & 61    & 193   &  470    & 2241    & 151   & 123   & 319   & 689     & x \\

		\midrule

		\multirow{2}{*}{$Shims$} & \multirow{2}{*}{240s}  & $f$ & 8.54 & 12.26 & 13.23 & 14.53   & 13.42   & 11.83 & 17.00 & 18.21 &  20.45  & 18.00   & 17.78 & 25.25 & 26.38 & 28.98   & 26.08 \\ % 
		&         & {time (s)}                             & 1    & 2     & 6     & 17      & 109     & 1     & 2     & 8     &  26     & 160     & 2     & 3     & 13    & 37      & 229   \\

		\bottomrule
	\end{tabular}
	\normalsize
\end{table}

The actual RAM consumption of $Gurobi$\/ was over 8.5 GB, and all of $Shims$\/'s executions consumed at most 1.5 GB of RAM.

\section{Conclusions}
\label{conclusions}

In this work, we modelled and solved a real air transport problem named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem}\/ (ACLP+RPDP). For the first time in the literature, a {\it NP-hard}\/ problem that involves {\it simultaneously}\/ pallet assembly, load balancing, and route planning is addressed, where the cost-effectiveness of transport is maximized. Currently, there is no commercial software available for this problem.

We adopted some simplifications that are not critical, but that allowed for an unprecedented solution to this problem considering more than two nodes. In practical cases, the number $K$\/ of nodes, excluding the base, is small ($K \leq 6$), each of which has hundreds of items to be shipped. Considering a real aircraft, we have developed some node-by-node solutions. This allows us to test two strategies: find a solution using the shortest tours, or enumerate all $K!$\/ tours and choose the best. The complete process can be executed quickly on a handheld computer, offering good results and reducing stress for the transport planners.

As validation, we carried out tests in several real air transport scenarios. In these cases, the solution process can establish, in less than four minutes, a flight tour for a single aircraft with a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximising the total score, and minimising fuel consumption along the planned route, which is beneficial to reduce carbon emissions. This output is an essential part of airlift: it guarantees flight safety, makes ground operations more efficient, and makes sure that each item gets to its right destination.

Our main contributions were the mathematical modelling of ACLP+RPDP, involving four well-known and interconnected {\it NP-hard}\/ sub-problems, a complete process to solve it, and a new heuristic named {\it Shims}\/ that offers fast node solutions with good quality. The method of this work is not exclusive to aircraft and airports: it can be adapted to ships and ports, vehicles and warehouses, or wagons and railways, provided that their practical cases are similar to those considered here. In these situations, it would be necessary to make some changes in the model: for example, modify the load balancing constraints and consider the available space in vehicles or wagons instead of pallets.

Because we applied this heuristic process to all $K!$\/ tours, perhaps it seems that our methodology is not generalizable. This is not true; we did this because {\it Shims}\/ is very fast and $K$\/ is small in practical cases of air transport. In other contexts where $K$\/ may have larger values, this heuristic process remains valid. It is enough to apply it only to some tours that offer a better expectation of a good result, for example, on the TSP's shortest path or other similar ones.

As this is ongoing research, we thought about some possible future improvements: consider more than one aircraft, implement parallel algorithms in some steps of the solution to improve computational efficiency, and model 3-D items.

\section*{CRediT authorship contribution statement}

\textbf{Antonio Celio Pereira de Mesquita:} Conceptualization, Methodology, Software, Writing - original draft preparation, Investigation, Validation. \textbf{Carlos Alberto Alonso Sanches:} Conceptualization, Methodology, Resources, Supervision, Writing - reviewing \& editing.

\section*{Declaration of competing interest}

The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

\section*{Data availability}

The dataset and code used in this work are available in {\tt github.com/celiomesquita/ACLP\_RPDP\_P}.

\section*{Acknowledgments}

This research was partially supported by the S\~{a}o Paulo Research Foundation (FAPESP), grant 2022/05803-3.

\section*{References}

\bibliographystyle{apalike}

\bibliography{bib}

\end{document}

### 2024-01-25 common.py

@@ -43,9 +43,10 @@ class Config(object):
             
 class Node(object):
     def __init__(self, id, tau):
-        self.ID      = id
-        # self.next    = 0
-        self.tau     = tau # node sum of torques
+        self.ID   = id
+        self.tLim = 1 # 1s
+        self.Vol  = 0.0
+        self.tau  = tau # node sum of torques
         self.ICAO = CITIES[0]
         if id < len(CITIES):
             self.ICAO = CITIES[id]
@@ -227,7 +228,7 @@ def loadDistances(fname):
     fname = os.path.abspath(fname)
 
     output_string = ""
-    str_list = fname.split("iRace/")
+    str_list = fname.split("tunning/")
     for element in str_list:
         output_string += element
 
@@ -386,7 +387,7 @@ def loadNodeItems(scenario, instance, node, unatended, surplus): # unatended, fu
     fname = os.path.abspath(fname)
 
     output_string = ""
-    str_list = fname.split("iRace/")
+    str_list = fname.split("tunning/")
     for element in str_list:
         output_string += element
 
@@ -405,7 +406,8 @@ def loadNodeItems(scenario, instance, node, unatended, surplus): # unatended, fu
             v   = float(cols[2])
             frm =   int(cols[3])
             to  =   int(cols[4])
-            if frm == node.ID and to in unatended:          
+            if frm == node.ID and to in unatended:
+                node.Vol += v       
                 items.append( Item(id, -1, w, s, v, frm, to) ) # P:-1 item, -2: consolidated
                 id += 1
 
@@ -428,7 +430,7 @@ def loadNodeItems(scenario, instance, node, unatended, surplus): # unatended, fu
     # avgAttr /= len(items)
     # print(f"avgAttr = {avgAttr:.3f}")
 
-    return items
+    return items, node
 
 
 def setPalletsDestinations(items, pallets, nodes, k, L_k):

### 2024-01-25 main.py

@@ -14,14 +14,14 @@ import tabu
 import grasp
 import noise2
 
-from plots import TTT
+# from plots import TTT
 
 # from py3Djanet import Packer, Item, Bin
 
 # from py3Druiz import Packer, Item, Bin
 
 
-def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplusFolder, tipo, numOptDict, rampDistCG, afterDict, beforeDict, eta1_vol, eta2_vol):
+def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, folder, tipo, numOptDict, rampDistCG, afterDict, beforeDict, eta1_vol, eta2_vol):
     """
     Solves one tour
     """
@@ -42,7 +42,25 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplusF
 
     k0 = len(tour.nodes)-1 # the base index on return
 
+    # a first tour iteration to calculate the total volume to be tested for inclusion the tour nodes.
+    tourVol = 0.0
+    tourTime = 0
     for k, node in enumerate(tour.nodes):  # solve each node sequentialy
+
+        # L_k destination nodes set
+        unattended = [n.ID for n in tour.nodes[k+1:]]
+
+        # node max volume is updated
+        _, node = common.loadNodeItems(scenario, inst, node, unattended, folder)
+
+        tourVol += node.Vol
+        tourTime += secBreak
+
+    # a second tour iteration solving node-by-node
+    for k, node in enumerate(tour.nodes):  # solve each node sequentialy
+
+        # the solving node time is proportional to its relative sum of candidate items volumes
+        node.tLim = (node.Vol/tourVol) * tourTime
 
         next = tour.nodes[k0]
         if k < k0:
@@ -64,13 +82,14 @@ def solveTour(scenario, inst, pi, tour, method, pallets, cfg, secBreak, surplusF
         unattended = [n.ID for n in tour.nodes[k+1:]]
 
         # load items parameters from this node and problem instance, that go to unnatended
-        items = common.loadNodeItems(scenario, inst, node, unattended, surplusFolder)
+        items, _ = common.loadNodeItems(scenario, inst, node, unattended, folder)
+
 
         if k > 0 and k < k0: # not in the base
 
             # load consolidated generated in the previous node
             # prevNode = tour.nodes[k-1]
-            # cons = common.loadNodeCons(surplusFolder, scenario, inst, pi, prevNode, numItems )
[…]

### 2024-01-26 latex/eswa.tex
Experiments. One more table to go.

@@
 
-%		\State $limit \gets 0.92$  \label{shims:limit}
-
 		\For{$i \gets 1$ to $m$}  \label{shims:eta1_a}
@@
 
+	\item In the {\it composition phase}\/ (lines \ref{shims:beginsecond}-\ref{shims:endsecond}), a set of shims named $Set$\/ is created for each pallet $i$, where each shim is formed by a set of edges in the range $[\eta_1,\eta_2]$, whose total volume is limited by $slack_i$.
+In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it First-Fit Decreasing}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \citep{JohnsonGarey1985}.
+Basically, shims are created by accumulating the following edges, taking $slack_i$\/ as a limit.
+
+	\item In the {\it selection phase}\/ (lines \ref{shims:beginthird}-\ref{shims:endthird}), the best shim in $Set$\/ is chosen.
+Initially, two shims are found: $sh_w$\/ with larger weight and $sh_v$\/ with larger volume.
+Between the two, the one with the highest score will be chosen, and its edges will be inserted into $E_{N_{\pi_k}}$.
+
+\end{itemize}
+
 {\color{blue}
@@
 }
-
-	\item In the {\it composition phase}\/ (lines \ref{shims:beginsecond}-\ref{shims:endsecond}), a set of shims named $Set$\/ is created for each pallet $i$, where each shim is formed by a set of edges in the range $[\eta_1,\eta_2]$, whose total volume is limited by $slack_i$.
-In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it First-Fit Decreasing}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \citep{JohnsonGarey1985}.
-Basically, shims are created by accumulating the following edges, taking $slack_i$\/ as a limit.
-
-	\item In the {\it selection phase}\/ (lines \ref{shims:beginthird}-\ref{shims:endthird}), the best shim in $Set$\/ is chosen.
-Initially, two shims are found: $sh_w$\/ with larger weight and $sh_v$\/ with larger volume.
-Between the two, the one with the highest score will be chosen, and its edges will be inserted into $E_{N_{\pi_k}}$.
-
-\end{itemize}
 
@@
 
+{\color{blue}
+	After the items were generated, some experiments with the \textit{iRace} tool were executed to determine the values for parameters $threshold_1$ and $threshold_2$ used in \textit{Shims}, whose results are presented in Table \ref{tab:irace}.
+
+	\begin{table}[H]
+		\centering
+		\caption{\textit{iRace} testing results}  \label{tab:irace}
+		\footnotesize
+		\begin{tabular}{cccc}
+			\toprule
+			\textbf{surplus} & \boldm{$threshold_1$} & \boldm{$threshold_2$} & \textbf{time (min)}\\
+			\midrule
+			1.2& 0.8621 & 1.0539 & 47 \\
+			1.5& 0.9199 & 1.1399 & 59 \\
+			2.0& 0.9617 & 1.5706 & 63 \\						
+			\bottomrule
+		\end{tabular}
+		\normalsize
+	\end{table}
+
+	For these tests, we considered the odd instances (1, 3, 5, 7) as the \textbf{training} set and the even instances (2, 4, 6) as the \textbf{testing} set for the \textit{iRace} runs.
+We supplied \textit{iRace} for $threshold_1$ determination, the range [0.8, 1.0], and for $threshold_2$, the range [1.0, 2.0].
+The number of iterations in each experiment was limited to 3000 executions for \textit{iRace} to have plenty of data for its statistical tests.
+More detail about these configurations may be found on the \textit{iRace} user guide in {\tt cran.r-project.org/web/packages/irace/irace.pdf}.
+
+}
+
 \subsection{Obtained solutions: quality and runtimes}
@@
 		\toprule
-		{\specialcell{{\bf Tested}\\{\bf tours}}}&$method$          & $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
-		\toprule
-		\multirow{4}{*}{Shortest} &\multirow{2}{*}{$Gurobi$} & $f$ & 8.52    & 11.70   & 13.07   & 13.54 & 12.44     &  1.00\\
-		&                                                               &               {time (s)}  & 17      & 17      & 16      & 17    & 26        &  1.0 \\
+		{\specialcell{{\bf Tested}\\{\bf tours}}}&$method$&$scenario$&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&\specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		\toprule
+		\multirow{4}{*}{Shortest} &\multirow{2}{*}{$Gurobi$}  & $f$ & 8.61    & 11.56   & 13.67   & 13.87 & 12.48       &  0.9985 \\ % best 6028
+		&                              &               {time (s)}   & 26      & 22      & 28      & 28    & 25          &  1.0    \\ % longer 129
 		\cmidrule(lr){2-9}
-		&                                                               \multirow{2}{*}{$Shims$}    & $f$ & 8.54    & 11.63   & 12.97   & 13.43 & 12.36     &  0.99 \\
-		&                                                               &         {time (s)}        & 1       & 1       & 2       & 2     & 2         &  11.6 \\
-
-		\midrule
-
-		\multirow{4}{*}{All $K!$}&                                  \multirow{2}{*}{$Gurobi$} & $f$ & 8.52    & 12.25   & 13.32   & 14.61  & \textbf{x}        & 1.00 \\
-		&                                                               &               {time (s)}  & 25      & 36      & 121     & 297    & \textbf{x}        &  1.0 \\
+		&                           \multirow{2}{*}{$Shims$}  & $f$ & 8.61    & 11.56   & 13.46   & 13.93 & 12.51       &  0.9965 \\
+		&                             &                {time (s)}   & 1       & 1       & 1       & 1     & 1           &  25.8   \\
+
+		\midrule
+
+		\multirow{4}{*}{All $K!$}&  \multirow{2}{*}{$Gurobi$} & $f$ & 8.60    & 12.20   & 13.66   & 15.00  & {\bf x}    & 0.9998 \\ % best 4947
+		&                              &               {time (s)}   & 30      & 35      & 123     & 314    & {\bf x}    & 1.0 \\
 		\cmidrule(lr){2-9}	
-		&                                                               \multirow{2}{*}{$Shims$}    & $f$ & 8.54    & 12.26   & 13.23   & 14.53  & 13.42    & 0.99  \\
-		&                                                               &         {time (s)}        & 1       & 2       & 6       & 17     & 106      & 18.4   \\
+		&                           \multirow{2}{*}{$Shims$}  & $f$ & 8.61    & 12.20   & 13.46   & 14.99  & 13.35      & 0.9958  \\
+		&                              &               {time (s)}   & 1       & 1       & 4       & 10     & 51         & 7.49   \\
 		\bottomrule
@@
 		\toprule
-		\multirow{4}{*}{ Shortest } & \multirow{2}{*}{$Gurobi$} & $f$ & 11.80 & 16.74 & 18.04 & 18.86 & 16.92 & 1.00 \\
-		&&               {time (s)}                                                                  & 38    & 30    & 27    & 29    & 81    & 1.00  \\
+		\multirow{4}{*}{ Shortest } & \multirow{2}{*}{$Gurobi$} & $f$ & 11.76 & 16.29 & 18.02 & 19.42 & 16.35 & 0.9994 \\ % best 8189
+		&&               {time (s)}                                   & 51    & 39    & 40    & 36    & 38    & 1.0 \\ 
 		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{$Shims$}                                                                    & $f$ & 11.83 & 16.74 & 18.01 & 18.89 & 16.91 & 1.00  \\
-		&&         {time (s)}                                                                        & 1     & 2     & 2     & 3     & 3     & 20.5  \\
-
-		\midrule
-
-		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$}                                  & $f$ & 11.81 & 17.01 & 18.25 & 20.59 & 18.36   & 1.00 \\ 
-		&&               {time (s)}                                                                  & 53    & 61    & 193   & 470   & 2241    & 1.0  \\
+		&\multirow{2}{*}{$Shims$}                               & $f$ & 11.78 & 16.29 & 17.96 & 19.43 & 16.37 & 0.9993  \\
+		&&         {time (s)}                                         & 1     & 1     & 2     & 2     & 2     & 25.5  \\
+
+		\midrule
+
+		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$}   & $f$ & 11.76 & 16.37 & 18.01 & 20.95 & 17.60 & 0.9999 \\  % best 8469
+		&&               {time (s)}                                   & 64    & 58    & 195   & 472   & 2258  & 1.0  \\
 		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{$Shims$}                                                                    & $f$ & 11.83 & 17.00 & 18.21 & 20.45 & 18.00   & 0.99  \\ 
-		&&         {time (s)}                                                                        & 1     & 2     & 9     & 26    & 157     & 15.5 \\
+		&\multirow{2}{*}{$Shims$}                               & $f$ & 11.78 & 16.29 & 17.99 & 20.93 & 17.50 & 0.9976  \\ 
+		&&         {time (s)}                                         & 1     & 2     & 5     & 15    & 100   & 23.8 \\
 		\bottomrule
@@
 		\toprule
-		{\specialcell{{\bf Tested}\\{\bf tours}}}& $method$          & $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
-		\toprule
-		\multirow{4}{*}{Shortest} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.75 & 24.39 & 26.46 & 27.09 & 24.00 & 1.00 \\ 
-		&&                                                                               {time (s)}  & 162   & 95    & 74    & 68    & 67    & 1.0  \\
+		{\specialcell{{\bf Tested}\\{\bf tours}}}& $method$& $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		\toprule
+		\multirow{4}{*}{Shortest} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.89 & 24.96 & 26.15 & 27.52 & 23.85 & 0.9996 \\ % best 12042
+		&&                                              {time (s)}  & 183   & 87    & 73    & 79    & 71    & 1.0  \\
 		\cmidrule(lr){2-9}	
-		&                                                                   \multirow{2}{*}{$Shims$} & $f$ & 17.78 & 24.40 & 26.36 & 27.01 & 23.94 & 0.99  \\
-		&&                                                                                {time (s)} & 2     & 3     & 3     & 4     & 5     & 27.4  \\
-
-		\midrule
-
-		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.75 & 25.24 & 26.46 & 29.11 &  \textbf{x}   & 1.00 \\ 
-		&&                                                                                {time (s)} & 151   & 123   & 319   & 689   &  \textbf{x}   & 1.0  \\
+		&                            \multirow{2}{*}{$Shims$} & $f$ & 17.94 & 24.96 & 26.03 & 27.43 & 23.73 & 0.9972  \\
+		&&                                              {time (s)}  & 1     & 2     & 2     & 3     & 4     & 41.1  \\
+
+		\midrule
+
+		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.90 & 25.44 & 26.15 & 29.13 &{\bf x}& 0.9994 \\ % best 9867
+		&&                                               {time (s)} & 178   & 143   & 378   & 862   &{\bf x}& 1.0  \\
 		\cmidrule(lr){2-9}	
-		&                                                                \multirow{2}{*}{$Shims$}    & $f$ & 17.78 & 25.25 & 26.38 & 28.98 & 26.08 & 1.00  \\
-		&&                                                                                {time (s)} & 2     & 3     & 13    & 37    & 229   & 23.3  \\
+		&                         \multirow{2}{*}{$Shims$}    & $f$ & 17.94 & 25.45 & 26.14 & 28.84 & 26.22  & 0.9970  \\
+		&&                                               {time (s)} & 1     & 3     & 10    & 31    & 196    & 34.7  \\
 		\bottomrule
@@
 This occurs because the runtime limit per node is smaller and there tend to be more packed contents on the aircraft, reducing the space for allocating items and making the solution difficult.
-	\item When $Gurobi$\/ finishes, it finds the best solution, but the one obtained by $Shims$\/ reaches $99\%$\/ of that value.
-	\item $Shims$\/ always finds a solution, being 11 to 27 times faster.
+	\item When $Gurobi$\/ finishes, it finds the best solution, but the one obtained by $Shims$\/ reaches {\color{blue}$99.9\%$}\/ of that value.
+	\item $Shims$\/ always finds a solution, being {\color{blue}8 to 41} times faster.
 	\item All runtimes are much lower than the limit because the solution on many nodes can be fast.

### 2024-01-26 latex/eswa.tex
GA TSP solver included

@@
 In this section, we describe the context of the problem addressed in this work as well as the assumptions considered.
+
+{\color{blue}
+
+As for the size of the problem dealt with in this work, addressing the scale of the Travelling Salesperson Problem (TSP) in practical, real-world air transport contexts indeed reveals a fascinating and complex challenge.
+In the realm of academic research, TSP problems often explore theoretical models that can scale to hundreds or even larger numbers of cities.
+
+In practical terms, especially within the air transport sector, the size and complexity of a TSP can vary significantly depending on several factors:
+
+\textbf{Fleet and Route Planning}: For airline operations, the problem isn't just about finding the shortest path through a set of points but also involves scheduling, aircraft assignment, crew scheduling, and maintenance planning.
+This multi-dimensional complexity transforms a standard TSP into a more intricate problem in our study, the Pickup and Delivery Problem (PDP).
+
+\textbf{Operational Constraints}: Factors like weather, air traffic control restrictions, and the availability of take-off and landing slots make real-world TSP scenarios in air transport even more challenging.
+These constraints add layers of complexity that significantly depart from the theoretical TSP model.
+
+\textbf{Dynamic Nature}: Unlike the static nature of classical TSP, real-world problems in air transport are highly dynamic.
+Routes need to be optimized continuously as conditions change, requiring algorithms that can adapt in real-time or near-real-time.
+
+In practical air transport scenarios, the size of a TSP can therefore be less about the sheer number of nodes and more about the complexity and dynamic nature of the problem, which in our case (a single aircraft in a single day) is rarely expected to be more than seven nodes, based on the experience of the first author when working on an operations control centre.
+}
 
@@
 More detail about these configurations may be found on the \textit{iRace} user guide in {\tt cran.r-project.org/web/packages/irace/irace.pdf}.
-
 }
@@
 For more details, see {\tt www.support.gurobi.com}.
+
+\begin{table}[H]
+{\color{blue}
+	\centering
+	\caption{Overall results}  \label{tab:mr}
+	\footnotesize
+	\begin{tabular}{cccc}	
+	\toprule
+	{\bf Method} & {\bf Best scenarios} & {\bf Worst scenarios} & {\bf Worst run times (min)}  \\
+	\midrule
+	NMO           & 4             & 5    & {\color{red}60}  \\
+	ACO           & 2, 3          & 4, 5 & {\color{red}25}, {\color{red}61}  \\
+	GRASP         & 1             & 4, 5 & {\color{red}28}, {\color{red}55}  \\
+	TS            & -             & 5    & {\color{red}44}  \\
+	\midrule
+	{\it Gurobi}  & 1, 2, 3, 4    & 5    & {\color{red}did not solve}  \\
+	{\it Shims}   & 1, 2, 3, 4, 5 & -    & {\bf 3.26} \\
+	\bottomrule
+	\end{tabular}
+\normalsize
+}
+\end{table}
+
+\subsubsection{Shims results}
 
@@
 We can observe more cases where $Gurobi$\/ fails, even in smaller scenarios.
-When $Gurobi$\/ finishes, $Shims$\/ finds a solution of similar quality ($98\%$\/ or better).
+When $Gurobi$\/ finishes, $Shims$\/ finds a solution of similar quality ($99\%$\/ or better).
 In all cases, $Shims$\/ finds a good solution in less than 4 minutes.
@@
 
-		\multirow{10}{*}{$Gurobi$} & \multirow{2}{*}{240s} & $f$ & 8.53 & 12.24 & 13.34 &  x      &  x      & 11.81 & 17.02 & 18.31 &  x      &  x      & 17.75 & 25.24 & x     & x       & x \\ 
-		&               &  {time (s)}                              & 41   & 43    & 123   &  x      &  x      & 50    & 62    & 181   &  x      &  x      & 166   & 140   & x     & x       & x \\ 
+		\multirow{10}{*}{$Gurobi$} & \multirow{2}{*}{240s} & $f$ & 8.53 & 12.24 & 13.34 &  {\bf x}& {\bf x} & 11.81 & 17.02 & 18.31 & {\bf x} & {\bf x} & 17.75 & 25.24 & {\bf x} & {\bf x} & {\bf x} \\ 
+		&               &  {time (s)}                            & 41   & 43    & 123   &  {\bf x}& {\bf x} & 50    & 62    & 181   & {\bf x} & {\bf x} & 166   & 140   & {\bf x} & {\bf x} & {\bf x} \\ 
 		\cmidrule(lr){2-18}
 
-		&                              \multirow{2}{*}{1200s} & $f$ & 8.52 & 12.25 & 13.31 & 14.67   & x       & 11.81 & 17.02 & 18.25 &  20.64  &  x      & 17.75 & 25.24 & 26.49 & 27.97   & x \\ 
-		&        &  {time (s)}                             & 39   & 41    & 129   & 217     & x       & 52    & 61    & 190   &  304    &  x      & 161   & 139   & 384   & 579     & x \\ 
+		&                           \multirow{2}{*}{1200s} & $f$ & 8.52 & 12.25 & 13.31 & 14.67   & {\bf x} & 11.81 & 17.02 & 18.25 &  20.64  & {\bf x} & 17.75 & 25.24 & 26.49 & 27.97   & {\bf x} \\ 
+		&        &  {time (s)}                                   & 39   & 41    & 129   & 217     & {\bf x} & 52    & 61    & 190   &  304    & {\bf x} & 161   & 139   & 384   & 579     & {\bf x} \\ 
 		\cmidrule(lr){2-18}                              
-		& \multirow{2}{*}{2400s}& $f$                               & 8.53 & 12.25 & 13.30 & 14.59   & 13.59   & 11.80 & 17.02 & 18.25 &  20.60  &  x      & 17.75 & 25.24 & 26.46 & 29.10   & x \\
-		&        &  {time (s)}                             & 33   & 39    & 127   & 313     & 1444    & 51    & 62    & 188   &  327    &  x      & 158   & 127   & 381   & 543     & x \\
+		&                           \multirow{2}{*}{2400s} & $f$ & 8.60 & 12.21 & 13.67 & 15.00   & 13.41   & 11.77 & 16.38 & 18.02 & 20.94   & {\bf x} & 17.75 & 25.24 & 26.46 & 29.10   & {\bf x} \\
+		&        &  {time (s)}                                   & 26   & 38    & 134   & 310     & 1520    & 46    & 60    & 206   & 474     & {\bf x} & 158   & 127   & 381   & 543     & {\bf x} \\
 		\cmidrule(lr){2-18}
 
-		& \multirow{2}{*}{3600s}& $f$                               & 8.52 & 12.25 & 13.32 & 14.61   & x       & 11.81 & 17.01 & 18.25 &  20.59  & 18.36   & 17.75 & 25.24 & 26.46 & 29.11   & x \\ 
-		&        &  {time (s)}                             & 25   & 36    & 121   & 297     & x       & 53    & 61    & 193   &  470    & 2241    & 151   & 123   & 319   & 689     & x \\
-
-		\midrule
-
-		\multirow{2}{*}{$Shims$} & \multirow{2}{*}{240s}  & $f$ & 8.54 & 12.26 & 13.23 & 14.53   & 13.42   & 11.83 & 17.00 & 18.21 &  20.45  & 18.00   & 17.78 & 25.25 & 26.38 & 28.98   & 26.08 \\ % 
-		&         & {time (s)}                             & 1    & 2     & 6     & 17      & 109     & 1     & 2     & 8     &  26     & 160     & 2     & 3     & 13    & 37      & 229   \\
+		&                           \multirow{2}{*}{3600s} & $f$ & 8.60 & 12.20 & 13.66 & 15.00   & {\bf x} & 11.76 & 16.37 & 18.01 &  20.95  & 17.60   & 17.90 & 25.44 & 26.15 & 29.13   & {\bf x} \\ 
+		&        &  {time (s)}                                   & 30   & 35    & 123   & 314     & {\bf x} & 64    & 58    & 195   &  472    & 2258    & 178   & 143   & 378   & 862     & {\bf x} \\
+
+		\midrule
+
+		\multirow{2}{*}{$Shims$} &  \multirow{2}{*}{240s}  & $f$ & 8.61 & 12.20 & 13.46 & 14.99   & 13.35   & 11.78 & 16.29 & 17.99 &  20.93  & 17.50   & 17.94 & 25.45 & 26.14 & 28.84   & 26.22 \\ % 
+		&         & {time (s)}                                   & 1    & 2     & 4     & 10      & 51      & 1     & 2     & 5     &  15     & 100     & 1     & 3     & 10    & 31      & 196   \\
 
@@
 The actual RAM consumption of $Gurobi$\/ was over 8.5 GB, and all of $Shims$\/'s executions consumed at most 1.5 GB of RAM.
+
+{\color{blue}
+\subsection{{\it Shims} validation with more nodes}
+
+To validate the effectiveness of the {\it Shims} heuristic, we solve a 15-node TSP containing the 15 main Brazilian airports.
+For this task, we implemented a TSP solution method based on the Genetic Algorithm (GA) that returns the 20\% best tours as input to the ACLP-RPDP solved by {\it Shims}.
+
+}
 

### 2024-01-26 latex/responses.tex
GA TSP solver included

\documentclass{article}
\usepackage[latin1]{inputenc}
\usepackage{textcomp}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{color, colortbl}

\begin{document}
\thispagestyle{empty}

\begin{center}

{\Large\bf Air cargo load and route planning\\ in pickup and delivery operations}\\
\vspace{0.5cm}
{\em Antonio Celio Pereira de Mesquita \\ Carlos Alberto Alonso Sanches}
\end{center}

The authors express their gratitude to the three anonymous reviewers for their comments and criticism, which helped to improve the content and presentation of this paper. We carefully considered all comments from the reviewers, and proceeded to the revision of the text.

We do understand that now our paper achieves the high standard quality of this journal. As will be presented below, we followed the requests from reviewers.

\section{\#1-a: The entire editorial decision letter}

\footnotesize

Ms. Ref. No.: ESWA-D-23-16263\\
Title: Air cargo load and route planning in pickup and delivery operations \\
Expert Systems With Applications \\

Dear Carlos,\\

As Editor, I'm writing this editorial decision letter on your paper submission.  If you are interested in submitting a revised version, please read through this entire editorial decision letter carefully and take all actions seriously in order to avoid any delay in the review process of your revised manuscript submission. You need to upload a 'Detailed Response to Reviewers' in the EM system with the following sections while submitting the revised manuscript. Please note that the Required Sections (Section \#1-a, Section \#1-b, Section \#2-a, Section \#2-b, Section \#3-a, and Section \#3-b) defined clearly in this editorial decision letter with specific Compliance Requirements (part of this editorial decision letter) must be clearly labeled and included in the 'Detailed Response to Reviewers'.  The Required Sections must be clearly placed before the revised manuscript. NOTE: Section \#1-(a) MUST contain the COMPLETE text covered in this email letter from Editor-in-Chief, rather than just only the first paragraph of this email letter from Editor-in-Chief. Please note that we will NOT admit any revised manuscript submission for further review if any of the Required Sections is incomplete or any non-compliance of the ESWA authors' guidelines in the PDF file of the revised manuscript that you approve in EM system. So you need to take both (1) Required Sections and (2) Compliance Requirements seriously to avoid any delay in the review process of your revised manuscript. The font size of the Required Sections should be consistent and readable. All required sections should not be embedded. The Required Sections must be placed clearly BEFORE the revised manuscript and in the order listed as followings:

REQUIRED SECTIONS: Please confirm that your submission includes ALL of  the required sections.

\begin{itemize}

\item Section \#1-a: Including the entire editorial decision letter (i.e., complete text covered in this email letter) from Editor

\item Section \#1-b: Including your responses to Editor

\item Section \#2-a: Including the entire comments made by the Associate Editor

\item Section \#2-b: Including your Point-to-Point responses to the Associate Editor

\item Section \#3-a: Including the entire comments made by the Reviewer

\item Section \#3-b. Including your Point-to-Point responses to the Reviewer.

\end{itemize}

COMPLIANCE REQUIREMENTS:  Please confirm that your submission completely complies with the compliance requirements.

Please note that prior to admitting the revised submission to the next rigorous review process, all paper submissions must completely comply with ESWA Guide for Authors (see details at \url{https://www.elsevier.com/journals/expert-systems-with-applications/0957-4174/guide-for-authors}).  These include at least the following Compliance Requirements:

A) Authorship policies - Please also note that ESWA takes authorship very seriously and all paper submissions MUST completely comply with all of the following three policies on authorship (clearly stated in the questionnaire responses in EM system) prior to a rigorous peer review process:

A)-1: The corresponding author needs to enter the full names, full affiliation with country and email address of every contributing author in EM online system.  It is also mandatory that every contributing coauthor must be listed in EM at submission.

A)-2: It is mandatory that the full names, full affiliation with country and email address of every contributing author must be included in title (authorship) page of the manuscript. The first page of the manuscript should contain the title of the paper, and the full name, full affiliation with country and email address of every contributing author. The second page of the manuscript should begin with the paper abstract. Note that cover letter is not title (authorship) page.

A)-3: The authorship information in EM system must be consistent with the authorship information on the title (authorship) page of the manuscript.

B)  Guidelines of reference style and reference list - Citations in the text should follow the referencing style used by the American Psychological Association (APA).

B)-1: Reference Style: Citations in the text should follow the referencing style used by the American Psychological Association. You are referred to the Publication Manual of the American Psychological Association, Sixth Edition, ISBN 978-1-4338-0561-5.  APA's in-text citations require the author's last name and the year of publication. You should cite publications in the text, for example, (Smith, 2020).  However, you should not use [Smith, 2020]. Note: There should be no [1], [2], [3], etc in your manuscript.

B)-2: Reference List:

References should be arranged first alphabetically by the surname of the first author followed by initials of the author's given name, and then further sorted chronologically if necessary. More than one reference from the same author(s) in the same year must be identified by the letters 'a', 'b', 'c', etc., placed after the year of publication. For example, Van der Geer, J., Hanraads, J. A. J., \& Lupton, R. A. (2010). The art of writing a scientific article. Journal of Scientific Communications, 163, 51-59. \url{https://doi.org/10.1016/j.Sc.2010.00372}. Note: There should be no [1], [2], [3], etc in your references list.

C) Highlights guidelines - There should be a maximum of 85 characters, including spaces, per Highlight in the Highlights section. Please kindly read this guideline carefully - the guideline does NOT say there should be a maximum of 85 words per Highlight.  It says there should be a maximum of 85 characters per Highlight.  As examples, the word "impact" consists of 6 characters; the word "significance" consists of 12 characters. Only include 3 to 5 Highlights. Minimum number is 3, and maximum number is 5.

D) Please pay attention to the new journal policy asking for institutional emails in all submissions. Every author must have an institutional email in both, the submission platform and manuscript.

NOTE: Your paper submission will be returned to authors and will NOT be admitted to further review if the revised paper fails to completely comply with the ESWA Authorship policies, ESWA guidelines of reference style and reference list, Highlights guidelines. You need to take the Compliance Requirements seriously to avoid any delay in the review process of your revised manuscript.

To submit a Complete and Compliance revision, please go to \url{https://www.editorialmanager.com/eswa/} and login as an Author.a

Your username is: \url{alonso@ita.br}

If you need to retrieve password details, please go to:
click here to reset your password

On your Main Menu page is a folder entitled "Submissions Needing Revision". You will find your submission record there.

When you are ready to submit your revised paper, please upload the source files of your revised paper (if in Word, please upload the .doc or docx file; if in .TeX, please upload the .TeX, style) as these would be required in Production if your manuscript gets accepted

Please note that we allow 21 days for the first author revision and 14 days for any additional author revisions that are required.

Note: While submitting the revised manuscript, please double check the author names provided in the submission so that authorship related changes are made in the revision stage. If your manuscript is accepted, any authorship change will involve approval from co-authors and respective editor handling the submission and this may cause a significant delay in publishing your manuscript.Please also check which journals do you select for publication. There is no change back under the condition that the manuscript is accepted

Research Elements (optional)
This journal encourages you to share research objects - including your raw data, methods, protocols, software, hardware and more - which support your original research article in a Research Elements journal. Research Elements are open access, multidisciplinary, peer-reviewed journals which make the objects associated with your research more discoverable, trustworthy and promote replicability and reproducibility. As open access journals, there may be an Article Publishing Charge if your paper is accepted for publication. Find out more about the Research Elements journals at
\url{https://www.elsevier.com/authors/tools-and-resources/research-elements-journals?dgcid=ec_em_research_elements_email}.\\

Yours sincerely,\\

Joaquín Torres-Sospedra, Ph.D.

Editor

Expert Systems With Applications

\normalsize

\section{\#1-b: Our responses to Editor}

Agradecemos e seguimos todas as orientações do Editor:

\begin{itemize}
\item Incluímos neste texto todas as {\bf Required Sections};

\item A submissão do nosso artigo revisado atende todas as {\bf Compliance Requirements}.

\end{itemize}

\section{\#2-a: The entire comments made by the Associate Editor}

De acordo com a mensagem recebida, parece-nos que os comentários do Associate Editor são os seguintes:

\vspace{0.5cm}

Reviewers' comments:

\begin{itemize}

\item [1.] More analysis shall be given to discuss the experimental results.
\item [2.] More newly-published related references shall be cited and dicussed in Introduction.
\item [3.] More experiments hope to be added to support the conclusion.
\end{itemize}

{\color{blue}
New experiments were conducted due to two changes suggested by the reviewers:

\begin{itemize}
\item [1.] Node solution times depended on the total volume of candidate items to be loaded in each node in relation to the total items volume to be loaded in the tour.
\item [2.] A new process of determining the {\it Shims} volume thresholds for $eta_1$ and $eta_2$.
\end{itemize}
}

\section{\#2-b: Our Point-to-Point responses to the Associate Editor}

\paragraph{Revision:}
More analysis shall be given to discuss the experimental results. 
\begin{center}
	\begin{minipage}{10cm}
		\textit{Como pode ser observado logo abaixo, aumentamos substancialmente os dados experimentais apresentados. Consequentemente, houve também um sensível aumento na discussão sobre eles.}
	\end{minipage}
\end{center}

\paragraph{Revision:}
More newly-published related references shall be cited and dicussed in Introduction.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Incluímos e comentamos novas referências (procurar trabalhos recentes sobre TSP e PDP em aeronaves).}
	\end{minipage}
\end{center}

\paragraph{Revision:}
More experiments hope to be added to support the conclusion.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Incluímos novos resultados experimentais (novas figuras, testes, etc.). Concretamente: desempenhos obtidos com outras metaheurísticas; definição empírica de $\eta_1$\/ and $\eta_2$\/ no algoritmo {\it Shims}; limite de tempo variável em cada nó, de acordo com o volume de carga previsto em cada nó; resolução de casos com um maior número de nós}.

{\color{blue}
We included Table 10 with the results of the {\it iRace} tool, a software package that helps to automatically fine-tune algorithm parameters. We fine-tuned parameters $threshold_1$\/ and $threshold_2$\/, the relative volumes of each pallet used to determine the indexes  $\eta_1$\/ and $\eta_2$ for the {\it Shims} heuristic.
}
	\end{minipage}
\end{center}

\section{\#3-a: The entire comments made by the Reviewers}

Indicamos a seguir, em cada subseção, os comentários feitos pelos três revisores.

\subsection{Reviewer \#1}

\begin{itemize}

\item [1] In section 2, Literature review in the first and second paragraphs is too simple. Some analysis needs to be given to introduce the research status.

\item [2] In the manuscript, any figures about algorithm simulation results cannot be found. Some figures and analysis are suggested to be presented to show the superiority of the used algorithm.

\item [3] Also, any figures about experiments are not given in the manuscript. "By using a portable computer, our strategy quickly found practical solutions to a wide range of real problems in much less than operationally acceptable time."
Could you please show the results with some figures?

\item [4] Too many old references are cited.

\end{itemize}

\subsection{Reviewer \#2}

This paper addresses the optimization of air cargo load planning and routing for pickup and delivery operations. It introduces a novel problem called the Air Cargo Load Planning with Routing Pickup and Delivery Problem (ACLP+RPDP), which is mathematically modeled to tackle the challenges of pallet assembly, load balancing, and route planning in air transport.

The problem raised in this paper is innovative and applicable, and the proposed algorithm demonstrates effectiveness in solving the problem. However, there are several issues that need to be addressed:

\begin{itemize}

\item [1.] The experimental section lacks effectiveness analysis, where the specific contributions of each strategy are not thoroughly verified and analyzed through separate experiments.

\item[2.] The comparative experiment should include both well-known and novel algorithms for a comprehensive evaluation.

\item [3.] The algorithm is missing a complexity analysis, which is essential to better understand its efficiency and potential limitations.

\item [4.] The inclusion of more figures to illustrate the flow, principle, and results of the algorithm would greatly enhance the clarity and visual representation.

\item [5.] The format of the article needs further adjustment, particularly in addressing the existing large blank spaces on pages 7 and 10.

\end{itemize}

\subsection{Reviewer \#3}

This paper models the ACLP+RPDP of a real case, and develops a problem-solver. The experiments have demonstrated the contributions. However, there are some issues that should be addressed:

\begin{itemize}

\item [1:] In section 1, the authors claim that they solve an air cargo problem involving simultaneously APP, WBP, PDP, and TSP. But, there is not clear description of the importance. In other words, the motivation should be enhanced.

\item [2:] The literature review is simple. Although some studies focus on the same sub-problem in ACLPP, they still show differences in designing modelling and problem-solver. Apparently, authors should supplement them.

\item [3:] In section 5, the running time at each node is fixed. Whether different nodes have different requirements for running time.

\item [4:] In section 5.2.2, $\eta_1$\/ and $\eta_2$\/ are set by authors. It is better to study them by experiments.

\item [5:] The experimental settings are not complete.

\item [6:] To the best of our knowledge, the genetic algorithm is a good method for solving ACLPP. The method is not compared in the experiments.

\item [7:] With the increase of nodes, what impact will be brought in solving the problem. The experiments and analysis should be supplemented.

\end{itemize}

\section{\#3-b: Our Point-to-Point responses to the Reviewers}

Em cada subseção, responderemos os comentários de cada um dos três revisores.

\subsection{Reviewer \#1}

\paragraph{Revision:}
In section 2, Literature review in the first and second paragraphs is too simple. Some analysis needs to be given to introduce the research status.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Aumentamos a revisão da literatura, descrevendo os trabalhos com maior detalhe. Acrescentamos também alguns trabalhos mais recentes. Citar quais são esses trabalhos...}
	\end{minipage}
\end{center}

\paragraph{Revision:}
In the manuscript, any figures about algorithm simulation results cannot be found. Some figures and analysis are suggested to be presented to show the superiority of the used algorithm.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Incluímos figuras comparativas, mostrando o melhor desempenho da heurística {\it Shims}. Citar quais são essas figuras...}
	\end{minipage}
\end{center}

\paragraph{Revision:}
Also, any figures about experiments are not given in the manuscript. "By using a portable computer, our strategy quickly found practical solutions to a wide range of real problems in much less than operationally acceptable time." Could you please show the results with some figures?
\begin{center}
	\begin{minipage}{10cm}
		\textit{Nas novas figuras incluídas no artigo, mostramos o desempenho dos algoritmos desenvolvidos. Citar quais são essas figuras...}
	\end{minipage}
\end{center}

\paragraph{Revision:}
Too many old references are cited.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Incluímos referências mais recentes (citá-las). De qualquer modo, preferimos manter as antigas, para assim termos um levantamento histórico da evolução da pesquisa neste campo.}
	\end{minipage}
\end{center}

\subsection{Reviewer \#2}

\paragraph{Revision:}
The experimental section lacks effectiveness analysis, where the specific contributions of each strategy are not thoroughly verified and analyzed through separate experiments.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Talvez não tenha ficado claro no texto original, mas três estratégias foram imprescindíveis para que pudéssemos obter alguma solução nas instâncias criadas: (1) considerar $K<m$\ para que cada pallet recebesse uma destinação predefinida; (2) em cada tour, calcular uma solução nó a nó; (3) em cada nó intermediário do tour considerado, realocar os packed contents minimizando o torque na aeronave. Sem a adoção destas três estratégias, não conseguimos realizar nenhum experimento, mesmo em instâncias menores. A única estratégia que permitiu diferentes experimentos e comparações foi a heurística utilizada na solução nó a nó. Procuramos deixar isto mais claro na nossa conclusão.}
	\end{minipage}
\end{center}

\paragraph{Revision:}
The comparative experiment should include both well-known and novel algorithms for a comprehensive evaluation.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Como estamos lidando com um novo problema, não dispomos de outras resoluções disponíveis na literatura. Na verdade, temos apenas meta-heurísticas bem conhecidas para a solução nó a nó, no qual o torque da aeronave é minimizado. Por este motivo, incluímos novas tabelas comparativas com os resultados obtidos nesses casos.}
	\end{minipage}
\end{center}

\paragraph{Revision:}
The algorithm is missing a complexity analysis, which is essential to better understand its efficiency and potential limitations.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Agradecemos este comentário. Incluímos a análise de complexidade do algoritmo apresentado no nosso artigo.}
	\end{minipage}
\end{center}

\paragraph{Revision:}
The inclusion of more figures to illustrate the flow, principle, and results of the algorithm would greatly enhance the clarity and visual representation.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Agradecemos esta sugestão. Acrescentamos novas figuras e tabelas que ilustram melhor nossos resultados. Citar quais são...}
	\end{minipage}
\end{center}

\paragraph{Revision:}
The format of the article needs further adjustment, particularly in addressing the existing large blank spaces on pages 7 and 10.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Estes espaços surgiram devido à formatação do \LaTeX, e deverão ser eliminados na versão final.}
	\end{minipage}
\end{center}

\subsection{Reviewer \#3}

\paragraph{Revision:}
In section 1, the authors claim that they solve an air cargo problem involving simultaneously APP, WBP, PDP, and TSP. But, there is not clear description of the importance. In other words, the motivation should be enhanced.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Acrescentamos novas motivações para a resolução do ACLP+RPDP, que envolve simultaneamente 4 sub-problemas {\it NP-hard}.}
	\end{minipage}
\end{center}

\paragraph{Revision:}
The literature review is simple. Although some studies focus on the same sub-problem in ACLPP, they still show differences in designing modelling and problem-solver. Apparently, authors should supplement them.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Agradecemos esta sugestão. Incluímos mais informações na revisão da literatura, indicando as modelagens e os métodos de resolução.}
	\end{minipage}
\end{center}

\paragraph{Revision:}
In section 5, the running time at each node is fixed. Whether different nodes have different requirements for running time.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Agradecemos esta sugestão. Entre os novos resultados apresentados, incluímos uma resolução na qual o tempo dedicado a cada nó é proporcional à carga disponível para embarque.}
	\end{minipage}
\end{center}

\paragraph{Revision:}
In section 5.2.2, $\eta_1$\/ and $\eta_2$\/ are set by authors. It is better to study them by experiments.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Estudamos a viabilidade de definir empiricamente os valores dessas constantes do algoritmo Shims. Explicar...}
	\end{minipage}
\end{center}

\paragraph{Revision:}
The experimental settings are not complete.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Não entendemos este comentário... De qualquer modo, aumentamos a seção de experimentos do nosso artigo, procurando definir todas as variáveis envolvidas. Caso reste algum ponto obscuro, pedimos ao revisor que o especifique, por favor.}
	\end{minipage}
\end{center}

\paragraph{Revision:}
To the best of our knowledge, the genetic algorithm is a good method for solving ACLPP. The method is not compared in the experiments.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Analisamos as resoluções do ACLPP através de algoritmos genéticos. Explicar...}
	\end{minipage}
\end{center}

\paragraph{Revision:}
With the increase of nodes, what impact will be brought in solving the problem. The experiments and analysis should be supplemented.
\begin{center}
	\begin{minipage}{10cm}
		\textit{Agradecemos esta sugestão. Incluímos resultados da resolução de casos com um maior número $K$\/ de nós, mantendo a suposição $K < m$, necessária para a definição prévia da destinação dos pallets. Para isso, utilizamos uma heurística conhecida para o TSP, aplicando nosso algoritmo apenas nesse tour.}
	\end{minipage}
\end{center}

\end{document}

### 2024-01-26 common.py

@@ -9,6 +9,9 @@ import random
 
 CITIES = ["GRU", "GIG", "SSA", "CNF", "CWB", "BSB", "REC"]
 
+# 15 cities for Shims validation
+#CITIES = ["GRU", "GIG", "CGH", "BSB", "CNF", "POA", "REC", "SSA", "FOR", "CWB", "MAO", "BEL", "FLN", "VIX", "GYN"]
+
 # tour is pi in the mathematical formulation
 class Tour(object):
     def __init__(self, nodes, cost):

### 2024-01-26 latex/eswa.tex
tsp_deap working fine

@@
 
-		\multirow{10}{*}{$Gurobi$} & \multirow{2}{*}{240s} & $f$ & 8.53 & 12.24 & 13.34 &  {\bf x}& {\bf x} & 11.81 & 17.02 & 18.31 & {\bf x} & {\bf x} & 17.75 & 25.24 & {\bf x} & {\bf x} & {\bf x} \\ 
-		&               &  {time (s)}                            & 41   & 43    & 123   &  {\bf x}& {\bf x} & 50    & 62    & 181   & {\bf x} & {\bf x} & 166   & 140   & {\bf x} & {\bf x} & {\bf x} \\ 
+		\multirow{10}{*}{$Gurobi$} & \multirow{2}{*}{240s} & $f$ & 8.60 & 12.20 & 13.67 & {\bf x} & {\bf x} & 11.77 & 16.38 & 18.10 & {\bf x} & {\bf x} & 17.89 & 25.42 & {\bf x} & {\bf x} & {\bf x} \\ 
+		&               &  {time (s)}                            & 31   & 35    & 124   & {\bf x} & {\bf x} & 52    & 59    & 200   & {\bf x} & {\bf x} & 188   & 145   & {\bf x} & {\bf x} & {\bf x} \\ 
 		\cmidrule(lr){2-18}
 
-		&                           \multirow{2}{*}{1200s} & $f$ & 8.52 & 12.25 & 13.31 & 14.67   & {\bf x} & 11.81 & 17.02 & 18.25 &  20.64  & {\bf x} & 17.75 & 25.24 & 26.49 & 27.97   & {\bf x} \\ 
-		&        &  {time (s)}                                   & 39   & 41    & 129   & 217     & {\bf x} & 52    & 61    & 190   &  304    & {\bf x} & 161   & 139   & 384   & 579     & {\bf x} \\ 
+		&                           \multirow{2}{*}{1200s} & $f$ & 8.61 & 12.20 & 13.31 & 10.00   & {\bf x} & 11.77 & 17.02 & 18.25 &  20.64  & {\bf x} & 17.75 & 25.24 & 26.49 & 27.97   & {\bf x} \\ 
+		&        &  {time (s)}                                   & 28   & 37    & 129   & 320     & {\bf x} & 46    & 61    & 190   &  304    & {\bf x} & 161   & 139   & 384   & 579     & {\bf x} \\ 
 		\cmidrule(lr){2-18}                              
-		&                           \multirow{2}{*}{2400s} & $f$ & 8.60 & 12.21 & 13.67 & 15.00   & 13.41   & 11.77 & 16.38 & 18.02 & 20.94   & {\bf x} & 17.75 & 25.24 & 26.46 & 29.10   & {\bf x} \\
-		&        &  {time (s)}                                   & 26   & 38    & 134   & 310     & 1520    & 46    & 60    & 206   & 474     & {\bf x} & 158   & 127   & 381   & 543     & {\bf x} \\
+		&                           \multirow{2}{*}{2400s} & $f$ & 8.60 & 12.21 & 13.67 & 15.00   & 13.41   & 11.77 & 16.37 & 18.03 & 20.95   & {\bf x} & 17.89 & 25.44 & 26.15 & 29.13   & {\bf x} \\
+		&        &  {time (s)}                                   & 26   & 38    & 134   & 310     & 1520    & 46    & 60    & 199   & 461     & {\bf x} & 164   & 140   & 383   & 786     & {\bf x} \\
 		\cmidrule(lr){2-18}
@@
 For this task, we implemented a TSP solution method based on the Genetic Algorithm (GA) that returns the 20\% best tours as input to the ACLP-RPDP solved by {\it Shims}.
+
+We implemented the GA with DEAP (Distributed Evolutionary Algorithms in Python) , a evolutionary computation framework available in {\tt github.com/deap/deap}.
+More details can be found in \citep{DEAP_JMLR2012}.
+
+DEAP took 3 seconds to solve a 15-node TSP.
+
+\citep{PeerlinckAmy2022MFEO} applied DEAP to solve a Multi-Objective Knapsack Problem. This algorithm is adapted to solve the ACLPP in each node of this work's solution.
 

### 2024-01-26 ga_deap.py

@@ -1,0 +1,55 @@
+import numpy as np
+from deap import base, creator, tools, algorithms
+
+# Example MKP data
+# Values of items
+values = np.array([60, 100, 120])
+# Weights of items in each dimension
+weights = np.array([[10, 20, 30],  # Dimension 1 weights (volumes)
+                    [20, 15, 25]]) # Dimension 2 weights (weights)
+# Capacities of the knapsack in each dimension
+#                      m3   kg
+capacities = np.array([50, 50])
+
+# Genetic Algorithm setup                          (volumes, weights)
+creator.create("FitnessMax", base.Fitness, weights=(1.0,))
+creator.create("Individual", list, fitness=creator.FitnessMax)
+
+toolbox = base.Toolbox()
+toolbox.register("attr_bool", np.random.randint, 0, 2)
+toolbox.register("individual", tools.initRepeat, creator.Individual, toolbox.attr_bool, n=len(values))
+toolbox.register("population", tools.initRepeat, list, toolbox.individual)
+
+def evalMKP(individual):
+    weight_tot = np.dot(weights, individual)
+    value_tot = np.dot(values, individual)
+    # Check if the weight is greater than the capacity for any dimension
+    if any(weight_tot > capacities):
+        return 0,  # Return 0 if the individual exceeds capacity constraints
+    return value_tot,
+
+toolbox.register("evaluate", evalMKP)
+toolbox.register("mate", tools.cxTwoPoint)
+toolbox.register("mutate", tools.mutFlipBit, indpb=0.05)
+toolbox.register("select", tools.selTournament, tournsize=3)
+
+# Genetic Algorithm parameters
+population_size = 50
+crossover_probability = 0.7
+mutation_probability = 0.2
+number_of_generations = 100
+
+# Solve the MKP
+population = toolbox.population(n=population_size)
+
+stats = tools.Statistics(lambda ind: ind.fitness.values)
+stats.register("avg", np.mean)
+stats.register("max", np.max)
+
+best = tools.HallOfFame(1)
+
+algorithms.eaSimple(population, toolbox, cxpb=crossover_probability, mutpb=mutation_probability, 
+                    ngen=number_of_generations, stats=stats, halloffame=best, verbose=True)
+
+# Best solution
+print(f"Best Individual = {best[0]}, Value = {best[0].fitness.values[0]}")

### 2024-01-27 latex/eswa.tex
Items for the 15 nodes generated.

@@
 
-In practical air transport scenarios, the size of a TSP can therefore be less about the sheer number of nodes and more about the complexity and dynamic nature of the problem, which in our case (a single aircraft in a single day) is rarely expected to be more than seven nodes, based on the experience of the first author when working on an operations control centre.
+In practical air transport scenarios, the size of a TSP can be less about the sheer number of nodes and more about the complexity and dynamic nature of the problem, which in our case (a single aircraft in a single day) is rarely expected to be more than seven nodes.
+Despite this, at the end of this work, we intend to submit the best method found to a more challenging 15-node problem as a validation test.
 }
@@
 
+	\item {\color{blue} Lately, the time limit to solve a node will be adjusted to be proportional to the sum of volumes of the candidate items of the node. }
+
 \end{itemize}
@@
 
-In this algorithm, we use five values for $scenario$, according to Tables \ref{tab:costs} and \ref{tab:scenarios}, which define the number $K$ of destinations, the set $L$\/ of nodes, the costs $C$\/, and the shortest tours $\pi_{TSP1}$\/ and $\pi_{TSP2}$.
+In this algorithm, we use {\color{blue}six} values for $scenario$, according to Tables \ref{tab:costs} and \ref{tab:scenarios}, which define the number $K$ of destinations, the set $L$\/ of nodes, the costs $C$\/, and the shortest tours $\pi_{TSP1}$\/ and $\pi_{TSP2}$.
+
+{\color{blue}
+Scenario 6 will be used only for the best method validation.	
+}
 
@@
 		5 & 6    & \{$0,1,2,3,4,5,6$\}   & 0 4 1 2 6 5 3 0 & 0 3 5 6 2 1 4 0 \\
+		6 & 15   & \{$0,1,2,3,4,5,6,7,8,9,10,11,12,13,14$\}   & not known & not known \\
 		\bottomrule

### 2024-01-27 common.py

@@ -8,7 +8,7 @@ from time import time
 import random
 import pandas as pd
 
-CITIES = ["GRU", "GIG", "SSA", "CNF", "CWB", "BSB", "REC"]
+# CITIES = ["GRU", "GIG", "SSA", "CNF", "CWB", "BSB", "REC"]
 
 # 15 cities for Shims validation
 #CITIES = ["GRU", "GIG", "CGH", "BSB", "CNF", "POA", "REC", "SSA", "FOR", "CWB", "MAO", "BEL", "FLN", "VIX", "GYN"]
@@ -30,8 +30,8 @@ class Config(object):
     def __init__(self, scenario):
         self.weiCap = 0
         self.volCap = 0
-        self.numNodes = {0:3,  1:3,  2:3,    3:4,    4:5,    5:6,    6:7    }[scenario]
-        self.Sce      = {0:1,  1:1,  2:2,    3:3,    4:4,    5:5,    6:6    }[scenario]
+        self.numNodes = {0:3, 1:3, 2:3, 3:4, 4:5, 5:6, 6:7, 7:15 }[scenario]
+        self.Sce      = {0:1, 1:1, 2:2, 3:3, 4:4, 5:5, 6:6, 7:7  }[scenario]
 
         self.size       = "smaller"
         self.numPallets = 7
@@ -46,14 +46,14 @@ class Config(object):
             self.kmCost     = 4.9 
             
 class Node(object):
-    def __init__(self, id, tau):
+    def __init__(self, id, icao):
         self.ID   = id
         self.tLim = 1 # 1s
         self.Vol  = 0.0
-        self.tau  = tau # node sum of torques
-        self.ICAO = CITIES[0]
-        if id < len(CITIES):
-            self.ICAO = CITIES[id]
+        # self.tau  = tau # node sum of torques
+        self.ICAO = icao
+        # if id < len(CITIES):
+        #     self.ICAO = CITIES[id]
 
 class Item(object):
     """
@@ -241,10 +241,10 @@ def loadDistances(fname):
     df = pd.read_csv(fname, sep=' ', index_col=0)
 
     distances = df.values.tolist()
+
+    cities = df.columns.to_list()
      
-    # with open(fname, 'r') as f:
-    #     distances = [ [ float(num) for num in line.split(' ') ] for line in f ] 
-    return distances 
+    return distances, cities
 
 def factorial(x):
     result = 1

### 2024-01-27 latex/eswa.tex
Included a map of Brazil

@@
 To validate the effectiveness of the {\it Shims} heuristic, we solve a 15-node TSP containing the 15 main Brazilian airports.
-For this task, we implemented a TSP solution method based on the Genetic Algorithm (GA) that returns the 20\% best tours as input to the ACLP-RPDP solved by {\it Shims}.
+For this task, we implemented a TSP solution method based on the Genetic Algorithm (GA) that returns a population of 100 tours as input to the ACLP-RPDP solved by {\it Shims}.
+The GA run time was around three seconds, and the total run time was around 33 seconds.
 
@@
 
-DEAP took 3 seconds to solve a 15-node TSP.
-
 \citep{PeerlinckAmy2022MFEO} applied DEAP to solve a Multi-Objective Knapsack Problem. This algorithm is adapted to solve the ACLPP in each node of this work's solution.
 
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.25]{brazil.png}
+	\caption{ {\it Shims/GA} best tour (may not be optimal)}
+	\label{fig:brazil}
+\end{figure}
 }

### 2024-01-27 main.py

@@ -349,11 +349,12 @@ if __name__ == "__main__":
     iRace_testing = False
 
     # scenarios = [2,3,4,5,6] # represent 1,2,3,4,5 in the article
-    scenarios = [2]
+    scenarios = [7] # 7: 15-node problem
+    # scenarios = [2]
 
-    # folder = "surplus20"  # 1.2
+    folder = "surplus20"  # 1.2
     
-    folder = "surplus50"  # 1.5
+    # folder = "surplus50"  # 1.5
 
     # folder = "surplus100" # 2.0
 
@@ -402,13 +403,15 @@ if __name__ == "__main__":
     # tipo = "KP"
     tipo = "FFD"
 
-    dists, _ = common.loadDistances("./params/distances7.txt") # dists, cities
+    distances_file = "./params/distances7.txt"
+    if scenarios[0] == 7:
+        distances_file = "./params/distances15.txt"
+
+    dists, _ = common.loadDistances(distances_file) # dists, cities
     costs = [[0.0 for _ in dists] for _ in dists]
 
     for scenario in scenarios:
 
-        instances = [1,2,3,4,5,6,7]
-        
         if testing:
             instances = [1]
 
@@ -416,6 +419,13 @@ if __name__ == "__main__":
             instances = [iRace_instance]
 
         cfg = common.Config(scenario)
+
+        instances = [1,2,3,4,5,6,7]
+
+        print(f"{cfg.numNodes} nodes")
+
+        if cfg.numNodes > 7: # Shims validation
+            instances = [1]
 
         # time limit per node
         secBreak = timeLimit/common.factorial(cfg.numNodes)
@@ -438,8 +448,8 @@ if __name__ == "__main__":
             cfg.weiCap = cfg.payload
 
         perc = 1.0
-        # if cfg.numNodes > 3:
-            # perc = 0.25
+        if cfg.numNodes > 3:
+            perc = 0.25 # discard the worst tours
 
         instanceTime = 0.
         instanceTime2 = 0. # with 3D packing
@@ -457,15 +467,22 @@ if __name__ == "__main__":
             # leastSC = 0. # minimum cost score/cost relation
             bestAV = 0.
             bestAT = 0.
-            # tours = common.getTours(cfg.numNodes-1, costs, perc)
-            tours = tsp_deap.getTours("./params/distances7.txt", cfg.numNodes)
 
-            print( len(tours), len(tours[0].nodes) )
+
+            if cfg.numNodes <= 7:
+                # permutation of nodes - TSP solution with all tours
[…]

### 2024-01-27 latex/eswa.tex
Brazil figura changed

@@
 \end{figure}
+
+The airports cited in Figure \ref{fig:brazil} are: São Paulo (GRU), Congonhas (CGH), Curitiba (CWB), Porto Alegre (POA), Florianópolis (FLN),  Vitória (VIX), Salvador (SSA), Recife (REC), Fortaleza (FOR), Belém (BEL), Manaus (MAO), Brasília (BSB), Goiânia (GYN), Confins (CNF), and Rio de Janeiro (GIG), returning to GRU.
+
 }

### 2024-01-27 tsp_deap.py

@@ -22,6 +22,20 @@ Vitória-Eurico de Aguiar Salles Airport (VIX)
 Goiânia-Santa Genoveva Airport (GYN)
 """
 
+def rotate(tour_nodes, base_node):
+    """
+    Shifts the elements of the list such that the specified base_node becomes the first element.
+    The rest of the list follows in the original order.
+    If the base_node is not found in the list, the original list is returned.
+    """
+    if base_node in tour_nodes:
+        index = tour_nodes.index(base_node)
+        return tour_nodes[index:] + tour_nodes[:index]
+    else:
+        return tour_nodes
+
+
+
 def getTours(distances_file, numNodes):   
 
     # Load the distance matrix from the file
@@ -32,8 +46,6 @@ def getTours(distances_file, numNodes):
     distance_matrix = distance_matrix.head(numNodes)
 
     distance_matrix = pd.DataFrame(distance_matrix, columns = cities)
-
-    # print(cities)
 
     # Convert the DataFrame to a numpy array for easier manipulation
     distance_matrix_np = distance_matrix.to_numpy()
@@ -102,10 +114,16 @@ def getTours(distances_file, numNodes):
 
         for nid in tour:
             tnodes.append(nodes[nid])
-            
+
         tours.append( common.Tour(tnodes, best_distances[i]) )
 
     return tours # list of tours
+
+def list_to_string(lst):
+    """
+    Converts a list into a string with elements separated by spaces.
+    """
+    return ' '.join(map(str, lst))
 
 if __name__ == "__main__":
 
@@ -114,19 +132,29 @@ if __name__ == "__main__":
     # distances_file = './params/distances15.txt'
     distances_file = './params/distances7.txt'
 
-    tours = getTours(distances_file, 3)
+    tours = getTours(distances_file, 7)
+
+    base_node = "GRU"
 
     for tour in tours:
-        origin = tour.nodes[0]
-        print(f"{origin.ICAO} ", end='')
-        prev = tour.nodes[0]
-        for j, node in enumerate(tour.nodes):
-            if j > 0:
-                print(f"{node.ICAO} ", end='')
-                prev = node
+
+        icaos = []
+        for n in tour.nodes:
+            icaos.append(n.ICAO)
+
+        icaos = rotate(icaos, base_node)
+
+        print(icaos)
+
+        # origin = tour.nodes[0]
+        # stour = f"{origin.ICAO} "
+        # prev = tour.nodes[0]
[…]

### 2024-01-27 latex/responses.tex
ga_deap_mmkp not working propperly

@@
 \usepackage{color, colortbl}
+
+\usepackage{natbib}
+\bibliographystyle{elsarticle-harv}
+
+\usepackage{graphicx}
+\usepackage{float} % to position figures and tables
+
+\newcommand{\boldm}[1] {\mathversion{bold}#1\mathversion{normal}}
+
+\usepackage{booktabs}
 
@@
 		\textit{Agradecemos esta sugestão. Entre os novos resultados apresentados, incluímos uma resolução na qual o tempo dedicado a cada nó é proporcional à carga disponível para embarque.}
+		{\color{blue}
+
+An important experiment inclusion was regarding the time limit to solve a node.
+This was adjusted to be proportional to the sum of the volumes of the candidate items in the node, as each node has a different number of items to be embarked.
+
+This inclusion may have contributed to the slight improvement in the \textit{Shims} results.
+
+		}
+
 	\end{minipage}
@@
 		\textit{Estudamos a viabilidade de definir empiricamente os valores dessas constantes do algoritmo Shims. Explicar...}
+
+{\color{blue}
+	After the items were generated, some experiments with the \textit{iRace} tool were executed to determine the values for parameters $threshold_1$ and $threshold_2$ used in \textit{Shims}, whose results are presented in Table \ref{tab:irace}.
+
+	\begin{table}[H]
+		\centering
+		\caption{\textit{iRace} testing results}  \label{tab:irace}
+		\footnotesize
+		\begin{tabular}{cccc}
+			\toprule
+			\textbf{surplus} & \boldm{$threshold_1$} & \boldm{$threshold_2$} & \textbf{time (min)}\\
+			\midrule
+			1.2& 0.8621 & 1.0539 & 47 \\
+			1.5& 0.9199 & 1.1399 & 59 \\
+			2.0& 0.9617 & 1.5706 & 63 \\						
+			\bottomrule
+		\end{tabular}
+		\normalsize
+	\end{table}
+
+	For these tests, we considered the odd instances (1, 3, 5, 7) as the \textbf{training} set and the even instances (2, 4, 6) as the \textbf{testing} set for the \textit{iRace} runs.
+We supplied \textit{iRace} for $threshold_1$ determination, the range [0.8, 1.0], and for $threshold_2$, the range [1.0, 2.0].
+The number of iterations in each experiment was limited to 3000 executions for \textit{iRace} to have plenty of data for its statistical tests.
+More detail about these configurations may be found on the \textit{iRace} user guide in {\tt cran.r-project.org/web/packages/irace/irace.pdf}.
+}
+
 	\end{minipage}
@@
 Caso reste algum ponto obscuro, pedimos ao revisor que o especifique, por favor.}
+
+{\color{blue}
+
+An important experiment inclusion was regarding the time limit to solve a node.
+This was adjusted to be proportional to the sum of the volumes of the candidate items in the node, as each node has a different number of items to be embarked.
+
+}
+
 	\end{minipage}
@@
 		\textit{Analisamos as resoluções do ACLPP através de algoritmos genéticos. Explicar...}
+
+		{\color{blue}
+	We implemented the genetic algorithm with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework available at {\tt github.com/deap/deap}.
+More details can be found in \citep{DEAP_JMLR2012}.
+
+	We used this package to solve a 15-node TSP with success, whose run time was less than 3s, and generated near-optimal solutions.
+
+	\citep{PeerlinckAmy2022MFEO} applied DEAP to solve a multi-objective knapsack problem. We adapted their algorithm to solve the ACLPP in each node of this work's solution.
+
+Our algorithm is not complete, as we did not include the constraints regarding the weight and balance problem or the pickup and delivery problem.
+
+In the first attempt, considering only the weight and volume constraints, our implementation of DEAP managed to solve problems up to 100 items to be allocated to 18 pallets in less than 10 seconds.
+
+When we included the item count constraint to prevent each item from being allocated to more than one pallet, it was able to solve only small problems (20 items on 2 pallets).
+
+We increased the number of items to 100, and the number of pallets to 18, but, even adjusting the GA parameters to more generations (2400) and more individuals (1200), DEAP did not generate any feasible solution.
+The run time of this attempt was around 5 minutes.
+
+For the crossover probability, we tested the range [0.6 ? 0.8], and for the mutation probability, we tested from 0.01 to 0.1 with no success.
+
+Up to this point in time, we had no success solving the ACLPP with DEAP.
+
+The code implemented is available in {\tt //github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/ga\_deap\_mmkp.py}.
+
+		}
 	\end{minipage}
@@
 Para isso, utilizamos uma heurística conhecida para o TSP, aplicando nosso algoritmo apenas nesse tour.}
+
+		{\color{blue}
+
+To validate the effectiveness of the {\it Shims} heuristic, we solve a 15-node TSP containing the 15 main Brazilian airports.
+For this task, we implemented a TSP solution method based on the Genetic Algorithm (GA) that returns a population of 100 tours as input to the ACLP-RPDP solved by {\it Shims}.
+The GA run time was around three seconds, and the total run time was around 33 seconds.
+
+We implemented the GA with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework available in {\tt github.com/deap/deap}.
+More details can be found in \citep{DEAP_JMLR2012}.
+
+The code implemented is available in {\tt //github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/tsp\_deap.py}.
+
+	\begin{figure}[H]
+		\centering
+		\caption{\label{fig:brazil} {\it Shims/GA} best tour (may not be optimal)}
+		\includegraphics[scale=0.3]{brazil.png}
+	\end{figure}
+
+\vspace{3mm}
+
+The airports cited in Figure \ref{fig:brazil}  are: São Paulo (GRU), Congonhas (CGH), Curitiba (CWB), Porto Alegre (POA), Florianópolis (FLN),  Vitória (VIX), Salvador (SSA), Recife (REC), Fortaleza (FOR), Belém (BEL), Manaus (MAO), Brasília (BSB), Goiânia (GYN), Confins (CNF), and Rio de Janeiro (GIG), returning to GRU.
+		}
 	\end{minipage}
@@
 
+\bibliographystyle{apalike}
+
+\bibliography{bib}
+
 \end{document}

### 2024-01-28 latex/responses.tex
GA DEAP for ACLPP not working

@@
 
-	\citep{PeerlinckAmy2022MFEO} applied DEAP to solve a multi-objective knapsack problem. We adapted their algorithm to solve the ACLPP in each node of this work's solution.
+	\vspace{3mm}
 
-Our algorithm is not complete, as we did not include the constraints regarding the weight and balance problem or the pickup and delivery problem.
+	\citep{PeerlinckAmy2022MFEO} applied DEAP to solve a multi-objective knapsack problem.
+
+	\vspace{3mm}
+
+	Our GA implementation is not complete. Due to performance issues, we did not include the constraints regarding the Weight and Balance Problem or the Pickup and Delivery Problem.
 
 	In the first attempt, considering only the weight and volume constraints, our implementation of DEAP managed to solve problems up to 100 items to be allocated to 18 pallets in less than 10 seconds.	
-
 	When we included the item count constraint to prevent each item from being allocated to more than one pallet, it was able to solve only small problems (20 items on 2 pallets).
@@
 
-For the crossover probability, we tested the range [0.6 ? 0.8], and for the mutation probability, we tested from 0.01 to 0.1 with no success.
+	For the crossover probability, we tested the range [0.6 - 0.8], and for the mutation probability, we tested from 0.01 to 0.1 with no success.
 
-Up to this point in time, we had no success solving the ACLPP with DEAP.
+	\vspace{3mm}
 
@@
 
+	\vspace{3mm}
+
+Large chromosome sizes, like 18,000-dimension arrays, can pose significant challenges in using Genetic Algorithms (GA).
+
+The computational complexity is significant, as each generation requires evaluating individual fitness, which can be resource-intensive.
+
+Large chromosomes can slow GA convergence due to the vast search space, potentially requiring more generations to find optimal solutions.
+
+Efficient encoding and decoding of the problem into the chromosome and decoding solutions is crucial for a 18,000-dimension array. 
+Also, the design of the fitness function becomes more critical as chromosome size increases, as it needs to guide the evolutionary process without being overly computationally demanding.
+
+\vspace{3mm}
+
+Up to this point in time, we had no success solving the ACLPP with DEAP because it included the palletization planning of 200 to 1000 items per node in a set of 18 aircraft pallets.
 		}
@@
 
+\vspace{3mm}
+
 The code implemented is available in {\tt //github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/tsp\_deap.py}.
+
+\vspace{3mm}
 
@@
 		\centering
-		\caption{\label{fig:brazil} {\it Shims/GA} best tour (may not be optimal)}
+		\caption{\label{fig:brazil} {\it Shims/GA} best ACLP+RPDP tour}
 		\includegraphics[scale=0.3]{brazil.png}

### 2024-01-28 ga_deap_mmkp.py

@@ -17,7 +17,8 @@ def Solve(items_file, node_ID):
     # filter items on the base
     items_matrix_np = items_matrix_np[((items_matrix_np[:,3] == node_ID))]
 
-    items_matrix_np = items_matrix_np[:100,:]
+    # num_items_test = 250
+    # items_matrix_np = items_matrix_np[:num_items_test,:]
 
     #  w    s    v  from to
     items_weights = items_matrix_np[ :,0]
@@ -25,6 +26,10 @@ def Solve(items_file, node_ID):
     items_volumes = items_matrix_np[ :,2]
 
     #               CG     Vol   Weight TO
+    # capacities = [( 14.89, 14.8, 4500, -1), # pallet 0
+    #               ( 14.89, 14.8, 4500, -1), # pallet 1
+    #               (-17.57,  7.0, 3000, -1)] # pallet 17
+
     capacities = [( 14.89, 14.8, 4500, -1), # pallet 0
                   ( 14.89, 14.8, 4500, -1), # pallet 1
                   ( 11.47, 14.8, 4500, -1),
@@ -43,7 +48,7 @@ def Solve(items_file, node_ID):
                   (-13.17, 14.8, 4500, -1),
                   (-17.57, 10.0, 3000, -1),
                   (-17.57,  7.0, 3000, -1)] # pallet 17
-
+    
     n_items     = len(items_scores)
     n_pallets = len(capacities)
 
@@ -56,6 +61,7 @@ def Solve(items_file, node_ID):
         volume_used = [0] * n_pallets
         item_count  = [0] * n_items
 
+        penalty = 0
         for i in range(n_items):
             for j in range(n_pallets):
                 if individual[i * n_pallets + j] == 1: # if item "i" is selected for pallet "j"
@@ -66,14 +72,24 @@ def Solve(items_file, node_ID):
         
                 # Check if an item is placed in more than one pallet
                 if item_count[i] > 1:
-                    return 0,  # Penalize if an item is in more than one pallet
+                    penalty += 200*(1 - item_count[i])
+
+        if penalty < 0:
+            return score_sum + penalty,  # Penalize if are items in more than one pallet, returns a tuple
             
         # Check if the weights and volumes are within the capacities for each pallet
+        # returns a penalty (negative score) proportional to the exceed value
         for j in range(n_pallets):
-            if volume_used[j] > capacities[j][1] or weight_used[j] > capacities[j][2]:
[…]

### 2024-01-28 latex/responses.tex
Latex files updated

@@
 \usepackage{natbib}
-\bibliographystyle{elsarticle-harv}
+%\bibliographystyle{elsarticle-harv}
+\bibliographystyle{apalike}
 
@@
 
-{\color{blue}
-New experiments were conducted due to two changes suggested by the reviewers:
-
-\begin{itemize}
-\item [1.] Node solution times depended on the total volume of candidate items to be loaded in each node in relation to the total items volume to be loaded in the tour.
-\item [2.] A new process of determining the {\it Shims} volume thresholds for $eta_1$ and $eta_2$.
-\end{itemize}
-}
-
 \section{\#2-b: Our Point-to-Point responses to the Associate Editor}
@@
 {\color{blue}
+	New experiments were conducted due to two changes suggested by the reviewers:
+
+	\begin{itemize}
+		\item [1.] Node solution times depended on the total volume of candidate items to be loaded in each node in relation to the total items volume to be loaded in the tour.
+		\item [2.] A new process of determining the {\it Shims} volume thresholds for $eta_1$ and $eta_2$.
+	\end{itemize}
+
 	We included Table 10 with the results of the {\it iRace} tool, a software package that helps to automatically fine-tune algorithm parameters.
@@
 
-This inclusion may have contributed to the slight improvement in the \textit{Shims} results.
+This inclusion certainly contributed to the slight improvement observed in the results.
 
@@
 {\color{blue}
-	After the items were generated, some experiments with the \textit{iRace} tool were executed to determine the values for parameters $threshold_1$ and $threshold_2$ used in \textit{Shims}, whose results are presented in Table \ref{tab:irace}.
+	After the items were generated, some experiments with the \textit{iRace} tool were executed to determine the values for volume limits $threshold_1$ and $threshold_2$ used in \textit{Shims}, whose results are presented in Table \ref{tab:irace}.
 
@@
 			\toprule
-			\textbf{surplus} & \boldm{$threshold_1$} & \boldm{$threshold_2$} & \textbf{time (min)}\\
+
+			\boldm{$surplus$} & \boldm{$threshold_1$} & \boldm{$threshold_2$} & \textbf{run time}\\
+			                 & for $\eta_1$          & for $\eta_2$          & (min) \\
 			\midrule
@@
 The number of iterations in each experiment was limited to 3000 executions for \textit{iRace} to have plenty of data for its statistical tests.
-More detail about these configurations may be found on the \textit{iRace} user guide in {\tt cran.r-project.org/web/packages/irace/irace.pdf}.
+More detail about these configurations may be found on the \textit{iRace} user guide in \url{https://cran.r-project.org/web/packages/irace/irace.pdf}.
 }
@@
 		{\color{blue}
-	We implemented the genetic algorithm with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework available at {\tt github.com/deap/deap}.
+	We implemented the genetic algorithm with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework available at \url{https://github.com/deap/deap}.
 More details can be found in \citep{DEAP_JMLR2012}.
-
-	We used this package to solve a 15-node TSP with success, whose run time was less than 3s, and generated near-optimal solutions.
 
@@
 
-	Our GA implementation is not complete. Due to performance issues, we did not include the constraints regarding the Weight and Balance Problem or the Pickup and Delivery Problem.
+	Our GA implementation is not complete. Due to performance issues, we did not include the constraints regarding the Weight and Balance (WBP) or the Pickup and Delivery (PDP) problems.
 
@@
 
-	We increased the number of items to 100, and the number of pallets to 18, but, even adjusting the GA parameters to more generations (2400) and more individuals (1200), DEAP did not generate any feasible solution.
-The run time of this attempt was around 5 minutes.
+	We increased the number of items to 150, and the number of pallets to 18, but, even adjusting the GA parameters to more generations (2400) and more individuals (1200), DEAP did not generate any feasible solution.
+The run time of this attempt was around 10 minutes.
 
@@
 
-	The code implemented is available in {\tt //github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/ga\_deap\_mmkp.py}.
+	The code implemented is available in \url{https://github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/ga\_deap\_mmkp.py}.
 
@@
 
-Large chromosome sizes, like 18,000-dimension arrays, can pose significant challenges in using Genetic Algorithms (GA).
+	The palletization planning of 200 to 1000 items per node in a set of 18 aircraft pallets may be an inadequate problem to be solved with GA.
+
+Large chromosome sizes, like 18,000-dimension arrays, can pose significant challenges in using genetic algorithms.
 
@@
 
-Up to this point in time, we had no success solving the ACLPP with DEAP because it included the palletization planning of 200 to 1000 items per node in a set of 18 aircraft pallets.
+We hope that these considerations help justify why we did not include GA as a candidate method to solve the ACLP+RPDP.
 		}
@@
 
-We implemented the GA with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework available in {\tt github.com/deap/deap}.
+We implemented the GA with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework available in \url{https://github.com/deap/deap}.
 More details can be found in \citep{DEAP_JMLR2012}.
@@
 
-The code implemented is available in {\tt //github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/tsp\_deap.py}.
+The code implemented is available in \url{https://github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/tsp\_deap.py}.
 
@@
 The airports cited in Figure \ref{fig:brazil}  are: São Paulo (GRU), Congonhas (CGH), Curitiba (CWB), Porto Alegre (POA), Florianópolis (FLN), Vitória (VIX), Salvador (SSA), Recife (REC), Fortaleza (FOR), Belém (BEL), Manaus (MAO), Brasília (BSB), Goiânia (GYN), Confins (CNF), and Rio de Janeiro (GIG), returning to GRU.
+
 }
@@
 
-\bibliographystyle{apalike}
-
 \bibliography{bib}

### 2024-01-28 latex/eswa.tex
ga_deap_mmkp not working propperly
GA DEAP for ACLPP not working
Latex files updated
More comments included in the responses

@@
 \usepackage{natbib}
-\bibliographystyle{elsarticle-harv}
+%\bibliographystyle{elsarticle-harv}
 
@@
 
-Recently, \citep{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP), and {\it Weight and Balance Problem} (WBP).
+A few years ago, \citep{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP), and {\it Weight and Balance Problem} (WBP).
 Several aspects were considered: item characteristics to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs); when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how total cargo weight is balanced; etc.
+
+{\color{blue}
+
+Recently, \citep{zhao2023} developed integer programming solution that address ACLPP and WBP separately, impacting payload optimization and aircraft centre of gravity (CG).
+The authors argue that, with improvements in computer processing power, joint combinatorial optimization of ACPP and WBP is now feasible.
+They propose three integer linear programming models: a bi-objective optimization model (BOM), a combinatorial optimization model (COM), and an improved combinatorial optimization model (IOM).
+The models aim to determine the maximum loading capacity and lowest CG deviation from a specified target CG.
+
+The study uses Gurobi to solve four scenarios with various conditional metrics for three models.
+Results show that the BOM has the fastest solution speed, but the CG deviation is the largest, making the results unacceptable in some cases.
+The COM has the longest solution time, making it difficult to tolerate in practice.
+The IOM offers the best optimization solution, despite taking a little longer to solve computationally than the BOM.
+
+Using a Boeing 777F, four scenarios and 25 cases were tested, finding that the BOM's variable feasible domain is limited, the optimization result is insufficient, and the CG deviation is sometimes too large.
+The COM takes too long to solve, making it difficult to meet the needs of the actual scene.
+The IOM accelerates the solving process compared to the COM, but some scenarios still require significant computation time.
+
+}
 
@@
 
-\begin{table}[htb]
+\begin{table}[htp]
 
@@
 
-\begin{table}[h]
+\begin{table}[htp]
 	\centering
@@
 
+\vspace{-2mm}
+
 \begin{itemize}
@@
 
-\begin{figure}[h]
+\begin{figure}[htp]
 	\centering
@@
 We also make the following assumptions:
+
+\vspace{-2mm}
 
@@
 
-\begin{table}[ht]
+\begin{table}[htp]
 	\centering
@@
 
-\begin{table}[H]
+\begin{table}[htp]
 	\centering
@@
 
+\vspace{-2mm}
+
 \begin{itemize}
+	\setlength\itemsep{-0.3em}
 
@@
 
-\begin{algorithm}[H]
+\begin{algorithm}[htp]
 
@@
 In the generation of test instances, we use two types of random selections:
+
+\vspace{-2mm}
+
 \begin{itemize}
@@
 		\toprule
-			\textbf{surplus} & \boldm{$threshold_1$} & \boldm{$threshold_2$} & \textbf{time (min)}\\
+
+		\boldm{$surplus$} & \boldm{$threshold_1$} & \boldm{$threshold_2$} & \textbf{run time}\\
+		& for $\eta_1$          & for $\eta_2$          & (min) \\
 		\midrule
@@
 \begin{table}[H]
-{\color{blue}
-	\centering
-	\caption{Overall results}  \label{tab:mr}
+
+	\centering
+	\caption{ \color{blue}Overall results}  \label{tab:mr}
 	\footnotesize
@@
 \normalsize
-}
+
 \end{table}
@@
 To facilitate the comparison between both, we added a last column in the tables where two values are indicated:
+
+\vspace{-2mm}
 
@@
 From these data, we can draw some conclusions:
+
+\vspace{-2mm}
 
@@
 
-\begin{figure}
-	\centering
-	\includegraphics[scale=0.25]{brazil.png}
-	\caption{ {\it Shims/GA} best tour (may not be optimal)}
-	\label{fig:brazil}
+	\begin{figure}[H]
+		\centering
+		\caption{\label{fig:brazil} {\it Shims/GA} best ACLP+RPDP tour \\
+			\tiny{Solution for scenario 7, instance 1}}
+		\fbox{\includegraphics[scale=0.30]{brazil.png}}
 	\end{figure}
@@
 
+It is important to remember that this is not the shortest tour, but the one with the best relationship with score, cost, and fuel consumption.
+
+}
+
+{\color{purple}
+
+	It is also important to highlight that three strategies were essential for us to obtain a solution in the instances created:
+
+	\vspace{-2mm}
+
+	\begin{itemize}
+		\setlength\itemsep{-0.3em}
+		\item[(1)] consider $K<m$\ so that each pallet received a predefined destination;
+		\item[(2)] on each tour, calculate a node-by-node solution; and
+		\item[(3)] at each intermediate node of the considered tour, relocate the packed contents, minimizing the torque on the aircraft.
+	\end{itemize}
+
+	\vspace{-2mm}
+
+	Without adopting these three strategies, we were unable to carry out any experiments, even in smaller instances.
 }

### 2024-01-28 latex/responses.tex
More comments included in the responses

@@
 		\textit{Como pode ser observado logo abaixo, aumentamos substancialmente os dados experimentais apresentados. Consequentemente, houve também um sensível aumento na discussão sobre eles.}
+
+		{\color{blue}As can be seen below, we have substantially increased the experimental data presented.
+Consequently, there was also a noticeable increase in discussion about them.
+Especially about the overall improvement caused by the run time limit proportional to the volumes of items in each node and the additional experiments to validate the best method found.
+		}
 	\end{minipage}
@@
 		\textit{Incluímos e comentamos novas referências (procurar trabalhos recentes sobre TSP e PDP em aeronaves).}
+
+		{\color{blue}
+		We included the work of \citep{zhao2023}, that developed an air cargo loading plan that addresses the Air Cargo Palletization Problem (ACPP) and the Aircraft Weight and Balance Problem (WBP) separately, impacting payload optimization and aircraft center of gravity (CG).
+The authors propose three integer linear programming models: a bi-objective optimization model (BOM), a combinatorial optimization model (COM), and an improved combinatorial optimization model (IOM).
+The models aim to determine the maximum loading capacity and lowest CG deviation from a specified target CG.
+
+		The study uses Gurobi to solve four scenarios with various conditional metrics for the three models.
+Results show that the BOM has the fastest solution speed, but the CG deviation is the largest, making the results unacceptable in some cases.
+The COM takes too long to solve, making it difficult to meet the needs of the actual scene.
+The IOM accelerates the solving process, but some scenarios still require significant computation time.
+		}
+
 	\end{minipage}
@@
 		\textit{Aumentamos a revisão da literatura, descrevendo os trabalhos com maior detalhe. Acrescentamos também alguns trabalhos mais recentes. Citar quais são esses trabalhos...}
+
+		{\color{blue}
+	We included the work of \citep{zhao2023}, that developed an air cargo loading plan that addresses the Air Cargo Palletization Problem (ACPP) and the Aircraft Weight and Balance Problem (WBP) separately, impacting payload optimization and aircraft center of gravity (CG).
+The authors propose three integer linear programming models: a bi-objective optimization model (BOM), a combinatorial optimization model (COM), and an improved combinatorial optimization model (IOM).
+The models aim to determine the maximum loading capacity and lowest CG deviation from a specified target CG.
+
+	The study uses Gurobi to solve four scenarios with various conditional metrics for the three models.
+Results show that the BOM has the fastest solution speed, but the CG deviation is the largest, making the results unacceptable in some cases.
+The COM takes too long to solve, making it difficult to meet the needs of the actual scene.
+The IOM accelerates the solving process, but some scenarios still require significant computation time.
+}
+
 	\end{minipage}
@@
 		\textit{Incluímos figuras comparativas, mostrando o melhor desempenho da heurística {\it Shims}. Citar quais são essas figuras...}
+
+		{\color{blue}
+		We included a map showing a 15-node tour as to validate the effectiveness of the {\it Shims} heuristic, we solve a 15-node TSP containing the 15 main Brazilian airports.
+		For this task, we implemented a TSP solution method based on the Genetic Algorithm (GA) that returns a population of 100 tours as input to the ACLP-RPDP solved by {\it Shims}.
+		}
 	\end{minipage}
@@
 		\textit{Nas novas figuras incluídas no artigo, mostramos o desempenho dos algoritmos desenvolvidos. Citar quais são essas figuras...}
+
+		{\color{blue}
+The GA run time to solve a 15-node TSP was around three seconds, and the total run time (including the ACLP+RPDP solution by Shims) was around 33 seconds, which we believe is adequate for an operational scenario and using a common hand-held computer.
+
+A figure was included showing the solution tour.
+	}
 	\end{minipage}
@@
 		\textit{Incluímos referências mais recentes (citá-las). De qualquer modo, preferimos manter as antigas, para assim termos um levantamento histórico da evolução da pesquisa neste campo.}
+
+		{\color{blue}
+		We included a more recent reference, \citep{zhao2023}, that demonstrates how hard it is to solve the ACLP and WBP, separately or together.
+
+		In any case, we prefer to keep the old ones so that we have a historical survey of the evolution of research in this field.	
+		}
 	\end{minipage}
@@
 Procuramos deixar isto mais claro na nossa conclusão.}
+
+		{\color{blue}
+		It may not have been clear in the original text, but three strategies were essential for us to obtain a solution in the instances created:
+
+		\begin{itemize}
+		\item[(1)] consider $K<m$\ so that each pallet received a predefined destination;
+
+		\item[(2)] on each tour, calculate a node-by-node solution; and
+
+		\item[(3)] at each intermediate node of the considered tour, relocate the packed contents, minimizing the torque on the aircraft.
+		\end{itemize}
+
+		Without adopting these three strategies, we were unable to carry out any experiments, even in smaller instances.
+The only strategy that allowed different experiments and comparisons was the heuristic used in the node-by-node solution.
+We seek to make this clearer in the discussion of the results and in our conclusion.
+		}
 	\end{minipage}
@@
 Por este motivo, incluímos novas tabelas comparativas com os resultados obtidos nesses casos.}
+
+		{\color{blue}
+		As we are dealing with a new problem, we do not have other solutions available in the literature.
+In fact, we only have well-known meta-heuristics for the node-to-node solution, in which the aircraft torque is minimized.
+For this reason, we have included new comparative tables with the results obtained in these cases.
+		}
 	\end{minipage}
@@
 
-	The code implemented is available in \url{https://github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/ga\_deap\_mmkp.py}.
+	The implemented code is available in \url{https://github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/ga\_deap\_mmkp.py}.
 
@@
 
-The code implemented is available in \url{https://github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/tsp\_deap.py}.
+The implemented code is available in \url{https://github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/tsp\_deap.py}.
 

### 2024-01-28 latex/eswa.tex
Main flow diagram

@@
 \usetikzlibrary{positioning}
+\usetikzlibrary{shapes.arrows}
 
@@
 \geometry{a4paper}
+
+% lets get the exact color
+\usepackage{xcolor}
+\definecolor{myblue}{HTML}{B0D7FF}
+
+\tikzset{
+	mybox/.style={rectangle,
+		draw,
+		fill= myblue,
+		rounded corners,
+		minimum width=2cm,
+		inner sep=5pt,
+		align=left,
+		minimum height=1cm
+	},
+	myarrow/.style={draw=black,
+		fill=white,
+		minimum width=0.6cm,
+		single arrow
+	},
+	longarrow/.style={draw=none,
+		shading=axis,
+		left color=white,
+		right color=myblue,
+		minimum width=0.6cm,
+		single arrow,
+		anchor=east
+	}
+}
 
@@
 Scenario 6 will be used only for the best method validation.
+
+Figure \ref{fig:flow} depicts the Algorithm \ref{alg:main} for easy understanding of the main flow.
+
 }
+
+\begin{figure}[htp]
+
+	\centering
+	\begin{tikzpicture}
+
+		%% Make the boxes
+		\node [mybox] (first)                        {Select:\\ 1. scenario\\2. surplus\\3. time limit};
+		\node [mybox, right=0.7cm of first] (second) {Config:\\4. parameters\\5. cost matrix\\6. pallets\\7. items};
+		\node [mybox, right=0.7cm of second] (third) {8. Define tour\\     schema:\\- All tours\\ \ \ \ (K-1)!\\ \ \ \ or\\- Solve a TSP\\ \ \ \ (set of tours))};
+		\node [mybox, right=0.7cm of third] (fourth) {Solve ACLPP+RPDP:\\9. Many tours\\ \ \ \ (node-by-node)\\10. Select best tour};
+
+		%% Draw arrows between them
+		\node[myarrow] at ([xshift=8pt]first.east) {\phantom{\tiny{aaa}}};
+		\node[myarrow] at ([xshift=8pt]second.east) {\phantom{\tiny{aaa}}};
+		\node[myarrow] at ([xshift=8pt]third.east) {\phantom{\tiny{aaa}}};
+
+	\end{tikzpicture} 
+	\caption{Solution main flow}
+	\label{fig:flow}
+\end{figure}
 

### 2024-01-28 latex/responses.tex
Main flow diagram

@@
 \usepackage{booktabs}
+
+\usepackage{tikz}
+\usetikzlibrary{calc}
+\usetikzlibrary{positioning}
+\usetikzlibrary{shapes.arrows}
+% lets get the exact color
+\usepackage{xcolor}
+\definecolor{myblue}{HTML}{B0D7FF}
+
+\tikzset{
+	mybox/.style={rectangle,
+		draw,
+		fill= myblue,
+		rounded corners,
+		minimum width=2cm,
+		inner sep=5pt,
+		align=left,
+		minimum height=1cm
+	},
+	myarrow/.style={draw=black,
+		fill=white,
+		minimum width=0.6cm,
+		single arrow
+	},
+	longarrow/.style={draw=none,
+		shading=axis,
+		left color=white,
+		right color=myblue,
+		minimum width=0.6cm,
+		single arrow,
+		anchor=east
+	}
+}
 
@@
 		\textit{Agradecemos esta sugestão. Acrescentamos novas figuras e tabelas que ilustram melhor nossos resultados. Citar quais são...}
+
+		{\color{blue}
+		We included the figure below to depict the main algorithm for easy understanding of the main flow.
+		}
+
+		\begin{tikzpicture}
+
+			\centering
+
+		%% Make the boxes
+\node [mybox] (first)                        {Select:\\ 1. scenario\\2. surplus\\3. time limit};
+\node [mybox, right=0.7cm of first] (second) {Config:\\4. parameters\\5. cost matrix\\6. pallets\\7. items};
+\node [mybox, right=0.7cm of second] (third) {8. Define tour\\     schema:\\- All tours\\ \ \ \ (K-1)!\\ \ \ \ or\\- Solve a TSP\\ \ \ \ (set of tours))};
+\node [mybox, right=0.7cm of third] (fourth) {Solve ACLPP+RPDP:\\9. Many tours\\ \ \ \ (node-by-node)\\10. Select best tour};
+
+%% Draw arrows between them
+\node[myarrow] at ([xshift=8pt]first.east) {\phantom{\tiny{aaa}}};
+\node[myarrow] at ([xshift=8pt]second.east) {\phantom{\tiny{aaa}}};
+\node[myarrow] at ([xshift=8pt]third.east) {\phantom{\tiny{aaa}}};
+
+		\end{tikzpicture}
+
+		{\color{blue}
+		We also included a map of Brazil with the 15 main busy airports and a tour for the reader to get an idea of how close to optimal the solution may be.
+		}
+
 	\end{minipage}

### 2024-01-28 latex/eswa.tex
Main flow description added

@@
 Scenario 6 will be used only for the best method validation.
-
-Figure \ref{fig:flow} depicts the Algorithm \ref{alg:main} for easy understanding of the main flow.
-
 }
@@
 		%% Make the boxes
-		\node [mybox] (first)                        {Select:\\ 1. scenario\\2. surplus\\3. time limit};
-		\node [mybox, right=0.7cm of first] (second) {Config:\\4. parameters\\5. cost matrix\\6. pallets\\7. items};
-		\node [mybox, right=0.7cm of second] (third) {8. Define tour\\     schema:\\- All tours\\ \ \ \ (K-1)!\\ \ \ \ or\\- Solve a TSP\\ \ \ \ (set of tours))};
-		\node [mybox, right=0.7cm of third] (fourth) {Solve ACLPP+RPDP:\\9. Many tours\\ \ \ \ (node-by-node)\\10. Select best tour};
+		\node [mybox] (first)                        { \small {\bf Select:}\\ \small 1. scenario\\ \small 2. surplus\\ \small 3. time limit};
+		\node [mybox, right=0.7cm of first] (second) { \small {\bf Config:}\\ \small 4. parameters\\ \small 5. cost matrix\\ \small 6. pallets\\ \small 7. items};
+		\node [mybox, right=0.7cm of second] (third) { \small 8.
+{\bf Define tour}\\  \small      schema:\\ \small - All tours\\ \ \ \ (K-1)!\\ \ \ \  \small or\\ \small - Solve a TSP\\ \small  \ \ \ (set of tours))};
+		\node [mybox, right=0.7cm of third] (fourth) { \small {\bf Solve ACLPP+RPDP:}\\ \small 9. Many tours\\ \small  \ \ \ (node-by-node)\\ \small 10. Select best tour};
 
@@
 \end{figure}
+
+{\color{blue}
+Figure \ref{fig:flow} depicts the Algorithm \ref{alg:main} for easy understanding of the main flow.
+(1.) Select a scenario from Table \ref{tab:scenarios}.
+(2.) Select a volume surplus (1.2, 1.5, or 2.0).
+(3.) Establish a time limit (240s, 1200s, 2400s, or 3600s).
+(4.) Configure the parameters presented in the mathematical formulation.
+(5.) Load the distance matrix and calculate the matrix cost according to the aircraft characteristics.
+(6.) Load the pallet configuration parameters.
+(7.) Load the item instance (all items to be transported) into memory.
+(8.) Choose a tour schema.
+Until 7 nodes, it is practical to generate (K-1)!
+tours.
+For a larger number of nodes, we recommend using a TSP heuristic that generates a set of similar suboptimal tours.
+(9.) and (10.) solve all tours and select the best.
+}
 

### 2024-01-29 latex/eswa.tex
brazil.png updated

@@
 		\centering
-		\includegraphics[scale=0.17]{nodes.png}
+		\includegraphics[scale=0.21]{nodes.png}
 		\captionof{figure}{A route with 7 airports}
@@
 
-\textbf{Fleet and Route Planning}: For airline operations, the problem isn't just about finding the shortest path through a set of points but also involves scheduling, aircraft assignment, crew scheduling, and maintenance planning.
-This multi-dimensional complexity transforms a standard TSP into a more intricate problem in our study, the Pickup and Delivery Problem (PDP).
+\textbf{Fleet and Route Planning}: For airline operations, the problem isn't just about finding the shortest path through a set of points; it also involves scheduling, aircraft assignment, crew scheduling, maintenance planning, and delivering the right cargo to the right client.
+This last operation task complexity transforms a standard TSP into a more intricate problem in our study, the Pickup and Delivery Problem (PDP).
 
 \textbf{Operational Constraints}: Factors like weather, air traffic control restrictions, and the availability of take-off and landing slots make real-world TSP scenarios in air transport even more challenging.
-These constraints add layers of complexity that significantly depart from the theoretical TSP model.
+Although we do not deal with these constraints, they add layers of complexity that significantly depart from the theoretical TSP model.
 
@@
 In practical air transport scenarios, the size of a TSP can be less about the sheer number of nodes and more about the complexity and dynamic nature of the problem, which in our case (a single aircraft in a single day) is rarely expected to be more than seven nodes.
-Despite this, at the end of this work, we intend to submit the best method found to a more challenging 15-node problem as a validation test.
+Despite this, at the end of this work, we to submit the best method found to a more challenging 15-node problem as a validation test.
 }
@@
 For this reason, a maximum runtime $tmax$\/ per tour is established.
-In the tests carried out in this work, $tmax \leq 3600s$.
+In the tests carried out in this work, {\color{blue} $tmax \leq [240s, 1200s, 2400s, 3600s]$}.
 
@@
 
-	\item {\color{blue} Lately, the time limit to solve a node will be adjusted to be proportional to the sum of volumes of the candidate items of the node. }
+	\item {\color{blue} Lately, the time limit to solve a node will be adjusted to be proportional to the sum of volumes of the candidate items of the node in relation to the sum of volumes of the entire tour.
+}
 
@@
 		\node [mybox, right=0.7cm of first] (second) { \small {\bf Config:}\\ \small 4. parameters\\ \small 5. cost matrix\\ \small 6. pallets\\ \small 7. items};
-		\node [mybox, right=0.7cm of second] (third) { \small 8.
-{\bf Define tour}\\  \small      schema:\\ \small - All tours\\ \ \ \ (K-1)!\\ \ \ \  \small or\\ \small - Solve a TSP\\ \small  \ \ \ (set of tours))};
+		\node [mybox, right=0.7cm of second] (third) { \small {\bf Define tour}\\  \small {\bf schema:}\\ \small 8. All K! tours\\ \ \ \  \small or\\ \small 8. Solve a TSP\\ \small  \ \ \ (set of tours))};
 		\node [mybox, right=0.7cm of third] (fourth) { \small {\bf Solve ACLPP+RPDP:}\\ \small 9. Many tours\\ \small  \ \ \ (node-by-node)\\ \small 10. Select best tour};
@@
 (8.) Choose a tour schema.
-Until 7 nodes, it is practical to generate (K-1)!
+Until 7 nodes, it is practical to generate K!
 tours.

### 2024-01-29 latex/responses.tex
Main flow description added
brazil.png updated

@@
 		{\color{blue}
-		We included the figure below to depict the main algorithm for easy understanding of the main flow.
+		We included the figure below and its description to depict the main algorithm for easy understanding of the main flow.
 		}
@@
 		%% Make the boxes
-\node [mybox] (first)                        {Select:\\ 1. scenario\\2. surplus\\3. time limit};
-\node [mybox, right=0.7cm of first] (second) {Config:\\4. parameters\\5. cost matrix\\6. pallets\\7. items};
-\node [mybox, right=0.7cm of second] (third) {8. Define tour\\     schema:\\- All tours\\ \ \ \ (K-1)!\\ \ \ \ or\\- Solve a TSP\\ \ \ \ (set of tours))};
-\node [mybox, right=0.7cm of third] (fourth) {Solve ACLPP+RPDP:\\9. Many tours\\ \ \ \ (node-by-node)\\10. Select best tour};
+\node [mybox] (first)                        { \small {\bf Select:}\\ \small 1. scenario\\ \small 2. surplus\\ \small 3. time limit};
+\node [mybox, right=0.7cm of first] (second) { \small {\bf Config:}\\ \small 4. parameters\\ \small 5. cost matrix\\ \small 6. pallets\\ \small 7. items};
+\node [mybox, right=0.7cm of second] (third) { \small {\bf Define tour}\\  \small {\bf schema:}\\ \small 8. All K! tours\\ \ \ \  \small or\\ \small 8. Solve a TSP\\ \small  \ \ \ (set of tours))};
+\node [mybox, right=0.7cm of third] (fourth) { \small {\bf Solve ACLPP+RPDP:}\\ \small 9. Many tours\\ \small  \ \ \ (node-by-node)\\ \small 10. Select best tour};
 

### 2024-01-29 latex/eswa.tex
GA and tmax

@@
 
+{\color{violet}
 	\item As the number of nodes is small, we have the possibility to check all possible tours, selecting the one that provides the best value for the objective function.
-For this reason, a maximum runtime $tmax$\/ per tour is established.
-In the tests carried out in this work, {\color{blue} $tmax \leq [240s, 1200s, 2400s, 3600s]$}.
-
-	\item At each node, we will use a MIP solver or heuristics to check which method offers the best result within the runtime limit.
-As there are $K!$ tours, each of them with $K+1$\/ legs, each solution per node will have a runtime limit equal to $tmax/(K+1)!$.
-
-	\item Just for speculation, we will especially test the two tours with the shortest distance, corresponding to the TSP solutions at the nodes considered.
-In these cases, the runtime limit per node will be $tmax/(K+1)$.
-
-	\item {\color{blue} Lately, the time limit to solve a node will be adjusted to be proportional to the sum of volumes of the candidate items of the node in relation to the sum of volumes of the entire tour.
+For this reason, a maximum time per solution ($tmax$) is established.
+In the tests carried out in this work, these times were in the range [240s, 1200s, 2400s, 3600s].
+
+	\item At each node, we will use a MIP solver or heuristics to check which method offers the best result within the time limit for each node.
+
+	\item The time limit to solve a node will be adjusted to be proportional to the sum of volumes of the candidate items of the node in relation to the sum of volumes of the entire tour.
 }
@@
 	\begin{algorithmic}[1]
-			\State {\bf ACLP+RPDP} \,\, {\it in:}~$scenario, surplus, tmax$ \, \, {\it out:}~$answer1, answer2$
+			\State {\bf ACLP+RPDP} \,\, {\it in:}~$scenario, surplus, tmax, tsp, ntours$ \, \, {\it out:}~$answer$
 
@@
 
-			\State Let $\pi_{TSP1}$\/ and $\pi_{TSP2}$ be the shortest tours ({\it cfr.} Table \ref{tab:scenarios})  \label{main:shortest}
-
 			\State $N \gets ItemsGeneration(scenario, surplus)$ \label{main:items}
 
+		{\color{violet}
+			\If {$tsp$ = True}
+
+				\State $tmax \gets tmax/ntours$ \Comment{max. tour time to solve the number of tours from the TSP heuristic}
+
+			\Else
+
+				\State $tmax \gets tmax/K!$
+
+			\EndIf
+		}
 			\For {each $method$}  \label{main:method}
 
-				\State $f_1 \gets SolveTour(\pi_{TSP1}, L, M, C, N, method, tmax)$  \label{main:f1}
-
-				\State $f_2 \gets SolveTour(\pi_{TSP2}, L, M, C, N, method, tmax)$  \label{main:f2}
-
-				\State $answer1[scenario,surplus,method] \gets max( f_1, f_2  )$  \label{main:a1}
-
 				\For {each $\pi \in S_K$} \label{main:loop1}
 
-					\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, tmax/K!)$ \label{main:fpi}
+					\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, tmax)$ \label{main:fpi}
 
@@
 
-				\State $answer2[scenario,surplus,method] \gets \max f_{\pi}$ \label{main:a2}
+				\State $answer[scenario,surplus,method] \gets \max f_{\pi}$ \label{main:a2}
 			\EndFor
@@
 	\caption{Testing scenarios}  \label{tab:scenarios}
-	\begin{tabular}{c c c c c}
-		\toprule
-		{\bf Scenario} & $K$ & $L$ & $\pi_{TSP1}$ & $\pi_{TSP2}$ \\		
-		\midrule
-		1 & 2    & \{$0,1,2$\}           & 0 1 2 0  & 0 2 1 0 \\
-		2 & 3    & \{$0,1,2,3$\}         & 0 1 2 3 0 & 0 3 2 1 0\\
-		3 & 4    & \{$0,1,2,3,4$\}       & 0 4 1 2 3 0 & 0 3 2 1 4 0 \\
-		4 & 5    & \{$0,1,2,3,4,5$\}     & 0 4 1 2 5 3 0 & 0 3 5 2 1 4 0 \\
-		5 & 6    & \{$0,1,2,3,4,5,6$\}   & 0 4 1 2 6 5 3 0 & 0 3 5 6 2 1 4 0 \\
-		6 & 15   & \{$0,1,2,3,4,5,6,7,8,9,10,11,12,13,14$\}   & not known & not known \\
+	\begin{tabular}{c c c }
+		\toprule
+		{\bf Scenario} & $K$ & $L$  \\		
+		\midrule
+		1 & 2    & \{$0,1,2$\}            \\
+		2 & 3    & \{$0,1,2,3$\}         \\
+		3 & 4    & \{$0,1,2,3,4$\}        \\
+		4 & 5    & \{$0,1,2,3,4,5$\}      \\
+		5 & 6    & \{$0,1,2,3,4,5,6$\}    \\
+		6 & 15   & \{$0,1,2,3,4,5,6,7,8,9,10,11,12,13,14$\}    \\
 		\bottomrule
@@
 
-The best results corresponding to the shortest tours are stored in $answer1$\/ (line \ref{main:a1}), and those obtained by testing all $K!$\/ tours are stored in $answer2$\/ (line \ref{main:a2}).
+The best results obtained by testing all tours are stored in $answer$\/ (line \ref{main:a2}).
 
@@
 
-$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $tmax$\/ is the runtime limit that will be distributed among the $K+1$\/ legs of the tour.
+$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $tmax$\/ is the {\color{blue}runtime limit of the tour.}
 
@@
 
+		{\color{violet}
+
+		\State $V_{\pi}  \gets 0$ \label{tour:Vol} \Comment{tour volume}
+
+		\For {$k \gets 0$ to $K$} \label{tour:loop0}
+
+			\State $v_{\pi_k}  \gets \sum_{j=1}^{n_k} v_j$
+			\State $V_{k} \gets V_{k} + v_{\pi_k}$
+
+		\EndFor
+
+	}
+
 		\For {$k \gets 0$ to $K$} \label{tour:loop1}
+
+		\State $tnode_{\pi_k} = (v_{\pi_k}/V_{\pi}) * tmax$ \Comment{{\color{blue} node time limit calculation}}
+
 		\State $L_{\pi_k} \gets L - \{\pi_0,\pi_1,\ldots,\pi_k\}$  \label{tour:lk1}	 \Comment{the set of remaining nodes is updated}	
@@
 
-		\State $G_2 \gets SolveNode(method, \pi_k, G_1, tmax/(K+1))$  \label{tour:SolveNode}
+		\State $G_2 \gets SolveNode(method, \pi_k, G_1, tnode_{\pi_k})$  \label{tour:SolveNode}
 
@@
 One of the main objectives of this work was to find a quick heuristic that offers a good-quality solution for the node-by-node problem.
-Taking this into account, we design algorithms based on four known meta-heuristics: {\it Ant Colony Optimization} (ACO) \citep{Dorigo1992,DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \citep{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \citep{Glover1986}, and {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \citep{FeoResende1989}.
-We considered several ideas from the literature \citep{NiarFreville1997,Fidanova2006,Alonso2019, Zhan2020}, and we were careful to use the same data structures and procedures in all implementations to enforce fair results comparison.
+Taking this into account, we design algorithms based on four known meta-heuristics: {\it Ant Colony Optimization} (ACO) \citep{Dorigo1992,DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \citep{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \citep{Glover1986}, {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \citep{FeoResende1989}, and the Genetic Algorithm (GA) \citep{holland1992adaptation}.
+We considered several ideas from the literature \citep{NiarFreville1997, Fidanova2006, Alonso2019, Zhan2020, shah2020genetic}, and we were careful to use the same data structures and procedures in all implementations to enforce fair results comparison.
 
@@
 
-{\color{blue}The arguments $threshold_1$ and $threshold_2$ are the volume thresholds for $\eta_1$ and $\eta_2$ calculations, respectively.
-
-The argument $tmax$ is the maximum time for Shims to solve a node.
-It is proportional to the sum of the candidate items volumes in each node, in relation to the sum of all nodes $tmax$, and the maximum time to solve the tour.
+{\color{violet}
+
+	The arguments $threshold_1$ and $threshold_2$ are the volume thresholds for $\eta_1$ and $\eta_2$ calculations, respectively.
+
+The argument $tmax$ is the maximum time for {\it Shims} to solve a node.
+It is proportional to the sum of the candidate items volumes in each node, in relation to the sum of all nodes volumes, and the maximum time to solve the tour.
 
@@
 
-We ran Algorithm \ref{alg:main} considering the five scenarios from Table \ref{tab:scenarios}, three values for  $surplus$\/ from $\{1.2, 1.5, 2.0\}$, four values for $tmax$\/ from $\{240s, 1200s, 2400s, 3600s\}$, and six different $method$\/ for the node-by-node solution: $Gurobi$\/ (see \ref{solver}), ACO, NMO, TS, GRASP, and $Shims$\/ (Algorithm \ref{alg:shims}).
+We ran Algorithm \ref{alg:main} considering the five scenarios from Table \ref{tab:scenarios}, three values for  $surplus$\/ from $\{1.2, 1.5, 2.0\}$, four values for $tmax$\/ from $\{240s, 1200s, 2400s, 3600s\}$, and seven different methods for the node-by-node solution: $Gurobi$\/ (see \ref{solver}), ACO, NMO, TS, GRASP, GA and $Shims$\/ (Algorithm \ref{alg:shims}).
 
@@
 	TS            & -             & 5    & {\color{red}44}  \\
+	GA            & -             & 1,2,3,4,5  & {\color{red}did not solve}  \\
 	\midrule
@@
 \end{table}
+
+{\color{violet}
+
+As to the genetic algorithm exclusion from the comparisons tests, we must consider that:
+\vspace{-2mm}
+
+\begin{itemize}
+	\setlength\itemsep{-0.3em}	
+	\item The palletization planning of 200 to 1000 items per node in a set of 18 aircraft pallets may be an inadequate problem to be solved with GA, because large chromosome sizes, like 18,000-dimension arrays, can pose significant challenges for each generation to evaluate individual fitness, which demonstrated to be resource-intensive.
+	\item Large chromosomes can slow GA convergence due to the vast search space, requiring more generations to find optimal solutions.
+	\item Efficient encoding and decoding of the problem into the chromosome and decoding solutions is crucial for a 18,000-dimension array. 
+	\item Also, the design of the fitness function becomes more critical as chromosome size increases, as it needs to guide the evolutionary process without being overly computationally demanding.
+\end{itemize}
+
+}
 

### 2024-01-30 latex/eswa.tex
eswa updated and sent to Carlos

@@
 \captionsetup{labelfont=bf}
-\captionsetup{skip=0pt}
+\captionsetup{skip=2pt}
 
@@
 \usepackage{xcolor}
-\definecolor{myblue}{HTML}{B0D7FF}
+\definecolor{myblue}{HTML}{D6EAF8} % A9CCE3  B0D7FF
 
@@
 	myarrow/.style={draw=black,
+		fill=white,
+		minimum width=0.6cm,
+		single arrow
+	},
+	downarrow/.style={draw=black,
 	fill=white,
@@
 	\centering
+	\fbox{
 	\begin{tikzpicture}
@@
 		%% Make the boxes
-		\node [mybox] (first)                        { \small {\bf Select:}\\ \small 1. scenario\\ \small 2. surplus\\ \small 3. time limit};
-		\node [mybox, right=0.7cm of first] (second) { \small {\bf Config:}\\ \small 4. parameters\\ \small 5. cost matrix\\ \small 6. pallets\\ \small 7. items};
-		\node [mybox, right=0.7cm of second] (third) { \small {\bf Define tour}\\  \small {\bf schema:}\\ \small 8. All K! tours\\ \ \ \  \small or\\ \small 8. Solve a TSP\\ \small  \ \ \ (set of tours))};
-		\node [mybox, right=0.7cm of third] (fourth) { \small {\bf Solve ACLPP+RPDP:}\\ \small 9. Many tours\\ \small  \ \ \ (node-by-node)\\ \small 10. Select best tour};
+		\node [mybox] (first)                        { \small {\bf Select:}\\ 
+			\small 1. scenario\\ 
+			\small 2. surplus\\ 
+			\small 3. time limit};
+
+		\node [mybox, right=0.7cm of first] (second) { \small {\bf Config:}\\ 
+			\small 4. parameters\\ \small 5. cost matrix\\ 
+			\small 6. pallets\\ 
+			\small 7. items};
+
+		\node [mybox, right=0.7cm of second] (third) { \small {\bf Define tour}\\  
+			\small {\bf schema:}\\ 
+			\small 8. All K! tours\\ \ \ \  
+			\small or\\ 
+			\small 8. Solve a TSP\\ 
+			\small  \ \ \ (set of tours))};
+
+		\node [mybox, right=0.7cm of third] (fourth) { \small {\bf Solve ACLPP+RPDP:}\\ 
+			\small 9. Many tours{\bf *}\\ 
+			\small 10. Select best tour};
+
+		\node [mybox, below=0.7cm of third] (fifth) { \small {\bf *Solve a tour:}\\
+			\small \ \ \ {\it - For each node:}\\
+			\small \ \ \ \ \ {a. Unload some pallets}\\
+			\small \ \ \ \ \ {b. Arrange remaining pallets to balance cargo}\\
+			\small \ \ \ \ \ {c. Set empty pallets destinations}\\
+			\small \ \ \ \ \ {d. \it For each pallet:}\\
+			\small \ \ \ \ \ \ \ \ \ {Solve an ACLPP with WBP and PDP constraints}\\
+			\small \ \ \ \ \ {e. Calculate node score and torque}\\
+			\small \ \ \ {\it - Return tour Score/Cost}
+
+	};
 
@@
 	\end{tikzpicture} 
+}
 	\caption{Solution main flow}
@@
 
+{\color{violet}
+
 One of the main objectives of this work was to find a quick heuristic that offers a good-quality solution for the node-by-node problem.
@@
 We considered several ideas from the literature \citep{NiarFreville1997, Fidanova2006, Alonso2019, Zhan2020, shah2020genetic}, and we were careful to use the same data structures and procedures in all implementations to enforce fair results comparison.
+
+As \citep{PeerlinckAmy2022MFEO} applied the DEAP (Distributed Evolutionary Algorithms in Python) to solve a Multi-Objective Knapsack Problem, we also decided to apply this evolutionary package to implement a GA solution for the ACLPP.
+
+}
 
@@
 In $\eta_2$\/ index, this same accumulated volume reaches $V_i \times  {\color{blue}threshold_2}$.
-{\color{blue}The values of these thresholds were defined empirically by the {\it iRace}\/ tool \citep{LopezIbanezManuel2016}}.
+{\color{blue}The values of these thresholds were defined empirically by the {\it irace}\/ tool \citep{LopezIbanezManuel2016}}.
 
@@
 
-{\color{blue}
-	iRace (Iterated Racing for Automatic Algorithm Configuration) streamlines the tuning of optimization algorithms by iteratively testing and eliminating weaker parameter configurations through statistical comparisons.
-It focuses computational resources on the most promising settings.
-This method significantly cuts down the time and effort needed to find optimal or near-optimal parameters, proving effective across various application domains.
+{\color{violet}
+	{\it irace} (Iterated Racing for Automatic Algorithm Configuration) streamlines the tuning of optimization algorithms by iteratively testing and eliminating weaker parameter configurations through statistical comparisons.
+Its application may be understood, from a Machine Learning standpoint, as the task of identifying optimal parameter values for solving new problem instances by leveraging knowledge gained from a set of training problem examples.
+This method significantly cuts down on the time and effort needed to find optimal or near-optimal parameters, proving effective across various application domains.
 }
@@
 	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$, where $i_1$ and $i_2$ are integer numbers;
-	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
+{\color{violet}
+\item $Roulette()$: a proportional selection function that is biased through $P$ (Table \ref{tab:weights}).
+}
 \end{itemize}
@@
 {\color{blue}
-	After the items were generated, some experiments with the \textit{iRace} tool were executed to determine the values for parameters $threshold_1$ and $threshold_2$ used in \textit{Shims}, whose results are presented in Table \ref{tab:irace}.
+	After the items were generated, some experiments with the \textit{irace} tool were executed to determine the values for parameters $threshold_1$ and $threshold_2$ used in \textit{Shims}, whose results are presented in Table \ref{tab:irace}.
 
@@
 	\centering
-	\caption{\textit{iRace} testing results}  \label{tab:irace}
+	\caption{\textit{irace} testing results}  \label{tab:irace}
 	\footnotesize
@@
 
-	For these tests, we considered the odd instances (1, 3, 5, 7) as the \textbf{training} set and the even instances (2, 4, 6) as the \textbf{testing} set for the \textit{iRace} runs.
-We supplied \textit{iRace} for $threshold_1$ determination, the range [0.8, 1.0], and for $threshold_2$, the range [1.0, 2.0].
-The number of iterations in each experiment was limited to 3000 executions for \textit{iRace} to have plenty of data for its statistical tests.
-More detail about these configurations may be found on the \textit{iRace} user guide in {\tt cran.r-project.org/web/packages/irace/irace.pdf}.
+	For these tests, we considered the odd instances (1, 3, 5, 7) as the \textbf{training} set and the even instances (2, 4, 6) as the \textbf{testing} set for the \textit{irace} runs.
+We supplied \textit{irace} for $threshold_1$ determination, the range [0.8, 1.0], and for $threshold_2$, the range [1.0, 2.0].
+The number of iterations in each experiment was limited to 3000 executions for \textit{irace} to have plenty of data for its statistical tests.
+More detail about these configurations may be found on the \textit{irace} user guide in {\tt cran.r-project.org/web/packages/irace/irace.pdf}.
 }
@@
 
-\citep{PeerlinckAmy2022MFEO} applied DEAP to solve a Multi-Objective Knapsack Problem. This algorithm is adapted to solve the ACLPP in each node of this work's solution.
-
 	\begin{figure}[H]
@@
 
+{\color{blue}
 We adopted some simplifications that are not critical, but that allowed for an unprecedented solution to this problem considering more than two nodes.
@@
 Considering a real aircraft, we have developed some node-by-node solutions.
-This allows us to test two strategies: find a solution using the shortest tours, or enumerate all $K!$\/ tours and choose the best.
-The complete process can be executed quickly on a handheld computer, offering good results and reducing stress for the transport planners.
+The complete process can be executed quickly on a handheld computer, offering good results and reducing stress for the transport planners.}
 

### 2024-01-31 latex/eswa.tex
Main flow updated

@@
 	myarrow/.style={draw=black,
-		fill=white,
-		minimum width=0.6cm,
-		single arrow
-	},
-	downarrow/.style={draw=black,
 		fill=white,
@@
 
+		\path[->] (third) edge (fifth);
+
 	\end{tikzpicture} 
@@
 	\label{fig:flow}
+\end{figure}
+
+\begin{figure}[htp]
+	\centering
+	\fbox{
+		\begin{tikzpicture}
+			\node [mybox] (one) { \small {\bf Input data:}\\
+				\small \ \ \ {Aircraft parameters}\\
+				\small \ \ \ {Airports and distances}\\
+				\small \ \ \ {$scenario$ \ \ ($K$ and $L$)}\\
+				\small \ \ \ {$surplus = \sum_{j=1}^{n_k} v_j$/$\sum_{i=1}^{m} V_i$ }\\
+				\small \ \ \ {$tmax$ \ \ (overall runtime limit)}\\
+				\small {{\bf Requirement:} $K < m$}
+			};
+
+			\node [mybox, below=0.7cm of one] (two) { \small {\bf Tours definition:}\\
+				\small \ \ \ {{\bf if} $K \leq 6$}\\
+				\small \ \ \ \ \ \ {$ntours = 2$ \ \ (optimal TSP solutions)}\\
+				\small \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\it or}\\				
+				\small \ \ \ \ \ \ {$ntours = K!$ \ \ (all possible tours)}\\
+				\small \ \ \ {\bf else}\\
+				\small \ \ \ \ \ \ {$ntours = 100$ \ \ (best tours obtained with TSP heuristic)}
+			};
+
+			\node [mybox, below=0.7cm of two] (three) { \small {\bf On each tour of the $ntours$:}\\
+				\small \ \ \ {Calculation of a node-by-node solution (divide runtime according to shipment volumes)}\\
+				\small \ \ \ {On each node of current tour:}\\
+				\small \ \ \ \ \ \ \ {- Relocate packed contents with torque optimization}\\
+				\small \ \ \ \ \ \ \ {- Predefine pallet destinations according to shipment volumes}\\
+				\small \ \ \ \ \ \ \ {- Find a node solution (MIP, Shims, NMO, ACO, GRASP, TS, GA) }\\
+				\small \ \ \ \ \ \ \ {- Accumulate values obtained from score, torque and cost}
+			};
+
+		\node[myarrow, rotate=-90] at ([yshift=-8pt]two.south) {\phantom{\tiny{aaa}}};
+		\node[myarrow, rotate=-90] at ([yshift=-8pt]one.south) {\phantom{\tiny{aaa}}};		
+		\end{tikzpicture} 
+	}
+	\caption{Solution process}
+	\label{fig:process}
 \end{figure}

### 2024-01-31 latex/eswa.tex
Article update

@@
 The IOM accelerates the solving process compared to the COM, but some scenarios still require significant computation time.
+
+}
+
+{\color{violet}
+
+Inefficient air transport plans can lead to unnecessary costs, extra routes, longer distances, and incorrect destinations.
+Unbalanced cargo increases fuel consumption due to altered aircraft pitch angles.
+
+A study by \cite[p.
+22]{gwh2007} highlights the risk of weight- and balance-related accidents on cargo flights compared to passenger flights.
+So, balancing cargo is crucial for safe aerial transportation, as an improperly positioned centre of gravity can result in dangerous take-off and landing conditions and stall recovery issues.
+
+Another important issue is managing fuel consumption.
+To get an idea of the influence of CG displacement on fuel consumption, \cite[p.
+140]{MongeauBes2003} report that {\it "a displacement of the CG of less than 75 cm in a long-range aircraft yields, over a 10,000 km flight, a saving of 4,000 kg of fuel"}.
+This reinforces the importance of balancing cargo.
+
+Despite technological advancements, many airlines still rely on manual aircraft loading and balancing, using simple computer-aided tools for CG computation and constraint testing.
+Loadmasters often create load plans manually, which is time-consuming and can lead to flight delays.
+
+As to the right delivery, an organisation's purchasing department may acquire supplies in the main capitals of the country and abroad to meet the company's needs, and it wants these materials to be delivered to the requesting organisations within the stipulated deadlines.
+
+A rational decision-making process is crucial to avoid creating inefficient or unsafe transport plans, considering the high costs of fuel, maintenance, operation, outsourcing expenses, potential operational impairments, and safety risks due to unbalanced cargo.
+
+Solving the ACLPP+RPDP is vital for optimising strategic scores of what must be transported, saving time and effort in loading, ensuring safety and balance, ensuring correct pickups and deliveries, and finding the best routes considering fuel use due to potential cargo imbalance.
 

### 2024-02-01 ga_deap_mmkp.py
New experiments with the shortest tours

@@ -17,8 +17,8 @@ def Solve(items_file, node_ID):
     # filter items on the base
     items_matrix_np = items_matrix_np[((items_matrix_np[:,3] == node_ID))]
 
-    # num_items_test = 250
-    # items_matrix_np = items_matrix_np[:num_items_test,:]
+    num_items_test = 250
+    items_matrix_np = items_matrix_np[:num_items_test,:]
 
     #  w    s    v  from to
     items_weights = items_matrix_np[ :,0]

### 2024-02-02 latex/eswa.tex
New experiments with the shortest tours
Tables  12, 13 and 14 updated
Plots

@@
 
-\begin{table}[H]
+\begin{table}[htp] % update this
 	\centering
@@
 		\toprule
-		\multirow{4}{*}{Shortest} &\multirow{2}{*}{$Gurobi$}  & $f$ & 8.61    & 11.56   & 13.67   & 13.87 & 12.48       &  0.9985 \\ % best 6028
-		&                              &               {time (s)}   & 26      & 22      & 28      & 28    & 25          &  1.0    \\ % longer 129
+		\multirow{4}{*}{Shortest} &\multirow{2}{*}{$Gurobi$}  & $f$ & 8.53    & 11.79   & 13.14   & 13.52 & 12.36       &  0.9998 \\ % best 5935
+		&                              &               {time (s)}   & 29      & 28      & 25      & 27    & 27          &  1.0    \\ % longer 136
 		\cmidrule(lr){2-9}
-		&                           \multirow{2}{*}{$Shims$}  & $f$ & 8.61    & 11.56   & 13.46   & 13.93 & 12.51       &  0.9965 \\
-		&                             &                {time (s)}   & 1       & 1       & 1       & 1     & 1           &  25.8   \\
+		&                           \multirow{2}{*}{$Shims$}  & $f$ & 8.54    & 11.78   & 13.06   & 13.51 & 12.34       &  0.9980 \\
+		&                             &                {time (s)}   & 1       & 1       & 1       & 1     & 2           &  22.7   \\
 
@@
 		\toprule
-		\multirow{4}{*}{ Shortest } & \multirow{2}{*}{$Gurobi$} & $f$ & 11.76 & 16.29 & 18.02 & 19.42 & 16.35 & 0.9994 \\ % best 8189
-		&&               {time (s)}                                   & 51    & 39    & 40    & 36    & 38    & 1.0 \\ 
+		\multirow{4}{*}{ Shortest } & \multirow{2}{*}{$Gurobi$} & $f$ & 11.83 & 16.73 & 18.07 & 18.83 & 16.86 & 0.9996 \\ % best 8235
+		&&               {time (s)}                                   & 55    & 64    & 39    & 40    & 88    & 1.0    \\ 
 		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{$Shims$}                               & $f$ & 11.78 & 16.29 & 17.96 & 19.43 & 16.37 & 0.9993  \\
-		&&         {time (s)}                                         & 1     & 1     & 2     & 2     & 2     & 25.5  \\
-
-		\midrule
-
-		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$}   & $f$ & 11.76 & 16.37 & 18.01 & 20.95 & 17.60 & 0.9999 \\  % best 8469
-		&&               {time (s)}                                   & 64    & 58    & 195   & 472   & 2258  & 1.0  \\
+		&\multirow{2}{*}{$Shims$}                               & $f$ & 11.85 & 16.72 & 18.05 & 18.80 & 16.87 & 0.9993  \\
+		&&         {time (s)}                                         & 1     & 1     & 2     & 2     & 2     & 35.8  \\
+
+		\midrule
+
+		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$}   & $f$ & 11.83 & 16.93 & 18.40 & 20.95 & 17.60 & 0.9999 \\  % best 8469
+		&&               {time (s)}                                   & 63    & 59    & 195   & 472   & 2258  & 1.0  \\
 		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{$Shims$}                               & $f$ & 11.78 & 16.29 & 17.99 & 20.93 & 17.50 & 0.9976  \\ 
+		&\multirow{2}{*}{$Shims$}                               & $f$ & 11.85 & 16.91 & 18.36 & 20.93 & 17.50 & 0.9976  \\ 
 		&&         {time (s)}                                         & 1     & 2     & 5     & 15    & 100   & 23.8 \\
@@
 		\toprule
-		\multirow{4}{*}{Shortest} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.89 & 24.96 & 26.15 & 27.52 & 23.85 & 0.9996 \\ % best 12042
-		&&                                              {time (s)}  & 183   & 87    & 73    & 79    & 71    & 1.0  \\
+		\multirow{4}{*}{Shortest} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.70 & 24.20 & 26.39 & 27.17 & 24.20 & 0.9995 \\ % best 11972
+		&&                                              {time (s)}  & 168   & 98    & 79    & 70    & 72    & 1.0  \\
 		\cmidrule(lr){2-9}	
-		&                            \multirow{2}{*}{$Shims$} & $f$ & 17.94 & 24.96 & 26.03 & 27.43 & 23.73 & 0.9972  \\
-		&&                                              {time (s)}  & 1     & 2     & 2     & 3     & 4     & 41.1  \\
-
-		\midrule
-
-		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.90 & 25.44 & 26.15 & 29.13 &{\bf x}& 0.9994 \\ % best 9867
+		&                            \multirow{2}{*}{$Shims$} & $f$ & 17.74 & 24.22 & 26.32 & 27.07 & 23.13 & 0.9896  \\
+		&&                                              {time (s)}  & 1     & 2     & 2     & 3     & 4     & 40.6  \\
+
+		\midrule
+
+		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.90 & 25.44 & 26.51 & 29.13 &{\bf x}& 0.9994 \\ % best 9867
 		&&                                               {time (s)} & 178   & 143   & 378   & 862   &{\bf x}& 1.0  \\
 		\cmidrule(lr){2-9}	
-		&                         \multirow{2}{*}{$Shims$}    & $f$ & 17.94 & 25.45 & 26.14 & 28.84 & 26.22  & 0.9970  \\
+		&                         \multirow{2}{*}{$Shims$}    & $f$ & 17.94 & 25.45 & 26.44 & 28.84 & 26.22  & 0.9970  \\
 		&&                                               {time (s)} & 1     & 3     & 10    & 31    & 196    & 34.7  \\
@@
 
+\begin{table}[htp]
+	\centering
+	\caption{Shims solutions with $surplus = 1.2$\/ and $tmax =  240s$}  \label{tab:15}
+	\footnotesize
+	\begin{tabular}{cccc}
+		\toprule
+$scenario$ & {\bf 7} & {\bf 8} & {\bf 9} \\
+Num. nodes & 15 & 14 & 13 \\
+		\toprule
+$f$        & 50.15 & 47.00 & 47.18    \\
+{time (s)} & 32     & 31     & 29        \\
+Occupancy rate & 0.89     & 0.84     & 0.88        \\
+		\bottomrule
+	\end{tabular}
+	\normalsize
+\end{table}
+
+\begin{center}
+	%\fbox{
+	\begin{tikzpicture}
+		\begin{axis}[xmin=6,ymin=200,
+			enlarge y limits={upper,value=0.2},
+			xlabel style={align=center},
+			xlabel={Number of nodes},
+			ylabel={Time (s)},
+			xtick={  5,   6, 7,  8,  9, 10, 11, 12, 13, 14, 15, 16},
+			ytick={250,350,450,550,650,750,850,950,1050,1150,1250},						
+			]
+			\addplot[
+			blue,
+			thick,
+			mark=*,
+			mark options={fill=blue},
+			visualization depends on=\thisrow{alignment} \as \alignment,
+			nodes near coords, % Place nodes near each coordinate
+			point meta=explicit symbolic, % The meta data used in the nodes is not explicitly provided and not numeric
+			every node near coord/.style={anchor=\alignment} % Align each coordinate at the anchor 40 degrees clockwise from the right edge
+			] table [% Provide data as a table
+			meta index=2 % the meta data is found in the third column
+			] {
+				x   y    label                  alignment
+				7   273   {\footnotesize 92.6}  -10
+				8   360   {\footnotesize 107.5} -10
+				9   468   {\footnotesize 107.5} -10
+				10  560   {\footnotesize 112.3} -10
+				11  660   {\footnotesize 92.8}  -10
+				12  756   {\footnotesize 91.2}  -10
+				13  884   {\footnotesize 96.8}  -10
+				14  1008   {\footnotesize 106.3} -10									
+				15  1170   {\footnotesize 99.8}  -10									
+			};
+		\end{axis}
+	\end{tikzpicture}
+	%}
+	\captionof{figure}{Shims performance}
+\end{center}
+
+\begin{center}
+	%\fbox{
+		\begin{tikzpicture}
+			\begin{axis}[xmin=6,ymin=600,
+				enlarge y limits={upper,value=0.2},
+				xlabel style={align=center},
+				xlabel={Number of nodes},
+				ylabel={Time (s)},
+				xtick={  5,  6,   7,   8,   9,  10,  11,  12,  13,  14,  15,  16},
+				ytick={700,860,1020,1180,1340,1500,1660,1820,1980,2140,2300,2460},						
+				]
+				\addplot[
+				blue,
+				thick,
+				mark=*,
+				mark options={fill=blue},
+				visualization depends on=\thisrow{alignment} \as \alignment,
+				nodes near coords, % Place nodes near each coordinate
+				point meta=explicit symbolic, % The meta data used in the nodes is not explicitly provided and not numeric
+				every node near coord/.style={anchor=\alignment} % Align each coordinate at the anchor 40 degrees clockwise from the right edge
+				] table [% Provide data as a table
+				meta index=2 % the meta data is found in the third column
+				] {
+					x   y    label                 alignment
+					7   708  {\footnotesize 91.9}  -10
+					8   837  {\footnotesize 108.2} -10
+					9   987  {\footnotesize 108.7} -10
+					10  1160 {\footnotesize 111.2} -10
+					11  1354 {\footnotesize 78.0}  -10
+					12  1574 {\footnotesize 84.1}  -10
+					13  1814 {\footnotesize 93.3}  -10
+					14  2083 {\footnotesize 104.8} -10									
+					15  2383 {\footnotesize 96.1}  -10									
+				};
+			\end{axis}
+		\end{tikzpicture}
+		%}
+	\captionof{figure}{Shims performance (total time)}
+\end{center}
+
 \section{Conclusions}

### 2024-02-02 gen_items.py

@@ -104,10 +104,10 @@ def genItems(sce, inst, surplus):
 if __name__ == "__main__":
 
     # scenarios = [1,2,3,4,5,6]
-    scenarios = [7] # 15 nodes for Shims validation
+    scenarios = [7,8,9,10,11,12,13,14]
 
-    instances = [1,2,3,4,5,6,7]
-    #instances = [1]
+    instances = [2,3,4,5,6,7]
+    # instances = [1]
 
     # surplus = 1.2
     # surplus = 1.5

### 2024-02-02 latex/eswa.tex
New time computation

@@
 		\begin{tikzpicture}
-			\begin{axis}[xmin=6,ymin=600,
+			\begin{axis}[xmin=6,ymin=80,
 				enlarge y limits={upper,value=0.2},
@@
 				xtick={  5,  6,   7,   8,   9,  10,  11,  12,  13,  14,  15,  16},
-				ytick={700,860,1020,1180,1340,1500,1660,1820,1980,2140,2300,2460},						
+				ytick={80,100,120,140,160,180,200,220,240,260,280,300},						
 				]
@@
 					x   y    label                 alignment
-					7   708  {\footnotesize 91.9}  -10
-					8   837  {\footnotesize 108.2} -10
-					9   987  {\footnotesize 108.7} -10
-					10  1160 {\footnotesize 111.2} -10
-					11  1354 {\footnotesize 78.0}  -10
-					12  1574 {\footnotesize 84.1}  -10
-					13  1814 {\footnotesize 93.3}  -10
-					14  2083 {\footnotesize 104.8} -10									
-					15  2383 {\footnotesize 96.1}  -10									
+					7   99  {\footnotesize 91.9}  -10
+					8   126  {\footnotesize 108.2} -10
+					9   145  {\footnotesize 108.7} -10
+					10  168 {\footnotesize 111.2} -10
+					11  191 {\footnotesize 78.0}  -10
+					12  215 {\footnotesize 84.1}  -10
+					13  239 {\footnotesize 93.3}  -10
+					14  264 {\footnotesize 104.8} -10									
+					15  293 {\footnotesize 96.1}  -10									
 				};

### 2024-02-02 main.py

@@ -340,23 +340,24 @@ def writeResults(method, scenario, folder, fvalue, elapsed):
 
 if __name__ == "__main__":
 
-    startTime = time.perf_counter()
+    mainStart = time.perf_counter()
 
     plot      = False
+
     testing   = False
+    # testing   = True
+
     # shortest = True # 2 shortest tours
     shortest = False # All K!
     iRace_testing = False
 
     # scenarios = [2,3,4,5,6] # represent 1,2,3,4,5 in the article
-    # scenarios = [6] # 7: 15-node problem
+    # scenarios = [3]
     # scenarios = [7,8,9]
     scenarios = [6,7,8,9,10,11,12,13,14]
 
     # folder = "surplus20"  # 1.2
-    
     # folder = "surplus50"  # 1.5
-
     folder = "surplus100" # 2.0
 
     iRace_scenario = 2
@@ -407,13 +408,15 @@ if __name__ == "__main__":
     tipo = "FFD"
 
     distances_file = "./params/distances7.txt"
-    if scenarios[0] == 6:
+    if scenarios[0] >= 6:
         distances_file = "./params/distances15.txt"
 
     dists, _ = common.loadDistances(distances_file) # dists, cities
     costs = [[0.0 for _ in dists] for _ in dists]
 
     for scenario in scenarios:
+
+        instances = [1,2,3,4,5,6,7]
 
         if testing:
             instances = [1]
@@ -423,13 +426,7 @@ if __name__ == "__main__":
 
         cfg = common.Config(scenario)
 
-        instances = [1,2,3,4,5,6,7]
-        # instances = [1]
-
         print(f"\n{cfg.numNodes} nodes")
-
-        if cfg.numNodes > 7: # Shims validation
-            instances = [1]
 
         for i, cols in enumerate(dists):
             for j, dist in enumerate(cols):
@@ -458,39 +455,38 @@ if __name__ == "__main__":
         if shortest: # the 2 shortest tours
             tourTime = timeLimit/2
 
-        instanceTime = 0.
-        instanceTime2 = 0. # with 3D packing
-        instanceSC   = 0. # score/cost relation
-        leastSC      = 0.
-        worstTime    = 0
-
         numOptDict = {"numOpt":0}
         afterDict  = {"value":0.}
         beforeDict = {"value":0.}
 
+        numInst = float(len(instances))
 
-        for inst in instances:  
+        avgInstTime       = 0
+        instBestAvgVol    = 0.
[…]

### 2024-02-02 latex/eswa.tex
New experiments

@@
 		%}
+	\captionof{figure}{Shims performance (new time computation)}
+\end{center}
+
+	\begin{center}
+		%\fbox{
+			\begin{tikzpicture}
+				\begin{axis}[xmin=6,ymin=600,
+					enlarge y limits={upper,value=0.2},
+					xlabel style={align=center},
+					xlabel={Number of nodes},
+					ylabel={Time (s)},
+					xtick={  5,  6,   7,   8,   9,  10,  11,  12,  13,  14,  15,  16},
+					ytick={600,1850,3100,4350,5600,6850,8100,9350,10600,11850,13100,14350},						
+					]
+					\addplot[
+					blue,
+					thick,
+					mark=*,
+					mark options={fill=blue},
+					visualization depends on=\thisrow{alignment} \as \alignment,
+					nodes near coords, % Place nodes near each coordinate
+					point meta=explicit symbolic, % The meta data used in the nodes is not explicitly provided and not numeric
+					every node near coord/.style={anchor=\alignment} % Align each coordinate at the anchor 40 degrees clockwise from the right edge
+					] table [% Provide data as a table
+					meta index=2 % the meta data is found in the third column
+					] {
+						x   y    label                 alignment
+						7   692  {\footnotesize 91.9}  -10
+						8   1570  {\footnotesize 108.2} -10
+						9   2580  {\footnotesize 108.7} -10
+						10  3754 {\footnotesize 111.2} -10
+						11  5086 {\footnotesize 78.0}  -10
+						12  6589 {\footnotesize 84.1}  -10
+						13  8528 {\footnotesize 93.3}  -10
+						14  10101 {\footnotesize 104.8} -10									
+						15  12150 {\footnotesize 96.1}  -10									
+					};
+				\end{axis}
+			\end{tikzpicture}
+			%}
 		\captionof{figure}{Shims performance (total time)}

### 2024-02-04 latex/ESWA.tex
Time complexity

--- 2024-02-02 latex/eswa.tex
@@
 	},
+	downarrow/.style={draw=black,
+	fill=white,
+	minimum width=0.6cm,
+	single arrow
+	},
 	longarrow/.style={draw=none,
@@
 This enables other risks, such as improper delivery, excessive fuel burn, and possible safety issues due to cargo imbalance, as well as a longer than necessary turn-around time. This {\it NP-hard}\/ problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modelled using standardised pallets in fixed positions.
-We developed a strategy to solve this problem, considering historical transport data from some Brazilian hub networks, and performed several experiments with a commercial solver, four known meta-heuristics, and a new heuristic designed specifically for this problem.
+We developed a strategy to solve this problem, considering historical transport data from some Brazilian hub networks, and performed several experiments with a commercial solver, {\color{blue}five} known meta-heuristics, and a new heuristic designed specifically for this problem.
 By using a portable computer, our strategy quickly found practical solutions to a wide range of real problems in much less than operationally acceptable time.
@@
 
-A few years ago, \citep{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP), and {\it Weight and Balance Problem} (WBP).
+A few years ago, \cite{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP), and {\it Weight and Balance Problem} (WBP).
 Several aspects were considered: item characteristics to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs); when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how total cargo weight is balanced; etc.
-
-{\color{blue}
-
-Recently, \citep{zhao2023} developed integer programming solution that address ACLPP and WBP separately, impacting payload optimization and aircraft centre of gravity (CG).
-The authors argue that, with improvements in computer processing power, joint combinatorial optimization of ACPP and WBP is now feasible.
-They propose three integer linear programming models: a bi-objective optimization model (BOM), a combinatorial optimization model (COM), and an improved combinatorial optimization model (IOM).
-The models aim to determine the maximum loading capacity and lowest CG deviation from a specified target CG.
-
-The study uses Gurobi to solve four scenarios with various conditional metrics for three models.
-Results show that the BOM has the fastest solution speed, but the CG deviation is the largest, making the results unacceptable in some cases.
-The COM has the longest solution time, making it difficult to tolerate in practice.
-The IOM offers the best optimization solution, despite taking a little longer to solve computationally than the BOM.
-
-Using a Boeing 777F, four scenarios and 25 cases were tested, finding that the BOM's variable feasible domain is limited, the optimization result is insufficient, and the CG deviation is sometimes too large.
-The COM takes too long to solve, making it difficult to meet the needs of the actual scene.
-The IOM accelerates the solving process compared to the COM, but some scenarios still require significant computation time.
-
-}
-
-{\color{violet}
-
-Inefficient air transport plans can lead to unnecessary costs, extra routes, longer distances, and incorrect destinations.
-Unbalanced cargo increases fuel consumption due to altered aircraft pitch angles.
-
-A study by \cite[p.
-22]{gwh2007} highlights the risk of weight- and balance-related accidents on cargo flights compared to passenger flights.
-So, balancing cargo is crucial for safe aerial transportation, as an improperly positioned centre of gravity can result in dangerous take-off and landing conditions and stall recovery issues.
-
-Another important issue is managing fuel consumption.
-To get an idea of the influence of CG displacement on fuel consumption, \cite[p.
-140]{MongeauBes2003} report that {\it "a displacement of the CG of less than 75 cm in a long-range aircraft yields, over a 10,000 km flight, a saving of 4,000 kg of fuel"}.
-This reinforces the importance of balancing cargo.
-
-Despite technological advancements, many airlines still rely on manual aircraft loading and balancing, using simple computer-aided tools for CG computation and constraint testing.
-Loadmasters often create load plans manually, which is time-consuming and can lead to flight delays.
-
-As to the right delivery, an organisation's purchasing department may acquire supplies in the main capitals of the country and abroad to meet the company's needs, and it wants these materials to be delivered to the requesting organisations within the stipulated deadlines.
-
-A rational decision-making process is crucial to avoid creating inefficient or unsafe transport plans, considering the high costs of fuel, maintenance, operation, outsourcing expenses, potential operational impairments, and safety risks due to unbalanced cargo.
-
-Solving the ACLPP+RPDP is vital for optimising strategic scores of what must be transported, saving time and effort in loading, ensuring safety and balance, ensuring correct pickups and deliveries, and finding the best routes considering fuel use due to potential cargo imbalance.
-
-}
 
@@
 Delays in shipping essential items such as respirators to critical areas highlighted the need for optimised solutions.
+
+Inefficient air transport plans can lead to unnecessary costs, extra routes, longer distances, and incorrect destinations.
+Unbalanced cargo increases fuel consumption due to altered aircraft pitch angles, increasing the risk of weight- and balance-related accidents.
+Balancing cargo is crucial for safe aerial transportation, as an improperly positioned centre of gravity (CG) can result in dangerous take-off and landing conditions and stall recovery issues.
+Despite technological advancements, many airlines still rely on manual aircraft loading and balancing, which can lead to flight delays.
+A rational decision-making process is essential to avoid creating inefficient or unsafe transport plans, considering high costs of fuel, maintenance, operation, outsourcing expenses, potential operational impairments, and safety risks due to unbalanced cargo.
+Solving this problem is vital for optimizing strategic scores, saving time and effort in loading, ensuring safety and balance, correct pickups and deliveries, and finding the best routes considering fuel use due to potential cargo imbalance.
 
@@
 The vast majority of operational research applied to air cargo is focused on challenges related to WBP, that is, the distribution of items on pallets to ensure load balancing.
-We can mention: \citep{LarsenMikkelsen1979}; \citep{Brosh1981}; \citep{Kevin1992}; \citep{Heidelberg1998}; \citep{fok2004optimizing}; \citep{KaluznyBohdanL2009Oalb}; \citep{Verstichel2011}; \citep{Limbourg2012}; \citep{RoesenerBarnes2016}; \citep{YangLiuGao2018}; \citep{zhao2021}; \citep{MiguelOptimalAP}.
-
-Other authors have addressed pallet assembly (APP) on aircraft, possibly also considering load balancing (WBP): \citep{MongeauBes2003}; \citep{Chan2006}; \citep{RoesenerHall2014}; \citep{Vancroonemburg2014}; \citep{Paquay2016}; \citep{Paquay2018}; \citep{wong2020}; \citep{eugene2021}; \citep{zhao2023}.
+We can mention: \cite{LarsenMikkelsen1979}; \cite{Brosh1981}; \cite{Kevin1992}; \cite{Heidelberg1998}; \cite{fok2004optimizing}; \cite{KaluznyBohdanL2009Oalb}; \cite{Verstichel2011}; \cite{Limbourg2012}; \cite{RoesenerBarnes2016}; \cite{YangLiuGao2018}; \cite{zhao2021}; \cite{MiguelOptimalAP}.
+
+Other authors have addressed pallet assembly (APP) on aircraft, possibly also considering load balancing (WBP): \cite{MongeauBes2003}; \cite{Chan2006}; \cite{RoesenerHall2014}; \cite{Vancroonemburg2014}; \cite{Paquay2016}; \cite{Paquay2018}; \cite{wong2020}; \cite{eugene2021}; \cite{zhao2023}.
 
 In all these works, there is a great diversity of scenarios and solutions: some consider items in two dimensions, and others in three dimensions; some used integer programming, and others developed specific heuristics.
-
-\citep{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem.
+Among the most recent, we can highlight:
+\begin{itemize}
+
+	\item \cite{RoesenerBarnes2016} proposed a heuristic to solve the {\it Dynamic Airlift Loading Problem} (DALP).
+Given a set of palletized cargo items that require transport between two nodes in a given time frame, the objective of this problem is to select an efficient subset of aircraft, partition the pallets into aircraft loads, and assign them to allowable positions in those aircraft.
+
+	\item \cite{Paquay2016} presented a mathematical model to optimise the loading of heterogeneous 3D boxes on pallets with a truncated parallelepiped format.
+Its objective is to maximise the volume used in containers, considering load balancing constraints, the presence of fragile items, and the possibility of rotating these boxes.
+\cite{Paquay2018} developed some heuristics to solve this problem.
+
+	\item \cite{YangLiuGao2018} modelled the air transport problem as a 2D packing problem and presented a heuristic for its optimisation in several aircraft, considering load balancing to minimise fuel consumption.
+
+	\item \cite{wong2020} developed a mathematical model and a tool based on mixed integer programming for optimising cargo in aircraft with different pallet configurations.
+Balance constraints and the presence of dangerous items were considered.
+\cite{eugene2021} integrated this tool into a digital simulation model with a visualisation and validation system based on sensors that alert about load deviations.
+
+	\item \cite{zhao2021} proposed a model for WBP based on Mixed Integer Programming (MIP).
+Instead of focusing on the CG deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
+
+	\item \cite{zhao2023} proposed three integer linear programming models: a bi-objective optimisation model (BOM), a combinatorial optimisation model (COM), and an improved combinatorial optimisation model (IOM).
+Its objectives were the maximum possible loading and the smallest CG deviation.
+Four scenarios with multiple metrics are resolved for the {\it Boeing 777F}.
+The results showed that the BOM has the fastest solution speed, but the CG deviation is the worst; in several cases, the CG deviation is unacceptable.
+COM has the longest solution time, which is difficult to tolerate in practice.
+
+\end{itemize}
+
+\cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem.
 The authors demonstrated that this problem is {\it NP-hard}.
 Although it is innovative, strong simplifications were imposed by these authors: in relation to loading, APP was ignored; regarding routing, it is assumed that a predefined tour plan is restricted to only two legs.
-Referring directly to this work, \citep{BrandtStefan2019} comment: {\it However, not even these sub-problems are acceptably solved for real-world problem sizes, or models omit some practically relevant constraints}.
+Referring directly to this work, \cite{BrandtStefan2019} comment: {\it However, not even these sub-problems are acceptably solved for real-world problem sizes, or models omit some practically relevant constraints}.
 
@@
 		\midrule
-		\citep{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$ \\
-		\citep{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$ \\
-		\citep{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ \\
-		\citep{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$ \\
-		\citep{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\citep{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
-		\citep{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$ \\
-		\citep{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$ \\
-		\citep{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
-		\citep{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\citep{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
-		\citep{Vancroonemburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\citep{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\citep{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
-		\citep{Paquay2016,Paquay2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$ \\
-		\citep{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$ \\
-		\citep{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$  \\
-		\citep{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$  \\
-		\citep{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
-		\citep{zhao2023} & $\bigstar$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$  \\
-		\citep{MiguelOptimalAP} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$  \\
+		\cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$ \\
+		\cite{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$ \\
+		\cite{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ \\
+		\cite{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$ \\
+		\cite{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
+		\cite{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
+		\cite{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$ \\
+		\cite{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$ \\
+		\cite{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
+		\cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
+		\cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
+		\cite{Vancroonemburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
+		\cite{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
+		\cite{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
+		\cite{Paquay2016,Paquay2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$ \\
+		\cite{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$ \\
+		\cite{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$  \\
+		\cite{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$  \\
+		\cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
+		\cite{zhao2023} & $\bigstar$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$  \\
+		\cite{MiguelOptimalAP} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$  \\
 		{\bf This article}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$  \\
@@
 
-{\color{blue}
-
-As for the size of the problem dealt with in this work, addressing the scale of the Travelling Salesperson Problem (TSP) in practical, real-world air transport contexts indeed reveals a fascinating and complex challenge.
-In the realm of academic research, TSP problems often explore theoretical models that can scale to hundreds or even larger numbers of cities.
-
-In practical terms, especially within the air transport sector, the size and complexity of a TSP can vary significantly depending on several factors:
-
-\textbf{Fleet and Route Planning}: For airline operations, the problem isn't just about finding the shortest path through a set of points; it also involves scheduling, aircraft assignment, crew scheduling, maintenance planning, and delivering the right cargo to the right client.
-This last operation task complexity transforms a standard TSP into a more intricate problem in our study, the Pickup and Delivery Problem (PDP).
-
-\textbf{Operational Constraints}: Factors like weather, air traffic control restrictions, and the availability of take-off and landing slots make real-world TSP scenarios in air transport even more challenging.
-Although we do not deal with these constraints, they add layers of complexity that significantly depart from the theoretical TSP model.
-
-\textbf{Dynamic Nature}: Unlike the static nature of classical TSP, real-world problems in air transport are highly dynamic.
-Routes need to be optimized continuously as conditions change, requiring algorithms that can adapt in real-time or near-real-time.
-
-In practical air transport scenarios, the size of a TSP can be less about the sheer number of nodes and more about the complexity and dynamic nature of the problem, which in our case (a single aircraft in a single day) is rarely expected to be more than seven nodes.
-Despite this, at the end of this work, we to submit the best method found to a more challenging 15-node problem as a validation test.
-}
-
 \subsection{Operational premises}
@@
 As we are dealing with an extremely complex and diverse problem, we decided to establish some simplifying characteristics:
-
-\vspace{-2mm}
 
@@
 $p_i$\/ are pallets, $1\leq i \leq 18$, whose weight and volume limits are $W_i$\/ and $V_i$, respectively.
-$D_i^{long}$\/ and $D_i^{lat}$\/ are, respectively, the longitudinal and lateral distances of each pallet centroids to the aircraft centre of gravity (CG) along both axes.
+$D_i^{long}$\/ and $D_i^{lat}$\/ are, respectively, the longitudinal and lateral distances of each pallet centroids to the aircraft CG along both axes.
 These distances will be used in the calculation of the torque, referring to the items allocated on each pallet.
@@
 We also make the following assumptions:
-
-\vspace{-2mm}
 
@@
 Once the assumptions and the mathematical modelling are presented, it is possible to see that ACLP+RPDP is {\it NP-hard}.
-In a similar way to \citep{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$\/ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in node 0, and no items in node 1.
+In a similar way to \cite{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$\/ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in node 0, and no items in node 1.
 Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is {\it NP-complete}.
-For more details, see \citep[p.~6]{LurkinSchyns2015}.
+For more details, see \cite[p.~6]{LurkinSchyns2015}.
 
@@
 
-In practical cases, we know that a common aircraft has $m=18$\/ pallets, flight itineraries have $K \leq 6$\/ nodes plus the base, and each node has hundreds of items to be shipped.
-We also know that missions with fewer nodes are more frequent than longer ones.
-Under these circumstances, we can adopt some important strategies:
-
-\vspace{-2mm}
-
-\begin{itemize}
-	\setlength\itemsep{-0.3em}
-
-	\item We consider that the number of destinations is smaller than the number of pallets, that is, $K < m$.
-We also avoid the trivial case where $K=1$.
-With this premise, we can preset the destinations of the pallets at each shipping node, reserving for each destination a number of pallets proportional to the volume available.
-We could have used another criterion, but it was observed in the experiments that the volume is more constrictive in airlift.
-
-{\color{violet}
-	\item As the number of nodes is small, we have the possibility to check all possible tours, selecting the one that provides the best value for the objective function.
-For this reason, a maximum time per solution ($tmax$) is established.
-In the tests carried out in this work, these times were in the range [240s, 1200s, 2400s, 3600s].
-
-	\item At each node, we will use a MIP solver or heuristics to check which method offers the best result within the time limit for each node.
-
-	\item The time limit to solve a node will be adjusted to be proportional to the sum of volumes of the candidate items of the node in relation to the sum of volumes of the entire tour.
-}
-
-\end{itemize}
-
-Our complete strategy is summarized in Algorithm \ref{alg:main}.
-
-\begin{algorithm}[htp]
-
-	\caption{Solving the ACLP+RPDP}  \label{alg:main}
-
-	\begin{algorithmic}[1]
-			\State {\bf ACLP+RPDP} \,\, {\it in:}~$scenario, surplus, tmax, tsp, ntours$ \, \, {\it out:}~$answer$
-
-			\State Let $M$ be the set of pallets ({\it cfr.} Table \ref{tab:larger}) \label{main:M}
-
-			\State Let $K$, $L$\/ and $C$ be according to the $scenario$ ({\it cfr.} Tables \ref{tab:costs}, \ref{tab:larger} and \ref{tab:scenarios}) \label{main:KLC}
-
-			\State $N \gets ItemsGeneration(scenario, surplus)$ \label{main:items}
-
-		{\color{violet}
-			\If {$tsp$ = True}
-
-				\State $tmax \gets tmax/ntours$ \Comment{max. tour time to solve the number of tours from the TSP heuristic}
-
-			\Else
-
-				\State $tmax \gets tmax/K!$
-
-			\EndIf
-		}
-			\For {each $method$}  \label{main:method}
-
-				\For {each $\pi \in S_K$} \label{main:loop1}
-
-					\State $f_{\pi} \gets SolveTour(\pi, L, M, C, N, method, tmax)$ \label{main:fpi}
-
-				\EndFor \label{main:loop2}
-
-				\State $answer[scenario,surplus,method] \gets \max f_{\pi}$ \label{main:a2}
-			\EndFor
-	\end{algorithmic}
-
-\end{algorithm}
-
-In this algorithm, we use {\color{blue}six} values for $scenario$, according to Tables \ref{tab:costs} and \ref{tab:scenarios}, which define the number $K$ of destinations, the set $L$\/ of nodes, the costs $C$\/, and the shortest tours $\pi_{TSP1}$\/ and $\pi_{TSP2}$.
-
-{\color{blue}
-Scenario 6 will be used only for the best method validation.
-}
-
 \begin{figure}[htp]
-
-	\centering
-	\fbox{
-	\begin{tikzpicture}
-
-		%% Make the boxes
-		\node [mybox] (first)                        { \small {\bf Select:}\\ 
-			\small 1. scenario\\ 
-			\small 2. surplus\\ 
-			\small 3. time limit};
-
-		\node [mybox, right=0.7cm of first] (second) { \small {\bf Config:}\\ 
-			\small 4. parameters\\ \small 5. cost matrix\\ 
-			\small 6. pallets\\ 
-			\small 7. items};
-
-		\node [mybox, right=0.7cm of second] (third) { \small {\bf Define tour}\\  
-			\small {\bf schema:}\\ 
-			\small 8. All K! tours\\ \ \ \  
-			\small or\\ 
-			\small 8. Solve a TSP\\ 
-			\small  \ \ \ (set of tours))};
-
-		\node [mybox, right=0.7cm of third] (fourth) { \small {\bf Solve ACLPP+RPDP:}\\ 
-			\small 9. Many tours{\bf *}\\ 
-			\small 10. Select best tour};
-
-		\node [mybox, below=0.7cm of third] (fifth) { \small {\bf *Solve a tour:}\\
-			\small \ \ \ {\it - For each node:}\\
-			\small \ \ \ \ \ {a. Unload some pallets}\\
-			\small \ \ \ \ \ {b. Arrange remaining pallets to balance cargo}\\
-			\small \ \ \ \ \ {c. Set empty pallets destinations}\\
-			\small \ \ \ \ \ {d. \it For each pallet:}\\
-			\small \ \ \ \ \ \ \ \ \ {Solve an ACLPP with WBP and PDP constraints}\\
-			\small \ \ \ \ \ {e. Calculate node score and torque}\\
-			\small \ \ \ {\it - Return tour Score/Cost}
-
-	};
-
-		%% Draw arrows between them
-		\node[myarrow] at ([xshift=8pt]first.east) {\phantom{\tiny{aaa}}};
-		\node[myarrow] at ([xshift=8pt]second.east) {\phantom{\tiny{aaa}}};
-		\node[myarrow] at ([xshift=8pt]third.east) {\phantom{\tiny{aaa}}};
-
-		\path[->] (third) edge (fifth);
-
-	\end{tikzpicture} 
-}
-	\caption{Solution main flow}
-	\label{fig:flow}
-\end{figure}
-
-\begin{figure}[htp]
-	\centering
-	\fbox{
+	\centering
+	\fbox{ \small
 		\begin{tikzpicture}
@@
 				\small \ \ \ {Airports and distances}\\
-				\small \ \ \ {$scenario$ \ \ ($K$ and $L$)}\\
+				\small \ \ \ {$scenario$ \ \ ($K$, $L$\/ and $C$)}\\
+				\small \ \ \ {Items available for shipment at each airport}\\						
 				\small \ \ \ {$surplus = \sum_{j=1}^{n_k} v_j$/$\sum_{i=1}^{m} V_i$ }\\		
 				\small \ \ \ {$tmax$ \ \ (overall runtime limit)}\\
-				\small {{\bf Requirement:} $K < m$}
+				\small {{\bf Requirement:} $1 < K < m$}
 			};
 
-			\node [mybox, below=0.7cm of one] (two) { \small {\bf Tours definition:}\\
+			\node [mybox, below=0.7cm of one] (two) { \small {\bf Number of tested tours:}\\
 				\small \ \ \ {{\bf if} $K \leq 6$}\\
@@
 				\small \ \ \ {\bf else}\\
-				\small \ \ \ \ \ \ {$ntours = 100$ \ \ (best tours obtained with TSP heuristic)}
+				\small \ \ \ \ \ \ {$ntours = 100$ \ \ (best tours obtained with a TSP heuristic)}
 			};
 
-			\node [mybox, below=0.7cm of two] (three) { \small {\bf On each tour of the $ntours$:}\\
-				\small \ \ \ {Calculation of a node-by-node solution (divide runtime according to shipment volumes)}\\
-				\small \ \ \ {On each node of current tour:}\\
-				\small \ \ \ \ \ \ \ {- Relocate packed contents with torque optimization}\\
-				\small \ \ \ \ \ \ \ {- Predefine pallet destinations according to shipment volumes}\\
+			\node [mybox, below=0.7cm of two] (three) { \small {\bf On each tour of $ntours$:}\\
+				\small \ \ \ {Calculate a node-by-node solution (divide runtime according to shipment volumes):}\\
+				\small \ \ \ \ \ \ \ {- Reallocate packed contents with torque optimization}\\
+				\small \ \ \ \ \ \ \ {- Preset pallet destinations according to shipment volumes}\\
 				\small \ \ \ \ \ \ \ {- Find a node solution (MIP, Shims, NMO, ACO, GRASP, TS, GA) }\\
-				\small \ \ \ \ \ \ \ {- Accumulate values obtained from score, torque and cost}
+				\small \ \ \ \ \ \ \ {- Accumulate values obtained from score, torque and cost}\\
+				\small \ \ \ \ \ \ \ {- Go to the next node on the current tour}
 			};
@@
 
-{\color{blue}
-Figure \ref{fig:flow} depicts the Algorithm \ref{alg:main} for easy understanding of the main flow.
-(1.) Select a scenario from Table \ref{tab:scenarios}.
-(2.) Select a volume surplus (1.2, 1.5, or 2.0).
-(3.) Establish a time limit (240s, 1200s, 2400s, or 3600s).
-(4.) Configure the parameters presented in the mathematical formulation.
-(5.) Load the distance matrix and calculate the matrix cost according to the aircraft characteristics.
-(6.) Load the pallet configuration parameters.
-(7.) Load the item instance (all items to be transported) into memory.
-(8.) Choose a tour schema.
-Until 7 nodes, it is practical to generate K!
-tours.
-For a larger number of nodes, we recommend using a TSP heuristic that generates a set of similar suboptimal tours.
-(9.) and (10.) solve all tours and select the best.
-}
+In practical cases, we know that a common aircraft has $m=18$\/ pallets, flight itineraries have $K \leq 6$\/ nodes plus the base, and each node has hundreds of items to be shipped.
+We also know that missions with fewer nodes are more frequent than longer ones.
+Under these circumstances, we can adopt some important strategies summarized in Figure \ref{fig:process}:
+
+\begin{itemize}
+
+	\item We consider that the number of destinations is smaller than the number of pallets ($K < m$), and we avoid the trivial case where $K=1$.
+With this premise, we can preset the destinations of the pallets at each shipping node, reserving a number of pallets proportional to the volume available for each destination.
+We could have used another criterion, but it was observed in the experiments that the volume is more constrictive in airlift.
+
+	\item An important parameter is the number $ntours$\/ of tours tested.
+In practical cases where $K \leq 6$, we have the possibility to check all possible tours ($ntours = K!$).
+In this situation, as $K$\/ is small, we can also specially analyse the two optimal solutions of the corresponding TSP ($ntours = 2$).
+Finally, in cases where $K>6$, we will use a heuristic to select $100$\/ tours of low length ($ntours=100$), and search among them for the one that provides the best value for the objective function.
+
+	\item To compare the performance of each strategy, an overall runtime limit $tmax$\/ is established and divided by $ntours$\/ tours.
+In turn, the runtime limit for each tour will be distributed among its nodes in proportion to the volume available for boarding.
+
+	\item At each node of a tour, the packed contents that remain on board are reallocated on pallets in order to minimize torque on the aircraft.
+This calculation is done quickly using a MIP solver.
+Then, the destinations of the pallets are previously defined in proportion to the shipment volume.
+Finally, considering the runtime limit of each node, we will use a MIP solver and five well-known meta-heuristics to find the best allocation of shipping items: {\it Ant Colony Optimization} (ACO), {\it Noising Method Optimization } (NMO), {\it Tabu Search} (TS), {\it Greedy Randomized Adaptive Search Procedure} (GRASP), and {\it Genetic Algorithm} (GA).
+
+	\item We will generate benchmarks using the $surplus$\/ parameter, which is a value in $\{1.2, 1.5, 2.0\}$.
+It corresponds, at each node $k$, to the ratio between the sum of the volumes of the items and the load capacity of the pallets ($surplus = \sum_{j=1}^{n_k} v_j$/$\sum_{i=1}^{m} V_i$).
+This parameter allows us to verify the different behaviour of each method, according to $scenario$\/ and the quantity of items available for shipment.
+
+	\item We will do tests by varying the number $K$\/ of destinations, the set $L$\/ of nodes, and the costs $C$.
+Each group of values tested is called $scenario$, according to Tables \ref{tab:costs} and \ref{tab:scenarios}, where $1<K \leq 6$.
+After finding the method with the best performance in node-by-node solution, we will test it in solving cases with $K > 6$.
+
+\end{itemize}
+
+Algorithm \ref{alg:main} is the main program of this solution process, which parameters are $scenario$, $surplus$, $tmax$, and $ntours$.
+The input data is obtained from Tables \ref{tab:costs}, \ref{tab:larger} and \ref{tab:scenarios} (lines \ref{main:M}-\ref{main:KLC}).
+
+\begin{algorithm}[htp]
+
+	\caption{Solving the ACLP+RPDP}  \label{alg:main}
+
+	\begin{algorithmic}[1]
+			\State {\bf ACLP+RPDP} \,\, {\it in:}~$scenario, surplus, tmax, ntours$ \, \, {\it out:}~$answer$
+
+			\State Let $M$ be the set of pallets ({\it cfr.} Table \ref{tab:larger}) \label{main:M}
+
+			\State Let $K$, $L$\/ and $C$ be according to the $scenario$ ({\it cfr.} Tables \ref{tab:costs}, \ref{tab:larger} and \ref{tab:scenarios}) \label{main:KLC}
+
+			\State $N \gets ItemsGeneration(scenario, surplus)$ \label{main:items}
+
+			\For {each $method$}  \label{main:method}
+
+				\For {$x \gets 1$ to $ntours$} \label{main:loop1}
+
+					\State $f(S_K[x]) \gets SolveTour(S_K[x], L, M, C, N, method, tmax/ntours)$ \label{main:fpi}
+
+				\EndFor \label{main:loop2}
+
+				\State $answer[scenario,surplus,method] \gets \max f(S_K[x])$ \label{main:a2}
+			\EndFor
+	\end{algorithmic}
+
+\end{algorithm}
+
+$surplus$ is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).
+$tmax$\/ is the runtime limit, which will be distributed among the tours (line \ref{main:fpi}).
+$method$\/ corresponds to a MIP solver or a heuristic to the node-by-node solution $SolveTour$, which will be presented in subsection \ref{methods}.
+The best results obtained by testing all tours are stored in $answer$\/ (line \ref{main:a2}), which is the output of this algorithm.
+
+{\color{red}$\pi_k$ are tours} {\color{blue}$\pi$ are tours}, that is, permutations with the $K$\/ nodes.
+When $ntours=2$, $\pi_1$\/ and $\pi_2$\/ are the optimal solutions of the corresponding TSP.
+When $ntours=K!$, $\pi_k$\/ is the $k^{th}$ permutation of $S_K$.
+Finally, when $ntours=100$, $\pi_1, \ldots, \pi_{100}$\/ are the best TSP solutions provided by a heuristic.
 
@@
 		5 & 6    & \{$0,1,2,3,4,5,6$\}    \\
-		6 & 15   & \{$0,1,2,3,4,5,6,7,8,9,10,11,12,13,14$\}    \\
 		\bottomrule
@@
 
-$surplus$\/ is a value in $\{1.2, 1.5, 2.0\}$, which corresponds, at each node $k$, to the ratio between the sum of the volumes of the items  and the load capacity of the pallets ($surplus = \sum_{j=1}^{n_k} v_j$/$\sum_{i=1}^{m} V_i$).
-This parameter allows us to verify the different behaviour of each $method$, according to $scenario$\/ and the quantity of items available for shipment.
-It is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).
-
-$tmax$\/ is the runtime limit, which will be distributed among the tours (lines \ref{main:f1}, \ref{main:f2} and \ref{main:fpi}).
-$method$\/ corresponds to a MIP solver or a heuristic to the node-by-node solution $SolveTour$, which will be presented in subsection \ref{methods}.
-
-The best results obtained by testing all tours are stored in $answer$\/ (line \ref{main:a2}).
-
-Next, we will present two subsections: in the first we explain how $SolveTour$\/ is executed, presetting the destinations of the pallets.
-In the second we will present the heuristics developed for node-by-node solutions.
+Next, we will present two subsections: in the first we explain how $SolveTour$\/ is executed. In the second we will present the heuristics developed for node-by-node solutions.
 
@@
 
-$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $tmax$\/ is the {\color{blue}runtime limit of the tour.}
+$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $tmax$\/ is the runtime limit of this tour.
 
@@
 
-		{\color{violet}
-
-		\State $V_{\pi}  \gets 0$ \label{tour:Vol} \Comment{tour volume}
-
-		\For {$k \gets 0$ to $K$} \label{tour:loop0}
-
-			\State $v_{\pi_k}  \gets \sum_{j=1}^{n_k} v_j$
-			\State $V_{k} \gets V_{k} + v_{\pi_k}$
-
-		\EndFor
-
-	}
-
 		\For {$k \gets 0$ to $K$} \label{tour:loop1}
-
-		\State $tnode_{\pi_k} = (v_{\pi_k}/V_{\pi}) * tmax$ \Comment{{\color{blue} node time limit calculation}}
+		\State $t_{\pi_k} = (\sum_{j=1}^{n_k} v_j/ \sum_{k=0}^K \sum_{j=1}^{n_k} v_j) * tmax$ \label{tour:tnode}	 \Comment{runtime limit proportional to the shipment volume}
 
@@
 
-		\State $G_2 \gets SolveNode(method, \pi_k, G_1, tnode_{\pi_k})$  \label{tour:SolveNode}
+		\State $G_2 \gets SolveNode(method, \pi_k, G_1, t_{\pi_k})$  \label{tour:SolveNode}
 
@@
 After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ flights (lines \ref{tour:loop1}-\ref{tour:loop2}).
-Initially, the set $L_{\pi_k}$\/ of remaining nodes is updated (line \ref{tour:lk1}), and the pallet destinations are unset (line \ref{tour:-1}).
-
-When the aircraft is at the base, the initial graph $G_1$\/ is empty, and there are no packed contents \ref{tour:g11}.
+Initially, the runtime limit $t_{\pi_k}$\/ for each node is calculated (line \ref{tour:tnode}), the set $L_{\pi_k}$\/ of remaining nodes is updated (line \ref{tour:lk1}), and the pallet destinations are unset (line \ref{tour:-1}).
+
+When the aircraft is at the base, the initial graph $G_1$\/ is empty, and there are no packed contents {\color{blue}(line \ref{tour:g11})}.
 Otherwise, $UpdatePacked$\/ (line \ref{tour:dest}) returns the set of packed contents that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation.
@@
 
-{\color{violet}
-
 One of the main objectives of this work was to find a quick heuristic that offers a good-quality solution for the node-by-node problem.
-Taking this into account, we design algorithms based on four known meta-heuristics: {\it Ant Colony Optimization} (ACO) \citep{Dorigo1992,DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \citep{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \citep{Glover1986}, {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \citep{FeoResende1989}, and the Genetic Algorithm (GA) \citep{holland1992adaptation}.
-We considered several ideas from the literature \citep{NiarFreville1997, Fidanova2006, Alonso2019, Zhan2020, shah2020genetic}, and we were careful to use the same data structures and procedures in all implementations to enforce fair results comparison.
-
-As \citep{PeerlinckAmy2022MFEO} applied the DEAP (Distributed Evolutionary Algorithms in Python) to solve a Multi-Objective Knapsack Problem, we also decided to apply this evolutionary package to implement a GA solution for the ACLPP.
-
-}
+Taking this into account, we design algorithms based on five known meta-heuristics: {\it Ant Colony Optimization} (ACO) \citep{Dorigo1992,DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \citep{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \citep{Glover1986}, {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \citep{FeoResende1989}, and {\it Genetic Algorithm} (GA) \citep{holland1992adaptation}.
+We considered several ideas from the literature \citep{NiarFreville1997, Fidanova2006, Alonso2019, Zhan2020, shah2020genetic,PeerlinckAmy2022MFEO}, and we were careful to use the same data structures and procedures in all implementations to enforce fair results comparison.
 
 However, the heuristic that presented better solutions was none of the previous ones.
-In this subsection, we will present a new heuristic for the node-by-node problem, called $Shims$.
+In this subsection, we will present a new heuristic for the node-by-node problem, called {\it Shims}.
 Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}).
@@
 
-Considering only the items that can be shipped at the node $\pi_k$, Figure \ref{fig:whip} represents $n_{\pi_k}$\/ possible edges $e_{ij}$\/ of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order.
-Initially, $Shims$\/ builds a greedy solution for the pallet $i$\/ selecting edges up to index $\eta_1$ ({\it greedy phase}).
+Considering only the items that can be shipped at node $\pi_k$, Figure \ref{fig:whip} represents $n_{\pi_k}$\/ possible edges $e_{ij}$\/ of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order.
+Initially, {\it Shims}\/ builds a greedy solution for the pallet $i$\/ selecting edges up to index $\eta_1$ ({\it greedy phase}).
 Then, with the edges between $\eta_1$\/ and $\eta_2$, it elaborates different possible complements ({\it composition phase}), including later the best ones in the same pallet ({\it selection phase}).
-$Shims$\/ is depicted in Algorithm \ref{alg:shims}.
+{\it Shims}\/ is depicted in Algorithm \ref{alg:shims}.
 
 \begin{algorithm}[H]
-	\caption{$Shims$\/ heuristic at node $\pi_k$}  \label{alg:shims}
+\small
+	\caption{{\it Shims}\/ heuristic at node $\pi_k$}  \label{alg:shims}
 	\begin{algorithmic}[1]
 
-		\State {\bf SolveNode} \,\, {\it in: $Shims, \pi_k, G, tmax, {\color{blue}threshold_1, threshold_2}$} \, \, {\it out: $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$}
+		\State {\bf SolveNode} \,\, {\it in: {\it Shims}, $\pi_k, G, tmax, level_1, level_2$} \, \, {\it out: $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$}
 
@@
 
-		\If{ ($E_{N_{\pi_k}} \cup \{e_{ij}\}$ is feasible) {\bf and} ($vol_i \leq V_i \times {\color{blue}threshold_1}$) {\bf and} ($| \tau_{\pi_k} + w_j \times D_i^{long} | \leq W_{max} \times limit^{CG}_{long}$)}
+		\If{ ($E_{N_{\pi_k}} \cup \{e_{ij}\}$ is feasible) {\bf and} ($vol_i \leq V_i \times level_1$) {\bf and} ($| \tau_{\pi_k} + w_j \times D_i^{long} | \leq W_{max} \times limit^{CG}_{long}$)}
 
@@
 		\EndIf
-		\Until ($vol_i > V_i \times {\color{blue}threshold_1}$) {\bf or} ($\eta_1>n_{\pi_k}$) \label{shims:endfirst}
+		\Until ($vol_i > V_i \times level_1$) {\bf or} ($\eta_1>n_{\pi_k}$) \label{shims:endfirst}
 
@@
 		\State $\eta_2 \gets \eta_1$  \label{shims:eta2a}
-%		\While{($\eta_2 \leq n_{\pi_k}$) {\bf and} ($ vol_i < (1 + 2 \times (1-limit)) \times V_i$)}
-{\color{blue}
-		\While{($\eta_2 \leq n_{\pi_k}$) {\bf and} ($ vol_i < V_i \times {\color{blue}threshold_2}$)}
+		\While{($\eta_2 \leq n_{\pi_k}$) {\bf and} ($ vol_i < V_i \times level_2$)}
 
@@
 		\EndWhile \label{shims:endsecond}
-	}
+
 		\State $vol \gets 0$; $b \gets 1$; $shims_b \gets \varnothing$; $Set \gets \{ shims_b \}$  \label{shims:beginthird}
@@
 
-{\color{violet}
-
-	The arguments $threshold_1$ and $threshold_2$ are the volume thresholds for $\eta_1$ and $\eta_2$ calculations, respectively.
-
-The argument $tmax$ is the maximum time for {\it Shims} to solve a node.
-It is proportional to the sum of the candidate items volumes in each node, in relation to the sum of all nodes volumes, and the maximum time to solve the tour.
-
-}
+$tmax$\/ is the runtime limit for {\it Shims}, and $level_1$\/ and $level_2$\/ are volume thresholds for the indices $\eta_1$ and $\eta_2$, respectively.
 
@@
 
-\vspace{-2mm}
-
-\begin{itemize}
-
-	\setlength\itemsep{-0.3em}
+\begin{itemize}
 
 	\item In the {\it greedy phase}\/ (lines \ref{shims:pallets}-\ref{shims:endfirst}), a partial solution for each pallet $i$\/ is constructed by adding edges following $\theta_{ij}$\/ order.
-	The $\eta_1$\/ index corresponds to the accumulated volume equal to $V_i \times {\color{blue}threshold_1}$.
-In $\eta_2$\/ index, this same accumulated volume reaches $V_i \times  {\color{blue}threshold_2}$.
-{\color{blue}The values of these thresholds were defined empirically by the {\it irace}\/ tool \citep{LopezIbanezManuel2016}}.
+The indices $\eta_1$\/ and $\eta_2$\/ refer to the accumulated volumes $V_i \times level_1$\/ and $V_i \times level_2$, respectively, which were defined empirically by the {\it irace}\/ tool \citep{LopezIbanezManuel2016}.
+We will explain this in detail in the subsection \ref{items}.
 
@@
 
-{\color{violet}
-	{\it irace} (Iterated Racing for Automatic Algorithm Configuration) streamlines the tuning of optimization algorithms by iteratively testing and eliminating weaker parameter configurations through statistical comparisons.
-Its application may be understood, from a Machine Learning standpoint, as the task of identifying optimal parameter values for solving new problem instances by leveraging knowledge gained from a set of training problem examples.
-This method significantly cuts down on the time and effort needed to find optimal or near-optimal parameters, proving effective across various application domains.
+{\color{blue}
+\subsection{Complexity analysis}
+
+These are the steps to solve the ACLP+RPDP with the complexity analysis:
+
+\subsubsection{Input data}
+
+Load $K$ lines of text to compose a distances matrix for $K$ airports, the $m$ packed contents from previous nodes, and $n$ items parameters from text files are of complexities $O(K)$, $O(m)$, and $O(n)$, respectively:
+
+	\begin{itemize}
+		\item $O(K + m + n)$
+	\end{itemize}
+
+\subsubsection{Number of tested tours ($T$)}
+
+\begin{itemize}
+\item If $K <= 6$, determine the number of tours through permutation.
+To get permutation of nodes as tours has a complexity of $O(T = K!
+\cdot K)$, as the number of nodes is permutated (factorial) to generate tours.
+
+	\begin{itemize}
+	\item $ O(K + m + n) + O(T = K! \cdot K)$ or
+	\item $ O(K + m + n) + O(T = 2)$, if the we are solving the two optimal TSP tours.
+	\end{itemize}
+
+\item If $K > 6$, we obtain 100 tours from a heuristic TSP solution (see its time complexity analysis in subsubsection \ref{ga_tsp}):
+
+	\begin{itemize}
+		\item $O(K + m + n) + O(T = 100)$
+	\end{itemize}
+\end{itemize}
+
+\subsubsection{Solving $T$ tours:}
+
+\begin{itemize}
+	\item $O(K + m + n) + O(T)$
+\end{itemize}
+
+a) each tour with $K$ nodes:
+\begin{itemize}
+	\item $O(K + m + n) + O(T \cdot K)$
+\end{itemize}
+
+b) in each node we have to reset (to eliminate the risk of previous destinations assigned to a pallet) $O(K \cdot m)$ and set the empty pallets destinations: $O(K \cdot m + K \cdot m) = O(2 \cdot K \cdot m) = O(K \cdot m)$
+\begin{itemize}
+	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m))$
+\end{itemize}
+
+c) To calculate the time limit to solve each node, it is necessary to determine the potential total volume for each node.: $O(K \cdot n)$
+\begin{itemize}
+	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n))$
+\end{itemize}
+
+d) The time complexity to put the packet contents destined to the next nodes in the tour on board: $O(m)$.
+\begin{itemize}
+	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m))$
+\end{itemize}
+
+d) The time complexity to optimize the pallets positions to minimize the CG deviation: $O(m^2)$ ($m$ pallets in $m$ positions)
+\begin{itemize}
+	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2))$
+\end{itemize}
+
+e) Plus the time complexity of the method used to solve an ACLPP with the aircraft partially loaded: $O(C)$
+\begin{itemize}
+	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + C))$
+\end{itemize}
+
+\subsubsection{Considering the Shims algorithm: $O(C) = O(Shims)$}
+
+a) the time complexity for sorting the pallets in non-descending order of distances to the CG:
+\begin{itemize}
+	\item $O(Shims) = m \cdot log(m)$
+\end{itemize}
+
+b) As each pallet is solved at a time, the overall Shims complexity is multiplied by $m$.
+\begin{itemize}
+	\item $O(Shims) = m.(m \cdot log(m))$
+\end{itemize}
+
+c) The greedy phase has $O(n)$.
+\begin{itemize}
+	\item $O(Shims) = m \cdot (m \cdot log(m) + n)$
+\end{itemize}
+
+d) The composition phase has $O(n)$ added to the First-Fit Decreasing algorithm time complexity $O(m \cdot log(m))$
+\begin{itemize}
+	\item $O(Shims) = m \cdot (m \cdot log(m) + n + n + m \cdot log(m))$
+	\item $O(Shims) = m \cdot (2n + 2 \cdot m \cdot log(m)) = m \cdot (n + m \cdot log(m))$
+\end{itemize}
+
+The overall time complexity to solve the ACLP+RPDP with Shims is:
+\begin{itemize}
+	\item $O(K+m+n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot log(m))))$
+\end{itemize}
+
+The overall time complexity provided combines two distinct parts: $O(K+m+n)$ and $O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot \log(m))))$.
+Analysing each part separately and then considering their combined implications provides insight into the algorithm's performance characteristics.
+
+\subsubsection{Complexity analysis considerations}
+
+\textbf{First Part:} $O(K+m+n)$
+
+This component of the time complexity suggests a linear dependency on the sizes of three separate inputs or parameters, denoted by $K$, $m$, and $n$.
+These are sequential steps, each influenced by each of these parameters size.
+This implies that as any of $K$, $m$, or $n$ increases, the time required by the algorithm increases linearly with respect to the largest of these parameters, the number of items to be transported ($n$).
+
+\textbf{Second Part:} $O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot \log(m))))$
+
+This portion is considerably more complex, indicating an algorithm with nested dependencies on several variables ($T$, $K$, $m$, and $n$):
+
+\begin{itemize}
+	\item $T \cdot K$: repeated iterations $T$ for each of $K$ nodes, nested loops where inner operations are contingent upon $K$.
+	\item $K \cdot m + K \cdot n$: These operations scale with both $K$ and each of $m$ and $n$, indicating processes that involve all pairs of $K$ with $m$ and $n$.
+	\item $m + m^2$: operations with linear and quadratic dependencies on the number of pallets $m$, indicating that as $m$ grows, the complexity increases more significantly due to the quadratic term $m^2$.
+	\item $m \cdot (n + m \cdot \log(m))$: Combines linear scaling with $m$ and $n$ and a logarithmic term $m \cdot \log(m)$, because the algorithm uses divide-and-conquer strategies on $m$ elements, the Shims heuristics.
+\end{itemize}
+
+\textbf{Combined Implications}
+
+When combining these parts, the linear terms in the first part ($O(K+m+n)$) are overshadowed by the more complex and potentially rapidly increasing terms of the second part, especially due to the presence of $m^2$ and the product $T \cdot K \cdot (...)$.
+The quadratic ($m^2$) and logarithmic ($m \cdot \log(m)$) terms indicate that the algorithm's efficiency decreases significantly as the number of pallets $m$ increases.
+
+The overall complexity suggests an algorithm that performs efficiently with small values of $K$, $m$, and $n$ but may become substantially more resource-intensive as these parameters increase.
+The number tours $T$ indicates that the entire complex operation is executed $T$ times, further amplifying the effect of the inner complexities.
+
+Optimizing such an algorithm involves minimizing the repetition count ($T$), reducing the complexity of operations dependent on $K$, $m$, and $n$.
+
+\subsubsection{Time Complexity Analysis of the Genetic Algorithm for TSP}
+\label{ga_tsp}
+
+When $K > 6$, we obtain 100 tours from a heuristic TSP solution with the Genetic Algorithm (GA), which addresses optimization problems by evolving a population of candidate solutions through selection, crossover (recombination), and mutation over multiple generations, aiming to improve the fitness of solutions.
+
+The time complexity of applying Genetic Algorithms to solve the TSP is influenced by several key factors, including the number of nodes (\(K\)), population size (\(P\)), number of generations (\(G\)), and the complexities of the crossover (\(C(K)\)), mutation (\(M(K)\)), and selection (\(S(K)\)) operations.
+
+Given these parameters, we can express the overall time complexity as:
+\begin{itemize}
+	\item $O(G \cdot (P \cdot (S(K) + P \cdot (C(K) + M(K)))))$
+\end{itemize}
+
+{\bf Detailed Explanation}
+\begin{itemize}
+	\item The selection operation's complexity, \(S(K)\), may depend on evaluating the fitness of each solution, which in turn relies on computing the total distance of a path through \(K\) nodes.
+	\item The complexities of the crossover and mutation operations, \(C(K)\) and \(M(K)\), are functions of \(K\) because they involve manipulating sequences representing paths through the \(K\) nodes.
+	\item Each generation involves selecting solutions based on fitness, generating new solutions through crossover, and applying mutations, leading to the multiplication by \(G\).
+\end{itemize}
+
+{\bf Impact of the number of nodes (\(K\))}
+\begin{itemize}
+	\item The direct involvement of \(K\) in the complexities of \(C(K)\), \(M(K)\), and \(S(K)\) highlights how the number of nodes affects the algorithm's performance.
+Specifically, operations that manipulate or evaluate paths become more complex as \(K\) increases.
+	\item The encoding of solutions and the calculation of fitness values, essential for the selection process, inherently depend on \(K\), emphasizing its role in the GA's computational demands.
+\end{itemize}
+
 }
@@
 		\toprule
-		$x$ & $P$ & $low$ ($kg$) & $high$ ($kg$) \\		
+		$item$ & $P$ & $low$ ($kg$) & $high$ ($kg$) \\		
 		\midrule
@@
 
-\vspace{-2mm}
-
-\begin{itemize}
-	\setlength\itemsep{-0.3em}
+\begin{itemize}
+
 	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$, where $i_1$ and $i_2$ are integer numbers;
-{\color{violet}
-\item $Roulette()$: a proportional selection function that is biased through $P$ (Table \ref{tab:weights}).
-}
+
+	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
+
 \end{itemize}
@@
 		\Until{$to_j \neq k$}
-		\State $x = Roulette()$ \Comment{biased through $P$ ({\it cfr.} Table \ref{tab:weights})} \label{ig:weight1} 
+		\State $x = Roulette(item)$ biased through $P$ \label{ig:weight1} \Comment{From Table \ref{tab:weights}}	 
 		\State $w_j \gets RandomInt(low(x), high(x))$        \label{ig:weight2}		
@@
 
-{\color{blue}
-	After the items were generated, some experiments with the \textit{irace} tool were executed to determine the values for parameters $threshold_1$ and $threshold_2$ used in \textit{Shims}, whose results are presented in Table \ref{tab:irace}.
+To determine the parameters $level_1$\/ and $level_2$, used by \textit{Shims}, we previously carried out some experiments with the tool {\it irace}\/ \citep{LopezIbanezManuel2016}, the results of which are presented in Table \ref {tab:irace}.
+In these tests, of every 7 instances generated for each value of $surplus$, 4 were used as the training set and 3 as the testing set.
+We provided the ranges [0.8, 1.0] and [1.0, 2.0] for $level_1$\/ and $level_2$, respectively.
+In each experiment, there was a maximum of 3,000 runs so that \textit{irace}\/ would have enough data for its statistical tests.
+For more details, see {\tt cran.r-project.org/web/packages/irace/}.
 
@@
 	\caption{\textit{irace} testing results}  \label{tab:irace}
-	\footnotesize
 	\begin{tabular}{cccc}
@@
 
-		\boldm{$surplus$} & \boldm{$threshold_1$} & \boldm{$threshold_2$} & \textbf{run time}\\
-		& for $\eta_1$          & for $\eta_2$          & (min) \\
+		$surplus$ & $level_1$ & $level_2$ & runtime (min)\\
+
 		\midrule
@@
 
-	For these tests, we considered the odd instances (1, 3, 5, 7) as the \textbf{training} set and the even instances (2, 4, 6) as the \textbf{testing} set for the \textit{irace} runs.
-We supplied \textit{irace} for $threshold_1$ determination, the range [0.8, 1.0], and for $threshold_2$, the range [1.0, 2.0].
-The number of iterations in each experiment was limited to 3000 executions for \textit{irace} to have plenty of data for its statistical tests.
-More detail about these configurations may be found on the \textit{irace} user guide in {\tt cran.r-project.org/web/packages/irace/irace.pdf}.
-}
-
 \subsection{Obtained solutions: quality and runtimes}
@@
 In the tests performed, we used a 64-bit, 16 GB, 3.6 GHz, eight-core processor with {\it Linux Ubuntu 22.04.1 LTS 64-bit} as the operational system and {\it Python 3.10.4}\/ as the programming language.
-We also used the well-known solver $Gurobi$\/ ({\tt www.gurobi.com}), version 9.5.2.
-
-We ran Algorithm \ref{alg:main} considering the five scenarios from Table \ref{tab:scenarios}, three values for  $surplus$\/ from $\{1.2, 1.5, 2.0\}$, four values for $tmax$\/ from $\{240s, 1200s, 2400s, 3600s\}$, and seven different methods for the node-by-node solution: $Gurobi$\/ (see \ref{solver}), ACO, NMO, TS, GRASP, GA and $Shims$\/ (Algorithm \ref{alg:shims}).
-
-For $Gurobi$\/ to be able to solve the largest possible number of tests without memory overflow, we set its parameter {\it MIPgap}\/ to 1\%.
+We also used the well-known solver {\it Gurobi}\/ ({\tt www.gurobi.com}), version 9.5.2.
+
+We will first present the results obtained in practical cases, that is, when $1<K\leq6$. Next, we will show how the same {\it Shims}\/ heuristic remains robust when $K>6$.
+
+\subsubsection{Results when $1<K\leq6$}
+
+We ran Algorithm \ref{alg:main} considering the 5 scenarios from Table \ref{tab:scenarios}, 3 values for  $surplus$\/ from $\{1.2, 1.5, 2.0\}$, 4 values for $tmax$\/ from $\{240s, 1200s, 2400s, 3600s\}$, and 7 different methods for the node-by-node solution: {\it Gurobi}\/ (see \ref{solver}), ACO, NMO, TS, GRASP, GA and {\it Shims}\/ (Algorithm \ref{alg:shims}).
+
+For {\it Gurobi}\/ to be able to solve the largest possible number of tests without memory overflow, we set its parameter {\it MIPgap}\/ to 1\%.
 This shortens its runtime, in addition to ensuring that its objective function $f$\/ is at most 1\% of the optimal solution.
@@
 
+For each $scenario$, $surplus$\/ and $tmax$\/ tested, 7 different instances were generated.
+Table \ref{tab:overall} succinctly shows the overall performance of the methods for the node-by-node solution.
+Only {\it Shims}\/ found a solution for all scenarios, as well as being the fastest.
+
 \begin{table}[H]
@@
 	\centering
-	\caption{ \color{blue}Overall results}  \label{tab:mr}
-	\footnotesize
+	\caption{Overall results}  \label{tab:overall}
+	\small
 	\begin{tabular}{cccc}	
 	\toprule
-	{\bf Method} & {\bf Best scenarios} & {\bf Worst scenarios} & {\bf Worst run times (min)}  \\
-	\midrule
-	NMO           & 4             & 5    & {\color{red}60}  \\
-	ACO           & 2, 3          & 4, 5 & {\color{red}25}, {\color{red}61}  \\
-	GRASP         & 1             & 4, 5 & {\color{red}28}, {\color{red}55}  \\
-	TS            & -             & 5    & {\color{red}44}  \\
-	GA            & -             & 1,2,3,4,5  & {\color{red}did not solve}  \\
-	\midrule
-	{\it Gurobi}  & 1, 2, 3, 4    & 5    & {\color{red}did not solve}  \\
-	{\it Shims}   & 1, 2, 3, 4, 5 & -    & {\bf 3.26} \\
+	{\bf Method} & {\bf Best scenarios} & {\bf Worst scenarios} & {\bf Worst runtimes (min)}  \\
+	\midrule
+	NMO           & 4       & 5    & 60  \\
+	ACO           & 2, 3    & 4, 5 & 25, 61  \\
+	GRASP         & 1       & 4, 5 & 28, 55  \\
+	TS            & -       & 5    & 44  \\
+	GA            & -       & 1, 2, 3, 4, 5  & did not solve  \\
+	\midrule
+	{\it Gurobi}  & 1, 2, 3, 4    & 5    & did not solve  \\
+	{\it Shims}   & 1, 2, 3, 4, 5 & -    & 3.26 \\
 	\bottomrule
@@
 
-{\color{violet}
-
-As to the genetic algorithm exclusion from the comparisons tests, we must consider that:
-\vspace{-2mm}
-
-\begin{itemize}
-	\setlength\itemsep{-0.3em}	
-	\item The palletization planning of 200 to 1000 items per node in a set of 18 aircraft pallets may be an inadequate problem to be solved with GA, because large chromosome sizes, like 18,000-dimension arrays, can pose significant challenges for each generation to evaluate individual fitness, which demonstrated to be resource-intensive.
-	\item Large chromosomes can slow GA convergence due to the vast search space, requiring more generations to find optimal solutions.
-	\item Efficient encoding and decoding of the problem into the chromosome and decoding solutions is crucial for a 18,000-dimension array. 
-	\item Also, the design of the fitness function becomes more critical as chromosome size increases, as it needs to guide the evolutionary process without being overly computationally demanding.
-\end{itemize}
-
-}
-
-\subsubsection{Shims results}
-
-Among the heuristics, we will present only the results obtained by $Shims$, which had the best performance.
-For each $scenario$, $surplus$\/ and $tmax$, seven different instances were generated.
-We will present the average of the objective function $f$\/ and the runtime of $Gurobi$\/ and $Shims$\/ in these instances.
+For this reason, we will present only the results obtained by {\it Shims}, with the average of the objective function $f$\/ and the runtime of {\it Gurobi}\/ and {\it Shims}.
 To facilitate the comparison between both, we added a last column in the tables where two values are indicated:
 
-\vspace{-2mm}
-
-\begin{itemize}
-	\setlength\itemsep{-0.3em}
+\begin{itemize}
+
 	\item {\bf Normalized}: value between 0 and 1, which corresponds to the ratio between the sum of $f$\/ values obtained by the method in all scenarios and the sum of the best values obtained by both methods in all scenarios.
 The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
+
 	\item {\bf Speed-up}: ratio of the sums of the runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
-\end{itemize}
-
-We also indicate the two adopted strategies: dedicating all the processing time to the two shortest tours or distributing it among all $K!$\/ tours.
-
+
+\end{itemize}
+
+We also indicate the two adopted strategies: dedicating all the processing time to the 2 shortest tours or distributing it among all $K!$\/ tours.
 The results obtained with $tmax = 3600s$, which is the highest tested runtime limit, are in Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, with $surplus$\/ values of $1.2$, $1.5$\/ and $2.0$, respectively.
-We indicate with an {\bf x} the cases where $Gurobi$\/ did not find a feasible solution within this runtime limit or had to be aborted due to high random-access memory (RAM) usage.
-
-\begin{table}[htp] % update this
+We indicate with an {\bf x} the cases where {\it Gurobi}\/ did not find a feasible solution within this runtime limit or had to be aborted due to high random-access memory (RAM) usage.
+
+\begin{table}[htp] 
 	\centering
@@
 		\toprule
-		{\specialcell{{\bf Tested}\\{\bf tours}}}&$method$&$scenario$&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&\specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
-		\toprule
-		\multirow{4}{*}{Shortest} &\multirow{2}{*}{$Gurobi$}  & $f$ & 8.53    & 11.79   & 13.14   & 13.52 & 12.36       &  0.9998 \\ % best 5935
+		$ntours$&$method$&$scenario$&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&\specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		\toprule
+		\multirow{4}{*}{2} &\multirow{2}{*}{{\it Gurobi}}  & $f$ & 8.53    & 11.79   & 13.14   & 13.52 & 12.36       &  0.9998 \\ % best 5935
 		&                              &               {time (s)}   & 29      & 28      & 25      & 27    & 27          &  1.0    \\ % longer 136
 		\cmidrule(lr){2-9}
-		&                           \multirow{2}{*}{$Shims$}  & $f$ & 8.54    & 11.78   & 13.06   & 13.51 & 12.34       &  0.9980 \\
+		&                           \multirow{2}{*}{{\it Shims}}  & $f$ & 8.54    & 11.78   & 13.06   & 13.51 & 12.34       &  0.9980 \\
 		&                             &                {time (s)}   & 1       & 1       & 1       & 1     & 2           &  22.7   \\
@@
 
-		\multirow{4}{*}{All $K!$}&  \multirow{2}{*}{$Gurobi$} & $f$ & 8.60    & 12.20   & 13.66   & 15.00  & {\bf x}    & 0.9998 \\ % best 4947
+		\multirow{4}{*}{$K!$}&  \multirow{2}{*}{{\it Gurobi}} & $f$ & 8.60    & 12.20   & 13.66   & 15.00  & {\bf x}    & 0.9998 \\ % best 4947
 		&                              &               {time (s)}   & 30      & 35      & 123     & 314    & {\bf x}    & 1.0 \\
 		\cmidrule(lr){2-9}	
-		&                           \multirow{2}{*}{$Shims$}  & $f$ & 8.61    & 12.20   & 13.46   & 14.99  & 13.35      & 0.9958  \\
+		&                           \multirow{2}{*}{{\it Shims}}  & $f$ & 8.61    & 12.20   & 13.46   & 14.99  & 13.35      & 0.9958  \\
 		&                              &               {time (s)}   & 1       & 1       & 4       & 10     & 51         & 7.49   \\
@@
 		\toprule
-		{\specialcell{{\bf Tested}\\{\bf tours}}} & $method$          & $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
-		\toprule
-		\multirow{4}{*}{ Shortest } & \multirow{2}{*}{$Gurobi$} & $f$ & 11.83 & 16.73 & 18.07 & 18.83 & 16.86 & 0.9996 \\ % best 8235
+		$ntours$ & $method$          & $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		\toprule
+		\multirow{4}{*}{ 2 } & \multirow{2}{*}{{\it Gurobi}} & $f$ & 11.83 & 16.73 & 18.07 & 18.83 & 16.86 & 0.9996 \\ % best 8235
 		&&               {time (s)}                                   & 55    & 64    & 39    & 40    & 88    & 1.0    \\ 
 		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{$Shims$}                               & $f$ & 11.85 & 16.72 & 18.05 & 18.80 & 16.87 & 0.9993  \\
+		&\multirow{2}{*}{{\it Shims}}                               & $f$ & 11.85 & 16.72 & 18.05 & 18.80 & 16.87 & 0.9993  \\
 		&&         {time (s)}                                         & 1     & 1     & 2     & 2     & 2     & 35.8  \\
@@
 
-		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$}   & $f$ & 11.83 & 16.93 & 18.40 & 20.95 & 17.60 & 0.9999 \\  % best 8469
+		\multirow{4}{*}{$K!$} & \multirow{2}{*}{{\it Gurobi}}   & $f$ & 11.83 & 16.93 & 18.40 & 20.95 & 17.60 & 0.9999 \\  % best 8469
 		&&               {time (s)}                                   & 63    & 59    & 195   & 472   & 2258  & 1.0  \\
 		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{$Shims$}                               & $f$ & 11.85 & 16.91 & 18.36 & 20.93 & 17.50 & 0.9976  \\ 
+		&\multirow{2}{*}{{\it Shims}}                               & $f$ & 11.85 & 16.91 & 18.36 & 20.93 & 17.50 & 0.9976  \\ 
 		&&         {time (s)}                                         & 1     & 2     & 5     & 15    & 100   & 23.8 \\
@@
 		\toprule
-		{\specialcell{{\bf Tested}\\{\bf tours}}}& $method$& $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
-		\toprule
-		\multirow{4}{*}{Shortest} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.70 & 24.20 & 26.39 & 27.17 & 24.20 & 0.9995 \\ % best 11972
+		$ntours$& $method$& $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
+		\toprule
+		\multirow{4}{*}{2} & \multirow{2}{*}{{\it Gurobi}} & $f$ & 17.70 & 24.20 & 26.39 & 27.17 & 24.20 & 0.9995 \\ % best 11972
 		&&                                              {time (s)}  & 168   & 98    & 79    & 70    & 72    & 1.0  \\
 		\cmidrule(lr){2-9}	
-		&                            \multirow{2}{*}{$Shims$} & $f$ & 17.74 & 24.22 & 26.32 & 27.07 & 23.13 & 0.9896  \\
+		&                            \multirow{2}{*}{{\it Shims}} & $f$ & 17.74 & 24.22 & 26.32 & 27.07 & 23.13 & 0.9896  \\
 		&&                                              {time (s)}  & 1     & 2     & 2     & 3     & 4     & 40.6  \\
@@
 
-		\multirow{4}{*}{All $K!$} & \multirow{2}{*}{$Gurobi$} & $f$ & 17.90 & 25.44 & 26.51 & 29.13 &{\bf x}& 0.9994 \\ % best 9867
+		\multirow{4}{*}{$K!$} & \multirow{2}{*}{{\it Gurobi}} & $f$ & 17.90 & 25.44 & 26.51 & 29.13 &{\bf x}& 0.9994 \\ % best 9867
 		&&                                               {time (s)} & 178   & 143   & 378   & 862   &{\bf x}& 1.0  \\
 		\cmidrule(lr){2-9}	
-		&                         \multirow{2}{*}{$Shims$}    & $f$ & 17.94 & 25.45 & 26.44 & 28.84 & 26.22  & 0.9970  \\
+		&                         \multirow{2}{*}{{\it Shims}}    & $f$ & 17.94 & 25.45 & 26.44 & 28.84 & 26.22  & 0.9970  \\
 		&&                                               {time (s)} & 1     & 3     & 10    & 31    & 196    & 34.7  \\
@@
 
-\vspace{-2mm}
-
-\begin{itemize}
-	\setlength\itemsep{-0.3em}
+\begin{itemize}
+
 	\item The strategy of testing all $K!$\/ tours often provide a better-quality solution, even with less time on each node.
 This shows that the four sub-problems are interconnected in such a way that it is not enough to solve them separately.
-	\item $Gurobi$\/ fails in some cases when $scenario=5$\/ and the strategy is to check all $K!$\/ tours.
+
+	\item {\it Gurobi}\/ fails in some cases when $scenario=5$\/ and the strategy is to check all $K!$\/ tours.
 This occurs because the runtime limit per node is smaller and there tend to be more packed contents on the aircraft, reducing the space for allocating items and making the solution difficult.
-	\item When $Gurobi$\/ finishes, it finds the best solution, but the one obtained by $Shims$\/ reaches {\color{blue}$99.9\%$}\/ of that value.
-	\item $Shims$\/ always finds a solution, being {\color{blue}8 to 41} times faster.
+
+	\item When {\it Gurobi}\/ finishes, it finds the best solution, but the one obtained by {\it Shims}\/ reaches at least $98.96\%$\/ of that value.
+Considering only the strategy of testing all $K!$ tours, this value increases to $99.58\%$.
+
+	\item {\it Shims}\/ always finds a solution, being 7 to 40 times faster.
+
 	\item All runtimes are much lower than the limit because the solution on many nodes can be fast.
-Anyway, in all the tests performed, the maximum time spent by $Shims$\/ did not reach 4 minutes.
-On the other hand, when $scenario=5$\/ and $surplus=1.5$, $Gurobi$\/ spent almost 40 minutes.
+Anyway, in all the tests performed, the maximum time spent by {\it Shims}\/ did not reach 4 minutes.
+On the other hand, when $scenario=5$\/ and $surplus=1.5$, {\it Gurobi}\/ spent almost 40 minutes.
+
 \end{itemize}
@@
 Table \ref{tab:100time} shows the results obtained with the strategy of testing the $K!$\/ tours in all scenarios with different $tmax$.
-We can observe more cases where $Gurobi$\/ fails, even in smaller scenarios.
-When $Gurobi$\/ finishes, $Shims$\/ finds a solution of similar quality ($99\%$\/ or better).
-In all cases, $Shims$\/ finds a good solution in less than 4 minutes.
+We can observe more cases where {\it Gurobi}\/ fails, even in smaller scenarios.
+When {\it Gurobi}\/ finishes, {\it Shims}\/ finds a solution of similar quality ($99\%$\/ or better).
+In all cases, {\it Shims}\/ finds a solution in less than 4 minutes.
 
@@
 
-		\multirow{10}{*}{$Gurobi$} & \multirow{2}{*}{240s} & $f$ & 8.60 & 12.20 & 13.67 & {\bf x} & {\bf x} & 11.77 & 16.38 & 18.10 & {\bf x} & {\bf x} & 17.89 & 25.42 & {\bf x} & {\bf x} & {\bf x} \\ 
+		\multirow{10}{*}{{\it Gurobi}} & \multirow{2}{*}{240s} & $f$ & 8.60 & 12.20 & 13.67 & {\bf x} & {\bf x} & 11.77 & 16.38 & 18.10 & {\bf x} & {\bf x} & 17.89 & 25.42 & {\bf x} & {\bf x} & {\bf x} \\ 
 		&               &  {time (s)}                            & 31   & 35    & 124   & {\bf x} & {\bf x} & 52    & 59    & 200   & {\bf x} & {\bf x} & 188   & 145   & {\bf x} & {\bf x} & {\bf x} \\ 
@@
 
-		\multirow{2}{*}{$Shims$} &  \multirow{2}{*}{240s}  & $f$ & 8.61 & 12.20 & 13.46 & 14.99   & 13.35   & 11.78 & 16.29 & 17.99 &  20.93  & 17.50   & 17.94 & 25.45 & 26.14 & 28.84   & 26.22 \\ % 
+		\multirow{2}{*}{{\it Shims}} &  \multirow{2}{*}{240s}  & $f$ & 8.61 & 12.20 & 13.46 & 14.99   & 13.35   & 11.78 & 16.29 & 17.99 &  20.93  & 17.50   & 17.94 & 25.45 & 26.14 & 28.84   & 26.22 \\ % 
 		&         & {time (s)}                                   & 1    & 2     & 4     & 10      & 51      & 1     & 2     & 5     &  15     & 100     & 1     & 3     & 10    & 31      & 196   \\
@@
 
-The actual RAM consumption of $Gurobi$\/ was over 8.5 GB, and all of $Shims$\/'s executions consumed at most 1.5 GB of RAM.
-
-{\color{blue}
-\subsection{{\it Shims} validation with more nodes}
-
-To validate the effectiveness of the {\it Shims} heuristic, we solve a 15-node TSP containing the 15 main Brazilian airports.
-For this task, we implemented a TSP solution method based on the Genetic Algorithm (GA) that returns a population of 100 tours as input to the ACLP-RPDP solved by {\it Shims}.
-The GA run time was around three seconds, and the total run time was around 33 seconds.
-
-We implemented the GA with DEAP (Distributed Evolutionary Algorithms in Python) , a evolutionary computation framework available in {\tt github.com/deap/deap}.
-More details can be found in \citep{DEAP_JMLR2012}.
-
-	\begin{figure}[H]
-		\centering
-		\caption{\label{fig:brazil} {\it Shims/GA} best ACLP+RPDP tour \\
-			\tiny{Solution for scenario 7, instance 1}}
-		\fbox{\includegraphics[scale=0.30]{brazil.png}}
-	\end{figure}
-
-The airports cited in Figure \ref{fig:brazil} are: São Paulo (GRU), Congonhas (CGH), Curitiba (CWB), Porto Alegre (POA), Florianópolis (FLN), Vitória (VIX), Salvador (SSA), Recife (REC), Fortaleza (FOR), Belém (BEL), Manaus (MAO), Brasília (BSB), Goiânia (GYN), Confins (CNF), and Rio de Janeiro (GIG), returning to GRU.
-
-It is important to remember that this is not the shortest tour, but the one with the best relationship with score, cost, and fuel consumption.
-
-}
-
-{\color{purple}
-
-	It is also important to highlight that three strategies were essential for us to obtain a solution in the instances created:
-
-	\vspace{-2mm}
-
-	\begin{itemize}
-		\setlength\itemsep{-0.3em}
-		\item[(1)] consider $K<m$\ so that each pallet received a predefined destination;
-		\item[(2)] on each tour, calculate a node-by-node solution; and
-		\item[(3)] at each intermediate node of the considered tour, relocate the packed contents, minimizing the torque on the aircraft.
-	\end{itemize}
-
-	\vspace{-2mm}
-
-	Without adopting these three strategies, we were unable to carry out any experiments, even in smaller instances.
-}
+The actual RAM consumption of {\it Gurobi}\/ was over 8.5 GB, and all of {\it Shims}\/'s executions consumed at most 1.5 GB of RAM.
+
+\subsubsection{Results when $K>6$}
+
+These last results that we will show here do not correspond to practical cases of air transport, as tours where $K>6$ very rarely occur.
+However, it is possible to see that the {\it Shims} heuristic maintains robust behaviour as the number of nodes grows, that is, it could be adapted to similar contexts
+({\color{red}vehicles and warehouses} {\color{blue}ships and trucks})
+, for example), where there may be more nodes.
+
+In the tests carried out, we considered real data from the 15 main Brazilian airports.
+For this task, we implemented a TSP heuristic based on GA that returned 100 tours in approximately 33 seconds.
+We implemented this heuristic with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework available in {\tt github.com/deap/deap}.
+For more details, see \citep{DEAP_JMLR2012}.
+
+Figure \ref{fig:performance} shows the runtime curve of {\it Shims}\/ as the number of nodes increases, indicating at each point the value found for the objective function $f$.
+Runtime and $f$\/ are the averages obtained from 7 instances generated with $surplus=2.0$\/ for each value of $K$.
+In Figure \ref{fig:morenodes}, we indicate one of the tours found by the TSP heuristic when $K=15$.
 
 \begin{table}[htp]
-	\centering
-	\caption{Shims solutions with $surplus = 1.2$\/ and $tmax =  240s$}  \label{tab:15}
+
+	\begin{minipage}{0.58\linewidth}
+	\centering
+	\begin{tikzpicture}
 	\footnotesize
-	\begin{tabular}{cccc}
-		\toprule
-$scenario$ & {\bf 7} & {\bf 8} & {\bf 9} \\
-Num. nodes & 15 & 14 & 13 \\
-		\toprule
-$f$        & 50.15 & 47.00 & 47.18    \\
-{time (s)} & 32     & 31     & 29        \\
-Occupancy rate & 0.89     & 0.84     & 0.88        \\
-		\bottomrule
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\begin{center}
-	%\fbox{
-	\begin{tikzpicture}
 		\begin{axis}[xmin=6,ymin=200,
@@
 			xlabel style={align=center},
-			xlabel={Number of nodes},
-			ylabel={Time (s)},
+			xlabel={Number $K$\/ of nodes},
+			ylabel={Runtime (s)},
 			xtick={ 5, 6, 7,  8,  9, 10, 11, 12, 13, 14, 15, 16},
@@
 				x   y    label                  alignment
-				7   273   {\footnotesize 92.6}  -10
-				8   360   {\footnotesize 107.5} -10
-				9   468   {\footnotesize 107.5} -10
-				10  560   {\footnotesize 112.3} -10
-				11  660   {\footnotesize 92.8}  -10
-				12  756   {\footnotesize 91.2}  -10
-				13  884   {\footnotesize 96.8}  -10
-				14  1008   {\footnotesize 106.3} -10									
-				15  1170   {\footnotesize 99.8}  -10									
+				7   273   {\footnotesize 92.6}  0
+				8   360   {\footnotesize 107.5} 0
+				9   468   {\footnotesize 107.5}  0
+				10  560   {\footnotesize 112.3}  0
+				11  660   {\footnotesize 92.8}  0
+				12  756   {\footnotesize 91.2} 0
+				13  884   {\footnotesize 96.8}  0
+				14  1008   {\footnotesize 106.3} 0									
+				15  1170   {\footnotesize 99.8}  0									
 			};
@@
 	\end{tikzpicture}
-	%}
-	\captionof{figure}{Shims performance}
-\end{center}
-
-\begin{center}
-	%\fbox{
-		\begin{tikzpicture}
-			\begin{axis}[xmin=6,ymin=80,
-				enlarge y limits={upper,value=0.2},
-				xlabel style={align=center},
-				xlabel={Number of nodes},
-				ylabel={Time (s)},
-				xtick={  5,  6,   7,   8,   9,  10,  11,  12,  13,  14,  15,  16},
-				ytick={80,100,120,140,160,180,200,220,240,260,280,300},						
-				]
-				\addplot[
-				blue,
-				thick,
-				mark=*,
-				mark options={fill=blue},
-				visualization depends on=\thisrow{alignment} \as \alignment,
-				nodes near coords, % Place nodes near each coordinate
-				point meta=explicit symbolic, % The meta data used in the nodes is not explicitly provided and not numeric
-				every node near coord/.style={anchor=\alignment} % Align each coordinate at the anchor 40 degrees clockwise from the right edge
-				] table [% Provide data as a table
-				meta index=2 % the meta data is found in the third column
-				] {
-					x   y    label                 alignment
-					7   99  {\footnotesize 91.9}  -10
-					8   126  {\footnotesize 108.2} -10
-					9   145  {\footnotesize 108.7} -10
-					10  168 {\footnotesize 111.2} -10
-					11  191 {\footnotesize 78.0}  -10
-					12  215 {\footnotesize 84.1}  -10
-					13  239 {\footnotesize 93.3}  -10
-					14  264 {\footnotesize 104.8} -10									
-					15  293 {\footnotesize 96.1}  -10									
-				};
-			\end{axis}
-		\end{tikzpicture}
-		%}
-	\captionof{figure}{Shims performance (total time)}
-\end{center}
+	\captionof{figure}{{\it Shims}\/ performance with $surplus=2.0$}
+	\label{fig:performance}
+	\end{minipage}\hfill 
+	\begin{minipage}{0.42\linewidth}
+		\centering
+		\includegraphics[scale=0.293]{brazil.png}
+		\captionof{figure}{A tour where $K=15$}
+		\label{fig:morenodes}
+	\end{minipage}
+\end{table}
 
@@
 
-{\color{blue}
 We adopted some simplifications that are not critical, but that allowed for an unprecedented solution to this problem considering more than two nodes.
@@
 Considering a real aircraft, we have developed some node-by-node solutions.
-The complete process can be executed quickly on a handheld computer, offering good results and reducing stress for the transport planners.}
+The complete process can be executed quickly on a handheld computer, offering good results and reducing stress for the transport planners.
 

### 2024-02-04 latex/TSP_complexity.tex
Time complexity

\documentclass{article}
\usepackage{amsmath}

\begin{document}

	\title{Revised Time Complexity Analysis of Solving TSP with Genetic Algorithms Including the Number of Cities}
	\author{}
	\date{}
	\maketitle

	\section{Introduction}
	The Traveling Salesman Problem (TSP) seeks the shortest path visiting each of a set of cities exactly once and returning to the origin city. Genetic Algorithms (GAs) offer a heuristic approach to find approximate solutions for the TSP by mimicking natural evolutionary processes.

	\section{Genetic Algorithm Overview}
	Genetic Algorithms address optimization problems by evolving a population of candidate solutions through selection, crossover (recombination), and mutation over multiple generations, aiming to improve the fitness of solutions.

	\section{Time Complexity Analysis Including the Number of Cities (\(N\))}
	The time complexity of applying Genetic Algorithms to solve the TSP is influenced by several key factors, including the number of cities (\(N\)), population size (\(P\)), number of generations (\(G\)), and the complexities of the crossover (\(C(N)\)), mutation (\(M(N)\)), and selection (\(S(N)\)) operations.

	Given these parameters, we can express the overall time complexity as:
	\begin{equation}
		O(G \cdot (P \cdot (S(N) + P \cdot (C(N) + M(N)))))
	\end{equation}

	\subsection{Detailed Explanation}
	\begin{itemize}
		\item The selection operation's complexity, \(S(N)\), may depend on evaluating the fitness of each solution, which in turn relies on computing the total distance of a path through \(N\) cities.
		\item The complexities of the crossover and mutation operations, \(C(N)\) and \(M(N)\), are functions of \(N\) because they involve manipulating sequences representing paths through the \(N\) cities.
		\item Each generation involves selecting solutions based on fitness, generating new solutions through crossover, and applying mutations, leading to the multiplication by \(G\).
	\end{itemize}

	\section{Impact of the Number of Cities (\(N\))}
	\begin{itemize}
		\item The direct involvement of \(N\) in the complexities of \(C(N)\), \(M(N)\), and \(S(N)\) highlights how the number of cities affects the algorithm's performance. Specifically, operations that manipulate or evaluate paths become more complex as \(N\) increases.
		\item The encoding of solutions and the calculation of fitness values, essential for the selection process, inherently depend on \(N\), emphasizing its role in the GA's computational demands.
	\end{itemize}

	\section{Conclusion}
	Incorporating \(N\) into the time complexity analysis of Genetic Algorithms for solving the TSP provides a more comprehensive understanding of the factors influencing the algorithm's efficiency. The revised notation underscores the significance of the number of cities, revealing its impact on the complexity of selection, crossover, and mutation operations. This analysis demonstrates the importance of considering \(N\) when evaluating the performance and scalability of Genetic Algorithms in the context of the TSP.

\end{document}

### 2024-02-04 latex/responses.tex
Time complexity

--- 2024-01-28 latex/responses.tex
@@
+% !TeX spellcheck = en_GB
 \documentclass{article}
@@
 
-Agradecemos e seguimos todas as orientações do Editor:
+We appreciate and comply with all the Editor's instructions::
 
 \begin{itemize}
-\item Incluímos neste texto todas as {\bf Required Sections};
+\item We have included all {\bf Required Sections} in this text, responding individually to each of the reviewers' comments.
 
-\item A submissão do nosso artigo revisado atende todas as {\bf Compliance Requirements}.
+\item Submission of this revised article meets all {\bf Compliance Requirements}. Regarding {\it Reference Style}, we use {\tt bibliographystyle\{apalike\}},\\ available in \LaTeX.
 
@@
 
-De acordo com a mensagem recebida, parece-nos que os comentários do Associate Editor são os seguintes:
+According to the message received, it appears that the Associate Editor's comments are as follows:
 
@@
 \item [1.] More analysis shall be given to discuss the experimental results.
-\item [2.] More newly-published related references shall be cited and dicussed in Introduction.
+\item [2.] More newly-published related references shall be cited and discussed in Introduction.
 \item [3.] More experiments hope to be added to support the conclusion.
@@
 	\begin{minipage}{10cm}
-		\textit{Como pode ser observado logo abaixo, aumentamos substancialmente os dados experimentais apresentados. Consequentemente, houve também um sensível aumento na discussão sobre eles.}
-
-		{\color{blue}As can be seen below, we have substantially increased the experimental data presented.
-Consequently, there was also a noticeable increase in discussion about them.
-Especially about the overall improvement caused by the run time limit proportional to the volumes of items in each node and the additional experiments to validate the best method found.
-		}
+		\textit{As we substantially increased the experimental data presented in the article (see the third answer in this section), there were also greater discussions about the results obtained.
+In any case, we emphasise that we are willing to clarify any result that is not sufficiently discussed; to do this, reviewers simply need to explicitly indicate what they are about.}
 	\end{minipage}
@@
 \paragraph{Revision:}
-More newly-published related references shall be cited and dicussed in Introduction.
+More newly-published related references shall be cited and discussed in Introduction.
 \begin{center}
 	\begin{minipage}{10cm}
-		\textit{Incluímos e comentamos novas referências (procurar trabalhos recentes sobre TSP e PDP em aeronaves).}
-
-		{\color{blue}
-		We included the work of \citep{zhao2023}, that developed an air cargo loading plan that addresses the Air Cargo Palletization Problem (ACPP) and the Aircraft Weight and Balance Problem (WBP) separately, impacting payload optimization and aircraft center of gravity (CG).
-The authors propose three integer linear programming models: a bi-objective optimization model (BOM), a combinatorial optimization model (COM), and an improved combinatorial optimization model (IOM).
-The models aim to determine the maximum loading capacity and lowest CG deviation from a specified target CG.
-
-		The study uses Gurobi to solve four scenarios with various conditional metrics for the three models.
-Results show that the BOM has the fastest solution speed, but the CG deviation is the largest, making the results unacceptable in some cases.
-The COM takes too long to solve, making it difficult to meet the needs of the actual scene.
-The IOM accelerates the solving process, but some scenarios still require significant computation time.
-		}
+		\textit{In Section 2, we highlight the most recent references and concisely discuss their main aspects.
+To the best of our knowledge, we have not failed to cite any recent and relevant publications related to the topic of this work.
+This can be confirmed through the bibliography of \citep{zhao2023}, published less than a year ago.}
 
@@
 	\begin{minipage}{10cm}
-		\textit{Incluímos novos resultados experimentais (novas figuras, testes, etc.).
-Concretamente: desempenhos obtidos com outras metaheurísticas; definição empírica de $\eta_1$\/ and $\eta_2$\/ no algoritmo {\it Shims}; limite de tempo variável em cada nó, de acordo com o volume de carga previsto em cada nó; resolução de casos com um maior número de nós}.
+		\textit{We substantially increased the experimental data: practical determination of Shims parameters (Table 10); distribution of the runtime among nodes in proportion to the volume available for shipment; performance comparison of five meta-heuristics in the node-by-node solution (Table 11); application and results of Shims in contexts with a greater number of nodes (subsubsection 6.2.2).
+If reviewers request any more specific experiments, we are willing to perform them.}
 
-{\color{blue}
-	New experiments were conducted due to two changes suggested by the reviewers:
-
-	\begin{itemize}
-		\item [1.] Node solution times depended on the total volume of candidate items to be loaded in each node in relation to the total items volume to be loaded in the tour.
-		\item [2.] A new process of determining the {\it Shims} volume thresholds for $eta_1$ and $eta_2$.
-	\end{itemize}
-
-	We included Table 10 with the results of the {\it iRace} tool, a software package that helps to automatically fine-tune algorithm parameters.
-We fine-tuned parameters $threshold_1$\/ and $threshold_2$\/, the relative volumes of each pallet used to determine the indexes  $\eta_1$\/ and $\eta_2$ for the {\it Shims} heuristic.
-}
 	\end{minipage}
@@
 
-Indicamos a seguir, em cada subseção, os comentários feitos pelos três revisores.
+We indicate below, in each subsection, the comments made by the three reviewers.
 
@@
 
-Em cada subseção, responderemos os comentários de cada um dos três revisores.
+In each subsection, we will respond to comments from each of the three reviewers.
 
@@
 	\begin{minipage}{10cm}
-		\textit{Aumentamos a revisão da literatura, descrevendo os trabalhos com maior detalhe. Acrescentamos também alguns trabalhos mais recentes. Citar quais são esses trabalhos...}
-
-		{\color{blue}
-	We included the work of \citep{zhao2023}, that developed an air cargo loading plan that addresses the Air Cargo Palletization Problem (ACPP) and the Aircraft Weight and Balance Problem (WBP) separately, impacting payload optimization and aircraft center of gravity (CG).
-The authors propose three integer linear programming models: a bi-objective optimization model (BOM), a combinatorial optimization model (COM), and an improved combinatorial optimization model (IOM).
-The models aim to determine the maximum loading capacity and lowest CG deviation from a specified target CG.
-
-	The study uses Gurobi to solve four scenarios with various conditional metrics for the three models.
-Results show that the BOM has the fastest solution speed, but the CG deviation is the largest, making the results unacceptable in some cases.
-The COM takes too long to solve, making it difficult to meet the needs of the actual scene.
-The IOM accelerates the solving process, but some scenarios still require significant computation time.
-}
+		\textit{We appreciate this suggestion.
+We include in Section 2 a brief description of the six most recent and relevant articles on the research topic.
+We also commented on how the recent article by \cite{zhao2023} alerted us to potential performance problems in the solutions offered for air transport problems.}
 
@@
 	\begin{minipage}{10cm}
-		\textit{Incluímos figuras comparativas, mostrando o melhor desempenho da heurística {\it Shims}. Citar quais são essas figuras...}
-
-		{\color{blue}
-		We included a map showing a 15-node tour as to validate the effectiveness of the {\it Shims} heuristic, we solve a 15-node TSP containing the 15 main Brazilian airports.
-		For this task, we implemented a TSP solution method based on the Genetic Algorithm (GA) that returns a population of 100 tours as input to the ACLP-RPDP solved by {\it Shims}.
-		}
+		\textit{As we commented in Sections 1 and 2 (see Table 2), we are dealing with a problem that has never been studied.
+Only \cite{LurkinSchyns2015} had addressed something similar but still quite different, as it did not produce a flight plan.
+In other words, no work has ever been published that has solved any instance of this problem.
+Therefore, developing a heuristic that presents solutions for it can be understood as a demonstration of superiority.
+On the other hand, we present the simulation results of our algorithm in Tables 12, 13, 14, and 15, and we now include Figures 6 and 7.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Nas novas figuras incluídas no artigo, mostramos o desempenho dos algoritmos desenvolvidos. Citar quais são essas figuras...}
-
-		{\color{blue}
-The GA run time to solve a 15-node TSP was around three seconds, and the total run time (including the ACLP+RPDP solution by Shims) was around 33 seconds, which we believe is adequate for an operational scenario and using a common hand-held computer.
-
-A figure was included showing the solution tour.
-	}
+		\textit{The solution for each ACLP+RPDP instance consists of 3 plans: the flight route; items that were delivered and shipped to each node on this route; and the allocation of items and packed contents on pallets at each node on this route.
+As it seems very difficult to express these plans through figures, we prefer to show two simpler indicators of the results obtained in each instance: objective function value and algorithm runtime.
+Some of these results were presented in Tables 12, 13, 14, and 15, and we now include Figures 6 and 7.
+Specifically, Figure 7 shows the flight plan of an instance in which $K=15$.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Incluímos referências mais recentes (citá-las). De qualquer modo, preferimos manter as antigas, para assim termos um levantamento histórico da evolução da pesquisa neste campo.}
-
-		{\color{blue}
-		We included a more recent reference, \citep{zhao2023}, that demonstrates how hard it is to solve the ACLP and WBP, separately or together.
-
-		In any case, we prefer to keep the old ones so that we have a historical survey of the evolution of research in this field.	
-		}
+		\textit{We believe it is appropriate to maintain the old references, as they constitute a historical survey of the evolution of research in this field.
+On the other hand, if reviewers prefer to omit some of them, we don't see a problem with that.
+At the same time, we seem to cite all recent relevant work, similar to \cite{zhao2023}, which is an article less than a year old.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Talvez não tenha ficado claro no texto original, mas três estratégias foram imprescindíveis para que pudéssemos obter alguma solução nas instâncias criadas: (1) considerar $K<m$\ para que cada pallet recebesse uma destinação predefinida; (2) em cada tour, calcular uma solução nó a nó; (3) em cada nó intermediário do tour considerado, realocar os packed contents minimizando o torque na aeronave.
-Sem a adoção destas três estratégias, não conseguimos realizar nenhum experimento, mesmo em instâncias menores.
-A única estratégia que permitiu diferentes experimentos e comparações foi a heurística utilizada na solução nó a nó.
-Procuramos deixar isto mais claro na nossa conclusão.}
-
-		{\color{blue}
+		\textit{
 		It may not have been clear in the original text, but three strategies were essential for us to obtain a solution in the instances created:
-
 		\begin{itemize}
-		\item[(1)] consider $K<m$\ so that each pallet received a predefined destination;
-
-		\item[(2)] on each tour, calculate a node-by-node solution; and
-
-		\item[(3)] at each intermediate node of the considered tour, relocate the packed contents, minimizing the torque on the aircraft.
+		\item[(1)] consider $K<m$\/ so that each pallet received a preset destination;
+		\item[(2)] on each tour, calculate a node-by-node solution;
+		\item[(3)] at each intermediate node of the considered tour, reallocate the packed contents, minimising the torque on the aircraft.
 		\end{itemize}
-
-		Without adopting these three strategies, we were unable to carry out any experiments, even in smaller instances.
-The only strategy that allowed different experiments and comparisons was the heuristic used in the node-by-node solution.
-We seek to make this clearer in the discussion of the results and in our conclusion.
-		}
+		Without adopting these three strategies, we would not be able to carry out any experiments, even in small instances.
+To make this clearer, we have included Figure 3, which describes the entire resolution process.
+The only strategy that allows different experiments and comparisons is the node-by-node solution.
+For this reason, we included Table 11, showing the effectiveness of each method for this problem.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Como estamos lidando com um novo problema, não dispomos de outras resoluções disponíveis na literatura.
-Na verdade, temos apenas meta-heurísticas bem conhecidas para a solução nó a nó, no qual o torque da aeronave é minimizado.
-Por este motivo, incluímos novas tabelas comparativas com os resultados obtidos nesses casos.}
-
-		{\color{blue}
-		As we are dealing with a new problem, we do not have other solutions available in the literature.
-In fact, we only have well-known meta-heuristics for the node-to-node solution, in which the aircraft torque is minimized.
-For this reason, we have included new comparative tables with the results obtained in these cases.
-		}
+		\textit{
+		As we are dealing with a new problem, we have no other solutions available in the literature.
+In fact, we only have well-known meta-heuristics for the node-by-node solution, in which the score/torque ratio is maximized.
+Table 11 shows the results obtained by NMO, ACO, GRASP, TS, and GA applied to this problem.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Agradecemos este comentário. Incluímos a análise de complexidade do algoritmo apresentado no nosso artigo.}
+		\textit{We appreciate this comment. In subsection 5.3, we include the time complexity analysis of the resolution process of an ACLP+RPDP instance.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Agradecemos esta sugestão. Acrescentamos novas figuras e tabelas que ilustram melhor nossos resultados. Citar quais são...}
+		\textit{We appreciate this suggestion.
+We reorganised the initial text of Section 5 and added Figure 3, which describes the entire resolution process.
+We also include Figures 6 and 7, which illustrate an important case.
+We consider that these changes facilitated the understanding of the proposed algorithm.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Estes espaços surgiram devido à formatação do \LaTeX, e deverão ser eliminados na versão final.}
+		\textit{These blank spaces appeared due to \LaTeX\/ formatting and should be eliminated in the final version.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Acrescentamos novas motivações para a resolução do ACLP+RPDP, que envolve simultaneamente 4 sub-problemas {\it NP-hard}.}
+		\textit{We added a new paragraph in Section 1, highlighting the motivations for ACLP+RPDP resolution: ``Inefficient air transport plans can lead to unnecessary costs, extra routes, longer distances, and incorrect destinations.
+Unbalanced cargo increases fuel consumption due to altered aircraft pitch angles, increasing the risk of weight- and balance-related accidents.
+Balancing cargo is crucial for safe aerial transportation, as an improperly positioned centre of gravity (CG) can result in dangerous take-off and landing conditions and stall recovery issues.
+Despite technological advancements, many airlines still rely on manual aircraft loading and balancing, which can lead to flight delays.
+A rational decision-making process is essential to avoid creating inefficient or unsafe transport plans, considering the high costs of fuel, maintenance, operation, outsourcing expenses, potential operational impairments, and safety risks due to unbalanced cargo.
+Solving this problem is vital for optimising strategic scores, saving time and effort in loading, ensuring safety and balance, correct pickups and deliveries, and finding the best routes considering fuel use due to potential cargo imbalance''.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Agradecemos esta sugestão. Incluímos mais informações na revisão da literatura, indicando as modelagens e os métodos de resolução.}
+		\textit{We appreciate this suggestion.
+We include more information about the main articles in this field of research, indicating their modelling, the scenarios considered, and the resolution methods adopted.
+It seems to us that this complementation has made the contribution of our work clearer.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Agradecemos esta sugestão. Entre os novos resultados apresentados, incluímos uma resolução na qual o tempo dedicado a cada nó é proporcional à carga disponível para embarque.}
-		{\color{blue}
-
-An important experiment inclusion was regarding the time limit to solve a node.
-This was adjusted to be proportional to the sum of the volumes of the candidate items in the node, as each node has a different number of items to be embarked.
-
-This inclusion certainly contributed to the slight improvement observed in the results.
-
-		}
-
+		\textit{We appreciate this suggestion, as we hadn't really thought about this possibility.
+We redid all our experiments, adopting, at each tour node, a runtime limit proportional to the load available for boarding.
+We found that there was a small improvement, which is why we incorporated this strategy into the resolution process.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Estudamos a viabilidade de definir empiricamente os valores dessas constantes do algoritmo Shims. Explicar...}
-
-{\color{blue}
-	After the items were generated, some experiments with the \textit{iRace} tool were executed to determine the values for volume limits $threshold_1$ and $threshold_2$ used in \textit{Shims}, whose results are presented in Table \ref{tab:irace}.
-
-	\begin{table}[H]
-		\centering
-		\caption{\textit{iRace} testing results}  \label{tab:irace}
-		\footnotesize
-		\begin{tabular}{cccc}
-			\toprule
-
-			\boldm{$surplus$} & \boldm{$threshold_1$} & \boldm{$threshold_2$} & \textbf{run time}\\
-			                 & for $\eta_1$          & for $\eta_2$          & (min) \\
-			\midrule
-			1.2& 0.8621 & 1.0539 & 47 \\
-			1.5& 0.9199 & 1.1399 & 59 \\
-			2.0& 0.9617 & 1.5706 & 63 \\						
-			\bottomrule
-		\end{tabular}
-		\normalsize
-	\end{table}
-
-	For these tests, we considered the odd instances (1, 3, 5, 7) as the \textbf{training} set and the even instances (2, 4, 6) as the \textbf{testing} set for the \textit{iRace} runs.
-We supplied \textit{iRace} for $threshold_1$ determination, the range [0.8, 1.0], and for $threshold_2$, the range [1.0, 2.0].
-The number of iterations in each experiment was limited to 3000 executions for \textit{iRace} to have plenty of data for its statistical tests.
-More detail about these configurations may be found on the \textit{iRace} user guide in \url{https://cran.r-project.org/web/packages/irace/irace.pdf}.
-}
-
+		\textit{We appreciate this suggestion.
+In subsection 6.1 and Table 10, we explain how the {\bf irace} \citep{LopezIbanezManuel2016} tool was adopted to empirically define these parameters of Shims, described in Algorithm 6.
+There was indeed an improvement in the results, which were incorporated into Tables 12, 13, 14, and 15.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Não entendemos este comentário...
-De qualquer modo, aumentamos a seção de experimentos do nosso artigo, procurando definir todas as variáveis envolvidas.
-Caso reste algum ponto obscuro, pedimos ao revisor que o especifique, por favor.}
-
-{\color{blue}
-
-An important experiment inclusion was regarding the time limit to solve a node.
-This was adjusted to be proportional to the sum of the volumes of the candidate items in the node, as each node has a different number of items to be embarked.
-
-}
-
+		\textit{Unfortunately, we do not understand this comment.
+Anyway, we have expanded Section 6, defining all the variables involved in the experiments carried out.
+If there is an unclear point, we ask the reviewer to please specify it.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Analisamos as resoluções do ACLPP através de algoritmos genéticos. Explicar...}
-
-		{\color{blue}
-	We implemented the genetic algorithm with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework available at \url{https://github.com/deap/deap}.
-More details can be found in \citep{DEAP_JMLR2012}.
-
-	\vspace{3mm}
-
-	\citep{PeerlinckAmy2022MFEO} applied DEAP to solve a multi-objective knapsack problem.
-
-	\vspace{3mm}
-
-	Our GA implementation is not complete. Due to performance issues, we did not include the constraints regarding the Weight and Balance (WBP) or the Pickup and Delivery (PDP) problems.
-
-	In the first attempt, considering only the weight and volume constraints, our implementation of DEAP managed to solve problems up to 100 items to be allocated to 18 pallets in less than 10 seconds.	
-	When we included the item count constraint to prevent each item from being allocated to more than one pallet, it was able to solve only small problems (20 items on 2 pallets).
-
-	We increased the number of items to 150, and the number of pallets to 18, but, even adjusting the GA parameters to more generations (2400) and more individuals (1200), DEAP did not generate any feasible solution.
-The run time of this attempt was around 10 minutes.
-
-	For the crossover probability, we tested the range [0.6 - 0.8], and for the mutation probability, we tested from 0.01 to 0.1 with no success.
-
-	\vspace{3mm}
-
-	The implemented code is available in \url{https://github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/ga\_deap\_mmkp.py}.
-
-	\vspace{3mm}
-
-	The palletization planning of 200 to 1000 items per node in a set of 18 aircraft pallets may be an inadequate problem to be solved with GA.
-
-Large chromosome sizes, like 18,000-dimension arrays, can pose significant challenges in using genetic algorithms.
-
-The computational complexity is significant, as each generation requires evaluating individual fitness, which can be resource-intensive.
-
-Large chromosomes can slow GA convergence due to the vast search space, potentially requiring more generations to find optimal solutions.
-
-Efficient encoding and decoding of the problem into the chromosome and decoding solutions is crucial for a 18,000-dimension array. 
-Also, the design of the fitness function becomes more critical as chromosome size increases, as it needs to guide the evolutionary process without being overly computationally demanding.
-
-\vspace{3mm}
-
-We hope that these considerations help justify why we did not include GA as a candidate method to solve the ACLP+RPDP.
-		}
+		\textit{We appreciate this suggestion.
+We used DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework \citep{DEAP_JMLR2012}, to compare GA as one of the heuristics in the node-by-node solution.
+In our first version, considering only weight and volume constraints, our implementation was able to solve problems of up to 100 items being allocated on 18 pallets in less than 10 seconds.
+When we included the item count constraint to prevent the same item from being allocated to more than one pallet, we were able to resolve only small problems (20 items on 2 pallets).
+We increased the number of items to 150 and the number of pallets to 18, but even adjusting the GA parameters for more generations (2,400) and more individuals (1,200), DEAP did not generate any viable solutions.
+The runtime for this attempt was about 10 minutes.
+These results are included in Table 11.
+On the other hand, we successfully used DEAP in developing a TSP heuristic for cases where $K>6$, as can be seen in subsubsection 6.2.2.}
 	\end{minipage}
@@
 	\begin{minipage}{10cm}
-		\textit{Agradecemos esta sugestão.
-Incluímos resultados da resolução de casos com um maior número $K$\/ de nós, mantendo a suposição $K < m$, necessária para a definição prévia da destinação dos pallets.
-Para isso, utilizamos uma heurística conhecida para o TSP, aplicando nosso algoritmo apenas nesse tour.}
-
-{\color{blue}
-
-To validate the effectiveness of the {\it Shims} heuristic, we solve a 15-node TSP containing the 15 main Brazilian airports.
-For this task, we implemented a TSP solution method based on the Genetic Algorithm (GA) that returns a population of 100 tours as input to the ACLP-RPDP solved by {\it Shims}.
-The GA run time was around three seconds, and the total run time was around 33 seconds.
-
-We implemented the GA with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework available in \url{https://github.com/deap/deap}.
-More details can be found in \citep{DEAP_JMLR2012}.
-
-\vspace{3mm}
-
-The implemented code is available in \url{https://github.com/celiomesquita/ACLP\_RPDP\_P/blob/main/tsp\_deap.py}.
-
-\vspace{3mm}
-
-	\begin{figure}[H]
-		\centering
-		\caption{\label{fig:brazil} {\it Shims/GA} best ACLP+RPDP tour}
-		\includegraphics[scale=0.3]{brazil.png}
-	\end{figure}
-
-\vspace{3mm}
-
-The airports cited in Figure \ref{fig:brazil}  are: São Paulo (GRU), Congonhas (CGH), Curitiba (CWB), Porto Alegre (POA), Florianópolis (FLN), Vitória (VIX), Salvador (SSA), Recife (REC), Fortaleza (FOR), Belém (BEL), Manaus (MAO), Brasília (BSB), Goiânia (GYN), Confins (CNF), and Rio de Janeiro (GIG), returning to GRU.
-
-}
+		\textit{We appreciate this suggestion.
+We included tests for solving cases with $K>6$, maintaining the requirement $K < m$, necessary to previously define the destination of the pallets.
+The results are described in subsubsection 6.2.2 and in Figures 6 and 7.}
 	\end{minipage}
 \end{center}
+
+\bibliographystyle{apalike}
 

### 2024-02-04 latex/ESWA_complexity.tex
Time compexity update

--- 2024-02-04 latex/ESWA.tex
@@
 \begin{document}
-
-\begin{frontmatter}
-
-	\title{Air cargo load and route planning in pickup and delivery operations}
-
-	\author{A.C.P.~Mesquita}
-	\ead{celio@ita.br}
-
-	\author{C.A.A.~Sanches\corref{cor1}}
-	\ead{alonso@ita.br}
-	\cortext[cor1]{Corresponding author.}
-
-	\address {Instituto Tecnol\'{o}gico de Aeron\'{a}utica - DCTA/ITA/IEC\\
-		Pra\c{c}a Mal. Eduardo Gomes, 50\\
-		S\~{a}o Jos\'{e} dos Campos - SP - 12.228-900 - Brazil}
-
-\begin{abstract}
-In the aerial pickup and delivery of goods in a distribution network, transport aviation faces risks of load imbalance due to the urgency required for loading, immediate take-off, and mission accomplishment.
-Transport planners deal with trip itineraries, prioritisation of items, building up pallets, and balanced loading, but there are no commercially available systems that can integrally assist in all these requirements.
-This enables other risks, such as improper delivery, excessive fuel burn, and possible safety issues due to cargo imbalance, as well as a longer than necessary turn-around time. This {\it NP-hard}\/ problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modelled using standardised pallets in fixed positions.
-We developed a strategy to solve this problem, considering historical transport data from some Brazilian hub networks, and performed several experiments with a commercial solver, {\color{blue}five} known meta-heuristics, and a new heuristic designed specifically for this problem.
-By using a portable computer, our strategy quickly found practical solutions to a wide range of real problems in much less than operationally acceptable time.
-\end{abstract}
-
-\begin{keyword}
-	OR in Airlines \sep Load Planning \sep Air Palletization \sep Weight and Balance \sep Pickup and Delivery \sep Vehicle Routing
-\end{keyword}
-
-\end{frontmatter}
-
-\section{Introduction}\label{sec:Introduction}
-
-The aviation industry adapts during global crises to keep supply chains moving.
-Air cargo provided complex expertise and the ability to access diverse destinations, delivering essential goods such as medicines, vaccine supplies, testing kits and other necessities with exceptional speed.
-This mode of transportation has become a preferred choice for governments, corporations and global companies in urgent need of transportation solutions.
-
-Air cargo services are specially designed for organisations that require customised transportation, handle sensitive goods, or serve remote locations with limited routes.
-Air carriers typically use high-capacity cargo planes for economies of scale.
-Many cargo airlines have worldwide networks spread across destinations around the world.
-
-A few years ago, \cite{BrandtStefan2019} defined the {\it Air Cargo Load Planning Problem} (ACLPP) as four sub-problems: {\it Aircraft Configuration Problem} (ACP), {\it Build-up Scheduling Problem} (BSP), {\it Air Cargo Palletization Problem} (APP), and {\it Weight and Balance Problem} (WBP).
-Several aspects were considered: item characteristics to be transported (dimensions, scores, dangerousness, etc.); types and quantities of {\it unit load devices} (ULDs); when these pallets are assembled; how items are allocated to pallets; in which positions these pallets are to be placed; how total cargo weight is balanced; etc.
-
-However, it is crucial to highlight that there are still other important challenges in air cargo transport that go beyond the definition of ACLPP, especially with regard to routes, and pickup and delivery at each destination.
-In this context, at least two more important sub-problems can be considered: simultaneous pickup and delivery at each node, called the {\it Pickup and Delivery Problem} (PDP), and searching for the best benefit-cost route, which is a special case of the {\it Travelling Salesman Problem} (TSP).
-
-The problem addressed in this work is the definition of the route of an aircraft that loads and unloads hundreds of items in up to seven hubs, dealing with weight, volume, and balance constraints, and maximising the benefit-cost ratio.
-This challenge has become even more acute during the COVID-19 pandemic due to the need for urgent medical supplies.
-Delays in shipping essential items such as respirators to critical areas highlighted the need for optimised solutions.
-
-Inefficient air transport plans can lead to unnecessary costs, extra routes, longer distances, and incorrect destinations.
-Unbalanced cargo increases fuel consumption due to altered aircraft pitch angles, increasing the risk of weight- and balance-related accidents.
-Balancing cargo is crucial for safe aerial transportation, as an improperly positioned centre of gravity (CG) can result in dangerous take-off and landing conditions and stall recovery issues.
-Despite technological advancements, many airlines still rely on manual aircraft loading and balancing, which can lead to flight delays.
-A rational decision-making process is essential to avoid creating inefficient or unsafe transport plans, considering high costs of fuel, maintenance, operation, outsourcing expenses, potential operational impairments, and safety risks due to unbalanced cargo.
-Solving this problem is vital for optimizing strategic scores, saving time and effort in loading, ensuring safety and balance, correct pickups and deliveries, and finding the best routes considering fuel use due to potential cargo imbalance.
-
-This problem is extremely complex, as it deals with different objectives: defining a route that visits all hubs; maximising the items transported along this route, prioritising the essential ones; making sure items reach the correct destinations; ensuring aircraft safety constraints; and saving fuel so that the flight is sustainable.
-
-We have developed a heuristic process that can be run on a handheld computer, quickly providing a good solution to real instances of this problem.
-Solutions consist of flight itineraries, pickup and delivery plans, and the allocation of items onto pallets, ensuring the load-balancing constraints.
-Our method also reduces the stresses that transport planners are subject to, as they have to deal with extensive information in a short time frame.
-
-To the best of our knowledge, this is the first time that an air cargo problem involving simultaneously APP, WBP, PDP, and TSP has been addressed.
-This new problem is named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem} (ACLP+RPDP).
-As we will describe in our mathematical modelling, these four sub-problems appear in an interconnected way in ACLP+RPDP and therefore cannot be solved independently.
-
-As a real case study, we consider a crucial network for the {\it Brazilian Air Force}, as can be seen in Table \ref{tab:costs} and Figure \ref{fig:nodes}.
-Although there are other airports of interest, these nodes were chosen due to their high demand.
-Other Brazilian airports tend to have smaller transport requests, which are generally met less expensively by cabotage, rail, or road transport.
-
-\begin{table}[htp]
-
-	\begin{minipage}{0.58\linewidth}
-
-		\caption{Brazilian airports distances ($km$)} \label{tab:costs}
-		\centering
-		\footnotesize
-
-		\newcolumntype{C}{>{\centering\arraybackslash}p{0.07\textwidth}}
-		\newcolumntype{R}{>{\raggedleft\arraybackslash}p{0.07\textwidth}}
-
-		\begin{tabular}{C R R R R R R R}
-			\toprule
-			IATA*   & GRU   & GIG   & SSA   & CNF   & CWB   & BSB   & REC \\	
-			\midrule	
-			GRU     & 0	    &343	&1,439   &504    &358    &866    &2,114\\
-			GIG	    & 343	&0	    &1,218   &371    &677    &935    &1,876\\
-			SSA	    & 1,439	&1,218	&0	    &938    &1,788   &1,062   &676\\
-			CNF	    & 504	&371	&938	&0	    &851    &606    &1,613\\
-			CWB	    & 358	&677	&1,788	&851	&0	    &1,084   &2,462\\
-			BSB	    & 866	&935	&1,062	&606	&1,084	&0	    &1,658\\
-			REC	    & 2,114	&1,876	&676	&1,613	&2,462	&1,658	&0\\
-			\bottomrule
-			\multicolumn{8}{c}{*International Air Transport Association}\\
-			\multicolumn{8}{c}{\small\textsuperscript{Source: www.airportdistancecalculator.com}}\\
-		\end{tabular}
-		\normalsize
-
-	\end{minipage}\hfill 
-	\begin{minipage}{0.42\linewidth}
-		\centering
-		\includegraphics[scale=0.21]{nodes.png}
-		\captionof{figure}{A route with 7 airports}
-		\label{fig:nodes}
-	\end{minipage}
-\end{table}
-
-This article is organised into six more sections.
-In Section \ref{literature}, we make the literature review.
-In Section \ref{assumptions}, we present the context and requirements of ACLP+RPDP.
-In Section \ref{modelling}, we describe its mathematical modelling.
-In Section \ref{algorithms}, we describe the developed algorithms, whose results are presented in Section \ref{results}.
-Finally, our conclusions are in Section \ref{conclusions}.
-
-\section{Literature review}
-\label{literature}
-
-The vast majority of operational research applied to air cargo is focused on challenges related to WBP, that is, the distribution of items on pallets to ensure load balancing.
-We can mention: \cite{LarsenMikkelsen1979}; \cite{Brosh1981}; \cite{Kevin1992}; \cite{Heidelberg1998}; \cite{fok2004optimizing}; \cite{KaluznyBohdanL2009Oalb}; \cite{Verstichel2011}; \cite{Limbourg2012}; \cite{RoesenerBarnes2016}; \cite{YangLiuGao2018}; \cite{zhao2021}; \cite{MiguelOptimalAP}.
-
-Other authors have addressed pallet assembly (APP) on aircraft, possibly also considering load balancing (WBP): \cite{MongeauBes2003}; \cite{Chan2006}; \cite{RoesenerHall2014}; \cite{Vancroonemburg2014}; \cite{Paquay2016}; \cite{Paquay2018}; \cite{wong2020}; \cite{eugene2021}; \cite{zhao2023}.
-
-In all these works, there is a great diversity of scenarios and solutions: some consider items in two dimensions, and others in three dimensions; some used integer programming, and others developed specific heuristics.
-Among the most recent, we can highlight:
-\begin{itemize}
-
-	\item \cite{RoesenerBarnes2016} proposed a heuristic to solve the {\it Dynamic Airlift Loading Problem} (DALP).
-Given a set of palletized cargo items that require transport between two nodes in a given time frame, the objective of this problem is to select an efficient subset of aircraft, partition the pallets into aircraft loads, and assign them to allowable positions in those aircraft.
-
-	\item \cite{Paquay2016} presented a mathematical model to optimise the loading of heterogeneous 3D boxes on pallets with a truncated parallelepiped format.
-Its objective is to maximise the volume used in containers, considering load balancing constraints, the presence of fragile items, and the possibility of rotating these boxes.
-\cite{Paquay2018} developed some heuristics to solve this problem.
-
-	\item \cite{YangLiuGao2018} modelled the air transport problem as a 2D packing problem and presented a heuristic for its optimisation in several aircraft, considering load balancing to minimise fuel consumption.
-
-	\item \cite{wong2020} developed a mathematical model and a tool based on mixed integer programming for optimising cargo in aircraft with different pallet configurations.
-Balance constraints and the presence of dangerous items were considered.
-\cite{eugene2021} integrated this tool into a digital simulation model with a visualisation and validation system based on sensors that alert about load deviations.
-
-	\item \cite{zhao2021} proposed a model for WBP based on Mixed Integer Programming (MIP).
-Instead of focusing on the CG deviation, the authors consider the original CG envelope of the aircraft, with a linearization method for its non-linear constraints.
-
-	\item \cite{zhao2023} proposed three integer linear programming models: a bi-objective optimisation model (BOM), a combinatorial optimisation model (COM), and an improved combinatorial optimisation model (IOM).
-Its objectives were the maximum possible loading and the smallest CG deviation.
-Four scenarios with multiple metrics are resolved for the {\it Boeing 777F}.
-The results showed that the BOM has the fastest solution speed, but the CG deviation is the worst; in several cases, the CG deviation is unacceptable.
-COM has the longest solution time, which is difficult to tolerate in practice.
-
-\end{itemize}
-
-\cite{LurkinSchyns2015} is the only work that simultaneously addresses an air cargo (WBP) and a flight itinerary (PDP) sub-problem.
-The authors demonstrated that this problem is {\it NP-hard}.
-Although it is innovative, strong simplifications were imposed by these authors: in relation to loading, APP was ignored; regarding routing, it is assumed that a predefined tour plan is restricted to only two legs.
-Referring directly to this work, \cite{BrandtStefan2019} comment: {\it However, not even these sub-problems are acceptably solved for real-world problem sizes, or models omit some practically relevant constraints}.
-
-Table \ref{tab:sa} lists this literature and the involved sub-problems.
-We also indicate whether the dimensions of the items were considered ({\bf 2D} or {\bf 3D}) and which solution method was used: heuristic search methods ({\bf Heu}), integer programming ({\bf Int}), or linear programming ({\bf Lin}).
-
-\begin{table}[htp]
-	\centering
-	\caption{Air cargo transport: literature, sub-problems and features}  \label{tab:sa}
-	\footnotesize
-	\begin{tabular}{r|cccc|ccccc}
-		\toprule
-		Work & {\bf APP}  & {\bf WBP}  &  {\bf PDP}   &{\bf TSP}   & {\bf 2D}  & {\bf 3D}  & {\bf Heu}  & {\bf Int}  & {\bf Lin} \\
-		\midrule
-		\cite{LarsenMikkelsen1979}  & $.$        & $\bigstar$ & $.$          & $.$        & $.$       & $.$       & $\bigstar$ & $.$        &  $.$ \\
-		\cite{Brosh1981}  & $.$ & $\bigstar$  & $.$   & $.$ & $.$ & $.$   & $.$  & $.$  &  $\bigstar$ \\
-		\cite{Kevin1992}  & $.$ & $\bigstar$  & $.$ & $.$ & $.$ &$.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{Heidelberg1998}  & $.$ & $\bigstar$  & $.$ & $.$ & $\bigstar$ &$.$   & $\bigstar$  & $.$  &  $.$ \\
-		\cite{MongeauBes2003}    & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{fok2004optimizing} & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
-		\cite{Chan2006}  & $\bigstar$ & $.$    & $.$ & $.$ & $.$ & $\bigstar$  & $\bigstar$  & $.$  &  $.$ \\
-		\cite{KaluznyBohdanL2009Oalb}  & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$ &$.$  & $.$  & $\bigstar$  &  $.$ \\
-		\cite{Verstichel2011}   & $.$ & $\bigstar$    & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\	
-		\cite{Limbourg2012} & $.$ & $\bigstar$  & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{RoesenerHall2014}  & $\bigstar$ & $\bigstar$  & $.$  & $.$ & $.$ & $\bigstar$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{Vancroonemburg2014}  & $\bigstar$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{LurkinSchyns2015} & $.$ & $\bigstar$  & $\bigstar$ & $.$  & $.$ & $.$   & $.$  & $\bigstar$  &  $.$ \\
-		\cite{RoesenerBarnes2016}  & $.$ & $\bigstar$   & $.$ & $.$ & $.$ & $.$   & $\bigstar$  & $.$  &  $.$ \\
-		\cite{Paquay2016,Paquay2018}  & $\bigstar$ & $\bigstar$ & $.$ & $.$ & $.$ & $\bigstar$ & $\bigstar$  & $\bigstar$ & $.$ \\
-		\cite{YangLiuGao2018} & $.$ & $\bigstar$  & $.$  & $.$ & $\bigstar$  & $.$ & $\bigstar$ & $.$  & $.$ \\
-		\cite{wong2020} & $\bigstar$  & $\bigstar$  & $.$  & $.$   & $.$  & $.$ & $.$ & $\bigstar$  & $.$  \\
-		\cite{eugene2021} & $\bigstar$ & $\bigstar$ & $.$  & $.$   & $.$ & $.$ & $.$ & $\bigstar$  & $.$  \\
-		\cite{zhao2021} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$ \\
-		\cite{zhao2023} & $\bigstar$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$  \\
-		\cite{MiguelOptimalAP} & $.$ & $\bigstar$ & $.$  & $.$  & $.$ & $.$ & $.$  & $\bigstar$ &  $.$  \\
-		{\bf This article}   & $\bigstar$ & $\bigstar$  & $\bigstar$& $\bigstar$ & $.$ & $.$ & $\bigstar$ & $\bigstar$   &  $.$  \\
-		\bottomrule 
-	\end{tabular}
-	\normalsize 
-\end{table}
-
-As can be seen, none of these papers address air cargo palletization and load balancing with route optimisation in a multi-leg transport plan for a single aircraft.
-Our work is the first to address a real air transport problem in which APP, WBP, PDP and TSP arise in an interconnected way.
-
-\section{Context and assumptions}
-\label{assumptions}
-
-In this section, we describe the context of the problem addressed in this work as well as the assumptions considered.
-
-\subsection{Operational premises}
-
-As we are dealing with an extremely complex and diverse problem, we decided to establish some simplifying characteristics:
-
-\begin{itemize}
-	\setlength\itemsep{-0.3em}
-
-	\item At each node of the tour, the items to be allocated are characterized by weight, volume, scores, and previously known destinations.
-We leave the consideration of 2D or 3D items to a future work.
-
-	\item We considered a unique pallet type: the {\it 463L Master Pallet}, a common size platform for bundling and moving air cargo.
-It is the primary air cargo pallet for more than 70 Air Forces and many air transport companies.
-This pallet has a capacity of $4,500 kg$ and $13.7 m^3$, which may be limited by its position along the cargo bay.
-It is equipped for locking into cargo aircraft rail systems, and includes tie-down rings to secure nets and cargo loads, which in total weighs $140 kg$.
-For more information, see {\tt www.463LPallet.com}.
-
-	\item All items allocated on a pallet must have the same destination.
-A pallet which has not yet reached its destination may receive more items, although it is known that these operations of removing restraining nets increase handling time and the risk of improper delivery.
-We do not consider oversized cargo in this work, but only cargo items that fit on these pallets.
-
-	\item Finally, as we are interested in minimizing fuel costs, we disregarded others costs not directly associated with aircraft flight, such as handling.
-
-\end{itemize}
-
-Throughout this text, we call {\it packed content}\/ a set of items of the same destination stacked on a pallet and covered with a restraining net.
-It is considered a single item, having the same attributes as its components, whose values are the sum of individual scores, weights, and volumes.
-To ensure accuracy in pickup and delivery operations, packed content must remain on board until its destination.
-
-\subsection{Aircraft parameters and load balancing}
-
-We consider real-world scenarios, where Table \ref{tab:larger} shows the aircraft parameters.
-$p_i$\/ are pallets, $1\leq i \leq 18$, whose weight and volume limits are $W_i$\/ and $V_i$, respectively.
-$D_i^{long}$\/ and $D_i^{lat}$\/ are, respectively, the longitudinal and lateral distances of each pallet centroids to the aircraft CG along both axes.
-These distances will be used in the calculation of the torque, referring to the items allocated on each pallet.
-In this aircraft, as the ramp has an inclination of $25^{\circ}$, we made the necessary corrections in $D_i^{long}$, $W_i$\/ and $V_i$\/ of the corresponding pallets ($p_1$, $p_2$, $p_3$, and $p_4$).
-
-\begin{table}[htp]
-	\centering
-	\caption{Aircraft parameters}  \label{tab:larger}
-	\footnotesize
-	\begin{tabular}{c | c c c c c c c c c}
-		\toprule
-		& \multicolumn{3}{c}{$Payload$: 75,000kg} & \multicolumn{3}{c}{$limit^{CG}_{long}$: $1.170m$} &
-		\multicolumn{3}{c}{$limit^{CG}_{lat}$: $0.19m$} \\
-		\midrule
-		\multirow{2}{*}{{\boldm{$p_i$}}}  & $p_{17}$ & $p_{15}$ & $p_{13}$ & $p_{11}$ & $p_{9}$ & $p_{7}$ & $p_{5}$ & $p_{3}$ & $p_{1}$ \\
-		& $p_{18}$ & $p_{16}$ & $p_{14}$ & $p_{12}$ & $p_{10}$ & $p_{8}$ & $p_{6}$ & $p_{4}$ & $p_{2}$ \\
-		\midrule 
-		\multirow{2}{*}{\boldm{$D_i^{long}$} ($m$)} & -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\
-		& -17.57 & -13.17 & -8.77 & -4.40 & 0 & 4.40 & 8.77 & 11.47 & 14.89 \\			
-		\midrule 
-		\multirow{2}{*}{\boldm{$D_i^{lat}$} ($m$)}  & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 & 1.32 \\
-		& -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 & -1.32 \\	
-		\midrule
-		{\boldm{$W_i$}} ($kg$)      &   4,500   &    4,500  &   4,500   &  4,500    & 4,500     & 4,500     & 4,500     & 3,000    & 3,000   \\
-		{\boldm{$V_i$}} ($m^3$)   &   14.8   &   14.8   &  14.8    &  14.8    & 14.8     & 14.8     & 14.8     & 10.0    & 7.0 \\	
-		\midrule	
-		\textbf{Fuel cost}  & \multicolumn{9}{c}{ $c_d$ = US\$ $4.90/km$ }	 \\
-		\midrule
-		\textbf{Fuel consumption rate}  & \multicolumn{9}{c}{ $c_g = 5\%$} \\		
-		\midrule
-		\textbf{Maximum weight}  & \multicolumn{9}{c}{ $W_{max} = \sum_i W_i = 75,000kg$} \\
-		\bottomrule
-	\end{tabular}
-	\normalsize 
-\end{table}
-
-This aircraft spends $c_d$\/ dollars per kilometre flown and can carry up to $W_{max}$\/ of cargo distributed on the pallets.
-The fuel penalty $c_g$\/ is the percentage of cost increase due to the CG deviation on the longitudinal axis, estimated at 5.0\%.
-It is important to consider that $c_g$\/ tends to zero as the aircraft attitude tends to be level.
-As the CG deviation varies from $0$\/ to $limit^{CG}_{long}$, this fuel penalty varies from $0$\/ to $c_g$.
-
-The torque applied to the aircraft must keep its CG in the operational range, which corresponds to a fixed percentage of the {\it Mean Aerodynamic Chord} \footnote{Chord is the distance between the leading and trailing edges of the wing, measured parallel to the normal airflow over the wing.
-The average length of the chord is known as the {\it Mean Aerodynamic Chord} (MAC).} which is considered $1.17m$ for the aircraft of this work (see Figure \ref{fig:lateral}).
-
-\begin{figure}[htp]
-	\centering
-	\includegraphics[scale=0.22]{lateral.png}
-	\caption{Aircraft longitudinal cut, where red lines are pallets positions}
-	\label{fig:lateral}
-\end{figure}
-
-We also make the following assumptions:
-
-\begin{itemize}
-	\setlength\itemsep{-0.3em}
-	\item on each pallet, the items are distributed in such a way that their CG coincides with the centroid of the pallet, because builders are well-trained to do so;
-	\item the CG of the total load must be at a maximum longitudinal distance of $limit^{CG}_{long}$\/ from the CG of the aircraft;
-	\item the CG of the total load must be at a maximum lateral distance of $limit^{CG}_{lat}$\/ from the CG of the aircraft;
-	\item the pallets are distributed in two identical rows (with odd and even indices, respectively), and the centroid of $p_i$\/ is at a distance $D^{lat}_{i}$\/ from the centreline of the aircraft;
-	\item when there are items or packed contents in $p_i$, the common destination of this load will be assigned to variable $T_i$. 
-\end{itemize}
-
-\section{The mathematical modelling}
-\label{modelling}
-
-In this section, we present the mathematical modelling of ACLP+RPDP in Tables \ref{tab:struc}, \ref{tab:graph}, \ref{tab:functions}, and \ref{tab:constraints}, with their corresponding descriptions.
-
-In Table \ref{tab:struc}, we describe the problem structure: nodes and their permutations, distances and associated costs, pallets characteristics, items available for shipment at each node, and packed contents shipped.
-The item $j$\/ in node $k$\/ has score $s_j$, weight $w_j$, volume $v_j$, and destination $to_j \in L_k$.
-Similarly, the packed content $q$, that remains on board at node $k$, has total weight $w_q$, total volume $v_q$, and destination $to_q \in L_k$.
-Packed contents that were destined to node $k$\/ are unloaded when the aircraft arrives there; that is, they are not considered in $Q_k$.
-
-\begin{table}[htp]
-	\centering
-	\footnotesize
-	\caption{Problem structure}  \label{tab:struc}
-	\begin{tabular}{l l}
-		\toprule
-		\textbf{Notation}	& \textbf{Description} \\
-		\midrule
-		$L = \{ 0, 1, \ldots, K \}$ & Set of $K+1$\/ nodes of the tour, where node 0 is the base \\
-		\midrule
-		$\pi$               & A permutation between nodes 1, \ldots, $K$\\
-		\midrule
-		$S_K$               & Set of $K!$\/ permutations \\
-		\midrule
-		$\pi_k$             & The $k^{th}$\/ node of tour $\pi$, $1 \leq k \leq K$\\
-		\midrule
-		Tour $\pi$          & $\{0,\pi_1,\ldots, \pi_K, 0\}$ \\
-		& For ease of notation, $\pi_0 = \pi_{K+1} = 0$\\
-		\midrule
-		$L_k$               & Set of remaining nodes of tour at node $k$, $0 \leq k \leq K$\\
-		& By definition, $L_0 = L$\\
-		\midrule
-		$d(a,b)$            & Distance from node $a$\/ to node $b$, where $0 \leq a, b \leq K$ \\
-		&   By definition, $d(a,a) = 0, \forall a$\\
-		\midrule
-		$C=\left[c_{a,b}\right]$    & Cost matrix of flights, where $c_{a,b} = c_d \times d(a,b)$\\
-		\midrule
-		$M = \{1, \ldots, m \}$  & Set of $m$\/ pallets in specific positions within the aircraft \\
-		&  See Table \ref{tab:larger}, where $m=18$ \\
-		\midrule
-		$N_k = \{1, \ldots, n_k \}$ & Set of $n_k$\/ items available for loading at node $k$, $1 \leq j \leq n_k$, $0 \leq k \leq K$\\
-		\midrule
-		$N = \bigcup_{0 \leq k \leq K} N_k$ & Set of items in all nodes along a tour\\
-		\midrule
-		$Q_k = \{1, \ldots, m_k \}$ & Set of $m_k \leq m$\/ packed contents at node $k$, $1 \leq q \leq m_k$, $0 \leq k \leq K$ \\ 
-		& By definition, $m_0 = 0$ and $Q_0 = \varnothing$\\
-		\bottomrule
-	\end{tabular}
-	\normalsize
-\end{table}
-
-Table \ref{tab:graph} contains decision variables and the ACLP+RPDP allocation graph.
-
-\begin{table}[htp]
-	\centering
-	\footnotesize
-	\caption{Decision variables and allocation graph}  \label{tab:graph}
-	\begin{tabular}{l l}
-		\toprule
-		\textbf{Notation}	& \textbf{Description} \\
-		\midrule
-		$X_{ij}^{\pi_k}$\/ and $Y_{iq}^{\pi_k}$ & Binary variables, where $1 \leq i \leq m$, $1 \leq j \leq n_{\pi_k}$, $1 \leq q \leq m_{\pi_k}$\/ and $0 \leq k \leq K$\\
-		\midrule
-		$X_{ij}^{\pi_k} = 1$ & If item $j$\/ at node ${\pi_k}$\/ is assigned to pallet $i$, and 0 otherwise\\
-		\midrule
-		$Y_{iq}^{\pi_k} = 1$ & If packed content $q$\/ at node ${\pi_k}$\/ is assigned to pallet $i$, and 0 otherwise\\
-		\midrule
-		$T_i^{\pi_k} \in L_{\pi_k}$ &  Destination of items and packed contents assigned to pallet $i$\ at node ${\pi_k}$\\
-		\midrule
-		$G_{\pi_k}(V_{\pi_k}, E_{\pi_k})$ & Allocation graph at node ${\pi_k}$\\
-		\midrule
-		$V_{\pi_k} = M \cup N_{\pi_k} \cup Q_{\pi_k}$ & Allocation graph vertices at node ${\pi_k}$: pallets, items and packet contents\\
-		\midrule
-		$E_{N_{\pi_k}}$ & Allocation graph edges at node ${\pi_k}$, corresponding to shipped items\\
-		\midrule
-		$E_{Q_{\pi_k}}$ & Allocation graph edges at node ${\pi_k}$, corresponding to packed contents\\
-		\midrule
-		$E_{\pi_k} = E_{N_{\pi_k}} \cup E_{Q_{\pi_k}}$ & Allocation graph edges at node ${\pi_k}$\\
-		\midrule
-		$(i, j) \in E_{N_{\pi_k}}$ & If $X_{ij}^{\pi_k} = 1$, where $i$\/ is a pallet and $j$\/ is a item at node ${\pi_k}$ \\
-		\midrule
-		$(i, q) \in E_{Q_{\pi_k}}$ & If $Y_{iq}^{\pi_k} = 1$, where $i$\/ is a pallet and $q$\/ is a packed content at node ${\pi_k}$\\
-		\bottomrule
-	\end{tabular}
-	\normalsize
-\end{table}
-
-The calculus functions of ACLP+RPDP are described in Table \ref{tab:functions}.
-
-\begin{table}[htp]
-	\centering
-	\footnotesize
-	\caption{Calculus functions} \label{tab:functions}
-	\begin{tabular}{ c l }
-		\toprule
-		\textbf{Function} & \textbf{Description} \\
-		\midrule
-		(\ref{eq:scores}) & Total score of transported items throughout tour $\pi$\\
-		\midrule
-		(\ref{eq:tau})    & Longitudinal torque applied by loaded pallets at node ${\pi_k}$\\
-		\midrule
-		(\ref{eq:costs})  & Total cost of fuel on tour $\pi$\/ (distances and CG longitudinal deviations) \\
-		\midrule
-		(\ref{eq:departing}) & Set of not visited nodes at node ${\pi_k}$\\
-		\midrule
-		(\ref{eq:LatItem}) & Lateral torque at node ${\pi_k}$ (shipped items) \\
-		\midrule
-		(\ref{eq:LatPacked}) & Lateral torque at node ${\pi_k}$ (packed contents)\\
-		\midrule
-		(\ref{eq1}) & Objective function of ACLP+RPDP\\
-		\bottomrule
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\begin{equation} \label{eq:scores}
-	\tilde{s}_\pi = \sum_{k=0}^{K} \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times s_j
-\end{equation}
-
-\begin{equation} \label{eq:tau}
-	\tau_{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times \Big ( \sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{long}; \ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:costs}
-	\tilde{c}_{\pi} = \sum_{k=0}^{K} \Big [ c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|) \Big ] 
-\end{equation}
-
-\begin{equation} \label{eq:departing}
-	L_{\pi_k} = L_{\pi_{k-1}} - \{\pi_k\}; \ k \in \{1, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:LatItem}
-	\epsilon_{\pi_k}^t = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{j=1}^{n_{\pi_k}} \Big ( X_{ij}^{\pi_k} \times w_j \times (i\%2) - X_{ij}^{\pi_k} \times w_j \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}
-\end{equation}
-
-\begin{equation} \label{eq:LatPacked}
-	\epsilon_{\pi_k}^a = \sum_{i=1}^{m} \Big [ D_i^{lat} \times \sum_{q=1}^{m_{\pi_k}} \Big ( Y_{iq}^{\pi_k} \times w_q \times (i\%2) - Y_{iq}^{\pi_k} \times w_q \times (i+1)\%2 \Big ) \Big ] \Big / W_{max} \times limit^{CG}_{lat}
-\end{equation}
-
-\begin{equation} \label{eq1}
-	\max_{\pi \in S_K} f_\pi = \tilde{s}_\pi/\tilde{c}_\pi
-\end{equation}
-
-Longitudinal (\ref{eq:tau}) and lateral torques (\ref{eq:LatItem}, \ref{eq:LatPacked}) are calculated in proportion to the highest torque supported by the aircraft.
-As there are two rows of pallets, one on each side of the centerline, we use the operator {\it modulo}\/ ($\%$) to calculate lateral torques.
-In our experiments, we found that the magnitude of these lateral torques was always minimal, so we decided to ignore them in the fuel consumption (\ref{eq:costs}).
-The objective of ACLP+RPDP (\ref{eq1}) is to find a permutation $\pi \in S_K$ with the corresponding allocation of items on pallets at each node that maximises the function $f_\pi = \tilde{s}_\pi/\tilde{c}_\pi$.
-
-Finally, ACLP+RPDP constraints related to each node ${\pi_k}$\/ are described in Table \ref{tab:constraints}.
-
-\begin{table}[htp]
-	\centering
-	\footnotesize
-	\caption{Constraints} \label{tab:constraints}
-	\begin{tabular}{ c l }
-		\toprule
-		\textbf{Constraint} & \textbf{Description} \\
-		\midrule
-		(\ref{eq:torqlong}, \ref{eq:torqlat}) & Longitudinal and lateral torques must be within aircraft limits \\
-		\midrule
-		(\ref{eq:app2}, \ref{eq:app3}) & Items allocated to each pallet cannot exceed its weight and volume limits\\
-		\midrule
-		(\ref{eq:app4}) & At most, each item is associated with a single pallet \\
-		\midrule
-		(\ref{eq:app5}) & Packed contents that have not yet reached their destination must remain on board\\
-		\midrule
-		(\ref{eq22}, \ref{eq23}) & Items allocated on the same pallet must have the same destinations\\
-		\midrule
-		(\ref{eq24}, \ref{eq25}) & If there is a packed content on the pallet, it must also have the same destination as other items\\
-
-		\bottomrule
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\begin{equation} \label{eq:torqlong}
-	| \tau_{\pi_k} | \leq 1;\ k \in \{0, \ldots, K\}
-\end{equation}
-
-\begin{equation} \label{eq:torqlat}
-	| \epsilon_{\pi_k}^t + \epsilon_{\pi_k}^a| \leq  1
-\end{equation}
-
-\begin{equation} \label{eq:app2}
-	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}
-\end{equation}
-
-\begin{equation} \label{eq:app3}
-	\sum_{j=1}^{n_{\pi_k}} X_{ij}^{\pi_k} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq}^{\pi_k} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}
-\end{equation}
-
-\begin{equation} \label{eq:app4}
-	\sum_{i=1}^{m} X_{ij}^{\pi_k} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:app5}
-	\sum_{i=1}^{m} Y_{iq}^{\pi_k} = 1;\ to_q \in L_{\pi_k}; \ q \in \{1, \ldots, m_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq22}
-	X_{ij}^{\pi_k} <= X_{ij}^{\pi_k} \times (T_i^{\pi_k} - to_j + 1); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq23}
-	X_{ij}^{\pi_k} <= X_{ij}^{\pi_k} \times (to_j - T_i^{\pi_k} + 1); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq24}
-	Y_{iq}^{\pi_k} <= Y_{iq}^{\pi_k} \times (T_i^{\pi_k} - to_q + 1); \ i \in \{1, \ldots, m\}; \ q \in \{1, \ldots, m_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq25}
-	Y_{iq}^{\pi_k} <= Y_{iq}^{\pi_k} \times (to_q - T_i^{\pi_k} + 1); \ i \in \{1, \ldots, m\}; \ q \in \{1, \ldots, m_{\pi_k}\}
-\end{equation}
-
-Once the assumptions and the mathematical modelling are presented, it is possible to see that ACLP+RPDP is {\it NP-hard}.
-In a similar way to \cite{LurkinSchyns2015}, consider the simple case where $K=1$\/ (one leg), $m=2$\/ (two pallets around the aircraft CG), $2n$\/ sufficiently light items with same scores in node 0, and no items in node 1.
-Under these conditions, through polynomial reductions for the {\it Set-Partition Problem}, it is possible to demonstrate that the decision problem associated with ACLP+RPDP is {\it NP-complete}.
-For more details, see \cite[p.~6]{LurkinSchyns2015}.
-
-\section{Solution strategies}
-\label{algorithms}
-
-Throughout our research, we have thoughtfully described ACLP+RPDP in standard MIP format and found that no solver can handle its practical cases in a feasible time.
-Thus, as ACLP+RPDP is highly complex, involving four intractable and interconnected sub-problems, we decided to focus only on {\it real cases}, developing quick node-by-node solutions, not necessarily optimal, but which would allow us to obtain a complete tour.
-
-\begin{figure}[htp]
-	\centering
-	\fbox{ \small
-		\begin{tikzpicture}
-			\node [mybox] (one) { \small {\bf Input data:}\\
-				\small \ \ \ {Aircraft parameters}\\
-				\small \ \ \ {Airports and distances}\\
-				\small \ \ \ {$scenario$ \ \ ($K$, $L$\/ and $C$)}\\
-				\small \ \ \ {Items available for shipment at each airport}\\						
-				\small \ \ \ {$surplus = \sum_{j=1}^{n_k} v_j$/$\sum_{i=1}^{m} V_i$ }\\		
-				\small \ \ \ {$tmax$ \ \ (overall runtime limit)}\\
-				\small {{\bf Requirement:} $1 < K < m$}
-			};
-
-			\node [mybox, below=0.7cm of one] (two) { \small {\bf Number of tested tours:}\\
-				\small \ \ \ {{\bf if} $K \leq 6$}\\
-				\small \ \ \ \ \ \ {$ntours = 2$ \ \ (optimal TSP solutions)}\\
-				\small \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ {\it or}\\				
-				\small \ \ \ \ \ \ {$ntours = K!$ \ \ (all possible tours)}\\
-				\small \ \ \ {\bf else}\\
-				\small \ \ \ \ \ \ {$ntours = 100$ \ \ (best tours obtained with a TSP heuristic)}
-			};
-
-			\node [mybox, below=0.7cm of two] (three) { \small {\bf On each tour of $ntours$:}\\
-				\small \ \ \ {Calculate a node-by-node solution (divide runtime according to shipment volumes):}\\
-				\small \ \ \ \ \ \ \ {- Reallocate packed contents with torque optimization}\\
-				\small \ \ \ \ \ \ \ {- Preset pallet destinations according to shipment volumes}\\
-				\small \ \ \ \ \ \ \ {- Find a node solution (MIP, Shims, NMO, ACO, GRASP, TS, GA) }\\
-				\small \ \ \ \ \ \ \ {- Accumulate values obtained from score, torque and cost}\\
-				\small \ \ \ \ \ \ \ {- Go to the next node on the current tour}
-			};
-
-			\node[myarrow, rotate=-90] at ([yshift=-8pt]two.south) {\phantom{\tiny{aaa}}};
-			\node[myarrow, rotate=-90] at ([yshift=-8pt]one.south) {\phantom{\tiny{aaa}}};		
-		\end{tikzpicture} 
-	}
-	\caption{Solution process}
-	\label{fig:process}
-\end{figure}
-
-In practical cases, we know that a common aircraft has $m=18$\/ pallets, flight itineraries have $K \leq 6$\/ nodes plus the base, and each node has hundreds of items to be shipped.
-We also know that missions with fewer nodes are more frequent than longer ones.
-Under these circumstances, we can adopt some important strategies summarized in Figure \ref{fig:process}:
-
-\begin{itemize}
-
-	\item We consider that the number of destinations is smaller than the number of pallets ($K < m$), and we avoid the trivial case where $K=1$.
-With this premise, we can preset the destinations of the pallets at each shipping node, reserving a number of pallets proportional to the volume available for each destination.
-We could have used another criterion, but it was observed in the experiments that the volume is more constrictive in airlift.
-
-	\item An important parameter is the number $ntours$\/ of tours tested.
-In practical cases where $K \leq 6$, we have the possibility to check all possible tours ($ntours = K!$).
-In this situation, as $K$\/ is small, we can also specially analyse the two optimal solutions of the corresponding TSP ($ntours = 2$).
-Finally, in cases where $K>6$, we will use a heuristic to select $100$\/ tours of low length ($ntours=100$), and search among them for the one that provides the best value for the objective function.
-
-	\item To compare the performance of each strategy, an overall runtime limit $tmax$\/ is established and divided by $ntours$\/ tours.
-In turn, the runtime limit for each tour will be distributed among its nodes in proportion to the volume available for boarding.
-
-	\item At each node of a tour, the packed contents that remain on board are reallocated on pallets in order to minimize torque on the aircraft.
-This calculation is done quickly using a MIP solver.
-Then, the destinations of the pallets are previously defined in proportion to the shipment volume.
-Finally, considering the runtime limit of each node, we will use a MIP solver and five well-known meta-heuristics to find the best allocation of shipping items: {\it Ant Colony Optimization} (ACO), {\it Noising Method Optimization } (NMO), {\it Tabu Search} (TS), {\it Greedy Randomized Adaptive Search Procedure} (GRASP), and {\it Genetic Algorithm} (GA).
-
-	\item We will generate benchmarks using the $surplus$\/ parameter, which is a value in $\{1.2, 1.5, 2.0\}$.
-It corresponds, at each node $k$, to the ratio between the sum of the volumes of the items and the load capacity of the pallets ($surplus = \sum_{j=1}^{n_k} v_j$/$\sum_{i=1}^{m} V_i$).
-This parameter allows us to verify the different behaviour of each method, according to $scenario$\/ and the quantity of items available for shipment.
-
-	\item We will do tests by varying the number $K$\/ of destinations, the set $L$\/ of nodes, and the costs $C$.
-Each group of values tested is called $scenario$, according to Tables \ref{tab:costs} and \ref{tab:scenarios}, where $1<K \leq 6$.
-After finding the method with the best performance in node-by-node solution, we will test it in solving cases with $K > 6$.
-
-\end{itemize}
-
-Algorithm \ref{alg:main} is the main program of this solution process, which parameters are $scenario$, $surplus$, $tmax$, and $ntours$.
-The input data is obtained from Tables \ref{tab:costs}, \ref{tab:larger} and \ref{tab:scenarios} (lines \ref{main:M}-\ref{main:KLC}).
-
-\begin{algorithm}[htp]
-
-	\caption{Solving the ACLP+RPDP}  \label{alg:main}
-
-	\begin{algorithmic}[1]
-			\State {\bf ACLP+RPDP} \,\, {\it in:}~$scenario, surplus, tmax, ntours$ \, \, {\it out:}~$answer$
-
-			\State Let $M$ be the set of pallets ({\it cfr.} Table \ref{tab:larger}) \label{main:M}
-
-			\State Let $K$, $L$\/ and $C$ be according to the $scenario$ ({\it cfr.} Tables \ref{tab:costs}, \ref{tab:larger} and \ref{tab:scenarios}) \label{main:KLC}
-
-			\State $N \gets ItemsGeneration(scenario, surplus)$ \label{main:items}
-
-			\For {each $method$}  \label{main:method}
-
-				\For {$x \gets 1$ to $ntours$} \label{main:loop1}
-
-					\State $f(S_K[x]) \gets SolveTour(S_K[x], L, M, C, N, method, tmax/ntours)$ \label{main:fpi}
-
-				\EndFor \label{main:loop2}
-
-				\State $answer[scenario,surplus,method] \gets \max f(S_K[x])$ \label{main:a2}
-			\EndFor
-	\end{algorithmic}
-
-\end{algorithm}
-
-$surplus$ is passed to $ItemsGeneration$\/ (line \ref{main:items}), responsible for creating the items to be shipped, which will be presented in the next section (Algorithm \ref{alg:itemsgen}).
-$tmax$\/ is the runtime limit, which will be distributed among the tours (line \ref{main:fpi}).
-$method$\/ corresponds to a MIP solver or a heuristic to the node-by-node solution $SolveTour$, which will be presented in subsection \ref{methods}.
-The best results obtained by testing all tours are stored in $answer$\/ (line \ref{main:a2}), which is the output of this algorithm.
-
-{\color{red}$\pi_k$ are tours} {\color{blue}$\pi$ are tours}, that is, permutations with the $K$\/ nodes.
-When $ntours=2$, $\pi_1$\/ and $\pi_2$\/ are the optimal solutions of the corresponding TSP.
-When $ntours=K!$, $\pi_k$\/ is the $k^{th}$ permutation of $S_K$.
-Finally, when $ntours=100$, $\pi_1, \ldots, \pi_{100}$\/ are the best TSP solutions provided by a heuristic.
-
-\begin{table}[h]
-	\centering
-	\caption{Testing scenarios}  \label{tab:scenarios}
-	\begin{tabular}{c c c }
-		\toprule
-		{\bf Scenario} & $K$ & $L$  \\		
-		\midrule
-		1 & 2    & \{$0,1,2$\}            \\
-		2 & 3    & \{$0,1,2,3$\}         \\
-		3 & 4    & \{$0,1,2,3,4$\}        \\
-		4 & 5    & \{$0,1,2,3,4,5$\}      \\
-		5 & 6    & \{$0,1,2,3,4,5,6$\}    \\
-		\bottomrule
-	\end{tabular}
-\end{table}
-
-Next, we will present two subsections: in the first we explain how $SolveTour$\/ is executed. In the second we will present the heuristics developed for node-by-node solutions.
-
-\subsection{SolveTour algorithm}
-\label{tour}
-
-As we commented in the previous subsection, we will adopt the strategy of presetting the destinations of each pallet throughout the tour.
-This is feasible in practical cases where $1<K<m$.
-For this, each pallet $i$\/ also has a field $T^k_i$, $0\leq k \leq K$, which stores its next destination after being loaded at node $k$.
-For this reason, $T^k_i \in L_k$, $1 \leq i \leq m$, $0\leq k \leq K$.
-
-$SolveTour$\/ is described in Algorithm \ref{alg:tour}, where $\pi$\/ is a permutation of the nodes (excluding the base) that defines the order of visits in this tour, $method$\/ corresponds to a MIP solver or a heuristic for solving the node-by-node problems, and $tmax$\/ is the runtime limit of this tour.
-
-\begin{algorithm}[H]
-	\caption{Solving the tour $\pi$ with $method$}  \label{alg:tour}
-	\begin{algorithmic}[1]
-
-		\State {\bf SolveTour} \,\, {\it in: $\pi, L, M, C, N, method$, $tmax$} \, \, {\it out: $score / cost$}
-
-		\State $\pi_0     \gets 0$ \label{tour:pi1} \Comment{all tours start and end at the base}
-		\State $\pi_{K+1} \gets 0$ \label{tour:pi2}
-
-		\State $score \gets 0$ \label{tour:score}
-		\State $cost  \gets 0$ \label{tour:cost}
-
-		\For {$k \gets 0$ to $K$} \label{tour:loop1}
-		\State $t_{\pi_k} = (\sum_{j=1}^{n_k} v_j/ \sum_{k=0}^K \sum_{j=1}^{n_k} v_j) * tmax$ \label{tour:tnode}	 \Comment{runtime limit proportional to the shipment volume}
-
-		\State $L_{\pi_k} \gets L - \{\pi_0,\pi_1,\ldots,\pi_k\}$  \label{tour:lk1}	 \Comment{the set of remaining nodes is updated}	
-		\State $T_i^{\pi_k} \gets -1$, $1 \leq i \leq m$ \label{tour:-1}             \Comment{the pallet destination is unset}
-
-		\If {$k = 0$}
-			\State Let $G_1(M \cup N_0, \varnothing)$ \label{tour:g11}               \Comment{no packed contents at the base}
-		\Else
-			\State $E_{Q_{\pi_k}}, M \gets UpdatePacked(M, Q_{\pi_k}, \pi_k)$ \label{tour:dest}			
-			\State Let $G_1(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$ \label{tour:g12}
-		\EndIf  \label{tour:lk2}
-
-		\State $M \gets SetPalletsDestinations(M, \pi_k )$ \label{tour:dest2}
-
-		\State $G_2 \gets SolveNode(method, \pi_k, G_1, t_{\pi_k})$  \label{tour:SolveNode}
-
-		\State $s, \tau \gets ScoreAndTorque(\pi_k, G_2)$ \label{tour:analyse}
-
-		\State $score \gets score + s$ \label{tour:score2}
-		\State $cost  \gets cost  + c_{\pi_k,\pi_{k+1}} \times (1 + c_g \times |\tau|)$ \label{tour:cost2}
-
-		\EndFor  \label{tour:loop2}
-
-	\end{algorithmic}
-\end{algorithm}
-
-As we mentioned in the previous section, all tours start and end at the base $0$\/ (lines \ref{tour:pi1}-\ref{tour:pi2}).
-After initializing the score and cost values (lines \ref{tour:score}-\ref{tour:cost}), there is a loop for the $K+1$\/ flights (lines \ref{tour:loop1}-\ref{tour:loop2}).
-Initially, the runtime limit $t_{\pi_k}$\/ for each node is calculated (line \ref{tour:tnode}), the set $L_{\pi_k}$\/ of remaining nodes is updated (line \ref{tour:lk1}), and the pallet destinations are unset (line \ref{tour:-1}).
-
-When the aircraft is at the base, the initial graph $G_1$\/ is empty, and there are no packed contents {\color{blue}(line \ref{tour:g11})}.
-Otherwise, $UpdatePacked$\/ (line \ref{tour:dest}) returns the set of packed contents that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation.
-This allocation is stored in graph $G_1$\/ (line \ref{tour:g12}).
-
-$SetPalletsDestinations$\/ (line \ref{tour:dest2}) presets the destination of each pallet based on the volume demands of the current node without changing the pallet's destination with packed contents.
-
-Finally, $SolveNode$\/ includes the edges corresponding to the items shipped at the current node, returning the graph $G_2$\/ (line \ref{tour:dest2}).
-The score and the CG deviation of $G_2$\/ are calculated (line \ref{tour:analyse}) and accumulated (lines \ref{tour:score2}-\ref{tour:cost2}), allowing the final result of this tour as output.
-
-$UpdatePacked$, described in Algorithm \ref{alg:cons}, finds the best packed-pallet allocation, in terms of CG deviation, for the packed contents that remain on board.
-
-\begin{algorithm}[H]
-	\caption{Updating the packed contents that remain boarded at node $\pi_k$}  \label{alg:cons}
-	\begin{algorithmic}[1]
-
-		\State {\bf UpdatePacked} \,\, {\it in: ${M, Q_{\pi_k}, \pi_k}$} \, \, {\it out: $E_{Q_{\pi_k}}, M$}
-
-		\State $E_{Q_{\pi_k}} \gets MinCGDeviation(E_{Q_{\pi_k}})$   \label{cons:mincg}
-
-		\For{$i \gets 1$ to $m$} \label{cons:Ybegin}
-
-		\For{$q \gets 1$ to $m_{\pi_k}$}
-
-		\State $T_i^{\pi_k} \gets -1$
-
-		\If{$(i, q) \in E_{Q_{\pi_k}}$} 
-		\State $T_i^{\pi_k} \gets to_q$   \label{cons:dest} \Comment{reassign pallet destinations}
-		\EndIf
-		\EndFor		
-		\EndFor \label{cons:Yend}
-
-	\end{algorithmic}
-\end{algorithm}
-
-$MinCGDeviation$\/ (line \ref{cons:mincg}) relocates the packed contents on the pallets, minimizing torque and ensuring that they all remain on board, one packed content on each pallet.
-It is run through a MIP solver with the objective function (\ref{eq:torque}) and the constraints (\ref{eq:embarked}) and (\ref{eq:one}).
-As there are few variables, $E_{Q_{\pi_k}}$\/ is obtained in less than $30$\/ milliseconds.
-Finally, the destination of each pallet with packed content is updated (lines \ref{cons:Ybegin}-\ref{cons:Yend}).
-
-\begin{equation} \label{eq:torque}
-	\min f =  \Big | \sum_{i=1}^{m} \sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \times w_q \times D_i^{long}  \Big |
-\end{equation}
-
-\begin{equation} \label{eq:embarked}
-	\sum_{i=1}^{m} Y^k_{iq} = 1;\ q \in \{1,\ldots,m_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:one}
-	\sum_{q=1}^{m_{\pi_k}} Y^k_{iq} \leq 1;\ i \in \{1,\ldots,m\}
-\end{equation}
-
-$SetPalletsDestinations$, which sets the pallets destination not yet defined, is described in Algorithm \ref{alg:dest}.
-
-\begin{algorithm}[H]
-	\caption{Setting pallets destination based on the items to be embarked at node $\pi_k$}  \label{alg:dest}
-	\begin{algorithmic}[1]
-
-		\State {\bf SetPalletsDestinations} \,\, {\it in: $M, \pi_k$} \, \, {\it out: $M$}
-
-		\State $vol_x \gets  0$, $x \in L_{\pi_k}$  \label{dest:vector1}
-
-		\State $max \gets 0$  \Comment{destination with maximum volume demand}
-
-		\State $total \gets 0$
-
-		\For{$j \gets 1$ to $n_{\pi_k}$}
-
-		\If {$to_j \in L_{\pi_k}$}
-
-		\State $vol_{to_j} \gets vol_{to_j} + v_j$ 
-		\State $total \gets total + v_j$
-
-		\If {$vol_{to_j} > vol_{max}$}
-
-		\State $max \gets to_j$ \label{dest:max} 
-		\EndIf
-
-		\EndIf
-
-		\EndFor
-
-		\For{$x \in L_{\pi_k}$} \label{dest:propor1}
-
-		\If {$vol_{x} \neq 0$}
-
-		\State $needed \gets \max \{1, \lfloor{ (m-m_{\pi_k}) \times vol_{x}/total}\rfloor \}$  
-		\label{dest:needed}
-		\State $np \gets 0$
-
-		\For{$i \gets 1$ to $m$}
-
-		\If{($np < needed$) {\bf and} ($T_i^{\pi_k} = -1$)}  
-		\State $T_i^{\pi_k} \gets x$
-		\State $np \gets np + 1$ \Comment{number of necessary pallets to node $x$}
-		\EndIf
-
-		\EndFor
-		\EndIf
-
-		\EndFor \label{dest:propor2}
-
-		\For{$i \gets 1$ to $m$} \label{dest:max1}
-		\If{$T_i^{\pi_k} \gets -1$} 
-		\State $T_i^{\pi_k} \gets max$  \Comment{any remaining pallet is assigned to the maximum demand destination}
-		\EndIf
-		\EndFor \label{dest:max2}
-
-	\end{algorithmic}
-\end{algorithm}
-
-$vol$\/ stores the demand volume of items destined for the non-visited nodes (line \ref{dest:vector1}).
-The destination of empty pallets is defined proportionally to the volume of items to be embarked (lines \ref{dest:propor1}-\ref{dest:propor2}).
-$max$\/ is the destination with maximum volume demand (line \ref{dest:max}), and $needed$\/ is the number of necessary pallets to node $x$ (line \ref{dest:needed}).
-The destination with the maximum volume defines any remaining pallets (lines \ref{dest:max1}-\ref{dest:max2}).
-
-$ScoreAndTorque$, described in Algorithm \ref{alg:eval}, evaluates the allocation graph $G$\/ generated by $SolveNode$\/ at node $\pi_k$\/ and returns the corresponding cargo score and aircraft torque.
-
-\begin{algorithm}[H]
-	\caption{Cargo score and aircraft torque}  \label{alg:eval}
-	\begin{algorithmic}[1]
-
-		\State {\bf ScoreAndTorque} \,\, {\it in: $\pi_k, G$} \, \, {\it out: $s, \tau$}
-
-		\State Let $G(V_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$
-
-		\State $s \gets 0$
-		\State $\tau_i \gets 0$, $1 \leq i \leq m$
-		\For{$i \gets 1$ to $m$} \label{eval:loop1}
-
-		\For{$j \gets 1$ to $n_{\pi_k}$}
-
-		\If{$X_{ij}^{\pi_k} = 1$}
-		\State $s \gets s + s_j$ \label{eval:score1} \Comment{accumulates cargo score}
-		\State $\tau_i \gets \tau_i + w_j \times D_i^{long}$  \label{eval:eps1} \Comment{accumulates aircraft torque}
-		\EndIf
-		\EndFor
-
-		\For{$q \gets 1$ to $m_{\pi_k}$}
-		\If{$Y_{iq}^{\pi_k} = 1$}
-		\State $s \gets s + s_q$   \label{eval:score2} \Comment{accumulates cargo score}
-		\State $\tau_i \gets \tau_i + w_q \times D_i^{long}$ \label{eval:eps2} \Comment{accumulates aircraft torque}
-		\EndIf
-		\EndFor	   
-		\EndFor 		 \label{eval:loop2}
-		\State $\tau \gets  \sum_{i=1}^{m} \tau_i / (W_{max} \times limit^{CG}_{long})$ \label{eval:torque} \Comment{final calculation of the aircraft torque}
-
-	\end{algorithmic}
-\end{algorithm}
-
-Algorithm \ref{alg:eval}\/ consists of a loop that goes through all the pallets (lines \ref{eval:loop1}-\ref{eval:loop2}), accumulating the scores (lines \ref{eval:score1} and \ref{eval:score2}) and the torques (lines \ref{eval:eps1} and \ref{eval:eps2}) of the shipped items, allowing the final calculation of the aircraft torque (line \ref{eval:torque}).
-
-\subsection{Node-by-node solutions}
-\label{methods}
-
-In this subsection, we present two implementations of $SolveNode$\/ algorithm: with a MIP solver and with heuristics.
-
-\subsubsection{Node-by-node solutions with a MIP solver}
-\label{solver}
-
-Our strategy adopted in $SolveTour$\/ defines the values of some variables: the set of nodes to be visited is updated, the packed contents that remain on board are reallocated to minimize the CG deviation, and the pallet's destinations are determined according to the volume of items available for shipment.
-
-In this way, the mathematical model for $SolveNode(MIP,\pi_k,G, tmax)$\/ becomes simpler, which finds an allocation of available items at the node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$.
-Thus, we use a MIP solver with a runtime limit $tmax$\/ at the node $\pi_k$ to maximize the objective function (\ref{eq:obj2}) with the calculus equations (\ref{eq:newf}) to (\ref{eq:costs2}), subject to the constraints (\ref{eq:tau2}) to (\ref{eq:ifB}).
-The binary variables $X_{ij}$\/ and $Y_{iq}$\/ define the sets of edges $E_{N_{\pi_k}}$\/ and $E_{Q_{\pi_k}}$, respectively, included in graph $G$.
-
-\begin{equation} \label{eq:obj2}
-	\max f= \tilde{s} / \tilde{c}
-\end{equation}
-
-\begin{equation} \label{eq:newf}
-	\tilde{s} = \sum_{i=1}^{m} \sum_{j=1}^{n_{\pi_k}} X_{ij} \times s_j
-\end{equation}
-
-\begin{equation} 
-	\tau_{\pi_k} = \sum_{i=1}^{m} \Big [ D_i^{long} \times (\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j +  \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q) \Big ] \Big / W_{max} \times limit^{CG}_{long}
-\end{equation}
-
-\begin{equation} \label{eq:costs2}
-	\tilde{c} =  c_{\pi_k, \pi_{k+1}}\times(1+c_g\times|\tau_{\pi_k}|)
-\end{equation}
-
-\begin{equation} \label{eq:tau2}
-	|\tau_{\pi_k}| \leq 1
-\end{equation}
-
-\begin{equation} 
-	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times w_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times w_q  \leq W_i; \ i \in \{1, \ldots, m\}
-\end{equation}
-
-\begin{equation} 
-	\sum_{j=1}^{n_{\pi_k}} X_{ij} \times v_j + \sum_{q=1}^{m_{\pi_k}} Y_{iq} \times v_q  \leq\ V_i; \ i \in \{1, \ldots, m\}
-\end{equation}
-
-\begin{equation} 
-	\sum_{i=1}^{m} X_{ij} \leq 1; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:if2}
-	X_{ij} = 0;\ to_j \notin L_{\pi_k}; \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:ifA}
-	X_{ij} \leq X_{ij} \times (T_i^{\pi_k} - to_j + 1); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-\begin{equation} \label{eq:ifB}
-	X_{ij} \leq X_{ij} \times (to_j - T_i^{\pi_k} + 1 ); \ i \in \{1, \ldots, m\}; \ j \in \{1, \ldots, n_{\pi_k}\}
-\end{equation}
-
-The constraints (\ref{eq:ifA}) and (\ref{eq:ifB}) are equivalent to $X_{ij} =1$\/ if $to_j = T_i^{\pi_k}$, and $X_{ij} =0$\/ otherwise .
-
-\subsubsection{Node-by-node solutions with heuristics}
-
-One of the main objectives of this work was to find a quick heuristic that offers a good-quality solution for the node-by-node problem.
-Taking this into account, we design algorithms based on five known meta-heuristics: {\it Ant Colony Optimization} (ACO) \citep{Dorigo1992,DorigoManiezzoColorni1996}, {\it Noising Method Optimization} (NMO) \citep{CharonHudry1993,CharonHudry2001,Zhan2020}, {\it Tabu Search} (TS) \citep{Glover1986}, {\it Greedy Randomized Adaptive Search Procedure} (GRASP) \citep{FeoResende1989}, and {\it Genetic Algorithm} (GA) \citep{holland1992adaptation}.
-We considered several ideas from the literature \citep{NiarFreville1997, Fidanova2006, Alonso2019, Zhan2020, shah2020genetic,PeerlinckAmy2022MFEO}, and we were careful to use the same data structures and procedures in all implementations to enforce fair results comparison.
-
-However, the heuristic that presented better solutions was none of the previous ones.
-In this subsection, we will present a new heuristic for the node-by-node problem, called {\it Shims}.
-Like in mechanics, shims are collections of spacers to fill gaps, which may be composed of parts with different thicknesses (see Figure \ref{fig:shims}).
-This strategy is based on a practical observation: usually, subsets of smaller and lighter items are saved for later adjustments to the remaining available space.
-
-The selection of edges for $E_{ N_{\pi_k} }$ uses the {\it edge attractiveness}\/ $\theta_{ij}$\/ (\ref{eq:edge}), which can be understood as the tendency to allocate the item $j$\/ to the pallet $i$\/ at the node $\pi_k$.
-It is directly proportional to the score, and inversely to the volume and the torque of each item.
-
-\begin{equation} \label{eq:edge}
-	\theta_{ij} = \frac{ s_j }{ v_j }\ \times \Big ( 1 - \frac{ w_j \times |D_i^{long}| }{ \max w_j \times\  \max | D_i^{long} | } \Big );\ i \in \{1,\ldots,m\},\ j \in \{1,\ldots,n_{\pi_k}\}
-\end{equation}
-
-\begin{table}[H]
-
-	\begin{minipage}{0.08\linewidth}
-
-	\end{minipage}\hfill
-	\begin{minipage}{0.34\linewidth}
-
-		\includegraphics[scale=0.25]{shims.png}
-		\captionof{figure}{{\it Shims} of various thicknesses}
-		\small\textsuperscript{Source: www.mscdirect.com/product/details/70475967}
-		\label{fig:shims}
-
-	\end{minipage}\hfill
-	\begin{minipage}{0.58\linewidth}
-		\scriptsize
-		\begin{tikzpicture}[scale=0.75, samples=100]
-			\filldraw[fill=blue!3!white, draw=black] (0, 0) rectangle (12.5, 5.5);
-			\draw[->] (.2, .6) -- coordinate (x axis mid) (12, .6);
-			\node at (5, 0.3) {$\eta_1$};
-			\node at (5, 2.5) {|};
-			\node at (6, 2.5) {\small{shims}};
-			\node at (7, 2.5) {|};
-			\node at (7, 0.3) {$\eta_2$};
-			\node at (0.3, 5) {$\theta_{ij}$};
-			\draw[->] (.6, .2) -- coordinate (y axis mid) (0.6, 5.3);
-			\node at (12, 0.3) {$e_{ij}$};
-			\draw[smooth, domain = 0.09:2, color=black, thick] plot (.3+1/\x,{4.2+log2(\x)});
-			\draw[->, dashed] (5, 0.6)--(5, 2.0);
-			\draw[->, dashed] (7, 0.6)--(7, 1.5);
-		\end{tikzpicture}
-
-		\captionof{figure}{$n_{\pi_k}$\/ possible edges $e_{ij}$\/ sorted by $\theta_{ij}$\/ in non-ascending order}
-		\label{fig:whip}		
-	\end{minipage}
-\end{table}
-
-Considering only the items that can be shipped at node $\pi_k$, Figure \ref{fig:whip} represents $n_{\pi_k}$\/ possible edges $e_{ij}$\/ of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order.
-Initially, {\it Shims}\/ builds a greedy solution for the pallet $i$\/ selecting edges up to index $\eta_1$ ({\it greedy phase}).
-Then, with the edges between $\eta_1$\/ and $\eta_2$, it elaborates different possible complements ({\it composition phase}), including later the best ones in the same pallet ({\it selection phase}).
-{\it Shims}\/ is depicted in Algorithm \ref{alg:shims}.
-
-\begin{algorithm}[H]
-\small
-	\caption{{\it Shims}\/ heuristic at node $\pi_k$}  \label{alg:shims}
-	\begin{algorithmic}[1]
-
-		\State {\bf SolveNode} \,\, {\it in: {\it Shims}, $\pi_k, G, tmax, level_1, level_2$} \, \, {\it out: $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}} \cup E_{N_{\pi_k}})$}
-
-		\State $T_{begin} \gets$ current system time
-
-		\State Let $G(M \cup N_{\pi_k} \cup Q_{\pi_k}, E_{Q_{\pi_k}})$   \label{shims:initQ}
-
-		\State Sort $M$ by $|D_i^{long}|$ in non-descending order \label{shims:pallets}
-
-		\State $E_{N_{\pi_k}} \gets \varnothing$
-
-		\State $\tau_{max} \gets W_{max} \times limit^{CG}_{long}$
-
-		\For{$i \gets 1$ to $m$}  \label{shims:eta1_a}
-
-		\State $\tau_{\pi_k} \gets \sum_{(i,q)\in E_{Q_{\pi_k}}} w_q \times D_i^{long}$ 
-		\State $vol_i \gets \sum_{(i,q)\in E_{Q_{\pi_k}}} v_q$
-
-		\State Let $E$\/ be an array of $n_{\pi_k}$ possibles edges of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order \label{shims:possible}
-		\State $\eta_1 \gets 1$  
-		\Repeat
-		\State $e_{ij} \gets E_{\eta_1}$
-
-		\If{ ($E_{N_{\pi_k}} \cup \{e_{ij}\}$ is feasible) {\bf and} ($vol_i \leq V_i \times level_1$) {\bf and} ($| \tau_{\pi_k} + w_j \times D_i^{long} | \leq W_{max} \times limit^{CG}_{long}$)}
-
-		\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup \{e_{ij}\}$
-
-		\State $vol_i \gets vol_i + v_j$
-
-		\State $\tau_{\pi_k} \gets \tau_{\pi_k} + w_j \times D_i^{long}$
-
-		\State $\eta_1 \gets \eta_1 + 1$
-
-		\EndIf
-		\Until ($vol_i > V_i \times level_1$) {\bf or} ($\eta_1>n_{\pi_k}$) \label{shims:endfirst}
-
-		\State $slack_i \gets V_i - vol_i$  \label{shims:beginsecond}
-		\State $\eta_2 \gets \eta_1$  \label{shims:eta2a}
-		\While{($\eta_2 \leq n_{\pi_k}$) {\bf and} ($ vol_i < V_i \times level_2$)}
-
-		\State $e_{ij} \gets E_{\eta_2}$
-
-		\State $vol_i \gets vol_i + v_j$
-		\State $\eta_2 \gets \eta_2 + 1$  \label{shims:eta2} 
-		\EndWhile \label{shims:endsecond}
-
-		\State $vol \gets 0$; $b \gets 1$; $shims_b \gets \varnothing$; $Set \gets \{ shims_b \}$  \label{shims:beginthird}
-
-		\For{$x \gets \eta_1$ to $\eta_2$} \label{edges_indexes}
-
-		\If{ $T_{current} - T_{begin} > tmax$}
-		\State {\bf break}
-		\EndIf
-
-		\State $NewShims \gets$ {\bf True} \label{new_shims}
-		\State $e_{ij} \gets E_x$
-
-		\For{$shims \in Set$} \label{shims_set}
-
-		\If {($e_{ij} \not\in (E_{N_{\pi_k}} \cup shims))$ {\bf and} ($e_{ij}$ is feasible)  {\bf and} ($(v_j + vol) \leq slack_i$)}
-
-		\State $shims \gets shims \cup \{e_{ij}\}$
-		\State $vol \gets vol + v_j$
-		\State $NewShims \gets$ {\bf False} \label{new_shims_false}
-
-		\State {\bf break}  
-		\EndIf
-
-		\EndFor
-
-		\If{$NewShims$} \label{new_shims2}
-		\State $vol \gets 0$; $b \gets b + 1$;  $shims_b \gets \{e_{ij}\}$
-		\State $Set \gets Set \cup \{ shims_b \}$
-		\EndIf
-
-		\EndFor
-
-		\State $sh_w \gets shims$, where $shims \in Set$ and $\sum_{e_{ij} \in shims} w_j$\/ is maximum  \label{best_weight}		
-		\State $sh_v \gets shims$, where $shims \in Set$ and $\sum_{e_{ij} \in shims} v_j$\/ is maximum \label{best_volume}
-		\State $sh_{best} \gets shims$, where $shims \in \{sh_w, sh_v\}$\/ and $\sum_{e_{ij} \in shims} s_j$\/ is maximum \label{best_score}
-		\State $E_{N_{\pi_k}} \gets E_{N_{\pi_k}} \cup sh_{best}$ \label{shims:endthird}
-
-		\EndFor
-
-	\end{algorithmic}
-\end{algorithm}
-
-$tmax$\/ is the runtime limit for {\it Shims}, and $level_1$\/ and $level_2$\/ are volume thresholds for the indices $\eta_1$ and $\eta_2$, respectively.
-
-Initially, $Q_{\pi_k}$\/ (line \ref{shims:initQ}) corresponds to the packed contents that remain on board.
-It is important to remember that $E_{Q_{\pi_k}}$\/ and $M$\/ were modified by the procedure $UpdatePacked(M, Q_{\pi_k}, \pi_k)$ and the procedure $SetPalletsDestinations(M, \pi_k)$.
-Then, the pallets $i$\/ are considered in non-descending order of $|D_i^{long}|$.
-
-For each pallet $i$, its $n_{\pi_k}$\/ possible edges $e_{ij}$\/ are considered in non-increasing order of $\theta_{ij}$:
-
-\begin{itemize}
-
-	\item In the {\it greedy phase}\/ (lines \ref{shims:pallets}-\ref{shims:endfirst}), a partial solution for each pallet $i$\/ is constructed by adding edges following $\theta_{ij}$\/ order.
-The indices $\eta_1$\/ and $\eta_2$\/ refer to the accumulated volumes $V_i \times level_1$\/ and $V_i \times level_2$, respectively, which were defined empirically by the {\it irace}\/ tool \citep{LopezIbanezManuel2016}.
-We will explain this in detail in the subsection \ref{items}.
-
-	\item In the {\it composition phase}\/ (lines \ref{shims:beginsecond}-\ref{shims:endsecond}), a set of shims named $Set$\/ is created for each pallet $i$, where each shim is formed by a set of edges in the range $[\eta_1,\eta_2]$, whose total volume is limited by $slack_i$.
-In this phase, the heuristic that provided the best results, both in terms of time and quality, is based on {\it First-Fit Decreasing}, which is an approximation algorithm for the {\it Bin Packing Problem}\/ \citep{JohnsonGarey1985}.
-Basically, shims are created by accumulating the following edges, taking $slack_i$\/ as a limit.
-
-	\item In the {\it selection phase}\/ (lines \ref{shims:beginthird}-\ref{shims:endthird}), the best shim in $Set$\/ is chosen.
-Initially, two shims are found: $sh_w$\/ with larger weight and $sh_v$\/ with larger volume.
-Between the two, the one with the highest score will be chosen, and its edges will be inserted into $E_{N_{\pi_k}}$.
-
-\end{itemize}
 
@@
 
-\section{Implementation and results}
-\label{results}
-
-This section is composed of two parts: the generation of the test instances and the results obtained in our implementation.
-
-\subsection{Instances generation}
-\label{items}
-
-As we are dealing with a new problem that until now had not been modelled in the literature, we have to create our own benchmarks.
-For this, we based it on the characteristics of real airlifts carried out by the {\it Brazilian Air Force}, as described below.
-
-In the delivery of supplies carried out in Brazil from 2008 to 2010, 23\% of the items weighed between $10kg$ and $20kg$, 22\% from $21kg$ to $40kg$, 24\% from $41kg$ to $80kg$, 23\% from $81kg$ to $200kg$, and 8\% between $201kg$ and $340kg$.
-These five groups of items are described in Table \ref{tab:weights}, where $P$\/ represents the group probability.
-On the other hand, the average density of these items is approximately $246 kg/m^3$.
-
-\begin{table}[h]
-	\centering
-	\caption{Items weight distribution}  \label{tab:weights}
-	\begin{tabular}{c c c c }
-		\toprule
-		$item$ & $P$ & $low$ ($kg$) & $high$ ($kg$) \\		
-		\midrule
-		1              & 0.23           & 10  & 20  \\
-		2              & 0.22           & 21  & 40  \\
-		3              & 0.24           & 41  & 80  \\		
-		4              & 0.23           & 81  & 200 \\
-		5              & 0.08           & 201 & 340 \\
-		\bottomrule
-	\end{tabular}
-\end{table}
-
-In the generation of test instances, we use two types of random selections:
-
-\begin{itemize}
-
-	\item $RandomInt(i_1,i_2)$: randomly selects a integer number in $[i_1,i_2]$, where $i_1$ and $i_2$ are integer numbers;
-
-	\item $Roulette(set)$ biased through $\phi$: selects an element from $set$, where the probability of each element is proportional to the value of a given function $\phi$\/ defined on $set$.
-
-\end{itemize}
-
-The procedure $ItemsGeneration$, which generates $N$ (all items to be moved among the nodes), is described in Algorithm \ref{alg:itemsgen}.
-
-\begin{algorithm}[H]
-	\caption{Generating items}  \label{alg:itemsgen}
-	\begin{algorithmic}[1]
-		\State {\bf ItemsGeneration} \,\, {\it in: $scenario, surplus$} \, \, {\it out: $N$}
-		\State Let $L$ be the set of nodes and $M$ the set of pallets  \label{ig:LM}
-		\State $limit \gets surplus \times \sum_{i=1}^{m} V_i$ \label{ig:extended}
-		\For{$k \gets 0$ to $K$}
-		\State $N_k \gets \varnothing$
-		\State $j \gets 0$
-		\State $vol \gets 0$	\label{ig:totals}	
-		\While{$vol < limit$}
-		\State $j \gets j+1$
-		\State Let $t_j^k$\/ be the item $j$\/ at the node $k$
-		\Repeat
-		\State $to_j \gets RandomInt(0, K)$ \label{ig:dest}
-		\Until{$to_j \neq k$}
-		\State $x = Roulette(item)$ biased through $P$ \label{ig:weight1} \Comment{From Table \ref{tab:weights}}	 
-		\State $w_j \gets RandomInt(low(x), high(x))$        \label{ig:weight2}		
-		\State $s_j \gets \round{100 \times (1 - \log_{10}(RandomInt(1, 9)))} $ \label{ig:score}
-		\State $v_j \gets w_j / RandomInt(148, 344)$ \label{ig:volume}
-		\State $vol \gets vol + v_j$ 
-		\State $N_k \gets N_k \cup \{t_j^k\}$ 
-		\EndWhile
-		\EndFor
-		\State $N \gets \bigcup_{0 \leq k \leq K} N_k$
-	\end{algorithmic}
-\end{algorithm}
-
-$scenario$\/ defines $L$\/ and $M$\/ (line \ref{ig:LM}), and the argument $surplus$\/ sets a limit on the total volume of items at each node (line \ref{ig:extended}).
-To avoid simply loading all items, we use $surplus \in \{1.2,\ 1.5,\ 2.0\}$\/.
-This also represents more instances for tests in each scenario.
-
-For each generated $t^k_j$ item, its destination is randomly selected (line \ref{ig:dest}), its weight has a distribution according to Table \ref{tab:weights} (lines \ref{ig:weight1}-\ref{ig:weight2}), its score varies $100$\/ (highest) and $5$\/ (lowest) according to a logarithmic scale (line \ref{ig:score}), and its volume is randomly defined from the density, where we allow a variation of 40\% around the average density of $246 kg/m^3$\/ (line \ref{ig:volume}).
-
-To determine the parameters $level_1$\/ and $level_2$, used by \textit{Shims}, we previously carried out some experiments with the tool {\it irace}\/ \citep{LopezIbanezManuel2016}, the results of which are presented in Table \ref {tab:irace}.
-In these tests, of every 7 instances generated for each value of $surplus$, 4 were used as the training set and 3 as the testing set.
-We provided the ranges [0.8, 1.0] and [1.0, 2.0] for $level_1$\/ and $level_2$, respectively.
-In each experiment, there was a maximum of 3,000 runs so that \textit{irace}\/ would have enough data for its statistical tests.
-For more details, see {\tt cran.r-project.org/web/packages/irace/}.
-
-	\begin{table}[H]
-	\centering
-	\caption{\textit{irace} testing results}  \label{tab:irace}
-	\begin{tabular}{cccc}
-		\toprule
-
-		$surplus$ & $level_1$ & $level_2$ & runtime (min)\\
-
-		\midrule
-		1.2& 0.8621 & 1.0539 & 47 \\
-		1.5& 0.9199 & 1.1399 & 59 \\
-		2.0& 0.9617 & 1.5706 & 63 \\						
-		\bottomrule
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\subsection{Obtained solutions: quality and runtimes}
-
-In the tests performed, we used a 64-bit, 16 GB, 3.6 GHz, eight-core processor with {\it Linux Ubuntu 22.04.1 LTS 64-bit} as the operational system and {\it Python 3.10.4}\/ as the programming language.
-We also used the well-known solver {\it Gurobi}\/ ({\tt www.gurobi.com}), version 9.5.2.
-
-We will first present the results obtained in practical cases, that is, when $1<K\leq6$. Next, we will show how the same {\it Shims}\/ heuristic remains robust when $K>6$.
-
-\subsubsection{Results when $1<K\leq6$}
-
-We ran Algorithm \ref{alg:main} considering the 5 scenarios from Table \ref{tab:scenarios}, 3 values for  $surplus$\/ from $\{1.2, 1.5, 2.0\}$, 4 values for $tmax$\/ from $\{240s, 1200s, 2400s, 3600s\}$, and 7 different methods for the node-by-node solution: {\it Gurobi}\/ (see \ref{solver}), ACO, NMO, TS, GRASP, GA and {\it Shims}\/ (Algorithm \ref{alg:shims}).
-
-For {\it Gurobi}\/ to be able to solve the largest possible number of tests without memory overflow, we set its parameter {\it MIPgap}\/ to 1\%.
-This shortens its runtime, in addition to ensuring that its objective function $f$\/ is at most 1\% of the optimal solution.
-For more details, see {\tt www.support.gurobi.com}.
-
-For each $scenario$, $surplus$\/ and $tmax$\/ tested, 7 different instances were generated.
-Table \ref{tab:overall} succinctly shows the overall performance of the methods for the node-by-node solution.
-Only {\it Shims}\/ found a solution for all scenarios, as well as being the fastest.
-
-\begin{table}[H]
-
-	\centering
-	\caption{Overall results}  \label{tab:overall}
-	\small
-	\begin{tabular}{cccc}	
-	\toprule
-	{\bf Method} & {\bf Best scenarios} & {\bf Worst scenarios} & {\bf Worst runtimes (min)}  \\
-	\midrule
-	NMO           & 4       & 5    & 60  \\
-	ACO           & 2, 3    & 4, 5 & 25, 61  \\
-	GRASP         & 1       & 4, 5 & 28, 55  \\
-	TS            & -       & 5    & 44  \\
-	GA            & -       & 1, 2, 3, 4, 5  & did not solve  \\
-	\midrule
-	{\it Gurobi}  & 1, 2, 3, 4    & 5    & did not solve  \\
-	{\it Shims}   & 1, 2, 3, 4, 5 & -    & 3.26 \\
-	\bottomrule
-	\end{tabular}
-\normalsize
-
-\end{table}
-
-For this reason, we will present only the results obtained by {\it Shims}, with the average of the objective function $f$\/ and the runtime of {\it Gurobi}\/ and {\it Shims}.
-To facilitate the comparison between both, we added a last column in the tables where two values are indicated:
-
-\begin{itemize}
-
-	\item {\bf Normalized}: value between 0 and 1, which corresponds to the ratio between the sum of $f$\/ values obtained by the method in all scenarios and the sum of the best values obtained by both methods in all scenarios.
-The higher the value of {\bf Normalized}, the closer the method approached the best solutions found.
-
-	\item {\bf Speed-up}: ratio of the sums of the runtimes of all scenarios and the sum of the method runtimes in all scenarios. The method with the highest {\bf Speed-up}\/ is the fastest.
-
-\end{itemize}
-
-We also indicate the two adopted strategies: dedicating all the processing time to the 2 shortest tours or distributing it among all $K!$\/ tours.
-The results obtained with $tmax = 3600s$, which is the highest tested runtime limit, are in Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, with $surplus$\/ values of $1.2$, $1.5$\/ and $2.0$, respectively.
-We indicate with an {\bf x} the cases where {\it Gurobi}\/ did not find a feasible solution within this runtime limit or had to be aborted due to high random-access memory (RAM) usage.
-
-\begin{table}[htp] 
-	\centering
-	\caption{Solutions with $surplus = 1.2$\/ and $tmax =  3600s$}  \label{tab:20}
-	\footnotesize
-	\begin{tabular}{ccccccccc}
-		\toprule
-		$ntours$&$method$&$scenario$&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&\specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
-		\toprule
-		\multirow{4}{*}{2} &\multirow{2}{*}{{\it Gurobi}}  & $f$ & 8.53    & 11.79   & 13.14   & 13.52 & 12.36       &  0.9998 \\ % best 5935
-		&                              &               {time (s)}   & 29      & 28      & 25      & 27    & 27          &  1.0    \\ % longer 136
-		\cmidrule(lr){2-9}
-		&                           \multirow{2}{*}{{\it Shims}}  & $f$ & 8.54    & 11.78   & 13.06   & 13.51 & 12.34       &  0.9980 \\
-		&                             &                {time (s)}   & 1       & 1       & 1       & 1     & 2           &  22.7   \\
-
-		\midrule
-
-		\multirow{4}{*}{$K!$}&  \multirow{2}{*}{{\it Gurobi}} & $f$ & 8.60    & 12.20   & 13.66   & 15.00  & {\bf x}    & 0.9998 \\ % best 4947
-		&                              &               {time (s)}   & 30      & 35      & 123     & 314    & {\bf x}    & 1.0 \\
-		\cmidrule(lr){2-9}	
-		&                           \multirow{2}{*}{{\it Shims}}  & $f$ & 8.61    & 12.20   & 13.46   & 14.99  & 13.35      & 0.9958  \\
-		&                              &               {time (s)}   & 1       & 1       & 4       & 10     & 51         & 7.49   \\
-		\bottomrule
-
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\begin{table}[htp]
-	\centering
-	\caption{Solutions with $surplus = 1.5$\/ and $tmax =  3600s$}  \label{tab:50}
-	\footnotesize
-	\begin{tabular}{ccccccccc}
-		\toprule
-		$ntours$ & $method$          & $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
-		\toprule
-		\multirow{4}{*}{ 2 } & \multirow{2}{*}{{\it Gurobi}} & $f$ & 11.83 & 16.73 & 18.07 & 18.83 & 16.86 & 0.9996 \\ % best 8235
-		&&               {time (s)}                                   & 55    & 64    & 39    & 40    & 88    & 1.0    \\ 
-		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{{\it Shims}}                               & $f$ & 11.85 & 16.72 & 18.05 & 18.80 & 16.87 & 0.9993  \\
-		&&         {time (s)}                                         & 1     & 1     & 2     & 2     & 2     & 35.8  \\
-
-		\midrule
-
-		\multirow{4}{*}{$K!$} & \multirow{2}{*}{{\it Gurobi}}   & $f$ & 11.83 & 16.93 & 18.40 & 20.95 & 17.60 & 0.9999 \\  % best 8469
-		&&               {time (s)}                                   & 63    & 59    & 195   & 472   & 2258  & 1.0  \\
-		\cmidrule(lr){2-9}	
-		&\multirow{2}{*}{{\it Shims}}                               & $f$ & 11.85 & 16.91 & 18.36 & 20.93 & 17.50 & 0.9976  \\ 
-		&&         {time (s)}                                         & 1     & 2     & 5     & 15    & 100   & 23.8 \\
-		\bottomrule
-
-	\end{tabular}
-	\normalsize
-\end{table}
-
-\begin{table}[htp]
-	\centering
-	\caption{Solutions with $surplus = 2.0$\/ and $tmax =  3600s$}  \label{tab:100}
-	\footnotesize
-	\begin{tabular}{ccccccccc}
-		\toprule
-		$ntours$& $method$& $scenario$ & {\bf 1} & {\bf 2} & {\bf 3} & {\bf 4} & {\bf 5} & \specialcell{{\bf Normalized}\\{\bf Speed-up}}   \\
-		\toprule
-		\multirow{4}{*}{2} & \multirow{2}{*}{{\it Gurobi}} & $f$ & 17.70 & 24.20 & 26.39 & 27.17 & 24.20 & 0.9995 \\ % best 11972
-		&&                                              {time (s)}  & 168   & 98    & 79    & 70    & 72    & 1.0  \\
-		\cmidrule(lr){2-9}	
-		&                            \multirow{2}{*}{{\it Shims}} & $f$ & 17.74 & 24.22 & 26.32 & 27.07 & 23.13 & 0.9896  \\
-		&&                                              {time (s)}  & 1     & 2     & 2     & 3     & 4     & 40.6  \\
-
-		\midrule
-
-		\multirow{4}{*}{$K!$} & \multirow{2}{*}{{\it Gurobi}} & $f$ & 17.90 & 25.44 & 26.51 & 29.13 &{\bf x}& 0.9994 \\ % best 9867
-		&&                                               {time (s)} & 178   & 143   & 378   & 862   &{\bf x}& 1.0  \\
-		\cmidrule(lr){2-9}	
-		&                         \multirow{2}{*}{{\it Shims}}    & $f$ & 17.94 & 25.45 & 26.44 & 28.84 & 26.22  & 0.9970  \\
-		&&                                               {time (s)} & 1     & 3     & 10    & 31    & 196    & 34.7  \\
-		\bottomrule
-
-	\end{tabular}
-	\normalsize
-\end{table}
-
-From these data, we can draw some conclusions:
-
-\begin{itemize}
-
-	\item The strategy of testing all $K!$\/ tours often provide a better-quality solution, even with less time on each node.
-This shows that the four sub-problems are interconnected in such a way that it is not enough to solve them separately.
-
-	\item {\it Gurobi}\/ fails in some cases when $scenario=5$\/ and the strategy is to check all $K!$\/ tours.
-This occurs because the runtime limit per node is smaller and there tend to be more packed contents on the aircraft, reducing the space for allocating items and making the solution difficult.
-
-	\item When {\it Gurobi}\/ finishes, it finds the best solution, but the one obtained by {\it Shims}\/ reaches at least $98.96\%$\/ of that value.
-Considering only the strategy of testing all $K!$ tours, this value increases to $99.58\%$.
-
-	\item {\it Shims}\/ always finds a solution, being 7 to 40 times faster.
-
-	\item All runtimes are much lower than the limit because the solution on many nodes can be fast.
-Anyway, in all the tests performed, the maximum time spent by {\it Shims}\/ did not reach 4 minutes.
-On the other hand, when $scenario=5$\/ and $surplus=1.5$, {\it Gurobi}\/ spent almost 40 minutes.
-
-\end{itemize}
-
-Table \ref{tab:100time} shows the results obtained with the strategy of testing the $K!$\/ tours in all scenarios with different $tmax$.
-We can observe more cases where {\it Gurobi}\/ fails, even in smaller scenarios.
-When {\it Gurobi}\/ finishes, {\it Shims}\/ finds a solution of similar quality ($99\%$\/ or better).
-In all cases, {\it Shims}\/ finds a solution in less than 4 minutes.
-
-\begin{table}[H]
-	\centering
-	\caption{Solutions testing all $K!$\/ tours with different runtime limits}  \label{tab:100time}
-	\scriptsize
-	\setlength{\tabcolsep}{3.5pt}
-	\begin{tabular}{ccc|ccccc|ccccc|ccccc}
-		\toprule
-		&             \multicolumn{2}{r}{$surplus$} &\multicolumn{5}{c}{\bf 1.2}&\multicolumn{5}{c}{\bf 1.5}&\multicolumn{5}{c}{\bf 2.0} \\
-		\midrule
-		$method$& {$tmax$} & $scenario$  &{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5}&{\bf 1}&{\bf 2}&{\bf 3}&{\bf 4}&{\bf 5} \\
-		\toprule
-
-		\multirow{10}{*}{{\it Gurobi}} & \multirow{2}{*}{240s} & $f$ & 8.60 & 12.20 & 13.67 & {\bf x} & {\bf x} & 11.77 & 16.38 & 18.10 & {\bf x} & {\bf x} & 17.89 & 25.42 & {\bf x} & {\bf x} & {\bf x} \\ 
-		&               &  {time (s)}                            & 31   & 35    & 124   & {\bf x} & {\bf x} & 52    & 59    & 200   & {\bf x} & {\bf x} & 188   & 145   & {\bf x} & {\bf x} & {\bf x} \\ 
-		\cmidrule(lr){2-18}
-
-		&                           \multirow{2}{*}{1200s} & $f$ & 8.61 & 12.20 & 13.31 & 10.00   & {\bf x} & 11.77 & 17.02 & 18.25 &  20.64  & {\bf x} & 17.75 & 25.24 & 26.49 & 27.97   & {\bf x} \\ 
-		&        &  {time (s)}                                   & 28   & 37    & 129   & 320     & {\bf x} & 46    & 61    & 190   &  304    & {\bf x} & 161   & 139   & 384   & 579     & {\bf x} \\ 
-		\cmidrule(lr){2-18}                              
-		&                           \multirow{2}{*}{2400s} & $f$ & 8.60 & 12.21 & 13.67 & 15.00   & 13.41   & 11.77 & 16.37 & 18.03 & 20.95   & {\bf x} & 17.89 & 25.44 & 26.15 & 29.13   & {\bf x} \\
-		&        &  {time (s)}                                   & 26   & 38    & 134   & 310     & 1520    & 46    & 60    & 199   & 461     & {\bf x} & 164   & 140   & 383   & 786     & {\bf x} \\
-		\cmidrule(lr){2-18}
-
-		&                           \multirow{2}{*}{3600s} & $f$ & 8.60 & 12.20 & 13.66 & 15.00   & {\bf x} & 11.76 & 16.37 & 18.01 &  20.95  & 17.60   & 17.90 & 25.44 & 26.15 & 29.13   & {\bf x} \\ 
-		&        &  {time (s)}                                   & 30   & 35    & 123   & 314     & {\bf x} & 64    & 58    & 195   &  472    & 2258    & 178   & 143   & 378   & 862     & {\bf x} \\
-
-		\midrule
-
-		\multirow{2}{*}{{\it Shims}} &  \multirow{2}{*}{240s}  & $f$ & 8.61 & 12.20 & 13.46 & 14.99   & 13.35   & 11.78 & 16.29 & 17.99 &  20.93  & 17.50   & 17.94 & 25.45 & 26.14 & 28.84   & 26.22 \\ % 
-		&         & {time (s)}                                   & 1    & 2     & 4     & 10      & 51      & 1     & 2     & 5     &  15     & 100     & 1     & 3     & 10    & 31      & 196   \\
-
-		\bottomrule
-	\end{tabular}
-	\normalsize
-\end{table}
-
-The actual RAM consumption of {\it Gurobi}\/ was over 8.5 GB, and all of {\it Shims}\/'s executions consumed at most 1.5 GB of RAM.
-
-\subsubsection{Results when $K>6$}
-
-These last results that we will show here do not correspond to practical cases of air transport, as tours where $K>6$ very rarely occur.
-However, it is possible to see that the {\it Shims} heuristic maintains robust behaviour as the number of nodes grows, that is, it could be adapted to similar contexts
-({\color{red}vehicles and warehouses} {\color{blue}ships and trucks})
-, for example), where there may be more nodes.
-
-In the tests carried out, we considered real data from the 15 main Brazilian airports.
-For this task, we implemented a TSP heuristic based on GA that returned 100 tours in approximately 33 seconds.
-We implemented this heuristic with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework available in {\tt github.com/deap/deap}.
-For more details, see \citep{DEAP_JMLR2012}.
-
-Figure \ref{fig:performance} shows the runtime curve of {\it Shims}\/ as the number of nodes increases, indicating at each point the value found for the objective function $f$.
-Runtime and $f$\/ are the averages obtained from 7 instances generated with $surplus=2.0$\/ for each value of $K$.
-In Figure \ref{fig:morenodes}, we indicate one of the tours found by the TSP heuristic when $K=15$.
-
-\begin{table}[htp]
-
-	\begin{minipage}{0.58\linewidth}
-	\centering
-	\begin{tikzpicture}
-	\footnotesize
-		\begin{axis}[xmin=6,ymin=200,
-			enlarge y limits={upper,value=0.2},
-			xlabel style={align=center},
-			xlabel={Number $K$\/ of nodes},
-			ylabel={Runtime (s)},
-			xtick={ 5, 6, 7,  8,  9, 10, 11, 12, 13, 14, 15, 16},
-			ytick={250,350,450,550,650,750,850,950,1050,1150,1250},						
-			]
-			\addplot[
-			blue,
-			thick,
-			mark=*,
-			mark options={fill=blue},
-			visualization depends on=\thisrow{alignment} \as \alignment,
-			nodes near coords, % Place nodes near each coordinate
-			point meta=explicit symbolic, % The meta data used in the nodes is not explicitly provided and not numeric
-			every node near coord/.style={anchor=\alignment} % Align each coordinate at the anchor 40 degrees clockwise from the right edge
-			] table [% Provide data as a table
-			meta index=2 % the meta data is found in the third column
-			] {
-				x   y    label                  alignment
-				7   273   {\footnotesize 92.6}  0
-				8   360   {\footnotesize 107.5} 0
-				9   468   {\footnotesize 107.5}  0
-				10  560   {\footnotesize 112.3}  0
-				11  660   {\footnotesize 92.8}  0
-				12  756   {\footnotesize 91.2} 0
-				13  884   {\footnotesize 96.8}  0
-				14  1008   {\footnotesize 106.3} 0									
-				15  1170   {\footnotesize 99.8}  0									
-			};
-		\end{axis}
-	\end{tikzpicture}
-	\captionof{figure}{{\it Shims}\/ performance with $surplus=2.0$}
-	\label{fig:performance}
-	\end{minipage}\hfill 
-	\begin{minipage}{0.42\linewidth}
-		\centering
-		\includegraphics[scale=0.293]{brazil.png}
-		\captionof{figure}{A tour where $K=15$}
-		\label{fig:morenodes}
-	\end{minipage}
-\end{table}
-
-\section{Conclusions}
-\label{conclusions}
-
-In this work, we modelled and solved a real air transport problem named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem}\/ (ACLP+RPDP).
-For the first time in the literature, a {\it NP-hard}\/ problem that involves {\it simultaneously}\/ pallet assembly, load balancing, and route planning is addressed, where the cost-effectiveness of transport is maximized.
-Currently, there is no commercial software available for this problem.
-
-We adopted some simplifications that are not critical, but that allowed for an unprecedented solution to this problem considering more than two nodes.
-In practical cases, the number $K$\/ of nodes, excluding the base, is small ($K \leq 6$), each of which has hundreds of items to be shipped.
-Considering a real aircraft, we have developed some node-by-node solutions.
-The complete process can be executed quickly on a handheld computer, offering good results and reducing stress for the transport planners.
-
-As validation, we carried out tests in several real air transport scenarios.
-In these cases, the solution process can establish, in less than four minutes, a flight tour for a single aircraft with a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximising the total score, and minimising fuel consumption along the planned route, which is beneficial to reduce carbon emissions.
-This output is an essential part of airlift: it guarantees flight safety, makes ground operations more efficient, and makes sure that each item gets to its right destination.
-
-Our main contributions were the mathematical modelling of ACLP+RPDP, involving four well-known and interconnected {\it NP-hard}\/ sub-problems, a complete process to solve it, and a new heuristic named {\it Shims}\/ that offers fast node solutions with good quality.
-The method of this work is not exclusive to aircraft and airports: it can be adapted to ships and ports, vehicles and warehouses, or wagons and railways, provided that their practical cases are similar to those considered here.
-In these situations, it would be necessary to make some changes in the model: for example, modify the load balancing constraints and consider the available space in vehicles or wagons instead of pallets.
-
-Because we applied this heuristic process to all $K!$\/ tours, perhaps it seems that our methodology is not generalizable.
-This is not true; we did this because {\it Shims}\/ is very fast and $K$\/ is small in practical cases of air transport.
-In other contexts where $K$\/ may have larger values, this heuristic process remains valid.
-It is enough to apply it only to some tours that offer a better expectation of a good result, for example, on the TSP's shortest path or other similar ones.
-
-As this is ongoing research, we thought about some possible future improvements: consider more than one aircraft, implement parallel algorithms in some steps of the solution to improve computational efficiency, and model 3-D items.
-
-\section*{CRediT authorship contribution statement}
-
-\textbf{Antonio Celio Pereira de Mesquita:} Conceptualization, Methodology, Software, Writing - original draft preparation, Investigation, Validation.
-\textbf{Carlos Alberto Alonso Sanches:} Conceptualization, Methodology, Resources, Supervision, Writing - reviewing \& editing.
-
-\section*{Declaration of competing interest}
-
-The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.
-
-\section*{Data availability}
-
-The dataset and code used in this work are available in {\tt github.com/celiomesquita/ACLP\_RPDP\_P}.
-
-\section*{Acknowledgments}
-
-This research was partially supported by the S\~{a}o Paulo Research Foundation (FAPESP), grant 2022/05803-3.
-
-\section*{References}
-
-\bibliographystyle{apalike}
-
-\bibliography{bib}
-
 \end{document}

### 2024-02-04 latex/ESWA_complexity.tex
Time complexity analysis update

@@
 
-When $K > 6$, we obtain 100 tours from a heuristic TSP solution with the Genetic Algorithm (GA), which addresses optimization problems by evolving a population of candidate solutions through selection, crossover (recombination), and mutation over multiple generations, aiming to improve the fitness of solutions.
+When $K > 6$, we obtain 100 tours from a heuristic TSP solution with the Genetic Algorithm (GA), which addresses the TSP by evolving a population of candidate tours through selection, crossover (recombination), and mutation over multiple generations, aiming to minimize tour length.
 
-The time complexity of applying Genetic Algorithms to solve the TSP is influenced by several key factors, including the number of nodes (\(K\)), population size (\(P\)), number of generations (\(G\)), and the complexities of the crossover (\(C(K)\)), mutation (\(M(K)\)), and selection (\(S(K)\)) operations.
+The time complexity of this method is influenced by the number of nodes (\(K\)), the population size (\(P\)), the number of generations (\(G\)), and the complexities of the crossover (\(C(K)\)), mutation (\(M(K)\)), and selection (\(S(K)\)) operations.
 
@@
 
-{\bf Detailed Explanation}
-\begin{itemize}
-	\item The selection operation's complexity, \(S(K)\), may depend on evaluating the fitness of each solution, which in turn relies on computing the total distance of a path through \(K\) nodes.
-	\item The complexities of the crossover and mutation operations, \(C(K)\) and \(M(K)\), are functions of \(K\) because they involve manipulating sequences representing paths through the \(K\) nodes.
-	\item Each generation involves selecting solutions based on fitness, generating new solutions through crossover, and applying mutations, leading to the multiplication by \(G\).
-\end{itemize}
+The selection complexity, \(S(K)\), depends on evaluating the path length of each tour.
+The complexities of the crossover and mutation operations, \(C(K)\) and \(M(K)\), are functions of \(K\) because they involve manipulating sequences of nodes.
 
-{\bf Impact of the number of nodes (\(K\))}
-\begin{itemize}
-	\item The direct involvement of \(K\) in the complexities of \(C(K)\), \(M(K)\), and \(S(K)\) highlights how the number of nodes affects the algorithm's performance.
+Each generation involves selecting tours based on shorter distances, generating new tours through crossover, and applying mutations, leading to the multiplication by \(G\).
+
+The direct involvement of \(K\) in the complexities of \(C(K)\), \(M(K)\), and \(S(K)\) highlights how the number of nodes affects the algorithm's performance.
 Specifically, operations that manipulate or evaluate paths become more complex as \(K\) increases.
-	\item The encoding of solutions and the calculation of fitness values, essential for the selection process, inherently depend on \(K\), emphasizing its role in the GA's computational demands.
-\end{itemize}
+
+The encoding of tours and the calculation of path length have greater influence in the GA's computational demands.
 

### 2024-02-05 latex/ESWA_complexity.tex
Time complexity updated

@@
 
-Load $K$ lines of text to compose a distances matrix for $K$ airports, the $m$ packed contents from previous nodes, and $n$ items parameters from text files are of complexities $O(K)$, $O(m)$, and $O(n)$, respectively:
+Load $K$ lines of text to compose a distances matrix for $K$ airports, the $m$ packed contents from previous nodes, and $n$ items parameters from text files are of complexities $\mathcal{O}(K)$, $\mathcal{O}(m)$, and $\mathcal{O}(n)$, respectively:
 
 	\begin{itemize}
-		\item $O(K + m + n)$
+		\item $\mathcal{O}(K + m + n)$
 	\end{itemize}
@@
 \item If $K <= 6$, determine the number of tours through permutation.
-To get permutation of nodes as tours has a complexity of $O(T = K!
-\cdot K)$, as the number of nodes is permutated (factorial) to generate tours.
+To get permutation of nodes as tours has a complexity of $\mathcal{O}(T = K!)$, as the nodes are permutated (factorial) to generate tours.
 
 	\begin{itemize}
-	\item $ O(K + m + n) + O(T = K! \cdot K)$ or
-	\item $ O(K + m + n) + O(T = 2)$, if the we are solving the two optimal TSP tours.
+	\item $ \mathcal{O}(K + m + n) + \mathcal{O}(T = K!)$ or
+	\item $ \mathcal{O}(K + m + n) + \mathcal{O}(T = 2)$, if the we are solving the two optimal TSP tours.
 	\end{itemize}
@@
 	\begin{itemize}
-		\item $O(K + m + n) + O(T = 100)$
+		\item $\mathcal{O}(K + m + n) + \mathcal{O}(T = 100)$
 	\end{itemize}
@@
 \begin{itemize}
-	\item $O(K + m + n) + O(T)$
+	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T)$
 \end{itemize}
 
-a) each tour with $K$ nodes:
+a) For each tour with $K$ nodes:
 \begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K)$
+	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K)$
 \end{itemize}
 
-b) in each node we have to reset (to eliminate the risk of previous destinations assigned to a pallet) $O(K \cdot m)$ and set the empty pallets destinations: $O(K \cdot m + K \cdot m) = O(2 \cdot K \cdot m) = O(K \cdot m)$
+b) in each node we have to reset (to eliminate the risk of previous destinations assigned to a pallet) $\mathcal{O}(K \cdot m)$ and set the empty pallets destinations: $\mathcal{O}(K \cdot m + K \cdot m) = \mathcal{O}(2 \cdot K \cdot m) = \mathcal{O}(K \cdot m)$
 \begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m))$
+	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K \cdot (K \cdot m))$
 \end{itemize}
 
-c) To calculate the time limit to solve each node, it is necessary to determine the potential total volume for each node.: $O(K \cdot n)$
+c) To calculate the time limit to solve each node, it is necessary to determine the potential total volume for each node.: $\mathcal{O}(K \cdot n)$
 \begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n))$
+	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n))$
 \end{itemize}
 
-d) The time complexity to put the packet contents destined to the next nodes in the tour on board: $O(m)$.
+d) The time complexity to put the packet contents destined to the next nodes in the tour on board: $\mathcal{O}(m)$.
 \begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m))$
+	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m))$
 \end{itemize}
 
-d) The time complexity to optimize the pallets positions to minimize the CG deviation: $O(m^2)$ ($m$ pallets in $m$ positions)
+d) The time complexity to optimize the pallets positions to minimize the CG deviation: $\mathcal{O}(m^2)$ ($m$ pallets in $m$ positions)
 \begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2))$
+	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2))$
 \end{itemize}
 
-e) Plus the time complexity of the method used to solve an ACLPP with the aircraft partially loaded: $O(C)$
+e) Plus the time complexity of the method used to solve an ACLPP with the aircraft partially loaded: $\mathcal{O}(C)$
 \begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + C))$
+	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + C))$
 \end{itemize}
 
-\subsubsection{Considering the Shims algorithm: $O(C) = O(Shims)$}
+\subsubsection{Considering the Shims algorithm: $\mathcal{O}(C) = \mathcal{O}(Shims)$}
 
@@
 \begin{itemize}
-	\item $O(Shims) = m \cdot log(m)$
+	\item $\mathcal{O}(Shims) = m \cdot log(m)$
 \end{itemize}
@@
 \begin{itemize}
-	\item $O(Shims) = m.(m \cdot log(m))$
+	\item $\mathcal{O}(Shims) = m.(m \cdot log(m))$
 \end{itemize}
 
-c) The greedy phase has $O(n)$.
+c) The greedy phase has $\mathcal{O}(n)$.
 \begin{itemize}
-	\item $O(Shims) = m \cdot (m \cdot log(m) + n)$
+	\item $\mathcal{O}(Shims) = m \cdot (m \cdot log(m) + n)$
 \end{itemize}
 
-d) The composition phase has $O(n)$ added to the First-Fit Decreasing algorithm time complexity $O(m \cdot log(m))$
+d) The composition phase has $\mathcal{O}(n)$ added to the First-Fit Decreasing algorithm time complexity $\mathcal{O}(m \cdot log(m))$
 \begin{itemize}
-	\item $O(Shims) = m \cdot (m \cdot log(m) + n + n + m \cdot log(m))$
-	\item $O(Shims) = m \cdot (2n + 2 \cdot m \cdot log(m)) = m \cdot (n + m \cdot log(m))$
+	\item $\mathcal{O}(Shims) = m \cdot (m \cdot log(m) + n + n + m \cdot log(m))$
+	\item $\mathcal{O}(Shims) = m \cdot (2n + 2 \cdot m \cdot log(m)) = m \cdot (n + m \cdot log(m))$
 \end{itemize}
@@
 \begin{itemize}
-	\item $O(K+m+n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot log(m))))$
+	\item $\mathcal{O}(K+m+n) + \mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot log(m))))$
 \end{itemize}
 
-The overall time complexity provided combines two distinct parts: $O(K+m+n)$ and $O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot \log(m))))$.
+The overall time complexity provided combines two distinct parts: $\mathcal{O}(K+m+n)$ and $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot \log(m))))$.
 Analysing each part separately and then considering their combined implications provides insight into the algorithm's performance characteristics.
@@
 
-\textbf{First Part:} $O(K+m+n)$
+\textbf{First Part:} $\mathcal{O}(K+m+n)$
 
@@
 
-\textbf{Second Part:} $O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot \log(m))))$
+\textbf{Second Part:} $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot \log(m))))$
 
@@
 
-When combining these parts, the linear terms in the first part ($O(K+m+n)$) are overshadowed by the more complex and potentially rapidly increasing terms of the second part, especially due to the presence of $m^2$ and the product $T \cdot K \cdot (...)$.
+When combining these parts, the linear terms in the first part ($\mathcal{O}(K+m+n)$) are overshadowed by the more complex and potentially rapidly increasing terms of the second part, especially due to the presence of $m^2$ and the product $T \cdot K \cdot (...)$.
 The quadratic ($m^2$) and logarithmic ($m \cdot \log(m)$) terms indicate that the algorithm's efficiency decreases significantly as the number of pallets $m$ increases.
+
+So, the overall time complexity may be simplified by relaxing the first part to:
+\begin{itemize}
+	\item $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot log(m))))$
+\end{itemize}
 
@@
 \begin{itemize}
-	\item $O(G \cdot (P \cdot (S(K) + P \cdot (C(K) + M(K)))))$
+	\item $\mathcal{O}(G \cdot (P \cdot (S(K) + P \cdot (C(K) + M(K)))))$
 \end{itemize}

### 2024-02-05 latex/ESWA_complexity_2.tex
Complexity analysis

--- 2024-02-05 latex/ESWA_complexity.tex
@@
 
-Load $K$ lines of text to compose a distances matrix for $K$ airports, the $m$ packed contents from previous nodes, and $n$ items parameters from text files are of complexities $\mathcal{O}(K)$, $\mathcal{O}(m)$, and $\mathcal{O}(n)$, respectively:
+Load $K$ lines of text to compose a distances matrix for $K$ airports, the $m$ packed contents from previous nodes, and $n$ items parameters from text files are of complexities $\mathcal{O}(K)$, $\mathcal{O}(m)$, and $\mathcal{O}(n)$, respectively.
 
-	\begin{itemize}
-		\item $\mathcal{O}(K + m + n)$
-	\end{itemize}
+Although there is some time complexity to load these input data into memory, we will consider them already loaded in memory, not including this parcel in the overall ACLP+RPDP solution time complexity.
 
@@
 \item If $K <= 6$, determine the number of tours through permutation.
-To get permutation of nodes as tours has a complexity of $\mathcal{O}(T = K!)$, as the nodes are permutated (factorial) to generate tours.
+To get permutations of nodes as tours has a complexity of $\mathcal{O}(T = K!)$, as the nodes are permutated (factorial) to generate tours.
 
 	\begin{itemize}
-	\item $ \mathcal{O}(K + m + n) + \mathcal{O}(T = K!)$ or
-	\item $ \mathcal{O}(K + m + n) + \mathcal{O}(T = 2)$, if the we are solving the two optimal TSP tours.
+	\item $ \mathcal{O}(T = K!)$ or
+	\item $ \mathcal{O}(T = 2)$, if the we are solving the two optimal TSP tours.
 	\end{itemize}
@@
 	\begin{itemize}
-		\item $\mathcal{O}(K + m + n) + \mathcal{O}(T = 100)$
+		\item $\mathcal{O}(T = 100)$
 	\end{itemize}
@@
 \begin{itemize}
-	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T)$
+	\item $\mathcal{O}(T)$
 \end{itemize}
@@
 \begin{itemize}
-	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K)$
+	\item $\mathcal{O}(T \cdot K)$
 \end{itemize}
 
-b) in each node we have to reset (to eliminate the risk of previous destinations assigned to a pallet) $\mathcal{O}(K \cdot m)$ and set the empty pallets destinations: $\mathcal{O}(K \cdot m + K \cdot m) = \mathcal{O}(2 \cdot K \cdot m) = \mathcal{O}(K \cdot m)$
+b) in each node we have to reset pallets destinations, to eliminate the risk of previous destinations assigned to a pallet, $\mathcal{O}(K \cdot m)$ and set the empty pallets destinations: $\mathcal{O}(K \cdot m + K \cdot m \cdot n)$
 \begin{itemize}
-	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K \cdot (K \cdot m))$
+	\item $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot m \cdot n))$
 \end{itemize}
@@
 \begin{itemize}
-	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n))$
+	\item $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot m \cdot n + K \cdot n))$
 \end{itemize}
@@
 \begin{itemize}
-	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m))$
+	\item $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot m \cdot n + K \cdot n + m))$
 \end{itemize}
@@
 \begin{itemize}
-	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2))$
+	\item $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot m \cdot n + K \cdot n + m + m^2))$
 \end{itemize}
@@
 \begin{itemize}
-	\item $\mathcal{O}(K + m + n) + \mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + C))$
+	\item $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot m \cdot n + K \cdot n + m + m^2 + C))$
 \end{itemize}
@@
 \begin{itemize}
-	\item $\mathcal{O}(K+m+n) + \mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot log(m))))$
+	\item $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot m \cdot n + K \cdot n + m + m^2 + m \cdot (n + m \cdot log(m))))$	
 \end{itemize}
-
-The overall time complexity provided combines two distinct parts: $\mathcal{O}(K+m+n)$ and $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot \log(m))))$.
-Analysing each part separately and then considering their combined implications provides insight into the algorithm's performance characteristics.
 
@@
 
-\textbf{First Part:} $\mathcal{O}(K+m+n)$
-
-This component of the time complexity suggests a linear dependency on the sizes of three separate inputs or parameters, denoted by $K$, $m$, and $n$.
-These are sequential steps, each influenced by each of these parameters size.
-This implies that as any of $K$, $m$, or $n$ increases, the time required by the algorithm increases linearly with respect to the largest of these parameters, the number of items to be transported ($n$).
-
-\textbf{Second Part:} $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot \log(m))))$
-
-This portion is considerably more complex, indicating an algorithm with nested dependencies on several variables ($T$, $K$, $m$, and $n$):
+This complexity analysis indicates an algorithm with nested dependencies on several variables ($T$, $K$, $m$, and $n$):
 
@@
 	\item $T \cdot K$: repeated iterations $T$ for each of $K$ nodes, nested loops where inner operations are contingent upon $K$.
-	\item $K \cdot m + K \cdot n$: These operations scale with both $K$ and each of $m$ and $n$, indicating processes that involve all pairs of $K$ with $m$ and $n$.
+	\item $K \cdot m + K \cdot m \cdot n + K \cdot n$: These operations scale with both $K$ and each of $m$ and $n$, indicating processes that involve all pairs of $K$ with $m$ and $n$.
 	\item $m + m^2$: operations with linear and quadratic dependencies on the number of pallets $m$, indicating that as $m$ grows, the complexity increases more significantly due to the quadratic term $m^2$.
@@
 
-\textbf{Combined Implications}
-
-When combining these parts, the linear terms in the first part ($\mathcal{O}(K+m+n)$) are overshadowed by the more complex and potentially rapidly increasing terms of the second part, especially due to the presence of $m^2$ and the product $T \cdot K \cdot (...)$.
-The quadratic ($m^2$) and logarithmic ($m \cdot \log(m)$) terms indicate that the algorithm's efficiency decreases significantly as the number of pallets $m$ increases.
-
-So, the overall time complexity may be simplified by relaxing the first part to:
-\begin{itemize}
-	\item $\mathcal{O}(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot log(m))))$
-\end{itemize}
+The quadratic ($m^2$) and logarithmic ($m \cdot \log(m)$) terms indicate that the algorithm's efficiency decreases significantly as the number of pallets ($m$) increases.
 
 The overall complexity suggests an algorithm that performs efficiently with small values of $K$, $m$, and $n$ but may become substantially more resource-intensive as these parameters increase.
-The number tours $T$ indicates that the entire complex operation is executed $T$ times, further amplifying the effect of the inner complexities.
+The number of tours $T$ indicates that the entire complex operation is executed $T$ times, further amplifying the effect of the inner complexities.
 
-Optimizing such an algorithm involves minimizing the repetition count ($T$), reducing the complexity of operations dependent on $K$, $m$, and $n$.
+Optimising such an algorithm involves minimising the repetition count ($T$), reducing the complexity of operations dependent on $K$, $m$, and $n$.
+As a suggestion to deal with this (reduce the number of tours to be tested), we propose a TSP solution with the Genetics Algorithm, whose complexity is analysed in sub-subsection \ref{ga_tsp}.
 
@@
 
-When $K > 6$, we obtain 100 tours from a heuristic TSP solution with the Genetic Algorithm (GA), which addresses the TSP by evolving a population of candidate tours through selection, crossover (recombination), and mutation over multiple generations, aiming to minimize tour length.
+When $K > 6$, we obtain 100 tours from a heuristic TSP solution with the Genetic Algorithm (GA), which addresses the TSP by evolving a population of candidate tours through selection, crossover (recombination), and mutation over multiple generations, aiming to minimise tour length.
 
-The time complexity of this method is influenced by the number of nodes (\(K\)), the population size (\(P\)), the number of generations (\(G\)), and the complexities of the crossover (\(C(K)\)), mutation (\(M(K)\)), and selection (\(S(K)\)) operations.
+The number of nodes (K), population size (P), number of generations (G), and complexity of the crossover (C(K)), mutation (M(K)), and selection (S(K)) operations all have an impact on the method's time complexity.
 
@@
 
-The selection complexity, \(S(K)\), depends on evaluating the path length of each tour.
-The complexities of the crossover and mutation operations, \(C(K)\) and \(M(K)\), are functions of \(K\) because they involve manipulating sequences of nodes.
+The selection complexity, $S(K)$, depends on evaluating the path length of each tour.
+The complexities of the crossover and mutation operations, $C(K)$ and $M(K)$, are functions of $K$ because they involve manipulating sequences of nodes.
 
-Each generation involves selecting tours based on shorter distances, generating new tours through crossover, and applying mutations, leading to the multiplication by \(G\).
+Each generation involves selecting tours based on shorter distances, generating new tours through crossover, and applying mutations, leading to the multiplication by $G$.
 
-The direct involvement of \(K\) in the complexities of \(C(K)\), \(M(K)\), and \(S(K)\) highlights how the number of nodes affects the algorithm's performance.
-Specifically, operations that manipulate or evaluate paths become more complex as \(K\) increases.
+The direct involvement of $K$ in the complexities of $C(K)$, $M(K)$, and $S(K)$ highlights how the number of nodes affects the algorithm's performance.
+Specifically, operations that manipulate or evaluate paths become more complex as $K$ increases.
 
-The encoding of tours and the calculation of path length have greater influence in the GA's computational demands.
+The encoding of tours and the calculation of path length have greater influence on the GA's computational demands.
 

### 2024-02-06 latex/ESWA_carlos.tex
Update

--- 2024-02-04 latex/ESWA.tex
@@
 This enables other risks, such as improper delivery, excessive fuel burn, and possible safety issues due to cargo imbalance, as well as a longer than necessary turn-around time. This {\it NP-hard}\/ problem, named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem (ACLP+RPDP)}, is mathematically modelled using standardised pallets in fixed positions.
-We developed a strategy to solve this problem, considering historical transport data from some Brazilian hub networks, and performed several experiments with a commercial solver, {\color{blue}five} known meta-heuristics, and a new heuristic designed specifically for this problem.
+We developed a strategy to solve this problem, considering historical transport data from some Brazilian hub networks, and performed several experiments with a commercial solver, five known meta-heuristics, and a new heuristic designed specifically for this problem.
 By using a portable computer, our strategy quickly found practical solutions to a wide range of real problems in much less than operationally acceptable time.
@@
 In this context, at least two more important sub-problems can be considered: simultaneous pickup and delivery at each node, called the {\it Pickup and Delivery Problem} (PDP), and searching for the best benefit-cost route, which is a special case of the {\it Travelling Salesman Problem} (TSP).
-
-The problem addressed in this work is the definition of the route of an aircraft that loads and unloads hundreds of items in up to seven hubs, dealing with weight, volume, and balance constraints, and maximising the benefit-cost ratio.
-This challenge has become even more acute during the COVID-19 pandemic due to the need for urgent medical supplies.
-Delays in shipping essential items such as respirators to critical areas highlighted the need for optimised solutions.
 
@@
 
+The problem addressed in this work is the definition of the route of an aircraft that loads and unloads hundreds of items in several hubs, dealing with weight, volume, and balance constraints, and maximising the benefit-cost ratio.
+This challenge has become even more acute during the COVID-19 pandemic due to the need for urgent medical supplies.
+Delays in shipping essential items such as respirators to critical areas highlighted the need for optimised solutions.
+
 This problem is extremely complex, as it deals with different objectives: defining a route that visits all hubs; maximising the items transported along this route, prioritising the essential ones; making sure items reach the correct destinations; ensuring aircraft safety constraints; and saving fuel so that the flight is sustainable.
@@
 
-		\caption{Brazilian airports distances ($km$)} \label{tab:costs}
+		\caption{Distances between some Brazilian airports ($km$)} \label{tab:costs}
 		\centering
@@
 Among the most recent, we can highlight:
+
 \begin{itemize}
@@
 
-	\item \cite{zhao2023} proposed three integer linear programming models: a bi-objective optimisation model (BOM), a combinatorial optimisation model (COM), and an improved combinatorial optimisation model (IOM).
-Its objectives were the maximum possible loading and the smallest CG deviation.
-Four scenarios with multiple metrics are resolved for the {\it Boeing 777F}.
-The results showed that the BOM has the fastest solution speed, but the CG deviation is the worst; in several cases, the CG deviation is unacceptable.
-COM has the longest solution time, which is difficult to tolerate in practice.
+	\item \cite{zhao2023} presented three models that use integer programming for air cargo planning and weight balance optimization: bi-objective optimization (BOM), combinatorial optimization (COM), and enhanced combinatorial optimization (IOM).
+Considering a Boeing 777F in several scenarios, the tests revealed performance problems: BOM is fast, but produces large CG deviation; COM offers accurate optimization, but with impractical runtimes; IOM provides a balanced solution, improving speed over COM, but requiring high computational demands in some cases.
+Although IOM stands out for its effectiveness, all models face trade-offs between speed, accuracy and computational efficiency.
+This work alerted us to potential performance issues in solution methods.
 
@@
 
-Throughout this text, we call {\it packed content}\/ a set of items of the same destination stacked on a pallet and covered with a restraining net.
+Throughout this text, we call {\it packed content}\/ (see Figure \ref{fig:packed}) a set of items of the same destination stacked on a pallet and covered with a restraining net.
 It is considered a single item, having the same attributes as its components, whose values are the sum of individual scores, weights, and volumes.
 To ensure accuracy in pickup and delivery operations, packed content must remain on board until its destination.
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[scale=0.3]{packed.jpg}
+	\caption{A {\it packed content}\/ on 463L pallet inside a {\it Boeing C-17}}
+	\small\textsuperscript{Source: From Wikimedia Commons, the free media repository}
+	\label{fig:packed}
+\end{figure}
 
@@
 
-\section{Solution strategies}
+\section{Solution process}
 \label{algorithms}
@@
 Throughout our research, we have thoughtfully described ACLP+RPDP in standard MIP format and found that no solver can handle its practical cases in a feasible time.
-Thus, as ACLP+RPDP is highly complex, involving four intractable and interconnected sub-problems, we decided to focus only on {\it real cases}, developing quick node-by-node solutions, not necessarily optimal, but which would allow us to obtain a complete tour.
+Thus, as ACLP+RPDP is highly complex, involving four intractable and interconnected sub-problems, we decided to focus on {\it real cases}, developing quick node-by-node solutions, not necessarily optimal, but which would allow us to obtain a complete tour.
 
@@
 				\small \ \ \ {\bf else}\\
-				\small \ \ \ \ \ \ {$ntours = 100$ \ \ (best tours obtained with a TSP heuristic)}
+				\small \ \ \ \ \ \ {$ntours = 100$ \ \ (tours obtained with a TSP heuristic)}
 			};
@@
 \begin{algorithm}[htp]
-
-	\caption{Solving the ACLP+RPDP}  \label{alg:main}
+	\small
+	\caption{Solving ACLP+RPDP}  \label{alg:main}
 
@@
 
-				\For {$x \gets 1$ to $ntours$} \label{main:loop1}
-
-					\State $f(S_K[x]) \gets SolveTour(S_K[x], L, M, C, N, method, tmax/ntours)$ \label{main:fpi}
+				\For {$i \gets 1$ to $ntours$} \label{main:loop1}
+
+					\State $f_i \gets SolveTour(\pi[i], L, M, C, N, method, tmax/ntours)$ \label{main:fpi}
 
@@
 
-				\State $answer[scenario,surplus,method] \gets \max f(S_K[x])$ \label{main:a2}
+				\State $answer[scenario,surplus,method] \gets \max f_i$ \label{main:a2}
 			\EndFor
@@
 
-{\color{red}$\pi_k$ are tours} {\color{blue}$\pi$ are tours}, that is, permutations with the $K$\/ nodes.
-When $ntours=2$, $\pi_1$\/ and $\pi_2$\/ are the optimal solutions of the corresponding TSP.
-When $ntours=K!$, $\pi_k$\/ is the $k^{th}$ permutation of $S_K$.
-Finally, when $ntours=100$, $\pi_1, \ldots, \pi_{100}$\/ are the best TSP solutions provided by a heuristic.
+$\pi[\ ]$ is a vector of tours indexed from 1 to $ntours$.
+When $ntours=2$, $\pi[1]$\/ and $\pi[2]$\/ are the optimal solutions of the corresponding TSP.
+When $ntours=K!$, $\pi[i]$\/ is the $i^{th}$ permutation of $S_K$.
+Finally, when $ntours=100$, $\pi[1], \ldots, \pi[100]$\/ are the solutions provided by a TSP heuristic.
 
@@
 \begin{algorithm}[H]
-	\caption{Solving the tour $\pi$ with $method$}  \label{alg:tour}
+	\small
+	\caption{Solving tour $\pi$ with $method$}  \label{alg:tour}
 	\begin{algorithmic}[1]
@@
 
-When the aircraft is at the base, the initial graph $G_1$\/ is empty, and there are no packed contents {\color{blue}(line \ref{tour:g11})}.
+When the aircraft is at the base, the initial graph $G_1$\/ is empty, and there are no packed contents (line \ref{tour:g11}).
 Otherwise, $UpdatePacked$\/ (line \ref{tour:dest}) returns the set of packed contents that have not yet reached their destination and remain on board, rearranging them on the pallets to minimize CG deviation.
@@
 \begin{algorithm}[H]
+	\small
 	\caption{Updating the packed contents that remain boarded at node $\pi_k$}  \label{alg:cons}
@@
 \begin{algorithm}[H]
+	\small
 	\caption{Setting pallets destination based on the items to be embarked at node $\pi_k$}  \label{alg:dest}
@@
 		\State {\bf SetPalletsDestinations} \,\, {\it in: $M, \pi_k$} \, \, {\it out: $M$}
-
 		\State $vol_x \gets  0$, $x \in L_{\pi_k}$  \label{dest:vector1}
-
 		\State $max \gets 0$  \Comment{destination with maximum volume demand}
-
 		\State $total \gets 0$ 
-
 		\For{$j \gets 1$ to $n_{\pi_k}$}
-
 		\If {$to_j \in L_{\pi_k}$} 
-
 		\State $vol_{to_j} \gets vol_{to_j} + v_j$ 
 		\State $total \gets total + v_j$ 
-
 		\If {$vol_{to_j} > vol_{max}$}
-
 		\State $max \gets to_j$ \label{dest:max} 
 		\EndIf
-
 		\EndIf
-
 		\EndFor
-
 		\For{$x \in L_{\pi_k}$} \label{dest:propor1}
-
 		\If {$vol_{x} \neq 0$}
-
 		\State $needed \gets \max \{1, \lfloor{ (m-m_{\pi_k}) \times vol_{x}/total}\rfloor \}$  
@@
 		\State $np \gets 0$
-
 		\For{$i \gets 1$ to $m$}
-
 		\If{($np < needed$) {\bf and} ($T_i^{\pi_k} = -1$)}  
@@
 		\EndIf
-
 		\EndFor
 		\EndIf 
-
 		\EndFor \label{dest:propor2}
-
 		\For{$i \gets 1$ to $m$} \label{dest:max1}
@@
 \begin{algorithm}[H]
+	\small
 	\caption{Cargo score and aircraft torque}  \label{alg:eval}
@@
 
-In this way, the mathematical model for $SolveNode(MIP,\pi_k,G, tmax)$\/ becomes simpler, which finds an allocation of available items at the node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$.
-Thus, we use a MIP solver with a runtime limit $tmax$\/ at the node $\pi_k$ to maximize the objective function (\ref{eq:obj2}) with the calculus equations (\ref{eq:newf}) to (\ref{eq:costs2}), subject to the constraints (\ref{eq:tau2}) to (\ref{eq:ifB}).
+In this way, the mathematical model for $SolveNode(MIP,\pi_k,G, tmax)$\/ becomes simpler, which finds an allocation of available items at node $\pi_k$\/ using previously defined values of $L_{\pi_k}$, $T_i^{\pi_k}$, and $a^{\pi_k}_q$.
+Thus, we use a MIP solver with a runtime limit $tmax$\/ at node $\pi_k$ to maximize the objective function (\ref{eq:obj2}) with the calculus equations (\ref{eq:newf}) to (\ref{eq:costs2}), subject to the constraints (\ref{eq:tau2}) to (\ref{eq:ifB}).
 The binary variables $X_{ij}$\/ and $Y_{iq}$\/ define the sets of edges $E_{N_{\pi_k}}$\/ and $E_{Q_{\pi_k}}$, respectively, included in graph $G$.
@@
 
-The selection of edges for $E_{ N_{\pi_k} }$ uses the {\it edge attractiveness}\/ $\theta_{ij}$\/ (\ref{eq:edge}), which can be understood as the tendency to allocate the item $j$\/ to the pallet $i$\/ at the node $\pi_k$.
+The selection of edges for $E_{ N_{\pi_k} }$ uses the {\it edge attractiveness}\/ $\theta_{ij}$\/ (\ref{eq:edge}), which can be understood as the tendency to allocate item $j$\/ to pallet $i$\/ at node $\pi_k$.
 It is directly proportional to the score, and inversely to the volume and the torque of each item.
@@
 
-Considering only the items that can be shipped at node $\pi_k$, Figure \ref{fig:whip} represents $n_{\pi_k}$\/ possible edges $e_{ij}$\/ of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order.
-Initially, {\it Shims}\/ builds a greedy solution for the pallet $i$\/ selecting edges up to index $\eta_1$ ({\it greedy phase}).
+Considering only the items that can be shipped at node $\pi_k$, Figure \ref{fig:whip} represents $n_{\pi_k}$\/ possible edges $e_{ij}$\/ of pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order.
+Initially, {\it Shims}\/ builds a greedy solution for pallet $i$\/ selecting edges up to index $\eta_1$ ({\it greedy phase}).
 Then, with the edges between $\eta_1$\/ and $\eta_2$, it elaborates different possible complements ({\it composition phase}), including later the best ones in the same pallet ({\it selection phase}).
@@
 
-		\State Let $E$\/ be an array of $n_{\pi_k}$ possibles edges of the pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order \label{shims:possible}
+		\State Let $E$\/ be an array of $n_{\pi_k}$ possibles edges of pallet $i$\/ sorted by $\theta_{ij}$\/ in non-ascending order \label{shims:possible}
 		\State $\eta_1 \gets 1$  
@@
 
-$tmax$\/ is the runtime limit for {\it Shims}, and $level_1$\/ and $level_2$\/ are volume thresholds for the indices $\eta_1$ and $\eta_2$, respectively.
-
+$tmax$\/ is the runtime limit for {\it Shims}, and $level_1$\/ and $level_2$\/ are volume thresholds for indices $\eta_1$ and $\eta_2$, respectively.
 Initially, $Q_{\pi_k}$\/ (line \ref{shims:initQ}) corresponds to the packed contents that remain on board.
 It is important to remember that $E_{Q_{\pi_k}}$\/ and $M$\/ were modified by the procedure $UpdatePacked(M, Q_{\pi_k}, \pi_k)$ and the procedure $SetPalletsDestinations(M, \pi_k)$.
-Then, the pallets $i$\/ are considered in non-descending order of $|D_i^{long}|$.
+Then, the pallets are considered in non-descending order of $|D_i^{long}|$.
 
@@
 	\item In the {\it greedy phase}\/ (lines \ref{shims:pallets}-\ref{shims:endfirst}), a partial solution for each pallet $i$\/ is constructed by adding edges following $\theta_{ij}$\/ order.
-The indices $\eta_1$\/ and $\eta_2$\/ refer to the accumulated volumes $V_i \times level_1$\/ and $V_i \times level_2$, respectively, which were defined empirically by the {\it irace}\/ tool \citep{LopezIbanezManuel2016}.
-We will explain this in detail in the subsection \ref{items}.
+Indices $\eta_1$\/ and $\eta_2$\/ refer to the accumulated volumes $V_i \times level_1$\/ and $V_i \times level_2$, respectively, which were defined empirically by the {\it irace}\/ tool \citep{LopezIbanezManuel2016}.
+We will explain this in subsection \ref{items}.
 
@@
 
-{\color{blue}
-\subsection{Complexity analysis}
-
-These are the steps to solve the ACLP+RPDP with the complexity analysis:
-
-\subsubsection{Input data}
-
-Load $K$ lines of text to compose a distances matrix for $K$ airports, the $m$ packed contents from previous nodes, and $n$ items parameters from text files are of complexities $O(K)$, $O(m)$, and $O(n)$, respectively:
+\subsection{Time complexity using {\it Shims}}
+
+We finish this section with the analysis of the time complexity of our ACLP+RPDP solution, considering the use of the {\it Shims}\/ heuristic.
+
+In this process, when $K>6$\/ (that is, in unusual cases of air transportation), we need a GA-based TSP heuristic to generate 100 tours of size $K$.
+This heuristic can be chosen and calibrated to be fast enough: specifically, in the tests we will present in the next section, it took just 33 seconds with $K=15$.
+For this reason, we will not perform its complexity analysis.
+We will also not consider the time spent by the {\it irace}\/ tool, in defining parameters for {\it Shims}, as it is only executed once in the calibration of our method.
+
+Let $n = \max n_k$.
+Without loss of generality, we assume that $m = \mathcal{O}(n)$ and $\max m_k = \mathcal{O}(n)$.
+Therefore, to read the input data, Algorithm \ref{alg:main} requires time $\mathcal{O}(m + K^2 + K*n + m*\max m_k) = \mathcal{O}( n*m)$, considering $K^2 = \mathcal{O}(m*n)$.
+Since $ntours \leq 6!$, it is enough to calculate the complexity of Algorithm \ref{alg:tour}.
 
 \begin{itemize}
-		\item $O(K + m + n)$
+
+	\item Its initial variables are $\mathcal{O}(K*n)$.
+
+	\item Algorithm \ref{alg:cons} minimizes CG deviation and updates pallets.
+	We can consider that the MIP solver spends time $\mathcal{O}(m^2) = O(m*n)$, as it needs to allocate at most $m$\/ packed contents to $m$\/ pallets.
+The final loop of Algorithm \ref{alg:cons} has the same complexity.
+
+	\item Algorithm \ref{alg:dest} takes $\mathcal{O}(n+K+m)$ time.
+
+	\item {\it Shims}, described in Algorithm \ref{alg:shims}, spends $\mathcal{O}(m*\log m)$ in ordering the pallets and then performs a loop with $m$\/ iterations.
+Each of these iterations spends $\mathcal{O}(n*\log n)$\/ on ordering the edges by the value of $\theta_{ij}$, and $\mathcal{O}(n)$\/ on the others statements.
+So its total time is $\mathcal{O}(m*n*\log n)$.
+
+	\item Algorithm \ref{alg:eval} takes $\mathcal{O}(m*n)$ time.
+
 \end{itemize}
 
-\subsubsection{Number of tested tours ($T$)}
-
-\begin{itemize}
-\item If $K <= 6$, determine the number of tours through permutation.
-To get permutation of nodes as tours has a complexity of $O(T = K!
-\cdot K)$, as the number of nodes is permutated (factorial) to generate tours.
-
-	\begin{itemize}
-	\item $ O(K + m + n) + O(T = K! \cdot K)$ or
-	\item $ O(K + m + n) + O(T = 2)$, if the we are solving the two optimal TSP tours.
-	\end{itemize}
-
-\item If $K > 6$, we obtain 100 tours from a heuristic TSP solution (see its time complexity analysis in subsubsection \ref{ga_tsp}):
-
-	\begin{itemize}
-		\item $O(K + m + n) + O(T = 100)$
-	\end{itemize}
-\end{itemize}
-
-\subsubsection{Solving $T$ tours:}
-
-\begin{itemize}
-	\item $O(K + m + n) + O(T)$
-\end{itemize}
-
-a) each tour with $K$ nodes:
-\begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K)$
-\end{itemize}
-
-b) in each node we have to reset (to eliminate the risk of previous destinations assigned to a pallet) $O(K \cdot m)$ and set the empty pallets destinations: $O(K \cdot m + K \cdot m) = O(2 \cdot K \cdot m) = O(K \cdot m)$
-\begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m))$
-\end{itemize}
-
-c) To calculate the time limit to solve each node, it is necessary to determine the potential total volume for each node.: $O(K \cdot n)$
-\begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n))$
-\end{itemize}
-
-d) The time complexity to put the packet contents destined to the next nodes in the tour on board: $O(m)$.
-\begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m))$
-\end{itemize}
-
-d) The time complexity to optimize the pallets positions to minimize the CG deviation: $O(m^2)$ ($m$ pallets in $m$ positions)
-\begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2))$
-\end{itemize}
-
-e) Plus the time complexity of the method used to solve an ACLPP with the aircraft partially loaded: $O(C)$
-\begin{itemize}
-	\item $O(K + m + n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + C))$
-\end{itemize}
-
-\subsubsection{Considering the Shims algorithm: $O(C) = O(Shims)$}
-
-a) the time complexity for sorting the pallets in non-descending order of distances to the CG:
-\begin{itemize}
-	\item $O(Shims) = m \cdot log(m)$
-\end{itemize}
-
-b) As each pallet is solved at a time, the overall Shims complexity is multiplied by $m$.
-\begin{itemize}
-	\item $O(Shims) = m.(m \cdot log(m))$
-\end{itemize}
-
-c) The greedy phase has $O(n)$.
-\begin{itemize}
-	\item $O(Shims) = m \cdot (m \cdot log(m) + n)$
-\end{itemize}
-
-d) The composition phase has $O(n)$ added to the First-Fit Decreasing algorithm time complexity $O(m \cdot log(m))$
-\begin{itemize}
-	\item $O(Shims) = m \cdot (m \cdot log(m) + n + n + m \cdot log(m))$
-	\item $O(Shims) = m \cdot (2n + 2 \cdot m \cdot log(m)) = m \cdot (n + m \cdot log(m))$
-\end{itemize}
-
-The overall time complexity to solve the ACLP+RPDP with Shims is:
-\begin{itemize}
-	\item $O(K+m+n) + O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot log(m))))$
-\end{itemize}
-
-The overall time complexity provided combines two distinct parts: $O(K+m+n)$ and $O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot \log(m))))$.
-Analysing each part separately and then considering their combined implications provides insight into the algorithm's performance characteristics.
-
-\subsubsection{Complexity analysis considerations}
-
-\textbf{First Part:} $O(K+m+n)$
-
-This component of the time complexity suggests a linear dependency on the sizes of three separate inputs or parameters, denoted by $K$, $m$, and $n$.
-These are sequential steps, each influenced by each of these parameters size.
-This implies that as any of $K$, $m$, or $n$ increases, the time required by the algorithm increases linearly with respect to the largest of these parameters, the number of items to be transported ($n$).
-
-\textbf{Second Part:} $O(T \cdot K \cdot (K \cdot m + K \cdot n + m + m^2 + m \cdot (n + m \cdot \log(m))))$
-
-This portion is considerably more complex, indicating an algorithm with nested dependencies on several variables ($T$, $K$, $m$, and $n$):
-
-\begin{itemize}
-	\item $T \cdot K$: repeated iterations $T$ for each of $K$ nodes, nested loops where inner operations are contingent upon $K$.
-	\item $K \cdot m + K \cdot n$: These operations scale with both $K$ and each of $m$ and $n$, indicating processes that involve all pairs of $K$ with $m$ and $n$.
-	\item $m + m^2$: operations with linear and quadratic dependencies on the number of pallets $m$, indicating that as $m$ grows, the complexity increases more significantly due to the quadratic term $m^2$.
-	\item $m \cdot (n + m \cdot \log(m))$: Combines linear scaling with $m$ and $n$ and a logarithmic term $m \cdot \log(m)$, because the algorithm uses divide-and-conquer strategies on $m$ elements, the Shims heuristics.
-\end{itemize}
-
-\textbf{Combined Implications}
-
-When combining these parts, the linear terms in the first part ($O(K+m+n)$) are overshadowed by the more complex and potentially rapidly increasing terms of the second part, especially due to the presence of $m^2$ and the product $T \cdot K \cdot (...)$.
-The quadratic ($m^2$) and logarithmic ($m \cdot \log(m)$) terms indicate that the algorithm's efficiency decreases significantly as the number of pallets $m$ increases.
-
-The overall complexity suggests an algorithm that performs efficiently with small values of $K$, $m$, and $n$ but may become substantially more resource-intensive as these parameters increase.
-The number tours $T$ indicates that the entire complex operation is executed $T$ times, further amplifying the effect of the inner complexities.
-
-Optimizing such an algorithm involves minimizing the repetition count ($T$), reducing the complexity of operations dependent on $K$, $m$, and $n$.
-
-\subsubsection{Time Complexity Analysis of the Genetic Algorithm for TSP}
-\label{ga_tsp}
-
-When $K > 6$, we obtain 100 tours from a heuristic TSP solution with the Genetic Algorithm (GA), which addresses optimization problems by evolving a population of candidate solutions through selection, crossover (recombination), and mutation over multiple generations, aiming to improve the fitness of solutions.
-
-The time complexity of applying Genetic Algorithms to solve the TSP is influenced by several key factors, including the number of nodes (\(K\)), population size (\(P\)), number of generations (\(G\)), and the complexities of the crossover (\(C(K)\)), mutation (\(M(K)\)), and selection (\(S(K)\)) operations.
-
-Given these parameters, we can express the overall time complexity as:
-\begin{itemize}
-	\item $O(G \cdot (P \cdot (S(K) + P \cdot (C(K) + M(K)))))$
-\end{itemize}
-
-{\bf Detailed Explanation}
-\begin{itemize}
-	\item The selection operation's complexity, \(S(K)\), may depend on evaluating the fitness of each solution, which in turn relies on computing the total distance of a path through \(K\) nodes.
-	\item The complexities of the crossover and mutation operations, \(C(K)\) and \(M(K)\), are functions of \(K\) because they involve manipulating sequences representing paths through the \(K\) nodes.
-	\item Each generation involves selecting solutions based on fitness, generating new solutions through crossover, and applying mutations, leading to the multiplication by \(G\).
-\end{itemize}
-
-{\bf Impact of the number of nodes (\(K\))}
-\begin{itemize}
-	\item The direct involvement of \(K\) in the complexities of \(C(K)\), \(M(K)\), and \(S(K)\) highlights how the number of nodes affects the algorithm's performance.
-Specifically, operations that manipulate or evaluate paths become more complex as \(K\) increases.
-	\item The encoding of solutions and the calculation of fitness values, essential for the selection process, inherently depend on \(K\), emphasizing its role in the GA's computational demands.
-\end{itemize}
-
-}
+Therefore, we can conclude that the total time of Algorithm \ref{alg:tour} is $\mathcal{O}(K*m*n*\log n)$, which is the time complexity of the ACLP+RPDP solution process .
 
@@
 \begin{algorithm}[H]
+	\small
 	\caption{Generating items}  \label{alg:itemsgen}
@@
 
-To determine the parameters $level_1$\/ and $level_2$, used by \textit{Shims}, we previously carried out some experiments with the tool {\it irace}\/ \citep{LopezIbanezManuel2016}, the results of which are presented in Table \ref {tab:irace}.
+To determine the parameters $level_1$\/ and $level_2$\/ used by \textit{Shims}, we previously carried out some experiments with the {\it irace}\/ tool \citep{LopezIbanezManuel2016}, the results of which are presented in Table \ref {tab:irace}.
 In these tests, of every 7 instances generated for each value of $surplus$, 4 were used as the training set and 3 as the testing set.
@@
 	\centering
-	\caption{\textit{irace} testing results}  \label{tab:irace}
+	\caption{\textit{irace}\/ results}  \label{tab:irace}
 	\begin{tabular}{cccc}
@@
 
-We will first present the results obtained in practical cases, that is, when $1<K\leq6$. Next, we will show how the same {\it Shims}\/ heuristic remains robust when $K>6$.
+We will first present the results obtained in practical cases, that is, when $1<K\leq6$. Next, we will show how {\it Shims}\/ remains robust when $K>6$.
 
@@
 
-We ran Algorithm \ref{alg:main} considering the 5 scenarios from Table \ref{tab:scenarios}, 3 values for  $surplus$\/ from $\{1.2, 1.5, 2.0\}$, 4 values for $tmax$\/ from $\{240s, 1200s, 2400s, 3600s\}$, and 7 different methods for the node-by-node solution: {\it Gurobi}\/ (see \ref{solver}), ACO, NMO, TS, GRASP, GA and {\it Shims}\/ (Algorithm \ref{alg:shims}).
+We ran Algorithm \ref{alg:main} considering the 5 scenarios from Table \ref{tab:scenarios}, 3 values for  $surplus$\/ from $\{1.2, 1.5, 2.0\}$, 4 values for $tmax$\/ from $\{240s, 1200s, 2400s, 3600s\}$, and 7 different methods for the node-by-node solution: ACO, NMO, TS, GRASP, GA, {\it Gurobi}\/ (subsection \ref{solver}), and {\it Shims}\/ (Algorithm \ref{alg:shims}).
 
@@
 
-We also indicate the two adopted strategies: dedicating all the processing time to the 2 shortest tours or distributing it among all $K!$\/ tours.
+We also indicate the adopted strategies: dedicating all the processing time to the $ntours=2$ shortest tours or distributing it among all $ntours=K!$\/ tours.
 The results obtained with $tmax = 3600s$, which is the highest tested runtime limit, are in Tables \ref{tab:20}, \ref{tab:50} and \ref{tab:100}, with $surplus$\/ values of $1.2$, $1.5$\/ and $2.0$, respectively.
@@
 
-These last results that we will show here do not correspond to practical cases of air transport, as tours where $K>6$ very rarely occur.
-However, it is possible to see that the {\it Shims} heuristic maintains robust behaviour as the number of nodes grows, that is, it could be adapted to similar contexts
-({\color{red}vehicles and warehouses} {\color{blue}ships and trucks})
-, for example), where there may be more nodes.
-
-In the tests carried out, we considered real data from the 15 main Brazilian airports.
-For this task, we implemented a TSP heuristic based on GA that returned 100 tours in approximately 33 seconds.
-We implemented this heuristic with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework available in {\tt github.com/deap/deap}.
-For more details, see \citep{DEAP_JMLR2012}.
-
-Figure \ref{fig:performance} shows the runtime curve of {\it Shims}\/ as the number of nodes increases, indicating at each point the value found for the objective function $f$.
+These last results do not correspond to practical cases of air transport, as tours where $K>6$ very rarely occur.
+However, it is possible to see that the {\it Shims}\/ maintains robust behaviour as the number of nodes grows, that is, it could be adapted to similar contexts (ships and trucks, for example), where there may be more nodes.
+
+Considering real data from the 15 main Brazilian airports, we implemented a GA-based TSP heuristic that returned 100 tours in approximately 33 seconds.
+We implemented this heuristic with DEAP (Distributed Evolutionary Algorithms in Python), an evolutionary computation framework.
+For more details, see \citep{DEAP_JMLR2012} and {\tt github.com/deap/deap}.
+
+Figure \ref{fig:performance} shows the runtime curve of {\it Shims}\/ as the number $K$\/ of nodes increases, indicating at each point the value found for the objective function $f$.
 Runtime and $f$\/ are the averages obtained from 7 instances generated with $surplus=2.0$\/ for each value of $K$.
-In Figure \ref{fig:morenodes}, we indicate one of the tours found by the TSP heuristic when $K=15$.
+In Figure \ref{fig:morenodes}, we indicate one of the tours found by this TSP heuristic when $K=15$.
 
@@
 In this work, we modelled and solved a real air transport problem named {\it Air Cargo Load Planning with Routing, Pickup, and Delivery Problem}\/ (ACLP+RPDP).
-For the first time in the literature, a {\it NP-hard}\/ problem that involves {\it simultaneously}\/ pallet assembly, load balancing, and route planning is addressed, where the cost-effectiveness of transport is maximized.
+For the first time in the literature, a {\it NP-hard}\/ problem that involves {\it simultaneously}\/ pallet assembly, load balancing, route planning, and pick up and delivery is addressed, where the cost-effectiveness of transport is maximized.
 Currently, there is no commercial software available for this problem.
 
-We adopted some simplifications that are not critical, but that allowed for an unprecedented solution to this problem considering more than two nodes.
-In practical cases, the number $K$\/ of nodes, excluding the base, is small ($K \leq 6$), each of which has hundreds of items to be shipped.
-Considering a real aircraft, we have developed some node-by-node solutions.
-The complete process can be executed quickly on a handheld computer, offering good results and reducing stress for the transport planners.
-
-As validation, we carried out tests in several real air transport scenarios.
-In these cases, the solution process can establish, in less than four minutes, a flight tour for a single aircraft with a good distribution of load on pallets to be put in the cargo bay, enforcing the loaded aircraft balance, maximising the total score, and minimising fuel consumption along the planned route, which is beneficial to reduce carbon emissions.
+We adopted some simplifications that are not critical, but that allowed for an unprecedented solution to this problem considering several nodes.
+In practical cases, there are hundreds of items to be shipped at each node, and the number of nodes, excluding the base, is smaller than the number of pallets.
+Considering a real aircraft, we have developed node-by-node solutions such that the complete process can be executed quickly on a handheld computer, offering good results and reducing stress for the transport planners.
+
+As validation, we carried out tests in several scenarios.
+In the real cases, the solution process can establish, in less than four minutes, a flight itinerary for a single aircraft with a good distribution of load on pallets at each node of the tour, enforcing the weight balance, maximising the total score, and minimising fuel consumption along the planned route, which is beneficial to reduce carbon emissions.
 This output is an essential part of airlift: it guarantees flight safety, makes ground operations more efficient, and makes sure that each item gets to its right destination.
@@
 Our main contributions were the mathematical modelling of ACLP+RPDP, involving four well-known and interconnected {\it NP-hard}\/ sub-problems, a complete process to solve it, and a new heuristic named {\it Shims}\/ that offers fast node solutions with good quality.
-The method of this work is not exclusive to aircraft and airports: it can be adapted to ships and ports, vehicles and warehouses, or wagons and railways, provided that their practical cases are similar to those considered here.
-In these situations, it would be necessary to make some changes in the model: for example, modify the load balancing constraints and consider the available space in vehicles or wagons instead of pallets.
-
-Because we applied this heuristic process to all $K!$\/ tours, perhaps it seems that our methodology is not generalizable.
-This is not true; we did this because {\it Shims}\/ is very fast and $K$\/ is small in practical cases of air transport.
-In other contexts where $K$\/ may have larger values, this heuristic process remains valid.
-It is enough to apply it only to some tours that offer a better expectation of a good result, for example, on the TSP's shortest path or other similar ones.
+
+We also show that this solution process remains valid in other contexts where there may be a greater number of nodes.
+This way, our method is not exclusive to aircraft and airports: it can be adapted to ships and ports, trucks and warehouses, or wagons and railways.
+In these situations, it would be necessary to make some changes in the model: for example, modify the load balancing constraints and consider parts of the available space as pallets.
 
